### 2022-08-12 notes.tex
some notes, and a bit of agda code

\documentclass[12pt]{article}
%AMS-TeX packages
\usepackage{tikz}
\usepackage[utf8]{inputenc}
\usepackage{amssymb,amsmath,amsthm,stmaryrd, wasysym} 
%geometry (sets margin) and other useful packages
\usepackage[margin=1.25in]{geometry}
\usepackage{graphicx,ctable,booktabs}
\usepackage{tikz-cd}
\usepackage{mathpartir}

\usepackage[sort&compress,square,comma,authoryear]{natbib}
\bibliographystyle{plainnat}

\newcommand{\leftmultimap}{\mathop{\rotatebox[origin=c]{180}{$\multimap$}}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\Lang}{\textrm{Lang}}
\newcommand{\Grammar}{\textrm{Grammar}}
\newcommand{\Action}{\textrm{Action}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newtheorem{definition}{Definition}

\newcommand\rsepimp{\mathbin{-\mkern-6mu*}}
\newcommand\lsepimp{\mathbin{\rotatebox[origin=c]{180}{$-\mkern-6mu*$}}}
\newcommand\sepconj{\mathbin{*}}
\newcommand\bunch{\mathcal W}

\begin{document}
\title{Monoidal Toposes of Grammars and Semantic Actions}
\author{Max S. New}
\maketitle

\begin{abstract}
  We show that formal grammars (of unrestricted complexity) can be
  modeled as objects of a monoidal topos and algebraic operations on
  languages can be lifted to universal constructions in this
  category. Further, this can be extended to a monoidal topos of
  formal grammars equipped with a semantic action that produces a
  semantic value.

  Since the grammars are of unrestricted complexity, this topos can in
  addition to modeling regular and context-free languages, also
  constructs such as the \emph{Brzozowski derivative} of a language,
  and familiar facts about this derivative are derivable from
  manipulation of universal properties.
\end{abstract}

\section{Languages, Grammars and Semantic Actions}

A \emph{language} over an alphabet $\Sigma$ is a subset of strings
$\Sigma^*$. A \emph{grammar} $g$ is a syntactic object that presents a
language. There are many descriptions of formal grammars: regular
expressions, context-free grammars, context-sensitive etc. The problem
of \emph{language recognition} is to take a string $w \in \Sigma^*$
and a grammar $g$ and determine if $w$ is a member of the language
defined by $g$.

In programming applications such as compilation or communication over
a network, we are not just interested in the problem of language
\emph{recognition}, determining whether or not a word is in a
language, but in the problem of \emph{parsing}, the attempt to produce
a \emph{parse tree} from the string. We can think of the parsing as a
``constructive'' version of the language recognition problem: the
parse tree tells us not just \emph{whether} a word is a member of the
language, but also \emph{why}. A language is defined as a subset of
$\Sigma^*$, which is equivalent to a \emph{predicate} on $\Sigma^*$:
\[ L : \Sigma^* \to \Omega \]
where $\Omega$ is the set of \emph{truth values}\footnote{classically,
this is equivalent to any two element set, but constructively to
assume this set is boolean would require the language is
decidable.}. We can categorify this situation by replacing the
lattice of truth values $\Omega$ with the topos of sets $\Set$:
\[ g : \Sigma^* \to \Set \]
That is, a language is a \emph{functor} from $\Sigma^*$ (viewed as a
discrete category) to the category of sets and functions.
Functors from $\Sigma^*$ to $\Set$ are called \emph{presheaves} over
$\Sigma^*$ and the category of such functors is extremely
well-behaved. We show in the remainder of this note that this provides
us with a rich language for constructing languages algebraically.
\begin{quote}
  A grammar is a presheaf over the set of strings
\end{quote}
Normally a formal grammar is defined to be a syntactic object in some
formal system such as regular expressions or context-free grammars.
%
We see each of these notions of formal grammar as being a
\emph{presentation} of the semantic notion of grammar. In Lawvere's
terminology, we might call the typical formal grammars as
\emph{subjective} and our semantic notion an \emph{objective grammar}. 
%
These different classes of formal grammars can be described as
inductively generated subcategories closed under certain constructions
definable on all grammars.

Note that a presheaf $P : \Sigma^* \to \Set$ over a discrete category
like $\Sigma^*$ is equivalent to simply a set $\int P$ of ``parse
trees'' with a function $\pi : \int P \to \Sigma^*$ that shows what
string each parse tree is a parse of. This is a generalization of the
fact that a language can be thought of as a subset $L \subseteq
\Sigma^*$. We will use the presentation in terms of presheaves
throughout.

Returning to the parsing problem, the output of a parser usually isn't
the literal parse tree, but instead the result of some \emph{semantic
actions} run on the parse tree, which usually serve to remove some
syntactic details that are unnecessary for later processing.
%
Given a grammar in the guise of a presheaf $g$ over strings, we define
a semantic action to be a set $S$ of with a function that produces a
semantic element from any parse $a : \prod_{w \in \Sigma^*} g_w \to
S$. The semantic actions can also be arranged into a structured
category, as the category of semantic actions can be defined by the
following pullback, a construction in topos theory known as
\emph{Artin gluing}:

% https://q.uiver.app/?q=WzAsNCxbMCwxLCJcXFNldCJdLFsxLDEsIlxcR3JhbW1hciJdLFsxLDAsIlxcR3JhbW1hcl5cXHRvIl0sWzAsMCwiXFxBY3Rpb24iXSxbMywwXSxbMCwxXSxbMywyXSxbMiwxXSxbMywxLCIiLDEseyJzdHlsZSI6eyJuYW1lIjoiY29ybmVyIn19XV0=
\[\begin{tikzcd}
	\Action & {\Grammar^\to} \\
	\Set & \Grammar
	\arrow[from=1-1, to=2-1]
	\arrow[from=2-1, to=2-2]
	\arrow[from=1-1, to=1-2]
	\arrow[from=1-2, to=2-2]
	\arrow["\lrcorner"{anchor=center, pos=0.125}, draw=none, from=1-1, to=2-2]
\end{tikzcd}\]

The observations we present here are fairly mundane but we intend that
they should form the beginnings of a general theory of grammars that
can be extended to more structured forms such as type-checking.

\section{Language Combinators in Monoidal Toposes}

Next, we describe the structures present in the categories $\Grammar$
and $\Action$ described above and how they relate to familiar aspects
of formal grammars and parsing.

First, we should consider what the morphisms are.

\begin{definition}
  A language $L$ is included in another $L'$, written $L \subseteq L'$
  when every string in $L$ is in $L'$.

  A morphism of formal grammars $f : g \to g'$ is a natural
  transformation. More concretely, it is a family of functions
  \[ \prod_{w \in \Sigma^*} g_w \to g_w' \]

  A morphism of semantic actions $\phi : (a,g,X) \to (a', g', X')$ is
  a pair of a morphism of grammars $f : g \to g'$ and a function $s :
  X \to X'$ such that for every $w \in \Sigma^*$ and parse $p \in g_w$
  \[ s(a(p)) = a'(f_w(p)) \]
\end{definition}

Each of these notions is a refinement of the previous: a morphism of
grammars implies an inclusion of their underlying languages, and by
definition a morphism of semantic actions contains a morphism of the
underlying grammars. This is formalized in the fact that there are
functors

% https://q.uiver.app/?q=WzAsMyxbMSwwLCJcXEdyYW1tYXIiXSxbMiwwLCJcXExhbmciXSxbMCwwLCJcXEFjdGlvbiJdLFsyLDAsInxcXGNkb3R8Il0sWzAsMSwifHxcXGNkb3R8fCJdXQ==
\[\begin{tikzcd}
	\Action & \Grammar & \Lang
	\arrow["{|\cdot|}", from=1-1, to=1-2]
	\arrow["{||\cdot||}", from=1-2, to=1-3]
\end{tikzcd}\]

In fact these functors both have fully faithful right adjoints
exhibiting these as reflective subcategories. In this sense, the
theory of semantic actions subsumes the theory of grammars subsumes
the theory of languages.

Next, we can define the most basic possible grammars: the grammars
that accept exactly one string.
%
Since $\Grammar$ is a presheaf category, the \emph{Yoneda embedding}
defines for each string $w$, the representable presheaf $Y w$. Since
$\Sigma^*$ is a set, this takes a particularly simple form, there is a
single parse precisely when the input string is $w$:
\[ (Y w)_w' = \{ * | w = w' \} \]
%
This extends to a trivial semantic action.

Next, since $\Sigma^*$ is not just a set but a \emph{monoid} with
concatenation of strings\footnote{in fact the free monoid}, by the
\emph{Day convolution} structure we get that the presheaf category
$\Grammar$ has a \emph{monoidal} structure. The monoidal product is
the \emph{concatenation} of grammars:
\[ (g \sepconj g')w = \sum_{w_1w_2 = w} g_{w_1} \times g_{w_2}\]
In other words, a parse of the sequenced grammar $g \sepconj g'$
consists of a choice of splitting the input string into a prefix and a
suffix paired with a $g$-parse of the prefix and a $g'$ parse of the
suffix. This can be extended to a monoidal structure on semantic
actions where the semantic data is the product
\[ (a, g, X) \sepconj (a', g', X') = ((a\sepconj a')(p,p') = (a(p), a'(p'), g \sepconj g', X \times X'))\]
We use the separating conjunction symbol because the semantics
(presheaves over a monoidal category) are similar, but note that the
monoidal structure is \emph{not} symmetric in that $g \sepconj g'$ is
usually very different from $g' \sepconj g$.

This has a unit as well, which is the language of just the empty
string (with trivial parse tree/semantic action):
\[ I w = w = \epsilon \]

Since $\Sigma^*$ is a free monoid, every word $w$ is a concatenation
of finitely many characters $c \in \Sigma$. Given $w = c_1c_2\cdots$
we can now choose between two different representations:
\[ Y(c_1c_2\cdots) \qquad \textrm{or} \qquad Yc_1 \sepconj Yc_2 \sepconj \cdots\]
Since the Yoneda embedding preserves the monoidal structure up to
isomorphism, these are isomorphic as grammars and semantic actions as
well.

Next, $\Lang$ has all limits and colimits. Coproducts give us the
``union'' of languages, with initial object the empty language, and
products give us the intersection, with the terminal object being the
language that accepts all strings, but with no information in the
parse.
\[ (g_1 + g_2) w = (g_1 w) + (g_2 w) \]
\[ 0 w = \emptyset \]
\[ (g_1 \times g_2) w =  g_1 w \times g_2 w\]
\[ 1 w =  1 \]

We can also define inductively defined grammars (and so access regular
and context free languages) as colimits, for instance the Kleene star
operation is the colimit of the $\omega$-diagram
\[ 0 \to (I + g * 0) \to I + g \sepconj (I + g \sepconj 0) \to \cdots \]

\subsection{Closed Structure}

By general properties of presheaf categories, the category of
languages has \emph{closed} structure with respect to the monoidal
products of sequencing $\sepconj$ and product $\times$.

First, the product has a right (multi-variable) adjoints:
\[ (L_1 \times L_2 \to  L_3) \cong L_1 \to (L_2 \Rightarrow L_3) \cong L_2 \to (L_1 \Rightarrow L_3) \]

Defined by
\[ (L \Rightarrow L')w = \textrm{Hom}((Y w \times L), L') = Lw \to L'w \]
Informally, $(L \Rightarrow L')w$ is constructive \emph{implication}
of languages: a parse of $w$ is a function that takes $L$-parses of
$w$ to $L'$-parses of $w$

Next, the sequencing operator $\sepconj$, since it is not symmetric, has
two right multi-variable adjoints:
\[ (L_1 \sepconj L_2 \to L_3) \cong (L_1 \to L_3 \lsepimp L_2) \cong (L_2 \to L_2 \rsepimp L_3) \]

Defined by
\[ (L \multimap L')w = \Pi_{w_l} L w_l \to L'(w_lw) \]
\[ (L' \leftmultimap L)w = \Pi_{w_r} L w_r \to L'(ww_r) \]

A parse of $(L \multimap L')w$ is a kind of ``partial'' parse of $L'$:
given any parse $L w_l$, we can get a complete parse $L' (w_lw)$. So
if $(L \multimap L') w$ successfully parses (and $L$ is non-empty),
then we have succesfully parsed a suffix of $L'$, that when
concatenated with a prefix that can by parsed as $L$ will yield a full
parse of $L'$.

Dually, the left hom $(L' \leftmultimap L)$ is a partial parse from
the right of the string.

These left and right hom operations can be used to define the
\emph{derivative} of a language with respect to a character or
word. If we focus on left-to-right parsing this would be:
\[ d^l_c(L) = Yc \multimap L \]
Why?
\[ (Yc \multimap L)w = \Pi_{w_l} \{ 1 | c = w_l \} \to L(w_lw) \cong L(cw) \]
I.e., a parse of $d^l_c(L)$ is precisely a parse of the string with
$c$ prepended.
%
This allows us, for any particular string $w$, to reduce the problem
of parsing $w$ with a grammar $g$ to the problem of producing a parse
of an empty string for the grammar $d^l_w(g)$. Then the typical
derivation of the Brzozowski derivative of a regular language follows
from abstract nonsense?

If we reduce to language recognition, this is a de Morgan-dual of the
``quotient'' of a formal language.
\[ (L_1 / L_2) w \iff \exists w' \in L_2. L_1(ww') \iff \neg (\forall w' \in L_2. \neg L_1(ww')) \iff \neg ((\neg L_1 \leftmultimap L_2) w) \]

%% A grammar $g$ is
%% \begin{enumerate}
%% \item \emph{total} if for every $w$ there is at least one parse
%% \item \emph{deterministic} if for every $w$ there is at most one parse
%% \item \emph{empty} if for every $w$ there are no parses
%% \end{enumerate}

%% %% \begin{definition}
%% %%   A grammar $g$ is \emph{uniquely decidable} if there exists a
%% %%   ``complement'' $g'$ such that $g \vee g'$ is total and
%% %%   deterministic.
%% %% \end{definition}

%% In other words, $g \vee g' \cong \top$

%% \subsection{Endofunctors}

%% TODO

%% \section{Associativity, Precedence, and Ordering using Category-valued Presheaves}

%% The presheaf of sets model of grammars allows us to reason about not
%% just whether a string parses, but also about \emph{ambiguity} of a
%% grammar: whether there exists \emph{more than one} parse. This is
%% modeled by the grammar $g$ having more than one parse of the same
%% string $w$. We can use this as above to reason about when a grammar is
%% ambiguous or not, and if the grammar is unambiguous, we have a clear,
%% deterministic specification for the parser.

%% But what about parsing \emph{ambiguous} grammars? We can simply say
%% that a parser should produce any parse, or perhaps all of them, but in
%% practice many parser generators and programming languages supporting
%% infix/mixfix notation allow for \emph{associativity} and
%% \emph{precedence} declarations. These are described by providing an
%% ambiguous grammar and disambiguating not by rewriting the grammar, but
%% instead by expressing a \emph{preference} between the multiple
%% possible parses. For instance, one of the most common sources of
%% ambiguity in parsing is in associativity and precedence of binary
%% operators. For an example of associativity, a grammar for subtraction
%% might initially be written as:
%% \begin{mathpar}
%%   \begin{array}{rcl}
%%     E &::= & E - E \,|\,(E)\,|\,Number\\
%%   \end{array}
%% \end{mathpar}

%% But then we note that there are multiple, semantically different
%% parses of $3 - 4 - 5$: trees semantically equivalent to $(3 - 4) - 5$
%% or $3 - (4 - 5)$. We can rule out the right-nested version by
%% \emph{factoring} the grammar:
%% \begin{mathpar}
%%   \begin{array}{rcl}
%%     E &::= & E - L \,|\,L\\
%%     L &::= & ( E ) \,|\, Number
%%   \end{array}
%% \end{mathpar}
%% This grammar has fewer parses than the previous: now $3 - 4 - 5$ will
%% only parse as a left-nested term since $4 - 5$ does not match the
%% ``leaf'' production $L$. Instead, most languages declare that $-$ is a
%% left associative operator. We can think of this as expressing a
%% \emph{preference} for certain parses over others, specifically the tree
%% % https://q.uiver.app/?q=WzAsNSxbMiwyLCJFXzIiXSxbMywxLCJFXzMiXSxbMiwwLCItIl0sWzEsMSwiLSJdLFswLDIsIkVfMSJdLFsyLDFdLFszLDRdLFszLDBdLFsyLDNdXQ==
%% \[\begin{tikzcd}
%% 	&& {-} \\
%% 	& {-} && {E_3} \\
%% 	{E_1} && {E_2}
%% 	\arrow[from=1-3, to=2-4]
%% 	\arrow[from=2-2, to=3-1]
%% 	\arrow[from=2-2, to=3-3]
%% 	\arrow[from=1-3, to=2-2]
%% \end{tikzcd}\]
%% is prefered over
%% % https://q.uiver.app/?q=WzAsNSxbMSwwLCItIl0sWzAsMSwiRV8xIl0sWzIsMSwiLSJdLFsxLDIsIkVfMiJdLFszLDIsIkVfMyJdLFswLDFdLFswLDJdLFsyLDNdLFsyLDRdXQ==
%% \[\begin{tikzcd}
%% 	& {-} \\
%% 	{E_1} && {-} \\
%% 	& {E_2} && {E_3}
%% 	\arrow[from=1-2, to=2-1]
%% 	\arrow[from=1-2, to=2-3]
%% 	\arrow[from=2-3, to=3-2]
%% 	\arrow[from=2-3, to=3-4]
%% \end{tikzcd}\]
%% We can express this as an \emph{ordering}, where the prefered tree is
%% ``bigger''.

%% Fortunately, we can express this with a small change. Rather than
%% taking functors from $\Sigma^*$ to \emph{sets} we can take functors to
%% \emph{partially ordered sets}, or even more generally into
%% \emph{categories}. So a grammar $g$ defines for each string $w$ a
%% category $g w$ whose objects are parses and for any two parses
%% $p_1,p_2 \in g w$, we have a set $(g w)(p_1,p_2)$ of ``reasons to
%% prefer $p_2$ over $p_1$''. Furthermore, preference is transitive and
%% reflexive, with associativity and unit of transitive reasoning.

%% Then the parsing problem can be rephrased as, given a syntactic
%% description of a grammar $g$ and a string $w$, what is the \emph{best}
%% parse in $g w$, if any? In categorical language, can we construct a
%% terminal object of $g w$?

%% \section{Ordered Separation Logic for Formal Grammars}

%% Next, based on our algebraic constructions in the previous section, we
%% describe a \emph{polymorphic ordered separation logic} for
%% grammars. While this may seem a surprising application of separation
%% logic, it is quite natural when considering the semantics: Day
%% convolution of presheaves over a \emph{commutative} monoid is one of
%% the oldest semantic interpretations of separation logic proofs. For
%% parsing, we are taking presheaves over a non-commutative monoid (if
%% $|\Sigma| \ge 2$). This non-commutativity is obvious for instance in
%% regular expressions: $r_1r_2$ almost never presents the same language
%% as $r_2r_1$.

%% This consists of three judgments.
%% \begin{enumerate}
%% \item An \emph{open} grammar $g$ parameterized by a context of grammar
%%   variables $\Gamma = \gamma_1,\ldots$:
%%   \[ \Gamma \vdash g \]
%% \item For each $\Gamma$ a judgment of \emph{parse terms} of a grammar
%%   $g$ parameterized by a ``bunch'' of parsed words $\bunch$, where $g$
%%   and $\bunch$ are both parameterized by $\Gamma$:
%%   \[ \Gamma | \bunch \vdash p : g \]
%% \end{enumerate}

%% The intended denotational semantics is
%% \begin{enumerate}
%% \item A grammar $\Gamma \vdash g$ denotes a (mixed-variance?) functor
%%   $\sem{g} : \Lang^{|\Gamma|} \to \Lang$
%% \item A parse term $\Gamma | \mathcal W \vdash p : g$ denotes a
%%   natural transformation $\mathcal W \Rightarrow g$ where a bunch $\mathcal W$
%%   is interpreted appropriately using the two monoidal structures on
%%   $\Lang$:
%%   \begin{align*}
%%     \sem{\cdot_\epsilon} &= \epsilon\\
%%     \sem{\mathcal W_m,w : g} &= \sem{\mathcal W_m}\sepconj \sem{g}\\
%%     \sem{\mathcal W_m,\mathcal W_a} &= \sem{\mathcal W_m}\sepconj \sem{\mathcal W_a}\\
%%     \sem{\cdot_\top} &= \top\\
%%     \sem{\mathcal W_a;\mathcal W_m} &= \sem{\mathcal W_a} \times \sem{\mathcal W_m}        
%%   \end{align*}
%% \end{enumerate}

%% To get a feel for the differences between the multiplicative and
%% additive structure here are some examples.
%% A parse term
%% \[ \cdot_\epsilon \vdash p : g\]
%% denotes a $g$-parse of the empty string, whereas a term
%% \[ \cdot_\top \vdash p : g\]
%% denotes a function from strings $w \in \Sigma^*$ to $g$-parses of $w$.
%% The grammar $g \sepconj g'$ presents the language that parses words
%% $w$ that can be split as $w_1w_2$ with a pair of parses $gw_1$ and
%% $g'w_2$. On the other hand, $g \times g'$ is the intersection of the
%% two languages: a parse of a word $w$ is a pair of a $g$ parse and a
%% $g'$ parse of the entire word.  $g \times 1$ is equivalent to $g$,
%% whereas a parse of $w$ using $g \sepconj 1$ is a $g$-parse of a prefix
%% $w'$ of $w$, and correspondingly $1 \sepconj g$ parses words with a
%% $g$ suffix. On the other hand $\epsilon \times g$ is the restriction
%% of $g$ to only the parses of the empty string.

%% \begin{figure}
%%   \begin{mathpar}
%%     \inferrule
%%     {\Delta, \gamma \vdash g \and \textrm{where $\gamma$ strictly positive}}
%%     {\Delta \vdash \mu \gamma. g}

%%   \end{mathpar}
%%   \caption{Ordered Separation Logic}
%% \end{figure}

%% Given any string $w$ over our alphabet, there is a corresponding
%% grammar $\lceil w \rceil$ that accepts precisely $w$ with a single
%% parse. Then we can formulate the \emph{parsing problem} as a
%% \emph{program synthesis} problem:
%% \begin{quote}
%%   Given a closed grammar $\cdot \vdash g$ and a word $w$, attempt to
%%   construct a parse $x : \lceil w \rceil \vdash p : g$
%% \end{quote}
%% Or, even more naturally, we can break w up into its constituent
%% characters:
%% \begin{quote}
%%   Given a closed grammar $\cdot \vdash g$ and a word $c_1\ldots_n$,
%%   construct a parse $x_1 : \lceil c_1 \rceil,\ldots_n \vdash p
%%   : g$
%% \end{quote}

%% \begin{figure}
%%   \begin{mathpar}
%%     \inferrule{g \in \mathcal G}{\mathcal G \vdash g}

%%   \end{mathpar}
%%   \caption{Context-free Expressions}
%% \end{figure}

%% \section{Semantic Actions and Grammar/Parser Combinators}

%% Finally, we arrive at the last main component of a parser, the
%% so-called ``semantic actions''. While the constructive parsing
%% approach does produce a parse \emph{tree}, it is not usually the case
%% that this parse tree is \emph{the same} as the desired typed of
%% ASTs. Instead, we need only that there is a \emph{function} from parse
%% trees to ASTs. In general this might fail to be injective (such as
%% forgetting whether parentheses were used) or surjective (if there are
%% internal compiler forms that do not directly correspond to source
%% programs).

%% We can formalize this by noting that there is a functor, the constant
%% functor $\Delta$ from $\Set$ to $\Set^{\Sigma^*}$ that sends a set $A$
%% to the constant presheaf $\Delta(A)_w = A$. Then we can take the comma
%% category $\Set^{\Sigma^*}/K$. The objects of this category are pairs
%% of a grammar $g$ and a set $A$ with a function
%% \[ \prod_{w \in \Sigma^*} g_w \to A \]

%% Then the projection functor $cod : \Set^{\Sigma^*}/\Delta \to \Set$ is
%% an opfibration. The pushforward is the ``map'' operation on
%% grammars. We can equivalently view an opfibration as a functor from
%% $\Set$ to $\Set^{\Sigma^*}$

%% While the presheaf presentation directly captures the notion of a
%% parse tree, in a program the parse tree described by a grammar is
%% often massaged to express precedence data and so the tree structure of
%% a parse is more complex than the actual data being parsed. For
%% instance, an AST data type for a single binary arithmetic operation
%% can be represented as a binary tree with numbers at the leaves, such
%% as the following Haskell/Agda-like data definition:
%% \begin{verbatim}
%% data Tree where
%%   op   : Tree -> Tree -> Tree
%%   leaf : Num  -> Tree
%% \end{verbatim}

%% If the single operator op is defined to be left-associative (such as
%% subtraction/division), a typical grammar for parsing these expressions
%% would be defined by \emph{factoring} the grammar into two mutually
%% recursive non-terminals, \emph{E} for expression and \emph{L} for
%% leaf:
%% \begin{mathpar}
%%   \begin{array}{rcl}
%%     E &::= & E - L \,|\,L\\
%%     L &::= & ( E ) \,|\, Number
%%   \end{array}
%% \end{mathpar}

%% This roughly corresponds to the following mutually recursive datatypes:
%% \begin{verbatim}
%% data Expr where
%%   op   : Expr -> Leaf -> Expr
%%   leaf : Leaf -> Expr

%% data Leaf where
%%   parens : Expr -> Leaf
%%   num    : Number -> Leaf
%% \end{verbatim}

%% I.e., an \texttt{Expr} is a non-empty list of \texttt{Expr + Number}.

%% This mismatch is usually handled by \emph{semantic actions}, a kind of
%% fold over the parse tree that is compositionally expressed alongside
%% the definition of the parser itself. This can be seen as a manually
%% performed optimization to make up for the compiler's inability to
%% \emph{fuse} the two functions of parsing (an unfold) with constructing
%% the intended AST (a fold). However, semantic actions become more
%% useful when we have a \emph{compositional} language of grammars with
%% corresponding semantic actions, such as in parser combinator
%% libraries.

%% Parser combinator libraries are typically based on some notion of
%% backtracking error monad: a term of type \texttt{Parser Sigma A} is a
%% parser of strings from the alphabet Sigma that on correct parses produces
%% an associated value of type \texttt{A}.
%% %
%% Intuitively, this should be semantically (though not performantly)
%% isomorphic to a grammar $g$ paired with a function from $g$-parses to
%% $A$.

\section{Potential Extensions}

What we have presented so far is quite simple. We highlight a few
possible avenues for exploration.

\subsection{Internal Language of Monoidal Toposes}

A powerful tool that toposes provides is the use of \emph{internal
languages} which in the case of a topos can interpret all of
intuitionistic set theory or type theory. When working with specific
toposes it is fruitful to extend the type theory with new axioms and
structures based on the structure present in the category.

For a \emph{monoidal topos} this would be some form of
\emph{separation type theory}, a proof-relevant form of ordered
separation logic. Is there any syntax we can use off the shelf?

Since the topos of semantic actions is constructed by Artin gluing, it
contains the category of $\Set$ as an open subtopos and $\Grammar$ as
its closed complement. This structure can be axiomatized as an
abstract proposition $\Phi$ which we can think of as ``when $\Phi$ is
true we are in the subtopos $\Set$.''  Sterling and Harper have termed
this situation a ``phase distinction''. From $\Phi$, one can define
open and closed modalities $\fullmoon$ and $\newmoon$ that ``project
out'' the underlying set and grammar from a semantic action.
%
Then we might be able to internally describe the pushforward of an
action $(a,g,X)$ along a function $f : X \to Y$ as follows. First we
assume we have a set $A$ (corresponding to $(a,g,X)$) and a
$\Phi$-modal set $Y$ (corresponding to $Y$). Then given any $f : A \to
Y$ we can define the pushforward as the refinement type
\[ [ \Phi \hookrightarrow y : Y | \newmoon (\sum_{x:A} f(x) = b) ]\]

\subsection{More General Parsing}

Our goal with this theory is to have a higher explanatory power for
real life parsing tools than existing theory of langauges. We consider
several ad-hoc extensions of langauges and how they might be modeled
semantically.

First, almost all parser generators support \emph{precedence} rules
for modularly defining the binding of different binary operators.
This technically goes outside of the notion of grammar, though can be
explained by translation to ordinary grammars.
%
Instead we can model this semantically by defining a \emph{precedence
grammar} as a functor from strings to preordered sets. Then the
parsing problem is not to find a matching parse, but instead find a
\emph{greatest} parse.

Another extension is to describe the parsing of morphisms of a
category other than strings. This can be used to model common ad hoc
extensions of parsing such as the end of string character \$ by
parsing arrows from $\bullet$ to $\top$ in a category with two
objects:
\[% https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZARgBoAGAXVJADcBDAGwFcYkQAdDnCNEAX1LpMufIRTkK1Ok1bsuAI2aNGMHAOkwoAc3hFQAMwBOEALZIyIHkkkyWbRCAAkAoSGNmLNa4kuMIvESSBkxwMNKM9AowjAAKIngE7EZY2gAW6jQM9uwAxq6GJua+3hBeIP6BKMQAHMGh4TSR0XEJYsmpGSBZsg4g+fyU-EA
\begin{tikzcd}
\bullet \arrow[r, "\$"] \arrow["w"', loop, distance=2em, in=305, out=235] \arrow["w'"', loop, distance=2em, in=125, out=55] & \top
\end{tikzcd}\]
Here the endomorphisms of $\bullet$ are the strings and a single arrow
\$ allows you to transition to the end of string.  We could extend our
model by viewing a grammar over a category $\mathcal C$ as a functor
$|\mathcal C^\to| \to \Set$, which would mean a grammar gives you a
set of parses for all hom sets. This only has a partial monoidal
structure on the base category, but that is enough to define a
monoidal structure on the category of presheaves.

Finally, the applications of most interest to us are the extensions to
tree grammars, or better yet, grammars for trees with binding. For
example, given a polynomial functor $P$ describing a type of
``operations'', we could define a tree grammar as a functor from the
free operad of the polynomial to set $|P^*| \to \Set$, where each
operation of the polynomial would induce its own version of the
separating conjunction on tree grammars.

\end{document}

### 2022-08-14 notes.tex
eureka!?

@@
 
-\section{Language Combinators in Monoidal Toposes}
+\section{Grammar Combinators in Monoidal Toposes}
 
@@
 
+\section{Internal Language for the Topos of Semantic Actions}
+
+A powerful tool for expressing constructions in a topos is the use of
+the \emph{internal language}, which for a generic topos is a form of
+extensional dependent type theory.
+%
+To extract maximum benefit from the internal language, we add axioms
+and judgments to type theory for the specific topos models we are
+working with.
+%
+For the toposes of grammars and semantic actions we have a plethora of
+structure, we highlight the three most relevant:
+
+\begin{itemize}
+\item Both are toposes so models of extensional dependent type theory
+\item They are models of the (non-dependent) ordered logic of bunched
+  implications in that they are non-symmetric monoidal closed as well
+  as Cartesian closed.
+\item The topos of semantic actions, being constructed by Artin
+  gluing, contains a proposition $\Phi$ that presents the categories
+  $\Set$ and $\Grammar$ as open and complementary closed
+  subtoposes. Sterling and Harper refer to this situation as a ``phase
+  distinction''.
+\end{itemize}
+
+Combining the internal language of dependent type theory with bunched
+implications or even just monoidal structure is tricky. The reason is
+that all constructions need to be stable under pullback to be used in
+dependent type theory, but our monoidal structure is not stable under
+pullback, as is typical.
+%
+Fortunately, the phase distinction allows for a fruitful combination
+of these aspects, called \emph{quantitative type theory}
+\cite{atkey-qtt,mcbride-nuttin}.
+%
+We will need a non-symmetric variant.
+
 \section{Potential Extensions}
@@
 possible avenues for exploration.
-
-\subsection{Internal Language of Monoidal Toposes}
-
-A powerful tool that toposes provides is the use of \emph{internal
-languages} which in the case of a topos can interpret all of
-intuitionistic set theory or type theory. When working with specific
-toposes it is fruitful to extend the type theory with new axioms and
-structures based on the structure present in the category.
-
-For a \emph{monoidal topos} this would be some form of
-\emph{separation type theory}, a proof-relevant form of ordered
-separation logic. Is there any syntax we can use off the shelf?
-
-Since the topos of semantic actions is constructed by Artin gluing, it
-contains the category of $\Set$ as an open subtopos and $\Grammar$ as
-its closed complement. This structure can be axiomatized as an
-abstract proposition $\Phi$ which we can think of as ``when $\Phi$ is
-true we are in the subtopos $\Set$.''  Sterling and Harper have termed
-this situation a ``phase distinction''. From $\Phi$, one can define
-open and closed modalities $\fullmoon$ and $\newmoon$ that ``project
-out'' the underlying set and grammar from a semantic action.
-%
-Then we might be able to internally describe the pushforward of an
-action $(a,g,X)$ along a function $f : X \to Y$ as follows. First we
-assume we have a set $A$ (corresponding to $(a,g,X)$) and a
-$\Phi$-modal set $Y$ (corresponding to $Y$). Then given any $f : A \to
-Y$ we can define the pushforward as the refinement type
-\[ [ \Phi \hookrightarrow y : Y | \newmoon (\sum_{x:A} f(x) = b) ]\]
 
@@
 
+\bibliographystyle{plain}
+\bibliography{grammar}
+
 \end{document}

### 2022-08-15 notes.tex
description of inductive grammars.

@@
 \newtheorem{definition}{Definition}
+\newtheorem{theorem}{Theorem}
 
@@
 
-This has a unit as well, which is the language of just the empty
-string (with trivial parse tree/semantic action):
+This has a unit as well, which is the grammar with a unique parse of
+the empty string (with trivial semantic action):
 \[ I w = w = \epsilon \]
@@
 
-Next, $\Lang$ has all limits and colimits. Coproducts give us the
-``union'' of languages, with initial object the empty language, and
-products give us the intersection, with the terminal object being the
-language that accepts all strings, but with no information in the
-parse.
+Next, $\Grammar$ has all limits and colimits. Binary coproducts give
+us the ``sum'' of grammars, i.e., a parse is a parse in one of the two
+languages. The initial object is the empty grammar, the grammar that
+has no parses.
 \[ (g_1 + g_2) w = (g_1 w) + (g_2 w) \]
 \[ 0 w = \emptyset \]
+The semantic actions ``lie over'' the corresponding coproduct in
+$\Set$. These correspond to the union and empty languages in $\Lang$.
+
+Products give us a kind of intersection of grammars, where a parse
+tree is a parse from each language, with the terminal object being
+the language that accepts all strings with a unique parse.
 \[ (g_1 \times g_2) w =  g_1 w \times g_2 w\]
 \[ 1 w =  1 \]
+These lie over the product and unit semantic actions.
 
-We can also define inductively defined grammars (and so access regular
-and context free languages) as colimits, for instance the Kleene star
-operation is the colimit of the $\omega$-diagram
-\[ 0 \to (I + g * 0) \to I + g \sepconj (I + g \sepconj 0) \to \cdots \]
+Inductively defined grammars should be initial algebras of functors $F
+: \Grammar \to \Grammar$. To define regular and context free grammars
+it should suffice to use Ad\'amek's construction generalizing Kleene's
+fixed point theorem to categories. That is, if $F$ is
+$\omega$-co-continuous, i.e., preserves $\omega$-colimits, then it has
+an initial algebra, which can be constructed as the colimit of
+\[ F^0(0) \to F^1(0) \to F^2(0) \to \cdots \]
+
+Then, for example the Kleene star of a grammar $g$ is given by the
+initial algebra of the functor $F(\gamma) = I + g \sepconj \gamma$,
+which is $\omega$-co-continuous because $\sepconj$ and $+$ are
+$\omega$-co-continuous.
+
+So far the interesting structure has mostly been in the category of
+grammars, and the corresponding sets of parses have been isomorphic to
+the set of parses. The simple combinator that we need is a ``map'' for
+semantic actions, that says given a semantic action $A$ that produces
+$X$ values, and a function $f : X \to Y$, we can push forward and get
+an action $f_*A$ that produces $Y$ values. If $A = (a, g, X)$ then we
+can define this simply as
+\[ f_*(a,g,X) = (f \circ a, g, Y) \]
 
@@
 By general properties of presheaf categories, the category of
-languages has \emph{closed} structure with respect to the monoidal
+grammars has \emph{closed} structure with respect to the monoidal
 products of sequencing $\sepconj$ and product $\times$.
 
-First, the product has a right (multi-variable) adjoints:
-\[ (L_1 \times L_2 \to  L_3) \cong L_1 \to (L_2 \Rightarrow L_3) \cong L_2 \to (L_1 \Rightarrow L_3) \]
+First, the product has a right (multi-variable) adjoint:
+\[ (g_1 \times g_2 \to  g_3) \cong g_1 \to (g_2 \Rightarrow g_3) \cong g_2 \to (g_1 \Rightarrow g_3) \]
 
 Defined by
-\[ (L \Rightarrow L')w = \textrm{Hom}((Y w \times L), L') = Lw \to L'w \]
-Informally, $(L \Rightarrow L')w$ is constructive \emph{implication}
-of languages: a parse of $w$ is a function that takes $L$-parses of
-$w$ to $L'$-parses of $w$
+\[ (g \Rightarrow g')w = \textrm{Hom}((Y w \times g), g') = gw \to g'w \]
+Informally, $(g \Rightarrow g')w$ is constructive \emph{implication}
+of languages: a parse of $w$ is a function that takes $g$-parses of
+$w$ to $g'$-parses of $w$
 
@@
 two right multi-variable adjoints:
-\[ (L_1 \sepconj L_2 \to L_3) \cong (L_1 \to L_3 \lsepimp L_2) \cong (L_2 \to L_2 \rsepimp L_3) \]
+\[ (g_1 \sepconj g_2 \to g_3) \cong (g_1 \to g_3 \lsepimp g_2) \cong (g_2 \to g_2 \rsepimp g_3) \]
 
 Defined by
-\[ (L \multimap L')w = \Pi_{w_l} L w_l \to L'(w_lw) \]
-\[ (L' \leftmultimap L)w = \Pi_{w_r} L w_r \to L'(ww_r) \]
+\[ (g \rsepimp g')w = \Pi_{w_l} g w_l \to g'(w_lw) \]
+\[ (g' \lsepimp g)w = \Pi_{w_r} g w_r \to g'(ww_r) \]
 
-A parse of $(L \multimap L')w$ is a kind of ``partial'' parse of $L'$:
-given any parse $L w_l$, we can get a complete parse $L' (w_lw)$. So
-if $(L \multimap L') w$ successfully parses (and $L$ is non-empty),
-then we have succesfully parsed a suffix of $L'$, that when
-concatenated with a prefix that can by parsed as $L$ will yield a full
-parse of $L'$.
+A parse of $(g \rsepimp g')w$ is a kind of ``partial'' parse of $g'$:
+given any parse $g w_l$, we can get a complete parse $g' (w_lw)$. So
+if $(g \rsepimp g') w$ successfully parses (and $g$ is non-empty),
+then we have succesfully parsed a suffix of $g'$, that when
+concatenated with a prefix that can by parsed as $g$ will yield a full
+parse of $g'$.
+%
+This Day convolution closed structure is used in separation logic.
 
-Dually, the left hom $(L' \leftmultimap L)$ is a partial parse from
+Dually, the left hom $(g' \lsepimp g)$ is a partial parse from
 the right of the string.
@@
 word. If we focus on left-to-right parsing this would be:
-\[ d^l_c(L) = Yc \multimap L \]
+\[ d^l_c(g) = Yc \rsepimp g \]
 Why?
-\[ (Yc \multimap L)w = \Pi_{w_l} \{ 1 | c = w_l \} \to L(w_lw) \cong L(cw) \]
-I.e., a parse of $d^l_c(L)$ is precisely a parse of the string with
+\[ (Yc \rsepimp g)w = \Pi_{w_l} \{ 1 | c = w_l \} \to g(w_lw) \cong g(cw) \]
+I.e., a parse of $d^l_c(g)$ is precisely a parse of the string with
 $c$ prepended.
@@
 ``quotient'' of a formal language.
-\[ (L_1 / L_2) w \iff \exists w' \in L_2. L_1(ww') \iff \neg (\forall w' \in L_2. \neg L_1(ww')) \iff \neg ((\neg L_1 \leftmultimap L_2) w) \]
+\[ (g_1 / g_2) w \iff \exists w' \in g_2. g_1(ww') \iff \neg (\forall w' \in g_2. \neg g_1(ww')) \iff \neg ((\neg g_1 \lsepimp g_2) w) \]
 
-%% A grammar $g$ is
-%% \begin{enumerate}
-%% \item \emph{total} if for every $w$ there is at least one parse
-%% \item \emph{deterministic} if for every $w$ there is at most one parse
-%% \item \emph{empty} if for every $w$ there are no parses
-%% \end{enumerate}
+\section{Application: Derivative of a Regular Semantic Action}
 
-%% %% \begin{definition}
-%% %%   A grammar $g$ is \emph{uniquely decidable} if there exists a
-%% %%   ``complement'' $g'$ such that $g \vee g'$ is total and
-%% %%   deterministic.
-%% %% \end{definition}
+Let $A$ be a semantic action inductively built using the constructions
+of separating conjunction, unit, binary and empty coproducts,
+characters, Kleene star, and pushforward.
 
-%% In other words, $g \vee g' \cong \top$
+Claim:
+\begin{theorem}
+  If $A$ is a regular semantic action, then for any character $c$,
+  \[ d_c(A) = Yc \rsepimp A \]
+  is isomorphic to a regular semantic action.
+\end{theorem}
+\begin{proof}
+  By induction on the definition of $A$.
+  \begin{enumerate}
+  \item $A = I$, then $Y c\rsepimp I \cong 0$
+  \item $A = A_1 \sepconj A_2$, then
+    \begin{align*}
+      Yc \rsepimp A_1 \sepconj A_2
+      \cong ((Yc \rsepimp A_1) \sepconj A_2) + ((A_1 \times I) \sepconj (Yc \rsepimp A_2))
+    \end{align*}
+  \item $A = f_*A'$, then
+    \begin{align*}
+      Y c \rsepimp f_*A \cong g_*(Yc \rsepimp A)
+    \end{align*}
+    Where $g : (1 \Rightarrow X) \to (1 \Rightarrow Y)$ is $g(h)(x) =
+    f(h(x))$.
+  \item TODO: finish
+  \end{enumerate}
+\end{proof}
 
@@
 
+%% We can define some concepts on grammars using the internal language.
+
+%% A grammar $g$ is
+%% \begin{enumerate}
+%% \item \emph{total} if for every $w$ there is at least one
+%%   parse. Equivalently, $g$ has a global element.
+%% \item \emph{deterministic} if for every $w$ there is at most one
+%%   parse. In the internal language, $g$ is a proposition.
+%% \item \emph{empty} if for every $w$ there are no parses. In the
+%%   internal language, $g \cong 0$
+%% \end{enumerate}
+
+%% \begin{definition}
+%%   A grammar $g$ is \emph{decidable} if its propositional truncation
+%%   $||g||$ is a complemented truth value.
+%% \end{definition}
+
+%% \begin{definition}
+%%   A grammar $g$ is \emph{uniquely decidable} if there exists a
+%%   ``complement'' $\neg g$ such that $g \vee \neg g$ is total and
+%%   deterministic.
+%% \end{definition}
+
+%% In other words, $g \vee g' \cong \top$
+
 A powerful tool for expressing constructions in a topos is the use of

### 2022-08-22 notes.tex
idea for extension to single-sorted tree and abt grammars

@@
 
-\section{Application: Derivative of a Regular Semantic Action}
+%% \section{Application: Derivative of a Regular Semantic Action}
 
-Let $A$ be a semantic action inductively built using the constructions
-of separating conjunction, unit, binary and empty coproducts,
-characters, Kleene star, and pushforward.
+%% Let $A$ be a semantic action inductively built using the constructions
+%% of separating conjunction, unit, binary and empty coproducts,
+%% characters, Kleene star, and pushforward.
 
-Claim:
-\begin{theorem}
-  If $A$ is a regular semantic action, then for any character $c$,
-  \[ d_c(A) = Yc \rsepimp A \]
-  is isomorphic to a regular semantic action.
-\end{theorem}
-\begin{proof}
-  By induction on the definition of $A$.
-  \begin{enumerate}
-  \item $A = I$, then $Y c\rsepimp I \cong 0$
-  \item $A = A_1 \sepconj A_2$, then
-    \begin{align*}
-      Yc \rsepimp A_1 \sepconj A_2
-      \cong ((Yc \rsepimp A_1) \sepconj A_2) + ((A_1 \times I) \sepconj (Yc \rsepimp A_2))
-    \end{align*}
-  \item $A = f_*A'$, then
-    \begin{align*}
-      Y c \rsepimp f_*A \cong g_*(Yc \rsepimp A)
-    \end{align*}
-    Where $g : (1 \Rightarrow X) \to (1 \Rightarrow Y)$ is $g(h)(x) =
-    f(h(x))$.
-  \item TODO: finish
-  \end{enumerate}
-\end{proof}
+%% Claim:
+%% \begin{theorem}
+%%   If $A$ is a regular semantic action, then for any character $c$,
+%%   \[ d_c(A) = Yc \rsepimp A \]
+%%   is isomorphic to a regular semantic action.
+%% \end{theorem}
+%% \begin{proof}
+%%   By induction on the definition of $A$.
+%%   \begin{enumerate}
+%%   \item $A = I$, then $Y c\rsepimp I \cong 0$
+%%   \item $A = A_1 \sepconj A_2$, then
+%%     \begin{align*}
+%%       Yc \rsepimp A_1 \sepconj A_2
+%%       \cong ((Yc \rsepimp A_1) \sepconj A_2) + ((A_1 \times I) \sepconj (Yc \rsepimp A_2))
+%%     \end{align*}
+%%   \item $A = f_*A'$, then
+%%     \begin{align*}
+%%       Y c \rsepimp f_*A \cong g_*(Yc \rsepimp A)
+%%     \end{align*}
+%%     Where $g : (1 \Rightarrow X) \to (1 \Rightarrow Y)$ is $g(h)(x) =
+%%     f(h(x))$.
+%%   \item TODO: finish
+%%   \end{enumerate}
+%% \end{proof}
 
@@
 
+\section{Category Grammars}
+
+As a pre-cursor to multi-sorted tree grammars, we can also consider
+\emph{free category} grammars.
+
+\section{Tree Grammars}
+
+One advantage to our semantic approach is that the generalizations to
+tree parsing, multiple sorts etc, are clear: rather than working in a
+topos built from presheaves over strings, we simply take presheaves
+over sets of trees. The difficulty then is in adapting our axiomatic
+approach.
+
+We will take an approach parameterized in the notion of tree over
+which we are defining the grammar.
+\begin{definition}
+  A notion of single-sorted tree is a cartesian multicategory
+  $\mathcal T$ with one object.
+\end{definition}
+Usually, the notion of tree will be \emph{freely generated} from a
+finite set of operations, but allowing for non-trivial equalities
+allows tree grammars to subsume string grammars. However, if we allow
+for non-free tree grammars, we should probably allow for ``monoid
+grammars''.
+
+\begin{definition}
+  For any notion of tree $\mathcal T$, we define the topos of
+  $\mathcal T$-grammars as the topos $\Set^{|Arr(\mathcal T)|}$.
+\end{definition}
+
+The set of strings has a \emph{monoid} structure, and the category of
+string grammars inherits a \emph{monoidal} category structure from the
+Day convolution.
+%
+Similarly, a free multicategory comes with \emph{non-associative}
+operations of various arities. In the case of a binary operation that
+is not associative this is called a \emph{magma}, and so the topos of
+grammars would inherit a kind of \emph{magmoidal category} structure.
+
+For each operation $o : \mathcal T(n)$, we get an associated
+separating conjunction operation $\sepconj_o : G(T)^n \to G(T)$ defined by
+\[ \sepconj_o(g_1,\ldots)_t = \sum_{o(t_1,\ldots) = t} g_1(t_1) \times \cdots\]
+and similarly, for each $i \in [0,n)$ we get an associated right
+adjoint $(g_1,\ldots,\hat g_i,\ldots)\rsepimp_{o,i} g'$ defined by
+\[ ((g_1,\ldots,\hat g_i,\ldots)\rsepimp_{o,i} g')_t = \prod_{t_1} g_1 t_1 \to \cdots \hat g_i \to \cdots g'(o(t_1,\ldots,t,\ldots))\]
+where $\hat g_i$ indicates that the $i$th place is omitted.
+
+Focusing only on the conjunction operations, what we see here is that
+our notion of tree is a (free) algebraic theory $\mathcal T$ and our
+category of grammars is a kind of \emph{psuedo-algebra} of the theory
+that defines our notion of tree.
+
+\section{Abstract Binding Tree Grammars}
+
+Generalizing further, type systems are a kind of grammar over trees
+with binding structure. Just as notions of tree can be defined as free
+algebraic theories, notions of tree with binding structure can be
+defined as free \emph{second order} algebraic theories.
+
+\begin{definition}
+  A notion of tree with binding $\mathcal T$ is a cartesian multicategory all of
+  whose objects are exponentials $o \to^n o$ for some distinguished
+  object $o$.
+\end{definition}
+
+\begin{definition}
+  Given a notion of tree with binding $\mathcal T$, the topos of
+  grammars is simply $\Set^{|Arr(\mathcal T)|}$.
+
+  But there are many sorts here...
+\end{definition}
+
+Again the interesting question is what the appropriate notion of
+``separating conjunction'' is for this notion of grammar. Following
+the pattern thus far, the categorical structure we want on the
+category of grammars should be given by a pseudo-algebra of the notion
+of theory the grammars are defined over. An algebra of a 2nd-order
+theory is given by an exponentiable object in a category with
+cartesian products with an interpretation of all the operations.
+
+In this case this means that we want to interpret the 2nd-order
+operations as functors whose domain is \emph{a category of
+endofunctors}.
+
+An example should be illuminating. One of the canonical examples of a
+second-order theory is the untyped lambda calculus. This is a single
+sorted theory with two operations :
+
+\begin{enumerate}
+\item \texttt{app : (0, 0)}
+\item \texttt{lam : (1)}
+\end{enumerate}
+
+That is, application is applied to two subterms with no introduced
+bindings, whereas lambda is applied to a subterm with an additional
+binding introduced.  As morphisms in a 2nd order algebraic theory
+viewed as a cartesian multicategory these are in the hom sets:
+
+\begin{enumerate}
+\item \texttt{app : Hom(o, o; o)}
+\item \texttt{lam : Hom(o -> o; o)}
+\end{enumerate}
+
+To give the topos of grammars $G(\mathcal T)$ an algebra structure, we
+define a functor of two arguments for \texttt{app}:
+\[ \textrm{app}(g_1,g_2)_t = \sum_{\texttt{app}(t_1,t_2)} g_1 t_1 \times g_2 t_2 \]
+and a functor whose argument is an endofunctor for \texttt{lam}:
+\[ \textrm{lam}(F)_t = \sum_{\texttt{lam}(x.s)} F(Y x)_s \]
+this doesn't quite make sense...
+%% Something something second-order algebraic theory..
+
+Intuitively, the functor takes in a grammar for the variable and
+outputs a grammar over terms with that free variable in them.
+
 \section{Potential Extensions}

### 2022-08-22 notes.tex
getting a little too speculative

@@
 \newcommand{\Set}{\textrm{Set}}
+\newcommand{\FinSet}{\mathbb F}
+\newcommand{\Cat}{\textrm{Cat}}
 \newcommand{\Lang}{\textrm{Lang}}
@@
 \begin{definition}
-  A notion of tree with binding $\mathcal T$ is a cartesian multicategory all of
-  whose objects are exponentials $o \to^n o$ for some distinguished
-  object $o$.
+  A notion of tree with binding $\mathcal T$ is a cartesian
+  multicategory all of whose objects are exponentials $o \to^n o$ for
+  some distinguished generator object $o$.
 \end{definition}
 
-\begin{definition}
-  Given a notion of tree with binding $\mathcal T$, the topos of
-  grammars is simply $\Set^{|Arr(\mathcal T)|}$.
+The ``trees'' themselves are the morphisms of the multicategory of the
+form
+\[ \mathcal T(\cdot; o) \]
 
-  But there are many sorts here...
-\end{definition}
+The ``constructors'' of this notion of are given by the morphisms of
+the multicategory, in which can all be given in the form:
+\[ c \in \mathcal T((o \to^{n_1} o), \ldots_{p}; o) \]
+A constructor $c$ of this form has $p$ subterms each of which has
+bindings for $n_i$ new variables.
 
-Again the interesting question is what the appropriate notion of
-``separating conjunction'' is for this notion of grammar. Following
-the pattern thus far, the categorical structure we want on the
-category of grammars should be given by a pseudo-algebra of the notion
-of theory the grammars are defined over. An algebra of a 2nd-order
-theory is given by an exponentiable object in a category with
-cartesian products with an interpretation of all the operations.
+So what category do our grammars live in?
+%
+In analogy with the string and tree cases, our category of grammars
+should be some kind of \emph{pseudo}-algebra of the theory in $\Cat$.
+%
+However, unlike the string and tree case, even ``single sorted''
+second-order theories behave more like multi-sorted theories, in that
+a model of a second-order theory needs to interpret not just the
+generator $o$, but also all $n-ary$ exponentials $o \to^n o$ as
+categories.
+%
+This suggests we define for each natural number $n$, a category of
+grammars over trees with $n$ variables:
+\[ \Grammar(n) = \Set^{\mathcal T(o,\ldots_n;o)} \]
+This varies functorially in $n$, defining a functor
+\[ \Grammar(-) : \FinSet \to \Cat \]
+where $\FinSet$ is a skeleton of the category of finite sets and
+arbitrary functions, i.e., the action of ``renaming'' the variables.
 
-In this case this means that we want to interpret the 2nd-order
-operations as functors whose domain is \emph{a category of
-endofunctors}.
+Then we get for each operation $f \in \mathcal T((o^{n_1} \to o,
+\ldots_p; o^m \to o))$ a functor $\Grammar(f) : \Grammar(n_1) \times
+\cdots_p \to \Grammar(m)$ defined as a kind of separating conjunction:
+\[ \Grammar(f)(g_1,\ldots_p)(t) = \sum_{f(\vec x_1. s_1,\ldots_p) = t} (g_1(\vec x_1.s_1)) \times \cdots \]
+Does that even make sense??
 
-An example should be illuminating. One of the canonical examples of a
-second-order theory is the untyped lambda calculus. This is a single
-sorted theory with two operations :
+Maybe we can combine these into one using Grothendieck construction?
+Not sure it has a nice internal language though.
+\[ \Grammar = \int_{[n] : \FinSet} \Grammar(n) \]
 
-\begin{enumerate}
-\item \texttt{app : (0, 0)}
-\item \texttt{lam : (1)}
-\end{enumerate}
+%% Then we can construct a category of grammars for trees with binding as
+%% the Grothendieck construction:
+%% Where objects are pairs of a finite cardinal and a grammar over trees
+%% with that many variables, with a morphism being a pair of a renaming
+%% and then a parse transformer between a grammar and a ``renamed''
+%% grammar.
 
-That is, application is applied to two subterms with no introduced
-bindings, whereas lambda is applied to a subterm with an additional
-binding introduced.  As morphisms in a 2nd order algebraic theory
-viewed as a cartesian multicategory these are in the hom sets:
+%% Again the interesting question is what the appropriate notion of
+%% ``separating conjunction'' is for this notion of grammar. Following
+%% the pattern thus far, the categorical structure we want on the
+%% category of grammars should be given by a pseudo-algebra of the notion
+%% of theory the grammars are defined over. An algebra of a 2nd-order
+%% theory is given by an exponentiable object in a category with
+%% cartesian products with an interpretation of all the operations.
 
-\begin{enumerate}
-\item \texttt{app : Hom(o, o; o)}
-\item \texttt{lam : Hom(o -> o; o)}
-\end{enumerate}
+%% In this case this means that we want to interpret the 2nd-order
+%% operations as functors whose domain is \emph{a category of
+%% endofunctors}.
 
-To give the topos of grammars $G(\mathcal T)$ an algebra structure, we
-define a functor of two arguments for \texttt{app}:
-\[ \textrm{app}(g_1,g_2)_t = \sum_{\texttt{app}(t_1,t_2)} g_1 t_1 \times g_2 t_2 \]
-and a functor whose argument is an endofunctor for \texttt{lam}:
-\[ \textrm{lam}(F)_t = \sum_{\texttt{lam}(x.s)} F(Y x)_s \]
-this doesn't quite make sense...
-%% Something something second-order algebraic theory..
+%% An example should be illuminating. One of the canonical examples of a
+%% second-order theory is the untyped lambda calculus. This is a single
+%% sorted theory with two operations :
 
-Intuitively, the functor takes in a grammar for the variable and
-outputs a grammar over terms with that free variable in them.
+%% \begin{enumerate}
+%% \item \texttt{app : (0, 0)}
+%% \item \texttt{lam : (1)}
+%% \end{enumerate}
+
+%% That is, application is applied to two subterms with no introduced
+%% bindings, whereas lambda is applied to a subterm with an additional
+%% binding introduced.  As morphisms in a 2nd order algebraic theory
+%% viewed as a cartesian multicategory these are in the hom sets:
+
+%% \begin{enumerate}
+%% \item \texttt{app : Hom(o, o; o)}
+%% \item \texttt{lam : Hom(o -> o; o)}
+%% \end{enumerate}
+
+%% To give the topos of grammars $G(\mathcal T)$ an algebra structure, we
+%% define a functor of two arguments for \texttt{app}:
+%% \[ \textrm{app}(g_1,g_2)_t = \sum_{\texttt{app}(t_1,t_2)} g_1 t_1 \times g_2 t_2 \]
+%% and a functor whose argument is an endofunctor for \texttt{lam}:
+%% \[ \textrm{lam}(F)_t = \sum_{\texttt{lam}(x.s)} F(Y x)_s \]
+%% this doesn't quite make sense...
+%% %% Something something second-order algebraic theory..
+
+%% % Examples: Schemes' let vs letrec vs let*
+
+%% Intuitively, the functor takes in a grammar for the variable and
+%% outputs a grammar over terms with that free variable in them.
 

### 2023-12-01 paper/paper.tex
intro to a paper?

\documentclass[acmsmall,anonymous,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

\begin{document}
\title{Formal Grammars as Types in Non-commutative Bunched Type Theory}
\author{Steven Schaefer}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{stschaef@umich.edu}

% TODO: add Pedro's info

\author{Max S. New}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{maxsnew@umich.edu}

\begin{abstract}
  We propose the category of functions from strings to sets as a
  universal syntax-independent model of formal grammars, generalizing
  the set of strings model of formal language theory to account for
  ambiguity. The morphisms of this category give a notion of parse
  transformer, and isomorphism corresponds to \emph{strong
  equivalence} of grammars. We observe that the standard algebraic
  constructions of regular expressions and their extension to
  mu-regular expressions correspond to universal constructions in this
  category, which we axiomatize as a structure we call a 2-Kleene
  Algebra.

\end{abstract}

\section{A Syntax-Independent Notion of Syntax}

The theory of formal languages and parsing is one of the oldest and
most thoroughly developed areas of theoretical computer science. A
prominent topic in the 1950s and 60s led to a series of remarkable
developments: Chomsky's hierarchy of grammar formalisms, practical
algorithms for parsing of regular and context-free grammars and
variants, as well as implementations of practical tools for the
generation of efficient parsers.

A central notion is that of a \emph{formal language} $L$ over an
alphabet $\Sigma$ as simply a \emph{subset} of the set of strings $L
\subseteq \Sigma^*$. This definition is especially useful as it gives
a semantics to formal grammars that is completely independent of any
particular syntactic grammar formalism: any new notion of grammar can
be given a semantics in this common framework and it provides a
precise mathematical speicification for implementing a
\emph{recognizer} of a language and comparing the power of different
formalisms by demonstrating what languages can be formalized within
it. This also allows for completely \emph{language theoretic}
formulations of language classes: e.g., the regular \emph{languages}
can be characterized as those that have finitely many derivatives
\cite{brzozowski-or-myhill-nerode-idk}. This notion is remarkable
because it makes no reference to any specific syntactic formalism for
defining languages such as regular expressions or finite automata.

Formal language theory alone is not sufficient as a specification of a
parser: a language $L \subseteq \Sigma^*$ is equivalently defined as
its indicator function $\chi_L : \Sigma^* \to \Prop$ mapping each
string $w$ to the proposition that it is in the language $w \in
L$. This can then serve as the specification for a language
\emph{recognizer}: a program $r$ of type $\Sigma^* \to \Bool$ such
that $r(w) = \texttt{true}$ if and only if $\chi_L(w)$, but
recognition is insufficient for most tasks for which the theory of
parsing was developed: the production of \emph{semantic} structures
from synactic descriptions in linguistics, compilation or
deserialization. The output of a parser is not just a boolean but a
\emph{parse tree}\footnote{such parse trees are usually not
materialized in memory, as the construction of the parse tree is fused
with application of semantic actions, which can be seen as a kind of
fold over the generated tree.}.

Due to this limitation, parsers are typically specified not by a
formal language, but by some \emph{formal grammar}, which specifies
not just \emph{which} strings are in the language but specifies what
are the \emph{parse trees} for the string. However unlike formal
language theory, there is no common universal notion of what
constitutes a formal grammar, but rather many syntactic systems such
as Chomskyan(?)  grammars, Thue systems, Montague grammars, etc which
are all syntactic presentations of the same underlying idea of an
abstract specification for parsing. The most common of these is the
Chomsky hierarchy of \emph{generative} grammars which specify parse
trees using \emph{derivations} of the string from a set of rewrite
rules, the parse tree encoding which rules were used.

Our first contribution is conceptual: we propose a simple,
syntax-independent definition of \emph{formal grammar}:
\begin{definition}
  A \emph{formal grammar} over an alphabet $\Sigma$ is a function
  $\Sigma^* \to \Set$.
\end{definition}
We say that a grammar $G$ associates to every string $w$ the set $Gw$
of \emph{parses} of that string. While we will work informally in this
paper, the collection $\Set$ can be formalized in many different
logical foundations: as a universe in type theory, a proper class in
set theory, etc. We argue that this is already a common intuition
latent in much prior work on grammars, especially when comparing
different formalisms for equivalence such as \cite{??}. This notion is
also completely natural in the context of constructive type theories
such as Agda which until recently did not include a universe of
propositions, and so for example in Elliott (\cite{??}) a formal
language was defined as a function to the universe $\Type$ with little
comment.

Every formal grammar $G$ induces a formal language, which can be seen
most easily by using the characteristic function formulation: we can
``squash'' any set into the proposition that it is inhabited, so
$\chi_G(w) = ||G(w)||$ which defines a subset $L_G = \{ w \in
\Sigma^*| G(w) \mathrm{inhabited}\}$.

Formal languages naturally arrange themselves into a partial order by
using the subset inclusion as the ordering. Formal grammars naturally
arrange themselves into a \emph{category} where a morphism $\alpha : G
\to H$ is given by a \emph{family} of functions
\[ \alpha^w : Gw \to Hw \]
for every string $w \in \Sigma^*$. The intuition is that a morphism of
grammars is a \emph{translation of parses}: a $G$-parse of $w$ gets
transformed to an $H$-parse of $w$. This then induces a notion of
\emph{isomorphism} of grammars: two grammars are equivalent if there
is a bijective translation of parses, precisely capturing the notion
of a \emph{strong equivalence} of grammars.

This category-theoretic framework can be further used to describe and
compare different notions of formal grammar. We can define a
\emph{notion of grammar} to be a category (or perhaps just groupoid)
paired with a functor into $\Set^{\Sigma^*}$. This naturally forms a
2-category the slice 2-category over $\Set^{\Sigma^*}$ and equivalence
in this 2-category defines a sensible notion of \emph{strong
equivalence of grammar formalisms} (TODO: check details).

The category of grammars is quite rich in structure, which can be most
easily observed from the fact that it is equivalently defined as the
category of functors $\Set^{\Sigma^*}$ where $\Sigma^*$ is here viewed
as a discrete category, that is, the objects are strings and only
morphisms are identity morphisms. Such functor categories, often
called presheaf categories, are incredibly rich in structure, which we
will exploit in section \ref{blah} to succinctly define the semantics
of regular and context-free grammars.

\section{Future work}

\subsection{Beyond Strings}

While parsing typically refers to the production of semantic objects
from their description as strings, many tasks in programming can be
viewed as parsing of more structured objects such as trees, trees with
binding structure or graphs. Fundamental to programming language
implementation is \emph{type checking}, analogous to language
recognition, or more generally \emph{typed elaboration}, analogous to
parsing, which produces a semantic object in addition to performing
some analysis.

\end{document}

### 2023-12-06 paper/paper.tex
started related work

@@
 \email{maxsnew@umich.edu}
+
+\author{Pedro H. Azevedo de Amorim}
+\affiliation{
+  \department{Department of Computer Science}
+  \institution{University of Oxford}
+  \country{UK}
+}
+\email{pedro.azevedo.de.amorim@cs.ox.ac.uk}
 
@@
 
+\section{Related work}
+
+\paragraph{Kleene Algebra}
+
+Since the early works in the theory of formal languages, Kleene
+algebras have played an important role in its development. They
+generalize the operations known from regular languages by introducing
+operations generalizing language composition, language union and the
+Kleene star.  More generally, they are defined as inequational theory
+where the inequality is meant to capture language containment. This
+theory is extremely successful, having found applications in algebraic
+path problems, theory of programming languages, compiler optimizations
+and more.
+
+A frequently fruitful research direction is exploring varying
+extensions of Kleene algebras, Kleene algebra with tests (KAT) being
+one of the most notable ones. Our approach is radically different from
+most extensions, which usually aim at modifying or adding new
+operations to Kleene algebras, but still keeping it as an inequational
+theory. By adopting a category-theoretic treatment and allowing the
+``order structure'' to encode more information than merely
+inequalities, we were able to extend Kleene algebra to reason about
+parsing as well.
+
 \section{Future work}

### 2023-12-12 paper/paper.tex
Kleene categories

@@
 \usepackage{fancyvrb}
+
+\newcommand{\cat}[1]{\mathbf{#1}}
 
@@
 
+\section{Kleene Category}
+
+An important tool from the theory of formal languages was the development
+of Kleene algebras, which serve as a theoretical substrate to studying
+various kinds of formal languages. Formally, they are a set equipped
+with three operations and two constants satisfying the axioms of
+\ref{fig:axioms}.
+
+\begin{figure}
+  \begin{align*}
+    x + (y + z) &= (x + y) + z & x + y &= y + x\\
+    x + 0 &= x & x + x &= x\\
+    x(yz) &= (xy)z & x1 &= 1x = x\\
+    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
+    x0 &= 0x = x & & \\
+    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
+     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
+  \end{align*}
+  \label{fig:axioms}
+  \caption{Kleene algebra axioms}
+\end{figure}
+
+The addition operation can be used to define a partial order structure
+$a \leq b$ if $a + b = b$. In the theory of formal languages this
+order structure can be used to model language containment. What we
+want to do in this section is to categorify the concept of Kleene
+algebra and build on top of this new categorical concept as the
+theoretical basis to the formal theory of parsing.
+
+\begin{definition}
+  A Kleene category is a distributive monoidal category $\cat{K}$
+  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
+  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
+  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
+  X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogue isomorphism
+  for $G_{A,B}$ also holds.
+\end{definition}
+
+\begin{example}
+  Every Kleene algebra, seen a posetal category, is a Kleene category.
+\end{example}
+
+\begin{example}
+  The opposite category of every Kleene category is a model of a variant of
+  conjunctive ordered logic, where the Kleene star plays the role of the ``of
+  course'' modality from substructural logics.
+\end{example}
+
+The proposed axioms are a direct translation of the Kleene algebra
+axioms to a categorical setting. Its most unusual aspect is the
+axiomatization of the Kleene star as a family os initial algebras
+satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
+more structure, then these isomorphisms hold ``for free''.
+
+\begin{theorem}
+  \label{th:kleeneclosed}
+  Let $\cat{K}$ be a Kleene category such that it is also monoidal
+  closed.  Then, the initial algebras isomorphisms hold automatically.
+\end{theorem}
+\begin{proof}
+  We prove this by the unicity (up-to isomorphism) of initial
+  algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
+  (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
+  X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
+  (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
+    1}(X))$.
+
+  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
+  we want to show that there is a unique algebra morphism $h : B
+  \otimes \mu X.\, F_{A,1} \to Y$. We can show existence and uniqueness
+  by showing that the diagram on the left commute if, and only if,
+  the diagram on the right commutes:
+
+  This equivalence follows by using the adjunction structure given
+  by the monoidal closed structure of $\cat{K}$. A completely analogous
+  argument for $G_{A,B}$ also holds.
+\end{proof}
+
+This result feels similar in spirit to the definition of action
+algebras, which are algebras where the product also has adjoint
+operations which results in the Kleene star being more easily
+axiomatized \cite{kozen1994}.
+
+%% \begin{theorem}
+%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
+%%   The functor category $[A, \cat{K}]$.
+%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
+%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
+%% \end{theorem}
+
+\begin{theorem}
+  If $\cat{C}$ is a locally small monoidal category, then the category
+  $\cat{Set}^{\cat{C}}$ is a Kleene category.
+\end{theorem}
+\begin{proof}
+  The coproduct structure exists since it is a presheaf category. The
+  monoidal structure is defined using the Day convolution and,
+  furthermore, this monoidal structure has a closed structure as well.
+
+  From this closed structure it follows that it is a distributive
+  monoidal category, since left adjoint functors preserve colimits. As
+  for the Kleene star, since presheaf categories admit small colimits,
+  the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
+  defined as the filtered colimit of the diagrams:
+
+  From Theorem~\ref{th:kleeneclosed} it follows that
+  these initial algebras satisfy the required isomorphisms.
+\end{proof}
+
+\begin{corollary}
+  For every alphabet $\Sigma$, the presheaf category $\cat{Set}^{\cat{\Sigma^*}}$
+  is a Kleene category.
+\end{corollary}
+\begin{proof}
+  Note that string concatenation makes the discrete category
+  $\Sigma^*$ a strict monoidal category.
+\end{proof}
+
 \section{Related work}

### 2023-12-14 paper/paper.tex
draft comments, typo fixes, and some comments

@@
 \newcommand{\cat}[1]{\mathbf{#1}}
+\newcommand{\Set}{\mathbf{Set}}
+
+\newif\ifdraft
+\drafttrue
+\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven}: #1]}\fi}
+\newcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max}: #1]}\fi}
+\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro}: #1]}\fi}
 
@@
 of regular and context-free grammars.
+\steven{Topoi model dependent type theory.
+This dependence is seemingly useful for also reasoning about context sensitivity, but this doesn't work for \textit{reasons}, right?}
 
@@
 
+\steven{What is the ``of course'' modality? Is this ``$!$'' in linear logic?}
+
 The proposed axioms are a direct translation of the Kleene algebra
 axioms to a categorical setting. Its most unusual aspect is the
-axiomatization of the Kleene star as a family os initial algebras
+axiomatization of the Kleene star as a family of initial algebras
 satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has

### 2023-12-27 paper/paper.tex
some notes on canonicity

@@
-\documentclass[acmsmall,anonymous,review,screen]{acmart}
+\documentclass[sigconf,anonymous,review,screen]{acmart}
 \usepackage{mathpartir}
@@
 \usepackage{fancyvrb}
+\usepackage{stmaryrd}
 
@@
 \newcommand{\Set}{\mathbf{Set}}
+\newcommand{\Type}{\mathbf{Type}}
+\newcommand{\Prop}{\mathbf{Prop}}
+\newcommand{\Bool}{\mathbf{Bool}}
 
@@
 \newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven}: #1]}\fi}
-\newcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max}: #1]}\fi}
+\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max}: #1]}\fi}
 \newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro}: #1]}\fi}
@@
 
-% TODO: add Pedro's info
-
 \author{Max S. New}
@@
   Algebra.
+\end{abstract}
 
-\end{abstract}
+\maketitle
 
@@
 
+\section{Universal Constructions in the Category of Grammars}
+
+Predicate BI: bicartesian closed, monoidal biclosed, omega-colimits
+and therefore certain initial algebras
+
+\section{A Bunched Type Theory for Grammars}
+
+\subsection{Semantics}
+
+Set types $A$ and set contexts $\Delta$ denotes sets, and
+grammar types $\gamma$ and grammar contexts $\Gamma$ in context
+$\Delta$ denote families of grammars $(\Set^{\Sigma^*})^\Delta$.  A
+
+\subsection{Canonicity}
+
+Let $\phi$ be a propositional BI formula over $\Sigma$. Then there are
+two obvious notions of what constitutes a ``parse'' of a string w
+according to the grammar $\phi$:
+\begin{enumerate}
+\item On the one hand we have the semantics just defined, $\llbracket \phi \rrbracket w$
+\item On the other hand, we can view the string $w$ as a formula
+  $\lceil w \rceil$ and define a parse to be an equivalence classes of
+  proofs $x:\lceil w \rceil \vdash \phi$
+\end{enumerate}
+
+It is not difficult to see that at least for the ``strictly positive''
+formulae (those featuring only the positive connectives
+$\vee,\bot,\epsilon,\otimes,\mu$ and the generators $c$) that every
+element $t \in \llbracket \phi \rrbracket w$ is a kind of tree and
+that the nodes of the tree correspond precisely to the introduction
+forms of the language. However it is far less clear that every proof
+$x:\lceil w \rceil \vdash p : \phi$ corresponds to some sequence of
+introduction forms since proofs can include elimination forms as
+well. So we want to prove a kind of \emph{canonicity} result for the
+calculus, a generalization of the disjunction theorem of constructive
+logic:
+
+\begin{enumerate}
+\item Every proof $\lceil w \rceil \vdash p : \phi \otimes \psi$ is equal to $(q,r)$ for some $q,r$
+\item Every proof $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
+\item Every proof $\lceil w \rceil \vdash p : \phi \vee \psi$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
+\item Every proof $\lceil w \rceil \vdash p : c$ is equal to $x:c$
+\item Every proof $\lceil w \rceil \vdash p : \mu X. \phi$ is equal to $\textrm{roll}(q)$ where $q : \phi(\mu X.\phi/X)$
+\end{enumerate}
+
+To prove this result we will use a simple Kripke logical relation.
+
+First, let $L$ be the category of BI formulae and proofs (quotiented
+by $\beta\eta$ equality). Define a functor $N : L \to \Set^{\Sigma^*}$ by
+\[ N(\phi)(w) = L(w,\phi) \]
+
+Then define the gluing category $\mathcal G$ as the comma category
+$\Set^{\Sigma^*}/N$. That is, an object of this category is a pair of
+a formula $\phi \in L$ and an object $S \in \mathcal
+P(\Sigma^*)/N(\phi)$. We can then use the equivalence $\mathcal
+P(\Sigma^*)/N(\phi) \cong \mathcal P(\int N(\phi))$ to get a simple
+description of such an $S$: it is simply a family of sets indexed by
+proofs $L(w,\phi)$:
+\[ \prod_{w\in\Sigma^*} L(w,\phi) \to \Set \]
+This category clearly comes with a projection functor $\pi : \mathcal
+G \to \mathcal L$ and then our goal is to define a section by using
+the universal property of $\mathcal L$.
+
+To this end we define
+\begin{enumerate}
+\item $(\phi, S) \otimes (\psi, T) = (\phi \otimes \psi, S\otimes T)$ where
+  \[ (S \otimes T)(w, p) = (w_1w_2 = w) \times (q_1,q_2 = p) \times S\,w_1\,q_1 \times T\,w_2\,q_2\]
+\item $(\phi, S) \multimap (\psi, T) = (\phi \multimap \psi, S \multimap T)$ where
+  \[ (S \multimap T)(w,p) = w' \to q \to S\,w'\,q \to T (ww') (p\,q) \]
+\item $\mu X. ??$ ??
+\end{enumerate}
+
+\section{Logical Characterizations of Language Classes}
+
+Our bunched type theory
+
+Predicate BI with existential quantifiers is already enough to define
+formulae that correspond to arbitrary recursively enumerable languages
+since we can simulate the tape of a Turing machine using two
+stacks. I.e., we can define a set $S$ of lists of tape symbols, and
+then define a formula that first ``copies'' the input onto one stack, then executes
+
+\[ \mu \textrm{Copy} : S \to *. \lambda s. (\epsilon \wedge T\,s\,[])\bigvee_{c \in \Sigma} c (S\,{c::s}) \]
+
+\section{Related Work}
+
+TODO
+
+%% That paper with Fritz Henglein about regexes as types
+
+%% Vaughan Pratt on the residual operator
 \section{Future work}

### 2023-12-29 paper/paper.tex
categorical KA

--- 2023-12-12 paper/paper.tex
@@
 \newcommand{\cat}[1]{\mathbf{#1}}
+\newcommand{\Set}{\cat{Set}}
+\newcommand{\lto}{\multimap}
 
@@
 
-An important tool from the theory of formal languages was the development
-of Kleene algebras, which serve as a theoretical substrate to studying
-various kinds of formal languages. Formally, they are a set equipped
-with three operations and two constants satisfying the axioms of
-\ref{fig:axioms}.
+Kleene algebras are an important tool in the theory of regular
+languages. More broadly, they serve as a theoretical substrate to
+studying various kinds of formal languages. Formally, they are a tuple
+$(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is a set, $+$ and $\cdot$
+are binary operations over $A$, $(-)^*$ is a function over $A$, and
+$1$ and $0$ are constants. These structures satisfy the axoims depicted
+in Figure~\ref{fig:axioms}.
 
@@
 
-The addition operation can be used to define a partial order structure
-$a \leq b$ if $a + b = b$. In the theory of formal languages this
-order structure can be used to model language containment. What we
-want to do in this section is to categorify the concept of Kleene
-algebra and build on top of this new categorical concept as the
-theoretical basis to the formal theory of parsing.
+The addition operation can be used to define the partial order
+structure $a \leq b$ if $a + b = b$. In the theory of formal languages
+this order structure can be used to model language containment. In this
+section, we want to categorify the concept of Kleene algebra and
+build on top of it in order to define an abstract theory of parsing.
+We start by defining \emph{Kleene categories}.
 
@@
 
+As a sanity check, note that Kleene algebras are indeed examples of
+Kleene categories.
+
 \begin{example}
@@
 \end{example}
+
+An unexpected example comes from the theory of substructural logics.
 
@@
 
-The proposed axioms are a direct translation of the Kleene algebra
-axioms to a categorical setting. Its most unusual aspect is the
-axiomatization of the Kleene star as a family os initial algebras
-satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
-more structure, then these isomorphisms hold ``for free''.
+This proposed definition is sensible. After all, the proposed axioms
+are a direct translation of the Kleene algebra axioms to a categorical
+setting. Its most convoluted aspect is the axiomatization of the Kleene
+star as a family os initial algebras satisfying certain
+isomorphisms. If the Kleene category $\cat{K}$ has more structure,
+then these isomorphisms hold ``for free''.
 
@@
 operations which results in the Kleene star being more easily
-axiomatized \cite{kozen1994}.
+axiomatized \cite{kozen1994}. We are now ready to prove that our
+concept of formal grammars fit nicely within our categorical
+framework. We start by presenting a well-known construction
+from presheaf categories.
+
+\begin{definition}
+  Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
+  two functors $\cat{C} \to \Set$. Their Day convolution tensor
+  product is defined as the following coend formula:
+  \[
+  (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z) 
+  \]
+  Dually, its internal hom is given by the following end formula:
+  \[
+  (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
+  \]
+\end{definition}
+
+\begin{lemma}[\cite{day1970}]
+  Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
+  monoidal closed.
+\end{lemma}
 
@@
 %% \end{theorem}
-
 \begin{theorem}
-  If $\cat{C}$ is a locally small monoidal category, then the category
-  $\cat{Set}^{\cat{C}}$ is a Kleene category.
+  If $\cat{C}$ is a locally small monoidal category, then
+  $\Set^{\cat{C}}$ is a Kleene category.
 \end{theorem}
 \begin{proof}
-  The coproduct structure exists since it is a presheaf category. The
-  monoidal structure is defined using the Day convolution and,
-  furthermore, this monoidal structure has a closed structure as well.
 
-  From this closed structure it follows that it is a distributive
-  monoidal category, since left adjoint functors preserve colimits. As
-  for the Kleene star, since presheaf categories admit small colimits,
+  By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
+  is a presheaf category, it has coproducts, and since the tensor
+  is a left adjoint, it preserves colimits and, therefore, it is
+  a distributive category.
+
+  As for the Kleene star, since presheaf categories admit small colimits,
   the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
@@
 
-  From Theorem~\ref{th:kleeneclosed} it follows that
-  these initial algebras satisfy the required isomorphisms.
+  From Theorem~\ref{th:kleeneclosed} it follows that these initial
+  algebras satisfy the required isomorphisms and this concludes the
+  proof.
 \end{proof}
@@
 \begin{corollary}
-  For every alphabet $\Sigma$, the presheaf category $\cat{Set}^{\cat{\Sigma^*}}$
+  For every alphabet $\Sigma$, the presheaf category $\Set^{\cat{\Sigma^*}}$
   is a Kleene category.
@@
 \begin{proof}
-  Note that string concatenation makes the discrete category
-  $\Sigma^*$ a strict monoidal category.
+  Note that string concatenation and the empty string make the
+  discrete category $\Sigma^*$ a strict monoidal category.
 \end{proof}

### 2023-12-29 paper/paper.tex
added comment re:glued model

@@
 \newcommand{\cat}[1]{\mathbf{#1}}
+\newcommand{\lto}{\multimap}
 \newcommand{\Set}{\mathbf{Set}}
@@
   conjunctive ordered logic, where the Kleene star plays the role of the ``of
-  course'' modality from substructural logics.
+  course'' modality from substructural logics which allows hypotheses to
+  be discarded or duplicated.
 \end{example}
@@
 operations which results in the Kleene star being more easily
-axiomatized \cite{kozen1994}. We are now ready to prove that our
-concept of formal grammars fit nicely within our categorical
-framework. We start by presenting a well-known construction
-from presheaf categories.
+axiomatized \cite{kozen1994}.
+
+We are now ready to prove that our concept of formal grammars fits
+nicely within our categorical framework. We start by presenting a
+well-known construction from presheaf categories.
 
@@
   By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
-  is a presheaf category, it has coproducts, and since the tensor
-  is a left adjoint, it preserves colimits and, therefore, it is
+  is a presheaf category, it has coproducts. Furthermore, the tensor
+  is a left adjoint, i.e. it preserves colimits and, therefore, it is
   a distributive category.
@@
 \end{proof}
+
+Much like in the posetal case, the abstract structure of a 2-Kleene
+algebra is expressive enough to synthetically reason about formal
+languages. A significant difference between them is that while Kleene
+algebras can reason about language containment, 2-Kleene algebras can
+reason about parsing as well.
+
+For the rest of the paper we will use the presheaf category
+$\Set^{\cat{\Sigma^*}}$ as a concrete model that will serve as our
+guide when expanding abstract formalism presented in this section so
+that it can handle more classes of languages beyond regular ones.
 
@@
 To this end we define
+\pedro{I'm pretty sure that these definitions have been somewhat presented
+  in ``Glueing and orthogonality for models of linear logic'' - Hyland and Schalk.
+We should probably cite them so that we don't have to rederive their constructions and proofs here.}
 \begin{enumerate}

### 2024-01-09 paper/paper.tex
Added typing relations

@@
 \newcommand{\lto}{\multimap}
+\newcommand{\tol}{\rotatebox[origin=c]{180}{$\lto$}}
 \newcommand{\Set}{\mathbf{Set}}
@@
 \newcommand{\Bool}{\mathbf{Bool}}
+
+\newcommand{\subst}[3]{#1\{#2/#3\}}
+\newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
+\newcommand{\lamb}[2]{\lambda #1.\, #2}
+\newcommand{\dlamb}[2]{\hat{\lambda} #1.\, #2}
+\newcommand{\app}[2]{#1 \, #2}
+\newcommand{\PiTy}[3]{\Pi #1 : #2.\, #3}
+\newcommand{\SigTy}[3]{\Sigma #1 : #2.\, #3}
+\newcommand{\LinPiTy}[3]{\widebar\Pi #1 : #2.\, #3}
+\newcommand{\LinSigTy}[3]{\widebar\Sigma #1 : #2.\, #3}
+\newcommand{\amp}{\mathrel{\&}}
+\newcommand{\GrTy}{\mathsf{Gr}}
+
+\newcommand{\ctxwff}[1]{#1 \,\, \mathsf{ok}}
+\newcommand{\ctxwffjdg}[2]{#1 \vdash #2 \,\, \mathsf{type}}
+\newcommand{\linctxwff}[2]{#1 \vdash #2 \,\, \mathsf{ok}}
+\newcommand{\linctxwffjdg}[2]{#1 \vdash #2 \,\, \mathsf{linear}}
 
@@
 
+\subsection{Syntax}
+
+\begin{figure}
+  \label{fig:structjdg}
+  \begin{mathpar}
+    \inferrule{~}{\ctxwff \cdot}
+    \and
+    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}
+
+    \\
+
+    \inferrule{~}{\linctxwff \Gamma \cdot}
+    \and
+    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, x : A}}
+
+    \\
+
+    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}
+
+    \and
+
+    \inferrule{\Gamma \vdash A : L_i}{\linctxwffjdg \Gamma A}
+
+    \\
+
+    \inferrule{\Gamma \vdash X \equiv Y : U_i}{\ctxwffjdg \Gamma {X\equiv Y}}
+
+    \and
+
+    \inferrule{\Gamma \vdash A \equiv B : L_i}{\linctxwffjdg \Gamma {A \equiv B}}
+
+  \end{mathpar}
+  \caption{Structural judgements}
+\end{figure}
+
+\begin{figure}
+  \label{fig:typewf}
+  \begin{mathpar}
+    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
+ %   
+    \and
+%
+    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
+%
+    \\
+%
+    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
+%
+    \and
+%
+    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
+%
+    \\
+%
+    \inferrule{~}{\Gamma \vdash 1 : U_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
+%
+    \\
+%
+    \inferrule{~}{\Gamma \vdash I : L_i}
+ %   
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
+%
+    \\
+%
+    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
+%
+    \\
+%
+    \inferrule{~}{\Gamma \vdash \top : L_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \\ \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
+%
+    \\
+%
+    \inferrule{~}{\Gamma \vdash \GrTy : L_1}
+%
+    \and
+%
+    \mprset { fraction ={===}}
+    \inferrule{\Gamma \vdash X : L_1}{\Gamma \vdash : \GrTy}
+%
+    \and
+%
+    \mprset { fraction ={---}}
+    \inferrule{\Gamma, x : \GrTy \vdash A : L_i}{\Gamma \vdash \mu x.\, A : L_i}
+  \end{mathpar}
+  \caption{Type well-formedness}
+\end{figure}
+
+\begin{figure}
+  \label{fig:inttyping}
+  \begin{mathpar}
+  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
+  %
+  \and
+  %
+  \inferrule{~}{\Gamma \vdash () : 1}
+  %
+  \\
+  %
+  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e; : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
+  %
+  \and
+%
+  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
+  %
+  \\
+  %
+  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
+  \and
+  \inferrule{\ctxwffjdg \Gamma {\PiTy x X Y} \\ \Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
+  %
+  \\
+  %
+  \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
+  %
+  \\
+  %
+  \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
+  \and
+  \inferrule{\Gamma \mid \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
+  \end{mathpar}
+  \caption{Intuitionistic typing}
+\end{figure}
+
+\begin{figure}
+  \label{fig:linsyntax}
+  \begin{mathpar}
+    \inferrule{~}{\Gamma \mid a : A \vdash a : A}
+    \and
+    \inferrule{\Gamma \mid \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma \mid \Delta \vdash e : A}
+    %
+    \\
+    %
+    \inferrule{~}{\Gamma \mid \cdot \vdash () : I}
+    \and
+    \inferrule{\Gamma \mid \Delta \vdash e : I \\ \Gamma \mid \Delta' \vdash e' : C}{\Gamma \mid \Delta,\Delta' \vdash \letin {()} e {e'} : C}
+    %
+    \\
+    %
+    \inferrule{\Gamma \mid \Delta \vdash e : A \\ \Gamma \mid \Delta' \vdash e' : B}{\Gamma \mid \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
+    %
+    \\
+    %
+    \inferrule{\Gamma \mid \Delta \vdash e : A \otimes B \\ \Gamma \mid \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma \mid  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
+    \\
+    %
+    \inferrule{\Gamma \mid \Delta, a : A \vdash e : B}{\Gamma \mid \Delta \vdash \lamb a e : A\lto B}
+    \and
+    \inferrule{\Gamma \mid \Delta \vdash e : A \lto B \\ \Gamma \mid \Delta' \vdash e' : A}{\Gamma \mid \Delta, \Delta' \vdash \app e {e'} : A}
+    \\
+    %
+    \inferrule{\Gamma \mid a : A, \Delta \vdash e : B}{\Gamma \mid \Delta \vdash \lamb a e : A\tol B}
+    \and
+    \inferrule{\Gamma \mid \Delta \vdash e : A \tol B \\ \Gamma \mid \Delta' \vdash e' : A}{\Gamma \mid \Delta, \Delta' \vdash \app e {e'} : A}
+    %
+    \\
+    %
+    \inferrule{\Gamma, x : X \mid \Delta  \vdash e : A}{\Gamma \mid \Delta \vdash \dlamb x e : \LinPiTy x X A}
+    \and
+    \inferrule{\Gamma \mid \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma \mid \Delta \vdash \app e {e'} : \subst A {e'} x}
+    %
+    \\
+    %
+    \inferrule{~}{\Gamma \mid \Delta \vdash () : \top}
+    %
+    \\
+    %
+    \inferrule{\Gamma \mid \Delta \vdash e_1 : A_1 \quad \Gamma \mid \delta \vdash e_2 : A_2}{\Gamma \mid \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
+    \and
+    \inferrule{\Gamma \mid \Delta \vdash e : A_1 \amp A_2 }{\Gamma \mid \Delta \vdash \pi_i \, e : A_i}
+    %
+    \\
+    %
+    \inferrule{\Gamma \vdash e : X \quad \Gamma \mid \Delta \vdash e' : \subst A e x}{\Gamma \mid \Delta \vdash (e, e') : \LinSigTy x X A}
+    %
+    \\
+    %
+    \inferrule{\Gamma \mid \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X \mid \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma\mid \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
+    %
+    \\
+    %
+    \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma \mid \cdot \vdash \mathsf{G}^{-1}\, e : A}
+    %
+    \\
+    %
+    \inferrule{~}{\Gamma\mid \cdot \vdash \varepsilon : \GrTy}
+    \and
+    \inferrule{\sigma \in \Sigma}{\Gamma \mid \cdot \vdash \sigma : \GrTy}
+    %
+    \\
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \mathsf{fold}\, t : \mu X.\, A}
+    \and
+    \inferrule{~}{\Gamma; \Delta \vdash \mathsf{unfold}\, t : \mu X.\, A}
+  \end{mathpar}
+  \caption{Linear typing}
+\end{figure}
+
+\pedro{Need to figure out the rules for inductive types and write down
+the judgemental equality rules}
 \subsection{Semantics}
@@
 \pedro{I'm pretty sure that these definitions have been somewhat presented
-  in ``Glueing and orthogonality for models of linear logic'' - Hyland and Schalk.
+  in ``Glueing and orthogonality for models of linear logic'' by Hyland and Schalk.
 We should probably cite them so that we don't have to rederive their constructions and proofs here.}

### 2024-01-11 paper/paper.tex
some notes from over break

@@
 \newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro}: #1]}\fi}
+\newcommand{\pipe}{\,|\,}
 
@@
 \begin{document}
-\title{Formal Grammars as Types in Non-commutative Bunched Type Theory}
+\title{Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Bunched Type Theory}
 \author{Steven Schaefer}
@@
 \begin{abstract}
-  We propose the category of functions from strings to sets as a
-  universal syntax-independent model of formal grammars, generalizing
-  the set of strings model of formal language theory to account for
-  ambiguity. The morphisms of this category give a notion of parse
-  transformer, and isomorphism corresponds to \emph{strong
-  equivalence} of grammars. We observe that the standard algebraic
-  constructions of regular expressions and their extension to
-  mu-regular expressions correspond to universal constructions in this
-  category, which we axiomatize as a structure we call a 2-Kleene
-  Algebra.
+  We propose a semantic framework for the study of formal
+  grammars. First, we provide a syntax-independent notion of formal
+  grammar as a function from strings to sets, generalizing the
+  familiar notion of a language as a set of strings. The set of such
+  grammars naturally form a very rich category whose notion of
+  isomorphism corresponds to strong equivalence of grammars, and for
+  which many language-theoretic constructs can be defined using
+  universal properties.
+
+  Based on these category-theoretic constructions we propose a new
+  syntactic formalism for formal grammars: a two-level non-commutative
+  bunched dependent type theory whose types can be interpreted as
+  grammars, and terms as parse transformers. The non-dependent
+  fragment of this type theory is already enough to express
+  regular/context-free expressions as well as finite automata,
+  generalizing the language theoretic formalisms such as Kleene
+  algebras to incorporate language ambiguity. By incorporating a
+  restricted form of type dependency, we can further express indexed
+  grammars, pushdown automata and Turing machines. This provides a new
+  form of ``logical characterization'' of grammar classes based on a
+  substructural logic rather than ordinary first-order logic. Further,
+  the type theory can not only express the grammars as types, but
+  equivalences between grammars and parsers as terms of the grammar.
+
+  We give this type theory a semantics in our category of grammars and
+  prove a canonicity theorem that shows that every term in a context
+  corresponding to a fixed string is equal to a term in a canonical
+  form encoding a parse tree of that string.
 \end{abstract}
@@
 \emph{recognizer}: a program $r$ of type $\Sigma^* \to \Bool$ such
-that $r(w) = \texttt{true}$ if and only if $\chi_L(w)$, but
+that $r(w) = \texttt{true}$ if and only if $\chi_L(w)$ holds, but
 recognition is insufficient for most tasks for which the theory of
@@
 
+\paragraph{TODO: more stuff}
+
+%% That paper with Fritz Henglein about regexes as types
+
+%% Vaughan Pratt on the residual operator
+
 \section{Universal Constructions in the Category of Grammars}
@@
 \subsection{Syntax}
+Judgments: $\Delta$ set ctx, $\Delta \vdash A$ set, $\Delta \vdash M :
+A$ term, $\Delta \vdash \Gamma$ grammar ctx, $\Delta \vdash G$
+grammar, $\Delta\pipe \Gamma \vdash p : G$ parse term.
 
@@
 
-Set types $A$ and set contexts $\Delta$ denotes sets, and
-grammar types $\gamma$ and grammar contexts $\Gamma$ in context
-$\Delta$ denote families of grammars $(\Set^{\Sigma^*})^\Delta$.  A
+Propositional fragment: monoidal biclosed category that is also
+bicartesian closed and has $\omega$-colimits. Call such a BIC.
+
+Set theoretic fragment: category with families $\mathcal C$ + $\Pi,\Sigma,W$.
+
+Combination: category with families $\mathcal C$ + a functor $L : C
+\to \textrm{BIC}$.
+
+Grammatical fragment: functor from $\mathcal C$ to the category of BI-models
+
+Example: use $\Set$ as the CwF, use the functor $\Delta \mapsto
+(\Set^{\Sigma^*})^\Delta$ as the model of the propositional fragment
+using the pointwise-Day convolution monoidal structure.
 
@@
 
-Let $\phi$ be a propositional BI formula over $\Sigma$. Then there are
+Let $\cdot \vdash G$ be a grammar. Then there are
 two obvious notions of what constitutes a ``parse'' of a string w
-according to the grammar $\phi$:
+according to the grammar $G$:
 \begin{enumerate}
-\item On the one hand we have the semantics just defined, $\llbracket \phi \rrbracket w$
-\item On the other hand, we can view the string $w$ as a formula
-  $\lceil w \rceil$ and define a parse to be an equivalence classes of
-  proofs $x:\lceil w \rceil \vdash \phi$
+\item On the one hand we have the semantics just defined, $\llbracket G \rrbracket w$
+\item On the other hand, we can view the string $w$ as a grammar
+  $\lceil w \rceil$ and define a parse to be a parse terms $\cdot\pipe
+  x:\lceil w \rceil \vdash p : G$ modulo $\beta\eta$ equivalence.
 \end{enumerate}
@@
 formulae (those featuring only the positive connectives
-$\vee,\bot,\epsilon,\otimes,\mu$ and the generators $c$) that every
-element $t \in \llbracket \phi \rrbracket w$ is a kind of tree and
-that the nodes of the tree correspond precisely to the introduction
-forms of the language. However it is far less clear that every proof
-$x:\lceil w \rceil \vdash p : \phi$ corresponds to some sequence of
-introduction forms since proofs can include elimination forms as
-well. So we want to prove a kind of \emph{canonicity} result for the
+$0,+,\epsilon,\otimes,\mu$ and the generators $c$) that every element
+$t \in \llbracket G \rrbracket w$ is a kind of tree and that the nodes
+of the tree correspond precisely to the introduction forms of the
+type. However it is far less obvious that \emph{every} proof $x:\lceil
+w \rceil \vdash p : \phi$ is equal to some sequence of introduction
+forms since proofs can include elimination forms as well. To show that
+this is indeed the case we give a \emph{canonicity} result for the
 calculus, a generalization of the disjunction theorem of constructive
 logic:
-
 \begin{enumerate}
-\item Every proof $\lceil w \rceil \vdash p : \phi \otimes \psi$ is equal to $(q,r)$ for some $q,r$
-\item Every proof $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
-\item Every proof $\lceil w \rceil \vdash p : \phi \vee \psi$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
-\item Every proof $\lceil w \rceil \vdash p : c$ is equal to $x:c$
-\item Every proof $\lceil w \rceil \vdash p : \mu X. \phi$ is equal to $\textrm{roll}(q)$ where $q : \phi(\mu X.\phi/X)$
+\item Every term $\lceil w \rceil \vdash p : G + H$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
+\item There are no terms $\lceil w \rceil \vdash p : 0$
+\item If there is a term $\lceil w \rceil \vdash p : c$ then $w = c$ and $p = x$.
+\item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
+\item Every term $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
+\item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
+\item Every term $\lceil w \rceil \vdash p : \mu X. G$ is equal to $\textrm{roll}(q)$ where $q : G(\mu X.G/X)$
+\item Every term $\lceil w \rceil \vdash p : (x:A) \times G$ is equal
+  to $(M,q)$ where $\cdot \vdash M : A$
 \end{enumerate}
 
-To prove this result we will use a simple Kripke logical relation.
+To prove this result we will use a logical families construction by
+constructing a second model. First, the category with families will be
+the category of logical families over set contexts/types
+$\Delta$/$A$. Then the propositional portion will be defined by
+mapping a logical family $\hat \Gamma \to \Gamma$
 
@@
 
-\section{Related Work}
-
-TODO
-
-%% That paper with Fritz Henglein about regexes as types
-
-%% Vaughan Pratt on the residual operator
 \section{Future work}
@@
 parsing, which produces a semantic object in addition to performing
-some analysis.
+some analysis. A type system could then be interpreted as a \emph{tree
+grammar}. Our definition of string grammars as functors from
+$\Sigma^*$ to $\Set$ can then be easily adapted to functors to $\Set$
+from the set of \emph{trees} freely generated form some signature of
+nodes. Then just as the set of string grammars inherits a monoidal
+structure from the free monoid of strings using the Day convolution,
+this set of tree grammars inherits a kind of weak algebraic structure
+corresponding to the tree constructors. This suggests an even more
+unusual form of bunched type theory, where the monoidal concatenation
+and unit are replaced with context constructors corresponding to the
+tree constructors.
 

### 2024-01-11 paper/paper.tex
fix some rules of the syntax

@@
 \newcommand{\lto}{\multimap}
-\newcommand{\tol}{\rotatebox[origin=c]{180}{$\lto$}}
+\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
 \newcommand{\Set}{\mathbf{Set}}
@@
 \newcommand{\lamb}[2]{\lambda #1.\, #2}
+\newcommand{\lamblto}[2]{\lambda^{{\lto}} #1.\, #2}
+\newcommand{\lambtol}[2]{\lambda^{{\tol}} #1.\, #2}
 \newcommand{\dlamb}[2]{\hat{\lambda} #1.\, #2}
 \newcommand{\app}[2]{#1 \, #2}
+\newcommand{\applto}[2]{#1 \mathop{{}^{\lto}} #2}
+\newcommand{\apptol}[2]{#1 \mathop{{}^{\tol}} #2}
 \newcommand{\PiTy}[3]{\Pi #1 : #2.\, #3}
@@
 \begin{document}
-\title{Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Bunched Type Theory}
+\title{Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}
 \author{Steven Schaefer}
@@
   Based on these category-theoretic constructions we propose a new
-  syntactic formalism for formal grammars: a two-level non-commutative
-  bunched dependent type theory whose types can be interpreted as
-  grammars, and terms as parse transformers. The non-dependent
-  fragment of this type theory is already enough to express
-  regular/context-free expressions as well as finite automata,
-  generalizing the language theoretic formalisms such as Kleene
-  algebras to incorporate language ambiguity. By incorporating a
-  restricted form of type dependency, we can further express indexed
+  syntactic formalism for formal grammars: a version of dependent
+  linear-non-linear type theory in which the tensor product is not
+  commutatiev. The linear types can be interpreted as grammars, and
+  linear terms as parse transformers. The non-dependent fragment of
+  this type theory is already enough to express regular/context-free
+  expressions as well as finite automata, with the isomorphism between
+  regular languages and automata definable as linear terms. This
+  generalizes the language theoretic formalisms such as Kleene
+  algebras to incorporate ambiguity. By incorporating a dependency of
+  linear types on non-linear types, we can further express indexed
   grammars, pushdown automata and Turing machines. This provides a new
@@
 
-  We give this type theory a semantics in our category of grammars and
+  We give this type theory a semantics in the category of grammars and
   prove a canonicity theorem that shows that every term in a context
@@
     \and
-    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, x : A}}
+    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}
 
@@
 %
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
+%
     \\
@@
 %
-    \inferrule{~}{\Gamma \vdash \GrTy : L_1}
-%
-    \and
-%
-    \mprset { fraction ={===}}
-    \inferrule{\Gamma \vdash X : L_1}{\Gamma \vdash : \GrTy}
-%
-    \and
-%
-    \mprset { fraction ={---}}
-    \inferrule{\Gamma, x : \GrTy \vdash A : L_i}{\Gamma \vdash \mu x.\, A : L_i}
+    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
   \end{mathpar}
@@
   %
-  \and
+  \\\
   %
@@
   %
+  \and
+  %
+  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
+  %
   \\
 %
-  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e; : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
+  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
   %
@@
   %
-  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
-  %
-  \\
-  %
   \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
   \and
-  \inferrule{\ctxwffjdg \Gamma {\PiTy x X Y} \\ \Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
+  \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
   %
-  \\
+  \and
   %
@@
     \and
-    \inferrule{\Gamma \mid \Delta \vdash e : I \\ \Gamma \mid \Delta' \vdash e' : C}{\Gamma \mid \Delta,\Delta' \vdash \letin {()} e {e'} : C}
+    \inferrule{\Gamma \mid \Delta \vdash e : I \\ \Gamma \mid \Delta_1',\Delta_2' \vdash e' : C}{\Gamma \mid \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
     %
@@
     %
-    \inferrule{\Gamma \mid \Delta, a : A \vdash e : B}{\Gamma \mid \Delta \vdash \lamb a e : A\lto B}
+    \inferrule{\Gamma \mid \Delta, a : A \vdash e : B}{\Gamma \mid \Delta \vdash \lamblto a e : A\lto B}
     \and
-    \inferrule{\Gamma \mid \Delta \vdash e : A \lto B \\ \Gamma \mid \Delta' \vdash e' : A}{\Gamma \mid \Delta, \Delta' \vdash \app e {e'} : A}
+    \inferrule{\Gamma \mid \Delta' \vdash e' : A \\ \Gamma \mid \Delta \vdash e : A \lto B}{\Gamma \mid \Delta', \Delta \vdash \applto {e'} {e} : B}
     \\
     %
-    \inferrule{\Gamma \mid a : A, \Delta \vdash e : B}{\Gamma \mid \Delta \vdash \lamb a e : A\tol B}
+    \inferrule{\Gamma \mid a : A, \Delta \vdash e : B}{\Gamma \mid \Delta \vdash \lambtol a e : B\tol A}
     \and
-    \inferrule{\Gamma \mid \Delta \vdash e : A \tol B \\ \Gamma \mid \Delta' \vdash e' : A}{\Gamma \mid \Delta, \Delta' \vdash \app e {e'} : A}
+    \inferrule{\Gamma \mid \Delta \vdash e : B \tol A \\ \Gamma \mid \Delta' \vdash e' : A}{\Gamma \mid \Delta, \Delta' \vdash \apptol e {e'} : B}
     %
@@
     \inferrule{\Gamma \mid \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma \mid \Delta \vdash \app e {e'} : \subst A {e'} x}
+    %
+    \\
+    %
+    \inferrule{\Gamma \vdash e : X \quad \Gamma \mid \Delta \vdash e' : \subst A e x}{\Gamma \mid \Delta \vdash (e, e') : \LinSigTy x X A}
+    %
+    \\
+    %
+    \inferrule{\Gamma \mid \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X \mid \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma\mid \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
     %
@@
     %
-    \inferrule{\Gamma \vdash e : X \quad \Gamma \mid \Delta \vdash e' : \subst A e x}{\Gamma \mid \Delta \vdash (e, e') : \LinSigTy x X A}
-    %
-    \\
-    %
-    \inferrule{\Gamma \mid \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X \mid \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma\mid \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
-    %
-    \\
-    %
     \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma \mid \cdot \vdash \mathsf{G}^{-1}\, e : A}
@@
     %
-    \inferrule{~}{\Gamma\mid \cdot \vdash \varepsilon : \GrTy}
+    \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
     \and
-    \inferrule{\sigma \in \Sigma}{\Gamma \mid \cdot \vdash \sigma : \GrTy}
-    %
-    \\
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \mathsf{fold}\, t : \mu X.\, A}
-    \and
-    \inferrule{~}{\Gamma; \Delta \vdash \mathsf{unfold}\, t : \mu X.\, A}
+    \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
   \end{mathpar}

### 2024-01-11 paper/paper.tex
sketch some of canonicity

@@
 
+\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
 \newcommand{\cat}[1]{\mathbf{#1}}
@@
 \newcommand{\Set}{\mathbf{Set}}
+\newcommand{\Gr}{\mathbf{Gr}}
 \newcommand{\Type}{\mathbf{Type}}
@@
 
-\newcommand{\subst}[3]{#1\{#2/#3\}}
+\newcommand{\simulsubst}[2]{#1\{#2\}}
+\newcommand{\subst}[3]{\simulsubst {#1} {#2/#3}}
 \newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
@@
 the judgemental equality rules}
+
 \subsection{Semantics}
+
+For illustrative purposes, we give a brief overview of the concrete semantics in sets and grammars:
+\begin{enumerate}
+\item A non-linear context $\Gamma$ denotes a set $\sem{\Gamma}$
+\item A non-linear type $\Gamma \vdash X : U_i$ denotes a family of sets $\sem{X} : \sem{\Gamma} \to \Set_i$
+\item A non-linear term $\Gamma \vdash e : X$ denotes a section $\sem{e} : \Pi(\gamma:\sem{\Gamma})\,\sem{X}\,\gamma$
+\item Linear contexts $\Gamma \vdash \Delta$ and types $\Gamma \vdash A : L_i$ both denote families of grammars $\sem{\Gamma}\to \Gr_i$
+\item A linear term $\Gamma; \Delta \vdash e : A$ denotes a family of parse transformers $\sem{e} : \Pi(\gamma:\sem{\Gamma})\Pi(w:\Sigma^*)\,\sem{\Delta}\gamma w \Rightarrow \sem{A}\gamma w)$
+\end{enumerate}
+Most of the non-linear semantics is standard, we show just the $G$ type and the linear semantics:
+\begin{enumerate}
+\item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
+\item $\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+\item $\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+\item $\sem{I} \gamma w = \{ () \pipe w = \epsilon \}$
+\item $\sem{A \otimes B} \gamma w = \Sigma(w1,w2:\Sigma^*) w1w1 = w \wedge \sem{A} \gamma w_1 \times \sem{B} \gamma w_2$
+\item $\sem{A \lto B} \gamma w = \Pi(w_a:\Sigma^*) A \gamma w_a \Rightarrow B\gamma (w_aw)$
+\item $\sem{B \tol A} \gamma w = \Pi(w_a:\Sigma^*) A \gamma w_a \Rightarrow B\gamma (ww_a)$
+\item $\sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x)$
+\end{enumerate}
 
@@
 
-Let $\cdot \vdash G$ be a grammar. Then there are
+Let $\cdot \vdash A$ be a closed linear type. Then there are
 two obvious notions of what constitutes a ``parse'' of a string w
-according to the grammar $G$:
+according to the grammar $A$:
 \begin{enumerate}
-\item On the one hand we have the semantics just defined, $\llbracket G \rrbracket w$
-\item On the other hand, we can view the string $w$ as a grammar
-  $\lceil w \rceil$ and define a parse to be a parse terms $\cdot\pipe
-  x:\lceil w \rceil \vdash p : G$ modulo $\beta\eta$ equivalence.
+\item On the one hand we have the semantics just defined, $\llbracket A \rrbracket \cdot w$
+\item On the other hand, we can view the string $w = c_1c_2\cdots$ as
+  a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
+  define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
+  \lceil w \rceil \vdash e : G$.
 \end{enumerate}
-
 It is not difficult to see that at least for the ``strictly positive''
 formulae (those featuring only the positive connectives
-$0,+,\epsilon,\otimes,\mu$ and the generators $c$) that every element
-$t \in \llbracket G \rrbracket w$ is a kind of tree and that the nodes
-of the tree correspond precisely to the introduction forms of the
-type. However it is far less obvious that \emph{every} proof $x:\lceil
-w \rceil \vdash p : \phi$ is equal to some sequence of introduction
-forms since proofs can include elimination forms as well. To show that
-this is indeed the case we give a \emph{canonicity} result for the
-calculus, a generalization of the disjunction theorem of constructive
-logic:
+$0,+,I,\otimes,\mu$ and the generators $c$) that every element $t \in
+\llbracket A \rrbracket w$ is a kind of tree and that the nodes of the
+tree correspond precisely to the introduction forms of the
+type. However it is far less obvious that \emph{every} linear term
+$\lceil w \rceil \vdash p : \phi$ is equal to some sequence of
+introduction forms since proofs can include elimination forms as
+well. To show that this is indeed the case we give a \emph{canonicity}
+result for the calculus, a generalization of the disjunction theorem
+of constructive logic:
 \begin{enumerate}
@@
 
-To prove this result we will use a logical families construction by
-constructing a second model. First, the category with families will be
+To prove this result we will use a logical families model. We give a
+brief overview of this model concretely:
+\begin{enumerate}
+\item A context $\Gamma$ denotes a family of sets indexed by closing substitutions $\hat\Gamma : (\cdot \vdash \Gamma) \Rightarrow \Set_i$
+\item A type $\Gamma \vdash X : U_i$ denotes a family of sets $\hat X : \Pi(\gamma:\cdot \vdash \Gamma) \hat\Gamma \Rightarrow (\cdot \vdash \simulsubst X \gamma) \Rightarrow \Set_i$
+\item A term $\Gamma \vdash e : X$ denotes a section $\hat e : \Pi(\gamma)\Pi(\hat\gamma)\hat X \gamma \hat\gamma (\simulsubst e \gamma)$
+\item A linear type $\Gamma \vdash A : L_i$ denotes a family of grammars $\hat A : \Pi(\gamma:\cdot\vdash\Gamma)\,\hat\Gamma \Rightarrow \Pi(w:\Sigma^*) (\cdot;\lceil w\rceil \vdash A[\gamma])\Rightarrow \Set_i$, and the denotation of a linear context $\Delta$ is similar.
+\item A linear term $\Gamma;\Delta \vdash e : A$ denotes a function \[\hat e : \Pi(\gamma)\Pi(\hat\gamma)\Pi(w)\Pi(\delta : \lceil w \rceil \vdash \simulsubst \Delta \gamma) \hat\Delta \gamma \hat\gamma \delta \Rightarrow \hat A \gamma \hat\gamma w {(\simulsubst {\simulsubst e \gamma} \delta)}\]
+\end{enumerate}
+And some of the constructions:
+\begin{enumerate}
+\item $\widehat {(G A)} \gamma \hat\gamma e = \hat A \gamma \hat\gamma \varepsilon (G^{-1}e)$
+\item $\widehat {(A \otimes B)} \gamma \hat\gamma w e = \Sigma(w_Aw_B = w)\Sigma(e_A)\Sigma(e_B) (e_A,e_B) = e \wedge \hat A \gamma \hat \gamma w_A e_A \times \hat B \gamma \hat \gamma w_B e_B$
+\item $\widehat {(A \lto B)} \gamma \hat\gamma w e = \Pi(w_A)\Pi(e_A) \hat A \gamma\hat\gamma w_A e_A \Rightarrow \hat B \gamma\hat\gamma (ww_A) (\applto {e_A} e)$
+\end{enumerate}
+
+First, the category with families will be
 the category of logical families over set contexts/types

### 2024-01-15 paper/paper.tex
sketch of the gluing construction for canonicity

@@
 
+\newcommand{\gluedNL}{{\mathcal G}_S}
+\newcommand{\gluedNLUniv}{{\mathcal G}_{S,i}}
+\newcommand{\gluedL}{{\mathcal G}_L}
+
 \newcommand{\simulsubst}[2]{#1\{#2\}}
@@
 \newcommand{\lambtol}[2]{\lambda^{{\tol}} #1.\, #2}
-\newcommand{\dlamb}[2]{\hat{\lambda} #1.\, #2}
+\newcommand{\dlamb}[2]{\overline{\lambda} #1.\, #2}
 \newcommand{\app}[2]{#1 \, #2}
@@
 \begin{document}
+
 \title{Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}
@@
   linear-non-linear type theory in which the tensor product is not
-  commutatiev. The linear types can be interpreted as grammars, and
+  commutative. The linear types can be interpreted as grammars, and
   linear terms as parse transformers. The non-dependent fragment of
@@
 
-\section{Universal Constructions in the Category of Grammars}
+%% \section{Universal Constructions in the Category of Grammars}
 
-Predicate BI: bicartesian closed, monoidal biclosed, omega-colimits
-and therefore certain initial algebras
+%% Predicate BI: bicartesian closed, monoidal biclosed, omega-colimits
+%% and therefore certain initial algebras
 
-\section{A Bunched Type Theory for Grammars}
+\section{Non-commutative Linear-Non-Linear Type Theory as a Syntax for Grammars}
 
@@
 
-\subsection{Semantics}
+\subsection{Semantics in Grammars}
 
@@
 \begin{enumerate}
-\item On the one hand we have the semantics just defined, $\llbracket A \rrbracket \cdot w$
+\item On the one hand we have the set-theoretic semantics just
+  defined, $\llbracket A \rrbracket \cdot w$
 \item On the other hand, we can view the string $w = c_1c_2\cdots$ as
@@
 \end{enumerate}
-It is not difficult to see that at least for the ``strictly positive''
+It is not difficult to see that at least for the ``purely positive''
 formulae (those featuring only the positive connectives
-$0,+,I,\otimes,\mu$ and the generators $c$) that every element $t \in
-\llbracket A \rrbracket w$ is a kind of tree and that the nodes of the
-tree correspond precisely to the introduction forms of the
-type. However it is far less obvious that \emph{every} linear term
-$\lceil w \rceil \vdash p : \phi$ is equal to some sequence of
-introduction forms since proofs can include elimination forms as
-well. To show that this is indeed the case we give a \emph{canonicity}
-result for the calculus, a generalization of the disjunction theorem
-of constructive logic:
+$0,+,I,\otimes,\mu, \overline\Sigma,c$ and the generators $c$) that
+every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
+that the nodes of the tree correspond precisely to the introduction
+forms of the type. However it is far less obvious that \emph{every}
+linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
+sequence of introduction forms since proofs can include elimination
+forms as well. To show that this is indeed the case we give a
+\emph{canonicity} result for the calculus: that the parses for .
+
+\begin{definition}
+  A non-linear type $X$ is purely positive if it is built up using
+  only finite sums, finite products and least fixed points.
+
+  A linear type is purely positive if it is built up using only finite
+  sums, tensor products, generators $c$, least fixed points and linear
+  sigma types over purely positive non-linear types.
+\end{definition}
+
+\begin{definition}
+  %% Let $X$ be a closed non-linear type. The closed elements $\textrm{Cl}(X)$ of $X$ are the definitional equivalence classes of terms $\cdot \vdash e : X$.
+
+  Let $A$ be a closed linear type. The nerve $N(A)$ is a presheaf on
+  strings that takes a string $w$ to the definitional equivalence
+  classes of terms $\cdot; \lceil w\rceil \vdash e: N(A)$.
+\end{definition}
+
+\begin{theorem}[Canonicity]
+  Let $A$ be a closed, purely positive linear type. Then there is an
+  isomorphism between $\llbracket A\rrbracket$ and $N(A)$.
+\end{theorem}
+\begin{proof}
+  We outline the proof here, more details are in the appendix. The
+  proof proceeds first by a standard logical families construction
+  that combines canonicity arguments for dependent type theory
+  \cite{coquand,etc} with logical relations constructions for linear
+  types \cite{hylandschalk}. It is easy to see by induction that the
+  logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
+  \rrbracket$ and the fundamental lemma proves that the projection
+  morphism $p : \hat A \to N(A)$ has a section, the canonicalization
+  procedure. Then we establish again by induction that
+  canonicalization is also a retraction by observing that introduction
+  forms are taken to constructors.
+\end{proof}
+
 \begin{enumerate}
@@
 
-\section{Logical Characterizations of Language Classes}
-
-Our bunched type theory
+\section{Type-theoretic Characterizations of Language Classes}
 
@@
 
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{references}
+
+\clearpage
+\appendix
+
+\section{Semantics in Sets and Grammars}
+
+A model of Grammar Type Theory is a category with families equipped
+with a functor $L : \mathcal C \to \textrm{GrLog}$ from the category
+of contexts to the category of grammar logic models and strict morphisms.
+
+First, we give the set theoretic model. We use the standard category
+with families structure given by the category of sets and a cumulative
+tower of universes.
+%
+Next, we define $L$ as follows:
+\[ L(\Gamma) = \mathcal P \Sigma^*/\textrm{Disc} \Gamma \]
+Where $\textrm{Disc} \Gamma$ is the presheaf that is constantly $\Gamma$.
+Note that
+\[ \mathcal P \Sigma^*/\textrm{Disc} \Gamma \cong (\Set/\Gamma)/\textrm{Disc} \Sigma^*\]
+where $\textrm{Disc} \Sigma^*$ is the projection $\Gamma \times \Sigma^*
+\to \Gamma$.  Then by the fundamental theorem of topos thoery, $\Set/\Gamma$ is
+a topos. Since $\textrm{Disc} : \Set \to \Set/\Gamma$ is monoidal,
+$\textrm{Disc} \Sigma^*$ carries a monoid structure in $\Set/\Gamma$ and
+this induces a monoidal closed structure on $(\Set/\Gamma)/\textrm{Disc}
+\Sigma^*$. Again by the fundamental theorem, it also has finite
+products and coproducts and a natural numbers object, and so can
+interpret least fixed points.\max{More details about constructing initial algebras?}
+
+Then we can define the remaining type structure as follows. Unraveling
+the above, a non-linear type $\Gamma \vdash X$ is interpreted as an
+object of $\Set/\sem{\Gamma}$ and a linear type $\Gamma \vdash A$ is
+interpreted as an object of $\mathcal P\Sigma^*/\textrm{Disc}\sem{\Gamma}$. We
+define them below in indexed style:
+\begin{enumerate}
+\item $\sem{c} \gamma w = (w = c)$
+\item $\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+\item $\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+\item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
+\item $\sem{L_i}\gamma = (\mathcal P_i\Sigma^*/\textrm{Disc}\sem{\Gamma})_0$
+\end{enumerate}
+
+\section{Canonicity}
+
+For the canonicity argument, we construct a second ``glued'' model
+that contains a projection morphism to the syntactic model.
+
+Let $S$ be the syntactic category with families for the type
+theory. Define the glued category with families to be the comma category
+% https://q.uiver.app/#q=WzAsNSxbMCwwLCJcXG1hdGhjYWwgRyJdLFswLDIsIlMiXSxbMiwyLCJcXFNldCJdLFsyLDAsIlxcU2V0Il0sWzEsMSwiXFxzd2Fycm93Il0sWzMsMl0sWzEsMl0sWzAsMV0sWzAsM11d
+\[\begin{tikzcd}[ampersand replacement=\&]
+	{\gluedNL} \&\& \Set \\
+	\& \rotatebox[origin=c]{-45}{$\Downarrow$} \\
+	S \&\& \Set
+	\arrow[Rightarrow, no head, from=1-3, to=3-3]
+	\arrow["{S(\cdot,-)}"', from=3-1, to=3-3]
+	\arrow[from=1-1, to=3-1]
+	\arrow[from=1-1, to=1-3]
+\end{tikzcd}\]
+This is a standard glued category for canonicity of dependent type
+theory, and carries a category with families structure that we will
+reuse from prior work for the (non-linear) universes,
+$\Pi/\Sigma/\textrm{Id}$ \cite{coquand}. Concretely an object of
+$\mathcal G$ is equivalent to a pair of a syntactic context $\Gamma$,
+and a family $\hat \Gamma$ over (equivalence classes of) closing
+substitutions for $\Gamma$.
+
+Next we turn to interpreting the linear types and morphisms. For this
+we define a functor $\gluedL : \gluedNL \to \textrm{MonCat}$ where
+$\textrm{MonCat}$ is the category\footnote{in univalent foundations
+this would be the category of \emph{strict} monoidal categories.} of
+monoidal categories and lax monoidal functors.  We construct
+$\gluedL(p : \hat \Gamma \to S(\cdot,\Gamma))$ as the following comma
+category:
+
+% https://q.uiver.app/#q=WzAsNixbMCwwLCJcXGdsdWVkTkwocCkiXSxbMCwyLCJMX1xcR2FtbWEiXSxbMSwyLCJcXG1hdGhjYWwgUCBcXFNpZ21hXiovXFx0ZXh0cm17RGlzY30oUyhcXGNkb3QsXFxHYW1tYSkpIl0sWzIsMiwiXFxtYXRoY2FsIFAgXFxTaWdtYV4qL1xcdGV4dHJte0Rpc2N9KFxcaGF0XFxHYW1tYSkiXSxbMiwwLCJcXG1hdGhjYWwgUCBcXFNpZ21hXiovXFx0ZXh0cm17RGlzY30oXFxoYXRcXEdhbW1hKSJdLFsxLDEsIlxccm90YXRlYm94W29yaWdpbj1jXXstNDV9e1xcRG93bmFycm93fSJdLFs0LDMsIiIsMCx7ImxldmVsIjoyLCJzdHlsZSI6eyJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsNF0sWzAsMV0sWzEsMiwiTiIsMl0sWzIsMywiXFx0ZXh0cm17RGlzY30ocF4qKSIsMl1d
+\[\begin{tikzcd}[ampersand replacement=\&]
+	{\gluedNL(p)} \&\& {\mathcal P \Sigma^*/\textrm{Disc}(\hat\Gamma)} \\
+	\& {\rotatebox[origin=c]{-45}{$\Downarrow$}} \\
+	{L_\Gamma} \& {\mathcal P \Sigma^*/\textrm{Disc}(S(\cdot,\Gamma))} \& {\mathcal P \Sigma^*/\textrm{Disc}(\hat\Gamma)}
+	\arrow[Rightarrow, no head, from=1-3, to=3-3]
+	\arrow[from=1-1, to=1-3]
+	\arrow[from=1-1, to=3-1]
+	\arrow["N"', from=3-1, to=3-2]
+	\arrow["{\textrm{Disc}(p^*)}"', from=3-2, to=3-3]
+\end{tikzcd}\]
+
+Where $N : L_\Gamma \to {\mathcal P
+  \Sigma^*/\textrm{Disc}(S(\cdot,\Gamma))}$ is the ``dependent nerve''
+of the inclusion $\lceil\cdot\rceil : \Sigma^* \to L_\Gamma$ of
+strings into linear contexts defined in indexed style as
+\[ N(\Gamma \vdash A) \gamma w = \lceil w \rceil \vdash \cdot :  \simulsubst A \gamma \]
+Then observe that $N$ and ${\textrm{Disc}(p^*)}$ are lax monoidal and
+${\mathcal P \Sigma^*/\textrm{Disc}(\hat\Gamma)}$ is a topos so the
+criteria for Hyland and Schalk's gluing construction apply, defining a
+biclosed monoidal structure on $\gluedNL(p)$ as well as finite
+(co-)products. Further, these constructions are all stable under
+reindexing in $\gluedNL$, and so interpret the dependent linear typing
+constructions for $I,\otimes,0,\oplus,\top,\amp,\lto,\tol$.
+
+Concretely, an object of $\gluedNL(p)$ is equivalent to a pair of a
+non-linear type $\Gamma \vdash A$ and a formal grammar $\hat A$
+indexed by $\hat \Gamma$ with a projection function from $\hat A$ to
+$N(A)$. We can think of this presheaf $\hat A$ as the presheaf of
+canonical forms of $\simulsubst A \gamma$ for closing substitutions
+$\gamma$ in contexts of the form $\lceil w \rceil$.
+
+All that remains in this model is to define the semantics of the
+non-standard connectives. We define them in an indexed style. First we
+define the interpretation of a non-linear type $\Gamma \vdash X$ as a
+family of sets $\hat X$ indexed by a closing substitution $\gamma :
+S(\cdot,\Gamma)$, a semantic substitution $\hat \gamma : \hat
+\Gamma(\gamma)$ and a closed term $\cdot \vdash e : \simulsubst A
+\gamma$. Second, we define a linear type $\Gamma \vdash A$ as a family
+$\hat A$ of grammars indexed by $\gamma,\hat\gamma$ as above along
+with a morphism $\pi : \hat A \gamma \hat \gamma \to N(\simulsubst A
+\gamma)$.
+
+\begin{enumerate}
+\item $\hat{c} \gamma \hat \gamma w = (w = c)$ and $\pi(*) = \lceil c \rceil = x_0:c \vdash x_0:c$
+\item $\widehat{(\LinSigTy x X A)} \gamma \hat \gamma w = \Sigma(\cdot \vdash e : \simulsubst X \gamma)\Sigma(\hat x: \hat X \gamma \hat \gamma e) \hat{A}(\gamma,e)(\hat\gamma,\hat x) w$
+\item $\sem{\LinPiTy x X A} \gamma w = \Sigma(x:\simulsubst X \gamma;\lceil w\rceil \vdash e:\simulsubst A \gamma)\Pi(\cdot \vdash e' : \simulsubst X \gamma)\Pi(\hat x:\hat{X}\gamma\hat\gamma x)\Sigma(\hat a : \hat A (\gamma,e')(\hat\gamma,\hat x) \times (\pi(\hat a) = \subst e {e'} x))$
+  and $\pi(e,f) = \dlamb x.e$
+\item $\widehat{G A} \gamma \hat \gamma e = \Sigma(\hat a: \hat{A} \gamma \hat\gamma \varepsilon) (p\gamma\hat\gamma \varepsilon(\hat a) = \mathsf{G}^{-1}e)$
+\item $\sem{L_i}\gamma = \gluedNLUniv(\gamma)$
+\end{enumerate}
+
 \end{document}

### 2024-01-15 paper/paper.tex
sketch of the canonicity proof

@@
 
+Thus we have defined a second model $\gluedNL,\gluedL$ of the type
+theory, with a morphism to the syntactic model. By the initiality of
+the syntactic model, there is a morphism back to the glued model which
+is a section of this projection. For purely positive types, we can
+observe by induction on the type structure that the total space of the
+family is isomorphic to the set-theoretic semantics. Further, the
+projection above defines a section of the families. Finally, by
+induction on formulae again we see that this projection is also a
+section and therefore an isomorphism, establishing the canonicity
+result as stated in Section\ref{section:canonicity}.
 \end{document}

### 2024-01-18 paper/paper.tex
Thompson's construction almost done

@@
 \usepackage{fancyvrb}
+\usepackage{hyperref}
+\usepackage{cleveref}
 \usepackage{stmaryrd}
@@
 of regular and context-free grammars.
-\steven{Topoi model dependent type theory.
-This dependence is seemingly useful for also reasoning about context sensitivity, but this doesn't work for \textit{reasons}, right?}
 
@@
 
-\steven{What is the ``of course'' modality? Is this ``$!$'' in linear logic?}
-
 The proposed axioms are a direct translation of the Kleene algebra
@@
 \section{Non-commutative Linear-Non-Linear Type Theory as a Syntax for Grammars}
-
+\label{sec:typetheory}
 \subsection{Syntax}
@@
     %
-    \inferrule{\Gamma \mid \Delta \vdash e_1 : A_1 \quad \Gamma \mid \delta \vdash e_2 : A_2}{\Gamma \mid \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
+    \inferrule{\Gamma \mid \Delta \vdash e_1 : A_1 \quad \Gamma \mid \Delta \vdash e_2 : A_2}{\Gamma \mid \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
     \and
@@
 formulae (those featuring only the positive connectives
-$0,+,I,\otimes,\mu, \overline\Sigma,c$ and the generators $c$) that
+$0,+,I,\otimes,\mu, \overline\Sigma,c$) that
 every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
@@
 \section{Type-theoretic Characterizations of Language Classes}
-
-Predicate BI with existential quantifiers is already enough to define
+Classically, formal language theory is closely related to automata theory.
+In the styles of Chomsky and Sch\"utzenberger, there is a hierarchy of language classes and the types of automata that recognize them.
+Inside of the type theory of \cref{sec:typetheory}, we can characterize these language classes; moreover, we demonstrate equivalence of each class to the associated automaton as an isomorphism within our logic.
+
+Note, predicate BI with existential quantifiers is already enough to define
 formulae that correspond to arbitrary recursively enumerable languages
@@
 
+We reflect on this expressive power to motivate why we put heavy syntactic restrictions on our presentation of certain types of automata.
+For instance, when defining a pushdown automata, we ought to be very careful about how the set representing the stack can be accessed and added to.
+Careless definitions may accidentally endow us with too much expressive power, launching us further up in the Chomsky hierarchy.
+To begin, let us focus only on language classes, and their associated automata whose parses are definable without any mention to the inuitionistic context.
+In the context of automata, this refers to the class of automata definable without access to any memory --- i.e.\ finite state machines.
+
+\subsection{Finite State Machines}
+We may encode a finite state machine via a particular syntactic form for grammars.
+A \emph{nondeterministic finite automaton} (NFA) is defined as a mutual \footnote{For convenience we refer to a mutual fixed point here, even though we only take single fixed points as primitive.
+As in Beki\'c's theorem, we may decompose our mutual fixed points into a series of single fixed points.} fixed point in the following manner,
+
+\steven{Is there a better way to present this?}
+\[
+  NFA ::= \mu \begin{pmatrix}
+        Y_{s} = \langle Q \rangle \\
+        Y_{1} = \langle Q \rangle \\
+        \vdots \\
+        Y_{e} = I
+        \end{pmatrix}~\textbf{in}~Y_{s}
+\]
+where
+\[
+  Q ::= c \otimes Y ~|~ Y ~|~ Q \oplus Q
+\]
+and $Y \in \{ Y_{s}, Y_{1}, \dots, Y_{N} \}$.
+Intuitively, we read each $Y_{j}$ as the label for a state, and the definitions in form $Q$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$ an $\epsilon$-transition.
+Without loss of generality, $Y_{s}$ is reserved for a single distinguished start state, and $Y_{e}$ is reserved for a single distinguished accepting state.
+A disjunction\footnote{There is an admissible notion of coproduct making use of the primitive dependent pairing --- i.e.
+$A \oplus B := \LinPiTy b \Bool {\textbf{if}~b~A~B}$.
+We will utilize this without much comment to represent the usual notion of alternation of grammars.} of transitions in an NFA grammar corresponds to multiple outgoing edges of a given state in an automaton.
+
+Let $\hat{Y}_{k}$ refer to the syntactic term on the right hand side of the definition of $Y_{k}$ inside of an NFA $N$.
+We say that $A$ is a \emph{deterministic finite automaton} (DFA) if there are no $\epsilon$-transitions and for every definition $\hat{Y}_{k}$ the $c_{j}$ are distinct guards for each produced state.
+\steven{Is this the right notion of determinism?
+Perhaps, it is is better to define determinism as something like: there is at most one derivation for all $j, k$, for all $a \in \Sigma$ we have $x : Y_{k}, y : a \vdash z : Y_{j}$.}
+
+With this setting for finite automata, we can now internalize some classical theorems inside of our formal system.
+First, let us describe a version of Thompson's construction \cite{??} where we show that any regular expression can be recognized by some NFA\@.
+Moreover, we will show that this NFA is strongly equivalent to the original grammar.
+
+\subsection{Equivalence Between Regular Expressions and Finite Automata}
+
+\steven{Define regular expressions}
+\steven{Move proofs to appendix, leave the exposition and the theorem statements.}
+
+\subsubsection{Grammar to NFA}
+
+We will define the recognizing NFA of a regular expression inductively. To handle the atomic grammars --- literals and the empty grammar --- we define the following,
+\[
+  NFA(c) =
+  \mu \begin{pmatrix}
+        Y_{s} = c \otimes Y_{e} \\
+        Y_{e} = I
+  \end{pmatrix}~\textbf{in}~Y_{s}
+\]
+\[
+  NFA(I) =
+  \mu \begin{pmatrix}
+        Y_{s} = Y_{e} \\
+        Y_{e} = I
+  \end{pmatrix}~\textbf{in}~Y_{s}
+\]
+
+The isomorphism between the grammars $c$ and $NFA(c)$ are witnessed by the following derivations,
+\[
+  \inferrule{p : c \vdash p : c \\ \cdot \vdash () : I}{p : c \vdash p \otimes () : Y_{s} = c \otimes I}
+\]
+and
+\[
+  \inferrule{q : Y_{s} = c \otimes I \vdash q : Y_{s} \\ p : c, e : I \vdash p : c }{q : Y_{s} \vdash \letin {p \otimes e} q p : c}
+\]
+
+Similar terms are used to show the equivalence between $I$ and $NFA(I)$.
+
+We can now inductively define some NFA combinators and show that they preserve grammar equivalence.
+First begin with the disjunction of two NFAs.
+hat is, assuming that $g \cong NFA(g)$ and $g' \cong NFA(g')$, we define an NFA that recognizes $g \oplus g'$.
+Let the states in $NFA(g)$ be represented by $X$'s and the states in $NFA(g')$ be represented by $Z$'s,
+
+\[
+  NFA(g \oplus g') =
+  \mu \begin{pmatrix}
+        Y_{s} =  X_{s}' \oplus Z_{s}' \\
+        X_{s}', X_{1}', \dots = \dots \\
+        X'_{e} = Y_{e} \\
+        Z_{s}', Z_{1}, \dots = \dots \\
+        Z_{e}' = Y_{e} \\
+        Y_{e} = I
+  \end{pmatrix}~\textbf{in}~Y_{s}
+\]
+
+For each defintion $X_{j}$ in $NFA(g)$ we create an analogous definition $X_{j}'$ inside of $NFA(g \oplus g')$ with the syntactic equality $X_{j}'$ of $X_{j}$ for $j \neq e$, except we replaec $X_{k}$ with $X_{k}'$ in each production.
+Lastly, $X_{e}'$ has an $\epsilon$-transition to the accepting state $Y_{e}$.
+Likewise, we build analogous defintions for the states $Z_{j}$ in $NFA(g')$, then we flatten all of these defintions into one mutally recursive list.
+
+To show the strong equivalence of $NFA(g \oplus g')$ and $g \oplus g'$, we will demonstrate an isomorphism between them with two proof terms in our logic that mutually invert each other.
+To define a map $NFA(g \oplus g') \to g \oplus g'$ we will use a multi-fold over all the mutually recursive defintions of $NFA(g \oplus g')$.
+Even though multiple folding is not taken as primitive, using Beki\'c's theorem we can define a family of multifold terms to serve as elimination forms for our admissible multiple fixed points.
+That is, rules of the form,
+
+\[
+  \inferrule{\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n})~\textbf{in}~X_{k} \\
+             \Gamma ; x_{1} : \simulsubst {A_{1}}{B_{1}/X_{1},\dots,B_{n}/X_{n}} \vdash e_{1} : B_{1} \\
+             \dots \\
+             \Gamma ; x_{n} : \simulsubst {A_{n}}{B_{1}/X_{1},\dots,B_{n}/X_{n}} \vdash e_{n} : B_{n}
+  }{\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
+\]
+
+Let the substitution $\{B_{1}/X_{1},\dots,B_{n}/X_{n} \}$ be given by $\gamma$.
+Additionally, let the substitution $\{X_{s}'/X_{s}, X_{e}'/X_{e}, X_{1}'/X_{1}, \dots \}$ be given by $\delta_{X}$.
+Likewise define $\delta_{Z}$.
+We can then characterize the map from $NFA(g \oplus g')$ with the following term,
+
+\[
+  \inferrule{
+    \Delta \vdash e : NFA(g \oplus g') \\
+    y_{s} : \simulsubst {(X_{s}' \oplus Z'_{s})} \gamma \vdash p_{s} : g \oplus g' \\
+    x_{s}' : \simulsubst {(\simulsubst {X_{s}} {\delta_{X}})}{\gamma} \vdash p_{x,s} : g \\
+    x_{j}' : \simulsubst {(\simulsubst {X_{j}} {\delta_{X}})}{\gamma} \vdash p_{x,j} : X_{j} \text{ where } j \not \in \{s, e\} \\
+    x_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{x, e} : X_{e} \\
+    z_{s}' : \simulsubst {(\simulsubst {Z_{s}} {\delta_{Z}})}{\gamma} \vdash p_{z,s} : g' \\
+    z_{j}' : \simulsubst {(\simulsubst {Z_{j}} {\delta_{Z}})}{\gamma} \vdash p_{z,j} : Z_{j} \text{ where } j \not \in \{s, e\} \\
+    z_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{z,e} : Z_{e} \\
+    y_{e} : \simulsubst I {\gamma} \vdash p_{e} : I
+  }{\Delta \vdash \mathsf{mfold}(e) : g \oplus g'}
+\]
+
+For each of the variables above, we simplify the type and provide the actual proof term.
+For instance, $\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$, so the derivation for $p_{s}$ is just the identity.
+We do this for all types and terms.
+
+Let $\hat{X}_{k}$ refer to the syntactic term on the right hand side of the definition of $X_{k}$ inside of $NFA(g)$.
+Likewise define $\hat{Z}_{k}$.
+Thus, syntactically $X_{k}' = \simulsubst {\hat{X}_{k}} {\delta_{X}}$ for $k \neq e$.
+
+\begin{itemize}
+  \item $y_{s}$ --- $\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$ and $p_{s} = y_{s} : g \oplus g'$
+  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = X_{s}$ and $p_{x,s} = \phi^{-1}(x_{s}') : g$
+  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = X_{j}$ and $p_{x, j} = x_{j}'$
+  \item $x_{e}'$ --- $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{x,e} = x_{e}' : X_{e} = I$
+  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = Z_{s}$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
+  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = Z_{j}$ and $p_{z, j} = z_{j}'$
+  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = z_{e}' : Z_{e} = I$
+  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
+\end{itemize}
+
+where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively. The map in the other direction is give as,
+
+\[
+\inferrule{p_{1} : g \vdash \textbf{in}_{X_{s}'}(\phi(p_{1})) : NFA(g \oplus g') \\ p_{2} : g' \vdash \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) : NFA(g \oplus g') \\ p : g \oplus g' \vdash p : g \oplus g'}{p : g \oplus g' \vdash \textbf{case}~p \{\textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\ \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) \} : NFA(g \oplus g')}
+\]
+
+It is evident that the maps $NFA(g \oplus g') \to g \oplus g'$ and $g \oplus g' \to NFA(g \oplus g')$ form an isomorphism.
+
+\begin{align*}
+  \textbf{case}~\mathsf{mfold}(e) \{\textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\ \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) \}
+  & = e
+\end{align*}
+
+% \begin{align*}
+%   \begin{aligned}
+%     \textbf{case}~\mathsf{mfold}(e) \{\textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\ \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2}))
+%   \end{aligned}
+%   & =
+%   \begin{aligned}
+%     \textbf{case}~(&  \textbf{case}~ e \{ \\
+%     & \textbf{in}_{X_{s}'}u \mapsto \mathsf{mfold}(\textbf{in}_{X_{s}'}u), \\ & \textbf{in}_{Z_{s}'}v \mapsto \mathsf{mfold}(\textbf{in}_{Z_{s}'}v) \\ \}) \{ \\ & \textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\  & \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) \}
+%   \end{aligned} \\
+%   & = \textbf{case}~e \{
+%     \textbf{in}_{X_{s}'}u \mapsto \textbf{in}_{X_{s}'}(\phi(\textsf{mfold}(\textbf{in}_{{X_{s}'}}u))) \\
+%     \textbf{in}_{Z_{s}'} v \mapsto
+%     \}
+    %   \end{align*}
+% This doesn't seem worth it.
+But you argue formally that commuting the case statements as such is done with the eta rule.
+Then you get everything to cancel down to \phi(\phi^{-1}(...)) (or the same with \psi)
+
+To see this consider a case distinction on $e : NFA(g \oplus g')$.
+\steven{I could type out the equality proof term, but it blows up in size and doesn't seem like a useful way to capture this iso.
+I have a commented out version of it above.
+Going to refactor tomorrow to something more sane, as these as are the crux of the proofs of equality} \steven{In any case, the NFA-to-grammar direction of these isos should all follow from some $\eta$-rule on the structure of the type for the start state, right?} Without loss of generality assume that $e = \textbf{in}_{X_{s}'} u$.
+Then, $\mathsf{mfold}(e) = \textbf{in}_{g}(\phi^{-1}u)$ and passing this through the above case statement results in $\textbf{in}_{{X_{s}'}}(\phi(\phi^{{-1}}u)) = \textbf{in}_{{X_{s}'}}(u) = e$.
+Further,
+
+\begin{align*}
+  \mathsf{mfold}(\textbf{case}~p \{\textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\ \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) \})
+  & = p
+\end{align*}
+
+for very similar reasons. \steven{How do you actually unpack what this fold term does?}
+
+We now construct a combinator for the sequencing of two NFAs.
+
+\[
+  NFA(g \otimes g') = \mu
+  \begin{pmatrix}
+    Y_{s} = X_{s}' \\
+    X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
+    X_{e}' = Z_{s}' \\
+    Z_{j}' = \simulsubst {\hat{Z}_{j}} {\delta_{Z}}, j \neq e \\
+    Z_{e}' = Y_{e} \\
+    Y_{e} = I
+  \end{pmatrix}
+\]
+
+Just as before, we use a multifold to define the map to $g \otimes g'$. Here the table of terms is as follows,
+
+\begin{itemize}
+  \item $y_{s}$ --- $\simulsubst {X_{s}'} \gamma = g \otimes g'$ and $p_{s} = y_{s} $
+  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = X_{s} \otimes g'$ and $p_{x,s} = \letin {u \otimes v} {x_{s}'} {\phi^{-1}(u) \otimes v} : g \otimes g'$
+  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = X_{j} \otimes g'$ and $p_{x, j} = x_{j}' : X_{j} \otimes g'$
+  \item $x_{e}'$ --- $\simulsubst {Z_{s}'}{\gamma} = g'$ and $p_{x,e} = x_{e}' : g'$
+  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = Z_{s}$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
+  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = Z_{j}$ and $p_{z, j} = z_{j}'$
+  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = z_{e}' : Z_{e} = I$
+  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
+\end{itemize}
+
+The map in the other direction is given as,
+
+\[
+  \inferrule{
+    p : g \otimes g' \vdash p : g \otimes g' \\
+    p_{1} : g, p_{2} : g' \vdash \phi(p_{1}) \otimes \psi(p_{2}) : g \otimes g'}
+  {p : g \otimes g' \vdash \letin {p_{1} \otimes p_{2}} {p} {\phi(p_{1}) \otimes \psi(p_{2})} : NFA(g \otimes g')}
+\]
+
+Again, these are clearly isomorphic.
+
+Finally, we build an automata combinator to represent the Kleene star\footnote{Which is another admissible definition via use of the fixed point constructor}.
+
+\[
+  NFA(g^{*}) = \mu
+  \begin{pmatrix}
+    Y_{s} = X_{s}' \oplus Y_{e} \\
+    X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
+    X_{e}' = X_{s}' \oplus Y_{e} \\
+    Y_{e} = I
+  \end{pmatrix}
+\]
+
+We fold over these defintions to get a parse of $g^{*}$,
+\begin{itemize}
+  \item $y_{s}$ --- $\simulsubst {(X_{s}' \oplus Y_{e})} \gamma = g^{*} \oplus I$ and $p_{s} = \textbf{case}~y_{s} \{ \textbf{in}_{g^{*}} u \mapsto u, \textbf{in}_{I} v \mapsto \mathsf{cons}_{*I} v \} : g^{*}$
+  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = X_{s} \otimes g^{*}$ and $p_{x,s} = \letin {u \otimes p_{*}} {x_{s}'} {\mathsf{cons}_{*}(\phi^{-1}(u), p_{*})} : g^{*} $ \steven{slight abuse of notation.
+Not sure how to best capture the Kleene star constructors}
+  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = X_{j} \otimes g^{*}$ and $p_{x, j} = : X_{j} \otimes g^{*}$
+  \item $x_{e}'$ --- $\simulsubst {(X_{s}' \oplus Y_{e})}{\gamma} = g^{*} \oplus I$ and $p_{x,e} = \textbf{case}~x_{e}' \{ \textbf{in}_{g^{*}} u \mapsto u, \textbf{in}_{I} v \mapsto \mathsf{cons}_{*I} v \} : g^{*}$
+  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
+\end{itemize}
+
+and to derive a parse of $NFA(g^{*})$ from a parse of $g*$,
+
+\[
+  \inferrule{p : g^{*} \vdash p : g^{*} \\ \cdot \vdash \textbf{in}_{{Y_{e}}}(()) : NFA(g^{*}) \\ u : g, v : g^{*} \vdash \textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v))) : NFA(g^{*}) }{p : g^{*} \vdash \mathsf{fold}_{*}(\textbf{in}_{{Y_{e}}}(()), \textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v)))) : NFA(g^{*}) }
+\]
+
+This completes our inductive proof that $NFA(g) \cong g$ for all regular expressions $g$.
+
+% \subsubsection{NFA to Grammar}
+
+% \steven{This section and following are basically todos for rn}
+
+% We now seek to take an NFA and derive the regular expression recognized by it.
+To this end we weaken our syntactic form of an NFA to that a \emph{generalized nondeterministic finite automata} (GNFA) --- where guards on transition can be arbitrary regexps instead of single characters.
+We then repeatedly take away states from the NFA to give an increasingly more complete algebraic description of the regular expression inside of the guards.
+We repeat until there is only a single transition between start and accepting state, and the guard on the single transition denotes the regular expression that is equivalent to the NFA.
+
+% \[
+%   GNFA ::= \mu \begin{pmatrix}
+%         Y_{s} = \langle Q \rangle \\
+%         Y_{1} = \langle Q \rangle \\
+%         \vdots \\
+%         Y_{e} = I
+%         \end{pmatrix}~\textbf{in}~Y_{s}
+% \]
+% where
+% \[
+%   Q ::= \langle RegExp \rangle \otimes Y ~|~ Y ~|~ Q \oplus Q
+% \]
+% \[
+%   RegExp ::= c ~|~ I ~|~ RegExp \oplus RegExp ~|~ RegExp \otimes RegExp ~|~ RegExp^{*}
+% \]
+
+% This is shown inductively by showing that the equivalence of a GNFA to a copy of itself with a general state --- i.e.\ non-start, non-accepting --- removed.
+Consider GNFA $A$ and remove state $Y_{r}$ to achieve $A'$.
+Let $\delta_{Y}$ denote the substitution that replaces $Y_{k}'/Y_{k}$ for $k \neq r$, and $\gamma$ denote the substitution $\hat{Y}_{r}/Y_{r}$.
+
+% \[
+%   A = \mu \begin{pmatrix}
+%             Y_{s} = \hat{Y}_{s} \\
+%             Y_{r} = \hat{Y}_{r} \\
+%             Y_{j} = \hat{Y}_{j}, j \not \in \{s, r, e\} \\
+%             Y_{e} = I
+%           \end{pmatrix}~\textbf{in}~Y_{s}
+% \]
+% \[
+%   A' = \mu \begin{pmatrix}
+%             Y_{s}' = \simulsubst {\hat{Y}_{s}} {\delta_{Y} \circ \gamma} \\
+%             Y_{j}' = \simulsubst {\hat{Y}_{j}} {\delta_{Y} \circ \gamma}, j \not \in \{s, e\} \\
+%             Y_{e}' = I
+%           \end{pmatrix}~\textbf{in}~Y_{s}'
+% \]
+
+% We can pretty easily translate between parses of $A$ and $A'$.
+This can be seen by looking at the structure of the terms $\hat{Y}_{k}$ and $\simulsubst {\hat{Y}_{k}} {\delta_{Y} \circ \gamma}$.
+For terms $\hat{Y}_{k}$ that don't have $Y_{r}$ as a subterm, the action of $\gamma$ is trivial --- so $\simulsubst {\hat{Y}_{k}} {\delta_{Y} \circ \gamma} = \simulsubst {\hat{Y}_{k}} {\delta_{Y}}$.
+For these $Y_{k}$ it is trivial to define a fold from $A$ to $A'$, or vice versa as we are simply renaming terms.
+For $\hat{Y}_{k}$ containing $Y_{r}$ as a subterm, it is not as trivial.
+Instead, each occurence of $Y_{r}$ is replaced with an equivalent definition.
+
+% Folding from $A$ to $A'$,
+
+% \begin{itemize}
+%   \item $y_{j}, j \not \in \{r, e\}$ and $Y_{r} \not \in \hat{Y}_{k}$ --- $\simulsubst {\hat{Y}_{k}} {\delta_{Y} \circ \Gamma} = \simulsubst {\hat{Y}_{k}} {\delta_{Y}}$ and $p_{s} = y_{j} : Y_{j}'$
+%   \item $y_{j}, j \not \in \{r, e\}$ and $Y_{r} \in \hat{Y}_{k}$ ---
+%   \item $y_{e}$ --- $\simulsubst I {\delta \circ \gamma} = I$ and $p_{e} = y_{e} : I$
+% \end{itemize}
+
+\steven{An aside on some axioms}
+
+Something like $A \cong (I \& A) \oplus \sum_{c} c \otimes (c \lto A)$
+
+For this taylor iso to hold, we need the map $c \otimes (c \lto A) \vdash (c \otimes \top) \amp A$ to be an iso.
+Specifically, this term needs an iso $\phi : c \otimes (c \lto A) \to (c \otimes \top) \amp A$ defined below,
+
+\[
+p : c \otimes (c \lto A) \vdash \phi(p) = \letin {p_{1} \otimes p_{2}} p {(p_{1} \otimes \top, p_{1}^{\lto}p_{2})} : (c \otimes \top) \amp A
+\]
+
+Also add the axiom that $\top \cong (\sum_{c \in \mathcal{A}} c)^{*}$
+
+If we have the above, we can then show the following isos for some alphabet $\mathcal{A}$,
+\begin{align*}
+  A
+  & \cong \top \amp A \\
+  & \cong (\Sigma_{c \in \mathcal{A}})^{*}\amp A \\
+  & \cong (I \oplus (\Sigma_{c} c \otimes \top)) \amp A \\
+  & \cong (I \amp A) \oplus ((\Sigma_{c} c \otimes \top)) \amp A) \\
+  & \cong (I \amp A) \oplus (\Sigma_{c }c \otimes (c \lto A)) & \text{using } \phi^{-1}
+\end{align*}
+
+From here we need to also show that derivatives of regular languages are regular (look at Brzozowksi's proof) and that regular languages have finitely many derivatives (potentially with some size constraint on the derivative?).
+Then the last iso in the chain above can be used to build a DFA equivalent to $A$ using Brzozowksi's algorithm.
+
+Need to also add in axioms for $\amp$ distributing over $\oplus$ and $\lto$ distributing over $\oplus$. And that $c \amp c' \vdash 0$ for $c \neq c'$
+
+\subsection{DFAs and NFAs}
+
+All DFAs are NFAs, so NFAs are at least as powerful as DFAs.
+It is perhaps surprising that DFAs are as powerful as NFAs too!
+We can internalize the classic powerset construction to transform an NFA into a DFA that recognizes the same language.
+Note however that this loses some information.
+Throughout this process we may lose data such as the ambiguity of the grammar.
+Effectively we are forgetting the structure of our proof terms.
+Under the ``grammars as types'' lens, we are erasing the structure underlying the inhabitants of the type to simply mere existence --- this is a form of propostional truncation.
+
+\steven{Worth spending some more time on the propositional truncation idea}
+
+\steven{Put powerset construction here}
+
+\subsection{TODO More Complicated Grammars}
+Want to put PDA, CFGs, CSGs here. Basically, use this section to highlight the usefulness of the intuitionistic part of the contexts
+
 \section{Future work}
+
+\subsection{Implementation}
 

### 2024-01-20 paper/paper.tex
comments

@@
 that it can handle more classes of languages beyond regular ones.
+
+\paragraph{Beyond Simple Types}
+
+\begin{definition}
+  A model is a locally Cartesian category equipped with an internal Kleene category.
+\end{definition}
+
+\begin{theorem}
+  The presheaf category $\Set^{\cat{\Sigma^*}}$ is a model.
+\end{theorem}
 
@@
 To this end we define
-\pedro{I'm pretty sure that these definitions have been somewhat presented
-  in ``Glueing and orthogonality for models of linear logic'' by Hyland and Schalk.
-We should probably cite them so that we don't have to rederive their constructions and proofs here.}
 \begin{enumerate}
@@
 I have a commented out version of it above.
-Going to refactor tomorrow to something more sane, as these as are the crux of the proofs of equality} \steven{In any case, the NFA-to-grammar direction of these isos should all follow from some $\eta$-rule on the structure of the type for the start state, right?} Without loss of generality assume that $e = \textbf{in}_{X_{s}'} u$.
+Going to refactor tomorrow to something more sane, as these as are the crux of the proofs of equality} \steven{In any case, the NFA-to-grammar direction of these isos should all follow from some $\eta$-rule on the structure of the type for the start state, right?}
+\pedro{What I like doing in these cases is either presenting the high-level structure of the proof, e.g., ``by applying the rules <...> we can conclude'', or present the full proof in the appendix.
+  Since in this case it feels like it won't be a particularly illuminating proof, I suggest doing
+the former.}
+Without loss of generality assume that $e = \textbf{in}_{X_{s}'} u$.
 Then, $\mathsf{mfold}(e) = \textbf{in}_{g}(\phi^{-1}u)$ and passing this through the above case statement results in $\textbf{in}_{{X_{s}'}}(\phi(\phi^{{-1}}u)) = \textbf{in}_{{X_{s}'}}(u) = e$.
@@
 This completes our inductive proof that $NFA(g) \cong g$ for all regular expressions $g$.
+
+\pedro{I feel like that for the sake of organizing the paper, since the proof is non-trivial,
+there should be a theorem environment with an accompanying proof environment}
 

### 2024-01-22 paper/paper.tex
submitted abstract

@@
   algebras to incorporate ambiguity. By incorporating a dependency of
-  linear types on non-linear types, we can further express indexed
-  grammars, pushdown automata and Turing machines. This provides a new
-  form of ``logical characterization'' of grammar classes based on a
-  substructural logic rather than ordinary first-order logic. Further,
-  the type theory can not only express the grammars as types, but
-  equivalences between grammars and parsers as terms of the grammar.
+  linear types on non-linear types, we can express richer grammar
+  formalisms such as indexed grammars, pushdown automata and Turing
+  machines. This provides a new form of ``logical characterization''
+  of grammar classes based on a substructural logic rather than
+  ordinary first-order logic. Further, the type theory can not only
+  express the grammars as types, but equivalences between grammars and
+  parsers as terms of the grammar.
 

### 2024-01-22 paper/paper.tex
start work on examples section

@@
 
-\section{Type-theoretic Characterizations of Language Classes}
+\section{Type Theoretic Formal Grammar Theory}
+
+Now that we have developed a type theoretic syntax for formal grammars
+and parse transformers, we show how many common concepts of formal
+language theory and grammars are naturally expressed using type
+theoretic notions, and some constructions and proofs can be carried
+out in our type theory as well.
+
+Firstly, regular expressions and their extension to context-free
+expressions are naturally expressed as a limited syntax of linear
+types. That is, regular expressions are precisely the linear types
+constructed using only $c, I, \otimes, 0, \oplus$ and the Kleene star,
+which can be defined as $A^* = \mu x. I \oplus A \otimes X$. The
+context-free expressions are similarly expressible by allowing not
+just the Kleene star, but unrestricted usage of $\mu$. Remember,
+however, that these are a notion of regular expressions that maintain
+a notion of ambiguous grammar. That is, $A \oplus A$ is not considered
+to be the same as $A$. While there does not appear to be an analogous
+weakening of this definition to naturally describe the
+context-sensitive languages, we will see shortly that the indexing of
+linear types by non-linear types allows for capturing a notion of
+context in grammars.
+
+In addition to expressing algebraic presentations of grammars, the
+type-theoretic syntax allows for naturally expressing grammars that
+capture the behavior of automata. \max{sync this up with Steven's
+  stuff, and add stuff about PDAs and Turing Machines}.
+
+A key construct in formal language theory is the Brzozowski derivative
+\cite{brzozowski}. Given a string $w$ the (left) derivative of a
+language $L$ with respect to $w$, written $\partial_w L$ is the language
+\[ (\partial_wL ) = \{ ww' \pipe w' \in L \}\]
+This semantics is given using the linear implication $w \tol
+L$. Expressing Brzozowski's calculation of the derivative of a regular
+expression relies on additional axioms to the type theory.\max{axioms
+  to carry out Brzozowski's derivative calculations plz.}
+
+Using the $G$ modality, we can internalize some of the external
+reasoning about grammars into the language of non-linear types. Recall
+that the denotation of $G A$ is the set of parses of the empty
+string. As is well known in formal language theory, all the parsing
+behavior of a grammar can be reduced to reasoning about parsing of the
+empty string by using the derivative of a grammar. For instance, the
+set of parses for a fixed string $w$ can be defined as $G(w \tol A)$
+and so the type of all parses of a grammar can be defined as
+$\Sigma(w:\textrm{String}) G(w \tol A)$. This can be used to define
+internally a notion of \emph{ambiguity} of a grammar. A grammar $A$ is
+ambiguous if for every string $w$, the type $G(w \tol A)$ of
+$w-$parses is a \emph{proposition} in the sense that all terms in $G(w
+\tol A)$ are equal.\max{I think we should add a propositional
+  truncation to the type theory.}  Similarly, the type of \emph{total
+parsers} can be defined as $G(\top \tol A)$. A total parser encodes a
+function that selects a parse for every possible string. A
+\emph{semi-parser} is then defined as $G(\top \tol A \oplus \top)$,
+where the parser is allowed to fail for any reason. This has the
+downside that the always failing semi-parser $\lambda _. \textrm{in}_r
+()$ is valid for any grammar. If we were working with a full
+\emph{bunched} type theory, we would be able to express the more
+precise type of a \emph{deciding} parser as $G(\top \tol A \oplus \neg
+A)$ where $\neg A = A \Rightarrow 0$ uses the intuitionistic
+implication.
+
 Classically, formal language theory is closely related to automata theory.
@@
 
-\subsection{TODO More Complicated Grammars}
+\subsection{Indexed Automata}
+
+We can generalize to more powerful notions of automata by utilizing
+dependency of linear types on non-linear types. Let $S$ be a finite
+non-linear type and define $\textrm{List}\,S = \mu x. 1 + S \times
+x$. Then a pushdown automaton with $S$ as the stack alphabet can be
+defined as a mutual fixed point.
+
 Want to put PDA, CFGs, CSGs here. Basically, use this section to highlight the usefulness of the intuitionistic part of the contexts

### 2024-01-22 paper/paper.tex
put cases in appendix

--- 2024-01-20 paper/paper.tex
@@
 In the styles of Chomsky and Sch\"utzenberger, there is a hierarchy of language classes and the types of automata that recognize them.
-Inside of the type theory of \cref{sec:typetheory}, we can characterize these language classes; moreover, we demonstrate equivalence of each class to the associated automaton as an isomorphism within our logic.
-
-Note, predicate BI with existential quantifiers is already enough to define
-formulae that correspond to arbitrary recursively enumerable languages
+The type theory of \cref{sec:typetheory} serves as a syntax to reason about formal grammars; moreover, we describe notions of automata into this formal system via syntactic forms that describe an automaton as a grammar.
+This method allows us to internalize many classic automata-theoretic results as proof terms in our type theory.
+
+Note, a predicate logic over LNL with existential quantifiers is already enough to define
+formulae that correspond to arbitrary recursively enumerable languages ---
 since we can simulate the tape of a Turing machine using two
@@
 
-We reflect on this expressive power to motivate why we put heavy syntactic restrictions on our presentation of certain types of automata.
-For instance, when defining a pushdown automata, we ought to be very careful about how the set representing the stack can be accessed and added to.
+We reflect on this expressive power to motivate why we put heavy syntactic restrictions on our presentation of certain classes of automata.
+For instance, when defining a pushdown automata, we could use the intuitionistic context to represent the stack, but we ought to be very careful about how this set is accessed and added to.
 Careless definitions may accidentally endow us with too much expressive power, launching us further up in the Chomsky hierarchy.
-To begin, let us focus only on language classes, and their associated automata whose parses are definable without any mention to the inuitionistic context.
-In the context of automata, this refers to the class of automata definable without access to any memory --- i.e.\ finite state machines.
+To begin, let us focus only on language classes, and their associated automata, whose parses are definable without any mention to the inuitionistic context.
+When viewed as a machine, this refers to the class of automata definable without access to any memory --- i.e.\ finite state machines.
 
 \subsection{Finite State Machines}
-We may encode a finite state machine via a particular syntactic form for grammars.
-A \emph{nondeterministic finite automaton} (NFA) is defined as a mutual \footnote{For convenience we refer to a mutual fixed point here, even though we only take single fixed points as primitive.
+A finite state machine may be expressed via a particular syntactic form for grammars.
+A \emph{nondeterministic finite automaton} (NFA) is defined as a mutual\footnote{For convenience we refer to a mutual fixed point here, even though we only take single fixed points as primitive.
 As in Beki\'c's theorem, we may decompose our mutual fixed points into a series of single fixed points.} fixed point in the following manner,
 
-\steven{Is there a better way to present this?}
 \[
@@
 and $Y \in \{ Y_{s}, Y_{1}, \dots, Y_{N} \}$.
-Intuitively, we read each $Y_{j}$ as the label for a state, and the definitions in form $Q$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$ an $\epsilon$-transition.
+Intuitively, we read each $Y_{j}$ as the label for a state, and the definitions in form $Q$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$ denotes an $\epsilon$-transition.
 Without loss of generality, $Y_{s}$ is reserved for a single distinguished start state, and $Y_{e}$ is reserved for a single distinguished accepting state.
@@
 Let $\hat{Y}_{k}$ refer to the syntactic term on the right hand side of the definition of $Y_{k}$ inside of an NFA $N$.
-We say that $A$ is a \emph{deterministic finite automaton} (DFA) if there are no $\epsilon$-transitions and for every definition $\hat{Y}_{k}$ the $c_{j}$ are distinct guards for each produced state.
-\steven{Is this the right notion of determinism?
+We say that $N$ is a \emph{deterministic finite automaton} (DFA) if there are no $\epsilon$-transitions and for every definition $\hat{Y}_{k}$ the $c_{j}$ are distinct guards for each produced state.
+% \steven{Is this the right notion of determinism?
 Perhaps, it is is better to define determinism as something like: there is at most one derivation for all $j, k$, for all $a \in \Sigma$ we have $x : Y_{k}, y : a \vdash z : Y_{j}$.}
@@
 With this setting for finite automata, we can now internalize some classical theorems inside of our formal system.
-First, let us describe a version of Thompson's construction \cite{??} where we show that any regular expression can be recognized by some NFA\@.
+First, let us describe a version of Thompson's construction \cite{?} where we show that any regular expression can be recognized by some NFA\@.
 Moreover, we will show that this NFA is strongly equivalent to the original grammar.
 
-\subsection{Equivalence Between Regular Expressions and Finite Automata}
-
-\steven{Define regular expressions}
-\steven{Move proofs to appendix, leave the exposition and the theorem statements.}
-
-\subsubsection{Grammar to NFA}
-
-We will define the recognizing NFA of a regular expression inductively. To handle the atomic grammars --- literals and the empty grammar --- we define the following,
+\subsection{Equivalence Between Regular Grammars and Finite Automata}
+
+The class of \emph{regular} grammars is given as follows:
+
+\[
+Reg ::= 0 ~|~ I ~|~ c ~|~ Reg \oplus Reg ~|~ Reg \otimes Reg ~|~ Reg^{*}
+\]
+where $c \in \mathcal{A}$ is a character literal drawn from a fixed alphabet, and the Kleene star $\cdot^{*}$ is derived as an admissible operation using the fixed point operator --- $g^{*} = \mu X.I \oplus (g \otimes X)$.
+\steven{This definition is maybe fair to assume.
+Also maybe better suited for a section earlier in the paper}
+
+\begin{theorem}
+  \label{thm:thompson}
+  For $g$ a regular grammar, there is an NFA $N$ that recognizes the same language as $g$. Further, $N$ and $g$ are isomorphic.
+\end{theorem}
+
+\begin{proof}
+  We construct the recognizing NFA $N$ inductively. To handle the atomic grammars --- literals and the empty grammar --- we define the following,
+
 \[
@@
 First begin with the disjunction of two NFAs.
-hat is, assuming that $g \cong NFA(g)$ and $g' \cong NFA(g')$, we define an NFA that recognizes $g \oplus g'$.
-Let the states in $NFA(g)$ be represented by $X$'s and the states in $NFA(g')$ be represented by $Z$'s,
+That is, assuming that $g \cong NFA(g)$ and $g' \cong NFA(g')$, we define an NFA that recognizes $g \oplus g'$.
+Let the states in $NFA(g)$ be represented by $X$'s and the states in $NFA(g')$ be represented by $Z$'s.
+Further let $\hat{X}_{k}$ refer to the syntactic term on the right hand side of the definition of $X_{k}$ inside of $NFA(g)$.
+Likewise define each $\hat{Z}_{k}$.
 
@@
         Y_{s} =  X_{s}' \oplus Z_{s}' \\
-        X_{s}', X_{1}', \dots = \dots \\
+        X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
         X'_{e} = Y_{e} \\
-        Z_{s}', Z_{1}, \dots = \dots \\
+        Z_{j}' = \simulsubst {\hat{Z}_{j}} {\delta_{Z}}, j \neq e \\
         Z_{e}' = Y_{e} \\
@@
 \]
-
-For each defintion $X_{j}$ in $NFA(g)$ we create an analogous definition $X_{j}'$ inside of $NFA(g \oplus g')$ with the syntactic equality $X_{j}'$ of $X_{j}$ for $j \neq e$, except we replaec $X_{k}$ with $X_{k}'$ in each production.
+where the substitution $\{X_{s}'/X_{s}, X_{e}'/X_{e}, X_{1}'/X_{1}, \dots \}$ is given by $\delta_{X}$. Likewise define $\delta_{Z}$.
+In the construction of $NFA(g \oplus g')$ we replicate each defintion $X_{j}$ in $NFA(g)$ with an analogous definition $X_{j}'$ inside of $NFA(g \oplus g')$, except we replace $X_{k}$ with $X_{k}'$ in each production.
 Lastly, $X_{e}'$ has an $\epsilon$-transition to the accepting state $Y_{e}$.
 Likewise, we build analogous defintions for the states $Z_{j}$ in $NFA(g')$, then we flatten all of these defintions into one mutally recursive list.
+This has the effect of disjunctively tying together $NFA(g)$ and $NFA(g')$ into one automaton.
 
@@
 To define a map $NFA(g \oplus g') \to g \oplus g'$ we will use a multi-fold over all the mutually recursive defintions of $NFA(g \oplus g')$.
-Even though multiple folding is not taken as primitive, using Beki\'c's theorem we can define a family of multifold terms to serve as elimination forms for our admissible multiple fixed points.
+Note, even though multiple folding is not taken as primitive, using Beki\'c's theorem we can define a family of multifold terms to serve as elimination forms for our admissible multiple fixed points.
 That is, rules of the form,
@@
   \inferrule{\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n})~\textbf{in}~X_{k} \\
-             \Gamma ; x_{1} : \simulsubst {A_{1}}{B_{1}/X_{1},\dots,B_{n}/X_{n}} \vdash e_{1} : B_{1} \\
+             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
              \dots \\
-             \Gamma ; x_{n} : \simulsubst {A_{n}}{B_{1}/X_{1},\dots,B_{n}/X_{n}} \vdash e_{n} : B_{n}
+             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
   }{\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
@@
 
-Let the substitution $\{B_{1}/X_{1},\dots,B_{n}/X_{n} \}$ be given by $\gamma$.
-Additionally, let the substitution $\{X_{s}'/X_{s}, X_{e}'/X_{e}, X_{1}'/X_{1}, \dots \}$ be given by $\delta_{X}$.
-Likewise define $\delta_{Z}$.
-We can then characterize the map from $NFA(g \oplus g')$ with the following term,
+are admissible, where the substitution $\{B_{1}/X_{1},\dots,B_{n}/X_{n} \}$ is given by $\gamma$.
+In this instance, we wish to simultaneously fold over the mutually recursive definitions comprising $NFA(g \oplus g')$.
+We may think of this as an eliminator that replaces each state inside of $NFA(g \oplus g')$ with a tail-recursive call that will ultimately return a term of type $g \oplus g'$.
+The structure of such a derivation is given below,
 
@@
 For instance, $\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$, so the derivation for $p_{s}$ is just the identity.
-We do this for all types and terms.
-
-Let $\hat{X}_{k}$ refer to the syntactic term on the right hand side of the definition of $X_{k}$ inside of $NFA(g)$.
-Likewise define $\hat{Z}_{k}$.
-Thus, syntactically $X_{k}' = \simulsubst {\hat{X}_{k}} {\delta_{X}}$ for $k \neq e$.
-
+The rest of the terms are given below,
 \begin{itemize}
@@
 
-where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively. The map in the other direction is give as,
+where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively.
+This finshes the construction of a parse transformer $NFA(g \oplus g') \to g \oplus g'$ The map in the other direction is given the derivation,
 
@@
 
-To see this consider a case distinction on $e : NFA(g \oplus g')$.
-\steven{I could type out the equality proof term, but it blows up in size and doesn't seem like a useful way to capture this iso.
+% \steven{I could type out the equality proof term, but it blows up in size and doesn't seem like a useful way to capture this iso.
 I have a commented out version of it above.
 Going to refactor tomorrow to something more sane, as these as are the crux of the proofs of equality} \steven{In any case, the NFA-to-grammar direction of these isos should all follow from some $\eta$-rule on the structure of the type for the start state, right?}
-\pedro{What I like doing in these cases is either presenting the high-level structure of the proof, e.g., ``by applying the rules <...> we can conclude'', or present the full proof in the appendix.
-  Since in this case it feels like it won't be a particularly illuminating proof, I suggest doing
-the former.}
-Without loss of generality assume that $e = \textbf{in}_{X_{s}'} u$.
+% \pedro{What I like doing in these cases is either presenting the high-level structure of the proof, e.g., ``by applying the rules <...> we can conclude'', or present the full proof in the appendix.
+  % Since in this case it feels like it won't be a particularly illuminating proof, I suggest doing
+% the former.}
+    %
+
+By breaking into a case distinction on $e$ and then applying the $\eta$-rule for coproducts, the above proof will simplify into two cases where $\phi$ will be inverted by $\phi^{{-1}}$ and $\psi$ will be inverted by $\psi^{-1}$, ultimately cancelling to give the identity map.
+The other direction below simplifies in a similar manner.
+\steven{Most concise list for this?}
+
+% Without loss of generality assume that $e = \textbf{in}_{X_{s}'} u$.
 Then, $\mathsf{mfold}(e) = \textbf{in}_{g}(\phi^{-1}u)$ and passing this through the above case statement results in $\textbf{in}_{{X_{s}'}}(\phi(\phi^{{-1}}u)) = \textbf{in}_{{X_{s}'}}(u) = e$.
@@
 
-for very similar reasons. \steven{How do you actually unpack what this fold term does?}
-
-We now construct a combinator for the sequencing of two NFAs.
-
-\[
-  NFA(g \otimes g') = \mu
-  \begin{pmatrix}
-    Y_{s} = X_{s}' \\
-    X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
-    X_{e}' = Z_{s}' \\
-    Z_{j}' = \simulsubst {\hat{Z}_{j}} {\delta_{Z}}, j \neq e \\
-    Z_{e}' = Y_{e} \\
-    Y_{e} = I
-  \end{pmatrix}
-\]
-
-Just as before, we use a multifold to define the map to $g \otimes g'$. Here the table of terms is as follows,
-
-\begin{itemize}
-  \item $y_{s}$ --- $\simulsubst {X_{s}'} \gamma = g \otimes g'$ and $p_{s} = y_{s} $
-  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = X_{s} \otimes g'$ and $p_{x,s} = \letin {u \otimes v} {x_{s}'} {\phi^{-1}(u) \otimes v} : g \otimes g'$
-  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = X_{j} \otimes g'$ and $p_{x, j} = x_{j}' : X_{j} \otimes g'$
-  \item $x_{e}'$ --- $\simulsubst {Z_{s}'}{\gamma} = g'$ and $p_{x,e} = x_{e}' : g'$
-  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = Z_{s}$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
-  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = Z_{j}$ and $p_{z, j} = z_{j}'$
-  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = z_{e}' : Z_{e} = I$
-  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
-\end{itemize}
-
-The map in the other direction is given as,
-
-\[
-  \inferrule{
-    p : g \otimes g' \vdash p : g \otimes g' \\
-    p_{1} : g, p_{2} : g' \vdash \phi(p_{1}) \otimes \psi(p_{2}) : g \otimes g'}
-  {p : g \otimes g' \vdash \letin {p_{1} \otimes p_{2}} {p} {\phi(p_{1}) \otimes \psi(p_{2})} : NFA(g \otimes g')}
-\]
-
-Again, these are clearly isomorphic.
-
-Finally, we build an automata combinator to represent the Kleene star\footnote{Which is another admissible definition via use of the fixed point constructor}.
-
-\[
-  NFA(g^{*}) = \mu
-  \begin{pmatrix}
-    Y_{s} = X_{s}' \oplus Y_{e} \\
-    X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
-    X_{e}' = X_{s}' \oplus Y_{e} \\
-    Y_{e} = I
-  \end{pmatrix}
-\]
-
-We fold over these defintions to get a parse of $g^{*}$,
-\begin{itemize}
-  \item $y_{s}$ --- $\simulsubst {(X_{s}' \oplus Y_{e})} \gamma = g^{*} \oplus I$ and $p_{s} = \textbf{case}~y_{s} \{ \textbf{in}_{g^{*}} u \mapsto u, \textbf{in}_{I} v \mapsto \mathsf{cons}_{*I} v \} : g^{*}$
-  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = X_{s} \otimes g^{*}$ and $p_{x,s} = \letin {u \otimes p_{*}} {x_{s}'} {\mathsf{cons}_{*}(\phi^{-1}(u), p_{*})} : g^{*} $ \steven{slight abuse of notation.
-Not sure how to best capture the Kleene star constructors}
-  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = X_{j} \otimes g^{*}$ and $p_{x, j} = : X_{j} \otimes g^{*}$
-  \item $x_{e}'$ --- $\simulsubst {(X_{s}' \oplus Y_{e})}{\gamma} = g^{*} \oplus I$ and $p_{x,e} = \textbf{case}~x_{e}' \{ \textbf{in}_{g^{*}} u \mapsto u, \textbf{in}_{I} v \mapsto \mathsf{cons}_{*I} v \} : g^{*}$
-  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
-\end{itemize}
-
-and to derive a parse of $NFA(g^{*})$ from a parse of $g*$,
-
-\[
-  \inferrule{p : g^{*} \vdash p : g^{*} \\ \cdot \vdash \textbf{in}_{{Y_{e}}}(()) : NFA(g^{*}) \\ u : g, v : g^{*} \vdash \textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v))) : NFA(g^{*}) }{p : g^{*} \vdash \mathsf{fold}_{*}(\textbf{in}_{{Y_{e}}}(()), \textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v)))) : NFA(g^{*}) }
-\]
-
-This completes our inductive proof that $NFA(g) \cong g$ for all regular expressions $g$.
-
-\pedro{I feel like that for the sake of organizing the paper, since the proof is non-trivial,
-there should be a theorem environment with an accompanying proof environment}
+This concludes the disjunctive combination of two NFAs. Because they follow a very similar style of argumentation, we include the cases for sequencing and Kleene star in \cref{sec:appendixnfa}.
+
+\end{proof}
 
@@
 result as stated in Section\ref{section:canonicity}.
+
+\section{NFA stuff}
+\label{sec:appendixnfa}
+
+We provide the final two cases of \cref{thm:thompson}.
+
+\begin{proof}
+We now construct a combinator for the sequencing of two NFAs.
+
+\[
+  NFA(g \otimes g') = \mu
+  \begin{pmatrix}
+    Y_{s} = X_{s}' \\
+    X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
+    X_{e}' = Z_{s}' \\
+    Z_{j}' = \simulsubst {\hat{Z}_{j}} {\delta_{Z}}, j \neq e \\
+    Z_{e}' = Y_{e} \\
+    Y_{e} = I
+  \end{pmatrix}
+\]
+
+Just as before, we use a multifold to define the map to $g \otimes g'$. Here the table of terms is as follows,
+
+\begin{itemize}
+  \item $y_{s}$ --- $\simulsubst {X_{s}'} \gamma = g \otimes g'$ and $p_{s} = y_{s} $
+  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = X_{s} \otimes g'$ and $p_{x,s} = \letin {u \otimes v} {x_{s}'} {\phi^{-1}(u) \otimes v} : g \otimes g'$
+  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = X_{j} \otimes g'$ and $p_{x, j} = x_{j}' : X_{j} \otimes g'$
+  \item $x_{e}'$ --- $\simulsubst {Z_{s}'}{\gamma} = g'$ and $p_{x,e} = x_{e}' : g'$
+  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = Z_{s}$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
+  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = Z_{j}$ and $p_{z, j} = z_{j}'$
+  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = z_{e}' : Z_{e} = I$
+  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
+\end{itemize}
+
+The map in the other direction is given as,
+
+\[
+  \inferrule{
+    p : g \otimes g' \vdash p : g \otimes g' \\
+    p_{1} : g, p_{2} : g' \vdash \phi(p_{1}) \otimes \psi(p_{2}) : g \otimes g'}
+  {p : g \otimes g' \vdash \letin {p_{1} \otimes p_{2}} {p} {\phi(p_{1}) \otimes \psi(p_{2})} : NFA(g \otimes g')}
+\]
+
+Again, these are clearly isomorphic.
+
+Finally, we build an automata combinator to represent the Kleene star\footnote{Which is another admissible definition via use of the fixed point constructor}.
+
+\[
+  NFA(g^{*}) = \mu
+  \begin{pmatrix}
+    Y_{s} = X_{s}' \oplus Y_{e} \\
+    X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
+    X_{e}' = X_{s}' \oplus Y_{e} \\
+    Y_{e} = I
+  \end{pmatrix}
+\]
+
+We fold over these defintions to get a parse of $g^{*}$,
+\begin{itemize}
+  \item $y_{s}$ --- $\simulsubst {(X_{s}' \oplus Y_{e})} \gamma = g^{*} \oplus I$ and $p_{s} = \textbf{case}~y_{s} \{ \textbf{in}_{g^{*}} u \mapsto u, \textbf{in}_{I} v \mapsto \mathsf{cons}_{*I} v \} : g^{*}$
+  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = X_{s} \otimes g^{*}$ and $p_{x,s} = \letin {u \otimes p_{*}} {x_{s}'} {\mathsf{cons}_{*}(\phi^{-1}(u), p_{*})} : g^{*} $ \steven{slight abuse of notation.
+Not sure how to best capture the Kleene star constructors}
+  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = X_{j} \otimes g^{*}$ and $p_{x, j} = : X_{j} \otimes g^{*}$
+  \item $x_{e}'$ --- $\simulsubst {(X_{s}' \oplus Y_{e})}{\gamma} = g^{*} \oplus I$ and $p_{x,e} = \textbf{case}~x_{e}' \{ \textbf{in}_{g^{*}} u \mapsto u, \textbf{in}_{I} v \mapsto \mathsf{cons}_{*I} v \} : g^{*}$
+  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
+\end{itemize}
+
+and to derive a parse of $NFA(g^{*})$ from a parse of $g*$,
+
+\[
+  \inferrule{p : g^{*} \vdash p : g^{*} \\ \cdot \vdash \textbf{in}_{{Y_{e}}}(()) : NFA(g^{*}) \\ u : g, v : g^{*} \vdash \textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v))) : NFA(g^{*}) }{p : g^{*} \vdash \mathsf{fold}_{*}(\textbf{in}_{{Y_{e}}}(()), \textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v)))) : NFA(g^{*}) }
+\]
+
+This completes our inductive proof that $NFA(g) \cong g$ for all regular expressions $g$.
+
+\end{proof}
+
 \end{document}

### 2024-01-22 paper/paper.tex
judgemental equality plus Brzozowski

--- 2024-01-22 paper/paper.tex
@@
   Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
-  we want to show that there is a unique algebra morphism $h : B
-  \otimes \mu X.\, F_{A,1} \to Y$. We can show existence and uniqueness
-  by showing that the diagram on the left commute if, and only if,
-  the diagram on the right commutes:
+  we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,1} \to B \lto Y$. We can show existence and
+  uniqueness by showing that the diagram on top commutes if, and
+  only if, the diagram on the bottom commutes:
+
+% https://q.uiver.app/#q=WzAsOCxbMCwwLCJCICsgQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFsyLDAsIkIgKyBZIFxcb3RpbWVzIEEiXSxbMiwyLCJZIl0sWzAsMiwiQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDMsIjEgKyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDUsIlxcbXUgWC5cXCwgMSArIFggXFxvdGltZXMgQSJdLFsyLDMsIjEgKyAoIEIgXFxsdG8gWSkgXFxvdGltZXMgQSJdLFsyLDUsIkIgXFxsdG8gWSJdLFswLDEsImlkICsgKGlkIFxcb3RpbWVzIGg7IGV2KSBcXG90aW1lcyBpZF9BIl0sWzEsMiwiW2YsZ10iXSxbMywyLCJpZCBcXG90aW1lcyBoOyBldiIsMl0sWzAsMywiW2lkIFxcb3RpbWVzIGgsIGlkIFxcb3RpbWVzIHRsXSIsMl0sWzQsNiwiaWQgKyAoaCBcXG90aW1lcyBYKVxcb3RpbWVzIGlkIl0sWzUsNywiaCIsMl0sWzYsNywiW2YnLCBnJ10iXSxbNCw1LCJbaGQsIHRsXSIsMl1d
+\[\begin{tikzcd}
+	{B + B \otimes (\mu X.\, 1 + X \otimes A)} && {B + Y \otimes A} \\
+	\\
+	{B \otimes (\mu X.\, 1 + X \otimes A)} && Y \\
+	{1 + (\mu X.\, 1 + X \otimes A)} && {1 + ( B \lto Y) \otimes A} \\
+	\\
+	{\mu X.\, 1 + X \otimes A} && {B \lto Y}
+	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
+	\arrow["{[f,g]}", from=1-3, to=3-3]
+	\arrow["{id \otimes h; ev}"', from=3-1, to=3-3]
+	\arrow["{[id \otimes h, id \otimes tl]}"', from=1-1, to=3-1]
+	\arrow["{id + (h \otimes X)\otimes id}", from=4-1, to=4-3]
+	\arrow["h"', from=6-1, to=6-3]
+	\arrow["{[f', g']}", from=4-3, to=6-3]
+	\arrow["{[hd, tl]}"', from=4-1, to=6-1]
+\end{tikzcd}\]
 
@@
 
-Much like in the posetal case, the abstract structure of a 2-Kleene
-algebra is expressive enough to synthetically reason about formal
+Much like in the posetal case, the abstract structure of a Kleene
+category is expressive enough to synthetically reason about formal
 languages. A significant difference between them is that while Kleene
-algebras can reason about language containment, 2-Kleene algebras can
+algebras can reason about language containment, Kleene categories can
 reason about parsing as well.
@@
 
+Though Kleene categories are expressive enough to reason about concepts
+that are outside of reach of Kleene algebras, their simply-typed nature
+makes them not so expressive from a type theory point of view.
+
+Since we are interested in design a type theory for reasoning about various
+properties of parsers, we will assume that our Kleene category is defined
+internal to a model of linear depedendently typed theory, which leads us
+to the notion of models we are interested in:
+
 \begin{definition}
-  A model is a locally Cartesian category equipped with an internal Kleene category.
+  A model is a locally Cartesian category with \pedro{add the linear logic connectives here} equipped with an internal Kleene category.
 \end{definition}
@@
     %
+    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
+    %
+    \and
+    %
     \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
@@
   \caption{Type well-formedness}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
+%
+    \and
+%
+    \inferrule{~}{}
+%
+    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
+%   
+    \and
+%
+    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
+    %
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
+%
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C} 
+    %
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
+\end{mathpar}
+  \caption{Judgemental equality}
 \end{figure}
@@
   to carry out Brzozowski's derivative calculations plz.}
+
+\begin{align*}
+  c\tol c &\equiv I\\
+  c\tol d &\equiv 0\\
+  c\tol I &\equiv 0\\
+  c\tol 0 &\equiv 0\\
+  c\tol (A \otimes B) &\equiv (c\tol A) \otimes B + \nu(A) \otimes (c\tol B)\\
+  c\tol A^* &\equiv (c\tol A) \otimes A^*\\
+  c\tol (A + B) &\equiv (c\tol A) + (c\tol B)
+\end{align*}
+
+\pedro{The type family $\nu$ above is equal to $I$ if its argument
+  parses the empty string and zero otherwise. In order to define
+  it we should probably add an extra axiom, right?}
 

### 2024-01-22 paper/paper.tex
added brzozowski axioms

--- 2024-01-22 paper/paper.tex
@@
 
-\section{Type-theoretic Characterizations of Language Classes}
-Classically, formal language theory is closely related to automata theory.
-In the styles of Chomsky and Sch\"utzenberger, there is a hierarchy of language classes and the types of automata that recognize them.
-The type theory of \cref{sec:typetheory} serves as a syntax to reason about formal grammars; moreover, we describe notions of automata into this formal system via syntactic forms that describe an automaton as a grammar.
-This method allows us to internalize many classic automata-theoretic results as proof terms in our type theory.
+\section{Type Theoretic Formal Grammar Theory}
+
+Now that we have developed a type theoretic syntax for formal grammars
+and parse transformers, we show how many common concepts of formal
+language theory and grammars are naturally expressed using type
+theoretic notions, and some constructions and proofs can be carried
+out in our type theory as well.
+
+Firstly, regular expressions and their extension to context-free
+expressions are naturally expressed as a limited syntax of linear
+types. That is, regular expressions are precisely the linear types
+constructed using only $c, I, \otimes, 0, \oplus$ and the Kleene star,
+which can be defined as $A^* = \mu x. I \oplus A \otimes X$. The
+context-free expressions are similarly expressible by allowing not
+just the Kleene star, but unrestricted usage of $\mu$. Remember,
+however, that these are a notion of regular expressions that maintain
+a notion of ambiguous grammar. That is, $A \oplus A$ is not considered
+to be the same as $A$. While there does not appear to be an analogous
+weakening of this definition to naturally describe the
+context-sensitive languages, we will see shortly that the indexing of
+linear types by non-linear types allows for capturing a notion of
+context in grammars.
+
+In addition to expressing algebraic presentations of grammars, the
+type-theoretic syntax allows for naturally expressing grammars that
+capture the behavior of automata. \max{sync this up with Steven's
+  stuff, and add stuff about PDAs and Turing Machines}.
+
+A key construct in formal language theory is the Brzozowski derivative
+\cite{brzozowski}. Given a string $w$ the (left) derivative of a
+language $L$ with respect to $w$, written $\partial_w L$ is the language
+\[ (\partial_wL ) = \{ ww' \pipe w' \in L \}\]
+This semantics is given using the linear implication $w \tol
+L$. Expressing Brzozowski's calculation of the derivative of a regular
+expression relies on additional axioms to the type theory.\max{axioms
+  to carry out Brzozowski's derivative calculations plz.}
+
+\steven{The axioms that I think we need:
+
+$c, c'$ characters. $g$ any grammar.
+
+\begin{gather}
+  c \lto c \cong I \\
+  c \lto c' \cong 0, c \neq c' \\
+  c \lto I \cong 0 \\
+  c \lto g \otimes h \cong \left( (c \lto g) \otimes h \right) \oplus \left( (g \amp I) \otimes (c \lto h) \right) \\
+  I \lto c \cong c \\
+  c \lto (g \oplus h) \cong (c \lto g) \oplus (c \lto h) \\
+  c \lto (c' \lto g) \cong (c' \otimes c) \lto g
+\end{gather}
+
+For point $(5)$, Brzozowski has $I \lto g \cong g$ for all grammars $g$, but I don't think that holds for us because we want to preserve ambiguity.
+We could instead have something like $I \lto g \cong (g \amp I)^{*} \otimes g$.
+For our purposes the character presentation seems sufficient, as the derivative reasoning Brzozowski uses only takes derivatives with respect to strings.
+
+We want $c \lto g^{*} \cong (g \amp I)^{*} \otimes (c \lto g) \otimes g^{*}$, but this is derivable utilizing axiom $(4)$ and $g^{*} \cong I \oplus (g \otimes g^{*})$.
+
+To make the DFA proof go through I think we additionally need the map $c \otimes (c \lto A) \vdash (c \otimes \top) \amp A$ to be an iso.
+Specifically, this term needs an iso $\phi : c \otimes (c \lto A) \to (c \otimes \top) \amp A$ defined below,
+
+\[
+p : c \otimes (c \lto A) \vdash \phi(p) = \letin {p_{1} \otimes p_{2}} p {(p_{1} \otimes \top, p_{1}^{\lto}p_{2})} : (c \otimes \top) \amp A
+\]
+
+}
+
+Using the $G$ modality, we can internalize some of the external
+reasoning about grammars into the language of non-linear types. Recall
+that the denotation of $G A$ is the set of parses of the empty
+string. As is well known in formal language theory, all the parsing
+behavior of a grammar can be reduced to reasoning about parsing of the
+empty string by using the derivative of a grammar. For instance, the
+set of parses for a fixed string $w$ can be defined as $G(w \tol A)$
+and so the type of all parses of a grammar can be defined as
+$\Sigma(w:\textrm{String}) G(w \tol A)$. This can be used to define
+internally a notion of \emph{ambiguity} of a grammar. A grammar $A$ is
+ambiguous \steven{This is an unambiguous grammar, right?} if for every string $w$, the type $G(w \tol A)$ of
+$w-$parses is a \emph{proposition} in the sense that all terms in $G(w
+\tol A)$ are equal.\max{I think we should add a propositional
+  truncation to the type theory.}  Similarly, the type of \emph{total
+parsers} can be defined as $G(\top \tol A)$. A total parser encodes a
+function that selects a parse for every possible string. A
+\emph{semi-parser} is then defined as $G(\top \tol A \oplus \top)$,
+where the parser is allowed to fail for any reason. This has the
+downside that the always failing semi-parser $\lambda _. \textrm{in}_r
+()$ is valid for any grammar. If we were working with a full
+\emph{bunched} type theory, we would be able to express the more
+precise type of a \emph{deciding} parser as $G(\top \tol A \oplus \neg
+A)$ where $\neg A = A \Rightarrow 0$ uses the intuitionistic
+implication.
+
+Classically, formal language theory is closely related to automata theory. In the styles of Chomsky and
+Sch\"utzenberger, there is a hierarchy of language classes and the types of automata that recognize them. Inside of the
+type theory of \cref{sec:typetheory}, we can characterize these language classes; moreover, we demonstrate equivalence
+of each class to the associated automaton as an isomorphism within our logic.
 
@@
 
-We reflect on this expressive power to motivate why we put heavy syntactic restrictions on our presentation of certain classes of automata.
-For instance, when defining a pushdown automata, we could use the intuitionistic context to represent the stack, but we ought to be very careful about how this set is accessed and added to.
-Careless definitions may accidentally endow us with too much expressive power, launching us further up in the Chomsky hierarchy.
-To begin, let us focus only on language classes, and their associated automata, whose parses are definable without any mention to the inuitionistic context.
-When viewed as a machine, this refers to the class of automata definable without access to any memory --- i.e.\ finite state machines.
+We reflect on this expressive power to motivate why we put heavy syntactic restrictions on our presentation of certain
+classes of automata. For instance, when defining a pushdown automata, we could use the intuitionistic context to
+represent the stack, but we ought to be very careful about how this set is accessed and added to. Careless definitions
+may accidentally endow us with too much expressive power, launching us further up in the Chomsky hierarchy. To begin,
+let us focus only on language classes, and their associated automata, whose parses are definable without any mention to
+the inuitionistic context. When viewed as a machine, this refers to the class of automata definable without access to
+any memory --- i.e.\ finite state machines.
 
 \subsection{Finite State Machines}
-A finite state machine may be expressed via a particular syntactic form for grammars.
-A \emph{nondeterministic finite automaton} (NFA) is defined as a mutual\footnote{For convenience we refer to a mutual fixed point here, even though we only take single fixed points as primitive.
-As in Beki\'c's theorem, we may decompose our mutual fixed points into a series of single fixed points.} fixed point in the following manner,
+A finite state machine may be expressed via a particular syntactic form for grammars. A \emph{nondeterministic finite automaton} (NFA) is defined as a
+mutual\footnote{For convenience we refer to a mutual fixed point here, even though we
+  only take single fixed points as primitive. As in Beki\'c's theorem, we may decompose our mutual fixed points into a series of single fixed points.}
+fixed point in the following manner,
 
@@
 \]
-and $Y \in \{ Y_{s}, Y_{1}, \dots, Y_{N} \}$.
-Intuitively, we read each $Y_{j}$ as the label for a state, and the definitions in form $Q$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$ denotes an $\epsilon$-transition.
-Without loss of generality, $Y_{s}$ is reserved for a single distinguished start state, and $Y_{e}$ is reserved for a single distinguished accepting state.
-A disjunction\footnote{There is an admissible notion of coproduct making use of the primitive dependent pairing --- i.e.
+and $Y \in \{ Y_{s}, Y_{1}, \dots, Y_{N} \}$. Intuitively, we read each $Y_{j}$ as the label for a state, and the
+definitions in form $Q$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$
+denotes an $\epsilon$-transition. Without loss of generality, $Y_{s}$ is reserved for a single distinguished start
+state, and $Y_{e}$ is reserved for a single distinguished accepting state. A
+disjunction\footnote{There is an admissible notion of coproduct making use of the primitive dependent pairing --- i.e.
 $A \oplus B := \LinPiTy b \Bool {\textbf{if}~b~A~B}$.
-We will utilize this without much comment to represent the usual notion of alternation of grammars.} of transitions in an NFA grammar corresponds to multiple outgoing edges of a given state in an automaton.
+We will utilize this without much comment to represent the usual notion of alternation of grammars.}
+of transitions in an NFA grammar corresponds to multiple outgoing edges of a given state in an automaton.
 
 Let $\hat{Y}_{k}$ refer to the syntactic term on the right hand side of the definition of $Y_{k}$ inside of an NFA $N$.
-We say that $N$ is a \emph{deterministic finite automaton} (DFA) if there are no $\epsilon$-transitions and for every definition $\hat{Y}_{k}$ the $c_{j}$ are distinct guards for each produced state.
+We say that $N$ is a \emph{deterministic finite automaton} (DFA) if there are no $\epsilon$-transitions and for every
+definition $\hat{Y}_{k}$ the $c_{j}$ are distinct guards for each produced state.
+
 % \steven{Is this the right notion of determinism?
@@
 
-The class of \emph{regular} grammars is given as follows:
-
-\[
-Reg ::= 0 ~|~ I ~|~ c ~|~ Reg \oplus Reg ~|~ Reg \otimes Reg ~|~ Reg^{*}
-\]
-where $c \in \mathcal{A}$ is a character literal drawn from a fixed alphabet, and the Kleene star $\cdot^{*}$ is derived as an admissible operation using the fixed point operator --- $g^{*} = \mu X.I \oplus (g \otimes X)$.
+% The class of \emph{regular} grammars is given as follows:
+
+% \[
+% Reg ::= 0 ~|~ I ~|~ c ~|~ Reg \oplus Reg ~|~ Reg \otimes Reg ~|~ Reg^{*}
+% \]
+% where $c \in \mathcal{A}$ is a character literal drawn from a fixed alphabet, and the Kleene star $\cdot^{*}$ is derived as an admissible operation using the fixed point operator --- $g^{*} = \mu X.I \oplus (g \otimes X)$.
 \steven{This definition is maybe fair to assume.
@@
   \label{thm:thompson}
-  For $g$ a regular grammar, there is an NFA $N$ that recognizes the same language as $g$. Further, $N$ and $g$ are isomorphic.
+  For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same language as $g$. Further, $N$ and $g$ are isomorphic.
 \end{theorem}
@@
 
-We can now inductively define some NFA combinators and show that they preserve grammar equivalence.
-First begin with the disjunction of two NFAs.
-That is, assuming that $g \cong NFA(g)$ and $g' \cong NFA(g')$, we define an NFA that recognizes $g \oplus g'$.
-Let the states in $NFA(g)$ be represented by $X$'s and the states in $NFA(g')$ be represented by $Z$'s.
-Further let $\hat{X}_{k}$ refer to the syntactic term on the right hand side of the definition of $X_{k}$ inside of $NFA(g)$.
-Likewise define each $\hat{Z}_{k}$.
+We can now inductively define some NFA combinators and show that they preserve grammar equivalence. First begin with
+the disjunction of two NFAs. That is, assuming that $g \cong NFA(g)$ and $g' \cong NFA(g')$, we define an NFA that
+recognizes $g \oplus g'$. Let the states in $NFA(g)$ be represented by $X$'s and the states in $NFA(g')$ be represented
+by $Z$'s. Further let $\hat{X}_{k}$ refer to the syntactic term on the right hand side of the definition of $X_{k}$
+inside of $NFA(g)$. Likewise define each $\hat{Z}_{k}$.
 
@@
 \]
-where the substitution $\{X_{s}'/X_{s}, X_{e}'/X_{e}, X_{1}'/X_{1}, \dots \}$ is given by $\delta_{X}$. Likewise define $\delta_{Z}$.
-In the construction of $NFA(g \oplus g')$ we replicate each defintion $X_{j}$ in $NFA(g)$ with an analogous definition $X_{j}'$ inside of $NFA(g \oplus g')$, except we replace $X_{k}$ with $X_{k}'$ in each production.
-Lastly, $X_{e}'$ has an $\epsilon$-transition to the accepting state $Y_{e}$.
-Likewise, we build analogous defintions for the states $Z_{j}$ in $NFA(g')$, then we flatten all of these defintions into one mutally recursive list.
-This has the effect of disjunctively tying together $NFA(g)$ and $NFA(g')$ into one automaton.
-
-To show the strong equivalence of $NFA(g \oplus g')$ and $g \oplus g'$, we will demonstrate an isomorphism between them with two proof terms in our logic that mutually invert each other.
-To define a map $NFA(g \oplus g') \to g \oplus g'$ we will use a multi-fold over all the mutually recursive defintions of $NFA(g \oplus g')$.
-Note, even though multiple folding is not taken as primitive, using Beki\'c's theorem we can define a family of multifold terms to serve as elimination forms for our admissible multiple fixed points.
-That is, rules of the form,
+where the substitution $\{X_{s}'/X_{s}, X_{e}'/X_{e}, X_{1}'/X_{1}, \dots \}$ is given by $\delta_{X}$. Likewise define
+$\delta_{Z}$.
+In the construction of $NFA(g \oplus g')$ we replicate each defintion $X_{j}$ in $NFA(g)$ with an analogous definition
+$X_{j}'$ inside of $NFA(g \oplus g')$, except we replace $X_{k}$ with $X_{k}'$ in each production. Lastly, $X_{e}'$ has
+an $\epsilon$-transition to the accepting state $Y_{e}$. Likewise, we build analogous defintions for the states
+$Z_{j}$ in $NFA(g')$, then we flatten all of these defintions into one mutally recursive list. This has the effect
+of disjunctively tying together $NFA(g)$ and $NFA(g')$ into one automaton.
+
+To show the strong equivalence of $NFA(g \oplus g')$ and $g \oplus g'$, we will demonstrate an isomorphism between them
+with two proof terms in our logic that mutually invert each other. To define a map $NFA(g \oplus g') \to g \oplus g'$
+we will use a multi-fold over all the mutually recursive defintions of $NFA(g \oplus g')$. Note, even though multiple
+folding is not taken as primitive, using Beki\'c's theorem we can define a family of multifold terms to serve as
+elimination forms for our admissible multiple fixed points. That is, rules of the form,
 
@@
 
-are admissible, where the substitution $\{B_{1}/X_{1},\dots,B_{n}/X_{n} \}$ is given by $\gamma$.
-In this instance, we wish to simultaneously fold over the mutually recursive definitions comprising $NFA(g \oplus g')$.
-We may think of this as an eliminator that replaces each state inside of $NFA(g \oplus g')$ with a tail-recursive call that will ultimately return a term of type $g \oplus g'$.
-The structure of such a derivation is given below,
+are admissible, where the substitution $\{B_{1}/X_{1},\dots,B_{n}/X_{n} \}$ is given by $\gamma$. In this instance, we
+wish to simultaneously fold over the mutually recursive definitions comprising $NFA(g \oplus g')$. We may think of this
+as an eliminator that replaces each state inside of $NFA(g \oplus g')$ with a tail-recursive call that will ultimately
+return a term of type $g \oplus g'$. The structure of such a derivation is given below,
 
@@
 
-For each of the variables above, we simplify the type and provide the actual proof term.
-For instance, $\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$, so the derivation for $p_{s}$ is just the identity.
+For each of the variables above, we simplify the type and provide the actual proof term. For instance,
+$\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$, so the derivation for $p_{s}$ is just the identity.
 The rest of the terms are given below,
@@
 where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively.
-This finshes the construction of a parse transformer $NFA(g \oplus g') \to g \oplus g'$ The map in the other direction is given the derivation,
+This finshes the construction of a parse transformer $NFA(g \oplus g') \to g \oplus g'$.
+The map in the other direction is given the derivation,
 
@@
 
-By breaking into a case distinction on $e$ and then applying the $\eta$-rule for coproducts, the above proof will simplify into two cases where $\phi$ will be inverted by $\phi^{{-1}}$ and $\psi$ will be inverted by $\psi^{-1}$, ultimately cancelling to give the identity map.
-The other direction below simplifies in a similar manner.
+By breaking into a case distinction on $e$ and then applying the $\eta$-rule for coproducts, the above proof will
+simplify into two cases where $\phi$ will be inverted by $\phi^{{-1}}$ and $\psi$ will be inverted by $\psi^{-1}$,
+ultimately cancelling to give the identity map. The other direction below simplifies in a similar manner.
 \steven{Most concise list for this?}
@@
 
-This concludes the disjunctive combination of two NFAs. Because they follow a very similar style of argumentation, we include the cases for sequencing and Kleene star in \cref{sec:appendixnfa}.
+This concludes the disjunctive combination of two NFAs.
+Because they follow a very similar style of argumentation, we include the cases for sequencing and
+Kleene star in \cref{sec:appendixnfa}.
 
@@
 
-For this taylor iso to hold, we need the map $c \otimes (c \lto A) \vdash (c \otimes \top) \amp A$ to be an iso.
-Specifically, this term needs an iso $\phi : c \otimes (c \lto A) \to (c \otimes \top) \amp A$ defined below,
-
-\[
-p : c \otimes (c \lto A) \vdash \phi(p) = \letin {p_{1} \otimes p_{2}} p {(p_{1} \otimes \top, p_{1}^{\lto}p_{2})} : (c \otimes \top) \amp A
-\]
-
 Also add the axiom that $\top \cong (\sum_{c \in \mathcal{A}} c)^{*}$
@@
 
-From here we need to also show that derivatives of regular languages are regular (look at Brzozowksi's proof) and that regular languages have finitely many derivatives (potentially with some size constraint on the derivative?).
-Then the last iso in the chain above can be used to build a DFA equivalent to $A$ using Brzozowksi's algorithm.
+From here we need to also show that derivatives of regular languages are regular (look at Brzozowski's proof) and that regular languages have finitely many derivatives (potentially with some size constraint on the derivative?).
+Then the last iso in the chain above can be used to build a DFA equivalent to $A$ using Brzozowski's algorithm.
 
@@
 
-\section{NFA stuff}
+\section{Equivalence Betweeen Regular Grammars and Finite Automata}
 \label{sec:appendixnfa}

### 2024-01-23 paper/paper.tex

@@
 }
+\begin{align*}
+  c\tol c &\equiv I\\
+  c\tol d &\equiv 0\\
+  c\tol I &\equiv 0\\
+  c\tol 0 &\equiv 0\\
+  c\tol (A \otimes B) &\equiv (c\tol A) \otimes B + \nu(A) \otimes (c\tol B)\\
+  c\tol A^* &\equiv (c\tol A) \otimes A^*\\
+  c\tol (A + B) &\equiv (c\tol A) + (c\tol B)
+\end{align*}
+
+\pedro{The type family $\nu$ above is equal to $I$ if its argument
+  parses the empty string and zero otherwise. In order to define
+  it we should probably add an extra axiom, right?}
 

### 2024-01-23 paper/paper.tex
fix merge conflicts, delete redundant axioms

@@
 
-\steven{The axioms that I think we need:
-
-$c, c'$ characters. $g$ any grammar.
-
-\begin{gather}
-  c \lto c \cong I \\
-  c \lto c' \cong 0, c \neq c' \\
-  c \lto I \cong 0 \\
-  c \lto g \otimes h \cong \left( (c \lto g) \otimes h \right) \oplus \left( (g \amp I) \otimes (c \lto h) \right) \\
-  I \lto c \cong c \\
-  c \lto (g \oplus h) \cong (c \lto g) \oplus (c \lto h) \\
-  c \lto (c' \lto g) \cong (c' \otimes c) \lto g
-\end{gather}
-
-For point $(5)$, Brzozowski has $I \lto g \cong g$ for all grammars $g$, but I don't think that holds for us because we want to preserve ambiguity.
-We could instead have something like $I \lto g \cong (g \amp I)^{*} \otimes g$.
-For our purposes the character presentation seems sufficient, as the derivative reasoning Brzozowski uses only takes derivatives with respect to strings.
-
-We want $c \lto g^{*} \cong (g \amp I)^{*} \otimes (c \lto g) \otimes g^{*}$, but this is derivable utilizing axiom $(4)$ and $g^{*} \cong I \oplus (g \otimes g^{*})$.
-
-To make the DFA proof go through I think we additionally need the map $c \otimes (c \lto A) \vdash (c \otimes \top) \amp A$ to be an iso.
-Specifically, this term needs an iso $\phi : c \otimes (c \lto A) \to (c \otimes \top) \amp A$ defined below,
-
-\[
-p : c \otimes (c \lto A) \vdash \phi(p) = \letin {p_{1} \otimes p_{2}} p {(p_{1} \otimes \top, p_{1}^{\lto}p_{2})} : (c \otimes \top) \amp A
-\]
-
-}
 \begin{align*}
@@
   it we should probably add an extra axiom, right?}
-
+\steven{$\nu$ can be derived using $\amp$, $\nu(A) = I \amp A$}
+
+\steven{In addition to Pedro's axioms:
+
+$c, c'$ characters. $g$ any grammar.
+
+\begin{gather}
+  I \lto c \cong c \\
+  c \lto (c' \lto g) \cong (c' \otimes c) \lto g
+\end{gather}
+
+For axiom $(1)$, Brzozowski has $I \lto g \cong g$ for all grammars $g$, but I don't think that holds for us because we want to preserve ambiguity.
+We could instead have something like $I \lto g \cong (g \amp I)^{*} \otimes g$.
+For our purposes the character presentation seems sufficient, as the derivative reasoning Brzozowski uses only takes derivatives with respect to strings.
+
+Axiom $(2)$ might be admissible, but I'm unsure.
+
+I think for Kleene star we need $c \lto g^{*} \cong (g \amp I)^{*} \otimes (c \lto g) \otimes g^{*}$.
+This should derivable utilizing the product axiom and $g^{*} \cong I \oplus (g \otimes g^{*})$, or is that sort of isomorphism opaque to us right now.
+
+To make the DFA proof go through I think we additionally need to axiomatically make the map $c \otimes (c \lto A) \vdash (c \otimes \top) \amp A$ to be an iso.
+Specifically, this term needs an iso $\phi : c \otimes (c \lto A) \to (c \otimes \top) \amp A$ defined below,
+
+\[
+p : c \otimes (c \lto A) \vdash \phi(p) = \letin {p_{1} \otimes p_{2}} p {(p_{1} \otimes \top, p_{1}^{\lto}p_{2})} : (c \otimes \top) \amp A
+\]
+
+but this last bit isn't just Brzozowski's definitions
+
+}
 Using the $G$ modality, we can internalize some of the external
@@
 
-\steven{An aside on some axioms}
-
-Something like $A \cong (I \& A) \oplus \sum_{c} c \otimes (c \lto A)$
-
-Also add the axiom that $\top \cong (\sum_{c \in \mathcal{A}} c)^{*}$
-
-If we have the above, we can then show the following isos for some alphabet $\mathcal{A}$,
-\begin{align*}
-  A
-  & \cong \top \amp A \\
-  & \cong (\Sigma_{c \in \mathcal{A}})^{*}\amp A \\
-  & \cong (I \oplus (\Sigma_{c} c \otimes \top)) \amp A \\
-  & \cong (I \amp A) \oplus ((\Sigma_{c} c \otimes \top)) \amp A) \\
-  & \cong (I \amp A) \oplus (\Sigma_{c }c \otimes (c \lto A)) & \text{using } \phi^{-1}
-\end{align*}
-
-From here we need to also show that derivatives of regular languages are regular (look at Brzozowski's proof) and that regular languages have finitely many derivatives (potentially with some size constraint on the derivative?).
+% \steven{An aside on some axioms}
+
+% Something like $A \cong (I \& A) \oplus \sum_{c} c \otimes (c \lto A)$
+
+% Also add the axiom that $\top \cong (\sum_{c \in \mathcal{A}} c)^{*}$
+
+% If we have the above, we can then show the following isos for some alphabet $\mathcal{A}$,
+% \begin{align*}
+%   A
+%   & \cong \top \amp A \\
+%   & \cong (\Sigma_{c \in \mathcal{A}})^{*}\amp A \\
+%   & \cong (I \oplus (\Sigma_{c} c \otimes \top)) \amp A \\
+%   & \cong (I \amp A) \oplus ((\Sigma_{c} c \otimes \top)) \amp A) \\
+%   & \cong (I \amp A) \oplus (\Sigma_{c }c \otimes (c \lto A)) & \text{using } \phi^{-1}
+% \end{align*}
+
+% From here we need to also show that derivatives of regular languages are regular (look at Brzozowski's proof) and that regular languages have finitely many derivatives (potentially with some size constraint on the derivative?).
 Then the last iso in the chain above can be used to build a DFA equivalent to $A$ using Brzozowski's algorithm.
 
-Need to also add in axioms for $\amp$ distributing over $\oplus$ and $\lto$ distributing over $\oplus$. And that $c \amp c' \vdash 0$ for $c \neq c'$
+% Need to also add in axioms for $\amp$ distributing over $\oplus$ and $\lto$ distributing over $\oplus$. And that $c \amp c' \vdash 0$ for $c \neq c'$
 
@@
 \steven{Worth spending some more time on the propositional truncation idea}
-
-\steven{Put powerset construction here}
 

### 2024-01-24 paper/paper.tex
NFA-tuple to grammar

@@
 \subsection{Finite State Machines}
-A finite state machine may be expressed via a particular syntactic form for grammars. A \emph{nondeterministic finite automaton} (NFA) is defined as a
+A \emph{nondeterministic finite automaton} (NFA) is a finite state machine where transitions are labeled with characters from a fixed alphabet $\Sigma$.
+These are often represented formally as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$,
+
+\begin{itemize}
+  \item $Q$ a finite set of states
+  \item $\Sigma$ a fixed, finite alphabet
+  \item $\delta : Q \times \Sigma \to \mathcal{P}(Q)$ a labelled transition function
+  \item $q_{0}$ a start state
+  \item $F$ a set of accepting states
+\end{itemize}
+
+Intuitively, this can be thought of like a directed graph with nodes in $Q$ with an edge $q \overset{c}{\to} q'$ whenever $q' \in \delta(q, c)$.
+Note that transitions in an NFA may be labelled with the empty string $\epsilon$.
+
+As a gramamar in our type theory, an NFA may be encoded as a
 mutual\footnote{For convenience we refer to a mutual fixed point here, even though we
   only take single fixed points as primitive. As in Beki\'c's theorem, we may decompose our mutual fixed points into a series of single fixed points.}
-fixed point in the following manner,
+fixed point in the following manner:
+
+An NFA has finitely many states, so we may number them $Q = \{q_{1}, \dots, q_{N}\}$. For each $q_{j}$ we introduce
+a mutually recursive term $Y_{j}$ to represent the states of the NFA within this grammar presentation.
+We represent the transition function $\delta$ in the right hand side of each $Y_{j}$ definition.
+For character $c \in \Sigma \cup \{\epsilon\}$ and state $q_{j} \in Q$, we add a summand --- as a coproduct\footnote{There is an admissible notion of coproduct making use of the primitive dependent pairing --- i.e.
+$A \oplus B := \LinPiTy b \Bool {\textbf{if}~b~A~B}$.
+We will utilize this without much comment to represent the usual notion of alternation of grammars.} --- to the defintion of $Y_{j}$ for every state in the image of the transition function,
+
+\[
+Y_{j} = \bigoplus_{c \in \Sigma \cup \{\epsilon\}} \bigoplus_{q_{i} \in \delta(q_{j}, c)} c \otimes Y_{i}
+\]
+
+For convenience, refer to syntactic term on the right hand side of $Y_{j}$ as $\hat{Y}_{j}$.
+Further, we rename the start state to be represented by the term $Y_{s}$.
+Additionally, and without loss of generality, we assume that there is a single accepting state $Y_{e}$ in our encoding.
+This assupmtion is justified through the use of $\epsilon$-transitions.
+For each $q_{j} \in F$, we add an $\epsilon$-transition from $Y_{j}$ to $Y_{e}$.
+
+The above translation totals to give us a syntactic form for NFAs as grammars,
 
@@
   NFA ::= \mu \begin{pmatrix}
-        Y_{s} = \langle Q \rangle \\
-        Y_{1} = \langle Q \rangle \\
-        \vdots \\
+        Y_{s} = \langle \hat{Y} \rangle \\
+        Y_{j} = \langle \hat{Y} \rangle, j \not \in \{s, e\} \\
         Y_{e} = I
-        \end{pmatrix}~\textbf{in}~Y_{s}
+        \end{pmatrix}.Y_{s}
 \]
@@
 \[
-  Q ::= c \otimes Y ~|~ Y ~|~ Q \oplus Q
-\]
-and $Y \in \{ Y_{s}, Y_{1}, \dots, Y_{N} \}$. Intuitively, we read each $Y_{j}$ as the label for a state, and the
-definitions in form $Q$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$
-denotes an $\epsilon$-transition. Without loss of generality, $Y_{s}$ is reserved for a single distinguished start
-state, and $Y_{e}$ is reserved for a single distinguished accepting state. A
-disjunction\footnote{There is an admissible notion of coproduct making use of the primitive dependent pairing --- i.e.
-$A \oplus B := \LinPiTy b \Bool {\textbf{if}~b~A~B}$.
-We will utilize this without much comment to represent the usual notion of alternation of grammars.}
-of transitions in an NFA grammar corresponds to multiple outgoing edges of a given state in an automaton.
-
-Let $\hat{Y}_{k}$ refer to the syntactic term on the right hand side of the definition of $Y_{k}$ inside of an NFA $N$.
-We say that $N$ is a \emph{deterministic finite automaton} (DFA) if there are no $\epsilon$-transitions and for every
-definition $\hat{Y}_{k}$ the $c_{j}$ are distinct guards for each produced state.
-
-% \steven{Is this the right notion of determinism?
-Perhaps, it is is better to define determinism as something like: there is at most one derivation for all $j, k$, for all $a \in \Sigma$ we have $x : Y_{k}, y : a \vdash z : Y_{j}$.}
+  \hat{Y} ::= c \otimes Y ~|~ Y ~|~ \hat{Y} \oplus \hat{Y}
+\]
+
+and each $Y \in \{ Y_{s}, Y_{e}, Y_{1}, \dots, Y_{N} \}$. Intuitively, we read each $Y_{j}$ as the label for a state, and the
+definitions of form $\hat{Y}$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$
+denotes an $\epsilon$-transition. A
+disjunction of transitions in an NFA grammar corresponds to multiple outgoing edges of a given state in an automaton.
+
+Above we give a translation from the tuple presentation of an NFA to a fixed point grammar in our type theory.
+We don't provide any argument that this translation is done correctly; however, we hope that the construction is straightforward and lightweight enough to demonstrate our formal system as a natural and effective syntax for reasoning about both grammars and automata.
 
 With this setting for finite automata, we can now internalize some classical theorems inside of our formal system.
-First, let us describe a version of Thompson's construction \cite{?} where we show that any regular expression can be recognized by some NFA\@.
-Moreover, we will show that this NFA is strongly equivalent to the original grammar.
 
@@
 
-\begin{theorem}
+First, let us describe a version of Thompson's construction \cite{?} where we construct an NFA that recognizes a given regular expression.
+Moreover, we will show that this NFA is strongly equivalent to the original grammar.
+
+This construction
+
+\begin{theorem}[Thompson]
   \label{thm:thompson}
@@
         Y_{e} = I
-  \end{pmatrix}~\textbf{in}~Y_{s}
+  \end{pmatrix}.Y_{s}
 \]
@@
 We can now inductively define some NFA combinators and show that they preserve grammar equivalence. First begin with
-the disjunction of two NFAs. That is, assuming that $g \cong NFA(g)$ and $g' \cong NFA(g')$, we define an NFA that
-recognizes $g \oplus g'$. Let the states in $NFA(g)$ be represented by $X$'s and the states in $NFA(g')$ be represented
+the disjunction of two NFAs. For regular grammars $g$ and $g'$, assume that $g \cong NFA(g)$ and $g' \cong NFA(g')$, and we will define an NFA that
+recognizes $g \oplus g'$.
+
+Let the states in $NFA(g)$ be represented by $X$'s and the states in $NFA(g')$ be represented
 by $Z$'s. Further let $\hat{X}_{k}$ refer to the syntactic term on the right hand side of the definition of $X_{k}$
@@
         Y_{e} = I
-  \end{pmatrix}~\textbf{in}~Y_{s}
-\]
-where the substitution $\{X_{s}'/X_{s}, X_{e}'/X_{e}, X_{1}'/X_{1}, \dots \}$ is given by $\delta_{X}$. Likewise define
-$\delta_{Z}$.
+  \end{pmatrix}.Y_{s}
+\]
+where the substitution $\{X_{s}'/X_{s}, X_{e}'/X_{e}, X_{1}'/X_{1}, \dots \}$ is given by $\delta_{X}$ --- and $\delta_{Z}$ constructed similarly.
+
 In the construction of $NFA(g \oplus g')$ we replicate each defintion $X_{j}$ in $NFA(g)$ with an analogous definition
@@
 we will use a multi-fold over all the mutually recursive defintions of $NFA(g \oplus g')$. Note, even though multiple
-folding is not taken as primitive, using Beki\'c's theorem we can define a family of multifold terms to serve as
-elimination forms for our admissible multiple fixed points. That is, rules of the form,
-
-\[
-  \inferrule{\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n})~\textbf{in}~X_{k} \\
+fixed points are not taken as primitive, using Beki\'c's theorem we can define admissible multiple fixed points types.
+Below we give rules to form $\mathsf{cons}$ and $\mathsf{mfold}$ terms to serve as the introduction and elimination rules, respectively, for multi-term fixed points,
+
+\[
+  \inferrule
+  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
+  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}) . X_{k}}
+\]
+where $\sigma$ is the substitution that unrolls the mutually recursive definition one level by replacing each $X_{k}$ with the fixed point,
+\steven{I'm not sure this is right. TODO sanity check it with a small example}
+\begin{align*}
+  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}) . X_{1} / X_{1} , \dots, \\
+  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}) . X_{n} / X_{n} \}
+\end{align*}
+
+\[
+  \inferrule{\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
              \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
@@
 
-are admissible, where the substitution $\{B_{1}/X_{1},\dots,B_{n}/X_{n} \}$ is given by $\gamma$. In this instance, we
+where the substitution $\gamma$ represents $\{B_{1}/X_{1},\dots,B_{n}/X_{n} \}$.
+
+When defining a derivation from $NFA(g \oplus g')$ to $g \oplus g'$, we
 wish to simultaneously fold over the mutually recursive definitions comprising $NFA(g \oplus g')$. We may think of this
@@
 where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively.
-This finshes the construction of a parse transformer $NFA(g \oplus g') \to g \oplus g'$.
+This finshes the construction of a parse transformer from $NFA(g \oplus g')$ to $g \oplus g'$.
 The map in the other direction is given the derivation,

### 2024-01-24 paper/paper.tex
update multifixed point inference rules

@@
 To show the strong equivalence of $NFA(g \oplus g')$ and $g \oplus g'$, we will demonstrate an isomorphism between them
-with two proof terms in our logic that mutually invert each other. To define a map $NFA(g \oplus g') \to g \oplus g'$
+with two proof terms in our logic that mutually invert each other. To define a map from $NFA(g \oplus g')$ to $g \oplus g'$
 we will use a multi-fold over all the mutually recursive defintions of $NFA(g \oplus g')$. Note, even though multiple
-fixed points are not taken as primitive, using Beki\'c's theorem we can define admissible multiple fixed points types.
-Below we give rules to form $\mathsf{cons}$ and $\mathsf{mfold}$ terms to serve as the introduction and elimination rules, respectively, for multi-term fixed points,
+fixed points are not taken as primitive, using Beki\'c's theorem we can define an admissible axiom schema to define multiple fixed points types.
+Below we give rules to form $\mathsf{cons}$ and $\mathsf{mfold}$ terms to serve as the introduction and elimination rules, respectively, for multi-term fixed points.
+First, the introduction rule,
 
@@
   {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
-  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}) . X_{k}}
-\]
-where $\sigma$ is the substitution that unrolls the mutually recursive definition one level by replacing each $X_{k}$ with the fixed point,
-\steven{I'm not sure this is right. TODO sanity check it with a small example}
+  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
+\]
+where $\sigma$ is the substitution that unrolls the mutually recursive definition one level by replacing each $X_{k}$ with the fixed point $\mu(X_{j} = A_{j}).X_{k}$,
+\steven{I'm not sure the cons rule is right. TODO sanity check it with a small example}
+
 \begin{align*}
-  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}) . X_{1} / X_{1} , \dots, \\
-  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}) . X_{n} / X_{n} \}
+  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \\
+  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
 \end{align*}
 
+and the elimination rule,
 \[
@@
 as an eliminator that replaces each state inside of $NFA(g \oplus g')$ with a tail-recursive call that will ultimately
-return a term of type $g \oplus g'$. The structure of such a derivation is given below,
+return a term of type $g \oplus g'$. The structure of such a derivation instatiates the above elimination rule as follow,
 
@@
     x_{s}' : \simulsubst {(\simulsubst {X_{s}} {\delta_{X}})}{\gamma} \vdash p_{x,s} : g \\
-    x_{j}' : \simulsubst {(\simulsubst {X_{j}} {\delta_{X}})}{\gamma} \vdash p_{x,j} : X_{j} \text{ where } j \not \in \{s, e\} \\
+    x_{j}' : \simulsubst {(\simulsubst {X_{j}} {\delta_{X}})}{\gamma} \vdash p_{x,j} : X_{j} \text{~where } j \not \in \{s, e\} \\
     x_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{x, e} : X_{e} \\
     z_{s}' : \simulsubst {(\simulsubst {Z_{s}} {\delta_{Z}})}{\gamma} \vdash p_{z,s} : g' \\
-    z_{j}' : \simulsubst {(\simulsubst {Z_{j}} {\delta_{Z}})}{\gamma} \vdash p_{z,j} : Z_{j} \text{ where } j \not \in \{s, e\} \\
+    z_{j}' : \simulsubst {(\simulsubst {Z_{j}} {\delta_{Z}})}{\gamma} \vdash p_{z,j} : Z_{j} \text{~where } j \not \in \{s, e\} \\
     z_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{z,e} : Z_{e} \\
@@
 where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively.
-This finshes the construction of a parse transformer from $NFA(g \oplus g')$ to $g \oplus g'$.
-The map in the other direction is given the derivation,
-
-\[
-\inferrule{p_{1} : g \vdash \textbf{in}_{X_{s}'}(\phi(p_{1})) : NFA(g \oplus g') \\ p_{2} : g' \vdash \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) : NFA(g \oplus g') \\ p : g \oplus g' \vdash p : g \oplus g'}{p : g \oplus g' \vdash \textbf{case}~p \{\textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\ \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) \} : NFA(g \oplus g')}
+This finshes the construction of a parse transformer from $NFA(g \oplus g')$ to $ g \oplus g'$.
+The map in the other direction is given below in a couple stages. First, we show how to turn a parse of $NFA(g)$ into a parse of $NFA(g \oplus g')$.
+
+\[
+  \inferrule
+  {
+    \Gamma ; \Delta \vdash u : NFA(g) = \mu(X_{j} = \hat{X}_{j}).X_{s} \\
+    \Gamma ; x_{s} : \simulsubst {X_{s}}{\gamma} = X_{s}' \vdash \mathsf{cons}(\textbf{in}_{X_{s}'}(x_{s})) : \mu(\dots).Y_{s} = NFA(g \oplus g')  \\
+    \Gamma ; x_{j} : \simulsubst {X_{j}}{\gamma} = X_{j}' \vdash x_{j} : X_{j}', j \not \in \{s, e\} \\
+    \Gamma ; x_{e} : \simulsubst {X_{e}}{\gamma} = I \vdash x_{e} : I = X_{e}'
+  }
+  {\Gamma ; \Delta \vdash \mathsf{mfold}(u) : NFA(g \oplus g')}
+\]
+where the dots in the above derivation take on the mutually recursive defintions present in $NFA(g \oplus g')$.
+\steven{Need a better way to display this without blowing up the definitions everywhere} Denote the above embedding of $NFA(g)$ into $NFA(g \oplus g')$ via $\alpha(u) := \mathsf{mfold}(u) : NFA(g \oplus g')$.
+Similarly define an embedding $\beta$ for an embedding of $NFA(g')$.
+We can then define the derivation from $g \oplus g'$ to $NFA(g \oplus g')$ by composing the maps isomorphisms $g \overset{\phi}{\cong} NFA(g)$ and $g \overset{\psi}{\cong} NFA(g')$ with $\alpha$ and $\beta$, respectively.
+
+\[
+  \inferrule
+  {
+    p : g \oplus g' \vdash p : g \oplus g' \\
+    p_{1} : g \vdash \alpha(\phi(p_{1})) : NFA(g \oplus g') \\
+    p_{2} : g' \vdash \beta(\psi(p_{2})) : NFA(g \oplus g')}
+  {
+    p : g \oplus g' \vdash \textbf{case}~p \{
+    \textbf{in}_{g} p_{1} \mapsto \alpha(\phi(p_{1})), \\
+    \textbf{in}_{g'} p_{2} \mapsto \beta(\psi(p_{2})) \} : NFA(g \oplus g')}
 \]

### 2024-01-25 paper/paper.tex
equational reasoning for disjunct forward

@@
 \]
-where $\sigma$ is the substitution that unrolls the mutually recursive definition one level by replacing each $X_{k}$ with the fixed point $\mu(X_{j} = A_{j}).X_{k}$,
+where $\sigma$ is the substitution that unrolls the mutually recursive definition one level by replacing each $X_{k}$ with the fixed point $\mu(X_{j} = A_{j}, \forall j).X_{k}$,
 \steven{I'm not sure the cons rule is right. TODO sanity check it with a small example}
@@
 as an eliminator that replaces each state inside of $NFA(g \oplus g')$ with a tail-recursive call that will ultimately
-return a term of type $g \oplus g'$. The structure of such a derivation instatiates the above elimination rule as follow,
+return a term of type $g \oplus g'$.
+If folding over a term $e$, refer to the output of this multifold as $s_{\oplus}(e)$.
+The structure of such a derivation instatiates the above elimination rule as follow,
 
@@
     y_{e} : \simulsubst I {\gamma} \vdash p_{e} : I
-  }{\Delta \vdash \mathsf{mfold}(e) : g \oplus g'}
+  }{\Delta \vdash s_{\oplus}(e) := \mathsf{mfold}(e) : g \oplus g'}
 \]
@@
   \item $y_{s}$ --- $\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$ and $p_{s} = y_{s} : g \oplus g'$
-  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = X_{s}$ and $p_{x,s} = \phi^{-1}(x_{s}') : g$
-  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = X_{j}$ and $p_{x, j} = x_{j}'$
-  \item $x_{e}'$ --- $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{x,e} = x_{e}' : X_{e} = I$
-  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = Z_{s}$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
-  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = Z_{j}$ and $p_{z, j} = z_{j}'$
-  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = z_{e}' : Z_{e} = I$
+  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = \mu(X_{i} = \hat{X}_{i}, \forall i).X_{s} = NFA(g)$ and $p_{x,s} = \phi^{-1}(x_{s}') : g$
+  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = \mu(X_{i} = \hat{X}_{i}, \forall i).X_{j}$ and $p_{x, j} = x_{j}' : \mu(X_{i} = \hat{X}_{i}, \forall i).
+X_{j}$
+  \item $x_{e}'$ --- $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{x,e} = x_{e}' : \mu(X_{i} = \hat{X}_{i}, \forall i).X_{e} = I$
+  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = \mu(Z_{i} = \hat{Z}_{i}, \forall i).Z_{s} = NFA(g')$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
+  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = \mu(Z_{i} = \hat{Z}_{i}, \forall i).Z_{j}$ and $p_{z, j} = z_{j}' : \mu(Z_{i} = \hat{Z}_{i}, \forall i).Z_{j}$
+  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = \mathsf{cons}_{Z_{e}}(z_{e}') : \mu(Z_{i} = \hat{Z}_{i}, \forall i). Z_{e} = I$
   \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
@@
 
+\steven{Need to add argument that justifies why the $X_{j}$s are allowed to be eliminated via the identity.
+A priori they could be interfered with by taking an $X_{s}$ that needs to be eliminated.
+I think it hinges on a syntactic restriction where there are no backedges to $X_{s}$ in the NFA.Even if there were, I think I can get away with a case distinction entailment instead of the one written}
+
 where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively.
-This finshes the construction of a parse transformer from $NFA(g \oplus g')$ to $ g \oplus g'$.
-The map in the other direction is given below in a couple stages. First, we show how to turn a parse of $NFA(g)$ into a parse of $NFA(g \oplus g')$.
+This finshes the construction of a parse transformer $s_{\oplus}$ from $NFA(g \oplus g')$ to $ g \oplus g'$.
+The map $r_{\oplus}$ in the other direction is given below in a couple stages. First, we show how to turn a parse of $NFA(g)$ into a parse of $NFA(g \oplus g')$.
 
@@
   {
-    \Gamma ; \Delta \vdash u : NFA(g) = \mu(X_{j} = \hat{X}_{j}).X_{s} \\
+    \Gamma ; \Delta \vdash u : NFA(g) = \mu(X_{j} = \hat{X}_{j}, \forall j).X_{s} \\
     \Gamma ; x_{s} : \simulsubst {X_{s}}{\gamma} = X_{s}' \vdash \mathsf{cons}(\textbf{in}_{X_{s}'}(x_{s})) : \mu(\dots).Y_{s} = NFA(g \oplus g')  \\
@@
 where the dots in the above derivation take on the mutually recursive defintions present in $NFA(g \oplus g')$.
-\steven{Need a better way to display this without blowing up the definitions everywhere} Denote the above embedding of $NFA(g)$ into $NFA(g \oplus g')$ via $\alpha(u) := \mathsf{mfold}(u) : NFA(g \oplus g')$.
+\steven{Need a better way to display these defs without blowing up the $\mu$ terms} Denote the above embedding of $NFA(g)$ into $NFA(g \oplus g')$ via $\alpha(u) := \mathsf{mfold}(u) : NFA(g \oplus g')$.
 Similarly define an embedding $\beta$ for an embedding of $NFA(g')$.
-We can then define the derivation from $g \oplus g'$ to $NFA(g \oplus g')$ by composing the maps isomorphisms $g \overset{\phi}{\cong} NFA(g)$ and $g \overset{\psi}{\cong} NFA(g')$ with $\alpha$ and $\beta$, respectively.
+We can then define the derivation from $g \oplus g'$ to $NFA(g \oplus g')$ by composing the isomorphisms $g \overset{\phi}{\cong} NFA(g)$ and $g \overset{\psi}{\cong} NFA(g')$ with maps $\alpha$ and $\beta$, respectively.
 
@@
   {
-    p : g \oplus g' \vdash \textbf{case}~p \{
+    p : g \oplus g' \vdash r_{\oplus}(p) := \textbf{case}~p \{
     \textbf{in}_{g} p_{1} \mapsto \alpha(\phi(p_{1})), \\
@@
 
-It is evident that the maps $NFA(g \oplus g') \to g \oplus g'$ and $g \oplus g' \to NFA(g \oplus g')$ form an isomorphism.
-
-\begin{align*}
-  \textbf{case}~\mathsf{mfold}(e) \{\textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\ \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) \}
-  & = e
-\end{align*}
-
-% \begin{align*}
-%   \begin{aligned}
-%     \textbf{case}~\mathsf{mfold}(e) \{\textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\ \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2}))
-%   \end{aligned}
-%   & =
-%   \begin{aligned}
-%     \textbf{case}~(&  \textbf{case}~ e \{ \\
-%     & \textbf{in}_{X_{s}'}u \mapsto \mathsf{mfold}(\textbf{in}_{X_{s}'}u), \\ & \textbf{in}_{Z_{s}'}v \mapsto \mathsf{mfold}(\textbf{in}_{Z_{s}'}v) \\ \}) \{ \\ & \textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\  & \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) \}
-%   \end{aligned} \\
-%   & = \textbf{case}~e \{
-%     \textbf{in}_{X_{s}'}u \mapsto \textbf{in}_{X_{s}'}(\phi(\textsf{mfold}(\textbf{in}_{{X_{s}'}}u))) \\
-%     \textbf{in}_{Z_{s}'} v \mapsto
-%     \}
-    %   \end{align*}
-% This doesn't seem worth it.
-But you argue formally that commuting the case statements as such is done with the eta rule.
-Then you get everything to cancel down to \phi(\phi^{-1}(...)) (or the same with \psi)
-
-% \steven{I could type out the equality proof term, but it blows up in size and doesn't seem like a useful way to capture this iso.
-I have a commented out version of it above.
-Going to refactor tomorrow to something more sane, as these as are the crux of the proofs of equality} \steven{In any case, the NFA-to-grammar direction of these isos should all follow from some $\eta$-rule on the structure of the type for the start state, right?}
-% \pedro{What I like doing in these cases is either presenting the high-level structure of the proof, e.g., ``by applying the rules <...> we can conclude'', or present the full proof in the appendix.
-  % Since in this case it feels like it won't be a particularly illuminating proof, I suggest doing
-% the former.}
-    %
-
-By breaking into a case distinction on $e$ and then applying the $\eta$-rule for coproducts, the above proof will
-simplify into two cases where $\phi$ will be inverted by $\phi^{{-1}}$ and $\psi$ will be inverted by $\psi^{-1}$,
-ultimately cancelling to give the identity map. The other direction below simplifies in a similar manner.
-\steven{Most concise list for this?}
-
-% Without loss of generality assume that $e = \textbf{in}_{X_{s}'} u$.
-Then, $\mathsf{mfold}(e) = \textbf{in}_{g}(\phi^{-1}u)$ and passing this through the above case statement results in $\textbf{in}_{{X_{s}'}}(\phi(\phi^{{-1}}u)) = \textbf{in}_{{X_{s}'}}(u) = e$.
-Further,
-
-\begin{align*}
-  \mathsf{mfold}(\textbf{case}~p \{\textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\ \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) \})
-  & = p
-\end{align*}
-
-This concludes the disjunctive combination of two NFAs.
-Because they follow a very similar style of argumentation, we include the cases for sequencing and
-Kleene star in \cref{sec:appendixnfa}.
+We include the proof that maps $s_{\oplus}$ and $r_{\oplus}$ form an isomorphism in \cref{sec:appendixnfa}. This concludes the disjunctive combination of two NFAs.
+Because they follow a very similar style of argumentation, we also include the cases for sequencing and
+Kleene star in \cref{sec:appendixnfa}. Totaling over all of these cases, this completes the proof.
 
@@
 
-We provide the final two cases of \cref{thm:thompson}.
-
+Below we finish the proof of \cref{thm:thompson}.
 \begin{proof}
+
+  First, we finish the provide the proof that the maps $s_{\oplus}$ and $r_{\oplus}$ provided in the disjunctive case do indeed mutually invert each other.
+For brevity in this proof, we will omit the list of mutually recursive bindings when appropriate.
+That is, write $\mu X_{s}'$ instead of $\mu(Y_{s},X_{s}',X_{j}',\dots).
+X_{s}'$.
+
+  Given $e : NFA(g \oplus g')$,
+
+\begin{align*}
+  r_{\oplus}(s_{\oplus}(e))
+  & =_{\mu \eta} r_{\oplus}(s_{\oplus}(\mathsf{cons}(e'))), \text{~where }e':\mu X_{s}' \oplus \mu Z_{s}' \\
+  &
+    \begin{aligned}
+      =_{\oplus \eta} r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{case}~e'~\{ & \\
+    \textbf{in}_{\mu X_{s}'} u \mapsto \textbf{in}_{\mu X_{s}'} u, & \\
+    \textbf{in}_{\mu Z_{s}'} v \mapsto \textbf{in}_{\mu Z_{s}'} v
+    \})))
+    \end{aligned} \\
+  &
+    \begin{aligned}
+    =_{\oplus \eta}
+    \textbf{case}~e'~\{ & \\
+    & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{in}_{\mu X_{s}'} u))), \\
+    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v))) \\
+    \}))
+    \end{aligned} \\
+  &
+    \begin{aligned}
+    =
+    \textbf{case}~e'~\{ & \\
+    & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(\mathsf{mfold}(\mathsf{cons}(\textbf{in}_{\mu X_{s}'} u))), \\
+    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(\mathsf{mfold}(\mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v))) \\
+    \}
+    \end{aligned} \\
+  &
+    \begin{aligned}
+    =_{\mu \beta}
+    \textbf{case}~e'~\{ & \\
+    & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(\textbf{in}_{g} \phi^{-1}(\mathsf{cons}_{\mu(X_{i}=\hat{X}_{i}).X_{s}}u')), \\
+    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(\textbf{in}_{g'} \psi^{-1}(\mathsf{cons}_{\mu(Z_{i}=\hat{Z}_{i}).Z_{s}}v')) \\
+      \} \\
+      & \text{~for }u' : \mu X_{s}', v' : \mu Z_{s}'
+    \end{aligned} \\
+  &
+    \begin{aligned}
+    =_{\oplus \beta}
+    \textbf{case}~e'~\{ & \\
+    & \textbf{in}_{\mu X_{s}'} u \mapsto \alpha \phi \phi^{-1}(\mathsf{cons}(u')), \\
+    & \textbf{in}_{\mu Z_{s}'} v \mapsto \beta \psi \psi^{-1}(\mathsf{cons}(v')) \\
+      \}
+    \end{aligned} \\
+  &
+    \begin{aligned}
+    =
+    \textbf{case}~e'~\{ & \\
+    & \textbf{in}_{\mu X_{s}'} u \mapsto \alpha (\mathsf{cons}(u')), \\
+    & \textbf{in}_{\mu Z_{s}'} v \mapsto \beta (\mathsf{cons}(v')) \\
+      \}
+    \end{aligned} \\
+   &
+    \begin{aligned}
+    =_{\mu \beta}
+    \textbf{case}~e'~\{ & \\
+    & \textbf{in}_{\mu X_{s}'} u \mapsto \mathsf{cons}(\textbf{in}_{\mu X_{s}'} u) , \\
+    & \textbf{in}_{\mu Z_{s}'} v \mapsto \mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v) \\
+      \}
+    \end{aligned} \\
+  & =_{\oplus \eta} \mathsf{cons}(e') \\
+  & =_{\mu \beta} e
+\end{align*}
+
 We now construct a combinator for the sequencing of two NFAs.

### 2024-01-25 paper/paper.tex
equational reasoning for or, seq, and some of *

@@
   Given $e : NFA(g \oplus g')$,
+
+  \steven{There a weird spacing issues between these align envs. Maybe wrap these in a figure env so that spacing is better}
 
@@
 
+Likewise for $p : g \oplus g'$,
+
+\steven{This first line is a weird alignment fix, touch up}
+\begin{align*}
+  s_{\oplus}(r_{\oplus}(p))
+  &
+  = s_{\oplus}(r_{\oplus}(p)) \\
+   & \begin{aligned}
+    =_{\oplus \eta}
+      \textbf{case}~p~\{
+    \textbf{in}_{g}p_{1} \mapsto s_{\oplus}(r_{\oplus}(\textbf{in}_{g}p_{1})), \\
+    \textbf{in}_{g'}p_{2} \mapsto s_{\oplus}(r_{\oplus}(\textbf{in}_{g'}p_{2})) \}
+    \end{aligned} \\
+   & \begin{aligned}
+    =_{\oplus \beta}
+      \textbf{case}~p~\{
+    \textbf{in}_{g}p_{1} \mapsto s_{\oplus}(\alpha \phi p_{1}), \\
+    \textbf{in}_{g'}p_{2} \mapsto s_{\oplus}(\beta \psi p_{2}) \}
+    \end{aligned} \\
+   & \begin{aligned}
+    =
+      \textbf{case}~p~\{
+    \textbf{in}_{g}p_{1} \mapsto \mathsf{mfold}(\alpha \phi p_{1}), \\
+    \textbf{in}_{g'}p_{2} \mapsto \mathsf{mfold}(\beta \psi p_{2}) \}
+    \end{aligned} \\
+    & \begin{aligned}
+    =_{\mu \beta}
+      \textbf{case}~p~\{
+    \textbf{in}_{g}p_{1} \mapsto \textbf{in}_{g} (\phi^{-1} \phi p_{1}), \\
+    \textbf{in}_{g'}p_{2} \mapsto \textbf{in}_{g'}(\psi^{-1} \psi p_{2}) \}
+    \end{aligned} \\
+     & \begin{aligned}
+    =
+      \textbf{case}~p~\{
+    \textbf{in}_{g}p_{1} \mapsto \textbf{in}_{g} (p_{1}), \\
+    \textbf{in}_{g'}p_{2} \mapsto \textbf{in}_{g'}(p_{2}) \}
+    \end{aligned} \\
+  &
+    =_{\oplus \eta} p
+\end{align*}
+
 We now construct a combinator for the sequencing of two NFAs.
@@
 
-Just as before, we use a multifold to define the map to $g \otimes g'$. Here the table of terms is as follows,
+Just as before, we use a multifold to define the map $s_{\otimes}$ from $NFA(g \otimes g')$ to $g \otimes g'$. Here the table of terms is as follows,
 
@@
   \item $y_{s}$ --- $\simulsubst {X_{s}'} \gamma = g \otimes g'$ and $p_{s} = y_{s} $
-  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = X_{s} \otimes g'$ and $p_{x,s} = \letin {u \otimes v} {x_{s}'} {\phi^{-1}(u) \otimes v} : g \otimes g'$
-  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = X_{j} \otimes g'$ and $p_{x, j} = x_{j}' : X_{j} \otimes g'$
+  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = (\mu (X_{i}=\hat{X}_{i}).X_{s}) \otimes g' = NFA(g) \otimes g'$ and $p_{x,s} = \letin {u \otimes v} {x_{s}'} {\phi^{-1}(u) \otimes v} : g \otimes g'$
+  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = (\mu (X_{i}=\hat{X}_{i}).X_{j}) \otimes g'$ and $p_{x, j} = x_{j}' : (\mu (X_{i}=\hat{X}_{i}).X_{j}) \otimes g'$
   \item $x_{e}'$ --- $\simulsubst {Z_{s}'}{\gamma} = g'$ and $p_{x,e} = x_{e}' : g'$
-  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = Z_{s}$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
-  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = Z_{j}$ and $p_{z, j} = z_{j}'$
-  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = z_{e}' : Z_{e} = I$
+  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = \mu (Z_{i} = \hat{Z}_{i}).Z_{s} = NFA(g')$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
+  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = \mu (Z_{i} = \hat{Z}_{i}).Z_{j}$ and $p_{z, j} = z_{j}' : \mu (Z_{i} = \hat{Z}_{i}).Z_{j}$
+  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = \mathsf{cons}(z_{e}') : \mu (Z_{i} = \hat{Z}_{i}).Z_{e} = I$
   \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
@@
 
-The map in the other direction is given as,
+The map $r_{\otimes}$ in the other direction is given as,
 
@@
     p : g \otimes g' \vdash p : g \otimes g' \\
-    p_{1} : g, p_{2} : g' \vdash \phi(p_{1}) \otimes \psi(p_{2}) : g \otimes g'}
-  {p : g \otimes g' \vdash \letin {p_{1} \otimes p_{2}} {p} {\phi(p_{1}) \otimes \psi(p_{2})} : NFA(g \otimes g')}
-\]
-
-Again, these are clearly isomorphic.
+    p_{1} : g, p_{2} : g' \vdash \mathsf{cons}(\phi(p_{1}), \psi(p_{2})) : NFA(g \otimes g')}
+  {p : g \otimes g' \vdash \letin {p_{1} \otimes p_{2}} {p} {\mathsf{cons}(\phi(p_{1}), \psi(p_{2}))} : NFA(g \otimes g')}
+\]
+
+We show that $r_{\otimes}$ and $s_{\otimes}$ form an isomorhic pair,
+\steven{This proof doesn't exactly work}
+
+\begin{align*}
+r_{\otimes}(s_{\otimes}(e))
+  & =_{\mu \eta} r_{\otimes}(s_{\otimes}(\mathsf{cons}_{\mu X_{s}'}(e'))) \\
+  & = r_{\otimes}(\mathsf{mfold}(\mathsf{cons}_{\mu X_{s}'}(e'))) \\
+  & =_{\mu \beta} r_{\otimes}(\phi^{-1}(u) \otimes \psi^{-1}(v)) \\
+  & =_{\mu \beta} \mathsf{cons}(\phi \phi^{-1} u, \psi \psi^{-1} v) \\
+  & =_{\mu \eta} e
+\end{align*}
+
+\begin{align*}
+  s_{\otimes}(r_{\otimes}(p))
+  & =_{\otimes \eta} s_{\otimes}(r_{\otimes}(p_{g} \otimes p_{g'})) \\
+  & = s_{\otimes}(\mathsf{cons}(\phi p_{g}, \psi p_{g'})) \\
+  & =_{\mu \beta} \phi^{-1}\phi p_{g} \otimes \psi^{-1} \psi p_{g'} \\
+  & = p_{g} \otimes p_{g'} \\
+  & =_{\otimes \eta} p
+\end{align*}
 
@@
 
-We fold over these defintions to get a parse of $g^{*}$,
+Again we define $s_{*}$ as a fold over these defintions to get a parse of $g^{*}$,
 \begin{itemize}
   \item $y_{s}$ --- $\simulsubst {(X_{s}' \oplus Y_{e})} \gamma = g^{*} \oplus I$ and $p_{s} = \textbf{case}~y_{s} \{ \textbf{in}_{g^{*}} u \mapsto u, \textbf{in}_{I} v \mapsto \mathsf{cons}_{*I} v \} : g^{*}$
-  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = X_{s} \otimes g^{*}$ and $p_{x,s} = \letin {u \otimes p_{*}} {x_{s}'} {\mathsf{cons}_{*}(\phi^{-1}(u), p_{*})} : g^{*} $ \steven{slight abuse of notation.
-Not sure how to best capture the Kleene star constructors}
-  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = X_{j} \otimes g^{*}$ and $p_{x, j} = : X_{j} \otimes g^{*}$
-  \item $x_{e}'$ --- $\simulsubst {(X_{s}' \oplus Y_{e})}{\gamma} = g^{*} \oplus I$ and $p_{x,e} = \textbf{case}~x_{e}' \{ \textbf{in}_{g^{*}} u \mapsto u, \textbf{in}_{I} v \mapsto \mathsf{cons}_{*I} v \} : g^{*}$
+  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = (\mu(X_{i}=\hat{X_{i}}).X_{s}) \otimes g^{*} = NFA(g) \otimes g^{*}$ and $p_{x,s} = \letin {u \otimes p_{*}} {x_{s}'} {\mathsf{cons}_{*}(\phi^{-1}(u), p_{*})} : g^{*} $
+  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = (\mu(X_{i}=\hat{X_{i}}).X_{j}) \otimes g^{*}$ and $p_{x, j} = : (\mu(X_{i}=\hat{X_{i}}).X_{j}) \otimes g^{*}$
+  \item $x_{e}'$ --- $\simulsubst {(X_{s}' \oplus Y_{e})}{\gamma} = g^{*} \oplus I$ and $p_{x,e} = \mathsf{cons}_{X_{e}}(()) \otimes \textbf{case}~x_{e}' \{ \textbf{in}_{g^{*}} u \mapsto u, \textbf{in}_{I} v \mapsto \mathsf{cons}_{*I} v \} : (\mu(X_{i}=\hat{X_{i}}).X_{e}) \otimes g^{*}$
   \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
@@
 
-and to derive a parse of $NFA(g^{*})$ from a parse of $g*$,
-
-\[
-  \inferrule{p : g^{*} \vdash p : g^{*} \\ \cdot \vdash \textbf{in}_{{Y_{e}}}(()) : NFA(g^{*}) \\ u : g, v : g^{*} \vdash \textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v))) : NFA(g^{*}) }{p : g^{*} \vdash \mathsf{fold}_{*}(\textbf{in}_{{Y_{e}}}(()), \textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v)))) : NFA(g^{*}) }
-\]
+and to derive a parse of $NFA(g^{*})$ from a parse of $g^{*}$ we define $r_{*}$ as,
+
+\[
+  \inferrule{p : g^{*} \vdash p : g^{*} \\ \cdot \vdash \mathsf{cons}(\textbf{in}_{{Y_{e}}}(())) : NFA(g^{*}) \\ u : g, v : g^{*} \vdash \mathsf{cons}(\textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v)))) : NFA(g^{*}) }{p : g^{*} \vdash \mathsf{fold}_{*}(\mathsf{cons}(\textbf{in}_{{Y_{e}}}(())), \mathsf{cons}(\textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v)))) : NFA(g^{*}) }
+\]
+
+To see that $s_{*}$ and $r_{*}$ for an isomorphism,
+\steven{This proof also needs more work. Esp after 4th equality}
+\steven{Think I need to thread through a case distinction on what the $g^{*}$ term is. Putting a pin in this for now to type up the DFA stuff before meeting}
+\begin{align*}
+  r_{*}(s_{*}(e))
+  & =_{\mu \eta} r_{*}(s_{*}(\mathsf{cons}(e'))), \text{~where }e':\mu X_{s}' \oplus Y_{e} \\
+  &
+    \begin{aligned}
+      =_{\oplus \eta}
+      \textbf{case}~e'~\{ & \\
+      & \textbf{in}_{X_{s}'} u \mapsto r_{*}(s_{*}(\mathsf{cons}(\textbf{in}_{X_{s}'} u))), \\
+      & \textbf{in}_{Y_{e}} v \mapsto r_{*}(s_{*}(\mathsf{cons}(\textbf{in}_{Y_{e}} v))) \\
+      \}
+    \end{aligned} \\
+  &
+    \begin{aligned}
+      =_{\mu \beta}
+      \textbf{case}~e'~\{ & \\
+      & \textbf{in}_{X_{s}'} u \mapsto r_{*}(\mathsf{cons}_{*}(\phi^{-1} u, p_{*})), \\
+      & \textbf{in}_{Y_{e}} v \mapsto r_{*}(\mathsf{cons}_{*}(()) \\
+      \}
+    \end{aligned} \\
+  &
+    \begin{aligned}
+      =_{\mu \beta}
+      \textbf{case}~e'~\{ & \\
+      & \textbf{in}_{X_{s}'} u \mapsto \mathsf{cons}(\textbf{in}_{X_{s}'}(\phi(\mathsf{cons}_{*}(\phi^{-1}u,p_{*})))), \\
+      & \textbf{in}_{Y_{e}} v \mapsto \mathsf{cons}(\textbf{in}_{Y_{e}}(()) \\
+      \}
+    \end{aligned} \\
+  & = \dots \\
+  & = e
+\end{align*}
+
+\steven{for the empty case above, can I make an $\eta$ equality for all terms of type $I$ to argue $\textbf{in}(v) = \textbf{in} (())$}
+
+\begin{align*}
+  s_{*}(r_{*}(p))
+  & =_{\mu \eta} s_{*}(r_{*}(\mathsf{cons}(p_{e}, p_{*}))), \text{~where } p_{e} : I, p_{*} : g \otimes g^{*} \\
+  & =_{\mu \beta} s_{*}()
+  & = p
+\end{align*}
 

### 2024-01-25 paper/paper.tex
DFA section stub + sync before meeting

@@
 Intuitively, this can be thought of like a directed graph with nodes in $Q$ with an edge $q \overset{c}{\to} q'$ whenever $q' \in \delta(q, c)$.
-Note that transitions in an NFA may be labelled with the empty string $\epsilon$.
+Note that transitions in an NFA may be labelled with the empty string $\epsilon$ --- such transitions are referred to as \emph{$\epsilon$-transitions}.
 
@@
 mutual\footnote{For convenience we refer to a mutual fixed point here, even though we
-  only take single fixed points as primitive. As in Beki\'c's theorem, we may decompose our mutual fixed points into a series of single fixed points.}
+  only take single fixed points as primitive. As in Beki\'c's theorem, we may decompose our mutual fixed points into a series of single fixed points. So this operation is admissible.}
 fixed point in the following manner:
@@
 \]
+
+\steven{Somewhere mention that $\epsilon$ corresponds to $I$, characters $c \in \Sigma$ correspond to grammars $c$, etc}
 
@@
 where $\sigma$ is the substitution that unrolls the mutually recursive definition one level by replacing each $X_{k}$ with the fixed point $\mu(X_{j} = A_{j}, \forall j).X_{k}$,
-\steven{I'm not sure the cons rule is right. TODO sanity check it with a small example}
 
@@
 
-\subsection{DFAs and NFAs}
-
-All DFAs are NFAs, so NFAs are at least as powerful as DFAs.
-It is perhaps surprising that DFAs are as powerful as NFAs too!
-We can internalize the classic powerset construction to transform an NFA into a DFA that recognizes the same language.
-Note however that this loses some information.
-Throughout this process we may lose data such as the ambiguity of the grammar.
-Effectively we are forgetting the structure of our proof terms.
-Under the ``grammars as types'' lens, we are erasing the structure underlying the inhabitants of the type to simply mere existence --- this is a form of propostional truncation.
-
-\steven{Worth spending some more time on the propositional truncation idea}
+\subsection{Brzozowski Derivatives and Deterministic Finite Automata}
+\steven{this section sounds awkward and will be adjusted in a later editing pass}
+A \emph{deterministic finite automaton} (DFA) is an NFA whose transitions are deterministic.
+Meaning at every state $q \in Q$ there is exactly one transition for every character $c \in \Sigma$ --- i.e.\ the size of $\delta(q, c)$ is 1 and $\delta(q, \epsilon)$ is empty.
+It is clear that all DFAs happen to be NFAs, but it is perhaps surprising that the class of DFAs is as powerful as the class of NFAs when measured by the languages they recognize!
+\cref{thm:thompson} shows that the class of languages recognized by NFAs is exactly the regular languages.
+As we will soon see, the class of languages recognizable by DFAs is also the regular languages!
+
+There are a couple famous constructions we might use to show this, such as the powerset construction \cite{?} or Brzozowski's algorithm \cite{?}.
+Just as with Thompson's construction, we shall internalize a version of the latter as a proof term in our formal calculus.
+However, note that the construction of a DFA is lossy in some sense.
+That is, throughout this process we lose the ability to form an isomorphism between parses of a DFA for $g$ and parses of $g$ itself.
+Rather than express an isomorphism (strong equivalence) as a proof term, we shall construct a weak equivalence.
+This means we shall derive entailments $g \vdash DFA(g)$ and $DFA(g) \vdash g$ that do not necessarily invert each other.
+Semantically, this notion of weak equivalence between grammars $g$ and $g'$ means that $g$ and $g'$ both recognize the same language.
+
+Recall that a DFA is given as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$ in the same manner as an NFA.\ When encoding a DFA as a mutually recursive grammar, we use nearly the same process as when we encoded an NFA, with one subtle change.
+When we encoded NFAs, we were able to ensure we had a single accepting state by connecting all of the accepting states $q_{j} \in F$ to a single accepting state $Y_{e}$ with $\epsilon$-transitions.
+However, $\epsilon$-transitions are not permitted in a DFA, as they introduce a source of nondeterminism.
+Therefore we need a scheme to encode which states are accepting.
+The change we need to make is pretty lightweight, we add the grammar $I$ as a summand to distinguish which states are accepting.
+
+\[
+  DFA ::= \mu \begin{pmatrix}
+        Y_{s} = \langle \hat{Y} \rangle \\
+        Y_{j} = \langle \hat{Y} \rangle, j \not \in \{s, e\} \\
+        \end{pmatrix}.Y_{s}
+\]
+where
+\[
+\hat{Y} ::= \hat{Y_{0}} ~|~ I \oplus \hat{Y_{0}}
+\]
+\[
+  \hat{Y_{0}} ::= c \otimes Y ~|~ Y ~|~ \hat{Y_{0}} \oplus \hat{Y_{0}}, Y \in \{Y_{s}, Y_{1}, \dots, Y_{N}\}
+\]
+
+A state $Y_{j}$ in a DFA is accepting if and only if $\hat{Y}_{j}$ contains $I$ as a summand.
+
+some fluff to lead into the thoerem
+
+\begin{theorem}
+  For every regular grammar $g$, there is a DFA $D$ that is weakly equivalent to $g$.
+\end{theorem}
+
+\begin{proof}
+  \steven{This is currently word salad}
+
+  DFA construction
+
+  $DFA(g) \vdash g$ : just fold over the defs in $DFA(g)$ and have every $Y_{w}$ eliminate to $w \lto g$.
+This is doable bc when eliminating $Y_{w'}$ you will hold terms of type $a \otimes (w \lto g)$.
+These $w \lto g$ terms might not be exactly what you want, but they are weakly equivalent to some derivative that you do want!
+How do you know that they're weakly equivalent?
+if $Y_{w'}$ transitions to $Y_{w}$ via character $a$, then we know that $a \lto (w \lto g)$ is weakly equivalent to $a \lto (w' \lto g)$ \emph{by construction of the DFA}.
+This argument entirely hinges on the method by which we defined the DFA
+
+  $g \vdash DFA(g)$
+  Idea is to us $\top \cong \text{Char}^{*}$ and $A^{*} \cong \Sigma(n) A^{\otimes n}$ and $A \cong \top \amp A$ to turn a parse into a pair of a parse and a finite string.
+And then this finite string can be walked through the DFA to read off the parse, and you can show its accepting.
+Details currently a little shaky, and encapsulated in a bad picture of a whiteboard.
+\steven{Gonna has this out with Max in meeting}
+\end{proof}
 
@@
 defined as a mutual fixed point.
+
+\steven{Forms for PDAs and Turing Machines go here}
 

### 2024-01-26 paper/paper.tex
section 2

@@
 \newcommand{\Bool}{\mathbf{Bool}}
+\newcommand{\nat}{\mathbb{N}}
 
@@
   \end{align*}
+  \caption{Kleene algebra axioms}
   \label{fig:axioms}
-  \caption{Kleene algebra axioms}
 \end{figure}
@@
 The addition operation can be used to define the partial order
-structure $a \leq b$ if $a + b = b$. In the theory of formal languages
-this order structure can be used to model language containment. In this
-section, we want to categorify the concept of Kleene algebra and
-build on top of it in order to define an abstract theory of parsing.
-We start by defining \emph{Kleene categories}.
+structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
+formal languages, this order structure can be used to model language
+containment. In this section, we want to categorify the concept of
+Kleene algebra and build on top of it in order to define an abstract
+theory of parsing.  We start by defining \emph{Kleene categories}.
 
@@
 
-An unexpected example comes from the theory of substructural logics.
+This example illustrates how restricted Kleene algebras are in terms
+of reasoning about languages. By only having at most one morphism
+between objects, there is not a lot of information they can convey. In
+this case, the only information you get is language containment. We
+will see throughout this paper how the extra degrees of freedom
+granted by having more morphisms give you more algebraic structure for
+reasoning about languages.
+
+For the next example, we see an unexpected connection with the theory
+of substructural logics.
 
@@
 
-Though Kleene categories are expressive enough to reason about concepts
-that are outside of reach of Kleene algebras, their simply-typed nature
-makes them not so expressive from a type theory point of view.
-
-Since we are interested in design a type theory for reasoning about various
-properties of parsers, we will assume that our Kleene category is defined
-internal to a model of linear depedendently typed theory, which leads us
-to the notion of models we are interested in:
+Though Kleene categories are expressive enough to reason about
+concepts that are outside of reach of Kleene algebras, their
+simply-typed nature makes them not so expressive from a type theory
+point of view. This is limiting because type theories are tested
+syntactic frameworks for manipulating complicated categories
+structures while avoiding some issues common in category theory, such
+as coherence issues.
+
+With this in mind, we want to design a dependent type theory that
+builds on top of Kleene categories with the goal of internalizing
+various reasoning principles for languages and their parsers. This
+leads us to the abstract notion of model we are interested in
+capturing with our type theory.
 
 \begin{definition}
-  A model is a locally Cartesian category with \pedro{add the linear logic connectives here} equipped with an internal Kleene category.
+  A model is a locally Cartesian category with two hierarchies of
+  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
+  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
+  $U_i$ to be closed under dependent products and sum,
+  $L_i$ to be closed under the Kleene category connectives, 
+  dependent products, left and right closed structures, with
+  a type constructor $G : L_i \to U_i$ and a linear dependent sum
+  going the other direction.
 \end{definition}
@@
 %
+    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
+    %
+    \and
+    %
     \inferrule{~}{\Gamma \vdash \top : L_i}
@@
 %
-    \inferrule{\Gamma \vdash A : L_i \\ \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
+    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
 %
@@
 \begin{align*}
-  c\tol c &\equiv I\\
-  c\tol d &\equiv 0\\
-  c\tol I &\equiv 0\\
-  c\tol 0 &\equiv 0\\
-  c\tol (A \otimes B) &\equiv (c\tol A) \otimes B + \nu(A) \otimes (c\tol B)\\
-  c\tol A^* &\equiv (c\tol A) \otimes A^*\\
-  c\tol (A + B) &\equiv (c\tol A) + (c\tol B)
+  c\lto c &\equiv I\\
+  c\lto d &\equiv 0\\
+  c\lto I &\equiv 0\\
+  c\lto 0 &\equiv 0\\
+  c\lto (A \otimes B) &\equiv (c\lto A) \otimes B + \nu(A) \otimes (c\lto B)\\
+  c\lto A^* &\equiv (c\lto A) \otimes A^*\\
+  c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
 \end{align*}

### 2024-01-26 paper/paper.tex
DFA first half

@@
 L$. Expressing Brzozowski's calculation of the derivative of a regular
-expression relies on additional axioms to the type theory.\max{axioms
-  to carry out Brzozowski's derivative calculations plz.}
-
+expression relies on additional axioms to the type theory, presented in \cref{fig:brzozowskideriv}.
+
+\begin{figure}
+\label{fig:brzozowskideriv}
 \begin{align*}
@@
   c\lto 0 &\equiv 0\\
-  c\lto (A \otimes B) &\equiv (c\lto A) \otimes B + \nu(A) \otimes (c\lto B)\\
-  c\lto A^* &\equiv (c\lto A) \otimes A^*\\
+  c\lto (A \otimes B) & \equiv (c\lto A) \otimes B + (A \amp I) \otimes (c\lto B)\\
+  c\lto A^* &\equiv (A \amp I)^{*} \otimes (c \lto A) \otimes A^* \\
   c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
 \end{align*}
-
-\pedro{The type family $\nu$ above is equal to $I$ if its argument
-  parses the empty string and zero otherwise. In order to define
-  it we should probably add an extra axiom, right?}
-\steven{$\nu$ can be derived using $\amp$, $\nu(A) = I \amp A$}
-
-\steven{In addition to Pedro's axioms:
-
-$c, c'$ characters. $g$ any grammar.
-
-\begin{gather}
-  I \lto c \cong c \\
-  c \lto (c' \lto g) \cong (c' \otimes c) \lto g
-\end{gather}
-
-For axiom $(1)$, Brzozowski has $I \lto g \cong g$ for all grammars $g$, but I don't think that holds for us because we want to preserve ambiguity.
-We could instead have something like $I \lto g \cong (g \amp I)^{*} \otimes g$.
-For our purposes the character presentation seems sufficient, as the derivative reasoning Brzozowski uses only takes derivatives with respect to strings.
-
-Axiom $(2)$ might be admissible, but I'm unsure.
-
-I think for Kleene star we need $c \lto g^{*} \cong (g \amp I)^{*} \otimes (c \lto g) \otimes g^{*}$.
-This should derivable utilizing the product axiom and $g^{*} \cong I \oplus (g \otimes g^{*})$, or is that sort of isomorphism opaque to us right now.
-
-To make the DFA proof go through I think we additionally need to axiomatically make the map $c \otimes (c \lto A) \vdash (c \otimes \top) \amp A$ to be an iso.
-Specifically, this term needs an iso $\phi : c \otimes (c \lto A) \to (c \otimes \top) \amp A$ defined below,
-
-\[
-p : c \otimes (c \lto A) \vdash \phi(p) = \letin {p_{1} \otimes p_{2}} p {(p_{1} \otimes \top, p_{1}^{\lto}p_{2})} : (c \otimes \top) \amp A
-\]
-
-but this last bit isn't just Brzozowski's definitions
-
-}
+\caption{Equality for Brzozowski Derivatives}
+\end{figure}
+
 Using the $G$ modality, we can internalize some of the external
@@
 For character $c \in \Sigma \cup \{\epsilon\}$ and state $q_{j} \in Q$, we add a summand --- as a coproduct\footnote{There is an admissible notion of coproduct making use of the primitive dependent pairing --- i.e.
-$A \oplus B := \LinPiTy b \Bool {\textbf{if}~b~A~B}$.
+$A \oplus B := \LinSigTy b \Bool {\textbf{if}~b~A~B}$.
 We will utilize this without much comment to represent the usual notion of alternation of grammars.} --- to the defintion of $Y_{j}$ for every state in the image of the transition function,
@@
 
-\steven{Somewhere mention that $\epsilon$ corresponds to $I$, characters $c \in \Sigma$ correspond to grammars $c$, etc}
-
 and each $Y \in \{ Y_{s}, Y_{e}, Y_{1}, \dots, Y_{N} \}$. Intuitively, we read each $Y_{j}$ as the label for a state, and the
 definitions of form $\hat{Y}$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$
-denotes an $\epsilon$-transition. A
+denotes an $\epsilon$-transition.
+Note that here each $c$ represents a grammar that is inhabited with only a trivial parse of the single character $c$; likwise, and $\epsilon$-transtion corresponds to a transition with label $I$ and $I \otimes Y = Y$ because $I$ is the unit for the operation $\otimes$.
+A
 disjunction of transitions in an NFA grammar corresponds to multiple outgoing edges of a given state in an automaton.
@@
 Moreover, we will show that this NFA is strongly equivalent to the original grammar.
-
-This construction
 
@@
     y_{e} : \simulsubst I {\gamma} \vdash p_{e} : I
-  }{\Delta \vdash s_{\oplus}(e) := \mathsf{mfold}(e) : g \oplus g'}
+  }{\Delta \vdash s_{\oplus}(e) := \mathsf{mfold}(y_{s}.p_{s},\dots,y_{e}.p_{e})(e) : g \oplus g'}
 \]
@@
     \Gamma ; x_{s} : \simulsubst {X_{s}}{\gamma} = X_{s}' \vdash \mathsf{cons}(\textbf{in}_{X_{s}'}(x_{s})) : \mu(\dots).Y_{s} = NFA(g \oplus g')  \\
-    \Gamma ; x_{j} : \simulsubst {X_{j}}{\gamma} = X_{j}' \vdash x_{j} : X_{j}', j \not \in \{s, e\} \\
-    \Gamma ; x_{e} : \simulsubst {X_{e}}{\gamma} = I \vdash x_{e} : I = X_{e}'
+    \Gamma ; x_{j} : \simulsubst {X_{j}}{\gamma} = \mu(X_{j} = \hat{X}_{j}, \forall j).X_{j}' \vdash x_{j} : \mu(X_{j} = \hat{X}_{j}, \forall j).X_{j}', j \not \in \{s, e\} \\
+    \Gamma ; x_{e} : \simulsubst {X_{e}}{\gamma} = I \vdash \mathsf{cons}(x_{e}) : I = \mu(\dots).Y_{e} = \mu(\dots).X_{e}'
   }
-  {\Gamma ; \Delta \vdash \mathsf{mfold}(u) : NFA(g \oplus g')}
+  {\Gamma ; \Delta \vdash \mathsf{mfold}(x_{s}.\mathsf{cons}(\textbf{in}_{X_{s}'}(x_{s}), x_{1}.x_{1}, \dots, x_{e}.x_{e}))(u) : NFA(g \oplus g')}
 \]
@@
 Therefore we need a scheme to encode which states are accepting.
-The change we need to make is pretty lightweight, we add the grammar $I$ as a summand to distinguish which states are accepting.
+The changes we need to make is pretty lightweight.
+First, we add the grammar $I$ as a summand to distinguish which states are accepting.
+Second, each state must have a transition for every character in the alphabet.
 
@@
 \[
-  \hat{Y_{0}} ::= c \otimes Y ~|~ Y ~|~ \hat{Y_{0}} \oplus \hat{Y_{0}}, Y \in \{Y_{s}, Y_{1}, \dots, Y_{N}\}
+  \hat{Y_{0}} ::= \bigoplus_{c : \Sigma} c \otimes Y, Y \in \{Y_{s}, Y_{1}, \dots, Y_{N}\}
 \]
@@
 
-some fluff to lead into the thoerem
+In the definition of a DFA, we make use of a sum ranging over all characters in the alphabet.
+Viewed as a machine, this is intened to have the effect of being able to run any arbitrary string input through the automaton to reach some well-defined, but not necessarily accepting, state.
+In our type theory, this would amount to the following entailment being derivable,
+
+\[
+p : \top \vdash \mathsf{run}(p) : \mu \begin{pmatrix}
+        Y_{s} = \langle \hat{Y} \rangle \\
+        Y_{j} = \langle \hat{Y} \rangle, j \not \in \{s, e\} \\
+        \end{pmatrix}.\bigoplus_{k} Y_{k}
+\]
+
+for some term $\mathsf{run}(p)$ that might depend on the structure of the DFA in question.
+That is, any input should be runnable through the DFA to return \emph{some} state.
+This entailment holds in the intended model, but a priori this does not hold in the type theory.
+Noting this point motivates taking $\top \cong (\bigoplus_{c} c)^{*}$ as an axiom.
+This syntactically reflects the semantic notion that parses correspond to finite strings over a fixed alphabet.
+Such an axiom is especially justified when crafting syntax to reason about parses of finite input strings; further, we conjecture that without this axiom we may find models of the type theory that correspond to parses of infinite sequences of characters.
+In this paper, we shall focus on the expressivity of the syntax to reason about parses of strings, and thus we assume $\top \cong (\bigoplus_{c} c)^{*}$.
+
+Additionally, we assume that the derivation from $c \otimes (c \lto A)$ to $(c \otimes \top) \amp A$ is an isomorphism. Specifically, the term $\phi$ defined below is assumed to have an inverse,
+
+\[
+p : c \otimes (c \lto A) \vdash \phi(p) = \letin {p_{1} \otimes p_{2}} p {(p_{1} \otimes \top, p_{1}^{\lto}p_{2})} : (c \otimes \top) \amp A
+\]
+
+Axiomatically adding an inverse for $\phi$ doesn't greatly affect our semantics.
+In fact, it describes a property that is simply true in the intended string model: ``if a parse of character $c$ and then a parse of any word completes to a parse of $A$, then that word was actually in fact of type $c \lto A$''.
+This axiom, the one desribed in the paragraph above, and the ones given in \cref{fig:brzozowskideriv} are not meant to anything particularly deep.
+They are a means of reflecting the intended semantic properties in our syntax.
+
+Utilizing the reflected properties we can derive the following isomorphism,
+
+\begin{align*}
+  A
+  & \cong \top \amp A \\
+  & \cong (\bigoplus_{c} c)^{*} \amp A \\
+  & \cong (I \oplus \bigoplus_{c} (c \otimes \top)) \amp A \\
+  & \cong (I \amp A) \oplus \bigoplus_{c} ((c \otimes \top) \amp A) \\
+  & \cong (I \amp A) \oplus \bigoplus_{c} (c \otimes (c \lto A))
+\end{align*}
+
+Which effectively states that parses of a grammar $A$ are either empty or begin with some character from our fixed alphabet, a fact used in our proof of \cref{thm:dfa}.
 
 \begin{theorem}
+  \label{thm:dfa}
   For every regular grammar $g$, there is a DFA $D$ that is weakly equivalent to $g$.
@@
 \begin{proof}
-  \steven{This is currently word salad}
-
-  DFA construction
-
-  $DFA(g) \vdash g$ : just fold over the defs in $DFA(g)$ and have every $Y_{w}$ eliminate to $w \lto g$.
-This is doable bc when eliminating $Y_{w'}$ you will hold terms of type $a \otimes (w \lto g)$.
-These $w \lto g$ terms might not be exactly what you want, but they are weakly equivalent to some derivative that you do want!
-How do you know that they're weakly equivalent?
-if $Y_{w'}$ transitions to $Y_{w}$ via character $a$, then we know that $a \lto (w \lto g)$ is weakly equivalent to $a \lto (w' \lto g)$ \emph{by construction of the DFA}.
-This argument entirely hinges on the method by which we defined the DFA
-
-  $g \vdash DFA(g)$
-  Idea is to us $\top \cong \text{Char}^{*}$ and $A^{*} \cong \Sigma(n) A^{\otimes n}$ and $A \cong \top \amp A$ to turn a parse into a pair of a parse and a finite string.
-And then this finite string can be walked through the DFA to read off the parse, and you can show its accepting.
-Details currently a little shaky, and encapsulated in a bad picture of a whiteboard.
-\steven{Gonna has this out with Max in meeting}
+  First, note that regular grammars have finitely many derivatives, up to weak equivalence \cite{?}.
+That is, we can isolate a finite set of words $w_{1},\dots,w_{n}$ such that $w \lto g$ is weakly equivalent to $w_{j} \to g$ for some $g$.
+These $w_{j}$ may then serve as representatives for their weak equivalences classes.
+
+  The high-level idea underlying Brzozowski's DFA construction algorithm is that the states of the DFA are the derivative equivalence classes, and the transitions are formed by taking derivatives.
+If the derivative of $w \lto g$ with respect to a character $c$ is weakly equivalent to $w' \lto g$ then there is a transition from the state representing $w \lto g$ to the state representing $w' \lto g$.
+These pieces of data form the transition table form our DFA.\ Lastly, we make the state representing $w \lto g$ accepting if the type $I \amp (w \lto g)$ is inhabited.
+If it is inhabited the state for $w \lto g$ will have an $I$ as a summand, otherwise it will not.
+To represent this acceptance formally in our construction, use the function symbol $\mathsf{ac}(w) : 0 \oplus I$ to be $I$ if $w \lto g$ is inhabited and 0 otherwise.
+Similarly, $\mathsf{ac}(I)$ will represent the acceptance of the start state.
+
+  We may precompute all this info and then construct the DFA by introducing states labeled $Y_{j}$ for each representative $w \in \{w_{1}, \dots, w_{n}\}$.
+Letting $j$ range in $\{s, 1, \dots, n\}$ and taking $w_{s} = I$ we can describe all states of the DFA with a similar structure.
+
+  \[
+  DFA(g) =
+  \mu \begin{pmatrix}
+        Y_{j} = \mathsf{ac}(w_{j}) \oplus \bigoplus_{c} c \otimes Y_{n_{j,c}} \\
+  \end{pmatrix}.Y_{s}
+\]
+
+Note that by construction the target state of each transition, $Y_{n_{j,c}}$, is such that $c \lto (w_{j} \lto g)$ is weakly equivalent to $w_{n_{j, c}} \lto g$.
+
+To turn a term of type $DFA(g)$ into a term of type $g$, we simply fold over the definitions of $DFA(g)$. Again, the structure of such a derivation,
+
+\[
+  \inferrule
+  {
+    \Gamma ; \Delta \vdash e : DFA(g) = \mu(Y_{j} = \hat{Y}_{j}).Y_{s} \\
+    \Gamma ; x_{s} : \simulsubst {\mathsf{ac}(w_{j}) \oplus \bigoplus_{c} c \otimes Y_{n_{s,c}}} {\gamma} = \mathsf{ac}(w_{j}) \oplus \bigoplus_{c} c \otimes B_{n_{s,c}} \vdash e_{s} : B_{s} \\
+    \Gamma ;x_{j} : \simulsubst {\mathsf{ac}(w_{j}) \oplus \bigoplus_{c} c \otimes Y_{n_{j,c}}}{\gamma} = \mathsf{ac}(w_{j}) \oplus \bigoplus_{c} c \otimes B_{n_{j,c}} \vdash e_{j} : B_{j}
+  }
+  {\Gamma ; \Delta \vdash \mathsf{mfold}(x_{s}.e_{s}, x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{s}}
+\]
+
+When instantiating the above rule, we make each $B_{k} = w_{k} \lto g$.
+In this manner, $B_{s} = I \lto g = g$ which is our desired output type.
+Let us examine one of the entailments to see how this is achievable.
+From a term $x_{j} : \mathsf{ac}(w_{j}) \oplus \bigoplus_{c} c \otimes (w_{n_{j,c}} \lto g)$, we need to produce some $e_{j} : w_{j} \lto g$.
+Which we may do in the following manner,
+
+\begin{align*}
+    \begin{aligned}
+  e_{j} =
+    \phi(\textbf{case}~x_{j}~\{&  \textbf{in}_{\mathsf{ac}} u \mapsto \mathbf{in}_{I \amp (w_{j} \to g)} u, \\
+  & \textbf{in}_{c \otimes (w_{n_{j,c}} \lto g)} (v_{c} \otimes d_{j,c}) \mapsto \mathbf{in}_{c \otimes (w_{j} \lto g)}(v_{c} \otimes \alpha(d_{j,c}))
+  \})
+    \end{aligned}
+\end{align*}
+
+Where $\phi$ witnesses the isomorphism $g \cong (I \amp g) \oplus \bigoplus_{c} (c \otimes (c \lto g))$, and $\alpha$ is a function that acts a lookup table for replacing $d_{j,c} : w_{n_{j,c}} \lto g$ with an appropriate member of its weak equivalence class.
+In particular, when deriving $w_{j} \lto g$ we want $\alpha(d_{j,c}) : (w_{j} \otimes c) \lto g$.
+Finding such an element is indeed possible \emph{by construction of our DFA}.
+Recall that we added a transition between $Y_{j}$ and $Y_{n_{j,c}}$ exactly when $c \lto (w_{j} \lto g) = (w_{j} \otimes c) \lto g$ was weakly equivalent to $w_{n_{j,c}} \lto g$.
+That is, the multifold constructed above is well-defined, and will ultimately return a term of type $B_{s} = I \lto g = g$ so we do, in fact, have a derivation from $DFA(g)$ to $g$.
+
+Now we wish to build a derivation from $g$ to $DFA(g)$.
+\steven{fill in}
 \end{proof}
@@
 
-  \steven{There a weird spacing issues between these align envs. Maybe wrap these in a figure env so that spacing is better}
-
 \begin{align*}
@@
     \begin{aligned}
-      =_{\oplus \eta} r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{case}~e'~\{ & \\
-    \textbf{in}_{\mu X_{s}'} u \mapsto \textbf{in}_{\mu X_{s}'} u, & \\
-    \textbf{in}_{\mu Z_{s}'} v \mapsto \textbf{in}_{\mu Z_{s}'} v
+      =_{\oplus \eta} r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{case}~e'~\{
+    & \textbf{in}_{\mu X_{s}'} u \mapsto \textbf{in}_{\mu X_{s}'} u, \\
+    & \textbf{in}_{\mu Z_{s}'} v \mapsto \textbf{in}_{\mu Z_{s}'} v
     \})))
@@
     =_{\oplus \eta}
-    \textbf{case}~e'~\{ & \\
+    \textbf{case}~e'~\{
     & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{in}_{\mu X_{s}'} u))), \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v))) \\
+    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v)))
     \}))
@@
     =
-    \textbf{case}~e'~\{ & \\
+    \textbf{case}~e'~\{
     & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(\mathsf{mfold}(\mathsf{cons}(\textbf{in}_{\mu X_{s}'} u))), \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(\mathsf{mfold}(\mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v))) \\
+    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(\mathsf{mfold}(\mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v)))
     \}
@@
     =_{\mu \beta}
-    \textbf{case}~e'~\{ & \\
+    \textbf{case}~e'~\{
     & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(\textbf{in}_{g} \phi^{-1}(\mathsf{cons}_{\mu(X_{i}=\hat{X}_{i}).X_{s}}u')), \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(\textbf{in}_{g'} \psi^{-1}(\mathsf{cons}_{\mu(Z_{i}=\hat{Z}_{i}).Z_{s}}v')) \\
+    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(\textbf{in}_{g'} \psi^{-1}(\mathsf{cons}_{\mu(Z_{i}=\hat{Z}_{i}).Z_{s}}v'))
       \} \\
@@
     =_{\oplus \beta}
-    \textbf{case}~e'~\{ & \\
+    \textbf{case}~e'~\{
     & \textbf{in}_{\mu X_{s}'} u \mapsto \alpha \phi \phi^{-1}(\mathsf{cons}(u')), \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto \beta \psi \psi^{-1}(\mathsf{cons}(v')) \\
+    & \textbf{in}_{\mu Z_{s}'} v \mapsto \beta \psi \psi^{-1}(\mathsf{cons}(v'))
       \}
@@
     =
-    \textbf{case}~e'~\{ & \\
+    \textbf{case}~e'~\{
     & \textbf{in}_{\mu X_{s}'} u \mapsto \alpha (\mathsf{cons}(u')), \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto \beta (\mathsf{cons}(v')) \\
+    & \textbf{in}_{\mu Z_{s}'} v \mapsto \beta (\mathsf{cons}(v'))
       \}
@@
     =_{\mu \beta}
-    \textbf{case}~e'~\{ & \\
+    \textbf{case}~e'~\{
     & \textbf{in}_{\mu X_{s}'} u \mapsto \mathsf{cons}(\textbf{in}_{\mu X_{s}'} u) , \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto \mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v) \\
+    & \textbf{in}_{\mu Z_{s}'} v \mapsto \mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v)
       \}
@@
 
-\steven{This first line is a weird alignment fix, touch up}
 \begin{align*}
@@
   &
-  = s_{\oplus}(r_{\oplus}(p)) \\
-   & \begin{aligned}
     =_{\oplus \eta}
       \textbf{case}~p~\{
-    \textbf{in}_{g}p_{1} \mapsto s_{\oplus}(r_{\oplus}(\textbf{in}_{g}p_{1})), \\
-    \textbf{in}_{g'}p_{2} \mapsto s_{\oplus}(r_{\oplus}(\textbf{in}_{g'}p_{2})) \}
+    \begin{aligned} \\
+    & \textbf{in}_{g}p_{1} \mapsto s_{\oplus}(r_{\oplus}(\textbf{in}_{g}p_{1})), \\
+    & \textbf{in}_{g'}p_{2} \mapsto s_{\oplus}(r_{\oplus}(\textbf{in}_{g'}p_{2})) \}
     \end{aligned} \\
@@
       \textbf{case}~p~\{
-    \textbf{in}_{g}p_{1} \mapsto s_{\oplus}(\alpha \phi p_{1}), \\
-    \textbf{in}_{g'}p_{2} \mapsto s_{\oplus}(\beta \psi p_{2}) \}
+    & \textbf{in}_{g}p_{1} \mapsto s_{\oplus}(\alpha \phi p_{1}), \\
+    & \textbf{in}_{g'}p_{2} \mapsto s_{\oplus}(\beta \psi p_{2}) \}
     \end{aligned} \\
@@
       \textbf{case}~p~\{
-    \textbf{in}_{g}p_{1} \mapsto \mathsf{mfold}(\alpha \phi p_{1}), \\
-    \textbf{in}_{g'}p_{2} \mapsto \mathsf{mfold}(\beta \psi p_{2}) \}
+    & \textbf{in}_{g}p_{1} \mapsto \mathsf{mfold}(\alpha \phi p_{1}), \\
+    & \textbf{in}_{g'}p_{2} \mapsto \mathsf{mfold}(\beta \psi p_{2}) \}
     \end{aligned} \\
@@
       \textbf{case}~p~\{
-    \textbf{in}_{g}p_{1} \mapsto \textbf{in}_{g} (\phi^{-1} \phi p_{1}), \\
-    \textbf{in}_{g'}p_{2} \mapsto \textbf{in}_{g'}(\psi^{-1} \psi p_{2}) \}
+    & \textbf{in}_{g}p_{1} \mapsto \textbf{in}_{g} (\phi^{-1} \phi p_{1}), \\
+    & \textbf{in}_{g'}p_{2} \mapsto \textbf{in}_{g'}(\psi^{-1} \psi p_{2}) \}
     \end{aligned} \\
@@
       \textbf{case}~p~\{
-    \textbf{in}_{g}p_{1} \mapsto \textbf{in}_{g} (p_{1}), \\
-    \textbf{in}_{g'}p_{2} \mapsto \textbf{in}_{g'}(p_{2}) \}
+    & \textbf{in}_{g}p_{1} \mapsto \textbf{in}_{g} (p_{1}), \\
+    & \textbf{in}_{g'}p_{2} \mapsto \textbf{in}_{g'}(p_{2}) \}
     \end{aligned} \\
@@
 \end{align*}
+
+Thus $s_{\oplus}$ and $r_{\oplus}$ form the isomorphism $NFA(g \oplus g') \cong g \oplus g'$.
 
@@
     Y_{e} = I
-  \end{pmatrix}
+  \end{pmatrix}.Y_{s}
 \]
@@
   \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = (\mu (X_{i}=\hat{X}_{i}).X_{j}) \otimes g'$ and $p_{x, j} = x_{j}' : (\mu (X_{i}=\hat{X}_{i}).X_{j}) \otimes g'$
-  \item $x_{e}'$ --- $\simulsubst {Z_{s}'}{\gamma} = g'$ and $p_{x,e} = x_{e}' : g'$
+  \item $x_{e}'$ --- $\simulsubst {Z_{s}'}{\gamma} = g' = I \otimes g'$ and $p_{x,e} = \letin {mt \otimes p_{g'}} {\mathsf{cons}(mt) \otimes p_{g'}} x_{e}' : (\mu (X_{i}=\hat{X}_{i}).X_{e}) \otimes g'$
   \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = \mu (Z_{i} = \hat{Z}_{i}).Z_{s} = NFA(g')$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
@@
 
-The map $r_{\otimes}$ in the other direction is given as,
+The map $r_{\otimes}$ in the other direction is given again in two steps. Like the disjunctive case, we give maps $\alpha$ and $\beta$. First, $\beta$,
+
+\[
+  \inferrule
+  {
+    \Gamma; \Delta \vdash e : NFA(g') = \mu(Z_{j} = \hat{Z}_{j}).Z_{s} \\
+    \Gamma; z_{s} : \simulsubst {Z_{s}} \gamma = \mu Z_{s}' \vdash z_{s} : \mu Z_{s}' \\
+    \Gamma; z_{j} : \simulsubst {Z_{j}} \gamma = \mu Z_{j}' \vdash z_{j} : \mu Z_{j}' \\
+    \Gamma; z_{e} : \simulsubst {Z_{e}} \gamma = I \vdash \mathsf{cons}_{Z_{e}'}(\mathsf{cons}_{Y_{e}}(z_{e})) : \mu Z_{e}'
+  }
+  {\Gamma ; \Delta \vdash \beta(e) := \mathsf{mfold}(z_{s}.z_{s},\dots,z_{e}.\mathsf{cons}_{Z_{e}'}(\mathsf{cons}_{Y_{e}}(z_{e})))(e) : \mu Z_{s}
+  }
+\]
+
+Note that even the rule $\alpha$ shown below doesn't strictly follow the syntax for an $\mathsf{mfold}$, it can actually be shown to be admissible\footnote{To do so, one could define a new automata-like fixed point where the end state of NFA(g) is replaced with a term of type $\mu Z_{s}$}.
+This presentation is more compact and to the point,
+
+\[
+  \inferrule
+  {
+    \Gamma; \Delta \vdash e : NFA(g) \otimes \mu Z_{s}' = \mu(X_{j} = \hat{X}_{j}).X_{s} \otimes \mu Z_{s}' \\
+    \Gamma; x_{s} : \simulsubst {X_{s}} \gamma = \mu X_{s}' \vdash x_{s} : \mu X_{s}' \\
+    \Gamma; x_{j} : \simulsubst {X_{j}} \gamma = \mu X_{j}' \vdash x_{j} : \mu X_{j}' \\
+    \Gamma; x_{e} : \simulsubst {X_{e}} \gamma = I \otimes \mu Z_{s}' = \mu Z_{s}' \vdash \mathsf{cons}(x_{e}) : \mu X_{e}'
+  }
+  {
+    \Gamma ; \Delta \vdash \alpha(e) := \mathsf{mfold}(x_{s}.x_{s},x_{1}.x_{1},\dots,x_{e}.\mathsf{cons}(x_{e}))(e) : \mu X_{s}'
+  }
+\]
+
+We then leverage these $\alpha$ and $\beta$ to build $r_{\otimes}$,
 
@@
     p : g \otimes g' \vdash p : g \otimes g' \\
-    p_{1} : g, p_{2} : g' \vdash \mathsf{cons}(\phi(p_{1}), \psi(p_{2})) : NFA(g \otimes g')}
-  {p : g \otimes g' \vdash \letin {p_{1} \otimes p_{2}} {p} {\mathsf{cons}(\phi(p_{1}), \psi(p_{2}))} : NFA(g \otimes g')}
-\]
-
-We show that $r_{\otimes}$ and $s_{\otimes}$ form an isomorhic pair,
-\steven{This proof doesn't exactly work}
+    p_{1} : g, p_{2} : g' \vdash \mathsf{cons}(\alpha(\phi(p_{1}) \otimes \beta(\psi(p_{2})))) : NFA(g \otimes g')}
+  {p : g \otimes g' \vdash \letin {p_{1} \otimes p_{2}} {p} {\mathsf{cons}(\alpha(\phi(p_{1}) \otimes \beta(\psi(p_{2}))))} : NFA(g \otimes g')}
+\]
+
+We now show that $s_{\otimes}$ and $r_{\otimes}$ form an isomorphism.
+
+\steven{This proof is a little sketchy.
+I'm not sure how to best write that the terms with $m$ and $n$ arise via functorial action by the term in the third equality.
+Moreover, I'm not sure that this third equality term is the right one}
 
@@
 r_{\otimes}(s_{\otimes}(e))
-  & =_{\mu \eta} r_{\otimes}(s_{\otimes}(\mathsf{cons}_{\mu X_{s}'}(e'))) \\
-  & = r_{\otimes}(\mathsf{mfold}(\mathsf{cons}_{\mu X_{s}'}(e'))) \\
-  & =_{\mu \beta} r_{\otimes}(\phi^{-1}(u) \otimes \psi^{-1}(v)) \\
-  & =_{\mu \beta} \mathsf{cons}(\phi \phi^{-1} u, \psi \psi^{-1} v) \\
+  & =_{\mu \eta} r_{\otimes}(s_{\otimes}(\mathsf{cons}(e'))) \\
+  & = r_{\otimes}(\mathsf{mfold}(\mathsf{cons}(\mathsf{mfold}(y_{s}.y_{s},\dots,y_{e}.y_{e})(e)))) \\
+  &
+    \begin{aligned}
+    =_{\mu \beta}
+    r_{\otimes} (
+    & (\letin {u \otimes v} {x_{s}'} {\phi^{-1}(u) \otimes v}) \\
+    & (\mathsf{mfold}(y_{s}.y_{s},\dots,y_{e}.y_{e})(e)))
+    \end{aligned}
+     \\
+  & = r_{\otimes}(\phi^{-1}(m) \otimes \psi^{-1}(n)) \\
+  & = \mathsf{cons}(\alpha(\phi \phi^{-1} m \otimes \beta(\psi \psi^{-1} n))) \\
+  & = \mathsf{cons}(\alpha(m \otimes \beta(n))) \\
+  & = \mathsf{cons}(\mathsf{mfold}(x_{s}.x_{s}, \dots, x_{e} . \mathsf{cons}(x_{e}))(m \otimes \beta (n))) \\
+  & = \mathsf{cons}(\mathsf{mfold}(y_{s}.y_{s}, \dots, y_{e} . y_{e})(e)) \\
   & =_{\mu \eta} e
@@
   & =_{\otimes \eta} s_{\otimes}(r_{\otimes}(p_{g} \otimes p_{g'})) \\
-  & = s_{\otimes}(\mathsf{cons}(\phi p_{g}, \psi p_{g'})) \\
+  & = s_{\otimes}(\mathsf{cons}(\alpha(\phi p_{g} \otimes \beta(\psi p_{g'}))) \\
   & =_{\mu \beta} \phi^{-1}\phi p_{g} \otimes \psi^{-1} \psi p_{g'} \\
@@
 \end{align*}
+
+Thus $s_{\otimes}$ and $r_{\otimes}$ form the isomorphism $NFA(g \otimes g') \cong g \otimes g'$.
 

### 2024-01-26 paper/paper.tex
PDA defintion

@@
 the inuitionistic context. When viewed as a machine, this refers to the class of automata definable without access to
-any memory --- i.e.\ finite state machines.
+a stack/tape of memory --- i.e.\ finite state machines.
 
@@
 
-\steven{Need to add argument that justifies why the $X_{j}$s are allowed to be eliminated via the identity.
-A priori they could be interfered with by taking an $X_{s}$ that needs to be eliminated.
-I think it hinges on a syntactic restriction where there are no backedges to $X_{s}$ in the NFA.Even if there were, I think I can get away with a case distinction entailment instead of the one written}
+\steven{Question raised by Pranav: why doesn't the $X_{s}'$ being eliminated to $g$ screw up what the other $X_{j}'$s get eliminated too.
+Need to add an argument that justifies why the $X_{j}$s are allowed to be eliminated via the identity.
+A priori they could be interfered with by taking in an $X_{s}'$ that needs to be eliminated.
+
+I think the crux of this arugment hinges on a syntactic restriction where there are no backedges to $X_{s}$ in the NFA.\ This is true for the $\otimes$, $\oplus$ cases, unclear in the Kleene star case.
+Even if there were one, we might be able to get away with a case distinction entailment instead of the one written.}
 
@@
 \subsection{Brzozowski Derivatives and Deterministic Finite Automata}
-\steven{this section sounds awkward and will be adjusted in a later editing pass}
 A \emph{deterministic finite automaton} (DFA) is an NFA whose transitions are deterministic.
@@
 In our type theory, this would amount to the following entailment being derivable,
+
+\steven{This might be the wrong entailment, and I may instead want a propositional truncation. This distinction feels related to the hang ups in the second half of the DFA proof}
 
@@
 
-Now we wish to build a derivation from $g$ to $DFA(g)$.
-\steven{fill in}
+Now we wish to build a derivation from $g$ to $DFA(g)$. TODO proof.
+\steven{
+  I think there is a way to do this via extracting some string related to a parse.
+i.e.
+$p : g \vdash M : (\LinSigTy {n} {\mathbb{N}} {\text{Char}^{n}} \amp g)$.
+However, Max said that this style of argument --- or perhaps just pattern matching over the sigma type --- requires full bunched contexts.
+I don't totally see why.
+
+  The hope is to then morally run the DFA on the string, and you can show that it must be accepting by some of the properties used to construct the DFA.
+
+  An alternative to the above approach is to reflect the fact that grammars are weakly equivalent if they recognize all the same strings. Something like,
+
+  \[
+    \inferrule
+    {\Gamma \vdash M : \Pi_{w} G(w \lto A) \to G(w \to B)}
+    {\Gamma \vdash Sem(M) : G(A \lto B)}
+  \]
+
+  You then instantiate this rule for $A = g$, $B = DFA(g)$.
+This gives you access to the same kind of string I am trying to access above and you make the same kind of argument.
+I'm too tired to write the stub of that argument now, and I'm a little shaky on it, but that will be a goal for next week.
+}
 \end{proof}
@@
 We can generalize to more powerful notions of automata by utilizing
-dependency of linear types on non-linear types. Let $S$ be a finite
+dependency of linear types on non-linear types.
+In this section, we shall explore how we can perform these encodings for pushdown automata and Turing machines.
+In this paper, we don't provide any proofs of their equivalence to context-free or recursively-enumerable languages; however, to motivate this type theory as a natural and general syntax for reasoning about both languages and machines, we do show how to internalize these notions of automata.
+
+\subsubsection{Pushdown Automata}
+Let us recall that a nondeterministic \emph{pushdown automaton} (PDA) is a type of automaton that utilizes a stack of memory.
+Similarly to NFAs, they are often represented as 7-tuples $(Q, \Sigma, \Gamma, \delta, q_{0}, Z, F)$.
+Here $Q, \Sigma, q_{0},$ and $F$ all have the exact same meanings as with NFAs.
+Additionally PDAs now have,
+
+\begin{itemize}
+\item $\Gamma$ a fixed, finite stack alphabet. We interpret ``the stack'' as words over $\Gamma$
+\item $Z$ an initial stack symbol
+\item $\delta : Q \times (\Sigma \cup \{\epsilon\}) \times \Gamma \to Q \times \Gamma^{*}$ a transition function that consumes a state, input character, and stack character then returns a state and new stack
+\end{itemize}
+
+Just as with NFAs, we encode PDAs via a particular mutual fixed point type, except now we let these types make use of non-linear types. In particular, let $S$ be a finite
 non-linear type and define $\textrm{List}\,S = \mu x. 1 + S \times
@@
 
-\steven{Forms for PDAs and Turing Machines go here}
-
-Want to put PDA, CFGs, CSGs here. Basically, use this section to highlight the usefulness of the intuitionistic part of the contexts
+Like with NFAs, we introduce a label $Y_{j}$ for each $q_{j} \in Q$ but now the right hand side of each of their definition is a dependent type that takes into account the stack.
+We again say there is a single accepting state.
+The initial stacks $Z$ corresponds to the empty list $\mathsf{nil}$.
+Additionally, without loss of generality we say that every transition consumes the head of the stack --- as if a transition did not consume anything, you could pop and replace the head with the same element.
+
+Consider a transition in the PDA between states $Y_{i}$ and $Y_{j}$ with current stack $\omega_{0} \in \Gamma^{*}$ that consumes stack character $a$, consumes an input character $c$, and then pushes $\omega \in \Gamma^{*}$ onto the stack.
+We translate this into the definition of $Y_{i}$.
+$\omega_{0}$ will correspond to some current list $L$, and the production on the right hand side of $Y_{i}$ will take the form of a (partial) case distinction on the head of $L$.
+The case corresponding to $a$ will be a dependent type that encodes the pushing of $\omega$ onto the stack as appending some list $L_{0}$ to the tail of $L$.
+That is, the definition of $Y_{i}$ looks like,
+
+\[
+Y_{i} = \LinSigTy {L} {\mathrm{List}\,S} {\textbf{case}~L~\{(a, tl) \mapsto \LinSigTy {\mathsf{append}(L_{0}, tl)} {\mathrm{List}\,S} {c \otimes Y_{j}}, \dots\}}
+\]
+
+In total,
+
+\steven{idk why I'm giving these as meta-level grammars.
+It doesn't seem like an illuminating way to share the form for PDAs, but I can't think of anything cleaner.
+This can definitely be put more cleanly}
+
+\[
+  PDA ::= \mu
+  \begin{pmatrix}
+    Y_{s} = \langle \hat{P}_{s} \rangle \\
+    Y_{j} = \langle \hat{P} \rangle \\
+    Y_{e} = \LinSigTy {L} {\textrm{List}\,S} {I}
+  \end{pmatrix}.Y_{s}
+\]
+
+\[
+\hat{P} ::= \LinSigTy{L}{\textrm{List}\,S}{\textbf{case}~L~\{ \langle cases \rangle \}}
+\]
+\[
+\hat{P}_{s} ::= \LinSigTy{\mathsf{nil}}{\textrm{List}\,S}{\textbf{case}~L~\{ \langle cases \rangle \}}
+\]
+\[
+cases ::= case ~|~ cases, case
+\]
+\[
+case ::= (z, tl) \mapsto T_{\oplus}
+\]
+\[
+  T_{\oplus} ::= T ~|~ T \oplus T_{\oplus}
+\]
+\[
+T ::= \LinSigTy{L'}{\textrm{List}\,S}{c \otimes Y}
+\]
+
+\subsubsection{Turing Machines}
+We can encode a Turing machine nearly identically to how we encoded PDAs.
+Instead of having our encoding depend on a non-linear type of lists over $S$, we have it depend on a pair of type $\mathrm{List}\,S \times \mathrm{List}\,S$.
+As mentioned earlier \steven{Probably copy that earlier bit from the beginning of the section here and reword}, we can simulate the infinite tape of a Turing machine with two stacks.
+
+\steven{I'm satisfied giving this high-level intuition and not dumping a syntactic mess for the audience, but as always its useful to have explicit examples because the claim is that the syntax is natural and easy to use.}
 
@@
 To see that $s_{*}$ and $r_{*}$ for an isomorphism,
-\steven{This proof also needs more work. Esp after 4th equality}
-\steven{Think I need to thread through a case distinction on what the $g^{*}$ term is. Putting a pin in this for now to type up the DFA stuff before meeting}
+\steven{This proof also needs more work.
+Esp after 4th equality.
+Putting a pin in both directions of the Kleene star equality proof until I fully understand the issue with the $\otimes$ case.
+Each of these stems from an ignorance on $\mu$-type $\beta$ and $\eta$ rules}
 \begin{align*}
@@
 
-\steven{for the empty case above, can I make an $\eta$ equality for all terms of type $I$ to argue $\textbf{in}(v) = \textbf{in} (())$}
-
 \begin{align*}

### 2024-03-18 POPL/paper.tex
popl/prelim paper stub

--- 2024-01-26 paper/paper.tex
@@
-\documentclass[sigconf,anonymous,review,screen]{acmart}
+% -*- fill-column: 80; -*-
+\documentclass[sigconf,review,screen]{acmart}
 \usepackage{mathpartir}
@@
 \usepackage{stmaryrd}
+
+\usepackage{todonotes}
+\newcommand{\todoin}[1]{\todo[inline]{TODO:\@ #1}}
 
@@
 \begin{abstract}
-  We propose a semantic framework for the study of formal
-  grammars. First, we provide a syntax-independent notion of formal
-  grammar as a function from strings to sets, generalizing the
-  familiar notion of a language as a set of strings. The set of such
-  grammars naturally form a very rich category whose notion of
-  isomorphism corresponds to strong equivalence of grammars, and for
-  which many language-theoretic constructs can be defined using
-  universal properties.
+  We propose a semantic framework for the study of formal grammars. First, we
+  provide a syntax-independent notion of formal grammar as a function from
+  strings to sets of parse trees, generalizing the familiar notion of a language
+  as a set of strings. This point-of-view naturally endows a grammar with the
+  structure of a presheaf, which allows many language-theoretic constructs to be
+  defined via universal properties.
 
-  Based on these category-theoretic constructions we propose a new
-  syntactic formalism for formal grammars: a version of dependent
-  linear-non-linear type theory in which the tensor product is not
-  commutative. The linear types can be interpreted as grammars, and
-  linear terms as parse transformers. The non-dependent fragment of
-  this type theory is already enough to express regular/context-free
-  expressions as well as finite automata, with the isomorphism between
-  regular languages and automata definable as linear terms. This
-  generalizes the language theoretic formalisms such as Kleene
-  algebras to incorporate ambiguity. By incorporating a dependency of
-  linear types on non-linear types, we can express richer grammar
-  formalisms such as indexed grammars, pushdown automata and Turing
-  machines. This provides a new form of ``logical characterization''
-  of grammar classes based on a substructural logic rather than
-  ordinary first-order logic. Further, the type theory can not only
+  With these categorical constructions in mind, we propose a new syntactic
+  formalism for formal grammars: a version of dependent linear-non-linear type
+  theory where the tensor product is non-commutative. The linear types in this
+  theory are interpreted as grammars --- and linear terms as parse transformers.
+  The non-dependent fragment of this type theory is already expressive enough to
+  capture regular expressions, context-free expressions, and finite automata.
+  Moreover, the isomorphism between regular languages and finite automata is
+  definable as a proof term in our formalism. The
+  non-dependent fragment of this type theory is expressive enough to capture
+  both regular and context-free expressions, as well as finite automata already
+  regular expressions, context-free expressions, and finite automata. Moreover,
+  the is definable as a proof term in this formalism. Further, through use of
+  dependency of linear types on non-linear types, we can express richer grammar
+  formalisms such as indexed grammars, pushdown automata, and Turing machines.
+
+  Classically, many results in both language and automata theory are proved
+  informally with respect to some ambient foundation --- e.g. Kleene algebra, grammars,
+  parsers, semantic actions, etc. The type theory provided in
+  this paper serves as an intuitive and general framework to syntactically unify
+  these related ideas and internalize their
+  theorems formally. We give a new form of \emph{logical characterization} of
+  grammar classes based on a substructural logic rather than ordinary
+  first-order logic. Further, the type theory can not only
   express the grammars as types, but equivalences between grammars and
@@
   We give this type theory a semantics in the category of grammars and
-  prove a canonicity theorem that shows that every term in a context
-  corresponding to a fixed string is equal to a term in a canonical
-  form encoding a parse tree of that string.
+  prove a related canonicity theorem. Moreover, we keep an eye toward the
+  implementation of this type theory for correct-by-construction parsers and
+  parser combinators.
 \end{abstract}
@@
 \section{A Syntax-Independent Notion of Syntax}
+\todoin{Rework intro from LICS draft with a focus on implementation}
 
-The theory of formal languages and parsing is one of the oldest and
-most thoroughly developed areas of theoretical computer science. A
-prominent topic in the 1950s and 60s led to a series of remarkable
-developments: Chomsky's hierarchy of grammar formalisms, practical
-algorithms for parsing of regular and context-free grammars and
-variants, as well as implementations of practical tools for the
-generation of efficient parsers.
+\section{Kleene Categories}
+\todoin{Tell a story winding through the use of KAs, action algebras, etc to
+  motivate why we'd want categorical semantics for KCats}
+\todoin{Incoporate Kleene Category definition}
+\todoin{End with the punch that $\Set^{\Sigma^{*}}$ is a KCat}
 
-A central notion is that of a \emph{formal language} $L$ over an
-alphabet $\Sigma$ as simply a \emph{subset} of the set of strings $L
-\subseteq \Sigma^*$. This definition is especially useful as it gives
-a semantics to formal grammars that is completely independent of any
-particular syntactic grammar formalism: any new notion of grammar can
-be given a semantics in this common framework and it provides a
-precise mathematical speicification for implementing a
-\emph{recognizer} of a language and comparing the power of different
-formalisms by demonstrating what languages can be formalized within
-it. This also allows for completely \emph{language theoretic}
-formulations of language classes: e.g., the regular \emph{languages}
-can be characterized as those that have finitely many derivatives
-\cite{brzozowski-or-myhill-nerode-idk}. This notion is remarkable
-because it makes no reference to any specific syntactic formalism for
-defining languages such as regular expressions or finite automata.
-
-Formal language theory alone is not sufficient as a specification of a
-parser: a language $L \subseteq \Sigma^*$ is equivalently defined as
-its indicator function $\chi_L : \Sigma^* \to \Prop$ mapping each
-string $w$ to the proposition that it is in the language $w \in
-L$. This can then serve as the specification for a language
-\emph{recognizer}: a program $r$ of type $\Sigma^* \to \Bool$ such
-that $r(w) = \texttt{true}$ if and only if $\chi_L(w)$ holds, but
-recognition is insufficient for most tasks for which the theory of
-parsing was developed: the production of \emph{semantic} structures
-from synactic descriptions in linguistics, compilation or
-deserialization. The output of a parser is not just a boolean but a
-\emph{parse tree}\footnote{such parse trees are usually not
-materialized in memory, as the construction of the parse tree is fused
-with application of semantic actions, which can be seen as a kind of
-fold over the generated tree.}.
-
-Due to this limitation, parsers are typically specified not by a
-formal language, but by some \emph{formal grammar}, which specifies
-not just \emph{which} strings are in the language but specifies what
-are the \emph{parse trees} for the string. However unlike formal
-language theory, there is no common universal notion of what
-constitutes a formal grammar, but rather many syntactic systems such
-as Chomskyan(?)  grammars, Thue systems, Montague grammars, etc which
-are all syntactic presentations of the same underlying idea of an
-abstract specification for parsing. The most common of these is the
-Chomsky hierarchy of \emph{generative} grammars which specify parse
-trees using \emph{derivations} of the string from a set of rewrite
-rules, the parse tree encoding which rules were used.
-
-Our first contribution is conceptual: we propose a simple,
-syntax-independent definition of \emph{formal grammar}:
-\begin{definition}
-  A \emph{formal grammar} over an alphabet $\Sigma$ is a function
-  $\Sigma^* \to \Set$.
-\end{definition}
-We say that a grammar $G$ associates to every string $w$ the set $Gw$
-of \emph{parses} of that string. While we will work informally in this
-paper, the collection $\Set$ can be formalized in many different
-logical foundations: as a universe in type theory, a proper class in
-set theory, etc. We argue that this is already a common intuition
-latent in much prior work on grammars, especially when comparing
-different formalisms for equivalence such as \cite{??}. This notion is
-also completely natural in the context of constructive type theories
-such as Agda which until recently did not include a universe of
-propositions, and so for example in Elliott (\cite{??}) a formal
-language was defined as a function to the universe $\Type$ with little
-comment.
-
-Every formal grammar $G$ induces a formal language, which can be seen
-most easily by using the characteristic function formulation: we can
-``squash'' any set into the proposition that it is inhabited, so
-$\chi_G(w) = ||G(w)||$ which defines a subset $L_G = \{ w \in
-\Sigma^*| G(w) \mathrm{inhabited}\}$.
-
-Formal languages naturally arrange themselves into a partial order by
-using the subset inclusion as the ordering. Formal grammars naturally
-arrange themselves into a \emph{category} where a morphism $\alpha : G
-\to H$ is given by a \emph{family} of functions
-\[ \alpha^w : Gw \to Hw \]
-for every string $w \in \Sigma^*$. The intuition is that a morphism of
-grammars is a \emph{translation of parses}: a $G$-parse of $w$ gets
-transformed to an $H$-parse of $w$. This then induces a notion of
-\emph{isomorphism} of grammars: two grammars are equivalent if there
-is a bijective translation of parses, precisely capturing the notion
-of a \emph{strong equivalence} of grammars.
-
-This category-theoretic framework can be further used to describe and
-compare different notions of formal grammar. We can define a
-\emph{notion of grammar} to be a category (or perhaps just groupoid)
-paired with a functor into $\Set^{\Sigma^*}$. This naturally forms a
-2-category the slice 2-category over $\Set^{\Sigma^*}$ and equivalence
-in this 2-category defines a sensible notion of \emph{strong
-equivalence of grammar formalisms} (TODO: check details).
-
-The category of grammars is quite rich in structure, which can be most
-easily observed from the fact that it is equivalently defined as the
-category of functors $\Set^{\Sigma^*}$ where $\Sigma^*$ is here viewed
-as a discrete category, that is, the objects are strings and only
-morphisms are identity morphisms. Such functor categories, often
-called presheaf categories, are incredibly rich in structure, which we
-will exploit in section \ref{blah} to succinctly define the semantics
-of regular and context-free grammars.
-
-\section{Kleene Category}
-
-Kleene algebras are an important tool in the theory of regular
-languages. More broadly, they serve as a theoretical substrate to
-studying various kinds of formal languages. Formally, they are a tuple
-$(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is a set, $+$ and $\cdot$
-are binary operations over $A$, $(-)^*$ is a function over $A$, and
-$1$ and $0$ are constants. These structures satisfy the axoims depicted
-in Figure~\ref{fig:axioms}.
-
-\begin{figure}
-  \begin{align*}
-    x + (y + z) &= (x + y) + z & x + y &= y + x\\
-    x + 0 &= x & x + x &= x\\
-    x(yz) &= (xy)z & x1 &= 1x = x\\
-    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
-    x0 &= 0x = x & & \\
-    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
-     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
-  \end{align*}
-  \caption{Kleene algebra axioms}
-  \label{fig:axioms}
-\end{figure}
-
-The addition operation can be used to define the partial order
-structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
-formal languages, this order structure can be used to model language
-containment. In this section, we want to categorify the concept of
-Kleene algebra and build on top of it in order to define an abstract
-theory of parsing.  We start by defining \emph{Kleene categories}.
-
-\begin{definition}
-  A Kleene category is a distributive monoidal category $\cat{K}$
-  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
-  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
-  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
-  X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogue isomorphism
-  for $G_{A,B}$ also holds.
-\end{definition}
-
-As a sanity check, note that Kleene algebras are indeed examples of
-Kleene categories.
-
-\begin{example}
-  Every Kleene algebra, seen a posetal category, is a Kleene category.
-\end{example}
-
-This example illustrates how restricted Kleene algebras are in terms
-of reasoning about languages. By only having at most one morphism
-between objects, there is not a lot of information they can convey. In
-this case, the only information you get is language containment. We
-will see throughout this paper how the extra degrees of freedom
-granted by having more morphisms give you more algebraic structure for
-reasoning about languages.
-
-For the next example, we see an unexpected connection with the theory
-of substructural logics.
-
-\begin{example}
-  The opposite category of every Kleene category is a model of a variant of
-  conjunctive ordered logic, where the Kleene star plays the role of the ``of
-  course'' modality from substructural logics which allows hypotheses to
-  be discarded or duplicated.
-\end{example}
-
-The proposed axioms are a direct translation of the Kleene algebra
-axioms to a categorical setting. Its most unusual aspect is the
-axiomatization of the Kleene star as a family of initial algebras
-satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
-more structure, then these isomorphisms hold ``for free''.
-
-\begin{theorem}
-  \label{th:kleeneclosed}
-  Let $\cat{K}$ be a Kleene category such that it is also monoidal
-  closed.  Then, the initial algebras isomorphisms hold automatically.
-\end{theorem}
-\begin{proof}
-  We prove this by the unicity (up-to isomorphism) of initial
-  algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
-  (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
-  X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
-  (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
-    1}(X))$.
-
-  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
-  we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,1} \to B \lto Y$. We can show existence and
-  uniqueness by showing that the diagram on top commutes if, and
-  only if, the diagram on the bottom commutes:
-
-% https://q.uiver.app/#q=WzAsOCxbMCwwLCJCICsgQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFsyLDAsIkIgKyBZIFxcb3RpbWVzIEEiXSxbMiwyLCJZIl0sWzAsMiwiQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDMsIjEgKyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDUsIlxcbXUgWC5cXCwgMSArIFggXFxvdGltZXMgQSJdLFsyLDMsIjEgKyAoIEIgXFxsdG8gWSkgXFxvdGltZXMgQSJdLFsyLDUsIkIgXFxsdG8gWSJdLFswLDEsImlkICsgKGlkIFxcb3RpbWVzIGg7IGV2KSBcXG90aW1lcyBpZF9BIl0sWzEsMiwiW2YsZ10iXSxbMywyLCJpZCBcXG90aW1lcyBoOyBldiIsMl0sWzAsMywiW2lkIFxcb3RpbWVzIGgsIGlkIFxcb3RpbWVzIHRsXSIsMl0sWzQsNiwiaWQgKyAoaCBcXG90aW1lcyBYKVxcb3RpbWVzIGlkIl0sWzUsNywiaCIsMl0sWzYsNywiW2YnLCBnJ10iXSxbNCw1LCJbaGQsIHRsXSIsMl1d
-\[\begin{tikzcd}
-	{B + B \otimes (\mu X.\, 1 + X \otimes A)} && {B + Y \otimes A} \\
-	\\
-	{B \otimes (\mu X.\, 1 + X \otimes A)} && Y \\
-	{1 + (\mu X.\, 1 + X \otimes A)} && {1 + ( B \lto Y) \otimes A} \\
-	\\
-	{\mu X.\, 1 + X \otimes A} && {B \lto Y}
-	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
-	\arrow["{[f,g]}", from=1-3, to=3-3]
-	\arrow["{id \otimes h; ev}"', from=3-1, to=3-3]
-	\arrow["{[id \otimes h, id \otimes tl]}"', from=1-1, to=3-1]
-	\arrow["{id + (h \otimes X)\otimes id}", from=4-1, to=4-3]
-	\arrow["h"', from=6-1, to=6-3]
-	\arrow["{[f', g']}", from=4-3, to=6-3]
-	\arrow["{[hd, tl]}"', from=4-1, to=6-1]
-\end{tikzcd}\]
-
-  This equivalence follows by using the adjunction structure given
-  by the monoidal closed structure of $\cat{K}$. A completely analogous
-  argument for $G_{A,B}$ also holds.
-\end{proof}
-
-This result feels similar in spirit to the definition of action
-algebras, which are algebras where the product also has adjoint
-operations which results in the Kleene star being more easily
-axiomatized \cite{kozen1994}.
-
-We are now ready to prove that our concept of formal grammars fits
-nicely within our categorical framework. We start by presenting a
-well-known construction from presheaf categories.
-
-\begin{definition}
-  Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
-  two functors $\cat{C} \to \Set$. Their Day convolution tensor
-  product is defined as the following coend formula:
-  \[
-  (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z) 
-  \]
-  Dually, its internal hom is given by the following end formula:
-  \[
-  (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
-  \]
-\end{definition}
-
-\begin{lemma}[\cite{day1970}]
-  Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
-  monoidal closed.
-\end{lemma}
-
-%% \begin{theorem}
-%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
-%%   The functor category $[A, \cat{K}]$.
-%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
-%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
-%% \end{theorem}
-\begin{theorem}
-  If $\cat{C}$ is a locally small monoidal category, then
-  $\Set^{\cat{C}}$ is a Kleene category.
-\end{theorem}
-\begin{proof}
-
-  By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
-  is a presheaf category, it has coproducts. Furthermore, the tensor
-  is a left adjoint, i.e. it preserves colimits and, therefore, it is
-  a distributive category.
-
-  As for the Kleene star, since presheaf categories admit small colimits,
-  the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
-  defined as the filtered colimit of the diagrams:
-
-  From Theorem~\ref{th:kleeneclosed} it follows that these initial
-  algebras satisfy the required isomorphisms and this concludes the
-  proof.
-\end{proof}
-
-\begin{corollary}
-  For every alphabet $\Sigma$, the presheaf category $\Set^{\cat{\Sigma^*}}$
-  is a Kleene category.
-\end{corollary}
-\begin{proof}
-  Note that string concatenation and the empty string make the
-  discrete category $\Sigma^*$ a strict monoidal category.
-\end{proof}
-
-Much like in the posetal case, the abstract structure of a Kleene
-category is expressive enough to synthetically reason about formal
-languages. A significant difference between them is that while Kleene
-algebras can reason about language containment, Kleene categories can
-reason about parsing as well.
-
-For the rest of the paper we will use the presheaf category
-$\Set^{\cat{\Sigma^*}}$ as a concrete model that will serve as our
-guide when expanding abstract formalism presented in this section so
-that it can handle more classes of languages beyond regular ones.
-
-\paragraph{Beyond Simple Types}
-
-Though Kleene categories are expressive enough to reason about
-concepts that are outside of reach of Kleene algebras, their
-simply-typed nature makes them not so expressive from a type theory
-point of view. This is limiting because type theories are tested
-syntactic frameworks for manipulating complicated categories
-structures while avoiding some issues common in category theory, such
-as coherence issues.
-
-With this in mind, we want to design a dependent type theory that
-builds on top of Kleene categories with the goal of internalizing
-various reasoning principles for languages and their parsers. This
-leads us to the abstract notion of model we are interested in
-capturing with our type theory.
-
-\begin{definition}
-  A model is a locally Cartesian category with two hierarchies of
-  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
-  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
-  $U_i$ to be closed under dependent products and sum,
-  $L_i$ to be closed under the Kleene category connectives, 
-  dependent products, left and right closed structures, with
-  a type constructor $G : L_i \to U_i$ and a linear dependent sum
-  going the other direction.
-\end{definition}
-
-\begin{theorem}
-  The presheaf category $\Set^{\cat{\Sigma^*}}$ is a model.
-\end{theorem}
-
-\section{Related work}
-
-\paragraph{Kleene Algebra}
-
-Since the early works in the theory of formal languages, Kleene
-algebras have played an important role in its development. They
-generalize the operations known from regular languages by introducing
-operations generalizing language composition, language union and the
-Kleene star.  More generally, they are defined as inequational theory
-where the inequality is meant to capture language containment. This
-theory is extremely successful, having found applications in algebraic
-path problems, theory of programming languages, compiler optimizations
-and more.
-
-A frequently fruitful research direction is exploring varying
-extensions of Kleene algebras, Kleene algebra with tests (KAT) being
-one of the most notable ones. Our approach is radically different from
-most extensions, which usually aim at modifying or adding new
-operations to Kleene algebras, but still keeping it as an inequational
-theory. By adopting a category-theoretic treatment and allowing the
-``order structure'' to encode more information than merely
-inequalities, we were able to extend Kleene algebra to reason about
-parsing as well.
-
-\paragraph{TODO: more stuff}
-
-%% That paper with Fritz Henglein about regexes as types
-
-%% Vaughan Pratt on the residual operator
-
-%% \section{Universal Constructions in the Category of Grammars}
-
-%% Predicate BI: bicartesian closed, monoidal biclosed, omega-colimits
-%% and therefore certain initial algebras
-
-\section{Non-commutative Linear-Non-Linear Type Theory as a Syntax for Grammars}
-\label{sec:typetheory}
-\subsection{Syntax}
-Judgments: $\Delta$ set ctx, $\Delta \vdash A$ set, $\Delta \vdash M :
-A$ term, $\Delta \vdash \Gamma$ grammar ctx, $\Delta \vdash G$
-grammar, $\Delta\pipe \Gamma \vdash p : G$ parse term.
-
-\begin{figure}
-  \label{fig:structjdg}
-  \begin{mathpar}
-    \inferrule{~}{\ctxwff \cdot}
-    \and
-    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}
-
-    \\
-
-    \inferrule{~}{\linctxwff \Gamma \cdot}
-    \and
-    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}
-
-    \\
-
-    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}
-
-    \and
-
-    \inferrule{\Gamma \vdash A : L_i}{\linctxwffjdg \Gamma A}
-
-    \\
-
-    \inferrule{\Gamma \vdash X \equiv Y : U_i}{\ctxwffjdg \Gamma {X\equiv Y}}
-
-    \and
-
-    \inferrule{\Gamma \vdash A \equiv B : L_i}{\linctxwffjdg \Gamma {A \equiv B}}
-
-  \end{mathpar}
-  \caption{Structural judgements}
-\end{figure}
-
-\begin{figure}
-  \label{fig:typewf}
-  \begin{mathpar}
-    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
- %   
-    \and
-%
-    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
-%
-    \\
-%
-    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
-%
-    \and
-%
-    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
-%
-    \\
-%
-    \inferrule{~}{\Gamma \vdash 1 : U_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
-%
-    \\
-%
-    \inferrule{~}{\Gamma \vdash I : L_i}
- %   
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
-%
-    \\
-%
-    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
-%
-    \\
-%
-    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
-    %
-    \and
-    %
-    \inferrule{~}{\Gamma \vdash \top : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
-%
-    \\
-    %
-    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
-    %
-    \and
-    %
-    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
-  \end{mathpar}
-  \caption{Type well-formedness}
-\end{figure}
-
-\begin{figure}
-  \begin{mathpar}
-    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
-%
-    \and
-%
-    \inferrule{~}{}
-%
-    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
-%   
-    \and
-%
-    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
-    %
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
-%
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C} 
-    %
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
-\end{mathpar}
-  \caption{Judgemental equality}
-\end{figure}
-
-\begin{figure}
-  \label{fig:inttyping}
-  \begin{mathpar}
-  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
-  %
-  \\\
-  %
-  \inferrule{~}{\Gamma \vdash () : 1}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
-  %
-  \\
-%
-  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
-  \and
-  \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
-  %
-  \\
-  %
-  \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
-  \and
-  \inferrule{\Gamma \mid \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
-  \end{mathpar}
-  \caption{Intuitionistic typing}
-\end{figure}
-
-\begin{figure}
-  \label{fig:linsyntax}
-  \begin{mathpar}
-    \inferrule{~}{\Gamma \mid a : A \vdash a : A}
-    \and
-    \inferrule{\Gamma \mid \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma \mid \Delta \vdash e : A}
-    %
-    \\
-    %
-    \inferrule{~}{\Gamma \mid \cdot \vdash () : I}
-    \and
-    \inferrule{\Gamma \mid \Delta \vdash e : I \\ \Gamma \mid \Delta_1',\Delta_2' \vdash e' : C}{\Gamma \mid \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
-    %
-    \\
-    %
-    \inferrule{\Gamma \mid \Delta \vdash e : A \\ \Gamma \mid \Delta' \vdash e' : B}{\Gamma \mid \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
-    %
-    \\
-    %
-    \inferrule{\Gamma \mid \Delta \vdash e : A \otimes B \\ \Gamma \mid \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma \mid  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
-    \\
-    %
-    \inferrule{\Gamma \mid \Delta, a : A \vdash e : B}{\Gamma \mid \Delta \vdash \lamblto a e : A\lto B}
-    \and
-    \inferrule{\Gamma \mid \Delta' \vdash e' : A \\ \Gamma \mid \Delta \vdash e : A \lto B}{\Gamma \mid \Delta', \Delta \vdash \applto {e'} {e} : B}
-    \\
-    %
-    \inferrule{\Gamma \mid a : A, \Delta \vdash e : B}{\Gamma \mid \Delta \vdash \lambtol a e : B\tol A}
-    \and
-    \inferrule{\Gamma \mid \Delta \vdash e : B \tol A \\ \Gamma \mid \Delta' \vdash e' : A}{\Gamma \mid \Delta, \Delta' \vdash \apptol e {e'} : B}
-    %
-    \\
-    %
-    \inferrule{\Gamma, x : X \mid \Delta  \vdash e : A}{\Gamma \mid \Delta \vdash \dlamb x e : \LinPiTy x X A}
-    \and
-    \inferrule{\Gamma \mid \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma \mid \Delta \vdash \app e {e'} : \subst A {e'} x}
-    %
-    \\
-    %
-    \inferrule{\Gamma \vdash e : X \quad \Gamma \mid \Delta \vdash e' : \subst A e x}{\Gamma \mid \Delta \vdash (e, e') : \LinSigTy x X A}
-    %
-    \\
-    %
-    \inferrule{\Gamma \mid \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X \mid \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma\mid \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
-    %
-    \\
-    %
-    \inferrule{~}{\Gamma \mid \Delta \vdash () : \top}
-    %
-    \\
-    %
-    \inferrule{\Gamma \mid \Delta \vdash e_1 : A_1 \quad \Gamma \mid \Delta \vdash e_2 : A_2}{\Gamma \mid \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
-    \and
-    \inferrule{\Gamma \mid \Delta \vdash e : A_1 \amp A_2 }{\Gamma \mid \Delta \vdash \pi_i \, e : A_i}
-    %
-    \\
-    %
-    \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma \mid \cdot \vdash \mathsf{G}^{-1}\, e : A}
-    %
-    \\
-    %
-    \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
-    \and
-    \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
-  \end{mathpar}
-  \caption{Linear typing}
-\end{figure}
-
-\pedro{Need to figure out the rules for inductive types and write down
-  the judgemental equality rules}
-
-\subsection{Semantics in Grammars}
-
-For illustrative purposes, we give a brief overview of the concrete semantics in sets and grammars:
-\begin{enumerate}
-\item A non-linear context $\Gamma$ denotes a set $\sem{\Gamma}$
-\item A non-linear type $\Gamma \vdash X : U_i$ denotes a family of sets $\sem{X} : \sem{\Gamma} \to \Set_i$
-\item A non-linear term $\Gamma \vdash e : X$ denotes a section $\sem{e} : \Pi(\gamma:\sem{\Gamma})\,\sem{X}\,\gamma$
-\item Linear contexts $\Gamma \vdash \Delta$ and types $\Gamma \vdash A : L_i$ both denote families of grammars $\sem{\Gamma}\to \Gr_i$
-\item A linear term $\Gamma; \Delta \vdash e : A$ denotes a family of parse transformers $\sem{e} : \Pi(\gamma:\sem{\Gamma})\Pi(w:\Sigma^*)\,\sem{\Delta}\gamma w \Rightarrow \sem{A}\gamma w)$
-\end{enumerate}
-Most of the non-linear semantics is standard, we show just the $G$ type and the linear semantics:
-\begin{enumerate}
-\item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
-\item $\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-\item $\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-\item $\sem{I} \gamma w = \{ () \pipe w = \epsilon \}$
-\item $\sem{A \otimes B} \gamma w = \Sigma(w1,w2:\Sigma^*) w1w1 = w \wedge \sem{A} \gamma w_1 \times \sem{B} \gamma w_2$
-\item $\sem{A \lto B} \gamma w = \Pi(w_a:\Sigma^*) A \gamma w_a \Rightarrow B\gamma (w_aw)$
-\item $\sem{B \tol A} \gamma w = \Pi(w_a:\Sigma^*) A \gamma w_a \Rightarrow B\gamma (ww_a)$
-\item $\sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x)$
-\end{enumerate}
-
-Propositional fragment: monoidal biclosed category that is also
-bicartesian closed and has $\omega$-colimits. Call such a BIC.
-
-Set theoretic fragment: category with families $\mathcal C$ + $\Pi,\Sigma,W$.
-
-Combination: category with families $\mathcal C$ + a functor $L : C
-\to \textrm{BIC}$.
-
-Grammatical fragment: functor from $\mathcal C$ to the category of BI-models
-
-Example: use $\Set$ as the CwF, use the functor $\Delta \mapsto
-(\Set^{\Sigma^*})^\Delta$ as the model of the propositional fragment
-using the pointwise-Day convolution monoidal structure.
-
-\subsection{Canonicity}
-
-Let $\cdot \vdash A$ be a closed linear type. Then there are
-two obvious notions of what constitutes a ``parse'' of a string w
-according to the grammar $A$:
-\begin{enumerate}
-\item On the one hand we have the set-theoretic semantics just
-  defined, $\llbracket A \rrbracket \cdot w$
-\item On the other hand, we can view the string $w = c_1c_2\cdots$ as
-  a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
-  define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
-  \lceil w \rceil \vdash e : G$.
-\end{enumerate}
-It is not difficult to see that at least for the ``purely positive''
-formulae (those featuring only the positive connectives
-$0,+,I,\otimes,\mu, \overline\Sigma,c$) that
-every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
-that the nodes of the tree correspond precisely to the introduction
-forms of the type. However it is far less obvious that \emph{every}
-linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
-sequence of introduction forms since proofs can include elimination
-forms as well. To show that this is indeed the case we give a
-\emph{canonicity} result for the calculus: that the parses for .
-
-\begin{definition}
-  A non-linear type $X$ is purely positive if it is built up using
-  only finite sums, finite products and least fixed points.
-
-  A linear type is purely positive if it is built up using only finite
-  sums, tensor products, generators $c$, least fixed points and linear
-  sigma types over purely positive non-linear types.
-\end{definition}
-
-\begin{definition}
-  %% Let $X$ be a closed non-linear type. The closed elements $\textrm{Cl}(X)$ of $X$ are the definitional equivalence classes of terms $\cdot \vdash e : X$.
-
-  Let $A$ be a closed linear type. The nerve $N(A)$ is a presheaf on
-  strings that takes a string $w$ to the definitional equivalence
-  classes of terms $\cdot; \lceil w\rceil \vdash e: N(A)$.
-\end{definition}
-
-\begin{theorem}[Canonicity]
-  Let $A$ be a closed, purely positive linear type. Then there is an
-  isomorphism between $\llbracket A\rrbracket$ and $N(A)$.
-\end{theorem}
-\begin{proof}
-  We outline the proof here, more details are in the appendix. The
-  proof proceeds first by a standard logical families construction
-  that combines canonicity arguments for dependent type theory
-  \cite{coquand,etc} with logical relations constructions for linear
-  types \cite{hylandschalk}. It is easy to see by induction that the
-  logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
-  \rrbracket$ and the fundamental lemma proves that the projection
-  morphism $p : \hat A \to N(A)$ has a section, the canonicalization
-  procedure. Then we establish again by induction that
-  canonicalization is also a retraction by observing that introduction
-  forms are taken to constructors.
-\end{proof}
-
-\begin{enumerate}
-\item Every term $\lceil w \rceil \vdash p : G + H$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
-\item There are no terms $\lceil w \rceil \vdash p : 0$
-\item If there is a term $\lceil w \rceil \vdash p : c$ then $w = c$ and $p = x$.
-\item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
-\item Every term $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
-\item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
-\item Every term $\lceil w \rceil \vdash p : \mu X. G$ is equal to $\textrm{roll}(q)$ where $q : G(\mu X.G/X)$
-\item Every term $\lceil w \rceil \vdash p : (x:A) \times G$ is equal
-  to $(M,q)$ where $\cdot \vdash M : A$
-\end{enumerate}
-
-To prove this result we will use a logical families model. We give a
-brief overview of this model concretely:
-\begin{enumerate}
-\item A context $\Gamma$ denotes a family of sets indexed by closing substitutions $\hat\Gamma : (\cdot \vdash \Gamma) \Rightarrow \Set_i$
-\item A type $\Gamma \vdash X : U_i$ denotes a family of sets $\hat X : \Pi(\gamma:\cdot \vdash \Gamma) \hat\Gamma \Rightarrow (\cdot \vdash \simulsubst X \gamma) \Rightarrow \Set_i$
-\item A term $\Gamma \vdash e : X$ denotes a section $\hat e : \Pi(\gamma)\Pi(\hat\gamma)\hat X \gamma \hat\gamma (\simulsubst e \gamma)$
-\item A linear type $\Gamma \vdash A : L_i$ denotes a family of grammars $\hat A : \Pi(\gamma:\cdot\vdash\Gamma)\,\hat\Gamma \Rightarrow \Pi(w:\Sigma^*) (\cdot;\lceil w\rceil \vdash A[\gamma])\Rightarrow \Set_i$, and the denotation of a linear context $\Delta$ is similar.
-\item A linear term $\Gamma;\Delta \vdash e : A$ denotes a function \[\hat e : \Pi(\gamma)\Pi(\hat\gamma)\Pi(w)\Pi(\delta : \lceil w \rceil \vdash \simulsubst \Delta \gamma) \hat\Delta \gamma \hat\gamma \delta \Rightarrow \hat A \gamma \hat\gamma w {(\simulsubst {\simulsubst e \gamma} \delta)}\]
-\end{enumerate}
-And some of the constructions:
-\begin{enumerate}
-\item $\widehat {(G A)} \gamma \hat\gamma e = \hat A \gamma \hat\gamma \varepsilon (G^{-1}e)$
-\item $\widehat {(A \otimes B)} \gamma \hat\gamma w e = \Sigma(w_Aw_B = w)\Sigma(e_A)\Sigma(e_B) (e_A,e_B) = e \wedge \hat A \gamma \hat \gamma w_A e_A \times \hat B \gamma \hat \gamma w_B e_B$
-\item $\widehat {(A \lto B)} \gamma \hat\gamma w e = \Pi(w_A)\Pi(e_A) \hat A \gamma\hat\gamma w_A e_A \Rightarrow \hat B \gamma\hat\gamma (ww_A) (\applto {e_A} e)$
-\end{enumerate}
-
-First, the category with families will be
-the category of logical families over set contexts/types
-$\Delta$/$A$. Then the propositional portion will be defined by
-mapping a logical family $\hat \Gamma \to \Gamma$
-
-First, let $L$ be the category of BI formulae and proofs (quotiented
-by $\beta\eta$ equality). Define a functor $N : L \to \Set^{\Sigma^*}$ by
-\[ N(\phi)(w) = L(w,\phi) \]
-
-Then define the gluing category $\mathcal G$ as the comma category
-$\Set^{\Sigma^*}/N$. That is, an object of this category is a pair of
-a formula $\phi \in L$ and an object $S \in \mathcal
-P(\Sigma^*)/N(\phi)$. We can then use the equivalence $\mathcal
-P(\Sigma^*)/N(\phi) \cong \mathcal P(\int N(\phi))$ to get a simple
-description of such an $S$: it is simply a family of sets indexed by
-proofs $L(w,\phi)$:
-\[ \prod_{w\in\Sigma^*} L(w,\phi) \to \Set \]
-This category clearly comes with a projection functor $\pi : \mathcal
-G \to \mathcal L$ and then our goal is to define a section by using
-the universal property of $\mathcal L$.
-
-To this end we define
-\begin{enumerate}
-\item $(\phi, S) \otimes (\psi, T) = (\phi \otimes \psi, S\otimes T)$ where
-  \[ (S \otimes T)(w, p) = (w_1w_2 = w) \times (q_1,q_2 = p) \times S\,w_1\,q_1 \times T\,w_2\,q_2\]
-\item $(\phi, S) \multimap (\psi, T) = (\phi \multimap \psi, S \multimap T)$ where
-  \[ (S \multimap T)(w,p) = w' \to q \to S\,w'\,q \to T (ww') (p\,q) \]
-\item $\mu X. ??$ ??
-\end{enumerate}
-
-\section{Type Theoretic Formal Grammar Theory}
-
-Now that we have developed a type theoretic syntax for formal grammars
-and parse transformers, we show how many common concepts of formal
-language theory and grammars are naturally expressed using type
-theoretic notions, and some constructions and proofs can be carried
-out in our type theory as well.
-
-Firstly, regular expressions and their extension to context-free
-expressions are naturally expressed as a limited syntax of linear
-types. That is, regular expressions are precisely the linear types
-constructed using only $c, I, \otimes, 0, \oplus$ and the Kleene star,
-which can be defined as $A^* = \mu x. I \oplus A \otimes X$. The
-context-free expressions are similarly expressible by allowing not
-just the Kleene star, but unrestricted usage of $\mu$. Remember,
-however, that these are a notion of regular expressions that maintain
-a notion of ambiguous grammar. That is, $A \oplus A$ is not considered
-to be the same as $A$. While there does not appear to be an analogous
-weakening of this definition to naturally describe the
-context-sensitive languages, we will see shortly that the indexing of
-linear types by non-linear types allows for capturing a notion of
-context in grammars.
-
-In addition to expressing algebraic presentations of grammars, the
-type-theoretic syntax allows for naturally expressing grammars that
-capture the behavior of automata. \max{sync this up with Steven's
-  stuff, and add stuff about PDAs and Turing Machines}.
-
-A key construct in formal language theory is the Brzozowski derivative
-\cite{brzozowski}. Given a string $w$ the (left) derivative of a
-language $L$ with respect to $w$, written $\partial_w L$ is the language
-\[ (\partial_wL ) = \{ ww' \pipe w' \in L \}\]
-This semantics is given using the linear implication $w \tol
-L$. Expressing Brzozowski's calculation of the derivative of a regular
-expression relies on additional axioms to the type theory.\max{axioms
-  to carry out Brzozowski's derivative calculations plz.}
-
-\begin{align*}
-  c\lto c &\equiv I\\
-  c\lto d &\equiv 0\\
-  c\lto I &\equiv 0\\
-  c\lto 0 &\equiv 0\\
-  c\lto (A \otimes B) &\equiv (c\lto A) \otimes B + \nu(A) \otimes (c\lto B)\\
-  c\lto A^* &\equiv (c\lto A) \otimes A^*\\
-  c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
-\end{align*}
-
-\pedro{The type family $\nu$ above is equal to $I$ if its argument
-  parses the empty string and zero otherwise. In order to define
-  it we should probably add an extra axiom, right?}
-\steven{$\nu$ can be derived using $\amp$, $\nu(A) = I \amp A$}
-
-\steven{In addition to Pedro's axioms:
-
-$c, c'$ characters. $g$ any grammar.
-
-\begin{gather}
-  I \lto c \cong c \\
-  c \lto (c' \lto g) \cong (c' \otimes c) \lto g
-\end{gather}
-
-For axiom $(1)$, Brzozowski has $I \lto g \cong g$ for all grammars $g$, but I don't think that holds for us because we want to preserve ambiguity.
-We could instead have something like $I \lto g \cong (g \amp I)^{*} \otimes g$.
-For our purposes the character presentation seems sufficient, as the derivative reasoning Brzozowski uses only takes derivatives with respect to strings.
-
-Axiom $(2)$ might be admissible, but I'm unsure.
-
-I think for Kleene star we need $c \lto g^{*} \cong (g \amp I)^{*} \otimes (c \lto g) \otimes g^{*}$.
-This should derivable utilizing the product axiom and $g^{*} \cong I \oplus (g \otimes g^{*})$, or is that sort of isomorphism opaque to us right now.
-
-To make the DFA proof go through I think we additionally need to axiomatically make the map $c \otimes (c \lto A) \vdash (c \otimes \top) \amp A$ to be an iso.
-Specifically, this term needs an iso $\phi : c \otimes (c \lto A) \to (c \otimes \top) \amp A$ defined below,
-
-\[
-p : c \otimes (c \lto A) \vdash \phi(p) = \letin {p_{1} \otimes p_{2}} p {(p_{1} \otimes \top, p_{1}^{\lto}p_{2})} : (c \otimes \top) \amp A
-\]
-
-but this last bit isn't just Brzozowski's definitions
-
-}
-Using the $G$ modality, we can internalize some of the external
-reasoning about grammars into the language of non-linear types. Recall
-that the denotation of $G A$ is the set of parses of the empty
-string. As is well known in formal language theory, all the parsing
-behavior of a grammar can be reduced to reasoning about parsing of the
-empty string by using the derivative of a grammar. For instance, the
-set of parses for a fixed string $w$ can be defined as $G(w \tol A)$
-and so the type of all parses of a grammar can be defined as
-$\Sigma(w:\textrm{String}) G(w \tol A)$. This can be used to define
-internally a notion of \emph{ambiguity} of a grammar. A grammar $A$ is
-ambiguous \steven{This is an unambiguous grammar, right?} if for every string $w$, the type $G(w \tol A)$ of
-$w-$parses is a \emph{proposition} in the sense that all terms in $G(w
-\tol A)$ are equal.\max{I think we should add a propositional
-  truncation to the type theory.}  Similarly, the type of \emph{total
-parsers} can be defined as $G(\top \tol A)$. A total parser encodes a
-function that selects a parse for every possible string. A
-\emph{semi-parser} is then defined as $G(\top \tol A \oplus \top)$,
-where the parser is allowed to fail for any reason. This has the
-downside that the always failing semi-parser $\lambda _. \textrm{in}_r
-()$ is valid for any grammar. If we were working with a full
-\emph{bunched} type theory, we would be able to express the more
-precise type of a \emph{deciding} parser as $G(\top \tol A \oplus \neg
-A)$ where $\neg A = A \Rightarrow 0$ uses the intuitionistic
-implication.
-
-Classically, formal language theory is closely related to automata theory. In the styles of Chomsky and
-Sch\"utzenberger, there is a hierarchy of language classes and the types of automata that recognize them. Inside of the
-type theory of \cref{sec:typetheory}, we can characterize these language classes; moreover, we demonstrate equivalence
-of each class to the associated automaton as an isomorphism within our logic.
-
-Note, a predicate logic over LNL with existential quantifiers is already enough to define
-formulae that correspond to arbitrary recursively enumerable languages ---
-since we can simulate the tape of a Turing machine using two
-stacks. I.e., we can define a set $S$ of lists of tape symbols, and
-then define a formula that first ``copies'' the input onto one stack, then executes
-
-\[ \mu \textrm{Copy} : S \to *. \lambda s. (\epsilon \wedge T\,s\,[])\bigvee_{c \in \Sigma} c (S\,{c::s}) \]
-
-We reflect on this expressive power to motivate why we put heavy syntactic restrictions on our presentation of certain
-classes of automata. For instance, when defining a pushdown automata, we could use the intuitionistic context to
-represent the stack, but we ought to be very careful about how this set is accessed and added to. Careless definitions
-may accidentally endow us with too much expressive power, launching us further up in the Chomsky hierarchy. To begin,
-let us focus only on language classes, and their associated automata, whose parses are definable without any mention to
-the inuitionistic context. When viewed as a machine, this refers to the class of automata definable without access to
-any memory --- i.e.\ finite state machines.
-
-\subsection{Finite State Machines}
-A \emph{nondeterministic finite automaton} (NFA) is a finite state machine where transitions are labeled with characters from a fixed alphabet $\Sigma$.
-These are often represented formally as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$,
-
-\begin{itemize}
-  \item $Q$ a finite set of states
-  \item $\Sigma$ a fixed, finite alphabet
-  \item $\delta : Q \times \Sigma \to \mathcal{P}(Q)$ a labelled transition function
-  \item $q_{0}$ a start state
-  \item $F$ a set of accepting states
-\end{itemize}
-
-Intuitively, this can be thought of like a directed graph with nodes in $Q$ with an edge $q \overset{c}{\to} q'$ whenever $q' \in \delta(q, c)$.
-Note that transitions in an NFA may be labelled with the empty string $\epsilon$ --- such transitions are referred to as \emph{$\epsilon$-transitions}.
-
-As a gramamar in our type theory, an NFA may be encoded as a
-mutual\footnote{For convenience we refer to a mutual fixed point here, even though we
-  only take single fixed points as primitive. As in Beki\'c's theorem, we may decompose our mutual fixed points into a series of single fixed points. So this operation is admissible.}
-fixed point in the following manner:
-
-An NFA has finitely many states, so we may number them $Q = \{q_{1}, \dots, q_{N}\}$. For each $q_{j}$ we introduce
-a mutually recursive term $Y_{j}$ to represent the states of the NFA within this grammar presentation.
-We represent the transition function $\delta$ in the right hand side of each $Y_{j}$ definition.
-For character $c \in \Sigma \cup \{\epsilon\}$ and state $q_{j} \in Q$, we add a summand --- as a coproduct\footnote{There is an admissible notion of coproduct making use of the primitive dependent pairing --- i.e.
-$A \oplus B := \LinPiTy b \Bool {\textbf{if}~b~A~B}$.
-We will utilize this without much comment to represent the usual notion of alternation of grammars.} --- to the defintion of $Y_{j}$ for every state in the image of the transition function,
-
-\[
-Y_{j} = \bigoplus_{c \in \Sigma \cup \{\epsilon\}} \bigoplus_{q_{i} \in \delta(q_{j}, c)} c \otimes Y_{i}
-\]
-
-For convenience, refer to syntactic term on the right hand side of $Y_{j}$ as $\hat{Y}_{j}$.
-Further, we rename the start state to be represented by the term $Y_{s}$.
-Additionally, and without loss of generality, we assume that there is a single accepting state $Y_{e}$ in our encoding.
-This assupmtion is justified through the use of $\epsilon$-transitions.
-For each $q_{j} \in F$, we add an $\epsilon$-transition from $Y_{j}$ to $Y_{e}$.
-
-The above translation totals to give us a syntactic form for NFAs as grammars,
-
-\[
-  NFA ::= \mu \begin{pmatrix}
-        Y_{s} = \langle \hat{Y} \rangle \\
-        Y_{j} = \langle \hat{Y} \rangle, j \not \in \{s, e\} \\
-        Y_{e} = I
-        \end{pmatrix}.Y_{s}
-\]
-where
-\[
-  \hat{Y} ::= c \otimes Y ~|~ Y ~|~ \hat{Y} \oplus \hat{Y}
-\]
-
-\steven{Somewhere mention that $\epsilon$ corresponds to $I$, characters $c \in \Sigma$ correspond to grammars $c$, etc}
-
-and each $Y \in \{ Y_{s}, Y_{e}, Y_{1}, \dots, Y_{N} \}$. Intuitively, we read each $Y_{j}$ as the label for a state, and the
-definitions of form $\hat{Y}$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$
-denotes an $\epsilon$-transition. A
-disjunction of transitions in an NFA grammar corresponds to multiple outgoing edges of a given state in an automaton.
-
-Above we give a translation from the tuple presentation of an NFA to a fixed point grammar in our type theory.
-We don't provide any argument that this translation is done correctly; however, we hope that the construction is straightforward and lightweight enough to demonstrate our formal system as a natural and effective syntax for reasoning about both grammars and automata.
-
-With this setting for finite automata, we can now internalize some classical theorems inside of our formal system.
-
-\subsection{Equivalence Between Regular Grammars and Finite Automata}
-
-% The class of \emph{regular} grammars is given as follows:
-
-% \[
-% Reg ::= 0 ~|~ I ~|~ c ~|~ Reg \oplus Reg ~|~ Reg \otimes Reg ~|~ Reg^{*}
-% \]
-% where $c \in \mathcal{A}$ is a character literal drawn from a fixed alphabet, and the Kleene star $\cdot^{*}$ is derived as an admissible operation using the fixed point operator --- $g^{*} = \mu X.I \oplus (g \otimes X)$.
-\steven{This definition is maybe fair to assume.
-Also maybe better suited for a section earlier in the paper}
-
-First, let us describe a version of Thompson's construction \cite{?} where we construct an NFA that recognizes a given regular expression.
-Moreover, we will show that this NFA is strongly equivalent to the original grammar.
-
-This construction
-
-\begin{theorem}[Thompson]
-  \label{thm:thompson}
-  For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same language as $g$. Further, $N$ and $g$ are isomorphic.
-\end{theorem}
-
-\begin{proof}
-  We construct the recognizing NFA $N$ inductively. To handle the atomic grammars --- literals and the empty grammar --- we define the following,
-
-\[
-  NFA(c) =
-  \mu \begin{pmatrix}
-        Y_{s} = c \otimes Y_{e} \\
-        Y_{e} = I
-  \end{pmatrix}.Y_{s}
-\]
-\[
-  NFA(I) =
-  \mu \begin{pmatrix}
-        Y_{s} = Y_{e} \\
-        Y_{e} = I
-  \end{pmatrix}.Y_{s}
-\]
-
-The isomorphism between the grammars $c$ and $NFA(c)$ are witnessed by the following derivations,
-\[
-  \inferrule{p : c \vdash p : c \\ \cdot \vdash () : I}{p : c \vdash p \otimes () : Y_{s} = c \otimes I}
-\]
-and
-\[
-  \inferrule{q : Y_{s} = c \otimes I \vdash q : Y_{s} \\ p : c, e : I \vdash p : c }{q : Y_{s} \vdash \letin {p \otimes e} q p : c}
-\]
-
-Similar terms are used to show the equivalence between $I$ and $NFA(I)$.
-
-We can now inductively define some NFA combinators and show that they preserve grammar equivalence. First begin with
-the disjunction of two NFAs. For regular grammars $g$ and $g'$, assume that $g \cong NFA(g)$ and $g' \cong NFA(g')$, and we will define an NFA that
-recognizes $g \oplus g'$.
-
-Let the states in $NFA(g)$ be represented by $X$'s and the states in $NFA(g')$ be represented
-by $Z$'s. Further let $\hat{X}_{k}$ refer to the syntactic term on the right hand side of the definition of $X_{k}$
-inside of $NFA(g)$. Likewise define each $\hat{Z}_{k}$.
-
-\[
-  NFA(g \oplus g') =
-  \mu \begin{pmatrix}
-        Y_{s} =  X_{s}' \oplus Z_{s}' \\
-        X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
-        X'_{e} = Y_{e} \\
-        Z_{j}' = \simulsubst {\hat{Z}_{j}} {\delta_{Z}}, j \neq e \\
-        Z_{e}' = Y_{e} \\
-        Y_{e} = I
-  \end{pmatrix}.Y_{s}
-\]
-where the substitution $\{X_{s}'/X_{s}, X_{e}'/X_{e}, X_{1}'/X_{1}, \dots \}$ is given by $\delta_{X}$ --- and $\delta_{Z}$ constructed similarly.
-
-In the construction of $NFA(g \oplus g')$ we replicate each defintion $X_{j}$ in $NFA(g)$ with an analogous definition
-$X_{j}'$ inside of $NFA(g \oplus g')$, except we replace $X_{k}$ with $X_{k}'$ in each production. Lastly, $X_{e}'$ has
-an $\epsilon$-transition to the accepting state $Y_{e}$. Likewise, we build analogous defintions for the states
-$Z_{j}$ in $NFA(g')$, then we flatten all of these defintions into one mutally recursive list. This has the effect
-of disjunctively tying together $NFA(g)$ and $NFA(g')$ into one automaton.
-
-To show the strong equivalence of $NFA(g \oplus g')$ and $g \oplus g'$, we will demonstrate an isomorphism between them
-with two proof terms in our logic that mutually invert each other. To define a map from $NFA(g \oplus g')$ to $g \oplus g'$
-we will use a multi-fold over all the mutually recursive defintions of $NFA(g \oplus g')$. Note, even though multiple
-fixed points are not taken as primitive, using Beki\'c's theorem we can define an admissible axiom schema to define multiple fixed points types.
-Below we give rules to form $\mathsf{cons}$ and $\mathsf{mfold}$ terms to serve as the introduction and elimination rules, respectively, for multi-term fixed points.
-First, the introduction rule,
-
-\[
-  \inferrule
-  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
-  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
-\]
-where $\sigma$ is the substitution that unrolls the mutually recursive definition one level by replacing each $X_{k}$ with the fixed point $\mu(X_{j} = A_{j}, \forall j).X_{k}$,
-
-\begin{align*}
-  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \\
-  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
-\end{align*}
-
-and the elimination rule,
-\[
-  \inferrule{\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
-             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
-             \dots \\
-             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
-  }{\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
-\]
-
-where the substitution $\gamma$ represents $\{B_{1}/X_{1},\dots,B_{n}/X_{n} \}$.
-
-When defining a derivation from $NFA(g \oplus g')$ to $g \oplus g'$, we
-wish to simultaneously fold over the mutually recursive definitions comprising $NFA(g \oplus g')$. We may think of this
-as an eliminator that replaces each state inside of $NFA(g \oplus g')$ with a tail-recursive call that will ultimately
-return a term of type $g \oplus g'$.
-If folding over a term $e$, refer to the output of this multifold as $s_{\oplus}(e)$.
-The structure of such a derivation instatiates the above elimination rule as follow,
-
-\[
-  \inferrule{
-    \Delta \vdash e : NFA(g \oplus g') \\
-    y_{s} : \simulsubst {(X_{s}' \oplus Z'_{s})} \gamma \vdash p_{s} : g \oplus g' \\
-    x_{s}' : \simulsubst {(\simulsubst {X_{s}} {\delta_{X}})}{\gamma} \vdash p_{x,s} : g \\
-    x_{j}' : \simulsubst {(\simulsubst {X_{j}} {\delta_{X}})}{\gamma} \vdash p_{x,j} : X_{j} \text{~where } j \not \in \{s, e\} \\
-    x_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{x, e} : X_{e} \\
-    z_{s}' : \simulsubst {(\simulsubst {Z_{s}} {\delta_{Z}})}{\gamma} \vdash p_{z,s} : g' \\
-    z_{j}' : \simulsubst {(\simulsubst {Z_{j}} {\delta_{Z}})}{\gamma} \vdash p_{z,j} : Z_{j} \text{~where } j \not \in \{s, e\} \\
-    z_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{z,e} : Z_{e} \\
-    y_{e} : \simulsubst I {\gamma} \vdash p_{e} : I
-  }{\Delta \vdash s_{\oplus}(e) := \mathsf{mfold}(e) : g \oplus g'}
-\]
-
-For each of the variables above, we simplify the type and provide the actual proof term. For instance,
-$\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$, so the derivation for $p_{s}$ is just the identity.
-The rest of the terms are given below,
-\begin{itemize}
-  \item $y_{s}$ --- $\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$ and $p_{s} = y_{s} : g \oplus g'$
-  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = \mu(X_{i} = \hat{X}_{i}, \forall i).X_{s} = NFA(g)$ and $p_{x,s} = \phi^{-1}(x_{s}') : g$
-  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = \mu(X_{i} = \hat{X}_{i}, \forall i).X_{j}$ and $p_{x, j} = x_{j}' : \mu(X_{i} = \hat{X}_{i}, \forall i).
-X_{j}$
-  \item $x_{e}'$ --- $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{x,e} = x_{e}' : \mu(X_{i} = \hat{X}_{i}, \forall i).X_{e} = I$
-  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = \mu(Z_{i} = \hat{Z}_{i}, \forall i).Z_{s} = NFA(g')$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
-  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = \mu(Z_{i} = \hat{Z}_{i}, \forall i).Z_{j}$ and $p_{z, j} = z_{j}' : \mu(Z_{i} = \hat{Z}_{i}, \forall i).Z_{j}$
-  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = \mathsf{cons}_{Z_{e}}(z_{e}') : \mu(Z_{i} = \hat{Z}_{i}, \forall i). Z_{e} = I$
-  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
-\end{itemize}
-
-\steven{Need to add argument that justifies why the $X_{j}$s are allowed to be eliminated via the identity.
-A priori they could be interfered with by taking an $X_{s}$ that needs to be eliminated.
-I think it hinges on a syntactic restriction where there are no backedges to $X_{s}$ in the NFA.Even if there were, I think I can get away with a case distinction entailment instead of the one written}
-
-where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively.
-This finshes the construction of a parse transformer $s_{\oplus}$ from $NFA(g \oplus g')$ to $ g \oplus g'$.
-The map $r_{\oplus}$ in the other direction is given below in a couple stages. First, we show how to turn a parse of $NFA(g)$ into a parse of $NFA(g \oplus g')$.
-
-\[
-  \inferrule
-  {
-    \Gamma ; \Delta \vdash u : NFA(g) = \mu(X_{j} = \hat{X}_{j}, \forall j).X_{s} \\
-    \Gamma ; x_{s} : \simulsubst {X_{s}}{\gamma} = X_{s}' \vdash \mathsf{cons}(\textbf{in}_{X_{s}'}(x_{s})) : \mu(\dots).Y_{s} = NFA(g \oplus g')  \\
-    \Gamma ; x_{j} : \simulsubst {X_{j}}{\gamma} = X_{j}' \vdash x_{j} : X_{j}', j \not \in \{s, e\} \\
-    \Gamma ; x_{e} : \simulsubst {X_{e}}{\gamma} = I \vdash x_{e} : I = X_{e}'
-  }
-  {\Gamma ; \Delta \vdash \mathsf{mfold}(u) : NFA(g \oplus g')}
-\]
-where the dots in the above derivation take on the mutually recursive defintions present in $NFA(g \oplus g')$.
-\steven{Need a better way to display these defs without blowing up the $\mu$ terms} Denote the above embedding of $NFA(g)$ into $NFA(g \oplus g')$ via $\alpha(u) := \mathsf{mfold}(u) : NFA(g \oplus g')$.
-Similarly define an embedding $\beta$ for an embedding of $NFA(g')$.
-We can then define the derivation from $g \oplus g'$ to $NFA(g \oplus g')$ by composing the isomorphisms $g \overset{\phi}{\cong} NFA(g)$ and $g \overset{\psi}{\cong} NFA(g')$ with maps $\alpha$ and $\beta$, respectively.
-
-\[
-  \inferrule
-  {
-    p : g \oplus g' \vdash p : g \oplus g' \\
-    p_{1} : g \vdash \alpha(\phi(p_{1})) : NFA(g \oplus g') \\
-    p_{2} : g' \vdash \beta(\psi(p_{2})) : NFA(g \oplus g')}
-  {
-    p : g \oplus g' \vdash r_{\oplus}(p) := \textbf{case}~p \{
-    \textbf{in}_{g} p_{1} \mapsto \alpha(\phi(p_{1})), \\
-    \textbf{in}_{g'} p_{2} \mapsto \beta(\psi(p_{2})) \} : NFA(g \oplus g')}
-\]
-
-We include the proof that maps $s_{\oplus}$ and $r_{\oplus}$ form an isomorphism in \cref{sec:appendixnfa}. This concludes the disjunctive combination of two NFAs.
-Because they follow a very similar style of argumentation, we also include the cases for sequencing and
-Kleene star in \cref{sec:appendixnfa}. Totaling over all of these cases, this completes the proof.
-
-\end{proof}
-
-% \subsubsection{NFA to Grammar}
-
-% \steven{This section and following are basically todos for rn}
-
-% We now seek to take an NFA and derive the regular expression recognized by it.
-To this end we weaken our syntactic form of an NFA to that a \emph{generalized nondeterministic finite automata} (GNFA) --- where guards on transition can be arbitrary regexps instead of single characters.
-We then repeatedly take away states from the NFA to give an increasingly more complete algebraic description of the regular expression inside of the guards.
-We repeat until there is only a single transition between start and accepting state, and the guard on the single transition denotes the regular expression that is equivalent to the NFA.
-
-% \[
-%   GNFA ::= \mu \begin{pmatrix}
-%         Y_{s} = \langle Q \rangle \\
-%         Y_{1} = \langle Q \rangle \\
-%         \vdots \\
-%         Y_{e} = I
-%         \end{pmatrix}~\textbf{in}~Y_{s}
-% \]
-% where
-% \[
-%   Q ::= \langle RegExp \rangle \otimes Y ~|~ Y ~|~ Q \oplus Q
-% \]
-% \[
-%   RegExp ::= c ~|~ I ~|~ RegExp \oplus RegExp ~|~ RegExp \otimes RegExp ~|~ RegExp^{*}
-% \]
-
-% This is shown inductively by showing that the equivalence of a GNFA to a copy of itself with a general state --- i.e.\ non-start, non-accepting --- removed.
-Consider GNFA $A$ and remove state $Y_{r}$ to achieve $A'$.
-Let $\delta_{Y}$ denote the substitution that replaces $Y_{k}'/Y_{k}$ for $k \neq r$, and $\gamma$ denote the substitution $\hat{Y}_{r}/Y_{r}$.
-
-% \[
-%   A = \mu \begin{pmatrix}
-%             Y_{s} = \hat{Y}_{s} \\
-%             Y_{r} = \hat{Y}_{r} \\
-%             Y_{j} = \hat{Y}_{j}, j \not \in \{s, r, e\} \\
-%             Y_{e} = I
-%           \end{pmatrix}~\textbf{in}~Y_{s}
-% \]
-% \[
-%   A' = \mu \begin{pmatrix}
-%             Y_{s}' = \simulsubst {\hat{Y}_{s}} {\delta_{Y} \circ \gamma} \\
-%             Y_{j}' = \simulsubst {\hat{Y}_{j}} {\delta_{Y} \circ \gamma}, j \not \in \{s, e\} \\
-%             Y_{e}' = I
-%           \end{pmatrix}~\textbf{in}~Y_{s}'
-% \]
-
-% We can pretty easily translate between parses of $A$ and $A'$.
-This can be seen by looking at the structure of the terms $\hat{Y}_{k}$ and $\simulsubst {\hat{Y}_{k}} {\delta_{Y} \circ \gamma}$.
-For terms $\hat{Y}_{k}$ that don't have $Y_{r}$ as a subterm, the action of $\gamma$ is trivial --- so $\simulsubst {\hat{Y}_{k}} {\delta_{Y} \circ \gamma} = \simulsubst {\hat{Y}_{k}} {\delta_{Y}}$.
-For these $Y_{k}$ it is trivial to define a fold from $A$ to $A'$, or vice versa as we are simply renaming terms.
-For $\hat{Y}_{k}$ containing $Y_{r}$ as a subterm, it is not as trivial.
-Instead, each occurence of $Y_{r}$ is replaced with an equivalent definition.
-
-% Folding from $A$ to $A'$,
-
-% \begin{itemize}
-%   \item $y_{j}, j \not \in \{r, e\}$ and $Y_{r} \not \in \hat{Y}_{k}$ --- $\simulsubst {\hat{Y}_{k}} {\delta_{Y} \circ \Gamma} = \simulsubst {\hat{Y}_{k}} {\delta_{Y}}$ and $p_{s} = y_{j} : Y_{j}'$
-%   \item $y_{j}, j \not \in \{r, e\}$ and $Y_{r} \in \hat{Y}_{k}$ ---
-%   \item $y_{e}$ --- $\simulsubst I {\delta \circ \gamma} = I$ and $p_{e} = y_{e} : I$
-% \end{itemize}
-
-% \steven{An aside on some axioms}
-
-% Something like $A \cong (I \& A) \oplus \sum_{c} c \otimes (c \lto A)$
-
-% Also add the axiom that $\top \cong (\sum_{c \in \mathcal{A}} c)^{*}$
-
-% If we have the above, we can then show the following isos for some alphabet $\mathcal{A}$,
-% \begin{align*}
-%   A
-%   & \cong \top \amp A \\
-%   & \cong (\Sigma_{c \in \mathcal{A}})^{*}\amp A \\
-%   & \cong (I \oplus (\Sigma_{c} c \otimes \top)) \amp A \\
-%   & \cong (I \amp A) \oplus ((\Sigma_{c} c \otimes \top)) \amp A) \\
-%   & \cong (I \amp A) \oplus (\Sigma_{c }c \otimes (c \lto A)) & \text{using } \phi^{-1}
-% \end{align*}
-
-% From here we need to also show that derivatives of regular languages are regular (look at Brzozowski's proof) and that regular languages have finitely many derivatives (potentially with some size constraint on the derivative?).
-Then the last iso in the chain above can be used to build a DFA equivalent to $A$ using Brzozowski's algorithm.
-
-% Need to also add in axioms for $\amp$ distributing over $\oplus$ and $\lto$ distributing over $\oplus$. And that $c \amp c' \vdash 0$ for $c \neq c'$
-
-\subsection{Brzozowski Derivatives and Deterministic Finite Automata}
-\steven{this section sounds awkward and will be adjusted in a later editing pass}
-A \emph{deterministic finite automaton} (DFA) is an NFA whose transitions are deterministic.
-Meaning at every state $q \in Q$ there is exactly one transition for every character $c \in \Sigma$ --- i.e.\ the size of $\delta(q, c)$ is 1 and $\delta(q, \epsilon)$ is empty.
-It is clear that all DFAs happen to be NFAs, but it is perhaps surprising that the class of DFAs is as powerful as the class of NFAs when measured by the languages they recognize!
-\cref{thm:thompson} shows that the class of languages recognized by NFAs is exactly the regular languages.
-As we will soon see, the class of languages recognizable by DFAs is also the regular languages!
-
-There are a couple famous constructions we might use to show this, such as the powerset construction \cite{?} or Brzozowski's algorithm \cite{?}.
-Just as with Thompson's construction, we shall internalize a version of the latter as a proof term in our formal calculus.
-However, note that the construction of a DFA is lossy in some sense.
-That is, throughout this process we lose the ability to form an isomorphism between parses of a DFA for $g$ and parses of $g$ itself.
-Rather than express an isomorphism (strong equivalence) as a proof term, we shall construct a weak equivalence.
-This means we shall derive entailments $g \vdash DFA(g)$ and $DFA(g) \vdash g$ that do not necessarily invert each other.
-Semantically, this notion of weak equivalence between grammars $g$ and $g'$ means that $g$ and $g'$ both recognize the same language.
-
-Recall that a DFA is given as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$ in the same manner as an NFA.\ When encoding a DFA as a mutually recursive grammar, we use nearly the same process as when we encoded an NFA, with one subtle change.
-When we encoded NFAs, we were able to ensure we had a single accepting state by connecting all of the accepting states $q_{j} \in F$ to a single accepting state $Y_{e}$ with $\epsilon$-transitions.
-However, $\epsilon$-transitions are not permitted in a DFA, as they introduce a source of nondeterminism.
-Therefore we need a scheme to encode which states are accepting.
-The change we need to make is pretty lightweight, we add the grammar $I$ as a summand to distinguish which states are accepting.
-
-\[
-  DFA ::= \mu \begin{pmatrix}
-        Y_{s} = \langle \hat{Y} \rangle \\
-        Y_{j} = \langle \hat{Y} \rangle, j \not \in \{s, e\} \\
-        \end{pmatrix}.Y_{s}
-\]
-where
-\[
-\hat{Y} ::= \hat{Y_{0}} ~|~ I \oplus \hat{Y_{0}}
-\]
-\[
-  \hat{Y_{0}} ::= c \otimes Y ~|~ Y ~|~ \hat{Y_{0}} \oplus \hat{Y_{0}}, Y \in \{Y_{s}, Y_{1}, \dots, Y_{N}\}
-\]
-
-A state $Y_{j}$ in a DFA is accepting if and only if $\hat{Y}_{j}$ contains $I$ as a summand.
-
-some fluff to lead into the thoerem
-
-\begin{theorem}
-  For every regular grammar $g$, there is a DFA $D$ that is weakly equivalent to $g$.
-\end{theorem}
-
-\begin{proof}
-  \steven{This is currently word salad}
-
-  DFA construction
-
-  $DFA(g) \vdash g$ : just fold over the defs in $DFA(g)$ and have every $Y_{w}$ eliminate to $w \lto g$.
-This is doable bc when eliminating $Y_{w'}$ you will hold terms of type $a \otimes (w \lto g)$.
-These $w \lto g$ terms might not be exactly what you want, but they are weakly equivalent to some derivative that you do want!
-How do you know that they're weakly equivalent?
-if $Y_{w'}$ transitions to $Y_{w}$ via character $a$, then we know that $a \lto (w \lto g)$ is weakly equivalent to $a \lto (w' \lto g)$ \emph{by construction of the DFA}.
-This argument entirely hinges on the method by which we defined the DFA
-
-  $g \vdash DFA(g)$
-  Idea is to us $\top \cong \text{Char}^{*}$ and $A^{*} \cong \Sigma(n) A^{\otimes n}$ and $A \cong \top \amp A$ to turn a parse into a pair of a parse and a finite string.
-And then this finite string can be walked through the DFA to read off the parse, and you can show its accepting.
-Details currently a little shaky, and encapsulated in a bad picture of a whiteboard.
-\steven{Gonna has this out with Max in meeting}
-\end{proof}
-
-\subsection{Indexed Automata}
-
-We can generalize to more powerful notions of automata by utilizing
-dependency of linear types on non-linear types. Let $S$ be a finite
-non-linear type and define $\textrm{List}\,S = \mu x. 1 + S \times
-x$. Then a pushdown automaton with $S$ as the stack alphabet can be
-defined as a mutual fixed point.
-
-\steven{Forms for PDAs and Turing Machines go here}
-
-Want to put PDA, CFGs, CSGs here. Basically, use this section to highlight the usefulness of the intuitionistic part of the contexts
-
-\section{Future work}
-
-\subsection{Implementation}
-
-\subsection{Beyond Strings}
-
-While parsing typically refers to the production of semantic objects
-from their description as strings, many tasks in programming can be
-viewed as parsing of more structured objects such as trees, trees with
-binding structure or graphs. Fundamental to programming language
-implementation is \emph{type checking}, analogous to language
-recognition, or more generally \emph{typed elaboration}, analogous to
-parsing, which produces a semantic object in addition to performing
-some analysis. A type system could then be interpreted as a \emph{tree
-grammar}. Our definition of string grammars as functors from
-$\Sigma^*$ to $\Set$ can then be easily adapted to functors to $\Set$
-from the set of \emph{trees} freely generated form some signature of
-nodes. Then just as the set of string grammars inherits a monoidal
-structure from the free monoid of strings using the Day convolution,
-this set of tree grammars inherits a kind of weak algebraic structure
-corresponding to the tree constructors. This suggests an even more
-unusual form of bunched type theory, where the monoidal concatenation
-and unit are replaced with context constructors corresponding to the
-tree constructors.
-
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{references}
-
-\clearpage
-\appendix
-
-\section{Semantics in Sets and Grammars}
-
-A model of Grammar Type Theory is a category with families equipped
-with a functor $L : \mathcal C \to \textrm{GrLog}$ from the category
-of contexts to the category of grammar logic models and strict morphisms.
-
-First, we give the set theoretic model. We use the standard category
-with families structure given by the category of sets and a cumulative
-tower of universes.
-%
-Next, we define $L$ as follows:
-\[ L(\Gamma) = \mathcal P \Sigma^*/\textrm{Disc} \Gamma \]
-Where $\textrm{Disc} \Gamma$ is the presheaf that is constantly $\Gamma$.
-Note that
-\[ \mathcal P \Sigma^*/\textrm{Disc} \Gamma \cong (\Set/\Gamma)/\textrm{Disc} \Sigma^*\]
-where $\textrm{Disc} \Sigma^*$ is the projection $\Gamma \times \Sigma^*
-\to \Gamma$.  Then by the fundamental theorem of topos thoery, $\Set/\Gamma$ is
-a topos. Since $\textrm{Disc} : \Set \to \Set/\Gamma$ is monoidal,
-$\textrm{Disc} \Sigma^*$ carries a monoid structure in $\Set/\Gamma$ and
-this induces a monoidal closed structure on $(\Set/\Gamma)/\textrm{Disc}
-\Sigma^*$. Again by the fundamental theorem, it also has finite
-products and coproducts and a natural numbers object, and so can
-interpret least fixed points.\max{More details about constructing initial algebras?}
-
-Then we can define the remaining type structure as follows. Unraveling
-the above, a non-linear type $\Gamma \vdash X$ is interpreted as an
-object of $\Set/\sem{\Gamma}$ and a linear type $\Gamma \vdash A$ is
-interpreted as an object of $\mathcal P\Sigma^*/\textrm{Disc}\sem{\Gamma}$. We
-define them below in indexed style:
-\begin{enumerate}
-\item $\sem{c} \gamma w = (w = c)$
-\item $\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-\item $\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-\item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
-\item $\sem{L_i}\gamma = (\mathcal P_i\Sigma^*/\textrm{Disc}\sem{\Gamma})_0$
-\end{enumerate}
-
-\section{Canonicity}
-
-For the canonicity argument, we construct a second ``glued'' model
-that contains a projection morphism to the syntactic model.
-
-Let $S$ be the syntactic category with families for the type
-theory. Define the glued category with families to be the comma category
-% https://q.uiver.app/#q=WzAsNSxbMCwwLCJcXG1hdGhjYWwgRyJdLFswLDIsIlMiXSxbMiwyLCJcXFNldCJdLFsyLDAsIlxcU2V0Il0sWzEsMSwiXFxzd2Fycm93Il0sWzMsMl0sWzEsMl0sWzAsMV0sWzAsM11d
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{\gluedNL} \&\& \Set \\
-	\& \rotatebox[origin=c]{-45}{$\Downarrow$} \\
-	S \&\& \Set
-	\arrow[Rightarrow, no head, from=1-3, to=3-3]
-	\arrow["{S(\cdot,-)}"', from=3-1, to=3-3]
-	\arrow[from=1-1, to=3-1]
-	\arrow[from=1-1, to=1-3]
-\end{tikzcd}\]
-This is a standard glued category for canonicity of dependent type
-theory, and carries a category with families structure that we will
-reuse from prior work for the (non-linear) universes,
-$\Pi/\Sigma/\textrm{Id}$ \cite{coquand}. Concretely an object of
-$\mathcal G$ is equivalent to a pair of a syntactic context $\Gamma$,
-and a family $\hat \Gamma$ over (equivalence classes of) closing
-substitutions for $\Gamma$.
-
-Next we turn to interpreting the linear types and morphisms. For this
-we define a functor $\gluedL : \gluedNL \to \textrm{MonCat}$ where
-$\textrm{MonCat}$ is the category\footnote{in univalent foundations
-this would be the category of \emph{strict} monoidal categories.} of
-monoidal categories and lax monoidal functors.  We construct
-$\gluedL(p : \hat \Gamma \to S(\cdot,\Gamma))$ as the following comma
-category:
-
-% https://q.uiver.app/#q=WzAsNixbMCwwLCJcXGdsdWVkTkwocCkiXSxbMCwyLCJMX1xcR2FtbWEiXSxbMSwyLCJcXG1hdGhjYWwgUCBcXFNpZ21hXiovXFx0ZXh0cm17RGlzY30oUyhcXGNkb3QsXFxHYW1tYSkpIl0sWzIsMiwiXFxtYXRoY2FsIFAgXFxTaWdtYV4qL1xcdGV4dHJte0Rpc2N9KFxcaGF0XFxHYW1tYSkiXSxbMiwwLCJcXG1hdGhjYWwgUCBcXFNpZ21hXiovXFx0ZXh0cm17RGlzY30oXFxoYXRcXEdhbW1hKSJdLFsxLDEsIlxccm90YXRlYm94W29yaWdpbj1jXXstNDV9e1xcRG93bmFycm93fSJdLFs0LDMsIiIsMCx7ImxldmVsIjoyLCJzdHlsZSI6eyJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsNF0sWzAsMV0sWzEsMiwiTiIsMl0sWzIsMywiXFx0ZXh0cm17RGlzY30ocF4qKSIsMl1d
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{\gluedNL(p)} \&\& {\mathcal P \Sigma^*/\textrm{Disc}(\hat\Gamma)} \\
-	\& {\rotatebox[origin=c]{-45}{$\Downarrow$}} \\
-	{L_\Gamma} \& {\mathcal P \Sigma^*/\textrm{Disc}(S(\cdot,\Gamma))} \& {\mathcal P \Sigma^*/\textrm{Disc}(\hat\Gamma)}
-	\arrow[Rightarrow, no head, from=1-3, to=3-3]
-	\arrow[from=1-1, to=1-3]
-	\arrow[from=1-1, to=3-1]
-	\arrow["N"', from=3-1, to=3-2]
-	\arrow["{\textrm{Disc}(p^*)}"', from=3-2, to=3-3]
-\end{tikzcd}\]
-
-Where $N : L_\Gamma \to {\mathcal P
-  \Sigma^*/\textrm{Disc}(S(\cdot,\Gamma))}$ is the ``dependent nerve''
-of the inclusion $\lceil\cdot\rceil : \Sigma^* \to L_\Gamma$ of
-strings into linear contexts defined in indexed style as
-\[ N(\Gamma \vdash A) \gamma w = \lceil w \rceil \vdash \cdot :  \simulsubst A \gamma \]
-Then observe that $N$ and ${\textrm{Disc}(p^*)}$ are lax monoidal and
-${\mathcal P \Sigma^*/\textrm{Disc}(\hat\Gamma)}$ is a topos so the
-criteria for Hyland and Schalk's gluing construction apply, defining a
-biclosed monoidal structure on $\gluedNL(p)$ as well as finite
-(co-)products. Further, these constructions are all stable under
-reindexing in $\gluedNL$, and so interpret the dependent linear typing
-constructions for $I,\otimes,0,\oplus,\top,\amp,\lto,\tol$.
-
-Concretely, an object of $\gluedNL(p)$ is equivalent to a pair of a
-non-linear type $\Gamma \vdash A$ and a formal grammar $\hat A$
-indexed by $\hat \Gamma$ with a projection function from $\hat A$ to
-$N(A)$. We can think of this presheaf $\hat A$ as the presheaf of
-canonical forms of $\simulsubst A \gamma$ for closing substitutions
-$\gamma$ in contexts of the form $\lceil w \rceil$.
-
-All that remains in this model is to define the semantics of the
-non-standard connectives. We define them in an indexed style. First we
-define the interpretation of a non-linear type $\Gamma \vdash X$ as a
-family of sets $\hat X$ indexed by a closing substitution $\gamma :
-S(\cdot,\Gamma)$, a semantic substitution $\hat \gamma : \hat
-\Gamma(\gamma)$ and a closed term $\cdot \vdash e : \simulsubst A
-\gamma$. Second, we define a linear type $\Gamma \vdash A$ as a family
-$\hat A$ of grammars indexed by $\gamma,\hat\gamma$ as above along
-with a morphism $\pi : \hat A \gamma \hat \gamma \to N(\simulsubst A
-\gamma)$.
-
-\begin{enumerate}
-\item $\hat{c} \gamma \hat \gamma w = (w = c)$ and $\pi(*) = \lceil c \rceil = x_0:c \vdash x_0:c$
-\item $\widehat{(\LinSigTy x X A)} \gamma \hat \gamma w = \Sigma(\cdot \vdash e : \simulsubst X \gamma)\Sigma(\hat x: \hat X \gamma \hat \gamma e) \hat{A}(\gamma,e)(\hat\gamma,\hat x) w$
-\item $\sem{\LinPiTy x X A} \gamma w = \Sigma(x:\simulsubst X \gamma;\lceil w\rceil \vdash e:\simulsubst A \gamma)\Pi(\cdot \vdash e' : \simulsubst X \gamma)\Pi(\hat x:\hat{X}\gamma\hat\gamma x)\Sigma(\hat a : \hat A (\gamma,e')(\hat\gamma,\hat x) \times (\pi(\hat a) = \subst e {e'} x))$
-  and $\pi(e,f) = \dlamb x.e$
-\item $\widehat{G A} \gamma \hat \gamma e = \Sigma(\hat a: \hat{A} \gamma \hat\gamma \varepsilon) (p\gamma\hat\gamma \varepsilon(\hat a) = \mathsf{G}^{-1}e)$
-\item $\sem{L_i}\gamma = \gluedNLUniv(\gamma)$
-\end{enumerate}
-
-Thus we have defined a second model $\gluedNL,\gluedL$ of the type
-theory, with a morphism to the syntactic model. By the initiality of
-the syntactic model, there is a morphism back to the glued model which
-is a section of this projection. For purely positive types, we can
-observe by induction on the type structure that the total space of the
-family is isomorphic to the set-theoretic semantics. Further, the
-projection above defines a section of the families. Finally, by
-induction on formulae again we see that this projection is also a
-section and therefore an isomorphism, establishing the canonicity
-result as stated in Section\ref{section:canonicity}.
-
-\section{Equivalence Betweeen Regular Grammars and Finite Automata}
-\label{sec:appendixnfa}
-
-Below we finish the proof of \cref{thm:thompson}.
-\begin{proof}
-
-  First, we finish the provide the proof that the maps $s_{\oplus}$ and $r_{\oplus}$ provided in the disjunctive case do indeed mutually invert each other.
-For brevity in this proof, we will omit the list of mutually recursive bindings when appropriate.
-That is, write $\mu X_{s}'$ instead of $\mu(Y_{s},X_{s}',X_{j}',\dots).
-X_{s}'$.
-
-  Given $e : NFA(g \oplus g')$,
-
-  \steven{There a weird spacing issues between these align envs. Maybe wrap these in a figure env so that spacing is better}
-
-\begin{align*}
-  r_{\oplus}(s_{\oplus}(e))
-  & =_{\mu \eta} r_{\oplus}(s_{\oplus}(\mathsf{cons}(e'))), \text{~where }e':\mu X_{s}' \oplus \mu Z_{s}' \\
-  &
-    \begin{aligned}
-      =_{\oplus \eta} r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{case}~e'~\{ & \\
-    \textbf{in}_{\mu X_{s}'} u \mapsto \textbf{in}_{\mu X_{s}'} u, & \\
-    \textbf{in}_{\mu Z_{s}'} v \mapsto \textbf{in}_{\mu Z_{s}'} v
-    \})))
-    \end{aligned} \\
-  &
-    \begin{aligned}
-    =_{\oplus \eta}
-    \textbf{case}~e'~\{ & \\
-    & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{in}_{\mu X_{s}'} u))), \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v))) \\
-    \}))
-    \end{aligned} \\
-  &
-    \begin{aligned}
-    =
-    \textbf{case}~e'~\{ & \\
-    & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(\mathsf{mfold}(\mathsf{cons}(\textbf{in}_{\mu X_{s}'} u))), \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(\mathsf{mfold}(\mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v))) \\
-    \}
-    \end{aligned} \\
-  &
-    \begin{aligned}
-    =_{\mu \beta}
-    \textbf{case}~e'~\{ & \\
-    & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(\textbf{in}_{g} \phi^{-1}(\mathsf{cons}_{\mu(X_{i}=\hat{X}_{i}).X_{s}}u')), \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(\textbf{in}_{g'} \psi^{-1}(\mathsf{cons}_{\mu(Z_{i}=\hat{Z}_{i}).Z_{s}}v')) \\
-      \} \\
-      & \text{~for }u' : \mu X_{s}', v' : \mu Z_{s}'
-    \end{aligned} \\
-  &
-    \begin{aligned}
-    =_{\oplus \beta}
-    \textbf{case}~e'~\{ & \\
-    & \textbf{in}_{\mu X_{s}'} u \mapsto \alpha \phi \phi^{-1}(\mathsf{cons}(u')), \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto \beta \psi \psi^{-1}(\mathsf{cons}(v')) \\
-      \}
-    \end{aligned} \\
-  &
-    \begin{aligned}
-    =
-    \textbf{case}~e'~\{ & \\
-    & \textbf{in}_{\mu X_{s}'} u \mapsto \alpha (\mathsf{cons}(u')), \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto \beta (\mathsf{cons}(v')) \\
-      \}
-    \end{aligned} \\
-   &
-    \begin{aligned}
-    =_{\mu \beta}
-    \textbf{case}~e'~\{ & \\
-    & \textbf{in}_{\mu X_{s}'} u \mapsto \mathsf{cons}(\textbf{in}_{\mu X_{s}'} u) , \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto \mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v) \\
-      \}
-    \end{aligned} \\
-  & =_{\oplus \eta} \mathsf{cons}(e') \\
-  & =_{\mu \beta} e
-\end{align*}
-
-Likewise for $p : g \oplus g'$,
-
-\steven{This first line is a weird alignment fix, touch up}
-\begin{align*}
-  s_{\oplus}(r_{\oplus}(p))
-  &
-  = s_{\oplus}(r_{\oplus}(p)) \\
-   & \begin{aligned}
-    =_{\oplus \eta}
-      \textbf{case}~p~\{
-    \textbf{in}_{g}p_{1} \mapsto s_{\oplus}(r_{\oplus}(\textbf{in}_{g}p_{1})), \\
-    \textbf{in}_{g'}p_{2} \mapsto s_{\oplus}(r_{\oplus}(\textbf{in}_{g'}p_{2})) \}
-    \end{aligned} \\
-   & \begin{aligned}
-    =_{\oplus \beta}
-      \textbf{case}~p~\{
-    \textbf{in}_{g}p_{1} \mapsto s_{\oplus}(\alpha \phi p_{1}), \\
-    \textbf{in}_{g'}p_{2} \mapsto s_{\oplus}(\beta \psi p_{2}) \}
-    \end{aligned} \\
-   & \begin{aligned}
-    =
-      \textbf{case}~p~\{
-    \textbf{in}_{g}p_{1} \mapsto \mathsf{mfold}(\alpha \phi p_{1}), \\
-    \textbf{in}_{g'}p_{2} \mapsto \mathsf{mfold}(\beta \psi p_{2}) \}
-    \end{aligned} \\
-    & \begin{aligned}
-    =_{\mu \beta}
-      \textbf{case}~p~\{
-    \textbf{in}_{g}p_{1} \mapsto \textbf{in}_{g} (\phi^{-1} \phi p_{1}), \\
-    \textbf{in}_{g'}p_{2} \mapsto \textbf{in}_{g'}(\psi^{-1} \psi p_{2}) \}
-    \end{aligned} \\
-     & \begin{aligned}
-    =
-      \textbf{case}~p~\{
-    \textbf{in}_{g}p_{1} \mapsto \textbf{in}_{g} (p_{1}), \\
-    \textbf{in}_{g'}p_{2} \mapsto \textbf{in}_{g'}(p_{2}) \}
-    \end{aligned} \\
-  &
-    =_{\oplus \eta} p
-\end{align*}
-
-We now construct a combinator for the sequencing of two NFAs.
-
-\[
-  NFA(g \otimes g') = \mu
-  \begin{pmatrix}
-    Y_{s} = X_{s}' \\
-    X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
-    X_{e}' = Z_{s}' \\
-    Z_{j}' = \simulsubst {\hat{Z}_{j}} {\delta_{Z}}, j \neq e \\
-    Z_{e}' = Y_{e} \\
-    Y_{e} = I
-  \end{pmatrix}
-\]
-
-Just as before, we use a multifold to define the map $s_{\otimes}$ from $NFA(g \otimes g')$ to $g \otimes g'$. Here the table of terms is as follows,
-
-\begin{itemize}
-  \item $y_{s}$ --- $\simulsubst {X_{s}'} \gamma = g \otimes g'$ and $p_{s} = y_{s} $
-  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = (\mu (X_{i}=\hat{X}_{i}).X_{s}) \otimes g' = NFA(g) \otimes g'$ and $p_{x,s} = \letin {u \otimes v} {x_{s}'} {\phi^{-1}(u) \otimes v} : g \otimes g'$
-  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = (\mu (X_{i}=\hat{X}_{i}).X_{j}) \otimes g'$ and $p_{x, j} = x_{j}' : (\mu (X_{i}=\hat{X}_{i}).X_{j}) \otimes g'$
-  \item $x_{e}'$ --- $\simulsubst {Z_{s}'}{\gamma} = g'$ and $p_{x,e} = x_{e}' : g'$
-  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = \mu (Z_{i} = \hat{Z}_{i}).Z_{s} = NFA(g')$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
-  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = \mu (Z_{i} = \hat{Z}_{i}).Z_{j}$ and $p_{z, j} = z_{j}' : \mu (Z_{i} = \hat{Z}_{i}).Z_{j}$
-  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = \mathsf{cons}(z_{e}') : \mu (Z_{i} = \hat{Z}_{i}).Z_{e} = I$
-  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
-\end{itemize}
-
-The map $r_{\otimes}$ in the other direction is given as,
-
-\[
-  \inferrule{
-    p : g \otimes g' \vdash p : g \otimes g' \\
-    p_{1} : g, p_{2} : g' \vdash \mathsf{cons}(\phi(p_{1}), \psi(p_{2})) : NFA(g \otimes g')}
-  {p : g \otimes g' \vdash \letin {p_{1} \otimes p_{2}} {p} {\mathsf{cons}(\phi(p_{1}), \psi(p_{2}))} : NFA(g \otimes g')}
-\]
-
-We show that $r_{\otimes}$ and $s_{\otimes}$ form an isomorhic pair,
-\steven{This proof doesn't exactly work}
-
-\begin{align*}
-r_{\otimes}(s_{\otimes}(e))
-  & =_{\mu \eta} r_{\otimes}(s_{\otimes}(\mathsf{cons}_{\mu X_{s}'}(e'))) \\
-  & = r_{\otimes}(\mathsf{mfold}(\mathsf{cons}_{\mu X_{s}'}(e'))) \\
-  & =_{\mu \beta} r_{\otimes}(\phi^{-1}(u) \otimes \psi^{-1}(v)) \\
-  & =_{\mu \beta} \mathsf{cons}(\phi \phi^{-1} u, \psi \psi^{-1} v) \\
-  & =_{\mu \eta} e
-\end{align*}
-
-\begin{align*}
-  s_{\otimes}(r_{\otimes}(p))
-  & =_{\otimes \eta} s_{\otimes}(r_{\otimes}(p_{g} \otimes p_{g'})) \\
-  & = s_{\otimes}(\mathsf{cons}(\phi p_{g}, \psi p_{g'})) \\
-  & =_{\mu \beta} \phi^{-1}\phi p_{g} \otimes \psi^{-1} \psi p_{g'} \\
-  & = p_{g} \otimes p_{g'} \\
-  & =_{\otimes \eta} p
-\end{align*}
-
-Finally, we build an automata combinator to represent the Kleene star\footnote{Which is another admissible definition via use of the fixed point constructor}.
-
-\[
-  NFA(g^{*}) = \mu
-  \begin{pmatrix}
-    Y_{s} = X_{s}' \oplus Y_{e} \\
-    X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
-    X_{e}' = X_{s}' \oplus Y_{e} \\
-    Y_{e} = I
-  \end{pmatrix}
-\]
-
-Again we define $s_{*}$ as a fold over these defintions to get a parse of $g^{*}$,
-\begin{itemize}
-  \item $y_{s}$ --- $\simulsubst {(X_{s}' \oplus Y_{e})} \gamma = g^{*} \oplus I$ and $p_{s} = \textbf{case}~y_{s} \{ \textbf{in}_{g^{*}} u \mapsto u, \textbf{in}_{I} v \mapsto \mathsf{cons}_{*I} v \} : g^{*}$
-  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = (\mu(X_{i}=\hat{X_{i}}).X_{s}) \otimes g^{*} = NFA(g) \otimes g^{*}$ and $p_{x,s} = \letin {u \otimes p_{*}} {x_{s}'} {\mathsf{cons}_{*}(\phi^{-1}(u), p_{*})} : g^{*} $
-  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = (\mu(X_{i}=\hat{X_{i}}).X_{j}) \otimes g^{*}$ and $p_{x, j} = : (\mu(X_{i}=\hat{X_{i}}).X_{j}) \otimes g^{*}$
-  \item $x_{e}'$ --- $\simulsubst {(X_{s}' \oplus Y_{e})}{\gamma} = g^{*} \oplus I$ and $p_{x,e} = \mathsf{cons}_{X_{e}}(()) \otimes \textbf{case}~x_{e}' \{ \textbf{in}_{g^{*}} u \mapsto u, \textbf{in}_{I} v \mapsto \mathsf{cons}_{*I} v \} : (\mu(X_{i}=\hat{X_{i}}).X_{e}) \otimes g^{*}$
-  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
-\end{itemize}
-
-and to derive a parse of $NFA(g^{*})$ from a parse of $g^{*}$ we define $r_{*}$ as,
-
-\[
-  \inferrule{p : g^{*} \vdash p : g^{*} \\ \cdot \vdash \mathsf{cons}(\textbf{in}_{{Y_{e}}}(())) : NFA(g^{*}) \\ u : g, v : g^{*} \vdash \mathsf{cons}(\textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v)))) : NFA(g^{*}) }{p : g^{*} \vdash \mathsf{fold}_{*}(\mathsf{cons}(\textbf{in}_{{Y_{e}}}(())), \mathsf{cons}(\textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v)))) : NFA(g^{*}) }
-\]
-
-To see that $s_{*}$ and $r_{*}$ for an isomorphism,
-\steven{This proof also needs more work. Esp after 4th equality}
-\steven{Think I need to thread through a case distinction on what the $g^{*}$ term is. Putting a pin in this for now to type up the DFA stuff before meeting}
-\begin{align*}
-  r_{*}(s_{*}(e))
-  & =_{\mu \eta} r_{*}(s_{*}(\mathsf{cons}(e'))), \text{~where }e':\mu X_{s}' \oplus Y_{e} \\
-  &
-    \begin{aligned}
-      =_{\oplus \eta}
-      \textbf{case}~e'~\{ & \\
-      & \textbf{in}_{X_{s}'} u \mapsto r_{*}(s_{*}(\mathsf{cons}(\textbf{in}_{X_{s}'} u))), \\
-      & \textbf{in}_{Y_{e}} v \mapsto r_{*}(s_{*}(\mathsf{cons}(\textbf{in}_{Y_{e}} v))) \\
-      \}
-    \end{aligned} \\
-  &
-    \begin{aligned}
-      =_{\mu \beta}
-      \textbf{case}~e'~\{ & \\
-      & \textbf{in}_{X_{s}'} u \mapsto r_{*}(\mathsf{cons}_{*}(\phi^{-1} u, p_{*})), \\
-      & \textbf{in}_{Y_{e}} v \mapsto r_{*}(\mathsf{cons}_{*}(()) \\
-      \}
-    \end{aligned} \\
-  &
-    \begin{aligned}
-      =_{\mu \beta}
-      \textbf{case}~e'~\{ & \\
-      & \textbf{in}_{X_{s}'} u \mapsto \mathsf{cons}(\textbf{in}_{X_{s}'}(\phi(\mathsf{cons}_{*}(\phi^{-1}u,p_{*})))), \\
-      & \textbf{in}_{Y_{e}} v \mapsto \mathsf{cons}(\textbf{in}_{Y_{e}}(()) \\
-      \}
-    \end{aligned} \\
-  & = \dots \\
-  & = e
-\end{align*}
-
-\steven{for the empty case above, can I make an $\eta$ equality for all terms of type $I$ to argue $\textbf{in}(v) = \textbf{in} (())$}
-
-\begin{align*}
-  s_{*}(r_{*}(p))
-  & =_{\mu \eta} s_{*}(r_{*}(\mathsf{cons}(p_{e}, p_{*}))), \text{~where } p_{e} : I, p_{*} : g \otimes g^{*} \\
-  & =_{\mu \beta} s_{*}()
-  & = p
-\end{align*}
-
-This completes our inductive proof that $NFA(g) \cong g$ for all regular expressions $g$.
-
-\end{proof}
+\section{Automata}
 

### 2024-04-11 steven-prelim/paper.tex
paper sketch

--- 2024-03-18 POPL/paper.tex
@@
 % -*- fill-column: 80; -*-
-\documentclass[sigconf,review,screen]{acmart}
+\documentclass[sigconf,screen,nonacm]{acmart}
 \usepackage{mathpartir}
 \usepackage{tikz-cd}
+\usepackage{lipsum}
 \usepackage{enumitem}
@@
 
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-%% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
-  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
-\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-  June 03--05, 2018, Woodstock, NY}
-\acmPrice{15.00}
-\acmISBN{978-1-4503-XXXX-X/18/06}
-
 \begin{document}
 
-\title{Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}
+\title{CSE Preliminary Examination}
 \author{Steven Schaefer}
-\affiliation{
-  \department{Electrical Engineering and Computer Science}
+\affiliation{\department{Electrical Engineering and Computer Science}
   \institution{University of Michigan}
@@
 \email{stschaef@umich.edu}
+
+\maketitle
+\pagestyle{plain}
+
+\section*{Distribution of Work}
+\todoin{I need to describe what my contributions are, but it needs to sounds
+  more natrual than below}
+This work was conducted with Max New, one of my advisors, and Pedro H.A. de
+Amorim, a collaborator at Oxford. The majority of this paper is adapted from a
+draft written by Max, Pedro, and I for submission to LICS in January. We did not submit
+to LICS, and now we are targetting submission to POPL in July.
+
+This paper is an adaptation of my contributions to the LICS draft. In summary,
+these include the proofs presented, contribution to the formation of
+inference rules, choice of axioms, and an ongoing formalization effort in Agda of
+several of the included constructions.
+
+From the LICS draft I am omitting some work from Max and Pedro that may be
+relevant to this paper, although not directly necessary. Some of these omissions
+include a canonicity result for the type theory and a categorification of Kleene
+algebras.
+
+\todoin{I want to describe some cubical contributions as well, even if not
+  entirely relevant to this paper. Something
+  something resuable library code? After all, some of it is used in the formalization}
+
+\listoftodos{list of todos}
+
+\pagebreak
+
+\title{Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}
 
@@
 \begin{abstract}
-  We propose a semantic framework for the study of formal grammars. First, we
-  provide a syntax-independent notion of formal grammar as a function from
-  strings to sets of parse trees, generalizing the familiar notion of a language
+  In this paper, we
+  provide a syntax-independent notion of formal grammar. Semantically, we view a grammar as a function from
+  strings to sets of parse trees, generalizing from the familiar notion of a language
   as a set of strings. This point-of-view naturally endows a grammar with the
@@
   theory are interpreted as grammars --- and linear terms as parse transformers.
-  The non-dependent fragment of this type theory is already expressive enough to
-  capture regular expressions, context-free expressions, and finite automata.
-  Moreover, the isomorphism between regular languages and finite automata is
-  definable as a proof term in our formalism. The
-  non-dependent fragment of this type theory is expressive enough to capture
-  both regular and context-free expressions, as well as finite automata already
-  regular expressions, context-free expressions, and finite automata. Moreover,
-  the is definable as a proof term in this formalism. Further, through use of
-  dependency of linear types on non-linear types, we can express richer grammar
-  formalisms such as indexed grammars, pushdown automata, and Turing machines.
 
@@
   these related ideas and internalize their
-  theorems formally. We give a new form of \emph{logical characterization} of
+  theorems formally. We give a new \emph{logical characterization} of
   grammar classes based on a substructural logic rather than ordinary
-  first-order logic. Further, the type theory can not only
-  express the grammars as types, but equivalences between grammars and
-  parsers as terms of the grammar.
+  first-order logic.
+
+  The non-dependent fragment of this type theory is already expressive enough to
+  capture regular expressions, context-free expressions, and finite automata.
+  Moreover, several classic theorems --- such as the equivalence between regular languages and finite automata --- are
+  definable as proof terms in this formalism.
+  Further, through use of dependency of linear types on non-linear types, we can express richer grammar
+  formalisms such as indexed grammars, pushdown automata, and Turing machines.
 
@@
 \todoin{Rework intro from LICS draft with a focus on implementation}
+Formal language theory is among one of the oldest and most studied areas in
+theoretical computer science.
 
-\section{Kleene Categories}
-\todoin{Tell a story winding through the use of KAs, action algebras, etc to
-  motivate why we'd want categorical semantics for KCats}
-\todoin{Incoporate Kleene Category definition}
-\todoin{End with the punch that $\Set^{\Sigma^{*}}$ is a KCat}
+Informally, a grammar $G$ is often described generatively as a set of production rules for
+generating strings. Often this is presented as a finite set of nonterminal
+symbols $N$, a finite set of terminal symbols $\Sigma$, and a finite set of rewriting
+rules for substitution of nonterminals. In this setting, we continually apply
+rewrite rules until we are left only with a string of terminals
+$s \in \Sigma^{*}$. \todoin{cite chomsky}
 
-\section{Automata}
+One of the primary objects of study is that of a \emph{formal language} over
+a fixed alphabet $\Sigma$. A language $L$ is a \emph{subset} of strings
+$L \subseteq \Sigma^{*}$. This notion of formal language gives an intuitive
+set-based semantics for grammars. That is, we may interpret a grammar $G$ as the
+language $L(G)$ that is generated by it.
+
+\section{Non-commutative Linear-non-linear Type Theory as a Syntax for Formal Grammars}
+\missingfigure{Inference Rules}
+\missingfigure{Axioms}
+
+\section{Type Theoretic Formal Grammar Theory}
+
+\section{Automata as Grammars}
+
+\subsection{Finite Automata}
+
+\subsection{Equivalence Between Regular Grammars and Finite Automata}
+
+\subsection{Brzozowski Derivatives and Deterministic Finite Automata}
+\todoin{You should know complexity statements about all this classic stuff like
+  Brzozowski's alg}
+
+\subsection{Pushdown Automata}
+
+\subsection{Turing Machines}
+
+\section{Future Work}
+\subsection{Formalization}
+\todoin{proofs in agda}
+
+\subsection{Implementation of a Verfied Parser}
+\todoin{sell how this should be used for building verified code}
+\todoin{talk about the need for a verified parser because compcert yadda yadda}
+
+\subsection{Beyond Strings}
 

### 2024-04-11 steven-prelim/paper/paper.tex
steven prelim
NFA traces defined in paper

paper writing 4/12, mostly todos

some comments

emptynfa iso

literal nfa iso

nfa proof WIP

decprops are the worst

paper intro + presentation stub

syntax given

I desperately need to write the proofs now

more grammar examples

brzozowski derivatives added to prelim paper

Getting off of plane

max comments

longer intro

intro changes, automata inclusions

Spell check

still badly need to add to later sections. Must finish by Sun

fix typo

notes in paper and cover

Proof sketch progress

2nd half a lot better, still more to go

try to get a cou

% -*- fill-column: 80; -*-
\documentclass[acmsmall,screen,nonacm]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{lipsum}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{stmaryrd}
\usepackage{tikz}
\usetikzlibrary{automata, positioning, arrows, fit}

\usepackage{pdfpages}

\usepackage{todonotes}

\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\Sigma^{*}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\nat}{\mathbb{N}}

\newcommand{\gluedNL}{{\mathcal G}_S}
\newcommand{\gluedNLUniv}{{\mathcal G}_{S,i}}
\newcommand{\gluedL}{{\mathcal G}_L}

\newcommand{\simulsubst}[2]{#1\{#2\}}
\newcommand{\subst}[3]{\simulsubst {#1} {#2/#3}}
\newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
\newcommand{\lamb}[2]{\lambda #1.\, #2}
\newcommand{\lamblto}[2]{\lambda^{{\lto}} #1.\, #2}
\newcommand{\lambtol}[2]{\lambda^{{\tol}} #1.\, #2}
\newcommand{\dlamb}[2]{\overline{\lambda} #1.\, #2}
\newcommand{\app}[2]{#1 \, #2}
\newcommand{\applto}[2]{#1 \mathop{{}^{\lto}} #2}
\newcommand{\apptol}[2]{#1 \mathop{{}^{\tol}} #2}
\newcommand{\PiTy}[3]{\Pi #1 : #2.\, #3}
\newcommand{\SigTy}[3]{\Sigma #1 : #2.\, #3}
\newcommand{\LinPiTy}[3]{\widebar\Pi #1 : #2.\, #3}
\newcommand{\LinSigTy}[3]{\widebar\Sigma #1 : #2.\, #3}
\newcommand{\amp}{\mathrel{\&}}
\newcommand{\GrTy}{\mathsf{Gr}}

\newcommand{\ctxwff}[1]{#1 \,\, \mathsf{ok}}
\newcommand{\ctxwffjdg}[2]{#1 \vdash #2 \,\, \mathsf{type}}
\newcommand{\linctxwff}[2]{#1 \vdash #2 \,\, \mathsf{ok}}
\newcommand{\linctxwffjdg}[2]{#1 \vdash #2 \,\, \mathsf{linear}}

\newif\ifdraft
\drafttrue
\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven says}: #1]}\fi}
\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max says}: #1]}\fi}
\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro says}: #1]}\fi}
\newcommand{\pipe}{\,|\,}

\newcommand{\todoin}[1]{\ifdraft{\todo[inline]{TODO:\@ #1}}\fi}

\begin{document}

\pagestyle{plain}

\includepdf{cover.pdf}

\pagebreak

\ifdraft{
\listoftodos{list of todos}
\pagebreak}\fi

\title{Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}

\author{Steven Schaefer}
\affiliation{\department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{stschaef@umich.edu}

\author{Max S. New}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{maxsnew@umich.edu}

\author{Pedro H. Azevedo de Amorim}
\affiliation{
  \department{Department of Computer Science}
  \institution{University of Oxford}
  \country{UK}
}
\email{pedro.azevedo.de.amorim@cs.ox.ac.uk}

\makeatletter
\let\@authorsaddresses\@empty
\makeatother

\begin{abstract}
  In this paper, we give a type system in which we can
  internally carry out many classical constructions from
  the theory of formal grammars. Inspired by the common
  structure underlying various ideas in formal language
  theory, our type theory makes these relationships explicit
  and provides a unifying syntax to reason about formal
  grammars. We discuss application to the design and
  implementation of correct-by-construction parsers for use
  in real-world compilers.
\end{abstract}

\maketitle

\setcounter{page}{1}

\section{Introduction}
% idk some intro about verification in general, then build up to verifying compilers, and ultimately parsers
In the realm of software development, ensuring the correctness and reliability
of code is of the utmost importance. In practice, developers often write tests
to gain confidence that they have written a program correctly; however,
program correctness cannot be guaranteed through testing alone. Where testing is
incomplete, we may further use formal methods for generating mathematical proofs of
correctness for our software.

Importantly, any proof of correctness is relative to the formal specification given
for the program. For instance, even if one provides a verified artifact for
their system's source code, it is not a guarantee that this system will run as
expected on real machines because it is not taking into account hardware
semantics.
The developer puts a large degree of implicit trust in
their compiler to generate a binary that reflects the intended behavior of their
source code. To bridge this gap, the compilers themselves may be verified, such
as CompCert\cite{leroy_formal_2009} --- a verified C compiler ---
and CakeML\cite{kumar_cakeml_2014} --- a verified implementation of an ML-like language.

Unfortunately even the apparent assurances associated with
this verification were not enough, as even early versions of CompCert still
had bugs\cite{yangFindingUnderstandingBugs}. Despite significant effort
verifying the middle- and backends of their C compiler, Leroy et al.\ did not
initially prove the correctness of their compiler frontend. When considering
machine traces, the presence of
these parsing bugs can jeopardize the guarantees provided by other verification
efforts. That is, verification exists on a stack of trust, and even one break in
the chain of verification can lead to incorrect behavior of real machines.

Several years after its release, the CompCert project was later updated to use a
verified parser\cite{jourdanValidatingLRParsers2012}. This parser was generated
by Menhir --- an LR parser generator --- and verified
post-hoc. After-the-fact
verification efforts like this are
time-intensive and difficult to generalize. Significant person-hours are
invested in managing the idiosyncrasies of the parser-generator being used and
even the particular grammar describing the
language.

To aid in the development of verified parsers, the primary contribution
of this paper is
an type
theory whose types natively capture the semantics of formal
grammars. Types in this theory are grammars, and terms are
parse transformers. This kind of type theory naturally
requires (non-commutative) linear and dependent types.
Within this formalism, we may define parsers as terms
defined in the context of an input string. Moreover, by virtue of
typechecking, these parsers are \emph{correct-by-construction}. The intrinsic
verification gained from the type system therefore removes the need for a
post-hoc verification in a language like Coq, thus lowering the barrier to
compiler frontend verification.

Substructural logics, like linear and separation logic, offer precise control
over how values from a type may be used, in particular how many times a value
may be used in a derivation. This control enables \emph{resource-sensitive}
reasoning, where we may treat a resource as \emph{consumed} after usage. In the
context of parsing, we may treat the characters composing a string as finite resources
that are consumed at parse-time.

Dependent type theory extends the simply-typed $\lambda$-calculus by allowing type
definitions to contain the occurrence of a term. In lieu of set-based
foundations, this extension can then be taken as a foundation for mathematics.
The expressive power unlocked by this insight has enabled the development of
dependently-typed proof assistants
such as Coq or Agda. Within each of these languages, systems may be given very
precise functional specifications of their behavior.

In 1995, Benton et al.
\cite{bentonMixedLinearNonlinear1995} first gave categorical models of a logic
that synthesized linear and non-linear logic. Later, Krishnaswami et al.
\cite{krishnaswami_integrating_2015} built upon this work to develop a type
system that unifies both linear and
dependent types --- referred to as a \emph{linear-non-linear} (LNL) type theory. In
such settings, we have two sub-theories --- one linear and one non-linear. The
linear sub-theory embeds ordinary linear types without dependence, while the
non-linear sub-theory embeds standard Martin-L\"of type theory. The interplay of
linear types with dependence lives at the boundary between these two fragments.
Benton et al.\ gives an accounting of how we may jump between these
fragments via a monoidal adjunction $F \dashv G$ between linear terms and
non-linear terms.

To the end of defining real verified
parsers, in our LNL theory we define a parser as a term,

\crefname{equation}{}{}
\begin{equation}
  \label{eq:parser}
  w : \String \vdash \mathsf{parse}_{g}(w) : g \oplus \top
\end{equation}

We will investigate the precise meaning of judgments like \cref{eq:parser} in
\cref{sec:tt}, but first let's get an intuitive feel for the meaning behind such a
term. On the left of the turnstile occurs a single string $w$, we may think of
this describing what resources are linearly available to us. Using each piece of
our the context on the left of the turnstile, we build a term
$\mathsf{parse}_{g}(w)$ of type $g \oplus \top$. Notably, the type being
instantiated is a sum type, and the data of which side we inject into denotes
whether the string is accepted or rejected by the parser. That is all to say, a
parser takes in an input string and either matches it to the desired grammar or
rejects the string.

Practical parsers are often implemented via an abstract automaton. For instance,
a parser may translate a grammar into a state machine recognizes the
same language as the original grammar, as the automata
point-of-view lends itself to straightforward execution. The
strength of our syntax is that it allows us to describe
these automata as grammars themselves. Moreover, the
equivalence between the grammar and automata presentations
is derivable as a proof term within the theory. This allows
us access to simultaneously internalization of the equivalence
between grammars and automata as well as the ability to
export to automaton implementations of a parser that are executable.

To realize an executable implementation of our type theory,
we shallowly embed it in Agda. Because our implementation is
defined as a function in Agda, we can leverage the Agda
ecosystem for code generation. Further, in lieu of
implementing a complete language from scratch, we make use of the
Agda proof system as a setting to guarantee correctness of
our generate parsers.

Within this paper, we give our progress in defining a verified
parser for regular expressions. The construction of such a parser entails
the definition of a parser for deterministic finite automata (DFA) grammars,
and an equivalence proof between DFAs and regular expressions. We have thus far
completed the construction of a parser for DFAs $D$,

\[
  w : \String \vdash \mathsf{parse}_{D}(w) : D \oplus \top
\]

and half of the equivalence proof between DFAs and regular expressions, both
formalized in Agda. Upon completion of the second half of this proof, we would
have a system where a user could simply give a regular grammar and then extract
an intrinsically verified parser for it. In future work, we
hope to show an analogous construction involving pushdown
automata (PDAs) and context-free grammars.

The contributions of this paper include an LNL theory for reasoning about formal
grammars, the encoding of automata as formal grammars in this theory, and the
definition of parser terms for DFA grammars. The contents include,

\begin{itemize}
  \item A syntax-independent definition of formal grammar
  \item The formal rules governing our type system
  \item Interpretations of the type system for reasoning about formal grammars
        and formal languages
  \item The presentation of automata as grammars
  \item The definition of a parser within our type system
\end{itemize}

\section{A Syntax-Independent Notion of Syntax}
\label{sec:synindsyn}

% \max{this section doesn't flow naturally from the intro, which is all about the
%   type theory and parsing. This is where you want to outline in the intro and
%   possibly say here that we want to revisit ``formal grammar theory'' and then
%   we will make a type theory that has formal grammar theory as its semantics.}

In this section we will revisit some foundational ideas in formal
grammar theory, before giving a type
theory for which formal grammars serves as its semantics.

Formal languages and grammars are among some of the oldest and most studied areas in
theoretical computer science. The developments of the 1950s and 1960s gave rise
to expressive formalism for reasoning about grammars, and further practical
algorithms for constructing parsers for regular and context-free grammars.

A grammar $G$ is often described generatively as a set of production rules for
generating strings. Often this is presented as a finite set of nonterminal
symbols $N$, a finite set of terminal symbols $\Sigma$, and a finite set of rewriting
rules for substitution of nonterminals. In this setting, we continually apply
rewrite rules until we are left only with a string of terminals
$w \in \String$. We will give a more precise notion of
grammar in a moment, but for now let us take this intuitive view of grammars as
string generators.

One of the primary objects of study is that of a \emph{formal language} over
a fixed alphabet $\Sigma$. A language $L$ is a \emph{subset} of strings
$L \subseteq \String$. This notion of formal language gives an intuitive
set-based semantics for grammars. That is, we may interpret a grammar $G$ as the
language $L(G)$ that is generated by it. We are yet to give any precise
definition of what we mean by ``grammar'', although we note that any suitable
notion of grammar can be given a language semantics as above.

Formal languages serve as a basis for mathematically specifying a
\emph{recognizer} --- a
function $r : \String \to \Bool$ that determines if a string belongs to a
given language. However, they are insufficient for specifying a \emph{parser}
--- a function that decomposes a string into semantic components that adhere to
the structure of a formal grammar. In practice, parsers don't just tell you
if a string is valid or invalid. Instead, they emit structured data such as
abstract syntax trees (ASTs).

This motivates our first conceptual contribution, a simple and
syntax-independent notion of formal grammar:

\begin{definition}
  \label{def:grammar}
  A \emph{formal grammar} $G$ over a fixed alphabet $\Sigma$ is a function
  $G : \String \to \Set$.
\end{definition}

We say that a grammar $G$ associates to every string $w$ the set\footnote{We
  make little comment as to which foundations we use to encode sets. We argue\max{drop ``We argue''}
  that our construction is agnostic to the particular foundation, and so without
  loss of generality\max{incorrect use of ``without loss of generality''} we may take a class of small sets, a Grothendieck universe,
  a type theoretic universe, or any similar foundation. Here we choose a type
  theoretic universe.} $G w$ of parse
trees showing that $w$ matches $G$.

The simplest class of formal grammar on the Chomsky\max{first mention of Chomsky hierarchy, why not bring it up earlier when you are describing what a grammar is? Also cite}
hierarchy are the \emph{regular expressions}

\max{inaccurate reflection of common terminology. These are regular expressions,
  regular grammars usually means a CFG with a restriction on recursion that
  ensures you only allow Kleene star.}.

\begin{gather*}
  \mathsf{Char} \in \Sigma \\
  \mathsf{Reg} ::= \mathsf{Char}~|~I~|~\mathsf{Reg} \otimes \mathsf{Reg}~|~\mathsf{Reg} \oplus \mathsf{Reg}~|~\mathsf{Reg}^{*}
\end{gather*}

Here, we use a non-standard syntax inspired by linear logic. Disjunction of
regular expressions is written $\oplus$, and concatenation of regular
expressions is written like the linear tensor product $\otimes$. The
empty grammar is sometimes written in the same manner as the empty string,
$\varepsilon$, but we instead choose to represent the empty grammar by $I$. This
is also linear logic syntax to represent the unit $I$ of the operation $\otimes$.

To see an illustration of the notion of grammar given in \cref{def:grammar}, consider the regular expression
$G = (a \otimes b)^{*} \otimes a$ --- a sequence of 0 or more $ab$'s followed by
an $a$. We may think of $G$ as a function that takes in strings and outputs all
of the parse trees of that string. Passing the input string $baa$ through $G$
would output the empty set, because there are no $G$-parses of $baa$. However,
passing $aba$ through $G$ would output a singleton set, as there is only a
single $G$-parse of $aba$. Namely, the one that breaks up $aba$ as $(ab)a$.

Observe that for any input string $w$ for the above grammar $G$, $G(w)$ is
either empty or a singleton set. Either the string isn't parsed by $G$,
or it is parsed by $G$ for exactly one reason. That is, $G$ is
\emph{unambiguous}. If we instead consider the grammar
$H = (a \otimes b)^{*} \oplus (b \otimes a)^{*}$ there are now ambiguities. For
instance, what is the action of $H$ on the empty string $\varepsilon$?
The empty string $\varepsilon$ matches both $(a \otimes b)^{*}$ and $(b \otimes a)^{*}$, so
$\varepsilon$ can match either the left or the right pattern
of $H$. As a result, $H(\varepsilon)$ is a two-element set.

For every grammar, we can specify a formal language recognized by that grammar. By ranging over all input strings, we can
see that every formal grammar $G$ induces a formal language $L_{G}$,

\[
  L_{G} = \{ w \in \String : G w \text{ is inhabited} \}
\]

As they are just sets, formal languages naturally are
endowed with a partial
order via the subset inclusion ordering. Similarly, formal grammars are
naturally endowed with the structure of a \emph{category} where a morphism
$f : G \to H$ is given by a \emph{family} of functions:

\[
  f^{w} : Gw \to Hw, w \in \String
\]

Intuitively, we read a morphism of grammars as a \emph{parse transformer}: $f$
translates a
$G$-parse of $w$ into an $H$-parse of $w$. The notion of isomorphism induced by
this categorical structure encodes that two grammars are equivalent if there is
a bijective translation of parses. Isomorphisms in the category of
grammars are precisely Chomsky's notion of \emph{strong equivalence} of
grammars \cite{chom1963}.

The category of grammars is quite rich in structure. It can
equivalently be described as the functor category $\Set^{\String}$, where
$\String$ is a discrete category. That is, the objects in $\String$ are strings
and the only morphisms are identity morphisms. Such functor categories into
$\Set$ --- often called presheaf categories --- carry a remarkable amount of
structure. In fact, any presheaf category naturally models dependent type
theory, which we exploit to motivate the design of the formalism
presented in \cref{sec:tt}.

\section{A Type Theory to Reason About Formal Grammars}
\label{sec:tt}

\subsection{Syntax}
\label{subsec:syntax}

In this section we will walk through most of the syntax of our type theory. We
include the full syntax in \cref{sec:infer}.

The linear typing judgment in our syntax takes on the following schematic form,
\[
  \Gamma ; \Delta \vdash M : A
\]

Let us briefly digest what this judgment is trying to say. First, $A$ represents
a \emph{linear type} in our syntax. The intended semantics of these linear types
are formal grammars. That is, the linear typing system is designed to syntactically
reflect the behavior of formal grammars. For this reason, we may often
interchangeably use the terms ``linear type'' and ``grammar''. To see the
correspondence between types and grammars, note that the constructors for linear types
mirror operations on formal grammars. For example, we have linear type constructors
$\otimes$, $\oplus$, and $\amp$ which correspond to grammar concatenation,
disjunction, and conjunction, respectively. The other type constructors also have
corresponding grammar intuition which may be seen in \cref{subsubsec:other}.

$M$ is an inhabitant of type $A$, which we want to think of as a parse tree of
the grammar $A$. The core idea of this entire paper follows precisely from this
single correspondence: grammars are types, and the inhabitants of these types
are parse trees for the grammars!

$\Gamma$ represents a \emph{non-linear context} --- also referred to as an
\emph{intuitionistic} context. The non-linear fragment of the theory
corresponds to usual Martin-L\"of type theory. That is, the intuitionistic
contexts obey all of the usual structural rules --- contraction, weakening, and
exchange --- so we may use their contents as many times as we'd like.

On the other hand $\Delta$ represents a \emph{linear context}. These linear
contexts
behave substructurally --- in particular they have no contraction, weakening, or exchange
rule. This means that everything within $\Delta$ must be used \emph{exactly
  once} and \emph{in order of occurrence}. Thus, we can think of the linear
contexts as an ordered list of limited
resources. Once a resource is consumed, we cannot make reference to it again.
We may then
think of the variables in a linear context akin to building blocks for
constructing patterns over strings.

In the realm of parsing, our substructural
restrictions make it so that the linear contexts behave quite similarly to
formal grammars. For instance, linear context extension mirrors concatenation of
grammars which may be seen via the introduction rule for $\otimes$,

\[
  \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta , \Delta' \vdash e \otimes e' : A \otimes B}
\]

Just as two grammars may be concatenated, if we can build an $A$ with the
contents of $\Delta$, and a $B$ with the contents of $\Delta'$, then we we may
concatenate $\Delta$ and $\Delta'$ to build an inhabitant $A \otimes B$.

This is perhaps better illustrated through a concrete example. Consider the
regular expression $A = (a \oplus b) \otimes (b \oplus c)^{*}$ and the string
$w = abc$. We then have the following derivation that $w$ matches $A$,

\begin{equation}
  \label{eq:samplederiv}
  \inferrule
  {
    w_{1} : a \vdash \mathsf{inl}(w_{1}) : a \oplus b \\
    \inferrule
    {
      w_{2} : b \vdash \mathsf{inl}(w_{2}) : b \oplus c \\
      w_{3} : c \vdash \mathsf{inr}(w_{3}) : b \oplus c
    }
    {
      w_{2} : b, w_{3} :c \vdash \mathsf{cons}(w_{2},w_{3}) : (b \oplus c)^{*}
    }
  }
  {w_{1} : a, w_{2} : b, w_{3} : c \vdash \mathsf{inl}(w_{1}) \otimes \mathsf{cons}(w_{2}, w_{3}) : (a \oplus b) \otimes (b \oplus c)^{*} }
\end{equation}

When parsing $w$ into $A$ by hand, we perform the same process as reflected in
the derivation \cref{eq:samplederiv}. That is, we split the string $w$ up into pieces, one of which matches
$a \oplus b$, and and another that matches $(b \oplus c)^{*}$. When further
parsing this Kleene star bit, we break it up into smaller pieces each of which
matches the patterns $b \oplus c$ and then we combine them all into one parse of
the Kleene star $(b \oplus c)^{*}$. Note that this exact process is echoed in
the above proof tree.

\subsubsection{Dependence of the Linear on the Non-Linear}

We may now however ask, ``what are the $\Gamma$'s doing in the $\otimes$ introduction rule?''
Each linear context $\Delta$ is dependent on a non-linear context $\Gamma$, so
we may use variables drawn from $\Gamma$ when defining the types in $\Delta$.

In the grammar setting, this dependence lets
us use variables from regular dependent type theory to
keep track of data without \emph{consuming any resources}, so to speak. That is,
using a piece of a linear context consumes it, so we cannot linearly encode anything that
needs to be referenced more than once.

We will see a concrete instance of this dependence in \cref{subsubsec:pda}. In
that case, we encode the traces of a pushdown automaton as a grammar. When doing so,
we make use of our non-linear types to define a type of stacks. The fact that
our encoding of stacks is non-linear reflects
that a pushdown automaton may consume a character from the input string without
affecting the state of the stack.

\subsubsection{Other Type Constructors}
\label{subsubsec:other}

\paragraph{Brzozowksi Derivative}
In the 1950s Janusz Brzozowksi developed a notion of derivative for formal
languages \cite{brzozowskiDerivativesRegularExpressions1964}. The derivative of language $L$ by string $w$ is given as,

\[
  w^{-1}L = \{ w' \in \String : w \otimes w' \in L \}
\]

That is, the derivative of $L$ with respect to $w$ are precisely the strings
that complete to an $L$-parse when prefixed by a $w$.

Brzozowski initially gave an accounting of this operation for
generalized regular expressions, but later work demonstrates that the same
construction can be generalized to context free grammars \cite{mightParsingDerivativesFunctional2011}.

The ``magic wand'' from linear logic takes on the role of Brzozowski's
derivative in our syntax.

\begin{equation}
  \label{eq:linfun}
\inferrule{\Gamma ; a : A, \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A \lto B}
\end{equation}

We read this as saying that $\Delta$ constructs a $A \lto B$-parse when
prepending $\Delta$ with an $A$-parse completes to a $B$ parse. In this sense,
the type $A \lto B$ behaves like a linear function type. The only distinction
between $A \lto B$ and an intuitionistic function type is that linear functions
consume their input.

Note that the introduction rule presented in \cref{eq:linfun} does indeed
generalize Brzozowski's initial presentation of the derivative. First, to match
his initial presentation
let $A$ be the grammar accepting a single string $w$, which we may write as the
discrete grammar $\lceil w \rceil$.
Additionally, our construction lifts from a language-theoretic construction
to a grammar point-of-view.
In this sense, we carry around not just the data of what strings belong to
$\lceil w \rceil \lto B$ but also are endowed the data of a function that
constructs parse trees of $B$ from pairs of $\lceil w \rceil$-parses and
$(\lceil w \rceil \lto B)$-parses.

\paragraph{Top and Bottom}
We include linear versions of the unit and empty types. The
grammar $\top$ matches every input string, while the grammar
$\bot$ represents falsity and matches no strings. These two
are often used together to give syntactic shorthand to
grammars that encode propositions.

For instance, we may encode the proposition
$\mathsf{isVowel}$ as,

\[
  \mathsf{isVowel}(w) :=
    \begin{cases}
      \top, \quad w \in \{a, e, i, o, u\} \\
      \bot, \quad \text{otherwise}
    \end{cases}
\]

\paragraph{Fixed Point Operator}
We include a least fixed point operator on types, $\mu$, which is the means by
which we define recursive grammars.

Observe that we need not take the Kleene star as a primitive
grammar constructor, as it is definable as a fixed point.
The Kleene star of a grammar $g$ is given as,

\[
  g^{*} := \mu X . I \oplus (g \otimes X)
\]

\begin{figure}[h!]
\begin{mathpar}
  \inferrule
  {\Gamma ; \Delta \vdash p : I}
  {\Gamma ; \Delta \vdash \mathsf{nil} : g^{*}}

  \and

  \inferrule
  {\Gamma ; \Delta \vdash p : g \\ \Gamma ; \Delta' \vdash q
  : g^{*}}
  {\Gamma ; \Delta \vdash \mathsf{cons}(p , q) : g^{*}}

  \and

  \inferrule
  {
    \Gamma ; \Delta \vdash p : g^{*} \\
    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
    \Gamma ; x : g , y : h \vdash p_{\ast} : h
  }
  {\Gamma ; \Delta \vdash \mathsf{foldr}(p_{\varepsilon} , p_{\ast}) : g^{*}}
\end{mathpar}
\caption{Kleene Star Rules}
\label{fig:star}
\end{figure}

Likewise, $g^{*}$ has admissible introduction and
elimination rules, shown in \cref{fig:star}. Note that this
definition of $g^{*}$ and these
rules arbitrarily assigns a handedness to the Kleene star.
We could have just as well took it to be a fixed point of
$I \oplus (X \otimes g)$. In fact, the definitions are
equivalent, as the existence of the $\mathsf{foldl}$ term below
shows that $g^{*}$ is also a fixed point of
$I \oplus (X \otimes g)$.

\begin{equation}
  \label{eq:foldl}
  \inferrule
  {
    \Gamma ; \Delta \vdash p : g^{*} \\
    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
    \Gamma ; y : h , x : h \vdash p_{\ast} : h
  }
  {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
\end{equation}

In fact, the $\mathsf{foldl}$ term is defined using
$\mathsf{foldr}$ --- much in the same way one
may define a left fold over lists in terms of a right fold
in a functional programming language\footnote{The
  underlying trick is to fold over a list of functions
  instead of the original string. We curry each character $c$
  of the string into a function that concatenates $c$, and
  right fold over this list of linear functions. Because function
  application is left-associative, this results in a left
  fold over the original string. }.

We only take fixed points of a single variable as a
primitive operation in the type theory, but we may apply
Beki\`c's theorem \cite{Bekić1984} to define an admissible
notion of multivariate fixed point. This is particularly
useful for defining grammars that encode the states of an
automaton. In \cref{fig:multifix} we provide the
introduction and elimination principles for such a fixed
point, where $\sigma$ is the substitution that unrolls the
mutually recursive definitions one level. That is,

\begin{align*}
  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \\
  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
\end{align*}

\begin{figure}[h!]
\begin{mathpar}
  \inferrule
  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}

  \\

  \inferrule
  {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
             \dots \\
             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
  }
  {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
\end{mathpar}
\caption{Multi-fixed Points}
\label{fig:multifix}
\end{figure}

The most prevalent application of these multi-fixed points
is for defining the traces through automata, as we will see
later in \cref{eq:nfatrace}.

\todoin{with citation, talk somewhere about how CFGs are just regular + fixed points}
\max{this is a good point to give an overview of how regular expressions and context-free grammars/expressions are essentially sub-languages of our type system.}

\subsection{Semantics in Grammars}
\label{subsec:seming}
As mentioned earlier, the interpretation of grammars as functions from
$\String$ to $\Set$ serves as our intended semantics. When designing the syntax of this formalism, we were often
informed by this model. That is, there is an interpretation
$\llbracket \cdot \rrbracket$ of our syntax that provides grammar-based
semantics.

\begin{enumerate}
  \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
  \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
        $\sem X : \sem \Gamma \to \Set_{i}$
  \item A non-linear term $\Gamma \vdash e : X$ denotes a section
        $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$
  \item Linear contexts $\Gamma \vdash \Delta$ and types
        $\Gamma \vdash A : L_{i}$ both denote families of grammars
        $\sem \Gamma \to \Gr_{i}$
  \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
        transformers
        $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
\end{enumerate}

Most of the non-linear semantics is standard, so let us investigate in more
depth the linear semantics. The interpretation of each
connective is exactly what is suggested by the syntax, in
large part to the inspiration taken from this interpretation
when designing the type theory.

\begin{enumerate}
  % \item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
  \item $\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
        \begin{itemize}
          \item The semantics of the linear $\Pi$-type is
                indeed a dependent function out of the
                semantics of $X$.
        \end{itemize}
  \item $\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
        \begin{itemize}
          \item The semantics of the linear $\Sigma$-type is
                a dependent pair.
        \end{itemize}
  \item $\sem{I} \gamma w = \{ () \pipe w = \epsilon \}$
        \begin{itemize}
          \item The only string accepted by $I$ is the empty string.
        \end{itemize}
  \item
        $\sem{A \otimes B} \gamma w = \Sigma(w1,w2:\Sigma^*) \left( (w_{1}w_{2} = w) \wedge \sem{A} \gamma w_1 \times \sem{B} \gamma w_2 \right)$
        \begin{itemize}
          \item The string $w$ matches $A \otimes B$
                precisely when it may be split into two
                pieces such that the left one matches $A$
                and the right one matches $B$.
        \end{itemize}
  \item
        $\sem{A \lto B} \gamma w = \Pi(w_a:\Sigma^*) \left( A \gamma w_a \Rightarrow B\gamma (w_aw) \right)$
        \begin{itemize}
          \item The semantics of the linear function type is
                a dependent function that turns $A$-parses
                of a substring into $B$-parses of an
                extended string.
        \end{itemize}
  \item
        $\sem{B \tol A} \gamma w = \Pi(w_a:\Sigma^*) \left( A \gamma w_a \Rightarrow B\gamma (ww_a) \right)$
  \item $\sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x)$
        \begin{itemize}
          \item The semantics of the least fixed point
                operator is indeed the least fixed point of sets.
        \end{itemize}
\end{enumerate}

\subsection{Semantics in Languages}
Recall from \cref{sec:synindsyn} that every formal grammar induces a formal language. This corresponds to another
interpretation $L(\cdot)$ of our syntax. In particular, the semantics of a
linear type $A : L_{i}$
from \cref{subsec:seming} are augmented such that

\[
  L(A) = \{ w \in \String : \sem A w \text{ is nonempty} \}
\]

\subsubsection{Weak Equivalence}
Recall that an isomorphism of grammars defined a strong equivalence. We may
additionally define a \emph{weak equivalence} of grammars when two grammars
define the same language.

That is, $A$ and $B$ are weakly equivalent when $L(A) = L(B)$. Note that this is
a semantic notion, but this just as well captured syntactically by the existence
of two judgments $\Gamma ; x : A \vdash M : B$ and
$\Gamma ; x : B \vdash N : B$, which are not necessarily mutually inverse.

\section{Automata as Grammars}
Classically, formal language theory is closely related to the study of automata.
In the Chomsky hierarchy, each language class is associated to a class of
automata that serve as recognizers. Internal to the type theory of \cref{sec:tt},
we can characterize these language classes syntactically; moreover, we
demonstrate the equivalence of these language classes to its associated automata class as a
proof term within our logic.

Internalizing these equivalences is particularly useful for building a regular
expression parser. That is, it is relatively easy for use to write down a parser
for deterministic finite automata (DFAs). Then we may compose this parser with
a proof equivalence between DFAs and regular expressions to extract a verified
regular expression parser.

As we will see in \cref{subsubsec:tm}, we may very quickly encode arbitrary
recursively enumerable languages --- since we can simulate the
tape of Turing machine in the dependent fragment of the language. We reflect
here on this expressive power to motivate why we put stringent syntactic
restrictions on particular classes of automata. The type theory we present truly
is general enough to easily capture all of the Chomsky hierarchy.
To begin, let's focus on the simplest languages and their corresponding
automata --- regular expressions and finite automata.

\subsection{Non-deterministic Finite Automata}
\label{subsec:finiteaut}
\begin{figure}
  \begin{tikzpicture}[node distance = 25mm ]
    \node[state, initial, accepting] (1) {$1$};
    \node[state, below left of=1] (2) {$2$};
    \node[state, right of=2] (3) {$3$};

    \path[->] (1) edge[above] node{$b$} (2)
              (1) edge[below, bend right, left=0.3] node{$\epsilon$} (3)
              (2) edge[loop left] node{$a$} (2)
              (2) edge[below] node{$a, b$} (3)
              (3) edge[above, bend right, right=0.3] node{$a$} (1);
  \end{tikzpicture}
  \caption{An example NFA}
  \label{fig:NFA}
\end{figure}

Classically, a \emph{nondeterministic finite automaton} (NFA) is a finite state machine where
transitions are labeled with characters from a fixed alphabet $\Sigma$. These
are often represented formally as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$,

\begin{itemize}
  \item $Q$ a finite set of states
  \item $\Sigma$ a fixed, finite alphabet
  \item $\delta : Q \times (\Sigma \cup \{ \varepsilon\}) \to \mathcal{P}(Q)$ the labeled transition function
  \item $q_{0} \in Q$ the start state
  \item $F \subset Q$ a set of accepting states
\end{itemize}

Intuitively, this can be thought of like a directed graph with nodes in $Q$ with
an edge $q \overset{c}{\to} q'$ whenever $q' \in \delta(q, c)$. Note that
transitions in an NFA may be labeled with the empty string $\varepsilon$ --- such
transitions are referred to as \emph{$\varepsilon$-transitions}. We may see an
example of an NFA in \cref{fig:NFA}.

From an NFA, we may construct a grammar of traces as follows:

First, we define a mutual fixed point grammar that describes the traces through the NFA.\ Then, we have another
grammar that tells us if we're currently in an accepting state of the automaton.
A parse of the NFA grammar is then a pair of a trace and the data that we're in
an accepting state.

Consider the NFA $N$ pictured in \cref{fig:NFA}. There are three states, $1$, $2$,
and $3$. We introduce a non-linear type $Q$ with three inhabitants $q_{1}$,
$q_{2}$, and $q_{3}$ to represent each of these states, respectively. Given a
$q, q'$, we can
then define the type of traces from $q$ to $q'$. For instance, let's construct
the traces starting at $q_{1}$ and ending at $q_{2}$ as an example.

\begin{equation}
  \label{eq:nfatrace}
  \mathsf{Trace}_{N}(q_{1}, q_{2}) = \mu
  \begin{pmatrix}
    g_{q_{1}} := g_{q_{3}} \oplus ( b \otimes g_{q_{2}} ) \\
    g_{q_{2}} := ( a \otimes g_{q_{2}} ) \oplus ( a \otimes g_{q_{3}} ) \oplus ( b \otimes g_{q_{3}} ) \oplus I \\
    g_{q_{3}} := a \otimes g_{q_{1}} \\
  \end{pmatrix}. g_{q_{1}}
\end{equation}

We should read this as defining three mutually recursive grammars, one for each
state. The definitions of these mutually recursive grammars capture the
transitions of the automaton. To ensure that we only encode traces that end in
state 2, we only include the unit $I$ as a summand in
$g_{q_{2}}$. That is, by only including $I$ at this location, it makes it the
only place where derivations of $\mathsf{Trace}(q_{1}, q_{2})$ can terminate. We
can think of all of these definitions underneath
of the $\mu$ binder as bringing some local grammars into scope. With these local
grammars in scope we are ultimately constructing a term of type $g_{q_{1}}$ to denote
only traces starting in state 1.

To encode the acceptance criteria of $N$, we want to internalize a proposition
over each state of the NFA.\ That is, for each $q \in Q$ we define a term

\[
  \mathsf{acc}(q) := q \text{ is accepting}
\]

In this example, $\mathsf{acc}(q_{1}) = \top$ and
$\mathsf{acc}(q_{2}) = \mathsf{acc}(q_{3}) = \bot$. An accepting trace of the NFA $N$ is then given by the following dependent grammar,
\[
 \mathsf{AccTrace}_{N} := \LinSigTy q Q {\left( \mathsf{Trace}_{N}(q_{0} , q) \amp \mathsf{acc}(q) \right)}
\]

where $q_{0}$ is the initial state. That is, a trace is accepted by the NFA if
we can construct the trace, and the trace ends at an accepting state. This idea
is simple enough and aligns with how we intuitively treat these automata.

Generalizing over the above example, we want to define the type of traces as,

\[
  \mathsf{Trace}_{N}(q_{0}, q_{1}) = \mu
  \begin{pmatrix}
    g_{q} := \mathsf{Trans}(q), & q \neq q_{1} \\
    g_{q} := \mathsf{Trans}(q) \oplus I , & q = q_{1}
  \end{pmatrix}. g_{q_{0}}
\]

where $\mathsf{Trans}(q)$ is an iterated disjunction that describes which
transition you should take. For an NFA, $\mathsf{Trans}(q)$ must take on the
following syntactic form,

\begin{gather*}
 \mathsf{State} \in \{ g_{q} : q \in Q \} \\
 \mathsf{Char} \in \Sigma \\
 \mathsf{Trans}_{N}(q) ::= \mathsf{Char} \otimes \mathsf{State}~|~\mathsf{State}~|~\mathsf{Trans}_{N}(q) \oplus \mathsf{Trans}_{N}(q)
\end{gather*}

That is, $\mathsf{Trans}(q)$ is a disjunction of literals followed by grammars
that encode states.

When conducting proofs involving NFA grammars, we often to
need either construct terms of type $\mathsf{Trace}_{N}(q , q')$. To
this end, we give three admissible rules for constructing
traces,

\begin{figure}[h!]
  \label{fig:admissibleintro}
  \begin{mathpar}
    \inferrule
    {~}
    {\Gamma ; \cdot \vdash \mathsf{nil} : \mathsf{Trace}_N
      (q , q)}

    \and

    \inferrule
    {\Gamma ; \Delta \vdash M : \mathsf{Trace}_N
      (dst , q') \\
    \exists \text{ transition } src \overset{c}{\to} dst
    }
    {\Gamma ; x : c , \Delta \vdash \mathsf{cons}(M) : \mathsf{Trace}_N
    (src , q')}

    \and

    \inferrule
    {\Gamma ; \Delta \vdash M : \mathsf{Trace}_N (dst , q')
      \\
    \exists~\varepsilon\text{-transition } src
    \overset{\varepsilon}{\to} dst}
    {\Gamma ; \Delta \vdash \mathsf{\varepsilon cons}(M) : \mathsf{Trace}_N
    (src , q')}
  \end{mathpar}
  \caption{Admissible Trace Constructors}
\end{figure}

That is, we may use $\mathsf{nil}$ to terminate a trace that
begins and ends at state $q$. The rules $\mathsf{cons}$ and
$\mathsf{\varepsilon cons}$ are then used to inductively
stitch traces together when sound. For instance, we can read
the $\mathsf{cons}$ rule as saying that we can
create a trace coming from a state $src$ provided that we may first
transition via the character $c$ to state $dst$ and
inductively build a trace starting from $dst$. The
$\mathsf{\varepsilon cons}$ rule says something similar, but
with an $\varepsilon$-transition instead of a transition
labelled by a character. Recall that
these rules are \emph{admissible}. They are not strictly
necessary as primitives to conduct our proofs; however, they
do provide convenient shorthand notation for building terms
of type $\mathsf{Trace}_{N}(q , q')$.

Dual to constructing traces, we often want to construct
other terms in a context containing values of type
$\mathsf{Trace}_{N}(q , q')$. For this purpose, we make use
of the elimination principle for multiple-fixed points ---
which we write as $\mathsf{mfold}$ --- given
in \cref{subsubsec:other}.

Note that this general construction will readily generalize to other types of
automata. If we wanted to define say deterministic finite automata\footnote{DFAs
\emph{could} just be defined as NFAs that happen to be deterministic. This is
one way to do so, or you may choose to present the transition relation for the
automaton as a transition \emph{function} instead. Concerns like these don't
matter so much when defining things on paper, but at formalization time these
are important and can make some proofs much easier.}, pushdown automata, Turing
machines, etc, we just swap out the type of traces for a different, but very
similar, trace construction.

With this setting for finite automata, we can now internalize some classical theorems inside of our formal system.

\subsection{A DFA Parser}
\label{subsec:regexparser}

Just as we encoded traces of NFAs as grammars, we likewise
encode the traces of a DFA as grammars. The key difference
between NFAs and DFAs is \emph{determinism} --- meaning,
that in a state $q$ inside of DFA, given a character $c$ there
will be exactly one transition that we may take leaving $q$
with label $c$. For us, this changes the definition of valid
transitions for a DFA, instead of the definition of
$\mathsf{Trans}$ provided in \cref{subsec:finiteaut} DFAs
obey

\begin{gather*}
 \mathsf{State} \in \{ g_{q} : q \in Q \} \\
 \mathsf{Trans}(q) ::= \bigoplus_{c \in \Sigma} (c \otimes \mathsf{State})
\end{gather*}

Meaning, each state has a transition for every character.

We now wish to define a parser term for DFA grammars. In
particular, for a DFA $D$ we want to build a term,

\[
  w : \String \vdash \mathsf{parse}_{D} : \mathsf{AccTrace}_{D} \oplus \top
\]

where left injection into the output type denotes acceptance
by the parser, and right injection denotes rejection. To
build such a parser, it will suffice to construct a term

\[
  w : \String \vdash \mathsf{parse}_{D} : \LinSigTy q Q {\mathsf{Trace}_{D}(q_{0} , q)}
\]
This is because given a trace of a DFA, we may easily check
if we should accept or reject by simply testing
if the final state is accepting.

Because $w$ is a Kleene star of characters, we may construct
our desired $\mathsf{parse}_{D}$ term as a $\mathsf{foldl}$
over $w$. In the empty case, we just have the trace that
ends at the accepting state. In the recursive case, we
effectively add to our trace by transitioning one character
at a time, as we read them moving across $w$.

\todoin{Derivation tree for this construction}

Perhaps this derivation is not too surprising. All it says
is that a DFA may be ran by transitioning a single character
at a time, and then accepting or rejecting based on the
final state. This is exactly what DFAs did initially, so
what did we gain? Well, this has the benefit of our type
system to ensure its correctness. Moreover, this construction exports to an
intrinsically verified and executable DFA via Agda.

\subsection{Regular Grammars and DFAs}
\label{subsec:deriv}

In order to extend the DFA
parser from \cref{subsec:regexparser} to the construction of
a verified parser
generator for regular grammars we need to perform some
plumbing establishing an equivalence between regular
expressions and DFAs.

There are several routes we may hope to take in establishing
this equivalence. First, we could prove an equivalence
between NFAs and regular expressions, and separately prove
an equivalence between NFAs and DFAs.
In \cref{subsec:eqproofs}, we include a version of
Thompson's construction --- which established the
equivalence between regular grammars and DFAs. We may additionally
hope to internalize a variant of the powerset construction \cite{rabinFiniteAutomataTheir1959}
--- which takes as input an NFA and constructs a DFA that
recognizes the same language --- and combine the results of
Thompson's construction and the powerset constructions to give an equivalence
between regular expressions and DFAs.  This route is alluring, as it
internalizes several classic grammar-theoretic constructions. However, it may
necessitate extensions to the LNL theory, like
a propositional truncation, and we have not yet investigated
how this would interact with the existing types in the
theory. The addition of a propositional truncation may seem
harmless, but it is not always immediately clear how
distinct constructions will interact. For instance, when
exploring LNL models, Benton discovered that the synthesis
of linear and dependent types require a new presentation of
the $!$ modality from linear logic
\cite{bentonMixedLinearNonlinear1995}. That is all to say,
this is a work in progress and
it is not immediate that the addition of a propositional
truncation is adequate for establishing the weak equivalence
between NFAs and DFAs.

We may instead hope to internalize an equivalence between
regular grammars and DFAs by using Brzozowski derivatives to
directly create a DFA that is weakly equivalent to a given
regular expression, as described by Owens et al.
\cite{owensRegularexpressionDerivativesReexamined2009}.
One characterization of regular grammars is that they are
precisely those grammars which have finitely many inequivalent Brzozowski
derivatives
\cite{brzozowskiDerivativesRegularExpressions1964}.
The algorithm used by Owens takes in a
regular grammar and generates a DFA that recognizes the same
language, and the states in this DFA are the finitely many
derivative equivalence classes. We initially had a version of
this theorem very roughly internalized in the LNL theory.
To our taste, too much of this presentation relied on
meta-arguments that lived outside of
our formalism, and thus this particular phrasing of the
theorem did not translate well into formalization.

In any case, we believe
that revisiting these lines of thought will lead to a
satisfactory internalization of the equivalence between
regular grammars and DFAs, and thus would bridge the gap
between our DFA parser and a full regular expression parser.

\subsection{Equivalence Between Regular Grammars and Finite Automata}
\label{subsec:eqproofs}
In this section, we describe a version of Thompson's
construction \cite{thompsonProgrammingTechniquesRegular1968}
   where we construct an NFA that recognizes a given regular
expression. Moreover, we will show that this NFA is strongly equivalent to the
original grammar. Witnessing this construction in our syntax has two benefits
\begin{enumerate}
  \item It reinforces this high-level view that the syntax is a natural and
        general setting for formal grammar reasoning, as we demonstrate that
        this formal system subsume results from existing systems, and
  \item Following the
        development of Thompson's construction, we then need
        only establish the equivalence of NFAs and DFAs to
        complete the full regular expression parser
\end{enumerate}

\begin{theorem}[Thompson]
  \label{thm:thompson}
  For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same
  language as $g$.
\end{theorem}

We make a pretty straightforward adaptation of Thompson's theorem to our setting,

\begin{theorem}[Typed Thompson]
  \label{thm:typthompson}
  For $g$ a regular grammar $g$, there is an NFA $N$ such that $g$ is isomorphic
  to $\mathsf{AccTrace}_{N}$.
\end{theorem}

\begin{proof}[Proof Sketch]
  The following proof is currently formalized in Agda. Below
  we give a walk through the structure of the proof.

  Recall that regular grammars are inductively defined via
  disjunction, concatenation, and Kleene star over literals
  and the empty grammar. By induction over regular grammars,
  we will construct an NFA that is equivalent to $g$.

  First, define the recognizing NFA for the empty grammar
  $I$.

  \begin{figure}[h!]
  \begin{tikzpicture}[node distance = 25mm ]
    \node[state, initial] (1) {$1$};
    \node[state, right of=1, accepting] (2) {$2$};

    \path[->] (1) edge[below] node{$\varepsilon$} (2);
  \end{tikzpicture}
  \caption{$NFA(I)$}
  \label{fig:emptyNFA}
  \end{figure}

  The type of traces from the initial state of $NFA(I)$ to the single
  accepting state is given by,

  \[
    \mathsf{Trace}_{NFA(I)}(q_{1}, q_{2}) = \mu
      \begin{pmatrix}
         g_{q_{1}} := g_{q_{2}} \\
         g_{q_{2}} := I
      \end{pmatrix}. g_{q_{1}}
  \]

  The accepting traces through $NFA(I)$ are then described
  as,

  \[
    \mathsf{AccTrace}_{NFA(I)} = \LinSigTy q {\{1 , 2\}} {\left( \mathsf{Trace}_{NFA(I)}(q_{1} , q) \amp \mathsf{acc}(q) \right)}
  \]

  A quick inspection of \cref{fig:emptyNFA} reveals that the
  only reasonable choice for $q$ is $q_{2}$ --- because
  state 2 is accepting while state 1 is not. Therefore,

  \begin{align*}
    \mathsf{AccTrace}_{NFA(I)}
    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \amp \mathsf{acc}(q_{2}) \\
    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \amp \top \\
    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})
  \end{align*}

  From here, to prove
  $I \cong \mathsf{AccTrace}_{NFA(I)}$ it suffices to show
  $I \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$. Below we
  give two parse transformers, one from $I$ to
  $\mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$ and vice versa. The formalized
  artifact provides the low-level details that these are
  indeed mutually inverse.

  \[
    \inferrule
    {p : I \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
     \exists \text{~transition~} q_{1} \overset{\varepsilon}{\to} q_{2}
    }
    {p : I \vdash \mathsf{\varepsilon cons}(\mathsf{nil}) : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})}
  \]

  Let $\gamma$ be the substitution $\{ g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$,

  \[
    \inferrule
    {
      p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash p :
        \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \\
      x_{1} : \simulsubst {g_{q_{2}}} {\gamma} = I \vdash x_{1} : I \\
      x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
    }
    {
      p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.x_{1} , x_{2}.x_{2})(p) : I
    }
  \]

  This concludes the proof for the case of the empty
  grammar. Let's now walk through the construction for
  literal grammars. Given a character $c$, we construct an
  NFA that recognizes only the string containing the single
  character $c$ as,

  \begin{figure}[h!]
  \begin{tikzpicture}[node distance = 25mm ]
    \node[state, initial] (1) {$1$};
    \node[state, right of=1, accepting] (2) {$2$};

    \path[->] (1) edge[below] node{$c$} (2);
  \end{tikzpicture}
  \caption{$NFA(c)$}
  \label{fig:literalNFA}
  \end{figure}

  The automaton in \cref{fig:literalNFA} induces the
  following type of traces from $q_{1}$ to $q_{2}$.

  \[
    \mathsf{Trace}_{NFA(c)}(q_{1}, q_{2}) = \mu
      \begin{pmatrix}
         g_{q_{1}} := c \otimes g_{q_{2}} \\
         g_{q_{2}} := I
      \end{pmatrix}. g_{q_{1}}
  \]

  Through the same argument as the empty grammar, the only
  state that is accepting is $q_{2}$ and thus,

  \[
    \mathsf{AccTrace}_{NFA(c)} \cong \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})
  \]

  To show the desired isomorphism of $c \cong NFA(c)$ we
  make a similar argument as we did for the empty grammar
  $I$, except we leverage the $\mathsf{cons}$ rule instead
  of $\mathsf{\varepsilon cons}$. That is, the parse
  transformers in either direction are given as,

  \[
    \inferrule
    {\cdot \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
     \exists \text{~transition~} q_{1} \overset{c}{\to} q_{2}
    }
    {p : c \vdash \mathsf{cons}(\mathsf{nil}) :
      \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})}
  \]

  \[
    \inferrule
    {
      p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash p :
        \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \\
      x_{1} : \simulsubst {(c \otimes g_{q_{2}})} {\gamma} = c \otimes I \vdash \mathsf{unitR}(x_{1}) : c \\
      x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
    }
    {
      p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.\mathsf{unitR}(x_{1}) , x_{2}.x_{2})(p) : c
    }
  \]

  Where $\gamma$ is the substitution
  $\{ c \otimes g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$ and
  $\mathsf{unitR}$ is a witness to the isomorphism
  $c \otimes I \cong c$. Again, we may see that these do
  indeed mutually invert each other in the Agda code.

  It remains to show that the desired isomorphisms are
  preserved by $\otimes$, $\oplus$, and Kleene star. Here,
  we will give the argument for the disjunction case, the
  others are defined quite similarly.

  Given two NFAs $N$ and $M$, we may define a new NFA that
  encodes the disjunction of $N$ and $M$. Denote the
  internal states of $N$ by $q_{j}$'s and the internal states
  of $M$ by $r_{k}$'s,

  \begin{figure}[h!]
  \begin{tikzpicture}[node distance = 20mm ]
    \node[state] (2) {$q_{init}$};
    \node[state, initial, below left of=2] (1) {$init$};
    \node[state, below right of=1] (3) {$r_{init}$};
    \node[right of=2] (4) {$\cdots$};
    \node[right of=3] (5) {$\cdots$};
    % \node[state, below right of=1] (3) {$3$};
    % \node[state, below of=2] (4) {$4$};

    \path[->] (1) edge[below] node{$\varepsilon$} (2);
    \path[->] (1) edge[below] node{$\varepsilon$} (3);
    \path[->] (2) edge[below] node{} (4);
    \path[->] (3) edge[below] node{} (5);

    \node[label={[name=l] $N$}, draw,line width=2pt,rounded corners=5pt, fit=(2)(4)] {};
    \node[label={[name=l] $M$}, draw,line width=2pt,rounded corners=5pt, fit=(3)(5)] {};
  \end{tikzpicture}
  \caption{$N \oplus_{NFA} M$}
  \label{fig:disjunctionNFA}
  \end{figure}

  \cref{fig:disjunctionNFA} shows the process of
  disjunctively combining NFAs. Precisely, we add a single
  new state and we included copies of the states from each of $N$
  and $M$. The new state acts as the initial state and has
  $\varepsilon$-transitions to the initial states of $N$ and
  $M$. We include all of the internal transitions from $N$
  and $M$, and the accepting states of $N \oplus_{NFA} M$
  are exactly the accepting states in each subautomaton.

  Let $g$ and $g'$ be two regular grammars such that
  $g \cong NFA(g)$ and $g' \cong NFA(g')$. As a matter of
  notation\footnote{We shall similarly abuse notation for
    $\otimes$ and Kleene. That is, for a regular grammar
    $g$, when we write $NFA(g)$ we mean the NFA inductively
    built up with the NFA-analogues to the constructors that
    built up $g$.}, we will write $NFA(g \oplus g')$ for
  $NFA(g) \oplus_{NFA} NFA(g')$. The traces of
  $NFA(g \oplus g')$ are then given by,

  \[
    \mathsf{Trace}_{NFA(g \oplus g')}(src , dst) = \mu
      \begin{pmatrix}
        g_{init} := g_{q_{0}} \oplus g_{r_{0}} \\
        g_{q_{j}} := \mathsf{Trans}_{NFA(g)}(q_{j}) \oplus \mathsf{isDst}(q_{j}) \\
        g_{r_{k}} := \mathsf{Trans}_{NFA(g')}(r_{k}) \oplus \mathsf{isDst}(r_{k})
      \end{pmatrix}.g_{src}
    \]

  where $\mathsf{Trans}$ is used to echo the same syntactic
  definitions that appear in the $NFA(g)$ and $NFA(g')$.
  Also, $src$ and $dst$ may take on any value in
  $Q := \{init\} \cup \{q_{j}\} \cup \{r_{k}\}$, and
  $\mathsf{isDst}(q)$ checks if $q$ is equal to $dst$. Which
  is all to say, the traces of $NFA(g \oplus g')$ comprise
  either a trace of $NFA(g)$, or a trace of $NFA(g')$, and
  the transition coming out of $g_{init}$ determines which
  subautomaton we step into.

  The parse transformer from $g \oplus g'$ checks which side
  of the sum type we are on, then takes the appropriate step
  from $g_{init}$ in the automaton.

  \[
    \inferrule
    {
      u : g \vdash \iota (\phi (u)) : \mathsf{AccTrace}_{NFA(g \oplus g')} \\
      v : g' \vdash \iota' (\psi (v)) : \mathsf{AccTrace}_{NFA(g \oplus g')}
    }
    {p : g \oplus g' \vdash \mathsf{case}~p \{ \mathsf{inl}(u) \mapsto s , \mathsf{inr}(v) \mapsto r \} : \mathsf{AccTrace}_{NFA(g \oplus g')}}
  \]

  with $\iota$ and $\iota'$ as embeddings from $NFA(g)$ and
  $NFA(g')$, respectively, into $NFA(g \oplus g')$,
  $\phi: g \cong \mathsf{AccTrace}_{NFA(g)}$, and $\psi: g' \cong \mathsf{AccTrace}_{NFA(g')}$. On a
  high level, all this construction does is turn a parse of
  $g$ into a parse of $NFA(g)$ and then embeds that inside
  of the larger automaton $NFA(g \oplus g')$. Likewise for $g'$.

  In the other direction, recall that the data of an
  accepting trace for $NFA(g \oplus g')$ is a pair of a
  trace and a proof that
  the end state $q'$ of that trace is accepting. By
  multifolding over the first part of that pair, we turn the
  term of type
  $\mathsf{Trace}_{NFA(g \oplus g')}(init , q')$ into a
  trace of either of the subautomata,

  \[
    p : \mathsf{Trace}_{NFA(g \oplus g')}(init , q') \vdash \mathsf{mfold}_{NFA(g \oplus g')} : \mathsf{Trace}_{NFA(g)}(q_{0} , q') \oplus \mathsf{Trace}_{NFA(g')}(r_{0} , q')
  \]

  Additionally, we leverage the fact that the only accepting
  states for $NFA(g \oplus g')$ are those from the
  subautomata to extract that $q'$ must be an accepting
  state from a subautomaton.

  \[
    x : \mathsf{acc}_{NFA(g \oplus g')}(q') \vdash M : \mathsf{acc}_{NFA(g)}(q') \oplus \mathsf{acc}_{NFA(g')}(q')
  \]

  We then combine the trace and proof of acceptance into an
  accepting trace of one of the subautomata,

  \[
    p : \mathsf{AccTrace}_{NFA(g \oplus g')} \vdash N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')}
  \]

  Lastly, we then inductively use the isomorphisms $\phi$
  and $\psi$ to turn the accepting traces into a parse of
  $g$ or $g'$,

  \[
     N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')} \vdash \mathsf{case}~N~\{\mathsf{inl}(n) \mapsto \phi^{-1}(n), \mathsf{inr}(n') \mapsto \psi^{-1}(n')\} : g \oplus g'
  \]

  As discussed above, the other cases and low-level details
  of isomorphism are reserved for the formalization. This
  concludes the proof of our variant of Thompson's construction.
\end{proof}

\subsection{Other Automata}
\subsubsection{Pushdown Automata}
\label{subsubsec:pda}
A (nondeterministic) \emph{pushdown automaton} is an automaton that employs a
stack. Just like NFAs, they have transitions labeled with characters from a
fixed string alphabet $\Sigma$. Additionally, they maintain a stack of
characters drawn from a stack alphabet $\Gamma$. They are often represented
formally as a 7-tuple $(Q, \Sigma, \Gamma, \delta, q_{0}, Z, F)$,

\begin{itemize}
  \item $Q$ a finite set of states
  \item $\Sigma$ a fixed, finite string alphabet
  \item $S$ a fixed, finite stack alphabet
  \item
        $\delta \subset Q \times (\Sigma \cup \{ \varepsilon \}) \times S \to \mathcal{P}(Q \times S^{*})$
        the transition function
  \item $q_{0} \in Q$ the start state
  \item $Z \in S$ the initial stack symbol
  \item $F \subset Q$ the accepting states
\end{itemize}

We encode the traces of a pushdown automaton very similarly to those of an NFA,
except the transitions of a PDA are instead encoded via the linear-non-linear
$\Pi$-type. This is because of simply transitioning via character, a PDA must
also pop and push characters onto a stack, which is used as the argument to
these dependent functions.

Let $S$ be a non-linear type encoding the stack
alphabet, and build lists over $S$ as the (non-linear) least fixed-point
$\mathsf{List}(S) := \mu X . 1 + S \times X$. Then, the type of states for a
PDA $P$ with stack alphabet $S$ are given as a functions that takes in lists $L$,
and then makes a case distinction between possible transitions based on what was witnessed as $\mathsf{head}(L)$. The choice of transition
will then determine which character to transition by and what word $w$ should be
pushed onto the stack. The word that is added to the top of the stack is
appended to $\mathsf{tail}(L)$ and then we recursively step into another state
called on argument $w + \mathsf{tail}(L)$.

% \begin{gather*}
%   \mathsf{State} \in {g _{q} : q \in Q} \\
%   \mathsf{Word} \in \String \\
%   \mathsf{Char} \in \Sigma   \\
%   \mathsf{StackChar} \in \Gamma \\
% \end{gather*}
% \begin{align*}
%   \mathsf{Trans}_{P}(q) ::=~&
%                            \LinPiTy {(hd :: tl)} {\mathsf{List}(S)} {\left( \mathsf{Char} \otimes \mathsf{State}(\mathsf{Word} + tl) \right)}~| \\
%   & \LinPiTy {(hd :: tl)} {\mathsf{List}(S)} {\left(  \mathsf{State}(\mathsf{Word} + tl) \right)}~|~\mathsf{Trans}_{P}(q) \oplus \mathsf{Trans}_{P}(q)
% \end{align*}

% That is to say, when transitioning a PDA pops off the head $hd$ of the stack

\subsubsection{Turing Machines}
\label{subsubsec:tm}

In
\cref{subsubsec:pda}, we gave a grammar presentation of
traces through a PDA by using a non-linear type $S$
to encode the stack. We may similarly use pairs $S \times S$
to encode the tape of a Turing machine. With two stacks we can simulate the behavior of the
infinite tape of a Turing machine. The intuition behind this correspondence is
that the left half of the tape is on one stack, the right
half of the tape the other, and we treat the tops of stacks
like the head of the tape.

Due to computability limitations, we are not able to extract
a verified parser for all recursively enumerable languages.
Yet we should still reflect on the generality of our syntax,
as it allows a natural encoding of Turing machines.

\section{Future Work}
\label{sec:future}

\subsection{Implementation for Context-Free Grammars}
As suggested throughout the paper and briefly explored in
\cref{subsubsec:pda}, the first extension to the work in
this paper will be to bring analogous constructions to
context-free grammars and their accompanying pushdown automata.

\subsection{Beyond Strings}
\label{subsec:beyond}

While parsing typically refers to the generation of semantic
objects from string input, many tasks in programming can be
viewed as parsing of objects with more structure, such as
trees with binding structure or graphs. Fundamental to the
frontend of many
programming language implementations are type systems. In
particular, \emph{type checking}
--- analogous to language recognition --- and \emph{typed
  elaboration} --- analogous to parsing --- arise when
producing a semantic object subject to some analysis. Just
as our string grammars were given as functors from $\String$
to $\Set$, we envision adapting the same philosophy
to functors from $\String$ to \emph{trees} to craft a syntax
that natively captures typed elaboration. This suggests an
unusual sort of bunched type theory, where context extension
no longer resembles concatenation of strings but instead
takes on the form of tree constructors.

Thus far, our theory has proved useful for internalizing long-standing
grammar-theoretic constructions, but there has been decades of research
conducted since. A fruitful avenue for future work includes
testing if our formalism can also internalize more recently proposed grammar
mechanisms, such as the interval parsing grammars given by Zhang et al. \cite{zhangIntervalParsingGrammars2023}.

\newpage

\bibliographystyle{plain}
\bibliography{refs.bib}

\newpage

\appendix

\section{Inference Rules}
\label{sec:infer}

\begin{figure}[h!]
  \label{fig:structjdg}
  \begin{mathpar}
    \inferrule{~}{\ctxwff \cdot}
    \and
    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}

    \\

    \inferrule{~}{\linctxwff \Gamma \cdot}
    \and
    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}

    \\

    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}

    \and

    \inferrule{\Gamma \vdash A : L_i}{\linctxwffjdg \Gamma A}

    \\

    \inferrule{\Gamma \vdash X \equiv Y : U_i}{\ctxwffjdg \Gamma {X\equiv Y}}

    \and

    \inferrule{\Gamma \vdash A \equiv B : L_i}{\linctxwffjdg \Gamma {A \equiv B}}

  \end{mathpar}
  \caption{Structural judgments}
\end{figure}

\begin{figure}
  \label{fig:typewf}
  \begin{mathpar}
    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
 %
    \and
%
    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
%
    \and
%
    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash 1 : U_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash I : L_i}
 %
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
    %
    \and
    %
    \inferrule{~}{\Gamma \vdash \top : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
%
    \\
    %
    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
    %
    \and
    %
    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
  \end{mathpar}
  \caption{Type well-formedness}
\end{figure}

\begin{figure}
  \label{fig:jdgeq}
  \begin{mathpar}
    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
%
    \and
%
    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
%
    \and
%
    \inferrule{~}{}
%
    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
%
    \and
%
    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
%
    \and
%
    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
%
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
%
    \and
%
    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
\end{mathpar}
  \caption{Judgmental equality}
\end{figure}

\begin{figure}
  \label{fig:inttyping}
  \begin{mathpar}
  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
  %
  \\\
  %
  \inferrule{~}{\Gamma \vdash () : 1}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
  %
  \\
%
  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
  \and
  \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
  %
  \\
  %
  \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
  \and
  \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
  \end{mathpar}
  \caption{Intuitionistic typing}
\end{figure}

\begin{figure}
  \label{fig:linsyntax}
  \begin{mathpar}
    \inferrule{~}{\Gamma ; a : A \vdash a : A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
    \\
    %
    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta', \Delta \vdash \applto {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta, \Delta' \vdash \apptol e {e'} : B}
    %
    \\
    %
    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}{\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma ; \Delta \vdash \app e {e'} : \subst A {e'} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash e : X \quad \Gamma ; \Delta \vdash e' : \subst A e x}{\Gamma ; \Delta \vdash (e, e') : \LinSigTy x X A}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \Delta \vdash () : \top}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e_1 : A_1 \quad \Gamma ; \Delta \vdash e_2 : A_2}{\Gamma ; \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : A_1 \amp A_2 }{\Gamma ; \Delta \vdash \pi_i \, e : A_i}
    %
    \\
    %
    \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma ; \cdot \vdash \mathsf{G}^{-1}\, e : A}
    %
    \\
    %
    \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
    \and
    \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
  \end{mathpar}
  \caption{Linear typing}
\end{figure}
\end{document}

### 2024-05-23 notes/grammar-as-separation-logic.tex
Repo reorg

\documentclass[12pt]{article}
%AMS-TeX packages
\usepackage{tikz}
\usepackage[utf8]{inputenc}
\usepackage{amssymb,amsmath,amsthm,stmaryrd, wasysym} 
%geometry (sets margin) and other useful packages
\usepackage[margin=1.25in]{geometry}
\usepackage{graphicx,ctable,booktabs}
\usepackage{tikz-cd}
\usepackage{mathpartir}

\usepackage[sort&compress,square,comma,authoryear]{natbib}
\bibliographystyle{plainnat}

\newcommand{\leftmultimap}{\mathop{\rotatebox[origin=c]{180}{$\multimap$}}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\FinSet}{\mathbb F}
\newcommand{\Cat}{\textrm{Cat}}
\newcommand{\Lang}{\textrm{Lang}}
\newcommand{\Grammar}{\textrm{Grammar}}
\newcommand{\Action}{\textrm{Action}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}

\newcommand\rsepimp{\mathbin{-\mkern-6mu*}}
\newcommand\lsepimp{\mathbin{\rotatebox[origin=c]{180}{$-\mkern-6mu*$}}}
\newcommand\sepconj{\mathbin{*}}
\newcommand\bunch{\mathcal W}

\begin{document}
\title{Grammars as Separation Logic Programs}
\author{Max S. New}
\maketitle

\section{Formal Languages and Formal Grammars}

Formal language theory is one of the great success stories of
theoretical computer science as a discipline, providing the foundation
for parser generators etc. The fundamental notion of formal language
theory is the formal language. Given an alphabet $\mathcal A$, a
formal language is a subset of the set of strings $L \subseteq
\mathcal A^*$. The beauty of this mathematical definition is that this
is a universal notion of language: it provides a simple specification
that can be used as a single semantic domain for many disparate
grammar formalisms without a priori limiting the languages to be
recognizable or efficiently recognizable. While formal languages
provide a simple semantic domain, they have limitations as
specifications for \emph{parsers}, one of the most ubiquitous sorts of
programs. The reason is that fundamentally a formal language $L$,
being a subset, can only properly serve by itself as the specification
for a \emph{recognizer} which takes a string and outputs a boolean. A
correct recognizer with respect to a language $L$ is a computable
version of the indicator function for the subset. But parsers output
far richer data than merely single bits, they output some kind of
\emph{parse tree}, or in reality when combined with semantic actions
produce an output of an arbitrary data type. Further, formal grammar
formalisms themselves have a richer semantics than merely a formal
language: for instance, formal grammars may be \emph{ambiguous}
meaning there is more than one parse tree for a single input string,
something which cannot be captured in a subset semantics.

This brings us to our first definition:
\begin{definition}
  A grammar between strings over a type $\mathcal A$ with semantic
  values in a type $X$ is a function
  \[ \mathcal A^* \to X \to \textrm{Type} \]
\end{definition}
That is, fundamentally a grammar is simply a constructive relation
between strings and ``things''\footnote{thanks Wadler}. Given such a
grammar $G$, a parse tree for a string $w$ with semantic meaning $x$
is an element $G\,w\,x$. We will show that just as formal languages
provide a unified semantics for languages, this notion of grammar
provides a unified semantics for grammars.

Given a grammar, we can immediately specify total and partial parsers for the grammar:
\begin{definition}
  Let $G \in \Grammar\,{\mathcal A}\,X$. A \emph{total parser} for $G$ is a function
  \[ (w:\mathcal A^*) \to (x: X) \times G\,w\,x\]
  and a (decidably) \emph{partial parser} is a function
  \[ (w : \mathcal A^*) \to 1 + ((x: X) \times G\,w\,x)\]
\end{definition}
That is, a parser outputs a semantic value and a parse tree that
``proves'' that the semantic value is represented by the
string\footnote{in a practical implementation, generating the parse
tree at runtime is unnecessary. This could possibly be modeled using
modalities for \emph{runtime irrelevance}.}. We say a grammar is
\emph{unambiguous for parsing} if for every $w$, the type
\[ (x: X) \times G\,w\,x \]
is a sub-singleton: it has at most one element.

Dually, we can just as easily define given a grammar notions of \emph{printer}:
\begin{definition}
  Let $G \in \Grammar\,{\mathcal A}\,X$. A \emph{printer} for $G$ is a function
  \[ (x : X) \to (w : \mathcal A^*) \times G\,w\,x \]
\end{definition}
The notion of partial printer is less useful.

If we think of an abstract grammar as a kind of constructive relation,
then that inevitably leads us to the thesis of this article:
\begin{quote}
  Grammar formalisms are logic programming languages
\end{quote}
But the most interesting point of this is that the logic that grammar
formalisms are based on is a kind of \emph{separation
logic}. Separation logic is typically used to reason about the layout
of data in memory and there is an analogous intuition for grammars: if
our memory is a string of characters in $\mathcal A$, then a grammar
describes the ways in which semantic values can be represented in
``memory'' as strings. To demonstrate this connection, we show that
the grammars assemble into a cartesian closed category that is also
monoidal biclosed, i.e., a model of the logic of bunched implications.

\section{A Category of Grammars}

Since a grammar is parameterized by a type of characters and semantic
values, there is some choice in defining a category of grammars: do we
allow these parameter types to vary or stay fixed? For modeling
grammar formalisms, it is most natural to fix a single type of
characters $\mathcal A$ and allow the types of semantic values to
vary. Intuitively this is like the typical situation of fixing a type
of tokens that are output from a lexer but defining non-terminals that
parse into many different output types.

\begin{mathpar}
  Let $\mathcal A$ be a type, the category of $\mathcal A$-grammars
  $\Grammar\,\mathcal A$ has
  \begin{enumerate}
  \item as objects...
  \item morphisms...
  \item composition/identity...
  \end{enumerate}
\end{mathpar}

To construct the various structures on this category, we observe that
it can be constructed in the following fashion: $\Set^{\mathcal
  A^*}/\Delta$, a kind of \emph{Artin Gluing} construction on toposes.

\section{Grammars as Dependent Types}

We define a 2-level dependent type theory for specifying grammar and
their output types. For the remainder of this section fix an alphabet
$\mathcal A$.

\begin{mathpar}
\item A context $\Gamma$ denotes a set.
\item A substitution $\gamma : \Delta \to \Gamma$ denotes a function
\item A type $\Gamma \vdash A$ denotes a family of sets over $\Gamma$
  and substitution $A[\gamma]$ is restriction of the family along a
  function.
\item A term $\Gamma \vdash M : A$ denotes a section of the family and
  substitution $M[\gamma]$ is restriction of a section along a
  function.
\item A grammar context $\Gamma \pipe \hat \Gamma$ denotes a grammar
  $\Grammar\,{\mathcal A^*}\,\Gamma$ and substitution
  $\hat\Gamma[\gamma]$ denotes restriction of a grammar along a
  function.
\item A grammar substitution $\Gamma\pipe \hat \gamma : \hat \Delta \to \hat \Gamma$ denotes a family of functions
  \[ \hat \gamma : (w:\mathcal A^*) \to (\gamma : \Gamma) \to \hat \Delta\,w\,\gamma \to \hat \Gamma\,w\,\gamma \]
\item A grammar type $\Gamma
  \vdash \hat A$ denotes a pair of a family of sets $\Gamma \vdash A$
  and a family of grammars $(\gamma :\Gamma) \to \Grammar\,{\mathcal
    A^*}\,(A\gamma)$.
\item A grammar term $\Gamma\pipe\hat\Gamma\vdash \hat M : \hat A$ denotes 
\end{mathpar}

\section{Subsuming Grammar Formalisms}

\end{document}

### 2024-05-23 notes/grammars-over-categories.tex
Repo reorg

\documentclass[12pt]{article}
%AMS-TeX packages
\usepackage{tikz}
\usepackage[utf8]{inputenc}
\usepackage{amssymb,amsmath,amsthm,stmaryrd, wasysym}
\usepackage{mathtools}
%geometry (sets margin) and other useful packages
\usepackage[margin=1.25in]{geometry}
\usepackage{graphicx,ctable,booktabs}
\usepackage{tikz-cd}
\usepackage{mathpartir}

\usepackage[sort&compress,square,comma,authoryear]{natbib}
\bibliographystyle{plainnat}

\newcommand{\leftmultimap}{\mathop{\rotatebox[origin=c]{180}{$\multimap$}}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\FinSet}{\mathbb F}
\newcommand{\Cat}{\textrm{Cat}}
\newcommand{\Lang}{\textrm{Lang}}
\newcommand{\Grammar}{\textrm{Grammar}}
\newcommand{\Action}{\textrm{Action}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}

\newcommand\rsepimp{\mathbin{-\mkern-6mu*}}
\newcommand\lsepimp{\mathbin{\rotatebox[origin=c]{180}{$-\mkern-6mu*$}}}
\newcommand\sepconj{\mathbin{*}}
\newcommand\bunch{\mathcal W}

\begin{document}
\title{Grammars over Free Categories}
\author{Max S. New}
\maketitle

We consider the semantics of \emph{formal grammars} over free
categories as a generalization of the notion of a formal grammar over
a free monoid. Whereas free monoids are a mathematical formalization
of completely \emph{unstructured} data, free categories are a very
simple class of \emph{structured} data, and so formal grammars over
them provide a simple setting for describing grammars over
\emph{typed} expression languages: those whose syntax is entirely
sequential.

As an example, we might consider a simple Hoare logic over a typed
sequential programming language. \ldots TODO examples

A signature for a free category $\Sigma$ is a \emph{quiver}, also
known as a directed multigraph. Then this is completed to the free
category $F\Sigma$. If the quiver $\Sigma$ has exactly one object,
then the free category is the delooping of the free monoid on the set
of edges $\Sigma_e^*$. In this way, formal grammars over free monoids,
i.e., strings, will be a special case.

\section{A Bicategory of Grammars}

A formal grammar over a free monoid $\Sigma^*$ denotes a set of
\emph{parses} for each string. This is fruitfully modeled as an object
of the presheaf category $\Set^{\Sigma^*}$, where $\Sigma^*$ here is
viewed as the \emph{discrete} category on the elements of
$\Sigma^*$. In plain English: a grammar\footnote{Here we are taking a
maximally broad view of what a grammar can be, abstracted from any
syntactic notion of formal grammar. A Lawverian terminology would call
this an \emph{objective} grammar as opposed to the \emph{subjective}
grammars from specific grammar formalisms such as regular expressions,
CFGs etc.} is a function from strings to the set of parses of the
string.

Several important structures immediately follow from this definition:
\begin{enumerate}
\item Since $\Set^{\Sigma^*}$ is a category, we get a notion of
  \emph{morphism} of grammars: a function that translates parses of a
  string $w$ in the domain grammar to parses of the same string in the
  codomain grammar.
\item Since $\Set^{\Sigma^*}$ is a presheaf category, we have the
  \emph{Yoneda embedding} $\Sigma^* \to \Set^{\Sigma^*}$. This says
  for each word $w$ in the language there is a corresponding grammar
  $w$ that has a unique parse for $w$ and no parses for any other
  string.
\item Since $\Sigma^*$ is a monoid, $\Set^{\Sigma^*}$ is monoidal
  closed with the \emph{Day Convolution} monoidal closed
  structure. The monoidal product is the sequencing of grammars and
  the monoidal unit is the grammar that has a unique parse for only
  the empty string.
\item Additionally any presheaf category is a bi-cartesian closed
  category, which gives us conjunction, disjunction and implication of
  grammars.
\end{enumerate}
Note that because we are working with a free monoid, we only ever need
to apply the Yoneda embedding to individual generators as the Yoneda
embedding is a strong monoidal functor.

To generalize this to free categories, note that the elements of the
monoid $\Sigma^*$ are in the free category view \emph{morphisms} of
the free category: the endomorphisms of the single object
$F\Sigma(*,*)$. This suggests that a \emph{grammar} over a free
category be paramaterized by a source and target object, that is, we
have for each pair of objects $a,b$, a grammar over the morphisms
$F\Sigma(a,b)$, which we can define as $\Set^{F\Sigma(a,b)}$. We will
call this category $\Grammar(a,b)$.

\begin{enumerate}
\item First, a morphism between grammars is the same as before: a
  morphism from $\alpha: G \to H : \Grammar(a,b)$ is a family of
  functions $\alpha_f : G(f) \to H(f)$ for every $f : a \to b$.
\item Now for each morphism $f : a \to b$ in $F\Sigma(a,b)$ we have
  the Yoneda embedding as a grammar $Yf : \Grammar(a,b)$ that has a
  unique parse of only $f$, i.e., $(Yf)(g) = \{ * \,|\, f = g \}$. As
  above we only need to use this when $f$ is a generator.
\item We generalize the monoidal structure on grammars over strings to
  a \emph{biclosed bicategory} structure on grammars over typed
  strings. That is for each object $a$ we get an ``empty string''
  grammar $\epsilon_a : \Grammar(a,a)$ that matches only the identity
  morphism. Next, given grammars $G : \Grammar(a,b)$ and $H :
  \Grammar(b, c)$ we can define a \emph{composition} grammar $G
  \odot_b H : \Grammar(a,c)$ by
  \[ (G \odot_b H)f = \sum_{g: a \to b}\sum_{h: b \to c} Gg \times Hh \]
  Additionally, since the composition is (inherently) non-symmetric we get a covariant and contravariant hom structure:
  \[ \inferrule{G : \Grammar(c,b) \and H : \Grammar(a,b)}{G \triangleright_{b} H : \Grammar(a,c)} \]
  defined as
  \[ (G \triangleright_{b} H)(f : a \to c) = \prod_{g : c \to b} G(g) \to H(gf) \]
  and a dual
  \[ \inferrule{H : \Grammar(b,c)\and G : \Grammar(b,a) }{H \mathop{\prescript{}{b}\triangleleft} G : \Grammar(a,c)} \]
  defined as
  \[ (H \mathop{\prescript{}{b}\triangleleft} G)(f : a \to c) = \prod_{g: b \to a} G(g) \to H(fg) \]
\item Additionally, each hom-category is a bicartesian closed
  category, giving us analogous disjunction, conjunction and
  implication of grammars.
\end{enumerate}

\end{document}

### 2024-05-23 notes/kall.tex
Repo reorg

\documentclass[12pt,a4paper]{article}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amscd}
\usepackage{cmll}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage[latin2]{inputenc}
\usepackage{t1enc}
\usepackage[mathscr]{eucal}
\usepackage{indentfirst}
\usepackage{graphicx}
\usepackage{graphics}
\usepackage{pict2e}
\usepackage{epic}
\usepackage{syntax}
\usepackage{enumitem}
\usepackage[colorlinks]{hyperref}
\numberwithin{equation}{section}
\usepackage[margin=2.9cm]{geometry}
\usepackage{epstopdf}

\usepackage{tikz-cd}

\usepackage{ stmaryrd }

\newcommand{\cat}{%
\mathbf%
}

\newcommand{\sem}[1]{
  \llbracket #1 \rrbracket
}

\def\numset#1{{\\mathbb #1}}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{question}[theorem]{Question}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{Conj}[theorem]{Conjecture}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{?}[theorem]{Problem}
\newtheorem{example}[theorem]{Example}

\newcommand{\im}{\operatorname{im}}
\newcommand{\Hom}{{\rm{Hom}}}
\newcommand{\diam}{{\rm{diam}}}
\newcommand{\ovl}{\overline}
\newcommand{\norm}[1]{||#1||}
\newcommand{\test}[1]{\mathcal D(#1)}
\newcommand{\dist}[1]{\mathcal D'(#1)}
\newcommand{\diff}[1]{(#1, \mathcal P^U_{#1})}
\newcommand{\lto}{\multimap}

\newcommand{\set}[2]{\{#1 \, | \, #2 \}}

\newcommand{\lrang}[1]{ \langle #1 \rangle }

\newcommand{\M}{\mathcal{M}}

%Grammar
\newcommand{\smid}{\ \mid \ }
\newcommand{\defined}{\coloneqq}
\newcommand{\subst}[3]{#1\{#2 / #3\}}

%% types
\newcommand{\nat}{\mathbb{N}}
\newcommand{\R}{\mathbb{R}}

%% stlc
\newcommand{\lamb}[2]{\lambda #1. \ #2}
\newcommand{\app}[2]{#1 \ #2}

%% cbpv
\newcommand{\produce}{\mathsf{produce}\ }
\newcommand{\thunk}{\mathsf{thunk}\ }
\newcommand{\force}{\mathsf{force}\ }
\newcommand{\letbe}[3]{\mathsf{let}\ #1 \ \mathsf{be}\ #2 \ \mathsf{in}\ #3}
\newcommand{\mtox}[3]{#1 \ \mathsf{to} \ #2 . \ #3}
\newcommand{\consume}{\mathsf{consume}\ }
\newcommand{\supply}{\mathsf{supply}\ }
\newcommand{\supplyas}[3]{\supply #1 \ \mathsf{as} \ #2 \ \mathsf{in} \ #3}
\newcommand{\factory}{\mathsf{factory}\ }
% Typing rule environment

\begin{document}

\title{Proof-relevant Kleene algebra and parsing}

\author{Pedro Amorim}

\maketitle

\section{Kleene algebra}

A Kleene algebra (KA) $(K, \cdot, +, (-)^*, 0 , 1)$ is a set $K$ with binary operations $\cdot$ and $+$ and the Kleene star operation $(-)^*$ satisfying the axioms depicted in Figure~\ref{fig:axioms}. In particular $+$ must be idempotent ($a + a = a$) which allows us to equip $K$ with the partial order $a \leq b$ iff $a + b = b$. One of the main applications of Kleene algebra is reasoning about languages recognized by regular expressions. Intuitively, the partial order can be seen as an abstraction of language containment. In particular, testing the language membership of a word $\omega$ in a language $b$ is equivalent to proving $\omega \leq b$.

\begin{example}
  One of the most important examples of Kleene algebras is the set of regular languages over an alphabet $\Sigma$. In this case the product is language concatenation, the sum is language union, $1$ is the language with the empty string $\varepsilon$, $0$ is the empty language and $(-)^*$ is the Kleene star operation.
\end{example}

One of the main limitations of the inequational theory of Kleene algebras is that they cannot reason about language parsing, only language recognition. Indeed, when we get a proof that $\omega \leq b$, we
only know that the word is an element of the language $b$ and not how to parse it from $b$ nor if
there are multiple ways of parsing it.

From a categorical point of view, this limitation arises from the fact that the partial order does not contain enough information for capturing anything other than containment. We address this issue by proposing a ``proof-relevant'' variant of Kleene algebras that can reason about properties other than language containment. It would be interesting if this approach could be extended to variants of Kleene algebras. For instance, by considering a probabilistic variant of KA it might be possible to consider probabilistic notions of parsing.
%
\begin{figure}
  \begin{align*}
    x + (y + z) &= (x + y) + z & x + y &= y + x\\
    x + 0 &= x & x + x &= x\\
    x(yz) &= (xy)z & x1 &= 1x = x\\
    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
    x0 &= 0x = x & & \\
    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
  \end{align*}
  \label{fig:axioms}
  \caption{Kleene algebra axioms}
\end{figure}

\section{Categorifying Kleene algebra} In this section we define the categorification of Kleene algebras.

\begin{lemma}
  If $\cat{K}$ is a Kleene algebra seen as a posetal category then $\cat{K}$ has coproducts and an initial object $0$.% iff $a \leq b \iff a + b = b$.
\end{lemma}
\begin{proof}
  %% Assume that $\cat{K}$ has coproducts and an initial object $0$. We want to show that $a\leq b \iff a + b = b$. Assume that $a \leq b$. Furthermore, we have $b \leq b$. From the universal property of coproducts, we have $a + b \leq b$. On the other direction, we can show that $b \leq a + b$ from $0 \leq a$, $b \leq b$, $0 + b = b$ and functoriality of the coproduct.

  By assumption $\cat{K}$ has a least element $0$ and $a \leq b \iff a + b = b$. We want to show that $\cat{K}$ has coproducts and an initial object $0$. It is straightforward to show that $0$ is the initial object. To show that $\cat{K}$ also has coproducts, we must show that it has the injection morphisms $a \leq a + b$. This follows easily from the axiom $a = a + a$ and by the definition of $\leq$, To show the universal property, assume that $a \leq c$ and $b \leq c$. We can then show that $(a + b) + c = a + (b + c) = a + c = c$. This concludes the proof.
\end{proof}

The theorem above shows that the idempotent sum from Kleene algebras corresponds to coproducts over posetal categories with initial object.

\begin{lemma}
  For every Kleene algebra $\cat{K}$ seen as a category, the triple $(\cat{K}, \cdot, 1)$ is a monoidal category.
\end{lemma}
\begin{proof}
  Since products in Kleene algebras are monotonic in both arguments, $\cdot$ is a bifunctor and by the product laws, $1$ is a monoidal unit. The commutativity of the coherence diagrams follows from $\cat{K}$ being a poset, i.e. all parallel arrows are equal.
\end{proof}

Note that in general, and in most applications of Kleene algebras, the product is not commutative, which is why the monoidal structure is not symmetric. The distributivity of the sum over the product corresponds to distributive coproducts.

\begin{definition}
  If $\cat{C}$ is a monoidal cocartesian category, then it is said to be commutative if the canonical morphisms $(A \otimes B) + (A \otimes C) \to A \otimes (B + C)$ and
  $(A \otimes C) + (B \otimes C) \to (A + B) \otimes C$ are isomorphisms.
\end{definition}

Given this definition it is straightforward to prove the following lemma.

\begin{lemma}
  Every Kleene algebra $\cat{K}$ is a distributive category.
\end{lemma}

Now we turn to the Kleene star. By looking at the first two axioms it becomes clear that $a^*$ should be an algebra for the functors $F_1(x) = 1 + ax$ and $G_1(x) = 1 + xa$. Furthermore, consider the third axiom when $b = 1$. It basically says that $a^*$ is the initial $F_1$-algebra. In the general case, it says that for every $b$, $a^*b$ is the initial algebra for the functor $F_b(x) = b + ax$ and $ba^*$ is the initial algebra for the functor $G_b(x) = b + xa$.

\begin{question}
  Let $\cat{C}$ be a distributive monoidal category such that for every pair of objects $A$ and $B$ the functor $F_B(X) = B + A \otimes X$ admits an initial algebra, which is represented as $\mu X.\, F_B(X)$. Does it follow that $\mu X.\, F_B(X) \cong (\mu X.\, F_1(X)) \otimes B$?
\end{question}
\begin{proof}
  Assuming distributivity, we can equip $(\mu X.\, F_1(X)) \otimes B$ with an $F_B$-algebra structure as follows: $B + A \otimes (\mu X.\, F_1(X)) \otimes B \cong (1 + A \otimes \mu X.\, F_1(X)) \otimes B \xrightarrow{\alpha \otimes id_B } (\mu X.\, F_1(X)) \otimes B$, where $\alpha$ is the $F_1$-algebra
  structure of $(\mu X.\, F_1(X))$.

  It is unclear to me if we can show in the non-Cartesian case that the isomorphism does hold. If $\otimes$ is Cartesian, $(\mu X.\, F_1(X)) \otimes B$ are pairs of a list and an element of $B$, which is isomorphic to ``lists'' where the empty list is a $B$ element. I'm not so sure if this same analysis can be done in the monoidal case.
\end{proof}

We can now define the categorification of Kleene algebras.

\begin{definition}
  A proof-relevant Kleene algebra is a distributive monoidal category such that for every object $A$, the functor $F(X) = 1 + A \otimes X$ admits an initial algebra.
\end{definition}

\section{Examples}

We illustrate the applicability of our definition by providing a couple of non-trivial examples of proof-relevant Kleene algebra. The simplest models are, of course, Kleene algebras seen as posetal categories.

\subsection{Ordered logic and Kleene algebra}

Ordered logic is a kind of substructural logic which drops every structural rule, contrasting with linear logic which does not drop the exchange rule. For the sake of simplicity, we will consider the fragment of ordered logic which only has the conjunctive connectives and the structural modality $\oc$ that recovers the structural rules\footnote{even exchange?}. From a proof theory point of view, this modality is usually axiomatized as
%
\begin{mathpar}
  \oc A \vdash 1 \and \oc A \vdash A \and  \oc A \vdash \oc A \otimes \oc A
\end{mathpar}
%
%
\begin{lemma}
  \label{lem:ineq}
  The following inequalities hold in every Kleene algebra:
  \begin{itemize}
  \item $1 \leq a^*$
  \item $a \leq a^*$
  \item $a^* a^*\leq a^*$
  \end{itemize}
\end{lemma}
\begin{proof}
  We can prove $1 \leq a^*$ by $1 \leq 1 + 0 \leq 1 + aa^* \leq a^*$. Next, by a similar argument we can prove that $aa^* \leq a^*$ which allows us to prove $a \leq a\cdot 1 \leq a\cdot a^* \leq a^*$. Finally, we can prove $a^*a^*$ by first observing that by using the last inequality and the idempotency of addition we get $a^* + a\cdot a^* \leq a^*$ which, by applying the second family of axioms for the Kleene star allows us to conclude.
\end{proof}
Note, however, that by Lemma~\ref{lem:ineq}, the inequalities $1 \leq a^*$, $aa^* \leq a^*$ and $a^*a^* \leq a^*$ can all
be proved using the Kleene algebra axioms. This suggests that we can propose an alternate
axiomatization for the $\oc$ modality:

\begin{mathpar}
  \inferrule{ }{\oc A \vdash 1}

  \and

  \inferrule{ }{\oc A \vdash A \otimes \oc A}

  \and

  \inferrule{ }{\oc A \vdash A \otimes \oc A}

  \\

  \inferrule{\Gamma \vdash X \\ X \vdash B \\ X \vdash X \otimes A}{\Gamma \vdash \oc A \otimes B}

  \and

    \inferrule{\Gamma \vdash X \\ X \vdash B \\ X \vdash A \otimes X}{\Gamma \vdash B \otimes \oc A}

\end{mathpar}

A similar treatment of the exponential has already been explored in the literature but as they are presented above they violate the cut elimination property, though I am not sure if we care all that
much about it in this context. Though we still have to properly characterize what this new exponential is from a category theory point of view, we can informally define what is a model of ordered logic.

\begin{theorem}
  If $\cat{K}$ is a Kleene algebra then its opposite category $\cat{K}^{op}$ is a ordered logic model. 
\end{theorem}

%% \begin{question}
%%   With this new correspondence in mind, are there useful Kleene algebra extensions that come out of it? A simple one is understanding what happens if we consider models of ordered logic in which the functors $A \otimes -$ and $- \otimes A$ have right adjoints, i.e. left and right implications.

%%   Would this still be a notion of derivative? Of course, if anything it would be a generalization of Brzozowski's derivative, since it you can differentiate any regular expression with respect to any other regular expression.
%% \end{question}

\subsection{Functorial grammars}

Let $\Sigma$ be an alphabet and $\Sigma^*$ be the set of finite strings over it, which can also be seen as a discrete category.

\begin{theorem}
  The category $\cat{Set}}^{(\Sigma^*)^{op}}$ is a proof-relevant Kleene algebra.
\end{theorem}
\begin{proof}
  Since every presheaf category is a topos, it has coproducts. The monoidal product is given by the Day convolution. It is also known that in presheaf categories the Day convolution is part of a monoidal closed structure, which means that it is a left adjoint and, in turn, preserves colimits, making it a distributive monoidal category.
\end{proof}

In the proof above we still need the Kleene star, which can probably be constructed using the usual semantics of inductive types based on sequential colimits.

%% \section{Proof-relevant Kleene algebras}

%% Now that we have a better understanding of what is a categorical model for Kleene algebras, we can
%% start exploring what are some of its useful proof-relevant models. The canonical example, if $\cat{K}$ is a Kleene algebra, is the presheaf category $\cat{PSh(K)}$. By using the Day convolution this category is monoidal and, furthermore, it can be shown to be monoidal closed by general categorical arguments.

%% We do not know if the exponential modality can also be lifted to $\cat{PSh(K)}$. Since the yoneda embedding is a dense functor, it might be possible to lift $\oc$ by a density argument...

%% By unfolding the definitions, a functor $F : \cat{K}^{op} \to \cat{Set}$ assigns to each element of $\cat{K}$ a set and if $f : a \leq_{\cat{K}} b$ then $F(f): F(b) \to F(a)$ is a function. The representable functors $Y(a)$ are functors that map $b$ to the singleton set if $b \leq a$ and to the empty set otherwise.

%% \begin{question}
%%   An important result in the theory of formal languages is that regular expressions have the same expressive power as finite automata. Is there a generalization of this theorem for the non-poset case?
%% \end{question}

%% \begin{remark}
%%   In order to connect this to the functorial approach to parsing where parsers are functors $\Sigma^* \to \cat{Set}$, where $\Sigma^*$ is the discrete category over the free monoid over a finite set $\Sigma$, maybe we should consider the functor category $[\cat{K}^{op}, \cat{PSh}(\Sigma^*)]$.
%% \end{remark}

%% \begin{Def}
%%   A pre-KA $(K, +, \cdot, *)$ is a set equipped with binary operations $+$ and $\cdot$, and a unary operation $(-)^*$ satisfying the axioms:
%%   \begin{align*}
%%     & a + b = b + a\\
%%     & a + (b + c) = (a + b) + c\\
%%     & a + 0 = a\\
%%     & a + a = a\\
%%     & a \cdot b = b \cdot a\\
%%     & a \cdot 1 = a\\
%%     & 1 + aa^* \leq a^*\\
%%     & b + ax \leq x \implies ba^* \leq x
%%   \end{align*}
%% \end{Def}

%% \begin{theorem}
%%   Let $\cat{K}$ be a posetal category. $\cat{K}$ is a pre-KA iff it has coproducts $+$, an initial object $0$ and a symmetric monoidal product $(\cdot, 1)$, and for every object $b$ the functors $F_b(x) = b + ax$ admit initial algebras such that the initial algebra for $F_b$ is isomorphic to $ba^*$, where $a^*$ is the initial algebra for $F_1$.
%% \end{theorem}

%% \section{From KA to LL}

%% The axiomatisation of linear logic ensures that certain structural rules are only valid under the ! modality. Such rules are captured by the axioms $!A \vdash 1$ and $!A \vdash !A \otimes !A$. It is possible to show that the opposite category of every pre-KA $\cat{K}$ is a model of the exponential conjunctive fragment of linear logic (ECLL). This is a consequence of the inequalities $1 \leq a^*$, $a^* a ^* \leq a^*$, $a \leq a^*$ and $(a \leq b^* \implies a^* \leq b^*)$ being valid theorems in the inequeational theory of pre-KAs.

%% What we can get out of this is that we can give an alternative axiomatisation of ECLL by importing the pre-KA axioms, resulting in the following inference rules:

%% %% \begin{typing}
%% %%   \inferrule{ }{!A \vdash 1}

%% %%   \and

%% %%   \inferrule{ }{!A \vdash A \otimes !A}

%% %%   \and

%% %%   \inferrule{\Gamma \vdash X \\ X \vdash B \\ X \vdash X \otimes A}{\Gamma \vdash !A \otimes B}
%% %% \end{typing}

%% The last rule is not exactly the same as its pre-KA counterpart exclusively due to proof-theoretic reasons. Since $\Gamma$ may contain more than one hypothesis, if we were to import the pre-KA rule directly we would have to add the metatheoretic operation $\bigotimes \Gamma$ which folds over the context and tensors everything in it. On the other hand, as it was noted by other authors, the rule as it stands makes it impossible to validate the cut elimination theorem.

%% \section{From LL to KA}

%% It is also worth exploring what happens when going the other way: defining the Kleene star operator using the linear logic exponential axioms:

%% \begin{align*}
%%   & 1 + a + a^*a^* \leq a^*\\
%%   & a \leq b^* \implies a^* \leq b^*
%% \end{align*}

%% When looking at the models for these axioms we see that they can admit a certain infinitary flavor\footnote{I'm being a bit loose here: when working with infinite strings, the interaction of infinity and products gets a bit subtle}. A consequence of this remark is that the axioms above are not equivalent to the original ones. To see why that is the case, consider the language model over a finite alphabet. If we interpret $a^*$ as the set of finite and infinite concatenations of $a$, it is easy to see that the pre-KA axiom $b + ax \leq x \implies ba^* \leq x$ is not valid -- consider the case where $b=1$ and $x$ only contains finite strings.

\bibliographystyle{acm}
  % \bibliography{mybib}

\end{document}

### 2024-06-29 POPL/paper.tex
POPL Paper Stub

@@
 % -*- fill-column: 80; -*-
-\documentclass[sigconf,review,screen]{acmart}
+\documentclass[sigconf,anonymous,review,screen]{acmart}
 \usepackage{mathpartir}
 \usepackage{tikz-cd}
+\usepackage{lipsum}
 \usepackage{enumitem}
@@
 \usepackage{stmaryrd}
+\usepackage{tikz}
+\usetikzlibrary{automata, positioning, arrows, fit}
+
+\usepackage{pdfpages}
 
 \usepackage{todonotes}
-\newcommand{\todoin}[1]{\todo[inline]{TODO:\@ #1}}
 
@@
 \newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
+\newcommand{\String}{\Sigma^{*}}
 \newcommand{\Set}{\mathbf{Set}}
@@
 \drafttrue
-\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven}: #1]}\fi}
-\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max}: #1]}\fi}
-\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro}: #1]}\fi}
+\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven says}: #1]}\fi}
+\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max says}: #1]}\fi}
+\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro says}: #1]}\fi}
 \newcommand{\pipe}{\,|\,}
 
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-%% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
-  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
-\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-  June 03--05, 2018, Woodstock, NY}
-\acmPrice{15.00}
-\acmISBN{978-1-4503-XXXX-X/18/06}
+\newcommand{\todoin}[1]{\ifdraft{\todo[inline]{TODO:\@ #1}}\fi}
 
@@
 
-\title{Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}
+\pagestyle{plain}
+
+\pagebreak
+
+\ifdraft{
+\listoftodos{list of todos}
+\pagebreak}\fi
+
+\title{Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}
+
 \author{Steven Schaefer}
-\affiliation{
-  \department{Electrical Engineering and Computer Science}
+\affiliation{\department{Electrical Engineering and Computer Science}
   \institution{University of Michigan}
@@
 
+\makeatletter
+\let\@authorsaddresses\@empty
+\makeatother
+
 \begin{abstract}
-  We propose a semantic framework for the study of formal grammars. First, we
-  provide a syntax-independent notion of formal grammar as a function from
-  strings to sets of parse trees, generalizing the familiar notion of a language
-  as a set of strings. This point-of-view naturally endows a grammar with the
-  structure of a presheaf, which allows many language-theoretic constructs to be
-  defined via universal properties.
+  In this paper, we propose a new syntactic formalism for formal grammars: a
+  dependent linear/non-linear type theory in which the tensor product is
+  non-commutative. The linear types in this calculus may be interpreted as
+  grammars, and linear terms as parse transformers. This provides a logical
+  characterization of grammar classes based on a substructural logic rather than
+  ordinary first-order logic.
 
-  With these categorical constructions in mind, we propose a new syntactic
-  formalism for formal grammars: a version of dependent linear-non-linear type
-  theory where the tensor product is non-commutative. The linear types in this
-  theory are interpreted as grammars --- and linear terms as parse transformers.
-  The non-dependent fragment of this type theory is already expressive enough to
-  capture regular expressions, context-free expressions, and finite automata.
-  Moreover, the isomorphism between regular languages and finite automata is
-  definable as a proof term in our formalism. The
-  non-dependent fragment of this type theory is expressive enough to capture
-  both regular and context-free expressions, as well as finite automata already
-  regular expressions, context-free expressions, and finite automata. Moreover,
-  the is definable as a proof term in this formalism. Further, through use of
-  dependency of linear types on non-linear types, we can express richer grammar
-  formalisms such as indexed grammars, pushdown automata, and Turing machines.
+  Internally to our calculus, we carry out many classical constructions from
+  the theory of formal grammars. Inspired by the common
+  structure underlying various ideas in formal language
+  theory, our type theory makes these relationships explicit
+  and provides a unifying syntax to reason about formal
+  grammars.
 
-  Classically, many results in both language and automata theory are proved
-  informally with respect to some ambient foundation --- e.g. Kleene algebra, grammars,
-  parsers, semantic actions, etc. The type theory provided in
-  this paper serves as an intuitive and general framework to syntactically unify
-  these related ideas and internalize their
-  theorems formally. We give a new form of \emph{logical characterization} of
-  grammar classes based on a substructural logic rather than ordinary
-  first-order logic. Further, the type theory can not only
-  express the grammars as types, but equivalences between grammars and
-  parsers as terms of the grammar.
-
-  We give this type theory a semantics in the category of grammars and
-  prove a related canonicity theorem. Moreover, we keep an eye toward the
-  implementation of this type theory for correct-by-construction parsers and
-  parser combinators.
+  We give this type theory a semantics in the category of grammars and prove a
+  canonicity theorem showing that every term in a context corresponding to a
+  fixed string is equal to a term in a canonical form encoding a parse tree of
+  that string. Finally, we discuss the application to the design and
+  implementation of correct-by-contruction parsers.
 \end{abstract}
@@
 
-\section{A Syntax-Independent Notion of Syntax}
-\todoin{Rework intro from LICS draft with a focus on implementation}
+\setcounter{page}{1}
 
-\section{Kleene Categories}
-\todoin{Tell a story winding through the use of KAs, action algebras, etc to
-  motivate why we'd want categorical semantics for KCats}
-\todoin{Incoporate Kleene Category definition}
-\todoin{End with the punch that $\Set^{\Sigma^{*}}$ is a KCat}
+\section{Introduction}
+The notion of formal language is central to the theory of parsing. A
+\emph{formal language} $L$ over an alphabet $\Sigma$ is classically defined as a
+\emph{subset} of strings $L \subseteq \String$. This definition is
+especially useful as it gives a semantics to formal grammars that is completely
+independent of any particular syntactic grammar formalism. Any new
+notion of grammar can be given a language semantics and it provides a precise
+mathematical specification for implementing a \emph{recognizer} of a language.
 
-\section{Automata}
+That is, a recognizer should be extensionally equivalent to the indicator function
+$\chi_{L} : \String \to \Prop$ mapping each string $w$ to the the proposition
+that it is in the language $w \in L$. However, language recognition is
+insufficient for specifying a parser --- a function that decomposes a string
+into semantic components that adhere to the structure of a formal grammar. In
+practice, parsers do more than simply return whether a string is valid or
+invalid; instead, they emit structured \emph{parse trees}\footnote{Such parse
+  trees are usually not materialized in memory, as semantic actions often perform
+  computation over the trees while they are being constructed.}
 
+% To aid in the development of verified parsers, the primary contribution
+% of this paper is
+% an type
+% theory whose types natively capture the semantics of formal
+% grammars. Types in this theory are grammars, and terms are
+% parse transformers. This kind of type theory naturally
+% requires (non-commutative) linear and dependent types.
+% Within this formalism, we may define parsers as terms
+% defined in the context of an input string. Moreover, by virtue of
+% typechecking, these parsers are \emph{correct-by-construction}. The intrinsic
+% verification gained from the type system therefore removes the need for a
+% post-hoc verification in a language like Coq, thus lowering the barrier to
+% compiler frontend verification.
+
+% Substructural logics, like linear and separation logic, offer precise control
+% over how values from a type may be used, in particular how many times a value
+% may be used in a derivation. This control enables \emph{resource-sensitive}
+% reasoning, where we may treat a resource as \emph{consumed} after usage. In the
+% context of parsing, we may treat the characters composing a string as finite resources
+% that are consumed at parse-time.
+
+% Dependent type theory extends the simply-typed $\lambda$-calculus by allowing type
+% definitions to contain the occurrence of a term. In lieu of set-based
+% foundations, this extension can then be taken as a foundation for mathematics.
+% The expressive power unlocked by this insight has enabled the development of
+% dependently-typed proof assistants
+% such as Coq or Agda. Within each of these languages, systems may be given very
+% precise functional specifications of their behavior.
+
+% In 1995, Benton et al.
+% \cite{bentonMixedLinearNonlinear1995} first gave categorical models of a logic
+% that synthesized linear and non-linear logic. Later, Krishnaswami et al.
+% \cite{krishnaswami_integrating_2015} built upon this work to develop a type
+% system that unifies both linear and
+% dependent types --- referred to as a \emph{linear-non-linear} (LNL) type theory. In
+% such settings, we have two sub-theories --- one linear and one non-linear. The
+% linear sub-theory embeds ordinary linear types without dependence, while the
+% non-linear sub-theory embeds standard Martin-L\"of type theory. The interplay of
+% linear types with dependence lives at the boundary between these two fragments.
+% Benton et al.\ gives an accounting of how we may jump between these
+% fragments via a monoidal adjunction $F \dashv G$ between linear terms and
+% non-linear terms.
+
+% To the end of defining real verified
+% parsers, in our LNL theory we define a parser as a term,
+
+% \crefname{equation}{}{}
+% \begin{equation}
+%   \label{eq:parser}
+%   w : \String \vdash \mathsf{parse}_{g}(w) : g \oplus \top
+% \end{equation}
+
+% We will investigate the precise meaning of judgments like \cref{eq:parser} in
+% \cref{sec:tt}, but first let's get an intuitive feel for the meaning behind such a
+% term. On the left of the turnstile occurs a single string $w$, we may think of
+% this describing what resources are linearly available to us. Using each piece of
+% our the context on the left of the turnstile, we build a term
+% $\mathsf{parse}_{g}(w)$ of type $g \oplus \top$. Notably, the type being
+% instantiated is a sum type, and the data of which side we inject into denotes
+% whether the string is accepted or rejected by the parser. That is all to say, a
+% parser takes in an input string and either matches it to the desired grammar or
+% rejects the string.
+
+% Practical parsers are often implemented via an abstract automaton. For instance,
+% a parser may translate a grammar into a state machine recognizes the
+% same language as the original grammar, as the automata
+% point-of-view lends itself to straightforward execution. The
+% strength of our syntax is that it allows us to describe
+% these automata as grammars themselves. Moreover, the
+% equivalence between the grammar and automata presentations
+% is derivable as a proof term within the theory. This allows
+% us access to simultaneously internalization of the equivalence
+% between grammars and automata as well as the ability to
+% export to automaton implementations of a parser that are executable.
+
+% To realize an executable implementation of our type theory,
+% we shallowly embed it in Agda. Because our implementation is
+% defined as a function in Agda, we can leverage the Agda
+% ecosystem for code generation. Further, in lieu of
+% implementing a complete language from scratch, we make use of the
+% Agda proof system as a setting to guarantee correctness of
+% our generate parsers.
+
+% Within this paper, we give our progress in defining a verified
+% parser for regular expressions. The construction of such a parser entails
+% the definition of a parser for deterministic finite automata (DFA) grammars,
+% and an equivalence proof between DFAs and regular expressions. We have thus far
+% completed the construction of a parser for DFAs $D$,
+
+% \[
+%   w : \String \vdash \mathsf{parse}_{D}(w) : D \oplus \top
+% \]
+
+% and half of the equivalence proof between DFAs and regular expressions, both
+% formalized in Agda. Upon completion of the second half of this proof, we would
+% have a system where a user could simply give a regular grammar and then extract
+% an intrinsically verified parser for it. In future work, we
+% hope to show an analogous construction involving pushdown
+% automata (PDAs) and context-free grammars.
+
+% The contributions of this paper include an LNL theory for reasoning about formal
+% grammars, the encoding of automata as formal grammars in this theory, and the
+% definition of parser terms for DFA grammars. The contents include,
+
+% \begin{itemize}
+%   \item A syntax-independent definition of formal grammar
+%   \item The formal rules governing our type system
+%   \item Interpretations of the type system for reasoning about formal grammars
+%         and formal languages
+%   \item The presentation of automata as grammars
+%   \item The definition of a parser within our type system
+% \end{itemize}
+
+% \section{A Syntax-Independent Notion of Syntax}
+% \label{sec:synindsyn}
+
+% % \max{this section doesn't flow naturally from the intro, which is all about the
+% %   type theory and parsing. This is where you want to outline in the intro and
+% %   possibly say here that we want to revisit ``formal grammar theory'' and then
+% %   we will make a type theory that has formal grammar theory as its semantics.}
+
+% In this section we will revisit some foundational ideas in formal
+% grammar theory, before giving a type
+% theory for which formal grammars serves as its semantics.
+
+% Formal languages and grammars are among some of the oldest and most studied areas in
+% theoretical computer science. The developments of the 1950s and 1960s gave rise
+% to expressive formalism for reasoning about grammars, and further practical
+% algorithms for constructing parsers for regular and context-free grammars.
+
+% A grammar $G$ is often described generatively as a set of production rules for
+% generating strings. Often this is presented as a finite set of nonterminal
+% symbols $N$, a finite set of terminal symbols $\Sigma$, and a finite set of rewriting
+% rules for substitution of nonterminals. In this setting, we continually apply
+% rewrite rules until we are left only with a string of terminals
+% $w \in \String$. We will give a more precise notion of
+% grammar in a moment, but for now let us take this intuitive view of grammars as
+% string generators.
+
+% One of the primary objects of study is that of a \emph{formal language} over
+% a fixed alphabet $\Sigma$. A language $L$ is a \emph{subset} of strings
+% $L \subseteq \String$. This notion of formal language gives an intuitive
+% set-based semantics for grammars. That is, we may interpret a grammar $G$ as the
+% language $L(G)$ that is generated by it. We are yet to give any precise
+% definition of what we mean by ``grammar'', although we note that any suitable
+% notion of grammar can be given a language semantics as above.
+
+% Formal languages serve as a basis for mathematically specifying a
+% \emph{recognizer} --- a
+% function $r : \String \to \Bool$ that determines if a string belongs to a
+% given language. However, they are insufficient for specifying a \emph{parser}
+% --- a function that decomposes a string into semantic components that adhere to
+% the structure of a formal grammar. In practice, parsers don't just tell you
+% if a string is valid or invalid. Instead, they emit structured data such as
+% abstract syntax trees (ASTs).
+
+% This motivates our first conceptual contribution, a simple and
+% syntax-independent notion of formal grammar:
+
+% \begin{definition}
+%   \label{def:grammar}
+%   A \emph{formal grammar} $G$ over a fixed alphabet $\Sigma$ is a function
+%   $G : \String \to \Set$.
+% \end{definition}
+
+% We say that a grammar $G$ associates to every string $w$ the set\footnote{We
+%   make little comment as to which foundations we use to encode sets. We argue\max{drop ``We argue''}
+%   that our construction is agnostic to the particular foundation, and so without
+%   loss of generality\max{incorrect use of ``without loss of generality''} we may take a class of small sets, a Grothendieck universe,
+%   a type theoretic universe, or any similar foundation. Here we choose a type
+%   theoretic universe.} $G w$ of parse
+% trees showing that $w$ matches $G$.
+
+% The simplest class of formal grammar on the Chomsky\max{first mention of Chomsky hierarchy, why not bring it up earlier when you are describing what a grammar is? Also cite}
+% hierarchy are the \emph{regular expressions}
+
+% \max{inaccurate reflection of common terminology. These are regular expressions,
+%   regular grammars usually means a CFG with a restriction on recursion that
+%   ensures you only allow Kleene star.}.
+
+% \begin{gather*}
+%   \mathsf{Char} \in \Sigma \\
+%   \mathsf{Reg} ::= \mathsf{Char}~|~I~|~\mathsf{Reg} \otimes \mathsf{Reg}~|~\mathsf{Reg} \oplus \mathsf{Reg}~|~\mathsf{Reg}^{*}
+% \end{gather*}
+
+% Here, we use a non-standard syntax inspired by linear logic. Disjunction of
+% regular expressions is written $\oplus$, and concatenation of regular
+% expressions is written like the linear tensor product $\otimes$. The
+% empty grammar is sometimes written in the same manner as the empty string,
+% $\varepsilon$, but we instead choose to represent the empty grammar by $I$. This
+% is also linear logic syntax to represent the unit $I$ of the operation $\otimes$.
+
+% To see an illustration of the notion of grammar given in \cref{def:grammar}, consider the regular expression
+% $G = (a \otimes b)^{*} \otimes a$ --- a sequence of 0 or more $ab$'s followed by
+% an $a$. We may think of $G$ as a function that takes in strings and outputs all
+% of the parse trees of that string. Passing the input string $baa$ through $G$
+% would output the empty set, because there are no $G$-parses of $baa$. However,
+% passing $aba$ through $G$ would output a singleton set, as there is only a
+% single $G$-parse of $aba$. Namely, the one that breaks up $aba$ as $(ab)a$.
+
+% Observe that for any input string $w$ for the above grammar $G$, $G(w)$ is
+% either empty or a singleton set. Either the string isn't parsed by $G$,
+% or it is parsed by $G$ for exactly one reason. That is, $G$ is
+% \emph{unambiguous}. If we instead consider the grammar
+% $H = (a \otimes b)^{*} \oplus (b \otimes a)^{*}$ there are now ambiguities. For
+% instance, what is the action of $H$ on the empty string $\varepsilon$?
+% The empty string $\varepsilon$ matches both $(a \otimes b)^{*}$ and $(b \otimes a)^{*}$, so
+% $\varepsilon$ can match either the left or the right pattern
+% of $H$. As a result, $H(\varepsilon)$ is a two-element set.
+
+% For every grammar, we can specify a formal language recognized by that grammar. By ranging over all input strings, we can
+% see that every formal grammar $G$ induces a formal language $L_{G}$,
+
+% \[
+%   L_{G} = \{ w \in \String : G w \text{ is inhabited} \}
+% \]
+
+% As they are just sets, formal languages naturally are
+% endowed with a partial
+% order via the subset inclusion ordering. Similarly, formal grammars are
+% naturally endowed with the structure of a \emph{category} where a morphism
+% $f : G \to H$ is given by a \emph{family} of functions:
+
+% \[
+%   f^{w} : Gw \to Hw, w \in \String
+% \]
+
+% Intuitively, we read a morphism of grammars as a \emph{parse transformer}: $f$
+% translates a
+% $G$-parse of $w$ into an $H$-parse of $w$. The notion of isomorphism induced by
+% this categorical structure encodes that two grammars are equivalent if there is
+% a bijective translation of parses. Isomorphisms in the category of
+% grammars are precisely Chomsky's notion of \emph{strong equivalence} of
+% grammars \cite{chom1963}.
+
+% The category of grammars is quite rich in structure. It can
+% equivalently be described as the functor category $\Set^{\String}$, where
+% $\String$ is a discrete category. That is, the objects in $\String$ are strings
+% and the only morphisms are identity morphisms. Such functor categories into
+% $\Set$ --- often called presheaf categories --- carry a remarkable amount of
+% structure. In fact, any presheaf category naturally models dependent type
+% theory, which we exploit to motivate the design of the formalism
+% presented in \cref{sec:tt}.
+
+% \section{A Type Theory to Reason About Formal Grammars}
+% \label{sec:tt}
+
+% \subsection{Syntax}
+% \label{subsec:syntax}
+
+% In this section we will walk through most of the syntax of our type theory. We
+% include the full syntax in \cref{sec:infer}.
+
+% The linear typing judgment in our syntax takes on the following schematic form,
+% \[
+%   \Gamma ; \Delta \vdash M : A
+% \]
+
+% Let us briefly digest what this judgment is trying to say. First, $A$ represents
+% a \emph{linear type} in our syntax. The intended semantics of these linear types
+% are formal grammars. That is, the linear typing system is designed to syntactically
+% reflect the behavior of formal grammars. For this reason, we may often
+% interchangeably use the terms ``linear type'' and ``grammar''. To see the
+% correspondence between types and grammars, note that the constructors for linear types
+% mirror operations on formal grammars. For example, we have linear type constructors
+% $\otimes$, $\oplus$, and $\amp$ which correspond to grammar concatenation,
+% disjunction, and conjunction, respectively. The other type constructors also have
+% corresponding grammar intuition which may be seen in \cref{subsubsec:other}.
+
+% $M$ is an inhabitant of type $A$, which we want to think of as a parse tree of
+% the grammar $A$. The core idea of this entire paper follows precisely from this
+% single correspondence: grammars are types, and the inhabitants of these types
+% are parse trees for the grammars!
+
+% $\Gamma$ represents a \emph{non-linear context} --- also referred to as an
+% \emph{intuitionistic} context. The non-linear fragment of the theory
+% corresponds to usual Martin-L\"of type theory. That is, the intuitionistic
+% contexts obey all of the usual structural rules --- contraction, weakening, and
+% exchange --- so we may use their contents as many times as we'd like.
+
+% On the other hand $\Delta$ represents a \emph{linear context}. These linear
+% contexts
+% behave substructurally --- in particular they have no contraction, weakening, or exchange
+% rule. This means that everything within $\Delta$ must be used \emph{exactly
+%   once} and \emph{in order of occurrence}. Thus, we can think of the linear
+% contexts as an ordered list of limited
+% resources. Once a resource is consumed, we cannot make reference to it again.
+% We may then
+% think of the variables in a linear context akin to building blocks for
+% constructing patterns over strings.
+
+% In the realm of parsing, our substructural
+% restrictions make it so that the linear contexts behave quite similarly to
+% formal grammars. For instance, linear context extension mirrors concatenation of
+% grammars which may be seen via the introduction rule for $\otimes$,
+
+% \[
+%   \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta , \Delta' \vdash e \otimes e' : A \otimes B}
+% \]
+
+% Just as two grammars may be concatenated, if we can build an $A$ with the
+% contents of $\Delta$, and a $B$ with the contents of $\Delta'$, then we we may
+% concatenate $\Delta$ and $\Delta'$ to build an inhabitant $A \otimes B$.
+
+% This is perhaps better illustrated through a concrete example. Consider the
+% regular expression $A = (a \oplus b) \otimes (b \oplus c)^{*}$ and the string
+% $w = abc$. We then have the following derivation that $w$ matches $A$,
+
+% \begin{equation}
+%   \label{eq:samplederiv}
+%   \inferrule
+%   {
+%     w_{1} : a \vdash \mathsf{inl}(w_{1}) : a \oplus b \\
+%     \inferrule
+%     {
+%       w_{2} : b \vdash \mathsf{inl}(w_{2}) : b \oplus c \\
+%       w_{3} : c \vdash \mathsf{inr}(w_{3}) : b \oplus c
+%     }
+%     {
+%       w_{2} : b, w_{3} :c \vdash \mathsf{cons}(w_{2},w_{3}) : (b \oplus c)^{*}
+%     }
+%   }
+%   {w_{1} : a, w_{2} : b, w_{3} : c \vdash \mathsf{inl}(w_{1}) \otimes \mathsf{cons}(w_{2}, w_{3}) : (a \oplus b) \otimes (b \oplus c)^{*} }
+% \end{equation}
+
+% When parsing $w$ into $A$ by hand, we perform the same process as reflected in
+% the derivation \cref{eq:samplederiv}. That is, we split the string $w$ up into pieces, one of which matches
+% $a \oplus b$, and and another that matches $(b \oplus c)^{*}$. When further
+% parsing this Kleene star bit, we break it up into smaller pieces each of which
+% matches the patterns $b \oplus c$ and then we combine them all into one parse of
+% the Kleene star $(b \oplus c)^{*}$. Note that this exact process is echoed in
+% the above proof tree.
+
+% \subsubsection{Dependence of the Linear on the Non-Linear}
+
+% We may now however ask, ``what are the $\Gamma$'s doing in the $\otimes$ introduction rule?''
+% Each linear context $\Delta$ is dependent on a non-linear context $\Gamma$, so
+% we may use variables drawn from $\Gamma$ when defining the types in $\Delta$.
+
+% In the grammar setting, this dependence lets
+% us use variables from regular dependent type theory to
+% keep track of data without \emph{consuming any resources}, so to speak. That is,
+% using a piece of a linear context consumes it, so we cannot linearly encode anything that
+% needs to be referenced more than once.
+
+% We will see a concrete instance of this dependence in \cref{subsubsec:pda}. In
+% that case, we encode the traces of a pushdown automaton as a grammar. When doing so,
+% we make use of our non-linear types to define a type of stacks. The fact that
+% our encoding of stacks is non-linear reflects
+% that a pushdown automaton may consume a character from the input string without
+% affecting the state of the stack.
+
+% \subsubsection{Other Type Constructors}
+% \label{subsubsec:other}
+
+% \paragraph{Brzozowksi Derivative}
+% In the 1950s Janusz Brzozowksi developed a notion of derivative for formal
+% languages \cite{brzozowskiDerivativesRegularExpressions1964}. The derivative of language $L$ by string $w$ is given as,
+
+% \[
+%   w^{-1}L = \{ w' \in \String : w \otimes w' \in L \}
+% \]
+
+% That is, the derivative of $L$ with respect to $w$ are precisely the strings
+% that complete to an $L$-parse when prefixed by a $w$.
+
+% Brzozowski initially gave an accounting of this operation for
+% generalized regular expressions, but later work demonstrates that the same
+% construction can be generalized to context free grammars \cite{mightParsingDerivativesFunctional2011}.
+
+% The ``magic wand'' from linear logic takes on the role of Brzozowski's
+% derivative in our syntax.
+
+% \begin{equation}
+%   \label{eq:linfun}
+% \inferrule{\Gamma ; a : A, \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A \lto B}
+% \end{equation}
+
+% We read this as saying that $\Delta$ constructs a $A \lto B$-parse when
+% prepending $\Delta$ with an $A$-parse completes to a $B$ parse. In this sense,
+% the type $A \lto B$ behaves like a linear function type. The only distinction
+% between $A \lto B$ and an intuitionistic function type is that linear functions
+% consume their input.
+
+% Note that the introduction rule presented in \cref{eq:linfun} does indeed
+% generalize Brzozowski's initial presentation of the derivative. First, to match
+% his initial presentation
+% let $A$ be the grammar accepting a single string $w$, which we may write as the
+% discrete grammar $\lceil w \rceil$.
+% Additionally, our construction lifts from a language-theoretic construction
+% to a grammar point-of-view.
+% In this sense, we carry around not just the data of what strings belong to
+% $\lceil w \rceil \lto B$ but also are endowed the data of a function that
+% constructs parse trees of $B$ from pairs of $\lceil w \rceil$-parses and
+% $(\lceil w \rceil \lto B)$-parses.
+
+% \paragraph{Top and Bottom}
+% We include linear versions of the unit and empty types. The
+% grammar $\top$ matches every input string, while the grammar
+% $\bot$ represents falsity and matches no strings. These two
+% are often used together to give syntactic shorthand to
+% grammars that encode propositions.
+
+% For instance, we may encode the proposition
+% $\mathsf{isVowel}$ as,
+
+% \[
+%   \mathsf{isVowel}(w) :=
+%     \begin{cases}
+%       \top, \quad w \in \{a, e, i, o, u\} \\
+%       \bot, \quad \text{otherwise}
+%     \end{cases}
+% \]
+
+% \paragraph{Fixed Point Operator}
+% We include a least fixed point operator on types, $\mu$, which is the means by
+% which we define recursive grammars.
+
+% Observe that we need not take the Kleene star as a primitive
+% grammar constructor, as it is definable as a fixed point.
+% The Kleene star of a grammar $g$ is given as,
+
+% \[
+%   g^{*} := \mu X . I \oplus (g \otimes X)
+% \]
+
+% \begin{figure}[h!]
+% \begin{mathpar}
+%   \inferrule
+%   {\Gamma ; \Delta \vdash p : I}
+%   {\Gamma ; \Delta \vdash \mathsf{nil} : g^{*}}
+
+%   \and
+
+%   \inferrule
+%   {\Gamma ; \Delta \vdash p : g \\ \Gamma ; \Delta' \vdash q
+%   : g^{*}}
+%   {\Gamma ; \Delta \vdash \mathsf{cons}(p , q) : g^{*}}
+
+%   \and
+
+%   \inferrule
+%   {
+%     \Gamma ; \Delta \vdash p : g^{*} \\
+%     \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
+%     \Gamma ; x : g , y : h \vdash p_{\ast} : h
+%   }
+%   {\Gamma ; \Delta \vdash \mathsf{foldr}(p_{\varepsilon} , p_{\ast}) : g^{*}}
+% \end{mathpar}
+% \caption{Kleene Star Rules}
+% \label{fig:star}
+% \end{figure}
+
+% Likewise, $g^{*}$ has admissible introduction and
+% elimination rules, shown in \cref{fig:star}. Note that this
+% definition of $g^{*}$ and these
+% rules arbitrarily assigns a handedness to the Kleene star.
+% We could have just as well took it to be a fixed point of
+% $I \oplus (X \otimes g)$. In fact, the definitions are
+% equivalent, as the existence of the $\mathsf{foldl}$ term below
+% shows that $g^{*}$ is also a fixed point of
+% $I \oplus (X \otimes g)$.
+
+% \begin{equation}
+%   \label{eq:foldl}
+%   \inferrule
+%   {
+%     \Gamma ; \Delta \vdash p : g^{*} \\
+%     \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
+%     \Gamma ; y : h , x : h \vdash p_{\ast} : h
+%   }
+%   {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
+% \end{equation}
+
+% In fact, the $\mathsf{foldl}$ term is defined using
+% $\mathsf{foldr}$ --- much in the same way one
+% may define a left fold over lists in terms of a right fold
+% in a functional programming language\footnote{The
+%   underlying trick is to fold over a list of functions
+%   instead of the original string. We curry each character $c$
+%   of the string into a function that concatenates $c$, and
+%   right fold over this list of linear functions. Because function
+%   application is left-associative, this results in a left
+%   fold over the original string. }.
+
+% We only take fixed points of a single variable as a
+% primitive operation in the type theory, but we may apply
+% Beki\`c's theorem \cite{Bekić1984} to define an admissible
+% notion of multivariate fixed point. This is particularly
+% useful for defining grammars that encode the states of an
+% automaton. In \cref{fig:multifix} we provide the
+% introduction and elimination principles for such a fixed
+% point, where $\sigma$ is the substitution that unrolls the
+% mutually recursive definitions one level. That is,
+
+% \begin{align*}
+%   \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \\
+%   & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
+% \end{align*}
+
+% \begin{figure}[h!]
+% \begin{mathpar}
+%   \inferrule
+%   {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
+%   {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
+
+%   \\
+
+%   \inferrule
+%   {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
+%              \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
+%              \dots \\
+%              \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
+%   }
+%   {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
+% \end{mathpar}
+% \caption{Multi-fixed Points}
+% \label{fig:multifix}
+% \end{figure}
+
+% The most prevalent application of these multi-fixed points
+% is for defining the traces through automata, as we will see
+% later in \cref{eq:nfatrace}.
+
+% \todoin{with citation, talk somewhere about how CFGs are just regular + fixed points}
+% \max{this is a good point to give an overview of how regular expressions and context-free grammars/expressions are essentially sub-languages of our type system.}
+
+% \subsection{Semantics in Grammars}
+% \label{subsec:seming}
+% As mentioned earlier, the interpretation of grammars as functions from
+% $\String$ to $\Set$ serves as our intended semantics. When designing the syntax of this formalism, we were often
+% informed by this model. That is, there is an interpretation
+% $\llbracket \cdot \rrbracket$ of our syntax that provides grammar-based
+% semantics.
+
+% \begin{enumerate}
+%   \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
+%   \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
+%         $\sem X : \sem \Gamma \to \Set_{i}$
+%   \item A non-linear term $\Gamma \vdash e : X$ denotes a section
+%         $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$
+%   \item Linear contexts $\Gamma \vdash \Delta$ and types
+%         $\Gamma \vdash A : L_{i}$ both denote families of grammars
+%         $\sem \Gamma \to \Gr_{i}$
+%   \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
+%         transformers
+%         $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
+% \end{enumerate}
+
+% Most of the non-linear semantics is standard, so let us investigate in more
+% depth the linear semantics. The interpretation of each
+% connective is exactly what is suggested by the syntax, in
+% large part to the inspiration taken from this interpretation
+% when designing the type theory.
+
+% \begin{enumerate}
+%   % \item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
+%   \item $\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+%         \begin{itemize}
+%           \item The semantics of the linear $\Pi$-type is
+%                 indeed a dependent function out of the
+%                 semantics of $X$.
+%         \end{itemize}
+%   \item $\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+%         \begin{itemize}
+%           \item The semantics of the linear $\Sigma$-type is
+%                 a dependent pair.
+%         \end{itemize}
+%   \item $\sem{I} \gamma w = \{ () \pipe w = \epsilon \}$
+%         \begin{itemize}
+%           \item The only string accepted by $I$ is the empty string.
+%         \end{itemize}
+%   \item
+%         $\sem{A \otimes B} \gamma w = \Sigma(w1,w2:\Sigma^*) \left( (w_{1}w_{2} = w) \wedge \sem{A} \gamma w_1 \times \sem{B} \gamma w_2 \right)$
+%         \begin{itemize}
+%           \item The string $w$ matches $A \otimes B$
+%                 precisely when it may be split into two
+%                 pieces such that the left one matches $A$
+%                 and the right one matches $B$.
+%         \end{itemize}
+%   \item
+%         $\sem{A \lto B} \gamma w = \Pi(w_a:\Sigma^*) \left( A \gamma w_a \Rightarrow B\gamma (w_aw) \right)$
+%         \begin{itemize}
+%           \item The semantics of the linear function type is
+%                 a dependent function that turns $A$-parses
+%                 of a substring into $B$-parses of an
+%                 extended string.
+%         \end{itemize}
+%   \item
+%         $\sem{B \tol A} \gamma w = \Pi(w_a:\Sigma^*) \left( A \gamma w_a \Rightarrow B\gamma (ww_a) \right)$
+%   \item $\sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x)$
+%         \begin{itemize}
+%           \item The semantics of the least fixed point
+%                 operator is indeed the least fixed point of sets.
+%         \end{itemize}
+% \end{enumerate}
+
+% \subsection{Semantics in Languages}
+% Recall from \cref{sec:synindsyn} that every formal grammar induces a formal language. This corresponds to another
+% interpretation $L(\cdot)$ of our syntax. In particular, the semantics of a
+% linear type $A : L_{i}$
+% from \cref{subsec:seming} are augmented such that
+
+% \[
+%   L(A) = \{ w \in \String : \sem A w \text{ is nonempty} \}
+% \]
+
+% \subsubsection{Weak Equivalence}
+% Recall that an isomorphism of grammars defined a strong equivalence. We may
+% additionally define a \emph{weak equivalence} of grammars when two grammars
+% define the same language.
+
+% That is, $A$ and $B$ are weakly equivalent when $L(A) = L(B)$. Note that this is
+% a semantic notion, but this just as well captured syntactically by the existence
+% of two judgments $\Gamma ; x : A \vdash M : B$ and
+% $\Gamma ; x : B \vdash N : B$, which are not necessarily mutually inverse.
+
+% \section{Automata as Grammars}
+% Classically, formal language theory is closely related to the study of automata.
+% In the Chomsky hierarchy, each language class is associated to a class of
+% automata that serve as recognizers. Internal to the type theory of \cref{sec:tt},
+% we can characterize these language classes syntactically; moreover, we
+% demonstrate the equivalence of these language classes to its associated automata class as a
+% proof term within our logic.
+
+% Internalizing these equivalences is particularly useful for building a regular
+% expression parser. That is, it is relatively easy for use to write down a parser
+% for deterministic finite automata (DFAs). Then we may compose this parser with
+% a proof equivalence between DFAs and regular expressions to extract a verified
+% regular expression parser.
+
+% As we will see in \cref{subsubsec:tm}, we may very quickly encode arbitrary
+% recursively enumerable languages --- since we can simulate the
+% tape of Turing machine in the dependent fragment of the language. We reflect
+% here on this expressive power to motivate why we put stringent syntactic
+% restrictions on particular classes of automata. The type theory we present truly
+% is general enough to easily capture all of the Chomsky hierarchy.
+% To begin, let's focus on the simplest languages and their corresponding
+% automata --- regular expressions and finite automata.
+
+% \subsection{Non-deterministic Finite Automata}
+% \label{subsec:finiteaut}
+% \begin{figure}
+%   \begin{tikzpicture}[node distance = 25mm ]
+%     \node[state, initial, accepting] (1) {$1$};
+%     \node[state, below left of=1] (2) {$2$};
+%     \node[state, right of=2] (3) {$3$};
+
+%     \path[->] (1) edge[above] node{$b$} (2)
+%               (1) edge[below, bend right, left=0.3] node{$\epsilon$} (3)
+%               (2) edge[loop left] node{$a$} (2)
+%               (2) edge[below] node{$a, b$} (3)
+%               (3) edge[above, bend right, right=0.3] node{$a$} (1);
+%   \end{tikzpicture}
+%   \caption{An example NFA}
+%   \label{fig:NFA}
+% \end{figure}
+
+% Classically, a \emph{nondeterministic finite automaton} (NFA) is a finite state machine where
+% transitions are labeled with characters from a fixed alphabet $\Sigma$. These
+% are often represented formally as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$,
+
+% \begin{itemize}
+%   \item $Q$ a finite set of states
+%   \item $\Sigma$ a fixed, finite alphabet
+%   \item $\delta : Q \times (\Sigma \cup \{ \varepsilon\}) \to \mathcal{P}(Q)$ the labeled transition function
+%   \item $q_{0} \in Q$ the start state
+%   \item $F \subset Q$ a set of accepting states
+% \end{itemize}
+
+% Intuitively, this can be thought of like a directed graph with nodes in $Q$ with
+% an edge $q \overset{c}{\to} q'$ whenever $q' \in \delta(q, c)$. Note that
+% transitions in an NFA may be labeled with the empty string $\varepsilon$ --- such
+% transitions are referred to as \emph{$\varepsilon$-transitions}. We may see an
+% example of an NFA in \cref{fig:NFA}.
+
+% From an NFA, we may construct a grammar of traces as follows:
+
+% First, we define a mutual fixed point grammar that describes the traces through the NFA.\ Then, we have another
+% grammar that tells us if we're currently in an accepting state of the automaton.
+% A parse of the NFA grammar is then a pair of a trace and the data that we're in
+% an accepting state.
+
+% Consider the NFA $N$ pictured in \cref{fig:NFA}. There are three states, $1$, $2$,
+% and $3$. We introduce a non-linear type $Q$ with three inhabitants $q_{1}$,
+% $q_{2}$, and $q_{3}$ to represent each of these states, respectively. Given a
+% $q, q'$, we can
+% then define the type of traces from $q$ to $q'$. For instance, let's construct
+% the traces starting at $q_{1}$ and ending at $q_{2}$ as an example.
+
+% \begin{equation}
+%   \label{eq:nfatrace}
+%   \mathsf{Trace}_{N}(q_{1}, q_{2}) = \mu
+%   \begin{pmatrix}
+%     g_{q_{1}} := g_{q_{3}} \oplus ( b \otimes g_{q_{2}} ) \\
+%     g_{q_{2}} := ( a \otimes g_{q_{2}} ) \oplus ( a \otimes g_{q_{3}} ) \oplus ( b \otimes g_{q_{3}} ) \oplus I \\
+%     g_{q_{3}} := a \otimes g_{q_{1}} \\
+%   \end{pmatrix}. g_{q_{1}}
+% \end{equation}
+
+% We should read this as defining three mutually recursive grammars, one for each
+% state. The definitions of these mutually recursive grammars capture the
+% transitions of the automaton. To ensure that we only encode traces that end in
+% state 2, we only include the unit $I$ as a summand in
+% $g_{q_{2}}$. That is, by only including $I$ at this location, it makes it the
+% only place where derivations of $\mathsf{Trace}(q_{1}, q_{2})$ can terminate. We
+% can think of all of these definitions underneath
+% of the $\mu$ binder as bringing some local grammars into scope. With these local
+% grammars in scope we are ultimately constructing a term of type $g_{q_{1}}$ to denote
+% only traces starting in state 1.
+
+% To encode the acceptance criteria of $N$, we want to internalize a proposition
+% over each state of the NFA.\ That is, for each $q \in Q$ we define a term
+
+% \[
+%   \mathsf{acc}(q) := q \text{ is accepting}
+% \]
+
+% In this example, $\mathsf{acc}(q_{1}) = \top$ and
+% $\mathsf{acc}(q_{2}) = \mathsf{acc}(q_{3}) = \bot$. An accepting trace of the NFA $N$ is then given by the following dependent grammar,
+% \[
+%  \mathsf{AccTrace}_{N} := \LinSigTy q Q {\left( \mathsf{Trace}_{N}(q_{0} , q) \amp \mathsf{acc}(q) \right)}
+% \]
+
+% where $q_{0}$ is the initial state. That is, a trace is accepted by the NFA if
+% we can construct the trace, and the trace ends at an accepting state. This idea
+% is simple enough and aligns with how we intuitively treat these automata.
+
+% Generalizing over the above example, we want to define the type of traces as,
+
+% \[
+%   \mathsf{Trace}_{N}(q_{0}, q_{1}) = \mu
+%   \begin{pmatrix}
+%     g_{q} := \mathsf{Trans}(q), & q \neq q_{1} \\
+%     g_{q} := \mathsf{Trans}(q) \oplus I , & q = q_{1}
+%   \end{pmatrix}. g_{q_{0}}
+% \]
+
+% where $\mathsf{Trans}(q)$ is an iterated disjunction that describes which
+% transition you should take. For an NFA, $\mathsf{Trans}(q)$ must take on the
+% following syntactic form,
+
+% \begin{gather*}
+%  \mathsf{State} \in \{ g_{q} : q \in Q \} \\
+%  \mathsf{Char} \in \Sigma \\
+%  \mathsf{Trans}_{N}(q) ::= \mathsf{Char} \otimes \mathsf{State}~|~\mathsf{State}~|~\mathsf{Trans}_{N}(q) \oplus \mathsf{Trans}_{N}(q)
+% \end{gather*}
+
+% That is, $\mathsf{Trans}(q)$ is a disjunction of literals followed by grammars
+% that encode states.
+
+% When conducting proofs involving NFA grammars, we often to
+% need either construct terms of type $\mathsf{Trace}_{N}(q , q')$. To
+% this end, we give three admissible rules for constructing
+% traces,
+
+% \begin{figure}[h!]
+%   \label{fig:admissibleintro}
+%   \begin{mathpar}
+%     \inferrule
+%     {~}
+%     {\Gamma ; \cdot \vdash \mathsf{nil} : \mathsf{Trace}_N
+%       (q , q)}
+
+%     \and
+
+%     \inferrule
+%     {\Gamma ; \Delta \vdash M : \mathsf{Trace}_N
+%       (dst , q') \\
+%     \exists \text{ transition } src \overset{c}{\to} dst
+%     }
+%     {\Gamma ; x : c , \Delta \vdash \mathsf{cons}(M) : \mathsf{Trace}_N
+%     (src , q')}
+
+%     \and
+
+%     \inferrule
+%     {\Gamma ; \Delta \vdash M : \mathsf{Trace}_N (dst , q')
+%       \\
+%     \exists~\varepsilon\text{-transition } src
+%     \overset{\varepsilon}{\to} dst}
+%     {\Gamma ; \Delta \vdash \mathsf{\varepsilon cons}(M) : \mathsf{Trace}_N
+%     (src , q')}
+%   \end{mathpar}
+%   \caption{Admissible Trace Constructors}
+% \end{figure}
+
+% That is, we may use $\mathsf{nil}$ to terminate a trace that
+% begins and ends at state $q$. The rules $\mathsf{cons}$ and
+% $\mathsf{\varepsilon cons}$ are then used to inductively
+% stitch traces together when sound. For instance, we can read
+% the $\mathsf{cons}$ rule as saying that we can
+% create a trace coming from a state $src$ provided that we may first
+% transition via the character $c$ to state $dst$ and
+% inductively build a trace starting from $dst$. The
+% $\mathsf{\varepsilon cons}$ rule says something similar, but
+% with an $\varepsilon$-transition instead of a transition
+% labelled by a character. Recall that
+% these rules are \emph{admissible}. They are not strictly
+% necessary as primitives to conduct our proofs; however, they
+% do provide convenient shorthand notation for building terms
+% of type $\mathsf{Trace}_{N}(q , q')$.
+
+% Dual to constructing traces, we often want to construct
+% other terms in a context containing values of type
+% $\mathsf{Trace}_{N}(q , q')$. For this purpose, we make use
+% of the elimination principle for multiple-fixed points ---
+% which we write as $\mathsf{mfold}$ --- given
+% in \cref{subsubsec:other}.
+
+% Note that this general construction will readily generalize to other types of
+% automata. If we wanted to define say deterministic finite automata\footnote{DFAs
+% \emph{could} just be defined as NFAs that happen to be deterministic. This is
+% one way to do so, or you may choose to present the transition relation for the
+% automaton as a transition \emph{function} instead. Concerns like these don't
+% matter so much when defining things on paper, but at formalization time these
+% are important and can make some proofs much easier.}, pushdown automata, Turing
+% machines, etc, we just swap out the type of traces for a different, but very
+% similar, trace construction.
+
+% With this setting for finite automata, we can now internalize some classical theorems inside of our formal system.
+
+% \subsection{A DFA Parser}
+% \label{subsec:regexparser}
+
+% Just as we encoded traces of NFAs as grammars, we likewise
+% encode the traces of a DFA as grammars. The key difference
+% between NFAs and DFAs is \emph{determinism} --- meaning,
+% that in a state $q$ inside of DFA, given a character $c$ there
+% will be exactly one transition that we may take leaving $q$
+% with label $c$. For us, this changes the definition of valid
+% transitions for a DFA, instead of the definition of
+% $\mathsf{Trans}$ provided in \cref{subsec:finiteaut} DFAs
+% obey
+
+% \begin{gather*}
+%  \mathsf{State} \in \{ g_{q} : q \in Q \} \\
+%  \mathsf{Trans}(q) ::= \bigoplus_{c \in \Sigma} (c \otimes \mathsf{State})
+% \end{gather*}
+
+% Meaning, each state has a transition for every character.
+
+% We now wish to define a parser term for DFA grammars. In
+% particular, for a DFA $D$ we want to build a term,
+
+% \[
+%   w : \String \vdash \mathsf{parse}_{D} : \mathsf{AccTrace}_{D} \oplus \top
+% \]
+
+% where left injection into the output type denotes acceptance
+% by the parser, and right injection denotes rejection. To
+% build such a parser, it will suffice to construct a term
+
+% \[
+%   w : \String \vdash \mathsf{parse}_{D} : \LinSigTy q Q {\mathsf{Trace}_{D}(q_{0} , q)}
+% \]
+% This is because given a trace of a DFA, we may easily check
+% if we should accept or reject by simply testing
+% if the final state is accepting.
+
+% Because $w$ is a Kleene star of characters, we may construct
+% our desired $\mathsf{parse}_{D}$ term as a $\mathsf{foldl}$
+% over $w$. In the empty case, we just have the trace that
+% ends at the accepting state. In the recursive case, we
+% effectively add to our trace by transitioning one character
+% at a time, as we read them moving across $w$.
+
+% \todoin{Derivation tree for this construction}
+
+% Perhaps this derivation is not too surprising. All it says
+% is that a DFA may be ran by transitioning a single character
+% at a time, and then accepting or rejecting based on the
+% final state. This is exactly what DFAs did initially, so
+% what did we gain? Well, this has the benefit of our type
+% system to ensure its correctness. Moreover, this construction exports to an
+% intrinsically verified and executable DFA via Agda.
+
+% \subsection{Regular Grammars and DFAs}
+% \label{subsec:deriv}
+
+% In order to extend the DFA
+% parser from \cref{subsec:regexparser} to the construction of
+% a verified parser
+% generator for regular grammars we need to perform some
+% plumbing establishing an equivalence between regular
+% expressions and DFAs.
+
+% There are several routes we may hope to take in establishing
+% this equivalence. First, we could prove an equivalence
+% between NFAs and regular expressions, and separately prove
+% an equivalence between NFAs and DFAs.
+% In \cref{subsec:eqproofs}, we include a version of
+% Thompson's construction --- which established the
+% equivalence between regular grammars and DFAs. We may additionally
+% hope to internalize a variant of the powerset construction \cite{rabinFiniteAutomataTheir1959}
+% --- which takes as input an NFA and constructs a DFA that
+% recognizes the same language --- and combine the results of
+% Thompson's construction and the powerset constructions to give an equivalence
+% between regular expressions and DFAs.  This route is alluring, as it
+% internalizes several classic grammar-theoretic constructions. However, it may
+% necessitate extensions to the LNL theory, like
+% a propositional truncation, and we have not yet investigated
+% how this would interact with the existing types in the
+% theory. The addition of a propositional truncation may seem
+% harmless, but it is not always immediately clear how
+% distinct constructions will interact. For instance, when
+% exploring LNL models, Benton discovered that the synthesis
+% of linear and dependent types require a new presentation of
+% the $!$ modality from linear logic
+% \cite{bentonMixedLinearNonlinear1995}. That is all to say,
+% this is a work in progress and
+% it is not immediate that the addition of a propositional
+% truncation is adequate for establishing the weak equivalence
+% between NFAs and DFAs.
+
+% We may instead hope to internalize an equivalence between
+% regular grammars and DFAs by using Brzozowski derivatives to
+% directly create a DFA that is weakly equivalent to a given
+% regular expression, as described by Owens et al.
+% \cite{owensRegularexpressionDerivativesReexamined2009}.
+% One characterization of regular grammars is that they are
+% precisely those grammars which have finitely many inequivalent Brzozowski
+% derivatives
+% \cite{brzozowskiDerivativesRegularExpressions1964}.
+% The algorithm used by Owens takes in a
+% regular grammar and generates a DFA that recognizes the same
+% language, and the states in this DFA are the finitely many
+% derivative equivalence classes. We initially had a version of
+% this theorem very roughly internalized in the LNL theory.
+% To our taste, too much of this presentation relied on
+% meta-arguments that lived outside of
+% our formalism, and thus this particular phrasing of the
+% theorem did not translate well into formalization.
+
+% In any case, we believe
+% that revisiting these lines of thought will lead to a
+% satisfactory internalization of the equivalence between
+% regular grammars and DFAs, and thus would bridge the gap
+% between our DFA parser and a full regular expression parser.
+
+% \subsection{Equivalence Between Regular Grammars and Finite Automata}
+% \label{subsec:eqproofs}
+% In this section, we describe a version of Thompson's
+% construction \cite{thompsonProgrammingTechniquesRegular1968}
+%    where we construct an NFA that recognizes a given regular
+% expression. Moreover, we will show that this NFA is strongly equivalent to the
+% original grammar. Witnessing this construction in our syntax has two benefits
+% \begin{enumerate}
+%   \item It reinforces this high-level view that the syntax is a natural and
+%         general setting for formal grammar reasoning, as we demonstrate that
+%         this formal system subsume results from existing systems, and
+%   \item Following the
+%         development of Thompson's construction, we then need
+%         only establish the equivalence of NFAs and DFAs to
+%         complete the full regular expression parser
+% \end{enumerate}
+
+% \begin{theorem}[Thompson]
+%   \label{thm:thompson}
+%   For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same
+%   language as $g$.
+% \end{theorem}
+
+% We make a pretty straightforward adaptation of Thompson's theorem to our setting,
+
+% \begin{theorem}[Typed Thompson]
+%   \label{thm:typthompson}
+%   For $g$ a regular grammar $g$, there is an NFA $N$ such that $g$ is isomorphic
+%   to $\mathsf{AccTrace}_{N}$.
+% \end{theorem}
+
+% \begin{proof}[Proof Sketch]
+%   The following proof is currently formalized in Agda. Below
+%   we give a walk through the structure of the proof.
+
+%   Recall that regular grammars are inductively defined via
+%   disjunction, concatenation, and Kleene star over literals
+%   and the empty grammar. By induction over regular grammars,
+%   we will construct an NFA that is equivalent to $g$.
+
+%   First, define the recognizing NFA for the empty grammar
+%   $I$.
+
+%   \begin{figure}[h!]
+%   \begin{tikzpicture}[node distance = 25mm ]
+%     \node[state, initial] (1) {$1$};
+%     \node[state, right of=1, accepting] (2) {$2$};
+
+%     \path[->] (1) edge[below] node{$\varepsilon$} (2);
+%   \end{tikzpicture}
+%   \caption{$NFA(I)$}
+%   \label{fig:emptyNFA}
+%   \end{figure}
+
+%   The type of traces from the initial state of $NFA(I)$ to the single
+%   accepting state is given by,
+
+%   \[
+%     \mathsf{Trace}_{NFA(I)}(q_{1}, q_{2}) = \mu
+%       \begin{pmatrix}
+%          g_{q_{1}} := g_{q_{2}} \\
+%          g_{q_{2}} := I
+%       \end{pmatrix}. g_{q_{1}}
+%   \]
+
+%   The accepting traces through $NFA(I)$ are then described
+%   as,
+
+%   \[
+%     \mathsf{AccTrace}_{NFA(I)} = \LinSigTy q {\{1 , 2\}} {\left( \mathsf{Trace}_{NFA(I)}(q_{1} , q) \amp \mathsf{acc}(q) \right)}
+%   \]
+
+%   A quick inspection of \cref{fig:emptyNFA} reveals that the
+%   only reasonable choice for $q$ is $q_{2}$ --- because
+%   state 2 is accepting while state 1 is not. Therefore,
+
+%   \begin{align*}
+%     \mathsf{AccTrace}_{NFA(I)}
+%     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \amp \mathsf{acc}(q_{2}) \\
+%     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \amp \top \\
+%     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})
+%   \end{align*}
+
+%   From here, to prove
+%   $I \cong \mathsf{AccTrace}_{NFA(I)}$ it suffices to show
+%   $I \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$. Below we
+%   give two parse transformers, one from $I$ to
+%   $\mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$ and vice versa. The formalized
+%   artifact provides the low-level details that these are
+%   indeed mutually inverse.
+
+%   \[
+%     \inferrule
+%     {p : I \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
+%      \exists \text{~transition~} q_{1} \overset{\varepsilon}{\to} q_{2}
+%     }
+%     {p : I \vdash \mathsf{\varepsilon cons}(\mathsf{nil}) : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})}
+%   \]
+
+%   Let $\gamma$ be the substitution $\{ g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$,
+
+%   \[
+%     \inferrule
+%     {
+%       p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash p :
+%         \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \\
+%       x_{1} : \simulsubst {g_{q_{2}}} {\gamma} = I \vdash x_{1} : I \\
+%       x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
+%     }
+%     {
+%       p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.x_{1} , x_{2}.x_{2})(p) : I
+%     }
+%   \]
+
+%   This concludes the proof for the case of the empty
+%   grammar. Let's now walk through the construction for
+%   literal grammars. Given a character $c$, we construct an
+%   NFA that recognizes only the string containing the single
+%   character $c$ as,
+
+%   \begin{figure}[h!]
+%   \begin{tikzpicture}[node distance = 25mm ]
+%     \node[state, initial] (1) {$1$};
+%     \node[state, right of=1, accepting] (2) {$2$};
+
+%     \path[->] (1) edge[below] node{$c$} (2);
+%   \end{tikzpicture}
+%   \caption{$NFA(c)$}
+%   \label{fig:literalNFA}
+%   \end{figure}
+
+%   The automaton in \cref{fig:literalNFA} induces the
+%   following type of traces from $q_{1}$ to $q_{2}$.
+
+%   \[
+%     \mathsf{Trace}_{NFA(c)}(q_{1}, q_{2}) = \mu
+%       \begin{pmatrix}
+%          g_{q_{1}} := c \otimes g_{q_{2}} \\
+%          g_{q_{2}} := I
+%       \end{pmatrix}. g_{q_{1}}
+%   \]
+
+%   Through the same argument as the empty grammar, the only
+%   state that is accepting is $q_{2}$ and thus,
+
+%   \[
+%     \mathsf{AccTrace}_{NFA(c)} \cong \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})
+%   \]
+
+%   To show the desired isomorphism of $c \cong NFA(c)$ we
+%   make a similar argument as we did for the empty grammar
+%   $I$, except we leverage the $\mathsf{cons}$ rule instead
+%   of $\mathsf{\varepsilon cons}$. That is, the parse
+%   transformers in either direction are given as,
+
+%   \[
+%     \inferrule
+%     {\cdot \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
+%      \exists \text{~transition~} q_{1} \overset{c}{\to} q_{2}
+%     }
+%     {p : c \vdash \mathsf{cons}(\mathsf{nil}) :
+%       \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})}
+%   \]
+
+%   \[
+%     \inferrule
+%     {
+%       p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash p :
+%         \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \\
+%       x_{1} : \simulsubst {(c \otimes g_{q_{2}})} {\gamma} = c \otimes I \vdash \mathsf{unitR}(x_{1}) : c \\
+%       x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
+%     }
+%     {
+%       p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.\mathsf{unitR}(x_{1}) , x_{2}.x_{2})(p) : c
+%     }
+%   \]
+
+%   Where $\gamma$ is the substitution
+%   $\{ c \otimes g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$ and
+%   $\mathsf{unitR}$ is a witness to the isomorphism
+%   $c \otimes I \cong c$. Again, we may see that these do
+%   indeed mutually invert each other in the Agda code.
+
+%   It remains to show that the desired isomorphisms are
+%   preserved by $\otimes$, $\oplus$, and Kleene star. Here,
+%   we will give the argument for the disjunction case, the
+%   others are defined quite similarly.
+
+%   Given two NFAs $N$ and $M$, we may define a new NFA that
+%   encodes the disjunction of $N$ and $M$. Denote the
+%   internal states of $N$ by $q_{j}$'s and the internal states
+%   of $M$ by $r_{k}$'s,
+
+%   \begin{figure}[h!]
+%   \begin{tikzpicture}[node distance = 20mm ]
+%     \node[state] (2) {$q_{init}$};
+%     \node[state, initial, below left of=2] (1) {$init$};
+%     \node[state, below right of=1] (3) {$r_{init}$};
+%     \node[right of=2] (4) {$\cdots$};
+%     \node[right of=3] (5) {$\cdots$};
+%     % \node[state, below right of=1] (3) {$3$};
+%     % \node[state, below of=2] (4) {$4$};
+
+%     \path[->] (1) edge[below] node{$\varepsilon$} (2);
+%     \path[->] (1) edge[below] node{$\varepsilon$} (3);
+%     \path[->] (2) edge[below] node{} (4);
+%     \path[->] (3) edge[below] node{} (5);
+
+%     \node[label={[name=l] $N$}, draw,line width=2pt,rounded corners=5pt, fit=(2)(4)] {};
+%     \node[label={[name=l] $M$}, draw,line width=2pt,rounded corners=5pt, fit=(3)(5)] {};
+%   \end{tikzpicture}
+%   \caption{$N \oplus_{NFA} M$}
+%   \label{fig:disjunctionNFA}
+%   \end{figure}
+
+%   \cref{fig:disjunctionNFA} shows the process of
+%   disjunctively combining NFAs. Precisely, we add a single
+%   new state and we included copies of the states from each of $N$
+%   and $M$. The new state acts as the initial state and has
+%   $\varepsilon$-transitions to the initial states of $N$ and
+%   $M$. We include all of the internal transitions from $N$
+%   and $M$, and the accepting states of $N \oplus_{NFA} M$
+%   are exactly the accepting states in each subautomaton.
+
+%   Let $g$ and $g'$ be two regular grammars such that
+%   $g \cong NFA(g)$ and $g' \cong NFA(g')$. As a matter of
+%   notation\footnote{We shall similarly abuse notation for
+%     $\otimes$ and Kleene. That is, for a regular grammar
+%     $g$, when we write $NFA(g)$ we mean the NFA inductively
+%     built up with the NFA-analogues to the constructors that
+%     built up $g$.}, we will write $NFA(g \oplus g')$ for
+%   $NFA(g) \oplus_{NFA} NFA(g')$. The traces of
+%   $NFA(g \oplus g')$ are then given by,
+
+%   \[
+%     \mathsf{Trace}_{NFA(g \oplus g')}(src , dst) = \mu
+%       \begin{pmatrix}
+%         g_{init} := g_{q_{0}} \oplus g_{r_{0}} \\
+%         g_{q_{j}} := \mathsf{Trans}_{NFA(g)}(q_{j}) \oplus \mathsf{isDst}(q_{j}) \\
+%         g_{r_{k}} := \mathsf{Trans}_{NFA(g')}(r_{k}) \oplus \mathsf{isDst}(r_{k})
+%       \end{pmatrix}.g_{src}
+%     \]
+
+%   where $\mathsf{Trans}$ is used to echo the same syntactic
+%   definitions that appear in the $NFA(g)$ and $NFA(g')$.
+%   Also, $src$ and $dst$ may take on any value in
+%   $Q := \{init\} \cup \{q_{j}\} \cup \{r_{k}\}$, and
+%   $\mathsf{isDst}(q)$ checks if $q$ is equal to $dst$. Which
+%   is all to say, the traces of $NFA(g \oplus g')$ comprise
+%   either a trace of $NFA(g)$, or a trace of $NFA(g')$, and
+%   the transition coming out of $g_{init}$ determines which
+%   subautomaton we step into.
+
+%   The parse transformer from $g \oplus g'$ checks which side
+%   of the sum type we are on, then takes the appropriate step
+%   from $g_{init}$ in the automaton.
+
+%   \[
+%     \inferrule
+%     {
+%       u : g \vdash \iota (\phi (u)) : \mathsf{AccTrace}_{NFA(g \oplus g')} \\
+%       v : g' \vdash \iota' (\psi (v)) : \mathsf{AccTrace}_{NFA(g \oplus g')}
+%     }
+%     {p : g \oplus g' \vdash \mathsf{case}~p \{ \mathsf{inl}(u) \mapsto s , \mathsf{inr}(v) \mapsto r \} : \mathsf{AccTrace}_{NFA(g \oplus g')}}
+%   \]
+
+%   with $\iota$ and $\iota'$ as embeddings from $NFA(g)$ and
+%   $NFA(g')$, respectively, into $NFA(g \oplus g')$,
+%   $\phi: g \cong \mathsf{AccTrace}_{NFA(g)}$, and $\psi: g' \cong \mathsf{AccTrace}_{NFA(g')}$. On a
+%   high level, all this construction does is turn a parse of
+%   $g$ into a parse of $NFA(g)$ and then embeds that inside
+%   of the larger automaton $NFA(g \oplus g')$. Likewise for $g'$.
+
+%   In the other direction, recall that the data of an
+%   accepting trace for $NFA(g \oplus g')$ is a pair of a
+%   trace and a proof that
+%   the end state $q'$ of that trace is accepting. By
+%   multifolding over the first part of that pair, we turn the
+%   term of type
+%   $\mathsf{Trace}_{NFA(g \oplus g')}(init , q')$ into a
+%   trace of either of the subautomata,
+
+%   \[
+%     p : \mathsf{Trace}_{NFA(g \oplus g')}(init , q') \vdash \mathsf{mfold}_{NFA(g \oplus g')} : \mathsf{Trace}_{NFA(g)}(q_{0} , q') \oplus \mathsf{Trace}_{NFA(g')}(r_{0} , q')
+%   \]
+
+%   Additionally, we leverage the fact that the only accepting
+%   states for $NFA(g \oplus g')$ are those from the
+%   subautomata to extract that $q'$ must be an accepting
+%   state from a subautomaton.
+
+%   \[
+%     x : \mathsf{acc}_{NFA(g \oplus g')}(q') \vdash M : \mathsf{acc}_{NFA(g)}(q') \oplus \mathsf{acc}_{NFA(g')}(q')
+%   \]
+
+%   We then combine the trace and proof of acceptance into an
+%   accepting trace of one of the subautomata,
+
+%   \[
+%     p : \mathsf{AccTrace}_{NFA(g \oplus g')} \vdash N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')}
+%   \]
+
+%   Lastly, we then inductively use the isomorphisms $\phi$
+%   and $\psi$ to turn the accepting traces into a parse of
+%   $g$ or $g'$,
+
+%   \[
+%      N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')} \vdash \mathsf{case}~N~\{\mathsf{inl}(n) \mapsto \phi^{-1}(n), \mathsf{inr}(n') \mapsto \psi^{-1}(n')\} : g \oplus g'
+%   \]
+
+%   As discussed above, the other cases and low-level details
+%   of isomorphism are reserved for the formalization. This
+%   concludes the proof of our variant of Thompson's construction.
+% \end{proof}
+
+% \subsection{Other Automata}
+% \subsubsection{Pushdown Automata}
+% \label{subsubsec:pda}
+% A (nondeterministic) \emph{pushdown automaton} is an automaton that employs a
+% stack. Just like NFAs, they have transitions labeled with characters from a
+% fixed string alphabet $\Sigma$. Additionally, they maintain a stack of
+% characters drawn from a stack alphabet $\Gamma$. They are often represented
+% formally as a 7-tuple $(Q, \Sigma, \Gamma, \delta, q_{0}, Z, F)$,
+
+% \begin{itemize}
+%   \item $Q$ a finite set of states
+%   \item $\Sigma$ a fixed, finite string alphabet
+%   \item $S$ a fixed, finite stack alphabet
+%   \item
+%         $\delta \subset Q \times (\Sigma \cup \{ \varepsilon \}) \times S \to \mathcal{P}(Q \times S^{*})$
+%         the transition function
+%   \item $q_{0} \in Q$ the start state
+%   \item $Z \in S$ the initial stack symbol
+%   \item $F \subset Q$ the accepting states
+% \end{itemize}
+
+% We encode the traces of a pushdown automaton very similarly to those of an NFA,
+% except the transitions of a PDA are instead encoded via the linear-non-linear
+% $\Pi$-type. This is because of simply transitioning via character, a PDA must
+% also pop and push characters onto a stack, which is used as the argument to
+% these dependent functions.
+
+% Let $S$ be a non-linear type encoding the stack
+% alphabet, and build lists over $S$ as the (non-linear) least fixed-point
+% $\mathsf{List}(S) := \mu X . 1 + S \times X$. Then, the type of states for a
+% PDA $P$ with stack alphabet $S$ are given as a functions that takes in lists $L$,
+% and then makes a case distinction between possible transitions based on what was witnessed as $\mathsf{head}(L)$. The choice of transition
+% will then determine which character to transition by and what word $w$ should be
+% pushed onto the stack. The word that is added to the top of the stack is
+% appended to $\mathsf{tail}(L)$ and then we recursively step into another state
+% called on argument $w + \mathsf{tail}(L)$.
+
+% % \begin{gather*}
+% %   \mathsf{State} \in {g _{q} : q \in Q} \\
+% %   \mathsf{Word} \in \String \\
+% %   \mathsf{Char} \in \Sigma   \\
+% %   \mathsf{StackChar} \in \Gamma \\
+% % \end{gather*}
+% % \begin{align*}
+% %   \mathsf{Trans}_{P}(q) ::=~&
+% %                            \LinPiTy {(hd :: tl)} {\mathsf{List}(S)} {\left( \mathsf{Char} \otimes \mathsf{State}(\mathsf{Word} + tl) \right)}~| \\
+% %   & \LinPiTy {(hd :: tl)} {\mathsf{List}(S)} {\left(  \mathsf{State}(\mathsf{Word} + tl) \right)}~|~\mathsf{Trans}_{P}(q) \oplus \mathsf{Trans}_{P}(q)
+% % \end{align*}
+
+% % That is to say, when transitioning a PDA pops off the head $hd$ of the stack
+
+% \subsubsection{Turing Machines}
+% \label{subsubsec:tm}
+
+% In
+% \cref{subsubsec:pda}, we gave a grammar presentation of
+% traces through a PDA by using a non-linear type $S$
+% to encode the stack. We may similarly use pairs $S \times S$
+% to encode the tape of a Turing machine. With two stacks we can simulate the behavior of the
+% infinite tape of a Turing machine. The intuition behind this correspondence is
+% that the left half of the tape is on one stack, the right
+% half of the tape the other, and we treat the tops of stacks
+% like the head of the tape.
+
+% Due to computability limitations, we are not able to extract
+% a verified parser for all recursively enumerable languages.
+% Yet we should still reflect on the generality of our syntax,
+% as it allows a natural encoding of Turing machines.
+
+% \section{Future Work}
+% \label{sec:future}
+
+% \subsection{Implementation for Context-Free Grammars}
+% As suggested throughout the paper and briefly explored in
+% \cref{subsubsec:pda}, the first extension to the work in
+% this paper will be to bring analogous constructions to
+% context-free grammars and their accompanying pushdown automata.
+
+% \subsection{Beyond Strings}
+% \label{subsec:beyond}
+
+% While parsing typically refers to the generation of semantic
+% objects from string input, many tasks in programming can be
+% viewed as parsing of objects with more structure, such as
+% trees with binding structure or graphs. Fundamental to the
+% frontend of many
+% programming language implementations are type systems. In
+% particular, \emph{type checking}
+% --- analogous to language recognition --- and \emph{typed
+%   elaboration} --- analogous to parsing --- arise when
+% producing a semantic object subject to some analysis. Just
+% as our string grammars were given as functors from $\String$
+% to $\Set$, we envision adapting the same philosophy
+% to functors from $\String$ to \emph{trees} to craft a syntax
+% that natively captures typed elaboration. This suggests an
+% unusual sort of bunched type theory, where context extension
+% no longer resembles concatenation of strings but instead
+% takes on the form of tree constructors.
+
+% Thus far, our theory has proved useful for internalizing long-standing
+% grammar-theoretic constructions, but there has been decades of research
+% conducted since. A fruitful avenue for future work includes
+% testing if our formalism can also internalize more recently proposed grammar
+% mechanisms, such as the interval parsing grammars given by Zhang et al. \cite{zhangIntervalParsingGrammars2023}.
+
+% \newpage
+
+% \bibliographystyle{plain}
+% \bibliography{refs.bib}
+
+% \newpage
+
+% \appendix
+
+% \section{Inference Rules}
+% \label{sec:infer}
+
+% \begin{figure}[h!]
+%   \label{fig:structjdg}
+%   \begin{mathpar}
+%     \inferrule{~}{\ctxwff \cdot}
+%     \and
+%     \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}
+
+%     \\
+
+%     \inferrule{~}{\linctxwff \Gamma \cdot}
+%     \and
+%     \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}
+
+%     \\
+
+%     \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}
+
+%     \and
+
+%     \inferrule{\Gamma \vdash A : L_i}{\linctxwffjdg \Gamma A}
+
+%     \\
+
+%     \inferrule{\Gamma \vdash X \equiv Y : U_i}{\ctxwffjdg \Gamma {X\equiv Y}}
+
+%     \and
+
+%     \inferrule{\Gamma \vdash A \equiv B : L_i}{\linctxwffjdg \Gamma {A \equiv B}}
+
+%   \end{mathpar}
+%   \caption{Structural judgments}
+% \end{figure}
+
+% \begin{figure}
+%   \label{fig:typewf}
+%   \begin{mathpar}
+%     \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
+%  %
+%     \and
+% %
+%     \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
+% %
+%     \\
+% %
+%     \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
+% %
+%     \and
+% %
+%     \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
+% %
+%     \\
+% %
+%     \inferrule{~}{\Gamma \vdash 1 : U_i}
+% %
+%     \and
+% %
+%     \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
+% %
+%     \\
+% %
+%     \inferrule{~}{\Gamma \vdash I : L_i}
+%  %
+%     \and
+% %
+%     \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
+% %
+%     \and
+% %
+%     \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
+% %
+%     \and
+% %
+%     \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
+% %
+%     \\
+% %
+%     \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
+% %
+%     \and
+% %
+%     \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
+% %
+%     \\
+% %
+%     \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
+%     %
+%     \and
+%     %
+%     \inferrule{~}{\Gamma \vdash \top : L_i}
+% %
+%     \and
+% %
+%     \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
+% %
+%     \\
+%     %
+%     \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
+%     %
+%     \and
+%     %
+%     \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
+%   \end{mathpar}
+%   \caption{Type well-formedness}
+% \end{figure}
+
+% \begin{figure}
+%   \label{fig:jdgeq}
+%   \begin{mathpar}
+%     \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
+% %
+%     \and
+% %
+%     \inferrule{~}{}
+% %
+%     \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
+%     %
+%     \and
+%     %
+%     \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
+% %
+%     \and
+%     %
+%     \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
+%     %
+%     \and
+%     %
+%     \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
+% \end{mathpar}
+%   \caption{Judgmental equality}
+% \end{figure}
+
+% \begin{figure}
+%   \label{fig:inttyping}
+%   \begin{mathpar}
+%   \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
+%   %
+%   \and
+%   %
+%   \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
+%   %
+%   \\\
+%   %
+%   \inferrule{~}{\Gamma \vdash () : 1}
+%   %
+%   \and
+%   %
+%   \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
+%   %
+%   \\
+% %
+%   \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
+%   %
+%   \and
+%   %
+%   \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
+%   \and
+%   \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
+%   %
+%   \and
+%   %
+%   \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
+%   %
+%   \\
+%   %
+%   \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
+%   \and
+%   \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
+%   \end{mathpar}
+%   \caption{Intuitionistic typing}
+% \end{figure}
+
+% \begin{figure}
+%   \label{fig:linsyntax}
+%   \begin{mathpar}
+%     \inferrule{~}{\Gamma ; a : A \vdash a : A}
+%     \and
+%     \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
+%     %
+%     \\
+%     %
+%     \inferrule{~}{\Gamma ; \cdot \vdash () : I}
+%     \and
+%     \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
+%     %
+%     \\
+%     %
+%     \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
+%     %
+%     \\
+%     %
+%     \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
+%     \\
+%     %
+%     \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
+%     \and
+%     \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta', \Delta \vdash \applto {e'} {e} : B}
+%     \\
+%     %
+%     \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
+%     \and
+%     \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta, \Delta' \vdash \apptol e {e'} : B}
+%     %
+%     \\
+%     %
+%     \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}{\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
+%     \and
+%     \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma ; \Delta \vdash \app e {e'} : \subst A {e'} x}
+%     %
+%     \\
+%     %
+%     \inferrule{\Gamma \vdash e : X \quad \Gamma ; \Delta \vdash e' : \subst A e x}{\Gamma ; \Delta \vdash (e, e') : \LinSigTy x X A}
+%     %
+%     \\
+%     %
+%     \inferrule{\Gamma ; \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
+%     %
+%     \\
+%     %
+%     \inferrule{~}{\Gamma ; \Delta \vdash () : \top}
+%     %
+%     \\
+%     %
+%     \inferrule{\Gamma ; \Delta \vdash e_1 : A_1 \quad \Gamma ; \Delta \vdash e_2 : A_2}{\Gamma ; \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
+%     \and
+%     \inferrule{\Gamma ; \Delta \vdash e : A_1 \amp A_2 }{\Gamma ; \Delta \vdash \pi_i \, e : A_i}
+%     %
+%     \\
+%     %
+%     \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma ; \cdot \vdash \mathsf{G}^{-1}\, e : A}
+%     %
+%     \\
+%     %
+%     \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
+%     \and
+%     \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
+%   \end{mathpar}
+%   \caption{Linear typing}
+% \end{figure}
 \end{document}

### 2024-06-29 POPL/paper.tex
Abstract + introduction condensing

--- 2024-03-18 POPL/paper.tex
@@
 \usepackage{stmaryrd}
+\usepackage{tikz}
+\usetikzlibrary{automata, positioning, arrows, fit}
 
-\usepackage{todonotes}
-\newcommand{\todoin}[1]{\todo[inline]{TODO:\@ #1}}
+\usepackage{pdfpages}
 
@@
 \drafttrue
-\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven}: #1]}\fi}
-\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max}: #1]}\fi}
-\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro}: #1]}\fi}
+\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven says}: #1]}\fi}
+\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max says}: #1]}\fi}
+\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro says}: #1]}\fi}
 \newcommand{\pipe}{\,|\,}
-
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-%% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
-  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
-\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-  June 03--05, 2018, Woodstock, NY}
-\acmPrice{15.00}
-\acmISBN{978-1-4503-XXXX-X/18/06}
 
@@
 
-\title{Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}
+\pagestyle{plain}
+
+\pagebreak
+
+\title{Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}
+
 \author{Steven Schaefer}
-\affiliation{
-  \department{Electrical Engineering and Computer Science}
+\affiliation{\department{Electrical Engineering and Computer Science}
   \institution{University of Michigan}
@@
 
+\setcounter{page}{1}
+
+\section{Introduction}
+
 \section{A Syntax-Independent Notion of Syntax}
-\todoin{Rework intro from LICS draft with a focus on implementation}
+The notion of formal language is central to the theory of parsing. A
+\emph{formal language} $L$ over an alphabet $\Sigma$ is classically defined as a
+\emph{subset} of strings $L \subseteq \String$. This definition is
+especially useful as it gives a semantics to formal grammars that is completely
+independent of any particular syntactic grammar formalism. Any new
+notion of grammar can be given a language semantics and it provides a precise
+mathematical specification for implementing a \emph{recognizer} of a language.
 
-\section{Kleene Categories}
-\todoin{Tell a story winding through the use of KAs, action algebras, etc to
-  motivate why we'd want categorical semantics for KCats}
-\todoin{Incoporate Kleene Category definition}
-\todoin{End with the punch that $\Set^{\Sigma^{*}}$ is a KCat}
+That is, a recognizer should be extensionally equivalent to the indicator function
+$\chi_{L} : \String \to \Prop$ mapping each string $w$ to the the proposition
+that it is in the language $w \in L$. However, language recognition is
+insufficient for specifying a parser --- a function that decomposes a string
+into semantic components that adhere to the structure of a formal grammar. In
+practice, parsers do more than simply return whether a string is valid or
+invalid; instead, they emit structured \emph{parse trees}\footnote{Such parse
+  trees are usually not materialized in memory, as semantic actions often perform
+  computation over the trees while they are being constructed.}
 
-\section{Automata}
+Our first contribution is a novel and syntax-independent characterization of
+formal grammar:
 
+\begin{definition}
+  \label{def:grammar}
+  A \emph{formal grammar} $G$ over a fixed alphabet $\Sigma$ is a function
+  $G : \String \to \Set$.
+\end{definition}
+
+We say that a grammar $G$ associates to every string $w$ the set\footnote{We
+  make little comment as to which foundations are used to encode $\Set$. Our
+  construction is polymorphic in choicce of a proper class of small sets, a
+  Grothendieck universe, a type theoretic universe, or any similar foundation.} $G w$ of parse
+trees showing that $w$ matches $G$. This definition of formal grammar serves as
+a mathematical specification for a parser --- taking on the same role as
+languages do for recognizers. In this sense, our new characterization of
+grammars is just a proof-relevant generalization of the existing notion of
+formal language.
+
+By ranging over all input strings, notice that each grammar induces a language.
+For each formal grammar $G$, define the language of accepted strings $L_{G}$ as,
+
+\[ L_{G} = \{ w \in \String ~|~ G w~\text{is inhabited} \} \]
+
+As they a simply sets, formal languages are naturally endowed with a partial
+order via subset inclusion. Similarly, formal grammars are naturally endowed
+with the structure of a \emph{category}. A morphism $f : G \to H$ between two
+grammars is given by a family of functions
+
+\[ f^{w} : G w \to H w \]
+
+for every string $w \in \String$. Intuitively, we read a morphism of grammars as
+a \emph{parse transformer} : $f$ translates a $G$-parse of $w$ into an $H$-parse
+of $w$. The notion of isomorphism induced by
+this categorical structure encodes that two grammars are equivalent if there is
+a bijective translation of parses --- precisely Chomsky's notion of \emph{strong equivalence} of
+grammars \cite{chom1963}.
+
+The above category of grammars can equivalently be described as the functor
+category $\Set^{\String}$ --- where the set $\String$ is viewed as a discrete
+category. That is, the objects in $\String$ are strings and the only morphisms
+are identity morphisms. Such functor categories into $\Set$ --- often called
+presheaf categories --- carry a remarkable amount of structure.
+
+The category-theoretic framework can be further used to describe and compare
+different notions of formal grammar. For instance, Chomskyan generative
+grammars, Thue systems, Montague grammars, and so on are all distinct syntactic
+presentations of the same underlying idea of an abstract specification for
+parsing. Each of these is an instance of a general \emph{notion of formal
+  grammar} --- a category paired with a functor into
+$\Set^{\String}$. That is, they may all be interpreted in our category of grammars. In this sense, $\Set^{\String}$ provides a universal semantic
+domain to reason about parsing. The slice 2-category over $\Set^{\String}$ is
+then the appropriate setting to reason over notions of formal grammar.
+
+\steven{Check details for 2-cat of grammar formalisms}
+
+Our contributions are then:
+
+\steven{These sound awkward but I'm painting with broad strokes rn}
+\begin{itemize}
+  \item A syntax-independent definition of formal grammar
+  \item A dependent
+    linear/non-linear (LNL) type theory to syntactically capture the semantics of parsing
+  \item A categorification of Kleene Algebra
+  \item Interpretations of the LNL type system for reasoning about formal grammars and formal languages
+  \item The presentation of automata as grammars. By internalizing automata ---
+        the operational view of a parser --- we may extract executable parsers
+  \item A language implemented shallowly in Agda
+\end{itemize}
+
+\section{Kleene Category}
+\steven{No changes pasted from LICS draft}
+\steven{Are these now ``Chomsky categories''? or is that disinct?}
+
+Kleene algebras are an important tool in the theory of regular
+languages. More broadly, they serve as a theoretical substrate to
+studying various kinds of formal languages. Formally, they are a tuple
+$(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is a set, $+$ and $\cdot$
+are binary operations over $A$, $(-)^*$ is a function over $A$, and
+$1$ and $0$ are constants. These structures satisfy the axoims depicted
+in Figure~\ref{fig:axioms}.
+
+\begin{figure}
+  \begin{align*}
+    x + (y + z) &= (x + y) + z & x + y &= y + x\\
+    x + 0 &= x & x + x &= x\\
+    x(yz) &= (xy)z & x1 &= 1x = x\\
+    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
+    x0 &= 0x = x & & \\
+    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
+     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
+  \end{align*}
+  \caption{Kleene algebra axioms}
+  \label{fig:axioms}
+\end{figure}
+
+The addition operation can be used to define the partial order
+structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
+formal languages, this order structure can be used to model language
+containment. In this section, we want to categorify the concept of
+Kleene algebra and build on top of it in order to define an abstract
+theory of parsing.  We start by defining \emph{Kleene categories}.
+
+\begin{definition}
+  A Kleene category is a distributive monoidal category $\cat{K}$
+  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
+  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
+  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
+  X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogue isomorphism
+  for $G_{A,B}$ also holds.
+\end{definition}
+
+As a sanity check, note that Kleene algebras are indeed examples of
+Kleene categories.
+
+\begin{example}
+  Every Kleene algebra, seen a posetal category, is a Kleene category.
+\end{example}
+
+This example illustrates how restricted Kleene algebras are in terms
+of reasoning about languages. By only having at most one morphism
+between objects, there is not a lot of information they can convey. In
+this case, the only information you get is language containment. We
+will see throughout this paper how the extra degrees of freedom
+granted by having more morphisms give you more algebraic structure for
+reasoning about languages.
+
+For the next example, we see an unexpected connection with the theory
+of substructural logics.
+
+\begin{example}
+  The opposite category of every Kleene category is a model of a variant of
+  conjunctive ordered logic, where the Kleene star plays the role of the ``of
+  course'' modality from substructural logics which allows hypotheses to
+  be discarded or duplicated.
+\end{example}
+
+The proposed axioms are a direct translation of the Kleene algebra
+axioms to a categorical setting. Its most unusual aspect is the
+axiomatization of the Kleene star as a family of initial algebras
+satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
+more structure, then these isomorphisms hold ``for free''.
+
+\begin{theorem}
+  \label{th:kleeneclosed}
+  Let $\cat{K}$ be a Kleene category such that it is also monoidal
+  closed.  Then, the initial algebras isomorphisms hold automatically.
+\end{theorem}
+\begin{proof}
+  We prove this by the unicity (up-to isomorphism) of initial
+  algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
+  (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
+  X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
+  (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
+    1}(X))$.
+
+  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
+  we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,1} \to B \lto Y$. We can show existence and
+  uniqueness by showing that the diagram on top commutes if, and
+  only if, the diagram on the bottom commutes:
+
+% https://q.uiver.app/#q=WzAsOCxbMCwwLCJCICsgQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFsyLDAsIkIgKyBZIFxcb3RpbWVzIEEiXSxbMiwyLCJZIl0sWzAsMiwiQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDMsIjEgKyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDUsIlxcbXUgWC5cXCwgMSArIFggXFxvdGltZXMgQSJdLFsyLDMsIjEgKyAoIEIgXFxsdG8gWSkgXFxvdGltZXMgQSJdLFsyLDUsIkIgXFxsdG8gWSJdLFswLDEsImlkICsgKGlkIFxcb3RpbWVzIGg7IGV2KSBcXG90aW1lcyBpZF9BIl0sWzEsMiwiW2YsZ10iXSxbMywyLCJpZCBcXG90aW1lcyBoOyBldiIsMl0sWzAsMywiW2lkIFxcb3RpbWVzIGgsIGlkIFxcb3RpbWVzIHRsXSIsMl0sWzQsNiwiaWQgKyAoaCBcXG90aW1lcyBYKVxcb3RpbWVzIGlkIl0sWzUsNywiaCIsMl0sWzYsNywiW2YnLCBnJ10iXSxbNCw1LCJbaGQsIHRsXSIsMl1d
+\[\begin{tikzcd}
+	{B + B \otimes (\mu X.\, 1 + X \otimes A)} && {B + Y \otimes A} \\
+	\\
+	{B \otimes (\mu X.\, 1 + X \otimes A)} && Y \\
+	{1 + (\mu X.\, 1 + X \otimes A)} && {1 + ( B \lto Y) \otimes A} \\
+	\\
+	{\mu X.\, 1 + X \otimes A} && {B \lto Y}
+	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
+	\arrow["{[f,g]}", from=1-3, to=3-3]
+	\arrow["{id \otimes h; ev}"', from=3-1, to=3-3]
+	\arrow["{[id \otimes h, id \otimes tl]}"', from=1-1, to=3-1]
+	\arrow["{id + (h \otimes X)\otimes id}", from=4-1, to=4-3]
+	\arrow["h"', from=6-1, to=6-3]
+	\arrow["{[f', g']}", from=4-3, to=6-3]
+	\arrow["{[hd, tl]}"', from=4-1, to=6-1]
+\end{tikzcd}\]
+
+  This equivalence follows by using the adjunction structure given
+  by the monoidal closed structure of $\cat{K}$. A completely analogous
+  argument for $G_{A,B}$ also holds.
+\end{proof}
+
+This result feels similar in spirit to the definition of action
+algebras, which are algebras where the product also has adjoint
+operations which results in the Kleene star being more easily
+axiomatized \cite{kozen1994}.
+
+We are now ready to prove that our concept of formal grammars fits
+nicely within our categorical framework. We start by presenting a
+well-known construction from presheaf categories.
+
+\begin{definition}
+  Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
+  two functors $\cat{C} \to \Set$. Their Day convolution tensor
+  product is defined as the following coend formula:
+  \[
+  (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z)
+  \]
+  Dually, its internal hom is given by the following end formula:
+  \[
+  (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
+  \]
+\end{definition}
+
+\begin{lemma}[\cite{day1970}]
+  Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
+  monoidal closed.
+\end{lemma}
+
+%% \begin{theorem}
+%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
+%%   The functor category $[A, \cat{K}]$.
+%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
+%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
+%% \end{theorem}
+\begin{theorem}
+  If $\cat{C}$ is a locally small monoidal category, then
+  $\Set^{\cat{C}}$ is a Kleene category.
+\end{theorem}
+\begin{proof}
+
+  By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
+  is a presheaf category, it has coproducts. Furthermore, the tensor
+  is a left adjoint, i.e. it preserves colimits and, therefore, it is
+  a distributive category.
+
+  As for the Kleene star, since presheaf categories admit small colimits,
+  the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
+  defined as the filtered colimit of the diagrams:
+
+  From Theorem~\ref{th:kleeneclosed} it follows that these initial
+  algebras satisfy the required isomorphisms and this concludes the
+  proof.
+\end{proof}
+
+\begin{corollary}
+  For every alphabet $\Sigma$, the presheaf category $\Set^{\cat{\Sigma^*}}$
+  is a Kleene category.
+\end{corollary}
+\begin{proof}
+  Note that string concatenation and the empty string make the
+  discrete category $\Sigma^*$ a strict monoidal category.
+\end{proof}
+
+Much like in the posetal case, the abstract structure of a Kleene
+category is expressive enough to synthetically reason about formal
+languages. A significant difference between them is that while Kleene
+algebras can reason about language containment, Kleene categories can
+reason about parsing as well.
+
+For the rest of the paper we will use the presheaf category
+$\Set^{\cat{\Sigma^*}}$ as a concrete model that will serve as our
+guide when expanding abstract formalism presented in this section so
+that it can handle more classes of languages beyond regular ones.
+
+\paragraph{Beyond Simple Types}
+
+Though Kleene categories are expressive enough to reason about
+concepts that are outside of reach of Kleene algebras, their
+simply-typed nature makes them not so expressive from a type theory
+point of view. This is limiting because type theories are tested
+syntactic frameworks for manipulating complicated categories
+structures while avoiding some issues common in category theory, such
+as coherence issues.
+
+With this in mind, we want to design a dependent type theory that
+builds on top of Kleene categories with the goal of internalizing
+various reasoning principles for languages and their parsers. This
+leads us to the abstract notion of model we are interested in
+capturing with our type theory.
+
+\begin{definition}
+  A model is a locally Cartesian category with two hierarchies of
+  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
+  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
+  $U_i$ to be closed under dependent products and sum,
+  $L_i$ to be closed under the Kleene category connectives,
+  dependent products, left and right closed structures, with
+  a type constructor $G : L_i \to U_i$ and a linear dependent sum
+  going the other direction.
+\end{definition}
+
+\begin{theorem}
+  The presheaf category $\Set^{\cat{\Sigma^*}}$ is a model.
+\end{theorem}
+
+\section{Related work}
+
+\paragraph{Kleene Algebra}
+
+Since the early works in the theory of formal languages, Kleene
+algebras have played an important role in its development. They
+generalize the operations known from regular languages by introducing
+operations generalizing language composition, language union and the
+Kleene star.  More generally, they are defined as inequational theory
+where the inequality is meant to capture language containment. This
+theory is extremely successful, having found applications in algebraic
+path problems, theory of programming languages, compiler optimizations
+and more.
+
+A frequently fruitful research direction is exploring varying
+extensions of Kleene algebras, Kleene algebra with tests (KAT) being
+one of the most notable ones. Our approach is radically different from
+most extensions, which usually aim at modifying or adding new
+operations to Kleene algebras, but still keeping it as an inequational
+theory. By adopting a category-theoretic treatment and allowing the
+``order structure'' to encode more information than merely
+inequalities, we were able to extend Kleene algebra to reason about
+parsing as well.
+
+\paragraph{TODO: more stuff}
+
+%% That paper with Fritz Henglein about regexes as types
+
+%% Vaughan Pratt on the residual operator
+
+\section{Rest of paper}
+\steven{Everything below still needs to be edited from my prelim paper}
+\steven{Need to grab the canonicity proof from the LICS draft}
+
+Substructural logics, like linear and separation logic, offer precise control
+over how values from a type may be used, in particular how many times a value
+may be used in a derivation. This control enables \emph{resource-sensitive}
+reasoning, where we may treat a resource as \emph{consumed} after usage. In the
+context of parsing, we may treat the characters composing a string as finite resources
+that are consumed at parse-time.
+
+Dependent type theory extends the simply-typed $\lambda$-calculus by allowing type
+definitions to contain the occurrence of a term. In lieu of set-based
+foundations, this extension can then be taken as a foundation for mathematics.
+The expressive power unlocked by this insight has enabled the development of
+dependently-typed proof assistants
+such as Coq or Agda. Within each of these languages, systems may be given very
+precise functional specifications of their behavior.
+
+In 1995, Benton et al.
+\cite{bentonMixedLinearNonlinear1995} first gave categorical models of a logic
+that synthesized linear and non-linear logic. Later, Krishnaswami et al.
+\cite{krishnaswami_integrating_2015} built upon this work to develop a type
+system that unifies both linear and
+dependent types --- referred to as a \emph{linear-non-linear} (LNL) type theory. In
+such settings, we have two sub-theories --- one linear and one non-linear. The
+linear sub-theory embeds ordinary linear types without dependence, while the
+non-linear sub-theory embeds standard Martin-L\"of type theory. The interplay of
+linear types with dependence lives at the boundary between these two fragments.
+Benton et al.\ gives an accounting of how we may jump between these
+fragments via a monoidal adjunction $F \dashv G$ between linear terms and
+non-linear terms.
+
+To the end of defining real verified
+parsers, in our LNL theory we define a parser as a term,
+
+\crefname{equation}{}{}
+\begin{equation}
+  \label{eq:parser}
+  w : \String \vdash \mathsf{parse}_{g}(w) : g \oplus \top
+\end{equation}
+
+We will investigate the precise meaning of judgments like \cref{eq:parser} in
+\cref{sec:tt}, but first let's get an intuitive feel for the meaning behind such a
+term. On the left of the turnstile occurs a single string $w$, we may think of
+this describing what resources are linearly available to us. Using each piece of
+our the context on the left of the turnstile, we build a term
+$\mathsf{parse}_{g}(w)$ of type $g \oplus \top$. Notably, the type being
+instantiated is a sum type, and the data of which side we inject into denotes
+whether the string is accepted or rejected by the parser. That is all to say, a
+parser takes in an input string and either matches it to the desired grammar or
+rejects the string.
+
+Practical parsers are often implemented via an abstract automaton. For instance,
+a parser may translate a grammar into a state machine recognizes the
+same language as the original grammar, as the automata
+point-of-view lends itself to straightforward execution. The
+strength of our syntax is that it allows us to describe
+these automata as grammars themselves. Moreover, the
+equivalence between the grammar and automata presentations
+is derivable as a proof term within the theory. This allows
+us access to simultaneously internalization of the equivalence
+between grammars and automata as well as the ability to
+export to automaton implementations of a parser that are executable.
+
+To realize an executable implementation of our type theory,
+we shallowly embed it in Agda. Because our implementation is
+defined as a function in Agda, we can leverage the Agda
+ecosystem for code generation. Further, in lieu of
+implementing a complete language from scratch, we make use of the
+Agda proof system as a setting to guarantee correctness of
+our generate parsers.
+
+Within this paper, we give our progress in defining a verified
+parser for regular expressions. The construction of such a parser entails
+the definition of a parser for deterministic finite automata (DFA) grammars,
+and an equivalence proof between DFAs and regular expressions. We have thus far
+completed the construction of a parser for DFAs $D$,
+
+\[
+  w : \String \vdash \mathsf{parse}_{D}(w) : D \oplus \top
+\]
+
+and half of the equivalence proof between DFAs and regular expressions, both
+formalized in Agda. Upon completion of the second half of this proof, we would
+have a system where a user could simply give a regular grammar and then extract
+an intrinsically verified parser for it. In future work, we
+hope to show an analogous construction involving pushdown
+automata (PDAs) and context-free grammars.
+
+\section{A Syntax-Independent Notion of Syntax}
+\label{sec:synindsyn}
+
+% \max{this section doesn't flow naturally from the intro, which is all about the
+%   type theory and parsing. This is where you want to outline in the intro and
+%   possibly say here that we want to revisit ``formal grammar theory'' and then
+%   we will make a type theory that has formal grammar theory as its semantics.}
+
+In this section we will revisit some foundational ideas in formal
+grammar theory, before giving a type
+theory for which formal grammars serves as its semantics.
+
+Formal languages and grammars are among some of the oldest and most studied areas in
+theoretical computer science. The developments of the 1950s and 1960s gave rise
+to expressive formalism for reasoning about grammars, and further practical
+algorithms for constructing parsers for regular and context-free grammars.
+
+A grammar $G$ is often described generatively as a set of production rules for
+generating strings. Often this is presented as a finite set of nonterminal
+symbols $N$, a finite set of terminal symbols $\Sigma$, and a finite set of rewriting
+rules for substitution of nonterminals. In this setting, we continually apply
+rewrite rules until we are left only with a string of terminals
+$w \in \String$. We will give a more precise notion of
+grammar in a moment, but for now let us take this intuitive view of grammars as
+string generators.
+
+One of the primary objects of study is that of a \emph{formal language} over
+a fixed alphabet $\Sigma$. A language $L$ is a \emph{subset} of strings
+$L \subseteq \String$. This notion of formal language gives an intuitive
+set-based semantics for grammars. That is, we may interpret a grammar $G$ as the
+language $L(G)$ that is generated by it. We are yet to give any precise
+definition of what we mean by ``grammar'', although we note that any suitable
+notion of grammar can be given a language semantics as above.
+
+Formal languages serve as a basis for mathematically specifying a
+\emph{recognizer} --- a
+function $r : \String \to \Bool$ that determines if a string belongs to a
+given language. However, they are insufficient for specifying a \emph{parser}
+--- a function that decomposes a string into semantic components that adhere to
+the structure of a formal grammar. In practice, parsers don't just tell you
+if a string is valid or invalid. Instead, they emit structured data such as
+abstract syntax trees (ASTs).
+
+This motivates our first conceptual contribution, a simple and
+syntax-independent notion of formal grammar:
+
+\begin{definition}
+  \label{def:grammar}
+  A \emph{formal grammar} $G$ over a fixed alphabet $\Sigma$ is a function
+  $G : \String \to \Set$.
+\end{definition}
+
+We say that a grammar $G$ associates to every string $w$ the set\footnote{We
+  make little comment as to which foundations we use to encode sets. We argue\max{drop ``We argue''}
+  that our construction is agnostic to the particular foundation, and so without
+  loss of generality\max{incorrect use of ``without loss of generality''} we may take a class of small sets, a Grothendieck universe,
+  a type theoretic universe, or any similar foundation. Here we choose a type
+  theoretic universe.} $G w$ of parse
+trees showing that $w$ matches $G$.
+
+The simplest class of formal grammar on the Chomsky\max{first mention of Chomsky hierarchy, why not bring it up earlier when you are describing what a grammar is? Also cite}
+hierarchy are the \emph{regular expressions}
+
+\max{inaccurate reflection of common terminology. These are regular expressions,
+  regular grammars usually means a CFG with a restriction on recursion that
+  ensures you only allow Kleene star.}.
+
+\begin{gather*}
+  \mathsf{Char} \in \Sigma \\
+  \mathsf{Reg} ::= \mathsf{Char}~|~I~|~\mathsf{Reg} \otimes \mathsf{Reg}~|~\mathsf{Reg} \oplus \mathsf{Reg}~|~\mathsf{Reg}^{*}
+\end{gather*}
+
+Here, we use a non-standard syntax inspired by linear logic. Disjunction of
+regular expressions is written $\oplus$, and concatenation of regular
+expressions is written like the linear tensor product $\otimes$. The
+empty grammar is sometimes written in the same manner as the empty string,
+$\varepsilon$, but we instead choose to represent the empty grammar by $I$. This
+is also linear logic syntax to represent the unit $I$ of the operation $\otimes$.
+
+To see an illustration of the notion of grammar given in \cref{def:grammar}, consider the regular expression
+$G = (a \otimes b)^{*} \otimes a$ --- a sequence of 0 or more $ab$'s followed by
+an $a$. We may think of $G$ as a function that takes in strings and outputs all
+of the parse trees of that string. Passing the input string $baa$ through $G$
+would output the empty set, because there are no $G$-parses of $baa$. However,
+passing $aba$ through $G$ would output a singleton set, as there is only a
+single $G$-parse of $aba$. Namely, the one that breaks up $aba$ as $(ab)a$.
+
+Observe that for any input string $w$ for the above grammar $G$, $G(w)$ is
+either empty or a singleton set. Either the string isn't parsed by $G$,
+or it is parsed by $G$ for exactly one reason. That is, $G$ is
+\emph{unambiguous}. If we instead consider the grammar
+$H = (a \otimes b)^{*} \oplus (b \otimes a)^{*}$ there are now ambiguities. For
+instance, what is the action of $H$ on the empty string $\varepsilon$?
+The empty string $\varepsilon$ matches both $(a \otimes b)^{*}$ and $(b \otimes a)^{*}$, so
+$\varepsilon$ can match either the left or the right pattern
+of $H$. As a result, $H(\varepsilon)$ is a two-element set.
+
+\[
+  L_{G} = \{ w \in \String : G w \text{ is inhabited} \}
+\]
+
+As they are just sets, formal languages naturally are
+endowed with a partial
+order via the subset inclusion ordering. Similarly, formal grammars are
+naturally endowed with the structure of a \emph{category} where a morphism
+$f : G \to H$ is given by a \emph{family} of functions:
+
+\[
+  f^{w} : Gw \to Hw, w \in \String
+\]
+
+Intuitively, we read a morphism of grammars as a \emph{parse transformer}: $f$
+translates a
+$G$-parse of $w$ into an $H$-parse of $w$. The notion of isomorphism induced by
+this categorical structure encodes that two grammars are equivalent if there is
+a bijective translation of parses. Isomorphisms in the category of
+grammars are precisely Chomsky's notion of \emph{strong equivalence} of
+grammars \cite{chom1963}.
+
+The category of grammars is quite rich in structure. It can
+equivalently be described as the functor category $\Set^{\String}$, where
+$\String$ is a discrete category. That is, the objects in $\String$ are strings
+and the only morphisms are identity morphisms. Such functor categories into
+$\Set$ --- often called presheaf categories --- carry a remarkable amount of
+structure. In fact, any presheaf category naturally models dependent type
+theory, which we exploit to motivate the design of the formalism
+presented in \cref{sec:tt}.
+
+\section{A Type Theory to Reason About Formal Grammars}
+\label{sec:tt}
+
+\subsection{Syntax}
+\label{subsec:syntax}
+
+In this section we will walk through most of the syntax of our type theory. We
+include the full syntax in \cref{sec:infer}.
+
+The linear typing judgment in our syntax takes on the following schematic form,
+\[
+  \Gamma ; \Delta \vdash M : A
+\]
+
+Let us briefly digest what this judgment is trying to say. First, $A$ represents
+a \emph{linear type} in our syntax. The intended semantics of these linear types
+are formal grammars. That is, the linear typing system is designed to syntactically
+reflect the behavior of formal grammars. For this reason, we may often
+interchangeably use the terms ``linear type'' and ``grammar''. To see the
+correspondence between types and grammars, note that the constructors for linear types
+mirror operations on formal grammars. For example, we have linear type constructors
+$\otimes$, $\oplus$, and $\amp$ which correspond to grammar concatenation,
+disjunction, and conjunction, respectively. The other type constructors also have
+corresponding grammar intuition which may be seen in \cref{subsubsec:other}.
+
+$M$ is an inhabitant of type $A$, which we want to think of as a parse tree of
+the grammar $A$. The core idea of this entire paper follows precisely from this
+single correspondence: grammars are types, and the inhabitants of these types
+are parse trees for the grammars!
+
+$\Gamma$ represents a \emph{non-linear context} --- also referred to as an
+\emph{intuitionistic} context. The non-linear fragment of the theory
+corresponds to usual Martin-L\"of type theory. That is, the intuitionistic
+contexts obey all of the usual structural rules --- contraction, weakening, and
+exchange --- so we may use their contents as many times as we'd like.
+
+On the other hand $\Delta$ represents a \emph{linear context}. These linear
+contexts
+behave substructurally --- in particular they have no contraction, weakening, or exchange
+rule. This means that everything within $\Delta$ must be used \emph{exactly
+  once} and \emph{in order of occurrence}. Thus, we can think of the linear
+contexts as an ordered list of limited
+resources. Once a resource is consumed, we cannot make reference to it again.
+We may then
+think of the variables in a linear context akin to building blocks for
+constructing patterns over strings.
+
+In the realm of parsing, our substructural
+restrictions make it so that the linear contexts behave quite similarly to
+formal grammars. For instance, linear context extension mirrors concatenation of
+grammars which may be seen via the introduction rule for $\otimes$,
+
+\[
+  \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta , \Delta' \vdash e \otimes e' : A \otimes B}
+\]
+
+Just as two grammars may be concatenated, if we can build an $A$ with the
+contents of $\Delta$, and a $B$ with the contents of $\Delta'$, then we we may
+concatenate $\Delta$ and $\Delta'$ to build an inhabitant $A \otimes B$.
+
+This is perhaps better illustrated through a concrete example. Consider the
+regular expression $A = (a \oplus b) \otimes (b \oplus c)^{*}$ and the string
+$w = abc$. We then have the following derivation that $w$ matches $A$,
+
+\begin{equation}
+  \label{eq:samplederiv}
+  \inferrule
+  {
+    w_{1} : a \vdash \mathsf{inl}(w_{1}) : a \oplus b \\
+    \inferrule
+    {
+      w_{2} : b \vdash \mathsf{inl}(w_{2}) : b \oplus c \\
+      w_{3} : c \vdash \mathsf{inr}(w_{3}) : b \oplus c
+    }
+    {
+      w_{2} : b, w_{3} :c \vdash \mathsf{cons}(w_{2},w_{3}) : (b \oplus c)^{*}
+    }
+  }
+  {w_{1} : a, w_{2} : b, w_{3} : c \vdash \mathsf{inl}(w_{1}) \otimes \mathsf{cons}(w_{2}, w_{3}) : (a \oplus b) \otimes (b \oplus c)^{*} }
+\end{equation}
+
+When parsing $w$ into $A$ by hand, we perform the same process as reflected in
+the derivation \cref{eq:samplederiv}. That is, we split the string $w$ up into pieces, one of which matches
+$a \oplus b$, and and another that matches $(b \oplus c)^{*}$. When further
+parsing this Kleene star bit, we break it up into smaller pieces each of which
+matches the patterns $b \oplus c$ and then we combine them all into one parse of
+the Kleene star $(b \oplus c)^{*}$. Note that this exact process is echoed in
+the above proof tree.
+
+\subsubsection{Dependence of the Linear on the Non-Linear}
+
+We may now however ask, ``what are the $\Gamma$'s doing in the $\otimes$ introduction rule?''
+Each linear context $\Delta$ is dependent on a non-linear context $\Gamma$, so
+we may use variables drawn from $\Gamma$ when defining the types in $\Delta$.
+
+In the grammar setting, this dependence lets
+us use variables from regular dependent type theory to
+keep track of data without \emph{consuming any resources}, so to speak. That is,
+using a piece of a linear context consumes it, so we cannot linearly encode anything that
+needs to be referenced more than once.
+
+We will see a concrete instance of this dependence in \cref{subsubsec:pda}. In
+that case, we encode the traces of a pushdown automaton as a grammar. When doing so,
+we make use of our non-linear types to define a type of stacks. The fact that
+our encoding of stacks is non-linear reflects
+that a pushdown automaton may consume a character from the input string without
+affecting the state of the stack.
+
+\subsubsection{Other Type Constructors}
+\label{subsubsec:other}
+
+\paragraph{Brzozowksi Derivative}
+In the 1950s Janusz Brzozowksi developed a notion of derivative for formal
+languages \cite{brzozowskiDerivativesRegularExpressions1964}. The derivative of language $L$ by string $w$ is given as,
+
+\[
+  w^{-1}L = \{ w' \in \String : w \otimes w' \in L \}
+\]
+
+That is, the derivative of $L$ with respect to $w$ are precisely the strings
+that complete to an $L$-parse when prefixed by a $w$.
+
+Brzozowski initially gave an accounting of this operation for
+generalized regular expressions, but later work demonstrates that the same
+construction can be generalized to context free grammars \cite{mightParsingDerivativesFunctional2011}.
+
+The ``magic wand'' from linear logic takes on the role of Brzozowski's
+derivative in our syntax.
+
+\begin{equation}
+  \label{eq:linfun}
+\inferrule{\Gamma ; a : A, \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A \lto B}
+\end{equation}
+
+We read this as saying that $\Delta$ constructs a $A \lto B$-parse when
+prepending $\Delta$ with an $A$-parse completes to a $B$ parse. In this sense,
+the type $A \lto B$ behaves like a linear function type. The only distinction
+between $A \lto B$ and an intuitionistic function type is that linear functions
+consume their input.
+
+Note that the introduction rule presented in \cref{eq:linfun} does indeed
+generalize Brzozowski's initial presentation of the derivative. First, to match
+his initial presentation
+let $A$ be the grammar accepting a single string $w$, which we may write as the
+discrete grammar $\lceil w \rceil$.
+Additionally, our construction lifts from a language-theoretic construction
+to a grammar point-of-view.
+In this sense, we carry around not just the data of what strings belong to
+$\lceil w \rceil \lto B$ but also are endowed the data of a function that
+constructs parse trees of $B$ from pairs of $\lceil w \rceil$-parses and
+$(\lceil w \rceil \lto B)$-parses.
+
+\paragraph{Top and Bottom}
+We include linear versions of the unit and empty types. The
+grammar $\top$ matches every input string, while the grammar
+$\bot$ represents falsity and matches no strings. These two
+are often used together to give syntactic shorthand to
+grammars that encode propositions.
+
+For instance, we may encode the proposition
+$\mathsf{isVowel}$ as,
+
+\[
+  \mathsf{isVowel}(w) :=
+    \begin{cases}
+      \top, \quad w \in \{a, e, i, o, u\} \\
+      \bot, \quad \text{otherwise}
+    \end{cases}
+\]
+
+\paragraph{Fixed Point Operator}
+We include a least fixed point operator on types, $\mu$, which is the means by
+which we define recursive grammars.
+
+Observe that we need not take the Kleene star as a primitive
+grammar constructor, as it is definable as a fixed point.
+The Kleene star of a grammar $g$ is given as,
+
+\[
+  g^{*} := \mu X . I \oplus (g \otimes X)
+\]
+
+\begin{figure}[h!]
+\begin{mathpar}
+  \inferrule
+  {\Gamma ; \Delta \vdash p : I}
+  {\Gamma ; \Delta \vdash \mathsf{nil} : g^{*}}
+
+  \and
+
+  \inferrule
+  {\Gamma ; \Delta \vdash p : g \\ \Gamma ; \Delta' \vdash q
+  : g^{*}}
+  {\Gamma ; \Delta \vdash \mathsf{cons}(p , q) : g^{*}}
+
+  \and
+
+  \inferrule
+  {
+    \Gamma ; \Delta \vdash p : g^{*} \\
+    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
+    \Gamma ; x : g , y : h \vdash p_{\ast} : h
+  }
+  {\Gamma ; \Delta \vdash \mathsf{foldr}(p_{\varepsilon} , p_{\ast}) : g^{*}}
+\end{mathpar}
+\caption{Kleene Star Rules}
+\label{fig:star}
+\end{figure}
+
+Likewise, $g^{*}$ has admissible introduction and
+elimination rules, shown in \cref{fig:star}. Note that this
+definition of $g^{*}$ and these
+rules arbitrarily assigns a handedness to the Kleene star.
+We could have just as well took it to be a fixed point of
+$I \oplus (X \otimes g)$. In fact, the definitions are
+equivalent, as the existence of the $\mathsf{foldl}$ term below
+shows that $g^{*}$ is also a fixed point of
+$I \oplus (X \otimes g)$.
+
+\begin{equation}
+  \label{eq:foldl}
+  \inferrule
+  {
+    \Gamma ; \Delta \vdash p : g^{*} \\
+    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
+    \Gamma ; y : h , x : h \vdash p_{\ast} : h
+  }
+  {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
+\end{equation}
+
+In fact, the $\mathsf{foldl}$ term is defined using
+$\mathsf{foldr}$ --- much in the same way one
+may define a left fold over lists in terms of a right fold
+in a functional programming language\footnote{The
+  underlying trick is to fold over a list of functions
+  instead of the original string. We curry each character $c$
+  of the string into a function that concatenates $c$, and
+  right fold over this list of linear functions. Because function
+  application is left-associative, this results in a left
+  fold over the original string. }.
+
+We only take fixed points of a single variable as a
+primitive operation in the type theory, but we may apply
+Beki\`c's theorem \cite{Bekić1984} to define an admissible
+notion of multivariate fixed point. This is particularly
+useful for defining grammars that encode the states of an
+automaton. In \cref{fig:multifix} we provide the
+introduction and elimination principles for such a fixed
+point, where $\sigma$ is the substitution that unrolls the
+mutually recursive definitions one level. That is,
+
+\begin{align*}
+  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \\
+  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
+\end{align*}
+
+\begin{figure}[h!]
+\begin{mathpar}
+  \inferrule
+  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
+  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
+
+  \\
+
+  \inferrule
+  {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
+             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
+             \dots \\
+             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
+  }
+  {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
+\end{mathpar}
+\caption{Multi-fixed Points}
+\label{fig:multifix}
+\end{figure}
+
+The most prevalent application of these multi-fixed points
+is for defining the traces through automata, as we will see
+later in \cref{eq:nfatrace}.
+
+\todoin{with citation, talk somewhere about how CFGs are just regular + fixed points}
+\max{this is a good point to give an overview of how regular expressions and context-free grammars/expressions are essentially sub-languages of our type system.}
+
+\subsection{Semantics in Grammars}
+\label{subsec:seming}
+As mentioned earlier, the interpretation of grammars as functions from
+$\String$ to $\Set$ serves as our intended semantics. When designing the syntax of this formalism, we were often
+informed by this model. That is, there is an interpretation
+$\llbracket \cdot \rrbracket$ of our syntax that provides grammar-based
+semantics.
+
+\begin{enumerate}
+  \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
+  \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
+        $\sem X : \sem \Gamma \to \Set_{i}$
+  \item A non-linear term $\Gamma \vdash e : X$ denotes a section
+        $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$
+  \item Linear contexts $\Gamma \vdash \Delta$ and types
+        $\Gamma \vdash A : L_{i}$ both denote families of grammars
+        $\sem \Gamma \to \Gr_{i}$
+  \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
+        transformers
+        $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
+\end{enumerate}
+
+Most of the non-linear semantics is standard, so let us investigate in more
+depth the linear semantics. The interpretation of each
+connective is exactly what is suggested by the syntax, in
+large part to the inspiration taken from this interpretation
+when designing the type theory.
+
+\begin{enumerate}
+  % \item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
+  \item $\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+        \begin{itemize}
+          \item The semantics of the linear $\Pi$-type is
+                indeed a dependent function out of the
+                semantics of $X$.
+        \end{itemize}
+  \item $\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+        \begin{itemize}
+          \item The semantics of the linear $\Sigma$-type is
+                a dependent pair.
+        \end{itemize}
+  \item $\sem{I} \gamma w = \{ () \pipe w = \epsilon \}$
+        \begin{itemize}
+          \item The only string accepted by $I$ is the empty string.
+        \end{itemize}
+  \item
+        $\sem{A \otimes B} \gamma w = \Sigma(w1,w2:\Sigma^*) \left( (w_{1}w_{2} = w) \wedge \sem{A} \gamma w_1 \times \sem{B} \gamma w_2 \right)$
+        \begin{itemize}
+          \item The string $w$ matches $A \otimes B$
+                precisely when it may be split into two
+                pieces such that the left one matches $A$
+                and the right one matches $B$.
+        \end{itemize}
+  \item
+        $\sem{A \lto B} \gamma w = \Pi(w_a:\Sigma^*) \left( A \gamma w_a \Rightarrow B\gamma (w_aw) \right)$
+        \begin{itemize}
+          \item The semantics of the linear function type is
+                a dependent function that turns $A$-parses
+                of a substring into $B$-parses of an
+                extended string.
+        \end{itemize}
+  \item
+        $\sem{B \tol A} \gamma w = \Pi(w_a:\Sigma^*) \left( A \gamma w_a \Rightarrow B\gamma (ww_a) \right)$
+  \item $\sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x)$
+        \begin{itemize}
+          \item The semantics of the least fixed point
+                operator is indeed the least fixed point of sets.
+        \end{itemize}
+\end{enumerate}
+
+\subsection{Semantics in Languages}
+Recall from \cref{sec:synindsyn} that every formal grammar induces a formal language. This corresponds to another
+interpretation $L(\cdot)$ of our syntax. In particular, the semantics of a
+linear type $A : L_{i}$
+from \cref{subsec:seming} are augmented such that
+
+\[
+  L(A) = \{ w \in \String : \sem A w \text{ is nonempty} \}
+\]
+
+\subsubsection{Weak Equivalence}
+Recall that an isomorphism of grammars defined a strong equivalence. We may
+additionally define a \emph{weak equivalence} of grammars when two grammars
+define the same language.
+
+That is, $A$ and $B$ are weakly equivalent when $L(A) = L(B)$. Note that this is
+a semantic notion, but this just as well captured syntactically by the existence
+of two judgments $\Gamma ; x : A \vdash M : B$ and
+$\Gamma ; x : B \vdash N : B$, which are not necessarily mutually inverse.
+
+\section{Automata as Grammars}
+Classically, formal language theory is closely related to the study of automata.
+In the Chomsky hierarchy, each language class is associated to a class of
+automata that serve as recognizers. Internal to the type theory of \cref{sec:tt},
+we can characterize these language classes syntactically; moreover, we
+demonstrate the equivalence of these language classes to its associated automata class as a
+proof term within our logic.
+
+Internalizing these equivalences is particularly useful for building a regular
+expression parser. That is, it is relatively easy for use to write down a parser
+for deterministic finite automata (DFAs). Then we may compose this parser with
+a proof equivalence between DFAs and regular expressions to extract a verified
+regular expression parser.
+
+As we will see in \cref{subsubsec:tm}, we may very quickly encode arbitrary
+recursively enumerable languages --- since we can simulate the
+tape of Turing machine in the dependent fragment of the language. We reflect
+here on this expressive power to motivate why we put stringent syntactic
+restrictions on particular classes of automata. The type theory we present truly
+is general enough to easily capture all of the Chomsky hierarchy.
+To begin, let's focus on the simplest languages and their corresponding
+automata --- regular expressions and finite automata.
+
+\subsection{Non-deterministic Finite Automata}
+\label{subsec:finiteaut}
+\begin{figure}
+  \begin{tikzpicture}[node distance = 25mm ]
+    \node[state, initial, accepting] (1) {$1$};
+    \node[state, below left of=1] (2) {$2$};
+    \node[state, right of=2] (3) {$3$};
+
+    \path[->] (1) edge[above] node{$b$} (2)
+              (1) edge[below, bend right, left=0.3] node{$\epsilon$} (3)
+              (2) edge[loop left] node{$a$} (2)
+              (2) edge[below] node{$a, b$} (3)
+              (3) edge[above, bend right, right=0.3] node{$a$} (1);
+  \end{tikzpicture}
+  \caption{An example NFA}
+  \label{fig:NFA}
+\end{figure}
+
+Classically, a \emph{nondeterministic finite automaton} (NFA) is a finite state machine where
+transitions are labeled with characters from a fixed alphabet $\Sigma$. These
+are often represented formally as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$,
+
+\begin{itemize}
+  \item $Q$ a finite set of states
+  \item $\Sigma$ a fixed, finite alphabet
+  \item $\delta : Q \times (\Sigma \cup \{ \varepsilon\}) \to \mathcal{P}(Q)$ the labeled transition function
+  \item $q_{0} \in Q$ the start state
+  \item $F \subset Q$ a set of accepting states
+\end{itemize}
+
+Intuitively, this can be thought of like a directed graph with nodes in $Q$ with
+an edge $q \overset{c}{\to} q'$ whenever $q' \in \delta(q, c)$. Note that
+transitions in an NFA may be labeled with the empty string $\varepsilon$ --- such
+transitions are referred to as \emph{$\varepsilon$-transitions}. We may see an
+example of an NFA in \cref{fig:NFA}.
+
+From an NFA, we may construct a grammar of traces as follows:
+
+First, we define a mutual fixed point grammar that describes the traces through the NFA.\ Then, we have another
+grammar that tells us if we're currently in an accepting state of the automaton.
+A parse of the NFA grammar is then a pair of a trace and the data that we're in
+an accepting state.
+
+Consider the NFA $N$ pictured in \cref{fig:NFA}. There are three states, $1$, $2$,
+and $3$. We introduce a non-linear type $Q$ with three inhabitants $q_{1}$,
+$q_{2}$, and $q_{3}$ to represent each of these states, respectively. Given a
+$q, q'$, we can
+then define the type of traces from $q$ to $q'$. For instance, let's construct
+the traces starting at $q_{1}$ and ending at $q_{2}$ as an example.
+
+\begin{equation}
+  \label{eq:nfatrace}
+  \mathsf{Trace}_{N}(q_{1}, q_{2}) = \mu
+  \begin{pmatrix}
+    g_{q_{1}} := g_{q_{3}} \oplus ( b \otimes g_{q_{2}} ) \\
+    g_{q_{2}} := ( a \otimes g_{q_{2}} ) \oplus ( a \otimes g_{q_{3}} ) \oplus ( b \otimes g_{q_{3}} ) \oplus I \\
+    g_{q_{3}} := a \otimes g_{q_{1}} \\
+  \end{pmatrix}. g_{q_{1}}
+\end{equation}
+
+We should read this as defining three mutually recursive grammars, one for each
+state. The definitions of these mutually recursive grammars capture the
+transitions of the automaton. To ensure that we only encode traces that end in
+state 2, we only include the unit $I$ as a summand in
+$g_{q_{2}}$. That is, by only including $I$ at this location, it makes it the
+only place where derivations of $\mathsf{Trace}(q_{1}, q_{2})$ can terminate. We
+can think of all of these definitions underneath
+of the $\mu$ binder as bringing some local grammars into scope. With these local
+grammars in scope we are ultimately constructing a term of type $g_{q_{1}}$ to denote
+only traces starting in state 1.
+
+To encode the acceptance criteria of $N$, we want to internalize a proposition
+over each state of the NFA.\ That is, for each $q \in Q$ we define a term
+
+\[
+  \mathsf{acc}(q) := q \text{ is accepting}
+\]
+
+In this example, $\mathsf{acc}(q_{1}) = \top$ and
+$\mathsf{acc}(q_{2}) = \mathsf{acc}(q_{3}) = \bot$. An accepting trace of the NFA $N$ is then given by the following dependent grammar,
+\[
+ \mathsf{AccTrace}_{N} := \LinSigTy q Q {\left( \mathsf{Trace}_{N}(q_{0} , q) \amp \mathsf{acc}(q) \right)}
+\]
+
+where $q_{0}$ is the initial state. That is, a trace is accepted by the NFA if
+we can construct the trace, and the trace ends at an accepting state. This idea
+is simple enough and aligns with how we intuitively treat these automata.
+
+Generalizing over the above example, we want to define the type of traces as,
+
+\[
+  \mathsf{Trace}_{N}(q_{0}, q_{1}) = \mu
+  \begin{pmatrix}
+    g_{q} := \mathsf{Trans}(q), & q \neq q_{1} \\
+    g_{q} := \mathsf{Trans}(q) \oplus I , & q = q_{1}
+  \end{pmatrix}. g_{q_{0}}
+\]
+
+where $\mathsf{Trans}(q)$ is an iterated disjunction that describes which
+transition you should take. For an NFA, $\mathsf{Trans}(q)$ must take on the
+following syntactic form,
+
+\begin{gather*}
+ \mathsf{State} \in \{ g_{q} : q \in Q \} \\
+ \mathsf{Char} \in \Sigma \\
+ \mathsf{Trans}_{N}(q) ::= \mathsf{Char} \otimes \mathsf{State}~|~\mathsf{State}~|~\mathsf{Trans}_{N}(q) \oplus \mathsf{Trans}_{N}(q)
+\end{gather*}
+
+That is, $\mathsf{Trans}(q)$ is a disjunction of literals followed by grammars
+that encode states.
+
+When conducting proofs involving NFA grammars, we often to
+need either construct terms of type $\mathsf{Trace}_{N}(q , q')$. To
+this end, we give three admissible rules for constructing
+traces,
+
+\begin{figure}[h!]
+  \label{fig:admissibleintro}
+  \begin{mathpar}
+    \inferrule
+    {~}
+    {\Gamma ; \cdot \vdash \mathsf{nil} : \mathsf{Trace}_N
+      (q , q)}
+
+    \and
+
+    \inferrule
+    {\Gamma ; \Delta \vdash M : \mathsf{Trace}_N
+      (dst , q') \\
+    \exists \text{ transition } src \overset{c}{\to} dst
+    }
+    {\Gamma ; x : c , \Delta \vdash \mathsf{cons}(M) : \mathsf{Trace}_N
+    (src , q')}
+
+    \and
+
+    \inferrule
+    {\Gamma ; \Delta \vdash M : \mathsf{Trace}_N (dst , q')
+      \\
+    \exists~\varepsilon\text{-transition } src
+    \overset{\varepsilon}{\to} dst}
+    {\Gamma ; \Delta \vdash \mathsf{\varepsilon cons}(M) : \mathsf{Trace}_N
+    (src , q')}
+  \end{mathpar}
+  \caption{Admissible Trace Constructors}
+\end{figure}
+
+That is, we may use $\mathsf{nil}$ to terminate a trace that
+begins and ends at state $q$. The rules $\mathsf{cons}$ and
+$\mathsf{\varepsilon cons}$ are then used to inductively
+stitch traces together when sound. For instance, we can read
+the $\mathsf{cons}$ rule as saying that we can
+create a trace coming from a state $src$ provided that we may first
+transition via the character $c$ to state $dst$ and
+inductively build a trace starting from $dst$. The
+$\mathsf{\varepsilon cons}$ rule says something similar, but
+with an $\varepsilon$-transition instead of a transition
+labelled by a character. Recall that
+these rules are \emph{admissible}. They are not strictly
+necessary as primitives to conduct our proofs; however, they
+do provide convenient shorthand notation for building terms
+of type $\mathsf{Trace}_{N}(q , q')$.
+
+Dual to constructing traces, we often want to construct
+other terms in a context containing values of type
+$\mathsf{Trace}_{N}(q , q')$. For this purpose, we make use
+of the elimination principle for multiple-fixed points ---
+which we write as $\mathsf{mfold}$ --- given
+in \cref{subsubsec:other}.
+
+Note that this general construction will readily generalize to other types of
+automata. If we wanted to define say deterministic finite automata\footnote{DFAs
+\emph{could} just be defined as NFAs that happen to be deterministic. This is
+one way to do so, or you may choose to present the transition relation for the
+automaton as a transition \emph{function} instead. Concerns like these don't
+matter so much when defining things on paper, but at formalization time these
+are important and can make some proofs much easier.}, pushdown automata, Turing
+machines, etc, we just swap out the type of traces for a different, but very
+similar, trace construction.
+
+With this setting for finite automata, we can now internalize some classical theorems inside of our formal system.
+
+\subsection{A DFA Parser}
+\label{subsec:regexparser}
+
+Just as we encoded traces of NFAs as grammars, we likewise
+encode the traces of a DFA as grammars. The key difference
+between NFAs and DFAs is \emph{determinism} --- meaning,
+that in a state $q$ inside of DFA, given a character $c$ there
+will be exactly one transition that we may take leaving $q$
+with label $c$. For us, this changes the definition of valid
+transitions for a DFA, instead of the definition of
+$\mathsf{Trans}$ provided in \cref{subsec:finiteaut} DFAs
+obey
+
+\begin{gather*}
+ \mathsf{State} \in \{ g_{q} : q \in Q \} \\
+ \mathsf{Trans}(q) ::= \bigoplus_{c \in \Sigma} (c \otimes \mathsf{State})
+\end{gather*}
+
+Meaning, each state has a transition for every character.
+
+We now wish to define a parser term for DFA grammars. In
+particular, for a DFA $D$ we want to build a term,
+
+\[
+  w : \String \vdash \mathsf{parse}_{D} : \mathsf{AccTrace}_{D} \oplus \top
+\]
+
+where left injection into the output type denotes acceptance
+by the parser, and right injection denotes rejection. To
+build such a parser, it will suffice to construct a term
+
+\[
+  w : \String \vdash \mathsf{parse}_{D} : \LinSigTy q Q {\mathsf{Trace}_{D}(q_{0} , q)}
+\]
+This is because given a trace of a DFA, we may easily check
+if we should accept or reject by simply testing
+if the final state is accepting.
+
+Because $w$ is a Kleene star of characters, we may construct
+our desired $\mathsf{parse}_{D}$ term as a $\mathsf{foldl}$
+over $w$. In the empty case, we just have the trace that
+ends at the accepting state. In the recursive case, we
+effectively add to our trace by transitioning one character
+at a time, as we read them moving across $w$.
+
+\todoin{Derivation tree for this construction}
+
+Perhaps this derivation is not too surprising. All it says
+is that a DFA may be ran by transitioning a single character
+at a time, and then accepting or rejecting based on the
+final state. This is exactly what DFAs did initially, so
+what did we gain? Well, this has the benefit of our type
+system to ensure its correctness. Moreover, this construction exports to an
+intrinsically verified and executable DFA via Agda.
+
+\subsection{Regular Grammars and DFAs}
+\label{subsec:deriv}
+
+In order to extend the DFA
+parser from \cref{subsec:regexparser} to the construction of
+a verified parser
+generator for regular grammars we need to perform some
+plumbing establishing an equivalence between regular
+expressions and DFAs.
+
+There are several routes we may hope to take in establishing
+this equivalence. First, we could prove an equivalence
+between NFAs and regular expressions, and separately prove
+an equivalence between NFAs and DFAs.
+In \cref{subsec:eqproofs}, we include a version of
+Thompson's construction --- which established the
+equivalence between regular grammars and DFAs. We may additionally
+hope to internalize a variant of the powerset construction \cite{rabinFiniteAutomataTheir1959}
+--- which takes as input an NFA and constructs a DFA that
+recognizes the same language --- and combine the results of
+Thompson's construction and the powerset constructions to give an equivalence
+between regular expressions and DFAs.  This route is alluring, as it
+internalizes several classic grammar-theoretic constructions. However, it may
+necessitate extensions to the LNL theory, like
+a propositional truncation, and we have not yet investigated
+how this would interact with the existing types in the
+theory. The addition of a propositional truncation may seem
+harmless, but it is not always immediately clear how
+distinct constructions will interact. For instance, when
+exploring LNL models, Benton discovered that the synthesis
+of linear and dependent types require a new presentation of
+the $!$ modality from linear logic
+\cite{bentonMixedLinearNonlinear1995}. That is all to say,
+this is a work in progress and
+it is not immediate that the addition of a propositional
+truncation is adequate for establishing the weak equivalence
+between NFAs and DFAs.
+
+We may instead hope to internalize an equivalence between
+regular grammars and DFAs by using Brzozowski derivatives to
+directly create a DFA that is weakly equivalent to a given
+regular expression, as described by Owens et al.
+\cite{owensRegularexpressionDerivativesReexamined2009}.
+One characterization of regular grammars is that they are
+precisely those grammars which have finitely many inequivalent Brzozowski
+derivatives
+\cite{brzozowskiDerivativesRegularExpressions1964}.
+The algorithm used by Owens takes in a
+regular grammar and generates a DFA that recognizes the same
+language, and the states in this DFA are the finitely many
+derivative equivalence classes. We initially had a version of
+this theorem very roughly internalized in the LNL theory.
+To our taste, too much of this presentation relied on
+meta-arguments that lived outside of
+our formalism, and thus this particular phrasing of the
+theorem did not translate well into formalization.
+
+In any case, we believe
+that revisiting these lines of thought will lead to a
+satisfactory internalization of the equivalence between
+regular grammars and DFAs, and thus would bridge the gap
+between our DFA parser and a full regular expression parser.
+
+\subsection{Equivalence Between Regular Grammars and Finite Automata}
+\label{subsec:eqproofs}
+In this section, we describe a version of Thompson's
+construction \cite{thompsonProgrammingTechniquesRegular1968}
+   where we construct an NFA that recognizes a given regular
+expression. Moreover, we will show that this NFA is strongly equivalent to the
+original grammar. Witnessing this construction in our syntax has two benefits
+\begin{enumerate}
+  \item It reinforces this high-level view that the syntax is a natural and
+        general setting for formal grammar reasoning, as we demonstrate that
+        this formal system subsume results from existing systems, and
+  \item Following the
+        development of Thompson's construction, we then need
+        only establish the equivalence of NFAs and DFAs to
+        complete the full regular expression parser
+\end{enumerate}
+
+\begin{theorem}[Thompson]
+  \label{thm:thompson}
+  For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same
+  language as $g$.
+\end{theorem}
+
+We make a pretty straightforward adaptation of Thompson's theorem to our setting,
+
+\begin{theorem}[Typed Thompson]
+  \label{thm:typthompson}
+  For $g$ a regular grammar $g$, there is an NFA $N$ such that $g$ is isomorphic
+  to $\mathsf{AccTrace}_{N}$.
+\end{theorem}
+
+\begin{proof}[Proof Sketch]
+  The following proof is currently formalized in Agda. Below
+  we give a walk through the structure of the proof.
+
+  Recall that regular grammars are inductively defined via
+  disjunction, concatenation, and Kleene star over literals
+  and the empty grammar. By induction over regular grammars,
+  we will construct an NFA that is equivalent to $g$.
+
+  First, define the recognizing NFA for the empty grammar
+  $I$.
+
+  \begin{figure}[h!]
+  \begin{tikzpicture}[node distance = 25mm ]
+    \node[state, initial] (1) {$1$};
+    \node[state, right of=1, accepting] (2) {$2$};
+
+    \path[->] (1) edge[below] node{$\varepsilon$} (2);
+  \end{tikzpicture}
+  \caption{$NFA(I)$}
+  \label{fig:emptyNFA}
+  \end{figure}
+
+  The type of traces from the initial state of $NFA(I)$ to the single
+  accepting state is given by,
+
+  \[
+    \mathsf{Trace}_{NFA(I)}(q_{1}, q_{2}) = \mu
+      \begin{pmatrix}
+         g_{q_{1}} := g_{q_{2}} \\
+         g_{q_{2}} := I
+      \end{pmatrix}. g_{q_{1}}
+  \]
+
+  The accepting traces through $NFA(I)$ are then described
+  as,
+
+  \[
+    \mathsf{AccTrace}_{NFA(I)} = \LinSigTy q {\{1 , 2\}} {\left( \mathsf{Trace}_{NFA(I)}(q_{1} , q) \amp \mathsf{acc}(q) \right)}
+  \]
+
+  A quick inspection of \cref{fig:emptyNFA} reveals that the
+  only reasonable choice for $q$ is $q_{2}$ --- because
+  state 2 is accepting while state 1 is not. Therefore,
+
+  \begin{align*}
+    \mathsf{AccTrace}_{NFA(I)}
+    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \amp \mathsf{acc}(q_{2}) \\
+    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \amp \top \\
+    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})
+  \end{align*}
+
+  From here, to prove
+  $I \cong \mathsf{AccTrace}_{NFA(I)}$ it suffices to show
+  $I \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$. Below we
+  give two parse transformers, one from $I$ to
+  $\mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$ and vice versa. The formalized
+  artifact provides the low-level details that these are
+  indeed mutually inverse.
+
+  \[
+    \inferrule
+    {p : I \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
+     \exists \text{~transition~} q_{1} \overset{\varepsilon}{\to} q_{2}
+    }
+    {p : I \vdash \mathsf{\varepsilon cons}(\mathsf{nil}) : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})}
+  \]
+
+  Let $\gamma$ be the substitution $\{ g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$,
+
+  \[
+    \inferrule
+    {
+      p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash p :
+        \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \\
+      x_{1} : \simulsubst {g_{q_{2}}} {\gamma} = I \vdash x_{1} : I \\
+      x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
+    }
+    {
+      p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.x_{1} , x_{2}.x_{2})(p) : I
+    }
+  \]
+
+  This concludes the proof for the case of the empty
+  grammar. Let's now walk through the construction for
+  literal grammars. Given a character $c$, we construct an
+  NFA that recognizes only the string containing the single
+  character $c$ as,
+
+  \begin{figure}[h!]
+  \begin{tikzpicture}[node distance = 25mm ]
+    \node[state, initial] (1) {$1$};
+    \node[state, right of=1, accepting] (2) {$2$};
+
+    \path[->] (1) edge[below] node{$c$} (2);
+  \end{tikzpicture}
+  \caption{$NFA(c)$}
+  \label{fig:literalNFA}
+  \end{figure}
+
+  The automaton in \cref{fig:literalNFA} induces the
+  following type of traces from $q_{1}$ to $q_{2}$.
+
+  \[
+    \mathsf{Trace}_{NFA(c)}(q_{1}, q_{2}) = \mu
+      \begin{pmatrix}
+         g_{q_{1}} := c \otimes g_{q_{2}} \\
+         g_{q_{2}} := I
+      \end{pmatrix}. g_{q_{1}}
+  \]
+
+  Through the same argument as the empty grammar, the only
+  state that is accepting is $q_{2}$ and thus,
+
+  \[
+    \mathsf{AccTrace}_{NFA(c)} \cong \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})
+  \]
+
+  To show the desired isomorphism of $c \cong NFA(c)$ we
+  make a similar argument as we did for the empty grammar
+  $I$, except we leverage the $\mathsf{cons}$ rule instead
+  of $\mathsf{\varepsilon cons}$. That is, the parse
+  transformers in either direction are given as,
+
+  \[
+    \inferrule
+    {\cdot \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
+     \exists \text{~transition~} q_{1} \overset{c}{\to} q_{2}
+    }
+    {p : c \vdash \mathsf{cons}(\mathsf{nil}) :
+      \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})}
+  \]
+
+  \[
+    \inferrule
+    {
+      p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash p :
+        \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \\
+      x_{1} : \simulsubst {(c \otimes g_{q_{2}})} {\gamma} = c \otimes I \vdash \mathsf{unitR}(x_{1}) : c \\
+      x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
+    }
+    {
+      p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.\mathsf{unitR}(x_{1}) , x_{2}.x_{2})(p) : c
+    }
+  \]
+
+  Where $\gamma$ is the substitution
+  $\{ c \otimes g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$ and
+  $\mathsf{unitR}$ is a witness to the isomorphism
+  $c \otimes I \cong c$. Again, we may see that these do
+  indeed mutually invert each other in the Agda code.
+
+  It remains to show that the desired isomorphisms are
+  preserved by $\otimes$, $\oplus$, and Kleene star. Here,
+  we will give the argument for the disjunction case, the
+  others are defined quite similarly.
+
+  Given two NFAs $N$ and $M$, we may define a new NFA that
+  encodes the disjunction of $N$ and $M$. Denote the
+  internal states of $N$ by $q_{j}$'s and the internal states
+  of $M$ by $r_{k}$'s,
+
+  \begin{figure}[h!]
+  \begin{tikzpicture}[node distance = 20mm ]
+    \node[state] (2) {$q_{init}$};
+    \node[state, initial, below left of=2] (1) {$init$};
+    \node[state, below right of=1] (3) {$r_{init}$};
+    \node[right of=2] (4) {$\cdots$};
+    \node[right of=3] (5) {$\cdots$};
+    % \node[state, below right of=1] (3) {$3$};
+    % \node[state, below of=2] (4) {$4$};
+
+    \path[->] (1) edge[below] node{$\varepsilon$} (2);
+    \path[->] (1) edge[below] node{$\varepsilon$} (3);
+    \path[->] (2) edge[below] node{} (4);
+    \path[->] (3) edge[below] node{} (5);
+
+    \node[label={[name=l] $N$}, draw,line width=2pt,rounded corners=5pt, fit=(2)(4)] {};
+    \node[label={[name=l] $M$}, draw,line width=2pt,rounded corners=5pt, fit=(3)(5)] {};
+  \end{tikzpicture}
+  \caption{$N \oplus_{NFA} M$}
+  \label{fig:disjunctionNFA}
+  \end{figure}
+
+  \cref{fig:disjunctionNFA} shows the process of
+  disjunctively combining NFAs. Precisely, we add a single
+  new state and we included copies of the states from each of $N$
+  and $M$. The new state acts as the initial state and has
+  $\varepsilon$-transitions to the initial states of $N$ and
+  $M$. We include all of the internal transitions from $N$
+  and $M$, and the accepting states of $N \oplus_{NFA} M$
+  are exactly the accepting states in each subautomaton.
+
+  Let $g$ and $g'$ be two regular grammars such that
+  $g \cong NFA(g)$ and $g' \cong NFA(g')$. As a matter of
+  notation\footnote{We shall similarly abuse notation for
+    $\otimes$ and Kleene. That is, for a regular grammar
+    $g$, when we write $NFA(g)$ we mean the NFA inductively
+    built up with the NFA-analogues to the constructors that
+    built up $g$.}, we will write $NFA(g \oplus g')$ for
+  $NFA(g) \oplus_{NFA} NFA(g')$. The traces of
+  $NFA(g \oplus g')$ are then given by,
+
+  \[
+    \mathsf{Trace}_{NFA(g \oplus g')}(src , dst) = \mu
+      \begin{pmatrix}
+        g_{init} := g_{q_{0}} \oplus g_{r_{0}} \\
+        g_{q_{j}} := \mathsf{Trans}_{NFA(g)}(q_{j}) \oplus \mathsf{isDst}(q_{j}) \\
+        g_{r_{k}} := \mathsf{Trans}_{NFA(g')}(r_{k}) \oplus \mathsf{isDst}(r_{k})
+      \end{pmatrix}.g_{src}
+    \]
+
+  where $\mathsf{Trans}$ is used to echo the same syntactic
+  definitions that appear in the $NFA(g)$ and $NFA(g')$.
+  Also, $src$ and $dst$ may take on any value in
+  $Q := \{init\} \cup \{q_{j}\} \cup \{r_{k}\}$, and
+  $\mathsf{isDst}(q)$ checks if $q$ is equal to $dst$. Which
+  is all to say, the traces of $NFA(g \oplus g')$ comprise
+  either a trace of $NFA(g)$, or a trace of $NFA(g')$, and
+  the transition coming out of $g_{init}$ determines which
+  subautomaton we step into.
+
+  The parse transformer from $g \oplus g'$ checks which side
+  of the sum type we are on, then takes the appropriate step
+  from $g_{init}$ in the automaton.
+
+  \[
+    \inferrule
+    {
+      u : g \vdash \iota (\phi (u)) : \mathsf{AccTrace}_{NFA(g \oplus g')} \\
+      v : g' \vdash \iota' (\psi (v)) : \mathsf{AccTrace}_{NFA(g \oplus g')}
+    }
+    {p : g \oplus g' \vdash \mathsf{case}~p \{ \mathsf{inl}(u) \mapsto s , \mathsf{inr}(v) \mapsto r \} : \mathsf{AccTrace}_{NFA(g \oplus g')}}
+  \]
+
+  with $\iota$ and $\iota'$ as embeddings from $NFA(g)$ and
+  $NFA(g')$, respectively, into $NFA(g \oplus g')$,
+  $\phi: g \cong \mathsf{AccTrace}_{NFA(g)}$, and $\psi: g' \cong \mathsf{AccTrace}_{NFA(g')}$. On a
+  high level, all this construction does is turn a parse of
+  $g$ into a parse of $NFA(g)$ and then embeds that inside
+  of the larger automaton $NFA(g \oplus g')$. Likewise for $g'$.
+
+  In the other direction, recall that the data of an
+  accepting trace for $NFA(g \oplus g')$ is a pair of a
+  trace and a proof that
+  the end state $q'$ of that trace is accepting. By
+  multifolding over the first part of that pair, we turn the
+  term of type
+  $\mathsf{Trace}_{NFA(g \oplus g')}(init , q')$ into a
+  trace of either of the subautomata,
+
+  \[
+    p : \mathsf{Trace}_{NFA(g \oplus g')}(init , q') \vdash \mathsf{mfold}_{NFA(g \oplus g')} : \mathsf{Trace}_{NFA(g)}(q_{0} , q') \oplus \mathsf{Trace}_{NFA(g')}(r_{0} , q')
+  \]
+
+  Additionally, we leverage the fact that the only accepting
+  states for $NFA(g \oplus g')$ are those from the
+  subautomata to extract that $q'$ must be an accepting
+  state from a subautomaton.
+
+  \[
+    x : \mathsf{acc}_{NFA(g \oplus g')}(q') \vdash M : \mathsf{acc}_{NFA(g)}(q') \oplus \mathsf{acc}_{NFA(g')}(q')
+  \]
+
+  We then combine the trace and proof of acceptance into an
+  accepting trace of one of the subautomata,
+
+  \[
+    p : \mathsf{AccTrace}_{NFA(g \oplus g')} \vdash N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')}
+  \]
+
+  Lastly, we then inductively use the isomorphisms $\phi$
+  and $\psi$ to turn the accepting traces into a parse of
+  $g$ or $g'$,
+
+  \[
+     N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')} \vdash \mathsf{case}~N~\{\mathsf{inl}(n) \mapsto \phi^{-1}(n), \mathsf{inr}(n') \mapsto \psi^{-1}(n')\} : g \oplus g'
+  \]
+
+  As discussed above, the other cases and low-level details
+  of isomorphism are reserved for the formalization. This
+  concludes the proof of our variant of Thompson's construction.
+\end{proof}
+
+\subsection{Other Automata}
+\subsubsection{Pushdown Automata}
+\label{subsubsec:pda}
+A (nondeterministic) \emph{pushdown automaton} is an automaton that employs a
+stack. Just like NFAs, they have transitions labeled with characters from a
+fixed string alphabet $\Sigma$. Additionally, they maintain a stack of
+characters drawn from a stack alphabet $\Gamma$. They are often represented
+formally as a 7-tuple $(Q, \Sigma, \Gamma, \delta, q_{0}, Z, F)$,
+
+\begin{itemize}
+  \item $Q$ a finite set of states
+  \item $\Sigma$ a fixed, finite string alphabet
+  \item $S$ a fixed, finite stack alphabet
+  \item
+        $\delta \subset Q \times (\Sigma \cup \{ \varepsilon \}) \times S \to \mathcal{P}(Q \times S^{*})$
+        the transition function
+  \item $q_{0} \in Q$ the start state
+  \item $Z \in S$ the initial stack symbol
+  \item $F \subset Q$ the accepting states
+\end{itemize}
+
+We encode the traces of a pushdown automaton very similarly to those of an NFA,
+except the transitions of a PDA are instead encoded via the linear-non-linear
+$\Pi$-type. This is because of simply transitioning via character, a PDA must
+also pop and push characters onto a stack, which is used as the argument to
+these dependent functions.
+
+Let $S$ be a non-linear type encoding the stack
+alphabet, and build lists over $S$ as the (non-linear) least fixed-point
+$\mathsf{List}(S) := \mu X . 1 + S \times X$. Then, the type of states for a
+PDA $P$ with stack alphabet $S$ are given as a functions that takes in lists $L$,
+and then makes a case distinction between possible transitions based on what was witnessed as $\mathsf{head}(L)$. The choice of transition
+will then determine which character to transition by and what word $w$ should be
+pushed onto the stack. The word that is added to the top of the stack is
+appended to $\mathsf{tail}(L)$ and then we recursively step into another state
+called on argument $w + \mathsf{tail}(L)$.
+
+% \begin{gather*}
+%   \mathsf{State} \in {g _{q} : q \in Q} \\
+%   \mathsf{Word} \in \String \\
+%   \mathsf{Char} \in \Sigma   \\
+%   \mathsf{StackChar} \in \Gamma \\
+% \end{gather*}
+% \begin{align*}
+%   \mathsf{Trans}_{P}(q) ::=~&
+%                            \LinPiTy {(hd :: tl)} {\mathsf{List}(S)} {\left( \mathsf{Char} \otimes \mathsf{State}(\mathsf{Word} + tl) \right)}~| \\
+%   & \LinPiTy {(hd :: tl)} {\mathsf{List}(S)} {\left(  \mathsf{State}(\mathsf{Word} + tl) \right)}~|~\mathsf{Trans}_{P}(q) \oplus \mathsf{Trans}_{P}(q)
+% \end{align*}
+
+% That is to say, when transitioning a PDA pops off the head $hd$ of the stack
+
+\subsubsection{Turing Machines}
+\label{subsubsec:tm}
+
+In
+\cref{subsubsec:pda}, we gave a grammar presentation of
+traces through a PDA by using a non-linear type $S$
+to encode the stack. We may similarly use pairs $S \times S$
+to encode the tape of a Turing machine. With two stacks we can simulate the behavior of the
+infinite tape of a Turing machine. The intuition behind this correspondence is
+that the left half of the tape is on one stack, the right
+half of the tape the other, and we treat the tops of stacks
+like the head of the tape.
+
+Due to computability limitations, we are not able to extract
+a verified parser for all recursively enumerable languages.
+Yet we should still reflect on the generality of our syntax,
+as it allows a natural encoding of Turing machines.
+
+\section{Future Work}
+\label{sec:future}
+
+\subsection{Implementation for Context-Free Grammars}
+As suggested throughout the paper and briefly explored in
+\cref{subsubsec:pda}, the first extension to the work in
+this paper will be to bring analogous constructions to
+context-free grammars and their accompanying pushdown automata.
+
+\subsection{Beyond Strings}
+\label{subsec:beyond}
+
+While parsing typically refers to the generation of semantic
+objects from string input, many tasks in programming can be
+viewed as parsing of objects with more structure, such as
+trees with binding structure or graphs. Fundamental to the
+frontend of many
+programming language implementations are type systems. In
+particular, \emph{type checking}
+--- analogous to language recognition --- and \emph{typed
+  elaboration} --- analogous to parsing --- arise when
+producing a semantic object subject to some analysis. Just
+as our string grammars were given as functors from $\String$
+to $\Set$, we envision adapting the same philosophy
+to functors from $\String$ to \emph{trees} to craft a syntax
+that natively captures typed elaboration. This suggests an
+unusual sort of bunched type theory, where context extension
+no longer resembles concatenation of strings but instead
+takes on the form of tree constructors.
+
+Thus far, our theory has proved useful for internalizing long-standing
+grammar-theoretic constructions, but there has been decades of research
+conducted since. A fruitful avenue for future work includes
+testing if our formalism can also internalize more recently proposed grammar
+mechanisms, such as the interval parsing grammars given by Zhang et al. \cite{zhangIntervalParsingGrammars2023}.
+
+\newpage
+
+\bibliographystyle{plain}
+\bibliography{refs.bib}
+
+\newpage
+
+\appendix
+
+\section{Inference Rules}
+\label{sec:infer}
+
+\begin{figure}[h!]
+  \label{fig:structjdg}
+  \begin{mathpar}
+    \inferrule{~}{\ctxwff \cdot}
+    \and
+    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}
+
+    \\
+
+    \inferrule{~}{\linctxwff \Gamma \cdot}
+    \and
+    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}
+
+    \\
+
+    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}
+
+    \and
+
+    \inferrule{\Gamma \vdash A : L_i}{\linctxwffjdg \Gamma A}
+
+    \\
+
+    \inferrule{\Gamma \vdash X \equiv Y : U_i}{\ctxwffjdg \Gamma {X\equiv Y}}
+
+    \and
+
+    \inferrule{\Gamma \vdash A \equiv B : L_i}{\linctxwffjdg \Gamma {A \equiv B}}
+
+  \end{mathpar}
+  \caption{Structural judgments}
+\end{figure}
+
+\begin{figure}
+  \label{fig:typewf}
+  \begin{mathpar}
+    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
+ %
+    \and
+%
+    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
+%
+    \\
+%
+    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
+%
+    \and
+%
+    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
+%
+    \\
+%
+    \inferrule{~}{\Gamma \vdash 1 : U_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
+%
+    \\
+%
+    \inferrule{~}{\Gamma \vdash I : L_i}
+ %
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
+%
+    \\
+%
+    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
+%
+    \\
+%
+    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
+    %
+    \and
+    %
+    \inferrule{~}{\Gamma \vdash \top : L_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
+%
+    \\
+    %
+    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
+    %
+    \and
+    %
+    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
+  \end{mathpar}
+  \caption{Type well-formedness}
+\end{figure}
+
+\begin{figure}
+  \label{fig:jdgeq}
+  \begin{mathpar}
+    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
+%
+    \and
+%
+    \inferrule{~}{}
+%
+    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
+    %
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
+%
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
+    %
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
+\end{mathpar}
+  \caption{Judgmental equality}
+\end{figure}
+
+\begin{figure}
+  \label{fig:inttyping}
+  \begin{mathpar}
+  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
+  %
+  \\\
+  %
+  \inferrule{~}{\Gamma \vdash () : 1}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
+  %
+  \\
+%
+  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
+  \and
+  \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
+  %
+  \\
+  %
+  \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
+  \and
+  \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
+  \end{mathpar}
+  \caption{Intuitionistic typing}
+\end{figure}
+
+\begin{figure}
+  \label{fig:linsyntax}
+  \begin{mathpar}
+    \inferrule{~}{\Gamma ; a : A \vdash a : A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
+    %
+    \\
+    %
+    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
+    %
+    \\
+    %
+    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
+    %
+    \\
+    %
+    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
+    \\
+    %
+    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
+    \and
+    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta', \Delta \vdash \applto {e'} {e} : B}
+    \\
+    %
+    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta, \Delta' \vdash \apptol e {e'} : B}
+    %
+    \\
+    %
+    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}{\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma ; \Delta \vdash \app e {e'} : \subst A {e'} x}
+    %
+    \\
+    %
+    \inferrule{\Gamma \vdash e : X \quad \Gamma ; \Delta \vdash e' : \subst A e x}{\Gamma ; \Delta \vdash (e, e') : \LinSigTy x X A}
+    %
+    \\
+    %
+    \inferrule{\Gamma ; \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
+    %
+    \\
+    %
+    \inferrule{~}{\Gamma ; \Delta \vdash () : \top}
+    %
+    \\
+    %
+    \inferrule{\Gamma ; \Delta \vdash e_1 : A_1 \quad \Gamma ; \Delta \vdash e_2 : A_2}{\Gamma ; \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : A_1 \amp A_2 }{\Gamma ; \Delta \vdash \pi_i \, e : A_i}
+    %
+    \\
+    %
+    \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma ; \cdot \vdash \mathsf{G}^{-1}\, e : A}
+    %
+    \\
+    %
+    \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
+    \and
+    \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
+  \end{mathpar}
+  \caption{Linear typing}
+\end{figure}
 \end{document}

### 2024-07-07 POPL/paper.tex
scaffolding the type theory section

@@
 
-\section{Kleene Category}
-\steven{No changes pasted from LICS draft}
-\steven{Are these now ``Chomsky categories''? or is that disinct?}
+\section{Categorification of Formal Grammars}
+\steven{Idk if this section has the right title, but I wanted to change it if we are adding ``chomsky categories''}
 
@@
 \begin{definition}
-  A model is a locally Cartesian category with two hierarchies of
+  A \emph{Chomsky Category} is a locally Cartesian category with two hierarchies of
   universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
@@
 \section{Related work}
+
+\steven{TODO: move to related work section at the end}
 
@@
 %% Vaughan Pratt on the residual operator
-
-\section{Rest of paper}
-\steven{Everything below still needs to be edited from my prelim paper}
-\steven{Need to grab the canonicity proof from the LICS draft}
-
-Substructural logics, like linear and separation logic, offer precise control
+%%
+
+\section{Non-commutative Linear-non-linear Type Theory as a Syntax for Formal Grammars}
+\label{sec:tt}
+
+In 1995, Benton et al.
+\cite{bentonMixedLinearNonlinear1995} first gave categorical models of a logic
+that synthesized linear and non-linear logic. Substructural logics, like linear and separation logic, offer precise control
 over how values from a type may be used, in particular how many times a value
@@
 
-Dependent type theory extends the simply-typed $\lambda$-calculus by allowing type
-definitions to contain the occurrence of a term. In lieu of set-based
-foundations, this extension can then be taken as a foundation for mathematics.
-The expressive power unlocked by this insight has enabled the development of
-dependently-typed proof assistants
-such as Coq or Agda. Within each of these languages, systems may be given very
-precise functional specifications of their behavior.
-
-In 1995, Benton et al.
-\cite{bentonMixedLinearNonlinear1995} first gave categorical models of a logic
-that synthesized linear and non-linear logic. Later, Krishnaswami et al.
-\cite{krishnaswami_integrating_2015} built upon this work to develop a type
+Building off of Benton et al., Krishnaswami et al.
+\cite{krishnaswami_integrating_2015} developed a type
 system that unifies both linear and
 dependent types --- referred to as a \emph{linear-non-linear} (LNL) type theory. In
-such settings, we have two sub-theories --- one linear and one non-linear. The
+such settings, there are two sub-theories --- one linear and one non-linear. The
 linear sub-theory embeds ordinary linear types without dependence, while the
@@
 
-To the end of defining real verified
-parsers, in our LNL theory we define a parser as a term,
-
-\crefname{equation}{}{}
-\begin{equation}
-  \label{eq:parser}
-  w : \String \vdash \mathsf{parse}_{g}(w) : g \oplus \top
-\end{equation}
-
-We will investigate the precise meaning of judgments like \cref{eq:parser} in
-\cref{sec:tt}, but first let's get an intuitive feel for the meaning behind such a
-term. On the left of the turnstile occurs a single string $w$, we may think of
-this describing what resources are linearly available to us. Using each piece of
-our the context on the left of the turnstile, we build a term
-$\mathsf{parse}_{g}(w)$ of type $g \oplus \top$. Notably, the type being
-instantiated is a sum type, and the data of which side we inject into denotes
-whether the string is accepted or rejected by the parser. That is all to say, a
-parser takes in an input string and either matches it to the desired grammar or
-rejects the string.
-
-Practical parsers are often implemented via an abstract automaton. For instance,
-a parser may translate a grammar into a state machine recognizes the
-same language as the original grammar, as the automata
-point-of-view lends itself to straightforward execution. The
-strength of our syntax is that it allows us to describe
-these automata as grammars themselves. Moreover, the
-equivalence between the grammar and automata presentations
-is derivable as a proof term within the theory. This allows
-us access to simultaneously internalization of the equivalence
-between grammars and automata as well as the ability to
-export to automaton implementations of a parser that are executable.
-
-To realize an executable implementation of our type theory,
-we shallowly embed it in Agda. Because our implementation is
-defined as a function in Agda, we can leverage the Agda
-ecosystem for code generation. Further, in lieu of
-implementing a complete language from scratch, we make use of the
-Agda proof system as a setting to guarantee correctness of
-our generate parsers.
-
-Within this paper, we give our progress in defining a verified
-parser for regular expressions. The construction of such a parser entails
-the definition of a parser for deterministic finite automata (DFA) grammars,
-and an equivalence proof between DFAs and regular expressions. We have thus far
-completed the construction of a parser for DFAs $D$,
-
-\[
-  w : \String \vdash \mathsf{parse}_{D}(w) : D \oplus \top
-\]
-
-and half of the equivalence proof between DFAs and regular expressions, both
-formalized in Agda. Upon completion of the second half of this proof, we would
-have a system where a user could simply give a regular grammar and then extract
-an intrinsically verified parser for it. In future work, we
-hope to show an analogous construction involving pushdown
-automata (PDAs) and context-free grammars.
-
-\section{A Syntax-Independent Notion of Syntax}
-\label{sec:synindsyn}
-
-% \max{this section doesn't flow naturally from the intro, which is all about the
-%   type theory and parsing. This is where you want to outline in the intro and
-%   possibly say here that we want to revisit ``formal grammar theory'' and then
-%   we will make a type theory that has formal grammar theory as its semantics.}
-
-In this section we will revisit some foundational ideas in formal
-grammar theory, before giving a type
-theory for which formal grammars serves as its semantics.
-
-Formal languages and grammars are among some of the oldest and most studied areas in
-theoretical computer science. The developments of the 1950s and 1960s gave rise
-to expressive formalism for reasoning about grammars, and further practical
-algorithms for constructing parsers for regular and context-free grammars.
-
-A grammar $G$ is often described generatively as a set of production rules for
-generating strings. Often this is presented as a finite set of nonterminal
-symbols $N$, a finite set of terminal symbols $\Sigma$, and a finite set of rewriting
-rules for substitution of nonterminals. In this setting, we continually apply
-rewrite rules until we are left only with a string of terminals
-$w \in \String$. We will give a more precise notion of
-grammar in a moment, but for now let us take this intuitive view of grammars as
-string generators.
-
-One of the primary objects of study is that of a \emph{formal language} over
-a fixed alphabet $\Sigma$. A language $L$ is a \emph{subset} of strings
-$L \subseteq \String$. This notion of formal language gives an intuitive
-set-based semantics for grammars. That is, we may interpret a grammar $G$ as the
-language $L(G)$ that is generated by it. We are yet to give any precise
-definition of what we mean by ``grammar'', although we note that any suitable
-notion of grammar can be given a language semantics as above.
-
-Formal languages serve as a basis for mathematically specifying a
-\emph{recognizer} --- a
-function $r : \String \to \Bool$ that determines if a string belongs to a
-given language. However, they are insufficient for specifying a \emph{parser}
---- a function that decomposes a string into semantic components that adhere to
-the structure of a formal grammar. In practice, parsers don't just tell you
-if a string is valid or invalid. Instead, they emit structured data such as
-abstract syntax trees (ASTs).
-
-This motivates our first conceptual contribution, a simple and
-syntax-independent notion of formal grammar:
-
-\begin{definition}
-  \label{def:grammar}
-  A \emph{formal grammar} $G$ over a fixed alphabet $\Sigma$ is a function
-  $G : \String \to \Set$.
-\end{definition}
-
-We say that a grammar $G$ associates to every string $w$ the set\footnote{We
-  make little comment as to which foundations we use to encode sets. We argue\max{drop ``We argue''}
-  that our construction is agnostic to the particular foundation, and so without
-  loss of generality\max{incorrect use of ``without loss of generality''} we may take a class of small sets, a Grothendieck universe,
-  a type theoretic universe, or any similar foundation. Here we choose a type
-  theoretic universe.} $G w$ of parse
-trees showing that $w$ matches $G$.
-
-The simplest class of formal grammar on the Chomsky\max{first mention of Chomsky hierarchy, why not bring it up earlier when you are describing what a grammar is? Also cite}
-hierarchy are the \emph{regular expressions}
-
-\max{inaccurate reflection of common terminology. These are regular expressions,
-  regular grammars usually means a CFG with a restriction on recursion that
-  ensures you only allow Kleene star.}.
-
-\begin{gather*}
-  \mathsf{Char} \in \Sigma \\
-  \mathsf{Reg} ::= \mathsf{Char}~|~I~|~\mathsf{Reg} \otimes \mathsf{Reg}~|~\mathsf{Reg} \oplus \mathsf{Reg}~|~\mathsf{Reg}^{*}
-\end{gather*}
-
-Here, we use a non-standard syntax inspired by linear logic. Disjunction of
-regular expressions is written $\oplus$, and concatenation of regular
-expressions is written like the linear tensor product $\otimes$. The
-empty grammar is sometimes written in the same manner as the empty string,
-$\varepsilon$, but we instead choose to represent the empty grammar by $I$. This
-is also linear logic syntax to represent the unit $I$ of the operation $\otimes$.
-
-To see an illustration of the notion of grammar given in \cref{def:grammar}, consider the regular expression
-$G = (a \otimes b)^{*} \otimes a$ --- a sequence of 0 or more $ab$'s followed by
-an $a$. We may think of $G$ as a function that takes in strings and outputs all
-of the parse trees of that string. Passing the input string $baa$ through $G$
-would output the empty set, because there are no $G$-parses of $baa$. However,
-passing $aba$ through $G$ would output a singleton set, as there is only a
-single $G$-parse of $aba$. Namely, the one that breaks up $aba$ as $(ab)a$.
-
-Observe that for any input string $w$ for the above grammar $G$, $G(w)$ is
-either empty or a singleton set. Either the string isn't parsed by $G$,
-or it is parsed by $G$ for exactly one reason. That is, $G$ is
-\emph{unambiguous}. If we instead consider the grammar
-$H = (a \otimes b)^{*} \oplus (b \otimes a)^{*}$ there are now ambiguities. For
-instance, what is the action of $H$ on the empty string $\varepsilon$?
-The empty string $\varepsilon$ matches both $(a \otimes b)^{*}$ and $(b \otimes a)^{*}$, so
-$\varepsilon$ can match either the left or the right pattern
-of $H$. As a result, $H(\varepsilon)$ is a two-element set.
-
-\[
-  L_{G} = \{ w \in \String : G w \text{ is inhabited} \}
-\]
-
-As they are just sets, formal languages naturally are
-endowed with a partial
-order via the subset inclusion ordering. Similarly, formal grammars are
-naturally endowed with the structure of a \emph{category} where a morphism
-$f : G \to H$ is given by a \emph{family} of functions:
-
-\[
-  f^{w} : Gw \to Hw, w \in \String
-\]
-
-Intuitively, we read a morphism of grammars as a \emph{parse transformer}: $f$
-translates a
-$G$-parse of $w$ into an $H$-parse of $w$. The notion of isomorphism induced by
-this categorical structure encodes that two grammars are equivalent if there is
-a bijective translation of parses. Isomorphisms in the category of
-grammars are precisely Chomsky's notion of \emph{strong equivalence} of
-grammars \cite{chom1963}.
-
-The category of grammars is quite rich in structure. It can
-equivalently be described as the functor category $\Set^{\String}$, where
-$\String$ is a discrete category. That is, the objects in $\String$ are strings
-and the only morphisms are identity morphisms. Such functor categories into
-$\Set$ --- often called presheaf categories --- carry a remarkable amount of
-structure. In fact, any presheaf category naturally models dependent type
-theory, which we exploit to motivate the design of the formalism
-presented in \cref{sec:tt}.
-
-\section{A Type Theory to Reason About Formal Grammars}
-\label{sec:tt}
+\steven{Next paragraph is paper scaffolding for what I want to say and is total word salad rn}
+In this paper, we instantiate an LNL theory to axiomatize the notion of a Chomsky category, and because $\Set^{\String}$ forms a Chomsky category, using the type theory corresponds to interesting formal grammar proofs.
+Internalization of previous results + use to build correct-by-construction parsers.
+
+\steven{Expand here on how linearity has an intuitive parsing intrerpration, however working only in a simply typed system is not sufficient.
+i.e.
+dependence offers the right view on indexed grammars.
+}
 
@@
 In this section we will walk through most of the syntax of our type theory. We
-include the full syntax in \cref{sec:infer}.
+include the full syntax in \cref{sec:infer}. \steven{full syntax in appendix}
 
@@
 \end{theorem}
+\steven{In Agda as a weak equivalence right now}
 

### 2024-07-07 POPL/paper.tex
more paper organization

@@
 operations which results in the Kleene star being more easily
-axiomatized \cite{kozen1994}.
+axiomatized \steven{TODO cite}
+% \cite{kozen1994}.
 
@@
 
-\begin{lemma}[\cite{day1970}]
+\begin{lemma}[TODO cite
+  % \cite{day1970}
+  ]
   Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
@@
 \begin{definition}
-  A \emph{Chomsky Category} is a locally Cartesian category with two hierarchies of
+  A \emph{Chomsky category} is a locally Cartesian category with two hierarchies of
   universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
@@
 \end{theorem}
-
-\section{Related work}
-
-\steven{TODO: move to related work section at the end}
-
-\paragraph{Kleene Algebra}
-
-Since the early works in the theory of formal languages, Kleene
-algebras have played an important role in its development. They
-generalize the operations known from regular languages by introducing
-operations generalizing language composition, language union and the
-Kleene star.  More generally, they are defined as inequational theory
-where the inequality is meant to capture language containment. This
-theory is extremely successful, having found applications in algebraic
-path problems, theory of programming languages, compiler optimizations
-and more.
-
-A frequently fruitful research direction is exploring varying
-extensions of Kleene algebras, Kleene algebra with tests (KAT) being
-one of the most notable ones. Our approach is radically different from
-most extensions, which usually aim at modifying or adding new
-operations to Kleene algebras, but still keeping it as an inequational
-theory. By adopting a category-theoretic treatment and allowing the
-``order structure'' to encode more information than merely
-inequalities, we were able to extend Kleene algebra to reason about
-parsing as well.
-
-\paragraph{TODO: more stuff}
-
-%% That paper with Fritz Henglein about regexes as types
-
-%% Vaughan Pratt on the residual operator
-%%
 
@@
 
+The structrual judgements are shown in \Cref{fig:structjdg}, the typing well-formedness in \Cref{fig:typewf}, and the intuitionistic typing rules in \Cref{fig:inttyping}.
+There are all just as they appear in \cite{krishnaswami_integrating_2015}.
+
+Our contribution is the addition of base types drawn from the alphabet $\Sigma$ along with the omission of the exchange rule for linear typing judgements.
+
+\steven{TODO need to update syntax in these rules}
+
+\begin{figure}
+  \begin{mathpar}
+    \boxed{\ctxwff \Gamma}
+
+    \inferrule{~}{\ctxwff \cdot}
+    \and
+    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}
+
+    \\
+
+    \boxed{\linctxwff \Gamma \Delta}
+
+    \inferrule{~}{\linctxwff \Gamma \cdot}
+    \and
+    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}
+
+    \\
+
+    \boxed{\ctxwffjdg \Gamma X}
+
+    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}
+
+    \\
+
+    \boxed{\linctxwffjdg \Gamma A}
+
+    \inferrule{\Gamma \vdash A : L_i}{\linctxwffjdg \Gamma A}
+
+    \\
+
+    \boxed{\ctxwffjdg \Gamma {X \equiv Y}}
+
+    \inferrule{\Gamma \vdash X \equiv Y : U_i}{\ctxwffjdg \Gamma {X\equiv Y}}
+
+    \\
+
+    \boxed{\linctxwffjdg \Gamma {A \equiv B}}
+
+    \inferrule{\Gamma \vdash A \equiv B : L_i}{\linctxwffjdg \Gamma {A \equiv B}}
+
+  \end{mathpar}
+  \caption{Structural judgments}
+  \label{fig:structjdg}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
+ %
+    \and
+%
+    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
+%
+    \\
+%
+    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
+%
+    \and
+%
+    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
+%
+    \\
+%
+    \inferrule{~}{\Gamma \vdash 1 : U_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
+%
+    \\
+%
+    \inferrule{~}{\Gamma \vdash I : L_i}
+ %
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
+%
+    \\
+%
+    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
+%
+    \\
+%
+    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
+    %
+    \and
+    %
+    \inferrule{~}{\Gamma \vdash \top : L_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
+%
+    \\
+    %
+    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
+    %
+    \and
+    %
+    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
+  \end{mathpar}
+  \caption{Type well-formedness}
+  \label{fig:typewf}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+  \boxed{\Gamma \vdash x : X}
+
+  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
+  %
+  \\\
+  %
+  \inferrule{~}{\Gamma \vdash () : 1}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
+  %
+  \\
+%
+  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
+  \and
+  \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
+  %
+  \\
+  %
+  \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
+  \and
+  \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
+  \end{mathpar}
+  \caption{Intuitionistic typing}
+  \label{fig:inttyping}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \boxed{\Gamma ; \Delta \vdash a : A}
+
+    \inferrule{~}{\Gamma ; a : A \vdash a : A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
+    %
+    \\
+    %
+    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
+    %
+    \\
+    %
+    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
+    %
+    \\
+    %
+    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
+    \\
+    %
+    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
+    \and
+    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta', \Delta \vdash \applto {e'} {e} : B}
+    \\
+    %
+    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta, \Delta' \vdash \apptol e {e'} : B}
+    %
+    \\
+    %
+    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}{\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma ; \Delta \vdash \app e {e'} : \subst A {e'} x}
+    %
+    \\
+    %
+    \inferrule{\Gamma \vdash e : X \quad \Gamma ; \Delta \vdash e' : \subst A e x}{\Gamma ; \Delta \vdash (e, e') : \LinSigTy x X A}
+    %
+    \\
+    %
+    \inferrule{\Gamma ; \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
+    %
+    \\
+    %
+    \inferrule{~}{\Gamma ; \Delta \vdash () : \top}
+    %
+    \\
+    %
+    \inferrule{\Gamma ; \Delta \vdash e_1 : A_1 \quad \Gamma ; \Delta \vdash e_2 : A_2}{\Gamma ; \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : A_1 \amp A_2 }{\Gamma ; \Delta \vdash \pi_i \, e : A_i}
+    %
+    \\
+    %
+    \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma ; \cdot \vdash \mathsf{G}^{-1}\, e : A}
+    %
+    \\
+    %
+    \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
+    \and
+    \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
+  \end{mathpar}
+  \caption{Linear typing}
+  \label{fig:linsyntax}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \boxed{\Gamma \vdash e \equiv e' : X}
+
+    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
+%
+    \and
+%
+    \inferrule{~}{}
+%
+    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
+%
+    \and
+
+    \\
+
+    \boxed{\Gamma ; \Delta \vdash a \equiv a' : A}
+
+    \\
+%
+    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
+    %
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
+%
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
+    %
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
+\end{mathpar}
+  \caption{Judgmental equality}
+  \label{fig:jdgeq}
+\end{figure}
 \subsection{Syntax}
@@
 
-In this section we will walk through most of the syntax of our type theory. We
-include the full syntax in \cref{sec:infer}. \steven{full syntax in appendix}
-
-The linear typing judgment in our syntax takes on the following schematic form,
-\[
-  \Gamma ; \Delta \vdash M : A
-\]
-
-Let us briefly digest what this judgment is trying to say. First, $A$ represents
+\paragraph{Non-linear Types}
+
+The non-linear fragment of the theory
+corresponds to usual Martin-L\"of type theory. That is, the non-linear
+contexts obey all of the usual structural rules --- contraction, weakening, and
+exchange --- so we may use their contents as many times as we'd like.
+For the application of parsing, we think of these as primarily as indicies terms for indexed grammars.
+That is, they are as useful for their dependence.
+We may think of them as brining into scope definitions from ordinary mathematics to be used for indexing
+
+\paragraph{Linear Types}
+
+The linear typing judgment in our syntax takes on the following schematic form $\Gamma ; \Delta \vdash a : A$. First, $A$ represents
 a \emph{linear type} in our syntax. The intended semantics of these linear types
@@
 reflect the behavior of formal grammars. For this reason, we may often
-interchangeably use the terms ``linear type'' and ``grammar''. To see the
-correspondence between types and grammars, note that the constructors for linear types
-mirror operations on formal grammars. For example, we have linear type constructors
-$\otimes$, $\oplus$, and $\amp$ which correspond to grammar concatenation,
-disjunction, and conjunction, respectively. The other type constructors also have
-corresponding grammar intuition which may be seen in \cref{subsubsec:other}.
-
-$M$ is an inhabitant of type $A$, which we want to think of as a parse tree of
+interchangeably use the terms ``linear type'' and ``grammar''.
+
+The term $a$ is an inhabitant of type $A$, which we want to think of as a parse tree of
 the grammar $A$. The core idea of this entire paper follows precisely from this
 single correspondence: grammars are types, and the inhabitants of these types
-are parse trees for the grammars!
-
-$\Gamma$ represents a \emph{non-linear context} --- also referred to as an
-\emph{intuitionistic} context. The non-linear fragment of the theory
-corresponds to usual Martin-L\"of type theory. That is, the intuitionistic
-contexts obey all of the usual structural rules --- contraction, weakening, and
-exchange --- so we may use their contents as many times as we'd like.
-
-On the other hand $\Delta$ represents a \emph{linear context}. These linear
+are parse trees for the grammars.
+
+$\Gamma$ reprsents a non-linear context, while $\Delta$ represents a \emph{linear context}. These linear
 contexts
@@
 
-\todoin{with citation, talk somewhere about how CFGs are just regular + fixed points}
+% \todoin{with citation, talk somewhere about how CFGs are just regular + fixed points}
 \max{this is a good point to give an overview of how regular expressions and context-free grammars/expressions are essentially sub-languages of our type system.}
@@
 
+\subsection{Canonicity}
+
+Let $\cdot \vdash A$ be a closed linear type. Then there are
+two obvious notions of what constitutes a ``parse'' of a string w
+according to the grammar $A$:
+\begin{enumerate}
+\item On the one hand we have the set-theoretic semantics just
+  defined, $\llbracket A \rrbracket \cdot w$
+\item On the other hand, we can view the string $w = c_1c_2\cdots$ as
+  a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
+  define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
+  \lceil w \rceil \vdash e : G$.
+\end{enumerate}
+It is not difficult to see that at least for the ``purely positive''
+formulae (those featuring only the positive connectives
+$0,+,I,\otimes,\mu, \overline\Sigma,c$) that
+every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
+that the nodes of the tree correspond precisely to the introduction
+forms of the type. However it is far less obvious that \emph{every}
+linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
+sequence of introduction forms since proofs can include elimination
+forms as well. To show that this is indeed the case we give a
+\emph{canonicity} result for the calculus: that the parses for .
+
+\begin{definition}
+  A non-linear type $X$ is purely positive if it is built up using
+  only finite sums, finite products and least fixed points.
+
+  A linear type is purely positive if it is built up using only finite
+  sums, tensor products, generators $c$, least fixed points and linear
+  sigma types over purely positive non-linear types.
+\end{definition}
+
+\begin{definition}
+  %% Let $X$ be a closed non-linear type. The closed elements $\textrm{Cl}(X)$ of $X$ are the definitional equivalence classes of terms $\cdot \vdash e : X$.
+
+  Let $A$ be a closed linear type. The nerve $N(A)$ is a presheaf on
+  strings that takes a string $w$ to the definitional equivalence
+  classes of terms $\cdot; \lceil w\rceil \vdash e: N(A)$.
+\end{definition}
+
+\begin{theorem}[Canonicity]
+  Let $A$ be a closed, purely positive linear type. Then there is an
+  isomorphism between $\llbracket A\rrbracket$ and $N(A)$.
+\end{theorem}
+\begin{proof}
+  We outline the proof here, more details are in the appendix. The
+  proof proceeds first by a standard logical families construction
+  that combines canonicity arguments for dependent type theory
+  \cite{coquand,etc} with logical relations constructions for linear
+  types \cite{hylandschalk}. It is easy to see by induction that the
+  logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
+  \rrbracket$ and the fundamental lemma proves that the projection
+  morphism $p : \hat A \to N(A)$ has a section, the canonicalization
+  procedure. Then we establish again by induction that
+  canonicalization is also a retraction by observing that introduction
+  forms are taken to constructors.
+\end{proof}
+
+\begin{enumerate}
+\item Every term $\lceil w \rceil \vdash p : G + H$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
+\item There are no terms $\lceil w \rceil \vdash p : 0$
+\item If there is a term $\lceil w \rceil \vdash p : c$ then $w = c$ and $p = x$.
+\item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
+\item Every term $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
+\item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
+\item Every term $\lceil w \rceil \vdash p : \mu X. G$ is equal to $\textrm{roll}(q)$ where $q : G(\mu X.G/X)$
+\item Every term $\lceil w \rceil \vdash p : (x:A) \times G$ is equal
+  to $(M,q)$ where $\cdot \vdash M : A$
+\end{enumerate}
+
+To prove this result we will use a logical families model. We give a
+brief overview of this model concretely:
+\begin{enumerate}
+\item A context $\Gamma$ denotes a family of sets indexed by closing substitutions $\hat\Gamma : (\cdot \vdash \Gamma) \Rightarrow \Set_i$
+\item A type $\Gamma \vdash X : U_i$ denotes a family of sets $\hat X : \Pi(\gamma:\cdot \vdash \Gamma) \hat\Gamma \Rightarrow (\cdot \vdash \simulsubst X \gamma) \Rightarrow \Set_i$
+\item A term $\Gamma \vdash e : X$ denotes a section $\hat e : \Pi(\gamma)\Pi(\hat\gamma)\hat X \gamma \hat\gamma (\simulsubst e \gamma)$
+\item A linear type $\Gamma \vdash A : L_i$ denotes a family of grammars $\hat A : \Pi(\gamma:\cdot\vdash\Gamma)\,\hat\Gamma \Rightarrow \Pi(w:\Sigma^*) (\cdot;\lceil w\rceil \vdash A[\gamma])\Rightarrow \Set_i$, and the denotation of a linear context $\Delta$ is similar.
+\item A linear term $\Gamma;\Delta \vdash e : A$ denotes a function \[\hat e : \Pi(\gamma)\Pi(\hat\gamma)\Pi(w)\Pi(\delta : \lceil w \rceil \vdash \simulsubst \Delta \gamma) \hat\Delta \gamma \hat\gamma \delta \Rightarrow \hat A \gamma \hat\gamma w {(\simulsubst {\simulsubst e \gamma} \delta)}\]
+\end{enumerate}
+And some of the constructions:
+\begin{enumerate}
+\item $\widehat {(G A)} \gamma \hat\gamma e = \hat A \gamma \hat\gamma \varepsilon (G^{-1}e)$
+\item $\widehat {(A \otimes B)} \gamma \hat\gamma w e = \Sigma(w_Aw_B = w)\Sigma(e_A)\Sigma(e_B) (e_A,e_B) = e \wedge \hat A \gamma \hat \gamma w_A e_A \times \hat B \gamma \hat \gamma w_B e_B$
+\item $\widehat {(A \lto B)} \gamma \hat\gamma w e = \Pi(w_A)\Pi(e_A) \hat A \gamma\hat\gamma w_A e_A \Rightarrow \hat B \gamma\hat\gamma (ww_A) (\applto {e_A} e)$
+\end{enumerate}
+
+First, the category with families will be
+the category of logical families over set contexts/types
+$\Delta$/$A$. Then the propositional portion will be defined by
+mapping a logical family $\hat \Gamma \to \Gamma$
+
+First, let $L$ be the category of BI formulae and proofs (quotiented
+by $\beta\eta$ equality). Define a functor $N : L \to \Set^{\Sigma^*}$ by
+\[ N(\phi)(w) = L(w,\phi) \]
+
+Then define the gluing category $\mathcal G$ as the comma category
+$\Set^{\Sigma^*}/N$. That is, an object of this category is a pair of
+a formula $\phi \in L$ and an object $S \in \mathcal
+P(\Sigma^*)/N(\phi)$. We can then use the equivalence $\mathcal
+P(\Sigma^*)/N(\phi) \cong \mathcal P(\int N(\phi))$ to get a simple
+description of such an $S$: it is simply a family of sets indexed by
+proofs $L(w,\phi)$:
+\[ \prod_{w\in\Sigma^*} L(w,\phi) \to \Set \]
+This category clearly comes with a projection functor $\pi : \mathcal
+G \to \mathcal L$ and then our goal is to define a section by using
+the universal property of $\mathcal L$.
+
+To this end we define
+\begin{enumerate}
+\item $(\phi, S) \otimes (\psi, T) = (\phi \otimes \psi, S\otimes T)$ where
+  \[ (S \otimes T)(w, p) = (w_1w_2 = w) \times (q_1,q_2 = p) \times S\,w_1\,q_1 \times T\,w_2\,q_2\]
+\item $(\phi, S) \multimap (\psi, T) = (\phi \multimap \psi, S \multimap T)$ where
+  \[ (S \multimap T)(w,p) = w' \to q \to S\,w'\,q \to T (ww') (p\,q) \]
+\item $\mu X. ??$ ??
+\end{enumerate}
+
 \subsection{Semantics in Languages}
-Recall from \cref{sec:synindsyn} that every formal grammar induces a formal language. This corresponds to another
+% Recall from \cref{sec:synindsyn} that every formal grammar induces a formal language. This corresponds to another
 interpretation $L(\cdot)$ of our syntax. In particular, the semantics of a
@@
 
-\todoin{Derivation tree for this construction}
+% \todoin{Derivation tree for this construction}
 
@@
 
+\section{Related work}
+
+\paragraph{Kleene Algebra}
+
+Since the early works in the theory of formal languages, Kleene
+algebras have played an important role in its development. They
+generalize the operations known from regular languages by introducing
+operations generalizing language composition, language union and the
+Kleene star.  More generally, they are defined as inequational theory
+where the inequality is meant to capture language containment. This
+theory is extremely successful, having found applications in algebraic
+path problems, theory of programming languages, compiler optimizations
+and more.
+
+A frequently fruitful research direction is exploring varying
+extensions of Kleene algebras, Kleene algebra with tests (KAT) being
+one of the most notable ones. Our approach is radically different from
+most extensions, which usually aim at modifying or adding new
+operations to Kleene algebras, but still keeping it as an inequational
+theory. By adopting a category-theoretic treatment and allowing the
+``order structure'' to encode more information than merely
+inequalities, we were able to extend Kleene algebra to reason about
+parsing as well.
+
+\paragraph{TODO: more stuff}
+
+%% That paper with Fritz Henglein about regexes as types
+
+%% Vaughan Pratt on the residual operator
+%%
+
 \section{Future Work}
@@
 
-\newpage
-
-\appendix
-
-\section{Inference Rules}
-\label{sec:infer}
-
-\begin{figure}[h!]
-  \label{fig:structjdg}
-  \begin{mathpar}
-    \inferrule{~}{\ctxwff \cdot}
-    \and
-    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}
-
-    \\
-
-    \inferrule{~}{\linctxwff \Gamma \cdot}
-    \and
-    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}
-
-    \\
-
-    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}
-
-    \and
-
-    \inferrule{\Gamma \vdash A : L_i}{\linctxwffjdg \Gamma A}
-
-    \\
-
-    \inferrule{\Gamma \vdash X \equiv Y : U_i}{\ctxwffjdg \Gamma {X\equiv Y}}
-
-    \and
-
-    \inferrule{\Gamma \vdash A \equiv B : L_i}{\linctxwffjdg \Gamma {A \equiv B}}
-
-  \end{mathpar}
-  \caption{Structural judgments}
-\end{figure}
-
-\begin{figure}
-  \label{fig:typewf}
-  \begin{mathpar}
-    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
- %
-    \and
-%
-    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
-%
-    \\
-%
-    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
-%
-    \and
-%
-    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
-%
-    \\
-%
-    \inferrule{~}{\Gamma \vdash 1 : U_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
-%
-    \\
-%
-    \inferrule{~}{\Gamma \vdash I : L_i}
- %
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
-%
-    \\
-%
-    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
-%
-    \\
-%
-    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
-    %
-    \and
-    %
-    \inferrule{~}{\Gamma \vdash \top : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
-%
-    \\
-    %
-    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
-    %
-    \and
-    %
-    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
-  \end{mathpar}
-  \caption{Type well-formedness}
-\end{figure}
-
-\begin{figure}
-  \label{fig:jdgeq}
-  \begin{mathpar}
-    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
-%
-    \and
-%
-    \inferrule{~}{}
-%
-    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
-    %
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
-%
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
-    %
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
-\end{mathpar}
-  \caption{Judgmental equality}
-\end{figure}
-
-\begin{figure}
-  \label{fig:inttyping}
-  \begin{mathpar}
-  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
-  %
-  \\\
-  %
-  \inferrule{~}{\Gamma \vdash () : 1}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
-  %
-  \\
-%
-  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
-  \and
-  \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
-  %
-  \\
-  %
-  \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
-  \and
-  \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
-  \end{mathpar}
-  \caption{Intuitionistic typing}
-\end{figure}
-
-\begin{figure}
-  \label{fig:linsyntax}
-  \begin{mathpar}
-    \inferrule{~}{\Gamma ; a : A \vdash a : A}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
-    %
-    \\
-    %
-    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
-    %
-    \\
-    %
-    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
-    %
-    \\
-    %
-    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
-    \\
-    %
-    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
-    \and
-    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta', \Delta \vdash \applto {e'} {e} : B}
-    \\
-    %
-    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta, \Delta' \vdash \apptol e {e'} : B}
-    %
-    \\
-    %
-    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}{\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma ; \Delta \vdash \app e {e'} : \subst A {e'} x}
-    %
-    \\
-    %
-    \inferrule{\Gamma \vdash e : X \quad \Gamma ; \Delta \vdash e' : \subst A e x}{\Gamma ; \Delta \vdash (e, e') : \LinSigTy x X A}
-    %
-    \\
-    %
-    \inferrule{\Gamma ; \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
-    %
-    \\
-    %
-    \inferrule{~}{\Gamma ; \Delta \vdash () : \top}
-    %
-    \\
-    %
-    \inferrule{\Gamma ; \Delta \vdash e_1 : A_1 \quad \Gamma ; \Delta \vdash e_2 : A_2}{\Gamma ; \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : A_1 \amp A_2 }{\Gamma ; \Delta \vdash \pi_i \, e : A_i}
-    %
-    \\
-    %
-    \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma ; \cdot \vdash \mathsf{G}^{-1}\, e : A}
-    %
-    \\
-    %
-    \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
-    \and
-    \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
-  \end{mathpar}
-  \caption{Linear typing}
-\end{figure}
 \end{document}

### 2024-07-08 POPL/paper.tex
intro to syntax, intro editing, citations

@@
 
-\setcounter{page}{1}
-
-\section{Introduction}
-
 \section{A Syntax-Independent Notion of Syntax}
+\label{sec:intro}
+The theory of formal languages and parsing is one of the oldest and
+most thoroughly developed areas of theoretical computer science. A
+prominent topic in the 1950, 60s, and 70s led to a series of remarkable
+developments: Chomsky's hierarchy of grammar formalisms \cite{chomThreeModels1956}, practical
+algorithms for parsing of regular and context-free grammars and
+variants \cite{KNUTH1965607,Earley1970}, as well as implementations of practical tools for the
+generation of efficient parsers \cite{Johnsonyacc}.
+
 The notion of formal language is central to the theory of parsing. A
@@
   trees are usually not materialized in memory, as semantic actions often perform
-  computation over the trees while they are being constructed.}
+  computation over the trees while they are being constructed.}.
 
@@
 
-As they a simply sets, formal languages are naturally endowed with a partial
+As they are simply sets, formal languages are naturally endowed with a partial
 order via subset inclusion. Similarly, formal grammars are naturally endowed
@@
 different notions of formal grammar. For instance, Chomskyan generative
-grammars, Thue systems, Montague grammars, and so on are all distinct syntactic
+grammars, semi-Thue systems, Montague grammars, and so on are all distinct syntactic
 presentations of the same underlying idea of an abstract specification for
@@
   grammar} --- a category paired with a functor into
-$\Set^{\String}$. That is, they may all be interpreted in our category of grammars. In this sense, $\Set^{\String}$ provides a universal semantic
-domain to reason about parsing. The slice 2-category over $\Set^{\String}$ is
-then the appropriate setting to reason over notions of formal grammar.
+$\Set^{\String}$. That is, they may all be interpreted in our category of grammars. In this sense, $\Set^{\String}$ provides a universal setting to reason about parsing.
+The slice 2-category over $\Set^{\String}$ then provides the appropriate expressivity to prove equivalences between grammar formalism, such as the equivalence between semi-Thue systems and Chomsky's unresticted grammars.
 
@@
 
+This paper begins in \cref{sec:categorify} by giving a categorical interpretation of Kleene algebra.
+However, Kleene algebras are primarily concerned with the regular languages.
+With this in mind, we further abstract our categorical model from the connectives of regular grammars to those of context-free grammars and beyond.
+From there, in \cref{sec:tt} we provide a type theory to construct and reason about such formal grammars.
+In \cref{sec:semantics} we give interpretations of the syntax as grammars and as languages, and prove a canonicity theorem for the type theory.
+Finally, in \cref{sec:automata} we give a presentation of automata as grammars.
+With this view in mind, we may encode parsing algorithms and execute them entirely internal to our formalism.
 Our contributions are then:
 
-\steven{These sound awkward but I'm painting with broad strokes rn}
 \begin{itemize}
-  \item A syntax-independent definition of formal grammar
+  \item A universal semantic notion of formal grammar given by the category $\Set^{\String}$
+  \item Categorical structures that capture the concepts integral to formal grammar theory, and that are notably modeled by $\Set^{\String}$
   \item A dependent
-    linear/non-linear (LNL) type theory to syntactically capture the semantics of parsing
-  \item A categorification of Kleene Algebra
-  \item Interpretations of the LNL type system for reasoning about formal grammars and formal languages
-  \item The presentation of automata as grammars. By internalizing automata ---
-        the operational view of a parser --- we may extract executable parsers
-  \item A language implemented shallowly in Agda
+    linear/non-linear (LNL) type theory to syntactically capture the above categorical structures
+  \item An implementation of the language, which is emedded shallowly in Agda
 \end{itemize}
@@
 \section{Categorification of Formal Grammars}
+\label{sec:categorify}
 \steven{Idk if this section has the right title, but I wanted to change it if we are adding ``chomsky categories''}
@@
 operations which results in the Kleene star being more easily
-axiomatized \steven{TODO cite}
-% \cite{kozen1994}.
+axiomatized \cite{KOZEN1994366}.
 
@@
 
-\begin{lemma}[TODO cite
-  % \cite{day1970}
-  ]
+\begin{lemma}[\cite{day1970construction}]
   Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
@@
 leads us to the abstract notion of model we are interested in
-capturing with our type theory.
+capturing with our type theory: a \emph{Chomsky category}.
 
 \begin{definition}
-  A \emph{Chomsky category} is a locally Cartesian category with two hierarchies of
+  A Chomsky category is a locally Cartesian category with two hierarchies of
   universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
@@
 \begin{theorem}
-  The presheaf category $\Set^{\cat{\Sigma^*}}$ is a model.
+  The presheaf category $\Set^{\cat{\Sigma^*}}$ is a Chomsky category.
 \end{theorem}
+
+Further, the syntactic category of the theory in \cref{sec:tt} is manifestly a Chomsky category.
+\steven{This is true, right? That's like the entire point of having the syntax}
 
@@
 
-In 1995, Benton et al.
-\cite{bentonMixedLinearNonlinear1995} first gave categorical models of a logic
-that synthesized linear and non-linear logic. Substructural logics, like linear and separation logic, offer precise control
-over how values from a type may be used, in particular how many times a value
-may be used in a derivation. This control enables \emph{resource-sensitive}
-reasoning, where we may treat a resource as \emph{consumed} after usage. In the
-context of parsing, we may treat the characters composing a string as finite resources
-that are consumed at parse-time.
-
-Building off of Benton et al., Krishnaswami et al.
-\cite{krishnaswami_integrating_2015} developed a type
-system that unifies both linear and
-dependent types --- referred to as a \emph{linear-non-linear} (LNL) type theory. In
-such settings, there are two sub-theories --- one linear and one non-linear. The
-linear sub-theory embeds ordinary linear types without dependence, while the
-non-linear sub-theory embeds standard Martin-L\"of type theory. The interplay of
-linear types with dependence lives at the boundary between these two fragments.
-Benton et al.\ gives an accounting of how we may jump between these
-fragments via a monoidal adjunction $F \dashv G$ between linear terms and
-non-linear terms.
-
-\steven{Next paragraph is paper scaffolding for what I want to say and is total word salad rn}
-In this paper, we instantiate an LNL theory to axiomatize the notion of a Chomsky category, and because $\Set^{\String}$ forms a Chomsky category, using the type theory corresponds to interesting formal grammar proofs.
-Internalization of previous results + use to build correct-by-construction parsers.
+Omission of the structural inference rules of a deductive system, such as in
+linear logic \cite{GIRARD19871}, offers precise control over the number of times
+a value may be used in a derivation. This control enables
+\emph{resource-sensitive} reasoning, where we may treat a resource as
+\emph{consumed} after usage. This viewpoint is amenable to parsing applications,
+as we may treat characters of a string finite resources that are consumed at
+parse-time. One may then envision a linear type system where the types comprise
+formal grammars generated over some alphabet $\Sigma$, and the type constructors
+correspond precisely to inductive constructions on grammars --- such as
+conjunction, disjunction, and concatenation.
+
+Unfortunately, programming in a linear term language is often cumbersome and
+unintuitive for users. Code in such a language can become unnecessarily verbose
+when using Girard encodings to translate high-level reasoning into low-level
+linear terms. To alleviate this concern, in 1995 Benton et al.\ gave categorical
+models of a logic that synthesized linear and non-linear logic
+\cite{bentonMixedLinearNonlinear1995} --- appropriately referred to as a
+\emph{linear-non-linear} (LNL) system. This work is simply typed, so the
+boundary between linear and non-linear subtheories is entirely characterized via
+a monoidal adjunction $F \dashv G$ between linear terms and non-linear terms.
+
+\steven{Maybe there is more to say here, or perhaps later, that the only thing
+  to gain from a non-dependent non-linear fragment are the things that are
+  definiable via $G$. Except that these are not interesting, as they are
+  necessarily empty}
+
+\steven{Are these paragraphs recounting too much history?}
+
+Inside of an LNL system, linearity may be thought of as an option that users can
+choose to deliberately invoke at deliberate points in their developments in an
+otherwise intuitionistic system. However, if we are wishing to treat parsers as
+linear terms over input strings, the non-linear fragment of an LNL theory does
+not really assist in the development of parsers. It is instead the case that
+parsers may benefit from a \emph{dependence} on non-linear terms.
+
+\steven{Mention sepration logic at some point. In particular the separating conjunction and magic wand}
+
+Krishnaswami et al.\ equipped LNL systems with a restricted sense of dependent
+types \cite{krishnaswami_integrating_2015}. When defining dependent types in
+this setting dependence on linear terms is disallowed, but either linear or
+non-linear terms are allowed to depend on non-linear terms.
+
+The non-linear
+subtheory corresponds to Martin-L\"of type theory (MLTT), which can be taken
+as a foundation for mathematics. Dependence of linear terms on non-linear terms
+of this form effectively allows us to bring any ordinary definitions from math
+into scope as indicies and then define grammars relative to those indices. That
+is all to say, the LNL dependent types in the style of Krishnaswami et al.\
+allows us to recover the defintion of Aho's indexed grammars \cite{AhoIndexed} internal to our
+type theory.
+\steven{The above paragraph should maybe be put later. Also flows weirdly}
+
+In this paper, we instantiate an LNL theory with dependent types as a deductive
+setting to conduct parsing and formal grammar theory. In this system, linear
+types encode formal grammars built over an alphabet/(set of base types)
+$\Sigma$. The non-linear types are just MLTT/math defs. The syntax is in the
+following subsections
+
+\steven{Rephrase the above paragraph to summarize the intro of this section, as
+  well as preview the syntax that's to come in the following subsections.}
+
+\steven{Everything below here hasn't really been touched}
+
+\subsection{Syntax}
+
+\steven{Next paragraph is paper scaffolding for what I want to say and is total
+  word salad rn} In this paper, we instantiate an LNL theory to axiomatize the
+notion of a Chomsky category, and because $\Set^{\String}$ forms a Chomsky
+category, using the type theory corresponds to interesting formal grammar
+proofs. Internalization of previous results + use to build
+correct-by-construction parsers.
 
@@
 
-The structrual judgements are shown in \Cref{fig:structjdg}, the typing well-formedness in \Cref{fig:typewf}, and the intuitionistic typing rules in \Cref{fig:inttyping}.
-There are all just as they appear in \cite{krishnaswami_integrating_2015}.
-
-Our contribution is the addition of base types drawn from the alphabet $\Sigma$ along with the omission of the exchange rule for linear typing judgements.
+The structrual judgements are shown in \cref{fig:structjdg}, the typing
+well-formedness in \cref{fig:typewf}, and the intuitionistic typing rules in
+\cref{fig:inttyping}. There are all just as they appear in
+\cite{krishnaswami_integrating_2015}.
+
+Our contribution is the addition of base types drawn from the alphabet $\Sigma$
+along with the omission of the exchange rule for linear typing judgements.
 
@@
   \begin{mathpar}
+    \boxed{\Gamma \vdash X : U_{i}}
+
     \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
@@
     \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
-%
-    \and
 
@@
 
-    \\
-%
     \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
 %
-    \and
+    \\
 %
@@
 
+\section{Semantics}
+\label{sec:semantics}
 \subsection{Semantics in Grammars}
@@
 \section{Automata as Grammars}
+\label{sec:automata}
 Classically, formal language theory is closely related to the study of automata.
@@
 \section{Related work}
+\label{sec:related}
 

### 2024-07-09 POPL/paper.tex
syntax section minus a couple cases

@@
 
-Omission of the structural inference rules of a deductive system, such as in
-linear logic \cite{GIRARD19871}, offers precise control over the number of times
-a value may be used in a derivation. This control enables
+Omission of the structural rules of a deductive system, such as in
+linear logic \cite{GIRARD19871}, offers precise control over how a value is used
+in a derivation. Namely, linear logic omits the weakening and contraction rules
+to ensure that every value in context exactly once. This control enables
 \emph{resource-sensitive} reasoning, where we may treat a resource as
@@
 as we may treat characters of a string finite resources that are consumed at
-parse-time. One may then envision a linear type system where the types comprise
+parse-time. That is, once a particular substring has been parsed, a parser need not go
+over it again. One may then envision a linear type system where the types comprise
 formal grammars generated over some alphabet $\Sigma$, and the type constructors
 correspond precisely to inductive constructions on grammars --- such as
-conjunction, disjunction, and concatenation.
+conjunction, disjunction, concatenation, etc.
 
@@
 \steven{Maybe there is more to say here, or perhaps later, that the only thing
-  to gain from a non-dependent non-linear fragment are the things that are
+  to gain parsing-wise from a non-dependent non-linear fragment are the things that are
   definiable via $G$. Except that these are not interesting, as they are
@@
 
-In this paper, we instantiate an LNL theory with dependent types as a deductive
-setting to conduct parsing and formal grammar theory. In this system, linear
-types encode formal grammars built over an alphabet/(set of base types)
-$\Sigma$. The non-linear types are just MLTT/math defs. The syntax is in the
-following subsections
-
-\steven{Rephrase the above paragraph to summarize the intro of this section, as
-  well as preview the syntax that's to come in the following subsections.}
-
-\steven{Everything below here hasn't really been touched}
-
 \subsection{Syntax}
-
-\steven{Next paragraph is paper scaffolding for what I want to say and is total
-  word salad rn} In this paper, we instantiate an LNL theory to axiomatize the
-notion of a Chomsky category, and because $\Set^{\String}$ forms a Chomsky
-category, using the type theory corresponds to interesting formal grammar
-proofs. Internalization of previous results + use to build
-correct-by-construction parsers.
-
-\steven{Expand here on how linearity has an intuitive parsing intrerpration, however working only in a simply typed system is not sufficient.
-i.e.
-dependence offers the right view on indexed grammars.
-}
+\label{subsec:syntax}
+
+We instantiate an LNL theory with dependent types as a deductive
+setting to axiomatize the structure of a Chomsky category.
 
@@
 well-formedness in \cref{fig:typewf}, and the intuitionistic typing rules in
-\cref{fig:inttyping}. There are all just as they appear in
+\cref{fig:inttyping}. These are all just as they appear in
 \cite{krishnaswami_integrating_2015}.
-
-Our contribution is the addition of base types drawn from the alphabet $\Sigma$
-along with the omission of the exchange rule for linear typing judgements.
-
-\steven{TODO need to update syntax in these rules}
 
@@
 \end{figure}
-\subsection{Syntax}
-\label{subsec:syntax}
-
-\paragraph{Non-linear Types}
-
-The non-linear fragment of the theory
-corresponds to usual Martin-L\"of type theory. That is, the non-linear
-contexts obey all of the usual structural rules --- contraction, weakening, and
-exchange --- so we may use their contents as many times as we'd like.
-For the application of parsing, we think of these as primarily as indicies terms for indexed grammars.
-That is, they are as useful for their dependence.
-We may think of them as brining into scope definitions from ordinary mathematics to be used for indexing
-
-\paragraph{Linear Types}
-
-The linear typing judgment in our syntax takes on the following schematic form $\Gamma ; \Delta \vdash a : A$. First, $A$ represents
-a \emph{linear type} in our syntax. The intended semantics of these linear types
-are formal grammars. That is, the linear typing system is designed to syntactically
-reflect the behavior of formal grammars. For this reason, we may often
-interchangeably use the terms ``linear type'' and ``grammar''.
-
-The term $a$ is an inhabitant of type $A$, which we want to think of as a parse tree of
+
+\subsection{Formal Grammars as Linear Types}
+The linear typing judgment in our syntax takes on the following schematic form
+$\Gamma ; \Delta \vdash a : A$. First, $A$ represents a \emph{linear type} in
+our syntax. The intended semantics of these linear types are formal grammars.
+That is, the linear typing system is designed to syntactically reflect the
+behavior of formal grammars. For this reason, we may often interchangeably use
+the terms ``linear type'' and ``grammar''.
+
+The term $a$ is an inhabitant of type $A$, which is thought of as a parse tree of
 the grammar $A$. The core idea of this entire paper follows precisely from this
@@
 
-$\Gamma$ reprsents a non-linear context, while $\Delta$ represents a \emph{linear context}. These linear
-contexts
-behave substructurally --- in particular they have no contraction, weakening, or exchange
-rule. This means that everything within $\Delta$ must be used \emph{exactly
-  once} and \emph{in order of occurrence}. Thus, we can think of the linear
-contexts as an ordered list of limited
-resources. Once a resource is consumed, we cannot make reference to it again.
-We may then
-think of the variables in a linear context akin to building blocks for
-constructing patterns over strings.
-
-In the realm of parsing, our substructural
-restrictions make it so that the linear contexts behave quite similarly to
-formal grammars. For instance, linear context extension mirrors concatenation of
-grammars which may be seen via the introduction rule for $\otimes$,
-
-\[
-  \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta , \Delta' \vdash e \otimes e' : A \otimes B}
-\]
-
-Just as two grammars may be concatenated, if we can build an $A$ with the
-contents of $\Delta$, and a $B$ with the contents of $\Delta'$, then we we may
-concatenate $\Delta$ and $\Delta'$ to build an inhabitant $A \otimes B$.
-
-This is perhaps better illustrated through a concrete example. Consider the
-regular expression $A = (a \oplus b) \otimes (b \oplus c)^{*}$ and the string
-$w = abc$. We then have the following derivation that $w$ matches $A$,
-
-\begin{equation}
-  \label{eq:samplederiv}
-  \inferrule
-  {
-    w_{1} : a \vdash \mathsf{inl}(w_{1}) : a \oplus b \\
-    \inferrule
-    {
-      w_{2} : b \vdash \mathsf{inl}(w_{2}) : b \oplus c \\
-      w_{3} : c \vdash \mathsf{inr}(w_{3}) : b \oplus c
-    }
-    {
-      w_{2} : b, w_{3} :c \vdash \mathsf{cons}(w_{2},w_{3}) : (b \oplus c)^{*}
-    }
-  }
-  {w_{1} : a, w_{2} : b, w_{3} : c \vdash \mathsf{inl}(w_{1}) \otimes \mathsf{cons}(w_{2}, w_{3}) : (a \oplus b) \otimes (b \oplus c)^{*} }
-\end{equation}
-
-When parsing $w$ into $A$ by hand, we perform the same process as reflected in
-the derivation \cref{eq:samplederiv}. That is, we split the string $w$ up into pieces, one of which matches
-$a \oplus b$, and and another that matches $(b \oplus c)^{*}$. When further
-parsing this Kleene star bit, we break it up into smaller pieces each of which
-matches the patterns $b \oplus c$ and then we combine them all into one parse of
-the Kleene star $(b \oplus c)^{*}$. Note that this exact process is echoed in
-the above proof tree.
-
-\subsubsection{Dependence of the Linear on the Non-Linear}
-
-We may now however ask, ``what are the $\Gamma$'s doing in the $\otimes$ introduction rule?''
-Each linear context $\Delta$ is dependent on a non-linear context $\Gamma$, so
-we may use variables drawn from $\Gamma$ when defining the types in $\Delta$.
-
-In the grammar setting, this dependence lets
-us use variables from regular dependent type theory to
-keep track of data without \emph{consuming any resources}, so to speak. That is,
-using a piece of a linear context consumes it, so we cannot linearly encode anything that
-needs to be referenced more than once.
-
-We will see a concrete instance of this dependence in \cref{subsubsec:pda}. In
-that case, we encode the traces of a pushdown automaton as a grammar. When doing so,
-we make use of our non-linear types to define a type of stacks. The fact that
-our encoding of stacks is non-linear reflects
-that a pushdown automaton may consume a character from the input string without
-affecting the state of the stack.
-
-\subsubsection{Other Type Constructors}
-\label{subsubsec:other}
-
-\paragraph{Brzozowksi Derivative}
-In the 1950s Janusz Brzozowksi developed a notion of derivative for formal
-languages \cite{brzozowskiDerivativesRegularExpressions1964}. The derivative of language $L$ by string $w$ is given as,
-
-\[
-  w^{-1}L = \{ w' \in \String : w \otimes w' \in L \}
-\]
-
-That is, the derivative of $L$ with respect to $w$ are precisely the strings
-that complete to an $L$-parse when prefixed by a $w$.
-
-Brzozowski initially gave an accounting of this operation for
-generalized regular expressions, but later work demonstrates that the same
-construction can be generalized to context free grammars \cite{mightParsingDerivativesFunctional2011}.
-
-The ``magic wand'' from linear logic takes on the role of Brzozowski's
-derivative in our syntax.
-
-\begin{equation}
-  \label{eq:linfun}
-\inferrule{\Gamma ; a : A, \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A \lto B}
-\end{equation}
-
-We read this as saying that $\Delta$ constructs a $A \lto B$-parse when
-prepending $\Delta$ with an $A$-parse completes to a $B$ parse. In this sense,
-the type $A \lto B$ behaves like a linear function type. The only distinction
-between $A \lto B$ and an intuitionistic function type is that linear functions
-consume their input.
-
-Note that the introduction rule presented in \cref{eq:linfun} does indeed
-generalize Brzozowski's initial presentation of the derivative. First, to match
-his initial presentation
-let $A$ be the grammar accepting a single string $w$, which we may write as the
-discrete grammar $\lceil w \rceil$.
-Additionally, our construction lifts from a language-theoretic construction
-to a grammar point-of-view.
-In this sense, we carry around not just the data of what strings belong to
-$\lceil w \rceil \lto B$ but also are endowed the data of a function that
-constructs parse trees of $B$ from pairs of $\lceil w \rceil$-parses and
-$(\lceil w \rceil \lto B)$-parses.
-
-\paragraph{Top and Bottom}
-We include linear versions of the unit and empty types. The
-grammar $\top$ matches every input string, while the grammar
-$\bot$ represents falsity and matches no strings. These two
-are often used together to give syntactic shorthand to
-grammars that encode propositions.
-
-For instance, we may encode the proposition
-$\mathsf{isVowel}$ as,
-
-\[
-  \mathsf{isVowel}(w) :=
-    \begin{cases}
-      \top, \quad w \in \{a, e, i, o, u\} \\
-      \bot, \quad \text{otherwise}
-    \end{cases}
-\]
-
-\paragraph{Fixed Point Operator}
-We include a least fixed point operator on types, $\mu$, which is the means by
-which we define recursive grammars.
-
-Observe that we need not take the Kleene star as a primitive
-grammar constructor, as it is definable as a fixed point.
-The Kleene star of a grammar $g$ is given as,
-
-\[
-  g^{*} := \mu X . I \oplus (g \otimes X)
-\]
-
-\begin{figure}[h!]
-\begin{mathpar}
-  \inferrule
-  {\Gamma ; \Delta \vdash p : I}
-  {\Gamma ; \Delta \vdash \mathsf{nil} : g^{*}}
-
-  \and
-
-  \inferrule
-  {\Gamma ; \Delta \vdash p : g \\ \Gamma ; \Delta' \vdash q
-  : g^{*}}
-  {\Gamma ; \Delta \vdash \mathsf{cons}(p , q) : g^{*}}
-
-  \and
-
-  \inferrule
-  {
-    \Gamma ; \Delta \vdash p : g^{*} \\
-    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
-    \Gamma ; x : g , y : h \vdash p_{\ast} : h
-  }
-  {\Gamma ; \Delta \vdash \mathsf{foldr}(p_{\varepsilon} , p_{\ast}) : g^{*}}
-\end{mathpar}
-\caption{Kleene Star Rules}
-\label{fig:star}
-\end{figure}
-
-Likewise, $g^{*}$ has admissible introduction and
-elimination rules, shown in \cref{fig:star}. Note that this
-definition of $g^{*}$ and these
-rules arbitrarily assigns a handedness to the Kleene star.
-We could have just as well took it to be a fixed point of
-$I \oplus (X \otimes g)$. In fact, the definitions are
-equivalent, as the existence of the $\mathsf{foldl}$ term below
-shows that $g^{*}$ is also a fixed point of
-$I \oplus (X \otimes g)$.
-
-\begin{equation}
-  \label{eq:foldl}
-  \inferrule
-  {
-    \Gamma ; \Delta \vdash p : g^{*} \\
-    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
-    \Gamma ; y : h , x : h \vdash p_{\ast} : h
-  }
-  {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
-\end{equation}
-
-In fact, the $\mathsf{foldl}$ term is defined using
-$\mathsf{foldr}$ --- much in the same way one
-may define a left fold over lists in terms of a right fold
-in a functional programming language\footnote{The
-  underlying trick is to fold over a list of functions
-  instead of the original string. We curry each character $c$
-  of the string into a function that concatenates $c$, and
-  right fold over this list of linear functions. Because function
-  application is left-associative, this results in a left
-  fold over the original string. }.
-
-We only take fixed points of a single variable as a
-primitive operation in the type theory, but we may apply
-Beki\`c's theorem \cite{Bekić1984} to define an admissible
-notion of multivariate fixed point. This is particularly
-useful for defining grammars that encode the states of an
-automaton. In \cref{fig:multifix} we provide the
-introduction and elimination principles for such a fixed
-point, where $\sigma$ is the substitution that unrolls the
-mutually recursive definitions one level. That is,
-
-\begin{align*}
-  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \\
-  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
-\end{align*}
-
-\begin{figure}[h!]
-\begin{mathpar}
-  \inferrule
-  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
-  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
-
-  \\
-
-  \inferrule
-  {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
-             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
-             \dots \\
-             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
-  }
-  {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
-\end{mathpar}
-\caption{Multi-fixed Points}
-\label{fig:multifix}
-\end{figure}
-
-The most prevalent application of these multi-fixed points
-is for defining the traces through automata, as we will see
-later in \cref{eq:nfatrace}.
-
-% \todoin{with citation, talk somewhere about how CFGs are just regular + fixed points}
-\max{this is a good point to give an overview of how regular expressions and context-free grammars/expressions are essentially sub-languages of our type system.}
-
-\section{Semantics}
-\label{sec:semantics}
-\subsection{Semantics in Grammars}
-\label{subsec:seming}
-As mentioned earlier, the interpretation of grammars as functions from
-$\String$ to $\Set$ serves as our intended semantics. When designing the syntax of this formalism, we were often
-informed by this model. That is, there is an interpretation
-$\llbracket \cdot \rrbracket$ of our syntax that provides grammar-based
-semantics.
+$\Gamma$ reprsents a non-linear context, while $\Delta$ represents a
+\emph{linear context} dependent on $\Gamma$. These linear contexts behave
+substructurally. As stated earlier, they do are linear --- they do not obey
+weakening or contraction --- because a character is exhausted once read by a
+parsing procedure. Moreover, the characters in strings appear in the order in
+which we read them. We do not have the freedom to freely permute characters,
+therefore any type theory that is used to reason about formal grammars ought to
+omit the structural rule of exchange as well. This means that every variable
+within $\Delta$ must be used \emph{exactly once} and \emph{in order of
+  occurrence}. Thus, we can think of the linear contexts as an ordered list of
+limited resources. Once a resource is consumed, we cannot make reference to it
+again. Variables in a linear context then act like building
+blocks for constructing patterns over strings.
+
+We give the base types and type constructors for linear terms. As the
+interpretation of types as grammars in $\Set^{\String}$ serves as our intended
+semantics, we simultaneously give the interpreation $\sem \cdot$ of the
+semantics as grammars.
+
+Most of the non-linear semantics is standard, so let us investigate in more
+depth the linear semantics. The interpretation of each
+connective is exactly what is suggested by the syntax, in
+large part to the inspiration taken from this interpretation
+when designing the type theory.
 
@@
 
-Most of the non-linear semantics is standard, so let us investigate in more
-depth the linear semantics. The interpretation of each
-connective is exactly what is suggested by the syntax, in
-large part to the inspiration taken from this interpretation
-when designing the type theory.
-
-\begin{enumerate}
-  % \item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
-  \item $\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-        \begin{itemize}
-          \item The semantics of the linear $\Pi$-type is
-                indeed a dependent function out of the
-                semantics of $X$.
-        \end{itemize}
-  \item $\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-        \begin{itemize}
-          \item The semantics of the linear $\Sigma$-type is
-                a dependent pair.
-        \end{itemize}
-  \item $\sem{I} \gamma w = \{ () \pipe w = \epsilon \}$
-        \begin{itemize}
-          \item The only string accepted by $I$ is the empty string.
-        \end{itemize}
-  \item
-        $\sem{A \otimes B} \gamma w = \Sigma(w1,w2:\Sigma^*) \left( (w_{1}w_{2} = w) \wedge \sem{A} \gamma w_1 \times \sem{B} \gamma w_2 \right)$
-        \begin{itemize}
-          \item The string $w$ matches $A \otimes B$
+\paragraph{Linear Unit}
+First, the linear unit $I$ may be built in the empty linear
+context\footnote{When appropriate, statements like ``the empty linear context''
+  may be taken polymorphically over all non-linear contexts that they might
+  depend on.} $\cdot \dashv I$. $I$ serves as the unit for the operation
+$\otimes$ described below.
+
+As a grammar,
+
+\[
+  \sem {I} \gamma w = \{ () ~|~ w = \varepsilon \}
+\]
+
+That is, $I$ maps strings to the proposition that they are the empty string. $I$
+is clearly only inhabited by the empty string $\varepsilon$, for which the
+outputted set contains a single parse tree.
+
+\paragraph{Base Types}
+For each character $c$ in the alphabet $\Sigma$, we include a base type at the
+lowest universe level $c : L_{0}$.
+
+The grammar interpretation for characters is quite similar to that of $I$.
+\[
+  \sem {c} \gamma w = \{ () ~|~ w = c \}
+\]
+
+The grammar for the character $c$ likewise maps strings $w$ to the proposition that
+they are equal to $c$.
+
+\paragraph{Tensor Product}
+The tensor product of two linear types is the first place where the ordering on
+contexts really takes effect. That is, the tensor product of two types is
+non-commutative.
+When context $\Delta$ forms a linear term of
+type $A$, and $\Delta'$ forms a linear term of type $B$, then the context
+extension $\Delta, \Delta'$ forms a linear term of type $A \otimes B$.
+
+As stated above, the type $I$ serves as the unit for $\otimes$. That is, for all
+linear types $A$, the following equalities hold:
+
+\[ I \otimes A \equiv A \otimes I \equiv A \]
+
+In the grammar semantics,
+
+\[
+  \sem {A \otimes B} = \Sigma_{w_{1}, w_{2} : \String} (w_{1}w_{2} = w) \land \left(  \sem {A} \gamma w_{1} \times \sem {B} \gamma w_{2} \right)
+\]
+
+The string $w$ matches $A \otimes B$
 precisely when it may be split into two
@@
 and the right one matches $B$.
-        \end{itemize}
-  \item
-        $\sem{A \lto B} \gamma w = \Pi(w_a:\Sigma^*) \left( A \gamma w_a \Rightarrow B\gamma (w_aw) \right)$
-        \begin{itemize}
-          \item The semantics of the linear function type is
-                a dependent function that turns $A$-parses
-                of a substring into $B$-parses of an
-                extended string.
-        \end{itemize}
-  \item
-        $\sem{B \tol A} \gamma w = \Pi(w_a:\Sigma^*) \left( A \gamma w_a \Rightarrow B\gamma (ww_a) \right)$
-  \item $\sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x)$
-        \begin{itemize}
-          \item The semantics of the least fixed point
-                operator is indeed the least fixed point of sets.
-        \end{itemize}
-\end{enumerate}
-
-\subsection{Canonicity}
+
+\paragraph{Linear Function Types}
+The monoidal structure provided by $\otimes$ is both left and right closed and
+this is denoted by the left and right linear function types.
+
+\steven{Say more about the adjunctions}
+
+Further, for linear types $A$ and $B$ the left linear function type $A \lto B$
+enjoys an elimination principle similar to function application/modus ponens.
+
+\[
+\inferrule{\Delta \vdash A \otimes A \lto B}{\Delta \vdash B}
+\]
+
+The interpretation of linear function types at a string $w$ is a linear function
+that takes in parses of $A$ on a string $w_{a}$ and outputs parses of $B$ on the
+string $w_{a} w$.
+
+\[
+  \sem{A \lto B} \gamma w = \Pi(w_a:\String) \left( A \gamma w_a \Rightarrow B\gamma (w_a w) \right)
+\]
+
+That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
+complete to parses of $B$. In this manner, the linear function types generalize
+Brzozowksi's notion of derivative
+\cite{brzozowskiDerivativesRegularExpressions1964}.
+Brzozowski initially only gave an accounting of this operation for
+generalized regular expressions, but later work from Might et al.\ demonstrates that the same
+construction can be generalized to context free grammars
+\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
+types the same notion of derivative extends to the grammars of our type theory.
+
+Of course, all the above statements for the left function type also have
+corresponding analogues for the
+right-handed counterpart.
+
+\paragraph{LNL Dependent Types}
+Given a non-linear type $X$, we may form both the dependent product type
+$\LinPiTy x X A$ and the dependent pair type $\LinSigTy x X A$ as linear types
+where $x$ is free in $A$.
+
+$\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+
+The grammar semantics of the linear product type is
+indeed a dependent function out of the
+semantics of $X$.
+
+$\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+
+Likewise, the grammar semantics of the linear dependent pair type is a dependent
+pair in $\Set$.
+
+\steven{TODO example of sum as an admissible usage of $\LinSigTy b \Bool A$}
+
+\paragraph{Universal Type}
+The universal type $\top$ may be formed in any context.
+
+\[ \sem{\top} \gamma w = \{ \ast \}\]
+
+Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.
+
+\paragraph{Empty Type}
+The empty type $\bot$ has no inhabitants.
+
+\[ \sem {\bot} \gamma w = \emptyset \]
+\steven{TODO need to include a case for $\bot$ in the inference rules}
+
+\paragraph{Pair Type}
+Given a term of type $A$ in context $\Delta$ and a term of type $B$ \emph{also
+  in context} $\Delta$, we may form a pair in context $\Delta$ of type
+$A \ampersand B$.
+
+\[ \sem{A \amp B }\gamma w = \sem{A} \gamma w \times \sem {B} \gamma w \]
+
+The grammar semantics of $A \amp B$ denotes a pair of a parse of $A$ and a
+parse of $B$ \emph{of the same string}. In this sense, $\amp$ acts a
+conjunction between grammars --- meaning a string only matches $A \amp B$
+if it simlutaneously matches each of $A$ and $B$.
+
+\paragraph{The $G^{-1}$ Modality}
+Turn nonlinear into linear
+
+\steven{Call this $F$ or $G^{-1}$?}
+
+\steven{Why can't these map any nonlinear type into linear ones? instead of just
+those that are already linear types lifted by $G$?}
+
+\steven{TODO G inverse}
+  % \item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
+
+\paragraph{Recursive Types}
+isorecursive types
+\steven{TODO recursive types}
+  % \item $\sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x)$
+
+Observe that we need not take the Kleene star as a primitive
+grammar constructor, as it is definable as a fixed point.
+The Kleene star of a grammar $g$ is given as,
+
+\[
+  g^{*} := \mu X . I \oplus (g \otimes X)
+\]
+
+\begin{figure}[h!]
+\begin{mathpar}
+  \inferrule
+  {\Gamma ; \Delta \vdash p : I}
+  {\Gamma ; \Delta \vdash \mathsf{nil} : g^{*}}
+
+  \and
+
+  \inferrule
+  {\Gamma ; \Delta \vdash p : g \\ \Gamma ; \Delta' \vdash q
+  : g^{*}}
+  {\Gamma ; \Delta \vdash \mathsf{cons}(p , q) : g^{*}}
+
+  \and
+
+  \inferrule
+  {
+    \Gamma ; \Delta \vdash p : g^{*} \\
+    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
+    \Gamma ; x : g , y : h \vdash p_{\ast} : h
+  }
+  {\Gamma ; \Delta \vdash \mathsf{foldr}(p_{\varepsilon} , p_{\ast}) : g^{*}}
+\end{mathpar}
+\caption{Kleene Star Rules}
+\label{fig:star}
+\end{figure}
+
+Likewise, $g^{*}$ has admissible introduction and
+elimination rules, shown in \cref{fig:star}. Note that this
+definition of $g^{*}$ and these
+rules arbitrarily assigns a handedness to the Kleene star.
+We could have just as well took it to be a fixed point of
+$I \oplus (X \otimes g)$. In fact, the definitions are
+equivalent, as the existence of the $\mathsf{foldl}$ term below
+shows that $g^{*}$ is also a fixed point of
+$I \oplus (X \otimes g)$.
+
+\begin{equation}
+  \label{eq:foldl}
+  \inferrule
+  {
+    \Gamma ; \Delta \vdash p : g^{*} \\
+    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
+    \Gamma ; y : h , x : h \vdash p_{\ast} : h
+  }
+  {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
+\end{equation}
+
+In fact, the $\mathsf{foldl}$ term is defined using
+$\mathsf{foldr}$ --- much in the same way one
+may define a left fold over lists in terms of a right fold
+in a functional programming language\footnote{The
+  underlying trick is to fold over a list of functions
+  instead of the original string. We curry each character $c$
+  of the string into a function that concatenates $c$, and
+  right fold over this list of linear functions. Because function
+  application is left-associative, this results in a left
+  fold over the original string. }.
+
+We only take fixed points of a single variable as a
+primitive operation in the type theory, but we may apply
+Beki\`c's theorem \cite{Bekić1984} to define an admissible
+notion of multivariate fixed point. This is particularly
+useful for defining grammars that encode the states of an
+automaton. In \cref{fig:multifix} we provide the
+introduction and elimination principles for such a fixed
+point, where $\sigma$ is the substitution that unrolls the
+mutually recursive definitions one level. That is,
+
+\begin{align*}
+  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \\
+  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
+\end{align*}
+
+\begin{figure}[h!]
+\begin{mathpar}
+  \inferrule
+  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
+  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
+
+  \\
+
+  \inferrule
+  {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
+             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
+             \dots \\
+             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
+  }
+  {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
+\end{mathpar}
+\caption{Multi-fixed Points}
+\label{fig:multifix}
+\end{figure}
+
+The most prevalent application of these multi-fixed points
+is for defining the traces through automata, as we will see
+later in \cref{eq:nfatrace}.
+
+\subsection{Subtheories}
+\paragraph{Regular Expressions} Restricting to regular expressions with the
+obvious connectives forms a subtheory of our type theory. Perhaps there is
+something to say here about how this reduces to Kleene algebra performed inside
+of a Kleene cat?
+
+\steven{TODO regular expressions}
+
+\paragraph{$\mu$-Regular Expressions} Regex + fixed point equivalent to context
+free \cite{leiss}.
+
+\steven{TODO $\mu$ regular expressions}
+
+\steven{Every after here has not been touched}
+
+\section{Canonicity}
 
@@
 \end{enumerate}
-
-\subsection{Semantics in Languages}
-% Recall from \cref{sec:synindsyn} that every formal grammar induces a formal language. This corresponds to another
-interpretation $L(\cdot)$ of our syntax. In particular, the semantics of a
-linear type $A : L_{i}$
-from \cref{subsec:seming} are augmented such that
-
-\[
-  L(A) = \{ w \in \String : \sem A w \text{ is nonempty} \}
-\]
-
-\subsubsection{Weak Equivalence}
-Recall that an isomorphism of grammars defined a strong equivalence. We may
-additionally define a \emph{weak equivalence} of grammars when two grammars
-define the same language.
-
-That is, $A$ and $B$ are weakly equivalent when $L(A) = L(B)$. Note that this is
-a semantic notion, but this just as well captured syntactically by the existence
-of two judgments $\Gamma ; x : A \vdash M : B$ and
-$\Gamma ; x : B \vdash N : B$, which are not necessarily mutually inverse.
 

### 2024-07-09 POPL/paper.tex
spelling

@@
   that string. Finally, we discuss the application to the design and
-  implementation of correct-by-contruction parsers.
+  implementation of correct-by-construction parsers.
 \end{abstract}
@@
   make little comment as to which foundations are used to encode $\Set$. Our
-  construction is polymorphic in choicce of a proper class of small sets, a
+  construction is polymorphic in choice of a proper class of small sets, a
   Grothendieck universe, a type theoretic universe, or any similar foundation.} $G w$ of parse
@@
 $\Set^{\String}$. That is, they may all be interpreted in our category of grammars. In this sense, $\Set^{\String}$ provides a universal setting to reason about parsing.
-The slice 2-category over $\Set^{\String}$ then provides the appropriate expressivity to prove equivalences between grammar formalism, such as the equivalence between semi-Thue systems and Chomsky's unresticted grammars.
+The slice 2-category over $\Set^{\String}$ then provides the appropriate expressivity to prove equivalences between grammar formalism, such as the equivalence between semi-Thue systems and Chomsky's unrestricted grammars.
 
@@
     linear/non-linear (LNL) type theory to syntactically capture the above categorical structures
-  \item An implementation of the language, which is emedded shallowly in Agda
+  \item An implementation of the language, which is embedded shallowly in Agda
 \end{itemize}
@@
 \label{sec:categorify}
-\steven{Idk if this section has the right title, but I wanted to change it if we are adding ``chomsky categories''}
+\steven{Idk if this section has the right title, but I wanted to change it if we are adding ``Chomsky categories''}
 
@@
 are binary operations over $A$, $(-)^*$ is a function over $A$, and
-$1$ and $0$ are constants. These structures satisfy the axoims depicted
+$1$ and $0$ are constants. These structures satisfy the axioms depicted
 in Figure~\ref{fig:axioms}.
@@
   algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
-  X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogue isomorphism
+  X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogous isomorphism
   for $G_{A,B}$ also holds.
@@
   to gain parsing-wise from a non-dependent non-linear fragment are the things that are
-  definiable via $G$. Except that these are not interesting, as they are
+  definable via $G$. Except that these are not interesting, as they are
   necessarily empty}
@@
 
-\steven{Mention sepration logic at some point. In particular the separating conjunction and magic wand}
+\steven{Mention separation logic at some point. In particular the separating conjunction and magic wand}
 
@@
 of this form effectively allows us to bring any ordinary definitions from math
-into scope as indicies and then define grammars relative to those indices. That
+into scope as indices and then define grammars relative to those indices. That
 is all to say, the LNL dependent types in the style of Krishnaswami et al.\
-allows us to recover the defintion of Aho's indexed grammars \cite{AhoIndexed} internal to our
+allows us to recover the definition of Aho's indexed grammars \cite{AhoIndexed} internal to our
 type theory.
@@
 
-The structrual judgements are shown in \cref{fig:structjdg}, the typing
+The structural judgments are shown in \cref{fig:structjdg}, the typing
 well-formedness in \cref{fig:typewf}, and the intuitionistic typing rules in
@@
 
-$\Gamma$ reprsents a non-linear context, while $\Delta$ represents a
+$\Gamma$ represents a non-linear context, while $\Delta$ represents a
 \emph{linear context} dependent on $\Gamma$. These linear contexts behave
@@
 interpretation of types as grammars in $\Set^{\String}$ serves as our intended
-semantics, we simultaneously give the interpreation $\sem \cdot$ of the
+semantics, we simultaneously give the interpretation $\sem \cdot$ of the
 semantics as grammars.
@@
 conjunction between grammars --- meaning a string only matches $A \amp B$
-if it simlutaneously matches each of $A$ and $B$.
+if it simultaneously matches each of $A$ and $B$.
 

### 2024-07-09 POPL/paper.tex
single column
Added comments

@@
 % -*- fill-column: 80; -*-
-\documentclass[sigconf,anonymous,review,screen]{acmart}
+\documentclass[sigconf,anonymous,review,screen,acmsmall]{acmart}
 \usepackage{mathpartir}
@@
 
-\section{A Syntax-Independent Notion of Syntax}
+\section{Introduction}
 \label{sec:intro}
+
 The theory of formal languages and parsing is one of the oldest and
@@
 generation of efficient parsers \cite{Johnsonyacc}.
+
+\paragraph{Contributions.} This paper begins in \cref{sec:categorify} by giving a categorical interpretation of Kleene algebra.
+However, Kleene algebras are primarily concerned with the regular languages.
+With this in mind, we further abstract our categorical model from the connectives of regular grammars to those of context-free grammars and beyond.
+From there, in \cref{sec:tt} we provide a type theory to construct and reason about such formal grammars.
+In \cref{sec:semantics} we give interpretations of the syntax as grammars and as languages, and prove a canonicity theorem for the type theory.
+Finally, in \cref{sec:automata} we give a presentation of automata as grammars.
+With this view in mind, we may encode parsing algorithms and execute them entirely internal to our formalism.
+Our contributions are then:
+
+\pedro{Something good is coming up with a name to the type theory... This makes it easier to
+refer to it throughout the paper. Spitballing some suggestions GTT (grammar type theory --- conflicts with Max's Gradual Type Theory), KTT (Kleene type theory)...}
+
+\begin{itemize}
+  \item A universal semantic notion of formal grammar given by the category $\Set^{\String}$
+  \item Categorical structures that capture the concepts integral to formal grammar theory, and that are notably modeled by $\Set^{\String}$
+  \item A dependent
+    linear/non-linear (LNL) type theory to syntactically capture the above categorical structures
+  \item An implementation of the language, which is embedded shallowly in Agda
+\end{itemize}
+
+\section{A Syntax-Independent Notion of Syntax}
 
@@
 for every string $w \in \String$. Intuitively, we read a morphism of grammars as
-a \emph{parse transformer} : $f$ translates a $G$-parse of $w$ into an $H$-parse
+a \emph{parse transformer}: $f$ translates a $G$-parse of $w$ into an $H$-parse
 of $w$. The notion of isomorphism induced by
@@
 grammars \cite{chom1963}.
+
+\pedro{Written like this, it seems like this correspondence iso-strong
+  equivalence is a well-known fact, but this is not an obvious
+  observation.}
 
@@
 The slice 2-category over $\Set^{\String}$ then provides the appropriate expressivity to prove equivalences between grammar formalism, such as the equivalence between semi-Thue systems and Chomsky's unrestricted grammars.
+\pedro{why?}
 
@@
 
-This paper begins in \cref{sec:categorify} by giving a categorical interpretation of Kleene algebra.
-However, Kleene algebras are primarily concerned with the regular languages.
-With this in mind, we further abstract our categorical model from the connectives of regular grammars to those of context-free grammars and beyond.
-From there, in \cref{sec:tt} we provide a type theory to construct and reason about such formal grammars.
-In \cref{sec:semantics} we give interpretations of the syntax as grammars and as languages, and prove a canonicity theorem for the type theory.
-Finally, in \cref{sec:automata} we give a presentation of automata as grammars.
-With this view in mind, we may encode parsing algorithms and execute them entirely internal to our formalism.
-Our contributions are then:
-
-\begin{itemize}
-  \item A universal semantic notion of formal grammar given by the category $\Set^{\String}$
-  \item Categorical structures that capture the concepts integral to formal grammar theory, and that are notably modeled by $\Set^{\String}$
-  \item A dependent
-    linear/non-linear (LNL) type theory to syntactically capture the above categorical structures
-  \item An implementation of the language, which is embedded shallowly in Agda
-\end{itemize}
-
-\section{Categorification of Formal Grammars}
+\section{Categorifying Formal Grammars}
 \label{sec:categorify}
@@
 Kleene algebras are an important tool in the theory of regular
-languages. More broadly, they serve as a theoretical substrate to
-studying various kinds of formal languages. Formally, they are a tuple
-$(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is a set, $+$ and $\cdot$
-are binary operations over $A$, $(-)^*$ is a function over $A$, and
-$1$ and $0$ are constants. These structures satisfy the axioms depicted
-in Figure~\ref{fig:axioms}.
+languages. More broadly, through various extensions, they serve as a
+theoretical substrate to studying different kinds of formal languages.
+
+We posit that by seeing Kleene algebra through the lens of category
+theory, we gain some insight into the type-theoretic essence of formal
+grammars. We begin this section by providing a ``simply-typed''
+version of a categorification of Kleene algebra, which we call Kleene
+category, and then show define a more expressive class of categories,
+which we call Chomsky category, that will serve as the categorical
+semantics to our type theory.
+
+\subsection{Kleene categories}
+A Kleene algebra is a tuple $(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is
+a set, $+$ and $\cdot$ are binary operations over $A$, $(-)^*$ is a
+function over $A$, and $1$ and $0$ are constants. These structures
+satisfy the axioms depicted in Figure~\ref{fig:axioms}.
 
@@
 
-This result feels similar in spirit to the definition of action
-algebras, which are algebras where the product also has adjoint
-operations which results in the Kleene star being more easily
-axiomatized \cite{KOZEN1994366}.
-
-We are now ready to prove that our concept of formal grammars fits
-nicely within our categorical framework. We start by presenting a
-well-known construction from presheaf categories.
+Something surprising about this lemma is that it provides an alternative
+perspective on the observation that if a Kleene algebra has an
+residuation operation, also called action algebra \cite{kozen1994action},
+then the Kleene star admits a simpler axiomatization.
+
+Since we want Kleene categories to generalize our notion of formal
+grammars as presheaves $\String \to \Set$, we prove that they do
+indeed form a Kleene category. We start by presenting a well-known
+construction from presheaf categories.
 
@@
 algebras can reason about language containment, Kleene categories can
-reason about parsing as well.
+reason about parsing as well, as illustrated by the presheaf model
+above.
 
@@
 
-\paragraph{Beyond Simple Types}
+\subsection{Beyond Simple Types}
 
@@
 concepts that are outside of reach of Kleene algebras, their
-simply-typed nature makes them not so expressive from a type theory
-point of view. This is limiting because type theories are tested
-syntactic frameworks for manipulating complicated categories
+simply-typed nature makes them not so expressive from a type theoretic
+point of view. This is limiting because type theories are sucessful
+syntactic frameworks for manipulating complicated categorical
 structures while avoiding some issues common in category theory, such
@@
 
-With this in mind, we want to design a dependent type theory that
-builds on top of Kleene categories with the goal of internalizing
-various reasoning principles for languages and their parsers. This
-leads us to the abstract notion of model we are interested in
-capturing with our type theory: a \emph{Chomsky category}.
-
+With this in mind, we want to design a categorical semantics that
+builds on top of Kleene categories with the goal of extending them
+with dependent types and making them capable reasoning about languages
+and their parsers. This leads us to the abstract notion of model we
+are interested in capturing with our type theory: a \emph{Chomsky
+category}.
+
+\pa{We should probably define some of the words in this definition}
 \begin{definition}
@@
 \steven{This is true, right? That's like the entire point of having the syntax}
+\pedro{This is likely true, but if we explicitly say so, this warants a proof. I think that
+if we don't say anything about the syntactic category, reviewers won't mind.}
 
@@
 in a derivation. Namely, linear logic omits the weakening and contraction rules
-to ensure that every value in context exactly once. This control enables
+to ensure that every value in context is used exactly once. This control enables
 \emph{resource-sensitive} reasoning, where we may treat a resource as
 \emph{consumed} after usage. This viewpoint is amenable to parsing applications,
-as we may treat characters of a string finite resources that are consumed at
+as we may treat characters of a string as finite resources that are consumed at
 parse-time. That is, once a particular substring has been parsed, a parser need not go
@@
 
-Unfortunately, programming in a linear term language is often cumbersome and
-unintuitive for users. Code in such a language can become unnecessarily verbose
-when using Girard encodings to translate high-level reasoning into low-level
-linear terms. To alleviate this concern, in 1995 Benton et al.\ gave categorical
-models of a logic that synthesized linear and non-linear logic
-\cite{bentonMixedLinearNonlinear1995} --- appropriately referred to as a
-\emph{linear-non-linear} (LNL) system. This work is simply typed, so the
-boundary between linear and non-linear subtheories is entirely characterized via
-a monoidal adjunction $F \dashv G$ between linear terms and non-linear terms.
+Unfortunately, programming in a linear term language is often
+cumbersome and unintuitive for users. Code in such a language can
+become unnecessarily verbose when using Girard encodings to translate
+high-level reasoning into low-level linear terms. To alleviate this
+concern, in 1995 Benton et al.\ proposed an alternative categorical
+semantics of linear logic based on adjoint interactions between linear
+and non-linear logics \cite{bentonMixedLinearNonlinear1995} ---
+appropriately referred to as a \emph{linear-non-linear} (LNL)
+system. This work is simply typed, so the boundary between linear and
+non-linear subtheories is entirely characterized via a monoidal
+adjunction $F \dashv G$ between linear terms and non-linear terms.
 
@@
 parsers may benefit from a \emph{dependence} on non-linear terms.
+\pedro{This last sentence could use an example. Is there a bite-sized, well-known
+example we could mention here?}
 
@@
 types \cite{krishnaswami_integrating_2015}. When defining dependent types in
-this setting dependence on linear terms is disallowed, but either linear or
+this setting, dependence on linear terms is disallowed, but either linear or
 non-linear terms are allowed to depend on non-linear terms.
@@
 \steven{The above paragraph should maybe be put later. Also flows weirdly}
+
+\pedro{I would argue that both paragraphs above should either be put
+  somewhere else or deleted. I think that we should definitely mention
+  Neel's LNL system, but as it stands, it feels disconnected from what
+  came before and what comes next. A suggestion: the mention of Aho's
+  indexed grammars could be the example we use illustrating the
+  benefits of dependent grammars.}
 
@@
 
+\pedro{After reading this section after the Kleene category one, I
+  think that it makes much more sense to switch their order. It's
+  going to be easier to justify providing a categorical semantics once
+  we have a ``target'' of what we want to interpret.}
+
 \section{Canonicity}
-
-Let $\cdot \vdash A$ be a closed linear type. Then there are
+\pedro{It could be cool to come up with a catchy title to this section.
+Something like ``Parse Canonicity''}
+
+\pedro{Added the ``initial context'' paragraph below}
+
+Canonicity is an important metatheoretic theorem in the type theory
+literature.  It provides insight on the normal forms of terms and,
+therefore, on its computational aspects. Frequently, proving
+canonicity for boolean types, i.e. every closed term of type bool
+reduces to either true or false, is enough to justifying that the type
+theory being studied is well-behaved. In our case, however, since we
+want to connect our type theory to parsers, we must provide a more
+detailed account of canonicity.
+
+If $\cdot \vdash A$ is a closed linear type then there are
 two obvious notions of what constitutes a ``parse'' of a string w
@@
 \label{sec:automata}
+\pedro{The switch from the canonicity proof to this section feels a bit ``bumpy''.
+  Since they are at least technically related, it might be a good idea to say something
+  like ``The canonicity theorem from the previous section gives a tight connection
+  between terms in our type theory and the theory of formal languages. In this section
+we further explore these connections by using the fact that, clasically ...''}
+
 Classically, formal language theory is closely related to the study of automata.
@@
 recursively enumerable languages --- since we can simulate the
-tape of Turing machine in the dependent fragment of the language. We reflect
+tape of Turing machine in the dependent fragment of the language.We reflect
 here on this expressive power to motivate why we put stringent syntactic
@@
 automata --- regular expressions and finite automata.
+\pedro{The connection dependent types-Turing machines seems too important to only
+  be mentioned here. It should at least indirectly motivate the type theory being
+dependently typed.}
 
@@
 
+\pedro{I like the overall structure of this section! Showing how the type theory
+can encode and reason about different kinds of languages is very instructive!}
+
 \section{Related work}

### 2024-07-10 POPL/paper.tex
comments
addressing some of pedros concerns

@@
 \usepackage{fancyvrb}
+\usepackage{xspace}
 \usepackage{hyperref}
@@
 \newcommand{\Set}{\mathbf{Set}}
+\newcommand{\SemAct}{\mathbf{SemAct}}
 \newcommand{\Gr}{\mathbf{Gr}}
+\newcommand{\Grammar}{\mathbf{Gr}}
 \newcommand{\Type}{\mathbf{Type}}
@@
 \newcommand{\nat}{\mathbb{N}}
+\newcommand{\theoryname}{Grammar Type Theory\xspace}
+\newcommand{\theoryabbv}{GTT\xspace}
 
@@
 \begin{abstract}
-  In this paper, we propose a new syntactic formalism for formal grammars: a
+  In this paper we propose \theoryname~(\theoryabbv), a new syntactic formalism for formal grammars: a
   dependent linear/non-linear type theory in which the tensor product is
@@
 
-  Internally to our calculus, we carry out many classical constructions from
+  Internal to \theoryname, we carry out many classical constructions from
   the theory of formal grammars. Inspired by the common
@@
 
-  We give this type theory a semantics in the category of grammars and prove a
+  We give a semantics in the category of grammars and prove a
   canonicity theorem showing that every term in a context corresponding to a
@@
 
-\paragraph{Contributions.} This paper begins in \cref{sec:categorify} by giving a categorical interpretation of Kleene algebra.
-However, Kleene algebras are primarily concerned with the regular languages.
-With this in mind, we further abstract our categorical model from the connectives of regular grammars to those of context-free grammars and beyond.
-From there, in \cref{sec:tt} we provide a type theory to construct and reason about such formal grammars.
-In \cref{sec:semantics} we give interpretations of the syntax as grammars and as languages, and prove a canonicity theorem for the type theory.
-Finally, in \cref{sec:automata} we give a presentation of automata as grammars.
-With this view in mind, we may encode parsing algorithms and execute them entirely internal to our formalism.
-Our contributions are then:
-
-\pedro{Something good is coming up with a name to the type theory... This makes it easier to
-refer to it throughout the paper. Spitballing some suggestions GTT (grammar type theory --- conflicts with Max's Gradual Type Theory), KTT (Kleene type theory)...}
+\paragraph{Contributions.} This paper begins in \cref{sec:synindnotion} by
+giving a universal semantic characterization of formal grammar.
+From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type thoery
+that axiomatizes the underlying structures needed to conduct formal grammar
+theory. Then in \cref{sec:categorify}, we characterize the abstract categorical
+models of our syntax. Afterwards, we demonstrate several models of \theoryabbv,
+include one to prove a metatheoretic canonicity theorem. Finally, in
+\cref{sec:automata} we give a presentation of automata as grammars. With this
+view in mind, we may encode parsing algorithms and execute them entirely
+internal to our formalism. Our contributions are then:
 
 \begin{itemize}
-  \item A universal semantic notion of formal grammar given by the category $\Set^{\String}$
-  \item Categorical structures that capture the concepts integral to formal grammar theory, and that are notably modeled by $\Set^{\String}$
-  \item A dependent
-    linear/non-linear (LNL) type theory to syntactically capture the above categorical structures
+  \item A syntax-independent notion of grammar given by the category $\Grammar$
+  \item \theoryname: A dependent
+    linear-non-linear type theory meant to syntactically capture the latent structures
+    in $\Grammar$ that enable parsing
+  \item Categorical structures that capture the concepts integral to formal grammar theory, and that are notably modeled by $\Grammar$
   \item An implementation of the language, which is embedded shallowly in Agda
@@
 \section{A Syntax-Independent Notion of Syntax}
+\label{sec:synindnotion}
 
@@
 As they are simply sets, formal languages are naturally endowed with a partial
-order via subset inclusion. Similarly, formal grammars are naturally endowed
-with the structure of a \emph{category}. A morphism $f : G \to H$ between two
-grammars is given by a family of functions
+order via subset inclusion. Similarly, formal grammars naturally coalesce into a
+\emph{category}. Define $\Grammar$ to be the category of formal grammars. Whose
+objects are grammars, and whose morphisms are functions $f : G \to H$ between two
+grammars presented by a family of functions
 
@@
 for every string $w \in \String$. Intuitively, we read a morphism of grammars as
-a \emph{parse transformer}: $f$ translates a $G$-parse of $w$ into an $H$-parse
-of $w$. The notion of isomorphism induced by
-this categorical structure encodes that two grammars are equivalent if there is
-a bijective translation of parses --- precisely Chomsky's notion of \emph{strong equivalence} of
-grammars \cite{chom1963}.
-
-\pedro{Written like this, it seems like this correspondence iso-strong
-  equivalence is a well-known fact, but this is not an obvious
-  observation.}
-
-The above category of grammars can equivalently be described as the functor
-category $\Set^{\String}$ --- where the set $\String$ is viewed as a discrete
-category. That is, the objects in $\String$ are strings and the only morphisms
-are identity morphisms. Such functor categories into $\Set$ --- often called
+a \emph{parse transformer} --- i.e. $f$ translates a $G$-parse of $w$ into an $H$-parse
+of $w$. The induced notion of isomorphism in $\Grammar$ encodes that two grammars are equivalent if there is
+a bijective translation of parses. In other words, two grammars are isomorhpic
+if they capture the same structural descriptions of strings, which is
+precisely Chomsky's notion of \emph{strong equivalence} of
+grammars \cite{chom1963}. In works like
+\cite{yoshinaga2002formal}, this notion of strong equivalence is often treated
+like the appropriate choice of ``sameness'' for two grammars. When abstracting over the latent
+structure of formal grammars and describing it in the the language of
+categories, it is encouraging to see the same definition of equivalence arise
+naturally.
+
+Upon inspection of the above definition, $\Grammar$ may equivalently be described as the functor
+category $\Set^{\String}$ --- where the set $\String$ is viewed as a
+discrete\footnote{Objects in $\String$ comprise the set of all strings, and the
+  only morphisms are the identity morphisms.}
+category. In fact, in this paper we will take $\Grammar$ to be defined as the
+functor category $\Set^{\String}$. Such functor categories into $\Set$ --- often called
 presheaf categories --- carry a remarkable amount of structure.
@@
   grammar} --- a category paired with a functor into
-$\Set^{\String}$. That is, they may all be interpreted in our category of grammars. In this sense, $\Set^{\String}$ provides a universal setting to reason about parsing.
-The slice 2-category over $\Set^{\String}$ then provides the appropriate expressivity to prove equivalences between grammar formalism, such as the equivalence between semi-Thue systems and Chomsky's unrestricted grammars.
-\pedro{why?}
+$\Grammar$. That is, they may all be interpreted in our category of grammars. In this sense, $\Grammar$ provides a universal setting to reason about parsing.
 
 \steven{Check details for 2-cat of grammar formalisms}
+
+\section{\theoryname as a Syntax for Formal Grammars}
+\label{sec:tt}
+
+Omission of the structural rules of a deductive system, such as in
+linear logic \cite{GIRARD19871}, offers precise control over how a value is used
+in a derivation. Namely, linear logic omits the weakening and contraction rules
+to ensure that every value in context is used exactly once. This control enables
+\emph{resource-sensitive} reasoning, where we may treat a resource as
+\emph{consumed} after usage. This viewpoint is amenable to parsing applications,
+as we may treat characters of a string as finite resources that are consumed at
+parse-time. That is, once a particular substring has been parsed, a parser need not go
+over it again. One may then envision a linear type system where the types comprise
+formal grammars generated over some alphabet $\Sigma$, and the type constructors
+correspond precisely to inductive constructions on grammars --- such as
+conjunction, disjunction, concatenation, etc.
+
+Unfortunately, programming in a linear term language is often
+cumbersome and unintuitive for users. Code in such a language can
+become unnecessarily verbose when using Girard encodings to translate
+high-level reasoning into low-level linear terms. To alleviate this
+concern, in 1995 Benton et al.\ proposed an alternative categorical
+semantics of linear logic based on adjoint interactions between linear
+and non-linear logics \cite{bentonMixedLinearNonlinear1995} ---
+appropriately referred to as a \emph{linear-non-linear} (LNL)
+system. This work is simply typed, so the boundary between linear and
+non-linear subtheories is entirely characterized via a monoidal
+adjunction $F \dashv G$ between linear terms and non-linear terms.
+
+Inside of an LNL system, linearity may be thought of as an option that users can
+choose to deliberately invoke at deliberate points in their developments in an
+otherwise intuitionistic system. However, if we are wishing to treat parsers as
+linear terms over input strings, the non-linear fragment of an LNL theory does
+not really assist in the development of parsers. It is instead the case that
+parsers may benefit from a \emph{dependence} on non-linear terms.
+Through the approach described by Krishaswami et al.\ in
+\cite{krishnaswami_integrating_2015},
+we may define a restricted form of dependent types. In particular, dependence
+on linear terms
+is disallowed; however, through dependence of a linear term on a non-linear
+index, we recover the definition of Aho's indexed grammars \cite{AhoIndexed}
+internal to \theoryabbv.
+
+\subsection{Syntax}
+\label{subsec:syntax}
+
+Below we describe \theoryname~(\theoryabbv), an instance of an LNL theory with dependent types
+to serve as a deductive
+setting for formal grammar theory. \theoryabbv axiomatizes the necessary structure underlying $\Set^{\String}$ to
+specify and construct parsers.
+
+The structural judgments are shown in \cref{fig:structjdg}, the typing
+well-formedness in \cref{fig:typewf}, and the intuitionistic typing rules in
+\cref{fig:inttyping}. These are all just as they appear in
+\cite{krishnaswami_integrating_2015}. The treatment of the non-linear types is
+pretty standard, so below we focus on the linear syntax.
+
+\begin{figure}
+  \begin{mathpar}
+    \boxed{\ctxwff \Gamma}
+
+    \inferrule{~}{\ctxwff \cdot}
+    \and
+    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}
+
+    \\
+
+    \boxed{\linctxwff \Gamma \Delta}
+
+    \inferrule{~}{\linctxwff \Gamma \cdot}
+    \and
+    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}
+
+    \\
+
+    \boxed{\ctxwffjdg \Gamma X}
+
+    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}
+
+    \\
+
+    \boxed{\linctxwffjdg \Gamma A}
+
+    \inferrule{\Gamma \vdash A : L_i}{\linctxwffjdg \Gamma A}
+
+    \\
+
+    \boxed{\ctxwffjdg \Gamma {X \equiv Y}}
+
+    \inferrule{\Gamma \vdash X \equiv Y : U_i}{\ctxwffjdg \Gamma {X\equiv Y}}
+
+    \\
+
+    \boxed{\linctxwffjdg \Gamma {A \equiv B}}
+
+    \inferrule{\Gamma \vdash A \equiv B : L_i}{\linctxwffjdg \Gamma {A \equiv B}}
+
+  \end{mathpar}
+  \caption{Structural judgments}
+  \label{fig:structjdg}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \boxed{\Gamma \vdash X : U_{i}}
+
+    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
+ %
+    \and
+%
+    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
+%
+    \\
+%
+    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
+%
+    \and
+%
+    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
+%
+    \\
+%
+    \inferrule{~}{\Gamma \vdash 1 : U_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
+%
+    \\
+%
+    \inferrule{~}{\Gamma \vdash I : L_i}
+ %
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
+%
+    \\
+%
+    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
+%
+    \\
+%
+    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
+    %
+    \and
+    %
+    \inferrule{~}{\Gamma \vdash \top : L_i}
+%
+    \and
+%
+    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
+%
+    \\
+    %
+    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
+    %
+    \and
+    %
+    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
+  \end{mathpar}
+  \caption{Type well-formedness}
+  \label{fig:typewf}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+  \boxed{\Gamma \vdash x : X}
+
+  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
+  %
+  \\\
+  %
+  \inferrule{~}{\Gamma \vdash () : 1}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
+  %
+  \\
+%
+  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
+  \and
+  \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
+  %
+  \\
+  %
+  \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
+  \and
+  \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
+  \end{mathpar}
+  \caption{Intuitionistic typing}
+  \label{fig:inttyping}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \boxed{\Gamma ; \Delta \vdash a : A}
+
+    \inferrule{~}{\Gamma ; a : A \vdash a : A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
+    %
+    \\
+    %
+    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
+    %
+    \\
+    %
+    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
+    %
+    \\
+    %
+    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
+    \\
+    %
+    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
+    \and
+    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta', \Delta \vdash \applto {e'} {e} : B}
+    \\
+    %
+    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta, \Delta' \vdash \apptol e {e'} : B}
+    %
+    \\
+    %
+    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}{\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma ; \Delta \vdash \app e {e'} : \subst A {e'} x}
+    %
+    \\
+    %
+    \inferrule{\Gamma \vdash e : X \quad \Gamma ; \Delta \vdash e' : \subst A e x}{\Gamma ; \Delta \vdash (e, e') : \LinSigTy x X A}
+    %
+    \\
+    %
+    \inferrule{\Gamma ; \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
+    %
+    \\
+    %
+    \inferrule{~}{\Gamma ; \Delta \vdash () : \top}
+    %
+    \\
+    %
+    \inferrule{\Gamma ; \Delta \vdash e_1 : A_1 \quad \Gamma ; \Delta \vdash e_2 : A_2}{\Gamma ; \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : A_1 \amp A_2 }{\Gamma ; \Delta \vdash \pi_i \, e : A_i}
+    %
+    \\
+    %
+    \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma ; \cdot \vdash \mathsf{G}^{-1}\, e : A}
+    %
+    \\
+    %
+    \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
+    \and
+    \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
+  \end{mathpar}
+  \caption{Linear typing}
+  \label{fig:linsyntax}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \boxed{\Gamma \vdash e \equiv e' : X}
+
+    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
+%
+    \and
+%
+    \inferrule{~}{}
+%
+    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
+
+    \\
+
+    \boxed{\Gamma ; \Delta \vdash a \equiv a' : A}
+
+    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
+%
+    \\
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
+    %
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
+%
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
+    %
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
+\end{mathpar}
+  \caption{Judgmental equality}
+  \label{fig:jdgeq}
+\end{figure}
+
+\subsection{Formal Grammars as Linear Types}
+The linear typing judgment in our syntax takes on the following schematic form
+$\Gamma ; \Delta \vdash a : A$. First, $A$ represents a \emph{linear type} in
+our syntax. The intended semantics of these linear types are formal grammars.
+That is, the linear typing system is designed to syntactically reflect the
+behavior of formal grammars. For this reason, we may often interchangeably use
+the terms ``linear type'' and ``grammar''.
+
+The term $a$ is an inhabitant of type $A$, which is thought of as a parse tree of
+the grammar $A$. The core idea of this entire paper follows precisely from this
+single correspondence: grammars are types, and the inhabitants of these types
+are parse trees for the grammars.
+
+$\Gamma$ represents a non-linear context, while $\Delta$ represents a
+\emph{linear context} dependent on $\Gamma$. These linear contexts behave
+substructurally. As stated earlier, they do are linear --- they do not obey
+weakening or contraction --- because a character is exhausted once read by a
+parsing procedure. Moreover, the characters in strings appear in the order in
+which we read them. We do not have the freedom to freely permute characters,
+therefore any type theory that is used to reason about formal grammars ought to
+omit the structural rule of exchange as well. This means that every variable
+within $\Delta$ must be used \emph{exactly once} and \emph{in order of
+  occurrence}. Thus, we can think of the linear contexts as an ordered list of
+limited resources. Once a resource is consumed, we cannot make reference to it
+again. Variables in a linear context then act like building
+blocks for constructing patterns over strings.
+
+We give the base types and type constructors for linear terms. As the
+interpretation of types as grammars in $\Set^{\String}$ serves as our intended
+semantics, we simultaneously give the interpretation $\sem \cdot$ of the
+semantics as grammars.
+
+Most of the non-linear semantics is standard, so let us investigate in more
+depth the linear semantics. The interpretation of each
+connective is exactly what is suggested by the syntax, ude in
+large part to the inspiration taken from the grammar interpretation
+when designing \theoryabbv.
+
+\begin{enumerate}
+  \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
+  \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
+        $\sem X : \sem \Gamma \to \Set_{i}$
+  \item A non-linear term $\Gamma \vdash e : X$ denotes a section
+        $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$
+  \item Linear contexts $\Gamma \vdash \Delta$ and types
+        $\Gamma \vdash A : L_{i}$ both denote families of grammars
+        $\sem \Gamma \to \Gr_{i}$
+  \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
+        transformers
+        $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
+\end{enumerate}
+
+\paragraph{Linear Unit}
+First, the linear unit $I$ may be built in the empty linear
+context\footnote{When appropriate, statements like ``the empty linear context''
+  may be taken polymorphically over all non-linear contexts that they might
+  depend on.} $\cdot \dashv I$. $I$ serves as the unit for the operation
+$\otimes$ described below.
+
+As a grammar,
+
+\[
+  \sem {I} \gamma w = \{ () ~|~ w = \varepsilon \}
+\]
+
+That is, $I$ maps strings to the proposition that they are the empty string. $I$
+is clearly only inhabited by the empty string $\varepsilon$, for which the
+outputted set contains a single parse tree.
+
+\paragraph{Base Types}
+For each character $c$ in the alphabet $\Sigma$, we include a base type at the
+lowest universe level $c : L_{0}$.
+
+The grammar interpretation for characters is quite similar to that of $I$.
+\[
+  \sem {c} \gamma w = \{ () ~|~ w = c \}
+\]
+
+The grammar for the character $c$ likewise maps strings $w$ to the proposition that
+they are equal to $c$.
+
+\paragraph{Tensor Product}
+The tensor product of two linear types is the first place where the ordering on
+contexts really takes effect. That is, the tensor product of two types is
+non-commutative.
+When context $\Delta$ forms a linear term of
+type $A$, and $\Delta'$ forms a linear term of type $B$, then the context
+extension $\Delta, \Delta'$ forms a linear term of type $A \otimes B$.
+
+As stated above, the type $I$ serves as the unit for $\otimes$. That is, for all
+linear types $A$, the following equalities hold:
+
+\[ I \otimes A \equiv A \otimes I \equiv A \]
+
+In the grammar semantics,
+
+\[
+  \sem {A \otimes B} = \Sigma_{w_{1}, w_{2} : \String} (w_{1}w_{2} = w) \land \left(  \sem {A} \gamma w_{1} \times \sem {B} \gamma w_{2} \right)
+\]
+
+The string $w$ matches $A \otimes B$
+precisely when it may be split into two
+pieces such that the left one matches $A$
+and the right one matches $B$.
+
+\paragraph{Linear Function Types}
+The monoidal structure provided by $\otimes$ is both left and right closed and
+this is denoted by the left and right linear function types.
+
+\steven{Say more about the adjunctions}
+
+Further, for linear types $A$ and $B$ the left linear function type $A \lto B$
+enjoys an elimination principle similar to function application/modus ponens.
+
+\[
+\inferrule{\Delta \vdash A \otimes A \lto B}{\Delta \vdash B}
+\]
+
+The interpretation of linear function types at a string $w$ is a linear function
+that takes in parses of $A$ on a string $w_{a}$ and outputs parses of $B$ on the
+string $w_{a} w$.
+
+\[
+  \sem{A \lto B} \gamma w = \Pi(w_a:\String) \left( A \gamma w_a \Rightarrow B\gamma (w_a w) \right)
+\]
+
+That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
+complete to parses of $B$. In this manner, the linear function types generalize
+Brzozowksi's notion of derivative
+\cite{brzozowskiDerivativesRegularExpressions1964}.
+Brzozowski initially only gave an accounting of this operation for
+generalized regular expressions, but later work from Might et al.\ demonstrates that the same
+construction can be generalized to context free grammars
+\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
+types the same notion of derivative extends to the grammars of \theoryabbv.
+
+Of course, all the above statements for the left function type also have
+corresponding analogues for the
+right-handed counterpart.
+
+\paragraph{LNL Dependent Types}
+Given a non-linear type $X$, we may form both the dependent product type
+$\LinPiTy x X A$ and the dependent pair type $\LinSigTy x X A$ as linear types
+where $x$ is free in $A$.
+
+$\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+
+The grammar semantics of the linear product type is
+indeed a dependent function out of the
+semantics of $X$.
+
+$\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+
+Likewise, the grammar semantics of the linear dependent pair type is a dependent
+pair in $\Set$.
+
+\steven{TODO example of sum as an admissible usage of $\LinSigTy b \Bool A$}
+
+\paragraph{Universal Type}
+The universal type $\top$ may be formed in any context.
+
+\[ \sem{\top} \gamma w = \{ \ast \}\]
+
+Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.
+
+\paragraph{Empty Type}
+The empty type $\bot$ has no inhabitants.
+
+\[ \sem {\bot} \gamma w = \emptyset \]
+\steven{TODO need to include a case for $\bot$ in the inference rules}
+
+\paragraph{Pair Type}
+Given a term of type $A$ in context $\Delta$ and a term of type $B$ \emph{also
+  in context} $\Delta$, we may form a pair in context $\Delta$ of type
+$A \ampersand B$.
+
+\[ \sem{A \amp B }\gamma w = \sem{A} \gamma w \times \sem {B} \gamma w \]
+
+The grammar semantics of $A \amp B$ denotes a pair of a parse of $A$ and a
+parse of $B$ \emph{of the same string}. In this sense, $\amp$ acts a
+conjunction between grammars --- meaning a string only matches $A \amp B$
+if it simultaneously matches each of $A$ and $B$.
+
+\paragraph{The $G^{-1}$ Modality}
+Turn nonlinear into linear
+
+\steven{Call this $F$ or $G^{-1}$?}
+
+\steven{Why can't these map any nonlinear type into linear ones? instead of just
+those that are already linear types lifted by $G$?}
+
+\steven{TODO G inverse}
+  % \item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
+
+\paragraph{Recursive Types}
+isorecursive types
+\steven{TODO recursive types}
+  % \item $\sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x)$
+
+Observe that we need not take the Kleene star as a primitive
+grammar constructor, as it is definable as a fixed point.
+The Kleene star of a grammar $g$ is given as,
+
+\[
+  g^{*} := \mu X . I \oplus (g \otimes X)
+\]
+
+\begin{figure}[h!]
+\begin{mathpar}
+  \inferrule
+  {\Gamma ; \Delta \vdash p : I}
+  {\Gamma ; \Delta \vdash \mathsf{nil} : g^{*}}
+
+  \and
+
+  \inferrule
+  {\Gamma ; \Delta \vdash p : g \\ \Gamma ; \Delta' \vdash q
+  : g^{*}}
+  {\Gamma ; \Delta \vdash \mathsf{cons}(p , q) : g^{*}}
+
+  \and
+
+  \inferrule
+  {
+    \Gamma ; \Delta \vdash p : g^{*} \\
+    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
+    \Gamma ; x : g , y : h \vdash p_{\ast} : h
+  }
+  {\Gamma ; \Delta \vdash \mathsf{foldr}(p_{\varepsilon} , p_{\ast}) : g^{*}}
+\end{mathpar}
+\caption{Kleene Star Rules}
+\label{fig:star}
+\end{figure}
+
+Likewise, $g^{*}$ has admissible introduction and
+elimination rules, shown in \cref{fig:star}. Note that this
+definition of $g^{*}$ and these
+rules arbitrarily assigns a handedness to the Kleene star.
+We could have just as well took it to be a fixed point of
+$I \oplus (X \otimes g)$. In fact, the definitions are
+equivalent, as the existence of the $\mathsf{foldl}$ term below
+shows that $g^{*}$ is also a fixed point of
+$I \oplus (X \otimes g)$.
+
+\begin{equation}
+  \label{eq:foldl}
+  \inferrule
+  {
+    \Gamma ; \Delta \vdash p : g^{*} \\
+    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
+    \Gamma ; y : h , x : h \vdash p_{\ast} : h
+  }
+  {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
+\end{equation}
+
+In fact, the $\mathsf{foldl}$ term is defined using
+$\mathsf{foldr}$ --- much in the same way one
+may define a left fold over lists in terms of a right fold
+in a functional programming language\footnote{The
+  underlying trick is to fold over a list of functions
+  instead of the original string. We curry each character $c$
+  of the string into a function that concatenates $c$, and
+  right fold over this list of linear functions. Because function
+  application is left-associative, this results in a left
+  fold over the original string. }.
+
+We only take fixed points of a single variable as a
+primitive operation in the type theory, but we may apply
+Beki\`c's theorem \cite{Bekić1984} to define an admissible
+notion of multivariate fixed point. This is particularly
+useful for defining grammars that encode the states of an
+automaton. In \cref{fig:multifix} we provide the
+introduction and elimination principles for such a fixed
+point, where $\sigma$ is the substitution that unrolls the
+mutually recursive definitions one level. That is,
+
+\begin{align*}
+  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \\
+  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
+\end{align*}
+
+\begin{figure}[h!]
+\begin{mathpar}
+  \inferrule
+  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
+  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
+
+  \\
+
+  \inferrule
+  {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
+             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
+             \dots \\
+             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
+  }
+  {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
+\end{mathpar}
+\caption{Multi-fixed Points}
+\label{fig:multifix}
+\end{figure}
+
+The most prevalent application of these multi-fixed points
+is for defining the traces through automata, as we will see
+later in \cref{eq:nfatrace}.
+
+\subsection{Subtheories}
+\paragraph{Regular Expressions} We may realize regular expression as a subtheory
+of our language by restricting the type constructors to the
+linear unit, characters, $\otimes$, $\oplus$, and Kleene star. Classically, the
+languages recognized by these are exactly the regular languages --- the lowest
+on the Chomksy hierarchy.
+
+\paragraph{$\mu$-Regular Expressions} Similarly, we may restrict the connectives
+to be linear unit, characters, $\otimes$, $\oplus$, and arbitrary recursion.
+Instead of regular expressions these correspond to Lei{\ss}'s $\mu$-regular
+expressions, which are known to be equivalent to context-free grammars
+\cite{leiss}.
+
+\steven{Give this subtheory a name }
+
+\paragraph{Beyond Context-Free Grammars} The previous two subtheories induce as semantics
+regular grammars and context-free grammars, respectively; however, by including
+the LNL dependent types we may actually express the entirety of the Chomsky
+hierarchy. Through use of the LNL dependent types, we may encode indexed
+grammars, which are known to be properly between context-free and
+context-sensitive grammars within the Chomsky hierarchy \cite{AhoIndexed}. We
+will further see in \cref{subsubsec:tm} how to use dependence to encode Turing machines internal to our
+calculus, which of course induces a semantics of unrestricted grammars.
+
+\steven{TODO $\mu$ regular expressions}
 
@@
 \label{sec:categorify}
-\steven{Idk if this section has the right title, but I wanted to change it if we are adding ``Chomsky categories''}
-
+
+Under the Curry-Howard-Lambek correspondence, the programs of the simply-typed
+$\lambda$-calculus may be viewed under any one of the three equivalent views: as
+terms in a simple functional language, as proofs in an intuitionisitc
+propositional logic, or as morphisms in a Cartesian closed category. In the
+same vein, with the syntax provided, we aim to give a
+categorical accounting of the semantics of \theoryabbv. Let us begin by paring back
+the connectives in the language and only considering the subtheory induced by
+regular expressions.
+
+\steven{Added this transition to handle putting the categorical semantics after
+  the syntax. I think it sounds weird though}
+
+By giving algebraic laws of manipulation and equivalence to regular expressions,
 Kleene algebras are an important tool in the theory of regular
@@
 which we call Chomsky category, that will serve as the categorical
-semantics to our type theory.
-
-\subsection{Kleene categories}
+semantics to \theoryabbv.
+
+\subsection{Kleene Categories}
 A Kleene algebra is a tuple $(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is
@@
 
-\begin{lemma}[\cite{day1970construction}]
+\begin{lemma}[Day \cite{day1970construction}]
   Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
@@
 and their parsers. This leads us to the abstract notion of model we
-are interested in capturing with our type theory: a \emph{Chomsky
+are interested in capturing with \theoryabbv: a \emph{Chomsky
 category}.
@@
 
+\steven{Max suggests augmenting the definition of a chomsky category to
+  something like two categories $L$ and $U$, and $U$ is cartesian closed I don't
+  fully recall the rest.
+
+  My best guess is that you take $L$ a Kleene category with a hierarchy of universes
+  two, and you further require that $L$ is $Psh(U)$-enriched, except he
+  suggested a further adjective on these presheaves. Perhaps representability?
+}
+
 \begin{theorem}
@@
 
-Further, the syntactic category of the theory in \cref{sec:tt} is manifestly a Chomsky category.
-\steven{This is true, right? That's like the entire point of having the syntax}
+Further, the syntactic category of \theoryname is manifestly a Chomsky category.
+
 \pedro{This is likely true, but if we explicitly say so, this warants a proof. I think that
@@
 
-\section{Non-commutative Linear-non-linear Type Theory as a Syntax for Formal Grammars}
-\label{sec:tt}
-
-Omission of the structural rules of a deductive system, such as in
-linear logic \cite{GIRARD19871}, offers precise control over how a value is used
-in a derivation. Namely, linear logic omits the weakening and contraction rules
-to ensure that every value in context is used exactly once. This control enables
-\emph{resource-sensitive} reasoning, where we may treat a resource as
-\emph{consumed} after usage. This viewpoint is amenable to parsing applications,
-as we may treat characters of a string as finite resources that are consumed at
-parse-time. That is, once a particular substring has been parsed, a parser need not go
-over it again. One may then envision a linear type system where the types comprise
-formal grammars generated over some alphabet $\Sigma$, and the type constructors
-correspond precisely to inductive constructions on grammars --- such as
-conjunction, disjunction, concatenation, etc.
-
-Unfortunately, programming in a linear term language is often
-cumbersome and unintuitive for users. Code in such a language can
-become unnecessarily verbose when using Girard encodings to translate
-high-level reasoning into low-level linear terms. To alleviate this
-concern, in 1995 Benton et al.\ proposed an alternative categorical
-semantics of linear logic based on adjoint interactions between linear
-and non-linear logics \cite{bentonMixedLinearNonlinear1995} ---
-appropriately referred to as a \emph{linear-non-linear} (LNL)
-system. This work is simply typed, so the boundary between linear and
-non-linear subtheories is entirely characterized via a monoidal
-adjunction $F \dashv G$ between linear terms and non-linear terms.
-
-\steven{Maybe there is more to say here, or perhaps later, that the only thing
-  to gain parsing-wise from a non-dependent non-linear fragment are the things that are
-  definable via $G$. Except that these are not interesting, as they are
-  necessarily empty}
-
-\steven{Are these paragraphs recounting too much history?}
-
-Inside of an LNL system, linearity may be thought of as an option that users can
-choose to deliberately invoke at deliberate points in their developments in an
-otherwise intuitionistic system. However, if we are wishing to treat parsers as
-linear terms over input strings, the non-linear fragment of an LNL theory does
-not really assist in the development of parsers. It is instead the case that
-parsers may benefit from a \emph{dependence} on non-linear terms.
-\pedro{This last sentence could use an example. Is there a bite-sized, well-known
-example we could mention here?}
-
-\steven{Mention separation logic at some point. In particular the separating conjunction and magic wand}
-
-Krishnaswami et al.\ equipped LNL systems with a restricted sense of dependent
-types \cite{krishnaswami_integrating_2015}. When defining dependent types in
-this setting, dependence on linear terms is disallowed, but either linear or
-non-linear terms are allowed to depend on non-linear terms.
-
-The non-linear
-subtheory corresponds to Martin-L\"of type theory (MLTT), which can be taken
-as a foundation for mathematics. Dependence of linear terms on non-linear terms
-of this form effectively allows us to bring any ordinary definitions from math
-into scope as indices and then define grammars relative to those indices. That
-is all to say, the LNL dependent types in the style of Krishnaswami et al.\
-allows us to recover the definition of Aho's indexed grammars \cite{AhoIndexed} internal to our
-type theory.
-\steven{The above paragraph should maybe be put later. Also flows weirdly}
-
-\pedro{I would argue that both paragraphs above should either be put
-  somewhere else or deleted. I think that we should definitely mention
-  Neel's LNL system, but as it stands, it feels disconnected from what
-  came before and what comes next. A suggestion: the mention of Aho's
-  indexed grammars could be the example we use illustrating the
-  benefits of dependent grammars.}
-
-\subsection{Syntax}
-\label{subsec:syntax}
-
-We instantiate an LNL theory with dependent types as a deductive
-setting to axiomatize the structure of a Chomsky category.
-
-The structural judgments are shown in \cref{fig:structjdg}, the typing
-well-formedness in \cref{fig:typewf}, and the intuitionistic typing rules in
-\cref{fig:inttyping}. These are all just as they appear in
-\cite{krishnaswami_integrating_2015}.
-
-\begin{figure}
-  \begin{mathpar}
-    \boxed{\ctxwff \Gamma}
-
-    \inferrule{~}{\ctxwff \cdot}
-    \and
-    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}
-
-    \\
-
-    \boxed{\linctxwff \Gamma \Delta}
-
-    \inferrule{~}{\linctxwff \Gamma \cdot}
-    \and
-    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}
-
-    \\
-
-    \boxed{\ctxwffjdg \Gamma X}
-
-    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}
-
-    \\
-
-    \boxed{\linctxwffjdg \Gamma A}
-
-    \inferrule{\Gamma \vdash A : L_i}{\linctxwffjdg \Gamma A}
-
-    \\
-
-    \boxed{\ctxwffjdg \Gamma {X \equiv Y}}
-
-    \inferrule{\Gamma \vdash X \equiv Y : U_i}{\ctxwffjdg \Gamma {X\equiv Y}}
-
-    \\
-
-    \boxed{\linctxwffjdg \Gamma {A \equiv B}}
-
-    \inferrule{\Gamma \vdash A \equiv B : L_i}{\linctxwffjdg \Gamma {A \equiv B}}
-
-  \end{mathpar}
-  \caption{Structural judgments}
-  \label{fig:structjdg}
-\end{figure}
-
-\begin{figure}
-  \begin{mathpar}
-    \boxed{\Gamma \vdash X : U_{i}}
-
-    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
- %
-    \and
-%
-    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
-%
-    \\
-%
-    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
-%
-    \and
-%
-    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
-%
-    \\
-%
-    \inferrule{~}{\Gamma \vdash 1 : U_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
-%
-    \\
-%
-    \inferrule{~}{\Gamma \vdash I : L_i}
- %
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
-%
-    \\
-%
-    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
-%
-    \\
-%
-    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
-    %
-    \and
-    %
-    \inferrule{~}{\Gamma \vdash \top : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
-%
-    \\
-    %
-    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
-    %
-    \and
-    %
-    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
-  \end{mathpar}
-  \caption{Type well-formedness}
-  \label{fig:typewf}
-\end{figure}
-
-\begin{figure}
-  \begin{mathpar}
-  \boxed{\Gamma \vdash x : X}
-
-  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
-  %
-  \\\
-  %
-  \inferrule{~}{\Gamma \vdash () : 1}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
-  %
-  \\
-%
-  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
-  \and
-  \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
-  %
-  \\
-  %
-  \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
-  \and
-  \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
-  \end{mathpar}
-  \caption{Intuitionistic typing}
-  \label{fig:inttyping}
-\end{figure}
-
-\begin{figure}
-  \begin{mathpar}
-    \boxed{\Gamma ; \Delta \vdash a : A}
-
-    \inferrule{~}{\Gamma ; a : A \vdash a : A}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
-    %
-    \\
-    %
-    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
-    %
-    \\
-    %
-    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
-    %
-    \\
-    %
-    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
-    \\
-    %
-    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
-    \and
-    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta', \Delta \vdash \applto {e'} {e} : B}
-    \\
-    %
-    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta, \Delta' \vdash \apptol e {e'} : B}
-    %
-    \\
-    %
-    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}{\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma ; \Delta \vdash \app e {e'} : \subst A {e'} x}
-    %
-    \\
-    %
-    \inferrule{\Gamma \vdash e : X \quad \Gamma ; \Delta \vdash e' : \subst A e x}{\Gamma ; \Delta \vdash (e, e') : \LinSigTy x X A}
-    %
-    \\
-    %
-    \inferrule{\Gamma ; \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
-    %
-    \\
-    %
-    \inferrule{~}{\Gamma ; \Delta \vdash () : \top}
-    %
-    \\
-    %
-    \inferrule{\Gamma ; \Delta \vdash e_1 : A_1 \quad \Gamma ; \Delta \vdash e_2 : A_2}{\Gamma ; \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : A_1 \amp A_2 }{\Gamma ; \Delta \vdash \pi_i \, e : A_i}
-    %
-    \\
-    %
-    \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma ; \cdot \vdash \mathsf{G}^{-1}\, e : A}
-    %
-    \\
-    %
-    \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
-    \and
-    \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
-  \end{mathpar}
-  \caption{Linear typing}
-  \label{fig:linsyntax}
-\end{figure}
-
-\begin{figure}
-  \begin{mathpar}
-    \boxed{\Gamma \vdash e \equiv e' : X}
-
-    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
-%
-    \and
-%
-    \inferrule{~}{}
-%
-    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
-
-    \\
-
-    \boxed{\Gamma ; \Delta \vdash a \equiv a' : A}
-
-    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
-%
-    \\
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
-    %
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
-%
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
-    %
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
-\end{mathpar}
-  \caption{Judgmental equality}
-  \label{fig:jdgeq}
-\end{figure}
-
-\subsection{Formal Grammars as Linear Types}
-The linear typing judgment in our syntax takes on the following schematic form
-$\Gamma ; \Delta \vdash a : A$. First, $A$ represents a \emph{linear type} in
-our syntax. The intended semantics of these linear types are formal grammars.
-That is, the linear typing system is designed to syntactically reflect the
-behavior of formal grammars. For this reason, we may often interchangeably use
-the terms ``linear type'' and ``grammar''.
-
-The term $a$ is an inhabitant of type $A$, which is thought of as a parse tree of
-the grammar $A$. The core idea of this entire paper follows precisely from this
-single correspondence: grammars are types, and the inhabitants of these types
-are parse trees for the grammars.
-
-$\Gamma$ represents a non-linear context, while $\Delta$ represents a
-\emph{linear context} dependent on $\Gamma$. These linear contexts behave
-substructurally. As stated earlier, they do are linear --- they do not obey
-weakening or contraction --- because a character is exhausted once read by a
-parsing procedure. Moreover, the characters in strings appear in the order in
-which we read them. We do not have the freedom to freely permute characters,
-therefore any type theory that is used to reason about formal grammars ought to
-omit the structural rule of exchange as well. This means that every variable
-within $\Delta$ must be used \emph{exactly once} and \emph{in order of
-  occurrence}. Thus, we can think of the linear contexts as an ordered list of
-limited resources. Once a resource is consumed, we cannot make reference to it
-again. Variables in a linear context then act like building
-blocks for constructing patterns over strings.
-
-We give the base types and type constructors for linear terms. As the
-interpretation of types as grammars in $\Set^{\String}$ serves as our intended
-semantics, we simultaneously give the interpretation $\sem \cdot$ of the
-semantics as grammars.
-
-Most of the non-linear semantics is standard, so let us investigate in more
-depth the linear semantics. The interpretation of each
-connective is exactly what is suggested by the syntax, in
-large part to the inspiration taken from this interpretation
-when designing the type theory.
-
-\begin{enumerate}
-  \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
-  \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
-        $\sem X : \sem \Gamma \to \Set_{i}$
-  \item A non-linear term $\Gamma \vdash e : X$ denotes a section
-        $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$
-  \item Linear contexts $\Gamma \vdash \Delta$ and types
-        $\Gamma \vdash A : L_{i}$ both denote families of grammars
-        $\sem \Gamma \to \Gr_{i}$
-  \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
-        transformers
-        $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
-\end{enumerate}
-
-\paragraph{Linear Unit}
-First, the linear unit $I$ may be built in the empty linear
-context\footnote{When appropriate, statements like ``the empty linear context''
-  may be taken polymorphically over all non-linear contexts that they might
-  depend on.} $\cdot \dashv I$. $I$ serves as the unit for the operation
-$\otimes$ described below.
-
-As a grammar,
-
-\[
-  \sem {I} \gamma w = \{ () ~|~ w = \varepsilon \}
-\]
-
-That is, $I$ maps strings to the proposition that they are the empty string. $I$
-is clearly only inhabited by the empty string $\varepsilon$, for which the
-outputted set contains a single parse tree.
-
-\paragraph{Base Types}
-For each character $c$ in the alphabet $\Sigma$, we include a base type at the
-lowest universe level $c : L_{0}$.
-
-The grammar interpretation for characters is quite similar to that of $I$.
-\[
-  \sem {c} \gamma w = \{ () ~|~ w = c \}
-\]
-
-The grammar for the character $c$ likewise maps strings $w$ to the proposition that
-they are equal to $c$.
-
-\paragraph{Tensor Product}
-The tensor product of two linear types is the first place where the ordering on
-contexts really takes effect. That is, the tensor product of two types is
-non-commutative.
-When context $\Delta$ forms a linear term of
-type $A$, and $\Delta'$ forms a linear term of type $B$, then the context
-extension $\Delta, \Delta'$ forms a linear term of type $A \otimes B$.
-
-As stated above, the type $I$ serves as the unit for $\otimes$. That is, for all
-linear types $A$, the following equalities hold:
-
-\[ I \otimes A \equiv A \otimes I \equiv A \]
-
-In the grammar semantics,
-
-\[
-  \sem {A \otimes B} = \Sigma_{w_{1}, w_{2} : \String} (w_{1}w_{2} = w) \land \left(  \sem {A} \gamma w_{1} \times \sem {B} \gamma w_{2} \right)
-\]
-
-The string $w$ matches $A \otimes B$
-precisely when it may be split into two
-pieces such that the left one matches $A$
-and the right one matches $B$.
-
-\paragraph{Linear Function Types}
-The monoidal structure provided by $\otimes$ is both left and right closed and
-this is denoted by the left and right linear function types.
-
-\steven{Say more about the adjunctions}
-
-Further, for linear types $A$ and $B$ the left linear function type $A \lto B$
-enjoys an elimination principle similar to function application/modus ponens.
-
-\[
-\inferrule{\Delta \vdash A \otimes A \lto B}{\Delta \vdash B}
-\]
-
-The interpretation of linear function types at a string $w$ is a linear function
-that takes in parses of $A$ on a string $w_{a}$ and outputs parses of $B$ on the
-string $w_{a} w$.
-
-\[
-  \sem{A \lto B} \gamma w = \Pi(w_a:\String) \left( A \gamma w_a \Rightarrow B\gamma (w_a w) \right)
-\]
-
-That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
-complete to parses of $B$. In this manner, the linear function types generalize
-Brzozowksi's notion of derivative
-\cite{brzozowskiDerivativesRegularExpressions1964}.
-Brzozowski initially only gave an accounting of this operation for
-generalized regular expressions, but later work from Might et al.\ demonstrates that the same
-construction can be generalized to context free grammars
-\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
-types the same notion of derivative extends to the grammars of our type theory.
-
-Of course, all the above statements for the left function type also have
-corresponding analogues for the
-right-handed counterpart.
-
-\paragraph{LNL Dependent Types}
-Given a non-linear type $X$, we may form both the dependent product type
-$\LinPiTy x X A$ and the dependent pair type $\LinSigTy x X A$ as linear types
-where $x$ is free in $A$.
-
-$\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-
-The grammar semantics of the linear product type is
-indeed a dependent function out of the
-semantics of $X$.
-
-$\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-
-Likewise, the grammar semantics of the linear dependent pair type is a dependent
-pair in $\Set$.
-
-\steven{TODO example of sum as an admissible usage of $\LinSigTy b \Bool A$}
-
-\paragraph{Universal Type}
-The universal type $\top$ may be formed in any context.
-
-\[ \sem{\top} \gamma w = \{ \ast \}\]
-
-Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.
-
-\paragraph{Empty Type}
-The empty type $\bot$ has no inhabitants.
-
-\[ \sem {\bot} \gamma w = \emptyset \]
-\steven{TODO need to include a case for $\bot$ in the inference rules}
-
-\paragraph{Pair Type}
-Given a term of type $A$ in context $\Delta$ and a term of type $B$ \emph{also
-  in context} $\Delta$, we may form a pair in context $\Delta$ of type
-$A \ampersand B$.
-
-\[ \sem{A \amp B }\gamma w = \sem{A} \gamma w \times \sem {B} \gamma w \]
-
-The grammar semantics of $A \amp B$ denotes a pair of a parse of $A$ and a
-parse of $B$ \emph{of the same string}. In this sense, $\amp$ acts a
-conjunction between grammars --- meaning a string only matches $A \amp B$
-if it simultaneously matches each of $A$ and $B$.
-
-\paragraph{The $G^{-1}$ Modality}
-Turn nonlinear into linear
-
-\steven{Call this $F$ or $G^{-1}$?}
-
-\steven{Why can't these map any nonlinear type into linear ones? instead of just
-those that are already linear types lifted by $G$?}
-
-\steven{TODO G inverse}
-  % \item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
-
-\paragraph{Recursive Types}
-isorecursive types
-\steven{TODO recursive types}
-  % \item $\sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x)$
-
-Observe that we need not take the Kleene star as a primitive
-grammar constructor, as it is definable as a fixed point.
-The Kleene star of a grammar $g$ is given as,
-
-\[
-  g^{*} := \mu X . I \oplus (g \otimes X)
-\]
-
-\begin{figure}[h!]
-\begin{mathpar}
-  \inferrule
-  {\Gamma ; \Delta \vdash p : I}
-  {\Gamma ; \Delta \vdash \mathsf{nil} : g^{*}}
-
-  \and
-
-  \inferrule
-  {\Gamma ; \Delta \vdash p : g \\ \Gamma ; \Delta' \vdash q
-  : g^{*}}
-  {\Gamma ; \Delta \vdash \mathsf{cons}(p , q) : g^{*}}
-
-  \and
-
-  \inferrule
-  {
-    \Gamma ; \Delta \vdash p : g^{*} \\
-    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
-    \Gamma ; x : g , y : h \vdash p_{\ast} : h
-  }
-  {\Gamma ; \Delta \vdash \mathsf{foldr}(p_{\varepsilon} , p_{\ast}) : g^{*}}
-\end{mathpar}
-\caption{Kleene Star Rules}
-\label{fig:star}
-\end{figure}
-
-Likewise, $g^{*}$ has admissible introduction and
-elimination rules, shown in \cref{fig:star}. Note that this
-definition of $g^{*}$ and these
-rules arbitrarily assigns a handedness to the Kleene star.
-We could have just as well took it to be a fixed point of
-$I \oplus (X \otimes g)$. In fact, the definitions are
-equivalent, as the existence of the $\mathsf{foldl}$ term below
-shows that $g^{*}$ is also a fixed point of
-$I \oplus (X \otimes g)$.
-
-\begin{equation}
-  \label{eq:foldl}
-  \inferrule
-  {
-    \Gamma ; \Delta \vdash p : g^{*} \\
-    \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
-    \Gamma ; y : h , x : h \vdash p_{\ast} : h
-  }
-  {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
-\end{equation}
-
-In fact, the $\mathsf{foldl}$ term is defined using
-$\mathsf{foldr}$ --- much in the same way one
-may define a left fold over lists in terms of a right fold
-in a functional programming language\footnote{The
-  underlying trick is to fold over a list of functions
-  instead of the original string. We curry each character $c$
-  of the string into a function that concatenates $c$, and
-  right fold over this list of linear functions. Because function
-  application is left-associative, this results in a left
-  fold over the original string. }.
-
-We only take fixed points of a single variable as a
-primitive operation in the type theory, but we may apply
-Beki\`c's theorem \cite{Bekić1984} to define an admissible
-notion of multivariate fixed point. This is particularly
-useful for defining grammars that encode the states of an
-automaton. In \cref{fig:multifix} we provide the
-introduction and elimination principles for such a fixed
-point, where $\sigma$ is the substitution that unrolls the
-mutually recursive definitions one level. That is,
-
-\begin{align*}
-  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \\
-  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
-\end{align*}
-
-\begin{figure}[h!]
-\begin{mathpar}
-  \inferrule
-  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
-  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
-
-  \\
-
-  \inferrule
-  {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
-             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
-             \dots \\
-             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
-  }
-  {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
-\end{mathpar}
-\caption{Multi-fixed Points}
-\label{fig:multifix}
-\end{figure}
-
-The most prevalent application of these multi-fixed points
-is for defining the traces through automata, as we will see
-later in \cref{eq:nfatrace}.
-
-\subsection{Subtheories}
-\paragraph{Regular Expressions} Restricting to regular expressions with the
-obvious connectives forms a subtheory of our type theory. Perhaps there is
-something to say here about how this reduces to Kleene algebra performed inside
-of a Kleene cat?
-
-\steven{TODO regular expressions}
-
-\paragraph{$\mu$-Regular Expressions} Regex + fixed point equivalent to context
-free \cite{leiss}.
-
-\steven{TODO $\mu$ regular expressions}
-
-\steven{Every after here has not been touched}
-
-\pedro{After reading this section after the Kleene category one, I
-  think that it makes much more sense to switch their order. It's
-  going to be easier to justify providing a categorical semantics once
-  we have a ``target'' of what we want to interpret.}
-
-\section{Canonicity}
-\pedro{It could be cool to come up with a catchy title to this section.
-Something like ``Parse Canonicity''}
-
-\pedro{Added the ``initial context'' paragraph below}
-
+\steven{I agree that we don't want to say anything that opens unnecessary
+  questions for proofs we haven't written. However, it seems hard to make the
+  case that we have the right categorical model of the syntax if this statement
+  isn't true. By restating our definition of Chomksy category, this should be
+  obvious or a quick proof}
+
+\section{Nonstandard Models}
+\label{sec:nonstandard}
+
+\subsection{Semantic Actions}
+Returning to the problem of parsing, the output of a parse usually isn't the
+literal parse tree. Rather, the output is the result of some \emph{semantic
+  actions} ran on the parse tree, which usually serve to remove some syntactic
+details that are unnecessary for later processing.
+
+Given a grammar $G : \String \to \Set$, we define a semantic action to be a set
+$X$ with a function $f$ that produces a semantic element from any parse of $G$.
+
+\[
+  f : \PiTy w \String {G w \to X}
+\]
+
+Further, semantic actions can be arranged into a structured category.
+Define $\SemAct$ as the comma
+category $\Grammar / \Delta$, where $\Delta : \Set \to \Grammar$ defines a
+discrete presheaf. That is, for a set $X$, $\Delta (X)(w) = X$ for all
+$w \in \String$. As $\SemAct$ is defined as a comma category, it has a forgetul
+functor into $\Grammar$. That is, $\SemAct$ serves as a notion of formal
+grammar. Moreover, $\SemAct$ is a model of \theoryabbv.
+
+\steven{It being a notion of formal grammar is distinct from being a model. This
+probably warrants a proof}
+
+\steven{TODO semantic actions}
+
+\subsection{Parse Canonicity}
 Canonicity is an important metatheoretic theorem in the type theory
@@
 canonicity for boolean types, i.e. every closed term of type bool
-reduces to either true or false, is enough to justifying that the type
+reduces to either true or false, is enough to justify that the type
 theory being studied is well-behaved. In our case, however, since we
-want to connect our type theory to parsers, we must provide a more
+want to connect \theoryabbv to parsers, we must provide a more
 detailed account of canonicity.
@@
 
+\pedro{We should conclude this section by explaining the relevance of the canonicity
+theorem. Could also be done before stating the theorem.}
+
 \section{Automata as Grammars}
@@
 In the Chomsky hierarchy, each language class is associated to a class of
-automata that serve as recognizers. Internal to the type theory of \cref{sec:tt},
+automata that serve as recognizers. Internal to \theoryabbv,
 we can characterize these language classes syntactically; moreover, we

### 2024-07-10 POPL/paper.tex
got rid of binary ampersand

@@
 \usepackage{stmaryrd}
+\usepackage{scalerel}
 \usepackage{tikz}
@@
 \newcommand{\Bool}{\mathbf{Bool}}
+\newcommand{\true}{\mathbf{true}}
+\newcommand{\false}{\mathbf{false}}
 \newcommand{\nat}{\mathbb{N}}
@@
 \newcommand{\gluedL}{{\mathcal G}_L}
+
+\newcommand{\amp}{\mathrel{\&}}
+\newcommand{\pair}{\amp}
+\DeclareMathOperator*{\bigamp}{\scalerel*{\&}{\bigoplus}}
 
@@
 \newcommand{\apptol}[2]{#1 \mathop{{}^{\tol}} #2}
-\newcommand{\PiTy}[3]{\Pi #1 : #2.\, #3}
-\newcommand{\SigTy}[3]{\Sigma #1 : #2.\, #3}
-\newcommand{\LinPiTy}[3]{\widebar\Pi #1 : #2.\, #3}
-\newcommand{\LinSigTy}[3]{\widebar\Sigma #1 : #2.\, #3}
-\newcommand{\amp}{\mathrel{\&}}
+\newcommand{\PiTy}[3]{\textstyle\prod (#1 : #2). #3}
+\newcommand{\SigTy}[3]{\textstyle\sum (#1 : #2). #3}
+\newcommand{\LinPiTy}[3]{\textstyle\bigamp (#1 : #2). #3}
+\newcommand{\LinSigTy}[3]{\textstyle\bigoplus (#1 : #2). #3}
 \newcommand{\GrTy}{\mathsf{Gr}}
@@
 
-\title{Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}
+\title{Formal Grammars as Types in Non-Commutative Linear-Non-Linear Type Theory}
 
@@
 
-\paragraph{Contributions.} This paper begins in \cref{sec:synindnotion} by
+In this paper, we propose an intuitive and structured generalization of the
+theoretical machinery used for parsing. In conjunction with an updated
+point-of-view on formal grammars, we provide a domain-specific
+language to build and verify parsers.
+
+\steven{The above paragraph maybe isn't the right thing to say, but I want
+  something that ties the first paragraph to the contributions that follow}
+
+\paragraph{Contributions} This paper begins in \cref{sec:synindnotion} by
 giving a universal semantic characterization of formal grammar.
@@
 
-    \\
-
     \boxed{\linctxwffjdg \Gamma A}
@@
     \inferrule{\Gamma \vdash X \equiv Y : U_i}{\ctxwffjdg \Gamma {X\equiv Y}}
-
-    \\
 
@@
 
+    \boxed{\Gamma \vdash A : L_{i}}
+
     \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
@@
     \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
-%
-    \\
+
+    \and
 %
@@
 %
-    \and
+    \\
 %
@@
     \and
-%
-    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
-%
-    \\
-    %
+
+    \inferrule{~}{\Gamma \vdash \bot : L_i}
+
+    \\
     \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
@@
     \inferrule{~}{\Gamma ; \Delta \vdash () : \top}
-    %
-    \\
-    %
-    \inferrule{\Gamma ; \Delta \vdash e_1 : A_1 \quad \Gamma ; \Delta \vdash e_2 : A_2}{\Gamma ; \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : A_1 \amp A_2 }{\Gamma ; \Delta \vdash \pi_i \, e : A_i}
-    %
-    \\
+
+    \and
+
+    \inferrule{~}{\Gamma ; x : \bot \vdash \mathsf{absurd}_A : A}
+    %
+    \and
+    %
     %
@@
 
+$\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+
 The grammar semantics of the linear product type is
 indeed a dependent function out of the
-semantics of $X$.
-
-$\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-
-Likewise, the grammar semantics of the linear dependent pair type is a dependent
+semantics of $X$. Likewise, the grammar semantics of the linear dependent pair type is a dependent
 pair in $\Set$.
 
-\steven{TODO example of sum as an admissible usage of $\LinSigTy b \Bool A$}
+Note that even though we do not take the binary additive conjunction and
+disjunction as primitive, we may define them via these LNL dependent types. In
+particular, via a dependence on $\Bool$.
+
+\[
+  A \amp B := \LinPiTy b \Bool {C(b)}
+\]
+\[
+  A \oplus B := \LinSigTy b \Bool {C(b)},
+\]
+where $C(\true) = A, C(\false) = B$.
 
@@
 \paragraph{Empty Type}
-The empty type $\bot$ has no inhabitants.
+The empty type $\bot$ has no inhabitants. The elimination for the empty type
+witnesses the principle of explosion --- i.e. from a term of type $\bot$ we may
+introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.
 
 \[ \sem {\bot} \gamma w = \emptyset \]
-\steven{TODO need to include a case for $\bot$ in the inference rules}
-
-\paragraph{Pair Type}
-Given a term of type $A$ in context $\Delta$ and a term of type $B$ \emph{also
-  in context} $\Delta$, we may form a pair in context $\Delta$ of type
-$A \ampersand B$.
-
-\[ \sem{A \amp B }\gamma w = \sem{A} \gamma w \times \sem {B} \gamma w \]
-
-The grammar semantics of $A \amp B$ denotes a pair of a parse of $A$ and a
-parse of $B$ \emph{of the same string}. In this sense, $\amp$ acts a
-conjunction between grammars --- meaning a string only matches $A \amp B$
-if it simultaneously matches each of $A$ and $B$.
 
@@
 \begin{theorem}
-  The presheaf category $\Set^{\cat{\Sigma^*}}$ is a Chomsky category.
+  The presheaf category $\Grammar = \Set^{\cat{\Sigma^*}}$ is a Chomsky category.
 \end{theorem}
@@
 \[
- \mathsf{AccTrace}_{N} := \LinSigTy q Q {\left( \mathsf{Trace}_{N}(q_{0} , q) \amp \mathsf{acc}(q) \right)}
+ \mathsf{AccTrace}_{N} := \LinSigTy q Q {\left( \mathsf{Trace}_{N}(q_{0} , q) \pair \mathsf{acc}(q) \right)}
 \]
@@
   \[
-    \mathsf{AccTrace}_{NFA(I)} = \LinSigTy q {\{1 , 2\}} {\left( \mathsf{Trace}_{NFA(I)}(q_{1} , q) \amp \mathsf{acc}(q) \right)}
+    \mathsf{AccTrace}_{NFA(I)} = \LinSigTy q {\{1 , 2\}} {\left( \mathsf{Trace}_{NFA(I)}(q_{1} , q) \pair \mathsf{acc}(q) \right)}
   \]
@@
     \mathsf{AccTrace}_{NFA(I)}
-    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \amp \mathsf{acc}(q_{2}) \\
-    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \amp \top \\
+    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \pair \mathsf{acc}(q_{2}) \\
+    & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \pair \top \\
     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})

### 2024-07-10 POPL/paper.tex
gonna add some future/related work then sleep

@@
 \newcommand{\nat}{\mathbb{N}}
-\newcommand{\theoryname}{Grammar Type Theory\xspace}
-\newcommand{\theoryabbv}{GTT\xspace}
+\newcommand{\theoryname}{Grammar Theoretic Type Theory\xspace}
+\newcommand{\theoryabbv}{GTTT\xspace}
 
@@
 prominent topic in the 1950, 60s, and 70s led to a series of remarkable
-developments: Chomsky's hierarchy of grammar formalisms \cite{chomThreeModels1956}, practical
+developments: Chomsky's hierarchy of grammars \cite{chomThreeModels1956}, practical
 algorithms for parsing of regular and context-free grammars and
@@
 generation of efficient parsers \cite{Johnsonyacc}.
+
+\steven{I don't like the name \theoryname (\theoryabbv). Feel free to play with it though, as
+  its set up as a pair of macros}
 
@@
 
+\steven{I initially wanted to say ``universal semantic domain to reason about
+  parsing'', but these words have specific domain theory usage that aren't
+  appropriate here, right?}
+
 \steven{Check details for 2-cat of grammar formalisms}
@@
 parsers may benefit from a \emph{dependence} on non-linear terms.
-Through the approach described by Krishaswami et al.\ in
+Through the approach described by Krishnaswami et al.\ in
 \cite{krishnaswami_integrating_2015},
@@
 internal to \theoryabbv.
+
+\steven{The above is meant to walk the reader into accepting LNL as a natural
+  choice, but it doesn't really flow well to the beginning of the next subsection }
 
@@
 The monoidal structure provided by $\otimes$ is both left and right closed and
-this is denoted by the left and right linear function types.
-
-\steven{Say more about the adjunctions}
-
-Further, for linear types $A$ and $B$ the left linear function type $A \lto B$
-enjoys an elimination principle similar to function application/modus ponens.
+this is denoted by the left and right linear function types. Further,
+for linear types $A$ and $B$ the left linear function type $A \lto B$
+enjoys an elimination principle similar to function application or modus ponens.
 
@@
 \cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
-types the same notion of derivative extends to the grammars of \theoryabbv.
+types, the same notion of derivative extends to the grammars of \theoryabbv.
+Note, to ensure that the linear function types do indeed generalize Brzozowski
+derivative, we must include the equalities in \cref{fig:brzozowskideriv} as axioms.
 
@@
 right-handed counterpart.
+
+\begin{figure}
+\begin{align*}
+  c\lto c &\equiv I\\
+  c\lto d &\equiv \bot\\
+  c\lto I &\equiv \bot\\
+  c\lto 0 &\equiv \bot\\
+  c\lto (A \otimes B) & \equiv (c\lto A) \otimes B + (A \amp I) \otimes (c\lto B)\\
+  c\lto A^* &\equiv (A \amp I)^{*} \otimes (c \lto A) \otimes A^* \\
+  c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
+\end{align*}
+\caption{Equality for Brzozowski Derivatives}
+\label{fig:brzozowskideriv}
+\end{figure}
 
@@
 The empty type $\bot$ has no inhabitants. The elimination for the empty type
-witnesses the principle of explosion --- i.e. from a term of type $\bot$ we may
+witnesses the principle of explosion --- i.e.\ from a term of type $\bot$ we may
 introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.
@@
 
-\paragraph{The $G^{-1}$ Modality}
-Turn nonlinear into linear
-
-\steven{Call this $F$ or $G^{-1}$?}
-
-\steven{Why can't these map any nonlinear type into linear ones? instead of just
-those that are already linear types lifted by $G$?}
-
-\steven{TODO G inverse}
-  % \item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
+\paragraph{The $G$ Modality}
+The $G$ modality provides the embedding of linear types in the empty context
+into non-linear types. The introduction and elimination forms for $G A$ obey the
+same rules as given in \cite{bentonMixedLinearNonlinear1995} and \cite{krishnaswami_integrating_2015}.
+
+The left adjoint $F$ from non-linear types back to linear types may be defined
+as,
+
+\[
+  F X := \LinSigTy a A I
+\]
+
+\steven{Elaborate on $G$ more}
+
+The semantics of $G A$ are exactly the semantics of $A$ at the empty word $\varepsilon$.
+
+\[ \sem{G A} \gamma = \sem{A} \gamma \varepsilon \]
 
@@
 \subsection{Subtheories}
+Each choice of connective has direct implications on the expressivity of
+\theoryabbv. In particular, the set of connectives used will determine where the
+grammar semantics is placed in
+the Chomsky hierarchy.
+
 \paragraph{Regular Expressions} We may realize regular expression as a subtheory
@@
 \paragraph{$\mu$-Regular Expressions} Similarly, we may restrict the connectives
-to be linear unit, characters, $\otimes$, $\oplus$, and arbitrary recursion.
+to be linear unit, characters, $\otimes$, $\oplus$, and arbitrary recursion via
+left-fixed point rather than only Kleene star.
 Instead of regular expressions these correspond to Lei{\ss}'s $\mu$-regular
 expressions, which are known to be equivalent to context-free grammars
-\cite{leiss}.
-
-\steven{Give this subtheory a name }
+\cite{leiss,krishnaswami_typed_2019}.
 
@@
 calculus, which of course induces a semantics of unrestricted grammars.
-
-\steven{TODO $\mu$ regular expressions}
 
@@
 
-\section{Nonstandard Models}
-\label{sec:nonstandard}
+\section{Other Models}
+\label{sec:othermodels}
+\subsection{Language Semantics}
+Every grammar induces a language semantics. Also languages can be taken as a
+propositionally truncated view of the syntax. Logical equivalence should induce
+weak equivalence, and thus even give a syntactic way to reason about language equivalence.
+\steven{TODO language semantics}
 
@@
 want to connect \theoryabbv to parsers, we must provide a more
-detailed account of canonicity.
+detailed account of canonicity. In particular, we give a nonstandard semantics
+of \typeabbv that carries a proof of canonicity along with it.
 
@@
 \label{sec:automata}
-\pedro{The switch from the canonicity proof to this section feels a bit ``bumpy''.
-  Since they are at least technically related, it might be a good idea to say something
-  like ``The canonicity theorem from the previous section gives a tight connection
+
+The canonicity theorem from the previous section gives a tight connection
 between terms in our type theory and the theory of formal languages. In this section
-we further explore these connections by using the fact that, clasically ...''}
-
-Classically, formal language theory is closely related to the study of automata.
+we further explore these connections by using the fact that, clasically
+formal language theory is closely related to the study of automata.
 In the Chomsky hierarchy, each language class is associated to a class of
@@
 proof term within our logic.
-
-Internalizing these equivalences is particularly useful for building a regular
-expression parser. That is, it is relatively easy for use to write down a parser
-for deterministic finite automata (DFAs). Then we may compose this parser with
-a proof equivalence between DFAs and regular expressions to extract a verified
-regular expression parser.
-
-As we will see in \cref{subsubsec:tm}, we may very quickly encode arbitrary
-recursively enumerable languages --- since we can simulate the
-tape of Turing machine in the dependent fragment of the language.We reflect
-here on this expressive power to motivate why we put stringent syntactic
-restrictions on particular classes of automata. The type theory we present truly
-is general enough to easily capture all of the Chomsky hierarchy.
-To begin, let's focus on the simplest languages and their corresponding
-automata --- regular expressions and finite automata.
-\pedro{The connection dependent types-Turing machines seems too important to only
-  be mentioned here. It should at least indirectly motivate the type theory being
-dependently typed.}
 
@@
 
-Meaning, each state has a transition for every character.
+Meaning, each state has a transition for every character. \emph{Note:} the above use
+  of a disjunction over $\Sigma$ is a bit of abuse of notation.
+  As $\Sigma$ is a finite alphabet, we wish to think of this as an iterated
+  binary sum, or perhaps as a finitely indexed sum defined via linear dependent
+  sums in the same manner by which we defined binary sums.
 
@@
 
-% \todoin{Derivation tree for this construction}
-
 Perhaps this derivation is not too surprising. All it says
@@
 system to ensure its correctness. Moreover, this construction exports to an
-intrinsically verified and executable DFA via Agda.
-
-\subsection{Regular Grammars and DFAs}
+intrinsically verified and executable decision procedure for DFAs in \thoeryabbv
+embedded in Agda.
+
+We should note that the
+construction of a DFA parser requires the addition of an additional but seemingly innocuous
+axiom.
+
+\[
+  \top \cong \left( \bigoplus_{c : \Sigma} c \right)^{*}
+\]
+
+That is, the characters in the alphabet $\Sigma$ really are all of the
+characters. The necessity of this axiom arises when we go to take a transition,
+as we must ensure that the nexts character read has a corresponding transition
+--- which would be guaranteed by determinism provided there are no surprise characters.
+\steven{The above paragraph sounds weird}
+
+\subsection{Regular Expressions and DFAs}
 \label{subsec:deriv}
@@
 a verified parser
-generator for regular grammars we need to perform some
+generator for regular expression we need to perform some
 plumbing establishing an equivalence between regular
@@
 \end{theorem}
-\steven{In Agda as a weak equivalence right now}
-
 We make a pretty straightforward adaptation of Thompson's theorem to our setting,
@@
 \begin{proof}[Proof Sketch]
-  The following proof is currently formalized in Agda. Below
-  we give a walk through the structure of the proof.
-
   Recall that regular grammars are inductively defined via
@@
   give two parse transformers, one from $I$ to
-  $\mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$ and vice versa. The formalized
-  artifact provides the low-level details that these are
-  indeed mutually inverse.
+  $\mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$ and vice versa.
 
@@
 
-  \cref{fig:disjunctionNFA} shows the process of
+  \Cref{fig:disjunctionNFA} shows the process of
   disjunctively combining NFAs. Precisely, we add a single
@@
 
-  As discussed above, the other cases and low-level details
-  of isomorphism are reserved for the formalization. This
+  When you wish to show that these actually maps invert each other to form an isomorphism, we invoke the
+  universal properties of the $\mathsf{Trace}$ types. That is,
+  $\mathsf{Trace}_{\oplus NFA}$, $\mathsf{Trace}_{N}$, and $\mathsf{Trace}_{M}$
+  are \emph{initial} algebras to the the functors that define them. Thus, the
+  embedding of traces through $N$ into $\oplus NFA$ and then mapped back down to $N$
+  compose to a map from $\mathsf{Trace}_{N}$ to itself. Such a map is unique and
+  thus must be the identity. The same reasoning shows that the desired
+  composite do indeed invert each other to form an isomorphism.
+
+  As discussed above, the other cases and
+  follow through a similar argument. This
   concludes the proof of our variant of Thompson's construction.
@@
 We encode the traces of a pushdown automaton very similarly to those of an NFA,
-except the transitions of a PDA are instead encoded via the linear-non-linear
-$\Pi$-type. This is because of simply transitioning via character, a PDA must
+except the transitions of a PDA are instead encoded via the LNL product type
+$\bigamp$. This is because when simply transitioning via character, a PDA must
 also pop and push characters onto a stack, which is used as the argument to
-these dependent functions.
+these dependent functions. Thus, the appropriate formation of traces through a
+PDA is dependent on a stack.
 
@@
 
-In
-\cref{subsubsec:pda}, we gave a grammar presentation of
+Above we gave a grammar presentation of
 traces through a PDA by using a non-linear type $S$
@@
 half of the tape the other, and we treat the tops of stacks
-like the head of the tape.
-
-Due to computability limitations, we are not able to extract
-a verified parser for all recursively enumerable languages.
-Yet we should still reflect on the generality of our syntax,
-as it allows a natural encoding of Turing machines.
+like the head of the tape. We should reflect on the generality of our syntax, as
+it allows a natural encoding of Turing machines.
 

### 2024-07-10 POPL/paper.tex
future/related work stubs

@@
 % -*- fill-column: 80; -*-
-\documentclass[sigconf,anonymous,review,screen,acmsmall]{acmart}
+\documentclass[sigconf,review,anonymous,screen,acmsmall]{acmart}
 \usepackage{mathpartir}
@@
 
-%% That paper with Fritz Henglein about regexes as types
-
-%% Vaughan Pratt on the residual operator
-%%
+\steven{Comments from Max or Pedro to mention these:
+
+That paper with Fritz Henglein about regexes as types
+
+Vaughan Pratt on the residual operator
+
+}
+
+\steven{Mention Vakar's ``Syntax and Semantics of Linear Dependent Types'' as
+  related work?}
 
@@
 \label{sec:future}
+
+\steven{Rework future work}
+\steven{Mention the two-level linear,dependent compiler. ``A two level linear
+  dependent type theory''. it could be a good place to build off for a language
+  implementation that isn't just inference rules}
 

### 2024-07-10 POPL/paper.tex
build errors

@@
 
-\pa{We should probably define some of the words in this definition}
+\pedro{We should probably define some of the words in this definition}
 \begin{definition}
@@
 detailed account of canonicity. In particular, we give a nonstandard semantics
-of \typeabbv that carries a proof of canonicity along with it.
+of \theoryabbv that carries a proof of canonicity along with it.
 
@@
   that combines canonicity arguments for dependent type theory
-  \cite{coquand,etc} with logical relations constructions for linear
-  types \cite{hylandschalk}. It is easy to see by induction that the
+  TODO cite coquand
+  % \cite{coquand,etc}
+  with logical relations constructions for linear
+  types
+  TODO cite hylandschalk
+  % \cite{hylandschalk}
+  . It is easy to see by induction that the
   logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
@@
 which we write as $\mathsf{mfold}$ --- given
-in \cref{subsubsec:other}.
+in \cref{fig:multifix}.
 
@@
 system to ensure its correctness. Moreover, this construction exports to an
-intrinsically verified and executable decision procedure for DFAs in \thoeryabbv
+intrinsically verified and executable decision procedure for DFAs in \theoryabbv
 embedded in Agda.

### 2024-07-10 POPL/paper.tex
a better(?) abstract

@@
 \newcommand{\nat}{\mathbb{N}}
-\newcommand{\theoryname}{Grammar Theoretic Type Theory\xspace}
-\newcommand{\theoryabbv}{GTTT\xspace}
+\newcommand{\theoryname}{Grammar Type Theory\xspace}
+\newcommand{\theoryabbv}{GramTT\xspace}
 
@@
 \begin{abstract}
-  In this paper we propose \theoryname~(\theoryabbv), a new syntactic formalism for formal grammars: a
-  dependent linear/non-linear type theory in which the tensor product is
-  non-commutative. The linear types in this calculus may be interpreted as
-  grammars, and linear terms as parse transformers. This provides a logical
-  characterization of grammar classes based on a substructural logic rather than
-  ordinary first-order logic.
-
-  Internal to \theoryname, we carry out many classical constructions from
-  the theory of formal grammars. Inspired by the common
-  structure underlying various ideas in formal language
-  theory, our type theory makes these relationships explicit
-  and provides a unifying syntax to reason about formal
-  grammars.
-
-  We give a semantics in the category of grammars and prove a
-  canonicity theorem showing that every term in a context corresponding to a
-  fixed string is equal to a term in a canonical form encoding a parse tree of
-  that string. Finally, we discuss the application to the design and
-  implementation of correct-by-construction parsers.
+  We propose \theoryname~(\theoryabbv) a new type-theoretic calculus
+  for the specification of formal grammars and intrinsic verification
+  of parsers. \theoryabbv is a dependent non-commutative
+  linear/non-linear type theory where the linear types can be viewed
+  as formal grammars, and linear terms can be viewed as parse
+  transformers, of which parsers are a special case. We formalize this
+  interpretation by giving a denotational semantics of the theory in a
+  novel categorical semantics of formal grammars in a category of
+  families indexed by strings.
+
+  To demonstrate the usefulness of the calculus for formal grammar
+  theory, we show that different grammar formalisms \max{which ones do
+    we actually show?} can be expressed as linear types in
+  \theoryabbv. We give simple characterizations of the regular and
+  context-free grammars as expressible by restricting which
+  connectives of \theoryabbv are allowed, a form of type theoretic
+  characterization of these grammar classes. Further, we demonstrate
+  some central proofs of formal grammar theory can be intrinsically
+  verified: \emph{strong} equivalence between regular expressions and
+  NFAs is expressed by a syntactic \emph{isomorphism} definable in
+  \theoryabbv.
+
+  Finally, since \theoryabbv is a syntactic calculus, in addition to
+  the ``standard'' semantics of linear types as grammars, we
+  demonstrate two non-standard semantic interpretations. One is a
+  semantics where linear types are instead interpreted as
+  \emph{semantic actions}, rather than grammars, which are more common
+  in application than formal grammars proper. Lastly we use a
+  non-standard semantics to prove a \emph{canonicity} metatheorem that
+  shows that the interpretation of types as grammars is canonical: the
+  type-theoretic notion of a parse agrees with the expected semantic
+  notion.
 \end{abstract}

### 2024-07-10 POPL/paper.tex
down with DOS

@@
 
-The theory of formal languages and parsing is one of the oldest and
-most thoroughly developed areas of theoretical computer science. A
-prominent topic in the 1950, 60s, and 70s led to a series of remarkable
-developments: Chomsky's hierarchy of grammars \cite{chomThreeModels1956}, practical
-algorithms for parsing of regular and context-free grammars and
-variants \cite{KNUTH1965607,Earley1970}, as well as implementations of practical tools for the
-generation of efficient parsers \cite{Johnsonyacc}.
-
-\steven{I don't like the name \theoryname (\theoryabbv). Feel free to play with it though, as
-  its set up as a pair of macros}
-
+Parsing data from a serialized format is one of the most common tasks
+in all of computing. Accordingly, this is a quite well-studied problem
+spanning theoretical computer science and linguistics with such
+milestones as Chomsky's hierarchy of grammars
+\cite{chomThreeModels1956}, practical algorithms for parsing of
+regular and context-free grammars and variants
+\cite{KNUTH1965607,Earley1970}, as well as implementations of
+practical tools for the generation of efficient parsers
+\cite{Johnsonyacc}.
+%
+
+%% The theory of formal languages and parsing is one of the oldest and
+%% most thoroughly developed areas of theoretical computer science. A
+%% prominent topic in the 1950, 60s, and 70s led to a series of remarkable
+%% developments: 
 In this paper, we propose an intuitive and structured generalization of the

### 2024-07-10 POPL/paper.tex
Recursive type + correction of regular exp over grammar

@@
   theory, we show that different grammar formalisms \max{which ones do
-    we actually show?} can be expressed as linear types in
+    we actually show?} \steven{regular expressions, $\mu$-regular expressions
+    (or context free expressions), and different automata classes (finite,
+    pushdown, turing)} can be expressed as linear types in
   \theoryabbv. We give simple characterizations of the regular and
@@
 \paragraph{Recursive Types}
-isorecursive types
-\steven{TODO recursive types}
-  % \item $\sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x)$
+Recursive linear types may be defined via the least-fixed point operator $\mu$.
+The grammar semantics of which are the fixed-point of sets induced by the
+grammar semantics of $A$.
+
+\[ \sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x) \]
 
@@
 useful for defining grammars that encode the states of an
-automaton. In \cref{fig:multifix} we provide the
+automaton, as we will see in \cref{sec:automata}. In \cref{fig:multifix} we provide the
 introduction and elimination principles for such a fixed
@@
 \begin{align*}
-  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \\
-  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
+  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
 \end{align*}
 
-\begin{figure}[h!]
+\begin{figure}
 \begin{mathpar}
@@
 
-  \\
+  \and
 
@@
   {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
-             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
-             \dots \\
-             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
+             \Gamma ; x_{j} : \simulsubst {A_{j}}{\sigma} \vdash e_{j} : B_{j}\quad \forall j
   }
@@
 \end{figure}
-
-The most prevalent application of these multi-fixed points
-is for defining the traces through automata, as we will see
-later in \cref{eq:nfatrace}.
 
@@
 
-\steven{Added this transition to handle putting the categorical semantics after
+\steven{Added the above transition paragraph to handle putting the categorical semantics after
   the syntax. I think it sounds weird though}
@@
 That is, the characters in the alphabet $\Sigma$ really are all of the
-characters. The necessity of this axiom arises when we go to take a transition,
+characters. The necessity of this axiom arises when we go to take a transition
+within the DFA,
 as we must ensure that the nexts character read has a corresponding transition
@@
 
-\subsection{Equivalence Between Regular Grammars and Finite Automata}
+\subsection{Equivalence Between Regular Expressions and Finite Automata}
 \label{subsec:eqproofs}
@@
   \label{thm:typthompson}
-  For $g$ a regular grammar $g$, there is an NFA $N$ such that $g$ is isomorphic
+  For $g$ a regular expression $g$, there is an NFA $N$ such that $g$ is isomorphic
   to $\mathsf{AccTrace}_{N}$.
@@
 \begin{proof}[Proof Sketch]
-  Recall that regular grammars are inductively defined via
+  Recall that regular expressions are inductively defined via
   disjunction, concatenation, and Kleene star over literals
-  and the empty grammar. By induction over regular grammars,
+  and the empty grammar. By induction over regular expressions,
   we will construct an NFA that is equivalent to $g$.
@@
 half of the tape the other, and we treat the tops of stacks
-like the head of the tape. We should reflect on the generality of our syntax, as
-it allows a natural encoding of Turing machines.
+like the head of the tape.
 

### 2024-07-10 POPL/paper.tex
a very small amount of intro

@@
 %
+The final frontier for parsing research is the development of
+\emph{formally verified} parser implementations, which due to the
+pervasive nature of parsing would be a useful component of many
+formally verified software systems: compilers and servers most
+especially.
+%
+
+In this work we aim to refine formal language theory into a
+mathematical theory of formal grammars that can be used to develop
+\emph{correct-by-construction} parsers. Our approach can be summarized
+in the Curry-Howard-style slogan ``grammars as types'', and the
+addendum ``parsers as terms'' where since the types of the terms are
+grammars, this means not just parsers but intrinsically verified
+parsers as terms. That is, we develop a type theory with a novel
+denotational semantics in a category with abstract formal grammars as
+objects and \emph{parse transformers} as morphisms. We say these are
+\emph{abstract} formal grammars because they are not tied to any
+particular syntactic formalism such as Chomskyian grammars, regular
+expressions, Lambek calculus etc. The category we choose is a rich
+category: it is a topos as well as being monoidal closed, and these
+universal properties serve as the inspiration for our type theory.
+
+The type theory itself is a form of dependent linear-non-linear type
+theory: the linear types can be interpreted as grammars, whereas the
+non-linear types can be interpreted as ordinary sets. As in prior work
+on dependent linear-non-linear types, the linear types are allowed to
+depend on the non-linear types but not vice-versa \cite{dep-lnl}.\max{had to stop here to take care of baby}
 

### 2024-07-10 POPL/paper.tex
Added comments and modified some paragraphs

@@
 \cite{chomThreeModels1956}, practical algorithms for parsing of
-regular and context-free grammars and variants
+regular and context-free grammars, and variants
 \cite{KNUTH1965607,Earley1970}, as well as implementations of
@@
 on dependent linear-non-linear types, the linear types are allowed to
-depend on the non-linear types but not vice-versa \cite{dep-lnl}.\max{had to stop here to take care of baby}
+depend on the non-linear types but not vice-versa \cite{dep-lnl}.
+
+The substructural nature of \theoryabbv is well-aligned with the
+requirements intrinsic to parsing and to the theory of formal
+languages, where strings constitute a very clear notion of resource
+that cannot be duplicated, reordered or dropped.
 
@@
   something that ties the first paragraph to the contributions that follow}
+\pedro{As a rule of thumb, I think that it's good to avoid subjective and
+  pointed adjectives such as ``intuitive'', espcially if we don't have evidence
+  that corroborates our claim. This is the perfect excuse for
+an annoying reviewer to complain about our work.}
 
@@
 models of our syntax. Afterwards, we demonstrate several models of \theoryabbv,
-include one to prove a metatheoretic canonicity theorem. Finally, in
+including one to prove a metatheoretic canonicity theorem. Finally, in
 \cref{sec:automata} we give a presentation of automata as grammars. With this
@@
   appropriate here, right?}
-
+\pedro{If you want to keep ``semantic domain'', we could go for ``all-encompassing semantic domain''}
 \steven{Check details for 2-cat of grammar formalisms}
@@
 become unnecessarily verbose when using Girard encodings to translate
-high-level reasoning into low-level linear terms. To alleviate this
+high-level reasoning into low-level linear terms.\pedro{What encodings are these?
+We should add a citation here}
+To alleviate this
 concern, in 1995 Benton et al.\ proposed an alternative categorical
@@
 
+In this section we go through a tour of \theoryabbv. We start by presenting its syntax.
+In order to avoid a notion of raw-term, we present it in an intrinsically-typed fashion,
+where only well-typed terms are meaningful. Next, we show how it admits a model in $\Set^{\String}$
+where the linear types have natural interpretation in terms of formal languages --- for instance,
+the tensor corresponds to grammar concatenation. We conclude the section by showing how
+classes of grammars correspond to certain subtype-theories inside \theoryabbv.
+
+\pedro{Hopefully this serves as a segue between this and the syntax subsection}
 \subsection{Syntax}
@@
 to serve as a deductive
-setting for formal grammar theory. \theoryabbv axiomatizes the necessary structure underlying $\Set^{\String}$ to
+setting for formal grammar theory.
+\theoryabbv axiomatizes the necessary structure underlying $\Set^{\String}$ \pedro{We should probably stick to either $\Set^{\String}$ or $\mathbf{Gr}$ from now on}to
 specify and construct parsers.
@@
 The structural judgments are shown in \cref{fig:structjdg}, the typing
-well-formedness in \cref{fig:typewf}, and the intuitionistic typing rules in
-\cref{fig:inttyping}. These are all just as they appear in
-\cite{krishnaswami_integrating_2015}. The treatment of the non-linear types is
-pretty standard, so below we focus on the linear syntax.
+well-formedness in \cref{fig:typewf}, and the intuitionistic typing
+rules in \cref{fig:inttyping}. These are mostly just as they appear in
+\cite{krishnaswami_integrating_2015}. It has two main differences. The
+first one it the presence of two-distinct arrow types, one for adding
+variables to the beginning of contexts and another one for adding
+variables to the end of contexts. This is an adequate change in the
+context of grammars, which are inherently non-symmetric. The second
+change is the introduction of inductive types, which also have a
+natural grammar counterpart, which usually allows kinds of recursive
+behavior such as the Kleene star. The treatment of the non-linear
+types is pretty standard, so below we focus on the linear syntax.
 
@@
 \subsection{Formal Grammars as Linear Types}
+\label{sec:formaltype}
+
 The linear typing judgment in our syntax takes on the following schematic form
@@
 \emph{linear context} dependent on $\Gamma$. These linear contexts behave
-substructurally. As stated earlier, they do are linear --- they do not obey
+substructurally. As stated earlier, they are linear --- they do not obey
 weakening or contraction --- because a character is exhausted once read by a
@@
 
-Most of the non-linear semantics is standard, so let us investigate in more
-depth the linear semantics. The interpretation of each
-connective is exactly what is suggested by the syntax, ude in
-large part to the inspiration taken from the grammar interpretation
-when designing \theoryabbv.
+\pedro{The paragraphs above are a bit circular, what about the following alternative?}
+
+Now that we have specified the syntactic aspect of \theoryabbv, we must justify
+its connections to formal grammars in order to show that it is a sound methodology
+for reasoning about parsers. We do this by showing that $\mathbf{Gr}$ is a model
+to our type theory.
+
+Most of the non-linear semantics is standard \pedro{We should either cite something or add the non-linear semantics in the appendix}, so let us investigate in more
+depth the linear semantics.
 
@@
   \item A non-linear term $\Gamma \vdash e : X$ denotes a section
-        $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$
+        $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$ \pedro{we have not defined what sections are}
   \item Linear contexts $\Gamma \vdash \Delta$ and types
@@
 As a grammar,
-
+%
 \[
@@
 \]
-
+%
 That is, $I$ maps strings to the proposition that they are the empty string. $I$
@@
 The grammar interpretation for characters is quite similar to that of $I$.
+%
 \[
@@
 \]
-
+%
 The grammar for the character $c$ likewise maps strings $w$ to the proposition that
@@
 linear types $A$, the following equalities hold:
-
+%
 \[ I \otimes A \equiv A \otimes I \equiv A \]
-
+%
 In the grammar semantics,
-
+%
 \[
@@
 \]
-
+%
 The string $w$ matches $A \otimes B$
@@
 enjoys an elimination principle similar to function application or modus ponens.
-
+%
 \[
@@
 \]
-
+%
 The interpretation of linear function types at a string $w$ is a linear function
@@
 string $w_{a} w$.
-
+%
 \[
@@
 \]
-
+%
 That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
@@
 derivative, we must include the equalities in \cref{fig:brzozowskideriv} as axioms.
+\pedro{We should further explain why these equations are required, or at the very
+least give some intuition}
 
@@
 semantics of $X$. Likewise, the grammar semantics of the linear dependent pair type is a dependent
-pair in $\Set$.
+pair in $\Set$.\pedro{The connection to grammar semantics needs to be better explained. For instance,
+we should explain that disjunction corresponds to ``or'' of grammars}
 
@@
 particular, via a dependence on $\Bool$.
-
+%
 \[
@@
 The universal type $\top$ may be formed in any context.
-
+%
 \[ \sem{\top} \gamma w = \{ \ast \}\]
-
+%
 Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.
@@
 introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.
-
+%
 \[ \sem {\bot} \gamma w = \emptyset \]
-
+%
 \paragraph{The $G$ Modality}
@@
 as,
-
+%
 \[
@@
 \]
-
+%
 \steven{Elaborate on $G$ more}
 
+\pedro{Agreed :) maybe mention connections to persistent propositions from the
+separation logic literature}
+
 The semantics of $G A$ are exactly the semantics of $A$ at the empty word $\varepsilon$.
-
+%
 \[ \sem{G A} \gamma = \sem{A} \gamma \varepsilon \]
-
+%
+%
 \paragraph{Recursive Types}
@@
 grammar semantics of $A$.
-
+%
 \[ \sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x) \]
+%
+\pedro{We should justify, even if briefly, the existence of this
+  fixed-point}
 
@@
 
+\pedro{This is an extremely important point! We should definitely make
+  a reference to the Kleene algebra world, where these things do not
+  hold. I think that it might even be a good idea to add the
+  ``underlying trick'' to the main body of text. It's important to
+  connect our work to existing ideas and concerns from the theory of
+  formal languages.}
+
 \begin{equation}
@@
 
-Under the Curry-Howard-Lambek correspondence, the programs of the simply-typed
-$\lambda$-calculus may be viewed under any one of the three equivalent views: as
-terms in a simple functional language, as proofs in an intuitionisitc
-propositional logic, or as morphisms in a Cartesian closed category. In the
-same vein, with the syntax provided, we aim to give a
-categorical accounting of the semantics of \theoryabbv. Let us begin by paring back
-the connectives in the language and only considering the subtheory induced by
-regular expressions.
-
-\steven{Added the above transition paragraph to handle putting the categorical semantics after
-  the syntax. I think it sounds weird though}
-
-By giving algebraic laws of manipulation and equivalence to regular expressions,
-Kleene algebras are an important tool in the theory of regular
-languages. More broadly, through various extensions, they serve as a
+An important part in the study of type theories is characterizing what
+categorical structures they are modeling. In the case of \theoryabbv,
+due to its intended connection to formal language theory, we use
+Kleene algebras as a starting point in its categorical
+semantics. Being posets, there is an immediate way of categorifying
+them.
+
+Kleene algebras are an important tool in the theory of regular languages
+due to them being a bridge between algebraic reasoning and equivalence
+of regular expressions. More broadly, through various extensions, they serve as a
 theoretical substrate to studying different kinds of formal languages.
@@
 formal languages, this order structure can be used to model language
-containment. In this section, we want to categorify the concept of
-Kleene algebra and build on top of it in order to define an abstract
-theory of parsing.  We start by defining \emph{Kleene categories}.
+containment. In this section, we categorify the concept of Kleene
+algebra and build on top of it in order to define an abstract theory
+of parsing. We start by defining \emph{Kleene categories}.
 
@@
   Every Kleene algebra, seen a posetal category, is a Kleene category.
+  The product $\cdot$ is a monoidal product and the addition is a
+  least-upper bound, which corresponds to a coproduct. Lastly, the
+  axioms of the Kleene star have a direct correspondence to the
+  coherence conditions postulated by the initial algebras of Kleene
+  categories.
 \end{example}
 
-This example illustrates how restricted Kleene algebras are in terms
-of reasoning about languages. By only having at most one morphism
-between objects, there is not a lot of information they can convey. In
-this case, the only information you get is language containment. We
-will see throughout this paper how the extra degrees of freedom
-granted by having more morphisms give you more algebraic structure for
-reasoning about languages.
+This example provides a neat categorical justification to how
+restrictive Kleene algebras are in terms of reasoning about
+languages. By only having at most one morphism between objects, there
+is not a lot of information they can convey. In this case, the only
+information you get is language containment. As demonstrated by
+$\mathbf{Gr}$, the extra degrees of freedom granted by having more
+morphisms gives you more algebraic structure for reasoning about
+languages.
 
@@
 
-The proposed axioms are a direct translation of the Kleene algebra
-axioms to a categorical setting. Its most unusual aspect is the
+As we have seen, the proposed axioms are a direct translation of the
+Kleene algebra axioms to a categorical setting. Its most unusual aspect is the
 axiomatization of the Kleene star as a family of initial algebras
@@
   We prove this by the unicity (up-to isomorphism) of initial
-  algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
-  (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
-  X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
-  (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
-    1}(X))$.
+  algebras. Let $[hd, tl]: I + (\mu X.\, F_{A, I}(X)) \otimes A \to
+  (\mu X.\, F_{A, I}(X))$ be the initial algebra structure of $(\mu
+  X.\, F_{A, I}(X))$ and consider the map $[hd, tl] : B + B \otimes
+  (\mu X.\, F_{A, I}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
+    I}(X))$.
 
   Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
-  we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,1} \to B \lto Y$. We can show existence and
+  we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,I} \to B \lto Y$. We can show existence and
   uniqueness by showing that the diagram on top commutes if, and
@@
 \[\begin{tikzcd}
-	{B + B \otimes (\mu X.\, 1 + X \otimes A)} && {B + Y \otimes A} \\
-	\\
-	{B \otimes (\mu X.\, 1 + X \otimes A)} && Y \\
-	{1 + (\mu X.\, 1 + X \otimes A)} && {1 + ( B \lto Y) \otimes A} \\
-	\\
-	{\mu X.\, 1 + X \otimes A} && {B \lto Y}
+	{B + B \otimes (\mu X.\, I + X \otimes A)} && {B + Y \otimes A} \\
+	\\
+	{B \otimes (\mu X.\, I + X \otimes A)} && Y \\
+	{I + (\mu X.\, I + X \otimes A)} && {I + ( B \lto Y) \otimes A} \\
+	\\
+	{\mu X.\, I + X \otimes A} && {B \lto Y}
 	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
@@
   by the monoidal closed structure of $\cat{K}$. A completely analogous
-  argument for $G_{A,B}$ also holds.
+  argument for $G_{A,B}$ also holds. Furthermore, by generalizing the
+  construction of \Cref{sec:formaltype}, we can also show that from
+  the monoidal closed assumption it follows that $\mu X.\, F_{A, I}(X) \cong \mu X.\, G_{A, I}(X)$
 \end{proof}
@@
 
-For the rest of the paper we will use the presheaf category
-$\Set^{\cat{\Sigma^*}}$ as a concrete model that will serve as our
-guide when expanding abstract formalism presented in this section so
-that it can handle more classes of languages beyond regular ones.
-
 \subsection{Beyond Simple Types}
@@
 
+\pedro{I agree, then we should add the quick proof :) }
+
 \section{Other Models}
 \label{sec:othermodels}
+
+One of the powers of type theories is that it can be profitable to interpret them
+in various models. In this section, by using our just-defined Chomsky categories,
+we show how other useful concepts from formal language theory can also be organized
+as models of \theoryabbv. We illustrate this point by providing two examples that
+are closely related to the theory of formal languages: language equivalence and
+semantic actions. Furthermore, in order to justify how $\mathbf{Gr}$ relates to
+more traditional notions of parsing, we define a glued model that proves a
+canonicity property of grammar terms.
+
 \subsection{Language Semantics}
@@
 \subsection{Semantic Actions}
-Returning to the problem of parsing, the output of a parse usually isn't the
+Returning to the problem of parsing, the output of a parse usually is not the
 literal parse tree. Rather, the output is the result of some \emph{semantic
@@
 \steven{TODO semantic actions}
+
+\pedro{This is a very nice opportunity of showing off the supremacy of denotational
+  reasoning ;) We should probably prove the gluing lemma in the previous section
+  and apply it here and in the canonicity section. The actual proof might have to be
+  moved to the appendix, though}
 

### 2024-07-11 POPL/paper.tex
some comments

@@
 an annoying reviewer to complain about our work.}
+\max{I agree about ``intuitive'', try to stick to objective claims}
 
@@
 of $w$. The induced notion of isomorphism in $\Grammar$ encodes that two grammars are equivalent if there is
-a bijective translation of parses. In other words, two grammars are isomorhpic
+a bijective translation of parses. In other words, two grammars are isomorphic
 if they capture the same structural descriptions of strings, which is
@@
 
-Upon inspection of the above definition, $\Grammar$ may equivalently be described as the functor
-category $\Set^{\String}$ --- where the set $\String$ is viewed as a
-discrete\footnote{Objects in $\String$ comprise the set of all strings, and the
-  only morphisms are the identity morphisms.}
-category. In fact, in this paper we will take $\Grammar$ to be defined as the
-functor category $\Set^{\String}$. Such functor categories into $\Set$ --- often called
-presheaf categories --- carry a remarkable amount of structure.
+Upon inspection of the above definition, $\Grammar$ may equivalently
+be described as the functor category $\Set^{\String}$ --- where the
+set $\String$ is viewed as a discrete\footnote{Objects in $\String$
+comprise the set of all strings, and the only morphisms are the
+identity morphisms.}  category. In fact, in this paper we will take
+$\Grammar$ to be defined as the functor category
+$\Set^{\String}$. Such functor categories into $\Set$ --- often called
+presheaf categories --- carry a remarkable amount of structure, and we
+will demonstrate that these well-known constructions on presheaf
+categories provide us with familiar compositional constructions on
+grammars\max{where do we talk about the properties of this presheaf cat?}.
 
@@
 \steven{Check details for 2-cat of grammar formalisms}
+\max{this section kind of trails off at the end with a very vague aside about a $2$-category of grammar formalisms.
+Maybe this should go in future work/discussion?
+What is the message of this section?
+How does it tie in with the message of the paper?
+What even is the message of this paper?}
 
@@
 linear logic \cite{GIRARD19871}, offers precise control over how a value is used
-in a derivation. Namely, linear logic omits the weakening and contraction rules
+in a derivation. Namely\max{namely?}, linear logic omits the weakening and contraction rules
 to ensure that every value in context is used exactly once. This control enables
@@
 parse-time. That is, once a particular substring has been parsed, a parser need not go
-over it again. One may then envision a linear type system where the types comprise
+over it again\max{what about a backtracking parser?}. One may then envision a linear type system where the types comprise
 formal grammars generated over some alphabet $\Sigma$, and the type constructors
@@
 Unfortunately, programming in a linear term language is often
-cumbersome and unintuitive for users. Code in such a language can
+cumbersome and unintuitive for users.\max{You are arguing against our own type theory here...} Code in such a language can
 become unnecessarily verbose when using Girard encodings to translate
@@
 change is the introduction of inductive types, which also have a
-natural grammar counterpart, which usually allows kinds of recursive
+natural grammar counterpart\max{what is it then?}, which usually allows kinds of recursive
 behavior such as the Kleene star. The treatment of the non-linear
-types is pretty standard, so below we focus on the linear syntax.
+types is pretty\max{pretty?} standard, so below we focus on the linear syntax.
 
@@
 Note, to ensure that the linear function types do indeed generalize Brzozowski
-derivative, we must include the equalities in \cref{fig:brzozowskideriv} as axioms.
+derivative, we must include the equalities in \cref{fig:brzozowskideriv} as axioms\max{we must? why must we}.
 \pedro{We should further explain why these equations are required, or at the very

### 2024-07-11 POPL/paper.tex
first pass at discussion

@@
   To demonstrate the usefulness of the calculus for formal grammar
-  theory, we show that different grammar formalisms \max{which ones do
+  theory, we show that existing grammar formalisms --- e.g.\ regular
+  expressions, context free expressions, and their associated automata \max{which ones do
     we actually show?} \steven{regular expressions, $\mu$-regular expressions
     (or context free expressions), and different automata classes (finite,
-    pushdown, turing)} can be expressed as linear types in
+    pushdown, turing)} --- can be expressed as linear types in
   \theoryabbv. We give simple characterizations of the regular and
@@
 %
+\steven{Some ideas on what the unifying message throughout the paper should be:
+
+}
 
@@
 on dependent linear-non-linear types, the linear types are allowed to
-depend on the non-linear types but not vice-versa \cite{dep-lnl}.
+depend on the non-linear types but not vice-versa \cite{krishnaswami_integrating_2015}.
 
@@
 %% developments: 
-In this paper, we propose an intuitive and structured generalization of the
+In this paper, we propose a structured generalization of the
 theoretical machinery used for parsing. In conjunction with an updated
@@
   something that ties the first paragraph to the contributions that follow}
-\pedro{As a rule of thumb, I think that it's good to avoid subjective and
-  pointed adjectives such as ``intuitive'', espcially if we don't have evidence
-  that corroborates our claim. This is the perfect excuse for
-an annoying reviewer to complain about our work.}
-\max{I agree about ``intuitive'', try to stick to objective claims}
 
@@
   grammar} --- a category paired with a functor into
-$\Grammar$. That is, they may all be interpreted in our category of grammars. In this sense, $\Grammar$ provides a universal setting to reason about parsing.
-
-\steven{I initially wanted to say ``universal semantic domain to reason about
-  parsing'', but these words have specific domain theory usage that aren't
-  appropriate here, right?}
-\pedro{If you want to keep ``semantic domain'', we could go for ``all-encompassing semantic domain''}
-\steven{Check details for 2-cat of grammar formalisms}
+$\Grammar$. That is, they may all be interpreted in our category of grammars. In
+this sense, $\Grammar$ provides an
+all-encompassing semantic domain for parsing.
+
 \max{this section kind of trails off at the end with a very vague aside about a $2$-category of grammar formalisms.
@@
 linear logic \cite{GIRARD19871}, offers precise control over how a value is used
-in a derivation. Namely\max{namely?}, linear logic omits the weakening and contraction rules
+in a derivation. Linear logic omits the weakening and contraction rules
 to ensure that every value in context is used exactly once. This control enables
@@
 as we may treat characters of a string as finite resources that are consumed at
-parse-time. That is, once a particular substring has been parsed, a parser need not go
-over it again\max{what about a backtracking parser?}. One may then envision a linear type system where the types comprise
+parse-time.
+That is, when reading a string, the occurrence of any character is read once.
+Freely duplicating or dropping characters from a string changes the meaning of that string.
+\max{what about a backtracking parser?} \steven{Does the sentence sound better now?
+I've tried to avoid suggesting any particular parser design.
+Also, I think backtracking is a very interesting question to talk about in future work}.
+One may then envision a linear type system where the types comprise
 formal grammars generated over some alphabet $\Sigma$, and the type constructors
@@
 Unfortunately, programming in a linear term language is often
-cumbersome and unintuitive for users.\max{You are arguing against our own type theory here...} Code in such a language can
-become unnecessarily verbose when using Girard encodings to translate
+cumbersome and unintuitive for users who are unacustomed to the usage
+restrictions.\max{You are arguing against our own type theory here...} \steven{I
+  want to argue against a \emph{purely linear} calculus to motivate the
+  convience of LNL, but perhaps this isn't appropriate here, and I don't want to
+  make our own type theory sound bad.Should probably just cut the beginning of
+  this paragraph and jump into LNL, but then there's not an immediate reason for
+why one might want/need it over a pure linear system} Code in such a language can
+become unnecessarily verbose when ensuring every function written is indeed linear. It would be more ergonomic to allow users using Girard encodings to translate
 high-level reasoning into low-level linear terms.\pedro{What encodings are these?
@@
 to serve as a deductive
-setting for formal grammar theory.
-\theoryabbv axiomatizes the necessary structure underlying $\Set^{\String}$ \pedro{We should probably stick to either $\Set^{\String}$ or $\mathbf{Gr}$ from now on}to
+setting for formal grammar theory. \theoryabbv axiomatizes the necessary
+structure underlying $\Grammar$ to
 specify and construct parsers.
@@
 context of grammars, which are inherently non-symmetric. The second
-change is the introduction of inductive types, which also have a
-natural grammar counterpart\max{what is it then?}, which usually allows kinds of recursive
-behavior such as the Kleene star. The treatment of the non-linear
-types is pretty\max{pretty?} standard, so below we focus on the linear syntax.
+change is the introduction of inductive types, which allows for the definition
+of recursive grammars. The treatment of the non-linear
+types is standard, so below we focus on the linear syntax.
 
@@
 %
-    \\
+    \and
 %
@@
 
-    \\
+    \and
     \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
@@
   %
-  \\\
+  \and
   %
@@
   %
-  \\
+  \and
 %
@@
   %
-  \\
+  \and
   %
@@
     %
-    \\
+    \and
     %
@@
 We give the base types and type constructors for linear terms. As the
-interpretation of types as grammars in $\Set^{\String}$ serves as our intended
+interpretation of types as grammars in $\Grammar$ serves as our intended
 semantics, we simultaneously give the interpretation $\sem \cdot$ of the
@@
 \pedro{The paragraphs above are a bit circular, what about the following alternative?}
+\steven{I am in favor of cutting what's above, but we haven't introduced the
+  sytnax in the main body of the paper yet, even if the figures appeared
+  earlier. So what you write needs to take that into account}
 
@@
 
-  \and
+  \\
 
@@
 $I \oplus (X \otimes g)$.
-
-\pedro{This is an extremely important point! We should definitely make
-  a reference to the Kleene algebra world, where these things do not
-  hold. I think that it might even be a good idea to add the
-  ``underlying trick'' to the main body of text. It's important to
-  connect our work to existing ideas and concerns from the theory of
-  formal languages.}
+This is a marked point of difference from Kleene
+algebra with recursion, where the fixed points for the left and right variants
+of Kleene star need not agree \cite{leiss}.
 
@@
 may define a left fold over lists in terms of a right fold
-in a functional programming language\footnote{The
-  underlying trick is to fold over a list of functions
+in a functional programming language.
+The underlying trick is to fold over a list of linear functions
 instead of the original string. We curry each character $c$
@@
 application is left-associative, this results in a left
-  fold over the original string. }.
+fold over the original string.
 
@@
 That is, the characters in the alphabet $\Sigma$ really are all of the
-characters. The necessity of this axiom arises when we go to take a transition
+characters. The need for this axiom arises when we go to take a transition
 within the DFA,
-as we must ensure that the nexts character read has a corresponding transition
+as we must ensure that the next character of the string has a corresponding
+transition from the current state
 --- which would be guaranteed by determinism provided there are no surprise characters.
-\steven{The above paragraph sounds weird}
 
@@
 appended to $\mathsf{tail}(L)$ and then we recursively step into another state
-called on argument $w + \mathsf{tail}(L)$.
+called on argument $w \cdot \mathsf{tail}(L)$.
 
@@
 
-\section{Related work}
-\label{sec:related}
+\section{Discussion and Future Work}
+\label{sec:discussion}
+
+The core objects of study in the theory of formal grammars have stood stalwart
+for decades. Study of interesting extensions or restrictions of particular
+grammar formalisms has remained consistent line of research over the years.
+Occasionally, researchers will either explicitly create a type system for their
+formalism, or researchers will leverage the same underlying structure of their
+systems. In this paper, we have not
+done much \emph{per se}, other than gather these ideas, abstract over their
+commonalities, and give the common language between formalisms.
+
+\paragraph{Typed Approaches to Grammars}
+The usage of a simple type system to reason about regular expressions was
+introduced by Frisch and Cardelli in 2004 \cite{frischCardelli}, and later expanded
+by Hengelin and Nielsen in 2011 \cite{henglein_regular_2011} to handle proofs of regular expression
+containment.
+
+In 1992, while investigating the addition of arbitrary recursion to Kleene
+algebra \cite{leiss},
+Lei{\ss} used a least fixed-point operator instead of the Kleene star. Although
+he did not explicitly make use of a type system, or have access to the view of Frisch and
+Cardelli, Leiss had added the full power of inductive types to a type system of regular expressions.
+
+Pratt's work on the residuation of action algebras in \cite{prattActionLogicPure1991} closely mirrors
+our observation that Brzozowski derivatives mirror linear function types.
+Moreover, he shows that his residuals form a Galois connection with sequential
+composition. This is precisely a posetal --- or \emph{thin} --- version of the
+adjunction between
+linear function types and tensor product in \theoryabbv.
+
+In \cite{firsovCertifiedNormalizationContextFree2015}, Firsov and Uustalu
+constructed a verified normalization procedure to turn a
+context-free grammar into an equivalent grammar in Chomsky normal form.
+Combining their work with a CYK parser, they have built a verified context-free
+grammar parser, \emph{up to weak equivalence}. To upgrade from a weak
+equivalence to a strong equivalence, they suggest updating their grammar
+formalism to treat parse trees as first-class proofs of language membership.
+That is, without explicitly stating so, they suggest that they next natural step
+is to reason about grammars in a system where parse trees are terms, as we have
+given in this paper.
+
+\steven{Make the above paragraph shorter. Also idk if its worth including, but I
+like that they practically ask for a type theory}
+
+\steven{Mention Neel's typed algebraic approch paper here. Say it is similar at
+  first glance, as its parsing with types, but that he actually is representing
+  something quite different. Or is it so different that its not worth mentioning?}
 
 \paragraph{Kleene Algebra}
-
 Since the early works in the theory of formal languages, Kleene
@@
 
-\paragraph{TODO: more stuff}
-
-\steven{Comments from Max or Pedro to mention these:
-
-That paper with Fritz Henglein about regexes as types
-
-Vaughan Pratt on the residual operator
-
-}
-
-\steven{Mention Vakar's ``Syntax and Semantics of Linear Dependent Types'' as
-  related work?}
-
-\section{Future Work}
-\label{sec:future}
-
-\steven{Rework future work}
-\steven{Mention the two-level linear,dependent compiler. ``A two level linear
-  dependent type theory''. it could be a good place to build off for a language
-  implementation that isn't just inference rules}
-
-\subsection{Implementation for Context-Free Grammars}
-As suggested throughout the paper and briefly explored in
-\cref{subsubsec:pda}, the first extension to the work in
-this paper will be to bring analogous constructions to
-context-free grammars and their accompanying pushdown automata.
-
-\subsection{Beyond Strings}
-\label{subsec:beyond}
-
+\steven{Can we sketch here what an apt categorification of KATs might be? As a
+  poor first pass: perhaps a Kleene category with a full involutive
+  subcategory that corresponds to the Boolean aglebra?.}
+\steven{What can we say about internalizing Kleene algebra proofs or extensions?}
+\steven{any KA citations needed above?}
+
+\paragraph{Linearity with Dependent Types}
+\steven{I don't know if it worth discussing the different influences of
+  categorical model here? We take Neel's model and add to it as necessary}
+
+Vakar might be worth citing to discuss different characterizations of the
+notion of model \cite{vakarSyntaxSemanticsLinear2015}
+
+\cite{fu2023twolevellineardependenttype} give a compiler for a two-level
+linear dependent type theory. This could serve as a template for a
+compiler implementation of \theoryname that is not just inference rules
+embedded in agda.
+
+\paragraph{Implementations in Parsing}
+Talk about how we have a DFA parser in agda and want to to get out a regex
+parser. Then context free, etc
+
+future work related to building a usable correct-by-construction parser
+combinator/semantic action combinator library in this language.
+
+Contrast this
+potential library with existing verified parsers. Citations for some in prelim presentation
+
+\paragraph{Applications to Type Checking}
 While parsing typically refers to the generation of semantic
@@
 to $\Set$, we envision adapting the same philosophy
-to functors from $\String$ to \emph{trees} to craft a syntax
+to functors from $\String$ to some \emph{category of trees} to craft a syntax
 that natively captures typed elaboration. This suggests an
@@
 
-Thus far, our theory has proved useful for internalizing long-standing
-grammar-theoretic constructions, but there has been decades of research
-conducted since. A fruitful avenue for future work includes
-testing if our formalism can also internalize more recently proposed grammar
-mechanisms, such as the interval parsing grammars given by Zhang et al. \cite{zhangIntervalParsingGrammars2023}.
-
-\newpage
+\steven{condense the future work section?}
+\steven{Put the 2 category stuff here?}
+
+\paragraph{Other Future Work}
+Backtracking parsing via delimited continuations?
 

### 2024-07-11 POPL/paper.tex
Some ideas on themes of paper + spelling

@@
     (or context free expressions), and different automata classes (finite,
-    pushdown, turing)} --- can be expressed as linear types in
+    pushdown, Turing)} --- can be expressed as linear types in
   \theoryabbv. We give simple characterizations of the regular and
@@
 \steven{Some ideas on what the unifying message throughout the paper should be:
-
+we offer an abstraction over the formalisms used ad-hoc when doing formal
+language/grammar theory. Whether a user of these constructs recognizes it or not, there is some
+universality between all of their systems. We don't necessarily do much, per se, other
+than abstract over the commonalities and provide a common language.
+
+However, just because there isn't a ton of clever work in designing this syntax, as
+a lot of it does flow very naturally from the literature, that doesn't diminish
+the value of the syntax. In fact, it emboldens the syntax because we want to
+appeal to the idea that ``there couldn't have been any other choice for
+universal grammar''.
+
+That is all to say, our appeal is generalization. I think does appear
+throughout the paper, but is maybe too aimless of a message to be satisfying.
+Perhaps, we should also lean on applications more. Idk I'm gonna sleep now, and
+we'll talk in the morning
 }
@@
 \emph{abstract} formal grammars because they are not tied to any
-particular syntactic formalism such as Chomskyian grammars, regular
+particular syntactic formalism such as Chomskyan grammars, regular
 expressions, Lambek calculus etc. The category we choose is a rich
@@
 giving a universal semantic characterization of formal grammar.
-From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type thoery
+From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type theory
 that axiomatizes the underlying structures needed to conduct formal grammar
@@
 Unfortunately, programming in a linear term language is often
-cumbersome and unintuitive for users who are unacustomed to the usage
+cumbersome and unintuitive for users who are unaccustomed to the usage
 restrictions.\max{You are arguing against our own type theory here...} \steven{I
   want to argue against a \emph{purely linear} calculus to motivate the
-  convience of LNL, but perhaps this isn't appropriate here, and I don't want to
+  convenience of LNL, but perhaps this isn't appropriate here, and I don't want to
   make our own type theory sound bad.Should probably just cut the beginning of
@@
 \steven{I am in favor of cutting what's above, but we haven't introduced the
-  sytnax in the main body of the paper yet, even if the figures appeared
+  syntax in the main body of the paper yet, even if the figures appeared
   earlier. So what you write needs to take that into account}
@@
 languages recognized by these are exactly the regular languages --- the lowest
-on the Chomksy hierarchy.
+on the Chomsky hierarchy.
 
@@
 simply-typed nature makes them not so expressive from a type theoretic
-point of view. This is limiting because type theories are sucessful
+point of view. This is limiting because type theories are successful
 syntactic frameworks for manipulating complicated categorical
@@
 
-\steven{Max suggests augmenting the definition of a chomsky category to
-  something like two categories $L$ and $U$, and $U$ is cartesian closed I don't
+\steven{Max suggests augmenting the definition of a Chomsky category to
+  something like two categories $L$ and $U$, and $U$ is Cartesian closed I don't
   fully recall the rest.
@@
 
-\pedro{This is likely true, but if we explicitly say so, this warants a proof. I think that
+\pedro{This is likely true, but if we explicitly say so, this warrants a proof. I think that
 if we don't say anything about the syntactic category, reviewers won't mind.}
@@
   case that we have the right categorical model of the syntax if this statement
-  isn't true. By restating our definition of Chomksy category, this should be
+  isn't true. By restating our definition of Chomsky category, this should be
   obvious or a quick proof}
@@
 discrete presheaf. That is, for a set $X$, $\Delta (X)(w) = X$ for all
-$w \in \String$. As $\SemAct$ is defined as a comma category, it has a forgetul
+$w \in \String$. As $\SemAct$ is defined as a comma category, it has a forgetful
 functor into $\Grammar$. That is, $\SemAct$ serves as a notion of formal
@@
 between terms in our type theory and the theory of formal languages. In this section
-we further explore these connections by using the fact that, clasically
+we further explore these connections by using the fact that, classically
 formal language theory is closely related to the study of automata.
@@
 he did not explicitly make use of a type system, or have access to the view of Frisch and
-Cardelli, Leiss had added the full power of inductive types to a type system of regular expressions.
+Cardelli, Lei{\ss} had added the full power of inductive types to a type system of regular expressions.
 
@@
 
-\steven{Mention Neel's typed algebraic approch paper here. Say it is similar at
+\steven{Mention Neel's typed algebraic approach paper here. Say it is similar at
   first glance, as its parsing with types, but that he actually is representing
@@
   poor first pass: perhaps a Kleene category with a full involutive
-  subcategory that corresponds to the Boolean aglebra?.}
+  subcategory that corresponds to the Boolean algebra?.}
 \steven{What can we say about internalizing Kleene algebra proofs or extensions?}
@@
 compiler implementation of \theoryname that is not just inference rules
-embedded in agda.
+embedded in Agda.
 
 \paragraph{Implementations in Parsing}
-Talk about how we have a DFA parser in agda and want to to get out a regex
+Talk about how we have a DFA parser in Agda and want to to get out a regex
 parser. Then context free, etc
@@
 \paragraph{Other Future Work}
-Backtracking parsing via delimited continuations?
+Backtracking parsing: via (delimited) continuations? effect handlers?
 

### 2024-07-11 POPL/paper.tex
Added comments

@@
 internal to our formalism. Our contributions are then:
-
+%
 \begin{itemize}
@@
 mathematical specification for implementing a \emph{recognizer} of a language.
-
-That is, a recognizer should be extensionally equivalent to the indicator function
+In this section we propose a \emph{proof-relevant} generalization of formal
+languages that can accommodate richer aspects of formal language theory such as
+parsing. We begin by proposing a presheaf approach to grammars that encompasses
+other classes of grammars in a syntax-independent way.
+
+\paragraph{From Formal Languages to Formal Grammars}
+A recognizer should be extensionally equivalent to the indicator function
 $\chi_{L} : \String \to \Prop$ mapping each string $w$ to the the proposition
@@
 identity morphisms.}  category. In fact, in this paper we will take
-$\Grammar$ to be defined as the functor category
-$\Set^{\String}$. Such functor categories into $\Set$ --- often called
-presheaf categories --- carry a remarkable amount of structure, and we
-will demonstrate that these well-known constructions on presheaf
-categories provide us with familiar compositional constructions on
-grammars\max{where do we talk about the properties of this presheaf cat?}.
-
+$\Grammar$ to be defined as the functor category $\Set^{\String}$ ---
+also called presheaf category. Such presheaf categories carry a
+remarkable amount of structure.  Indeed they are a monoidal topos and,
+as such, are a model to a substructural dependently typed theory.
+
+\paragraph{From Grammars to Formal Grammars}
 The category-theoretic framework can be further used to describe and compare
@@
 What even is the message of this paper?}
+
+\pedro{Would building on this proposed structure make the message of the section clearer? In this case
+  we would have to be a bit clearer in terms of how familiar notions of grammar can be seen as categories with functor into $\Grammar$.}
 
@@
   this paragraph and jump into LNL, but then there's not an immediate reason for
-why one might want/need it over a pure linear system} Code in such a language can
+  why one might want/need it over a pure linear system}
+\pedro{I think that it's reasonable to simply say something like ``programming in a purely linear term language
+is limiting due to variable usage restrictions''. In this case we're not throwing our own TT under the bus}
+Code in such a language can
 become unnecessarily verbose when ensuring every function written is indeed linear. It would be more ergonomic to allow users using Girard encodings to translate
@@
 
-In this section we go through a tour of \theoryabbv. We start by presenting its syntax.
+In this section we go on a tour of \theoryabbv. We start by presenting its syntax.
 In order to avoid a notion of raw-term, we present it in an intrinsically-typed fashion,
@@
 the tensor corresponds to grammar concatenation. We conclude the section by showing how
-classes of grammars correspond to certain subtype-theories inside \theoryabbv.
+classes of grammars correspond to certain sub-type-theories inside \theoryabbv.
 
@@
   earlier. So what you write needs to take that into account}
+\pedro{In which case we should introduce it in the previous section. I suggest
+  adding this right after ``and the intuitionistic typing rules in fig 3''}
 
@@
 
-\section{Categorifying Formal Grammars}
+\section{Abstracting Formal Grammars}
 \label{sec:categorify}

### 2024-07-11 POPL/paper.tex
updates to intro

@@
 \cite{Johnsonyacc}.
-%
-The final frontier for parsing research is the development of
-\emph{formally verified} parser implementations, which due to the
-pervasive nature of parsing would be a useful component of many
+
+The theory of formal languages and parsing is one of the oldest and most
+thoroughly developed area of theoretical computer science.
+The central object of study is a \emph{formal language} $L$ over an alphabet
+$\Sigma$, which is classically defined as a
+\emph{subset} of strings $L \subseteq \String$. The expressivity of formal
+languages is sufficient for defining a \emph{recognizer} of a language $L$:
+a function $r : \String \to \Prop$ such that $r$ maps each string $w$ to the
+proposition that $w$ is in the language $w \in L$. This technique of language
+recognition can concisely specify lexical analysis, and even lead to verified
+lexer implementations \cite{egolfVerbatim,Ouedraogo_2023}.
+
+While there are ample thoery and tooling to handle the first stage of a language frontend, the final
+frontier for parsing research is the development of
+\emph{formally verified} parser implementations. Due to the
+ubiquitous nature of parsing, these would useful components of many
 formally verified software systems: compilers and servers most
-especially.
-%
-\steven{Some ideas on what the unifying message throughout the paper should be:
-we offer an abstraction over the formalisms used ad-hoc when doing formal
-language/grammar theory. Whether a user of these constructs recognizes it or not, there is some
-universality between all of their systems. We don't necessarily do much, per se, other
-than abstract over the commonalities and provide a common language.
-
-However, just because there isn't a ton of clever work in designing this syntax, as
-a lot of it does flow very naturally from the literature, that doesn't diminish
-the value of the syntax. In fact, it emboldens the syntax because we want to
-appeal to the idea that ``there couldn't have been any other choice for
-universal grammar''.
-
-That is all to say, our appeal is generalization. I think does appear
-throughout the paper, but is maybe too aimless of a message to be satisfying.
-Perhaps, we should also lean on applications more. Idk I'm gonna sleep now, and
-we'll talk in the morning
-}
-
-In this work we aim to refine formal language theory into a
-mathematical theory of formal grammars that can be used to develop
-\emph{correct-by-construction} parsers. Our approach can be summarized
-in the Curry-Howard-style slogan ``grammars as types'', and the
-addendum ``parsers as terms'' where since the types of the terms are
-grammars, this means not just parsers but intrinsically verified
-parsers as terms. That is, we develop a type theory with a novel
+especially. However, language recognition is insufficient for specifying a
+parser, and for this task much of the structured foundations from the theory of
+formal languages is rendered moot. Rather than returning a propositon-valued result, a parser produces a
+structured \emph{parse tree} adherent to a \emph{formal grammar}.
+The research on such grammars is often performed atop an ad-hoc choice of grammar
+formalism, and results are proven with respect to that formalism. Critically,
+the field is lacking a unifying theory across these different formalisms, even
+though many of the same
+compositional reason principles appear across distinct grammar formalisms.
+
+In this work, we aim to elucidate these similarities and
+provide a mathematical theory of formal grammars for constructing verified
+parsers. Our approach can be summarized with the
+Curry-Howard-style slogan: ``grammars as types''; and the addendum: ``parsers as
+terms''. That is, we develop a type theory with a novel
 denotational semantics in a category with abstract formal grammars as
@@
 expressions, Lambek calculus etc. The category we choose is a rich
-category: it is a topos as well as being monoidal closed, and these
+in structure: it is a topos as well as being monoidal closed, and these
 universal properties serve as the inspiration for our type theory.
@@
 languages, where strings constitute a very clear notion of resource
-that cannot be duplicated, reordered or dropped.
-
-%% The theory of formal languages and parsing is one of the oldest and
-%% most thoroughly developed areas of theoretical computer science. A
-%% prominent topic in the 1950, 60s, and 70s led to a series of remarkable
-%% developments: 
-In this paper, we propose a structured generalization of the
-theoretical machinery used for parsing. In conjunction with an updated
-point-of-view on formal grammars, we provide a domain-specific
-language to build and verify parsers.
-
-\steven{The above paragraph maybe isn't the right thing to say, but I want
-  something that ties the first paragraph to the contributions that follow}
+that cannot be duplicated, reordered, or dropped. Moreover, the constructive
+aspect of \theoryabbv ensures that verification of parsers written in the
+calculus are \emph{correct-by-construction}. The type system is powerful enough
+that derivations of a term type checking carry intrinsic proofs of correctness.
 
@@
 that axiomatizes the underlying structures needed to conduct formal grammar
-theory. Then in \cref{sec:categorify}, we characterize the abstract categorical
-models of our syntax. Afterwards, we demonstrate several models of \theoryabbv,
-including one to prove a metatheoretic canonicity theorem. Finally, in
-\cref{sec:automata} we give a presentation of automata as grammars. With this
-view in mind, we may encode parsing algorithms and execute them entirely
-internal to our formalism. Our contributions are then:
+theory. In \cref{sec:automata} we demonstrate the usage of
+\theoryname for defining parser terms and internalizing results from formal language theory.
+Then in \cref{sec:categorify}, we characterize the abstract categorical
+models of our syntax. Afterwards in \cref{sec:othermodels}, we demonstrate several models,
+including one to prove a metatheoretic canonicity theorem.
+Our contributions are then:
 %
@@
 
-The notion of formal language is central to the theory of parsing. A
-\emph{formal language} $L$ over an alphabet $\Sigma$ is classically defined as a
-\emph{subset} of strings $L \subseteq \String$. This definition is
+The notion of formal language is central to the theory of parsing.
+This definition is
 especially useful as it gives a semantics to formal grammars that is completely
@@
 \paragraph{From Formal Languages to Formal Grammars}
-A recognizer should be extensionally equivalent to the indicator function
-$\chi_{L} : \String \to \Prop$ mapping each string $w$ to the the proposition
-that it is in the language $w \in L$. However, language recognition is
-insufficient for specifying a parser --- a function that decomposes a string
-into semantic components that adhere to the structure of a formal grammar. In
-practice, parsers do more than simply return whether a string is valid or
-invalid; instead, they emit structured \emph{parse trees}\footnote{Such parse
-  trees are usually not materialized in memory, as semantic actions often perform
-  computation over the trees while they are being constructed.}.
 
@@
 types, the same notion of derivative extends to the grammars of \theoryabbv.
+
 Note, to ensure that the linear function types do indeed generalize Brzozowski
-derivative, we must include the equalities in \cref{fig:brzozowskideriv} as axioms\max{we must? why must we}.
+derivatives, we must include the equalities in \cref{fig:brzozowskideriv} as axioms\max{we must? why must we}.
 \pedro{We should further explain why these equations are required, or at the very
@@
   fixed-point}
+\steven{As a consequence of Knaster-Tarski}
 
@@
 calculus, which of course induces a semantics of unrestricted grammars.
-
-\section{Abstracting Formal Grammars}
-\label{sec:categorify}
-
-An important part in the study of type theories is characterizing what
-categorical structures they are modeling. In the case of \theoryabbv,
-due to its intended connection to formal language theory, we use
-Kleene algebras as a starting point in its categorical
-semantics. Being posets, there is an immediate way of categorifying
-them.
-
-Kleene algebras are an important tool in the theory of regular languages
-due to them being a bridge between algebraic reasoning and equivalence
-of regular expressions. More broadly, through various extensions, they serve as a
-theoretical substrate to studying different kinds of formal languages.
-
-We posit that by seeing Kleene algebra through the lens of category
-theory, we gain some insight into the type-theoretic essence of formal
-grammars. We begin this section by providing a ``simply-typed''
-version of a categorification of Kleene algebra, which we call Kleene
-category, and then show define a more expressive class of categories,
-which we call Chomsky category, that will serve as the categorical
-semantics to \theoryabbv.
-
-\subsection{Kleene Categories}
-A Kleene algebra is a tuple $(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is
-a set, $+$ and $\cdot$ are binary operations over $A$, $(-)^*$ is a
-function over $A$, and $1$ and $0$ are constants. These structures
-satisfy the axioms depicted in Figure~\ref{fig:axioms}.
-
-\begin{figure}
-  \begin{align*}
-    x + (y + z) &= (x + y) + z & x + y &= y + x\\
-    x + 0 &= x & x + x &= x\\
-    x(yz) &= (xy)z & x1 &= 1x = x\\
-    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
-    x0 &= 0x = x & & \\
-    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
-     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
-  \end{align*}
-  \caption{Kleene algebra axioms}
-  \label{fig:axioms}
-\end{figure}
-
-The addition operation can be used to define the partial order
-structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
-formal languages, this order structure can be used to model language
-containment. In this section, we categorify the concept of Kleene
-algebra and build on top of it in order to define an abstract theory
-of parsing. We start by defining \emph{Kleene categories}.
-
-\begin{definition}
-  A Kleene category is a distributive monoidal category $\cat{K}$
-  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
-  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
-  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
-  X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogous isomorphism
-  for $G_{A,B}$ also holds.
-\end{definition}
-
-As a sanity check, note that Kleene algebras are indeed examples of
-Kleene categories.
-
-\begin{example}
-  Every Kleene algebra, seen a posetal category, is a Kleene category.
-  The product $\cdot$ is a monoidal product and the addition is a
-  least-upper bound, which corresponds to a coproduct. Lastly, the
-  axioms of the Kleene star have a direct correspondence to the
-  coherence conditions postulated by the initial algebras of Kleene
-  categories.
-\end{example}
-
-This example provides a neat categorical justification to how
-restrictive Kleene algebras are in terms of reasoning about
-languages. By only having at most one morphism between objects, there
-is not a lot of information they can convey. In this case, the only
-information you get is language containment. As demonstrated by
-$\mathbf{Gr}$, the extra degrees of freedom granted by having more
-morphisms gives you more algebraic structure for reasoning about
-languages.
-
-For the next example, we see an unexpected connection with the theory
-of substructural logics.
-
-\begin{example}
-  The opposite category of every Kleene category is a model of a variant of
-  conjunctive ordered logic, where the Kleene star plays the role of the ``of
-  course'' modality from substructural logics which allows hypotheses to
-  be discarded or duplicated.
-\end{example}
-
-As we have seen, the proposed axioms are a direct translation of the
-Kleene algebra axioms to a categorical setting. Its most unusual aspect is the
-axiomatization of the Kleene star as a family of initial algebras
-satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
-more structure, then these isomorphisms hold ``for free''.
-
-\begin{theorem}
-  \label{th:kleeneclosed}
-  Let $\cat{K}$ be a Kleene category such that it is also monoidal
-  closed.  Then, the initial algebras isomorphisms hold automatically.
-\end{theorem}
-\begin{proof}
-  We prove this by the unicity (up-to isomorphism) of initial
-  algebras. Let $[hd, tl]: I + (\mu X.\, F_{A, I}(X)) \otimes A \to
-  (\mu X.\, F_{A, I}(X))$ be the initial algebra structure of $(\mu
-  X.\, F_{A, I}(X))$ and consider the map $[hd, tl] : B + B \otimes
-  (\mu X.\, F_{A, I}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
-    I}(X))$.
-
-  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
-  we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,I} \to B \lto Y$. We can show existence and
-  uniqueness by showing that the diagram on top commutes if, and
-  only if, the diagram on the bottom commutes:
-
-% https://q.uiver.app/#q=WzAsOCxbMCwwLCJCICsgQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFsyLDAsIkIgKyBZIFxcb3RpbWVzIEEiXSxbMiwyLCJZIl0sWzAsMiwiQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDMsIjEgKyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDUsIlxcbXUgWC5cXCwgMSArIFggXFxvdGltZXMgQSJdLFsyLDMsIjEgKyAoIEIgXFxsdG8gWSkgXFxvdGltZXMgQSJdLFsyLDUsIkIgXFxsdG8gWSJdLFswLDEsImlkICsgKGlkIFxcb3RpbWVzIGg7IGV2KSBcXG90aW1lcyBpZF9BIl0sWzEsMiwiW2YsZ10iXSxbMywyLCJpZCBcXG90aW1lcyBoOyBldiIsMl0sWzAsMywiW2lkIFxcb3RpbWVzIGgsIGlkIFxcb3RpbWVzIHRsXSIsMl0sWzQsNiwiaWQgKyAoaCBcXG90aW1lcyBYKVxcb3RpbWVzIGlkIl0sWzUsNywiaCIsMl0sWzYsNywiW2YnLCBnJ10iXSxbNCw1LCJbaGQsIHRsXSIsMl1d
-\[\begin{tikzcd}
-	{B + B \otimes (\mu X.\, I + X \otimes A)} && {B + Y \otimes A} \\
-	\\
-	{B \otimes (\mu X.\, I + X \otimes A)} && Y \\
-	{I + (\mu X.\, I + X \otimes A)} && {I + ( B \lto Y) \otimes A} \\
-	\\
-	{\mu X.\, I + X \otimes A} && {B \lto Y}
-	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
-	\arrow["{[f,g]}", from=1-3, to=3-3]
-	\arrow["{id \otimes h; ev}"', from=3-1, to=3-3]
-	\arrow["{[id \otimes h, id \otimes tl]}"', from=1-1, to=3-1]
-	\arrow["{id + (h \otimes X)\otimes id}", from=4-1, to=4-3]
-	\arrow["h"', from=6-1, to=6-3]
-	\arrow["{[f', g']}", from=4-3, to=6-3]
-	\arrow["{[hd, tl]}"', from=4-1, to=6-1]
-\end{tikzcd}\]
-
-  This equivalence follows by using the adjunction structure given
-  by the monoidal closed structure of $\cat{K}$. A completely analogous
-  argument for $G_{A,B}$ also holds. Furthermore, by generalizing the
-  construction of \Cref{sec:formaltype}, we can also show that from
-  the monoidal closed assumption it follows that $\mu X.\, F_{A, I}(X) \cong \mu X.\, G_{A, I}(X)$
-\end{proof}
-
-Something surprising about this lemma is that it provides an alternative
-perspective on the observation that if a Kleene algebra has an
-residuation operation, also called action algebra \cite{kozen1994action},
-then the Kleene star admits a simpler axiomatization.
-
-Since we want Kleene categories to generalize our notion of formal
-grammars as presheaves $\String \to \Set$, we prove that they do
-indeed form a Kleene category. We start by presenting a well-known
-construction from presheaf categories.
-
-\begin{definition}
-  Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
-  two functors $\cat{C} \to \Set$. Their Day convolution tensor
-  product is defined as the following coend formula:
-  \[
-  (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z)
-  \]
-  Dually, its internal hom is given by the following end formula:
-  \[
-  (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
-  \]
-\end{definition}
-
-\begin{lemma}[Day \cite{day1970construction}]
-  Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
-  monoidal closed.
-\end{lemma}
-
-%% \begin{theorem}
-%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
-%%   The functor category $[A, \cat{K}]$.
-%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
-%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
-%% \end{theorem}
-\begin{theorem}
-  If $\cat{C}$ is a locally small monoidal category, then
-  $\Set^{\cat{C}}$ is a Kleene category.
-\end{theorem}
-\begin{proof}
-
-  By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
-  is a presheaf category, it has coproducts. Furthermore, the tensor
-  is a left adjoint, i.e. it preserves colimits and, therefore, it is
-  a distributive category.
-
-  As for the Kleene star, since presheaf categories admit small colimits,
-  the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
-  defined as the filtered colimit of the diagrams:
-
-  From Theorem~\ref{th:kleeneclosed} it follows that these initial
-  algebras satisfy the required isomorphisms and this concludes the
-  proof.
-\end{proof}
-
-\begin{corollary}
-  For every alphabet $\Sigma$, the presheaf category $\Set^{\cat{\Sigma^*}}$
-  is a Kleene category.
-\end{corollary}
-\begin{proof}
-  Note that string concatenation and the empty string make the
-  discrete category $\Sigma^*$ a strict monoidal category.
-\end{proof}
-
-Much like in the posetal case, the abstract structure of a Kleene
-category is expressive enough to synthetically reason about formal
-languages. A significant difference between them is that while Kleene
-algebras can reason about language containment, Kleene categories can
-reason about parsing as well, as illustrated by the presheaf model
-above.
-
-\subsection{Beyond Simple Types}
-
-Though Kleene categories are expressive enough to reason about
-concepts that are outside of reach of Kleene algebras, their
-simply-typed nature makes them not so expressive from a type theoretic
-point of view. This is limiting because type theories are successful
-syntactic frameworks for manipulating complicated categorical
-structures while avoiding some issues common in category theory, such
-as coherence issues.
-
-With this in mind, we want to design a categorical semantics that
-builds on top of Kleene categories with the goal of extending them
-with dependent types and making them capable reasoning about languages
-and their parsers. This leads us to the abstract notion of model we
-are interested in capturing with \theoryabbv: a \emph{Chomsky
-category}.
-
-\pedro{We should probably define some of the words in this definition}
-\begin{definition}
-  A Chomsky category is a locally Cartesian category with two hierarchies of
-  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
-  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
-  $U_i$ to be closed under dependent products and sum,
-  $L_i$ to be closed under the Kleene category connectives,
-  dependent products, left and right closed structures, with
-  a type constructor $G : L_i \to U_i$ and a linear dependent sum
-  going the other direction.
-\end{definition}
-
-\steven{Max suggests augmenting the definition of a Chomsky category to
-  something like two categories $L$ and $U$, and $U$ is Cartesian closed I don't
-  fully recall the rest.
-
-  My best guess is that you take $L$ a Kleene category with a hierarchy of universes
-  two, and you further require that $L$ is $Psh(U)$-enriched, except he
-  suggested a further adjective on these presheaves. Perhaps representability?
-}
-
-\begin{theorem}
-  The presheaf category $\Grammar = \Set^{\cat{\Sigma^*}}$ is a Chomsky category.
-\end{theorem}
-
-Further, the syntactic category of \theoryname is manifestly a Chomsky category.
-
-\pedro{This is likely true, but if we explicitly say so, this warrants a proof. I think that
-if we don't say anything about the syntactic category, reviewers won't mind.}
-
-\steven{I agree that we don't want to say anything that opens unnecessary
-  questions for proofs we haven't written. However, it seems hard to make the
-  case that we have the right categorical model of the syntax if this statement
-  isn't true. By restating our definition of Chomsky category, this should be
-  obvious or a quick proof}
-
-\pedro{I agree, then we should add the quick proof :) }
-
-\section{Other Models}
-\label{sec:othermodels}
-
-One of the powers of type theories is that it can be profitable to interpret them
-in various models. In this section, by using our just-defined Chomsky categories,
-we show how other useful concepts from formal language theory can also be organized
-as models of \theoryabbv. We illustrate this point by providing two examples that
-are closely related to the theory of formal languages: language equivalence and
-semantic actions. Furthermore, in order to justify how $\mathbf{Gr}$ relates to
-more traditional notions of parsing, we define a glued model that proves a
-canonicity property of grammar terms.
-
-\subsection{Language Semantics}
-Every grammar induces a language semantics. Also languages can be taken as a
-propositionally truncated view of the syntax. Logical equivalence should induce
-weak equivalence, and thus even give a syntactic way to reason about language equivalence.
-\steven{TODO language semantics}
-
-\subsection{Semantic Actions}
-Returning to the problem of parsing, the output of a parse usually is not the
-literal parse tree. Rather, the output is the result of some \emph{semantic
-  actions} ran on the parse tree, which usually serve to remove some syntactic
-details that are unnecessary for later processing.
-
-Given a grammar $G : \String \to \Set$, we define a semantic action to be a set
-$X$ with a function $f$ that produces a semantic element from any parse of $G$.
-
-\[
-  f : \PiTy w \String {G w \to X}
-\]
-
-Further, semantic actions can be arranged into a structured category.
-Define $\SemAct$ as the comma
-category $\Grammar / \Delta$, where $\Delta : \Set \to \Grammar$ defines a
-discrete presheaf. That is, for a set $X$, $\Delta (X)(w) = X$ for all
-$w \in \String$. As $\SemAct$ is defined as a comma category, it has a forgetful
-functor into $\Grammar$. That is, $\SemAct$ serves as a notion of formal
-grammar. Moreover, $\SemAct$ is a model of \theoryabbv.
-
-\steven{It being a notion of formal grammar is distinct from being a model. This
-probably warrants a proof}
-
-\steven{TODO semantic actions}
-
-\pedro{This is a very nice opportunity of showing off the supremacy of denotational
-  reasoning ;) We should probably prove the gluing lemma in the previous section
-  and apply it here and in the canonicity section. The actual proof might have to be
-  moved to the appendix, though}
-
-\subsection{Parse Canonicity}
-Canonicity is an important metatheoretic theorem in the type theory
-literature.  It provides insight on the normal forms of terms and,
-therefore, on its computational aspects. Frequently, proving
-canonicity for boolean types, i.e. every closed term of type bool
-reduces to either true or false, is enough to justify that the type
-theory being studied is well-behaved. In our case, however, since we
-want to connect \theoryabbv to parsers, we must provide a more
-detailed account of canonicity. In particular, we give a nonstandard semantics
-of \theoryabbv that carries a proof of canonicity along with it.
-
-If $\cdot \vdash A$ is a closed linear type then there are
-two obvious notions of what constitutes a ``parse'' of a string w
-according to the grammar $A$:
-\begin{enumerate}
-\item On the one hand we have the set-theoretic semantics just
-  defined, $\llbracket A \rrbracket \cdot w$
-\item On the other hand, we can view the string $w = c_1c_2\cdots$ as
-  a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
-  define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
-  \lceil w \rceil \vdash e : G$.
-\end{enumerate}
-It is not difficult to see that at least for the ``purely positive''
-formulae (those featuring only the positive connectives
-$0,+,I,\otimes,\mu, \overline\Sigma,c$) that
-every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
-that the nodes of the tree correspond precisely to the introduction
-forms of the type. However it is far less obvious that \emph{every}
-linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
-sequence of introduction forms since proofs can include elimination
-forms as well. To show that this is indeed the case we give a
-\emph{canonicity} result for the calculus: that the parses for .
-
-\begin{definition}
-  A non-linear type $X$ is purely positive if it is built up using
-  only finite sums, finite products and least fixed points.
-
-  A linear type is purely positive if it is built up using only finite
-  sums, tensor products, generators $c$, least fixed points and linear
-  sigma types over purely positive non-linear types.
-\end{definition}
-
-\begin{definition}
-  %% Let $X$ be a closed non-linear type. The closed elements $\textrm{Cl}(X)$ of $X$ are the definitional equivalence classes of terms $\cdot \vdash e : X$.
-
-  Let $A$ be a closed linear type. The nerve $N(A)$ is a presheaf on
-  strings that takes a string $w$ to the definitional equivalence
-  classes of terms $\cdot; \lceil w\rceil \vdash e: N(A)$.
-\end{definition}
-
-\begin{theorem}[Canonicity]
-  Let $A$ be a closed, purely positive linear type. Then there is an
-  isomorphism between $\llbracket A\rrbracket$ and $N(A)$.
-\end{theorem}
-\begin{proof}
-  We outline the proof here, more details are in the appendix. The
-  proof proceeds first by a standard logical families construction
-  that combines canonicity arguments for dependent type theory
-  TODO cite coquand
-  % \cite{coquand,etc}
-  with logical relations constructions for linear
-  types
-  TODO cite hylandschalk
-  % \cite{hylandschalk}
-  . It is easy to see by induction that the
-  logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
-  \rrbracket$ and the fundamental lemma proves that the projection
-  morphism $p : \hat A \to N(A)$ has a section, the canonicalization
-  procedure. Then we establish again by induction that
-  canonicalization is also a retraction by observing that introduction
-  forms are taken to constructors.
-\end{proof}
-
-\begin{enumerate}
-\item Every term $\lceil w \rceil \vdash p : G + H$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
-\item There are no terms $\lceil w \rceil \vdash p : 0$
-\item If there is a term $\lceil w \rceil \vdash p : c$ then $w = c$ and $p = x$.
-\item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
-\item Every term $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
-\item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
-\item Every term $\lceil w \rceil \vdash p : \mu X. G$ is equal to $\textrm{roll}(q)$ where $q : G(\mu X.G/X)$
-\item Every term $\lceil w \rceil \vdash p : (x:A) \times G$ is equal
-  to $(M,q)$ where $\cdot \vdash M : A$
-\end{enumerate}
-
-To prove this result we will use a logical families model. We give a
-brief overview of this model concretely:
-\begin{enumerate}
-\item A context $\Gamma$ denotes a family of sets indexed by closing substitutions $\hat\Gamma : (\cdot \vdash \Gamma) \Rightarrow \Set_i$
-\item A type $\Gamma \vdash X : U_i$ denotes a family of sets $\hat X : \Pi(\gamma:\cdot \vdash \Gamma) \hat\Gamma \Rightarrow (\cdot \vdash \simulsubst X \gamma) \Rightarrow \Set_i$
-\item A term $\Gamma \vdash e : X$ denotes a section $\hat e : \Pi(\gamma)\Pi(\hat\gamma)\hat X \gamma \hat\gamma (\simulsubst e \gamma)$
-\item A linear type $\Gamma \vdash A : L_i$ denotes a family of grammars $\hat A : \Pi(\gamma:\cdot\vdash\Gamma)\,\hat\Gamma \Rightarrow \Pi(w:\Sigma^*) (\cdot;\lceil w\rceil \vdash A[\gamma])\Rightarrow \Set_i$, and the denotation of a linear context $\Delta$ is similar.
-\item A linear term $\Gamma;\Delta \vdash e : A$ denotes a function \[\hat e : \Pi(\gamma)\Pi(\hat\gamma)\Pi(w)\Pi(\delta : \lceil w \rceil \vdash \simulsubst \Delta \gamma) \hat\Delta \gamma \hat\gamma \delta \Rightarrow \hat A \gamma \hat\gamma w {(\simulsubst {\simulsubst e \gamma} \delta)}\]
-\end{enumerate}
-And some of the constructions:
-\begin{enumerate}
-\item $\widehat {(G A)} \gamma \hat\gamma e = \hat A \gamma \hat\gamma \varepsilon (G^{-1}e)$
-\item $\widehat {(A \otimes B)} \gamma \hat\gamma w e = \Sigma(w_Aw_B = w)\Sigma(e_A)\Sigma(e_B) (e_A,e_B) = e \wedge \hat A \gamma \hat \gamma w_A e_A \times \hat B \gamma \hat \gamma w_B e_B$
-\item $\widehat {(A \lto B)} \gamma \hat\gamma w e = \Pi(w_A)\Pi(e_A) \hat A \gamma\hat\gamma w_A e_A \Rightarrow \hat B \gamma\hat\gamma (ww_A) (\applto {e_A} e)$
-\end{enumerate}
-
-First, the category with families will be
-the category of logical families over set contexts/types
-$\Delta$/$A$. Then the propositional portion will be defined by
-mapping a logical family $\hat \Gamma \to \Gamma$
-
-First, let $L$ be the category of BI formulae and proofs (quotiented
-by $\beta\eta$ equality). Define a functor $N : L \to \Set^{\Sigma^*}$ by
-\[ N(\phi)(w) = L(w,\phi) \]
-
-Then define the gluing category $\mathcal G$ as the comma category
-$\Set^{\Sigma^*}/N$. That is, an object of this category is a pair of
-a formula $\phi \in L$ and an object $S \in \mathcal
-P(\Sigma^*)/N(\phi)$. We can then use the equivalence $\mathcal
-P(\Sigma^*)/N(\phi) \cong \mathcal P(\int N(\phi))$ to get a simple
-description of such an $S$: it is simply a family of sets indexed by
-proofs $L(w,\phi)$:
-\[ \prod_{w\in\Sigma^*} L(w,\phi) \to \Set \]
-This category clearly comes with a projection functor $\pi : \mathcal
-G \to \mathcal L$ and then our goal is to define a section by using
-the universal property of $\mathcal L$.
-
-To this end we define
-\begin{enumerate}
-\item $(\phi, S) \otimes (\psi, T) = (\phi \otimes \psi, S\otimes T)$ where
-  \[ (S \otimes T)(w, p) = (w_1w_2 = w) \times (q_1,q_2 = p) \times S\,w_1\,q_1 \times T\,w_2\,q_2\]
-\item $(\phi, S) \multimap (\psi, T) = (\phi \multimap \psi, S \multimap T)$ where
-  \[ (S \multimap T)(w,p) = w' \to q \to S\,w'\,q \to T (ww') (p\,q) \]
-\item $\mu X. ??$ ??
-\end{enumerate}
-
-\pedro{We should conclude this section by explaining the relevance of the canonicity
-theorem. Could also be done before stating the theorem.}
 
@@
 
+\section{Abstracting Formal Grammars}
+\label{sec:categorify}
+
+An important part in the study of type theories is characterizing what
+categorical structures they are modeling. In the case of \theoryabbv,
+due to its intended connection to formal language theory, we use
+Kleene algebras as a starting point in its categorical
+semantics. Being posets, there is an immediate way of categorifying
+them.
+
+Kleene algebras are an important tool in the theory of regular languages
+due to them being a bridge between algebraic reasoning and equivalence
+of regular expressions. More broadly, through various extensions, they serve as a
+theoretical substrate to studying different kinds of formal languages.
+
+We posit that by seeing Kleene algebra through the lens of category
+theory, we gain some insight into the type-theoretic essence of formal
+grammars. We begin this section by providing a ``simply-typed''
+version of a categorification of Kleene algebra, which we call Kleene
+category, and then show define a more expressive class of categories,
+which we call Chomsky category, that will serve as the categorical
+semantics to \theoryabbv.
+
+\subsection{Kleene Categories}
+A Kleene algebra is a tuple $(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is
+a set, $+$ and $\cdot$ are binary operations over $A$, $(-)^*$ is a
+function over $A$, and $1$ and $0$ are constants. These structures
+satisfy the axioms depicted in Figure~\ref{fig:axioms}.
+
+\begin{figure}
+  \begin{align*}
+    x + (y + z) &= (x + y) + z & x + y &= y + x\\
+    x + 0 &= x & x + x &= x\\
+    x(yz) &= (xy)z & x1 &= 1x = x\\
+    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
+    x0 &= 0x = x & & \\
+    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
+     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
+  \end{align*}
+  \caption{Kleene algebra axioms}
+  \label{fig:axioms}
+\end{figure}
+
+The addition operation can be used to define the partial order
+structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
+formal languages, this order structure can be used to model language
+containment. In this section, we categorify the concept of Kleene
+algebra and build on top of it in order to define an abstract theory
+of parsing. We start by defining \emph{Kleene categories}.
+
+\begin{definition}
+  A Kleene category is a distributive monoidal category $\cat{K}$
+  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
+  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
+  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
+  X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogous isomorphism
+  for $G_{A,B}$ also holds.
+\end{definition}
+
+As a sanity check, note that Kleene algebras are indeed examples of
+Kleene categories.
+
+\begin{example}
+  Every Kleene algebra, seen a posetal category, is a Kleene category.
+  The product $\cdot$ is a monoidal product and the addition is a
+  least-upper bound, which corresponds to a coproduct. Lastly, the
+  axioms of the Kleene star have a direct correspondence to the
+  coherence conditions postulated by the initial algebras of Kleene
+  categories.
+\end{example}
+
+This example provides a neat categorical justification to how
+restrictive Kleene algebras are in terms of reasoning about
+languages. By only having at most one morphism between objects, there
+is not a lot of information they can convey. In this case, the only
+information you get is language containment. As demonstrated by
+$\mathbf{Gr}$, the extra degrees of freedom granted by having more
+morphisms gives you more algebraic structure for reasoning about
+languages.
+
+For the next example, we see an unexpected connection with the theory
+of substructural logics.
+
+\begin{example}
+  The opposite category of every Kleene category is a model of a variant of
+  conjunctive ordered logic, where the Kleene star plays the role of the ``of
+  course'' modality from substructural logics which allows hypotheses to
+  be discarded or duplicated.
+\end{example}
+
+As we have seen, the proposed axioms are a direct translation of the
+Kleene algebra axioms to a categorical setting. Its most unusual aspect is the
+axiomatization of the Kleene star as a family of initial algebras
+satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
+more structure, then these isomorphisms hold ``for free''.
+
+\begin{theorem}
+  \label{th:kleeneclosed}
+  Let $\cat{K}$ be a Kleene category such that it is also monoidal
+  closed.  Then, the initial algebras isomorphisms hold automatically.
+\end{theorem}
+\begin{proof}
+  We prove this by the unicity (up-to isomorphism) of initial
+  algebras. Let $[hd, tl]: I + (\mu X.\, F_{A, I}(X)) \otimes A \to
+  (\mu X.\, F_{A, I}(X))$ be the initial algebra structure of $(\mu
+  X.\, F_{A, I}(X))$ and consider the map $[hd, tl] : B + B \otimes
+  (\mu X.\, F_{A, I}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
+    I}(X))$.
+
+  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
+  we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,I} \to B \lto Y$. We can show existence and
+  uniqueness by showing that the diagram on top commutes if, and
+  only if, the diagram on the bottom commutes:
+
+% https://q.uiver.app/#q=WzAsOCxbMCwwLCJCICsgQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFsyLDAsIkIgKyBZIFxcb3RpbWVzIEEiXSxbMiwyLCJZIl0sWzAsMiwiQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDMsIjEgKyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDUsIlxcbXUgWC5cXCwgMSArIFggXFxvdGltZXMgQSJdLFsyLDMsIjEgKyAoIEIgXFxsdG8gWSkgXFxvdGltZXMgQSJdLFsyLDUsIkIgXFxsdG8gWSJdLFswLDEsImlkICsgKGlkIFxcb3RpbWVzIGg7IGV2KSBcXG90aW1lcyBpZF9BIl0sWzEsMiwiW2YsZ10iXSxbMywyLCJpZCBcXG90aW1lcyBoOyBldiIsMl0sWzAsMywiW2lkIFxcb3RpbWVzIGgsIGlkIFxcb3RpbWVzIHRsXSIsMl0sWzQsNiwiaWQgKyAoaCBcXG90aW1lcyBYKVxcb3RpbWVzIGlkIl0sWzUsNywiaCIsMl0sWzYsNywiW2YnLCBnJ10iXSxbNCw1LCJbaGQsIHRsXSIsMl1d
+\[\begin{tikzcd}
+	{B + B \otimes (\mu X.\, I + X \otimes A)} && {B + Y \otimes A} \\
+	\\
+	{B \otimes (\mu X.\, I + X \otimes A)} && Y \\
+	{I + (\mu X.\, I + X \otimes A)} && {I + ( B \lto Y) \otimes A} \\
+	\\
+	{\mu X.\, I + X \otimes A} && {B \lto Y}
+	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
+	\arrow["{[f,g]}", from=1-3, to=3-3]
+	\arrow["{id \otimes h; ev}"', from=3-1, to=3-3]
+	\arrow["{[id \otimes h, id \otimes tl]}"', from=1-1, to=3-1]
+	\arrow["{id + (h \otimes X)\otimes id}", from=4-1, to=4-3]
+	\arrow["h"', from=6-1, to=6-3]
+	\arrow["{[f', g']}", from=4-3, to=6-3]
+	\arrow["{[hd, tl]}"', from=4-1, to=6-1]
+\end{tikzcd}\]
+
+  This equivalence follows by using the adjunction structure given
+  by the monoidal closed structure of $\cat{K}$. A completely analogous
+  argument for $G_{A,B}$ also holds. Furthermore, by generalizing the
+  construction of \Cref{sec:formaltype}, we can also show that from
+  the monoidal closed assumption it follows that $\mu X.\, F_{A, I}(X) \cong \mu X.\, G_{A, I}(X)$
+\end{proof}
+
+Something surprising about this lemma is that it provides an alternative
+perspective on the observation that if a Kleene algebra has an
+residuation operation, also called action algebra \cite{kozen1994action},
+then the Kleene star admits a simpler axiomatization.
+
+Since we want Kleene categories to generalize our notion of formal
+grammars as presheaves $\String \to \Set$, we prove that they do
+indeed form a Kleene category. We start by presenting a well-known
+construction from presheaf categories.
+
+\begin{definition}
+  Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
+  two functors $\cat{C} \to \Set$. Their Day convolution tensor
+  product is defined as the following coend formula:
+  \[
+  (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z)
+  \]
+  Dually, its internal hom is given by the following end formula:
+  \[
+  (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
+  \]
+\end{definition}
+
+\begin{lemma}[Day \cite{day1970construction}]
+  Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
+  monoidal closed.
+\end{lemma}
+
+%% \begin{theorem}
+%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
+%%   The functor category $[A, \cat{K}]$.
+%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
+%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
+%% \end{theorem}
+\begin{theorem}
+  If $\cat{C}$ is a locally small monoidal category, then
+  $\Set^{\cat{C}}$ is a Kleene category.
+\end{theorem}
+\begin{proof}
+
+  By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
+  is a presheaf category, it has coproducts. Furthermore, the tensor
+  is a left adjoint, i.e. it preserves colimits and, therefore, it is
+  a distributive category.
+
+  As for the Kleene star, since presheaf categories admit small colimits,
+  the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
+  defined as the filtered colimit of the diagrams:
+
+  From Theorem~\ref{th:kleeneclosed} it follows that these initial
+  algebras satisfy the required isomorphisms and this concludes the
+  proof.
+\end{proof}
+
+\begin{corollary}
+  For every alphabet $\Sigma$, the presheaf category $\Set^{\cat{\Sigma^*}}$
+  is a Kleene category.
+\end{corollary}
+\begin{proof}
+  Note that string concatenation and the empty string make the
+  discrete category $\Sigma^*$ a strict monoidal category.
+\end{proof}
+
+Much like in the posetal case, the abstract structure of a Kleene
+category is expressive enough to synthetically reason about formal
+languages. A significant difference between them is that while Kleene
+algebras can reason about language containment, Kleene categories can
+reason about parsing as well, as illustrated by the presheaf model
+above.
+
+\subsection{Beyond Simple Types}
+
+Though Kleene categories are expressive enough to reason about
+concepts that are outside of reach of Kleene algebras, their
+simply-typed nature makes them not so expressive from a type theoretic
+point of view. This is limiting because type theories are successful
+syntactic frameworks for manipulating complicated categorical
+structures while avoiding some issues common in category theory, such
+as coherence issues.
+
+With this in mind, we want to design a categorical semantics that
+builds on top of Kleene categories with the goal of extending them
+with dependent types and making them capable reasoning about languages
+and their parsers. This leads us to the abstract notion of model we
+are interested in capturing with \theoryabbv: a \emph{Chomsky
+category}.
+
+\pedro{We should probably define some of the words in this definition}
+\begin{definition}
+  A Chomsky category is a locally Cartesian category with two hierarchies of
+  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
+  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
+  $U_i$ to be closed under dependent products and sum,
+  $L_i$ to be closed under the Kleene category connectives,
+  dependent products, left and right closed structures, with
+  a type constructor $G : L_i \to U_i$ and a linear dependent sum
+  going the other direction.
+\end{definition}
+
+\steven{Max suggests augmenting the definition of a Chomsky category to
+  something like two categories $L$ and $U$, and $U$ is Cartesian closed I don't
+  fully recall the rest.
+
+  My best guess is that you take $L$ a Kleene category with a hierarchy of universes
+  two, and you further require that $L$ is $Psh(U)$-enriched, except he
+  suggested a further adjective on these presheaves. Perhaps representability?
+}
+
+\begin{theorem}
+  The presheaf category $\Grammar = \Set^{\cat{\Sigma^*}}$ is a Chomsky category.
+\end{theorem}
+
+Further, the syntactic category of \theoryname is manifestly a Chomsky category.
+
+\pedro{This is likely true, but if we explicitly say so, this warrants a proof. I think that
+if we don't say anything about the syntactic category, reviewers won't mind.}
+
+\steven{I agree that we don't want to say anything that opens unnecessary
+  questions for proofs we haven't written. However, it seems hard to make the
+  case that we have the right categorical model of the syntax if this statement
+  isn't true. By restating our definition of Chomsky category, this should be
+  obvious or a quick proof}
+
+\pedro{I agree, then we should add the quick proof :) }
+
+\section{Other Models}
+\label{sec:othermodels}
+
+One of the powers of type theories is that it can be profitable to interpret them
+in various models. In this section, by using our just-defined Chomsky categories,
+we show how other useful concepts from formal language theory can also be organized
+as models of \theoryabbv. We illustrate this point by providing two examples that
+are closely related to the theory of formal languages: language equivalence and
+semantic actions. Furthermore, in order to justify how $\mathbf{Gr}$ relates to
+more traditional notions of parsing, we define a glued model that proves a
+canonicity property of grammar terms.
+
+\subsection{Language Semantics}
+Every grammar induces a language semantics. Also languages can be taken as a
+propositionally truncated view of the syntax. Logical equivalence should induce
+weak equivalence, and thus even give a syntactic way to reason about language equivalence.
+\steven{TODO language semantics}
+
+\subsection{Semantic Actions}
+Returning to the problem of parsing, the output of a parse usually is not the
+literal parse tree. Rather, the output is the result of some \emph{semantic
+  actions} ran on the parse tree, which usually serve to remove some syntactic
+details that are unnecessary for later processing.
+
+Given a grammar $G : \String \to \Set$, we define a semantic action to be a set
+$X$ with a function $f$ that produces a semantic element from any parse of $G$.
+
+\[
+  f : \PiTy w \String {G w \to X}
+\]
+
+Further, semantic actions can be arranged into a structured category.
+Define $\SemAct$ as the comma
+category $\Grammar / \Delta$, where $\Delta : \Set \to \Grammar$ defines a
+discrete presheaf. That is, for a set $X$, $\Delta (X)(w) = X$ for all
+$w \in \String$. As $\SemAct$ is defined as a comma category, it has a forgetful
+functor into $\Grammar$. That is, $\SemAct$ serves as a notion of formal
+grammar. Moreover, $\SemAct$ is a model of \theoryabbv.
+
+\steven{It being a notion of formal grammar is distinct from being a model. This
+probably warrants a proof}
+
+\steven{TODO semantic actions}
+
+\pedro{This is a very nice opportunity of showing off the supremacy of denotational
+  reasoning ;) We should probably prove the gluing lemma in the previous section
+  and apply it here and in the canonicity section. The actual proof might have to be
+  moved to the appendix, though}
+
+\subsection{Parse Canonicity}
+Canonicity is an important metatheoretic theorem in the type theory
+literature.  It provides insight on the normal forms of terms and,
+therefore, on its computational aspects. Frequently, proving
+canonicity for boolean types, i.e. every closed term of type bool
+reduces to either true or false, is enough to justify that the type
+theory being studied is well-behaved. In our case, however, since we
+want to connect \theoryabbv to parsers, we must provide a more
+detailed account of canonicity. In particular, we give a nonstandard semantics
+of \theoryabbv that carries a proof of canonicity along with it.
+
+If $\cdot \vdash A$ is a closed linear type then there are
+two obvious notions of what constitutes a ``parse'' of a string w
+according to the grammar $A$:
+\begin{enumerate}
+\item On the one hand we have the set-theoretic semantics just
+  defined, $\llbracket A \rrbracket \cdot w$
+\item On the other hand, we can view the string $w = c_1c_2\cdots$ as
+  a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
+  define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
+  \lceil w \rceil \vdash e : G$.
+\end{enumerate}
+It is not difficult to see that at least for the ``purely positive''
+formulae (those featuring only the positive connectives
+$0,+,I,\otimes,\mu, \overline\Sigma,c$) that
+every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
+that the nodes of the tree correspond precisely to the introduction
+forms of the type. However it is far less obvious that \emph{every}
+linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
+sequence of introduction forms since proofs can include elimination
+forms as well. To show that this is indeed the case we give a
+\emph{canonicity} result for the calculus: that the parses for .
+
+\begin{definition}
+  A non-linear type $X$ is purely positive if it is built up using
+  only finite sums, finite products and least fixed points.
+
+  A linear type is purely positive if it is built up using only finite
+  sums, tensor products, generators $c$, least fixed points and linear
+  sigma types over purely positive non-linear types.
+\end{definition}
+
+\begin{definition}
+  %% Let $X$ be a closed non-linear type. The closed elements $\textrm{Cl}(X)$ of $X$ are the definitional equivalence classes of terms $\cdot \vdash e : X$.
+
+  Let $A$ be a closed linear type. The nerve $N(A)$ is a presheaf on
+  strings that takes a string $w$ to the definitional equivalence
+  classes of terms $\cdot; \lceil w\rceil \vdash e: N(A)$.
+\end{definition}
+
+\begin{theorem}[Canonicity]
+  Let $A$ be a closed, purely positive linear type. Then there is an
+  isomorphism between $\llbracket A\rrbracket$ and $N(A)$.
+\end{theorem}
+\begin{proof}
+  We outline the proof here, more details are in the appendix. The
+  proof proceeds first by a standard logical families construction
+  that combines canonicity arguments for dependent type theory
+  TODO cite coquand
+  % \cite{coquand,etc}
+  with logical relations constructions for linear
+  types
+  TODO cite hylandschalk
+  % \cite{hylandschalk}
+  . It is easy to see by induction that the
+  logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
+  \rrbracket$ and the fundamental lemma proves that the projection
+  morphism $p : \hat A \to N(A)$ has a section, the canonicalization
+  procedure. Then we establish again by induction that
+  canonicalization is also a retraction by observing that introduction
+  forms are taken to constructors.
+\end{proof}
+
+\begin{enumerate}
+\item Every term $\lceil w \rceil \vdash p : G + H$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
+\item There are no terms $\lceil w \rceil \vdash p : 0$
+\item If there is a term $\lceil w \rceil \vdash p : c$ then $w = c$ and $p = x$.
+\item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
+\item Every term $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
+\item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
+\item Every term $\lceil w \rceil \vdash p : \mu X. G$ is equal to $\textrm{roll}(q)$ where $q : G(\mu X.G/X)$
+\item Every term $\lceil w \rceil \vdash p : (x:A) \times G$ is equal
+  to $(M,q)$ where $\cdot \vdash M : A$
+\end{enumerate}
+
+To prove this result we will use a logical families model. We give a
+brief overview of this model concretely:
+\begin{enumerate}
+\item A context $\Gamma$ denotes a family of sets indexed by closing substitutions $\hat\Gamma : (\cdot \vdash \Gamma) \Rightarrow \Set_i$
+\item A type $\Gamma \vdash X : U_i$ denotes a family of sets $\hat X : \Pi(\gamma:\cdot \vdash \Gamma) \hat\Gamma \Rightarrow (\cdot \vdash \simulsubst X \gamma) \Rightarrow \Set_i$
+\item A term $\Gamma \vdash e : X$ denotes a section $\hat e : \Pi(\gamma)\Pi(\hat\gamma)\hat X \gamma \hat\gamma (\simulsubst e \gamma)$
+\item A linear type $\Gamma \vdash A : L_i$ denotes a family of grammars $\hat A : \Pi(\gamma:\cdot\vdash\Gamma)\,\hat\Gamma \Rightarrow \Pi(w:\Sigma^*) (\cdot;\lceil w\rceil \vdash A[\gamma])\Rightarrow \Set_i$, and the denotation of a linear context $\Delta$ is similar.
+\item A linear term $\Gamma;\Delta \vdash e : A$ denotes a function \[\hat e : \Pi(\gamma)\Pi(\hat\gamma)\Pi(w)\Pi(\delta : \lceil w \rceil \vdash \simulsubst \Delta \gamma) \hat\Delta \gamma \hat\gamma \delta \Rightarrow \hat A \gamma \hat\gamma w {(\simulsubst {\simulsubst e \gamma} \delta)}\]
+\end{enumerate}
+And some of the constructions:
+\begin{enumerate}
+\item $\widehat {(G A)} \gamma \hat\gamma e = \hat A \gamma \hat\gamma \varepsilon (G^{-1}e)$
+\item $\widehat {(A \otimes B)} \gamma \hat\gamma w e = \Sigma(w_Aw_B = w)\Sigma(e_A)\Sigma(e_B) (e_A,e_B) = e \wedge \hat A \gamma \hat \gamma w_A e_A \times \hat B \gamma \hat \gamma w_B e_B$
+\item $\widehat {(A \lto B)} \gamma \hat\gamma w e = \Pi(w_A)\Pi(e_A) \hat A \gamma\hat\gamma w_A e_A \Rightarrow \hat B \gamma\hat\gamma (ww_A) (\applto {e_A} e)$
+\end{enumerate}
+
+First, the category with families will be
+the category of logical families over set contexts/types
+$\Delta$/$A$. Then the propositional portion will be defined by
+mapping a logical family $\hat \Gamma \to \Gamma$
+
+First, let $L$ be the category of BI formulae and proofs (quotiented
+by $\beta\eta$ equality). Define a functor $N : L \to \Set^{\Sigma^*}$ by
+\[ N(\phi)(w) = L(w,\phi) \]
+
+Then define the gluing category $\mathcal G$ as the comma category
+$\Set^{\Sigma^*}/N$. That is, an object of this category is a pair of
+a formula $\phi \in L$ and an object $S \in \mathcal
+P(\Sigma^*)/N(\phi)$. We can then use the equivalence $\mathcal
+P(\Sigma^*)/N(\phi) \cong \mathcal P(\int N(\phi))$ to get a simple
+description of such an $S$: it is simply a family of sets indexed by
+proofs $L(w,\phi)$:
+\[ \prod_{w\in\Sigma^*} L(w,\phi) \to \Set \]
+This category clearly comes with a projection functor $\pi : \mathcal
+G \to \mathcal L$ and then our goal is to define a section by using
+the universal property of $\mathcal L$.
+
+To this end we define
+\begin{enumerate}
+\item $(\phi, S) \otimes (\psi, T) = (\phi \otimes \psi, S\otimes T)$ where
+  \[ (S \otimes T)(w, p) = (w_1w_2 = w) \times (q_1,q_2 = p) \times S\,w_1\,q_1 \times T\,w_2\,q_2\]
+\item $(\phi, S) \multimap (\psi, T) = (\phi \multimap \psi, S \multimap T)$ where
+  \[ (S \multimap T)(w,p) = w' \to q \to S\,w'\,q \to T (ww') (p\,q) \]
+\item $\mu X. ??$ ??
+\end{enumerate}
+
+\pedro{We should conclude this section by explaining the relevance of the canonicity
+theorem. Could also be done before stating the theorem.}
+
 \section{Discussion and Future Work}

### 2024-07-11 POPL/paper.tex
work on semantics sections

@@
 
-  When you wish to show that these actually maps invert each other to form an isomorphism, we invoke the
+  When you wish to show that these actually maps invert each other to form an isomorphism\max{jumbled sentence}, we invoke the
   universal properties of the $\mathsf{Trace}$ types. That is,
@@
 
-\section{Abstracting Formal Grammars}
+\section{Categorical Semantics of \theoryabbv}
 \label{sec:categorify}
 
-An important part in the study of type theories is characterizing what
-categorical structures they are modeling. In the case of \theoryabbv,
-due to its intended connection to formal language theory, we use
-Kleene algebras as a starting point in its categorical
-semantics. Being posets, there is an immediate way of categorifying
-them.
-
-Kleene algebras are an important tool in the theory of regular languages
-due to them being a bridge between algebraic reasoning and equivalence
-of regular expressions. More broadly, through various extensions, they serve as a
-theoretical substrate to studying different kinds of formal languages.
-
-We posit that by seeing Kleene algebra through the lens of category
-theory, we gain some insight into the type-theoretic essence of formal
-grammars. We begin this section by providing a ``simply-typed''
-version of a categorification of Kleene algebra, which we call Kleene
-category, and then show define a more expressive class of categories,
-which we call Chomsky category, that will serve as the categorical
-semantics to \theoryabbv.
-
-\subsection{Kleene Categories}
+As described in previous sections, \theoryabbv has a ``standard''
+interpretation where non-linear types denote sets, linear types denote
+formal grammars, non-linear terms denote functions and linear terms
+denote parse transformers.
+%
+Further, as is typical for type theory, every type constructor in the
+calculus is interpreted by using a \emph{universal construction} in
+either the category of sets or the category of grammars\footnote{the
+only exception is that the axioms for Brzozowski derivatives we
+considered do not follow solely from universal properties. In this
+section we focus on models that do not necessarily satisfy these
+axioms.}.
+%
+This leads to an immediate opportunity for generalization: the type
+theory has in addition to the standard grammar-theoretic
+interpretation, an interpretation in any categorical structure that
+exhibits the same universal constructions.
+%
+On the one hand this gives us a structured way to formalize the
+standard semantics precisely, but additionally it enables us to
+consider \emph{non-standard} models in the next section that point to
+further applications as well as meta-theoretic results.
+
+In developing this categorical semantics, we will start with the
+closest analogue from formal language theory: Kleene algebra.  Kleene
+algebras are an important tool in the theory of regular languages
+serving as a bridge between algebraic reasoning and equivalence of
+regular expressions. More broadly, through various extensions, they
+serve as a theoretical substrate to studying different kinds of formal
+languages.
+%
+We can then see that the ``regular fragment'' of our type theory
+(i.e., just characters, $\otimes$, $\oplus$, $0,1$ and a Kleene star)
+has models in what we call a \emph{Kleene category} a categorification
+of Kleene algebra from posets to categories.
+%
+We then further develop this into our final notion of model, which we
+call a \emph{Chomsky category} as it can model not just regular
+languages, but the full Chomsky hierarchy.
+
+\subsection{Kleene Algebra and Kleene Categories}
 A Kleene algebra is a tuple $(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is
@@
 \end{tikzcd}\]
-
   This equivalence follows by using the adjunction structure given
@@
 Much like in the posetal case, the abstract structure of a Kleene
-category is expressive enough to synthetically reason about formal
+category is expressive enough to synthetically reason about regular
 languages. A significant difference between them is that while Kleene
 algebras can reason about language containment, Kleene categories can
-reason about parsing as well, as illustrated by the presheaf model
-above.
-
-\subsection{Beyond Simple Types}
+reason about \emph{ambiguity}, \emph{strong equivalence} of grammars.
+
+\subsection{Chomsky Categories}
 
@@
 
-\section{Other Models}
+\section{Concrete Models of \theoryabbv}
 \label{sec:othermodels}

### 2024-11-08 paper/paper.tex
sync paper changes

--- 2024-01-26 paper/paper.tex
@@
-\documentclass[sigconf,anonymous,review,screen]{acmart}
+% -*- fill-column: 80; -*-
+\documentclass[review,anonymous,screen,acmsmall,nonacm]{acmart}
 \usepackage{mathpartir}
 \usepackage{tikz-cd}
+\usepackage{lipsum}
 \usepackage{enumitem}
@@
 \usepackage{fancyvrb}
+\usepackage{xspace}
+
+\definecolor{codegreen}{rgb}{0,0.6,0}
+\definecolor{codegray}{rgb}{0.5,0.5,0.5}
+\definecolor{codepurple}{rgb}{0.58,0,0.82}
+\definecolor{backcolour}{rgb}{0.95,0.95,0.92}
+
+\usepackage{listings}
+\lstdefinelanguage{Agda}{
+keywords={data, where, module, import, open, public,
+record, field, let, in, if, then, else, case, of, with,
+do, postulate, primitive, mutual, abstract, private,
+forall, exists, cong, set, prop, sort, Level, Data, Type,
+Renamer},
+morekeywords=[2]{Set, Type, Prop, moduleFuncs, instance, Named},
+sensitive=true,
+comment=[l]{--},
+morecomment=[s]{\{-}{-\}},
+morestring=[b]",
+mathescape=true,
+escapeinside={(*@}{@*)}
+}
+\lstset{
+language=Agda,
+basicstyle=\ttfamily\small,
+keywordstyle=\color{blue},
+keywordstyle=[2]\color{teal},
+identifierstyle=\color{black},
+commentstyle=\color{gray}\textit,
+stringstyle=\color{orange},
+numbers=none,
+numberstyle=\tiny\color{gray},
+stepnumber=1,
+numbersep=5pt,
+showstringspaces=false,
+tabsize=4,
+captionpos=b,
+breaklines=true,
+breakatwhitespace=false,
+rulecolor=\color{black},
+texcl=true,
+backgroundcolor=\color{backcolour},
+frame=single
+}
+
 \usepackage{hyperref}
-\usepackage{cleveref}
+\usepackage[capitalise]{cleveref}
 \usepackage{stmaryrd}
-
+\usepackage{scalerel}
+\usepackage{tikz}
+\usetikzlibrary{automata, positioning, arrows, fit}
+
+\usepackage{pdfpages}
+
+\newcommand{\SPF}{\mathsf{SPF}}
+\newcommand{\Var}{\mathsf{Var}}
+\newcommand{\K}{\mathsf{K}}
+\newcommand{\map}{\mathsf{map}}
+\newcommand{\roll}{\mathsf{roll}}
+\newcommand{\fold}{\mathsf{fold}}
+\newcommand{\inl}{\mathsf{inl}}
+\newcommand{\inr}{\mathsf{inr}}
 \newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
@@
 \newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
+\newcommand{\String}{\textbf{String}}
+\newcommand{\stringg}{\mathtt{String}}
+\newcommand{\charg}{\mathtt{Char}}
 \newcommand{\Set}{\mathbf{Set}}
+\newcommand{\Syn}{\mathbf{Synx}}
+\newcommand{\SemAct}{\mathbf{SemAct}}
 \newcommand{\Gr}{\mathbf{Gr}}
+\newcommand{\Grammar}{\mathbf{Gr}}
 \newcommand{\Type}{\mathbf{Type}}
 \newcommand{\Prop}{\mathbf{Prop}}
-\newcommand{\Bool}{\mathbf{Bool}}
+\newcommand{\Bool}{\mathtt{Bool}}
+\newcommand{\true}{\mathtt{true}}
+\newcommand{\false}{\mathtt{false}}
 \newcommand{\nat}{\mathbb{N}}
+\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
+\newcommand{\theoryabbv}{$\textrm{Lambek}^D$}
+\newcommand{\lnld}{$\textrm{LNL}_D$}
+
+\newcommand{\isTy}{\textrm{ type}}
+\newcommand{\isCtx}{\textrm{ ctx}}
+\newcommand{\isSmall}{\textrm{ small}}
+\newcommand{\isLinTy}{\textrm{ lin. type}}
+\newcommand{\isLinCtx}{\textrm{ lin. ctx.}}
+
+\newcommand{\quoteTy}[1]{\lceil{#1}\rceil}
+\newcommand{\unquoteTy}[1]{\lfloor{#1}\rfloor}
 
@@
 
+\newcommand{\amp}{\mathrel{\&}}
+\newcommand{\pair}{\amp}
+\DeclareMathOperator*{\bigamp}{\scalerel*{\&}{\bigoplus}}
+\DeclareMathOperator*{\bigwith}{\scalerel*{\&}{\bigoplus}}
+
+\newcommand{\bang}{~\textbf{!}~}
+% Tentatively using uparrow for linear to non-linear to be in line with
+% Pfennings adjoint functional programming
+\newcommand{\ltonl}[1]{~\uparrow #1}
+\newcommand{\nil}{\texttt{nil}}
+\newcommand{\ident}{\texttt{id}}
+\newcommand{\cons}{\texttt{cons}}
+\newcommand{\epscons}{\varepsilon\texttt{cons}}
+\newcommand{\data}{\mathsf{data}}
+\newcommand{\where}{\mathsf{where}}
+\newcommand{\Trace}{\texttt{Trace}}
+\newcommand{\stringquote}[1]{\texttt{\textquotesingle#1\textquotesingle}}
+\newcommand{\internalize}[1]{\lceil#1\rceil}
+
+\newcommand{\oplusinj}[2]{\sigma\,#1\,#2}
+\newcommand{\withprj}[2]{\pi\,#1\,#2}
+
 \newcommand{\simulsubst}[2]{#1\{#2\}}
 \newcommand{\subst}[3]{\simulsubst {#1} {#2/#3}}
+\newcommand{\el}{\mathsf{el}}
 \newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
@@
 \newcommand{\lambtol}[2]{\lambda^{{\tol}} #1.\, #2}
-\newcommand{\dlamb}[2]{\overline{\lambda} #1.\, #2}
+\newcommand{\dlamb}[2]{{\lambda}^{{\&}} #1.\, #2}
+\newcommand{\withlamb}[2]{{\lambda}^{{\&}} #1.\, #2}
 \newcommand{\app}[2]{#1 \, #2}
-\newcommand{\applto}[2]{#1 \mathop{{}^{\lto}} #2}
+\newcommand{\applto}[2]{#1 \, #2}
 \newcommand{\apptol}[2]{#1 \mathop{{}^{\tol}} #2}
-\newcommand{\PiTy}[3]{\Pi #1 : #2.\, #3}
-\newcommand{\SigTy}[3]{\Sigma #1 : #2.\, #3}
-\newcommand{\LinPiTy}[3]{\widebar\Pi #1 : #2.\, #3}
-\newcommand{\LinSigTy}[3]{\widebar\Sigma #1 : #2.\, #3}
-\newcommand{\amp}{\mathrel{\&}}
+\newcommand{\PiTy}[3]{\textstyle\prod (#1 : #2). #3}
+\newcommand{\SigTy}[3]{\textstyle\sum (#1 : #2). #3}
+\newcommand{\LinPiTy}[3]{\textstyle\bigamp (#1 : #2). #3}
+\newcommand{\LinPiTyLimit}[3]{\bigwith\limits_{#1 : #2} #3}
+\newcommand{\LinSigTy}[3]{\textstyle\bigoplus (#1 : #2). #3}
+\newcommand{\LinSigTyLimit}[3]{\bigoplus\limits_{#1 : #2} #3}
+%% \newcommand{\DepWith}[2]{{\textstyle\bigamp}\limits_{#1}{#2}
+%% \newcommand{\DepPlus}[2]{{\textstyle\bigoplus}\limits_{#1}{#2}
 \newcommand{\GrTy}{\mathsf{Gr}}
 
-\newcommand{\ctxwff}[1]{#1 \,\, \mathsf{ok}}
-\newcommand{\ctxwffjdg}[2]{#1 \vdash #2 \,\, \mathsf{type}}
-\newcommand{\linctxwff}[2]{#1 \vdash #2 \,\, \mathsf{ok}}
-\newcommand{\linctxwffjdg}[2]{#1 \vdash #2 \,\, \mathsf{linear}}
+\newcommand{\equalizer}[3]{\{#1\,|\,\applto {#2}{#1} = \applto{#3}{#1} \}}
+\newcommand{\equalizerin}[1]{\langle #1 \rangle}
+\newcommand{\equalizerpi}[1]{#1.\pi}
+
+\newcommand{\ctxwff}[1]{#1 \isCtx}
+\newcommand{\ctxwffjdg}[2]{#1 \vdash #2 \isTy}
+\newcommand{\linctxwff}[2]{#1 \vdash #2 \isLinCtx}
+\newcommand{\linctxwffjdg}[2]{#1 \vdash #2 \isLinTy}
+
+\newsavebox{\logoagdabox}
+\sbox{\logoagdabox}{%
+  %
+  \raisebox{-2pt}{\includegraphics[height=1em]{logo-agda.pdf}}%
+}
+\newcommand{\agdalogo}{%
+  \usebox{\logoagdabox}}%
 
@@
 \drafttrue
-\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven}: #1]}\fi}
-\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max}: #1]}\fi}
-\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro}: #1]}\fi}
+\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven says}: #1]}\fi}
+\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max says}: #1]}\fi}
+\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro says}: #1]}\fi}
+\newcommand{\nathan}[1]{\ifdraft{\color{green}[{\bf Nathan says}: #1]}\fi}
 \newcommand{\pipe}{\,|\,}
 
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-%% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
-  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
-\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-  June 03--05, 2018, Woodstock, NY}
-\acmPrice{15.00}
-\acmISBN{978-1-4503-XXXX-X/18/06}
-
 \begin{document}
 
-\title{Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}
+\pagestyle{plain}
+
+\pagebreak
+
+\title{Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus}
+
 \author{Steven Schaefer}
-\affiliation{
-  \department{Electrical Engineering and Computer Science}
+\affiliation{\department{Electrical Engineering and Computer Science}
   \institution{University of Michigan}
@@
 \email{stschaef@umich.edu}
+
+\author{Nathan Varner}
+\affiliation{\department{Electrical Engineering and Computer Science}
+  \institution{University of Michigan}
+  \country{USA}
+}
+\email{nvarner@umich.edu}
 
@@
 
+\makeatletter
+\let\@authorsaddresses\@empty
+\makeatother
+
 \begin{abstract}
-  We propose a semantic framework for the study of formal
-  grammars. First, we provide a syntax-independent notion of formal
-  grammar as a function from strings to sets, generalizing the
-  familiar notion of a language as a set of strings. The set of such
-  grammars naturally form a very rich category whose notion of
-  isomorphism corresponds to strong equivalence of grammars, and for
-  which many language-theoretic constructs can be defined using
-  universal properties.
-
-  Based on these category-theoretic constructions we propose a new
-  syntactic formalism for formal grammars: a version of dependent
-  linear-non-linear type theory in which the tensor product is not
-  commutative. The linear types can be interpreted as grammars, and
-  linear terms as parse transformers. The non-dependent fragment of
-  this type theory is already enough to express regular/context-free
-  expressions as well as finite automata, with the isomorphism between
-  regular languages and automata definable as linear terms. This
-  generalizes the language theoretic formalisms such as Kleene
-  algebras to incorporate ambiguity. By incorporating a dependency of
-  linear types on non-linear types, we can express richer grammar
-  formalisms such as indexed grammars, pushdown automata and Turing
-  machines. This provides a new form of ``logical characterization''
-  of grammar classes based on a substructural logic rather than
-  ordinary first-order logic. Further, the type theory can not only
-  express the grammars as types, but equivalences between grammars and
-  parsers as terms of the grammar.
-
-  We give this type theory a semantics in the category of grammars and
-  prove a canonicity theorem that shows that every term in a context
-  corresponding to a fixed string is equal to a term in a canonical
-  form encoding a parse tree of that string.
+  We propose \theoryname~(\theoryabbv), a domain-specific dependent type theory
+  for the definition of intrinsically verified parsers and verification of weak
+  and strong equivalences of formal grammars. The linear types in \theoryabbv
+  have a simple interpretation as formal grammars, with values of the types
+  interpreted as parse trees and general terms as a form of intrinsically
+  verified parse transformers, of which parsers being a special case. We
+  demonstrate the expressivity of this system by showing that the indexed
+  inductive linear types in this calculus can be used to encode many different
+  familiar grammar formalisms, including regular and context-free grammars but
+  also traces of various types of automata. We can then construct strong
+  equivalences between grammar and automata formalisms and construct verified
+  parsers as simple linear functional programs. All of our examples have been
+  implemented in a prototype implementation of \theoryabbv as a shallow
+  embedding in Agda.
+
+  Our shallow embedding is based on a denotational semantics of \theoryabbv
+  where non-linear types denote ordinary sets and linear types denote formal
+  grammars. To develop this semantics compositionally, we define a notion of
+  general notion of model we call a \emph{Chomsky hyperdoctrine}, a
+  generalization of the Kleene algebra approach to formal language theory. In
+  addition to our ``standard model'' where linear types denote formal grammars,
+  we demonstrate two non-standard models. The first interprets linear types as
+  semantic actions over formal grammars, suggesting future extensions of our
+  calculus. The second is used to structure a logical relations argument to
+  prove a canonicity metatheorem for the type theory, proving that all closed
+  terms of context-free linear types normalize to a parse tree.
 \end{abstract}
@@
 
-\section{A Syntax-Independent Notion of Syntax}
-
-The theory of formal languages and parsing is one of the oldest and
-most thoroughly developed areas of theoretical computer science. A
-prominent topic in the 1950s and 60s led to a series of remarkable
-developments: Chomsky's hierarchy of grammar formalisms, practical
-algorithms for parsing of regular and context-free grammars and
-variants, as well as implementations of practical tools for the
-generation of efficient parsers.
-
-A central notion is that of a \emph{formal language} $L$ over an
-alphabet $\Sigma$ as simply a \emph{subset} of the set of strings $L
-\subseteq \Sigma^*$. This definition is especially useful as it gives
-a semantics to formal grammars that is completely independent of any
-particular syntactic grammar formalism: any new notion of grammar can
-be given a semantics in this common framework and it provides a
-precise mathematical speicification for implementing a
-\emph{recognizer} of a language and comparing the power of different
-formalisms by demonstrating what languages can be formalized within
-it. This also allows for completely \emph{language theoretic}
-formulations of language classes: e.g., the regular \emph{languages}
-can be characterized as those that have finitely many derivatives
-\cite{brzozowski-or-myhill-nerode-idk}. This notion is remarkable
-because it makes no reference to any specific syntactic formalism for
-defining languages such as regular expressions or finite automata.
-
-Formal language theory alone is not sufficient as a specification of a
-parser: a language $L \subseteq \Sigma^*$ is equivalently defined as
-its indicator function $\chi_L : \Sigma^* \to \Prop$ mapping each
-string $w$ to the proposition that it is in the language $w \in
-L$. This can then serve as the specification for a language
-\emph{recognizer}: a program $r$ of type $\Sigma^* \to \Bool$ such
-that $r(w) = \texttt{true}$ if and only if $\chi_L(w)$ holds, but
-recognition is insufficient for most tasks for which the theory of
-parsing was developed: the production of \emph{semantic} structures
-from synactic descriptions in linguistics, compilation or
-deserialization. The output of a parser is not just a boolean but a
-\emph{parse tree}\footnote{such parse trees are usually not
-materialized in memory, as the construction of the parse tree is fused
-with application of semantic actions, which can be seen as a kind of
-fold over the generated tree.}.
-
-Due to this limitation, parsers are typically specified not by a
-formal language, but by some \emph{formal grammar}, which specifies
-not just \emph{which} strings are in the language but specifies what
-are the \emph{parse trees} for the string. However unlike formal
-language theory, there is no common universal notion of what
-constitutes a formal grammar, but rather many syntactic systems such
-as Chomskyan(?)  grammars, Thue systems, Montague grammars, etc which
-are all syntactic presentations of the same underlying idea of an
-abstract specification for parsing. The most common of these is the
-Chomsky hierarchy of \emph{generative} grammars which specify parse
-trees using \emph{derivations} of the string from a set of rewrite
-rules, the parse tree encoding which rules were used.
-
-Our first contribution is conceptual: we propose a simple,
-syntax-independent definition of \emph{formal grammar}:
+\section{Introduction}
+\label{sec:intro}
+
+\steven{TODO adapt section 1 and the abstract to reflect the rest of the paper}
+%% Outline
+%% - Why a framework for parser verification
+%% - what intrinsic verification gets us
+%% - the secret sauce: indexed linear inductive types
+%% - prototype implementation
+%% - overview of "abstract" formal grammars?
+
+Parsing data from a serialized format is one of the most common tasks
+in all of computing. Accordingly, this is a quite well-studied problem
+spanning theoretical computer science and linguistics with such
+milestones as Chomsky's hierarchy of grammars
+\cite{chomThreeModels1956}, practical algorithms for parsing of
+regular and context-free grammars, and variants
+\cite{KNUTH1965607,Earley1970}, as well as implementations of
+practical tools for the generation of efficient parsers
+\cite{Johnsonyacc}.
+
+The theory of formal languages and parsing is one of the oldest and most
+thoroughly developed area of theoretical computer science.
+The central object of study is a \emph{formal language} $L$ over an alphabet
+$\Sigma$, which is classically defined as a
+\emph{subset} of strings $L \subseteq \String = \Sigma^{*}$. The expressivity of formal
+languages is sufficient for defining a \emph{recognizer} of a language $L$:
+a function $r : \String \to \Prop$ such that $r$ maps each string $w$ to the
+proposition that $w$ is in the language $w \in L$. This technique of language
+recognition can concisely specify lexical analysis, and even lead to verified
+lexer implementations \cite{egolfVerbatim,Ouedraogo_2023}.
+
+While there are ample theory and tooling to handle the first stage of a language frontend, the final
+frontier for parsing research is the development of
+\emph{formally verified} parser implementations. Due to the
+ubiquitous nature of parsing, these would useful components of many
+formally verified software systems: compilers and servers most
+especially. However, language recognition is insufficient for specifying a
+parser, and for this task much of the structured foundations from the theory of
+formal languages is rendered moot. Rather than returning a proposition-valued result, a parser produces a
+structured \emph{parse tree} adherent to a \emph{formal grammar}.
+The research on such grammars is often performed atop an ad-hoc choice of grammar
+formalism, and results are proven with respect to that formalism. Critically,
+the field is lacking a unifying theory across these different formalisms, even
+though many of the same
+compositional reason principles appear across distinct grammar formalisms.
+
+In this work, we aim to elucidate these similarities and
+provide a mathematical theory of formal grammars for constructing verified
+parsers. Our approach can be summarized with the
+Curry-Howard-style slogan: ``grammars as types''; and the addendum: ``parsers as
+terms''. That is, we develop a type theory with a novel
+denotational semantics in a category with abstract formal grammars as
+objects and \emph{parse transformers} as morphisms. We say these are
+\emph{abstract} formal grammars because they are not tied to any
+particular syntactic formalism such as Chomskyan grammars, regular
+expressions, Lambek calculus etc. The category we choose is a rich
+in structure: it is a topos as well as being monoidal closed, and these
+universal properties serve as the inspiration for our type theory.
+
+The type theory itself is a form of dependent linear-non-linear type
+theory: the linear types can be interpreted as grammars, whereas the
+non-linear types can be interpreted as ordinary sets. As in prior work
+on dependent linear-non-linear types, the linear types are allowed to
+depend on the non-linear types but not vice-versa \cite{krishnaswami_integrating_2015}.
+
+The substructural nature of \theoryabbv is well-aligned with the
+requirements intrinsic to parsing and to the theory of formal
+languages, where strings constitute a very clear notion of resource
+that cannot be duplicated, reordered, or dropped. Moreover, the constructive
+aspect of \theoryabbv ensures that verification of parsers written in the
+calculus are \emph{correct-by-construction}. The type system is powerful enough
+that derivations of a term type checking carry intrinsic proofs of correctness.
+
+\paragraph{Contributions} This paper begins in \cref{sec:synindnotion} by
+giving a universal semantic characterization of formal grammar.
+From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type theory
+that axiomatizes the underlying structures needed to conduct formal grammar
+theory. In \cref{sec:automata} we demonstrate the usage of
+\theoryname for defining parser terms and internalizing results from formal language theory.
+Then in \cref{sec:categorify}, we characterize the abstract categorical
+models of our syntax. Afterwards in \cref{sec:othermodels}, we demonstrate several models,
+including one to prove a metatheoretic canonicity theorem.
+Our contributions are then:
+%
+\begin{itemize}
+  \item A syntax-independent notion of grammar given by the category $\Grammar$
+  \item \theoryname: A dependent
+    linear-non-linear type theory meant to syntactically capture the latent structures
+    in $\Grammar$ that enable parsing
+  \item Categorical structures that capture the concepts integral to formal grammar theory, and that are notably modeled by $\Grammar$
+  \item An implementation of the language, which is embedded shallowly in Agda
+\end{itemize}
+
+\subsection{A Syntax-Independent Notion of Syntax}
+\label{sec:synindnotion}
+
+The notion of formal language is central to the theory of parsing.
+This definition is
+especially useful as it gives a semantics to formal grammars that is completely
+independent of any particular syntactic grammar formalism. Any new
+notion of grammar can be given a language semantics and it provides a precise
+mathematical specification for implementing a \emph{recognizer} of a language.
+In this section we propose a \emph{proof-relevant} generalization of formal
+languages that can accommodate richer aspects of formal language theory such as
+parsing. We begin by proposing a presheaf approach to grammars that encompasses
+other classes of grammars in a syntax-independent way.
+
+\paragraph{From Formal Languages to Formal Grammars}
+
+Our first contribution is a novel and syntax-independent characterization of
+formal grammar:
+
 \begin{definition}
-  A \emph{formal grammar} over an alphabet $\Sigma$ is a function
-  $\Sigma^* \to \Set$.
+  \label{def:grammar}
+  A \emph{formal grammar} $G$ over a fixed alphabet $\Sigma$ is a function
+  $G : \String \to \Set$.
 \end{definition}
-We say that a grammar $G$ associates to every string $w$ the set $Gw$
-of \emph{parses} of that string. While we will work informally in this
-paper, the collection $\Set$ can be formalized in many different
-logical foundations: as a universe in type theory, a proper class in
-set theory, etc. We argue that this is already a common intuition
-latent in much prior work on grammars, especially when comparing
-different formalisms for equivalence such as \cite{??}. This notion is
-also completely natural in the context of constructive type theories
-such as Agda which until recently did not include a universe of
-propositions, and so for example in Elliott (\cite{??}) a formal
-language was defined as a function to the universe $\Type$ with little
-comment.
-
-Every formal grammar $G$ induces a formal language, which can be seen
-most easily by using the characteristic function formulation: we can
-``squash'' any set into the proposition that it is inhabited, so
-$\chi_G(w) = ||G(w)||$ which defines a subset $L_G = \{ w \in
-\Sigma^*| G(w) \mathrm{inhabited}\}$.
-
-Formal languages naturally arrange themselves into a partial order by
-using the subset inclusion as the ordering. Formal grammars naturally
-arrange themselves into a \emph{category} where a morphism $\alpha : G
-\to H$ is given by a \emph{family} of functions
-\[ \alpha^w : Gw \to Hw \]
-for every string $w \in \Sigma^*$. The intuition is that a morphism of
-grammars is a \emph{translation of parses}: a $G$-parse of $w$ gets
-transformed to an $H$-parse of $w$. This then induces a notion of
-\emph{isomorphism} of grammars: two grammars are equivalent if there
-is a bijective translation of parses, precisely capturing the notion
-of a \emph{strong equivalence} of grammars.
-
-This category-theoretic framework can be further used to describe and
-compare different notions of formal grammar. We can define a
-\emph{notion of grammar} to be a category (or perhaps just groupoid)
-paired with a functor into $\Set^{\Sigma^*}$. This naturally forms a
-2-category the slice 2-category over $\Set^{\Sigma^*}$ and equivalence
-in this 2-category defines a sensible notion of \emph{strong
-equivalence of grammar formalisms} (TODO: check details).
-
-The category of grammars is quite rich in structure, which can be most
-easily observed from the fact that it is equivalently defined as the
-category of functors $\Set^{\Sigma^*}$ where $\Sigma^*$ is here viewed
-as a discrete category, that is, the objects are strings and only
-morphisms are identity morphisms. Such functor categories, often
-called presheaf categories, are incredibly rich in structure, which we
-will exploit in section \ref{blah} to succinctly define the semantics
-of regular and context-free grammars.
-
-\section{Kleene Category}
-
-Kleene algebras are an important tool in the theory of regular
-languages. More broadly, they serve as a theoretical substrate to
-studying various kinds of formal languages. Formally, they are a tuple
-$(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is a set, $+$ and $\cdot$
-are binary operations over $A$, $(-)^*$ is a function over $A$, and
-$1$ and $0$ are constants. These structures satisfy the axoims depicted
-in Figure~\ref{fig:axioms}.
+
+\steven{TODO need to go through and ensure that $\Sigma^{*}$ and $\String$ are each used appropriately}
+
+We say that a grammar $G$ associates to every string $w$ the set\footnote{We
+  make little comment as to which foundations are used to encode $\Set$. Our
+  construction is polymorphic in choice of a proper class of small sets, a
+  Grothendieck universe, a type theoretic universe, or any similar foundation.} $G w$ of parse
+trees showing that $w$ matches $G$. This definition of formal grammar serves as
+a mathematical specification for a parser --- taking on the same role as
+languages do for recognizers. In this sense, our new characterization of
+grammars is just a proof-relevant generalization of the existing notion of
+formal language.
+
+By ranging over all input strings, notice that each grammar induces a language.
+For each formal grammar $G$, define the language of accepted strings $L_{G}$ as,
+
+\[ L_{G} = \{ w \in \String ~|~ G w~\text{is inhabited} \} \]
+
+As they are simply sets, formal languages are naturally endowed with a partial
+order via subset inclusion. Just, formal grammars naturally coalesce into a
+\emph{category}.
+
+Define $\Grammar$ to be the category of formal grammars. Whose
+objects are grammars, and whose morphisms are functions $f : G \to H$ between two
+grammars presented by a family of functions $f^{w} : G w \to H w$ for
+$w : \String$. Intuitively, we interpret these morphisms as parse
+\emph{transformers} --- as constructive translations of $G$-parses into $H$-parses. Throughout this paper, we will make frequent reference to this categorical
+structure on $\Grammar$.
+
+\steven{Above sounds awkward. I want to additionally say something here that signposts that we will give more
+  information about that structure without just listing it all here. }
+
+\paragraph{From Grammars to Formal Grammars}
+The category-theoretic framework can be further used to describe and compare
+different notions of formal grammar. For instance, Chomskyan generative
+grammars, semi-Thue systems, Montague grammars, and so on are all distinct syntactic
+presentations of the same underlying idea of an abstract specification for
+parsing. Each of these is an instance of a general \emph{notion of formal
+  grammar} --- a category paired with a functor into
+$\Grammar$. That is, they may all be interpreted in our category of grammars. In
+this sense, $\Grammar$ provides an
+all-encompassing semantic domain for parsing.
+
+Upon further inspection, $\Grammar$ may equivalently
+be described as the functor category $\Set^{\String}$ --- where the
+set $\String$ is viewed as a discrete\footnote{Objects in $\String$
+comprise the set of all strings, and the only morphisms are the
+identity morphisms.} category. In fact, in this paper we will take
+$\Grammar$ to be defined as the functor category $\Set^{\String}$. Such functor
+categories in $\Set$, called presheaf categories, carry a
+remarkable amount of structure.  Indeed they are a monoidal topos and,
+as such, are a model to a substructural dependently typed theory.
+
+Thus we
+present \theoryname (\theoryabbv) as a domain-specific type theory that
+abstracts over the structure of $\Grammar$. To begin, let us look at what some
+programs in \theoryabbv look like.
+
+\steven{TODO Reword this transition into the examples}
+
+% \max{this section kind of trails off at the end with a very vague aside about a $2$-category of grammar formalisms.
+Maybe this should go in future work/discussion?
+What is the message of this section?
+How does it tie in with the message of the paper?
+What even is the message of this paper?}
+
+% \pedro{Would building on this proposed structure make the message of the section clearer? In this case
+%   we would have to be a bit clearer in terms of how familiar notions of grammar can be seen as categories with functor into $\Grammar$.}
+
+\section{\theoryname by Example}
+\label{sec:type-theory-examples}
+To gain intuition for working in \theoryabbv, we begin with some illustrative
+examples drawn from the theory of formal languages. In particular, we
+demonstrate how to encode finite grammars, regular expressions, and finite
+automata in \theoryabbv. Each of our examples will be defined for strings over the three character alphabet
+$\Sigma = \{ \texttt{a} , \texttt{b}, \texttt{c} \}$.
+
+\newcommand{\A}{\texttt{A}}
+\newcommand{\B}{\texttt{B}}
+\newcommand{\I}{\texttt{I}}
+\newcommand{\f}{\texttt{f}}
+\newcommand{\g}{\texttt{g}}
+\renewcommand{\L}{\texttt{L}}
+\renewcommand{\a}{\texttt{a}}
+\renewcommand{\b}{\texttt{b}}
+\renewcommand{\c}{\texttt{c}}
+\newcommand{\w}{\texttt{w}}
+
+\paragraph{Finite Grammars}
+First consider finite grammars --- those built from base types via disjunctions and
+concatenations. The base types comprise characters drawn from the alphabet, the
+empty string, and the empty grammar.
+For instance, the grammar for the single character $\a$ has a single parse tree for the string
+\stringquote{a} and no parse trees at any other strings. The grammar $\I$ recognizes has a single
+parse tree for the empty string $\epsilon = \stringquote{}$ and no parses for any other strings.
+The final base type, the empty grammar $\bot$, has no parses for any string. We
+use type-theoretic syntax to represent disjunction $\oplus$ and concatenation
+$\otimes$ of
+grammars. Over an input string $\w$, a parse of the disjunction $\A \oplus \B$ is either
+a parse of $\A$ over the string $\w$ or a
+parse of $\B$ over the string $\w$. Similarly, $\w$ matches $\A \otimes \B$ if
+$\w$ can be split into two strings $\w_{\A}$ and $\w_{\B}$ that match $\A$
+and $\B$, respectively.
+
+For a grammar $\A$, a parse tree of a string $\w$ can be represented as
+a term of type $\A$ in the context $\internalize \w$, where
+$\internalize \w$ is a context with one variable for each character of
+$\w$. For example, to define a parse tree for \stringquote{ab}, we use
+the context $\internalize{\stringquote{ab}} = a : \a , b : \b$. Then the left of
+\cref{fig:fingram} gives a typing derivation that shows that $\inl(a
+\otimes b)$ defines a valid parse tree for the finite grammar $(\a \otimes \b)
+\oplus \c$. On the right of \cref{fig:fingram}, we give a linear $\lambda$-term
+that implements the program given in the derivation.
+
+\begin{figure}
+\begin{minipage}{0.5\textwidth}
+\begin{mathpar}
+  \inferrule
+  {
+    \inferrule
+    {
+      \inferrule
+      {~}
+      {a : \a \vdash a : \a}
+      \\
+      \inferrule
+      {~}
+      {b : \b \vdash b : \b}
+    }
+    {a : \a , b : \b \vdash a \otimes b : \a \otimes \b}
+  }
+  {a : \a , b : \b \vdash \texttt{f} := \inl(a \otimes b) : (\a \otimes \b) \oplus \c}
+\end{mathpar}
+\end{minipage}%
+\begin{minipage}{0.5\textwidth}
+\begin{lstlisting}
+f : (*@$\uparrow$@*)(a (*@$\otimes$@*) b (*@$\lto$@*) (a (*@$\otimes$@*) b) (*@$\oplus$@*) c)
+f (a' , b') = inl (a' , b')
+\end{lstlisting}
+\end{minipage}
+\caption{\stringquote{ac} matches $(a \oplus b) \otimes c$}
+\label{fig:fingram}
+\end{figure}
+For this interpretation of terms as parse trees to make sense, our
+calculus cannot allow for \emph{any} of the usual structural rules of
+a type theory: weakening, contraction and exchange. Weakening allows
+for variables to go unused, while contraction allows for the same
+variable to be used twice, but in a parse tree, every character must
+be accounted for exactly once. That is, we want to prevent the following
+erroneous derivations,
+\[
+  a : \a , b : \b \not \vdash a : \a \qquad a : \a , b : \b \not \vdash (a, a) : \a \otimes \a
+\]
+
+%
+Finally, the ordering of characters in a string cannot be ignored while
+parsing, so we omit the exchange rule because it would allow
+for variables in the context to be reordered,
+\[
+  a : \a , b : \b \not \vdash (b , a) : \b \otimes \a
+\]
+% otherwise we would have that $\inl(x\otimes y)$ is a valid
+% term in context $y:b,x:a$, but there should be no parses of $ba$ in
+% this grammar.
+
+\paragraph{Regular Expressions}
+Regular expressions can be encoded as types generated by base types,
+$\oplus$, and $\otimes$, and the Kleene star $(\cdot)^{\ast}$.  For a
+grammar $\A$, we can define the Kleene star $\A^{*}$ as a particular
+\emph{inductive linear type} of linear lists, as shown in
+\cref{fig:kleenestarinductive}. Here $\A^{*} : \L$ means we are defining
+a \emph{linear} type. $\A^{*}$ has two constructors: $\nil$, which
+builds a parse of type $\A^{*}$ from nothing; and $\cons$, which
+linearly consumes a parse of $\A$ and a parse of $\A^{*}$ and builds a
+parse of $\A^{*}$. This linear consumption is defined by the linear
+function type $\lto$. The linear function type $\A \lto \B$ defines functions that
+take in parses of $\A$ as input, \emph{consume} the input, and return a parse of
+$\B$ as output. The arrow, $\uparrow$, wrapping these
+constructors means that the constructors by themselves are not consumed
+upon usage, and so are \emph{non-linear} values themselves. That is, the
+names $\nil$ and $\cons$ are function symbols that may be reused as many times
+as we wish.
+
+\begin{figure}
+\begin{lstlisting}
+data A(*@$^*$@*) : L where
+  nil : (*@$\uparrow$@*)(A(*@$^*$@*))
+  cons : (*@$\uparrow$@*)(A (*@$\lto$@*) A(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
+\end{lstlisting}
+% \begin{align*}
+% \data &~A^{*} : L~\where\\
+%       & \nil : \ltonl {A^{*}} \\
+%       & \cons : \ltonl {(A \lto A^{*} \lto A^{*})}
+% \end{align*}
+\caption{Kleene Star as an inductive type}
+\label{fig:kleenestarinductive}
+\end{figure}
+
+Through repeated application of the Kleene star constructors, \cref{fig:kleenestarderivation} gives a derivation that shows \stringquote{ab} matches the
+regular expression $(\a^{*} \otimes \b) \oplus \c$. The leaves of the proof tree
+that mention the arrow $\uparrow$ describe a cast from a non-linear type to a linear type.
+For instance, the premise of the leaf involving $\nil$
+views $\nil : \ltonl {(\a^{*})}$ as the name of a
+constructor, and a constructor should be nonlinearly valued because we may call
+it several time. However, the conclusion of this leaf views $\nil : \a^{*}$ as a
+linear value. After we call the constructor it returns a value that may only be
+used a single time. In this paper such casts are often made implicitly, but this
+does not introduce any ambiguity.
+Even though we use $\nil$ to refer to both a constructor and the
+linear value produced by it
+, we differentiate between the two views by inspecting
+the type.
+
+\begin{figure}
+\begin{mathpar}
+\inferrule
+{
+  \inferrule
+  {
+    \inferrule
+    {
+      \inferrule
+      {
+        \inferrule
+        {
+          \inferrule
+          {~}
+          {\cdot \vdash \cons : \ltonl {(\a \lto \a^* \lto \a^*)}}
+        }
+        {\cdot \vdash \cons : \a \lto \a^* \lto \a^*}
+        \\
+        \inferrule
+        {~}
+        {a : \a \vdash a : \a}
+      }
+      {a : a \vdash \cons~a : \a^* \lto \a^*}
+      \\
+      \inferrule
+      {
+        \inferrule
+        {~}
+        {\cdot \vdash \nil : \ltonl {(\a^*)}}
+      }
+      {\cdot \vdash \nil : \a^*}
+    }
+    {a : \a \vdash \cons~a~\nil : \a^{*}}
+    \\
+    \inferrule
+    {~}
+    {b : \b \vdash b : \b}
+  }
+  {a : \a , b : \b \vdash (\cons~a~\nil) \otimes b : \a^{*} \otimes \b}
+}
+{a : \a , b : \b \vdash \texttt{g} := \inl ((\cons~a~\nil) \otimes b) : (\a^{*} \otimes \b) \oplus \c}
+\end{mathpar}
+\begin{lstlisting}
+g : (*@$\uparrow$@*)((a (*@$\otimes$@*) b) (*@$\lto$@*) (a(*@$^*$@*) (*@$\otimes$@*) b) (*@$\oplus$@*) c)
+g (a' , b') = inl (cons a' nil (*@$\otimes$@*) b')
+\end{lstlisting}
+\caption{\stringquote{ab} matches $(a^{*} \otimes b) \oplus c$}
+\label{fig:kleenestarderivation}
+\end{figure}
+
+We may also have derivations where the term in context is not simply a
+string of base types. In \cref{fig:kleeneabstractproof} we show that every parse
+of the grammar $(\A \otimes \A)^{*}$ induces a parse of $\A^{*}$ for an
+arbitrary grammar $\A$. The context $(\A \otimes \A)^{*}$ does not correspond directly to a string, so it is
+not quite appropriate
+to treat linear terms here as parse \textit{trees}.
+The
+context $a : (\A \otimes \A)^{*}$ does not contain concrete data to be parsed; rather, there may be many choices of string underlying the parse tree captured
+by the variable $a$. Thus, the term $\texttt{h}$ from
+\cref{fig:kleeneabstractproof} is not a parse of a string, and
+it is more
+appropriate to think of it as a parse \textit{transformer} --- a function from
+parses of $(\A \otimes \A)^{*}$ to parse of $\A^{*}$.
+
+$\texttt{h}$ is defined by recursion on terms of type $(\A \otimes \A)^{*}$.
+This recursion is expressed in the derivation tree by invoking the
+elimination principle for Kleene star, written
+as $\fold$. $\texttt{h}$ is more compactly presented in the pseudocode
+of \cref{fig:kleeneabstractproof} by pattern matching on the input and making an
+explicit recursive call in the body of its definition. When appropriate we will favor this code-like
+presentation of terms in our calculus, as it is more readable and more
+faithful to the implementation of \theoryabbv in Agda.
+
+\begin{figure}
+\begin{mathpar}
+  \inferrule
+  {
+    \inferrule
+    {~}
+    {\cdot \vdash \nil : \A^*}
+    \\
+    \inferrule
+    {
+      \inferrule
+      {
+        \inferrule
+        {~}
+        {a_1 : \A, a_2 : \A , a_{*} : \A^* \vdash \cons (a_1 , \cons(a_2 , a_{*})) : \A^*}
+      }
+      {a' : \A \otimes \A , a_{*} : \A^* \vdash g := \letin {a_1 \otimes a_2} {a'} {\cons (a_1 , \cons(a_2 , a_{*})) : \A^*}}
+    }
+    {\cdot \vdash f := \lamblto a {\lamblto {a_{*}} {\app {\app g a} {a_{*}} }} : (\A \otimes \A) \lto \A^* \lto \A^*}
+  }
+  {a : (\A \otimes \A)^* \vdash \texttt{h} := \fold(\nil , f)(a) : \A^* }
+\end{mathpar}
+\begin{lstlisting}
+h : (*@$\uparrow$@*)((A (*@$\otimes$@*) A)(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
+h nil = nil
+h (cons (a (*@$\otimes$@*) a') as) = cons a (cons a' (h as))
+\end{lstlisting}
+\caption{A parse transformer for abstract grammars}
+\label{fig:kleeneabstractproof}
+\end{figure}
+
+\paragraph{Non-deterministic Finite Automata}
+\newcommand{\s}{\texttt{s}}
+\newcommand{\0}{\texttt{0}}
+\newcommand{\1}{\texttt{1}}
+\newcommand{\2}{\texttt{2}}
+Regular expressions are a compact formalism for defining a formal grammar,
+however simply defining a regular expression --- such as
+$(\a^{*} \otimes \b) \oplus \c$ --- does not give any insight into how to parse
+it. Many parsing algorithms are based on a notion of automaton, and to verify
+these algorithms we need to represent their associated automata in
+\theoryabbv.
+
+Finite automata are precisely the class of machines that recognize regular
+expressions. \cref{fig:exampleNFA} shows a non-deterministic
+finite automaton (NFA) for the regular expression $(\a^{*} \otimes b) \oplus c$.
+\cref{fig:nfainductive} defines $\Trace$ an \emph{indexed} inductive linear type of
+traces through this automaton. Defining an indexed inductive type can be thought
+of like defining a family of mutually inductive types varying coherently over a
+type of indexing data. That is, \cref{fig:nfainductive} defines the linear types
+$\Trace~\0$,
+$\Trace~\1$, and $\Trace~\2$ mutually inductively. The index $\s : \texttt{Fin 3}$ picks out which
+state in the automaton a trace begins at --- where $\texttt{Fin 3}$ is the
+finite type containing inhabitants $\{\0 , \1 , \2\}$.
+There are three kinds of constructors for $\Trace$:
+(1) those that terminate traces, (2) those that correspond to transitions
+labeled by a character, and (3) those that correspond to transitions labeled by
+the empty string $\epsilon$. The constructor $\texttt{stop}$ terminates a trace
+in the accepting state $\2$. The constructors $\texttt{1to1}$, $\texttt{1to2}$,
+$\texttt{0to2}$ each define a labeled transition through the NFA, and each of
+these consumes a parse of the label's character and a trace beginning at the
+destination of a transition to produce a trace beginning at the source of a
+transition. The constructor $\texttt{0to1}$ behaves similarly, except its
+transition is labeled with the empty string $\epsilon$. Therefore,
+$\texttt{0to1}$ takes in a trace beginning a state $\1$ and returns a trace
+beginning at state $\0$ corresponding to the same underlying string.
+
+Just as \cref{fig:kleenestarderivation} shows the \stringquote{ab} matches
+$(a^{*} \otimes b) \oplus c$, \cref{fig:nfaderivation} demonstrates that \stringquote{ab} is
+accepted by the NFA representation of this regular expression. In particular, we
+show that \stringquote{ab} matches the grammar $\Trace~0$ because $\0$ is the
+initial state of the NFA.\ We may imagine defining parse transformers from
+$(\a^{*} \otimes b) \oplus c$ to $\Trace~\1$, and vice versa. The existence of
+such maps between the two types would demonstrate that the regular expression
+and the automaton capture the same language --- i.e.\ they recognize the same
+set of strings. As we will see in LATER SECTION, we can define parse
+transformers between regular expressions and their associated automaton.
+Moreover, we prove that these parse transformers define a bijective
+correspondence of parses over every input string. That is, in a strong sense the
+parse of a regular expression and the traces of their associated automata
+capture precisely the same data.
+
+\begin{figure}
+  \begin{tikzpicture}[node distance = 25mm ]
+    \node[state, initial] (0) {$\0$};
+    \node[state, below left of=0] (1) {$\1$};
+    \node[state, right of=1, accepting] (2) {$\2$};
+
+    \path[->] (0) edge[above] node{$\epsilon$} (1)
+              (0) edge[right] node{$\stringquote{c}$} (2)
+              (1) edge[loop left] node{$\stringquote{a}$} (1)
+              (1) edge[below] node{$\stringquote{b}$} (2);
+  \end{tikzpicture}
+  \caption{NFA for $(\a^{*} \otimes \b) \oplus \c$}
+  \label{fig:exampleNFA}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}
+data Trace : (s : Fin 3) (*@$\to$@*) L where
+  stop : (*@$\uparrow$@*)(Trace 2)
+  1to1 : (*@$\uparrow$@*)(a (*@$\lto$@*) Trace 1 (*@$\lto$@*) Trace 1)
+  1to2 : (*@$\uparrow$@*)(b (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 1)
+  0to2 : (*@$\uparrow$@*)(c (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 0)
+  0to1 : (*@$\uparrow$@*)(Trace 1 (*@$\lto$@*) Trace 0)
+\end{lstlisting}
+\caption{$N$ as an indexed inductive type}
+\label{fig:nfainductive}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}
+k : (*@$\uparrow$@*)((a (*@$\otimes$@*) b) (*@$\lto$@*)  Trace 0)
+k (a' , b') = 0to1 (1to1 a' (1to2 b' stop))
+\end{lstlisting}
+\caption{\stringquote{ab} is accepted by the NFA for $(a^{*} \otimes b) \oplus c$}
+\label{fig:nfaderivation}
+\end{figure}
+
+%% 3. Third example: NFA traces to show the indexed inductive types
+
+\section{Syntax and Typing for \theoryname}
+\label{sec:tt}
+
+%% OUTLINE
+%%
+%% 1. Overview of syntax: what are the judgments
+%% 2. Cover the linear types except for inductives
+%% 3. Cover non-linear types, mainly focus on the universes
+%% 4. Cover indexed inductive linear types
+%% 5. Cover the "axioms" we add
+
+\max{was this paragraph supposed to be deleted?}  Next we present the
+full details of the syntax and typing of \theoryabbv. The syntax of
+\theoryabbv is based on the dependent linear-non-linear (\lnld)
+calculus from \cite{dlnl}. As in \lnld, \theoryabbv includes both
+non-linear dependent types, as well as linear types, which are allowed
+to depend on the non-linear types, but not on other linear types.
+%
+The main point of departure from \lnld's design in that the linear
+typing is \emph{non-commutative}, i.e., that exchange is not an
+admissible structural rule. Furthermore, we add a general-purpose
+indexed inductive linear type connective, as well as an
+\emph{equalizer} type, which we will show allows us to perform
+inductive proofs of equalities between linear terms.
+%
+Finally, we add base types and axioms to the system specifically to
+model formal grammars and parsing.
+
+The formation rules for the judgments of \theoryabbv are shown in
+\Cref{fig:formation}. $\Gamma$ stand for non-linear contexts, $X,Y,Z$
+stand for non-linear types, $M,N$ stand for non-linear terms, these
+act as in an ordinary dependent type theory. $\Delta$ stands for
+linear contexts, $A,B,C$ for linear types and $e,f,g$ for linear
+terms. These contexts, types and terms are allowed to depend on an
+ambient non-linear context $\Gamma$, but note that linear types $A$
+cannot depend on any \emph{linear} variables in $\Delta$. We include
+definitional equality judgments for both kinds of type and term
+judgments as well. Additionally, we have a judgment $\Gamma\vdash X
+\isSmall$ which is used in the definition of universe types.
+
+\begin{figure}
+  \begin{mathpar}
+    \inferrule{~}{\ctxwff \Gamma}
+
+    \inferrule{\ctxwff \Gamma}{\ctxwffjdg \Gamma X}
+
+    \inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash X \isSmall}
+
+    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg \Gamma Y}{\ctxwffjdg \Gamma {X \equiv Y}}
+
+    \inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash M : X}
+
+    \inferrule{\Gamma \vdash M : X \and \Gamma \vdash N : X}{\Gamma \vdash M \equiv N : X}
+
+    \inferrule{\ctxwff \Gamma}{\linctxwff \Gamma \Delta}
+
+    \inferrule{\ctxwff \Gamma}{\Gamma \vdash A \isLinTy}
+
+    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \equiv B}
+
+    \inferrule{\Gamma \vdash \Delta \isLinCtx \and \Gamma \vdash A \isLinTy}{\Gamma;\Delta\vdash e : A}
+
+    \inferrule{\Gamma; \Delta \vdash e : A \and \Gamma;\Delta \vdash f : A}{\Gamma;\Delta\vdash e \equiv f : A}
+  \end{mathpar}
+  \caption{Formation Rules}
+  \label{fig:formation}
+\end{figure}
+
+\subsection{Non-linear Typing}
+
+The non-linear types of \theoryabbv include the standard dependent
+types for $\Pi,\Sigma$, extensional equality, natural numbers,
+booleans, unit and empty types. We include these rules in the
+appendix. We present the other non-linear type constructions in
+\Cref{fig:non-linear-stuff}. First, we include universe types $U$ of
+small non-linear types and $L$ of linear types. These are defined as
+``universes ala Coquand'' in that we define a judgment saying when a
+non-linear type is small and define the universe to internalize
+precisely this judgment. These universe types are needed so that we
+can define types by recursion on natural numbers. Next, we include a
+non-linear type $\ltonl A$ where $A$ is a linear type. The intuition
+for this type is that its elements are the linear terms that are
+``resource free'': its introduction rule says we can construct an
+$\ltonl A$ when we have a linear term of type $A$ with no free linear
+variables. This type is used extensively in our examples, playing a
+similar role to the $!$ modality of ordinary linear logic.
+
+%% \begin{figure}
+%%   \[
+%%   \begin{array}{rrcl}
+%%     \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
+%%     \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe
+%%                                                   \mathsf{Bool} \pipe
+%%                                                   \mathsf{Empty} \pipe
+%%                                                   \mathsf{Unit} \pipe
+%%                                                   \mathsf{SPF}\,X\pipe U \pipe L
+%%                                                   \pipe \ltonl A \pipe \sum\limits_{x:X} Y
+%%                                                   \pipe \prod\limits_{x:X} Y\pipe M =_A
+%%                                                   N
+%%   \end{array}
+%%   \]
+%%   \caption{Non-linear Types}
+%% \end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \inferrule{}{\Gamma \vdash U \isTy}\and
+    \inferrule
+    {\Gamma \vdash M : U}
+    {\Gamma \vdash \unquoteTy M \isTy}\and
+    \inferrule
+    {\Gamma \vdash X \isTy}
+    {\Gamma \vdash \quoteTy X : U}\and
+    \unquoteTy{\quoteTy{X}} \equiv X \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}
+
+    \inferrule{}{\Gamma \vdash L \isTy}\and
+    \inferrule
+    {\Gamma \vdash M : L}
+    {\Gamma \vdash \unquoteTy M \isLinTy}\and
+    \inferrule
+    {\Gamma \vdash A \isLinTy}
+    {\Gamma \vdash \quoteTy A : L}\and
+    \unquoteTy{\quoteTy{A}} \equiv A \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}
+
+    \inferrule
+    {\Gamma \vdash A \isLinTy}
+    {\Gamma \vdash \ltonl { A } \isTy}
+    \and
+    \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash e : \ltonl A}\and
+    \inferrule{\Gamma \vdash M : \ltonl A}{\Gamma ; \cdot \vdash M : A} \and
+  \end{mathpar}
+  \caption{Non-linear types (selection)}
+  \label{fig:non-linear-types}
+\end{figure}
+
+\subsection{Linear Typing}
+
+We give an overview of the linear types in \Cref{fig:linear-types} and
+terms in \Cref{fig:linear-terms}. The equational theory for these
+types is straightforward and included in the appendix. First, the
+linear variable rule says that a linear variable can be used if it is
+the \emph{only} variable in the context.
+
+First, we cover the ``multiplicative'' connectives of non-commutative
+linear logic. The linear unit and tensor product are standard for a
+non-commutative linear logic: when we construct a linear unit we
+cannot use any variables and when we construct a tensor product, the
+two sides must use disjoint variables, and the variables the left side
+of the product uses must be to the left in the context of the
+variables used by the right side of the tensor product. Because we are
+non-commutative, there are two function types: $A \lto B$ and $B \tol
+A$. The difference between these is that the introduction rule for $A
+\lto B$ adds a variable to the right side of the context, whereas the
+introduction rule for $B \tol A$ adds a variable to the left side of
+the context. In our experience, because by convention parsing
+algorithms parse from left-to-right, we rarely need to use the $B \tol
+A$ connective. One of the main uses of linear types is that, in
+combination with the $\uparrow$ connective, we can abstract over
+linear terms, in that the non-linear type $\ltonl{A \lto B}$ is a
+non-linear type of linear terms with one linear input $A$ and output
+$B$.
+
+Next, we have the ``additive'' connectives. For these, we use the
+non-linear types to define \emph{indexed} versions of the additive
+disjunction $\oplus$ and additive conjunction $\&$ of linear
+logic. These are like a linear version of the $\Sigma$ and $\Pi$
+connectives of ordinary dependent type theory. The indexed $\&$ is
+defined by a $\lambda$ and eliminated using projection. The indexed
+$\oplus$ is like a \emph{weak} $\Sigma$ type: it has an injection
+introduction rule $\sigma$, but its elimination rule is given by
+\emph{pattern matching} rather than projections. Note that we can
+define the more typical nullary and binary versions of these
+connectives by using indexing over the empty and boolean type
+respectively. We will freely use $\bot$ to refer to this empty
+discunction and $\top$ to refer to the empty conjunction. Lastly, we
+include a type $\equalizer {a}{f}{g}$ that we call the
+\emph{equalizer} of $f$ and $g$. We think of this type as the
+``subset'' of $A$ that satisfy the equation $f\,a\equiv g\,a$. If
+these were non-linear types, such a type could be constructed using
+the equality type as $\sum_{x:X} f\,x=_Y g\,x$, but this construction
+can't be used because it uses a \emph{dependent} equality type. We
+note that the equalizer type is not used in defining any of our
+parsers or formal grammars, but instead will be used shortly to derive
+an induction principle for proving equalities of functions whose
+domain is an inductive type.
+
+\begin{figure}
+  \begin{mathpar}
+    \inferrule{}{\Gamma \vdash I \isLinTy}\and
+    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \otimes B \isLinTy}\and
+    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \lto B \isLinTy}\and
+    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \tol B \isLinTy}\and
+    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigoplus\limits_{x:X} A}\and
+    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigamp\limits_{x:X} A}\and
+    \inferrule
+    {\Gamma \vdash f : \ltonl {(A \lto B)} \and \Gamma \vdash g : \ltonl { (A \lto B) }}
+    {\Gamma \vdash \equalizer {a}{f}{g} \isLinTy}
+  \end{mathpar}
+  \caption{Linear Types (selection)}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \inferrule{~}{\Gamma ; a : A \vdash a : A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
+    %
+    \\
+    %
+    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
+    %
+    \\
+    %
+    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
+    %
+    \and
+    %
+    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
+    \\
+    %
+    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
+    \and
+    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta, \Delta' \vdash \applto {e'} {e} : B}
+    \\
+    %
+    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}
+              {\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash M : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {e'} x}
+    %
+    \\
+    %
+    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A e x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
+    %
+    \and
+    %
+    \inferrule{\Gamma ; \Delta \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {\sigma\,x\,a} e {e'}: C}
+    %
+    \and
+    %
+    \inferrule
+    { \Gamma ; \Delta \vdash e : A \\
+      \Gamma ; \Delta \vdash \applto {f}{e} \equiv \applto {g}{e}}
+    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {e}{f}{g}}
+    %
+    \and
+    %
+    \inferrule
+    {\Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
+    {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
+  \end{mathpar}
+  \caption{Linear terms (selection)}
+  \label{fig:linsyntax}
+\end{figure}
+
+\subsection{Indexed Inductive Linear Types}
+
+Next, we introduce one of the most complex and important linear
+type constructors of our development, \emph{indexed inductive linear
+types} $\mu F$. To encode these, we first add a non-linear type
+$\SPF\,X$ of \emph{strictly-positive functorial} linear type
+expressions indexed by a non-linear type $X$. We give the rules for
+strictly-positive functors in \Cref{fig:spf}. First, we add operations
+for a recursive reference $\Var{x}$ the constant functor, $K$, tensor
+products and additive conjunction and disjunction. As an example,
+\max{TODO: make an example}.  Next we add an operation $\el(F)$ which
+interprets our syntactic codes as an actual type constructor, and
+define it for each of the operations given above. Lastly, we give a
+\emph{functorial action} $\map(F)$ for each of the codes.\max{TODO:
+  find a citation for this kind of approach. Container papers work
+  this way for instance. I think there is a linear container paper
+  that uses this appraoch}
+
+With this notion of strictly-positive functor so defined, in
+\Cref{fig:iilt} we next add a linear type constructor $\mu F$, which
+defines a family of mutually-recursive inductive types for the family
+of type expressions $F$. The introduction rule for this is
+$\mathsf{roll}$, which constructs an element of $\mu F\,x$ from its
+one-level unfolding. The elimination principle is defined by a
+$\mathsf{fold}$ operation. The equations ensure that this makes the
+family an \emph{initial algebra} for the type expressions. That is the
+$\beta$ rule says that a $\mathsf{fold}$ applied to a $\mathsf{roll}$
+is equivalent to $\mathsf{map}$ping the $\mathsf{fold}$ over all the
+sub-expressions, which means that $\mathsf{fold}$ is an algebra
+homomorphism from $\mathsf{roll}$ to $\mathsf{f}$. Then the $\eta$
+rule says that $\mathsf{fold}$ is the \emph{unique} such homomorphism.
+
+This definition as an initial algebra is well-understood semantically
+but the $\eta$ principle in particular is somewhat cumbersome to use
+directly in proofs. In dependent type theory, we would have a
+dependent \emph{elimination} principle, which justifies
+\emph{inductive proofs}. However since linear types do not support
+dependency on linear types, we cannot directly adapt this
+approach. However, if we are trying to prove that two morphisms out of
+a mutually recursive type are equal, we can use the \emph{equalizer}
+type to prove their equality by induction.\max{TODO: show how this is
+  done.}
 
@@
   \begin{align*}
-    x + (y + z) &= (x + y) + z & x + y &= y + x\\
-    x + 0 &= x & x + x &= x\\
-    x(yz) &= (xy)z & x1 &= 1x = x\\
-    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
-    x0 &= 0x = x & & \\
-    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
-     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
+    \mathsf{Var} &: X \to \SPF\,X\\
+    \mathsf{K} &: L \to \SPF\,X\\
+    \mathsf{\bigoplus} &: (Y \to \SPF\,X) \to \SPF\,X\\
+    \mathsf{\bigamp} &: (Y \to \SPF\,X) \to \SPF\,X\\
+    \mathsf{\otimes} &: \SPF\,X \to \SPF\,X \to \SPF\,X\\
+    \el &: \prod_{X:U}\SPF\,X \to (X \to L) \to L\\
+    \map &: \prod_{X:U}\prod_{F : \SPF\,X}\prod_{A,B:X\to L}(\prod_{x:X}\uparrow(A\,x\lto B\,x)) \to \uparrow(\el(F)(A) \lto \el(F)(B))\\\\
+    \el(\Var\,M) B &= B\,M\\
+    \el(\mathsf{K}\,A) B &= A\\
+    \el(\bigoplus A) B &= \bigoplus_{y:Y}\el(A y)B\\
+    \el(\bigamp A) B &= \bigamp_{y:Y}\el(A y)B\\
+    \el(A \otimes A') B &= \el(A)B \otimes \el(A')B\\\\
+    \map(\Var\,M)\,f &= f\,M\\
+    \map(\K\,A)\,f &= \lambda a. a\\
+    \map(\bigoplus A)\,f &= \lambda a. \letin{\oplusinj y {a_y}}{a}{\oplusinj y {\map(A\,y)\,f\,a_y}}\\
+    \map(\bigwith\,A)\,f &= \lambda a. \withlamb{y}{\map(A\,y)\,f\,(\withprj y a)}\\
+    \map(A \otimes A')\,f &= \lambda b. \letin{(a,a')}{b}{(\map(A)\,f\,a,\map(A')\,f\,a')}
   \end{align*}
-  \caption{Kleene algebra axioms}
-  \label{fig:axioms}
+  \begin{mathpar}
+  \end{mathpar}
+  \caption{Strictly Positive Functors}
+  \label{fig:spf}
 \end{figure}
 
-The addition operation can be used to define the partial order
-structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
-formal languages, this order structure can be used to model language
-containment. In this section, we want to categorify the concept of
-Kleene algebra and build on top of it in order to define an abstract
-theory of parsing.  We start by defining \emph{Kleene categories}.
+\begin{figure}
+  \begin{mathpar}
+    \mathsf{roll} : \prod_{X:U}\prod_{F:X \to \SPF\,X}\prod_{x:X}\ltonl{(\el(F\,x)(\mu\,F))}
+
+    %% \inferrule*[right=IndIntro]
+    %% {\Gamma \vdash A : X \to \textrm{SPF}\,X\and
+    %%  \Gamma \vdash M : X\and
+    %%  \Gamma; \Delta \vdash e : \textrm{el}(A\,M) (\mu\,A)}
+    %% {\Gamma; \Delta \vdash \mathsf{roll}\, e : \mu\,A\, M}
+
+    \mathsf{fold} : \prod_{X:U}\prod_{F:X\to\SPF\,X}\prod_{A:X \to L}
+    (\prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)})
+    \to \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}
+
+    \inferrule*[right=Ind$\beta$]
+    {}
+    {\fold\,F\,f\,x\,(\roll e) \equiv f\,x\,(\map(F\,x)\,(\fold\,F\,f))}
+
+    \inferrule*[right=Ind$\eta$]
+    {x:X;a:\el(F\,x)(\mu F) \vdash e'\,x\,(\roll a) \equiv f\,x\,(\map(F\,x)\,e')
+    }
+    {\fold\,F\,f\equiv e'}
+    %% {\Gamma,x:X; a:\mu\,A\,x \vdash e' : B\,x
+    %%   \and \Gamma;\Delta \vdash e : \mu A M\\\\
+    %%   \Gamma,x:X; a':\el(A x)(\mu A) \vdash
+    %%   e'[x/x][\roll(a')/a] = f[x/x][\map(A\,x)\,(y.a.e'[y/x][a/a])\,a/b] : B\,x
+    %% }
+    %% {\Gamma; \Delta \vdash \fold(A)(x.b.f)(M,e) = e'[M/x][e/a] : B\,M}
+  \end{mathpar}
+  \caption{Indexed Inductive Linear Types}
+  \label{fig:iilt}
+\end{figure}
+
+\subsection{Grammar-specific Additions}
+
+So far, our calculus is a generic combination of dependent types with
+non-commutative linear types. In order to carry out formal grammar
+theory and define parsers, however, we add a few carefully chosen base
+types, terms and axioms that allow us to reason about formal grammars.
+\max{TODO: more prose, also are these all of our axioms?}
+\begin{enumerate}
+\item (Characters) For each character of our alphabet $c \in \Sigma$, we add a linear type $c \isLinTy$.
+\item (Strings are total) The function $\ltonl{(\String \lto \top)}$ is an isomorphism.\max{for this need to define $\String$, and isomorphism, and possibly we should define $\top$ in more detail earlier?
+idk}
+\item (Distributivity of $\oplus,\&$) The function  $\ltonl{(\bigoplus\limits_{f : \prod_{x:X} Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y)}$ is an isomorphism.
+\item ($\ltonl$ preserves $\oplus$) The function $\sum_{x:X}\ltonl{(A\,x)} \to \ltonl{(\bigoplus\limits_{x:X}A\,x)}$ is an isomorphism.
+\end{enumerate}
+
+% \steven{TODO the intro to this section is to be condensed based on what is
+%   included in the examples section as motivation and intuition building}
+
+% Omission of the structural rules of a deductive system, such as in
+% linear logic \cite{GIRARD19871}, offers precise control over how a value is used
+% in a derivation. Linear logic omits the weakening and contraction rules
+% to ensure that every value in context is used exactly once. This control enables
+% \emph{resource-sensitive} reasoning, where we may treat a resource as
+% \emph{consumed} after usage. This viewpoint is amenable to parsing applications,
+% as we may treat characters of a string as finite resources that are consumed at
+% parse-time.
+That is, when reading a string, the occurrence of any character is read once.
+Freely duplicating or dropping characters from a string changes the meaning of that string.
+\max{what about a backtracking parser?}.
+One may then envision a linear type system where the types comprise
+% formal grammars generated over some alphabet $\Sigma$, and the type constructors
+% correspond precisely to inductive constructions on grammars --- such as
+% conjunction, disjunction, concatenation, etc.
+
+% Programming in a purely linear term language is limiting due to the variable
+% usage restrictions. Code in such a language can
+% become unnecessarily verbose when ensuring the linearity invariant.
+% To alleviate this
+% concern, in 1995 Benton et al.\ proposed an alternative categorical
+% semantics of linear logic based on adjoint interactions between linear
+% and non-linear logics \cite{bentonMixedLinearNonlinear1995} ---
+% appropriately referred to as a \emph{linear-non-linear} (LNL)
+% system. This work is simply typed, so the boundary between linear and
+% non-linear subtheories is entirely characterized via a monoidal
+% adjunction between linear terms and non-linear terms.
+
+% \steven{Describe this monoidal adjunction with $\ltonl A$ and the derivable
+%   version of $F$}
+
+% Inside of an LNL system, linearity may be thought of as an option that users can
+% choose to deliberately invoke at deliberate points in their developments in an
+% otherwise intuitionistic system. However, if we are wishing to treat parsers as
+% linear terms over input strings, the non-linear fragment of an LNL theory does
+% not really assist in the development of parsers. It is instead the case that
+% parsers may benefit from a \emph{dependence} on non-linear terms.
+% Through the approach described by Krishnaswami et al.\ in
+% \cite{krishnaswami_integrating_2015},
+% we may define a restricted form of dependent types.
+
+% In particular, dependence
+% on linear terms
+% is disallowed; however, through dependence of a linear term on a non-linear
+% index, we can recover the definition of Aho's indexed grammars \cite{AhoIndexed}
+% internal to \theoryabbv. Although, we can define strictly more things, as Aho's
+% indexed grammars provide restricted access to the index whereas we can do
+% whatever with it.
+
+% \steven{elaborate on indexed grammars in a more articulate way in the last sentence}
+% \steven{Put the aho stuff in the subtheory section}
+%
+
+% \paragraph{Regular Expressions}
+% We may realize regular expression as a subtheory
+% of our language by restricting the type constructors to the
+% linear unit, characters, $\otimes$, $\oplus$, and Kleene star. Classically, the
+% languages recognized by these are exactly the regular languages --- the lowest
+% on the Chomsky hierarchy.
+
+% \paragraph{$\mu$-Regular Expressions} Similarly, we may restrict the connectives
+% to be linear unit, characters, $\otimes$, $\oplus$, and arbitrary recursion via
+% left-fixed point rather than only Kleene star.
+% Instead of regular expressions these correspond to Lei{\ss}'s $\mu$-regular
+% expressions, which are known to be equivalent to context-free grammars
+% \cite{leiss,krishnaswami_typed_2019}.
+
+% \paragraph{Beyond Context-Free Grammars} The previous two subtheories induce as semantics
+% regular grammars and context-free grammars, respectively; however, by including
+% the LNL dependent types we may actually express the entirety of the Chomsky
+% hierarchy. Through use of the LNL dependent types, we may encode indexed
+% grammars, which are known to be properly between context-free and
+% context-sensitive grammars within the Chomsky hierarchy \cite{AhoIndexed}. We
+% will further see in \cref{subsubsec:tm} how to use dependence to encode Turing machines internal to our
+% calculus, which of course induces a semantics of unrestricted grammars.
+
+% \paragraph{Indexed Grammars}
+% The LNL dependent types can be used to encode Aho's indexed grammars
+% \cite{AhoIndexed} internal to \theoryabbv. Although, we can define strictly more
+% things with this non-linear index, as we have unrestricted access to the index
+% whereas Aho's grammars can only manipulate it in a particular way.
+
+% \steven{Elaborate on these, and note that they are bw context free and context
+%   sensitive on the Chomsky hierarchy}
+
+\section{Formal Grammar Theory in \theoryname}
+\label{sec:applications}
+This section explores the applications of \theoryabbv to conducting formal
+gramamr theory. We demonstrate that several classical notions and constructions
+integral to the theory of formal languages are faithfully represented
+in \theoryname. By encoding well-established formal grammar concepts, we ensure
+that our framework remains grounded in the
+foundational principles of formal language theory while opening the door to
+compositional formal verification of parsers.
+
+\paragraph{Equivalence}
+In \cite{chom1963}, Chomsky differentiates between the weak generative capacity
+of a formal grammar --- the set of strings accepted by it --- and its strong
+generative capacity --- the set of ``structural descriptions'', or parse trees,
+that it generates. Two grammars are said to be \emph{weakly equivalent} if they
+agree in their weak generative capacity, and likewise \emph{strongly equivalent}
+if they agree in their strong generative capacity. We aim to recover each of
+these notions of equivalence in \theoryabbv.
 
 \begin{definition}
-  A Kleene category is a distributive monoidal category $\cat{K}$
-  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
-  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
-  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
-  X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogue isomorphism
-  for $G_{A,B}$ also holds.
+  \label{def:weakequiv}
+  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exists parse
+  transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B \lto \A)}$.
 \end{definition}
 
-As a sanity check, note that Kleene algebras are indeed examples of
-Kleene categories.
-
-\begin{example}
-  Every Kleene algebra, seen a posetal category, is a Kleene category.
-\end{example}
-
-This example illustrates how restricted Kleene algebras are in terms
-of reasoning about languages. By only having at most one morphism
-between objects, there is not a lot of information they can convey. In
-this case, the only information you get is language containment. We
-will see throughout this paper how the extra degrees of freedom
-granted by having more morphisms give you more algebraic structure for
-reasoning about languages.
-
-For the next example, we see an unexpected connection with the theory
-of substructural logics.
-
-\begin{example}
-  The opposite category of every Kleene category is a model of a variant of
-  conjunctive ordered logic, where the Kleene star plays the role of the ``of
-  course'' modality from substructural logics which allows hypotheses to
-  be discarded or duplicated.
-\end{example}
-
-The proposed axioms are a direct translation of the Kleene algebra
-axioms to a categorical setting. Its most unusual aspect is the
-axiomatization of the Kleene star as a family of initial algebras
-satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
-more structure, then these isomorphisms hold ``for free''.
-
-\begin{theorem}
-  \label{th:kleeneclosed}
-  Let $\cat{K}$ be a Kleene category such that it is also monoidal
-  closed.  Then, the initial algebras isomorphisms hold automatically.
+\newcommand{\lang}{\texttt{Lang}}
+A parse transformer $\f : \ltonl{(\A \lto \B)}$ takes as input parses of $\A$ for string
+$\w$ and returns a parse of $\B$ for $\w$. Therefore, $\f$ demonstrates an
+inclusion of languages $\lang(\A) \subset \lang(\B)$. Similarly, the existence of
+$\g$ shows that $\lang(\B) \subset \lang(\A)$, and thus $\lang(\A) = \lang(\B)$.
+
+\steven{Need better notation for language that is cohesive throughout the paper}
+
+\begin{definition}
+  \label{def:strongequiv}
+  Grammars $\A$ and $\B$ are \emph{strongly equivalent} if there exists parse
+  transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B \lto \A)}$ such
+  that $\f \circ \g \equiv \ident$ and $\g \circ \f \equiv \ident$.
+\end{definition}
+
+For a string $\w$, \cref{def:strongequiv} induces a bijection between the
+$\A$-parses of $\w$ and the $\B$-parses of $\w$, thus demonstrating that $\A$
+and $\B$ agree in their strong generative capacity.
+
+\paragraph{Ambiguity}
+A formal grammar $\A$ is ambiguous if there is a string $\w$ for which $\A$ has
+several parse trees of $\w$.
+For example, $\a \oplus \a$ is ambiguous because
+$\stringquote{a}$ has two parse trees --- on either side of the disjunction. On
+the other hand, a formal grammar is unambiguous when there is at most
+one parse tree for any input string. These notions of ambiguity are entirely
+semantic, so we seek to find a property expressible in the syntax of \theoryabbv
+that reifies the semantic description of ambiguity.
+
+\begin{definition}
+  \label{def:unambig}
+  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
+  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
+  then $\f \equiv \g$
+\end{definition}
+\cref{def:unambig} can be read more intuitively as stating that $\A$ is
+unabmiguous if there is at most one way to transform parses of any other grammar
+$\B$ into parse of $\A$.
+
+Note that this definition of unambiguity is preserved by strong equivalence, so a common proof technique for showing $\A$ is unambiguous is by proving $\A$
+is strongly equivalent to an unambguous grammar $\B$. In this style of proof, we
+build the parse transformers $\f : \ltonl{(\A \lto \B)}$ and
+$\g : \ltonl{(\B \lto \A)}$, then we have the obligation to show
+$\f \circ \g \equiv \ident$ and $\g \circ \f \equiv \ident$. However, because $\B$ is
+unambiguous there is only one parse transformer $\ltonl{(\B \lto \B)}$, so
+$\f \circ \g : \ltonl{(\B \lto \B)}$ and $\ident : \ltonl{(\B \lto \B)}$ are equal.
+Thus, in this type of proof, we really only have the obligation of building each of $\f$
+and $\g$ and demonstrating that $\g \circ \f \equiv \ident$.
+
+% \paragraph{A Grammar of Strings}
+% Define the character grammar $\charg$ to be the disjunction over characters in the alphabet $\Sigma$,
+% $\charg = \LinSigTy {c} {\Sigma} c$. Then define a linear type of strings as a
+% linear list of characters,
+% $\stringg = \charg^{*}$.
+
+% In the grammars model, we can build a
+% strong equivalence between $\top$ and $\stringg$. hen building parsers, we may wish to enforce this
+% syntactically by including $\top \cong \stringg$ as an axiom.
+
+% Through inclusion of this axiom, $! : \stringg \to \top$ is an isomorphism, and
+% thus also a monomorphism. That is, $\stringg$ is unambiguous.
+% \max{we should have a section of the previous section which gives the axioms, this one and the distributivity axiom}
+
+% \paragraph{Parseability}
+% We say $A$ is \emph{parseable} if there exists a grammar $B$ and a map
+% $\top \to A \oplus B$. Taking $\top \cong \stringg$, a grammar $A$ is parseable
+% exactly when we may read an arbitrary string into a parse of $A$ or a $B$-valued
+% refutation.
+
+% \paragraph{Decidability}
+% In the case that $A$ is parseable with respect to $\neg A$ --- i.e.\
+% there's a map $\top \to A \oplus \neg A$ --- then call $A$ \emph{decidable}.
+
+\subsection{Regular Expressions and Finite Automata}
+%% 2. Regex, NFA, DFA
+In this section, we make the equivalences
+between regular expressions and finite automata precise by encoding two theorems
+from classical language theory: (1) every regular expression is strongly equivalent
+to the traces through some non-deterministic finite automaton (NFA); and, (2) for
+every NFA, there is some deterministic finite automaton (DFA) such that the type
+of traces through the NFA is weakly equivalent to the type of traces through the DFA.
+
+\newcommand{\states}{\mathtt{states}}
+\newcommand{\labelt}{\mathtt{label}}
+\newcommand{\transitions}{\texttt{transitions}}
+\newcommand{\epstransitions}{\epsilon\texttt{transitions}}
+\newcommand{\isAcc}{\mathtt{isAcc}}
+\newcommand{\init}{\mathtt{init}}
+\newcommand{\src}{\mathtt{src}}
+\newcommand{\dst}{\mathtt{dst}}
+\newcommand{\parse}{\mathtt{parse}}
+\newcommand{\print}{\mathtt{print}}
+\newcommand{\epssrc}{\epsilon\mathtt{src}}
+\newcommand{\epsdst}{\epsilon\mathtt{dst}}
+
+\newcommand{\N}{\texttt{N}}
+
+\paragraph{Regular Expressions} The theory of regular expressions emerges as the
+subtheory of \theoryabbv generated by
+$\otimes$, $\oplus$, and Kleene star over the types $\bot$, $I$, and characters drawn from $\Sigma$.
+
+\paragraph{Nondeterministic Finite Automata} In \cref{sec:type-theory-examples}, we saw
+one particular NFA corresponding to the regular expression
+$(a^{*} \otimes b) \oplus c$. The definition of the linear type of traces
+through this
+NFA, as
+presented in \cref{fig:nfainductive}, inspects the concrete structure of the
+NFA.\ When generalizing from this example to an arbitrary NFA, we do not have the
+luxury of inspecting a concrete defintion. Therefore, we abstract over the
+specification of an arbitrary NFA and define traces with respect to that
+specification.
+
+Let the data of an NFA $\N$ be given by the following nonlinear data,
+\begin{itemize}
+\item $\N.\states$ --- a finite set of states
+\item $\N.\init : \N.\states$ --- the initial state
+\item $\N.\isAcc : \N.\states \to \Bool$ --- a mapping marking which states are
+        accepting
+\item $\N.\transitions$ --- a finite set of labeled transitions
+\item $\N.\labelt : \N.\transitions \to \Sigma$ --- a mapping from $N.\transitions$ to the character labeling the transition
+\item $\N.\src , \N.\dst : \N.\transitions \to \N.\states$ --- mappings from
+        $\N.\transitions$ to the source and target states, respectively, of a labeled transition
+\item $\N.\epstransitions$ --- a finite set of $\varepsilon$-transitions
+\item $\N.\epssrc, \N.\epsdst : \N.\epstransitions \to \N.\states$ --- mappings
+        from $\N.\epstransitions$ to the source and target states of an $\epsilon$-transition
+\end{itemize}
+
+In \cref{fig:nfatrace} we define a linear type of traces
+through an NFA $\N$. $\Trace_{\N}$ is an inductive type indexed by the first
+state $\s : \N.\states$ of a trace and a Boolean $b : \Bool$ that marks if the
+trace ends in an accepting state. Traces in $\N$ may be built through one of three constructors. We
+may terminate a trace at an accepting state with the constructor $\nil$. If we
+had a trace beginning at the destination state of a transition, then we may use
+the $\cons$ constructor to linearly combine that trace with a parse of the label
+of the transition to build a trace beginning at the source of the transition.
+Finally, if we had a trace beginning at the destination of an
+$\epsilon$-transition then we may use $\epscons$ to pull it back along the
+$\epsilon$-transition and construct a trace beginning at the source of the $\epsilon$-transition.
+
+\begin{figure}
+\begin{lstlisting}
+data Trace(*@$_{\N}$@*) : (s : N.states) (b : Bool) (*@$\to$@*) L where
+  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) N.states ] Trace(*@$_{\N}$@*)(s , N.isAcc s))
+  cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.transitions ] &[ b (*@$\in$@*) Bool ]
+           (N.label(t) (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.dst(t) , b) (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.src(t) , b)))
+  (*@$\epsilon$@*)cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.(*@$\epsilon$@*)transitions ] &[ b (*@$\in$@*) Bool ]
+           (Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)dst(t) , b) (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)src(t) , b)))
+\end{lstlisting}
+\caption{Traces through an NFA $N$ as an indexed inductive type}
+\label{fig:nfatrace}
+\end{figure}
+
+\paragraph{Deterministic Finite Automata}
+\newcommand{\D}{\texttt{D}}
+Just as the above defines an NFA, we provide a specification for an arbitrary
+deterministic finite automaton.\ A DFA $D$ is
+given by the following data,
+
+\begin{itemize}
+\item $\D.\states$ --- a finite set of states
+\item $\D.\init : \D.\states$ --- the initial state
+  \item $\D.\isAcc : \D.\states \to \Bool$ --- a mapping marking which states
+        are accepting
+\item $\D.\delta : \Sigma \to \D.\states \to \D.\states$ --- a deterministic transition function. Given a character $\c$ and a state
+        $\s$, $\D.\delta~\c~\s$ is the state for which there is a transition $\s \overset{\c}{\to} (\D.\delta~\c~\s)$
+\end{itemize}
+
+\begin{figure}
+\begin{lstlisting}
+data Trace(*@$_{\D}$@*) : (s : D.states) (b : Bool) (*@$\to$@*) L where
+  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) D.states ] Trace(*@$_{\D}$@*)(s , D.isAcc s))
+  cons : (*@$\uparrow$@*)(&[ c (*@$\in$@*) (*@$\Sigma$@*) ] &[ s (*@$\in$@*) D.states ] &[ b (*@$\in$@*) Bool ]
+             (c (*@$\lto$@*) Trace(*@$_{\D}$@*)(D.(*@$\delta$@*) c s , b) (*@$\lto$@*) Trace(*@$_{\D}$@*)(s , b)))
+\end{lstlisting}
+\caption{Traces through the DFA $D$ as an indexed inductive type}
+\label{fig:dfatrace}
+\end{figure}
+
+$\Trace_{\D}$ --- the linear type of traces through $\D$ --- is given in
+\cref{fig:dfatrace}, and its is
+quite similar to the definition of $\Trace_{\N}$. These
+traces may be terminated in an accepting state $\s$ with the $\nil$ constructor.
+The $\cons$ constructor builds a trace out of state $\s$ by linearly combining
+a parse of some character $\c$ with a
+trace out of the state $\D.\delta~\c~\s$. The trace built with $\cons$ is
+accepting if and only if the trace out of $D.\delta_{c}(s)$ is accepting.
+
+Because DFAs are deterministic, we are able to prove that their type of traces
+are unambiguous and define a parser for them. In particular we show that
+$\LinPiTy {s} {D.\states} {\LinSigTy {b} {\Bool} \Trace_{D}(s , b)}$ is isomorphic to
+$\stringg$, which simultaneously constructs a parser and demonstrates that
+traces through $D$ is unambiguous. This isomorphism intuitively says that we pay
+unambiguously build traces through $D$ beginning at any state $s$, and further
+that from any trace through $D$ we can uniquely recover the string that built
+the trace.
+
+The first half of the isomorphism,
+$\parse_{D} : \ltonl {\left( \stringg \lto \LinPiTy {s} {D.\states} {\Trace_{D}(s)} \right)}$,
+is defined by induction on strings in \cref{fig:dfaparser}. If this string is empty, then $\parse_{D}$
+defines a linear function that terminates a trace at the input state $s$. If the
+string is nonempty, then $\parse_{D}$ walks forward from the input state $s$ by
+the character at the head of the string.
+
+\begin{figure}
+\begin{lstlisting}
+parse : (*@$\uparrow$@*)(String (*@$\lto$@*) &[ s (*@$\in$@*) D.states ] (*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*)(s , b))
+parse String.nil s = (*@$\sigma$@*) (D.isAcc s) (Trace(*@$_{\D}$@*).nil s)
+parse (String.cons ((*@$\sigma$@*) c a) w) s = let (*@$\sigma$@*) b t = parse w (D.(*@$\delta$@*) c s) in
+                           (*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c s b a t)
+\end{lstlisting}
+\caption{Parser for DFA traces}
+\label{fig:dfaparser}
+\end{figure}
+
+The second half of the isomorphism,
+$\print_{D} : \ltonl {\left( \LinPiTy {s} {D.\states} {\Trace_{D}(s)} \lto \stringg \right)}$,
+is given in \cref{fig:printdfa} by induction on traces. If the trace is defined
+via $\nil$, then $\print_{D}$ returns the empty string. Otherwise, if the trace
+is defined by $\cons$ then $\parse_{D}$ appends the character $c$ to the output
+string and recurses.
+
+\begin{figure}
+\begin{lstlisting}
+print : (s : D.states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*)(s , b)) (*@$\lto$@*) String)
+print .s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).nil s)) = String.nil
+print .s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c (D.(*@$\delta$@*) c s) b a trace)) =
+  String.cons ((*@$\sigma$@*) c a) (print (D.(*@$\delta$@*) c s) ((*@$\sigma$@*) b trace))
+\end{lstlisting}
+% \[
+% \inferrule
+% {
+% }
+% {tr : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool } {\Trace_{D}(s, b)}} \vdash asdf : \stringg}
+% \]
+\caption{Printer for DFA traces}
+\label{fig:printdfa}
+\end{figure}
+
+\steven{TODO make the agda logo clickable, and mention its meaning somewhere}
+\newcommand{\Agda}{\agdalogo}
+\begin{theorem}[Determinization, \Agda]
+  Given an NFA $N$, there exists a DFA $D$ such that
+  $\Trace_{N}~\N.\init~\true$ is weakly equivalent to $\Trace_{D}~\D.\init~\true$.
 \end{theorem}
+\newcommand{\X}{\texttt{X}}
+\newcommand{\Y}{\texttt{Y}}
+\newcommand{\x}{\texttt{x}}
+\newcommand{\y}{\texttt{y}}
 \begin{proof}
-  We prove this by the unicity (up-to isomorphism) of initial
-  algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
-  (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
-  X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
-  (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
-    1}(X))$.
-
-  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
-  we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,1} \to B \lto Y$. We can show existence and
-  uniqueness by showing that the diagram on top commutes if, and
-  only if, the diagram on the bottom commutes:
-
-% https://q.uiver.app/#q=WzAsOCxbMCwwLCJCICsgQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFsyLDAsIkIgKyBZIFxcb3RpbWVzIEEiXSxbMiwyLCJZIl0sWzAsMiwiQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDMsIjEgKyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDUsIlxcbXUgWC5cXCwgMSArIFggXFxvdGltZXMgQSJdLFsyLDMsIjEgKyAoIEIgXFxsdG8gWSkgXFxvdGltZXMgQSJdLFsyLDUsIkIgXFxsdG8gWSJdLFswLDEsImlkICsgKGlkIFxcb3RpbWVzIGg7IGV2KSBcXG90aW1lcyBpZF9BIl0sWzEsMiwiW2YsZ10iXSxbMywyLCJpZCBcXG90aW1lcyBoOyBldiIsMl0sWzAsMywiW2lkIFxcb3RpbWVzIGgsIGlkIFxcb3RpbWVzIHRsXSIsMl0sWzQsNiwiaWQgKyAoaCBcXG90aW1lcyBYKVxcb3RpbWVzIGlkIl0sWzUsNywiaCIsMl0sWzYsNywiW2YnLCBnJ10iXSxbNCw1LCJbaGQsIHRsXSIsMl1d
+  We present a variant of Rabin and Scotts's classic powerset construction \cite{rabinFiniteAutomataTheir1959}. First, we build
+  the DFA $\D$. Define the
+  states of $\D$ to be the $\mathbb{P}_{\varepsilon}(\N.\states)$ --- the type
+  of $\epsilon$-closed\footnote{A subset of states $\X$ is $\epsilon$-closed if
+    for every $\s \in \X$ and $\epsilon$-transition $\s \overset{\epsilon}{\to} \s'$ we have $\s' \in \X$.} subsets of
+  $\N.\states$. A subset is accepting in $\D$ if it contains an accepting state
+  from $\N$. Construct the initial state of $\D$ as the
+  $\epsilon$-closure of $\N.\init$. Lastly define the transtion function
+  of $\D$ to send the subset $X$ under the character $\c$ to the
+  $\epsilon$-closure of all the states reachable from $X$ via a transition
+  labelled with the character $\c$.
+
+  We demonstrate the weak equivalence between $\Trace_{\N}~\N.\init~\true$ and
+  $\Trace_{\D}~\D.\init~\true$ by constructing parse transformers between the two
+  grammars. To build the parse transformer
+  $\ltonl{(\Trace_{\N}~\N.\init~\true \lto \Trace_{\D}~\D.\init~\true)}$, it
+  suffices to construct the stronger term
+\newcommand{\NtoD}{\texttt{NtoD}}
+\newcommand{\DtoN}{\texttt{DtoN}}
+  \[
+    \NtoD : \ltonl {\left(\Trace_{\N}~\s~\true \lto \bigwith\limits_{\X : \D.\states} \bigwith\limits_{\texttt{sInX} : \X \ni \s} \Trace_{\D}~\X~\true \right)}
+  \]
+  that maps a trace in $\N$ from an arbitrary state $\s$ to a trace in $\D$
+  that may begin at any subset of states $\X$ that contains $\s$. $\NtoD$ may then be instantiated at
+  $\s = \N.\init$ and $\X = D.\mathsf{init}$ to get the desired
+  parse transformer.
+
+\begin{figure}
+\begin{lstlisting}
+NtoD : (s : N.states) (*@$\to$@*)
+       (*@$\uparrow$@*)(Trace(*@$_{\N}$@*) s true (*@$\lto$@*)
+           &[ X (*@$\in$@*) D.states ] &[ sInX (*@$\in$@*) (X (*@$\ni$@*) s) ] Trace(*@$_{\D}$@*) X true)
+NtoD .s (nil s) X sInX = nil s
+NtoD .(N.src trans) (N.cons trans .true a dstTrace) X sInX =
+  D.cons (N.label trans) X true a
+    (((NtoD s dstTrace).(*@$\pi$@*) (D.(*@$\delta$@*) (N.label trans) X)).(*@$\pi$@*) (N.dst trans))
+NtoD .(N.(*@$\epsilon$@*)src trans) ((*@$\epsilon$@*)cons (*@$\epsilon$@*)trans .true (*@$\epsilon$@*)DstTrace) X sInX =
+  ((NtoD s (*@$\epsilon$@*)DstTrace).(*@$\pi$@*) X).(*@$\pi$@*) (N.(*@$\epsilon$@*)dst trans)
+\end{lstlisting}
+\caption{Parse transformer from NFA traces to Powerset DFA traces}
+\label{fig:NtoD}
+\end{figure}
+
+Because the state $\s$ is now arbitrary, we can define $\NtoD$ via
+recursion on the trace through the NFA. Code for this is given in \cref{fig:NtoD}.
+If the NFA trace is terminating at the accepting state $\s$ and $\s \in \X$,
+then we return a trace in $\D$ that terminates at
+$\X$, which is accepting because it contains the accepting state $\s$.
+
+If the NFA trace is defined via $\cons$ for a labeled transition
+$\texttt{trans} : \N.\transitions$ such that $N.\src~\texttt{trans} \in \X$, then
+we first decompose the trace into a parse of $\N.\labelt~\texttt{trans}$ and a
+trace starting from $\N.\dst~\texttt{trans}$. We then recursively call $\NtoD$
+on this trace which returns a family of traces parametrized by $\epsilon$-closed
+subsets containing $N.\dst~\texttt{trans}$. We instantiate this family at the
+set of states
+$\D.\delta~(\N.\labelt~\texttt{trans})~X$. Because $\N.\dst~\texttt{trans}$ is
+reachable from $\N.\src~\texttt{trans}$ by a single transition labeled by
+$\N.\labelt~\texttt{trans}$ and $\N.\src~\texttt{trans} \in X$ we have that
+$\N.\dst~\texttt{trans} \in \D.\delta~(\N.\labelt~\texttt{trans})~X$. Therefore
+we can linearly combine the parse of the label and the trace built from the
+recursive call to get a trace in $\D$ rooted at $X$.
+
+Lastly, if the NFA trace is defined via $\epsilon\cons$ for an $\epsilon$-transition
+$\epsilon\texttt{trans} : \N.\epstransitions$ such that
+$\N.\epssrc~\epsilon\texttt{trans} \in \X$, then the construction is quite
+similar. In this case we again recursivley call $\NtoD$, except afterwards we
+instantiate the family of grammars at the set $X$ itself. We may do this because
+all states in $\D$ are taken to be $\epsilon$-closed, therefore
+$\N.\epsdst~\epsilon\texttt{trans} \in \X$ because
+$\N.\epssrc~\epsilon\texttt{trans}$ is in $X$. This concludes the construction of $\NtoD$.
+
+To construct a term from DFA traces to NFA traces, we similarly strengthen our
+induction hypothesis and build a parse transformer
+\[
+  \DtoN : \ltonl {\left( \Trace_{\D}~\X~\true \lto \bigoplus\limits_{s : \N.\mathsf{states}}\bigoplus\limits_{\texttt{sInX} : \X \ni \s} \Trace_{\N}~s~\true \right)}
+\]
+
+For a moment, let us turn our attention to the content behind the traces in $\D$.
+Traces from subset $\X$ to subset $\Y$ describes the \emph{existence} of some trace
+in the $\N$ from some $\x \in \X$ to some $\y \in \Y$. In general it is not clear how
+to extract out a choice of trace in $\N$, which is what $\DtoN$ is asking of us.
+However, we may further take the states and transitions of $\N$ to be \emph{ordered}.
+This allows us to
+induce an order on traces in $\D$ from $\X$ to $\Y$. From these orderings we can then
+define the necessary choice functions on each type by taking the smallest with
+respect to the ordering.
+
+We define $\DtoN$ by recursion on traces through $\D$. First if the trace stops at
+some accepting $\epsilon$-closed subset $\X$, then we may choose some state
+$\s \in X$ such that $\s$ is accepting and terminate the trace in $\N$ at $\s$.
+
+Next if the trace in $\D$ is defined via transition
+$\X \overset{\c}{\to} (\D.\delta~\c~\X)$ then we may choose some state in
+$\s \in \D.\delta~\c~\X$. Recursively, $\DtoN$ sends the trace rooted at
+$\D.\delta~\c~\X$ to one rooted at $\s$. By definition of $\D.\delta~\c~\X$,
+there exists some $\s' \in \X$ such that $\s'$ tranitions to $\s$ by some number of
+$\epsilon$-transitions followed by a transition labeled by $\c$.
+We order all
+such traces of this form and choose the smallest one, which is then linearly
+combined with the parse of $\c$ to build an $\N$ trace. This concludes the
+definition of $\DtoN$, thus showing that $\Trace_{\N}~\N.\init~\true$ and $\Trace_{\D}~\D.\init~\true$ are
+weakly equivalent.
+\end{proof}
+
+\newcommand{\R}{\texttt{R}}
+\begin{theorem}[Thompson's Construction, \Agda]
+  Given a regular expression $\R$, there exists an NFA $\N$ such that $\R$ is
+  strongly equivalent to $\Trace_{N}(\N.\init)$.
+\end{theorem}
+
+\begin{proof}
+\steven{TODO this proof needs to be rewritten. Likely through use of equalizers}
+We present a variant of Thompson's construction
+\cite{thompsonProgrammingTechniquesRegular1968} in which we recurse on the
+structure of $\R$ to construct a strongly equivalent NFAs. In this paper we will walk through
+some of the cases of this proof. The remaining cases
+follow
+similar reasoning and the details may be found in the formalized artifact.
+
+\begin{figure}
+  \begin{tikzpicture}[node distance = 17mm ]
+    \node[state, initial] (0) {$\0$};
+    \node[state, right of=0, accepting] (1) {$\1$};
+
+    \path[->] (0) edge[above] node{$\stringquote{c}$} (1);
+  \end{tikzpicture}
+  \caption{NFA $\N_{\c}$ for a single character $\c$}
+  \label{fig:litnfa}
+\end{figure}
+
+First handle the case where $R$ is a literal character $c$. The NFA $N_{c}$ pictured in
+\cref{fig:litnfa} is strongly equivalent to the grammar $c$. To demonstrate this
+strong equivalence, we build terms between $c$ and $\Trace_{N_{c}}(N_{c}.\init)$
+and show they are mutually inverse.
+
+Define $f : \ltonl {(c \lto \Trace_{N_{c}}(N_{c}.\init))}$ as follows,
+
+\[
+  a : c \vdash f := \cons(a, \nil) : \Trace_{N_{c}}(N_{c}.\init)
+\]
+
+Define $g : \ltonl {(\Trace_{N_{c}}(N_{c}.\init) \lto c)}$ via induction on the
+trace. Map traces beginning in state 1, which are necessarily built via $\nil$,
+to the linear unit $I$. Then we must handle traces beginning in state 0, which
+are necessarily built via $\cons$. We decompose such a trace into a parse of the
+character $c$ and a trace beginning in state 1. We inductively transform the
+trace beginning in state 1 into a term of type $I$. Finally, we use the right
+unitor to turn a term of type $c \otimes I$ into one of type $c$.
+
 \[\begin{tikzcd}
-	{B + B \otimes (\mu X.\, 1 + X \otimes A)} && {B + Y \otimes A} \\
-	\\
-	{B \otimes (\mu X.\, 1 + X \otimes A)} && Y \\
-	{1 + (\mu X.\, 1 + X \otimes A)} && {1 + ( B \lto Y) \otimes A} \\
-	\\
-	{\mu X.\, 1 + X \otimes A} && {B \lto Y}
-	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
-	\arrow["{[f,g]}", from=1-3, to=3-3]
-	\arrow["{id \otimes h; ev}"', from=3-1, to=3-3]
-	\arrow["{[id \otimes h, id \otimes tl]}"', from=1-1, to=3-1]
-	\arrow["{id + (h \otimes X)\otimes id}", from=4-1, to=4-3]
-	\arrow["h"', from=6-1, to=6-3]
-	\arrow["{[f', g']}", from=4-3, to=6-3]
-	\arrow["{[hd, tl]}"', from=4-1, to=6-1]
+    {c \otimes \mathsf{Trace}_{N_{c}}(1)} &
+    {c \otimes I} &
+    c
+	\arrow["{id\otimes g}", from=1-1, to=1-2]
+	\arrow["{}", from=1-2, to=1-3]
 \end{tikzcd}\]
 
-  This equivalence follows by using the adjunction structure given
-  by the monoidal closed structure of $\cat{K}$. A completely analogous
-  argument for $G_{A,B}$ also holds.
+Note that the grammar $c$ is unambiguous, so $g \circ f$ is equal to the
+identity on $c$. Additionally, we can show that $f$ carries the structure of an
+indexed algebra homomorphism for the family of functors defining the trace type
+$\Trace_{N_{c}}$. Because $g$ is defined via induction, it carries the structure
+of an algebra homomorphism as well. Thus, $f \circ g$ is also an algebra
+homomorphism. $f \circ g$ and the identity map are both then algebra
+homomorphisms from $\Trace_{N_{c}}$ to itself. Because $\Trace_{N_{c}}$ is the
+initial algebra for the relevant family of functors, there is only one
+homomorphism from it to itself and thus $f \circ g = \mathsf{id}$. Therefore,
+$f$ and $g$ are mutually inverse and the regular expression is strongly
+equivalent to the traces through the NFA in this case.
+
+Now consider the case where $R$ is the disjunction of two regular expressions
+$R_{1}$ and $R_{2}$. Inductively, we may build NFAs $N_{1}$ and $N_{2}$ that are
+strongly equivalent to $R_{1}$ and $R_{2}$, respectively. We define $N_{\oplus}$
+as a new NFA built as the disjunction of $N_{1}$ and $N_{2}$.
+$N_{\oplus}.\states$ comprises the states from $N_{1}$, the states from
+$N_{2}$, and a new marked initial state. The accepting states are precisely
+those that were accepting in the subautomata. The initial state has an
+$\varepsilon$-transition to each of the initial states of the subautomata. Once in
+$N_{1}.\init$, the downstream transitions are a copy of those in $N_{1}$,
+likewise for $N_{2}.\init$.  That is, $N_{\oplus}$ contains a copy of each
+$N_{1}$ and $N_{2}$, and we choose which copy to inhabit by deciding which
+$\varepsilon$-transition out of the initial state to take. A schematic
+representation of this automaton is given in \cref{fig:disjunctionnfa}.
+
+\begin{figure}
+  \begin{tikzpicture}[node distance = 17mm ]
+    \node[state, initial] (init) {$\init$};
+    \node[state, above right of=init] (init1) {$N_{1}.\init$};
+    \node[state, below right of=init] (init2) {$N_{2}.\init$};
+    \node[right of=init1] (dots1) {$\dots$};
+    \node[right of=init2] (dots2) {$\dots$};
+    \node[state, accepting, right of=dots1] (acc1) {$s_{1}$};
+    \node[state, accepting, right of=dots2] (acc2) {$s_{2}$};
+
+    \path[->] (init) edge[below] node{$\varepsilon$} (init1)
+              (init) edge[above] node{$\varepsilon$} (init2)
+              (init1) edge[above] node{} (dots1)
+              (init2) edge[above] node{} (dots2)
+              (dots1) edge[above] node{} (acc1)
+              (dots2) edge[above] node{} (acc2);
+  \end{tikzpicture}
+  \caption{NFA $N_{\oplus}$ as a disjunction of $N_{1}$ and $N_{2}$}
+  \label{fig:disjunctionnfa}
+\end{figure}
+
+Because $N_{1}$ and $N_{2}$ are each strongly equivalent to $R_{1}$ and $R_{2}$,
+it suffices to show that
+$\Trace_{N_{1}}(N_{1}.\init) \oplus \Trace_{N_{2}}(N_{2}.\init)$ is strongly
+equivalent to $\Trace_{N_{\oplus}}(N_{\oplus}.\init)$.
+
+Just like the case of a single character, when building the maps in either
+direction, we strengthen the induction hypothesis by mapping out of traces
+beginning with an arbitrary state then proceed by induction on traces.
+
+First, define the
+mapping out of $\Trace_{N_{\oplus}}(s)$. If $s$ is a state from $N_{1}$ then
+send the trace to a trace in $N_{1}$ beginning at $s$. Likewise, if $s$ is from
+$N_{2}$, then we map to a trace in $N_{2}$ beginning at $s$. If $s$ is
+$N_{\oplus}.\init$, then we map the trace to a disjunction of traces
+$\Trace_{N_{1}}(N_{1}.\init) \oplus \Trace_{N_{2}}(N_{2}.\init)$.
+
+In the other direction, we need to map a disjunction of traces through $N_{1}$
+or $N_{2}$ to a trace in $N_{\oplus}$. Without loss of generality, assume that
+the trace came from $N_{1}$. That is, assume we currently have a term of type
+$\Trace_{{N_{1}}}(N_{1}.\init)$ and we aim to construct a term of type
+$\Trace_{N_{\oplus}}(N_{\oplus}.\init)$. We induct on the trace $N$ trace an map
+it over to a trace beginning at the copy of $N_{1}.\init$ in $N_{\oplus}$, and
+then we conclude by using the $\epscons$ constructor to turn this trace into one
+beginning at $N_{\oplus}.\init$.
+
+We show that these maps mutually invert each other by the same sort of argument
+we used with the single character NFA.\ That is, we prove that the each map is a
+homomorphism of algebras and then appeal to the initiality of each of
+$\Trace_{N_{1}}$, $\Trace_{N_{2}}$, and $\Trace_{N_{\oplus}}$ to prove that the
+relevant maps compose to the identity.
+
+In this sketch, we elide the proof that these maps do indeed form homomorphisms.
+Further, we are eliding the needed NFAs for the cases where the regular
+expression is $I$, $\bot$, a Kleene star, or a concatenation. All of these
+elided details may be found in the Agda formalization.
 \end{proof}
 
-This result feels similar in spirit to the definition of action
-algebras, which are algebras where the product also has adjoint
-operations which results in the Kleene star being more easily
-axiomatized \cite{kozen1994}.
-
-We are now ready to prove that our concept of formal grammars fits
-nicely within our categorical framework. We start by presenting a
-well-known construction from presheaf categories.
+\subsection{The Dyck Grammar}
+%% 3. Dyck language
+\newcommand{\balanced}{\mathtt{balanced}}
+\newcommand{\Dyck}{\mathtt{Dyck}}
+\steven{This subsection is still a WIP}
+Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
+grammars (CFGs). The smallest nontrivial CFG is the Dyck grammar
+$\Dyck$ --- the grammar of balanced parentheses. Moreover, the Chomsky-Sch\"utzenberger
+representation theorem \cite{CHOMSKY1959118} demonstrates that $\Dyck$, and grammars like it,
+cut to the essence of what it means to be context free.
+
+$\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$
+and is defined in \cref{fig:dyckinductive}. The $\nil$ constructor shows that
+the empty string is balanced, and the $\balanced$ constrcutor builds a balanced
+parse by wrapping an already balanced parse in an additional set of parentheses
+then following it with another balanced parse.
+
+\begin{figure}
+\begin{lstlisting}
+data Dyck : L where
+  nil : (*@$\uparrow$@*)(Dyck)
+  balanced : (*@$\uparrow$@*)('(' (*@$\lto$@*) Dyck (*@$\lto$@*) ')' (*@$\lto$@*) Dyck (*@$\lto$@*) Dyck)
+\end{lstlisting}
+\caption{The Dyck grammar as an inductive linear type}
+\label{fig:dyckinductive}
+\end{figure}
+
+\steven{somewhere in here, sneak in the benchmark where the Dyck parser could
+  handle 10's of thousands of characters in about 30s}
+
+We construct a parser for $\Dyck$ by building a deterministic automaton $M$ such
+that $\Trace_{M}(M.\init) \cong \Dyck$.
+
+\Cref{fig:DyckAutomaton} presents a sketch of the automaton $M$. The traditional
+approach to parse grammars like $\Dyck$ is to employ a stack to count how many
+open parentheses have been seen so far. We may think of
+$M$ as performing the same task as a stack machine, except instead of storing
+the number of open parentheses in a distinguished stack of memory, $M$ operates
+by storing this data in the current state. This is made feasible by specifying
+$M$ almost identically to how we defined DFAs, except we omit the
+finiteness condition on the states of the automaton.
+
+\steven{Mention that $\Trace_{M}$ is iso to string. Mention the same for DFAs above }
+
+% Unlike the finite automata seen thus far, $M$ has infinitely many states --- one
+% for each $n : \mathbb{N}$ with an additional fail state. This diverges a bit
+% from classical language theory, where the recognizer for $D$ would be encoded as a
+% pushdown automaton with finitely many
+% states and stack of memory. Such PDAs are encodable in our theory, although they
+% are a little cumbersome to use. Moreover, they
+% seem overfit to classical reasoning as they don't preserve the grammar up to
+% strong equivalence, only weak equivalence in general. \steven{cite Lambek
+%   automata and state that is is a secondary goal to preserve as much structure as possible}
+% By moving the stack of memory (the nats) into the
+% state, we obtain infinitely many states, which is fine bc the states need only
+% be a type for us to interface with it.
+
+\steven{TODO sketch proof of equivalence. Mention that it is mechanized}
+
+\begin{figure}
+  \begin{tikzpicture}[node distance = 17mm ]
+    \node[state, initial, accepting] (0) {$0$};
+    \node[state, right of=0] (1) {$1$};
+    \node[state, right of=1] (2) {$2$};
+    \node[right of=2] (3) {$\dots$};
+    \node[state, below of=0] (fail) {$\mathsf{fail}$};
+
+    \path[->] (0) edge[above, bend left] node{\stringquote{(}} (1)
+              (1) edge[below, bend left] node{\stringquote{)}} (0)
+              (1) edge[above, bend left] node{\stringquote{(}} (2)
+              (2) edge[below, bend left] node{\stringquote{)}} (1)
+              (2) edge[above, bend left] node{\stringquote{(}} (3)
+              (3) edge[below, bend left] node{\stringquote{)}} (2)
+              (fail) edge[loop right] node{\stringquote{(}, \stringquote{)}} (fail)
+              (0) edge[left] node{\stringquote{)}} (fail);
+  \end{tikzpicture}
+  \caption{Automaton $M$ for the Dyck grammar}
+  \label{fig:DyckAutomaton}
+\end{figure}
+
+% \subsection{LL(1)}
+% %% 4. LL(1) parser
+
+% \steven{TODO impute LL(1) example, mention that its weak equivalence is mechanized}
+
+\subsection{Turing Machines}
+\newcommand{\Turing}{\mathtt{Turing}}
+\newcommand{\ttt}{\mathtt{t}}
+\newcommand{\T}{\mathtt{T}}
+\newcommand{\acc}{\mathtt{acc}}
+\newcommand{\rej}{\mathtt{rej}}
+\newcommand{\Tape}{\mathtt{Tape}}
+\newcommand{\parseTape}{\mathtt{parseTape}}
+\newcommand{\snoc}{\mathtt{snoc}}
+\steven{This subsection is still a WIP}
+%% 5. Expressiveness: Turing Machines
+The expressive power of \theoryabbv spans the entire Chomsky hierachy, as we may
+encode the acccepting traces of Turing machines as linear types.
+
+\steven{TODO introduce tape alphabet and blank symbol}
+
+The data of a Turing machine $\T$  is given by the following data,
+\begin{itemize}
+  \item $\T.\states$ --- a finite set of states
+  \item $\T.\init : \T.\states$ --- the initial state
+  \item $\T.\acc , \T.\rej : \T.\states$ --- distinguished accepting and
+        rejecting states, respectively, such that $\T.\acc \neq \T.\rej$
+  \item $\T.\delta : \T.\states \to \Gamma \to \T.\states \times \Gamma \times \Bool$
+        --- a deterministic transition function. Given the current state and the
+        head of the tape return the new state, the character written to the
+        tape, and which direction to shift ($\true$ for left, $\false$ for right)
+\end{itemize}
+
+The tape of a
+(one-sided) Turing machine is encoded as nonlinear function from the natural numbers to the tape
+alphabet $\Gamma$, $\Tape = \mathbb{N} \to \Gamma$. To initialize the tape of
+$\T$, we define the parse transformer
+$\parseTape : \ltonl {(\stringg \lto \LinSigTy {\ttt} {\Tape} {\Turing~\ttt})}$ in \cref{fig:parseTape}.
+
+\steven{TODO finish the description of Turing machines}
+
+\begin{figure}
+\begin{lstlisting}
+data Turing : (t : Tape) (*@$\to$@*) L where
+  nil : (*@$\uparrow$@*)(&[ t (*@$\in$@*) Tape ] Turing t)
+  snoc : (*@$\uparrow$@*)(&[ c (*@$\in$@*) (*@$\Sigma$@*) ] &[ t (*@$\in$@*) Tape ](Turing t (*@$\lto$@*) c (*@$\lto$@*) Turing (push c t)))
+\end{lstlisting}
+\caption{Inductive linear type of Turing machines initialized with tape $\ttt$}
+\label{fig:turinginductive}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}
+parseTape : (*@$\uparrow$@*)(String (*@$\lto$@*) (*@$\oplus$@*)[ t (*@$\in$@*) Tape ] Turing t)
+parseTape String.nil = (*@$\sigma$@*) ((*@$\lambda$@*) n . blank) (Turing.nil ((*@$\lambda$@*) n . blank))
+parseTape (String.cons ((*@$\sigma$@*) c a) w) =
+  let (*@$\sigma$@*) t Tur = parseTape w in (*@$\sigma$@*) (push c t) (snoc c t Tur a)
+\end{lstlisting}
+\caption{Initialization of the tape of Turing machine as a parse transformer}
+\label{fig:parseTape}
+\end{figure}
+
+We may then define a term of type
+$\mathsf{parseTape} : \ltonl {\left( \stringg \lto \LinSigTy {\ttt} {\Tape} {\Turing~\ttt} \right)}$.
+In this manner, all of the interesting computations that the Turing machine
+would do occur solely inside of the nonlinear index. That is,
+$\mathsf{parseTape}$ does not discern which strings are accepted by the Turing
+machine, nor does it do any computation. $\mathsf{parseTape}$ simply loads the
+input string into the initial spaces on the tape.
+
+Once the string has been loaded onto the tape, we may further define the
+following term to run the Turing machine.
+\[
+  \mathsf{run} : \ltonl {\left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \Turing(t) \lto \bigwith\limits_{\mathsf{fuel} : \mathbb{N}} \mathsf{Maybe} \left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \bigoplus\limits_{b : \Bool} \bigoplus\limits_{\mathsf{trace} : \mathsf{TuringTrace}(b , t)} \Turing(t) \right) \right)}
+\]
+
+where $\mathsf{TuringTrace}$ is a nonlinear type denoting terminating traces
+through the Turing machine. Note that $\mathsf{run}$ returns a $\mathsf{Maybe}$
+type in a fuel parameter. Turing machines may not terminate. If they do
+terminate within $\mathsf{fuel}$-many steps, then we return the trace along with
+the current tape and a boolean denoting if the trace is accepting or rejcting.
+If the trace does not terminate within the alloted amount of fuel, then we
+timeout and return nothing.
+
+\subsection{Semantic Actions}
+\steven{TODO write this section and give some small simple actions}
+All constructions discussed until now detail how a string may be parsed into a
+\emph{concrete} syntax tree; however, this is often not the representation that later
+compiler passes would take in. Instead, they often take in an \emph{abstract}
+syntax tree that forgets syntactic details that are superfluous to the later
+passes. For example, after using parentheses to disambiguate a parse of a string
+into a concrete syntax tree, we may remove these parentheses to have a cleaner
+structured representation of the underlying data. The procedure that takes in a
+concrete syntax tree and returns an abstract one is referred to as a
+\emph{semantic action}.
+
+Given a linear type $A$, the inhabitants of $A$ serve as concrete syntax. We may
+then chose a non-linear type $X$ that serves as abstract syntax. A semantic
+action can then be defined as $\square (A \lto \Delta X)$ where
+$\Delta X = \LinSigTy x X \top$.
+\steven{Note here that even though $\Delta$ is a map from nonlinear types into
+  linear ones, it is very different from $F$/whatever we call that half of the adjunction}
+
+% \steven{Everything in this section below this marker is outdated and needs to be
+% folded into the above or cut}
+
+% \steven{Need a different segue because canonicity is no longer before this}
+% The canonicity theorem from the previous section gives a tight connection
+% between terms in our type theory and the theory of formal languages. In this section
+% we further explore these connections by using the fact that, classically
+% formal language theory is closely related to the study of automata.
+% In the Chomsky hierarchy, each language class is associated to a class of
+% automata that serve as recognizers. Internal to \theoryabbv,
+% we can characterize these language classes syntactically; moreover, we
+% demonstrate the equivalence of these language classes to its associated automata class as a
+% proof term within our logic.
+
+% \subsection{Non-deterministic Finite Automata}
+% \label{subsec:finiteaut}
+% \begin{figure}
+%   \begin{tikzpicture}[node distance = 25mm ]
+%     \node[state, initial, accepting] (1) {$1$};
+%     \node[state, below left of=1] (2) {$2$};
+%     \node[state, right of=2] (3) {$3$};
+
+%     \path[->] (1) edge[above] node{$b$} (2)
+%               (1) edge[below, bend right, left=0.3] node{$\varepsilon$} (3)
+%               (2) edge[loop left] node{$a$} (2)
+%               (2) edge[below] node{$a, b$} (3)
+%               (3) edge[above, bend right, right=0.3] node{$a$} (1);
+%   \end{tikzpicture}
+%   \caption{An example NFA}
+%   \label{fig:NFA}
+% \end{figure}
+
+% Classically, a \emph{nondeterministic finite automaton} (NFA) is a finite state machine where
+% transitions are labeled with characters from a fixed alphabet $\Sigma$. These
+% are often represented formally as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$,
+
+% \begin{itemize}
+%   \item $Q$ a finite set of states
+%   \item $\Sigma$ a fixed, finite alphabet
+%   \item $\delta : Q \times (\Sigma \cup \{ \varepsilon\}) \to \mathcal{P}(Q)$ the labeled transition function
+%   \item $q_{0} \in Q$ the start state
+%   \item $F \subset Q$ a set of accepting states
+% \end{itemize}
+
+% \subsection{A DFA Parser}
+% \label{subsec:regexparser}
+
+% Just as we encoded traces of NFAs as grammars, we likewise
+% encode the traces of a DFA as grammars. The key difference
+% between NFAs and DFAs is \emph{determinism} --- meaning,
+% that in a state $q$ inside of DFA, given a character $c$ there
+% will be exactly one transition that we may take leaving $q$
+% with label $c$. For us, this changes the definition of valid
+% transitions for a DFA, instead of the definition of
+% $\mathsf{Trans}$ provided in \cref{subsec:finiteaut} DFAs
+% obey
+
+% \begin{gather*}
+%  \mathsf{State} \in \{ g_{q} : q \in Q \} \\
+%  \mathsf{Trans}(q) ::= \bigoplus_{c \in \Sigma} (c \otimes \mathsf{State})
+% \end{gather*}
+
+% Meaning, each state has a transition for every character. \emph{Note:} the above use
+%   of a disjunction over $\Sigma$ is a bit of abuse of notation.
+%   As $\Sigma$ is a finite alphabet, we wish to think of this as an iterated
+%   binary sum, or perhaps as a finitely indexed sum defined via linear dependent
+%   sums in the same manner by which we defined binary sums.
+
+% We now wish to define a parser term for DFA grammars. In
+% particular, for a DFA $D$ we want to build a term,
+
+% \[
+%   w : \String \vdash \mathsf{parse}_{D} : \mathsf{AccTrace}_{D} \oplus \top
+% \]
+
+% where left injection into the output type denotes acceptance
+% by the parser, and right injection denotes rejection. To
+% build such a parser, it will suffice to construct a term
+
+% \[
+%   w : \String \vdash \mathsf{parse}_{D} : \LinSigTy q Q {\mathsf{Trace}_{D}(q_{0} , q)}
+% \]
+% This is because given a trace of a DFA, we may easily check
+% if we should accept or reject by simply testing
+% if the final state is accepting.
+
+% Because $w$ is a Kleene star of characters, we may construct
+% our desired $\mathsf{parse}_{D}$ term as a $\mathsf{foldl}$
+% over $w$. In the empty case, we just have the trace that
+% ends at the accepting state. In the recursive case, we
+% effectively add to our trace by transitioning one character
+% at a time, as we read them moving across $w$.
+
+% Perhaps this derivation is not too surprising. All it says
+% is that a DFA may be ran by transitioning a single character
+% at a time, and then accepting or rejecting based on the
+% final state. This is exactly what DFAs did initially, so
+% what did we gain? Well, this has the benefit of our type
+% system to ensure its correctness. Moreover, this construction exports to an
+% intrinsically verified and executable decision procedure for DFAs in \theoryabbv
+% embedded in Agda.
+
+% We should note that the
+% construction of a DFA parser requires the addition of an additional but seemingly innocuous
+% axiom.
+
+% \[
+%   \top \cong \left( \bigoplus_{c : \Sigma} c \right)^{*}
+% \]
+
+% That is, the characters in the alphabet $\Sigma$ really are all of the
+% characters. The need for this axiom arises when we go to take a transition
+% within the DFA,
+% as we must ensure that the next character of the string has a corresponding
+% transition from the current state
+% --- which would be guaranteed by determinism provided there are no surprise characters.
+
+% \subsection{Regular Expressions and DFAs}
+% \label{subsec:deriv}
+
+% In order to extend the DFA
+% parser from \cref{subsec:regexparser} to the construction of
+% a verified parser
+% generator for regular expression we need to perform some
+% plumbing establishing an equivalence between regular
+% expressions and DFAs.
+
+% There are several routes we may hope to take in establishing
+% this equivalence. First, we could prove an equivalence
+% between NFAs and regular expressions, and separately prove
+% an equivalence between NFAs and DFAs.
+% In \cref{subsec:eqproofs}, we include a version of
+% Thompson's construction --- which established the
+% equivalence between regular grammars and DFAs. We may additionally
+% hope to internalize a variant of the powerset construction \cite{rabinFiniteAutomataTheir1959}
+% --- which takes as input an NFA and constructs a DFA that
+% recognizes the same language --- and combine the results of
+% Thompson's construction and the powerset constructions to give an equivalence
+% between regular expressions and DFAs.  This route is alluring, as it
+% internalizes several classic grammar-theoretic constructions. However, it may
+% necessitate extensions to the LNL theory, like
+% a propositional truncation, and we have not yet investigated
+% how this would interact with the existing types in the
+% theory. The addition of a propositional truncation may seem
+% harmless, but it is not always immediately clear how
+% distinct constructions will interact. For instance, when
+% exploring LNL models, Benton discovered that the synthesis
+% of linear and dependent types require a new presentation of
+% the $!$ modality from linear logic
+% \cite{bentonMixedLinearNonlinear1995}. That is all to say,
+% this is a work in progress and
+% it is not immediate that the addition of a propositional
+% truncation is adequate for establishing the weak equivalence
+% between NFAs and DFAs.
+
+% We may instead hope to internalize an equivalence between
+% regular grammars and DFAs by using Brzozowski derivatives to
+% directly create a DFA that is weakly equivalent to a given
+% regular expression, as described by Owens et al.
+% \cite{owensRegularexpressionDerivativesReexamined2009}.
+% One characterization of regular grammars is that they are
+% precisely those grammars which have finitely many inequivalent Brzozowski
+% derivatives
+% \cite{brzozowskiDerivativesRegularExpressions1964}.
+% The algorithm used by Owens takes in a
+% regular grammar and generates a DFA that recognizes the same
+% language, and the states in this DFA are the finitely many
+% derivative equivalence classes. We initially had a version of
+% this theorem very roughly internalized in the LNL theory.
+% To our taste, too much of this presentation relied on
+% meta-arguments that lived outside of
+% our formalism, and thus this particular phrasing of the
+% theorem did not translate well into formalization.
+
+% In any case, we believe
+% that revisiting these lines of thought will lead to a
+% satisfactory internalization of the equivalence between
+% regular grammars and DFAs, and thus would bridge the gap
+% between our DFA parser and a full regular expression parser.
+
+% \subsection{Equivalence Between Regular Expressions and Finite Automata}
+% \label{subsec:eqproofs}
+% In this section, we describe a version of Thompson's
+% construction \cite{thompsonProgrammingTechniquesRegular1968}
+%    where we construct an NFA that recognizes a given regular
+% expression. Moreover, we will show that this NFA is strongly equivalent to the
+% original grammar. Witnessing this construction in our syntax has two benefits
+% \begin{enumerate}
+%   \item It reinforces this high-level view that the syntax is a natural and
+%         general setting for formal grammar reasoning, as we demonstrate that
+%         this formal system subsume results from existing systems, and
+%   \item Following the
+%         development of Thompson's construction, we then need
+%         only establish the equivalence of NFAs and DFAs to
+%         complete the full regular expression parser
+% \end{enumerate}
+
+% \begin{theorem}[Thompson]
+%   \label{thm:thompson}
+%   For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same
+%   language as $g$.
+% \end{theorem}
+% We make a pretty straightforward adaptation of Thompson's theorem to our setting,
+
+% \begin{theorem}[Typed Thompson]
+%   \label{thm:typthompson}
+%   For $g$ a regular expression $g$, there is an NFA $N$ such that $g$ is isomorphic
+%   to $\mathsf{AccTrace}_{N}$.
+% \end{theorem}
+
+% \begin{proof}[Proof Sketch]
+%   Recall that regular expressions are inductively defined via
+%   disjunction, concatenation, and Kleene star over literals
+%   and the empty grammar. By induction over regular expressions,
+%   we will construct an NFA that is equivalent to $g$.
+
+%   First, define the recognizing NFA for the empty grammar
+%   $I$.
+
+%   \begin{figure}[h!]
+%   \begin{tikzpicture}[node distance = 25mm ]
+%     \node[state, initial] (1) {$1$};
+%     \node[state, right of=1, accepting] (2) {$2$};
+
+%     \path[->] (1) edge[below] node{$\varepsilon$} (2);
+%   \end{tikzpicture}
+%   \caption{$NFA(I)$}
+%   \label{fig:emptyNFA}
+%   \end{figure}
+
+%   The type of traces from the initial state of $NFA(I)$ to the single
+%   accepting state is given by,
+
+%   \[
+%     \mathsf{Trace}_{NFA(I)}(q_{1}, q_{2}) = \mu
+%       \begin{pmatrix}
+%          g_{q_{1}} := g_{q_{2}} \\
+%          g_{q_{2}} := I
+%       \end{pmatrix}. g_{q_{1}}
+%   \]
+
+%   The accepting traces through $NFA(I)$ are then described
+%   as,
+
+%   \[
+%     \mathsf{AccTrace}_{NFA(I)} = \LinSigTy q {\{1 , 2\}} {\left( \mathsf{Trace}_{NFA(I)}(q_{1} , q) \pair \mathsf{acc}(q) \right)}
+%   \]
+
+%   A quick inspection of \cref{fig:emptyNFA} reveals that the
+%   only reasonable choice for $q$ is $q_{2}$ --- because
+%   state 2 is accepting while state 1 is not. Therefore,
+
+%   \begin{align*}
+%     \mathsf{AccTrace}_{NFA(I)}
+%     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \pair \mathsf{acc}(q_{2}) \\
+%     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \pair \top \\
+%     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})
+%   \end{align*}
+
+%   From here, to prove
+%   $I \cong \mathsf{AccTrace}_{NFA(I)}$ it suffices to show
+%   $I \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$. Below we
+%   give two parse transformers, one from $I$ to
+%   $\mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$ and vice versa.
+
+%   \[
+%     \inferrule
+%     {p : I \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
+%      \exists \text{~transition~} q_{1} \overset{\varepsilon}{\to} q_{2}
+%     }
+%     {p : I \vdash \mathsf{\varepsilon cons}(\mathsf{nil}) : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})}
+%   \]
+
+%   Let $\gamma$ be the substitution $\{ g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$,
+
+%   \[
+%     \inferrule
+%     {
+%       p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash p :
+%         \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \\
+%       x_{1} : \simulsubst {g_{q_{2}}} {\gamma} = I \vdash x_{1} : I \\
+%       x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
+%     }
+%     {
+%       p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.x_{1} , x_{2}.x_{2})(p) : I
+%     }
+%   \]
+
+%   This concludes the proof for the case of the empty
+%   grammar. Let's now walk through the construction for
+%   literal grammars. Given a character $c$, we construct an
+%   NFA that recognizes only the string containing the single
+%   character $c$ as,
+
+%   \begin{figure}[h!]
+%   \begin{tikzpicture}[node distance = 25mm ]
+%     \node[state, initial] (1) {$1$};
+%     \node[state, right of=1, accepting] (2) {$2$};
+
+%     \path[->] (1) edge[below] node{$c$} (2);
+%   \end{tikzpicture}
+%   \caption{$NFA(c)$}
+%   \label{fig:literalNFA}
+%   \end{figure}
+
+%   The automaton in \cref{fig:literalNFA} induces the
+%   following type of traces from $q_{1}$ to $q_{2}$.
+
+%   \[
+%     \mathsf{Trace}_{NFA(c)}(q_{1}, q_{2}) = \mu
+%       \begin{pmatrix}
+%          g_{q_{1}} := c \otimes g_{q_{2}} \\
+%          g_{q_{2}} := I
+%       \end{pmatrix}. g_{q_{1}}
+%   \]
+
+%   Through the same argument as the empty grammar, the only
+%   state that is accepting is $q_{2}$ and thus,
+
+%   \[
+%     \mathsf{AccTrace}_{NFA(c)} \cong \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})
+%   \]
+
+%   To show the desired isomorphism of $c \cong NFA(c)$ we
+%   make a similar argument as we did for the empty grammar
+%   $I$, except we leverage the $\mathsf{cons}$ rule instead
+%   of $\mathsf{\varepsilon cons}$. That is, the parse
+%   transformers in either direction are given as,
+
+%   \[
+%     \inferrule
+%     {\cdot \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
+%      \exists \text{~transition~} q_{1} \overset{c}{\to} q_{2}
+%     }
+%     {p : c \vdash \mathsf{cons}(\mathsf{nil}) :
+%       \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})}
+%   \]
+
+%   \[
+%     \inferrule
+%     {
+%       p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash p :
+%         \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \\
+%       x_{1} : \simulsubst {(c \otimes g_{q_{2}})} {\gamma} = c \otimes I \vdash \mathsf{unitR}(x_{1}) : c \\
+%       x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
+%     }
+%     {
+%       p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.\mathsf{unitR}(x_{1}) , x_{2}.x_{2})(p) : c
+%     }
+%   \]
+
+%   Where $\gamma$ is the substitution
+%   $\{ c \otimes g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$ and
+%   $\mathsf{unitR}$ is a witness to the isomorphism
+%   $c \otimes I \cong c$. Again, we may see that these do
+%   indeed mutually invert each other in the Agda code.
+
+%   It remains to show that the desired isomorphisms are
+%   preserved by $\otimes$, $\oplus$, and Kleene star. Here,
+%   we will give the argument for the disjunction case, the
+%   others are defined quite similarly.
+
+%   Given two NFAs $N$ and $M$, we may define a new NFA that
+%   encodes the disjunction of $N$ and $M$. Denote the
+%   internal states of $N$ by $q_{j}$'s and the internal states
+%   of $M$ by $r_{k}$'s,
+
+%   \begin{figure}[h!]
+%   \begin{tikzpicture}[node distance = 20mm ]
+%     \node[state] (2) {$q_{init}$};
+%     \node[state, initial, below left of=2] (1) {$init$};
+%     \node[state, below right of=1] (3) {$r_{init}$};
+%     \node[right of=2] (4) {$\cdots$};
+%     \node[right of=3] (5) {$\cdots$};
+%     % \node[state, below right of=1] (3) {$3$};
+%     % \node[state, below of=2] (4) {$4$};
+
+%     \path[->] (1) edge[below] node{$\varepsilon$} (2);
+%     \path[->] (1) edge[below] node{$\varepsilon$} (3);
+%     \path[->] (2) edge[below] node{} (4);
+%     \path[->] (3) edge[below] node{} (5);
+
+%     \node[label={[name=l] $N$}, draw,line width=2pt,rounded corners=5pt, fit=(2)(4)] {};
+%     \node[label={[name=l] $M$}, draw,line width=2pt,rounded corners=5pt, fit=(3)(5)] {};
+%   \end{tikzpicture}
+%   \caption{$N \oplus_{NFA} M$}
+%   \label{fig:disjunctionNFA}
+%   \end{figure}
+
+%   \Cref{fig:disjunctionNFA} shows the process of
+%   disjunctively combining NFAs. Precisely, we add a single
+%   new state and we included copies of the states from each of $N$
+%   and $M$. The new state acts as the initial state and has
+%   $\varepsilon$-transitions to the initial states of $N$ and
+%   $M$. We include all of the internal transitions from $N$
+%   and $M$, and the accepting states of $N \oplus_{NFA} M$
+%   are exactly the accepting states in each subautomaton.
+
+%   Let $g$ and $g'$ be two regular grammars such that
+%   $g \cong NFA(g)$ and $g' \cong NFA(g')$. As a matter of
+%   notation\footnote{We shall similarly abuse notation for
+%     $\otimes$ and Kleene. That is, for a regular grammar
+%     $g$, when we write $NFA(g)$ we mean the NFA inductively
+%     built up with the NFA-analogues to the constructors that
+%     built up $g$.}, we will write $NFA(g \oplus g')$ for
+%   $NFA(g) \oplus_{NFA} NFA(g')$. The traces of
+%   $NFA(g \oplus g')$ are then given by,
+
+%   \[
+%     \mathsf{Trace}_{NFA(g \oplus g')}(src , dst) = \mu
+%       \begin{pmatrix}
+%         g_{init} := g_{q_{0}} \oplus g_{r_{0}} \\
+%         g_{q_{j}} := \mathsf{Trans}_{NFA(g)}(q_{j}) \oplus \mathsf{isDst}(q_{j}) \\
+%         g_{r_{k}} := \mathsf{Trans}_{NFA(g')}(r_{k}) \oplus \mathsf{isDst}(r_{k})
+%       \end{pmatrix}.g_{src}
+%     \]
+
+%   where $\mathsf{Trans}$ is used to echo the same syntactic
+%   definitions that appear in the $NFA(g)$ and $NFA(g')$.
+%   Also, $src$ and $dst$ may take on any value in
+%   $Q := \{init\} \cup \{q_{j}\} \cup \{r_{k}\}$, and
+%   $\mathsf{isDst}(q)$ checks if $q$ is equal to $dst$. Which
+%   is all to say, the traces of $NFA(g \oplus g')$ comprise
+%   either a trace of $NFA(g)$, or a trace of $NFA(g')$, and
+%   the transition coming out of $g_{init}$ determines which
+%   subautomaton we step into.
+
+%   The parse transformer from $g \oplus g'$ checks which side
+%   of the sum type we are on, then takes the appropriate step
+%   from $g_{init}$ in the automaton.
+
+%   \[
+%     \inferrule
+%     {
+%       u : g \vdash \iota (\phi (u)) : \mathsf{AccTrace}_{NFA(g \oplus g')} \\
+%       v : g' \vdash \iota' (\psi (v)) : \mathsf{AccTrace}_{NFA(g \oplus g')}
+%     }
+%     {p : g \oplus g' \vdash \mathsf{case}~p \{ \mathsf{inl}(u) \mapsto s , \mathsf{inr}(v) \mapsto r \} : \mathsf{AccTrace}_{NFA(g \oplus g')}}
+%   \]
+
+%   with $\iota$ and $\iota'$ as embeddings from $NFA(g)$ and
+%   $NFA(g')$, respectively, into $NFA(g \oplus g')$,
+%   $\phi: g \cong \mathsf{AccTrace}_{NFA(g)}$, and $\psi: g' \cong \mathsf{AccTrace}_{NFA(g')}$. On a
+%   high level, all this construction does is turn a parse of
+%   $g$ into a parse of $NFA(g)$ and then embeds that inside
+%   of the larger automaton $NFA(g \oplus g')$. Likewise for $g'$.
+
+%   In the other direction, recall that the data of an
+%   accepting trace for $NFA(g \oplus g')$ is a pair of a
+%   trace and a proof that
+%   the end state $q'$ of that trace is accepting. By
+%   multifolding over the first part of that pair, we turn the
+%   term of type
+%   $\mathsf{Trace}_{NFA(g \oplus g')}(init , q')$ into a
+%   trace of either of the subautomata,
+
+%   \[
+%     p : \mathsf{Trace}_{NFA(g \oplus g')}(init , q') \vdash \mathsf{mfold}_{NFA(g \oplus g')} : \mathsf{Trace}_{NFA(g)}(q_{0} , q') \oplus \mathsf{Trace}_{NFA(g')}(r_{0} , q')
+%   \]
+
+%   Additionally, we leverage the fact that the only accepting
+%   states for $NFA(g \oplus g')$ are those from the
+%   subautomata to extract that $q'$ must be an accepting
+%   state from a subautomaton.
+
+%   \[
+%     x : \mathsf{acc}_{NFA(g \oplus g')}(q') \vdash M : \mathsf{acc}_{NFA(g)}(q') \oplus \mathsf{acc}_{NFA(g')}(q')
+%   \]
+
+%   We then combine the trace and proof of acceptance into an
+%   accepting trace of one of the subautomata,
+
+%   \[
+%     p : \mathsf{AccTrace}_{NFA(g \oplus g')} \vdash N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')}
+%   \]
+
+%   Lastly, we then inductively use the isomorphisms $\phi$
+%   and $\psi$ to turn the accepting traces into a parse of
+%   $g$ or $g'$,
+
+%   \[
+%      N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')} \vdash \mathsf{case}~N~\{\mathsf{inl}(n) \mapsto \phi^{-1}(n), \mathsf{inr}(n') \mapsto \psi^{-1}(n')\} : g \oplus g'
+%   \]
+
+%   When you wish to show that these actually maps invert each other to form an isomorphism\max{jumbled sentence}, we invoke the
+%   universal properties of the $\mathsf{Trace}$ types. That is,
+%   $\mathsf{Trace}_{\oplus NFA}$, $\mathsf{Trace}_{N}$, and $\mathsf{Trace}_{M}$
+%   are \emph{initial} algebras to the the functors that define them. Thus, the
+%   embedding of traces through $N$ into $\oplus NFA$ and then mapped back down to $N$
+%   compose to a map from $\mathsf{Trace}_{N}$ to itself. Such a map is unique and
+%   thus must be the identity. The same reasoning shows that the desired
+%   composite do indeed invert each other to form an isomorphism.
+
+%   As discussed above, the other cases and
+%   follow through a similar argument. This
+%   concludes the proof of our variant of Thompson's construction.
+% \end{proof}
+
+% \subsection{Other Automata}
+% \subsubsection{Pushdown Automata}
+% \label{subsubsec:pda}
+% A (nondeterministic) \emph{pushdown automaton} is an automaton that employs a
+% stack. Just like NFAs, they have transitions labeled with characters from a
+% fixed string alphabet $\Sigma$. Additionally, they maintain a stack of
+% characters drawn from a stack alphabet $\Gamma$. They are often represented
+% formally as a 7-tuple $(Q, \Sigma, \Gamma, \delta, q_{0}, Z, F)$,
+
+% \begin{itemize}
+%   \item $Q$ a finite set of states
+%   \item $\Sigma$ a fixed, finite string alphabet
+%   \item $S$ a fixed, finite stack alphabet
+%   \item
+%         $\delta \subset Q \times (\Sigma \cup \{ \varepsilon \}) \times S \to \mathcal{P}(Q \times S^{*})$
+%         the transition function
+%   \item $q_{0} \in Q$ the start state
+%   \item $Z \in S$ the initial stack symbol
+%   \item $F \subset Q$ the accepting states
+% \end{itemize}
+
+% We encode the traces of a pushdown automaton very similarly to those of an NFA,
+% except the transitions of a PDA are instead encoded via the LNL product type
+% $\bigamp$. This is because when simply transitioning via character, a PDA must
+% also pop and push characters onto a stack, which is used as the argument to
+% these dependent functions. Thus, the appropriate formation of traces through a
+% PDA is dependent on a stack.
+
+% Let $S$ be a non-linear type encoding the stack
+% alphabet, and build lists over $S$ as the (non-linear) least fixed-point
+% $\mathsf{List}(S) := \mu X . 1 + S \times X$. Then, the type of states for a
+% PDA $P$ with stack alphabet $S$ are given as a functions that takes in lists $L$,
+% and then makes a case distinction between possible transitions based on what was witnessed as $\mathsf{head}(L)$. The choice of transition
+% will then determine which character to transition by and what word $w$ should be
+% pushed onto the stack. The word that is added to the top of the stack is
+% appended to $\mathsf{tail}(L)$ and then we recursively step into another state
+% called on argument $w \cdot \mathsf{tail}(L)$.
+
+% \begin{gather*}
+%   \mathsf{State} \in {g _{q} : q \in Q} \\
+%   \mathsf{Word} \in \String \\
+%   \mathsf{Char} \in \Sigma   \\
+%   \mathsf{StackChar} \in \Gamma \\
+% \end{gather*}
+% \begin{align*}
+%   \mathsf{Trans}_{P}(q) ::=~&
+%                            \LinPiTy {(hd :: tl)} {\mathsf{List}(S)} {\left( \mathsf{Char} \otimes \mathsf{State}(\mathsf{Word} + tl) \right)}~| \\
+%   & \LinPiTy {(hd :: tl)} {\mathsf{List}(S)} {\left(  \mathsf{State}(\mathsf{Word} + tl) \right)}~|~\mathsf{Trans}_{P}(q) \oplus \mathsf{Trans}_{P}(q)
+% \end{align*}
+
+% That is to say, when transitioning a PDA pops off the head $hd$ of the stack
+
+% \subsubsection{Turing Machines}
+% \label{subsubsec:tm}
+
+% Above we gave a grammar presentation of
+% traces through a PDA by using a non-linear type $S$
+% to encode the stack. We may similarly use pairs $S \times S$
+% to encode the tape of a Turing machine. With two stacks we can simulate the behavior of the
+% infinite tape of a Turing machine. The intuition behind this correspondence is
+% that the left half of the tape is on one stack, the right
+% half of the tape the other, and we treat the tops of stacks
+% like the head of the tape.
+
+% \pedro{I like the overall structure of this section! Showing how the type theory
+% can encode and reason about different kinds of languages is very instructive!}
+
+\section{Denotational Semantics}
+\label{sec:semantics-and-metatheory}
+
+To justify our assertion that \theoryabbv is a syntax for formal
+grammars and parse transformers, we will now define a
+\emph{denotational semantics} that makes this mathematically
+precise. To interpret our types of codes we assume a universe of
+\emph{small} sets.
+
+\subsection{Formal Grammars and Parse Transformers}
+
+First, we need to establish a precise definition of a formal grammar
+that accounts for the possibility of ambiguity. The most common
+definition of a formal grammar is as a generative grammars, defined by
+a set of non-terminals, with a specified start symbol and set of
+production rules. We instead use a more abstract formulation that is
+closer in spirit to the standard definition of a formal language:
 
 \begin{definition}
-  Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
-  two functors $\cat{C} \to \Set$. Their Day convolution tensor
-  product is defined as the following coend formula:
-  \[
-  (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z) 
-  \]
-  Dually, its internal hom is given by the following end formula:
-  \[
-  (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
-  \]
+  A \emph{formal language} $L$ is a function from strings to propositions.
+  A \emph{formal grammar} $A$ is a function from strings to small sets.
 \end{definition}
 
-\begin{lemma}[\cite{day1970}]
-  Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
-  monoidal closed.
-\end{lemma}
-
+We think of the grammar $A$ as taking a string to the set of all parse
+trees for that string. However since $A$ could be any function
+whatsoever there is no requirement that an element of $A(w)$ be a
+``tree'' in the usual sense. This definition provides a simple,
+syntax-independent definition of a grammar that can be used for any
+formalism: generative grammars, categorial grammars, or our own
+type-theoretic grammars. Note that the definition of a formal grammar
+is a generalization of the usual notion of formal language since a
+proposition can be equivalently defined as a subset of a one-element
+set. Then the difference between a formal grammar and a formal
+language is that formal grammars can be \emph{ambiguous} in that there
+can be more than one parse of the same string. Even for unambiguous
+grammars, we care not just about \emph{whether} a string has a parse
+tree, but \emph{which} parse tree it has, i.e., what the structure of
+the element of $A(w)$ is.
+
+The linear types of \theoryabbv can all be interpreted as formal
+grammars, but what about the linear terms? These can be interpreted as
+\emph{parse transformers}
+\begin{definition}
+  Let $L_1,L_2$ be formal languages. Then we say $L_1\subseteq L_2$ if
+  for every string $w$, $L_1(w)\Rightarrow L_2(w)$.
+
+  Let $A_1,A_2$ be formal grammars. Then a parse transformer $f$ from
+  $A_1$ to $A_2$ is a function assigning to each string $w$ a function
+  $f_w : A_1(w) \to A_2(w)$.
+\end{definition}
+First, note that parse transformers generalize language inclusion: if
+$A_1(w), A_2(w)$ are all subsets of a one-element set, then a parse
+transformer is equivalent to showing that $A_1(w) \subseteq A_2(w)$.
+In our denotational semantics, linear terms will be interpreted as
+such parse transformers, and the notions of unambiguous grammar,
+parseability, etc, introduced in \Cref{sec:concepts} can be verified
+to correspond to their intended meanings under this interpretation.
+
+Parse transformers have an obvious notion of composition and identity
+transformer. This defines a \emph{category}.
+\begin{definition}
+  Define $\Grammar$ to be the category whose objects are formal
+  grammars and morphisms are parse transformers.
+\end{definition}
+This category is equivalent to the slice category $\Set/\Sigma^*$ and
+as such is very well-behaved. It is complete, co-complete, cartesian
+closed and carries a monoidal biclosed structure. We will use these
+structures to model the linear types, terms and equalities in
+\theoryabbv.
 %% \begin{theorem}
-%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
-%%   The functor category $[A, \cat{K}]$.
-%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
-%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
+%%   The category $\Grammar$ is complete, co-complete and cartesian
+%%   closed. Furthermore, it is monoidal biclosed with the monoidal
+%%   product given by
+%%   \[ (A \otimes B) w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in A w_1 \wedge b \in B w_2\} \]
+%%   and unit by
+%%   \[ I w = \{ () \pipe w = \varepsilon \} \]
 %% \end{theorem}
-\begin{theorem}
-  If $\cat{C}$ is a locally small monoidal category, then
-  $\Set^{\cat{C}}$ is a Kleene category.
+
+%% \max{todo: need to decide if we need this section here or if it's in the intro already}.
+% The linear typing judgment in our syntax takes on the following schematic form
+% $\Gamma ; \Delta \vdash a : A$. First, $A$ represents a \emph{linear type} in
+% our syntax. The intended semantics of these linear types are formal grammars.
+% That is, the linear typing system is designed to syntactically reflect the
+% behavior of formal grammars. For this reason, we may often interchangeably use
+% the terms ``linear type'' and ``grammar''.
+
+% The term $a$ is an inhabitant of type $A$, which is thought of as a parse tree of
+% the grammar $A$. The core idea of this entire paper follows precisely from this
+% single correspondence: grammars are types, and the inhabitants of these types
+% are parse trees for the grammars.
+
+% $\Gamma$ represents a non-linear context, while $\Delta$ represents a
+% \emph{linear context} dependent on $\Gamma$. These linear contexts behave
+% substructurally. As stated earlier, they are linear --- they do not obey
+% weakening or contraction --- because a character is exhausted once read by a
+% parsing procedure. Moreover, the characters in strings appear in the order in
+% which we read them. We do not have the freedom to freely permute characters,
+% therefore any type theory that is used to reason about formal grammars ought to
+% omit the structural rule of exchange as well. This means that every variable
+% within $\Delta$ must be used \emph{exactly once} and \emph{in order of
+%   occurrence}. Thus, we can think of the linear contexts as an ordered list of
+% limited resources. Once a resource is consumed, we cannot make reference to it
+% again. Variables in a linear context then act like building
+% blocks for constructing patterns over strings.
+
+% We give the base types and type constructors for linear terms. As the
+% interpretation of types as grammars in $\Grammar$ serves as our intended
+% semantics, we simultaneously give the interpretation $\sem \cdot$ of the
+% semantics as grammars.
+
+% \pedro{The paragraphs above are a bit circular, what about the following alternative?}
+% \steven{I am in favor of cutting what's above, but we haven't introduced the
+%   syntax in the main body of the paper yet, even if the figures appeared
+%   earlier. So what you write needs to take that into account}
+% \pedro{In which case we should introduce it in the previous section. I suggest
+%   adding this right after ``and the intuitionistic typing rules in fig 3''}
+\subsection{Semantics}
+
+We can now define our denotational semantics.
+\begin{definition}[Grammar Semantics]
+  We define the following interpretations by mutual recursion on the
+  judgments of \theoryabbv:
+  \begin{enumerate}
+  \item For each non-linear context $\Gamma \isCtx$, we define a set $\sem \Gamma$.
+  \item For each non-linear type $\Gamma \vdash X \isTy$, and element
+    $\gamma \in \sem\Gamma$, we define a set $\sem X \gamma$.
+  \item For each linear type $\Gamma A \isLinTy$ and element $\gamma
+    \in \sem\Gamma$, we define a formal grammar $\sem{A}\gamma$. We
+    similarly define a formal grammar $\sem\Delta\gamma$ for each
+    linear context $\Gamma \Delta\isLinCtx$.
+  \item For each non-linear term $\Gamma \vdash M : X$ and $\gamma \in \sem{\Gamma}$, we define an element $\sem{M}\gamma \in \sem{X}\gamma$.
+  \item For each linear term $\Gamma; \Delta \vdash e : A$ and $\gamma \in \sem{\Gamma}$ we define a parse transformer from $\sem{\Delta}\gamma$ to $\sem{A}\gamma$.
+  \end{enumerate}
+  And we verify the following conditions:
+  \begin{enumerate}
+  \item If $\Gamma \vdash X \isSmall$, then $\sem X \gamma$ is a small set.
+  \item If $\Gamma \vdash X \cong X'$ then for every $\gamma$, $\sem{X}\gamma = \sem{X'}\gamma$
+  \item If $\Gamma \vdash A \cong A'$ then for every $\gamma$, $\sem{A}\gamma = \sem{A'}\gamma$
+  \item If $\Gamma \vdash M \cong M' : X$ then for every $\gamma$, $\sem{M}\gamma = \sem{M'}\gamma$
+  \item If $\Gamma;\Delta \vdash e \cong e' : A$ then for every $\gamma$, $\sem{e}\gamma = \sem{e'}\gamma$
+  \end{enumerate}
+\end{definition}
+
+For reasons of brevity, we present only a selection of cases in
+\Cref{fig:semantics}. The full details are in the appendix.\max{todo:
+  put the full details in the appendix}.
+\max{TODO: how to describe $\mu$ succinctly?}\max{make sure $\lto$ and $\tol$ are right}
+\begin{figure}
+  \begin{align*}
+    \sem{c}\gamma\,w &= \{ c | w = c\}\\
+    \sem{I}\gamma\,w &= \{ () \pipe w = \varepsilon \}\\
+    \sem{A \otimes B}\gamma\,w &= \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
+    \sem{\bigoplus_{x:X} A}\gamma\,w &= \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
+    \sem{\bigwith_{x:X} A}\gamma\,w &= \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w\\
+    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
+    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
+    \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
+    \sem{\SPF\,X}\gamma &= \{ F : \textrm{Functor}(\textrm{Grammar}^{\sem{X}\gamma}, \textrm{Grammar}) \pipe F \textrm{ preserves $\omega$-colimits.} \}\\
+    \sem{\el(F)}\gamma\,G &= \sem{A}\gamma\,G\\
+    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:adamek}}
+  \end{align*}
+  \caption{Grammar Semantics}
+\end{figure}
+
+\max{TODO: prose}
+
+\max{todo: spelling}
+\begin{theorem}[Adamek]
+  If a functor $F : \mathcal C \to \mathcal C$ preserves
+  $\omega$-colimits and $\mathcal C$ has an initial object and
+  $\omega$-colimits, then $F$ has an initial algebra given by the
+  colimit of the chain
+  \[ 0 \to F(0) \to F(F(0)) \to \cdots \]
 \end{theorem}
-\begin{proof}
-
-  By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
-  is a presheaf category, it has coproducts. Furthermore, the tensor
-  is a left adjoint, i.e. it preserves colimits and, therefore, it is
-  a distributive category.
-
-  As for the Kleene star, since presheaf categories admit small colimits,
-  the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
-  defined as the filtered colimit of the diagrams:
-
-  From Theorem~\ref{th:kleeneclosed} it follows that these initial
-  algebras satisfy the required isomorphisms and this concludes the
-  proof.
-\end{proof}
-
-\begin{corollary}
-  For every alphabet $\Sigma$, the presheaf category $\Set^{\cat{\Sigma^*}}$
-  is a Kleene category.
-\end{corollary}
-\begin{proof}
-  Note that string concatenation and the empty string make the
-  discrete category $\Sigma^*$ a strict monoidal category.
-\end{proof}
-
-Much like in the posetal case, the abstract structure of a Kleene
-category is expressive enough to synthetically reason about formal
-languages. A significant difference between them is that while Kleene
-algebras can reason about language containment, Kleene categories can
-reason about parsing as well.
-
-For the rest of the paper we will use the presheaf category
-$\Set^{\cat{\Sigma^*}}$ as a concrete model that will serve as our
-guide when expanding abstract formalism presented in this section so
-that it can handle more classes of languages beyond regular ones.
-
-\paragraph{Beyond Simple Types}
-
-Though Kleene categories are expressive enough to reason about
-concepts that are outside of reach of Kleene algebras, their
-simply-typed nature makes them not so expressive from a type theory
-point of view. This is limiting because type theories are tested
-syntactic frameworks for manipulating complicated categories
-structures while avoiding some issues common in category theory, such
-as coherence issues.
-
-With this in mind, we want to design a dependent type theory that
-builds on top of Kleene categories with the goal of internalizing
-various reasoning principles for languages and their parsers. This
-leads us to the abstract notion of model we are interested in
-capturing with our type theory.
-
-\begin{definition}
-  A model is a locally Cartesian category with two hierarchies of
-  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
-  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
-  $U_i$ to be closed under dependent products and sum,
-  $L_i$ to be closed under the Kleene category connectives, 
-  dependent products, left and right closed structures, with
-  a type constructor $G : L_i \to U_i$ and a linear dependent sum
-  going the other direction.
-\end{definition}
-
-\begin{theorem}
-  The presheaf category $\Set^{\cat{\Sigma^*}}$ is a model.
-\end{theorem}
-
-\section{Related work}
+
+%% That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
+%% complete to parses of $B$. In this manner, the linear function types generalize
+%% Brzozowksi's notion of derivative
+%% \cite{brzozowskiDerivativesRegularExpressions1964}.
+%% Brzozowski initially only gave an accounting of this operation for
+%% generalized regular expressions, but later work from Might et al.\ demonstrates that the same
+%% construction can be generalized to context free grammars
+%% \cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
+%% types, the same notion of derivative extends to the grammars of \theoryabbv.
+
+%% Note, to ensure that the linear function types do indeed generalize Brzozowski
+%% derivatives, we must include the equalities in \cref{fig:brzozowskideriv} as axioms\max{we must? why must we}.
+%% \pedro{We should further explain why these equations are required, or at the very
+%% least give some intuition}
+
+%% Of course, all the above statements for the left function type also have
+%% corresponding analogues for the
+%% right-handed counterpart.
+
+%% \begin{figure}
+%% \begin{align*}
+%%   c\lto c &\equiv I\\
+%%   c\lto d &\equiv \bot\\
+%%   c\lto I &\equiv \bot\\
+%%   c\lto 0 &\equiv \bot\\
+%%   c\lto (A \otimes B) & \equiv (c\lto A) \otimes B + (A \amp I) \otimes (c\lto B)\\
+%%   c\lto A^* &\equiv (A \amp I)^{*} \otimes (c \lto A) \otimes A^* \\
+%%   c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
+%% \end{align*}
+%% \caption{Equality for Brzozowski Derivatives}
+%% \label{fig:brzozowskideriv}
+%% \end{figure}
+
+%% \paragraph{LNL Dependent Types}
+%% \steven{talk about how binary sum and binary with are definable over $\Bool$}
+%% Given a non-linear type $X$, we may form both the dependent product type
+%% $\LinPiTy x X A$ and the dependent pair type $\LinSigTy x X A$ as linear types
+%% where $x$ is free in $A$.
+
+%% $\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+
+%% $\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
+
+%% The grammar semantics of the linear product type is
+%% indeed a dependent function out of the
+%% semantics of $X$. Likewise, the grammar semantics of the linear dependent pair type is a dependent
+%% pair in $\Set$.\pedro{The connection to grammar semantics needs to be better explained. For instance,
+%% we should explain that disjunction corresponds to ``or'' of grammars}
+
+%% Note that even though we do not take the binary additive conjunction and
+%% disjunction as primitive, we may define them via these LNL dependent types. In
+%% particular, via a dependence on $\Bool$.
+%% %
+%% \[
+%%   A \amp B := \LinPiTy b \Bool {C(b)}
+%% \]
+%% \[
+%%   A \oplus B := \LinSigTy b \Bool {C(b)},
+%% \]
+%% where $C(\true) = A, C(\false) = B$.
+
+%% \paragraph{Universal Type}\max{this is absolutely not the right terminology. Universal type means multiple other completely different things}
+%% \steven{Definable as nullary product}
+%% The universal type $\top$ may be formed in any context.
+%% %
+%% \[ \sem{\top} \gamma w = \{ \ast \}\]
+%% %
+%% Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.
+
+%% \paragraph{Empty Type}
+%% \steven{Definable as nullary sum}
+%% The empty type $\bot$ has no inhabitants. The elimination for the empty type
+%% witnesses the principle of explosion --- i.e.\ from a term of type $\bot$ we may
+%% introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.
+%% %
+%% \[ \sem {\bot} \gamma w = \emptyset \]
+%% %
+%% \paragraph{The $G$ Modality}
+%% \steven{Rename this}
+%% \steven{Compare to persistence modalities from separation logic}
+%% The $G$ modality provides the embedding of linear types in the empty context
+%% into non-linear types. The introduction and elimination forms for $G A$ obey the
+%% same rules as given in \cite{bentonMixedLinearNonlinear1995} and \cite{krishnaswami_integrating_2015}.
+
+%% The left adjoint $F$ from non-linear types back to linear types may be defined
+%% as,
+%% %
+%% \[
+%%   F X := \LinSigTy a A I
+%% \]
+%% %
+%% \steven{Elaborate on $G$ more and rename $G A$to $\ltonl A$ in this section}
+
+%% \pedro{Agreed :) maybe mention connections to persistent propositions from the
+%% separation logic literature}
+
+%% The semantics of $G A$ are exactly the semantics of $A$ at the empty word $\varepsilon$.
+%% %
+%% \[ \sem{G A} \gamma = \sem{A} \gamma \varepsilon \]
+%% %
+%% %
+%% \paragraph{Recursive Types}
+%% \steven{reflavor this section as ``indexed inductive'' instead of ``recursive'',
+%% and entirely rework with the new inductive definition}
+%% \steven{The non-indexed inductive are definable as being trivially indexed over unit}
+%% Recursive linear types may be defined via the least-fixed point operator $\mu$.
+% The grammar semantics of which are the fixed-point of sets induced by the
+% grammar semantics of $A$.
+% %
+% \[ \sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x) \]
+% %
+% \pedro{We should justify, even if briefly, the existence of this
+%   fixed-point}
+
+% Observe that we need not take the Kleene star as a primitive
+% grammar constructor, as it is definable as a fixed point.
+% The Kleene star of a grammar $g$ is given as,
+
+% \[
+%   g^{*} := \mu X . I \oplus (g \otimes X)
+% \]
+
+% \begin{figure}[h!]
+% \begin{mathpar}
+%   \inferrule
+%   {\Gamma ; \Delta \vdash p : I}
+%   {\Gamma ; \Delta \vdash \mathsf{nil} : g^{*}}
+
+%   \and
+
+%   \inferrule
+%   {\Gamma ; \Delta \vdash p : g \\ \Gamma ; \Delta' \vdash q
+%   : g^{*}}
+%   {\Gamma ; \Delta \vdash \mathsf{cons}(p , q) : g^{*}}
+
+%   \\
+
+%   \inferrule
+%   {
+%     \Gamma ; \Delta \vdash p : g^{*} \\
+%     \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
+%     \Gamma ; x : g , y : h \vdash p_{\ast} : h
+%   }
+%   {\Gamma ; \Delta \vdash \mathsf{foldr}(p_{\varepsilon} , p_{\ast}) : g^{*}}
+% \end{mathpar}
+% \caption{Kleene Star Rules}
+% \label{fig:star}
+% \end{figure}
+
+% Likewise, $g^{*}$ has admissible introduction and
+% elimination rules, shown in \cref{fig:star}. Note that this
+% definition of $g^{*}$ and these
+% rules arbitrarily assigns a handedness to the Kleene star.
+% We could have just as well took it to be a fixed point of
+% $I \oplus (X \otimes g)$. In fact, the definitions are
+% equivalent, as the existence of the $\mathsf{foldl}$ term below
+% shows that $g^{*}$ is also a fixed point of
+% $I \oplus (X \otimes g)$.
+% This is a marked point of difference from Kleene
+% algebra with recursion, where the fixed points for the left and right variants
+% of Kleene star need not agree \cite{leiss}.
+
+% \begin{equation}
+%   \label{eq:foldl}
+%   \inferrule
+%   {
+%     \Gamma ; \Delta \vdash p : g^{*} \\
+%     \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
+%% %     \Gamma ; y : h , x : h \vdash p_{\ast} : h
+%% %   }
+%% %   {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
+%% % \end{equation}
+
+%% % In fact, the $\mathsf{foldl}$ term is defined using
+%% % $\mathsf{foldr}$ --- much in the same way one
+%% % may define a left fold over lists in terms of a right fold
+%% % in a functional programming language.
+%% % The underlying trick is to fold over a list of linear functions
+%% % instead of the original string. We curry each character $c$
+%% % of the string into a function that concatenates $c$, and
+%% % right fold over this list of linear functions. Because function
+%% % application is left-associative, this results in a left
+%% % fold over the original string.
+
+%% % We only take fixed points of a single variable as a
+%% % primitive operation in the type theory, but we may apply
+%% % Beki\`c's theorem \cite{Bekić1984} to define an admissible
+%% % notion of multivariate fixed point. This is particularly
+%% % useful for defining grammars that encode the states of an
+%% % automaton, as we will see in \cref{sec:automata}. In \cref{fig:multifix} we provide the
+%% % introduction and elimination principles for such a fixed
+%% % point, where $\sigma$ is the substitution that unrolls the
+%% % mutually recursive definitions one level. That is,
+
+%% % \begin{align*}
+%% %   \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
+%% % \end{align*}
+
+%% % \begin{figure}
+%% % \begin{mathpar}
+%% %   \inferrule
+%% %   {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
+%% %   {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
+
+%% %   \and
+
+%% %   \inferrule
+%% %   {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
+%% %              \Gamma ; x_{j} : \simulsubst {A_{j}}{\sigma} \vdash e_{j} : B_{j}\quad \forall j
+%% %   }
+%% %   {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
+%% % \end{mathpar}
+%% % \caption{Multi-fixed Points}
+%% % \label{fig:multifix}
+%% % \end{figure}
+
+%% %% \section{Categorical Semantics of \theoryabbv}
+%% %% \label{sec:categorify}
+
+%% As described in previous sections, \theoryabbv has a ``standard''
+%% interpretation where non-linear types denote sets, linear types denote
+%% formal grammars, non-linear terms denote functions and linear terms
+%% denote parse transformers.
+%% %
+%% Further, as is typical for type theory, every type constructor in the
+%% calculus is interpreted by using a \emph{universal construction} in
+%% either the category of sets or the category of grammars\footnote{the
+%% only exception is that the axioms for Brzozowski derivatives we
+%% considered do not follow solely from universal properties. In this
+%% section we focus on models that do not necessarily satisfy these
+%% axioms.}.
+%% %
+%% This leads to an immediate opportunity for generalization: the type
+%% theory has in addition to the standard grammar-theoretic
+%% interpretation, an interpretation in any categorical structure that
+%% exhibits the same universal constructions.
+%% %
+%% On the one hand this gives us a structured way to formalize the
+%% standard semantics precisely, but additionally it enables us to
+%% consider \emph{non-standard} models in the next section that point to
+%% further applications as well as meta-theoretic results.
+
+%% In developing this categorical semantics, we will start with the
+%% closest analogue from formal language theory: Kleene algebra.  Kleene
+%% algebras are an important tool in the theory of regular languages
+%% serving as a bridge between algebraic reasoning and equivalence of
+%% regular expressions. More broadly, through various extensions, they
+%% serve as a theoretical substrate to studying different kinds of formal
+%% languages.
+%% %
+%% We can then see that the ``regular fragment'' of our type theory
+%% (i.e., just characters, $\otimes$, $\oplus$, $0,1$ and a Kleene star)
+%% has models in what we call a \emph{Kleene category} a categorification
+%% of Kleene algebra from posets to categories.
+%% %
+%% We then further develop this into our final notion of model, which we
+%% call a \emph{Chomsky category} as it can model not just regular
+%% languages, but the full Chomsky hierarchy.
+
+%% \subsection{Comparison to Kleene Algebra}
+%% \subsection{Kleene Algebra and Kleene Categories}
+%% A Kleene algebra is a tuple $(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is
+%% a set, $+$ and $\cdot$ are binary operations over $A$, $(-)^*$ is a
+%% function over $A$, and $1$ and $0$ are constants. These structures
+%% satisfy the axioms depicted in Figure~\ref{fig:axioms}.
+
+%% \begin{figure}
+%%   \begin{align*}
+%%     x + (y + z) &= (x + y) + z & x + y &= y + x\\
+%%     x + 0 &= x & x + x &= x\\
+%%     x(yz) &= (xy)z & x1 &= 1x = x\\
+%%     x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
+%%     x0 &= 0x = x & & \\
+%%     1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
+%%      b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
+%%   \end{align*}
+%%   \caption{Kleene algebra axioms}
+%%   \label{fig:axioms}
+%% \end{figure}
+
+%% The addition operation can be used to define the partial order
+%% structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
+%% formal languages, this order structure can be used to model language
+%% containment. In this section, we categorify the concept of Kleene
+%% algebra and build on top of it in order to define an abstract theory
+%% of parsing. We start by defining \emph{Kleene categories}.
+
+%% \begin{definition}
+%%   A Kleene category is a distributive monoidal category $\cat{K}$
+%%   such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
+%%   = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
+%%   algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
+%%   X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogous isomorphism
+%%   for $G_{A,B}$ also holds.
+%% \end{definition}
+
+%% As a sanity check, note that Kleene algebras are indeed examples of
+%% Kleene categories.
+
+%% \begin{example}
+%%   Every Kleene algebra, seen a posetal category, is a Kleene category.
+%%   The product $\cdot$ is a monoidal product and the addition is a
+%%   least-upper bound, which corresponds to a coproduct. Lastly, the
+%%   axioms of the Kleene star have a direct correspondence to the
+%%   coherence conditions postulated by the initial algebras of Kleene
+%%   categories.
+%% \end{example}
+
+%% This example provides a neat categorical justification to how
+%% restrictive Kleene algebras are in terms of reasoning about
+%% languages. By only having at most one morphism between objects, there
+%% is not a lot of information they can convey. In this case, the only
+%% information you get is language containment. As demonstrated by
+%% $\mathbf{Gr}$, the extra degrees of freedom granted by having more
+%% morphisms gives you more algebraic structure for reasoning about
+%% languages.
+
+%% For the next example, we see an unexpected connection with the theory
+%% of substructural logics.
+
+%% \begin{example}
+%%   The opposite category of every Kleene category is a model of a variant of
+%%   conjunctive ordered logic, where the Kleene star plays the role of the ``of
+%%   course'' modality from substructural logics which allows hypotheses to
+%%   be discarded or duplicated.
+%% \end{example}
+
+%% As we have seen, the proposed axioms are a direct translation of the
+%% Kleene algebra axioms to a categorical setting. Its most unusual aspect is the
+%% axiomatization of the Kleene star as a family of initial algebras
+%% satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
+%% more structure, then these isomorphisms hold ``for free''.
+
+%% \begin{theorem}
+%%   \label{th:kleeneclosed}
+%%   Let $\cat{K}$ be a Kleene category such that it is also monoidal
+%%   closed.  Then, the initial algebras isomorphisms hold automatically.
+%% \end{theorem}
+%% \begin{proof}
+%%   We prove this by the unicity (up-to isomorphism) of initial
+%%   algebras. Let $[hd, tl]: I + (\mu X.\, F_{A, I}(X)) \otimes A \to
+%%   (\mu X.\, F_{A, I}(X))$ be the initial algebra structure of $(\mu
+%%   X.\, F_{A, I}(X))$ and consider the map $[hd, tl] : B + B \otimes
+%%   (\mu X.\, F_{A, I}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
+%%     I}(X))$.
+
+%%   Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
+%%   we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,I} \to B \lto Y$. We can show existence and
+%%   uniqueness by showing that the diagram on top commutes if, and
+%%   only if, the diagram on the bottom commutes:
+
+%% % https://q.uiver.app/#q=WzAsOCxbMCwwLCJCICsgQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFsyLDAsIkIgKyBZIFxcb3RpbWVzIEEiXSxbMiwyLCJZIl0sWzAsMiwiQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDMsIjEgKyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDUsIlxcbXUgWC5cXCwgMSArIFggXFxvdGltZXMgQSJdLFsyLDMsIjEgKyAoIEIgXFxsdG8gWSkgXFxvdGltZXMgQSJdLFsyLDUsIkIgXFxsdG8gWSJdLFswLDEsImlkICsgKGlkIFxcb3RpbWVzIGg7IGV2KSBcXG90aW1lcyBpZF9BIl0sWzEsMiwiW2YsZ10iXSxbMywyLCJpZCBcXG90aW1lcyBoOyBldiIsMl0sWzAsMywiW2lkIFxcb3RpbWVzIGgsIGlkIFxcb3RpbWVzIHRsXSIsMl0sWzQsNiwiaWQgKyAoaCBcXG90aW1lcyBYKVxcb3RpbWVzIGlkIl0sWzUsNywiaCIsMl0sWzYsNywiW2YnLCBnJ10iXSxbNCw1LCJbaGQsIHRsXSIsMl1d
+%% \[\begin{tikzcd}
+%% 	{B + B \otimes (\mu X.\, I + X \otimes A)} && {B + Y \otimes A} \\
+%% 	\\
+%% 	{B \otimes (\mu X.\, I + X \otimes A)} && Y \\
+%% 	{I + (\mu X.\, I + X \otimes A)} && {I + ( B \lto Y) \otimes A} \\
+%% 	\\
+%% 	{\mu X.\, I + X \otimes A} && {B \lto Y}
+%% 	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
+%% 	\arrow["{[f,g]}", from=1-3, to=3-3]
+%% 	\arrow["{id \otimes h; ev}"', from=3-1, to=3-3]
+%% 	\arrow["{[id \otimes h, id \otimes tl]}"', from=1-1, to=3-1]
+%% 	\arrow["{id + (h \otimes X)\otimes id}", from=4-1, to=4-3]
+%% 	\arrow["h"', from=6-1, to=6-3]
+%% 	\arrow["{[f', g']}", from=4-3, to=6-3]
+%% 	\arrow["{[hd, tl]}"', from=4-1, to=6-1]
+%% \end{tikzcd}\]
+%%   This equivalence follows by using the adjunction structure given
+%%   by the monoidal closed structure of $\cat{K}$. A completely analogous
+%%   argument for $G_{A,B}$ also holds. Furthermore, by generalizing the
+%%   construction of \Cref{sec:formaltype}, we can also show that from
+%%   the monoidal closed assumption it follows that $\mu X.\, F_{A, I}(X) \cong \mu X.\, G_{A, I}(X)$
+%% \end{proof}
+
+%% Something surprising about this lemma is that it provides an alternative
+%% perspective on the observation that if a Kleene algebra has an
+%% residuation operation, also called action algebra \cite{kozen1994action},
+%% then the Kleene star admits a simpler axiomatization.
+
+%% Since we want Kleene categories to generalize our notion of formal
+%% grammars as presheaves $\String \to \Set$, we prove that they do
+%% indeed form a Kleene category. We start by presenting a well-known
+%% construction from presheaf categories.
+
+%% \begin{definition}
+%%   Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
+%%   two functors $\cat{C} \to \Set$. Their Day convolution tensor
+%%   product is defined as the following coend formula:
+%%   \[
+%%   (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z)
+%%   \]
+%%   Dually, its internal hom is given by the following end formula:
+%%   \[
+%%   (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
+%%   \]
+%% \end{definition}
+
+%% \begin{lemma}[Day \cite{day1970construction}]
+%%   Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
+%%   monoidal closed.
+%% \end{lemma}
+
+%% %% \begin{theorem}
+%% %%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
+%% %%   The functor category $[A, \cat{K}]$.
+%% %%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
+%% %%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
+%% %% \end{theorem}
+%% \begin{theorem}
+%%   If $\cat{C}$ is a locally small monoidal category, then
+%%   $\Set^{\cat{C}}$ is a Kleene category.
+%% \end{theorem}
+%% \begin{proof}
+
+%%   By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
+%%   is a presheaf category, it has coproducts. Furthermore, the tensor
+%%   is a left adjoint, i.e. it preserves colimits and, therefore, it is
+%%   a distributive category.
+
+%%   As for the Kleene star, since presheaf categories admit small colimits,
+%%   the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
+%%   defined as the filtered colimit of the diagrams:
+
+%%   From Theorem~\ref{th:kleeneclosed} it follows that these initial
+%%   algebras satisfy the required isomorphisms and this concludes the
+%%   proof.
+%% \end{proof}
+
+%% \begin{corollary}
+%%   For every alphabet $\Sigma$, the presheaf category $\Set^{\cat{\Sigma^*}}$
+%%   is a Kleene category.
+%% \end{corollary}
+%% \begin{proof}
+%%   Note that string concatenation and the empty string make the
+%%   discrete category $\Sigma^*$ a strict monoidal category.
+%% \end{proof}
+
+%% Much like in the posetal case, the abstract structure of a Kleene
+%% category is expressive enough to synthetically reason about regular
+%% languages. A significant difference between them is that while Kleene
+%% algebras can reason about language containment, Kleene categories can
+%% reason about \emph{ambiguity}, \emph{strong equivalence} of grammars.
+
+%% \subsection{Lambek Hyperdoctrines and Chomsky Hyperdoctrines}
+
+%% Though Kleene categories are expressive enough to reason about
+%% concepts that are outside of reach of Kleene algebras, their
+%% simply-typed nature makes them not so expressive from a type theoretic
+%% point of view. This is limiting because type theories are successful
+%% syntactic frameworks for manipulating complicated categorical
+%% structures while avoiding some issues common in category theory, such
+%% as coherence issues.
+
+%% With this in mind, we want to design a categorical semantics that
+%% builds on top of Kleene categories with the goal of extending them
+%% with dependent types and making them capable reasoning about languages
+%% and their parsers. This leads us to the abstract notion of model we
+%% are interested in capturing with \theoryabbv: a \emph{Chomsky
+%% category}.
+
+%% We do this in two stages: first we define a \emph{Lambek hyperdoctrine} to be a
+%% notion of model for the judgmental structure of \theoryabbv: that is we have
+%% notions of linear and non-linear type, contexts, substitution and terms, but do
+%% not assume any particular type constructions exist. Then we will define when a
+%% Lambek hyperdoctrine is a Chomsky hyperdoctrine, meaning it can interpret all
+%% the type formers of \theoryabbv and therefore arbitrary strength grammars.
+
+%% \begin{definition}[Lambek Hyperdoctrine]
+%%   A \emph{Lambek hyperdoctrine} consists of
+%%   \begin{enumerate}
+%%   \item A category $\mathcal C$ with a terminal object.
+%%   \item A category with families structure over $\mathcal C$.
+%%   \item A contravariant functor $L : \mathcal C^{o} \to \textrm{MultiCat}$ from
+%%     $\mathcal C$ to the category of Multicategories.
+%%   \end{enumerate}
+%% \end{definition}
+%% Here the objects of $\mathcal C$ model the non-linear contexts, and morphisms
+%% model non-linear substitutions. The category with families structure over
+%% $\mathcal C$ models the dependent non-linear types. Finally the
+%% ``hyperdoctrine'' of multicategories $L$ models the linear types and terms. The
+%% fact that this is functorial in $\mathcal C$ corresponds to the fact that linear
+%% types are all relative to a non-linear context and that linear variables and
+%% composition commute with non-linear substitution.
+
+%% We will have three main Lambek hyperdoctrines of interest in this paper: the
+%% ``standard model'' of sets and grammars, the ``syntactic model'' given by our
+%% type theory itself and lastly a gluing model introduced in
+%% Section~\ref{sec:canonicity} to prove the canonicity theorem.
+%% \begin{example}
+%%   \begin{enumerate}
+%%   \item Let $\mathcal C$ be the category of sets, equipped with its usual
+%%     category with families structures of families. Define $L : \mathcal C^{o}
+%%     \to \textrm{MultiCat}$ to map $L(X)$ to the representable multicategory
+%%     $(\Set^{\Sigma^*})^X$ of $X$-indexed families of grammars where the monoidal
+%%     structure is given pointwise by the Day convolution monoidal structure.
+%%   \item We can build a ``syntactic'' Lambek hyperdoctrine $\Syn(\Sigma)$ from the syntax
+%%     itself: the category $\mathcal C$ is given by non-linear contexts and
+%%     substitutions, the category with families by the non-linear types and terms
+%%     and the hyperdoctrine of multicategories by the linear types and terms.
+%%   \end{enumerate}
+%% \end{example}
+
+%% Next each type linear and non-linear type constructor corresponds to extra data
+%% on a Lambek hyperdoctrine.
+%% \begin{definition}
+%%   \label{def:chomsky-data}
+%%   \begin{enumerate}
+%%   \item Dependent type structure: standard (say extensional type theory, give a reference)
+%%   \item Universes
+%%   \item Inductive grammars?
+%%   \item Non-inductive Grammar constructors: standard monoidal category stuff
+%%   \end{enumerate}
+%% \end{definition}
+%% \begin{definition}[Chomsky Hyperdoctrine]
+%%   A Chomsky Hyperdoctrine is a Lambek hyperdoctrine equipped with all of the data in Definition~\ref{def:chomsky-data}.
+%% \end{definition}
+%% \pedro{We should probably define some of the words in this definition}
+%% \begin{definition}
+%%   A Chomsky category is a locally Cartesian category with two hierarchies of
+%%   universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
+%%   every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
+%%   $U_i$ to be closed under dependent products and sum,
+%%   $L_i$ to be closed under the Kleene category connectives,
+%%   dependent products, left and right closed structures, with
+%%   a type constructor $G : L_i \to U_i$ and a linear dependent sum
+%%   going the other direction.
+%% \end{definition}
+
+%% \steven{Max suggests augmenting the definition of a Chomsky category to
+%%   something like two categories $L$ and $U$, and $U$ is Cartesian closed I don't
+%%   fully recall the rest.
+
+%%   My best guess is that you take $L$ a Kleene category with a hierarchy of universes
+%%   two, and you further require that $L$ is $Psh(U)$-enriched, except he
+%%   suggested a further adjective on these presheaves. Perhaps representability?
+%% }
+
+%% \begin{theorem}
+%%   The presheaf category $\Grammar = \Set^{\cat{\Sigma^*}}$ is a Chomsky category.
+%% \end{theorem}
+
+%% Further, the syntactic category of \theoryname is manifestly a Chomsky category.
+
+%% \pedro{This is likely true, but if we explicitly say so, this warrants a proof. I think that
+%% if we don't say anything about the syntactic category, reviewers won't mind.}
+
+%% \steven{I agree that we don't want to say anything that opens unnecessary
+%%   questions for proofs we haven't written. However, it seems hard to make the
+%%   case that we have the right categorical model of the syntax if this statement
+%%   isn't true. By restating our definition of Chomsky category, this should be
+%%   obvious or a quick proof}
+
+%% \pedro{I agree, then we should add the quick proof :) }
+
+%% \subsection{Concrete Models of \theoryabbv}
+%% \label{sec:othermodels}
+
+%% \steven{Because we can define a version of semantic actions internally, we
+%%   shouldn't put it as a separate model}
+
+%% One of the powers of type theories is that it can be profitable to interpret them
+%% in various models. In this section, by using our just-defined Chomsky categories,
+%% we show how other useful concepts from formal language theory can also be organized
+%% as models of \theoryabbv. We illustrate this point by providing two examples that
+%% are closely related to the theory of formal languages: language equivalence and
+%% semantic actions. Furthermore, in order to justify how $\mathbf{Gr}$ relates to
+%% more traditional notions of parsing, we define a glued model that proves a
+%% canonicity property of grammar terms.
+
+%% \subsection{Language Semantics}
+%% Every grammar induces a language semantics. Also languages can be taken as a
+%% propositionally truncated view of the syntax. Logical equivalence should induce
+%% weak equivalence, and thus even give a syntactic way to reason about language equivalence.
+%% \steven{TODO language semantics}
+
+%% % \subsection{Semantic Actions}
+%% % \steven{Tentatively planning to cut this subsection for an internal representation
+%% %   of semantic actions}
+%% % Returning to the problem of parsing, the output of a parse usually is not the
+%% % literal parse tree. Rather, the output is the result of some \emph{semantic
+%% %   actions} ran on the parse tree, which usually serve to remove some syntactic
+%% % details that are unnecessary for later processing.
+
+%% % Given a grammar $G : \String \to \Set$, we define a semantic action to be a set
+%% % $X$ with a function $f$ that produces a semantic element from any parse of $G$.
+
+%% % \[
+%% %   f : \PiTy w \String {G w \to X}
+%% % \]
+
+%% % Further, semantic actions can be arranged into a structured category.
+%% % Define $\SemAct$ as the comma
+%% % category $\Grammar / \Delta$, where $\Delta : \Set \to \Grammar$ defines a
+%% % discrete presheaf. That is, for a set $X$, $\Delta (X)(w) = X$ for all
+%% % $w \in \String$. As $\SemAct$ is defined as a comma category, it has a forgetful
+%% % functor into $\Grammar$. That is, $\SemAct$ serves as a notion of formal
+%% % grammar. Moreover, $\SemAct$ is a model of \theoryabbv.
+
+%% % \steven{It being a notion of formal grammar is distinct from being a model. This
+%% % probably warrants a proof}
+
+%% % \steven{TODO semantic actions}
+
+%% % \pedro{This is a very nice opportunity of showing off the supremacy of denotational
+%% %   reasoning ;) We should probably prove the gluing lemma in the previous section
+%% %   and apply it here and in the canonicity section. The actual proof might have to be
+%% %   moved to the appendix, though}
+
+%% \subsection{Parse Canonicity}
+%% Canonicity is an important metatheoretic theorem in the type theory
+%% literature.  It provides insight on the normal forms of terms and,
+%% therefore, on its computational aspects. Frequently, proving
+%% canonicity for boolean types, i.e. every closed term of type bool
+%% reduces to either true or false, is enough to justify that the type
+%% theory being studied is well-behaved. In our case, however, since we
+%% want to connect \theoryabbv to parsers, we must provide a more
+%% detailed account of canonicity. In particular, we give a nonstandard semantics
+%% of \theoryabbv that carries a proof of canonicity along with it.
+
+%% If $\cdot \vdash A$ is a closed linear type then there are
+%% two obvious notions of what constitutes a ``parse'' of a string w
+%% according to the grammar $A$:
+%% \begin{enumerate}
+%% \item On the one hand we have the set-theoretic semantics just
+%%   defined, $\llbracket A \rrbracket \cdot w$
+%% \item On the other hand, we can view the string $w = c_1c_2\cdots$ as
+%%   a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
+%%   define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
+%%   \lceil w \rceil \vdash e : G$.
+%% \end{enumerate}
+%% It is not difficult to see that at least for the ``purely positive''
+%% formulae (those featuring only the positive connectives
+%% $0,+,I,\otimes,\mu, \overline\Sigma,c$) that
+%% every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
+%% that the nodes of the tree correspond precisely to the introduction
+%% forms of the type. However it is far less obvious that \emph{every}
+%% linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
+%% sequence of introduction forms since proofs can include elimination
+%% forms as well. To show that this is indeed the case we give a
+%% \emph{canonicity} result for the calculus: that the parses for .
+
+%% \begin{definition}
+%%   A non-linear type $X$ is purely positive if it is built up using
+%%   only finite sums, finite products and least fixed points.
+
+%%   A linear type is purely positive if it is built up using only finite
+%%   sums, tensor products, generators $c$, least fixed points and linear
+%%   sigma types over purely positive non-linear types.
+%% \end{definition}
+
+%% \begin{definition}
+%%   %% Let $X$ be a closed non-linear type. The closed elements $\textrm{Cl}(X)$ of $X$ are the definitional equivalence classes of terms $\cdot \vdash e : X$.
+
+%%   Let $A$ be a closed linear type. The nerve $N(A)$ is a presheaf on
+%%   strings that takes a string $w$ to the definitional equivalence
+%%   classes of terms $\cdot; \lceil w\rceil \vdash e: N(A)$.
+%% \end{definition}
+
+%% \begin{theorem}[Canonicity]
+%%   Let $A$ be a closed, purely positive linear type. Then there is an
+%%   isomorphism between $\llbracket A\rrbracket$ and $N(A)$.
+%% \end{theorem}
+%% \begin{proof}
+%%   We outline the proof here, more details are in the appendix. The
+%%   proof proceeds first by a standard logical families construction
+%%   that combines canonicity arguments for dependent type theory
+%%   TODO cite coquand
+%%   % \cite{coquand,etc}
+%%   with logical relations constructions for linear
+%%   types
+%%   TODO cite hylandschalk
+%%   % \cite{hylandschalk}
+%%   . It is easy to see by induction that the
+%%   logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
+%%   \rrbracket$ and the fundamental lemma proves that the projection
+%%   morphism $p : \hat A \to N(A)$ has a section, the canonicalization
+%%   procedure. Then we establish again by induction that
+%%   canonicalization is also a retraction by observing that introduction
+%%   forms are taken to constructors.
+%% \end{proof}
+
+%% \begin{enumerate}
+%% \item Every term $\lceil w \rceil \vdash p : G + H$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
+%% \item There are no terms $\lceil w \rceil \vdash p : 0$
+%% \item If there is a term $\lceil w \rceil \vdash p : c$ then $w = c$ and $p = x$.
+%% \item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
+%% \item Every term $\lceil w \rceil \vdash p : \varepsilon$ is equal to $()$
+%% \item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
+%% \item Every term $\lceil w \rceil \vdash p : \mu X. G$ is equal to $\textrm{roll}(q)$ where $q : G(\mu X.G/X)$
+%% \item Every term $\lceil w \rceil \vdash p : (x:A) \times G$ is equal
+%%   to $(M,q)$ where $\cdot \vdash M : A$
+%% \end{enumerate}
+
+%% To prove this result we will use a logical families model. We give a
+%% brief overview of this model concretely:
+%% \begin{enumerate}
+%% \item A context $\Gamma$ denotes a family of sets indexed by closing substitutions $\hat\Gamma : (\cdot \vdash \Gamma) \Rightarrow \Set_i$
+%% \item A type $\Gamma \vdash X : U_i$ denotes a family of sets $\hat X : \Pi(\gamma:\cdot \vdash \Gamma) \hat\Gamma \Rightarrow (\cdot \vdash \simulsubst X \gamma) \Rightarrow \Set_i$
+%% \item A term $\Gamma \vdash e : X$ denotes a section $\hat e : \Pi(\gamma)\Pi(\hat\gamma)\hat X \gamma \hat\gamma (\simulsubst e \gamma)$
+%% \item A linear type $\Gamma \vdash A : L_i$ denotes a family of grammars $\hat A : \Pi(\gamma:\cdot\vdash\Gamma)\,\hat\Gamma \Rightarrow \Pi(w:\Sigma^*) (\cdot;\lceil w\rceil \vdash A[\gamma])\Rightarrow \Set_i$, and the denotation of a linear context $\Delta$ is similar.
+%% \item A linear term $\Gamma;\Delta \vdash e : A$ denotes a function \[\hat e : \Pi(\gamma)\Pi(\hat\gamma)\Pi(w)\Pi(\delta : \lceil w \rceil \vdash \simulsubst \Delta \gamma) \hat\Delta \gamma \hat\gamma \delta \Rightarrow \hat A \gamma \hat\gamma w {(\simulsubst {\simulsubst e \gamma} \delta)}\]
+%% \end{enumerate}
+%% And some of the constructions:
+%% \begin{enumerate}
+%% \item $\widehat {(G A)} \gamma \hat\gamma e = \hat A \gamma \hat\gamma \varepsilon (G^{-1}e)$
+%% \item $\widehat {(A \otimes B)} \gamma \hat\gamma w e = \Sigma(w_Aw_B = w)\Sigma(e_A)\Sigma(e_B) (e_A,e_B) = e \wedge \hat A \gamma \hat \gamma w_A e_A \times \hat B \gamma \hat \gamma w_B e_B$
+%% \item $\widehat {(A \lto B)} \gamma \hat\gamma w e = \Pi(w_A)\Pi(e_A) \hat A \gamma\hat\gamma w_A e_A \Rightarrow \hat B \gamma\hat\gamma (ww_A) (\applto {e_A} e)$
+%% \end{enumerate}
+
+%% First, the category with families will be
+%% the category of logical families over set contexts/types
+%% $\Delta$/$A$. Then the propositional portion will be defined by
+%% mapping a logical family $\hat \Gamma \to \Gamma$
+
+%% First, let $L$ be the category of BI formulae and proofs (quotiented
+%% by $\beta\eta$ equality). Define a functor $N : L \to \Set^{\Sigma^*}$ by
+%% \[ N(\phi)(w) = L(w,\phi) \]
+
+%% Then define the gluing category $\mathcal G$ as the comma category
+%% $\Set^{\Sigma^*}/N$. That is, an object of this category is a pair of
+%% a formula $\phi \in L$ and an object $S \in \mathcal
+%% P(\Sigma^*)/N(\phi)$. We can then use the equivalence $\mathcal
+%% P(\Sigma^*)/N(\phi) \cong \mathcal P(\int N(\phi))$ to get a simple
+%% description of such an $S$: it is simply a family of sets indexed by
+%% proofs $L(w,\phi)$:
+%% \[ \prod_{w\in\Sigma^*} L(w,\phi) \to \Set \]
+%% This category clearly comes with a projection functor $\pi : \mathcal
+%% G \to \mathcal L$ and then our goal is to define a section by using
+%% the universal property of $\mathcal L$.
+
+%% To this end we define
+%% \begin{enumerate}
+%% \item $(\phi, S) \otimes (\psi, T) = (\phi \otimes \psi, S\otimes T)$ where
+%%   \[ (S \otimes T)(w, p) = (w_1w_2 = w) \times (q_1,q_2 = p) \times S\,w_1\,q_1 \times T\,w_2\,q_2\]
+%% \item $(\phi, S) \multimap (\psi, T) = (\phi \multimap \psi, S \multimap T)$ where
+%%   \[ (S \multimap T)(w,p) = w' \to q \to S\,w'\,q \to T (ww') (p\,q) \]
+%% \item $\mu X. ??$ ??
+%% \end{enumerate}
+
+%% \pedro{We should conclude this section by explaining the relevance of the canonicity
+%% theorem. Could also be done before stating the theorem.}
+
+\section{Discussion and Future Work}
+\label{sec:discussion}
+\paragraph{Lambek Calculus and Categorial Grammar}
+In 1958 \cite{lambek1958mathematics}, Joachim Lambek introduced his syntactic
+calculus as a logical system for linguistic derivations. In fact, one may view
+the subtheory of our calculus generated over $\otimes, \lto, \tol$ to be a proof
+relevant presentation of Lambek's original system.
+
+\steven{Should discussion of this connection to lambek calculus be spread
+  throughout the paper?}
+
+Lambek later explicitly connects his syntactic calculus to the structure of a
+non-commutative biclosed monoidal category \cite{lambek1988categorial}. There is
+a rich line of subsequent research after these revelations by Lambek, including
+the study of pregroup grammars
+\cite{coeckeMathematicalFoundationsCompositional2010}, the presentation of
+abstract categorial grammars \cite{degrooteAbstractCategorialParsing2015}, and
+even Lambek calculus with dependent types \cite{luoLambekCalculusDependent}. As
+far as the authors can tell, these works seem to be primarily interested in natural
+language semantics in the style of Montague grammars, rather than parsing formal
+languages. That is, they seem to try to give a functorial semantic
+interpretation of natrual language.
+
+There seem to be some implementations from the abstract categorial grammar
+toolkit, but again they seem natural language focused and not up to the task of
+verified formal language parsing.
+
+Also mention here that \cite{luoLambekCalculusDependent} seems similar at first
+glance, but isn't as similar as it might look because the judgements and
+dependent types seem to
+take on a different interpretation of ours. Say something here like, ``as we
+understand it, it differs in x,y,z ways''
+
+\steven{TODO Edit the above}
+\steven{Even though the above sounds bad, are there lambek/categorial things that are missed?}
+
+%% 1. More practical: integrating it into a larger verified development
+%%    - verified imperative implementation: Lambek logic ala separation
+%%      logic?
+%% 2. Semantic actions
+%% 3. Type systems as tree grammars
+
+\paragraph{Typed Approaches to Regular Expressions}
+The usage of a simple type system to reason about regular expressions was
+introduced by Frisch and Cardelli in 2004 \cite{frischCardelli}, and later expanded
+by Hengelin and Nielsen in 2011 \cite{henglein_regular_2011} to handle proofs of regular expression
+containment. In 1992, while investigating the addition of arbitrary recursion to Kleene
+algebra \cite{leiss},
+Lei{\ss} used a least fixed-point operator instead of the Kleene star. Although
+he did not explicitly make use of a type system, or have access to the view of Frisch and
+Cardelli, Lei{\ss} had added the full power of inductive types to a type system
+of regular expressions.
+
+Pratt's work on the residuation of action algebras in \cite{prattActionLogicPure1991} closely mirrors
+our observation that Brzozowski derivatives mirror linear function types.
+Moreover, he shows that his residuals form a Galois connection with sequential
+composition. This is precisely a posetal --- or \emph{thin} --- version of the
+adjunction between
+linear function types and tensor product in \theoryabbv.
+
+Our primary observation in this paper can be summarized as saying that all of
+these above works can embed naturally into a rich, unifying theory.
+
+\steven{I'd like to synthesize the discussion of these works with the categorial
+grammar works, but I am not sure how. It seems as though there are these
+parallel tracks of research that have missed each other for decades}
+
+% In \cite{firsovCertifiedNormalizationContextFree2015}, Firsov and Uustalu
+% constructed a verified normalization procedure to turn a
+% context-free grammar into an equivalent grammar in Chomsky normal form.
+% Combining their work with a CYK parser, they have built a verified context-free
+% grammar parser, \emph{up to weak equivalence}. To upgrade from a weak
+% equivalence to a strong equivalence, they suggest updating their grammar
+% formalism to treat parse trees as first-class proofs of language membership.
+% That is, without explicitly stating so, they suggest that they next natural step
+% is to reason about grammars in a system where parse trees are terms, as we have
+% given in this paper.
+
+% \steven{Make the above paragraph shorter. Also idk if its worth including, but I
+% like that they practically ask for a type theory}
+
+\steven{Mention Neel's typed algebraic approach paper here. Say it is similar at
+  first glance, as its parsing with types, but that he actually is representing
+  something quite different. Or is it so different that its not worth mentioning?}
+
+\steven{TODO add lambek calculus/categorial grammars here}
 
 \paragraph{Kleene Algebra}
-
 Since the early works in the theory of formal languages, Kleene
@@
 
-\paragraph{TODO: more stuff}
-
-%% That paper with Fritz Henglein about regexes as types
-
-%% Vaughan Pratt on the residual operator
-
-%% \section{Universal Constructions in the Category of Grammars}
-
-%% Predicate BI: bicartesian closed, monoidal biclosed, omega-colimits
-%% and therefore certain initial algebras
-
-\section{Non-commutative Linear-Non-Linear Type Theory as a Syntax for Grammars}
-\label{sec:typetheory}
-\subsection{Syntax}
-Judgments: $\Delta$ set ctx, $\Delta \vdash A$ set, $\Delta \vdash M :
-A$ term, $\Delta \vdash \Gamma$ grammar ctx, $\Delta \vdash G$
-grammar, $\Delta\pipe \Gamma \vdash p : G$ parse term.
+\steven{Can we sketch here what an apt categorification of KATs might be? As a
+  poor first pass: perhaps a Kleene category with a full involutive
+  subcategory that corresponds to the Boolean algebra?.}
+\steven{What can we say about internalizing Kleene algebra proofs or extensions?}
+\steven{any KA citations needed above?}
+
+%%    - verified imperative implementation: Lambek logic ala separation
+%%      logic?
+\paragraph{Relation to Separation Logic}
+
+\theoryabbv is similar in spirit to separation logic. Semantically,
+they are closely related: linear types in \theoryabbv denote families
+of sets indexed by a monoid of strings, whereas separation logic
+formulae typically denote families of predicates indexed by an ordered
+partial commutative monoid of worlds. The monoidal structure in both
+cases is are instances of the category-theoretic notion of Day
+convolution monoidal structure. From a separation-logic perspective,
+our notion of memory is very primitive: a memory shape is just a
+string of characters and the state of the memory is never allowed to
+evolve.
+
+Syntactically, our type theory can be viewed as a kind of ``separation
+type theory'' in that we care not just about provability of formulae
+but the constructive content of our linear terms, as these act as
+parsers. The design of \theoryabbv is based on the $\textrm{LNL}_D$
+calculus which explicitly was designed to be a dependent
+type-theoretic version of separation logic. Many concepts of
+separation logic then have direct analogues in our system. The
+characters $c$ are analogous to the ``points-to'' formulae $l \mapsto
+x$ in separation logic. Our $\otimes,\lto,\tol$ are analogous to the
+separating conjunction and magic-wand. The analogue of the persistence
+modality $\square$ of separation logic is $\bigoplus_{\_:\uparrow A} I$,
+which ``zeros out'' all but the parses from the empty string. If we
+think of the portions of the string we are parsing as our notion of
+resource, then these make sense as the ``resource-free'' elements of
+the type $A$.
+
+This semantic connection to separation logic also suggests an avenue
+of future work: rather than working with a dependent type theory, we
+could instead make a program logic based on non-commutative separation
+logic for verifying imperative implementations of parsers. This could
+be accomplished by modifying an existing separation logic
+implementation or embedding the logic within \theoryabbv.
+
+%% \paragraph{Linearity with Dependent Types}
+%% \steven{I don't know if it worth discussing the different influences of
+%%   categorical model here? We take Neel's model and add to it as necessary}
+
+% Vakar might be worth citing to discuss different characterizations of the
+% notion of model \cite{vakarSyntaxSemanticsLinear2015}
+
+% \cite{fu2023twolevellineardependenttype} give a compiler for a two-level
+% linear dependent type theory. This could serve as a template for a
+% compiler implementation of \theoryname that is not just inference rules
+% embedded in Agda.
+
+\paragraph{Implementations in Parsing}
+\steven{Talk about future implementation-side work}
+
+\cite{fu2023twolevellineardependenttype} give a compiler for a two-level
+linear dependent type theory. This could serve as a template for a
+compiler implementation of \theoryname that is not just inference rules
+embedded in Agda.
+
+\paragraph{Type Checking and Elaboration}
+Our focus in this work has been on the verification of parsers for
+grammars over strings, but because \theoryabbv allows for the
+definition of arbitrarily powerful grammars, the system could also be
+used for more sophisticated semantic analysis such as scope checking
+or type checking. Alternatively, we could more directly encode type
+type systems as linear types in a modified version of \theoryabbv
+whose semantics is not in families of sets indexed by \emph{strings}
+but families of sets indexed by \emph{abstract syntax trees}.
+
+%% While parsing typically refers to the generation of semantic
+%% objects from string input, many tasks in programming can be
+%% viewed as parsing of objects with more structure, such as
+%% trees with binding structure or graphs. Fundamental to the
+%% frontend of many
+%% programming language implementations are type systems. In
+%% particular, \emph{type checking}
+%% --- analogous to language recognition --- and \emph{typed
+%%   elaboration} --- analogous to parsing --- arise when
+%% producing a semantic object subject to some analysis. Just
+%% as our string grammars were given as functors from $\String$
+%% to $\Set$, we envision adapting the same philosophy
+%% to functors from a \emph{category of trees} to $\Set$ to craft a syntax
+%% that natively captures typed elaboration. This suggests an
+%% unusual sort of bunched type theory, where context extension
+%% no longer resembles concatenation of strings but instead
+%% takes on the form of tree constructors.
+
+% type system              ~~ Formal Grammar
+% typing derivation        ~~ parse tree
+% algorithmic type system  ~~ LL(1), LR(1) grammar
+% Uniqueness of derivations ~~ unambiguous grammar
+% type elaborator           ~~ semantic actions
+
+\bibliographystyle{plain}
+\bibliography{refs.bib}
+
+\appendix
+
+\section{Elided Syntax}
+
+\max{TODO: add prose}
 
 \begin{figure}
-  \label{fig:structjdg}
   \begin{mathpar}
@@
     \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}
-
-    \\
+    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}
+
+    \boxed{\linctxwffjdg \Gamma A}
 
@@
     \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}
-
-    \\
-
-    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}
-
-    \and
-
-    \inferrule{\Gamma \vdash A : L_i}{\linctxwffjdg \Gamma A}
-
-    \\
-
-    \inferrule{\Gamma \vdash X \equiv Y : U_i}{\ctxwffjdg \Gamma {X\equiv Y}}
-
-    \and
-
-    \inferrule{\Gamma \vdash A \equiv B : L_i}{\linctxwffjdg \Gamma {A \equiv B}}
-
   \end{mathpar}
-  \caption{Structural judgements}
+  \caption{Context Rules}
 \end{figure}
 
-\begin{figure}
-  \label{fig:typewf}
-  \begin{mathpar}
-    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
- %   
-    \and
-%
-    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
-%
-    \\
-%
-    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
-%
-    \and
-%
-    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
-%
-    \\
-%
-    \inferrule{~}{\Gamma \vdash 1 : U_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
-%
-    \\
-%
-    \inferrule{~}{\Gamma \vdash I : L_i}
- %   
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
-%
-    \\
-%
-    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
-%
-    \\
-%
-    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
-    %
-    \and
-    %
-    \inferrule{~}{\Gamma \vdash \top : L_i}
-%
-    \and
-%
-    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
-%
-    \\
-    %
-    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
-    %
-    \and
-    %
-    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
-  \end{mathpar}
-  \caption{Type well-formedness}
-\end{figure}
+\max{TODO: add rules for non-linear types $\Pi,\Sigma$, equality etc}
 
@@
   \begin{mathpar}
-    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
-%
-    \and
-%
-    \inferrule{~}{}
-%
-    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
-%   
-    \and
-%
-    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
-    %
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
-%
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
-%
-    \and
-%
-    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C} 
-    %
-    \and
-    %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
-\end{mathpar}
-  \caption{Judgemental equality}
-\end{figure}
-
-\begin{figure}
-  \label{fig:inttyping}
-  \begin{mathpar}
+  \boxed{\Gamma \vdash M : X}
+
   \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
@@
   %
-  \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
-  %
-  \\\
+  \inferrule{\Gamma \vdash M : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash M : X}
+  %
+  \and
   %
@@
   %
-  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
-  %
-  \\
-%
-  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
-  \and
-  \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
-  %
-  \\
-  %
-  \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
-  \and
-  \inferrule{\Gamma \mid \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
+  \inferrule{\Gamma \vdash M : X \\ \Gamma \vdash N : \subst Y M x}{\Gamma \vdash (M, N) : \SigTy x X Y}
+  %
+  \and
+%
+  \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_1\, M : X}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_2\, M : \subst Y {\pi_1\, M} x}
+  \and
+  \inferrule{\Gamma, x : X \vdash M : Y}{\Gamma \vdash \lamb x M : \PiTy x X Y}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M : \PiTy x X Y \\ \Gamma \vdash N : X}{\Gamma \vdash \app M {N} : \subst Y {N} x}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M \equiv N : X}{\Gamma \vdash \mathsf{refl} : M =_X N}
+  \and
+  \textrm{TODO: inductive types (boolean, nat) with their eliminators allowed in }
   \end{mathpar}
   \caption{Intuitionistic typing}
+  \label{fig:inttyping}
 \end{figure}
@@
 \begin{figure}
-  \label{fig:linsyntax}
   \begin{mathpar}
-    \inferrule{~}{\Gamma \mid a : A \vdash a : A}
-    \and
-    \inferrule{\Gamma \mid \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma \mid \Delta \vdash e : A}
+    %
+    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
     %
     \\
-    %
-    \inferrule{~}{\Gamma \mid \cdot \vdash () : I}
-    \and
-    \inferrule{\Gamma \mid \Delta \vdash e : I \\ \Gamma \mid \Delta_1',\Delta_2' \vdash e' : C}{\Gamma \mid \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
-    %
+
+  \end{mathpar}
+  \caption{Linear Terms, extended}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \boxed{\Gamma \vdash M \equiv N : X}
+
+    \inferrule{\Gamma \vdash p : M =_X N}{\Gamma \vdash M \equiv N : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \app {(\lamb x M)} {N} \equiv \subst x M {N} : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash m \equiv \lamb x {\app M x} : \PiTy x X Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_1\, (M, N) \equiv M : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_2\, (M , N) \equiv N : \subst x {M} Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash M \equiv (\pi_1\, M, \pi_2\, M) : \SigTy x X Y}
+%
+    \and
+%
+    % \inferrule{~}{}
+%
+    \inferrule{~}{\Gamma \vdash M \equiv N : 1}
+%
+    % \and
+%
+    % \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
+
     \\
-    %
-    \inferrule{\Gamma \mid \Delta \vdash e : A \\ \Gamma \mid \Delta' \vdash e' : B}{\Gamma \mid \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
-    %
-    \\
-    %
-    \inferrule{\Gamma \mid \Delta \vdash e : A \otimes B \\ \Gamma \mid \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma \mid  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
-    \\
-    %
-    \inferrule{\Gamma \mid \Delta, a : A \vdash e : B}{\Gamma \mid \Delta \vdash \lamblto a e : A\lto B}
-    \and
-    \inferrule{\Gamma \mid \Delta' \vdash e' : A \\ \Gamma \mid \Delta \vdash e : A \lto B}{\Gamma \mid \Delta', \Delta \vdash \applto {e'} {e} : B}
-    \\
-    %
-    \inferrule{\Gamma \mid a : A, \Delta \vdash e : B}{\Gamma \mid \Delta \vdash \lambtol a e : B\tol A}
-    \and
-    \inferrule{\Gamma \mid \Delta \vdash e : B \tol A \\ \Gamma \mid \Delta' \vdash e' : A}{\Gamma \mid \Delta, \Delta' \vdash \apptol e {e'} : B}
-    %
-    \\
-    %
-    \inferrule{\Gamma, x : X \mid \Delta  \vdash e : A}{\Gamma \mid \Delta \vdash \dlamb x e : \LinPiTy x X A}
-    \and
-    \inferrule{\Gamma \mid \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma \mid \Delta \vdash \app e {e'} : \subst A {e'} x}
-    %
-    \\
-    %
-    \inferrule{\Gamma \vdash e : X \quad \Gamma \mid \Delta \vdash e' : \subst A e x}{\Gamma \mid \Delta \vdash (e, e') : \LinSigTy x X A}
-    %
-    \\
-    %
-    \inferrule{\Gamma \mid \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X \mid \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma\mid \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
-    %
-    \\
-    %
-    \inferrule{~}{\Gamma \mid \Delta \vdash () : \top}
-    %
-    \\
-    %
-    \inferrule{\Gamma \mid \Delta \vdash e_1 : A_1 \quad \Gamma \mid \Delta \vdash e_2 : A_2}{\Gamma \mid \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
-    \and
-    \inferrule{\Gamma \mid \Delta \vdash e : A_1 \amp A_2 }{\Gamma \mid \Delta \vdash \pi_i \, e : A_i}
-    %
-    \\
-    %
-    \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma \mid \cdot \vdash \mathsf{G}^{-1}\, e : A}
-    %
-    \\
-    %
-    \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
-    \and
-    \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
+
+    \boxed{\Gamma ; \Delta \vdash e \equiv e' : A}
+
+    % \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
+%
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e {e'} a : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : B \tol A}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {M} \equiv \subst {e} M x : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
+%
+    \and
+%
+%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
+% %
+%     \and
+%
+%     \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
+% %
+%     \and
+% %
+%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
+    %
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} e a : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} {e \otimes e'} e'' \equiv e'' \{ e/a , e'/b \} : C}
+%
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} e c : C}
+%
+    \and
+%
+    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(M, e)} {e'} \equiv e' \{ M/x , e/a \} : C}
+    %
+    \and
+    %
+    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
+    \and
+    %
+    \inferrule
+    {~}
+    {\Gamma ; \Delta \vdash \equalizerpi {\equalizerin {e}} \equiv e : A}
+    %
+    \and
+    %
+    \inferrule
+    {~}
+    {\Gamma ; \Delta \vdash \equalizerin {\equalizerpi {e}} \equiv e : A}
 \end{mathpar}
-  \caption{Linear typing}
+  \caption{Judgmental equality for linear terms}
+  \label{fig:jdgeq}
 \end{figure}
-
-\pedro{Need to figure out the rules for inductive types and write down
-  the judgemental equality rules}
-
-\subsection{Semantics in Grammars}
-
-For illustrative purposes, we give a brief overview of the concrete semantics in sets and grammars:
-\begin{enumerate}
-\item A non-linear context $\Gamma$ denotes a set $\sem{\Gamma}$
-\item A non-linear type $\Gamma \vdash X : U_i$ denotes a family of sets $\sem{X} : \sem{\Gamma} \to \Set_i$
-\item A non-linear term $\Gamma \vdash e : X$ denotes a section $\sem{e} : \Pi(\gamma:\sem{\Gamma})\,\sem{X}\,\gamma$
-\item Linear contexts $\Gamma \vdash \Delta$ and types $\Gamma \vdash A : L_i$ both denote families of grammars $\sem{\Gamma}\to \Gr_i$
-\item A linear term $\Gamma; \Delta \vdash e : A$ denotes a family of parse transformers $\sem{e} : \Pi(\gamma:\sem{\Gamma})\Pi(w:\Sigma^*)\,\sem{\Delta}\gamma w \Rightarrow \sem{A}\gamma w)$
-\end{enumerate}
-Most of the non-linear semantics is standard, we show just the $G$ type and the linear semantics:
-\begin{enumerate}
-\item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
-\item $\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-\item $\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-\item $\sem{I} \gamma w = \{ () \pipe w = \epsilon \}$
-\item $\sem{A \otimes B} \gamma w = \Sigma(w1,w2:\Sigma^*) w1w1 = w \wedge \sem{A} \gamma w_1 \times \sem{B} \gamma w_2$
-\item $\sem{A \lto B} \gamma w = \Pi(w_a:\Sigma^*) A \gamma w_a \Rightarrow B\gamma (w_aw)$
-\item $\sem{B \tol A} \gamma w = \Pi(w_a:\Sigma^*) A \gamma w_a \Rightarrow B\gamma (ww_a)$
-\item $\sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x)$
-\end{enumerate}
-
-Propositional fragment: monoidal biclosed category that is also
-bicartesian closed and has $\omega$-colimits. Call such a BIC.
-
-Set theoretic fragment: category with families $\mathcal C$ + $\Pi,\Sigma,W$.
-
-Combination: category with families $\mathcal C$ + a functor $L : C
-\to \textrm{BIC}$.
-
-Grammatical fragment: functor from $\mathcal C$ to the category of BI-models
-
-Example: use $\Set$ as the CwF, use the functor $\Delta \mapsto
-(\Set^{\Sigma^*})^\Delta$ as the model of the propositional fragment
-using the pointwise-Day convolution monoidal structure.
-
-\subsection{Canonicity}
-
-Let $\cdot \vdash A$ be a closed linear type. Then there are
-two obvious notions of what constitutes a ``parse'' of a string w
-according to the grammar $A$:
-\begin{enumerate}
-\item On the one hand we have the set-theoretic semantics just
-  defined, $\llbracket A \rrbracket \cdot w$
-\item On the other hand, we can view the string $w = c_1c_2\cdots$ as
-  a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
-  define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
-  \lceil w \rceil \vdash e : G$.
-\end{enumerate}
-It is not difficult to see that at least for the ``purely positive''
-formulae (those featuring only the positive connectives
-$0,+,I,\otimes,\mu, \overline\Sigma,c$) that
-every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
-that the nodes of the tree correspond precisely to the introduction
-forms of the type. However it is far less obvious that \emph{every}
-linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
-sequence of introduction forms since proofs can include elimination
-forms as well. To show that this is indeed the case we give a
-\emph{canonicity} result for the calculus: that the parses for .
-
-\begin{definition}
-  A non-linear type $X$ is purely positive if it is built up using
-  only finite sums, finite products and least fixed points.
-
-  A linear type is purely positive if it is built up using only finite
-  sums, tensor products, generators $c$, least fixed points and linear
-  sigma types over purely positive non-linear types.
-\end{definition}
-
-\begin{definition}
-  %% Let $X$ be a closed non-linear type. The closed elements $\textrm{Cl}(X)$ of $X$ are the definitional equivalence classes of terms $\cdot \vdash e : X$.
-
-  Let $A$ be a closed linear type. The nerve $N(A)$ is a presheaf on
-  strings that takes a string $w$ to the definitional equivalence
-  classes of terms $\cdot; \lceil w\rceil \vdash e: N(A)$.
-\end{definition}
-
-\begin{theorem}[Canonicity]
-  Let $A$ be a closed, purely positive linear type. Then there is an
-  isomorphism between $\llbracket A\rrbracket$ and $N(A)$.
-\end{theorem}
-\begin{proof}
-  We outline the proof here, more details are in the appendix. The
-  proof proceeds first by a standard logical families construction
-  that combines canonicity arguments for dependent type theory
-  \cite{coquand,etc} with logical relations constructions for linear
-  types \cite{hylandschalk}. It is easy to see by induction that the
-  logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
-  \rrbracket$ and the fundamental lemma proves that the projection
-  morphism $p : \hat A \to N(A)$ has a section, the canonicalization
-  procedure. Then we establish again by induction that
-  canonicalization is also a retraction by observing that introduction
-  forms are taken to constructors.
-\end{proof}
-
-\begin{enumerate}
-\item Every term $\lceil w \rceil \vdash p : G + H$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
-\item There are no terms $\lceil w \rceil \vdash p : 0$
-\item If there is a term $\lceil w \rceil \vdash p : c$ then $w = c$ and $p = x$.
-\item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
-\item Every term $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
-\item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
-\item Every term $\lceil w \rceil \vdash p : \mu X. G$ is equal to $\textrm{roll}(q)$ where $q : G(\mu X.G/X)$
-\item Every term $\lceil w \rceil \vdash p : (x:A) \times G$ is equal
-  to $(M,q)$ where $\cdot \vdash M : A$
-\end{enumerate}
-
-To prove this result we will use a logical families model. We give a
-brief overview of this model concretely:
-\begin{enumerate}
-\item A context $\Gamma$ denotes a family of sets indexed by closing substitutions $\hat\Gamma : (\cdot \vdash \Gamma) \Rightarrow \Set_i$
-\item A type $\Gamma \vdash X : U_i$ denotes a family of sets $\hat X : \Pi(\gamma:\cdot \vdash \Gamma) \hat\Gamma \Rightarrow (\cdot \vdash \simulsubst X \gamma) \Rightarrow \Set_i$
-\item A term $\Gamma \vdash e : X$ denotes a section $\hat e : \Pi(\gamma)\Pi(\hat\gamma)\hat X \gamma \hat\gamma (\simulsubst e \gamma)$
-\item A linear type $\Gamma \vdash A : L_i$ denotes a family of grammars $\hat A : \Pi(\gamma:\cdot\vdash\Gamma)\,\hat\Gamma \Rightarrow \Pi(w:\Sigma^*) (\cdot;\lceil w\rceil \vdash A[\gamma])\Rightarrow \Set_i$, and the denotation of a linear context $\Delta$ is similar.
-\item A linear term $\Gamma;\Delta \vdash e : A$ denotes a function \[\hat e : \Pi(\gamma)\Pi(\hat\gamma)\Pi(w)\Pi(\delta : \lceil w \rceil \vdash \simulsubst \Delta \gamma) \hat\Delta \gamma \hat\gamma \delta \Rightarrow \hat A \gamma \hat\gamma w {(\simulsubst {\simulsubst e \gamma} \delta)}\]
-\end{enumerate}
-And some of the constructions:
-\begin{enumerate}
-\item $\widehat {(G A)} \gamma \hat\gamma e = \hat A \gamma \hat\gamma \varepsilon (G^{-1}e)$
-\item $\widehat {(A \otimes B)} \gamma \hat\gamma w e = \Sigma(w_Aw_B = w)\Sigma(e_A)\Sigma(e_B) (e_A,e_B) = e \wedge \hat A \gamma \hat \gamma w_A e_A \times \hat B \gamma \hat \gamma w_B e_B$
-\item $\widehat {(A \lto B)} \gamma \hat\gamma w e = \Pi(w_A)\Pi(e_A) \hat A \gamma\hat\gamma w_A e_A \Rightarrow \hat B \gamma\hat\gamma (ww_A) (\applto {e_A} e)$
-\end{enumerate}
-
-First, the category with families will be
-the category of logical families over set contexts/types
-$\Delta$/$A$. Then the propositional portion will be defined by
-mapping a logical family $\hat \Gamma \to \Gamma$
-
-First, let $L$ be the category of BI formulae and proofs (quotiented
-by $\beta\eta$ equality). Define a functor $N : L \to \Set^{\Sigma^*}$ by
-\[ N(\phi)(w) = L(w,\phi) \]
-
-Then define the gluing category $\mathcal G$ as the comma category
-$\Set^{\Sigma^*}/N$. That is, an object of this category is a pair of
-a formula $\phi \in L$ and an object $S \in \mathcal
-P(\Sigma^*)/N(\phi)$. We can then use the equivalence $\mathcal
-P(\Sigma^*)/N(\phi) \cong \mathcal P(\int N(\phi))$ to get a simple
-description of such an $S$: it is simply a family of sets indexed by
-proofs $L(w,\phi)$:
-\[ \prod_{w\in\Sigma^*} L(w,\phi) \to \Set \]
-This category clearly comes with a projection functor $\pi : \mathcal
-G \to \mathcal L$ and then our goal is to define a section by using
-the universal property of $\mathcal L$.
-
-To this end we define
-\begin{enumerate}
-\item $(\phi, S) \otimes (\psi, T) = (\phi \otimes \psi, S\otimes T)$ where
-  \[ (S \otimes T)(w, p) = (w_1w_2 = w) \times (q_1,q_2 = p) \times S\,w_1\,q_1 \times T\,w_2\,q_2\]
-\item $(\phi, S) \multimap (\psi, T) = (\phi \multimap \psi, S \multimap T)$ where
-  \[ (S \multimap T)(w,p) = w' \to q \to S\,w'\,q \to T (ww') (p\,q) \]
-\item $\mu X. ??$ ??
-\end{enumerate}
-
-\section{Type Theoretic Formal Grammar Theory}
-
-Now that we have developed a type theoretic syntax for formal grammars
-and parse transformers, we show how many common concepts of formal
-language theory and grammars are naturally expressed using type
-theoretic notions, and some constructions and proofs can be carried
-out in our type theory as well.
-
-Firstly, regular expressions and their extension to context-free
-expressions are naturally expressed as a limited syntax of linear
-types. That is, regular expressions are precisely the linear types
-constructed using only $c, I, \otimes, 0, \oplus$ and the Kleene star,
-which can be defined as $A^* = \mu x. I \oplus A \otimes X$. The
-context-free expressions are similarly expressible by allowing not
-just the Kleene star, but unrestricted usage of $\mu$. Remember,
-however, that these are a notion of regular expressions that maintain
-a notion of ambiguous grammar. That is, $A \oplus A$ is not considered
-to be the same as $A$. While there does not appear to be an analogous
-weakening of this definition to naturally describe the
-context-sensitive languages, we will see shortly that the indexing of
-linear types by non-linear types allows for capturing a notion of
-context in grammars.
-
-In addition to expressing algebraic presentations of grammars, the
-type-theoretic syntax allows for naturally expressing grammars that
-capture the behavior of automata. \max{sync this up with Steven's
-  stuff, and add stuff about PDAs and Turing Machines}.
-
-A key construct in formal language theory is the Brzozowski derivative
-\cite{brzozowski}. Given a string $w$ the (left) derivative of a
-language $L$ with respect to $w$, written $\partial_w L$ is the language
-\[ (\partial_wL ) = \{ ww' \pipe w' \in L \}\]
-This semantics is given using the linear implication $w \tol
-L$. Expressing Brzozowski's calculation of the derivative of a regular
-expression relies on additional axioms to the type theory.\max{axioms
-  to carry out Brzozowski's derivative calculations plz.}
-
-\begin{align*}
-  c\lto c &\equiv I\\
-  c\lto d &\equiv 0\\
-  c\lto I &\equiv 0\\
-  c\lto 0 &\equiv 0\\
-  c\lto (A \otimes B) &\equiv (c\lto A) \otimes B + \nu(A) \otimes (c\lto B)\\
-  c\lto A^* &\equiv (c\lto A) \otimes A^*\\
-  c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
-\end{align*}
-
-\pedro{The type family $\nu$ above is equal to $I$ if its argument
-  parses the empty string and zero otherwise. In order to define
-  it we should probably add an extra axiom, right?}
-\steven{$\nu$ can be derived using $\amp$, $\nu(A) = I \amp A$}
-
-\steven{In addition to Pedro's axioms:
-
-$c, c'$ characters. $g$ any grammar.
-
-\begin{gather}
-  I \lto c \cong c \\
-  c \lto (c' \lto g) \cong (c' \otimes c) \lto g
-\end{gather}
-
-For axiom $(1)$, Brzozowski has $I \lto g \cong g$ for all grammars $g$, but I don't think that holds for us because we want to preserve ambiguity.
-We could instead have something like $I \lto g \cong (g \amp I)^{*} \otimes g$.
-For our purposes the character presentation seems sufficient, as the derivative reasoning Brzozowski uses only takes derivatives with respect to strings.
-
-Axiom $(2)$ might be admissible, but I'm unsure.
-
-I think for Kleene star we need $c \lto g^{*} \cong (g \amp I)^{*} \otimes (c \lto g) \otimes g^{*}$.
-This should derivable utilizing the product axiom and $g^{*} \cong I \oplus (g \otimes g^{*})$, or is that sort of isomorphism opaque to us right now.
-
-To make the DFA proof go through I think we additionally need to axiomatically make the map $c \otimes (c \lto A) \vdash (c \otimes \top) \amp A$ to be an iso.
-Specifically, this term needs an iso $\phi : c \otimes (c \lto A) \to (c \otimes \top) \amp A$ defined below,
-
-\[
-p : c \otimes (c \lto A) \vdash \phi(p) = \letin {p_{1} \otimes p_{2}} p {(p_{1} \otimes \top, p_{1}^{\lto}p_{2})} : (c \otimes \top) \amp A
-\]
-
-but this last bit isn't just Brzozowski's definitions
-
-}
-Using the $G$ modality, we can internalize some of the external
-reasoning about grammars into the language of non-linear types. Recall
-that the denotation of $G A$ is the set of parses of the empty
-string. As is well known in formal language theory, all the parsing
-behavior of a grammar can be reduced to reasoning about parsing of the
-empty string by using the derivative of a grammar. For instance, the
-set of parses for a fixed string $w$ can be defined as $G(w \tol A)$
-and so the type of all parses of a grammar can be defined as
-$\Sigma(w:\textrm{String}) G(w \tol A)$. This can be used to define
-internally a notion of \emph{ambiguity} of a grammar. A grammar $A$ is
-ambiguous \steven{This is an unambiguous grammar, right?} if for every string $w$, the type $G(w \tol A)$ of
-$w-$parses is a \emph{proposition} in the sense that all terms in $G(w
-\tol A)$ are equal.\max{I think we should add a propositional
-  truncation to the type theory.}  Similarly, the type of \emph{total
-parsers} can be defined as $G(\top \tol A)$. A total parser encodes a
-function that selects a parse for every possible string. A
-\emph{semi-parser} is then defined as $G(\top \tol A \oplus \top)$,
-where the parser is allowed to fail for any reason. This has the
-downside that the always failing semi-parser $\lambda _. \textrm{in}_r
-()$ is valid for any grammar. If we were working with a full
-\emph{bunched} type theory, we would be able to express the more
-precise type of a \emph{deciding} parser as $G(\top \tol A \oplus \neg
-A)$ where $\neg A = A \Rightarrow 0$ uses the intuitionistic
-implication.
-
-Classically, formal language theory is closely related to automata theory. In the styles of Chomsky and
-Sch\"utzenberger, there is a hierarchy of language classes and the types of automata that recognize them. Inside of the
-type theory of \cref{sec:typetheory}, we can characterize these language classes; moreover, we demonstrate equivalence
-of each class to the associated automaton as an isomorphism within our logic.
-
-Note, a predicate logic over LNL with existential quantifiers is already enough to define
-formulae that correspond to arbitrary recursively enumerable languages ---
-since we can simulate the tape of a Turing machine using two
-stacks. I.e., we can define a set $S$ of lists of tape symbols, and
-then define a formula that first ``copies'' the input onto one stack, then executes
-
-\[ \mu \textrm{Copy} : S \to *. \lambda s. (\epsilon \wedge T\,s\,[])\bigvee_{c \in \Sigma} c (S\,{c::s}) \]
-
-We reflect on this expressive power to motivate why we put heavy syntactic restrictions on our presentation of certain
-classes of automata. For instance, when defining a pushdown automata, we could use the intuitionistic context to
-represent the stack, but we ought to be very careful about how this set is accessed and added to. Careless definitions
-may accidentally endow us with too much expressive power, launching us further up in the Chomsky hierarchy. To begin,
-let us focus only on language classes, and their associated automata, whose parses are definable without any mention to
-the inuitionistic context. When viewed as a machine, this refers to the class of automata definable without access to
-any memory --- i.e.\ finite state machines.
-
-\subsection{Finite State Machines}
-A \emph{nondeterministic finite automaton} (NFA) is a finite state machine where transitions are labeled with characters from a fixed alphabet $\Sigma$.
-These are often represented formally as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$,
-
-\begin{itemize}
-  \item $Q$ a finite set of states
-  \item $\Sigma$ a fixed, finite alphabet
-  \item $\delta : Q \times \Sigma \to \mathcal{P}(Q)$ a labelled transition function
-  \item $q_{0}$ a start state
-  \item $F$ a set of accepting states
-\end{itemize}
-
-Intuitively, this can be thought of like a directed graph with nodes in $Q$ with an edge $q \overset{c}{\to} q'$ whenever $q' \in \delta(q, c)$.
-Note that transitions in an NFA may be labelled with the empty string $\epsilon$ --- such transitions are referred to as \emph{$\epsilon$-transitions}.
-
-As a gramamar in our type theory, an NFA may be encoded as a
-mutual\footnote{For convenience we refer to a mutual fixed point here, even though we
-  only take single fixed points as primitive. As in Beki\'c's theorem, we may decompose our mutual fixed points into a series of single fixed points. So this operation is admissible.}
-fixed point in the following manner:
-
-An NFA has finitely many states, so we may number them $Q = \{q_{1}, \dots, q_{N}\}$. For each $q_{j}$ we introduce
-a mutually recursive term $Y_{j}$ to represent the states of the NFA within this grammar presentation.
-We represent the transition function $\delta$ in the right hand side of each $Y_{j}$ definition.
-For character $c \in \Sigma \cup \{\epsilon\}$ and state $q_{j} \in Q$, we add a summand --- as a coproduct\footnote{There is an admissible notion of coproduct making use of the primitive dependent pairing --- i.e.
-$A \oplus B := \LinPiTy b \Bool {\textbf{if}~b~A~B}$.
-We will utilize this without much comment to represent the usual notion of alternation of grammars.} --- to the defintion of $Y_{j}$ for every state in the image of the transition function,
-
-\[
-Y_{j} = \bigoplus_{c \in \Sigma \cup \{\epsilon\}} \bigoplus_{q_{i} \in \delta(q_{j}, c)} c \otimes Y_{i}
-\]
-
-For convenience, refer to syntactic term on the right hand side of $Y_{j}$ as $\hat{Y}_{j}$.
-Further, we rename the start state to be represented by the term $Y_{s}$.
-Additionally, and without loss of generality, we assume that there is a single accepting state $Y_{e}$ in our encoding.
-This assupmtion is justified through the use of $\epsilon$-transitions.
-For each $q_{j} \in F$, we add an $\epsilon$-transition from $Y_{j}$ to $Y_{e}$.
-
-The above translation totals to give us a syntactic form for NFAs as grammars,
-
-\[
-  NFA ::= \mu \begin{pmatrix}
-        Y_{s} = \langle \hat{Y} \rangle \\
-        Y_{j} = \langle \hat{Y} \rangle, j \not \in \{s, e\} \\
-        Y_{e} = I
-        \end{pmatrix}.Y_{s}
-\]
-where
-\[
-  \hat{Y} ::= c \otimes Y ~|~ Y ~|~ \hat{Y} \oplus \hat{Y}
-\]
-
-\steven{Somewhere mention that $\epsilon$ corresponds to $I$, characters $c \in \Sigma$ correspond to grammars $c$, etc}
-
-and each $Y \in \{ Y_{s}, Y_{e}, Y_{1}, \dots, Y_{N} \}$. Intuitively, we read each $Y_{j}$ as the label for a state, and the
-definitions of form $\hat{Y}$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$
-denotes an $\epsilon$-transition. A
-disjunction of transitions in an NFA grammar corresponds to multiple outgoing edges of a given state in an automaton.
-
-Above we give a translation from the tuple presentation of an NFA to a fixed point grammar in our type theory.
-We don't provide any argument that this translation is done correctly; however, we hope that the construction is straightforward and lightweight enough to demonstrate our formal system as a natural and effective syntax for reasoning about both grammars and automata.
-
-With this setting for finite automata, we can now internalize some classical theorems inside of our formal system.
-
-\subsection{Equivalence Between Regular Grammars and Finite Automata}
-
-% The class of \emph{regular} grammars is given as follows:
-
-% \[
-% Reg ::= 0 ~|~ I ~|~ c ~|~ Reg \oplus Reg ~|~ Reg \otimes Reg ~|~ Reg^{*}
-% \]
-% where $c \in \mathcal{A}$ is a character literal drawn from a fixed alphabet, and the Kleene star $\cdot^{*}$ is derived as an admissible operation using the fixed point operator --- $g^{*} = \mu X.I \oplus (g \otimes X)$.
-\steven{This definition is maybe fair to assume.
-Also maybe better suited for a section earlier in the paper}
-
-First, let us describe a version of Thompson's construction \cite{?} where we construct an NFA that recognizes a given regular expression.
-Moreover, we will show that this NFA is strongly equivalent to the original grammar.
-
-This construction
-
-\begin{theorem}[Thompson]
-  \label{thm:thompson}
-  For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same language as $g$. Further, $N$ and $g$ are isomorphic.
-\end{theorem}
-
-\begin{proof}
-  We construct the recognizing NFA $N$ inductively. To handle the atomic grammars --- literals and the empty grammar --- we define the following,
-
-\[
-  NFA(c) =
-  \mu \begin{pmatrix}
-        Y_{s} = c \otimes Y_{e} \\
-        Y_{e} = I
-  \end{pmatrix}.Y_{s}
-\]
-\[
-  NFA(I) =
-  \mu \begin{pmatrix}
-        Y_{s} = Y_{e} \\
-        Y_{e} = I
-  \end{pmatrix}.Y_{s}
-\]
-
-The isomorphism between the grammars $c$ and $NFA(c)$ are witnessed by the following derivations,
-\[
-  \inferrule{p : c \vdash p : c \\ \cdot \vdash () : I}{p : c \vdash p \otimes () : Y_{s} = c \otimes I}
-\]
-and
-\[
-  \inferrule{q : Y_{s} = c \otimes I \vdash q : Y_{s} \\ p : c, e : I \vdash p : c }{q : Y_{s} \vdash \letin {p \otimes e} q p : c}
-\]
-
-Similar terms are used to show the equivalence between $I$ and $NFA(I)$.
-
-We can now inductively define some NFA combinators and show that they preserve grammar equivalence. First begin with
-the disjunction of two NFAs. For regular grammars $g$ and $g'$, assume that $g \cong NFA(g)$ and $g' \cong NFA(g')$, and we will define an NFA that
-recognizes $g \oplus g'$.
-
-Let the states in $NFA(g)$ be represented by $X$'s and the states in $NFA(g')$ be represented
-by $Z$'s. Further let $\hat{X}_{k}$ refer to the syntactic term on the right hand side of the definition of $X_{k}$
-inside of $NFA(g)$. Likewise define each $\hat{Z}_{k}$.
-
-\[
-  NFA(g \oplus g') =
-  \mu \begin{pmatrix}
-        Y_{s} =  X_{s}' \oplus Z_{s}' \\
-        X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
-        X'_{e} = Y_{e} \\
-        Z_{j}' = \simulsubst {\hat{Z}_{j}} {\delta_{Z}}, j \neq e \\
-        Z_{e}' = Y_{e} \\
-        Y_{e} = I
-  \end{pmatrix}.Y_{s}
-\]
-where the substitution $\{X_{s}'/X_{s}, X_{e}'/X_{e}, X_{1}'/X_{1}, \dots \}$ is given by $\delta_{X}$ --- and $\delta_{Z}$ constructed similarly.
-
-In the construction of $NFA(g \oplus g')$ we replicate each defintion $X_{j}$ in $NFA(g)$ with an analogous definition
-$X_{j}'$ inside of $NFA(g \oplus g')$, except we replace $X_{k}$ with $X_{k}'$ in each production. Lastly, $X_{e}'$ has
-an $\epsilon$-transition to the accepting state $Y_{e}$. Likewise, we build analogous defintions for the states
-$Z_{j}$ in $NFA(g')$, then we flatten all of these defintions into one mutally recursive list. This has the effect
-of disjunctively tying together $NFA(g)$ and $NFA(g')$ into one automaton.
-
-To show the strong equivalence of $NFA(g \oplus g')$ and $g \oplus g'$, we will demonstrate an isomorphism between them
-with two proof terms in our logic that mutually invert each other. To define a map from $NFA(g \oplus g')$ to $g \oplus g'$
-we will use a multi-fold over all the mutually recursive defintions of $NFA(g \oplus g')$. Note, even though multiple
-fixed points are not taken as primitive, using Beki\'c's theorem we can define an admissible axiom schema to define multiple fixed points types.
-Below we give rules to form $\mathsf{cons}$ and $\mathsf{mfold}$ terms to serve as the introduction and elimination rules, respectively, for multi-term fixed points.
-First, the introduction rule,
-
-\[
-  \inferrule
-  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
-  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
-\]
-where $\sigma$ is the substitution that unrolls the mutually recursive definition one level by replacing each $X_{k}$ with the fixed point $\mu(X_{j} = A_{j}, \forall j).X_{k}$,
-
-\begin{align*}
-  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \\
-  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
-\end{align*}
-
-and the elimination rule,
-\[
-  \inferrule{\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
-             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
-             \dots \\
-             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
-  }{\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
-\]
-
-where the substitution $\gamma$ represents $\{B_{1}/X_{1},\dots,B_{n}/X_{n} \}$.
-
-When defining a derivation from $NFA(g \oplus g')$ to $g \oplus g'$, we
-wish to simultaneously fold over the mutually recursive definitions comprising $NFA(g \oplus g')$. We may think of this
-as an eliminator that replaces each state inside of $NFA(g \oplus g')$ with a tail-recursive call that will ultimately
-return a term of type $g \oplus g'$.
-If folding over a term $e$, refer to the output of this multifold as $s_{\oplus}(e)$.
-The structure of such a derivation instatiates the above elimination rule as follow,
-
-\[
-  \inferrule{
-    \Delta \vdash e : NFA(g \oplus g') \\
-    y_{s} : \simulsubst {(X_{s}' \oplus Z'_{s})} \gamma \vdash p_{s} : g \oplus g' \\
-    x_{s}' : \simulsubst {(\simulsubst {X_{s}} {\delta_{X}})}{\gamma} \vdash p_{x,s} : g \\
-    x_{j}' : \simulsubst {(\simulsubst {X_{j}} {\delta_{X}})}{\gamma} \vdash p_{x,j} : X_{j} \text{~where } j \not \in \{s, e\} \\
-    x_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{x, e} : X_{e} \\
-    z_{s}' : \simulsubst {(\simulsubst {Z_{s}} {\delta_{Z}})}{\gamma} \vdash p_{z,s} : g' \\
-    z_{j}' : \simulsubst {(\simulsubst {Z_{j}} {\delta_{Z}})}{\gamma} \vdash p_{z,j} : Z_{j} \text{~where } j \not \in \{s, e\} \\
-    z_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{z,e} : Z_{e} \\
-    y_{e} : \simulsubst I {\gamma} \vdash p_{e} : I
-  }{\Delta \vdash s_{\oplus}(e) := \mathsf{mfold}(e) : g \oplus g'}
-\]
-
-For each of the variables above, we simplify the type and provide the actual proof term. For instance,
-$\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$, so the derivation for $p_{s}$ is just the identity.
-The rest of the terms are given below,
-\begin{itemize}
-  \item $y_{s}$ --- $\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$ and $p_{s} = y_{s} : g \oplus g'$
-  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = \mu(X_{i} = \hat{X}_{i}, \forall i).X_{s} = NFA(g)$ and $p_{x,s} = \phi^{-1}(x_{s}') : g$
-  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = \mu(X_{i} = \hat{X}_{i}, \forall i).X_{j}$ and $p_{x, j} = x_{j}' : \mu(X_{i} = \hat{X}_{i}, \forall i).
-X_{j}$
-  \item $x_{e}'$ --- $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{x,e} = x_{e}' : \mu(X_{i} = \hat{X}_{i}, \forall i).X_{e} = I$
-  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = \mu(Z_{i} = \hat{Z}_{i}, \forall i).Z_{s} = NFA(g')$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
-  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = \mu(Z_{i} = \hat{Z}_{i}, \forall i).Z_{j}$ and $p_{z, j} = z_{j}' : \mu(Z_{i} = \hat{Z}_{i}, \forall i).Z_{j}$
-  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = \mathsf{cons}_{Z_{e}}(z_{e}') : \mu(Z_{i} = \hat{Z}_{i}, \forall i). Z_{e} = I$
-  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
-\end{itemize}
-
-\steven{Need to add argument that justifies why the $X_{j}$s are allowed to be eliminated via the identity.
-A priori they could be interfered with by taking an $X_{s}$ that needs to be eliminated.
-I think it hinges on a syntactic restriction where there are no backedges to $X_{s}$ in the NFA.Even if there were, I think I can get away with a case distinction entailment instead of the one written}
-
-where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively.
-This finshes the construction of a parse transformer $s_{\oplus}$ from $NFA(g \oplus g')$ to $ g \oplus g'$.
-The map $r_{\oplus}$ in the other direction is given below in a couple stages. First, we show how to turn a parse of $NFA(g)$ into a parse of $NFA(g \oplus g')$.
-
-\[
-  \inferrule
-  {
-    \Gamma ; \Delta \vdash u : NFA(g) = \mu(X_{j} = \hat{X}_{j}, \forall j).X_{s} \\
-    \Gamma ; x_{s} : \simulsubst {X_{s}}{\gamma} = X_{s}' \vdash \mathsf{cons}(\textbf{in}_{X_{s}'}(x_{s})) : \mu(\dots).Y_{s} = NFA(g \oplus g')  \\
-    \Gamma ; x_{j} : \simulsubst {X_{j}}{\gamma} = X_{j}' \vdash x_{j} : X_{j}', j \not \in \{s, e\} \\
-    \Gamma ; x_{e} : \simulsubst {X_{e}}{\gamma} = I \vdash x_{e} : I = X_{e}'
-  }
-  {\Gamma ; \Delta \vdash \mathsf{mfold}(u) : NFA(g \oplus g')}
-\]
-where the dots in the above derivation take on the mutually recursive defintions present in $NFA(g \oplus g')$.
-\steven{Need a better way to display these defs without blowing up the $\mu$ terms} Denote the above embedding of $NFA(g)$ into $NFA(g \oplus g')$ via $\alpha(u) := \mathsf{mfold}(u) : NFA(g \oplus g')$.
-Similarly define an embedding $\beta$ for an embedding of $NFA(g')$.
-We can then define the derivation from $g \oplus g'$ to $NFA(g \oplus g')$ by composing the isomorphisms $g \overset{\phi}{\cong} NFA(g)$ and $g \overset{\psi}{\cong} NFA(g')$ with maps $\alpha$ and $\beta$, respectively.
-
-\[
-  \inferrule
-  {
-    p : g \oplus g' \vdash p : g \oplus g' \\
-    p_{1} : g \vdash \alpha(\phi(p_{1})) : NFA(g \oplus g') \\
-    p_{2} : g' \vdash \beta(\psi(p_{2})) : NFA(g \oplus g')}
-  {
-    p : g \oplus g' \vdash r_{\oplus}(p) := \textbf{case}~p \{
-    \textbf{in}_{g} p_{1} \mapsto \alpha(\phi(p_{1})), \\
-    \textbf{in}_{g'} p_{2} \mapsto \beta(\psi(p_{2})) \} : NFA(g \oplus g')}
-\]
-
-We include the proof that maps $s_{\oplus}$ and $r_{\oplus}$ form an isomorphism in \cref{sec:appendixnfa}. This concludes the disjunctive combination of two NFAs.
-Because they follow a very similar style of argumentation, we also include the cases for sequencing and
-Kleene star in \cref{sec:appendixnfa}. Totaling over all of these cases, this completes the proof.
-
-\end{proof}
-
-% \subsubsection{NFA to Grammar}
-
-% \steven{This section and following are basically todos for rn}
-
-% We now seek to take an NFA and derive the regular expression recognized by it.
-To this end we weaken our syntactic form of an NFA to that a \emph{generalized nondeterministic finite automata} (GNFA) --- where guards on transition can be arbitrary regexps instead of single characters.
-We then repeatedly take away states from the NFA to give an increasingly more complete algebraic description of the regular expression inside of the guards.
-We repeat until there is only a single transition between start and accepting state, and the guard on the single transition denotes the regular expression that is equivalent to the NFA.
-
-% \[
-%   GNFA ::= \mu \begin{pmatrix}
-%         Y_{s} = \langle Q \rangle \\
-%         Y_{1} = \langle Q \rangle \\
-%         \vdots \\
-%         Y_{e} = I
-%         \end{pmatrix}~\textbf{in}~Y_{s}
-% \]
-% where
-% \[
-%   Q ::= \langle RegExp \rangle \otimes Y ~|~ Y ~|~ Q \oplus Q
-% \]
-% \[
-%   RegExp ::= c ~|~ I ~|~ RegExp \oplus RegExp ~|~ RegExp \otimes RegExp ~|~ RegExp^{*}
-% \]
-
-% This is shown inductively by showing that the equivalence of a GNFA to a copy of itself with a general state --- i.e.\ non-start, non-accepting --- removed.
-Consider GNFA $A$ and remove state $Y_{r}$ to achieve $A'$.
-Let $\delta_{Y}$ denote the substitution that replaces $Y_{k}'/Y_{k}$ for $k \neq r$, and $\gamma$ denote the substitution $\hat{Y}_{r}/Y_{r}$.
-
-% \[
-%   A = \mu \begin{pmatrix}
-%             Y_{s} = \hat{Y}_{s} \\
-%             Y_{r} = \hat{Y}_{r} \\
-%             Y_{j} = \hat{Y}_{j}, j \not \in \{s, r, e\} \\
-%             Y_{e} = I
-%           \end{pmatrix}~\textbf{in}~Y_{s}
-% \]
-% \[
-%   A' = \mu \begin{pmatrix}
-%             Y_{s}' = \simulsubst {\hat{Y}_{s}} {\delta_{Y} \circ \gamma} \\
-%             Y_{j}' = \simulsubst {\hat{Y}_{j}} {\delta_{Y} \circ \gamma}, j \not \in \{s, e\} \\
-%             Y_{e}' = I
-%           \end{pmatrix}~\textbf{in}~Y_{s}'
-% \]
-
-% We can pretty easily translate between parses of $A$ and $A'$.
-This can be seen by looking at the structure of the terms $\hat{Y}_{k}$ and $\simulsubst {\hat{Y}_{k}} {\delta_{Y} \circ \gamma}$.
-For terms $\hat{Y}_{k}$ that don't have $Y_{r}$ as a subterm, the action of $\gamma$ is trivial --- so $\simulsubst {\hat{Y}_{k}} {\delta_{Y} \circ \gamma} = \simulsubst {\hat{Y}_{k}} {\delta_{Y}}$.
-For these $Y_{k}$ it is trivial to define a fold from $A$ to $A'$, or vice versa as we are simply renaming terms.
-For $\hat{Y}_{k}$ containing $Y_{r}$ as a subterm, it is not as trivial.
-Instead, each occurence of $Y_{r}$ is replaced with an equivalent definition.
-
-% Folding from $A$ to $A'$,
-
-% \begin{itemize}
-%   \item $y_{j}, j \not \in \{r, e\}$ and $Y_{r} \not \in \hat{Y}_{k}$ --- $\simulsubst {\hat{Y}_{k}} {\delta_{Y} \circ \Gamma} = \simulsubst {\hat{Y}_{k}} {\delta_{Y}}$ and $p_{s} = y_{j} : Y_{j}'$
-%   \item $y_{j}, j \not \in \{r, e\}$ and $Y_{r} \in \hat{Y}_{k}$ ---
-%   \item $y_{e}$ --- $\simulsubst I {\delta \circ \gamma} = I$ and $p_{e} = y_{e} : I$
-% \end{itemize}
-
-% \steven{An aside on some axioms}
-
-% Something like $A \cong (I \& A) \oplus \sum_{c} c \otimes (c \lto A)$
-
-% Also add the axiom that $\top \cong (\sum_{c \in \mathcal{A}} c)^{*}$
-
-% If we have the above, we can then show the following isos for some alphabet $\mathcal{A}$,
-% \begin{align*}
-%   A
-%   & \cong \top \amp A \\
-%   & \cong (\Sigma_{c \in \mathcal{A}})^{*}\amp A \\
-%   & \cong (I \oplus (\Sigma_{c} c \otimes \top)) \amp A \\
-%   & \cong (I \amp A) \oplus ((\Sigma_{c} c \otimes \top)) \amp A) \\
-%   & \cong (I \amp A) \oplus (\Sigma_{c }c \otimes (c \lto A)) & \text{using } \phi^{-1}
-% \end{align*}
-
-% From here we need to also show that derivatives of regular languages are regular (look at Brzozowski's proof) and that regular languages have finitely many derivatives (potentially with some size constraint on the derivative?).
-Then the last iso in the chain above can be used to build a DFA equivalent to $A$ using Brzozowski's algorithm.
-
-% Need to also add in axioms for $\amp$ distributing over $\oplus$ and $\lto$ distributing over $\oplus$. And that $c \amp c' \vdash 0$ for $c \neq c'$
-
-\subsection{Brzozowski Derivatives and Deterministic Finite Automata}
-\steven{this section sounds awkward and will be adjusted in a later editing pass}
-A \emph{deterministic finite automaton} (DFA) is an NFA whose transitions are deterministic.
-Meaning at every state $q \in Q$ there is exactly one transition for every character $c \in \Sigma$ --- i.e.\ the size of $\delta(q, c)$ is 1 and $\delta(q, \epsilon)$ is empty.
-It is clear that all DFAs happen to be NFAs, but it is perhaps surprising that the class of DFAs is as powerful as the class of NFAs when measured by the languages they recognize!
-\cref{thm:thompson} shows that the class of languages recognized by NFAs is exactly the regular languages.
-As we will soon see, the class of languages recognizable by DFAs is also the regular languages!
-
-There are a couple famous constructions we might use to show this, such as the powerset construction \cite{?} or Brzozowski's algorithm \cite{?}.
-Just as with Thompson's construction, we shall internalize a version of the latter as a proof term in our formal calculus.
-However, note that the construction of a DFA is lossy in some sense.
-That is, throughout this process we lose the ability to form an isomorphism between parses of a DFA for $g$ and parses of $g$ itself.
-Rather than express an isomorphism (strong equivalence) as a proof term, we shall construct a weak equivalence.
-This means we shall derive entailments $g \vdash DFA(g)$ and $DFA(g) \vdash g$ that do not necessarily invert each other.
-Semantically, this notion of weak equivalence between grammars $g$ and $g'$ means that $g$ and $g'$ both recognize the same language.
-
-Recall that a DFA is given as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$ in the same manner as an NFA.\ When encoding a DFA as a mutually recursive grammar, we use nearly the same process as when we encoded an NFA, with one subtle change.
-When we encoded NFAs, we were able to ensure we had a single accepting state by connecting all of the accepting states $q_{j} \in F$ to a single accepting state $Y_{e}$ with $\epsilon$-transitions.
-However, $\epsilon$-transitions are not permitted in a DFA, as they introduce a source of nondeterminism.
-Therefore we need a scheme to encode which states are accepting.
-The change we need to make is pretty lightweight, we add the grammar $I$ as a summand to distinguish which states are accepting.
-
-\[
-  DFA ::= \mu \begin{pmatrix}
-        Y_{s} = \langle \hat{Y} \rangle \\
-        Y_{j} = \langle \hat{Y} \rangle, j \not \in \{s, e\} \\
-        \end{pmatrix}.Y_{s}
-\]
-where
-\[
-\hat{Y} ::= \hat{Y_{0}} ~|~ I \oplus \hat{Y_{0}}
-\]
-\[
-  \hat{Y_{0}} ::= c \otimes Y ~|~ Y ~|~ \hat{Y_{0}} \oplus \hat{Y_{0}}, Y \in \{Y_{s}, Y_{1}, \dots, Y_{N}\}
-\]
-
-A state $Y_{j}$ in a DFA is accepting if and only if $\hat{Y}_{j}$ contains $I$ as a summand.
-
-some fluff to lead into the thoerem
-
-\begin{theorem}
-  For every regular grammar $g$, there is a DFA $D$ that is weakly equivalent to $g$.
-\end{theorem}
-
-\begin{proof}
-  \steven{This is currently word salad}
-
-  DFA construction
-
-  $DFA(g) \vdash g$ : just fold over the defs in $DFA(g)$ and have every $Y_{w}$ eliminate to $w \lto g$.
-This is doable bc when eliminating $Y_{w'}$ you will hold terms of type $a \otimes (w \lto g)$.
-These $w \lto g$ terms might not be exactly what you want, but they are weakly equivalent to some derivative that you do want!
-How do you know that they're weakly equivalent?
-if $Y_{w'}$ transitions to $Y_{w}$ via character $a$, then we know that $a \lto (w \lto g)$ is weakly equivalent to $a \lto (w' \lto g)$ \emph{by construction of the DFA}.
-This argument entirely hinges on the method by which we defined the DFA
-
-  $g \vdash DFA(g)$
-  Idea is to us $\top \cong \text{Char}^{*}$ and $A^{*} \cong \Sigma(n) A^{\otimes n}$ and $A \cong \top \amp A$ to turn a parse into a pair of a parse and a finite string.
-And then this finite string can be walked through the DFA to read off the parse, and you can show its accepting.
-Details currently a little shaky, and encapsulated in a bad picture of a whiteboard.
-\steven{Gonna has this out with Max in meeting}
-\end{proof}
-
-\subsection{Indexed Automata}
-
-We can generalize to more powerful notions of automata by utilizing
-dependency of linear types on non-linear types. Let $S$ be a finite
-non-linear type and define $\textrm{List}\,S = \mu x. 1 + S \times
-x$. Then a pushdown automaton with $S$ as the stack alphabet can be
-defined as a mutual fixed point.
-
-\steven{Forms for PDAs and Turing Machines go here}
-
-Want to put PDA, CFGs, CSGs here. Basically, use this section to highlight the usefulness of the intuitionistic part of the contexts
-
-\section{Future work}
-
-\subsection{Implementation}
-
-\subsection{Beyond Strings}
-
-While parsing typically refers to the production of semantic objects
-from their description as strings, many tasks in programming can be
-viewed as parsing of more structured objects such as trees, trees with
-binding structure or graphs. Fundamental to programming language
-implementation is \emph{type checking}, analogous to language
-recognition, or more generally \emph{typed elaboration}, analogous to
-parsing, which produces a semantic object in addition to performing
-some analysis. A type system could then be interpreted as a \emph{tree
-grammar}. Our definition of string grammars as functors from
-$\Sigma^*$ to $\Set$ can then be easily adapted to functors to $\Set$
-from the set of \emph{trees} freely generated form some signature of
-nodes. Then just as the set of string grammars inherits a monoidal
-structure from the free monoid of strings using the Day convolution,
-this set of tree grammars inherits a kind of weak algebraic structure
-corresponding to the tree constructors. This suggests an even more
-unusual form of bunched type theory, where the monoidal concatenation
-and unit are replaced with context constructors corresponding to the
-tree constructors.
-
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{references}
-
-\clearpage
-\appendix
-
-\section{Semantics in Sets and Grammars}
-
-A model of Grammar Type Theory is a category with families equipped
-with a functor $L : \mathcal C \to \textrm{GrLog}$ from the category
-of contexts to the category of grammar logic models and strict morphisms.
-
-First, we give the set theoretic model. We use the standard category
-with families structure given by the category of sets and a cumulative
-tower of universes.
-%
-Next, we define $L$ as follows:
-\[ L(\Gamma) = \mathcal P \Sigma^*/\textrm{Disc} \Gamma \]
-Where $\textrm{Disc} \Gamma$ is the presheaf that is constantly $\Gamma$.
-Note that
-\[ \mathcal P \Sigma^*/\textrm{Disc} \Gamma \cong (\Set/\Gamma)/\textrm{Disc} \Sigma^*\]
-where $\textrm{Disc} \Sigma^*$ is the projection $\Gamma \times \Sigma^*
-\to \Gamma$.  Then by the fundamental theorem of topos thoery, $\Set/\Gamma$ is
-a topos. Since $\textrm{Disc} : \Set \to \Set/\Gamma$ is monoidal,
-$\textrm{Disc} \Sigma^*$ carries a monoid structure in $\Set/\Gamma$ and
-this induces a monoidal closed structure on $(\Set/\Gamma)/\textrm{Disc}
-\Sigma^*$. Again by the fundamental theorem, it also has finite
-products and coproducts and a natural numbers object, and so can
-interpret least fixed points.\max{More details about constructing initial algebras?}
-
-Then we can define the remaining type structure as follows. Unraveling
-the above, a non-linear type $\Gamma \vdash X$ is interpreted as an
-object of $\Set/\sem{\Gamma}$ and a linear type $\Gamma \vdash A$ is
-interpreted as an object of $\mathcal P\Sigma^*/\textrm{Disc}\sem{\Gamma}$. We
-define them below in indexed style:
-\begin{enumerate}
-\item $\sem{c} \gamma w = (w = c)$
-\item $\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-\item $\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-\item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
-\item $\sem{L_i}\gamma = (\mathcal P_i\Sigma^*/\textrm{Disc}\sem{\Gamma})_0$
-\end{enumerate}
-
-\section{Canonicity}
-
-For the canonicity argument, we construct a second ``glued'' model
-that contains a projection morphism to the syntactic model.
-
-Let $S$ be the syntactic category with families for the type
-theory. Define the glued category with families to be the comma category
-% https://q.uiver.app/#q=WzAsNSxbMCwwLCJcXG1hdGhjYWwgRyJdLFswLDIsIlMiXSxbMiwyLCJcXFNldCJdLFsyLDAsIlxcU2V0Il0sWzEsMSwiXFxzd2Fycm93Il0sWzMsMl0sWzEsMl0sWzAsMV0sWzAsM11d
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{\gluedNL} \&\& \Set \\
-	\& \rotatebox[origin=c]{-45}{$\Downarrow$} \\
-	S \&\& \Set
-	\arrow[Rightarrow, no head, from=1-3, to=3-3]
-	\arrow["{S(\cdot,-)}"', from=3-1, to=3-3]
-	\arrow[from=1-1, to=3-1]
-	\arrow[from=1-1, to=1-3]
-\end{tikzcd}\]
-This is a standard glued category for canonicity of dependent type
-theory, and carries a category with families structure that we will
-reuse from prior work for the (non-linear) universes,
-$\Pi/\Sigma/\textrm{Id}$ \cite{coquand}. Concretely an object of
-$\mathcal G$ is equivalent to a pair of a syntactic context $\Gamma$,
-and a family $\hat \Gamma$ over (equivalence classes of) closing
-substitutions for $\Gamma$.
-
-Next we turn to interpreting the linear types and morphisms. For this
-we define a functor $\gluedL : \gluedNL \to \textrm{MonCat}$ where
-$\textrm{MonCat}$ is the category\footnote{in univalent foundations
-this would be the category of \emph{strict} monoidal categories.} of
-monoidal categories and lax monoidal functors.  We construct
-$\gluedL(p : \hat \Gamma \to S(\cdot,\Gamma))$ as the following comma
-category:
-
-% https://q.uiver.app/#q=WzAsNixbMCwwLCJcXGdsdWVkTkwocCkiXSxbMCwyLCJMX1xcR2FtbWEiXSxbMSwyLCJcXG1hdGhjYWwgUCBcXFNpZ21hXiovXFx0ZXh0cm17RGlzY30oUyhcXGNkb3QsXFxHYW1tYSkpIl0sWzIsMiwiXFxtYXRoY2FsIFAgXFxTaWdtYV4qL1xcdGV4dHJte0Rpc2N9KFxcaGF0XFxHYW1tYSkiXSxbMiwwLCJcXG1hdGhjYWwgUCBcXFNpZ21hXiovXFx0ZXh0cm17RGlzY30oXFxoYXRcXEdhbW1hKSJdLFsxLDEsIlxccm90YXRlYm94W29yaWdpbj1jXXstNDV9e1xcRG93bmFycm93fSJdLFs0LDMsIiIsMCx7ImxldmVsIjoyLCJzdHlsZSI6eyJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsNF0sWzAsMV0sWzEsMiwiTiIsMl0sWzIsMywiXFx0ZXh0cm17RGlzY30ocF4qKSIsMl1d
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{\gluedNL(p)} \&\& {\mathcal P \Sigma^*/\textrm{Disc}(\hat\Gamma)} \\
-	\& {\rotatebox[origin=c]{-45}{$\Downarrow$}} \\
-	{L_\Gamma} \& {\mathcal P \Sigma^*/\textrm{Disc}(S(\cdot,\Gamma))} \& {\mathcal P \Sigma^*/\textrm{Disc}(\hat\Gamma)}
-	\arrow[Rightarrow, no head, from=1-3, to=3-3]
-	\arrow[from=1-1, to=1-3]
-	\arrow[from=1-1, to=3-1]
-	\arrow["N"', from=3-1, to=3-2]
-	\arrow["{\textrm{Disc}(p^*)}"', from=3-2, to=3-3]
-\end{tikzcd}\]
-
-Where $N : L_\Gamma \to {\mathcal P
-  \Sigma^*/\textrm{Disc}(S(\cdot,\Gamma))}$ is the ``dependent nerve''
-of the inclusion $\lceil\cdot\rceil : \Sigma^* \to L_\Gamma$ of
-strings into linear contexts defined in indexed style as
-\[ N(\Gamma \vdash A) \gamma w = \lceil w \rceil \vdash \cdot :  \simulsubst A \gamma \]
-Then observe that $N$ and ${\textrm{Disc}(p^*)}$ are lax monoidal and
-${\mathcal P \Sigma^*/\textrm{Disc}(\hat\Gamma)}$ is a topos so the
-criteria for Hyland and Schalk's gluing construction apply, defining a
-biclosed monoidal structure on $\gluedNL(p)$ as well as finite
-(co-)products. Further, these constructions are all stable under
-reindexing in $\gluedNL$, and so interpret the dependent linear typing
-constructions for $I,\otimes,0,\oplus,\top,\amp,\lto,\tol$.
-
-Concretely, an object of $\gluedNL(p)$ is equivalent to a pair of a
-non-linear type $\Gamma \vdash A$ and a formal grammar $\hat A$
-indexed by $\hat \Gamma$ with a projection function from $\hat A$ to
-$N(A)$. We can think of this presheaf $\hat A$ as the presheaf of
-canonical forms of $\simulsubst A \gamma$ for closing substitutions
-$\gamma$ in contexts of the form $\lceil w \rceil$.
-
-All that remains in this model is to define the semantics of the
-non-standard connectives. We define them in an indexed style. First we
-define the interpretation of a non-linear type $\Gamma \vdash X$ as a
-family of sets $\hat X$ indexed by a closing substitution $\gamma :
-S(\cdot,\Gamma)$, a semantic substitution $\hat \gamma : \hat
-\Gamma(\gamma)$ and a closed term $\cdot \vdash e : \simulsubst A
-\gamma$. Second, we define a linear type $\Gamma \vdash A$ as a family
-$\hat A$ of grammars indexed by $\gamma,\hat\gamma$ as above along
-with a morphism $\pi : \hat A \gamma \hat \gamma \to N(\simulsubst A
-\gamma)$.
-
-\begin{enumerate}
-\item $\hat{c} \gamma \hat \gamma w = (w = c)$ and $\pi(*) = \lceil c \rceil = x_0:c \vdash x_0:c$
-\item $\widehat{(\LinSigTy x X A)} \gamma \hat \gamma w = \Sigma(\cdot \vdash e : \simulsubst X \gamma)\Sigma(\hat x: \hat X \gamma \hat \gamma e) \hat{A}(\gamma,e)(\hat\gamma,\hat x) w$
-\item $\sem{\LinPiTy x X A} \gamma w = \Sigma(x:\simulsubst X \gamma;\lceil w\rceil \vdash e:\simulsubst A \gamma)\Pi(\cdot \vdash e' : \simulsubst X \gamma)\Pi(\hat x:\hat{X}\gamma\hat\gamma x)\Sigma(\hat a : \hat A (\gamma,e')(\hat\gamma,\hat x) \times (\pi(\hat a) = \subst e {e'} x))$
-  and $\pi(e,f) = \dlamb x.e$
-\item $\widehat{G A} \gamma \hat \gamma e = \Sigma(\hat a: \hat{A} \gamma \hat\gamma \varepsilon) (p\gamma\hat\gamma \varepsilon(\hat a) = \mathsf{G}^{-1}e)$
-\item $\sem{L_i}\gamma = \gluedNLUniv(\gamma)$
-\end{enumerate}
-
-Thus we have defined a second model $\gluedNL,\gluedL$ of the type
-theory, with a morphism to the syntactic model. By the initiality of
-the syntactic model, there is a morphism back to the glued model which
-is a section of this projection. For purely positive types, we can
-observe by induction on the type structure that the total space of the
-family is isomorphic to the set-theoretic semantics. Further, the
-projection above defines a section of the families. Finally, by
-induction on formulae again we see that this projection is also a
-section and therefore an isomorphism, establishing the canonicity
-result as stated in Section\ref{section:canonicity}.
-
-\section{Equivalence Betweeen Regular Grammars and Finite Automata}
-\label{sec:appendixnfa}
-
-Below we finish the proof of \cref{thm:thompson}.
-\begin{proof}
-
-  First, we finish the provide the proof that the maps $s_{\oplus}$ and $r_{\oplus}$ provided in the disjunctive case do indeed mutually invert each other.
-For brevity in this proof, we will omit the list of mutually recursive bindings when appropriate.
-That is, write $\mu X_{s}'$ instead of $\mu(Y_{s},X_{s}',X_{j}',\dots).
-X_{s}'$.
-
-  Given $e : NFA(g \oplus g')$,
-
-  \steven{There a weird spacing issues between these align envs. Maybe wrap these in a figure env so that spacing is better}
-
-\begin{align*}
-  r_{\oplus}(s_{\oplus}(e))
-  & =_{\mu \eta} r_{\oplus}(s_{\oplus}(\mathsf{cons}(e'))), \text{~where }e':\mu X_{s}' \oplus \mu Z_{s}' \\
-  &
-    \begin{aligned}
-      =_{\oplus \eta} r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{case}~e'~\{ & \\
-    \textbf{in}_{\mu X_{s}'} u \mapsto \textbf{in}_{\mu X_{s}'} u, & \\
-    \textbf{in}_{\mu Z_{s}'} v \mapsto \textbf{in}_{\mu Z_{s}'} v
-    \})))
-    \end{aligned} \\
-  &
-    \begin{aligned}
-    =_{\oplus \eta}
-    \textbf{case}~e'~\{ & \\
-    & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{in}_{\mu X_{s}'} u))), \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(s_{\oplus}(\mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v))) \\
-    \}))
-    \end{aligned} \\
-  &
-    \begin{aligned}
-    =
-    \textbf{case}~e'~\{ & \\
-    & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(\mathsf{mfold}(\mathsf{cons}(\textbf{in}_{\mu X_{s}'} u))), \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(\mathsf{mfold}(\mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v))) \\
-    \}
-    \end{aligned} \\
-  &
-    \begin{aligned}
-    =_{\mu \beta}
-    \textbf{case}~e'~\{ & \\
-    & \textbf{in}_{\mu X_{s}'} u \mapsto r_{\oplus}(\textbf{in}_{g} \phi^{-1}(\mathsf{cons}_{\mu(X_{i}=\hat{X}_{i}).X_{s}}u')), \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto r_{\oplus}(\textbf{in}_{g'} \psi^{-1}(\mathsf{cons}_{\mu(Z_{i}=\hat{Z}_{i}).Z_{s}}v')) \\
-      \} \\
-      & \text{~for }u' : \mu X_{s}', v' : \mu Z_{s}'
-    \end{aligned} \\
-  &
-    \begin{aligned}
-    =_{\oplus \beta}
-    \textbf{case}~e'~\{ & \\
-    & \textbf{in}_{\mu X_{s}'} u \mapsto \alpha \phi \phi^{-1}(\mathsf{cons}(u')), \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto \beta \psi \psi^{-1}(\mathsf{cons}(v')) \\
-      \}
-    \end{aligned} \\
-  &
-    \begin{aligned}
-    =
-    \textbf{case}~e'~\{ & \\
-    & \textbf{in}_{\mu X_{s}'} u \mapsto \alpha (\mathsf{cons}(u')), \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto \beta (\mathsf{cons}(v')) \\
-      \}
-    \end{aligned} \\
-   &
-    \begin{aligned}
-    =_{\mu \beta}
-    \textbf{case}~e'~\{ & \\
-    & \textbf{in}_{\mu X_{s}'} u \mapsto \mathsf{cons}(\textbf{in}_{\mu X_{s}'} u) , \\
-    & \textbf{in}_{\mu Z_{s}'} v \mapsto \mathsf{cons}(\textbf{in}_{\mu Z_{s}'} v) \\
-      \}
-    \end{aligned} \\
-  & =_{\oplus \eta} \mathsf{cons}(e') \\
-  & =_{\mu \beta} e
-\end{align*}
-
-Likewise for $p : g \oplus g'$,
-
-\steven{This first line is a weird alignment fix, touch up}
-\begin{align*}
-  s_{\oplus}(r_{\oplus}(p))
-  &
-  = s_{\oplus}(r_{\oplus}(p)) \\
-   & \begin{aligned}
-    =_{\oplus \eta}
-      \textbf{case}~p~\{
-    \textbf{in}_{g}p_{1} \mapsto s_{\oplus}(r_{\oplus}(\textbf{in}_{g}p_{1})), \\
-    \textbf{in}_{g'}p_{2} \mapsto s_{\oplus}(r_{\oplus}(\textbf{in}_{g'}p_{2})) \}
-    \end{aligned} \\
-   & \begin{aligned}
-    =_{\oplus \beta}
-      \textbf{case}~p~\{
-    \textbf{in}_{g}p_{1} \mapsto s_{\oplus}(\alpha \phi p_{1}), \\
-    \textbf{in}_{g'}p_{2} \mapsto s_{\oplus}(\beta \psi p_{2}) \}
-    \end{aligned} \\
-   & \begin{aligned}
-    =
-      \textbf{case}~p~\{
-    \textbf{in}_{g}p_{1} \mapsto \mathsf{mfold}(\alpha \phi p_{1}), \\
-    \textbf{in}_{g'}p_{2} \mapsto \mathsf{mfold}(\beta \psi p_{2}) \}
-    \end{aligned} \\
-    & \begin{aligned}
-    =_{\mu \beta}
-      \textbf{case}~p~\{
-    \textbf{in}_{g}p_{1} \mapsto \textbf{in}_{g} (\phi^{-1} \phi p_{1}), \\
-    \textbf{in}_{g'}p_{2} \mapsto \textbf{in}_{g'}(\psi^{-1} \psi p_{2}) \}
-    \end{aligned} \\
-     & \begin{aligned}
-    =
-      \textbf{case}~p~\{
-    \textbf{in}_{g}p_{1} \mapsto \textbf{in}_{g} (p_{1}), \\
-    \textbf{in}_{g'}p_{2} \mapsto \textbf{in}_{g'}(p_{2}) \}
-    \end{aligned} \\
-  &
-    =_{\oplus \eta} p
-\end{align*}
-
-We now construct a combinator for the sequencing of two NFAs.
-
-\[
-  NFA(g \otimes g') = \mu
-  \begin{pmatrix}
-    Y_{s} = X_{s}' \\
-    X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
-    X_{e}' = Z_{s}' \\
-    Z_{j}' = \simulsubst {\hat{Z}_{j}} {\delta_{Z}}, j \neq e \\
-    Z_{e}' = Y_{e} \\
-    Y_{e} = I
-  \end{pmatrix}
-\]
-
-Just as before, we use a multifold to define the map $s_{\otimes}$ from $NFA(g \otimes g')$ to $g \otimes g'$. Here the table of terms is as follows,
-
-\begin{itemize}
-  \item $y_{s}$ --- $\simulsubst {X_{s}'} \gamma = g \otimes g'$ and $p_{s} = y_{s} $
-  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = (\mu (X_{i}=\hat{X}_{i}).X_{s}) \otimes g' = NFA(g) \otimes g'$ and $p_{x,s} = \letin {u \otimes v} {x_{s}'} {\phi^{-1}(u) \otimes v} : g \otimes g'$
-  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = (\mu (X_{i}=\hat{X}_{i}).X_{j}) \otimes g'$ and $p_{x, j} = x_{j}' : (\mu (X_{i}=\hat{X}_{i}).X_{j}) \otimes g'$
-  \item $x_{e}'$ --- $\simulsubst {Z_{s}'}{\gamma} = g'$ and $p_{x,e} = x_{e}' : g'$
-  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = \mu (Z_{i} = \hat{Z}_{i}).Z_{s} = NFA(g')$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
-  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = \mu (Z_{i} = \hat{Z}_{i}).Z_{j}$ and $p_{z, j} = z_{j}' : \mu (Z_{i} = \hat{Z}_{i}).Z_{j}$
-  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = \mathsf{cons}(z_{e}') : \mu (Z_{i} = \hat{Z}_{i}).Z_{e} = I$
-  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
-\end{itemize}
-
-The map $r_{\otimes}$ in the other direction is given as,
-
-\[
-  \inferrule{
-    p : g \otimes g' \vdash p : g \otimes g' \\
-    p_{1} : g, p_{2} : g' \vdash \mathsf{cons}(\phi(p_{1}), \psi(p_{2})) : NFA(g \otimes g')}
-  {p : g \otimes g' \vdash \letin {p_{1} \otimes p_{2}} {p} {\mathsf{cons}(\phi(p_{1}), \psi(p_{2}))} : NFA(g \otimes g')}
-\]
-
-We show that $r_{\otimes}$ and $s_{\otimes}$ form an isomorhic pair,
-\steven{This proof doesn't exactly work}
-
-\begin{align*}
-r_{\otimes}(s_{\otimes}(e))
-  & =_{\mu \eta} r_{\otimes}(s_{\otimes}(\mathsf{cons}_{\mu X_{s}'}(e'))) \\
-  & = r_{\otimes}(\mathsf{mfold}(\mathsf{cons}_{\mu X_{s}'}(e'))) \\
-  & =_{\mu \beta} r_{\otimes}(\phi^{-1}(u) \otimes \psi^{-1}(v)) \\
-  & =_{\mu \beta} \mathsf{cons}(\phi \phi^{-1} u, \psi \psi^{-1} v) \\
-  & =_{\mu \eta} e
-\end{align*}
-
-\begin{align*}
-  s_{\otimes}(r_{\otimes}(p))
-  & =_{\otimes \eta} s_{\otimes}(r_{\otimes}(p_{g} \otimes p_{g'})) \\
-  & = s_{\otimes}(\mathsf{cons}(\phi p_{g}, \psi p_{g'})) \\
-  & =_{\mu \beta} \phi^{-1}\phi p_{g} \otimes \psi^{-1} \psi p_{g'} \\
-  & = p_{g} \otimes p_{g'} \\
-  & =_{\otimes \eta} p
-\end{align*}
-
-Finally, we build an automata combinator to represent the Kleene star\footnote{Which is another admissible definition via use of the fixed point constructor}.
-
-\[
-  NFA(g^{*}) = \mu
-  \begin{pmatrix}
-    Y_{s} = X_{s}' \oplus Y_{e} \\
-    X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
-    X_{e}' = X_{s}' \oplus Y_{e} \\
-    Y_{e} = I
-  \end{pmatrix}
-\]
-
-Again we define $s_{*}$ as a fold over these defintions to get a parse of $g^{*}$,
-\begin{itemize}
-  \item $y_{s}$ --- $\simulsubst {(X_{s}' \oplus Y_{e})} \gamma = g^{*} \oplus I$ and $p_{s} = \textbf{case}~y_{s} \{ \textbf{in}_{g^{*}} u \mapsto u, \textbf{in}_{I} v \mapsto \mathsf{cons}_{*I} v \} : g^{*}$
-  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = (\mu(X_{i}=\hat{X_{i}}).X_{s}) \otimes g^{*} = NFA(g) \otimes g^{*}$ and $p_{x,s} = \letin {u \otimes p_{*}} {x_{s}'} {\mathsf{cons}_{*}(\phi^{-1}(u), p_{*})} : g^{*} $
-  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = (\mu(X_{i}=\hat{X_{i}}).X_{j}) \otimes g^{*}$ and $p_{x, j} = : (\mu(X_{i}=\hat{X_{i}}).X_{j}) \otimes g^{*}$
-  \item $x_{e}'$ --- $\simulsubst {(X_{s}' \oplus Y_{e})}{\gamma} = g^{*} \oplus I$ and $p_{x,e} = \mathsf{cons}_{X_{e}}(()) \otimes \textbf{case}~x_{e}' \{ \textbf{in}_{g^{*}} u \mapsto u, \textbf{in}_{I} v \mapsto \mathsf{cons}_{*I} v \} : (\mu(X_{i}=\hat{X_{i}}).X_{e}) \otimes g^{*}$
-  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
-\end{itemize}
-
-and to derive a parse of $NFA(g^{*})$ from a parse of $g^{*}$ we define $r_{*}$ as,
-
-\[
-  \inferrule{p : g^{*} \vdash p : g^{*} \\ \cdot \vdash \mathsf{cons}(\textbf{in}_{{Y_{e}}}(())) : NFA(g^{*}) \\ u : g, v : g^{*} \vdash \mathsf{cons}(\textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v)))) : NFA(g^{*}) }{p : g^{*} \vdash \mathsf{fold}_{*}(\mathsf{cons}(\textbf{in}_{{Y_{e}}}(())), \mathsf{cons}(\textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v)))) : NFA(g^{*}) }
-\]
-
-To see that $s_{*}$ and $r_{*}$ for an isomorphism,
-\steven{This proof also needs more work. Esp after 4th equality}
-\steven{Think I need to thread through a case distinction on what the $g^{*}$ term is. Putting a pin in this for now to type up the DFA stuff before meeting}
-\begin{align*}
-  r_{*}(s_{*}(e))
-  & =_{\mu \eta} r_{*}(s_{*}(\mathsf{cons}(e'))), \text{~where }e':\mu X_{s}' \oplus Y_{e} \\
-  &
-    \begin{aligned}
-      =_{\oplus \eta}
-      \textbf{case}~e'~\{ & \\
-      & \textbf{in}_{X_{s}'} u \mapsto r_{*}(s_{*}(\mathsf{cons}(\textbf{in}_{X_{s}'} u))), \\
-      & \textbf{in}_{Y_{e}} v \mapsto r_{*}(s_{*}(\mathsf{cons}(\textbf{in}_{Y_{e}} v))) \\
-      \}
-    \end{aligned} \\
-  &
-    \begin{aligned}
-      =_{\mu \beta}
-      \textbf{case}~e'~\{ & \\
-      & \textbf{in}_{X_{s}'} u \mapsto r_{*}(\mathsf{cons}_{*}(\phi^{-1} u, p_{*})), \\
-      & \textbf{in}_{Y_{e}} v \mapsto r_{*}(\mathsf{cons}_{*}(()) \\
-      \}
-    \end{aligned} \\
-  &
-    \begin{aligned}
-      =_{\mu \beta}
-      \textbf{case}~e'~\{ & \\
-      & \textbf{in}_{X_{s}'} u \mapsto \mathsf{cons}(\textbf{in}_{X_{s}'}(\phi(\mathsf{cons}_{*}(\phi^{-1}u,p_{*})))), \\
-      & \textbf{in}_{Y_{e}} v \mapsto \mathsf{cons}(\textbf{in}_{Y_{e}}(()) \\
-      \}
-    \end{aligned} \\
-  & = \dots \\
-  & = e
-\end{align*}
-
-\steven{for the empty case above, can I make an $\eta$ equality for all terms of type $I$ to argue $\textbf{in}(v) = \textbf{in} (())$}
-
-\begin{align*}
-  s_{*}(r_{*}(p))
-  & =_{\mu \eta} s_{*}(r_{*}(\mathsf{cons}(p_{e}, p_{*}))), \text{~where } p_{e} : I, p_{*} : g \otimes g^{*} \\
-  & =_{\mu \beta} s_{*}()
-  & = p
-\end{align*}
-
-This completes our inductive proof that $NFA(g) \cong g$ for all regular expressions $g$.
-
-\end{proof}
+\steven{Do we need inverse for $\ltonl A$ in judgemental equality figure?}
+\steven{Do we need to say anything about the equalizer proofs being unique in
+  judgemental equality figure?}
 

### 2024-11-08 paper/paper.tex
spelling

@@
 respectively. We will freely use $\bot$ to refer to this empty
-discunction and $\top$ to refer to the empty conjunction. Lastly, we
+disjunction and $\top$ to refer to the empty conjunction. Lastly, we
 include a type $\equalizer {a}{f}{g}$ that we call the
@@
   this way for instance. I think there is a linear container paper
-  that uses this appraoch}
+  that uses this approach}
 
@@
 This section explores the applications of \theoryabbv to conducting formal
-gramamr theory. We demonstrate that several classical notions and constructions
+grammar theory. We demonstrate that several classical notions and constructions
 integral to the theory of formal languages are faithfully represented
@@
 \cref{def:unambig} can be read more intuitively as stating that $\A$ is
-unabmiguous if there is at most one way to transform parses of any other grammar
+unambiguous if there is at most one way to transform parses of any other grammar
 $\B$ into parse of $\A$.
@@
 Note that this definition of unambiguity is preserved by strong equivalence, so a common proof technique for showing $\A$ is unambiguous is by proving $\A$
-is strongly equivalent to an unambguous grammar $\B$. In this style of proof, we
+is strongly equivalent to an unambiguous grammar $\B$. In this style of proof, we
 build the parse transformers $\f : \ltonl{(\A \lto \B)}$ and
@@
 NFA.\ When generalizing from this example to an arbitrary NFA, we do not have the
-luxury of inspecting a concrete defintion. Therefore, we abstract over the
+luxury of inspecting a concrete definition. Therefore, we abstract over the
 specification of an arbitrary NFA and define traces with respect to that
@@
 
-\steven{TODO make the agda logo clickable, and mention its meaning somewhere}
+\steven{TODO make the Agda logo clickable, and mention its meaning somewhere}
 \newcommand{\Agda}{\agdalogo}
@@
 \begin{proof}
-  We present a variant of Rabin and Scotts's classic powerset construction \cite{rabinFiniteAutomataTheir1959}. First, we build
+  We present a variant of Rabin and Scott's classic powerset construction \cite{rabinFiniteAutomataTheir1959}. First, we build
   the DFA $\D$. Define the
@@
   from $\N$. Construct the initial state of $\D$ as the
-  $\epsilon$-closure of $\N.\init$. Lastly define the transtion function
+  $\epsilon$-closure of $\N.\init$. Lastly define the transition function
   of $\D$ to send the subset $X$ under the character $\c$ to the
@@
 trace starting from $\N.\dst~\texttt{trans}$. We then recursively call $\NtoD$
-on this trace which returns a family of traces parametrized by $\epsilon$-closed
+on this trace which returns a family of traces parameterized by $\epsilon$-closed
 subsets containing $N.\dst~\texttt{trans}$. We instantiate this family at the
@@
 $\N.\epssrc~\epsilon\texttt{trans} \in \X$, then the construction is quite
-similar. In this case we again recursivley call $\NtoD$, except afterwards we
+similar. In this case we again recursively call $\NtoD$, except afterwards we
 instantiate the family of grammars at the set $X$ itself. We may do this because
@@
 $\D.\delta~\c~\X$ to one rooted at $\s$. By definition of $\D.\delta~\c~\X$,
-there exists some $\s' \in \X$ such that $\s'$ tranitions to $\s$ by some number of
+there exists some $\s' \in \X$ such that $\s'$ transitions to $\s$ by some number of
 $\epsilon$-transitions followed by a transition labeled by $\c$.
@@
 and is defined in \cref{fig:dyckinductive}. The $\nil$ constructor shows that
-the empty string is balanced, and the $\balanced$ constrcutor builds a balanced
+the empty string is balanced, and the $\balanced$ constructor builds a balanced
 parse by wrapping an already balanced parse in an additional set of parentheses
@@
 %% 5. Expressiveness: Turing Machines
-The expressive power of \theoryabbv spans the entire Chomsky hierachy, as we may
-encode the acccepting traces of Turing machines as linear types.
+The expressive power of \theoryabbv spans the entire Chomsky hierarchy, as we may
+encode the accepting traces of Turing machines as linear types.
 
@@
 terminate within $\mathsf{fuel}$-many steps, then we return the trace along with
-the current tape and a boolean denoting if the trace is accepting or rejcting.
-If the trace does not terminate within the alloted amount of fuel, then we
+the current tape and a boolean denoting if the trace is accepting or rejecting.
+If the trace does not terminate within the allotted amount of fuel, then we
 timeout and return nothing.
@@
 This category is equivalent to the slice category $\Set/\Sigma^*$ and
-as such is very well-behaved. It is complete, co-complete, cartesian
+as such is very well-behaved. It is complete, co-complete, Cartesian
 closed and carries a monoidal biclosed structure. We will use these
@@
 For reasons of brevity, we present only a selection of cases in
-\Cref{fig:semantics}. The full details are in the appendix.\max{todo:
+\Cref{fig:semantics}. The full details are in the appendix.\max{TODO:
   put the full details in the appendix}.
@@
 
-\max{todo: spelling}
+\max{TODO: spelling}
 \begin{theorem}[Adamek]
@@
 
-\steven{Should discussion of this connection to lambek calculus be spread
+\steven{Should discussion of this connection to Lambek calculus be spread
   throughout the paper?}
@@
 languages. That is, they seem to try to give a functorial semantic
-interpretation of natrual language.
+interpretation of natural language.
 
@@
 Also mention here that \cite{luoLambekCalculusDependent} seems similar at first
-glance, but isn't as similar as it might look because the judgements and
+glance, but isn't as similar as it might look because the judgments and
 dependent types seem to
@@
 \steven{TODO Edit the above}
-\steven{Even though the above sounds bad, are there lambek/categorial things that are missed?}
+\steven{Even though the above sounds bad, are there Lambek/categorial things that are missed?}
 
@@
 
-\steven{TODO add lambek calculus/categorial grammars here}
+\steven{TODO add Lambek calculus/categorial grammars here}
 
@@
 \end{figure}
-\steven{Do we need inverse for $\ltonl A$ in judgemental equality figure?}
+\steven{Do we need inverse for $\ltonl A$ in judgmental equality figure?}
 \steven{Do we need to say anything about the equalizer proofs being unique in
-  judgemental equality figure?}
+  judgmental equality figure?}
 

### 2024-11-08 paper/paper.tex
Turing machines

@@
 \newcommand{\ttt}{\mathtt{t}}
+\newcommand{\blank}{\mathtt{blank}}
+\newcommand{\push}{\mathtt{push}}
 \newcommand{\T}{\mathtt{T}}
@@
 \newcommand{\snoc}{\mathtt{snoc}}
-\steven{This subsection is still a WIP}
 %% 5. Expressiveness: Turing Machines
@@
 
-\steven{TODO introduce tape alphabet and blank symbol}
-
 The data of a Turing machine $\T$  is given by the following data,
 \begin{itemize}
+  \item $\Gamma$ --- the tape alphabet, a superset of the input alphabet $\Sigma$
+  \item $\blank : \Gamma$ --- a distinguished character that marks an empty cell on the tape
   \item $\T.\states$ --- a finite set of states
@@
         rejecting states, respectively, such that $\T.\acc \neq \T.\rej$
-  \item $\T.\delta : \T.\states \to \Gamma \to \T.\states \times \Gamma \times \Bool$
-        --- a deterministic transition function. Given the current state and the
-        head of the tape return the new state, the character written to the
-        tape, and which direction to shift ($\true$ for left, $\false$ for right)
+  \item $\T.\delta : \T.\states \times \Tape \times \mathbb{N} \to \T.\states \times \Tape \times \mathbb{N}$
+        --- a deterministic transition function. $\T.\delta$ reads the current state, the
+        current tape, and the index of the head of the tape and then returns the
+        updated state, tape, and head.
 \end{itemize}
-
-The tape of a
-(one-sided) Turing machine is encoded as nonlinear function from the natural numbers to the tape
-alphabet $\Gamma$, $\Tape = \mathbb{N} \to \Gamma$. To initialize the tape of
-$\T$, we define the parse transformer
-$\parseTape : \ltonl {(\stringg \lto \LinSigTy {\ttt} {\Tape} {\Turing~\ttt})}$ in \cref{fig:parseTape}.
-
-\steven{TODO finish the description of Turing machines}
 
@@
 \begin{lstlisting}
-data Turing : (t : Tape) (*@$\to$@*) L where
-  nil : (*@$\uparrow$@*)(&[ t (*@$\in$@*) Tape ] Turing t)
-  snoc : (*@$\uparrow$@*)(&[ c (*@$\in$@*) (*@$\Sigma$@*) ] &[ t (*@$\in$@*) Tape ](Turing t (*@$\lto$@*) c (*@$\lto$@*) Turing (push c t)))
+data Turing(*@$_{\T}$@*) : (t : Tape) (*@$\to$@*) L where
+  nil : (*@$\uparrow$@*)(&[ t (*@$\in$@*) Tape ] Turing(*@$_{\T}$@*) t)
+  snoc : (*@$\uparrow$@*)(&[ c (*@$\in$@*) (*@$\Sigma$@*) ] &[ t (*@$\in$@*) Tape ](Turing t (*@$\lto$@*) c (*@$\lto$@*) Turing(*@$_{\T}$@*) (push c t)))
 \end{lstlisting}
@@
 \begin{lstlisting}
-parseTape : (*@$\uparrow$@*)(String (*@$\lto$@*) (*@$\oplus$@*)[ t (*@$\in$@*) Tape ] Turing t)
+parseTape : (*@$\uparrow$@*)(String (*@$\lto$@*) (*@$\oplus$@*)[ t (*@$\in$@*) Tape ] Turing(*@$_{\T}$@*) t)
 parseTape String.nil = (*@$\sigma$@*) ((*@$\lambda$@*) n . blank) (Turing.nil ((*@$\lambda$@*) n . blank))
@@
 
-We may then define a term of type
-$\mathsf{parseTape} : \ltonl {\left( \stringg \lto \LinSigTy {\ttt} {\Tape} {\Turing~\ttt} \right)}$.
-In this manner, all of the interesting computations that the Turing machine
-would do occur solely inside of the nonlinear index. That is,
-$\mathsf{parseTape}$ does not discern which strings are accepted by the Turing
-machine, nor does it do any computation. $\mathsf{parseTape}$ simply loads the
-input string into the initial spaces on the tape.
-
-Once the string has been loaded onto the tape, we may further define the
-following term to run the Turing machine.
+The tape of a
+(one-sided) Turing machine is encoded as nonlinear function from the natural numbers to the tape
+alphabet $\Gamma$, $\Tape = \mathbb{N} \to \Gamma$.
+We define $\push~\c~\ttt : \Tape \to \Tape$ to increment the index of every
+character in the tape and insert $\c$ at index $0$. To run $\T$ on
+the input string $\w$, we write $\w$ to the leftmost entries of the tape and
+then repeatedly transition according to $\T.\delta$. If the machine enters the
+states $\T.\acc$ or $\T.\rej$, then the execution of $\T$ halts and either
+accepts or rejects accordingly.
+
+In \cref{fig:turinginductive}, we give an inductive linear type that describes
+Turing machines initialized with a given tape. However, unlike the types of
+traces for finite automata, the execution of a Turing machine is not described
+directly by the linear type system. We simply use the type $\Turing_{\T}$ to encode
+the initial state of the machine. To write the input string to the leftmost
+entries of the tape, we define the parse transformer
+$\parseTape : \ltonl {(\stringg \lto \LinSigTy {\ttt} {\Tape} {\Turing_{\T}~\ttt})}$
+in \cref{fig:parseTape}. Once the tape has been initialized, we capture the
+execution of $\T$ via the non-linear types.
+% data TuringTrace (b : Bool) : ⟨ Q ⟩ × Tape × Head → Type ℓ-zero where
+%   accept : ∀ t h → b ≡ true → TuringTrace b (acc , t , h)
+%   reject : ∀ t h → b ≡ false → TuringTrace b (rej , t , h)
+%   move : ∀ q t h →
+%     TuringTrace b (transition q t h) →
+%     TuringTrace b (q , t , h)
+\begin{figure}
+\begin{lstlisting}
+data Trace(*@$_{\T}$@*) : Bool (*@$\to$@*) T.states (*@$\times$@*) Tape (*@$\times$@*) (*@$\mathbb{N}$@*) (*@$\to$@*) U where
+  accept : (*@$\forall$@*) (t : Tape) (n : (*@$\mathbb{N}$@*)) (*@$\to$@*) Trace(*@$_{\T}$@*) true (T.acc , t , n)
+  reject : (*@$\forall$@*) (t : Tape) (n : (*@$\mathbb{N}$@*)) (*@$\to$@*) Trace(*@$_{\T}$@*) false (T.rej , t , n)
+  move : (*@$\forall$@*) (b : Bool) (s : T.states) (t : Tape) (n : (*@$\mathbb{N}$@*)) (*@$\to$@*)
+             Trace(*@$_{\T}$@*) b (T.(*@$\delta$@*) s t n) (*@$\to$@*) Trace(*@$_{\T}$@*) b (s , t , n)
+\end{lstlisting}
+\caption{Traces through a Turing machine as a non-linear inductive type}
+\label{fig:turingtrace}
+\end{figure}
+
+In \cref{fig:turingtrace} we define $\Trace_{\T}$ as a nonlinear type indexed by
+an acceptance Boolean and start configuration. A configuration of a Turing
+machine is the data of the start state, start tape, and start index for the head
+of tape. The
+traces of $\T$ either accept, reject, or transitions as described by $\T.\delta$.
+
+Once the input string has been loaded onto the tape, we execute $\T$ by defining
+a Turing machine interpreter of the following type,
+
 \[
-  \mathsf{run} : \ltonl {\left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \Turing(t) \lto \bigwith\limits_{\mathsf{fuel} : \mathbb{N}} \mathsf{Maybe} \left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \bigoplus\limits_{b : \Bool} \bigoplus\limits_{\mathsf{trace} : \mathsf{TuringTrace}(b , t)} \Turing(t) \right) \right)}
+  \texttt{run} : \ltonl {\left( \bigoplus\limits_{\ttt : \Tape} \Turing_{\T}~\ttt \lto \bigwith\limits_{\texttt{fuel} : \mathbb{N}} \texttt{Maybe} \left(  \bigoplus\limits_{\b : \Bool} \bigoplus\limits_{\s : \T.\states}  \bigoplus\limits_{\texttt{n} : \mathbb{N}} \bigoplus\limits_{\ttt : \Tape}  \bigoplus\limits_{\texttt{tr} : \Trace_{\T}~\b~(\s , \ttt , \texttt{n})} \Turing_{\T}~\ttt \right) \right)}
 \]
 
-where $\mathsf{TuringTrace}$ is a nonlinear type denoting terminating traces
-through the Turing machine. Note that $\mathsf{run}$ returns a $\mathsf{Maybe}$
-type in a fuel parameter. Turing machines may not terminate. If they do
-terminate within $\mathsf{fuel}$-many steps, then we return the trace along with
-the current tape and a boolean denoting if the trace is accepting or rejecting.
-If the trace does not terminate within the allotted amount of fuel, then we
-timeout and return nothing.
+$\mathtt{run}$ returns a result of $\mathtt{Maybe}$
+type varying over a fuel\footnote{Putting termination issues aside, we may omit
+  this fuel parameter. Although it is necessary when defining a Turing machine
+  interpreter in a total language like Agda.} parameter. It takes transitions at most
+$\texttt{fuel}$-many transitions in $\T$, and it returns a halting trace if one is
+found or timeouts otherwise. The data of this trace is encoded in the nonlinear
+index $\texttt{tr} : \Trace_{\T}~\b~(\s , \ttt , \texttt{n})$. Importantly, when
+$\texttt{tr}$ is an \emph{accepting} trace then the grammar
+\[
+\bigoplus\limits_{\s : \T.\states}  \bigoplus\limits_{\texttt{n} : \mathbb{N}} \bigoplus\limits_{\ttt : \Tape}  \bigoplus\limits_{\texttt{tr} : \Trace_{\T}~\true~(\s , \ttt , \texttt{n})} \Turing_{\T}~\ttt
+\]
+describes the strings that are accepted by $\T$. Therefore, we may encode
+arbitrary recusively enumerable languages within \theoryabbv.
 

### 2024-11-10 paper/paper.tex
font, gut Turing machine to be simpler reflection wip

--- 2024-11-08 paper/paper.tex
@@
 \usepackage{xspace}
+
+\usepackage[LGR,T1]{fontenc}
+\DeclareMathAlphabet{\mathgtt}{LGR}{cmtt}{m}{n}
+
+\renewcommand{\Gamma}{\mathgtt{G}}
+\renewcommand{\Delta}{\mathgtt{D}}
+
+\renewcommand{\familydefault}{\ttdefault}
+\usepackage{mathastext}
+\renewcommand{\familydefault}{\rmdefault}
 
@@
 \newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
-\newcommand{\theoryabbv}{$\textrm{Lambek}^D$}
+\newcommand{\theoryabbv}{$\textrm{Lambek}^D$~}
 \newcommand{\lnld}{$\textrm{LNL}_D$}
@@
 \newcommand{\Trace}{\texttt{Trace}}
-\newcommand{\stringquote}[1]{\texttt{\textquotesingle#1\textquotesingle}}
+\newcommand{\literal}[1]{\texttt{\textquotesingle#1\textquotesingle}}
+\newcommand{\stringquote}[1]{\texttt{\textquotedbl#1\textquotedbl}}
 \newcommand{\internalize}[1]{\lceil#1\rceil}
@@
 empty string, and the empty grammar.
-For instance, the grammar for the single character $\a$ has a single parse tree for the string
+For instance, the grammar corresponding to the character $a$ is given by $\literal a$ which
+has a single parse tree for the string
 \stringquote{a} and no parse trees at any other strings. The grammar $\I$ recognizes has a single
@@
 $\w$. For example, to define a parse tree for \stringquote{ab}, we use
-the context $\internalize{\stringquote{ab}} = a : \a , b : \b$. Then the left of
+the context $\internalize{\stringquote{ab}} = a : \literal a , b : \literal b$. Then the left of
 \cref{fig:fingram} gives a typing derivation that shows that $\inl(a
-\otimes b)$ defines a valid parse tree for the finite grammar $(\a \otimes \b)
-\oplus \c$. On the right of \cref{fig:fingram}, we give a linear $\lambda$-term
+\otimes b)$ defines a valid parse tree for the finite grammar
+$(\literal a \otimes \literal b)
+\oplus \literal c$. Underneath the derivation in \cref{fig:fingram}, we give a linear $\lambda$-term
 that implements the program given in the derivation.
@@
 \begin{figure}
-\begin{minipage}{0.5\textwidth}
 \begin{mathpar}
@@
       {~}
-      {a : \a \vdash a : \a}
+      {a : \literal a \vdash a : \literal a}
       \\
@@
       {~}
-      {b : \b \vdash b : \b}
+      {b : \literal b \vdash b : \literal b}
     }
-    {a : \a , b : \b \vdash a \otimes b : \a \otimes \b}
+    {a : \literal a , b : \literal b \vdash a \otimes b : \literal a \otimes
+      \literal b}
   }
-  {a : \a , b : \b \vdash \texttt{f} := \inl(a \otimes b) : (\a \otimes \b) \oplus \c}
+  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a \otimes b) :
+    (\literal a \otimes \literal b) \oplus \literal c}
 \end{mathpar}
-\end{minipage}%
-\begin{minipage}{0.5\textwidth}
 \begin{lstlisting}
-f : (*@$\uparrow$@*)(a (*@$\otimes$@*) b (*@$\lto$@*) (a (*@$\otimes$@*) b) (*@$\oplus$@*) c)
-f (a' , b') = inl (a' , b')
+f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
+f (a , b) = inl (a (*@$\otimes$@*) b)
 \end{lstlisting}
-\end{minipage}
-\caption{\stringquote{ac} matches $(a \oplus b) \otimes c$}
+\caption{\stringquote{ac} matches
+  $(\literal a \oplus \literal b) \otimes \literal c$}
 \label{fig:fingram}
@@
 \[
-  a : \a , b : \b \not \vdash a : \a \qquad a : \a , b : \b \not \vdash (a, a) : \a \otimes \a
+  a : \literal a , b : \literal b \not \vdash a : \literal a \qquad a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a
 \]
@@
 \[
-  a : \a , b : \b \not \vdash (b , a) : \b \otimes \a
+  a : \literal a , b : \literal b \not \vdash (b , a) : \literal b \otimes \literal a
 \]
@@
 Through repeated application of the Kleene star constructors, \cref{fig:kleenestarderivation} gives a derivation that shows \stringquote{ab} matches the
-regular expression $(\a^{*} \otimes \b) \oplus \c$. The leaves of the proof tree
+regular expression $({\literal a}^{*} \otimes \literal b) \oplus \literal c$. The leaves of the proof tree
 that mention the arrow $\uparrow$ describe a cast from a non-linear type to a linear type.
 For instance, the premise of the leaf involving $\nil$
-views $\nil : \ltonl {(\a^{*})}$ as the name of a
+views $\nil : \ltonl {({\literal a}^{*})}$ as the name of a
 constructor, and a constructor should be nonlinearly valued because we may call
-it several time. However, the conclusion of this leaf views $\nil : \a^{*}$ as a
+it several time. However, the conclusion of this leaf views $\nil : {\literal a}^{*}$ as a
 linear value. After we call the constructor it returns a value that may only be
@@
 
+\steven{TODO fig 3 sizing}
 \begin{figure}
 \begin{mathpar}
+\footnotesize
 \inferrule
@@
           {~}
-          {\cdot \vdash \cons : \ltonl {(\a \lto \a^* \lto \a^*)}}
+          {\cdot \vdash \cons : \ltonl {({\literal a} \lto {\literal a}^* \lto {\literal a}^*)}}
         }
-        {\cdot \vdash \cons : \a \lto \a^* \lto \a^*}
+        {\cdot \vdash \cons : {\literal a} \lto {\literal a}^* \lto {\literal a}^*}
         \\
@@
         {~}
-        {a : \a \vdash a : \a}
+        {a : {\literal a} \vdash a : {\literal a}}
       }
-      {a : a \vdash \cons~a : \a^* \lto \a^*}
+      {a : a \vdash \cons~a : {\literal a}^* \lto {\literal a}^*}
       \\
@@
         {~}
-        {\cdot \vdash \nil : \ltonl {(\a^*)}}
+        {\cdot \vdash \nil : \ltonl {({\literal a}^*)}}
       }
-      {\cdot \vdash \nil : \a^*}
+      {\cdot \vdash \nil : {\literal a}^*}
     }
-    {a : \a \vdash \cons~a~\nil : \a^{*}}
+    {a : {\literal a} \vdash \cons~a~\nil : {\literal a}^{*}}
     \\
@@
     {~}
-    {b : \b \vdash b : \b}
+    {b : {\literal b} \vdash b : {\literal b}}
   }
-  {a : \a , b : \b \vdash (\cons~a~\nil) \otimes b : \a^{*} \otimes \b}
+  {a : {\literal a} , b : {\literal b} \vdash (\cons~a~\nil) \otimes b : {\literal a}^{*} \otimes {\literal b}}
 }
-{a : \a , b : \b \vdash \texttt{g} := \inl ((\cons~a~\nil) \otimes b) : (\a^{*} \otimes \b) \oplus \c}
+{a : \literal a , b : \literal b \vdash \texttt{g} := \inl ((\cons~a~\nil)
+  \otimes b) : ({\literal a}^{*} \otimes \literal b) \oplus \literal c}
 \end{mathpar}
 \begin{lstlisting}
-g : (*@$\uparrow$@*)((a (*@$\otimes$@*) b) (*@$\lto$@*) (a(*@$^*$@*) (*@$\otimes$@*) b) (*@$\oplus$@*) c)
-g (a' , b') = inl (cons a' nil (*@$\otimes$@*) b')
+g : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*) ('a'(*@$^*$@*) (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
+g (a , b) = inl (cons a' nil (*@$\otimes$@*) b')
 \end{lstlisting}
-\caption{\stringquote{ab} matches $(a^{*} \otimes b) \oplus c$}
+\caption{\stringquote{ab} matches
+  $({\literal a}^{*} \otimes \literal b) \oplus \literal c$}
 \label{fig:kleenestarderivation}
@@
 string of base types. In \cref{fig:kleeneabstractproof} we show that every parse
-of the grammar $(\A \otimes \A)^{*}$ induces a parse of $\A^{*}$ for an
-arbitrary grammar $\A$. The context $(\A \otimes \A)^{*}$ does not correspond directly to a string, so it is
+of the grammar $(A \otimes A)^{*}$ induces a parse of $A^{*}$ for an
+arbitrary grammar $A$. The context $(A \otimes A)^{*}$ does not correspond directly to a string, so it is
 not quite appropriate
@@
 The
-context $a : (\A \otimes \A)^{*}$ does not contain concrete data to be parsed; rather, there may be many choices of string underlying the parse tree captured
+context $a : (A \otimes A)^{*}$ does not contain concrete data to be parsed; rather, there may be many choices of string underlying the parse tree captured
 by the variable $a$. Thus, the term $\texttt{h}$ from
@@
 
-$\texttt{h}$ is defined by recursion on terms of type $(\A \otimes \A)^{*}$.
+We define $\texttt{h}$ by recursion on terms of type $(\A \otimes \A)^{*}$.
 This recursion is expressed in the derivation tree by invoking the
 elimination principle for Kleene star, written
-as $\fold$. $\texttt{h}$ is more compactly presented in the pseudocode
+as $\fold$. The parse transformer $\texttt{h}$ is more compactly presented in the pseudocode
 of \cref{fig:kleeneabstractproof} by pattern matching on the input and making an
@@
 h nil = nil
-h (cons (a (*@$\otimes$@*) a') as) = cons a (cons a' (h as))
+h (cons (a1 (*@$\otimes$@*) a2) as) = cons a1 (cons a2 (h as))
 \end{lstlisting}
@@
 however simply defining a regular expression --- such as
-$(\a^{*} \otimes \b) \oplus \c$ --- does not give any insight into how to parse
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ --- does not give any insight into how to parse
 it. Many parsing algorithms are based on a notion of automaton, and to verify
@@
 expressions. \cref{fig:exampleNFA} shows a non-deterministic
-finite automaton (NFA) for the regular expression $(\a^{*} \otimes b) \oplus c$.
+finite automaton (NFA) for the regular expression
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$.
 \cref{fig:nfainductive} defines $\Trace$ an \emph{indexed} inductive linear type of
@@
 Just as \cref{fig:kleenestarderivation} shows the \stringquote{ab} matches
-$(a^{*} \otimes b) \oplus c$, \cref{fig:nfaderivation} demonstrates that \stringquote{ab} is
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$, \cref{fig:nfaderivation} demonstrates that \stringquote{ab} is
 accepted by the NFA representation of this regular expression. In particular, we
@@
 initial state of the NFA.\ We may imagine defining parse transformers from
-$(\a^{*} \otimes b) \oplus c$ to $\Trace~\1$, and vice versa. The existence of
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ to $\Trace~\1$, and vice versa. The existence of
 such maps between the two types would demonstrate that the regular expression
@@
   stop : (*@$\uparrow$@*)(Trace 2)
-  1to1 : (*@$\uparrow$@*)(a (*@$\lto$@*) Trace 1 (*@$\lto$@*) Trace 1)
-  1to2 : (*@$\uparrow$@*)(b (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 1)
-  0to2 : (*@$\uparrow$@*)(c (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 0)
+  1to1 : (*@$\uparrow$@*)('a' (*@$\lto$@*) Trace 1 (*@$\lto$@*) Trace 1)
+  1to2 : (*@$\uparrow$@*)('b' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 1)
+  0to2 : (*@$\uparrow$@*)('c' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 0)
   0to1 : (*@$\uparrow$@*)(Trace 1 (*@$\lto$@*) Trace 0)
@@
 \begin{lstlisting}
-k : (*@$\uparrow$@*)((a (*@$\otimes$@*) b) (*@$\lto$@*)  Trace 0)
-k (a' , b') = 0to1 (1to1 a' (1to2 b' stop))
+k : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*)  Trace 0)
+k (a , b) = 0to1 (1to1 a (1to2 b stop))
 \end{lstlisting}
-\caption{\stringquote{ab} is accepted by the NFA for $(a^{*} \otimes b) \oplus c$}
+\caption{\stringquote{ab} is accepted by the NFA for
+  $({\literal a}^{*} \otimes {\literal b }) \oplus \literal c$}
 \label{fig:nfaderivation}
@@
 %% \end{figure}
-
 \begin{figure}
@@
 is strongly equivalent to an unambiguous grammar $\B$. In this style of proof, we
-build the parse transformers $\f : \ltonl{(\A \lto \B)}$ and
+build the functions $\f : \ltonl{(\A \lto \B)}$ and
 $\g : \ltonl{(\B \lto \A)}$, then we have the obligation to show
 $\f \circ \g \equiv \ident$ and $\g \circ \f \equiv \ident$. However, because $\B$ is
-unambiguous there is only one parse transformer $\ltonl{(\B \lto \B)}$, so
+unambiguous there is only one function of type $\ltonl{(\B \lto \B)}$, so
 $\f \circ \g : \ltonl{(\B \lto \B)}$ and $\ident : \ltonl{(\B \lto \B)}$ are equal.
@@
 one particular NFA corresponding to the regular expression
-$(a^{*} \otimes b) \oplus c$. The definition of the linear type of traces
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$. The definition of the linear type of traces
 through this
@@
   cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.transitions ] &[ b (*@$\in$@*) Bool ]
-           (N.label(t) (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.dst(t) , b) (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.src(t) , b)))
+           ('N.label(t)' (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.dst(t) , b) (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.src(t) , b)))
   (*@$\epsilon$@*)cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.(*@$\epsilon$@*)transitions ] &[ b (*@$\in$@*) Bool ]
@@
 \item $\D.\delta : \Sigma \to \D.\states \to \D.\states$ --- a deterministic transition function. Given a character $\c$ and a state
-        $\s$, $\D.\delta~\c~\s$ is the state for which there is a transition $\s \overset{\c}{\to} (\D.\delta~\c~\s)$
+        $\s$, $\D.\delta~\c~\s$ is the state for which there is a transition $\s \overset{\stringquote{c}}{\to} (\D.\delta~\c~\s)$
 \end{itemize}
@@
   cons : (*@$\uparrow$@*)(&[ c (*@$\in$@*) (*@$\Sigma$@*) ] &[ s (*@$\in$@*) D.states ] &[ b (*@$\in$@*) Bool ]
-             (c (*@$\lto$@*) Trace(*@$_{\D}$@*)(D.(*@$\delta$@*) c s , b) (*@$\lto$@*) Trace(*@$_{\D}$@*)(s , b)))
+             ('c' (*@$\lto$@*) Trace(*@$_{\D}$@*)(D.(*@$\delta$@*) c s , b) (*@$\lto$@*) Trace(*@$_{\D}$@*)(s , b)))
 \end{lstlisting}
@@
 
-$\Trace_{\D}$ --- the linear type of traces through $\D$ --- is given in
-\cref{fig:dfatrace}, and its is
-quite similar to the definition of $\Trace_{\N}$. These
+$\Trace_{\D}$, the linear type of traces through $\D$, is given in
+\cref{fig:dfatrace}. These
 traces may be terminated in an accepting state $\s$ with the $\nil$ constructor.
@@
 trace out of the state $\D.\delta~\c~\s$. The trace built with $\cons$ is
-accepting if and only if the trace out of $D.\delta_{c}(s)$ is accepting.
+accepting if and only if the trace out of $D.\delta~c~s$ is accepting.
 
@@
 are unambiguous and define a parser for them. In particular we show that
-$\LinPiTy {s} {D.\states} {\LinSigTy {b} {\Bool} \Trace_{D}(s , b)}$ is isomorphic to
+$\LinPiTy {s} {D.\states} {\LinSigTy {b} {\Bool} \Trace_{D}(s , b)}$ is strongly
+equivalent to
 $\stringg$, which simultaneously constructs a parser and demonstrates that
-traces through $D$ is unambiguous. This isomorphism intuitively says that we pay
+traces through $D$ is unambiguous. Intuitively this strong equivalence says that we may
 unambiguously build traces through $D$ beginning at any state $s$, and further
@@
 
-The first half of the isomorphism,
+The first half of the equivalence,
 $\parse_{D} : \ltonl {\left( \stringg \lto \LinPiTy {s} {D.\states} {\Trace_{D}(s)} \right)}$,
@@
 
-The second half of the isomorphism,
+The second half of the equivalence,
 $\print_{D} : \ltonl {\left( \LinPiTy {s} {D.\states} {\Trace_{D}(s)} \lto \stringg \right)}$,
@@
 print : (s : D.states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*)(s , b)) (*@$\lto$@*) String)
-print .s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).nil s)) = String.nil
-print .s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c (D.(*@$\delta$@*) c s) b a trace)) =
+print s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).nil .s)) = String.nil
+print s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c (D.(*@$\delta$@*) c .s) b a trace)) =
   String.cons ((*@$\sigma$@*) c a) (print (D.(*@$\delta$@*) c s) ((*@$\sigma$@*) b trace))
@@
 NtoD .s (nil s) X sInX = nil s
-NtoD .(N.src trans) (N.cons trans .true a dstTrace) X sInX =
-  D.cons (N.label trans) X true a
-    (((NtoD s dstTrace).(*@$\pi$@*) (D.(*@$\delta$@*) (N.label trans) X)).(*@$\pi$@*) (N.dst trans))
-NtoD .(N.(*@$\epsilon$@*)src trans) ((*@$\epsilon$@*)cons (*@$\epsilon$@*)trans .true (*@$\epsilon$@*)DstTrace) X sInX =
-  ((NtoD s (*@$\epsilon$@*)DstTrace).(*@$\pi$@*) X).(*@$\pi$@*) (N.(*@$\epsilon$@*)dst trans)
+NtoD .(N.src t) (N.cons trans .true a dstTrace) X sInX =
+  D.cons (N.label t) X true a
+    (((NtoD s dstTrace).(*@$\pi$@*) (D.(*@$\delta$@*) (N.label t) X)).(*@$\pi$@*) (N.dst t))
+NtoD .(N.(*@$\epsilon$@*)src t) ((*@$\epsilon$@*)cons (*@$\epsilon$@*)trans .true (*@$\epsilon$@*)DstTrace) X sInX =
+  ((NtoD s (*@$\epsilon$@*)DstTrace).(*@$\pi$@*) X).(*@$\pi$@*) (N.(*@$\epsilon$@*)dst t)
 \end{lstlisting}
@@
 Next if the trace in $\D$ is defined via transition
-$\X \overset{\c}{\to} (\D.\delta~\c~\X)$ then we may choose some state in
+$\X \overset{\stringquote{c}}{\to} (\D.\delta~\c~\X)$ then we may choose some state in
 $\s \in \D.\delta~\c~\X$. Recursively, $\DtoN$ sends the trace rooted at
@@
 data Dyck : L where
-  nil : (*@$\uparrow$@*)(Dyck)
+  nil : (*@$\uparrow$@*) Dyck
   balanced : (*@$\uparrow$@*)('(' (*@$\lto$@*) Dyck (*@$\lto$@*) ')' (*@$\lto$@*) Dyck (*@$\lto$@*) Dyck)
@@
 
-\subsection{Turing Machines}
-\newcommand{\Turing}{\mathtt{Turing}}
-\newcommand{\ttt}{\mathtt{t}}
-\newcommand{\T}{\mathtt{T}}
-\newcommand{\acc}{\mathtt{acc}}
-\newcommand{\rej}{\mathtt{rej}}
-\newcommand{\Tape}{\mathtt{Tape}}
-\newcommand{\parseTape}{\mathtt{parseTape}}
-\newcommand{\snoc}{\mathtt{snoc}}
-\steven{This subsection is still a WIP}
-%% 5. Expressiveness: Turing Machines
-The expressive power of \theoryabbv spans the entire Chomsky hierarchy, as we may
-encode the accepting traces of Turing machines as linear types.
-
-\steven{TODO introduce tape alphabet and blank symbol}
-
-The data of a Turing machine $\T$  is given by the following data,
-\begin{itemize}
-  \item $\T.\states$ --- a finite set of states
-  \item $\T.\init : \T.\states$ --- the initial state
-  \item $\T.\acc , \T.\rej : \T.\states$ --- distinguished accepting and
-        rejecting states, respectively, such that $\T.\acc \neq \T.\rej$
-  \item $\T.\delta : \T.\states \to \Gamma \to \T.\states \times \Gamma \times \Bool$
-        --- a deterministic transition function. Given the current state and the
-        head of the tape return the new state, the character written to the
-        tape, and which direction to shift ($\true$ for left, $\false$ for right)
-\end{itemize}
-
-The tape of a
-(one-sided) Turing machine is encoded as nonlinear function from the natural numbers to the tape
-alphabet $\Gamma$, $\Tape = \mathbb{N} \to \Gamma$. To initialize the tape of
-$\T$, we define the parse transformer
-$\parseTape : \ltonl {(\stringg \lto \LinSigTy {\ttt} {\Tape} {\Turing~\ttt})}$ in \cref{fig:parseTape}.
-
-\steven{TODO finish the description of Turing machines}
-
-\begin{figure}
-\begin{lstlisting}
-data Turing : (t : Tape) (*@$\to$@*) L where
-  nil : (*@$\uparrow$@*)(&[ t (*@$\in$@*) Tape ] Turing t)
-  snoc : (*@$\uparrow$@*)(&[ c (*@$\in$@*) (*@$\Sigma$@*) ] &[ t (*@$\in$@*) Tape ](Turing t (*@$\lto$@*) c (*@$\lto$@*) Turing (push c t)))
-\end{lstlisting}
-\caption{Inductive linear type of Turing machines initialized with tape $\ttt$}
-\label{fig:turinginductive}
-\end{figure}
-
-\begin{figure}
-\begin{lstlisting}
-parseTape : (*@$\uparrow$@*)(String (*@$\lto$@*) (*@$\oplus$@*)[ t (*@$\in$@*) Tape ] Turing t)
-parseTape String.nil = (*@$\sigma$@*) ((*@$\lambda$@*) n . blank) (Turing.nil ((*@$\lambda$@*) n . blank))
-parseTape (String.cons ((*@$\sigma$@*) c a) w) =
-  let (*@$\sigma$@*) t Tur = parseTape w in (*@$\sigma$@*) (push c t) (snoc c t Tur a)
-\end{lstlisting}
-\caption{Initialization of the tape of Turing machine as a parse transformer}
-\label{fig:parseTape}
-\end{figure}
-
-We may then define a term of type
-$\mathsf{parseTape} : \ltonl {\left( \stringg \lto \LinSigTy {\ttt} {\Tape} {\Turing~\ttt} \right)}$.
-In this manner, all of the interesting computations that the Turing machine
-would do occur solely inside of the nonlinear index. That is,
-$\mathsf{parseTape}$ does not discern which strings are accepted by the Turing
-machine, nor does it do any computation. $\mathsf{parseTape}$ simply loads the
-input string into the initial spaces on the tape.
-
-Once the string has been loaded onto the tape, we may further define the
-following term to run the Turing machine.
+\subsection{Unrestricted Grammars}
+% \newcommand{\P}{\texttt{P}}
+\steven{Need to better introduce nonlinear type of strings. We haven't talked
+  about semantics nor have we included a nonlinear type of strings in the syntax}
+\newcommand{\stringNL}{String'}
+Let $\stringNL$ represent a nonlinear type of strings and let
+$P : \stringNL \to U$
+\steven{Or should $P$ be a semantic grammar?}
+
+The type of $P$ very nearly describes a formal grammar in \theoryabbv. That is,
+$P$ describes a mapping from strings to sets of parse trees; however, even
+though $P$ captures the same grammatical data $P$ is not
+a first-class linear type. We reflect this nonlinear notion of grammar as a
+linear type,
+
 \[
-  \mathsf{run} : \ltonl {\left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \Turing(t) \lto \bigwith\limits_{\mathsf{fuel} : \mathbb{N}} \mathsf{Maybe} \left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \bigoplus\limits_{b : \Bool} \bigoplus\limits_{\mathsf{trace} : \mathsf{TuringTrace}(b , t)} \Turing(t) \right) \right)}
+Reflect~P = \LinSigTyLimit {w} {\stringNL} {\LinSigTyLimit {x} {P~w} {~\lceil w \rceil}}
 \]
-
-where $\mathsf{TuringTrace}$ is a nonlinear type denoting terminating traces
-through the Turing machine. Note that $\mathsf{run}$ returns a $\mathsf{Maybe}$
-type in a fuel parameter. Turing machines may not terminate. If they do
-terminate within $\mathsf{fuel}$-many steps, then we return the trace along with
-the current tape and a boolean denoting if the trace is accepting or rejecting.
-If the trace does not terminate within the allotted amount of fuel, then we
-timeout and return nothing.
 

### 2024-11-10 paper/paper.tex
author order
generalize adamek

--- 2024-11-08 paper/paper.tex
@@
 
+\author{Pedro H. Azevedo de Amorim}
+\affiliation{
+  \department{Department of Computer Science}
+  \institution{University of Oxford}
+  \country{UK}
+}
+\email{pedro.azevedo.de.amorim@cs.ox.ac.uk}
+
 \author{Max S. New}
@@
 \email{maxsnew@umich.edu}
-
-\author{Pedro H. Azevedo de Amorim}
-\affiliation{
-  \department{Department of Computer Science}
-  \institution{University of Oxford}
-  \country{UK}
-}
-\email{pedro.azevedo.de.amorim@cs.ox.ac.uk}
 
@@
 \begin{theorem}[Adamek]
-  If a functor $F : \mathcal C \to \mathcal C$ preserves
-  $\omega$-colimits and $\mathcal C$ has an initial object and
-  $\omega$-colimits, then $F$ has an initial algebra given by the
+  If a functor $F : \mathcal C \to \mathcal C$ preserves all 
+  $\kappa$-colimits and $\mathcal C$ has an initial object and
+  $\kappa$-colimits, then $F$ has an initial algebra given by the
   colimit of the chain

### 2024-11-11 paper/paper.tex
draft of section 1,4. Removed dead code from comments

--- 2024-11-10 paper/paper.tex
@@
 \renewcommand{\Delta}{\mathgtt{D}}
+\renewcommand{\Sigma}{\mathgtt{S}}
+\renewcommand{\mu}{\mathgtt{m}}
 
@@
 
-\newcommand{\SPF}{\mathsf{SPF}}
-\newcommand{\Var}{\mathsf{Var}}
-\newcommand{\K}{\mathsf{K}}
-\newcommand{\map}{\mathsf{map}}
-\newcommand{\roll}{\mathsf{roll}}
-\newcommand{\fold}{\mathsf{fold}}
-\newcommand{\inl}{\mathsf{inl}}
-\newcommand{\inr}{\mathsf{inr}}
+\newcommand{\SPF}{\texttt{SPF}}
+\newcommand{\Var}{\texttt{Var}}
+\newcommand{\K}{\texttt{K}}
+\newcommand{\map}{\texttt{map}}
+\newcommand{\roll}{\texttt{roll}}
+\newcommand{\fold}{\texttt{fold}}
+\newcommand{\inl}{\texttt{inl}}
+\newcommand{\inr}{\texttt{inr}}
 \newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
@@
 \newcommand{\String}{\textbf{String}}
-\newcommand{\stringg}{\mathtt{String}}
+\newcommand{\stringg}{\texttt{String}}
 \newcommand{\charg}{\mathtt{Char}}
@@
   embedding in Agda.
-
-  Our shallow embedding is based on a denotational semantics of \theoryabbv
-  where non-linear types denote ordinary sets and linear types denote formal
-  grammars. To develop this semantics compositionally, we define a notion of
-  general notion of model we call a \emph{Chomsky hyperdoctrine}, a
-  generalization of the Kleene algebra approach to formal language theory. In
-  addition to our ``standard model'' where linear types denote formal grammars,
-  we demonstrate two non-standard models. The first interprets linear types as
-  semantic actions over formal grammars, suggesting future extensions of our
-  calculus. The second is used to structure a logical relations argument to
-  prove a canonicity metatheorem for the type theory, proving that all closed
-  terms of context-free linear types normalize to a parse tree.
 \end{abstract}
@@
 \label{sec:intro}
-
-\steven{TODO adapt section 1 and the abstract to reflect the rest of the paper}
-%% Outline
-%% - Why a framework for parser verification
-%% - what intrinsic verification gets us
-%% - the secret sauce: indexed linear inductive types
-%% - prototype implementation
-%% - overview of "abstract" formal grammars?
-
-Parsing data from a serialized format is one of the most common tasks
-in all of computing. Accordingly, this is a quite well-studied problem
-spanning theoretical computer science and linguistics with such
-milestones as Chomsky's hierarchy of grammars
-\cite{chomThreeModels1956}, practical algorithms for parsing of
-regular and context-free grammars, and variants
-\cite{KNUTH1965607,Earley1970}, as well as implementations of
-practical tools for the generation of efficient parsers
-\cite{Johnsonyacc}.
-
-The theory of formal languages and parsing is one of the oldest and most
-thoroughly developed area of theoretical computer science.
-The central object of study is a \emph{formal language} $L$ over an alphabet
-$\Sigma$, which is classically defined as a
-\emph{subset} of strings $L \subseteq \String = \Sigma^{*}$. The expressivity of formal
-languages is sufficient for defining a \emph{recognizer} of a language $L$:
-a function $r : \String \to \Prop$ such that $r$ maps each string $w$ to the
-proposition that $w$ is in the language $w \in L$. This technique of language
-recognition can concisely specify lexical analysis, and even lead to verified
-lexer implementations \cite{egolfVerbatim,Ouedraogo_2023}.
-
-While there are ample theory and tooling to handle the first stage of a language frontend, the final
-frontier for parsing research is the development of
-\emph{formally verified} parser implementations. Due to the
-ubiquitous nature of parsing, these would useful components of many
-formally verified software systems: compilers and servers most
-especially. However, language recognition is insufficient for specifying a
-parser, and for this task much of the structured foundations from the theory of
-formal languages is rendered moot. Rather than returning a proposition-valued result, a parser produces a
-structured \emph{parse tree} adherent to a \emph{formal grammar}.
-The research on such grammars is often performed atop an ad-hoc choice of grammar
-formalism, and results are proven with respect to that formalism. Critically,
-the field is lacking a unifying theory across these different formalisms, even
-though many of the same
-compositional reason principles appear across distinct grammar formalisms.
-
-In this work, we aim to elucidate these similarities and
-provide a mathematical theory of formal grammars for constructing verified
-parsers. Our approach can be summarized with the
-Curry-Howard-style slogan: ``grammars as types''; and the addendum: ``parsers as
-terms''. That is, we develop a type theory with a novel
-denotational semantics in a category with abstract formal grammars as
-objects and \emph{parse transformers} as morphisms. We say these are
-\emph{abstract} formal grammars because they are not tied to any
-particular syntactic formalism such as Chomskyan grammars, regular
-expressions, Lambek calculus etc. The category we choose is a rich
-in structure: it is a topos as well as being monoidal closed, and these
-universal properties serve as the inspiration for our type theory.
-
-The type theory itself is a form of dependent linear-non-linear type
+\steven{This intro sounds reallchoppy but follows the following outline.
+
+  \begin{itemize}
+  \item parsing correctness is imporant because parsing is
+  everywhere and failure
+          could even invalidate other proof efforts.
+  \item verified parsers should be
+          easier to write
+  \item Conal Elliott gives the right idea for what the foundation
+          of parser verification should be --- type theory
+
+  \end{itemize}
+          }
+Formal verification encompasses a broad suite of techniques one may use to
+mathematically guarantee the correctness of software systems. We can only
+reap the benefits of
+any proof effort if its underlying assumptions hold. For instance, verifying the
+functional correctness of an algorithm is no use if it is run on buggy hardware.
+Likewise, the promised guarantees of a program may be jeopardized if it is
+parsed incorrectly.
+
+Large verified developments, like CompCert \cite{leroy_formal_2009},
+invest enormous amounts of person-hours into proof development. In early
+versions of CompCert, the apparent
+assurances associated with their effort could indeed be invalidated through
+parser bugs \cite{yangFindingUnderstandingBugs}. Parsing is a ubiquitous task
+throughout computing and yet parsers are frequently unverified, even in
+otherwise verified systems. The CompCert team later
+developed a method for validating the correctness of LR parsers
+\cite{jourdanValidatingLRParsers2012} and then replaced their parser with one
+that is certified to be correct. Despite their success in  validating their C parser and other work in the
+generation of  verified parsers
+\cite{lasserCoStarVerifiedALL2021,firsovCertifiedNormalizationContextFree2015,danielssonTotalParserCombinators2010},
+it is still arduous and highly nontrivial to write and verify a
+parser. Further, distinct approaches to generating verified parsing
+implementations are often relative to a different grammar formalism. We believe
+that these sepearate grammar formalisms can all be unified into a
+syntax-independent notion of grammar, and that this notion of grammar can be
+axiomatized into a syntax that is useful for building verified parsers.
+
+\steven{I don't like how I make the cited work sound bad above}
+
+Elliott \cite{elliottSymbolicAutomaticDifferentiation2021}
+notes that grammars can be described as type-level predicates on strings that prove
+language membership. That is, a \emph{formal grammar} $A$ is a function
+$\String \to \Set$ such that for a string $w$, $A~w$ is the set of proofs
+showing that $w$ belongs to the language recognized by $A$. We adopt this
+definition of a formal grammar.
+
+In this paper, we present \theoryname (\theoryabbv) as a domain specific
+type theory for building verified parsers. The type theory itself is a form of dependent linear-non-linear type
 theory: the linear types can be interpreted as grammars, whereas the
@@
 
-\paragraph{Contributions} This paper begins in \cref{sec:synindnotion} by
-giving a universal semantic characterization of formal grammar.
+\paragraph{Contributions} This paper begins in \cref{sec:type-theory-examples}
+by giving small examples of programs in \theoryabbv so that we may build intuition.
 From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type theory
 that axiomatizes the underlying structures needed to conduct formal grammar
-theory. In \cref{sec:automata} we demonstrate the usage of
-\theoryname for defining parser terms and internalizing results from formal language theory.
-Then in \cref{sec:categorify}, we characterize the abstract categorical
-models of our syntax. Afterwards in \cref{sec:othermodels}, we demonstrate several models,
-including one to prove a metatheoretic canonicity theorem.
+theory. In \cref{sec:applicatoins} we demonstrate the
+applicablity of \theoryabbv for defining parser terms and representing results
+from formal language theory as proof terms in our calculus.
+Then in \cref{sec:semantics-and-metatheory}, we give a denotational semantics
+that makes precise the connection between \theoryabbv syntax and formal grammars.
+
 Our contributions are then:
@@
 \begin{itemize}
-  \item A syntax-independent notion of grammar given by the category $\Grammar$
-  \item \theoryname: A dependent
-    linear-non-linear type theory meant to syntactically capture the latent structures
-    in $\Grammar$ that enable parsing
-  \item Categorical structures that capture the concepts integral to formal grammar theory, and that are notably modeled by $\Grammar$
-  \item An implementation of the language, which is embedded shallowly in Agda
+  \item \theoryname (\theoryabbv): A dependent
+        linear-non-linear type theory for building verified parsers
+  \item An implementation of \theoryabbv, which is embedded shallowly in Cubical
+        Agda. Throughout this paper, we will use \agdalogo~ to mark results that
+        have been mechanized in our Agda development.
 \end{itemize}
-
-\subsection{A Syntax-Independent Notion of Syntax}
-\label{sec:synindnotion}
-
-The notion of formal language is central to the theory of parsing.
-This definition is
-especially useful as it gives a semantics to formal grammars that is completely
-independent of any particular syntactic grammar formalism. Any new
-notion of grammar can be given a language semantics and it provides a precise
-mathematical specification for implementing a \emph{recognizer} of a language.
-In this section we propose a \emph{proof-relevant} generalization of formal
-languages that can accommodate richer aspects of formal language theory such as
-parsing. We begin by proposing a presheaf approach to grammars that encompasses
-other classes of grammars in a syntax-independent way.
-
-\paragraph{From Formal Languages to Formal Grammars}
-
-Our first contribution is a novel and syntax-independent characterization of
-formal grammar:
-
-\begin{definition}
-  \label{def:grammar}
-  A \emph{formal grammar} $G$ over a fixed alphabet $\Sigma$ is a function
-  $G : \String \to \Set$.
-\end{definition}
-
-\steven{TODO need to go through and ensure that $\Sigma^{*}$ and $\String$ are each used appropriately}
-
-We say that a grammar $G$ associates to every string $w$ the set\footnote{We
-  make little comment as to which foundations are used to encode $\Set$. Our
-  construction is polymorphic in choice of a proper class of small sets, a
-  Grothendieck universe, a type theoretic universe, or any similar foundation.} $G w$ of parse
-trees showing that $w$ matches $G$. This definition of formal grammar serves as
-a mathematical specification for a parser --- taking on the same role as
-languages do for recognizers. In this sense, our new characterization of
-grammars is just a proof-relevant generalization of the existing notion of
-formal language.
-
-By ranging over all input strings, notice that each grammar induces a language.
-For each formal grammar $G$, define the language of accepted strings $L_{G}$ as,
-
-\[ L_{G} = \{ w \in \String ~|~ G w~\text{is inhabited} \} \]
-
-As they are simply sets, formal languages are naturally endowed with a partial
-order via subset inclusion. Just, formal grammars naturally coalesce into a
-\emph{category}.
-
-Define $\Grammar$ to be the category of formal grammars. Whose
-objects are grammars, and whose morphisms are functions $f : G \to H$ between two
-grammars presented by a family of functions $f^{w} : G w \to H w$ for
-$w : \String$. Intuitively, we interpret these morphisms as parse
-\emph{transformers} --- as constructive translations of $G$-parses into $H$-parses. Throughout this paper, we will make frequent reference to this categorical
-structure on $\Grammar$.
-
-\steven{Above sounds awkward. I want to additionally say something here that signposts that we will give more
-  information about that structure without just listing it all here. }
-
-\paragraph{From Grammars to Formal Grammars}
-The category-theoretic framework can be further used to describe and compare
-different notions of formal grammar. For instance, Chomskyan generative
-grammars, semi-Thue systems, Montague grammars, and so on are all distinct syntactic
-presentations of the same underlying idea of an abstract specification for
-parsing. Each of these is an instance of a general \emph{notion of formal
-  grammar} --- a category paired with a functor into
-$\Grammar$. That is, they may all be interpreted in our category of grammars. In
-this sense, $\Grammar$ provides an
-all-encompassing semantic domain for parsing.
-
-Upon further inspection, $\Grammar$ may equivalently
-be described as the functor category $\Set^{\String}$ --- where the
-set $\String$ is viewed as a discrete\footnote{Objects in $\String$
-comprise the set of all strings, and the only morphisms are the
-identity morphisms.} category. In fact, in this paper we will take
-$\Grammar$ to be defined as the functor category $\Set^{\String}$. Such functor
-categories in $\Set$, called presheaf categories, carry a
-remarkable amount of structure.  Indeed they are a monoidal topos and,
-as such, are a model to a substructural dependently typed theory.
-
-Thus we
-present \theoryname (\theoryabbv) as a domain-specific type theory that
-abstracts over the structure of $\Grammar$. To begin, let us look at what some
-programs in \theoryabbv look like.
-
-\steven{TODO Reword this transition into the examples}
-
-% \max{this section kind of trails off at the end with a very vague aside about a $2$-category of grammar formalisms.
-Maybe this should go in future work/discussion?
-What is the message of this section?
-How does it tie in with the message of the paper?
-What even is the message of this paper?}
-
-% \pedro{Would building on this proposed structure make the message of the section clearer? In this case
-%   we would have to be a bit clearer in terms of how familiar notions of grammar can be seen as categories with functor into $\Grammar$.}
 
@@
 
-\max{was this paragraph supposed to be deleted?}  Next we present the
-full details of the syntax and typing of \theoryabbv. The syntax of
+The syntax of
 \theoryabbv is based on the dependent linear-non-linear (\lnld)
-calculus from \cite{dlnl}. As in \lnld, \theoryabbv includes both
+calculus from \cite{krishnaswami_integrating_2015}. As in \lnld, \theoryabbv includes both
 non-linear dependent types, as well as linear types, which are allowed
@@
 strictly-positive functors in \Cref{fig:spf}. First, we add operations
-for a recursive reference $\Var{x}$ the constant functor, $K$, tensor
+for a recursive reference $\Var~x$ the constant functor, $K$, tensor
 products and additive conjunction and disjunction. As an example,
@@
 of type expressions $F$. The introduction rule for this is
-$\mathsf{roll}$, which constructs an element of $\mu F\,x$ from its
+$\roll$, which constructs an element of $\mu F\,x$ from its
 one-level unfolding. The elimination principle is defined by a
-$\mathsf{fold}$ operation. The equations ensure that this makes the
+$\fold$ operation. The equations ensure that this makes the
 family an \emph{initial algebra} for the type expressions. That is the
@@
 $\epsilon$-transition then we may use $\epscons$ to pull it back along the
-$\epsilon$-transition and construct a trace beginning at the source of the $\epsilon$-transition.
+$\epsilon$-transition and construct a trace beginning at the source of the
+$\epsilon$-transition. As a shorthand, write $Parse_{N}$ for the accepting
+traces out of $N.init$.
 
@@
 data Trace(*@$_{\N}$@*) : (s : N.states) (b : Bool) (*@$\to$@*) L where
-  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) N.states ] Trace(*@$_{\N}$@*)(s , N.isAcc s))
+  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) N.states ] Trace(*@$_{\N}$@*) s (N.isAcc s))
   cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.transitions ] &[ b (*@$\in$@*) Bool ]
-           ('N.label(t)' (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.dst(t) , b) (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.src(t) , b)))
+           ('N.label t' (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.dst t) b (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.src t) b))
   (*@$\epsilon$@*)cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.(*@$\epsilon$@*)transitions ] &[ b (*@$\in$@*) Bool ]
-           (Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)dst(t) , b) (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)src(t) , b)))
+           (Trace(*@$_{\N}$@*) (N.(*@$\epsilon$@*)dst t) b (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)src t) b))
 \end{lstlisting}
@@
 data Trace(*@$_{\D}$@*) : (s : D.states) (b : Bool) (*@$\to$@*) L where
-  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) D.states ] Trace(*@$_{\D}$@*)(s , D.isAcc s))
+  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) D.states ] Trace(*@$_{\D}$@*) s D.isAcc s)
   cons : (*@$\uparrow$@*)(&[ c (*@$\in$@*) (*@$\Sigma$@*) ] &[ s (*@$\in$@*) D.states ] &[ b (*@$\in$@*) Bool ]
-             ('c' (*@$\lto$@*) Trace(*@$_{\D}$@*)(D.(*@$\delta$@*) c s , b) (*@$\lto$@*) Trace(*@$_{\D}$@*)(s , b)))
+             ('c' (*@$\lto$@*) Trace(*@$_{\D}$@*) (D.(*@$\delta$@*) c s) b (*@$\lto$@*) Trace(*@$_{\D}$@*) s b))
 \end{lstlisting}
@@
 are unambiguous and define a parser for them. In particular we show that
-$\LinPiTy {s} {D.\states} {\LinSigTy {b} {\Bool} \Trace_{D}(s , b)}$ is strongly
+$\LinPiTy {s} {D.\states} {\LinSigTy {b} {\Bool} \Trace_{D}~s~b}$ is strongly
 equivalent to
@@
 The first half of the equivalence,
-$\parse_{D} : \ltonl {\left( \stringg \lto \LinPiTy {s} {D.\states} {\Trace_{D}(s)} \right)}$,
+$\parse_{D}$,
 is defined by induction on strings in \cref{fig:dfaparser}. If this string is empty, then $\parse_{D}$
 defines a linear function that terminates a trace at the input state $s$. If the
-string is nonempty, then $\parse_{D}$ walks forward from the input state $s$ by
+string is nonempty, then $\parse_{D}$ walks forward in $D$ from the input state $s$ by
 the character at the head of the string.
@@
 \begin{lstlisting}
-parse : (*@$\uparrow$@*)(String (*@$\lto$@*) &[ s (*@$\in$@*) D.states ] (*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*)(s , b))
-parse String.nil s = (*@$\sigma$@*) (D.isAcc s) (Trace(*@$_{\D}$@*).nil s)
-parse (String.cons ((*@$\sigma$@*) c a) w) s = let (*@$\sigma$@*) b t = parse w (D.(*@$\delta$@*) c s) in
+parse(*@$_D$@*) : (*@$\uparrow$@*)(String (*@$\lto$@*) &[ s (*@$\in$@*) D.states ] (*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*) s b)
+parse(*@$_D$@*) String.nil s = (*@$\sigma$@*) (D.isAcc s) (Trace(*@$_{\D}$@*).nil s)
+parse(*@$_D$@*) (String.cons ((*@$\sigma$@*) c a) w) s =
+   let (*@$\sigma$@*) b t = parse w (D.(*@$\delta$@*) c s) in
    (*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c s b a t)
@@
 The second half of the equivalence,
-$\print_{D} : \ltonl {\left( \LinPiTy {s} {D.\states} {\Trace_{D}(s)} \lto \stringg \right)}$,
+$\print_{D}$,
 is given in \cref{fig:printdfa} by induction on traces. If the trace is defined
 via $\nil$, then $\print_{D}$ returns the empty string. Otherwise, if the trace
-is defined by $\cons$ then $\parse_{D}$ appends the character $c$ to the output
+is defined by $\cons$ then $\parse_{D}$ appends the character from the most
+recent transition to the output
 string and recurses.
@@
 \begin{lstlisting}
-print : (s : D.states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*)(s , b)) (*@$\lto$@*) String)
-print s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).nil .s)) = String.nil
-print s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c (D.(*@$\delta$@*) c .s) b a trace)) =
-  String.cons ((*@$\sigma$@*) c a) (print (D.(*@$\delta$@*) c s) ((*@$\sigma$@*) b trace))
+print(*@$_D$@*) : (s : D.states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*) s b) (*@$\lto$@*) String)
+print(*@$_D$@*) s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).nil .s)) = String.nil
+print(*@$_D$@*) s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c (D.(*@$\delta$@*) c .s) b a trace)) =
+  String.cons ((*@$\sigma$@*) c a) (print(*@$_D$@*) (D.(*@$\delta$@*) c s) ((*@$\sigma$@*) b trace))
 \end{lstlisting}
-% \[
-% \inferrule
-% {
-% }
-% {tr : \LinPiTyLimit {s} {D.\states} {\LinSigTyLimit {b} {\Bool } {\Trace_{D}(s, b)}} \vdash asdf : \stringg}
-% \]
 \caption{Printer for DFA traces}
@@
 \begin{theorem}[Determinization, \Agda]
+  \label{thm:determinization}
   Given an NFA $N$, there exists a DFA $D$ such that
-  $\Trace_{N}~\N.\init~\true$ is weakly equivalent to $\Trace_{D}~\D.\init~\true$.
+  $Parse_{N}$ is weakly equivalent to $Parse_{D}$.
 \end{theorem}
@@
 
-  We demonstrate the weak equivalence between $\Trace_{\N}~\N.\init~\true$ and
-  $\Trace_{\D}~\D.\init~\true$ by constructing parse transformers between the two
+  We demonstrate the weak equivalence between $Parse_{N}$ and
+  $Parse_{D}$ by constructing parse transformers between the two
   grammars. To build the parse transformer
-  $\ltonl{(\Trace_{\N}~\N.\init~\true \lto \Trace_{\D}~\D.\init~\true)}$, it
-  suffices to construct the stronger term
+  $\ltonl{(Parse_{N} \lto Parse_{D})}$, we strengthen our inductive hypothesis
+  and build the term
 \newcommand{\NtoD}{\texttt{NtoD}}
@@
 If the NFA trace is defined via $\cons$ for a labeled transition
-$\texttt{trans} : \N.\transitions$ such that $N.\src~\texttt{trans} \in \X$, then
-we first decompose the trace into a parse of $\N.\labelt~\texttt{trans}$ and a
-trace starting from $\N.\dst~\texttt{trans}$. We then recursively call $\NtoD$
+$t : N.transitions$ such that $N.src~t \in \X$, then
+we first decompose the trace into a parse of $N.label~t$ and a
+trace starting from $N.dst~t$. We then recursively call $\NtoD$
 on this trace which returns a family of traces parameterized by $\epsilon$-closed
-subsets containing $N.\dst~\texttt{trans}$. We instantiate this family at the
+subsets containing $N.\dst~t$. We instantiate this family at the
 set of states
-$\D.\delta~(\N.\labelt~\texttt{trans})~X$. Because $\N.\dst~\texttt{trans}$ is
+$\D.\delta~(N.label~t)~X$. Because $\N.\dst~t$ is
 reachable from $\N.\src~\texttt{trans}$ by a single transition labeled by
-$\N.\labelt~\texttt{trans}$ and $\N.\src~\texttt{trans} \in X$ we have that
-$\N.\dst~\texttt{trans} \in \D.\delta~(\N.\labelt~\texttt{trans})~X$. Therefore
+$N.label~t$ and $N.src~t \in X$ we have that
+$N.dst~t \in \D.\delta~(N.label~t)~X$. Therefore
 we can linearly combine the parse of the label and the trace built from the
@@
 Lastly, if the NFA trace is defined via $\epsilon\cons$ for an $\epsilon$-transition
-$\epsilon\texttt{trans} : \N.\epstransitions$ such that
-$\N.\epssrc~\epsilon\texttt{trans} \in \X$, then the construction is quite
+$t : \N.\epstransitions$ such that
+$\N.\epssrc~t \in \X$, then the construction is quite
 similar. In this case we again recursively call $\NtoD$, except afterwards we
@@
 all states in $\D$ are taken to be $\epsilon$-closed, therefore
-$\N.\epsdst~\epsilon\texttt{trans} \in \X$ because
-$\N.\epssrc~\epsilon\texttt{trans}$ is in $X$. This concludes the construction of $\NtoD$.
+$\N.\epsdst~t \in \X$ because
+$\N.\epssrc~t$ is in $X$. This concludes the construction of $\NtoD$.
 
@@
 combined with the parse of $\c$ to build an $\N$ trace. This concludes the
-definition of $\DtoN$, thus showing that $\Trace_{\N}~\N.\init~\true$ and $\Trace_{\D}~\D.\init~\true$ are
+definition of $\DtoN$, thus showing that $Parse_{N}$ and $Parse_{D}$ are
 weakly equivalent.
@@
 
-First handle the case where $R$ is a literal character $c$. The NFA $N_{c}$ pictured in
+First handle the case where $R$ is a literal character $\literal c$. The NFA $N_{c}$ pictured in
 \cref{fig:litnfa} is strongly equivalent to the grammar $c$. To demonstrate this
-strong equivalence, we build terms between $c$ and $\Trace_{N_{c}}(N_{c}.\init)$
+strong equivalence, we build functions between $\literal c$ and $Parse_{N_{c}}$
 and show they are mutually inverse.
 
-Define $f : \ltonl {(c \lto \Trace_{N_{c}}(N_{c}.\init))}$ as follows,
-
-\[
-  a : c \vdash f := \cons(a, \nil) : \Trace_{N_{c}}(N_{c}.\init)
-\]
-
-Define $g : \ltonl {(\Trace_{N_{c}}(N_{c}.\init) \lto c)}$ via induction on the
-trace. Map traces beginning in state 1, which are necessarily built via $\nil$,
-to the linear unit $I$. Then we must handle traces beginning in state 0, which
-are necessarily built via $\cons$. We decompose such a trace into a parse of the
-character $c$ and a trace beginning in state 1. We inductively transform the
-trace beginning in state 1 into a term of type $I$. Finally, we use the right
-unitor to turn a term of type $c \otimes I$ into one of type $c$.
-
-\[\begin{tikzcd}
-    {c \otimes \mathsf{Trace}_{N_{c}}(1)} &
-    {c \otimes I} &
-    c
-	\arrow["{id\otimes g}", from=1-1, to=1-2]
-	\arrow["{}", from=1-2, to=1-3]
-\end{tikzcd}\]
-
-Note that the grammar $c$ is unambiguous, so $g \circ f$ is equal to the
-identity on $c$. Additionally, we can show that $f$ carries the structure of an
-indexed algebra homomorphism for the family of functors defining the trace type
-$\Trace_{N_{c}}$. Because $g$ is defined via induction, it carries the structure
-of an algebra homomorphism as well. Thus, $f \circ g$ is also an algebra
-homomorphism. $f \circ g$ and the identity map are both then algebra
-homomorphisms from $\Trace_{N_{c}}$ to itself. Because $\Trace_{N_{c}}$ is the
-initial algebra for the relevant family of functors, there is only one
-homomorphism from it to itself and thus $f \circ g = \mathsf{id}$. Therefore,
-$f$ and $g$ are mutually inverse and the regular expression is strongly
-equivalent to the traces through the NFA in this case.
-
+\begin{figure}
+\begin{lstlisting}
+A : (s : Fin 2) (*@$\to$@*) L
+A 0 = 'c'
+A 1 = I
+
+toN(*@$_c$@*) : (s : Fin 2) (*@$\to$@*) (*@$\uparrow$@*)(A s (*@$\lto$@*) Trace(*@$_{N_c}$@*) s true)
+toN(*@$_c$@*) 0 c = cons t true c (nil 1)
+toN(*@$_c$@*) 1 () = nil 1
+
+fromN(*@$_c$@*) : (s : Fin 2) (*@$\to$@*) (*@$\uparrow$@*)(Trace(*@$_{N_c}$@*) s true (*@$\lto$@*) A s)
+fromN(*@$_c$@*) 0 (cons t .true c tr) = let a (*@$\otimes$@*) () = c (*@$\otimes$@*) fromN(*@$_c$@*) 1 tr in a
+fromN(*@$_c$@*) 1 (nil .1) = ()
+\end{lstlisting}
+\caption{Parse transformer out of $Trace_{N_{c}}~s~true$}
+\label{fig:litNFAterms}
+\end{figure}
+
+In \cref{litNFATterms} we define a family of grammars $A$ and then prove that each $A~s$ is
+equivalent to the accepting traces in $N_{c}$ out of state $s$. Then the desired
+equivalence arises from taking $s = 0$.
+
+When
+defining $fromN_{c}$ we pattern match on the accepting traces beginning in each
+state. Out of state $0$ the only accepting traces are built via $cons$ over the
+single transition $t$, and out
+of state $1$ the only accepting trace is built via $nil$.
+
+Note that each $A~s$ is unambiguous, so $(fromN_{c}~s) \circ (toN_{c}~s)$ is equal to the
+identity on each $A~s$. To complete the proof of strong equivalence it remains
+to show that $(toN_{c}~s) \circ (fromN_{c}~s)$ is equal to the identity on
+$Trace~s~true$, which we show by using the equalizer type.
+
+\begin{figure}
+\begin{lstlisting}
+eqHom(*@$_{N_c}$@*) : (s : N(*@$_c$@*).state) (*@$\to$@*)
+      (*@$\uparrow$@*)(Trace(*@$_{N_c}$@*) s true (*@$\lto$@*) {tr | (toN(*@$_c$@*) s (*@$\circ$@*) fromN(*@$_{c}$@*) s) tr = id tr})
+\end{lstlisting}
+\caption{Signature for the equalizer term}
+\label{fig:litequalizer}
+\end{figure}
+
+The type signature for the family of functions, $eqHom_{N_{c}}$, into the equalizer type is given in
+\cref{fig:litequalizer}. By defining an $Trace_{N_{c}}$-algebra homomorphism
+into the equalizer, we enable a
+proof strategy for equality by induction on traces through $N_{c}$. First consider the traces
+out of state $1$. The only case to handle here is $nil 1$, and by definition
+$toN_{c}~1$ and $fromN_{c}~1$ invert each other.
+When defining $eqHom_{N_{c}}$ for traces out of state $0$, we must leverage an
+inductive hypothesis,
+
+\begin{align*}
+  (toN_{c}~0 \circ fromN_{c}~0) (cons~t~true~c~tr)
+  & \equiv (toN_{c}~0) (let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a) \\
+  & \equiv cons~t~true~(let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a)~(nil~1) \\
+  & \equiv cons~t~true~c~(nil~1) \\
+  & \equiv cons~t~true~c~tr \\
+\end{align*}
+
+The first two equalities are by defintion. The thrid equality holds because
+$\literal c$ is unambiguous and
+$let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a$ and $c$ have type
+$\literal c$. The final equality holds because our inductive hypothesis states
+that $nil~1 \equiv tr$. Thus, each $A s$ and $Trace_{N_{c}}~s~true$ are strongly
+equivalent --- in particular, $\literal c$ and $Parse_{N_{c}}$.
+
+\steven{Rewrite the disjunction case with equalizers}
 Now consider the case where $R$ is the disjunction of two regular expressions
@@
 it suffices to show that
-$\Trace_{N_{1}}(N_{1}.\init) \oplus \Trace_{N_{2}}(N_{2}.\init)$ is strongly
-equivalent to $\Trace_{N_{\oplus}}(N_{\oplus}.\init)$.
+$Parse_{N_{1}} \oplus Parse_{N_{2}}$ is strongly
+equivalent to $Parse_{N_{\oplus}}$.
 
@@
 \newcommand{\Dyck}{\mathtt{Dyck}}
-\steven{This subsection is still a WIP}
 Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
 grammars (CFGs). The smallest nontrivial CFG is the Dyck grammar
-$\Dyck$ --- the grammar of balanced parentheses. Moreover, the Chomsky-Sch\"utzenberger
-representation theorem \cite{CHOMSKY1959118} demonstrates that $\Dyck$, and grammars like it,
-cut to the essence of what it means to be context free.
+$\Dyck$ --- the grammar of balanced parentheses.
 
@@
 and is defined in \cref{fig:dyckinductive}. The $\nil$ constructor shows that
-the empty string is balanced, and the $\balanced$ constructor builds a balanced
+the empty string is balanced, and the $bal$ constructor builds a balanced
 parse by wrapping an already balanced parse in an additional set of parentheses
-then following it with another balanced parse.
-
+then following it with another balanced parse. We construct a parser for $\Dyck$ by building a deterministic automaton $M$ such
+that $Parse_{M}$ is strongly equivalent to $\Dyck$.
 \begin{figure}
@@
   nil : (*@$\uparrow$@*) Dyck
-  balanced : (*@$\uparrow$@*)('(' (*@$\lto$@*) Dyck (*@$\lto$@*) ')' (*@$\lto$@*) Dyck (*@$\lto$@*) Dyck)
+  bal : (*@$\uparrow$@*)('(' (*@$\lto$@*) Dyck (*@$\lto$@*) ')' (*@$\lto$@*) Dyck (*@$\lto$@*) Dyck)
 \end{lstlisting}
@@
 
-\steven{somewhere in here, sneak in the benchmark where the Dyck parser could
-  handle 10's of thousands of characters in about 30s}
-
-We construct a parser for $\Dyck$ by building a deterministic automaton $M$ such
-that $\Trace_{M}(M.\init) \cong \Dyck$.
-
 \Cref{fig:DyckAutomaton} presents a sketch of the automaton $M$. The traditional
@@
 open parentheses have been seen so far. We may think of
-$M$ as performing the same task as a stack machine, except instead of storing
+$M$ as performing the same task as a stack machine through slightly different means. Instead of storing
 the number of open parentheses in a distinguished stack of memory, $M$ operates
-by storing this data in the current state. This is made feasible by specifying
-$M$ almost identically to how we defined DFAs, except we omit the
-finiteness condition on the states of the automaton.
-
-\steven{Mention that $\Trace_{M}$ is iso to string. Mention the same for DFAs above }
-
-% Unlike the finite automata seen thus far, $M$ has infinitely many states --- one
-% for each $n : \mathbb{N}$ with an additional fail state. This diverges a bit
-% from classical language theory, where the recognizer for $D$ would be encoded as a
-% pushdown automaton with finitely many
-% states and stack of memory. Such PDAs are encodable in our theory, although they
-% are a little cumbersome to use. Moreover, they
-% seem overfit to classical reasoning as they don't preserve the grammar up to
-% strong equivalence, only weak equivalence in general. \steven{cite Lambek
-%   automata and state that is is a secondary goal to preserve as much structure as possible}
-% By moving the stack of memory (the nats) into the
-% state, we obtain infinitely many states, which is fine bc the states need only
-% be a type for us to interface with it.
-
-\steven{TODO sketch proof of equivalence. Mention that it is mechanized}
+by storing this data in the current state. $M$ is defined almost identically to how we defined DFAs, except we omit the
+finiteness condition on the states of the automaton. This similarity allows
+functions $parse_{M}$ and $print_{M}$ to be defined analogously to
+\cref{fig:dfaparser,fig:printdfa}, and so
+$\LinSigTy {s} {M.states} {\LinSigTy {b}{Bool} {Trace_{M}~s~b}}$ is likewise unambiguous.
+
+\begin{theorem}[\Agda]
+  \label{thm:dyck}
+  $Dyck$ and $Parse_{M}$ are strongly equivalent.
+\end{theorem}
+
+We omit the details of \cref{thm:dyck} in this paper, and the ful proof is
+mechanized in our Agda development. We can build a $Dyck$ parser by utilizing
+this artifact, and the
+runtime of the $Dyck$ parser is about 30 seconds for input that is 25,000
+characters long.
 
@@
     \node[right of=2] (3) {$\dots$};
-    \node[state, below of=0] (fail) {$\mathsf{fail}$};
+    \node[state, below of=0] (fail) {$fail$};
 
@@
 \subsection{Unrestricted Grammars}
-% \newcommand{\P}{\texttt{P}}
+\newcommand{\stringNL}{String'}
 \steven{Need to better introduce nonlinear type of strings. We haven't talked
   about semantics nor have we included a nonlinear type of strings in the syntax}
-\newcommand{\stringNL}{String'}
-Let $\stringNL$ represent a nonlinear type of strings and let
-$P : \stringNL \to U$
-\steven{Or should $P$ be a semantic grammar?}
-
-The type of $P$ very nearly describes a formal grammar in \theoryabbv. That is,
-$P$ describes a mapping from strings to sets of parse trees; however, even
-though $P$ captures the same grammatical data $P$ is not
-a first-class linear type. We reflect this nonlinear notion of grammar as a
-linear type,
+Define $\stringNL$ to be the non-linear list type over the alphabet $\Sigma$ and
+let $P : \stringNL \to U$ be a mapping from the set of strings to the universe of
+nonlinear types. The function $P$ describes a family of nonlinear types
+dependent on $\stringNL$.
+We view the image of $P$ as a set of parse trees for each input string, and thus
+$P$ carries the structure of a formal grammar. However, even though $P$ describes this grammatical data, it is not
+a first-class linear type in \theoryabbv. To address this gap, below we define a linear type that captures the same
+semantic grammar as $P$,
 
 \[
-Reflect~P = \LinSigTyLimit {w} {\stringNL} {\LinSigTyLimit {x} {P~w} {~\lceil w \rceil}}
+Reify~P = \LinSigTyLimit {w} {\stringNL} {\LinSigTyLimit {x} {P~w} {~\lceil w \rceil}}
 \]
 
+\steven{Probably put this ceiling business around the discussion of the linear
+  type of strings. Is this in section 3?}
+
+where $\lceil \stringquote{} \rceil = I$ and
+$\lceil c :: w \rceil = \literal c \otimes \lceil w \rceil$.
+
+This reification operation on functions $\stringNL \to U$ is incredibly expressive,
+as it allows to sidestep our linear typing connectives and utilize the whole of nonlinear
+dependent type theory to define a grammar. This power allows us to encode
+arbitrary grammars. In particular, given a
+Turing machine $Tur$ one may
+define a predicate $accepts : \stringNL \to U$ such that $accept~w$ is equal to $\top$ if
+$T$ accepts $w$ and equal to $\bot$ otherwise. Then, $Reify~accepts$ is a linear
+type that captures precisely the strings accepted by $Tur$. That is,
+$Lang(Reify~accepts)$ is recursively enumerable --- the most general class of
+languages in the Chomsky hierarchy.
+
 \subsection{Semantic Actions}
-\steven{TODO write this section and give some small simple actions}
-All constructions discussed until now detail how a string may be parsed into a
-\emph{concrete} syntax tree; however, this is often not the representation that later
-compiler passes would take in. Instead, they often take in an \emph{abstract}
-syntax tree that forgets syntactic details that are superfluous to the later
-passes. For example, after using parentheses to disambiguate a parse of a string
-into a concrete syntax tree, we may remove these parentheses to have a cleaner
-structured representation of the underlying data. The procedure that takes in a
+So far, each construction has been primarily concerned with the parsing of a
+string into a \emph{concrete} syntax tree. However, in a larger verified
+developement --- such as a compiler --- a parser will often emit an
+\emph{abstract} syntax tree that omits superflous syntactic details that aren't
+needed for later passes.  The procedure that takes in a
 concrete syntax tree and returns an abstract one is referred to as a
@@
 
-Given a linear type $A$, the inhabitants of $A$ serve as concrete syntax. We may
-then chose a non-linear type $X$ that serves as abstract syntax. A semantic
-action can then be defined as $\square (A \lto \Delta X)$ where
-$\Delta X = \LinSigTy x X \top$.
-\steven{Note here that even though $\Delta$ is a map from nonlinear types into
-  linear ones, it is very different from $F$/whatever we call that half of the adjunction}
-
-% \steven{Everything in this section below this marker is outdated and needs to be
-% folded into the above or cut}
-
-% \steven{Need a different segue because canonicity is no longer before this}
-% The canonicity theorem from the previous section gives a tight connection
-% between terms in our type theory and the theory of formal languages. In this section
-% we further explore these connections by using the fact that, classically
-% formal language theory is closely related to the study of automata.
-% In the Chomsky hierarchy, each language class is associated to a class of
-% automata that serve as recognizers. Internal to \theoryabbv,
-% we can characterize these language classes syntactically; moreover, we
-% demonstrate the equivalence of these language classes to its associated automata class as a
-% proof term within our logic.
-
-% \subsection{Non-deterministic Finite Automata}
-% \label{subsec:finiteaut}
-% \begin{figure}
-%   \begin{tikzpicture}[node distance = 25mm ]
-%     \node[state, initial, accepting] (1) {$1$};
-%     \node[state, below left of=1] (2) {$2$};
-%     \node[state, right of=2] (3) {$3$};
-
-%     \path[->] (1) edge[above] node{$b$} (2)
-%               (1) edge[below, bend right, left=0.3] node{$\varepsilon$} (3)
-%               (2) edge[loop left] node{$a$} (2)
-%               (2) edge[below] node{$a, b$} (3)
-%               (3) edge[above, bend right, right=0.3] node{$a$} (1);
-%   \end{tikzpicture}
-%   \caption{An example NFA}
-%   \label{fig:NFA}
-% \end{figure}
-
-% Classically, a \emph{nondeterministic finite automaton} (NFA) is a finite state machine where
-% transitions are labeled with characters from a fixed alphabet $\Sigma$. These
-% are often represented formally as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$,
-
-% \begin{itemize}
-%   \item $Q$ a finite set of states
-%   \item $\Sigma$ a fixed, finite alphabet
-%   \item $\delta : Q \times (\Sigma \cup \{ \varepsilon\}) \to \mathcal{P}(Q)$ the labeled transition function
-%   \item $q_{0} \in Q$ the start state
-%   \item $F \subset Q$ a set of accepting states
-% \end{itemize}
-
-% \subsection{A DFA Parser}
-% \label{subsec:regexparser}
-
-% Just as we encoded traces of NFAs as grammars, we likewise
-% encode the traces of a DFA as grammars. The key difference
-% between NFAs and DFAs is \emph{determinism} --- meaning,
-% that in a state $q$ inside of DFA, given a character $c$ there
-% will be exactly one transition that we may take leaving $q$
-% with label $c$. For us, this changes the definition of valid
-% transitions for a DFA, instead of the definition of
-% $\mathsf{Trans}$ provided in \cref{subsec:finiteaut} DFAs
-% obey
-
-% \begin{gather*}
-%  \mathsf{State} \in \{ g_{q} : q \in Q \} \\
-%  \mathsf{Trans}(q) ::= \bigoplus_{c \in \Sigma} (c \otimes \mathsf{State})
-% \end{gather*}
-
-% Meaning, each state has a transition for every character. \emph{Note:} the above use
-%   of a disjunction over $\Sigma$ is a bit of abuse of notation.
-%   As $\Sigma$ is a finite alphabet, we wish to think of this as an iterated
-%   binary sum, or perhaps as a finitely indexed sum defined via linear dependent
-%   sums in the same manner by which we defined binary sums.
-
-% We now wish to define a parser term for DFA grammars. In
-% particular, for a DFA $D$ we want to build a term,
-
-% \[
-%   w : \String \vdash \mathsf{parse}_{D} : \mathsf{AccTrace}_{D} \oplus \top
-% \]
-
-% where left injection into the output type denotes acceptance
-% by the parser, and right injection denotes rejection. To
-% build such a parser, it will suffice to construct a term
-
-% \[
-%   w : \String \vdash \mathsf{parse}_{D} : \LinSigTy q Q {\mathsf{Trace}_{D}(q_{0} , q)}
-% \]
-% This is because given a trace of a DFA, we may easily check
-% if we should accept or reject by simply testing
-% if the final state is accepting.
-
-% Because $w$ is a Kleene star of characters, we may construct
-% our desired $\mathsf{parse}_{D}$ term as a $\mathsf{foldl}$
-% over $w$. In the empty case, we just have the trace that
-% ends at the accepting state. In the recursive case, we
-% effectively add to our trace by transitioning one character
-% at a time, as we read them moving across $w$.
-
-% Perhaps this derivation is not too surprising. All it says
-% is that a DFA may be ran by transitioning a single character
-% at a time, and then accepting or rejecting based on the
-% final state. This is exactly what DFAs did initially, so
-% what did we gain? Well, this has the benefit of our type
-% system to ensure its correctness. Moreover, this construction exports to an
-% intrinsically verified and executable decision procedure for DFAs in \theoryabbv
-% embedded in Agda.
-
-% We should note that the
-% construction of a DFA parser requires the addition of an additional but seemingly innocuous
-% axiom.
-
-% \[
-%   \top \cong \left( \bigoplus_{c : \Sigma} c \right)^{*}
-% \]
-
-% That is, the characters in the alphabet $\Sigma$ really are all of the
-% characters. The need for this axiom arises when we go to take a transition
-% within the DFA,
-% as we must ensure that the next character of the string has a corresponding
-% transition from the current state
-% --- which would be guaranteed by determinism provided there are no surprise characters.
-
-% \subsection{Regular Expressions and DFAs}
-% \label{subsec:deriv}
-
-% In order to extend the DFA
-% parser from \cref{subsec:regexparser} to the construction of
-% a verified parser
-% generator for regular expression we need to perform some
-% plumbing establishing an equivalence between regular
-% expressions and DFAs.
-
-% There are several routes we may hope to take in establishing
-% this equivalence. First, we could prove an equivalence
-% between NFAs and regular expressions, and separately prove
-% an equivalence between NFAs and DFAs.
-% In \cref{subsec:eqproofs}, we include a version of
-% Thompson's construction --- which established the
-% equivalence between regular grammars and DFAs. We may additionally
-% hope to internalize a variant of the powerset construction \cite{rabinFiniteAutomataTheir1959}
-% --- which takes as input an NFA and constructs a DFA that
-% recognizes the same language --- and combine the results of
-% Thompson's construction and the powerset constructions to give an equivalence
-% between regular expressions and DFAs.  This route is alluring, as it
-% internalizes several classic grammar-theoretic constructions. However, it may
-% necessitate extensions to the LNL theory, like
-% a propositional truncation, and we have not yet investigated
-% how this would interact with the existing types in the
-% theory. The addition of a propositional truncation may seem
-% harmless, but it is not always immediately clear how
-% distinct constructions will interact. For instance, when
-% exploring LNL models, Benton discovered that the synthesis
-% of linear and dependent types require a new presentation of
-% the $!$ modality from linear logic
-% \cite{bentonMixedLinearNonlinear1995}. That is all to say,
-% this is a work in progress and
-% it is not immediate that the addition of a propositional
-% truncation is adequate for establishing the weak equivalence
-% between NFAs and DFAs.
-
-% We may instead hope to internalize an equivalence between
-% regular grammars and DFAs by using Brzozowski derivatives to
-% directly create a DFA that is weakly equivalent to a given
-% regular expression, as described by Owens et al.
-% \cite{owensRegularexpressionDerivativesReexamined2009}.
-% One characterization of regular grammars is that they are
-% precisely those grammars which have finitely many inequivalent Brzozowski
-% derivatives
-% \cite{brzozowskiDerivativesRegularExpressions1964}.
-% The algorithm used by Owens takes in a
-% regular grammar and generates a DFA that recognizes the same
-% language, and the states in this DFA are the finitely many
-% derivative equivalence classes. We initially had a version of
-% this theorem very roughly internalized in the LNL theory.
-% To our taste, too much of this presentation relied on
-% meta-arguments that lived outside of
-% our formalism, and thus this particular phrasing of the
-% theorem did not translate well into formalization.
-
-% In any case, we believe
-% that revisiting these lines of thought will lead to a
-% satisfactory internalization of the equivalence between
-% regular grammars and DFAs, and thus would bridge the gap
-% between our DFA parser and a full regular expression parser.
-
-% \subsection{Equivalence Between Regular Expressions and Finite Automata}
-% \label{subsec:eqproofs}
-% In this section, we describe a version of Thompson's
-% construction \cite{thompsonProgrammingTechniquesRegular1968}
-%    where we construct an NFA that recognizes a given regular
-% expression. Moreover, we will show that this NFA is strongly equivalent to the
-% original grammar. Witnessing this construction in our syntax has two benefits
-% \begin{enumerate}
-%   \item It reinforces this high-level view that the syntax is a natural and
-%         general setting for formal grammar reasoning, as we demonstrate that
-%         this formal system subsume results from existing systems, and
-%   \item Following the
-%         development of Thompson's construction, we then need
-%         only establish the equivalence of NFAs and DFAs to
-%         complete the full regular expression parser
-% \end{enumerate}
-
-% \begin{theorem}[Thompson]
-%   \label{thm:thompson}
-%   For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same
-%   language as $g$.
-% \end{theorem}
-% We make a pretty straightforward adaptation of Thompson's theorem to our setting,
-
-% \begin{theorem}[Typed Thompson]
-%   \label{thm:typthompson}
-%   For $g$ a regular expression $g$, there is an NFA $N$ such that $g$ is isomorphic
-%   to $\mathsf{AccTrace}_{N}$.
-% \end{theorem}
-
-% \begin{proof}[Proof Sketch]
-%   Recall that regular expressions are inductively defined via
-%   disjunction, concatenation, and Kleene star over literals
-%   and the empty grammar. By induction over regular expressions,
-%   we will construct an NFA that is equivalent to $g$.
-
-%   First, define the recognizing NFA for the empty grammar
-%   $I$.
-
-%   \begin{figure}[h!]
-%   \begin{tikzpicture}[node distance = 25mm ]
-%     \node[state, initial] (1) {$1$};
-%     \node[state, right of=1, accepting] (2) {$2$};
-
-%     \path[->] (1) edge[below] node{$\varepsilon$} (2);
-%   \end{tikzpicture}
-%   \caption{$NFA(I)$}
-%   \label{fig:emptyNFA}
-%   \end{figure}
-
-%   The type of traces from the initial state of $NFA(I)$ to the single
-%   accepting state is given by,
-
-%   \[
-%     \mathsf{Trace}_{NFA(I)}(q_{1}, q_{2}) = \mu
-%       \begin{pmatrix}
-%          g_{q_{1}} := g_{q_{2}} \\
-%          g_{q_{2}} := I
-%       \end{pmatrix}. g_{q_{1}}
-%   \]
-
-%   The accepting traces through $NFA(I)$ are then described
-%   as,
-
-%   \[
-%     \mathsf{AccTrace}_{NFA(I)} = \LinSigTy q {\{1 , 2\}} {\left( \mathsf{Trace}_{NFA(I)}(q_{1} , q) \pair \mathsf{acc}(q) \right)}
-%   \]
-
-%   A quick inspection of \cref{fig:emptyNFA} reveals that the
-%   only reasonable choice for $q$ is $q_{2}$ --- because
-%   state 2 is accepting while state 1 is not. Therefore,
-
-%   \begin{align*}
-%     \mathsf{AccTrace}_{NFA(I)}
-%     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \pair \mathsf{acc}(q_{2}) \\
-%     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \pair \top \\
-%     & \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})
-%   \end{align*}
-
-%   From here, to prove
-%   $I \cong \mathsf{AccTrace}_{NFA(I)}$ it suffices to show
-%   $I \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$. Below we
-%   give two parse transformers, one from $I$ to
-%   $\mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$ and vice versa.
-
-%   \[
-%     \inferrule
-%     {p : I \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
-%      \exists \text{~transition~} q_{1} \overset{\varepsilon}{\to} q_{2}
-%     }
-%     {p : I \vdash \mathsf{\varepsilon cons}(\mathsf{nil}) : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})}
-%   \]
-
-%   Let $\gamma$ be the substitution $\{ g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$,
-
-%   \[
-%     \inferrule
-%     {
-%       p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash p :
-%         \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \\
-%       x_{1} : \simulsubst {g_{q_{2}}} {\gamma} = I \vdash x_{1} : I \\
-%       x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
-%     }
-%     {
-%       p : \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.x_{1} , x_{2}.x_{2})(p) : I
-%     }
-%   \]
-
-%   This concludes the proof for the case of the empty
-%   grammar. Let's now walk through the construction for
-%   literal grammars. Given a character $c$, we construct an
-%   NFA that recognizes only the string containing the single
-%   character $c$ as,
-
-%   \begin{figure}[h!]
-%   \begin{tikzpicture}[node distance = 25mm ]
-%     \node[state, initial] (1) {$1$};
-%     \node[state, right of=1, accepting] (2) {$2$};
-
-%     \path[->] (1) edge[below] node{$c$} (2);
-%   \end{tikzpicture}
-%   \caption{$NFA(c)$}
-%   \label{fig:literalNFA}
-%   \end{figure}
-
-%   The automaton in \cref{fig:literalNFA} induces the
-%   following type of traces from $q_{1}$ to $q_{2}$.
-
-%   \[
-%     \mathsf{Trace}_{NFA(c)}(q_{1}, q_{2}) = \mu
-%       \begin{pmatrix}
-%          g_{q_{1}} := c \otimes g_{q_{2}} \\
-%          g_{q_{2}} := I
-%       \end{pmatrix}. g_{q_{1}}
-%   \]
-
-%   Through the same argument as the empty grammar, the only
-%   state that is accepting is $q_{2}$ and thus,
-
-%   \[
-%     \mathsf{AccTrace}_{NFA(c)} \cong \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})
-%   \]
-
-%   To show the desired isomorphism of $c \cong NFA(c)$ we
-%   make a similar argument as we did for the empty grammar
-%   $I$, except we leverage the $\mathsf{cons}$ rule instead
-%   of $\mathsf{\varepsilon cons}$. That is, the parse
-%   transformers in either direction are given as,
-
-%   \[
-%     \inferrule
-%     {\cdot \vdash \mathsf{nil} : \mathsf{Trace}(q_{2} , q_{2}) \\
-%      \exists \text{~transition~} q_{1} \overset{c}{\to} q_{2}
-%     }
-%     {p : c \vdash \mathsf{cons}(\mathsf{nil}) :
-%       \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2})}
-%   \]
-
-%   \[
-%     \inferrule
-%     {
-%       p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash p :
-%         \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \\
-%       x_{1} : \simulsubst {(c \otimes g_{q_{2}})} {\gamma} = c \otimes I \vdash \mathsf{unitR}(x_{1}) : c \\
-%       x_{2} : \simulsubst {I} {\gamma} = I \vdash x_{2} : I
-%     }
-%     {
-%       p : \mathsf{Trace}_{NFA(c)}(q_{1} , q_{2}) \vdash \mathsf{mfold}(x_{1}.\mathsf{unitR}(x_{1}) , x_{2}.x_{2})(p) : c
-%     }
-%   \]
-
-%   Where $\gamma$ is the substitution
-%   $\{ c \otimes g_{q_{2}} / g_{q_{1}}, I / g_{q_{2}} \}$ and
-%   $\mathsf{unitR}$ is a witness to the isomorphism
-%   $c \otimes I \cong c$. Again, we may see that these do
-%   indeed mutually invert each other in the Agda code.
-
-%   It remains to show that the desired isomorphisms are
-%   preserved by $\otimes$, $\oplus$, and Kleene star. Here,
-%   we will give the argument for the disjunction case, the
-%   others are defined quite similarly.
-
-%   Given two NFAs $N$ and $M$, we may define a new NFA that
-%   encodes the disjunction of $N$ and $M$. Denote the
-%   internal states of $N$ by $q_{j}$'s and the internal states
-%   of $M$ by $r_{k}$'s,
-
-%   \begin{figure}[h!]
-%   \begin{tikzpicture}[node distance = 20mm ]
-%     \node[state] (2) {$q_{init}$};
-%     \node[state, initial, below left of=2] (1) {$init$};
-%     \node[state, below right of=1] (3) {$r_{init}$};
-%     \node[right of=2] (4) {$\cdots$};
-%     \node[right of=3] (5) {$\cdots$};
-%     % \node[state, below right of=1] (3) {$3$};
-%     % \node[state, below of=2] (4) {$4$};
-
-%     \path[->] (1) edge[below] node{$\varepsilon$} (2);
-%     \path[->] (1) edge[below] node{$\varepsilon$} (3);
-%     \path[->] (2) edge[below] node{} (4);
-%     \path[->] (3) edge[below] node{} (5);
-
-%     \node[label={[name=l] $N$}, draw,line width=2pt,rounded corners=5pt, fit=(2)(4)] {};
-%     \node[label={[name=l] $M$}, draw,line width=2pt,rounded corners=5pt, fit=(3)(5)] {};
-%   \end{tikzpicture}
-%   \caption{$N \oplus_{NFA} M$}
-%   \label{fig:disjunctionNFA}
-%   \end{figure}
-
-%   \Cref{fig:disjunctionNFA} shows the process of
-%   disjunctively combining NFAs. Precisely, we add a single
-%   new state and we included copies of the states from each of $N$
-%   and $M$. The new state acts as the initial state and has
-%   $\varepsilon$-transitions to the initial states of $N$ and
-%   $M$. We include all of the internal transitions from $N$
-%   and $M$, and the accepting states of $N \oplus_{NFA} M$
-%   are exactly the accepting states in each subautomaton.
-
-%   Let $g$ and $g'$ be two regular grammars such that
-%   $g \cong NFA(g)$ and $g' \cong NFA(g')$. As a matter of
-%   notation\footnote{We shall similarly abuse notation for
-%     $\otimes$ and Kleene. That is, for a regular grammar
-%     $g$, when we write $NFA(g)$ we mean the NFA inductively
-%     built up with the NFA-analogues to the constructors that
-%     built up $g$.}, we will write $NFA(g \oplus g')$ for
-%   $NFA(g) \oplus_{NFA} NFA(g')$. The traces of
-%   $NFA(g \oplus g')$ are then given by,
-
-%   \[
-%     \mathsf{Trace}_{NFA(g \oplus g')}(src , dst) = \mu
-%       \begin{pmatrix}
-%         g_{init} := g_{q_{0}} \oplus g_{r_{0}} \\
-%         g_{q_{j}} := \mathsf{Trans}_{NFA(g)}(q_{j}) \oplus \mathsf{isDst}(q_{j}) \\
-%         g_{r_{k}} := \mathsf{Trans}_{NFA(g')}(r_{k}) \oplus \mathsf{isDst}(r_{k})
-%       \end{pmatrix}.g_{src}
-%     \]
-
-%   where $\mathsf{Trans}$ is used to echo the same syntactic
-%   definitions that appear in the $NFA(g)$ and $NFA(g')$.
-%   Also, $src$ and $dst$ may take on any value in
-%   $Q := \{init\} \cup \{q_{j}\} \cup \{r_{k}\}$, and
-%   $\mathsf{isDst}(q)$ checks if $q$ is equal to $dst$. Which
-%   is all to say, the traces of $NFA(g \oplus g')$ comprise
-%   either a trace of $NFA(g)$, or a trace of $NFA(g')$, and
-%   the transition coming out of $g_{init}$ determines which
-%   subautomaton we step into.
-
-%   The parse transformer from $g \oplus g'$ checks which side
-%   of the sum type we are on, then takes the appropriate step
-%   from $g_{init}$ in the automaton.
-
-%   \[
-%     \inferrule
-%     {
-%       u : g \vdash \iota (\phi (u)) : \mathsf{AccTrace}_{NFA(g \oplus g')} \\
-%       v : g' \vdash \iota' (\psi (v)) : \mathsf{AccTrace}_{NFA(g \oplus g')}
-%     }
-%     {p : g \oplus g' \vdash \mathsf{case}~p \{ \mathsf{inl}(u) \mapsto s , \mathsf{inr}(v) \mapsto r \} : \mathsf{AccTrace}_{NFA(g \oplus g')}}
-%   \]
-
-%   with $\iota$ and $\iota'$ as embeddings from $NFA(g)$ and
-%   $NFA(g')$, respectively, into $NFA(g \oplus g')$,
-%   $\phi: g \cong \mathsf{AccTrace}_{NFA(g)}$, and $\psi: g' \cong \mathsf{AccTrace}_{NFA(g')}$. On a
-%   high level, all this construction does is turn a parse of
-%   $g$ into a parse of $NFA(g)$ and then embeds that inside
-%   of the larger automaton $NFA(g \oplus g')$. Likewise for $g'$.
-
-%   In the other direction, recall that the data of an
-%   accepting trace for $NFA(g \oplus g')$ is a pair of a
-%   trace and a proof that
-%   the end state $q'$ of that trace is accepting. By
-%   multifolding over the first part of that pair, we turn the
-%   term of type
-%   $\mathsf{Trace}_{NFA(g \oplus g')}(init , q')$ into a
-%   trace of either of the subautomata,
-
-%   \[
-%     p : \mathsf{Trace}_{NFA(g \oplus g')}(init , q') \vdash \mathsf{mfold}_{NFA(g \oplus g')} : \mathsf{Trace}_{NFA(g)}(q_{0} , q') \oplus \mathsf{Trace}_{NFA(g')}(r_{0} , q')
-%   \]
-
-%   Additionally, we leverage the fact that the only accepting
-%   states for $NFA(g \oplus g')$ are those from the
-%   subautomata to extract that $q'$ must be an accepting
-%   state from a subautomaton.
-
-%   \[
-%     x : \mathsf{acc}_{NFA(g \oplus g')}(q') \vdash M : \mathsf{acc}_{NFA(g)}(q') \oplus \mathsf{acc}_{NFA(g')}(q')
-%   \]
-
-%   We then combine the trace and proof of acceptance into an
-%   accepting trace of one of the subautomata,
-
-%   \[
-%     p : \mathsf{AccTrace}_{NFA(g \oplus g')} \vdash N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')}
-%   \]
-
-%   Lastly, we then inductively use the isomorphisms $\phi$
-%   and $\psi$ to turn the accepting traces into a parse of
-%   $g$ or $g'$,
-
-%   \[
-%      N : \mathsf{AccTrace}_{NFA(g)} \oplus \mathsf{AccTrace}_{NFA(g')} \vdash \mathsf{case}~N~\{\mathsf{inl}(n) \mapsto \phi^{-1}(n), \mathsf{inr}(n') \mapsto \psi^{-1}(n')\} : g \oplus g'
-%   \]
-
-%   When you wish to show that these actually maps invert each other to form an isomorphism\max{jumbled sentence}, we invoke the
-%   universal properties of the $\mathsf{Trace}$ types. That is,
-%   $\mathsf{Trace}_{\oplus NFA}$, $\mathsf{Trace}_{N}$, and $\mathsf{Trace}_{M}$
-%   are \emph{initial} algebras to the the functors that define them. Thus, the
-%   embedding of traces through $N$ into $\oplus NFA$ and then mapped back down to $N$
-%   compose to a map from $\mathsf{Trace}_{N}$ to itself. Such a map is unique and
-%   thus must be the identity. The same reasoning shows that the desired
-%   composite do indeed invert each other to form an isomorphism.
-
-%   As discussed above, the other cases and
-%   follow through a similar argument. This
-%   concludes the proof of our variant of Thompson's construction.
-% \end{proof}
-
-% \subsection{Other Automata}
-% \subsubsection{Pushdown Automata}
-% \label{subsubsec:pda}
-% A (nondeterministic) \emph{pushdown automaton} is an automaton that employs a
-% stack. Just like NFAs, they have transitions labeled with characters from a
-% fixed string alphabet $\Sigma$. Additionally, they maintain a stack of
-% characters drawn from a stack alphabet $\Gamma$. They are often represented
-% formally as a 7-tuple $(Q, \Sigma, \Gamma, \delta, q_{0}, Z, F)$,
-
-% \begin{itemize}
-%   \item $Q$ a finite set of states
-%   \item $\Sigma$ a fixed, finite string alphabet
-%   \item $S$ a fixed, finite stack alphabet
-%   \item
-%         $\delta \subset Q \times (\Sigma \cup \{ \varepsilon \}) \times S \to \mathcal{P}(Q \times S^{*})$
-%         the transition function
-%   \item $q_{0} \in Q$ the start state
-%   \item $Z \in S$ the initial stack symbol
-%   \item $F \subset Q$ the accepting states
-% \end{itemize}
-
-% We encode the traces of a pushdown automaton very similarly to those of an NFA,
-% except the transitions of a PDA are instead encoded via the LNL product type
-% $\bigamp$. This is because when simply transitioning via character, a PDA must
-% also pop and push characters onto a stack, which is used as the argument to
-% these dependent functions. Thus, the appropriate formation of traces through a
-% PDA is dependent on a stack.
-
-% Let $S$ be a non-linear type encoding the stack
-% alphabet, and build lists over $S$ as the (non-linear) least fixed-point
-% $\mathsf{List}(S) := \mu X . 1 + S \times X$. Then, the type of states for a
-% PDA $P$ with stack alphabet $S$ are given as a functions that takes in lists $L$,
-% and then makes a case distinction between possible transitions based on what was witnessed as $\mathsf{head}(L)$. The choice of transition
-% will then determine which character to transition by and what word $w$ should be
-% pushed onto the stack. The word that is added to the top of the stack is
-% appended to $\mathsf{tail}(L)$ and then we recursively step into another state
-% called on argument $w \cdot \mathsf{tail}(L)$.
-
-% \begin{gather*}
-%   \mathsf{State} \in {g _{q} : q \in Q} \\
-%   \mathsf{Word} \in \String \\
-%   \mathsf{Char} \in \Sigma   \\
-%   \mathsf{StackChar} \in \Gamma \\
-% \end{gather*}
-% \begin{align*}
-%   \mathsf{Trans}_{P}(q) ::=~&
-%                            \LinPiTy {(hd :: tl)} {\mathsf{List}(S)} {\left( \mathsf{Char} \otimes \mathsf{State}(\mathsf{Word} + tl) \right)}~| \\
-%   & \LinPiTy {(hd :: tl)} {\mathsf{List}(S)} {\left(  \mathsf{State}(\mathsf{Word} + tl) \right)}~|~\mathsf{Trans}_{P}(q) \oplus \mathsf{Trans}_{P}(q)
-% \end{align*}
-
-% That is to say, when transitioning a PDA pops off the head $hd$ of the stack
-
-% \subsubsection{Turing Machines}
-% \label{subsubsec:tm}
-
-% Above we gave a grammar presentation of
-% traces through a PDA by using a non-linear type $S$
-% to encode the stack. We may similarly use pairs $S \times S$
-% to encode the tape of a Turing machine. With two stacks we can simulate the behavior of the
-% infinite tape of a Turing machine. The intuition behind this correspondence is
-% that the left half of the tape is on one stack, the right
-% half of the tape the other, and we treat the tops of stacks
-% like the head of the tape.
-
-% \pedro{I like the overall structure of this section! Showing how the type theory
-% can encode and reason about different kinds of languages is very instructive!}
+Given a linear type $A$ and a nonlinear type $X$, we define a semantic action to
+be a function $\ltonl {(A \lto \Delta X)}$, where
+$\Delta X = \LinSigTy {x} {X} {\top}$. That is, a semantic action is an
+elaboration from the concrete parses of $A$ into the semantic domain $X$. As
+written, these actions are quite general. As discussed above, we may define
+folds that are still grammatical in nature --- such as folding over a concrete
+parse tree and forgetting parentheses that aren't needed in later compiler
+passes. We may also define arbitrary computation over the structure of the parse
+tree, such as calculating the maximum nesting depth of parentheses for $Dyck$,
+returning the length the input string, or
+even an interpreter that evaluates arithmetic expressions at parse-time.
+
+\steven{What can I say about what we have in the Agda development here?}
 

### 2024-11-11 paper/paper.tex
some prose about dependent polynomial functors

--- 2024-11-10 paper/paper.tex
@@
 
-For reasons of brevity, we present only a selection of cases in
-\Cref{fig:semantics}. The full details are in the appendix.\max{TODO:
-  put the full details in the appendix}.
-\max{TODO: how to describe $\mu$ succinctly?}\max{make sure $\lto$ and $\tol$ are right}
+For reasons of brevity, we present only the cases related to the
+linear types in \Cref{fig:semantics}. The full details are in the
+appendix.\max{TODO: put the full details in the appendix}.  The
+grammar for a literal $c$ has a single parse precisely when the input
+string consists of the single character. The grammar for the unit
+similarly has a single parse for the empty string.  A parse of the
+tensor product $A \otimes B$ consists of a \emph{splitting} of the
+empty string into a prefix $w_1$ and suffix $w_2$ along with an $A$
+parse of $w_1$ and $B$ parse of $w_2$. A parse of $\bigoplus_{x:X} A$
+is a pair of an element of the set $X$ and a parse of $A(x)$, while
+dually a parse of $\bigwith_{x:X} A$ is a \emph{function} taking any
+$x:X$ to a parse of $A(x)$. A $w$-parse of $A \lto B$ is a function
+that takes an $A$ parse of some other string $w'$ to a $B$ parse of
+$ww'$, and $B \tol A$ is the same except the $B$ parse is for the
+reversed concatenation $w'w$.
+%
+The set $\uparrow A$ is the set of parse for the empty string for
+$A$. This definition means that $\sem{\uparrow (A \lto B)}$ (or $\sem{\uparrow (B \tol A)}$) is
+equivalent to the set of parse transformers:
+\[ \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\]
+Next, a parse in the equalizer $\equalizer{a : A}{f}{g}$ is defined as
+a parse in $\sem{A}$ that is mapped to the same parse by $f$ and
+$g$. Recall that $\sem{f}\gamma$ and $\sem{g}$ are elements of
+$\sem{\uparrow(A \lto B)}$, i.e., parse transformers.
+
+The universe $L$ of linear types is interpreted as the set of all
+(small) grammars.
+
 \begin{figure}
@@
     \sem{A \otimes B}\gamma\,w &= \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
+    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
+    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
     \sem{\bigoplus_{x:X} A}\gamma\,w &= \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
     \sem{\bigwith_{x:X} A}\gamma\,w &= \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w\\
-    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
-    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
     \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
-    \sem{\SPF\,X}\gamma &= \{ F : \textrm{Functor}(\textrm{Grammar}^{\sem{X}\gamma}, \textrm{Grammar}) \pipe F \textrm{ preserves $\omega$-colimits.} \}\\
-    \sem{\el(F)}\gamma\,G &= \sem{A}\gamma\,G\\
-    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:adamek}}
+    \sem{\equalizer{a:A}{f}{g}}\gamma\,w &= \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
+    \sem{L}\gamma &= \Grammar_0\\
+    \sem{\SPF\,X}\gamma &= \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
+    \sem{\el(F)}\gamma\,G &= \sem{F}\gamma\,G\\
+    \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
+    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:indexed-container}}\\
   \end{align*}
@@
 
-\max{TODO: prose}
-
-\max{TODO: spelling}
-\begin{theorem}[Adamek]
-  If a functor $F : \mathcal C \to \mathcal C$ preserves all 
-  $\kappa$-colimits and $\mathcal C$ has an initial object and
-  $\kappa$-colimits, then $F$ has an initial algebra given by the
-  colimit of the chain
-  \[ 0 \to F(0) \to F(F(0)) \to \cdots \]
+The most complex part of the semantics is the interpretation of
+strictly positive functors and indexed inductive linear types.  We
+interpret a strictly positive functor as a \emph{dependent polynomial
+functor} on the category of sets, also sometimes called an
+\emph{indexed container} \cite{dep-poly,indexed-container}.
+\begin{definition}
+  Let $I$ and $O$ be sets. A (dependent) \emph{polynomial} (of sets)
+  from $I$ to $O$ consists of a set of shapes $S$, a set of positions
+  $P$ and functions $I \leftarrow P \to S \to O$. The \emph{extension}
+  of a polynomial is a functor $\Set/I \to \Set/O$ defined by
+  \max{TODO: fill in details}.
+
+  A dependent polynomial functor from $I$ to $O$ is a functor $\Set/I
+  \to \Set/O$ that is nautrally isomorphic to the extension of a
+  dependent polynomial from $I$ to $O$.
+\end{definition}
+With this interpretation, $\el(F)$ and $\map(F)$ are just interpreted
+as the action of the functor on objects and morphisms,
+respectively. We interpret the constructors $\mathsf{K},\Var$, etc. on
+functors in the obvious way that matches the definitional behavior of
+$\el$ and $\map$. The non-trivial part of the construction is
+verifying that such functors are constructing corresponding dependent
+polynomials. The details are tedious but straightforward extension of
+prior work on dependent polynomials and indexed containers and we have
+verified the construction in Agda.
+
+The reason we use dependent polynomials functors on sets is that it is
+well-known that their initial algebras exist
+\cite{initial-algebras-of-dep-poly}. Additionally, initial algebras
+for dependent polynomials are straightforwardly constructed in Agda as
+an inductive type of ``IW'' trees, which is available as a
+construction in the cubical library of Agda \cite{iw,cubical-library}.
+\begin{theorem}
+  Every dependent polynomial functor from $I$ to $I$ has an initial
+  algebra.
 \end{theorem}
+Then an element $F \in \sem{X \to \SPF\,X}\gamma$ is an
+$\sem{X}\gamma$-indexed family of polynomial functors from
+$\sem{X}\gamma\times \Sigma^*$ to $\Sigma^*$, and taking the product
+of these constructs a polynomial functor from $\sem{X}\gamma\times
+\Sigma^*$ to itself. Then $\sem{\mu\,F}\gamma$ is defined to be the
+initial algebra of this functor, and the initial algebra structure is
+used to interpret $\roll, \fold$ and the corresponding axioms.
 
@@
 
+\paragraph{Inductive Linear Types}
+
 % In \cite{firsovCertifiedNormalizationContextFree2015}, Firsov and Uustalu

### 2024-11-11 paper/paper.tex
comments

--- 2024-11-11 paper/paper.tex
@@
 similar role to the $!$ modality of ordinary linear logic.
+\pedro{We should also reference the persistence modality $\square$
+from separation logic}
 
@@
 $B$.
+\pedro{Isn't this is a slightly incorrect desciption of $\ltonl{A \lto B}$?
+How is the description above distinct form $A \lto B$?}
 
@@
     %
-    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
+    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e' : C}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'} : C}
     \\
@@
 \end{figure}
+\pedro{Fixed tensor elimination rule (added output type)}
+\pedro{Perhaps we should modify the equalizer introduction rule.
+  As it stands, readers might be confused by the same symbol $e$
+  being used everywhere.}
 
@@
 \end{definition}
+\pedro{Up to this point, we have mentioned a couple of times parse transformers,
+  but we have never defined what they are. In particular, why does the definition
+  require the $\uparrow$ modality. I'm assuming that it's to properly deal with
+  linear variables, but these things are not obvious at first.}
 
@@
   \steven{TODO this proof needs to be rewritten. Likely through use of equalizers}
+  \pedro{Since this proof and the determinization one are well-known and we have
+    proved it in Agda, it's probably a good idea to include a much shorter proof sketch,
+    or move it to the appendix, though this would feel a bit superfluous.}
 We present a variant of Thompson's construction
@@
 $\Dyck$ --- the grammar of balanced parentheses.
+
+\pedro{Is it actually the smallest? How do you quantify size of nontrivial things?
+  Might be better to just say ``an interesting nontrivial CFG [...]''}
 
@@
 characters long.
+\pedro{It might be a good idea to mention explicitly in the introduction
+  that we are not claiming that our verified parsers are competitive with
+  the state of the art.}
 
@@
 \steven{What can I say about what we have in the Agda development here?}
+\pedro{I'm assuming that this section is still incomplete, so I'll wait
+for it to be complete in order to make more detailed comments.}
 
@@
 that accounts for the possibility of ambiguity. The most common
-definition of a formal grammar is as a generative grammars, defined by
+definition of a formal grammar is as generative grammars, defined by
 a set of non-terminals, with a specified start symbol and set of
@@
 grammars, but what about the linear terms? These can be interpreted as
-\emph{parse transformers}
+\emph{parse transformers}, which we now define.
 \begin{definition}
@@
 \end{definition}
+\pa{Do we actually need a definition of formal language morphism?
+In case we do, it should probably be its own definition.}
 First, note that parse transformers generalize language inclusion: if
@@
 
-Parse transformers have an obvious notion of composition and identity
-transformer. This defines a \emph{category}.
+Parse transformers can be composed: given two parse transformers $f$ and $g$,
+their composition is defined pointwise, i.e. $(f\circ g)_w = f_w \circ g_w$.
+Furthemore, given a formal grammar $A$, its identity transformer is $id_w =
+id_{A(w)}$, where $id_{A(w)}$ is the identity function on the set $A(w)$. This
+defines a \emph{category}.
 \begin{definition}

### 2024-11-12 paper/paper.tex
start work on syntactic metatheory and denotational semantics in appendix

--- 2024-11-11 paper/paper.tex
@@
 
+\newcommand{\Subst}{\textrm{Subst}}
 \newcommand{\SPF}{\mathsf{SPF}}
@@
 
-\section{Elided Syntax}
-
-\max{TODO: add prose}
+\section{Syntax}
+
+In this section we include the elided syntactic forms, as well as
+definitions and basic properties of linear and non-linear
+substitution.
 
@@
 
+\begin{definition}
+  The set of (non-linear) substitutions $\gamma \in
+  \textrm{Subst}(\Gamma,\Gamma')$ where $\Gamma \isCtx$ and $\Gamma'
+  \isCtx$ is defined by recursion on $\Gamma$:
+  \begin{align*}
+    \textrm{Subst}(\Gamma,\cdot) &= \{\cdot \}\\
+    \textrm{Subst}(\Gamma,\Gamma',x:A) &= \{ (\gamma,M/x) \pipe \gamma \in \textrm{Subst}(\Gamma,\Gamma') \wedge \Gamma \vdash M : A[\gamma] \}
+  \end{align*}
+  simultaneously with an action of substitution on types, terms,
+  etc. in the standard way.
+
+  By induction we can define an identity substitution $\textrm{id} :
+  \textrm{Subst}(\Gamma,\Gamma)$ which maps each variable to itself,
+  and composition of substitutions
+  \begin{align*}
+    \cdot \circ \gamma' &= \cdot\\
+    (\gamma,M/x) \circ \gamma' &= (\gamma\circ \gamma') , M[\gamma']/x
+  \end{align*}
+
+  It is straightforward, but laborious to establish that all forms in
+  the type theory that are parameterized by a $\Gamma$ support an
+  admissible action of substitution:
+  \begin{mathpar}
+    \inferrule
+    {\Gamma \vdash M : X \and \gamma \in \Subst(\Gamma',\Gamma)}
+    {\Gamma' \vdash M[\gamma] : X[\gamma]}
+
+    \inferrule
+    {\Gamma;\Delta \vdash e : A \and \gamma \in \Subst(\Gamma',\Gamma)}
+    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] : A[\gamma]}
+  \end{mathpar}
+  et cetera.
+\end{definition}
+
+\begin{definition}
+  Let $\Gamma \vdash \Delta \isLinCtx$ and $\Gamma \vdash \Delta'
+  \isLinCtx$. The set of linear substitutions $\Subst(\Delta',\Delta)$
+  is defined by recursion on $\Delta$:
+  \begin{align*}
+    \Subst(\Delta',\cdot) &= \{ \cdot \pipe \Delta' = \cdot \}\\
+    \Subst(\Delta',(\Delta,a:A)) &= \{ (\delta, e/a) \pipe \delta \in \Subst(\Delta_1, \Delta) \and \Delta_2 \vdash e : A \and \Delta' = \Delta_1,\Delta_2\}
+  \end{align*}
+
+  Given substitutions $\delta_1 \in \Subst(\Delta_1', \Delta_1)$ and
+  $\delta_2 \in \Subst(\Delta_2', \Delta_2)$, we can define a
+  substitution $\delta_1,\delta_2 \in
+  \Subst((\Delta_1',\Delta_2'),(\Delta_1,\Delta_2))$. Furthermore, for
+  any substitution $\delta \in \Subst(\Delta,(\Delta_1,\Delta_2))$, we
+  can deconstruct $\delta = \delta_1,\delta_2$ with $\delta_1 \in
+  \Subst(\Delta_1', \Delta_1)$ and $\delta_2 \in \Subst(\Delta_2',
+  \Delta_2)$.
+  %% Again for any $\Gamma \vdash \Delta$, we can define an identity
+  %% substitution mapping each variable to itself. Additionally we can define composition of 
+\end{definition}
+
+\begin{definition}
+  Given any $\Gamma ; \Delta \vdash e : A$ and $\delta \in
+  \Subst(\Delta',\Delta)$, we define the substitution $\Gamma; \Delta
+  \vdash e[\delta] : A$ as follows, frequently using the inversion
+  principle to split the substitution into constituent components.
+  \begin{align*}
+    a[e/a] &= e\\
+    (e_1,e_2)[\delta_1,\delta_2] &= (e_1[\delta_1], e_2[\delta_2])\\
+    (\letin {(a , b)} e {e'})[\delta_1,\delta_2,\delta_3] &= \letin{(a,b)} {e[\delta_2]}{e'[\delta_1,a/a,b/b,\delta_2]}\\
+    ()[\cdot] &= ()\\
+    \letin{()} e {e'}[\delta_1,\delta_2,\delta_3] &= \letin{()} {e[\delta_2]}{e'[\delta_1,a/a,b/b,\delta_2]}\\
+    (\lamblto {a} e)[\delta] &= \lamblto a {e[\delta,a/a]}\\
+    (\applto {e'} {e})[\delta_1,\delta_2] &= \applto {e'[\delta_1]} {e[\delta_2]}\\
+    (\lamblto {a} e)[\delta] &= \lamblto a {e[\delta,a/a]}\\
+    (\applto {e'} {e})[\delta_1,\delta_2] &= \applto {e'[\delta_1]} {e[\delta_2]}\\
+    (\lambtol {a} e)[\delta] &= \lamblto a {e[a/a,\delta]}\\
+    (\apptol {e'} {e})[\delta_1,\delta_2] &= \apptol {e'[\delta_1]} {e[\delta_2]}\\
+    (\dlamb x e)[\delta] &= \dlamb x {e[\delta]}\\
+    (e\,.\pi\,M)[\delta] &= (e[\delta]\,.\pi\,M)\\
+    (\sigma\,M\,e)[\delta] &= \sigma\,M\,e[\delta]\\
+    (\letin{\sigma\,x\,a}{e}{e'})[\delta_1,\delta_2,\delta_3] &= \letin{\sigma\,x\,a} {e[\delta_2]}{e'[\delta_1,\delta_3]}\\
+    (\equalizerin{e})[\delta] &= \equalizerin{e[\delta]}\\
+    (\equalizerpi{e})[\delta] &= \equalizerpi{e[\delta]}
+  \end{align*}
+\end{definition}
+
+\section{Denotational Semantics}
+
+Next we define the term semantics for linear terms.
+
+\begin{definition}
+  For any $\Gamma \vdash \Delta_1,\Delta_2 \isLinCtx$ and $\gamma \in
+  \sem{\Gamma}$ there is an isomorphism $m_{\Delta_1,\Delta_2} :
+  \sem{\Delta_1, \Delta_2}\gamma \cong \sem{\Delta_1}\gamma \otimes
+  \sem{\Delta_2}\gamma$.
+
+  This can be extended to a sequence of contexts of any length.
+\end{definition}
+
+\begin{mathpar}
+  \inferrule
+  {\Delta_1 \vdash e_1 : A_1 \and \Delta_2 \vdash e_2 : A_2}
+  {\sem{(e_1,e_2)}\gamma\,w\,\delta
+    = \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}
+  }
+
+  \inferrule
+  {\Delta_2 \vdash e : A \otimes B \and \Delta_1,a:A,b:B,\Delta_3 \vdash e' : C}
+  {\sem{\letin{(a,b)}{e}{e'}}\gamma\,w\,\delta =
+    \letin{(w_1,w_2w_3,\delta_1,(w_2,w_3,\delta_2,\delta_3))}{m_{\Delta_1,\Delta_2,\Delta_3}(\delta)}
+    \letin{(w_{2,1},w_{2,2},a,b)}{\sem{e}\gamma\,w_2\,\delta_2}
+    {\sem{e'}\gamma\,w\,(w_1,w_{2,1}w_{2,2}w_3, \delta_1, (w_{2,1},w_{2,2}w_3, a, (w_{2,2},w_3, b, \delta_3)))}
+    }
+\end{mathpar}
+\begin{align*}
+  \sem{(e_1,e_2)}\gamma\,w\,\delta &=
+  \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}\\
+\end{align*}
+
 \end{document}

### 2024-11-12 paper/paper.tex
wip on appendix

@@
 
-  By induction we can define an identity substitution $\textrm{id} :
-  \textrm{Subst}(\Gamma,\Gamma)$ which maps each variable to itself,
-  and composition of substitutions
-  \begin{align*}
-    \cdot \circ \gamma' &= \cdot\\
-    (\gamma,M/x) \circ \gamma' &= (\gamma\circ \gamma') , M[\gamma']/x
-  \end{align*}
+  \max{do we need identity and composition of substitutions and their functoriality?}
+  %% By induction we can define an identity substitution
+  %% $\textrm{Subst}(\Gamma,\Gamma)$ which maps each variable to
+  %% itself. We write this substitution simply as $\Gamma$.
+  %% \begin{align*}
+  %%   \cdot \circ \gamma' &= \cdot\\
+  %%   (\gamma,M/x) \circ \gamma' &= (\gamma\circ \gamma') , M[\gamma']/x
+  %% \end{align*}
 
   It is straightforward, but laborious to establish that all forms in
-  the type theory that are parameterized by a $\Gamma$ support an
-  admissible action of substitution:
+  the type theory that are parameterized by a non-linear context
+  $\Gamma$ support an admissible action of a substitution $\gamma \in
+  \Subst(\Gamma',\Gamma)$:
   \begin{mathpar}
     \inferrule
-    {\Gamma \vdash M : X \and \gamma \in \Subst(\Gamma',\Gamma)}
+    {\Gamma \vdash X \isTy}
+    {\Gamma' \vdash X[\gamma] \isTy}
+
+    \inferrule
+    {\Gamma \vdash X \isSmall}
+    {\Gamma' \vdash X[\gamma] \isSmall}
+
+    \inferrule
+    {\Gamma \vdash X \equiv Y}
+    {\Gamma' \vdash X[\gamma] \equiv Y[\gamma]}
+
+    \inferrule
+    {\Gamma \vdash M : X}
     {\Gamma' \vdash M[\gamma] : X[\gamma]}
@@
     \inferrule
-    {\Gamma;\Delta \vdash e : A \and \gamma \in \Subst(\Gamma',\Gamma)}
-    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] : A[\gamma]}
+    {\Gamma \vdash M \equiv N : X}
+    {\Gamma' \vdash M[\gamma] \equiv N[\gamma] : X[\gamma]}
+
+    \inferrule
+    {\Gamma \vdash \Delta \isLinCtx}
+    {\Gamma' \vdash \Delta[\gamma] \isLinCtx}
+
+    \inferrule
+    {\Gamma \vdash A \isLinTy}
+    {\Gamma' \vdash A[\gamma] \isLinTy}
+
+    \inferrule
+    {\Gamma \vdash A \equiv B}
+    {\Gamma' \vdash A[\gamma] \equiv B[\gamma]}
+
+    \inferrule
+    {\Gamma;\Delta \vdash e : A}
+    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] : A[\gamma]}j
+
+    \inferrule
+    {\Gamma;\Delta \vdash e \equiv f : A}
+    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] \equiv f[\gamma] : A[\gamma]}j
   \end{mathpar}
-  et cetera.
+
+  %% Further, these all satisfy identity and composition equations M[\Gamma] = M and M[\gamma][\gamma'] = M[(\gamma \circ \gamma')]
 \end{definition}
@@
   Given any $\Gamma ; \Delta \vdash e : A$ and $\delta \in
-  \Subst(\Delta',\Delta)$, we define the substitution $\Gamma; \Delta
-  \vdash e[\delta] : A$ as follows, frequently using the inversion
-  principle to split the substitution into constituent components.
+  \Subst(\Delta',\Delta)$, we define the action of the substitution on
+  $e$ as follows, frequently using the inversion principle to split
+  the substitution into constituent components.
   \begin{align*}
@@
   \end{align*}
+  By induction on linear term and equality judgments, we establish the
+  following admissible rules for $\delta \in \Subst(\Delta',\Delta)$:
+  \begin{mathpar}
+    \inferrule
+    {\Gamma; \Delta \vdash e : A}
+    {\Gamma; \Delta' \vdash e[\delta] : A}
+
+    \inferrule
+    {\Gamma; \Delta \vdash e \equiv f : A}
+    {\Gamma; \Delta' \vdash e[\delta] \equiv f[\delta'] : A}
+  \end{mathpar}
+
+  \max{identity and composition?}
 \end{definition}
@@
 
-Next we define the term semantics for linear terms.
-
 \begin{definition}
-  For any $\Gamma \vdash \Delta_1,\Delta_2 \isLinCtx$ and $\gamma \in
-  \sem{\Gamma}$ there is an isomorphism $m_{\Delta_1,\Delta_2} :
-  \sem{\Delta_1, \Delta_2}\gamma \cong \sem{\Delta_1}\gamma \otimes
-  \sem{\Delta_2}\gamma$.
-
-  This can be extended to a sequence of contexts of any length.
+  We define the unit grammar
+  \[ I\,w = \{ ()\pipe w = \varepsilon \}\]
+  and given grammars $A$ and $B$ we define their tensor product
+  \[ (A \otimes B)\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in A\,w_1 \wedge b \in B\,w_2 \} \]
+  This extends to a monoidal structure on the category of grammars.
+Note that we defined in the semantics of linear types $\sem{I}\gamma = I$ and $\sem{A \otimes B}\gamma = \sem{A}\gamma \otimes \sem{B} \gamma$
 \end{definition}
 
+\begin{definition}[Denotation of Linear Contexts]
+  The semantics of linear contexts $\Gamma \vdash \Delta \isLinCtx$ is
+  defined as follows:
+  \max{make sure this aligns with our interpretation of $\lto$/$\tol$}
+  \begin{align*}
+    \sem{\cdot}\,\gamma &= I \\
+    \sem{\Delta,x:A}\,\gamma &= \sem{\Delta}\,\gamma \otimes \sem{A}\gamma \\
+  \end{align*}
+\end{definition}
+
+We define denotations of linear terms in
+\Cref{fig:linear-term-semantics}. Note that the denotations interpret
+typing derivations, not raw terms, as the data of how contexts are
+split is needed in order to construct the correct associator
+functions.
+\max{todo: describe associators, define denotations, prove axioms}
+\begin{figure}
   \begin{mathpar}
@@
 \end{mathpar}
+  \caption{Denotations of linear terms}
+\end{figure}
+
+\begin{definition}
+  For any $\Gamma \vdash \Delta_1,\Delta_2 \isLinCtx$ and $\gamma \in
+  \sem{\Gamma}$ there is an isomorphism $m_{\Delta_1,\Delta_2} :
+  \sem{\Delta_1, \Delta_2}\gamma \cong \sem{\Delta_1}\gamma \otimes
+  \sem{\Delta_2}\gamma$.
+
+  This can be extended to a sequence of contexts of any length.
+\end{definition}
+
+\begin{lemma}[Equations for $\otimes$]
+
+\end{lemma}
+
+\begin{lemma}[Equations for $I$]
+
+\end{lemma}
+
+\begin{lemma}[Equations for $\lto$]
+
+\end{lemma}
+
+\begin{lemma}[Equations for $\tol$]
+
+\end{lemma}
+
 \begin{align*}

### 2024-11-12 paper/paper.tex
intro minor changes

--- 2024-11-11 paper/paper.tex
@@
 \label{sec:intro}
-\steven{This intro sounds reallchoppy but follows the following outline.
-
-  \begin{itemize}
-  \item parsing correctness is imporant because parsing is
-  everywhere and failure
-          could even invalidate other proof efforts.
-  \item verified parsers should be
-          easier to write
-  \item Conal Elliott gives the right idea for what the foundation
-          of parser verification should be --- type theory
-
-  \end{itemize}
-          }
 Formal verification encompasses a broad suite of techniques one may use to
@@
 Large verified developments, like CompCert \cite{leroy_formal_2009},
-invest enormous amounts of person-hours into proof development. In early
+invest a significant number of person-hours into proof development. In early
 versions of CompCert, the apparent
-assurances associated with their effort could indeed be invalidated through
+assurances associated with their effort were at risk of invalidation by
 parser bugs \cite{yangFindingUnderstandingBugs}. Parsing is a ubiquitous task
@@
 \cite{lasserCoStarVerifiedALL2021,firsovCertifiedNormalizationContextFree2015,danielssonTotalParserCombinators2010},
-it is still arduous and highly nontrivial to write and verify a
-parser. Further, distinct approaches to generating verified parsing
-implementations are often relative to a different grammar formalism. We believe
-that these sepearate grammar formalisms can all be unified into a
-syntax-independent notion of grammar, and that this notion of grammar can be
-axiomatized into a syntax that is useful for building verified parsers.
-
-\steven{I don't like how I make the cited work sound bad above}
-
-Elliott \cite{elliottSymbolicAutomaticDifferentiation2021}
-notes that grammars can be described as type-level predicates on strings that prove
+it is still arduous and highly nontrivial to write a verified
+parser. Further, approaches are sensitive to a choice of grammar and automata formalism to serve
+as a
+foundation, which can make it difficult to relate and interoperate different
+styles of verified parsers.
+
+In \cite{elliottSymbolicAutomaticDifferentiation2021}, Elliott describes formal
+grammar as type-level predicates on strings that prove
 language membership. That is, a \emph{formal grammar} $A$ is a function
 $\String \to \Set$ such that for a string $w$, $A~w$ is the set of proofs
-showing that $w$ belongs to the language recognized by $A$. We adopt this
-definition of a formal grammar.
-
-In this paper, we present \theoryname (\theoryabbv) as a domain specific
+showing that $w$ belongs to the language recognized by $A$. This typed approach
+to parsing serves a general setting that can simultaneuously interpret many distinct
+grammar formalisms. In this paper, we adopt Elliott's view on formal grammars and present a
+syntax that axiomatizes this semantic notion of grammar.
+
+We present \theoryname (\theoryabbv) as a domain specific
 type theory for building verified parsers. The type theory itself is a form of dependent linear-non-linear type
@@
 
+Parsers written in \theoryabbv take on a linear functional style, which makes
+them familiar to write and amenable to compositional verification techniques. We
+implement \theoryabbv as a shallowly embedded domain-specific langauge in the
+Cubical Agda proof assistant. Throughout this paper, we will use \agdalogo~ to
+mark results that are mechanized in our Agda development.
+
 \paragraph{Contributions} This paper begins in \cref{sec:type-theory-examples}
-by giving small examples of programs in \theoryabbv so that we may build intuition.
+by studying small example programs from \theoryabbv to build intuition.
 From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type theory
 that axiomatizes the underlying structures needed to conduct formal grammar
-theory. In \cref{sec:applicatoins} we demonstrate the
-applicablity of \theoryabbv for defining parser terms and representing results
-from formal language theory as proof terms in our calculus.
+theory. Later in \cref{sec:applications} we demonstrate the
+applicability of \theoryabbv for relating familiar grammar and automata
+formalisms as well as building concrete parsers.
 Then in \cref{sec:semantics-and-metatheory}, we give a denotational semantics
@@
         linear-non-linear type theory for building verified parsers
-  \item An implementation of \theoryabbv, which is embedded shallowly in Cubical
-        Agda. Throughout this paper, we will use \agdalogo~ to mark results that
-        have been mechanized in our Agda development.
+  \item A denotational grammars semantics for \theoryabbv
+  \item An implementation of \theoryabbv in Agda
 \end{itemize}
@@
 
-In \cref{litNFATterms} we define a family of grammars $A$ and then prove that each $A~s$ is
-equivalent to the accepting traces in $N_{c}$ out of state $s$. Then the desired
-equivalence arises from taking $s = 0$.
+In \cref{fig:litNFAterms} we define a family of grammars $A$ that serve as
+interpretations for each state in $N_{c}$. We then prove that each $A~s$ is
+equivalent to the accepting traces through $N_{c}$ out of state $s$.
 
@@
 
-Note that each $A~s$ is unambiguous, so $(fromN_{c}~s) \circ (toN_{c}~s)$ is equal to the
+Each $A~s$ is unambiguous, so $(fromN_{c}~s) \circ (toN_{c}~s)$ is equal to the
 identity on each $A~s$. To complete the proof of strong equivalence it remains
@@
 Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
-grammars (CFGs). The smallest nontrivial CFG is the Dyck grammar
+grammars (CFGs). An interesting nontrivial CFG is the Dyck grammar
 $\Dyck$ --- the grammar of balanced parentheses.
-
-\pedro{Is it actually the smallest? How do you quantify size of nontrivial things?
-  Might be better to just say ``an interesting nontrivial CFG [...]''}
 
@@
 \end{definition}
-\pa{Do we actually need a definition of formal language morphism?
+\pedro{Do we actually need a definition of formal language morphism?
 In case we do, it should probably be its own definition.}

### 2024-11-12 paper/paper.tex

@@
 
-\newcommand{\SPF}{\texttt{SPF}}
-\newcommand{\Var}{\texttt{Var}}
-\newcommand{\K}{\texttt{K}}
-\newcommand{\map}{\texttt{map}}
-\newcommand{\roll}{\texttt{roll}}
-\newcommand{\fold}{\texttt{fold}}
-\newcommand{\inl}{\texttt{inl}}
-\newcommand{\inr}{\texttt{inr}}
+\newcommand{\Subst}{\textrm{Subst}}
+\newcommand{\SPF}{\mathsf{SPF}}
+\newcommand{\Var}{\mathsf{Var}}
+\newcommand{\K}{\mathsf{K}}
+\newcommand{\map}{\mathsf{map}}
+\newcommand{\roll}{\mathsf{roll}}
+\newcommand{\fold}{\mathsf{fold}}
+\newcommand{\inl}{\mathsf{inl}}
+\newcommand{\inr}{\mathsf{inr}}
 \newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
@@
 
-\max{TODO: prose}
-
-\max{TODO: spelling}
-\begin{theorem}[Adamek]
-  If a functor $F : \mathcal C \to \mathcal C$ preserves
-  $\omega$-colimits and $\mathcal C$ has an initial object and
-  $\omega$-colimits, then $F$ has an initial algebra given by the
-  colimit of the chain
-  \[ 0 \to F(0) \to F(F(0)) \to \cdots \]
+The most complex part of the semantics is the interpretation of
+strictly positive functors and indexed inductive linear types.  We
+interpret a strictly positive functor as a \emph{dependent polynomial
+functor} on the category of sets, also sometimes called an
+\emph{indexed container} \cite{dep-poly,indexed-container}.
+\begin{definition}
+  Let $I$ and $O$ be sets. A (dependent) \emph{polynomial} (of sets)
+  from $I$ to $O$ consists of a set of shapes $S$, a set of positions
+  $P$ and functions $I \leftarrow P \to S \to O$. The \emph{extension}
+  of a polynomial is a functor $\Set/I \to \Set/O$ defined by
+  \max{TODO: fill in details}.
+
+  A dependent polynomial functor from $I$ to $O$ is a functor $\Set/I
+  \to \Set/O$ that is nautrally isomorphic to the extension of a
+  dependent polynomial from $I$ to $O$.
+\end{definition}
+With this interpretation, $\el(F)$ and $\map(F)$ are just interpreted
+as the action of the functor on objects and morphisms,
+respectively. We interpret the constructors $\mathsf{K},\Var$, etc. on
+functors in the obvious way that matches the definitional behavior of
+$\el$ and $\map$. The non-trivial part of the construction is
+verifying that such functors are constructing corresponding dependent
+polynomials. The details are tedious but straightforward extension of
+prior work on dependent polynomials and indexed containers and we have
+verified the construction in Agda.
+
+The reason we use dependent polynomials functors on sets is that it is
+well-known that their initial algebras exist
+\cite{initial-algebras-of-dep-poly}. Additionally, initial algebras
+for dependent polynomials are straightforwardly constructed in Agda as
+an inductive type of ``IW'' trees, which is available as a
+construction in the cubical library of Agda \cite{iw,cubical-library}.
+\begin{theorem}
+  Every dependent polynomial functor from $I$ to $I$ has an initial
+  algebra.
 \end{theorem}
+Then an element $F \in \sem{X \to \SPF\,X}\gamma$ is an
+$\sem{X}\gamma$-indexed family of polynomial functors from
+$\sem{X}\gamma\times \Sigma^*$ to $\Sigma^*$, and taking the product
+of these constructs a polynomial functor from $\sem{X}\gamma\times
+\Sigma^*$ to itself. Then $\sem{\mu\,F}\gamma$ is defined to be the
+initial algebra of this functor, and the initial algebra structure is
+used to interpret $\roll, \fold$ and the corresponding axioms.
 
@@
 
+\begin{definition}
+  The set of (non-linear) substitutions $\gamma \in
+  \textrm{Subst}(\Gamma,\Gamma')$ where $\Gamma \isCtx$ and $\Gamma'
+  \isCtx$ is defined by recursion on $\Gamma$:
+  \begin{align*}
+    \textrm{Subst}(\Gamma,\cdot) &= \{\cdot \}\\
+    \textrm{Subst}(\Gamma,\Gamma',x:A) &= \{ (\gamma,M/x) \pipe \gamma \in \textrm{Subst}(\Gamma,\Gamma') \wedge \Gamma \vdash M : A[\gamma] \}
+  \end{align*}
+  simultaneously with an action of substitution on types, terms,
+  etc. in the standard way.
+
+  \max{do we need identity and composition of substitutions and their functoriality?}
+  %% By induction we can define an identity substitution
+  %% $\textrm{Subst}(\Gamma,\Gamma)$ which maps each variable to
+  %% itself. We write this substitution simply as $\Gamma$.
+  %% \begin{align*}
+  %%   \cdot \circ \gamma' &= \cdot\\
+  %%   (\gamma,M/x) \circ \gamma' &= (\gamma\circ \gamma') , M[\gamma']/x
+  %% \end{align*}
+
+  It is straightforward, but laborious to establish that all forms in
+  the type theory that are parameterized by a non-linear context
+  $\Gamma$ support an admissible action of a substitution $\gamma \in
+  \Subst(\Gamma',\Gamma)$:
+  \begin{mathpar}
+    \inferrule
+    {\Gamma \vdash X \isTy}
+    {\Gamma' \vdash X[\gamma] \isTy}
+
+    \inferrule
+    {\Gamma \vdash X \isSmall}
+    {\Gamma' \vdash X[\gamma] \isSmall}
+
+    \inferrule
+    {\Gamma \vdash X \equiv Y}
+    {\Gamma' \vdash X[\gamma] \equiv Y[\gamma]}
+
+    \inferrule
+    {\Gamma \vdash M : X}
+    {\Gamma' \vdash M[\gamma] : X[\gamma]}
+
+    \inferrule
+    {\Gamma \vdash M \equiv N : X}
+    {\Gamma' \vdash M[\gamma] \equiv N[\gamma] : X[\gamma]}
+
+    \inferrule
+    {\Gamma \vdash \Delta \isLinCtx}
+    {\Gamma' \vdash \Delta[\gamma] \isLinCtx}
+
+    \inferrule
+    {\Gamma \vdash A \isLinTy}
+    {\Gamma' \vdash A[\gamma] \isLinTy}
+
+    \inferrule
+    {\Gamma \vdash A \equiv B}
+    {\Gamma' \vdash A[\gamma] \equiv B[\gamma]}
+
+    \inferrule
+    {\Gamma;\Delta \vdash e : A}
+    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] : A[\gamma]}j
+
+    \inferrule
+    {\Gamma;\Delta \vdash e \equiv f : A}
+    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] \equiv f[\gamma] : A[\gamma]}j
+  \end{mathpar}
+
+  %% Further, these all satisfy identity and composition equations M[\Gamma] = M and M[\gamma][\gamma'] = M[(\gamma \circ \gamma')]
+\end{definition}
+
+\begin{definition}
+  Let $\Gamma \vdash \Delta \isLinCtx$ and $\Gamma \vdash \Delta'
+  \isLinCtx$. The set of linear substitutions $\Subst(\Delta',\Delta)$
+  is defined by recursion on $\Delta$:
+  \begin{align*}
+    \Subst(\Delta',\cdot) &= \{ \cdot \pipe \Delta' = \cdot \}\\
+    \Subst(\Delta',(\Delta,a:A)) &= \{ (\delta, e/a) \pipe \delta \in \Subst(\Delta_1, \Delta) \and \Delta_2 \vdash e : A \and \Delta' = \Delta_1,\Delta_2\}
+  \end{align*}
+
+  Given substitutions $\delta_1 \in \Subst(\Delta_1', \Delta_1)$ and
+  $\delta_2 \in \Subst(\Delta_2', \Delta_2)$, we can define a
+  substitution $\delta_1,\delta_2 \in
+  \Subst((\Delta_1',\Delta_2'),(\Delta_1,\Delta_2))$. Furthermore, for
+  any substitution $\delta \in \Subst(\Delta,(\Delta_1,\Delta_2))$, we
+  can deconstruct $\delta = \delta_1,\delta_2$ with $\delta_1 \in
+  \Subst(\Delta_1', \Delta_1)$ and $\delta_2 \in \Subst(\Delta_2',
+  \Delta_2)$.
+  %% Again for any $\Gamma \vdash \Delta$, we can define an identity
+  %% substitution mapping each variable to itself. Additionally we can define composition of 
+\end{definition}
+
+\begin{definition}
+  Given any $\Gamma ; \Delta \vdash e : A$ and $\delta \in
+  \Subst(\Delta',\Delta)$, we define the action of the substitution on
+  $e$ as follows, frequently using the inversion principle to split
+  the substitution into constituent components.
+  \begin{align*}
+    a[e/a] &= e\\
+    (e_1,e_2)[\delta_1,\delta_2] &= (e_1[\delta_1], e_2[\delta_2])\\
+    (\letin {(a , b)} e {e'})[\delta_1,\delta_2,\delta_3] &= \letin{(a,b)} {e[\delta_2]}{e'[\delta_1,a/a,b/b,\delta_2]}\\
+    ()[\cdot] &= ()\\
+    \letin{()} e {e'}[\delta_1,\delta_2,\delta_3] &= \letin{()} {e[\delta_2]}{e'[\delta_1,a/a,b/b,\delta_2]}\\
+    (\lamblto {a} e)[\delta] &= \lamblto a {e[\delta,a/a]}\\
+    (\applto {e'} {e})[\delta_1,\delta_2] &= \applto {e'[\delta_1]} {e[\delta_2]}\\
+    (\lamblto {a} e)[\delta] &= \lamblto a {e[\delta,a/a]}\\
+    (\applto {e'} {e})[\delta_1,\delta_2] &= \applto {e'[\delta_1]} {e[\delta_2]}\\
+    (\lambtol {a} e)[\delta] &= \lamblto a {e[a/a,\delta]}\\
+    (\apptol {e'} {e})[\delta_1,\delta_2] &= \apptol {e'[\delta_1]} {e[\delta_2]}\\
+    (\dlamb x e)[\delta] &= \dlamb x {e[\delta]}\\
+    (e\,.\pi\,M)[\delta] &= (e[\delta]\,.\pi\,M)\\
+    (\sigma\,M\,e)[\delta] &= \sigma\,M\,e[\delta]\\
+    (\letin{\sigma\,x\,a}{e}{e'})[\delta_1,\delta_2,\delta_3] &= \letin{\sigma\,x\,a} {e[\delta_2]}{e'[\delta_1,\delta_3]}\\
+    (\equalizerin{e})[\delta] &= \equalizerin{e[\delta]}\\
+    (\equalizerpi{e})[\delta] &= \equalizerpi{e[\delta]}
+  \end{align*}
+  By induction on linear term and equality judgments, we establish the
+  following admissible rules for $\delta \in \Subst(\Delta',\Delta)$:
+  \begin{mathpar}
+    \inferrule
+    {\Gamma; \Delta \vdash e : A}
+    {\Gamma; \Delta' \vdash e[\delta] : A}
+
+    \inferrule
+    {\Gamma; \Delta \vdash e \equiv f : A}
+    {\Gamma; \Delta' \vdash e[\delta] \equiv f[\delta'] : A}
+  \end{mathpar}
+
+  \max{identity and composition?}
+\end{definition}
+
+\section{Denotational Semantics}
+
+\begin{definition}
+  We define the unit grammar
+  \[ I\,w = \{ ()\pipe w = \varepsilon \}\]
+  and given grammars $A$ and $B$ we define their tensor product
+  \[ (A \otimes B)\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in A\,w_1 \wedge b \in B\,w_2 \} \]
+  This extends to a monoidal structure on the category of grammars.
+Note that we defined in the semantics of linear types $\sem{I}\gamma = I$ and $\sem{A \otimes B}\gamma = \sem{A}\gamma \otimes \sem{B} \gamma$
+\end{definition}
+
+\begin{definition}[Denotation of Linear Contexts]
+  The semantics of linear contexts $\Gamma \vdash \Delta \isLinCtx$ is
+  defined as follows:
+  \max{make sure this aligns with our interpretation of $\lto$/$\tol$}
+  \begin{align*}
+    \sem{\cdot}\,\gamma &= I \\
+    \sem{\Delta,x:A}\,\gamma &= \sem{\Delta}\,\gamma \otimes \sem{A}\gamma \\
+  \end{align*}
+\end{definition}
+
+We define denotations of linear terms in
+\Cref{fig:linear-term-semantics}. Note that the denotations interpret
+typing derivations, not raw terms, as the data of how contexts are
+split is needed in order to construct the correct associator
+functions.
+\max{todo: describe associators, define denotations, prove axioms}
+\begin{figure}
+  \begin{mathpar}
+  \inferrule
+  {\Delta_1 \vdash e_1 : A_1 \and \Delta_2 \vdash e_2 : A_2}
+  {\sem{(e_1,e_2)}\gamma\,w\,\delta
+    = \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}
+  }
+
+  \inferrule
+  {\Delta_2 \vdash e : A \otimes B \and \Delta_1,a:A,b:B,\Delta_3 \vdash e' : C}
+  {\sem{\letin{(a,b)}{e}{e'}}\gamma\,w\,\delta =
+    \letin{(w_1,w_2w_3,\delta_1,(w_2,w_3,\delta_2,\delta_3))}{m_{\Delta_1,\Delta_2,\Delta_3}(\delta)}
+    \letin{(w_{2,1},w_{2,2},a,b)}{\sem{e}\gamma\,w_2\,\delta_2}
+    {\sem{e'}\gamma\,w\,(w_1,w_{2,1}w_{2,2}w_3, \delta_1, (w_{2,1},w_{2,2}w_3, a, (w_{2,2},w_3, b, \delta_3)))}
+    }
+\end{mathpar}
+  \caption{Denotations of linear terms}
+\end{figure}
+
+\begin{definition}
+  For any $\Gamma \vdash \Delta_1,\Delta_2 \isLinCtx$ and $\gamma \in
+  \sem{\Gamma}$ there is an isomorphism $m_{\Delta_1,\Delta_2} :
+  \sem{\Delta_1, \Delta_2}\gamma \cong \sem{\Delta_1}\gamma \otimes
+  \sem{\Delta_2}\gamma$.
+
+  This can be extended to a sequence of contexts of any length.
+\end{definition}
+
+\begin{lemma}[Equations for $\otimes$]
+
+\end{lemma}
+
+\begin{lemma}[Equations for $I$]
+
+\end{lemma}
+
+\begin{lemma}[Equations for $\lto$]
+
+\end{lemma}
+
+\begin{lemma}[Equations for $\tol$]
+
+\end{lemma}
+
+\begin{align*}
+  \sem{(e_1,e_2)}\gamma\,w\,\delta &=
+  \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}\\
+\end{align*}
+
 \end{document}

### 2024-11-12 paper/paper.tex
fix label typo
new draft of abstract and intro

@@
 \begin{abstract}
-  We propose \theoryname~(\theoryabbv), a domain-specific dependent type theory
-  for the definition of intrinsically verified parsers and verification of weak
-  and strong equivalences of formal grammars. The linear types in \theoryabbv
-  have a simple interpretation as formal grammars, with values of the types
-  interpreted as parse trees and general terms as a form of intrinsically
-  verified parse transformers, of which parsers being a special case. We
-  demonstrate the expressivity of this system by showing that the indexed
-  inductive linear types in this calculus can be used to encode many different
-  familiar grammar formalisms, including regular and context-free grammars but
-  also traces of various types of automata. We can then construct strong
-  equivalences between grammar and automata formalisms and construct verified
-  parsers as simple linear functional programs. All of our examples have been
-  implemented in a prototype implementation of \theoryabbv as a shallow
-  embedding in Agda.
+  We present \theoryname~(\theoryabbv), a domain-specific dependent
+  type theory for verified parsing and formal grammar theory. In
+  \theoryabbv, linear types are used as a syntax for formal grammars,
+  and parsers can be written as linear terms. The linear typing
+  restriction provides a form of intrinsic verification that a parser
+  yields only valid parse trees for the input string. We demonstrate
+  the expressivity of this system by showing that the combination of
+  inductive linear types and dependency on non-linear data can be used
+  to encode commonly used grammar formalisms such as regular and
+  context-free grammars as well as traces of various types of
+  automata. Using these encodings, we define parsers for regular
+  expressions using deterministic automata, as well as providing
+  examples parsers for LL(1) and LR(1) grammars.
+
+  We present a denotational semantics of our type theory that
+  interprets the types as a mathematical notion of formal
+  grammars. Based on this denotational semantics, we have made a
+  prototype implementation of \theoryabbv using a shallow embedding in
+  the Agda proof assistant. All of our examples parsers have been
+  implemented in this prototype implementation.
 \end{abstract}
@@
 \label{sec:intro}
-Formal verification encompasses a broad suite of techniques one may use to
-mathematically guarantee the correctness of software systems. We can only
-reap the benefits of
-any proof effort if its underlying assumptions hold. For instance, verifying the
-functional correctness of an algorithm is no use if it is run on buggy hardware.
-Likewise, the promised guarantees of a program may be jeopardized if it is
-parsed incorrectly.
-
-Large verified developments, like CompCert \cite{leroy_formal_2009},
-invest a significant number of person-hours into proof development. In early
-versions of CompCert, the apparent
-assurances associated with their effort were at risk of invalidation by
-parser bugs \cite{yangFindingUnderstandingBugs}. Parsing is a ubiquitous task
-throughout computing and yet parsers are frequently unverified, even in
-otherwise verified systems. The CompCert team later
-developed a method for validating the correctness of LR parsers
-\cite{jourdanValidatingLRParsers2012} and then replaced their parser with one
-that is certified to be correct. Despite their success in  validating their C parser and other work in the
-generation of  verified parsers
-\cite{lasserCoStarVerifiedALL2021,firsovCertifiedNormalizationContextFree2015,danielssonTotalParserCombinators2010},
-it is still arduous and highly nontrivial to write a verified
-parser. Further, approaches are sensitive to a choice of grammar and automata formalism to serve
-as a
-foundation, which can make it difficult to relate and interoperate different
-styles of verified parsers.
-
-In \cite{elliottSymbolicAutomaticDifferentiation2021}, Elliott describes formal
-grammar as type-level predicates on strings that prove
-language membership. That is, a \emph{formal grammar} $A$ is a function
-$\String \to \Set$ such that for a string $w$, $A~w$ is the set of proofs
-showing that $w$ belongs to the language recognized by $A$. This typed approach
-to parsing serves a general setting that can simultaneuously interpret many distinct
-grammar formalisms. In this paper, we adopt Elliott's view on formal grammars and present a
-syntax that axiomatizes this semantic notion of grammar.
-
-We present \theoryname (\theoryabbv) as a domain specific
-type theory for building verified parsers. The type theory itself is a form of dependent linear-non-linear type
-theory: the linear types can be interpreted as grammars, whereas the
-non-linear types can be interpreted as ordinary sets. As in prior work
-on dependent linear-non-linear types, the linear types are allowed to
-depend on the non-linear types but not vice-versa \cite{krishnaswami_integrating_2015}.
+Parsing structured data from untrusted input is a ubiquitous task in
+computing. Any formally verified software system that interacts with
+the outside world must contain some parsing component. For example, in
+an extensive experiment finding bugs in C compilers
+\cite{yangFindingUnderstandingBugs}, an early version of the formally
+verified CompCert C compiler only contained bugs in the then
+unverified parsing component \cite{leroy_formal_2009}. Bugs in parsers
+undermine the overall correctness theorem for a verified system: an
+incorrectly parsed C program will be compiled correctly but this is
+not very useful if it did not correctly correspond to the actual
+source program. Eventually, a correct parser was implemented using an
+automaton that is formally verified to implement an LR grammar
+\cite{jourdanValidatingLRParsers2012}.
+
+It is entirely understandable from an engineering perspective
+\emph{why} verified parsing was not part of the initial releases of
+CompCert: parsing algorithms and formal grammars are a complex area,
+featuring a variety of domain-specific formalisms such as context-free
+grammars and various automata. These formalisms have little relation
+to the main components of a verified compiler. For this reason, it is
+advantageous for verified parsers to be implemented using a reusable
+verified library, just as parser generators and regular expression
+matchers have done for many decades in unverified software.
+
+Prior approaches to verified parsing focus on verification of a
+particular grammar formalism such as \max{Steven write down what the
+  related work uses
+  here}\cite{lasserCoStarVerifiedALL2021,firsovCertifiedNormalizationContextFree2015,danielssonTotalParserCombinators2010}.
+Each new grammar formalism must then be extended with its own verified
+implementation.
+
+In this work, we present the design of \theoryname (\theoryabbv), a
+domain-specific language for formal verification of parsers. A key
+property is that \theoryabbv is an \emph{extensible} framework for
+verification of parsers in that it supports the definition of grammar
+formalisms of unrestricted complexity. That is, \theoryabbv is not a
+system for verifying \emph{one} type of grammar formalism, but instead
+is a domain-specific language in which many grammar formalisms and
+their verified parsers can be implemented. For example, \theoryabbv is
+not a verified parser generator that compiles regular expressions to
+deterministic finite automata, but is instead is a domain specific
+language \emph{for writing} such a verified parser generator.
+
+\max{Steven, is this accurate comparison?} The design of \theoryabbv
+is an extension of Joachim Lambek's \emph{categorial grammars}
+\cite{lambek}. Categorial grammars are a grammar formalism equivalent
+in expressive power to context-free grammars that in modern
+terminology would be considered a kind of \emph{non-commutative linear
+logic}, a version of linear logic where the tensor product is not
+commutative, reflecting the obvious property that the relative
+ordering of characters is significant in parsing problems. We extend
+non-commutative linear logic with two key components that increase its
+power to support arbitrarily powerful grammar formalisms: inductive
+linear types, as well as dependency of linear types on non-linear
+data. The resulting system has two kinds of types: non-linear types
+which model sets and linear types which model formal
+grammars. Crucially, the non-linear types and linear types are allowed
+to be dependent on non-linear types, but not on linear types. This
+combination has been used previously in the ``linear-non-linear
+dependent'' type theory with \emph{commutative} linear logic to model
+imperative programming \cite{krishnaswami_integrating_2015}.
 
@@
 that derivations of a term type checking carry intrinsic proofs of correctness.
-
 Parsers written in \theoryabbv take on a linear functional style, which makes
-them familiar to write and amenable to compositional verification techniques. We
-implement \theoryabbv as a shallowly embedded domain-specific langauge in the
-Cubical Agda proof assistant. Throughout this paper, we will use \agdalogo~ to
-mark results that are mechanized in our Agda development.
-
-\paragraph{Contributions} This paper begins in \cref{sec:type-theory-examples}
-by studying small example programs from \theoryabbv to build intuition.
-From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type theory
-that axiomatizes the underlying structures needed to conduct formal grammar
-theory. Later in \cref{sec:applications} we demonstrate the
-applicability of \theoryabbv for relating familiar grammar and automata
-formalisms as well as building concrete parsers.
-Then in \cref{sec:semantics-and-metatheory}, we give a denotational semantics
-that makes precise the connection between \theoryabbv syntax and formal grammars.
-
-Our contributions are then:
+them familiar to write and amenable to compositional verification techniques.
+
+To show the feasibility of our design, we have implemented \theoryabbv
+as a shallowly embedded domain-specific langauge in the Cubical Agda
+proof assistant \cite{agda,cubical}. We have implemented many example
+grammars and parsers in our system including regular expressions,
+non-deterministic and deterministic automata, as well as some example
+context-free grammars and parsers based on LL(1) and LR(1)
+automata. Throughout this paper, we will use \agdalogo~ to mark
+results that are mechanized in our Agda development.
+
+Our Agda prototype is based on a \emph{denotational semantics} of
+\theoryabbv. The core idea of the denotational semantics is
+\cite{elliottSymbolicAutomaticDifferentiation2021}, Elliott describes
+a formal grammar as type-level predicates on strings that prove
+language membership. That is, a \emph{formal grammar} $A$ is a
+function $\String \to \Set$ such that for a string $w$, $A~w$ is the
+set of ``proofs'' showing that $w$ belongs to the language recognized
+by $A$. We show that all linear types in \theoryabbv can be so
+interpreted as an abstract formal grammar in this sense, and that
+linear terms are a kind of \emph{parse transformer}, a function that
+takes a parse tree from one grammar to a parse tree in a different
+grammar but over the same underlying string.
+
+%% Formal verification encompasses a broad suite of techniques one may use to
+%% mathematically guarantee the correctness of software systems. We can only
+%% reap the benefits of
+%% any proof effort if its underlying assumptions hold. For instance, verifying the
+%% functional correctness of an algorithm is no use if it is run on buggy hardware.
+%% Likewise, the promised guarantees of a program may be jeopardized if it is
+%% parsed incorrectly.
+
+Our contributions are then:\max{This contribution section needs to be punchier}
 %
 \begin{itemize}
-  \item \theoryname (\theoryabbv): A dependent
-        linear-non-linear type theory for building verified parsers
-  \item A denotational grammars semantics for \theoryabbv
-  \item An implementation of \theoryabbv in Agda
+  \item The design of \theoryname (\theoryabbv): A dependent
+    linear-non-linear type theory for building verified parsers, which
+    extends prior work on dependent linear-non-linear type theory to
+    support inductive linear types
+  \item Demonstration of how to encode many common grammar formalisms
+    (regular expressions, (non-)deterministic automata, context-free
+    grammars) and parser formalisms within our type theory.
+  \item A prototype implementation of \theoryabbv in Agda with all
+    examples mechanized.
+  \item A denotational semantics for \theoryabbv that shows that the
+    parsers are in fact verified to be correct and soundness of the
+    equational theory.
 \end{itemize}
+
+This paper begins in \cref{sec:type-theory-examples} by studying small
+example programs from \theoryabbv to build intuition.  From there, in
+\cref{sec:tt} we provide the syntax, typing and equational theory of
+\theoryname. In \cref{sec:applications} we demonstrate the
+applicability of \theoryabbv for relating familiar grammar and
+automata formalisms as well as building concrete parsers.  Then in
+\cref{sec:semantics-and-metatheory}, we give a denotational semantics
+that makes precise the connection between \theoryabbv syntax and
+formal grammars. Finally in \Cref{sec:discussion} we discuss related
+and future work.
 
@@
 \newcommand{\states}{\mathtt{states}}
-\newcommand{\labelt}{\mathtt{label}}
+\newcommand{\labelt}{label}
 \newcommand{\transitions}{\texttt{transitions}}

### 2024-11-13 paper/paper.tex
edits to section 3

@@
 \label{sec:type-theory-examples}
-To gain intuition for working in \theoryabbv, we begin with some illustrative
-examples drawn from the theory of formal languages. In particular, we
-demonstrate how to encode finite grammars, regular expressions, and finite
-automata in \theoryabbv. Each of our examples will be defined for strings over the three character alphabet
-$\Sigma = \{ \texttt{a} , \texttt{b}, \texttt{c} \}$.
+To gain intuition for working in \theoryabbv, we begin with some
+illustrative examples drawn from the theory of formal languages. Each
+of our examples will be defined for strings over the three character
+alphabet $\Sigma = \{ \texttt{a} , \texttt{b}, \texttt{c} \}$.
 
@@
 empty string, and the empty grammar.
-For instance, the grammar corresponding to the character $a$ is given by $\literal a$ which
+For each character $a$ in the alphabet we have a type $\literal a$ which
 has a single parse tree for the string
-\stringquote{a} and no parse trees at any other strings. The grammar $\I$ recognizes has a single
+\stringquote{a} and no parse trees at any other strings. The grammar $\I$ has a single
 parse tree for the empty string $\epsilon = \stringquote{}$ and no parses for any other strings.
-The final base type, the empty grammar $\bot$, has no parses for any string. We
+The final base type, the empty grammar $0$, has no parses for any string. We
 use type-theoretic syntax to represent disjunction $\oplus$ and concatenation
@@
 
-For a grammar $\A$, a parse tree of a string $\w$ can be represented as
-a term of type $\A$ in the context $\internalize \w$, where
+For a type $\A$, a parse tree of a string $\w$ is represented as a
+term of type $\A$ in the context $\internalize \w$, where
 $\internalize \w$ is a context with one variable for each character of
 $\w$. For example, to define a parse tree for \stringquote{ab}, we use
-the context $\internalize{\stringquote{ab}} = a : \literal a , b : \literal b$. Then the left of
-\cref{fig:fingram} gives a typing derivation that shows that $\inl(a
-\otimes b)$ defines a valid parse tree for the finite grammar
-$(\literal a \otimes \literal b)
-\oplus \literal c$. Underneath the derivation in \cref{fig:fingram}, we give a linear $\lambda$-term
-that implements the program given in the derivation.
+the context $\internalize{\stringquote{ab}} = a : \literal a , b :
+\literal b$. On the left side of \cref{fig:fingram}, we give a typing
+derivation that shows that $\inl(a \otimes b)$ defines a valid parse
+tree for the finite grammar $(\literal a \otimes \literal b) \oplus
+\literal c$. Underneath the derivation in \cref{fig:fingram}, we give
+a linear $\lambda$-term that implements the program given in the
+derivation.\max{reverse the order of these?}
 
@@
 \end{figure}
-For this interpretation of terms as parse trees to make sense, our
+For this interpretation of parse trees as terms to make sense, our
 calculus cannot allow for \emph{any} of the usual structural rules of
-a type theory: weakening, contraction and exchange. Weakening allows
+type theory: weakening, contraction and exchange. Weakening allows
 for variables to go unused, while contraction allows for the same
@@
 $\oplus$, and $\otimes$, and the Kleene star $(\cdot)^{\ast}$.  For a
-grammar $\A$, we can define the Kleene star $\A^{*}$ as a particular
+grammar $\A$, we define the Kleene star $\A^{*}$ as a particular
 \emph{inductive linear type} of linear lists, as shown in
@@
 
-Through repeated application of the Kleene star constructors, \cref{fig:kleenestarderivation} gives a derivation that shows \stringquote{ab} matches the
-regular expression $({\literal a}^{*} \otimes \literal b) \oplus \literal c$. The leaves of the proof tree
-that mention the arrow $\uparrow$ describe a cast from a non-linear type to a linear type.
-For instance, the premise of the leaf involving $\nil$
-views $\nil : \ltonl {({\literal a}^{*})}$ as the name of a
-constructor, and a constructor should be nonlinearly valued because we may call
-it several time. However, the conclusion of this leaf views $\nil : {\literal a}^{*}$ as a
-linear value. After we call the constructor it returns a value that may only be
-used a single time. In this paper such casts are often made implicitly, but this
-does not introduce any ambiguity.
-Even though we use $\nil$ to refer to both a constructor and the
-linear value produced by it
-, we differentiate between the two views by inspecting
-the type.
+Through repeated application of the Kleene star constructors,
+\cref{fig:kleenestarderivation} gives a derivation that shows
+\stringquote{ab} matches the regular expression $({\literal a}^{*}
+\otimes \literal b) \oplus \literal c$. The leaves of the proof tree
+that mention the arrow $\uparrow$ describe a cast from a non-linear
+type to a linear type.  For instance, the premise of the leaf
+involving $\nil$ views $\nil : \ltonl {({\literal a}^{*})}$ as the
+name of a constructor, and a constructor should be nonlinearly valued
+because we may call it several times (or not at all). However, the
+conclusion of this leaf views $\nil : {\literal a}^{*}$ as a linear
+value, which in our syntax is an implicit coercion from a nonlinear
+value to a linear value. After we call the constructor it ``returns''
+a value that may only be used a single time.
 
@@
 We may also have derivations where the term in context is not simply a
-string of base types. In \cref{fig:kleeneabstractproof} we show that every parse
+string of literals. In \cref{fig:kleeneabstractproof} we show that every parse
 of the grammar $(A \otimes A)^{*}$ induces a parse of $A^{*}$ for an
@@
 not quite appropriate
-to treat linear terms here as parse \textit{trees}.
+to think of a linear term here as a parse \textit{trees}.
 The
@@
 of \cref{fig:kleeneabstractproof} by pattern matching on the input and making an
-explicit recursive call in the body of its definition. When appropriate we will favor this code-like
-presentation of terms in our calculus, as it is more readable and more
-faithful to the implementation of \theoryabbv in Agda.
+explicit recursive call in the body of its definition.
 
@@
 Regular expressions are a compact formalism for defining a formal grammar,
-however simply defining a regular expression --- such as
-$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ --- does not give any insight into how to parse
-it. Many parsing algorithms are based on a notion of automaton, and to verify
-these algorithms we need to represent their associated automata in
-\theoryabbv.
-
+but an expression such as
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ does not give a very operational perspective of how to parse
+it. For this reason, most parsers are based on compiling a grammar to a corresponding notion of automaton, which is readily implemented. To impleemnt
+these algorithms in \theoryabbv, we need a way to represent automata as types in the same way we can represent regular expressions.
+
+\max{combine these three figures into one?}
 Finite automata are precisely the class of machines that recognize regular
@@
 finite automaton (NFA) for the regular expression
-$({\literal a}^{*} \otimes \literal b) \oplus \literal c$.
-\cref{fig:nfainductive} defines $\Trace$ an \emph{indexed} inductive linear type of
-traces through this automaton. Defining an indexed inductive type can be thought
-of like defining a family of mutually inductive types varying coherently over a
-type of indexing data. That is, \cref{fig:nfainductive} defines the linear types
-$\Trace~\0$,
-$\Trace~\1$, and $\Trace~\2$ mutually inductively. The index $\s : \texttt{Fin 3}$ picks out which
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$, along with a type $\Trace$, an \emph{indexed} inductive linear type of traces through this automaton.
+Defining an indexed inductive type can be thought
+of as defining a family of mutually recursive inductive types, one for each element of the indexing type. Here $\Trace$ is 
+uses an index $\s : \texttt{Fin 3}$ which picks out which
 state in the automaton a trace begins at --- where $\texttt{Fin 3}$ is the
-finite type containing inhabitants $\{\0 , \1 , \2\}$.
-There are three kinds of constructors for $\Trace$:
-(1) those that terminate traces, (2) those that correspond to transitions
-labeled by a character, and (3) those that correspond to transitions labeled by
-the empty string $\epsilon$. The constructor $\texttt{stop}$ terminates a trace
-in the accepting state $\2$. The constructors $\texttt{1to1}$, $\texttt{1to2}$,
-$\texttt{0to2}$ each define a labeled transition through the NFA, and each of
-these consumes a parse of the label's character and a trace beginning at the
-destination of a transition to produce a trace beginning at the source of a
-transition. The constructor $\texttt{0to1}$ behaves similarly, except its
-transition is labeled with the empty string $\epsilon$. Therefore,
-$\texttt{0to1}$ takes in a trace beginning a state $\1$ and returns a trace
-beginning at state $\0$ corresponding to the same underlying string.
+finite type containing inhabitants $\{\0 , \1 , \2\}$. We can think of this as defining three mutually recursive inductive types $\Trace~\0$,
+$\Trace~\1$, and $\Trace~\2$.
+%
+There are three kinds of constructors for $\Trace$: (1) those that
+terminate traces, (2) those that correspond to transitions labeled by
+a character, and (3) those that correspond to transitions labeled by
+the empty string $\epsilon$. The constructor $\texttt{stop}$
+terminates a trace in the accepting state $\2$. The constructors
+$\texttt{1to1}$, $\texttt{1to2}$, $\texttt{0to2}$ each define a
+labeled transition through the NFA, and each of these consumes a parse
+of the label's character and a trace beginning at the destination of a
+transition to produce a trace beginning at the source of a
+transition. The constructor $\texttt{0to1}$ behaves similarly, except
+its transition is labeled with the empty string $\epsilon$. Therefore,
+$\texttt{0to1}$ takes in a trace beginning a state $\1$ and returns a
+trace beginning at state $\0$ corresponding to the same underlying
+string.
 
@@
 accepted by the NFA representation of this regular expression. In particular, we
-show that \stringquote{ab} matches the grammar $\Trace~0$ because $\0$ is the
-initial state of the NFA.\ We may imagine defining parse transformers from
-$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ to $\Trace~\1$, and vice versa. The existence of
-such maps between the two types would demonstrate that the regular expression
+show that \stringquote{ab} matches the grammar $\Trace~0$ since $\0$ is the
+initial state of the NFA.\ If we construct parse transformers from
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ to $\Trace~\0$, and vice versa, we would demonstrate that the regular expression
 and the automaton capture the same language --- i.e.\ they recognize the same
-set of strings. As we will see in LATER SECTION, we can define parse
+set of strings. As we will see in \cref{sec:applications}, we can indeed define parse
 transformers between regular expressions and their associated automaton.
@@
 parse of a regular expression and the traces of their associated automata
-capture precisely the same data.
+capture precisely the same data, i.e., that they are \emph{strongly equivalent} as grammar formalisms.
 
@@
   \end{tikzpicture}
-  \caption{NFA for $(\a^{*} \otimes \b) \oplus \c$}
-  \label{fig:exampleNFA}
-\end{figure}
-
-\begin{figure}
   \begin{lstlisting}
@@
 \end{lstlisting}
-\caption{$N$ as an indexed inductive type}
-\label{fig:nfainductive}
+  \caption{NFA for $(\a^{*} \otimes \b) \oplus \c$ and its corresponding type}
+  \label{fig:exampleNFA}
 \end{figure}
@@
 connectives by using indexing over the empty and boolean type
-respectively. We will freely use $\bot$ to refer to this empty
+respectively. We will freely use $0$ to refer to this empty
 disjunction and $\top$ to refer to the empty conjunction. Lastly, we
@@
 subtheory of \theoryabbv generated by
-$\otimes$, $\oplus$, and Kleene star over the types $\bot$, $I$, and characters drawn from $\Sigma$.
+$\otimes$, $\oplus$, and Kleene star over the types $0$, $I$, and characters drawn from $\Sigma$.
 
@@
 Further, we are eliding the needed NFAs for the cases where the regular
-expression is $I$, $\bot$, a Kleene star, or a concatenation. All of these
+expression is $I$, $0$, a Kleene star, or a concatenation. All of these
 elided details may be found in the Agda formalization.
@@
 define a predicate $accepts : \stringNL \to U$ such that $accept~w$ is equal to $\top$ if
-$T$ accepts $w$ and equal to $\bot$ otherwise. Then, $Reify~accepts$ is a linear
+$T$ accepts $w$ and equal to $0$ otherwise. Then, $Reify~accepts$ is a linear
 type that captures precisely the strings accepted by $Tur$. That is,
@@
 %%   c\lto c &\equiv I\\
-%%   c\lto d &\equiv \bot\\
-%%   c\lto I &\equiv \bot\\
-%%   c\lto 0 &\equiv \bot\\
+%%   c\lto d &\equiv 0\\
+%%   c\lto I &\equiv 0\\
+%%   c\lto 0 &\equiv 0\\
 %%   c\lto (A \otimes B) & \equiv (c\lto A) \otimes B + (A \amp I) \otimes (c\lto B)\\
@@
 %% \steven{Definable as nullary sum}
-%% The empty type $\bot$ has no inhabitants. The elimination for the empty type
-%% witnesses the principle of explosion --- i.e.\ from a term of type $\bot$ we may
+%% The empty type $0$ has no inhabitants. The elimination for the empty type
+%% witnesses the principle of explosion --- i.e.\ from a term of type $0$ we may
 %% introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.
 %% %
-%% \[ \sem {\bot} \gamma w = \emptyset \]
+%% \[ \sem {0} \gamma w = \emptyset \]
 %% %

### 2024-11-13 paper/paper.tex
start editing section 4. Done for the night

@@
 
-The syntax of
-\theoryabbv is based on the dependent linear-non-linear (\lnld)
-calculus from \cite{krishnaswami_integrating_2015}. As in \lnld, \theoryabbv includes both
-non-linear dependent types, as well as linear types, which are allowed
-to depend on the non-linear types, but not on other linear types.
-%
-The main point of departure from \lnld's design in that the linear
+The design of \theoryabbv is based the dependent
+linear-non-linear calculus (\lnld) and Lambek calculus, also known as
+non-commutative linear logic.
+\cite{krishnaswami_integrating_2015,lambekCalculus}. As in \lnld,
+\theoryabbv includes both non-linear dependent types, as well as
+linear types, which are allowed to depend on the non-linear types, but
+not on other linear types.
+%
+The main point of departure from \lnld's design is that, as in Lambek calculus \cite{lambekCalculus, the linear
 typing is \emph{non-commutative}, i.e., that exchange is not an
@@
 %
-Finally, we add base types and axioms to the system specifically to
-model formal grammars and parsing.
+Finally, while \lnld was enhanced with special connectives
+inspired by separation logic to model imperative programming, we
+instead add base types and axioms to the system specifically to model
+formal grammars and parsing.
 
@@
 small non-linear types and $L$ of linear types. These are defined as
-``universes ala Coquand'' in that we define a judgment saying when a
+universes ``ala Coquand'' in that we define a judgment saying when a
 non-linear type is small and define the universe to internalize
-precisely this judgment. These universe types are needed so that we
-can define types by recursion on natural numbers. Next, we include a
-non-linear type $\ltonl A$ where $A$ is a linear type. The intuition
-for this type is that its elements are the linear terms that are
-``resource free'': its introduction rule says we can construct an
-$\ltonl A$ when we have a linear term of type $A$ with no free linear
-variables. This type is used extensively in our examples, playing a
-similar role to the $!$ modality of ordinary linear logic.
-\pedro{We should also reference the persistence modality $\square$
-from separation logic}
+precisely this judgment \cite{universesAlaCoquand}. These universe
+types are needed so that we can define types by recursion on natural
+numbers. Next, we include a non-linear type $\ltonl A$ where $A$ is a
+linear type. The intuition for this type is that its elements are the
+linear terms that are ``resource free'': its introduction rule says we
+can construct an $\ltonl A$ when we have a linear term of type $A$
+with no free linear variables. This type is used extensively in our
+examples, playing a similar role to the $!$ modality of ordinary
+linear logic or the persistence modality $\square$ of separation logic
+\cite{linearlogic,persistencemodalityseparationlogic}.
 

### 2024-11-13 paper/paper.tex
fix typos in inferrule
some notes

@@
 \subsection{Non-linear Typing}
-
+\max{todo: include these rules in the appendix}
 The non-linear types of \theoryabbv include the standard dependent
@@
     \and
-    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash M : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {e'} x}
+    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash M : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {M} x}
     %
@@
     %
-    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A e x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
+    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A M x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
     %
@@
 
-\section{Denotational Semantics}
+\section{Denotational Semantics and Implementation}
 \label{sec:semantics-and-metatheory}

### 2024-11-13 paper/paper.tex
draft of syntax section

@@
 \newcommand{\String}{\textbf{String}}
+\newcommand{\Char}{\textbf{Char}}
 \newcommand{\stringg}{\texttt{String}}
@@
 %
-The main point of departure from \lnld's design is that, as in Lambek calculus \cite{lambekCalculus, the linear
+The main point of departure from \lnld's design is that, as in Lambek calculus \cite{lambekCalculus}, the linear
 typing is \emph{non-commutative}, i.e., that exchange is not an
@@
 \subsection{Non-linear Typing}
-\max{todo: include these rules in the appendix}
 The non-linear types of \theoryabbv include the standard dependent
 types for $\Pi,\Sigma$, extensional equality, natural numbers,
-booleans, unit and empty types. We include these rules in the
-appendix. We present the other non-linear type constructions in
+booleans, unit and empty types, and we assume function
+extensionality\cite{standard-type-theory-reference}. We present the
+other non-linear type constructions in
 \Cref{fig:non-linear-stuff}. First, we include universe types $U$ of
@@
 \end{figure}
+\max{todo: decide if we shoudl keep all of these rules or not}
 
@@
 First, we cover the ``multiplicative'' connectives of non-commutative
-linear logic. The linear unit and tensor product are standard for a
-non-commutative linear logic: when we construct a linear unit we
-cannot use any variables and when we construct a tensor product, the
-two sides must use disjoint variables, and the variables the left side
-of the product uses must be to the left in the context of the
-variables used by the right side of the tensor product. Because we are
-non-commutative, there are two function types: $A \lto B$ and $B \tol
-A$. The difference between these is that the introduction rule for $A
-\lto B$ adds a variable to the right side of the context, whereas the
-introduction rule for $B \tol A$ adds a variable to the left side of
-the context. In our experience, because by convention parsing
-algorithms parse from left-to-right, we rarely need to use the $B \tol
-A$ connective. One of the main uses of linear types is that, in
-combination with the $\uparrow$ connective, we can abstract over
-linear terms, in that the non-linear type $\ltonl{A \lto B}$ is a
-non-linear type of linear terms with one linear input $A$ and output
-$B$.
-\pedro{Isn't this is a slightly incorrect desciption of $\ltonl{A \lto B}$?
-How is the description above distinct form $A \lto B$?}
-
-Next, we have the ``additive'' connectives. For these, we use the
+linear logic. The linear unit ($\I$) and tensor product ($\otimes$)
+are standard for a non-commutative linear logic: when we construct a
+linear unit we cannot use any variables and when we construct a tensor
+product, the two sides must use disjoint variables, and the variables
+the left side of the product uses must be to the left in the context
+of the variables used by the right side of the tensor product. The
+elimination rules for unit and tensor are given by pattern
+matching. The pattern matching rules split the linear context into
+three pieces $\Delta_1,\Delta_2,\Delta_3$: the middle $\Delta_2$ is
+used by the scrutinee of the pattern match, and in the continuation
+this context is replaced by the variables brought into scope by the
+pattern match. This ensures that pattern matches maintain the proper
+ordering of resource usage.
+
+Because we are non-commutative, there are two function types: $A \lto
+B$ and $B \tol A$, which have similar $\lambda$ introduction forms and
+application elimination forms. The difference between these is that
+the introduction rule for $A \lto B$ adds a variable to the right side
+of the context, whereas the introduction rule (elided) for $B \tol A$ adds a
+variable to the left side of the context. In our experience, because
+by convention parsing algorithms parse from left-to-right, we rarely
+need to use the $B \tol A$ connective. As we have already seen, the
+$\lto$ connective is frequently used in conjunction with the
+$\uparrow$ connective so that we can abstract non-linearly over linear
+functions.
+
+Next, we cover the ``additive'' connectives. First, we use the
 non-linear types to define \emph{indexed} versions of the additive
-disjunction $\oplus$ and additive conjunction $\&$ of linear
-logic. These are like a linear version of the $\Sigma$ and $\Pi$
-connectives of ordinary dependent type theory. The indexed $\&$ is
-defined by a $\lambda$ and eliminated using projection. The indexed
-$\oplus$ is like a \emph{weak} $\Sigma$ type: it has an injection
-introduction rule $\sigma$, but its elimination rule is given by
-\emph{pattern matching} rather than projections. Note that we can
-define the more typical nullary and binary versions of these
-connectives by using indexing over the empty and boolean type
-respectively. We will freely use $0$ to refer to this empty
-disjunction and $\top$ to refer to the empty conjunction. Lastly, we
-include a type $\equalizer {a}{f}{g}$ that we call the
-\emph{equalizer} of $f$ and $g$. We think of this type as the
-``subset'' of $A$ that satisfy the equation $f\,a\equiv g\,a$. If
-these were non-linear types, such a type could be constructed using
-the equality type as $\sum_{x:X} f\,x=_Y g\,x$, but this construction
-can't be used because it uses a \emph{dependent} equality type. We
-note that the equalizer type is not used in defining any of our
-parsers or formal grammars, but instead will be used shortly to derive
-an induction principle for proving equalities of functions whose
-domain is an inductive type.
+disjunction $\oplus$ and additive conjunction $\&$ of linear logic,
+which can be thought of as linear versions of the $\Sigma$ and $\Pi$
+connectives of ordinary dependent type theory, respectively. The
+indexed $\&$ is defined by a $\lambda$ that brings a \emph{non-linear}
+variable into scope and eliminated using projection where the index
+specified is given by a non-linear term. The rules for indexed
+$\oplus$ are analogous to a \emph{weak} $\Sigma$ type: it has an
+injection introduction rule $\sigma$, but its elimination rule is
+given by \emph{pattern matching} rather than projections
+\cite{weakSigmatype}. We can define the more typical nullary and
+binary versions of these connectives by using indexing over the empty
+and boolean type respectively. We will freely use $0$ to refer to this
+empty disjunction and $\top$ to refer to the empty conjunction, and
+use infix $\oplus/\&$ for binary disjunction/conjunction.
+%
+In addition to the introduction and elimination rules for
+$\bigamp,\bigoplus$, our development requires an additional axiom that
+is not proveable from these rules: that additive conjunction
+\emph{distributes} over additive disjunction, e.g., in the finitary
+case that $0 \& A \cong 0$ and $(A + B) \& C \cong (A \& C) + (B \&
+C)$. In its most general form, the axiom says that the definable function
+${\bigoplus\limits_{f : \prod_{x:X}
+    Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto
+  \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y}$ has an
+inverse.
+
+Lastly, we include a type $\equalizer {a}{f}{g}$ that we call the
+\emph{equalizer} of linear functions $f$ and $g$. We think of this
+type as the ``subtype'' of elements of $A$ that satisfy the equation
+$f\,a\equiv g\,a$. Note that it is important here that $f, g$
+themselves are non-linearly used functions, as linear values cannot be
+used in a type.  Equalizer types are not needed for non-linear types
+since they can be constructed using the equality type as $\sum_{x:X}
+f\,x=_Y g\,x$, but this construction can't be used for linear types
+because it uses a \emph{dependent} version of the equality type, which
+we cannot define as a linear type. While the equalizer type is not
+used directly in defining any of our parsers or formal grammars, it is
+used for several proofs, allowing for inductive arguments about our
+indexed inductive types.
 
@@
   \begin{mathpar}
-    \inferrule{}{\Gamma \vdash I \isLinTy}\and
+    \inferrule{}{\Gamma \vdash \I \isLinTy}\and
     \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \otimes B \isLinTy}\and
@@
     \and
-    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
+    \inferrule{\Gamma ; \Delta_2 \vdash e : I \\ \Gamma ; \Delta_1,\Delta_3 \vdash e' : C}{\Gamma ; \Delta_1,\Delta_2,\Delta_3 \vdash \letin {()} e {e'} : C}
     %
@@
     %
-    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
-    %
-    \and
-    %
-    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e' : C}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'} : C}
+    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash (e , e') : A \otimes B}
+    %
+    \and
+    %
+    \inferrule{\Gamma ; \Delta_2 \vdash e : A \otimes B \\ \Gamma ; \Delta_1, a : A, b : B, \Delta_2 \vdash e' : C}{\Gamma ;  \Delta_1, \Delta_2, \Delta_3 \vdash \letin {(a , b)} e {e'} : C}
     \\
@@
     %
-    \inferrule{\Gamma ; \Delta \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {\sigma\,x\,a} e {e'}: C}
+    \inferrule{\Gamma ; \Delta_2 \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta_1, a : A, \Delta_3 \vdash e' : C}{\Gamma; \Delta_1, \Delta_2, \Delta_3 \vdash \letin {\sigma\,x\,a} e {e'}: C}
     %
@@
       \Gamma ; \Delta \vdash \applto {f}{e} \equiv \applto {g}{e}}
-    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {e}{f}{g}}
+    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {a}{f}{g}}
     %
@@
 \end{figure}
-\pedro{Fixed tensor elimination rule (added output type)}
-\pedro{Perhaps we should modify the equalizer introduction rule.
-  As it stands, readers might be confused by the same symbol $e$
-  being used everywhere.}
 
@@
 
-Next, we introduce one of the most complex and important linear
-type constructors of our development, \emph{indexed inductive linear
-types} $\mu F$. To encode these, we first add a non-linear type
-$\SPF\,X$ of \emph{strictly-positive functorial} linear type
-expressions indexed by a non-linear type $X$. We give the rules for
-strictly-positive functors in \Cref{fig:spf}. First, we add operations
-for a recursive reference $\Var~x$ the constant functor, $K$, tensor
-products and additive conjunction and disjunction. As an example,
-\max{TODO: make an example}.  Next we add an operation $\el(F)$ which
-interprets our syntactic codes as an actual type constructor, and
-define it for each of the operations given above. Lastly, we give a
-\emph{functorial action} $\map(F)$ for each of the codes.\max{TODO:
-  find a citation for this kind of approach. Container papers work
-  this way for instance. I think there is a linear container paper
-  that uses this approach}
-
-With this notion of strictly-positive functor so defined, in
-\Cref{fig:iilt} we next add a linear type constructor $\mu F$, which
-defines a family of mutually-recursive inductive types for the family
-of type expressions $F$. The introduction rule for this is
-$\roll$, which constructs an element of $\mu F\,x$ from its
-one-level unfolding. The elimination principle is defined by a
-$\fold$ operation. The equations ensure that this makes the
-family an \emph{initial algebra} for the type expressions. That is the
-$\beta$ rule says that a $\mathsf{fold}$ applied to a $\mathsf{roll}$
-is equivalent to $\mathsf{map}$ping the $\mathsf{fold}$ over all the
-sub-expressions, which means that $\mathsf{fold}$ is an algebra
-homomorphism from $\mathsf{roll}$ to $\mathsf{f}$. Then the $\eta$
-rule says that $\mathsf{fold}$ is the \emph{unique} such homomorphism.
+Next, we introduce the most complex and important linear type
+constructors of our development, \emph{indexed inductive linear
+types}. We encode these by adding a mechanism for constructing initial
+algebras of strictly positive functorial type expressions, following
+prior work on inductive types
+\cite{quantitativePolynomialFunctors,indexedContainers}. The syntax is
+given in \Cref{fig:iilt}. First, we add a non-linear type $\SPF\,X$ of
+\emph{strictly positive functorial} linear type expressions indexed by
+a non-linear type $X$. We think of the elements of this type as
+syntactic descriptions of linear types that are parameterized by
+$X$-many variables standing for linear types that are only used in
+strictly positive positions. Accordingly, the $\SPF\,X$ type supports
+an operation $\el$ that interprets it as such a type constructor, as
+well as an operator $\map$ that defines a functorial action on parse
+transformers. The $\SPF\,X$ type supports constructors for a reference
+$\Var~x$ to one of the linear type variables, a constant expression
+that doesn't mention any type variables $K$, as well as tensor
+products and additive conjunction and disjunction of type expressions.
+We additionally add equations in the appendix that say that the
+$\el$/$\map$ operations correspond to these descriptions of the
+constructors.
+
+\begin{figure}
+  \begin{footnotesize}
+    \begin{mathpar}
+      \inferrule{\Gamma \vdash X\isTy}{\Gamma \vdash \SPF\,X \isTy}\and
+      \inferrule{\Gamma \vdash X\isSmall}{\Gamma \vdash \SPF\,X \isSmall}\and
+    \el : \prod_{X:U}\SPF\,X \to (X \to L) \to L\and
+    \map : \prod_{X:U}\prod_{F : \SPF\,X}\prod_{A,B:X\to L}{\left(\prod_{x:X}\uparrow(\unquoteTy{A\,x}\lto \unquoteTy{B\,x})\right)} \to {\uparrow(\unquoteTy{\el(F)(A)} \lto \unquoteTy{\el(F)(B)})}\and
+    \mathsf{Var} : \prod_{X:U} X \to \SPF\,X\and
+    \mathsf{K} : \prod_{X:U} L \to \SPF\,X\and
+    \mathsf{\bigoplus} : \prod_{X:U}\prod_{Y:U}(Y \to \SPF\,X) \to \SPF\,X\and
+    \mathsf{\bigamp} : \prod_{X:U}\prod_{Y:U}(Y \to \SPF\,X) \to \SPF\,X\and
+    \mathsf{\otimes} : \prod_{X:U}\SPF\,X \to \SPF\,X \to \SPF\,X\and
+    \roll : \prod_{X:U}\prod_{F:X \to \SPF\,X}\prod_{x:X}\ltonl{(\el(F\,x)(\mu\,F))}\and
+    \fold : \prod_{X:U}\prod_{F:X\to\SPF\,X}\prod_{A:X \to L}
+    \left(\prod_{x:X}\ltonl{(\unquoteTy{\el(F\,x)(\unquoteTy{A})} \lto \unquoteTy{A\,x})}\right)
+    \to \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}\and
+
+    \inferrule*[right=Ind$\beta$]
+    {\Gamma \vdash f : \prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)} \and \Gamma; \Delta \vdash e : \el(F\,x)(\mu\,F)}
+    {\Gamma;\Delta \vdash \fold\,F\,f\,x\,(\roll e) \equiv f\,x\,(\map(F\,x)\,(\fold\,F\,f)) : A\,x}
+
+    \inferrule*[right=Ind$\eta$]
+    {\Gamma \vdash f : \prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)}
+     \and \Gamma \vdash e : \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}\\\\
+      \Gamma,x:X;a:\el(F\,x)(\mu F) \vdash e\,x\,(\roll a) \equiv f\,x\,(\map(F\,x)\,e) : A\,x }
+    {\fold\,F\,f\equiv e' : \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}}
+  \end{mathpar}
+  \end{footnotesize}
+  %% \begin{footnotesize}
+  %%   \begin{align*}
+  %%   \mathsf{Var} &: \prod_{X:U} X \to \SPF\,X\\
+  %%   \mathsf{K} &: \prod_{X:U} L \to \SPF\,X\\
+  %%   \mathsf{\bigoplus} &: \prod_{X:U}\prod_{Y:U}(Y \to \SPF\,X) \to \SPF\,X\\
+  %%   \mathsf{\bigamp} &: \prod_{X:U}\prod_{Y:U}(Y \to \SPF\,X) \to \SPF\,X\\
+  %%   \mathsf{\otimes} &: \prod_{X:U}\SPF\,X \to \SPF\,X \to \SPF\,X\\
+  %%   \el &: \prod_{X:U}\SPF\,X \to (X \to A) \to L\\
+  %%   \map &: \prod_{X:U}\prod_{F : \SPF\,X}\prod_{A,B:X\to L}{\left(\prod_{x:X}\uparrow(\unquoteTy{A\,x}\lto \unquoteTy{B\,x})\right)} \to {\uparrow(\unquoteTy{\el(F)(A)} \lto \unquoteTy{\el(F)(B)})}\\\\
+  %%   \roll &: \prod_{X:U}\prod_{F:X \to \SPF\,X}\prod_{x:X}\ltonl{(\el(F\,x)(\mu\,F))}\\
+  %%   \fold &: \prod_{X:U}\prod_{F:X\to\SPF\,X}\prod_{A:X \to L}
+  %%   \left(\prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)}\right)
+  %%   \to \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}
+  %% \end{align*}
+  %% \end{footnotesize}
+  \caption{Strictly Positive Functors and Indexed Inductive Linear Types}
+  \label{fig:iilt}
+\end{figure}
+
+Next, given a family of $X$-many strictly positive linear type
+expressions $F : X \to \SPF\,X$, we define a family $\mu F : X \to L$
+of $X$-many mutually recursive inductive types. The introduction rule
+for this is $\roll$, which constructs an element of $\mu F\,x$ from
+the one-level of the $x$th type expression. The elimination principle
+is defined by a mutual $\fold$ operation: given a family of output
+types $A$ indexed by $X$, we can define a family of functions from
+$\mu F\,x \multimap A\,x$ if you specify how to interpret all of the
+constructors as operations on $A$ values. We add $\beta\eta$ equations
+that specify that this makes the family $\mu F$ into an \emph{initial
+algebra} for the functor $\el(F)$. That is the $\beta$ rule says that
+a $\mathsf{fold}$ applied to a $\mathsf{roll}$ is equivalent to
+$\mathsf{map}$ping the $\mathsf{fold}$ over all the sub-expressions,
+which means that $\mathsf{fold}$ interprets all of the constructors
+homomorphically using the provided interpretation $\mathsf{f}$. Then
+the $\eta$ rule says that $\mathsf{fold}$ is the \emph{unique} such
+homomorphism, i.e. anything that satisfies the recurrence equation of
+the $\mathsf{fold}$ is equal to it.
 
@@
 directly in proofs. In dependent type theory, we would have a
-dependent \emph{elimination} principle, which justifies
-\emph{inductive proofs}. However since linear types do not support
-dependency on linear types, we cannot directly adapt this
-approach. However, if we are trying to prove that two morphisms out of
-a mutually recursive type are equal, we can use the \emph{equalizer}
-type to prove their equality by induction.\max{TODO: show how this is
-  done.}
-
-\begin{figure}
-  \begin{align*}
-    \mathsf{Var} &: X \to \SPF\,X\\
-    \mathsf{K} &: L \to \SPF\,X\\
-    \mathsf{\bigoplus} &: (Y \to \SPF\,X) \to \SPF\,X\\
-    \mathsf{\bigamp} &: (Y \to \SPF\,X) \to \SPF\,X\\
-    \mathsf{\otimes} &: \SPF\,X \to \SPF\,X \to \SPF\,X\\
-    \el &: \prod_{X:U}\SPF\,X \to (X \to L) \to L\\
-    \map &: \prod_{X:U}\prod_{F : \SPF\,X}\prod_{A,B:X\to L}(\prod_{x:X}\uparrow(A\,x\lto B\,x)) \to \uparrow(\el(F)(A) \lto \el(F)(B))\\\\
-    \el(\Var\,M) B &= B\,M\\
-    \el(\mathsf{K}\,A) B &= A\\
-    \el(\bigoplus A) B &= \bigoplus_{y:Y}\el(A y)B\\
-    \el(\bigamp A) B &= \bigamp_{y:Y}\el(A y)B\\
-    \el(A \otimes A') B &= \el(A)B \otimes \el(A')B\\\\
-    \map(\Var\,M)\,f &= f\,M\\
-    \map(\K\,A)\,f &= \lambda a. a\\
-    \map(\bigoplus A)\,f &= \lambda a. \letin{\oplusinj y {a_y}}{a}{\oplusinj y {\map(A\,y)\,f\,a_y}}\\
-    \map(\bigwith\,A)\,f &= \lambda a. \withlamb{y}{\map(A\,y)\,f\,(\withprj y a)}\\
-    \map(A \otimes A')\,f &= \lambda b. \letin{(a,a')}{b}{(\map(A)\,f\,a,\map(A')\,f\,a')}
-  \end{align*}
-  \begin{mathpar}
-  \end{mathpar}
-  \caption{Strictly Positive Functors}
-  \label{fig:spf}
-\end{figure}
-
-\begin{figure}
-  \begin{mathpar}
-    \mathsf{roll} : \prod_{X:U}\prod_{F:X \to \SPF\,X}\prod_{x:X}\ltonl{(\el(F\,x)(\mu\,F))}
-
-    %% \inferrule*[right=IndIntro]
-    %% {\Gamma \vdash A : X \to \textrm{SPF}\,X\and
-    %%  \Gamma \vdash M : X\and
-    %%  \Gamma; \Delta \vdash e : \textrm{el}(A\,M) (\mu\,A)}
-    %% {\Gamma; \Delta \vdash \mathsf{roll}\, e : \mu\,A\, M}
-
-    \mathsf{fold} : \prod_{X:U}\prod_{F:X\to\SPF\,X}\prod_{A:X \to L}
-    (\prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)})
-    \to \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}
-
-    \inferrule*[right=Ind$\beta$]
-    {}
-    {\fold\,F\,f\,x\,(\roll e) \equiv f\,x\,(\map(F\,x)\,(\fold\,F\,f))}
-
-    \inferrule*[right=Ind$\eta$]
-    {x:X;a:\el(F\,x)(\mu F) \vdash e'\,x\,(\roll a) \equiv f\,x\,(\map(F\,x)\,e')
-    }
-    {\fold\,F\,f\equiv e'}
-    %% {\Gamma,x:X; a:\mu\,A\,x \vdash e' : B\,x
-    %%   \and \Gamma;\Delta \vdash e : \mu A M\\\\
-    %%   \Gamma,x:X; a':\el(A x)(\mu A) \vdash
-    %%   e'[x/x][\roll(a')/a] = f[x/x][\map(A\,x)\,(y.a.e'[y/x][a/a])\,a/b] : B\,x
-    %% }
-    %% {\Gamma; \Delta \vdash \fold(A)(x.b.f)(M,e) = e'[M/x][e/a] : B\,M}
-  \end{mathpar}
-  \caption{Indexed Inductive Linear Types}
-  \label{fig:iilt}
-\end{figure}
+dependent \emph{elimination} principle, which can be used to
+implemetns functions by recursion as well as proofs by
+induction. However since linear types do not support dependency on
+linear types, we cannot directly adapt this approach. However, if we
+are trying to prove that two morphisms out of a mutually recursive
+type are equal, we can use the \emph{equalizer} type to prove their
+equality by induction. That is, if our goal is to prove two functions
+$f,g : \uparrow(\mu F\,x \lto A\,x)$ are equal, it suffices to
+implement a function $\mathsf{ind} : \uparrow(\mu F\,x \lto \equalizer
+{a} f g)$ such that $\textrm{ind}(a) \equiv a$. Then an
+inductive-style proof can be implemented by constructing
+$\mathsf{ind}$ using a $\fold$. This can all be justified using only
+the $\beta\eta$ principles for equalizers and inductive types, and
+this is how our most complex inductive proofs are implemented in Agda
+formalization.
 
@@
 
-So far, our calculus is a generic combination of dependent types with
-non-commutative linear types. In order to carry out formal grammar
-theory and define parsers, however, we add a few carefully chosen base
-types, terms and axioms that allow us to reason about formal grammars.
-\max{TODO: more prose, also are these all of our axioms?}
-\begin{enumerate}
-\item (Characters) For each character of our alphabet $c \in \Sigma$, we add a linear type $c \isLinTy$.
-\item (Strings are total) The function $\ltonl{(\String \lto \top)}$ is an isomorphism.\max{for this need to define $\String$, and isomorphism, and possibly we should define $\top$ in more detail earlier?
-idk}
-\item (Distributivity of $\oplus,\&$) The function  $\ltonl{(\bigoplus\limits_{f : \prod_{x:X} Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y)}$ is an isomorphism.
-\item ($\ltonl$ preserves $\oplus$) The function $\sum_{x:X}\ltonl{(A\,x)} \to \ltonl{(\bigoplus\limits_{x:X}A\,x)}$ is an isomorphism.
-\end{enumerate}
+So far, our calculus is a somewhat generic combination of dependent
+types with non-commutative linear types. In order to carry out formal
+grammar theory and define parsers, we need only add a few
+grammar-specific constructions. First for each character in our
+alphabet, we add a corresponding linear type $c$. We can then define a
+non-linear type $\Char$ as the disjunction of all of these characters,
+and define a type $\String$ as the Kleene star of $\Char$, i.e. as an
+inductive linear types. Then we add a function $\mathsf{read} :
+\ltonl{(\top\lto \String)}$ that says that we can extract the current
+string from the empty additive conjunction. Further, we add an axiom
+that this function is an isomorphism, i.e. that it is inverse to the
+unique function $\ltonl{(\String \lto \top)}$. In the next section, we
+will show how these basic principles are enough to provide a basis for
+verified parsing and formal grammar theory.
 
@@
 \section{Syntax}
+
+\begin{figure}
+  \begin{align*}
+    \el(\Var\,M) B &= B\,M\\
+    \el(\mathsf{K}\,A) B &= A\\
+    \el(\bigoplus A) B &= \bigoplus_{y:Y}\el(A y)B\\
+    \el(\bigamp A) B &= \bigamp_{y:Y}\el(A y)B\\
+    \el(A \otimes A') B &= \el(A)B \otimes \el(A')B\\\\
+    \map(\Var\,M)\,f &= f\,M\\
+    \map(\K\,A)\,f &= \lambda a. a\\
+    \map(\bigoplus A)\,f &= \lambda a. \letin{\oplusinj y {a_y}}{a}{\oplusinj y {\map(A\,y)\,f\,a_y}}\\
+    \map(\bigwith\,A)\,f &= \lambda a. \withlamb{y}{\map(A\,y)\,f\,(\withprj y a)}\\
+    \map(A \otimes A')\,f &= \lambda b. \letin{(a,a')}{b}{(\map(A)\,f\,a,\map(A')\,f\,a')}
+  \end{align*}
+  \caption{Strictly Positive Functors functorial actions}
+\end{figure}
 

### 2024-11-13 paper/paper.tex
missing analysis in the examples section

@@
 
+ %% \steven{TODO make the Agda logo clickable} %
+\newcommand{\Agda}{\agdalogo}
 \renewcommand{\Gamma}{\mathgtt{G}}
@@
 inductive linear types. Then we add a function $\mathsf{read} :
-\ltonl{(\top\lto \String)}$ that says that we can extract the current
-string from the empty additive conjunction. Further, we add an axiom
-that this function is an isomorphism, i.e. that it is inverse to the
-unique function $\ltonl{(\String \lto \top)}$. In the next section, we
-will show how these basic principles are enough to provide a basis for
-verified parsing and formal grammar theory.
+\ltonl{(\top\lto \String)}$ that intuitively ``reads'' the input
+string from the input and makes it available. It is important that the
+input type of $\mathsf{read}$ is $\top$, which can control any amount
+of resources, and not $\I$ which controls no resources. Further, we
+add an axiom that $\lambda s. \mathsf{read}(\top) \equiv \lambda
+s. s$, i.e., that if we have a string, but then throw it away and read
+it from the input, then in fact that is equivalent to the string we
+were given originally. This ensures that the elements of the $\String$
+type always stand for the actual input string in our reasoning. In the
+next section, we will show how these basic principles are enough to
+provide a basis for verified parsing and formal grammar theory.
 
@@
 
-\paragraph{Equivalence}
-In \cite{chom1963}, Chomsky differentiates between the weak generative capacity
-of a formal grammar --- the set of strings accepted by it --- and its strong
-generative capacity --- the set of ``structural descriptions'', or parse trees,
-that it generates. Two grammars are said to be \emph{weakly equivalent} if they
-agree in their weak generative capacity, and likewise \emph{strongly equivalent}
-if they agree in their strong generative capacity. We aim to recover each of
+\paragraph{Weak and Strong Equivalence}
+In the theory of formal grammars, there are two different notions of
+equivalence: weak generative capacity, meaning just which strings are
+accepted by the grammar, and \emph{strong} generative capacity, when
+the parse trees of the two grammars are isomorphic
+\cite{chom1963}. Using linear types as grammars, we can define both of
 these notions of equivalence in \theoryabbv.
@@
   \label{def:weakequiv}
-  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exists parse
-  transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B \lto \A)}$.
+  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exists
+  parse transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B
+    \lto \A)}$. $\A$ is a \emph{retract} of $\B$ if $\A$ and $\B$ are
+  weakly equivalent and $\lambda a. g(f(a)) \equiv \lambda a.a$. They
+  are \emph{strongly equivalent} if further the other composition is
+  the identity, i.e., $\lambda b. f(g(b)) \equiv \lambda b.b$.
 \end{definition}
-\pedro{Up to this point, we have mentioned a couple of times parse transformers,
-  but we have never defined what they are. In particular, why does the definition
-  require the $\uparrow$ modality. I'm assuming that it's to properly deal with
-  linear variables, but these things are not obvious at first.}
-
-\newcommand{\lang}{\texttt{Lang}}
-A parse transformer $\f : \ltonl{(\A \lto \B)}$ takes as input parses of $\A$ for string
-$\w$ and returns a parse of $\B$ for $\w$. Therefore, $\f$ demonstrates an
-inclusion of languages $\lang(\A) \subset \lang(\B)$. Similarly, the existence of
-$\g$ shows that $\lang(\B) \subset \lang(\A)$, and thus $\lang(\A) = \lang(\B)$.
-
-\steven{Need better notation for language that is cohesive throughout the paper}
-
-\begin{definition}
-  \label{def:strongequiv}
-  Grammars $\A$ and $\B$ are \emph{strongly equivalent} if there exists parse
-  transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B \lto \A)}$ such
-  that $\f \circ \g \equiv \ident$ and $\g \circ \f \equiv \ident$.
-\end{definition}
-
-For a string $\w$, \cref{def:strongequiv} induces a bijection between the
-$\A$-parses of $\w$ and the $\B$-parses of $\w$, thus demonstrating that $\A$
-and $\B$ agree in their strong generative capacity.
+
+That is, grammars are weakly equivalent if we can construct a parse
+tree of $B$ from a parse tree of $A$ and vice-versa, whereas they are
+strongly equivalent if these processes are inverse to each other. The
+definition of retract is a useful intermediate notion. In retrospect,
+our axiom about $\String$ says that $\String$ is a retract of $\top$.
+
+%% \newcommand{\lang}{\texttt{Lang}}
+%% A parse transformer $\f : \ltonl{(\A \lto \B)}$ takes as input parses of $\A$ for string
+%% $\w$ and returns a parse of $\B$ for $\w$. Therefore, $\f$ demonstrates an
+%% inclusion of languages $\lang(\A) \subset \lang(\B)$. Similarly, the existence of
+%% $\g$ shows that $\lang(\B) \subset \lang(\A)$, and thus $\lang(\A) = \lang(\B)$.
+
+%% \steven{Need better notation for language that is cohesive throughout the paper}
+
+%% \begin{definition}
+%%   \label{def:strongequiv}
+%%   Grammars $\A$ and $\B$ are \emph{strongly equivalent} if there exists parse
+%%   transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B \lto \A)}$ such
+%%   that $\f \circ \g \equiv \ident$ and $\g \circ \f \equiv \ident$.
+%% \end{definition}
+
+%% For a string $\w$, \cref{def:strongequiv} induces a bijection between the
+%% $\A$-parses of $\w$ and the $\B$-parses of $\w$, thus demonstrating that $\A$
+%% and $\B$ agree in their strong generative capacity.
 
 \paragraph{Ambiguity}
-A formal grammar $\A$ is ambiguous if there is a string $\w$ for which $\A$ has
-several parse trees of $\w$.
-For example, $\a \oplus \a$ is ambiguous because
-$\stringquote{a}$ has two parse trees --- on either side of the disjunction. On
-the other hand, a formal grammar is unambiguous when there is at most
-one parse tree for any input string. These notions of ambiguity are entirely
-semantic, so we seek to find a property expressible in the syntax of \theoryabbv
-that reifies the semantic description of ambiguity.
-
+A formal grammar $\A$ is ambiguous if there are multiple parse trees
+for the \emph{same} string $\w$.  For example, $\a \oplus \a$ is
+ambiguous because there are two parses of $\stringquote{a}$,
+constructed using $\inl$ and $\inr$. On the other hand, a formal
+grammar is unambiguous when there is at most one parse tree for any
+input string. We can capture this notion as a type in \theoryabbv in a
+clever way:
 \begin{definition}
@@
 \end{definition}
-\cref{def:unambig} can be read more intuitively as stating that $\A$ is
-unambiguous if there is at most one way to transform parses of any other grammar
-$\B$ into parse of $\A$.
-
-Note that this definition of unambiguity is preserved by strong equivalence, so a common proof technique for showing $\A$ is unambiguous is by proving $\A$
-is strongly equivalent to an unambiguous grammar $\B$. In this style of proof, we
-build the functions $\f : \ltonl{(\A \lto \B)}$ and
-$\g : \ltonl{(\B \lto \A)}$, then we have the obligation to show
-$\f \circ \g \equiv \ident$ and $\g \circ \f \equiv \ident$. However, because $\B$ is
-unambiguous there is only one function of type $\ltonl{(\B \lto \B)}$, so
-$\f \circ \g : \ltonl{(\B \lto \B)}$ and $\ident : \ltonl{(\B \lto \B)}$ are equal.
-Thus, in this type of proof, we really only have the obligation of building each of $\f$
-and $\g$ and demonstrating that $\g \circ \f \equiv \ident$.
+\cref{def:unambig} can be read more intuitively as stating that $\A$
+is unambiguous if there is at most one way to transform parses of any
+other grammar $\B$ into parse of $\A$. This notion of an unambiguous
+type is the analog for linear types of the definition of a (homotopy)
+\emph{proposition} in the terminology of homotopy type
+theory\cite{hott}. The most basic unambiguous types are $\top$ and
+$0$, and in a system of classical logic all unambiguous types would
+have to be equivalent to one of these, but with our axioms we can show
+also that $\I$ and literals $\literal c$ are unambiguous. To see this, first, we
+establish two useful properties of unambiguity.
+\begin{lemma}[\agdalogo]
+  If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
+  If a disjunction $\bigoplus\limits_{x:X} A(x)$ is unambiguous then
+  each $A(x)$ is unambiguous.
+\end{lemma}
+From the first principle, we can prove that $\String$ is unambiguous,
+since it is a retract of $\top$. In fact, observe that if $\A$ is a
+retract of $\B$ and $\B$ is unambiguous, then in fact $\A$ and $\B$
+are strongly equivalent, as the equation $\lambda b. f(g(b)) \equiv
+\lambda b. b$ follows because $\B$ is unambiguous. Therefore also
+$\String$ is strongly equivalent to $\top$. Next, since $\String$ is
+defined as a Kleene star, we can easily show that $\String \cong \I
+\oplus \Char \oplus (\Char \otimes \Char \otimes \String)$. Then by
+the second principle we have that $\I$ and $\Char$ and each literal
+$\literal c$ are unambiguous as well.
 
@@
 % there's a map $\top \to A \oplus \neg A$ --- then call $A$ \emph{decidable}.
+
+\paragraph{Partial and Total Parsers}
+
+We now turn to our main task, which is using our linear type system to
+implement verified parsers. Given a grammar defined as a linear type
+$A$, a first attempt at defining a parser would be to implement a
+function $\uparrow{(\String \lto A)}$. But since our linear functions
+must be total, this means that we can construct an $A$ parse for
+\emph{every} input string, which is impossible for most grammars of
+interest. Instead we might try to write a partial function as a
+$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option'' monad
+\cite{maybeMonad}. This allows for the possibility that the input
+string doesn't parse, but is far too weak as a specification: we can
+trivially implement a parser for any type by always returning
+$\inr$. The correct notion of a parser should be one that allows for
+failure, but only in the case that a parse cannot be constructed.
+\begin{definition}
+  A parser for a linear type $A$ is a function $\uparrow{(\String \lto
+    A \oplus A_{\neg})}$ where $A_{\neg}$ is a linear type that is
+  \emph{disjoint} from $A$ in that we can implement a function
+  $\uparrow{(A \& A_{\neg} \lto 0)}$
+\end{definition}
+Here we replace $\top$ in our partial parser type with a type
+$A_{\neg}$ that we can think of as a negation of $A$. The function
+$\uparrow{(A \& A_{\neg} \lto 0)}$ ensures that it is impossible for
+$A$ and $A_{\neg}$ to match the same input string. This means that in
+defining a parser, we will need to define a kind of negative grammar
+for strings that do not parse. Fortunately, we will see that
+deterministic automata naturally support such a notion with no
+additional effort: the negative grammar is simply the grammar for
+traces that end in a rejecting state.
+
+The main benefit of writing such a parser as a linear term, is that it
+looks just like an ordinary functional program into a sum type. We
+typically construct it by doing a fold over $\String$, which as we
+recall is defined as a type of linear lists of characters. But we get
+the \emph{soundness} of the parser completely for free from the
+typing: any $\inl$ parse that we return must correspond to a parse
+tree of the input string. Further if we verify the disjointness
+property we then also get the \emph{completeness} of the parser as
+well, that it never fails to generate an $A$ parse when it is
+possible.
+
+Our main proof tool then for constructing verified parsers is to show
+that a grammar $A$ is weakly equivalent to a grammar for a
+deterministic automaton. The parsers for deterministic automata can
+simply be implemented by stepping through the states of the
+automaton. The key is then the following simple observation:
+\begin{lemma}[\agdalogo]
+  If $\A$ is weakly equivalent to $\B$ then any parser for $\A$ can be
+  extended to a parser for $\B$.
+\end{lemma}
+Here we need both directions of the weak equivalence. We need $A \lto
+B$ to extend the parser from $\String \lto A \oplus A_{\neg}$ to
+$\String \lto B \oplus A_{\neg}$ but then we also need $B \lto A$ to
+establish that $A_{\neg}$ is disjoint from $B$.
 
@@
 
-\steven{TODO make the Agda logo clickable, and mention its meaning somewhere}
-\newcommand{\Agda}{\agdalogo}
 \begin{theorem}[Determinization, \Agda]
@@
   $P$ and functions $I \leftarrow P \to S \to O$. The \emph{extension}
-  of a polynomial is a functor $\Set/I \to \Set/O$ defined by
-  \max{TODO: fill in details}.
+  of a polynomial is a functor $\Set/I \to \Set/O$ defined as the
+  composite
+  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXFNldC9JIl0sWzMsMCwiXFxTZXQvTyJdLFsxLDAsIlxcU2V0L1AiXSxbMiwwLCJcXFNldC9TIl0sWzAsMiwiZl4qIl0sWzIsMywiXFxQaV9nIl0sWzMsMSwiXFxTaWdtYV9oIl1d
+  \[\begin{tikzcd}
+	    {\Set/I} & {\Set/P} & {\Set/S} & {\Set/O}
+	    \arrow["{f^*}", from=1-1, to=1-2]
+	    \arrow["{\Pi_g}", from=1-2, to=1-3]
+	    \arrow["{\Sigma_h}", from=1-3, to=1-4]
+  \end{tikzcd}\]
+  Where $f^*$ is the pullback functor along $f$; $\Pi_g$ is the depedent
+  product operation and $\Sigma_h$ is the dependent sum operation, which
+  are, respectively, the right and left adjoint of their pullback functors
+  $g^*$ and $h^*$.
 
@@
 where the inequality is meant to capture language containment.  This
-theory is extremely successful, having found applications in algebraic
-path problems, theory of programming languages, compiler optimizations
-and more.
+theory is extremely successful, having found applications in concurrency theory
+\cite{hoare2000concurrent}, software-defined networks \cite{anderson2014netkat},
+theory of programming languages \cite{angus2001kleene}, compiler optimizations
+\cite{kozen2000certification} and more.
 
@@
 extensions of Kleene algebras, Kleene algebra with tests (KAT) being
-one of the most notable ones. Our approach is radically different from
+one of the most notable ones \cite{kozen1997kleene}. Our approach is radically different from
 most extensions, which usually aim at modifying or adding new

### 2024-11-14 paper/paper.tex
big edits to examples section

--- 2024-11-13 paper/paper.tex
@@
 traces that end in a rejecting state.
-
-The main benefit of writing such a parser as a linear term, is that it
-looks just like an ordinary functional program into a sum type. We
-typically construct it by doing a fold over $\String$, which as we
-recall is defined as a type of linear lists of characters. But we get
-the \emph{soundness} of the parser completely for free from the
-typing: any $\inl$ parse that we return must correspond to a parse
-tree of the input string. Further if we verify the disjointness
-property we then also get the \emph{completeness} of the parser as
-well, that it never fails to generate an $A$ parse when it is
-possible.
-
-Our main proof tool then for constructing verified parsers is to show
-that a grammar $A$ is weakly equivalent to a grammar for a
-deterministic automaton. The parsers for deterministic automata can
-simply be implemented by stepping through the states of the
-automaton. The key is then the following simple observation:
 \begin{lemma}[\agdalogo]
+  If $A \oplus A_{\neg}$ is unambiguous, then $A$ and $A_{\neg}$ are
+  disjoint.
+\end{lemma}
+
+Writing a parser as a linear term in this way is an intrinsic
+verification of the \emph{soundness} of the parser completely for free
+from the typing: any $\inl$ parse that we return \emph{must}
+correspond to a parse tree of the input string. Further if we verify
+the disjointness property we then also get the \emph{completeness} of
+the parser as well, that it never fails to generate an $A$ parse when
+it is possible.
+
+Our main method for constructing verified parsers is to show that a
+grammar $A$ is weakly equivalent to a grammar for a deterministic
+automaton. Parsers for deterministic automata are simple to implement
+by stepping through the states of the automaton, with the rejecting
+traces serving as the negative grammar. This is sufficient due to the
+following:
+\begin{lemma}[\agdalogo]
+  \label{lem:wk-eqv-parse}
   If $\A$ is weakly equivalent to $\B$ then any parser for $\A$ can be
@@
 %% 2. Regex, NFA, DFA
-In this section, we make the equivalences
-between regular expressions and finite automata precise by encoding two theorems
-from classical language theory: (1) every regular expression is strongly equivalent
-to the traces through some non-deterministic finite automaton (NFA); and, (2) for
-every NFA, there is some deterministic finite automaton (DFA) such that the type
-of traces through the NFA is weakly equivalent to the type of traces through the DFA.
+
+In this section, we describe how to construct an intrinsically
+verified parser for regular expressions by compiling it to an NFA and
+then a DFA. That is, for each regular expression $A$, we construct an
+NFA $N(A)$ and a corresponding DFA $D(A)$ such that $A$ is strongly
+equivalent to the traces of $N(A)$ and weakly equivalent to the
+accepting traces of $D(A)$. Then we can easily construct a parser for
+traces of $D(A)$ and combine this using \Cref{lem:wk-eqv-parse} to get
+a verified regex parser.
 
@@
 
-\paragraph{Regular Expressions} The theory of regular expressions emerges as the
-subtheory of \theoryabbv generated by
-$\otimes$, $\oplus$, and Kleene star over the types $0$, $I$, and characters drawn from $\Sigma$.
-
-\paragraph{Nondeterministic Finite Automata} In \cref{sec:type-theory-examples}, we saw
-one particular NFA corresponding to the regular expression
-$({\literal a}^{*} \otimes \literal b) \oplus \literal c$. The definition of the linear type of traces
-through this
-NFA, as
-presented in \cref{fig:nfainductive}, inspects the concrete structure of the
-NFA.\ When generalizing from this example to an arbitrary NFA, we do not have the
-luxury of inspecting a concrete definition. Therefore, we abstract over the
-specification of an arbitrary NFA and define traces with respect to that
-specification.
-
-Let the data of an NFA $\N$ be given by the following nonlinear data,
-\begin{itemize}
-\item $\N.\states$ --- a finite set of states
-\item $\N.\init : \N.\states$ --- the initial state
-\item $\N.\isAcc : \N.\states \to \Bool$ --- a mapping marking which states are
-        accepting
-\item $\N.\transitions$ --- a finite set of labeled transitions
-\item $\N.\labelt : \N.\transitions \to \Sigma$ --- a mapping from $N.\transitions$ to the character labeling the transition
-\item $\N.\src , \N.\dst : \N.\transitions \to \N.\states$ --- mappings from
-        $\N.\transitions$ to the source and target states, respectively, of a labeled transition
-\item $\N.\epstransitions$ --- a finite set of $\varepsilon$-transitions
-\item $\N.\epssrc, \N.\epsdst : \N.\epstransitions \to \N.\states$ --- mappings
-        from $\N.\epstransitions$ to the source and target states of an $\epsilon$-transition
-\end{itemize}
-
-In \cref{fig:nfatrace} we define a linear type of traces
-through an NFA $\N$. $\Trace_{\N}$ is an inductive type indexed by the first
-state $\s : \N.\states$ of a trace and a Boolean $b : \Bool$ that marks if the
-trace ends in an accepting state. Traces in $\N$ may be built through one of three constructors. We
-may terminate a trace at an accepting state with the constructor $\nil$. If we
-had a trace beginning at the destination state of a transition, then we may use
-the $\cons$ constructor to linearly combine that trace with a parse of the label
-of the transition to build a trace beginning at the source of the transition.
-Finally, if we had a trace beginning at the destination of an
-$\epsilon$-transition then we may use $\epscons$ to pull it back along the
-$\epsilon$-transition and construct a trace beginning at the source of the
-$\epsilon$-transition. As a shorthand, write $Parse_{N}$ for the accepting
-traces out of $N.init$.
-
+A regular expressions in \theoryabbv is a linear type constructed
+using only the connectives $\literal c$, $0$, $\oplus$, $I$,
+$\otimes$, and Kleene star.  In \cref{sec:type-theory-examples}, we
+saw one particular NFA corresponding and its type of traces. More
+generally we define the linear type of traces as in
+\Cref{fig:nfatrace}
+
+%% The definition of the linear type of traces
+%% through this
+%% NFA, as
+%% presented in \cref{fig:nfainductive}, inspects the concrete structure of the
+%% NFA.\ When generalizing from this example to an arbitrary NFA, we do not have the
+%% luxury of inspecting a concrete definition. Therefore, we abstract over the
+%% specification of an arbitrary NFA and define traces with respect to that
+%% specification.
+
+%% Let the data of an NFA $\N$ be given by the following nonlinear data,
+%% \begin{itemize}
+%% \item $\N.\states$ --- a finite set of states
+%% \item $\N.\init : \N.\states$ --- the initial state
+%% \item $\N.\isAcc : \N.\states \to \Bool$ --- a mapping marking which states are
+%%         accepting
+%% \item $\N.\transitions$ --- a finite set of labeled transitions
+%% \item $\N.\labelt : \N.\transitions \to \Sigma$ --- a mapping from $N.\transitions$ to the character labeling the transition
+%% \item $\N.\src , \N.\dst : \N.\transitions \to \N.\states$ --- mappings from
+%%         $\N.\transitions$ to the source and target states, respectively, of a labeled transition
+%% \item $\N.\epstransitions$ --- a finite set of $\varepsilon$-transitions
+%% \item $\N.\epssrc, \N.\epsdst : \N.\epstransitions \to \N.\states$ --- mappings
+%%         from $\N.\epstransitions$ to the source and target states of an $\epsilon$-transition
+%% \end{itemize}
+
+In \cref{fig:nfatrace} we define a linear type of traces through an
+NFA $\N$. $\Trace_{\N}$ is an inductive type indexed by the starting
+state of the trace $\s : \N.\states$. Traces in $\N$ may be built
+through one of three constructors. We may terminate a trace at an
+accepting state with the constructor $\nil$. Here we use an Agda-style
+unicode syntax for $\bigamp$, as well as using the function arrow to
+mean a non-dependent version of $\bigamp$. If we had a trace beginning
+at the destination state of a transition, then we may use the $\cons$
+constructor to linearly combine that trace with a parse of the label
+of the transition to build a trace beginning at the source of the
+transition.  Finally, if we had a trace beginning at the destination
+of an $\epsilon$-transition then we may use $\epscons$ to pull it back
+along the $\epsilon$-transition and construct a trace beginning at the
+source of the $\epsilon$-transition. As a shorthand, write $Parse_{N}$
+for the accepting traces out of $N.init$.
+\newcommand{\D}{\texttt{D}}
 \begin{figure}
 \begin{lstlisting}
-data Trace(*@$_{\N}$@*) : (s : N.states) (b : Bool) (*@$\to$@*) L where
-  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) N.states ] Trace(*@$_{\N}$@*) s (N.isAcc s))
-  cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.transitions ] &[ b (*@$\in$@*) Bool ]
-           ('N.label t' (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.dst t) b (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.src t) b))
-  (*@$\epsilon$@*)cons : (*@$\uparrow$@*)(&[ t (*@$\in$@*) N.(*@$\epsilon$@*)transitions ] &[ b (*@$\in$@*) Bool ]
-           (Trace(*@$_{\N}$@*) (N.(*@$\epsilon$@*)dst t) b (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)src t) b))
+data Trace(*@$_{\N}$@*) : (s : N.states) (*@$\to$@*) L where
+  nil : (*@$\uparrow$@*)(&[ s : N.states ] N.isAcc (*@$\to$@*) Trace(*@$_{\N}$@*) s)
+  cons : (*@$\uparrow$@*)(&[ t : N.transitions ]
+    ('N.label t' (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.dst t) (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.src t)))
+  (*@$\epsilon$@*)cons : (*@$\uparrow$@*)(&[ t : N.(*@$\epsilon$@*)transitions ]
+    (Trace(*@$_{\N}$@*) (N.(*@$\epsilon$@*)dst t) (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)src t)))
+
+data Trace(*@$_{\D}$@*) : (s : D.states) (b : Bool) (*@$\to$@*) L where
+  nil : (*@$\uparrow$@*)(&[ s : D.states ] Trace(*@$_{\D}$@*) s D.isAcc s)
+  cons : (*@$\uparrow$@*)(&[ c : Char ] &[ s : D.states ] &[ b : Bool ]
+  ('c' (*@$\lto$@*) Trace(*@$_{\D}$@*) (D.(*@$\delta$@*) c s) b (*@$\lto$@*) Trace(*@$_{\D}$@*) s b))
 \end{lstlisting}
-\caption{Traces through an NFA $N$ as an indexed inductive type}
+\caption{Traces of an NFA $N$ and a DFA $D$}
 \label{fig:nfatrace}
@@
 
-\paragraph{Deterministic Finite Automata}
-\newcommand{\D}{\texttt{D}}
-Just as the above defines an NFA, we provide a specification for an arbitrary
-deterministic finite automaton.\ A DFA $D$ is
-given by the following data,
-
-\begin{itemize}
-\item $\D.\states$ --- a finite set of states
-\item $\D.\init : \D.\states$ --- the initial state
-  \item $\D.\isAcc : \D.\states \to \Bool$ --- a mapping marking which states
-        are accepting
-\item $\D.\delta : \Sigma \to \D.\states \to \D.\states$ --- a deterministic transition function. Given a character $\c$ and a state
-        $\s$, $\D.\delta~\c~\s$ is the state for which there is a transition $\s \overset{\stringquote{c}}{\to} (\D.\delta~\c~\s)$
-\end{itemize}
-
-\begin{figure}
-\begin{lstlisting}
-data Trace(*@$_{\D}$@*) : (s : D.states) (b : Bool) (*@$\to$@*) L where
-  nil : (*@$\uparrow$@*)(&[ s (*@$\in$@*) D.states ] Trace(*@$_{\D}$@*) s D.isAcc s)
-  cons : (*@$\uparrow$@*)(&[ c (*@$\in$@*) (*@$\Sigma$@*) ] &[ s (*@$\in$@*) D.states ] &[ b (*@$\in$@*) Bool ]
-             ('c' (*@$\lto$@*) Trace(*@$_{\D}$@*) (D.(*@$\delta$@*) c s) b (*@$\lto$@*) Trace(*@$_{\D}$@*) s b))
-\end{lstlisting}
-\caption{Traces through the DFA $D$ as an indexed inductive type}
-\label{fig:dfatrace}
-\end{figure}
-
-$\Trace_{\D}$, the linear type of traces through $\D$, is given in
-\cref{fig:dfatrace}. These
-traces may be terminated in an accepting state $\s$ with the $\nil$ constructor.
-The $\cons$ constructor builds a trace out of state $\s$ by linearly combining
-a parse of some character $\c$ with a
-trace out of the state $\D.\delta~\c~\s$. The trace built with $\cons$ is
-accepting if and only if the trace out of $D.\delta~c~s$ is accepting.
-
-Because DFAs are deterministic, we are able to prove that their type of traces
-are unambiguous and define a parser for them. In particular we show that
-$\LinPiTy {s} {D.\states} {\LinSigTy {b} {\Bool} \Trace_{D}~s~b}$ is strongly
-equivalent to
-$\stringg$, which simultaneously constructs a parser and demonstrates that
-traces through $D$ is unambiguous. Intuitively this strong equivalence says that we may
-unambiguously build traces through $D$ beginning at any state $s$, and further
-that from any trace through $D$ we can uniquely recover the string that built
-the trace.
-
-The first half of the equivalence,
+%% \paragraph{Deterministic Finite Automata}
+%% Just as the above defines an NFA, we provide a specification for an arbitrary
+%% deterministic finite automaton.\ A DFA $D$ is
+%% given by the following data,
+
+%% \begin{itemize}
+%% \item $\D.\states$ --- a finite set of states
+%% \item $\D.\init : \D.\states$ --- the initial state
+%%   \item $\D.\isAcc : \D.\states \to \Bool$ --- a mapping marking which states
+%%         are accepting
+%% \item $\D.\delta : \Sigma \to \D.\states \to \D.\states$ --- a deterministic transition function. Given a character $\c$ and a state
+%%         $\s$, $\D.\delta~\c~\s$ is the state for which there is a transition $\s \overset{\stringquote{c}}{\to} (\D.\delta~\c~\s)$
+%% \end{itemize}
+
+%% \begin{figure}
+
+%% \caption{Traces through the DFA $D$ as an indexed inductive type}
+%% \label{fig:dfatrace}
+%% \end{figure}
+
+$\Trace_{\D}$, the linear type of traces through $\D$, is given
+next. Unlike traces for an NFA, we parameterize this type additionally
+by a boolean which says whether the trace is accepting or
+rejecting. These traces may be terminated in an accepting state $\s$
+with the $\nil$ constructor.  The $\cons$ constructor builds a trace
+out of state $\s$ by linearly combining a parse of some character $\c$
+with a trace out of the state $\D.\delta~\c~\s$. The trace built with
+$\cons$ is accepting if and only if the trace out of $D.\delta~c~s$ is
+accepting.
+
+Because DFAs are deterministic, we are able to prove that their type
+of traces are unambiguous and define a parser for them. In particular
+we show that for any start state $s$, ${\LinSigTy {b} {\Bool}
+  \Trace_{D}~s~b}$ is a retract of $\stringg$. That is, first we
+construct a function $\parse_{D}$ that is a parser for
+$\Trace_D~{s}~{true}$, with $\Trace_D~{s}~{false}$ being the disjoint
+type used. The fact that these types are disjoint follows by showing
+that this function is part of a retraction, i.e., that there is only
+one way to trace through a deterministic automaton.
+
+The parser,
 $\parse_{D}$,
-is defined by induction on strings in \cref{fig:dfaparser}. If this string is empty, then $\parse_{D}$
+is defined by recursion on strings in \cref{fig:printdfa}. If this string is empty, then $\parse_{D}$
 defines a linear function that terminates a trace at the input state $s$. If the
@@
 the character at the head of the string.
+%
+The inverse, $\print_{D}$ is defined by recursion on traces. If the trace is defined
+via $\nil$, then $\print_{D}$ returns the empty string. Otherwise, if the trace
+is defined by $\cons$ then $\parse_{D}$ appends the character from the most
+recent transition to the output
+string and recurses.
+%
+We prove this is a retraction by induction on traces.
+
+\begin{theorem}[\agdalogo]
+  $\parse_D s$ is a parser for $\Trace_D~s~\true$.
+\end{theorem}
 
@@
 \begin{lstlisting}
-parse(*@$_D$@*) : (*@$\uparrow$@*)(String (*@$\lto$@*) &[ s (*@$\in$@*) D.states ] (*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*) s b)
+parse(*@$_D$@*) : (*@$\uparrow$@*)(String (*@$\lto$@*) &[ s : D.states ] (*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\D}$@*) s b)
 parse(*@$_D$@*) String.nil s = (*@$\sigma$@*) (D.isAcc s) (Trace(*@$_{\D}$@*).nil s)
@@
    (*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c s b a t)
-\end{lstlisting}
-\caption{Parser for DFA traces}
-\label{fig:dfaparser}
-\end{figure}
-
-The second half of the equivalence,
-$\print_{D}$,
-is given in \cref{fig:printdfa} by induction on traces. If the trace is defined
-via $\nil$, then $\print_{D}$ returns the empty string. Otherwise, if the trace
-is defined by $\cons$ then $\parse_{D}$ appends the character from the most
-recent transition to the output
-string and recurses.
-
-\begin{figure}
-\begin{lstlisting}
-print(*@$_D$@*) : (s : D.states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b (*@$\in$@*) Bool ] Trace(*@$_{\D}$@*) s b) (*@$\lto$@*) String)
+
+print(*@$_D$@*) : (s : D.states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\D}$@*) s b) (*@$\lto$@*) String)
 print(*@$_D$@*) s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).nil .s)) = String.nil
@@
 \end{lstlisting}
-\caption{Printer for DFA traces}
+\caption{Parser/Printer for DFA traces}
 \label{fig:printdfa}
@@
 
+Working backwards, we can then show the traces of an NFA are weakly
+equivalent to the traces of a DFA implementing a variant of Rabin and
+Scott's classic powerset construction
+\cite{rabinFiniteAutomataTheir1959}. Here we note that this is
+\emph{only} a weak equivalence and not a strong equivalence, as the
+DFA is unambiguous even if the NFA is not.
 \begin{theorem}[Determinization, \Agda]
@@
 \begin{proof}
-  We present a variant of Rabin and Scott's classic powerset construction \cite{rabinFiniteAutomataTheir1959}. First, we build
-  the DFA $\D$. Define the
-  states of $\D$ to be the $\mathbb{P}_{\varepsilon}(\N.\states)$ --- the type
+  Define the states of $\D$ to be the $\mathbb{P}_{\varepsilon}(\N.\states)$ --- the type
   of $\epsilon$-closed\footnote{A subset of states $\X$ is $\epsilon$-closed if
@@
   $\ltonl{(Parse_{N} \lto Parse_{D})}$, we strengthen our inductive hypothesis
-  and build the term
+  and build a term
 \newcommand{\NtoD}{\texttt{NtoD}}
@@
   parse transformer.
-
-\begin{figure}
-\begin{lstlisting}
-NtoD : (s : N.states) (*@$\to$@*)
-       (*@$\uparrow$@*)(Trace(*@$_{\N}$@*) s true (*@$\lto$@*)
-           &[ X (*@$\in$@*) D.states ] &[ sInX (*@$\in$@*) (X (*@$\ni$@*) s) ] Trace(*@$_{\D}$@*) X true)
-NtoD .s (nil s) X sInX = nil s
-NtoD .(N.src t) (N.cons trans .true a dstTrace) X sInX =
-  D.cons (N.label t) X true a
-    (((NtoD s dstTrace).(*@$\pi$@*) (D.(*@$\delta$@*) (N.label t) X)).(*@$\pi$@*) (N.dst t))
-NtoD .(N.(*@$\epsilon$@*)src t) ((*@$\epsilon$@*)cons (*@$\epsilon$@*)trans .true (*@$\epsilon$@*)DstTrace) X sInX =
-  ((NtoD s (*@$\epsilon$@*)DstTrace).(*@$\pi$@*) X).(*@$\pi$@*) (N.(*@$\epsilon$@*)dst t)
-\end{lstlisting}
-\caption{Parse transformer from NFA traces to Powerset DFA traces}
-\label{fig:NtoD}
-\end{figure}
-
-Because the state $\s$ is now arbitrary, we can define $\NtoD$ via
-recursion on the trace through the NFA. Code for this is given in \cref{fig:NtoD}.
-If the NFA trace is terminating at the accepting state $\s$ and $\s \in \X$,
-then we return a trace in $\D$ that terminates at
-$\X$, which is accepting because it contains the accepting state $\s$.
-
-If the NFA trace is defined via $\cons$ for a labeled transition
-$t : N.transitions$ such that $N.src~t \in \X$, then
-we first decompose the trace into a parse of $N.label~t$ and a
-trace starting from $N.dst~t$. We then recursively call $\NtoD$
-on this trace which returns a family of traces parameterized by $\epsilon$-closed
-subsets containing $N.\dst~t$. We instantiate this family at the
-set of states
-$\D.\delta~(N.label~t)~X$. Because $\N.\dst~t$ is
-reachable from $\N.\src~\texttt{trans}$ by a single transition labeled by
-$N.label~t$ and $N.src~t \in X$ we have that
-$N.dst~t \in \D.\delta~(N.label~t)~X$. Therefore
-we can linearly combine the parse of the label and the trace built from the
-recursive call to get a trace in $\D$ rooted at $X$.
-
-Lastly, if the NFA trace is defined via $\epsilon\cons$ for an $\epsilon$-transition
-$t : \N.\epstransitions$ such that
-$\N.\epssrc~t \in \X$, then the construction is quite
-similar. In this case we again recursively call $\NtoD$, except afterwards we
-instantiate the family of grammars at the set $X$ itself. We may do this because
-all states in $\D$ are taken to be $\epsilon$-closed, therefore
-$\N.\epsdst~t \in \X$ because
-$\N.\epssrc~t$ is in $X$. This concludes the construction of $\NtoD$.
+  Vice-versa,
+
+%% \begin{figure}
+%% \begin{lstlisting}
+%% NtoD : (s : N.states) (*@$\to$@*)
+%%        (*@$\uparrow$@*)(Trace(*@$_{\N}$@*) s true (*@$\lto$@*)
+%%            &[ X : D.states ] &[ sInX : (X (*@$\ni$@*) s) ] Trace(*@$_{\D}$@*) X true)
+%% NtoD .s (nil s) X sInX = nil s
+%% NtoD .(N.src t) (N.cons trans .true a dstTrace) X sInX =
+%%   D.cons (N.label t) X true a
+%%     (((NtoD s dstTrace).(*@$\pi$@*) (D.(*@$\delta$@*) (N.label t) X)).(*@$\pi$@*) (N.dst t))
+%% NtoD .(N.(*@$\epsilon$@*)src t) ((*@$\epsilon$@*)cons (*@$\epsilon$@*)trans .true (*@$\epsilon$@*)DstTrace) X sInX =
+%%   ((NtoD s (*@$\epsilon$@*)DstTrace).(*@$\pi$@*) X).(*@$\pi$@*) (N.(*@$\epsilon$@*)dst t)
+%% \end{lstlisting}
+%% \caption{Parse transformer from NFA traces to Powerset DFA traces}
+%% \label{fig:NtoD}
+%% \end{figure}
+
+%% Because the state $\s$ is now arbitrary, we can define $\NtoD$ via
+%% recursion on the trace through the NFA. Code for this is given in \cref{fig:NtoD}.
+%% If the NFA trace is terminating at the accepting state $\s$ and $\s \in \X$,
+%% then we return a trace in $\D$ that terminates at
+%% $\X$, which is accepting because it contains the accepting state $\s$.
+
+%% If the NFA trace is defined via $\cons$ for a labeled transition
+%% $t : N.transitions$ such that $N.src~t \in \X$, then
+%% we first decompose the trace into a parse of $N.label~t$ and a
+%% trace starting from $N.dst~t$. We then recursively call $\NtoD$
+%% on this trace which returns a family of traces parameterized by $\epsilon$-closed
+%% subsets containing $N.\dst~t$. We instantiate this family at the
+%% set of states
+%% $\D.\delta~(N.label~t)~X$. Because $\N.\dst~t$ is
+%% reachable from $\N.\src~\texttt{trans}$ by a single transition labeled by
+%% $N.label~t$ and $N.src~t \in X$ we have that
+%% $N.dst~t \in \D.\delta~(N.label~t)~X$. Therefore
+%% we can linearly combine the parse of the label and the trace built from the
+%% recursive call to get a trace in $\D$ rooted at $X$.
+
+%% Lastly, if the NFA trace is defined via $\epsilon\cons$ for an $\epsilon$-transition
+%% $t : \N.\epstransitions$ such that
+%% $\N.\epssrc~t \in \X$, then the construction is quite
+%% similar. In this case we again recursively call $\NtoD$, except afterwards we
+%% instantiate the family of grammars at the set $X$ itself. We may do this because
+%% all states in $\D$ are taken to be $\epsilon$-closed, therefore
+%% $\N.\epsdst~t \in \X$ because
+%% $\N.\epssrc~t$ is in $X$. This concludes the construction of $\NtoD$.
 
@@
 
-For a moment, let us turn our attention to the content behind the traces in $\D$.
-Traces from subset $\X$ to subset $\Y$ describes the \emph{existence} of some trace
-in the $\N$ from some $\x \in \X$ to some $\y \in \Y$. In general it is not clear how
-to extract out a choice of trace in $\N$, which is what $\DtoN$ is asking of us.
-However, we may further take the states and transitions of $\N$ to be \emph{ordered}.
-This allows us to
-induce an order on traces in $\D$ from $\X$ to $\Y$. From these orderings we can then
-define the necessary choice functions on each type by taking the smallest with
-respect to the ordering.
-
-We define $\DtoN$ by recursion on traces through $\D$. First if the trace stops at
-some accepting $\epsilon$-closed subset $\X$, then we may choose some state
-$\s \in X$ such that $\s$ is accepting and terminate the trace in $\N$ at $\s$.
-
-Next if the trace in $\D$ is defined via transition
-$\X \overset{\stringquote{c}}{\to} (\D.\delta~\c~\X)$ then we may choose some state in
-$\s \in \D.\delta~\c~\X$. Recursively, $\DtoN$ sends the trace rooted at
-$\D.\delta~\c~\X$ to one rooted at $\s$. By definition of $\D.\delta~\c~\X$,
-there exists some $\s' \in \X$ such that $\s'$ transitions to $\s$ by some number of
-$\epsilon$-transitions followed by a transition labeled by $\c$.
-We order all
-such traces of this form and choose the smallest one, which is then linearly
-combined with the parse of $\c$ to build an $\N$ trace. This concludes the
-definition of $\DtoN$, thus showing that $Parse_{N}$ and $Parse_{D}$ are
-weakly equivalent.
+%% For a moment, let us turn our attention to the content behind the traces in $\D$.
+%% Traces from subset $\X$ to subset $\Y$ describes the \emph{existence} of some trace
+%% in the $\N$ from some $\x \in \X$ to some $\y \in \Y$. In general it is not clear how
+%% to extract out a choice of trace in $\N$, which is what $\DtoN$ is asking of us.
+%% However, we may further take the states and transitions of $\N$ to be \emph{ordered}.
+%% This allows us to
+%% induce an order on traces in $\D$ from $\X$ to $\Y$. From these orderings we can then
+%% define the necessary choice functions on each type by taking the smallest with
+%% respect to the ordering.
+
+%% We define $\DtoN$ by recursion on traces through $\D$. First if the trace stops at
+%% some accepting $\epsilon$-closed subset $\X$, then we may choose some state
+%% $\s \in X$ such that $\s$ is accepting and terminate the trace in $\N$ at $\s$.
+
+%% Next if the trace in $\D$ is defined via transition
+%% $\X \overset{\stringquote{c}}{\to} (\D.\delta~\c~\X)$ then we may choose some state in
+%% $\s \in \D.\delta~\c~\X$. Recursively, $\DtoN$ sends the trace rooted at
+%% $\D.\delta~\c~\X$ to one rooted at $\s$. By definition of $\D.\delta~\c~\X$,
+%% there exists some $\s' \in \X$ such that $\s'$ transitions to $\s$ by some number of
+%% $\epsilon$-transitions followed by a transition labeled by $\c$.
+%% We order all
+%% such traces of this form and choose the smallest one, which is then linearly
+%% combined with the parse of $\c$ to build an $\N$ trace. This concludes the
+%% definition of $\DtoN$, thus showing that $Parse_{N}$ and $Parse_{D}$ are
+%% weakly equivalent.
 \end{proof}
 
+Finally, given any regular expression we can construct a
+\emph{strongly} equivalent NFA. While only weak equivalence is
+required to construct a parser, proving the strong equivalence shows
+that other aspects of formal grammar theory are also verifiable in
+\theoryabbv.
 \newcommand{\R}{\texttt{R}}
@@
 \end{theorem}
-
 \begin{proof}
-  \steven{TODO this proof needs to be rewritten. Likely through use of equalizers}
-  \pedro{Since this proof and the determinization one are well-known and we have
-    proved it in Agda, it's probably a good idea to include a much shorter proof sketch,
-    or move it to the appendix, though this would feel a bit superfluous.}
-We present a variant of Thompson's construction
-\cite{thompsonProgrammingTechniquesRegular1968} in which we recurse on the
-structure of $\R$ to construct a strongly equivalent NFAs. In this paper we will walk through
-some of the cases of this proof. The remaining cases
-follow
-similar reasoning and the details may be found in the formalized artifact.
-
-\begin{figure}
-  \begin{tikzpicture}[node distance = 17mm ]
-    \node[state, initial] (0) {$\0$};
-    \node[state, right of=0, accepting] (1) {$\1$};
-
-    \path[->] (0) edge[above] node{$\stringquote{c}$} (1);
-  \end{tikzpicture}
-  \caption{NFA $\N_{\c}$ for a single character $\c$}
-  \label{fig:litnfa}
-\end{figure}
-
-First handle the case where $R$ is a literal character $\literal c$. The NFA $N_{c}$ pictured in
-\cref{fig:litnfa} is strongly equivalent to the grammar $c$. To demonstrate this
-strong equivalence, we build functions between $\literal c$ and $Parse_{N_{c}}$
-and show they are mutually inverse.
-
-\begin{figure}
-\begin{lstlisting}
-A : (s : Fin 2) (*@$\to$@*) L
-A 0 = 'c'
-A 1 = I
-
-toN(*@$_c$@*) : (s : Fin 2) (*@$\to$@*) (*@$\uparrow$@*)(A s (*@$\lto$@*) Trace(*@$_{N_c}$@*) s true)
-toN(*@$_c$@*) 0 c = cons t true c (nil 1)
-toN(*@$_c$@*) 1 () = nil 1
-
-fromN(*@$_c$@*) : (s : Fin 2) (*@$\to$@*) (*@$\uparrow$@*)(Trace(*@$_{N_c}$@*) s true (*@$\lto$@*) A s)
-fromN(*@$_c$@*) 0 (cons t .true c tr) = let a (*@$\otimes$@*) () = c (*@$\otimes$@*) fromN(*@$_c$@*) 1 tr in a
-fromN(*@$_c$@*) 1 (nil .1) = ()
-\end{lstlisting}
-\caption{Parse transformer out of $Trace_{N_{c}}~s~true$}
-\label{fig:litNFAterms}
-\end{figure}
-
-In \cref{fig:litNFAterms} we define a family of grammars $A$ that serve as
-interpretations for each state in $N_{c}$. We then prove that each $A~s$ is
-equivalent to the accepting traces through $N_{c}$ out of state $s$.
-
-When
-defining $fromN_{c}$ we pattern match on the accepting traces beginning in each
-state. Out of state $0$ the only accepting traces are built via $cons$ over the
-single transition $t$, and out
-of state $1$ the only accepting trace is built via $nil$.
-
-Each $A~s$ is unambiguous, so $(fromN_{c}~s) \circ (toN_{c}~s)$ is equal to the
-identity on each $A~s$. To complete the proof of strong equivalence it remains
-to show that $(toN_{c}~s) \circ (fromN_{c}~s)$ is equal to the identity on
-$Trace~s~true$, which we show by using the equalizer type.
-
-\begin{figure}
-\begin{lstlisting}
-eqHom(*@$_{N_c}$@*) : (s : N(*@$_c$@*).state) (*@$\to$@*)
-      (*@$\uparrow$@*)(Trace(*@$_{N_c}$@*) s true (*@$\lto$@*) {tr | (toN(*@$_c$@*) s (*@$\circ$@*) fromN(*@$_{c}$@*) s) tr = id tr})
-\end{lstlisting}
-\caption{Signature for the equalizer term}
-\label{fig:litequalizer}
-\end{figure}
-
-The type signature for the family of functions, $eqHom_{N_{c}}$, into the equalizer type is given in
-\cref{fig:litequalizer}. By defining an $Trace_{N_{c}}$-algebra homomorphism
-into the equalizer, we enable a
-proof strategy for equality by induction on traces through $N_{c}$. First consider the traces
-out of state $1$. The only case to handle here is $nil 1$, and by definition
-$toN_{c}~1$ and $fromN_{c}~1$ invert each other.
-When defining $eqHom_{N_{c}}$ for traces out of state $0$, we must leverage an
-inductive hypothesis,
-
-\begin{align*}
-  (toN_{c}~0 \circ fromN_{c}~0) (cons~t~true~c~tr)
-  & \equiv (toN_{c}~0) (let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a) \\
-  & \equiv cons~t~true~(let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a)~(nil~1) \\
-  & \equiv cons~t~true~c~(nil~1) \\
-  & \equiv cons~t~true~c~tr \\
-\end{align*}
-
-The first two equalities are by defintion. The thrid equality holds because
-$\literal c$ is unambiguous and
-$let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a$ and $c$ have type
-$\literal c$. The final equality holds because our inductive hypothesis states
-that $nil~1 \equiv tr$. Thus, each $A s$ and $Trace_{N_{c}}~s~true$ are strongly
-equivalent --- in particular, $\literal c$ and $Parse_{N_{c}}$.
-
-\steven{Rewrite the disjunction case with equalizers}
-Now consider the case where $R$ is the disjunction of two regular expressions
-$R_{1}$ and $R_{2}$. Inductively, we may build NFAs $N_{1}$ and $N_{2}$ that are
-strongly equivalent to $R_{1}$ and $R_{2}$, respectively. We define $N_{\oplus}$
-as a new NFA built as the disjunction of $N_{1}$ and $N_{2}$.
-$N_{\oplus}.\states$ comprises the states from $N_{1}$, the states from
-$N_{2}$, and a new marked initial state. The accepting states are precisely
-those that were accepting in the subautomata. The initial state has an
-$\varepsilon$-transition to each of the initial states of the subautomata. Once in
-$N_{1}.\init$, the downstream transitions are a copy of those in $N_{1}$,
-likewise for $N_{2}.\init$.  That is, $N_{\oplus}$ contains a copy of each
-$N_{1}$ and $N_{2}$, and we choose which copy to inhabit by deciding which
-$\varepsilon$-transition out of the initial state to take. A schematic
-representation of this automaton is given in \cref{fig:disjunctionnfa}.
-
-\begin{figure}
-  \begin{tikzpicture}[node distance = 17mm ]
-    \node[state, initial] (init) {$\init$};
-    \node[state, above right of=init] (init1) {$N_{1}.\init$};
-    \node[state, below right of=init] (init2) {$N_{2}.\init$};
-    \node[right of=init1] (dots1) {$\dots$};
-    \node[right of=init2] (dots2) {$\dots$};
-    \node[state, accepting, right of=dots1] (acc1) {$s_{1}$};
-    \node[state, accepting, right of=dots2] (acc2) {$s_{2}$};
-
-    \path[->] (init) edge[below] node{$\varepsilon$} (init1)
-              (init) edge[above] node{$\varepsilon$} (init2)
-              (init1) edge[above] node{} (dots1)
-              (init2) edge[above] node{} (dots2)
-              (dots1) edge[above] node{} (acc1)
-              (dots2) edge[above] node{} (acc2);
-  \end{tikzpicture}
-  \caption{NFA $N_{\oplus}$ as a disjunction of $N_{1}$ and $N_{2}$}
-  \label{fig:disjunctionnfa}
-\end{figure}
-
-Because $N_{1}$ and $N_{2}$ are each strongly equivalent to $R_{1}$ and $R_{2}$,
-it suffices to show that
-$Parse_{N_{1}} \oplus Parse_{N_{2}}$ is strongly
-equivalent to $Parse_{N_{\oplus}}$.
-
-Just like the case of a single character, when building the maps in either
-direction, we strengthen the induction hypothesis by mapping out of traces
-beginning with an arbitrary state then proceed by induction on traces.
-
-First, define the
-mapping out of $\Trace_{N_{\oplus}}(s)$. If $s$ is a state from $N_{1}$ then
-send the trace to a trace in $N_{1}$ beginning at $s$. Likewise, if $s$ is from
-$N_{2}$, then we map to a trace in $N_{2}$ beginning at $s$. If $s$ is
-$N_{\oplus}.\init$, then we map the trace to a disjunction of traces
-$\Trace_{N_{1}}(N_{1}.\init) \oplus \Trace_{N_{2}}(N_{2}.\init)$.
-
-In the other direction, we need to map a disjunction of traces through $N_{1}$
-or $N_{2}$ to a trace in $N_{\oplus}$. Without loss of generality, assume that
-the trace came from $N_{1}$. That is, assume we currently have a term of type
-$\Trace_{{N_{1}}}(N_{1}.\init)$ and we aim to construct a term of type
-$\Trace_{N_{\oplus}}(N_{\oplus}.\init)$. We induct on the trace $N$ trace an map
-it over to a trace beginning at the copy of $N_{1}.\init$ in $N_{\oplus}$, and
-then we conclude by using the $\epscons$ constructor to turn this trace into one
-beginning at $N_{\oplus}.\init$.
-
-We show that these maps mutually invert each other by the same sort of argument
-we used with the single character NFA.\ That is, we prove that the each map is a
-homomorphism of algebras and then appeal to the initiality of each of
-$\Trace_{N_{1}}$, $\Trace_{N_{2}}$, and $\Trace_{N_{\oplus}}$ to prove that the
-relevant maps compose to the identity.
-
-In this sketch, we elide the proof that these maps do indeed form homomorphisms.
-Further, we are eliding the needed NFAs for the cases where the regular
-expression is $I$, $0$, a Kleene star, or a concatenation. All of these
-elided details may be found in the Agda formalization.
+We use a variant of Thompson's construction
+\cite{thompsonProgrammingTechniquesRegular1968}, showing that NFAs are, up to strong equivalence, closed under each type operation for regular expressions.
+
+%% In this paper we will walk through
+%% some of the cases of this proof. The remaining cases
+%% follow
+%% similar reasoning and the details may be found in the formalized artifact.
+
+%% \begin{figure}
+%%   \begin{tikzpicture}[node distance = 17mm ]
+%%     \node[state, initial] (0) {$\0$};
+%%     \node[state, right of=0, accepting] (1) {$\1$};
+
+%%     \path[->] (0) edge[above] node{$\stringquote{c}$} (1);
+%%   \end{tikzpicture}
+%%   \caption{NFA $\N_{\c}$ for a single character $\c$}
+%%   \label{fig:litnfa}
+%% \end{figure}
+
+%% First handle the case where $R$ is a literal character $\literal c$. The NFA $N_{c}$ pictured in
+%% \cref{fig:litnfa} is strongly equivalent to the grammar $c$. To demonstrate this
+%% strong equivalence, we build functions between $\literal c$ and $Parse_{N_{c}}$
+%% and show they are mutually inverse.
+
+%% \begin{figure}
+%% \begin{lstlisting}
+%% A : (s : Fin 2) (*@$\to$@*) L
+%% A 0 = 'c'
+%% A 1 = I
+
+%% toN(*@$_c$@*) : (s : Fin 2) (*@$\to$@*) (*@$\uparrow$@*)(A s (*@$\lto$@*) Trace(*@$_{N_c}$@*) s true)
+%% toN(*@$_c$@*) 0 c = cons t true c (nil 1)
+%% toN(*@$_c$@*) 1 () = nil 1
+
+%% fromN(*@$_c$@*) : (s : Fin 2) (*@$\to$@*) (*@$\uparrow$@*)(Trace(*@$_{N_c}$@*) s true (*@$\lto$@*) A s)
+%% fromN(*@$_c$@*) 0 (cons t .true c tr) = let a (*@$\otimes$@*) () = c (*@$\otimes$@*) fromN(*@$_c$@*) 1 tr in a
+%% fromN(*@$_c$@*) 1 (nil .1) = ()
+%% \end{lstlisting}
+%% \caption{Parse transformer out of $Trace_{N_{c}}~s~true$}
+%% \label{fig:litNFAterms}
+%% \end{figure}
+
+%% In \cref{fig:litNFAterms} we define a family of grammars $A$ that serve as
+%% interpretations for each state in $N_{c}$. We then prove that each $A~s$ is
+%% equivalent to the accepting traces through $N_{c}$ out of state $s$.
+
+%% When
+%% defining $fromN_{c}$ we pattern match on the accepting traces beginning in each
+%% state. Out of state $0$ the only accepting traces are built via $cons$ over the
+%% single transition $t$, and out
+%% of state $1$ the only accepting trace is built via $nil$.
+
+%% Each $A~s$ is unambiguous, so $(fromN_{c}~s) \circ (toN_{c}~s)$ is equal to the
+%% identity on each $A~s$. To complete the proof of strong equivalence it remains
+%% to show that $(toN_{c}~s) \circ (fromN_{c}~s)$ is equal to the identity on
+%% $Trace~s~true$, which we show by using the equalizer type.
+
+%% \begin{figure}
+%% \begin{lstlisting}
+%% eqHom(*@$_{N_c}$@*) : (s : N(*@$_c$@*).state) (*@$\to$@*)
+%%       (*@$\uparrow$@*)(Trace(*@$_{N_c}$@*) s true (*@$\lto$@*) {tr | (toN(*@$_c$@*) s (*@$\circ$@*) fromN(*@$_{c}$@*) s) tr = id tr})
+%% \end{lstlisting}
+%% \caption{Signature for the equalizer term}
+%% \label{fig:litequalizer}
+%% \end{figure}
+
+%% The type signature for the family of functions, $eqHom_{N_{c}}$, into the equalizer type is given in
+%% \cref{fig:litequalizer}. By defining an $Trace_{N_{c}}$-algebra homomorphism
+%% into the equalizer, we enable a
+%% proof strategy for equality by induction on traces through $N_{c}$. First consider the traces
+%% out of state $1$. The only case to handle here is $nil 1$, and by definition
+%% $toN_{c}~1$ and $fromN_{c}~1$ invert each other.
+%% When defining $eqHom_{N_{c}}$ for traces out of state $0$, we must leverage an
+%% inductive hypothesis,
+
+%% \begin{align*}
+%%   (toN_{c}~0 \circ fromN_{c}~0) (cons~t~true~c~tr)
+%%   & \equiv (toN_{c}~0) (let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a) \\
+%%   & \equiv cons~t~true~(let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a)~(nil~1) \\
+%%   & \equiv cons~t~true~c~(nil~1) \\
+%%   & \equiv cons~t~true~c~tr \\
+%% \end{align*}
+
+%% The first two equalities are by defintion. The thrid equality holds because
+%% $\literal c$ is unambiguous and
+%% $let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a$ and $c$ have type
+%% $\literal c$. The final equality holds because our inductive hypothesis states
+%% that $nil~1 \equiv tr$. Thus, each $A s$ and $Trace_{N_{c}}~s~true$ are strongly
+%% equivalent --- in particular, $\literal c$ and $Parse_{N_{c}}$.
+
+%% \steven{Rewrite the disjunction case with equalizers}
+%% Now consider the case where $R$ is the disjunction of two regular expressions
+%% $R_{1}$ and $R_{2}$. Inductively, we may build NFAs $N_{1}$ and $N_{2}$ that are
+%% strongly equivalent to $R_{1}$ and $R_{2}$, respectively. We define $N_{\oplus}$
+%% as a new NFA built as the disjunction of $N_{1}$ and $N_{2}$.
+%% $N_{\oplus}.\states$ comprises the states from $N_{1}$, the states from
+%% $N_{2}$, and a new marked initial state. The accepting states are precisely
+%% those that were accepting in the subautomata. The initial state has an
+%% $\varepsilon$-transition to each of the initial states of the subautomata. Once in
+%% $N_{1}.\init$, the downstream transitions are a copy of those in $N_{1}$,
+%% likewise for $N_{2}.\init$.  That is, $N_{\oplus}$ contains a copy of each
+%% $N_{1}$ and $N_{2}$, and we choose which copy to inhabit by deciding which
+%% $\varepsilon$-transition out of the initial state to take. A schematic
+%% representation of this automaton is given in \cref{fig:disjunctionnfa}.
+
+%% \begin{figure}
+%%   \begin{tikzpicture}[node distance = 17mm ]
+%%     \node[state, initial] (init) {$\init$};
+%%     \node[state, above right of=init] (init1) {$N_{1}.\init$};
+%%     \node[state, below right of=init] (init2) {$N_{2}.\init$};
+%%     \node[right of=init1] (dots1) {$\dots$};
+%%     \node[right of=init2] (dots2) {$\dots$};
+%%     \node[state, accepting, right of=dots1] (acc1) {$s_{1}$};
+%%     \node[state, accepting, right of=dots2] (acc2) {$s_{2}$};
+
+%%     \path[->] (init) edge[below] node{$\varepsilon$} (init1)
+%%               (init) edge[above] node{$\varepsilon$} (init2)
+%%               (init1) edge[above] node{} (dots1)
+%%               (init2) edge[above] node{} (dots2)
+%%               (dots1) edge[above] node{} (acc1)
+%%               (dots2) edge[above] node{} (acc2);
+%%   \end{tikzpicture}
+%%   \caption{NFA $N_{\oplus}$ as a disjunction of $N_{1}$ and $N_{2}$}
+%%   \label{fig:disjunctionnfa}
+%% \end{figure}
+
+%% Because $N_{1}$ and $N_{2}$ are each strongly equivalent to $R_{1}$ and $R_{2}$,
+%% it suffices to show that
+%% $Parse_{N_{1}} \oplus Parse_{N_{2}}$ is strongly
+%% equivalent to $Parse_{N_{\oplus}}$.
+
+%% Just like the case of a single character, when building the maps in either
+%% direction, we strengthen the induction hypothesis by mapping out of traces
+%% beginning with an arbitrary state then proceed by induction on traces.
+
+%% First, define the
+%% mapping out of $\Trace_{N_{\oplus}}(s)$. If $s$ is a state from $N_{1}$ then
+%% send the trace to a trace in $N_{1}$ beginning at $s$. Likewise, if $s$ is from
+%% $N_{2}$, then we map to a trace in $N_{2}$ beginning at $s$. If $s$ is
+%% $N_{\oplus}.\init$, then we map the trace to a disjunction of traces
+%% $\Trace_{N_{1}}(N_{1}.\init) \oplus \Trace_{N_{2}}(N_{2}.\init)$.
+
+%% In the other direction, we need to map a disjunction of traces through $N_{1}$
+%% or $N_{2}$ to a trace in $N_{\oplus}$. Without loss of generality, assume that
+%% the trace came from $N_{1}$. That is, assume we currently have a term of type
+%% $\Trace_{{N_{1}}}(N_{1}.\init)$ and we aim to construct a term of type
+%% $\Trace_{N_{\oplus}}(N_{\oplus}.\init)$. We induct on the trace $N$ trace an map
+%% it over to a trace beginning at the copy of $N_{1}.\init$ in $N_{\oplus}$, and
+%% then we conclude by using the $\epscons$ constructor to turn this trace into one
+%% beginning at $N_{\oplus}.\init$.
+
+%% We show that these maps mutually invert each other by the same sort of argument
+%% we used with the single character NFA.\ That is, we prove that the each map is a
+%% homomorphism of algebras and then appeal to the initiality of each of
+%% $\Trace_{N_{1}}$, $\Trace_{N_{2}}$, and $\Trace_{N_{\oplus}}$ to prove that the
+%% relevant maps compose to the identity.
+
+%% In this sketch, we elide the proof that these maps do indeed form homomorphisms.
+%% Further, we are eliding the needed NFAs for the cases where the regular
+%% expression is $I$, $0$, a Kleene star, or a concatenation. All of these
+%% elided details may be found in the Agda formalization.
 \end{proof}
 
-\subsection{The Dyck Grammar}
-%% 3. Dyck language
+\subsection{Context-free grammars}
+
+Next, we give two examples for parsing context-free
+grammars. Context-free grammars (CFG) can be encoded in our type
+theory in a similar way to regular expressions, as CFGs are equivalent
+to the formalism of \emph{$\mu$}-regular expressions, where the Kleene
+star is replaced by an arbitrary fixed point
+operation\cite{muregular}.
+
 \newcommand{\balanced}{\mathtt{balanced}}
 \newcommand{\Dyck}{\mathtt{Dyck}}
-Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
-grammars (CFGs). An interesting nontrivial CFG is the Dyck grammar
-$\Dyck$ --- the grammar of balanced parentheses.
-
-$\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$
-and is defined in \cref{fig:dyckinductive}. The $\nil$ constructor shows that
+A simple example of a CFG is the Dyck grammar of balanced
+parentheses, which we define in \Cref{fig:dyckinductive}
+$\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$. The $\nil$ constructor shows that
 the empty string is balanced, and the $bal$ constructor builds a balanced
@@
 
-\Cref{fig:DyckAutomaton} presents a sketch of the automaton $M$. The traditional
-approach to parse grammars like $\Dyck$ is to employ a stack to count how many
-open parentheses have been seen so far. We may think of
-$M$ as performing the same task as a stack machine through slightly different means. Instead of storing
-the number of open parentheses in a distinguished stack of memory, $M$ operates
-by storing this data in the current state. $M$ is defined almost identically to how we defined DFAs, except we omit the
-finiteness condition on the states of the automaton. This similarity allows
-functions $parse_{M}$ and $print_{M}$ to be defined analogously to
-\cref{fig:dfaparser,fig:printdfa}, and so
-$\LinSigTy {s} {M.states} {\LinSigTy {b}{Bool} {Trace_{M}~s~b}}$ is likewise unambiguous.
+The Dyck language is an example of an $\textrm{LL}(0)$ language, one
+that can be parsed top-down with no lookahead\cite{ll0}. This means we
+can implement it simply as an \emph{infinite} state deterministic
+automaton, in \Cref{fig:DyckAutomaton}. Here the state is a ``stack''
+counting how many open parentheses have been seen so far. Functions
+$parse_{M}$ and $print_{M}$ for this automaton can be defined
+analogously to \cref{fig:dfaparser,fig:printdfa}, and so $\LinSigTy
+{s} {M.states} {\LinSigTy {b}{Bool} {Trace_{M}~s~b}}$ is likewise
+unambiguous.
 
@@
   \label{thm:dyck}
-  $Dyck$ and $Parse_{M}$ are strongly equivalent.
+  $Dyck$ and $Parse_{M}$ are strongly equivalent. And therefore we can construct a parser for $\Dyck$.
 \end{theorem}
 
-We omit the details of \cref{thm:dyck} in this paper, and the ful proof is
-mechanized in our Agda development. We can build a $Dyck$ parser by utilizing
-this artifact, and the
-runtime of the $Dyck$ parser is about 30 seconds for input that is 25,000
-characters long.
-\pedro{It might be a good idea to mention explicitly in the introduction
-  that we are not claiming that our verified parsers are competitive with
-  the state of the art.}
+%% We can build a $Dyck$ parser by utilizing
+%% this artifact, and the
+%% runtime of the $Dyck$ parser is about 30 seconds for input that is 25,000
+%% characters long.
+%% \pedro{It might be a good idea to mention explicitly in the introduction
+%%   that we are not claiming that our verified parsers are competitive with
+%%   the state of the art.}
 
@@
 
+Our final example is of a simple grammar of arithmetic expressions
+with an associative operation. Here we take the alphabet to be $\{
+\stringquote{(}, \stringquote{)}, \stringquote{+} \} \cup \{ n | n \in
+\mathbb N \}$ given in \Cref{fig:dyckinductive}. This grammar is
+LR(1), and the corresponding automaton (which is too complex to
+include here) uses the $\&$ to guard transitions with lookahead
+information \max{Probably we should at least try to fit it in or put
+  it in the appendix.}
+
+\begin{theorem}[\agdalogo]
+  We construct a parser for $\mathsf{Exp}$ by showing it is weakly
+  equivalent to a deterministic automaton with lookahead.
+\end{theorem}
+\begin{figure}
+\begin{lstlisting}
+data Exp : L where
+  num : (*@$\uparrow$@*)(&[ n : Nat ] 'n' (*@$\lto$@*) Exp)
+  parens : (*@$\uparrow$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ('+' (*@$\otimes$@*) Atom *) ')' (*@$\lto$@*) Exp)
+\end{lstlisting}
+%% data Opening : Nat -> Bool -> L where
+%% data Closing : Nat -> Bool -> L where
+%% data Multiplying : Nat -> Bool -> L where
+%%   done : (*@$\uparrow$@*)(Multiplying 0 true)
+%%   times : (*@$\uparrow$@*)(&[ n : Nat ] &[ b : Bool] '*' -o Opening n b -o Multiplying n b)
+%%   unmatched : (*@$\uparrow$@*)(&[ n : Nat ] Multiplying (n + 1) false)
+%%   unexpected
+\caption{Binary operation expressions and a corresponding lookahead automaton}
+\label{fig:dyckinductive}
+\end{figure}
 % \subsection{LL(1)}
@@
 \newcommand{\stringNL}{String'}
-\steven{Need to better introduce nonlinear type of strings. We haven't talked
-  about semantics nor have we included a nonlinear type of strings in the syntax}
-Define $\stringNL$ to be the non-linear list type over the alphabet $\Sigma$ and
-let $P : \stringNL \to U$ be a mapping from the set of strings to the universe of
-nonlinear types. The function $P$ describes a family of nonlinear types
-dependent on $\stringNL$.
-We view the image of $P$ as a set of parse trees for each input string, and thus
-$P$ carries the structure of a formal grammar. However, even though $P$ describes this grammatical data, it is not
-a first-class linear type in \theoryabbv. To address this gap, below we define a linear type that captures the same
-semantic grammar as $P$,
-
+
+While we have shown only examples for context-free grammars, in fact
+arbitrarily complex grammars are encodable in \theoryabbv. To
+demonstrate this, we show that for any \emph{non-linear} function $P :
+\stringNL \to U$, where here $\stringNL$ is the \emph{non-linear} type
+of strings over the alphabet, we can construct a grammar whose parses
+correspond to $P$.
+%% Define $\stringNL$ to be the non-linear list type over the alphabet $\Sigma$ and
+%% let  be a mapping from the set of strings to the universe of
+%% nonlinear types. The function $P$ describes a family of nonlinear types
+%% dependent on $\stringNL$.
+%% We view the image of $P$ as a set of parse trees for each input string, and thus
+%% $P$ carries the structure of a formal grammar. However, even though $P$ describes this grammatical data, it is not
+%% a first-class linear type in \theoryabbv. 
 \[
@@
 \]
-
-\steven{Probably put this ceiling business around the discussion of the linear
-  type of strings. Is this in section 3?}
-
+%% \steven{Probably put this ceiling business around the discussion of the linear
+%%   type of strings. Is this in section 3?}
 where $\lceil \stringquote{} \rceil = I$ and
@@
 as it allows to sidestep our linear typing connectives and utilize the whole of nonlinear
-dependent type theory to define a grammar. This power allows us to encode
-arbitrary grammars. In particular, given a
+dependent type theory to define a grammar. For example, given a
 Turing machine $Tur$ one may
-define a predicate $accepts : \stringNL \to U$ such that $accept~w$ is equal to $\top$ if
+define a non-linear predicate $accepts : \stringNL \to U$ such that $accept~w$ is equal to $\top$ if
 $T$ accepts $w$ and equal to $0$ otherwise. Then, $Reify~accepts$ is a linear
@@
 languages in the Chomsky hierarchy.
-
-\subsection{Semantic Actions}
-So far, each construction has been primarily concerned with the parsing of a
-string into a \emph{concrete} syntax tree. However, in a larger verified
-developement --- such as a compiler --- a parser will often emit an
-\emph{abstract} syntax tree that omits superflous syntactic details that aren't
-needed for later passes.  The procedure that takes in a
-concrete syntax tree and returns an abstract one is referred to as a
-\emph{semantic action}.
-
-Given a linear type $A$ and a nonlinear type $X$, we define a semantic action to
-be a function $\ltonl {(A \lto \Delta X)}$, where
-$\Delta X = \LinSigTy {x} {X} {\top}$. That is, a semantic action is an
-elaboration from the concrete parses of $A$ into the semantic domain $X$. As
-written, these actions are quite general. As discussed above, we may define
-folds that are still grammatical in nature --- such as folding over a concrete
-parse tree and forgetting parentheses that aren't needed in later compiler
-passes. We may also define arbitrary computation over the structure of the parse
-tree, such as calculating the maximum nesting depth of parentheses for $Dyck$,
-returning the length the input string, or
-even an interpreter that evaluates arithmetic expressions at parse-time.
-
-\steven{What can I say about what we have in the Agda development here?}
-\pedro{I'm assuming that this section is still incomplete, so I'll wait
-for it to be complete in order to make more detailed comments.}
+\begin{theorem}[\Agda]
+  For any Turing machine, we can construct a grammar in \theoryabbv
+  that accepts the same language as the Turing machine.
+\end{theorem}
 
@@
 % type elaborator           ~~ semantic actions
+
+\subsection{Semantic Actions}
+
+Our verification has mainly focused on the verification that a parser
+outputs a correct concrete syntax tree for a grammar. However, in
+practice, parsers are combined with a \emph{semantic action} that
+emits an \emph{abstract} syntax tree that omits superflous syntactic
+details that aren't needed in later stages of the overall program. We
+can define a semantic action in \theoryabbv for a linear type $A$ with
+semantic outputs in a non-linear type $X$ to be a function $\ltonl {(A
+  \lto \bigoplus_{\_:X} \top)}$. That is, a semantic action is a
+function that produces a semantic element of $X$ from the concrete
+parses of $A$. In future work, we aim to study the question of
+verifying efficient implementations of parsers with semantic actions.
 

### 2024-11-14 paper/paper.tex
binop example

@@
 with an associative operation. Here we take the alphabet to be $\{
-\stringquote{(}, \stringquote{)}, \stringquote{+} \} \cup \{ n | n \in
-\mathbb N \}$ given in \Cref{fig:dyckinductive}. This grammar is
-LR(1), and the corresponding automaton (which is too complex to
-include here) uses the $\&$ to guard transitions with lookahead
-information \max{Probably we should at least try to fit it in or put
-  it in the appendix.}
+\stringquote{(}, \stringquote{)}, \stringquote{+} \stringquote{NUM}
+\}$. In \Cref{fig:binop-inductive} we define it using two mutually
+recursive types, corresponding to the two non-terminals we would use
+in a CFG syntax. The syntactic structure encodes that the binary
+operation is right associative. In the same figure, we define the
+traces of an automaton with one token of lookahead. The automaton has
+four different ``states'', each with access to a natural number
+``stack''. The ``opening'' state $O$ expects either an open paren, in
+which case it increments the stack and stays in the opening state, or
+sees a number and proceeds to the $D$ state. The ``done opening''
+state $D$ is where lookahead is used: if the next token will be a
+right paren, then we proceed to $C$, otherwise we proceed to $M$. Here
+$\mathsf{NotStartsWithRP}$ is defined as $\I \oplus (('(' \oplus '+'
+\oplus \mathsf{NUM}) \otimes \top)$. In the ``closing'' state $C$ if
+we observe a close paren we decrement the count and continue to the
+$D$ state. In the ``multiplying'' state $M$, we succeed if the string
+ends and the count is $0$, and if we see a plus we continue to the $O$
+state. Additionally, since the automaton need also parse all of the
+incorrect strings, we add additional failing cases. It is
+straightforward to implement a parser for this lookahead automaton,
+generalizing the approach for deterministic automata.
 
@@
   We construct a parser for $\mathsf{Exp}$ by showing it is weakly
-  equivalent to a deterministic automaton with lookahead.
+  equivalent to $\mathsf{O}~0~\true$ 
 \end{theorem}
@@
 data Exp : L where
-  num : (*@$\uparrow$@*)(&[ n : Nat ] 'n' (*@$\lto$@*) Exp)
-  parens : (*@$\uparrow$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ('+' (*@$\otimes$@*) Atom *) ')' (*@$\lto$@*) Exp)
+  done : (*@$\uparrow$@*)(Atom (*@$\lto$@*) Exp)
+  add : (*@$\uparrow$@*)(Atom (*@$\lto$@*) '+' (*@$\lto$@*) Exp (*@$\lto$@*) Exp)
+data Atom : L where
+  num : (*@$\uparrow$@*)(NUM (*@$\lto$@*) Atom)
+  parens : (*@$\uparrow$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ')' (*@$\lto$@*) Atom)
+
+data O : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
+  left : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O (n + 1) b (*@$\lto$@*) O n b)
+  num : (*@$\uparrow$@*)(&[ n ] &[ b ] NUM (*@$\lto$@*) DO n b (*@$\lto$@*) O n b)
+  done : (*@$\uparrow$@*)(&[ n ] O n false)
+  unexpected : (*@$\uparrow$@*)(&[ n ] (')' (*@$\oplus$@*) '+') (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) O n false)
+data D : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
+  lookAheadRP  : (*@$\uparrow$@*)(&[ n ] &[ b ] ((')' (*@$\otimes$@*) (*@$\top$@*)) (*@$\&$@*) C n b) (*@$\lto$@*) D n b)
+  lookAheadNot  : (*@$\uparrow$@*)(&[ n ] &[ b ] (NotStartsWithRP (*@$\&$@*) M n b) (*@$\lto$@*) D n b)
+data C : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
+  closeGood : (*@$\uparrow$@*)(&[ n ] &[ b ] ')' (*@$\lto$@*) D n b (*@$\lto$@*) C (n + 1) b)
+  closeBad : (*@$\uparrow$@*)(&[ n ] ')' (*@$\lto$@*) C 0 b)
+  done : (*@$\uparrow$@*)(&[ n ] C n false)
+  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) '+' (*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) C n false)
+data M : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
+  doneGood : (*@$\uparrow$@*)(M 0 true)
+  doneBad : (*@$\uparrow$@*)(&[ n ] M (n + 1) false)
+  add : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O n b (*@$\lto$@*) M n b)
+  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')'(*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) M n false)
 \end{lstlisting}
-%% data Opening : Nat -> Bool -> L where
-%% data Closing : Nat -> Bool -> L where
-%% data Multiplying : Nat -> Bool -> L where
-%%   done : (*@$\uparrow$@*)(Multiplying 0 true)
-%%   times : (*@$\uparrow$@*)(&[ n : Nat ] &[ b : Bool] '*' -o Opening n b -o Multiplying n b)
-%%   unmatched : (*@$\uparrow$@*)(&[ n : Nat ] Multiplying (n + 1) false)
-%%   unexpected
-\caption{Binary operation expressions and a corresponding lookahead automaton}
-\label{fig:dyckinductive}
+\caption{Associative arithmetic expressions and a corresponding lookahead automaton}
+\label{fig:binop-inductive}
 \end{figure}

### 2024-11-14 paper/paper.tex
disjointness of constructors

@@
 use infix $\oplus/\&$ for binary disjunction/conjunction.
-%
-In addition to the introduction and elimination rules for
-$\bigamp,\bigoplus$, our development requires an additional axiom that
-is not proveable from these rules: that additive conjunction
-\emph{distributes} over additive disjunction, e.g., in the finitary
-case that $0 \& A \cong 0$ and $(A + B) \& C \cong (A \& C) + (B \&
-C)$. In its most general form, the axiom says that the definable function
-${\bigoplus\limits_{f : \prod_{x:X}
-    Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto
-  \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y}$ has an
-inverse.
 
@@
 indexed inductive types.
+
+In addition to these type-theoretic principles, we need two additional
+axioms that do not generally hold in systems based on linear
+logic. First, we need that additive conjunction \emph{distributes}
+over additive disjunction, e.g., in the finitary case that $0 \& A
+\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most
+general form, the axiom says that the definable function
+${\bigoplus\limits_{f : \prod_{x:X}
+    Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto
+  \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y}$ has an
+inverse. Second, we need that the different constructors of
+$\bigoplus$ are disjoint. We add this by adding the axiom that for any
+$A : X \to L$ and $x \neq x' : X$ there is a function
+$\uparrow(\equalizer{b}{\sigma\,x \circ \pi_1}{\sigma\,x'\circ \pi_2}
+\lto 0)$ where $b: A(x) \& A(x')$, i.e., the grammar of pairs of an
+$a:A(x)$ and an $a': A(x')$ such that $\sigma\,x\,a = \sigma\,x'\,a'$
+is empty.
 
@@
 additional effort: the negative grammar is simply the grammar for
-traces that end in a rejecting state.
+traces that end in a rejecting state. This follows from the following
+principle, a consequence of our axiom that $\oplus$ constructors are
+disjoint.
 \begin{lemma}[\agdalogo]
@@
 \[ \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\]
-Next, a parse in the equalizer $\equalizer{a : A}{f}{g}$ is defined as
+Next, a parse in the equalizer $\equalizer{a}{f}{g}$ is defined as
 a parse in $\sem{A}$ that is mapped to the same parse by $f$ and
@@
     \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
-    \sem{\equalizer{a:A}{f}{g}}\gamma\,w &= \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
+    \sem{\equalizer{a}{f}{g}}\gamma\,w &= \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
     \sem{L}\gamma &= \Grammar_0\\

### 2024-11-14 paper/paper.tex
Appendix wip

@@
 \newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
+\newcommand{\semg}[1]{\sem{#1}\gamma}
 \newcommand{\cat}[1]{\mathbf{#1}}
@@
 \newcommand{\Grammar}{\mathbf{Gr}}
+\newcommand{\semcat}{\mathbf{C}}
 \newcommand{\Type}{\mathbf{Type}}
@@
     {~}
-    {\Gamma ; \Delta \vdash \equalizerin {\equalizerpi {e}} \equiv e : A}
+    {\Gamma ; \Delta \vdash \equalizerin {\equalizerpi {e}} \equiv e :
+      \equalizer e f g}
 \end{mathpar}
@@
 \section{Denotational Semantics}
-
-\begin{definition}
-  We define the unit grammar
-  \[ I\,w = \{ ()\pipe w = \varepsilon \}\]
-  and given grammars $A$ and $B$ we define their tensor product
-  \[ (A \otimes B)\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in A\,w_1 \wedge b \in B\,w_2 \} \]
-  This extends to a monoidal structure on the category of grammars.
-Note that we defined in the semantics of linear types $\sem{I}\gamma = I$ and $\sem{A \otimes B}\gamma = \sem{A}\gamma \otimes \sem{B} \gamma$
-\end{definition}
+\steven{Define a Kleene, Lambek, and Chomsky category}
+
+Let $\semcat$ be a Chomsky category.
+
+\steven{Finish the denotation of lin types figure for spf, el , map , mu}
+\begin{figure}
+  \begin{align*}
+    \semg{I} &= I \\
+    \semg{A \otimes B} &= \semg{A} \otimes \semg{B} \\
+    \semg{A \lto B}  &= \semg{A} \lto \semg{B} \\
+    \semg{B \tol A} &= \semg{B} \tol \semg{A} \\
+    \semg{\bigoplus_{x:X} A} &= \coprod_{x:X} \semg{A} \\
+    \semg{\bigwith_{x:X} A} &= \prod_{x : X} \semg{A} \\
+    \semg{\uparrow A} &= Hom(I,A) \\
+    \semg{\equalizer{a:A}{f}{g}} &= Eq\left( \semg{f} , \semg{g} \right) \\
+    \sem{L}\gamma &= \semcat_0\\
+    \sem{\SPF\,X}\gamma &= \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
+    \sem{\el(F)}\gamma\,G &= \sem{F}\gamma\,G\\
+    \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
+    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:indexed-container}}\\
+  \end{align*}
+  \caption{Denotation of Linear Types}
+\end{figure}
 
@@
 
-We define denotations of linear terms in
-\Cref{fig:linear-term-semantics}. Note that the denotations interpret
+\begin{definition}[Denotation of Linear Substitutions]
+  The semantics of a linear substitution
+  $\delta : \textrm{Subst}(\Delta' , \Delta)$ are given as maps
+  $\semg{\delta} : \semg{\Delta'} \to {\Delta}$. Define $\semg{\delta}$ by
+  recursion on $\delta$:
+  \begin{align*}
+    \semg{\cdot} &= id_{I} \\
+    \semg{\delta,e/a} &= \semg{\delta} \otimes \semg{e} \circ m_{\Delta_{1},\Delta_{2}}
+  \end{align*}
+where $\Delta_{2} \vdash e : A$ and $\Delta' = \Delta_{1},\Delta_{2}$.
+\end{definition}
+
+\begin{theorem}
+  For any $\Gamma \vdash \Delta_1,\Delta_2 \isLinCtx$ and $\gamma \in
+  \sem{\Gamma}$ there is a natural isomorphism $m_{\Delta_1,\Delta_2} :
+  \sem{\Delta_1, \Delta_2}\gamma \cong \sem{\Delta_1}\gamma \otimes
+  \sem{\Delta_2}\gamma$.
+
+  This can be extended to a sequence of contexts of any length.
+\end{theorem}
+\begin{proof}
+Construct $m_{\Delta_{1},\Delta_{2}}$ by recursion on $\Delta_{2}$.
+\begin{align*}
+  m_{\Delta_{1},\cdot} & = \rho^{-1} \\
+  m_{\Delta_{1},(\Delta_{2}, a : A)} & = \alpha \circ m_{\Delta_{1},\Delta_{2}} \otimes id
+\end{align*}
+\end{proof}
+
+\begin{lemma}
+  \label{lem:subst}
+  For each term $\Delta \vdash e : A$ and substitution
+  $\delta : \textrm{Subst}(\Delta', \Delta)$, the semantics of $\delta$ acting
+  on $e$ splits into the composition $\semg{e[\delta]} = \semg{e} \circ \semg{\delta}$.
+\end{lemma}
+
+\subsection{Grammar Semantics for Linear Terms}
+Here we define denotations of linear terms. Note that the denotations interpret
 typing derivations, not raw terms, as the data of how contexts are
 split is needed in order to construct the correct associator
-functions.
-\max{todo: describe associators, define denotations, prove axioms}
-\begin{figure}
-  \begin{mathpar}
-  \inferrule
-  {\Delta_1 \vdash e_1 : A_1 \and \Delta_2 \vdash e_2 : A_2}
-  {\sem{(e_1,e_2)}\gamma\,w\,\delta
-    = \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}
-  }
-
-  \inferrule
-  {\Delta_2 \vdash e : A \otimes B \and \Delta_1,a:A,b:B,\Delta_3 \vdash e' : C}
-  {\sem{\letin{(a,b)}{e}{e'}}\gamma\,w\,\delta =
-    \letin{(w_1,w_2w_3,\delta_1,(w_2,w_3,\delta_2,\delta_3))}{m_{\Delta_1,\Delta_2,\Delta_3}(\delta)}
-    \letin{(w_{2,1},w_{2,2},a,b)}{\sem{e}\gamma\,w_2\,\delta_2}
-    {\sem{e'}\gamma\,w\,(w_1,w_{2,1}w_{2,2}w_3, \delta_1, (w_{2,1},w_{2,2}w_3, a, (w_{2,2},w_3, b, \delta_3)))}
-    }
-\end{mathpar}
-  \caption{Denotations of linear terms}
-\end{figure}
-
-\begin{definition}
-  For any $\Gamma \vdash \Delta_1,\Delta_2 \isLinCtx$ and $\gamma \in
-  \sem{\Gamma}$ there is an isomorphism $m_{\Delta_1,\Delta_2} :
-  \sem{\Delta_1, \Delta_2}\gamma \cong \sem{\Delta_1}\gamma \otimes
-  \sem{\Delta_2}\gamma$.
-
-  This can be extended to a sequence of contexts of any length.
-\end{definition}
-
-\begin{lemma}[Equations for $\otimes$]
-
-\end{lemma}
-
-\begin{lemma}[Equations for $I$]
-
-\end{lemma}
-
-\begin{lemma}[Equations for $\lto$]
-
-\end{lemma}
-
-\begin{lemma}[Equations for $\tol$]
-
-\end{lemma}
-
+functions. Further, we demonstrate that the denotational semantics resprects the
+equational theory of \theoryabbv.
+
+\subsubsection{Variables}
+Note that the denotation of a singleton context $a : A$ is given as
+\[
+  \semg{a : A} = \semg{\cdot , a : A} = I \otimes \semg{A}
+\]
+
+So for $a: A \vdash a : A$, the denotation of a single variable term
+$\semg{a} : \semg{a : A} \to \semg{A}$ is given by the left unitor
+\[
+  \semg{a} = \lambda
+\]
+
+\subsubsection{Linear Unit}
+% I
+\paragraph{$I$-Introduction}
+$\semg{()} : \semg{\cdot} \to \semg{I}$.
+\[
+\semg{()} = id_{I}
+\]
+
+\paragraph{$I$-Elimination}
+$\semg{\letin {()} e e'} : \semg{\Delta'_{1},\Delta,\Delta'_{2}} \to \semg{C}$
+defined in the following diagram,
+
+\begin{center}
+\begin{tikzcd}
+  \semg{\Delta'_1,\Delta,\Delta'_2} \arrow[r , "m_{(\Delta'_1,\Delta),\Delta'_2}"] &
+  \semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}
+    \arrow[r , "m_{\Delta'_1,\Delta} \otimes id"] &
+  \left(\semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2}
+    \arrow[d , "(id \otimes \semg{e}) \otimes id"] \\
+    \semg{\Delta'_1,\Delta'_2} \arrow[d , "\semg{e'}"]
+  & \arrow[l , "m^{-1}_{\Delta'_1,\Delta'_2}"]
+    \semg{\Delta'_1} \otimes \semg{\Delta'_2}
+  & \left(\semg{\Delta'_1} \otimes I \right) \otimes \semg{\Delta'_2}
+  \arrow[l , "\rho \otimes id"] \\
+  \semg{C}
+\end{tikzcd}
+\end{center}
+
+We demonstrate that the denotations of the introduction and elimination forms
+for $I$ obey the $\beta$ and $\eta$ equalities for $I$.
+
+\paragraph{$I\beta$}
+Given
+$\Delta'_{1},\cdot,\Delta'_{2} \vdash e' : C$, the desired
+$\beta$ law is
+\[
+  \semg{\letin {()} {()} {e'}} = \semg{e}
+\]
+\begin{proof}
 \begin{align*}
-  \sem{(e_1,e_2)}\gamma\,w\,\delta &=
-  \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}\\
+  \semg{\letin {()} {()} {e'}}
+  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \semg{()}) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{(\Delta'_{1},\cdot),\Delta'_{2}} \\
+  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes id) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
+  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ \rho^{-1} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
+  & = \semg{e'}
 \end{align*}
+\end{proof}
+
+\paragraph{$I\eta$}
+Similarly, for $\Delta_{1}, a : A, \Delta_{3} \vdash e' : C$ and
+$\Delta_{2} \vdash e : I$ the desired $\eta$ law is
+\[
+\semg{\letin {()} {e} {e'[()/a]}} = \semg{e'[e/a]}
+\]
+However, through application of \cref{lem:subst} it suffices to handle the case
+where $e$ is a variable $a'$. That is,
+\begin{align*}
+\letin {()} {e} {e'[()/a]} &= (\letin {()} {a'} {e'[()/a]})[e/a'] \\
+e'[e/a] &= e'[a'/a][e/a]
+\end{align*}
+so without loss of generality we may take may take $e$ to be variable $a'$.
+
+\begin{proof}
+\begin{align*}
+  \semg{\letin {()} {a'} {e'[()/a]}}
+  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \semg{a'}) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
+  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \lambda) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
+  & = \semg{e'}
+\end{align*}
+
+\steven{Define generalized associators}
+
+Because
+$m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \lambda) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}}$
+is a composition of generalized associators from
+$\semg{\Delta'_{1},\Delta'_{2}}$ to itself, it is equal to the identity by the coherence theorem for monoidal categories.
+
+Further by \cref{lem:subst},
+\begin{align*}
+  \semg{e'[a'/a]}
+  &= \semg{e'} \circ \semg{a'/a} \\
+  &= \semg{e'}
+\end{align*}
+Again by the coherence theorem, $\semg{a'a} = id$. Thus, $\eta$ law for $I$
+holds in the denotational semantics.
+\end{proof}
+
+%tensor
+\subsubsection{Tensor}
+\paragraph{$\otimes$-Introduction}
+$\semg{(e_{1}, e_{2})} : \semg{\Delta,\Delta'} \to \semg{A \otimes B}$ is given
+by the diagram
+
+\begin{tikzcd}
+  \semg{\Delta,\Delta'} \arrow[r, "m_{\Delta,\Delta'}"] &
+  \semg{\Delta} \otimes \semg{\Delta'} \arrow[r, "\semg{e_1} \otimes \semg{e_2}"] &
+  \semg{A} \otimes \semg{G}
+\end{tikzcd}
+
+\paragraph{$\otimes$-Elimination}
+$\semg{\letin {(a,b)} e f} : \semg{\Delta'_{1},\Delta,\Delta'_{2}} \to \semg{C}$
+defined via the diagram,
+
+\begin{tikzcd}
+  \semg{\Delta'_1,\Delta,\Delta'_2} \arrow[r , "m_{(\Delta'_1,\Delta),\Delta'_2}"] &
+  \semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}
+    \arrow[r , "m_{\Delta'_1,\Delta} \otimes id"] &
+  \left(\semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2}
+    \arrow[d , "(id \otimes \semg{e}) \otimes id"] \\
+  \semg{c} & \arrow[l , "\semg{f}"]
+  \left( \left( \semg{\Delta'_1} \otimes \semg{A} \right) \otimes \semg{B}\right) \otimes \semg{\Delta'_2}
+  & \left(\semg{\Delta'_1} \otimes \left( \semg{A} \otimes \semg{B}\right) \right) \otimes \semg{\Delta'_2}
+    \arrow[l , "\alpha \otimes id"]
+\end{tikzcd}
+
+\paragraph{$\otimes\beta$}
+
+\paragraph{$\otimes\eta$}
+
+% linear function
+In the semantics $\semg{- \otimes A}$ and $\semg{A \lto -}$ are left adjoint, so
+for all $A$ and $B$ we have the natural isomorphism
+$\phi : Hom(\semg{\Delta} \otimes \semg{A}, \semg{B}) \to Hom(\semg{\Delta}, \semg{A \lto B})$.
+We use this adjunction to provide the denotation $\semg{\lamblto a e} : \semg{\Delta} \to \semg{A \lto B}$.
+
+\[
+\semg{\lamblto a e} = \phi\left( \semg{e} \right)
+\]
+
+The denotation for application of a $\lto$-function,
+$\semg{\applto {e'} {e}} : \semg{\Delta,\Delta'} \to \semg{B}$ defined by the diagram
+
+\begin{tikzcd}
+  \semg{\Delta,\Delta'} \arrow[r , "m_{\Delta,\Delta'}"] &
+  \semg{\Delta} \otimes \semg{\Delta'} \arrow[r , "id \otimes \semg{e}"] &
+  \semg{\Delta} \otimes \semg{A} \arrow[r , "\phi^{-1}\left( \semg{e'} \right)"]
+  &
+  \semg{B}
+\end{tikzcd}
+
+$\semg{\lambtol a e} : \semg{\Delta} \to \semg{B \tol A}$ defined using the
+natural isomorphism
+$\psi : Hom(\semg{A} \otimes \semg{\Delta}, \semg{B}) \to Hom(\semg{\Delta}, \semg{B \tol A})$
+acting on the following diagram
+
+\begin{tikzcd}
+  \semg{A} \otimes \semg{\Delta} \arrow[r , "\lambda^{-1} \otimes id"] &
+  \semg{a : A} \otimes \semg{\Delta} \arrow[r , "m^{-1}_{a : A , \Delta}"] &
+  \semg{a : A , \Delta} \arrow[r , "\semg{e}"] &
+  \semg{B}
+\end{tikzcd}
+
+The application of a $\tol$-function,
+$\semg{\apptol {e} {e'}} : \semg{\Delta',\Delta} \to \semg{B}$ defined by the diagram
+
+\begin{tikzcd}
+  \semg{\Delta',\Delta} \arrow[r , "m_{\Delta',\Delta}"] &
+  \semg{\Delta'} \otimes \semg{\Delta} \arrow[r , "\semg{e'} \otimes id"] &
+  \semg{A} \otimes \semg{\Delta} \arrow[r , "\phi^{-1}\left( \semg{e} \right)"] &
+  \semg{B}
+\end{tikzcd}
+
+% Products
+Product introduction,
+$\semg{\dlamb x e} : \semg{\Delta} \to \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)}$ is
+defined by the universal property of the product
+\[
+\semg{\dlamb x e} = \left( \sem{e}(\gamma , x) \right)_{(x : X)}
+\]
+
+Product elimination, $\semg{e.\pi~M} : \semg{\Delta} \to \sem{A}(\gamma , M)$
+
+\begin{tikzcd}
+  \semg{\Delta} \arrow[r , "\semg{e}"] &
+  \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)} \arrow[r , "\pi_M"] &
+  \sem{A}(\gamma , M)
+\end{tikzcd}
+
+% Sums
+Sum introduction,
+$\semg{\sigma~M~e} : \semg{\Delta} \to \LinSigTyLimit {x} {X} \sem{A}(\gamma, x)$
+
+\begin{tikzcd}
+  \semg{\Delta} \arrow[r , "\semg{e}"] &
+  \sem{A}(\gamma, M) \arrow[r , "i_M"] &
+  \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)}
+\end{tikzcd}
+
+Sum elimination,
+$\semg{\letin {\sigma~x~a} {e} {e'}} : \semg{\Delta'_{1},\Delta,\Delta'_{2}} \to \semg{C}$
+
+% https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAB124YBbAc2CcAIjAY46AcgD6ARlLDR4+exFjJUgEwBfEFtLpMufIRRzK1es1aIOXXgIVrpcx+K0ACThDw94nu-yCKorq2rr6IBjYeAREcjJUtIwsbJwMMABmOAAU-tyBruoyHl4+fpz5DsFqJewATlh8ABY4AJT+3li+cHn2Qari0mF6BtHGcaQaiZYpNpwAMlhgAMqNACoAnos8uO7AAB4ewAAaRxW8-SHOpHTuiO4AgsoDoVrZnHx0PDy3pO77rR0I0ihhiJhIk2myWstkWK3WWy6uwOR1OezSmRyvR4lycsjeHy+P3cfwBHTKcHYYHcNNp5xxDwJ7E+31+-3anAazTa5K65QCuMGmiZLOJpMB4VGRliKDICQs0NS7HSWVy9KqL2ctU63X8cNWfE222Rhz2aKC+WAjPezKJbIBtS5LQ57B1-MqgtekpBYxlyDkAGYoVYlZaAMIiu3i71RaXgshBhUhuYBAQR3SJGBQPjwIigDJ1CA8JBkEA4CBIORJZMgHhSYA2zWyZ4hVotvFhagMOgAI1EAAVQeMbCqcN6C0XK9Ry0gNEnZrX64VroUtKU+T0sFAQF3ewOhzKQKPx4Xi4g52WK4gAKzzmHZLe83Xq4AwLQut2b7e7vsMQe+kwj0xE9J0QANpyvAA2O82GQbFXwkSNWRJdl3Aoetsn2O53GOCUf33AC2GPYEJzPAAWCCkHA6sFx4AA9YAAFpigw5dmzoRAHjbNjOyPPc-wPQDiIiUikFvS8kAomiYR3Pjf3-OMiOAkjTyQaCJMQAB2LQKC0IA
+\begin{tikzcd}[column sep =large]
+{\semg{\Delta'_1,\Delta,\Delta'_2}} \arrow[r, "{m_{(\Delta'_1,\Delta),\Delta'_2}}"]                                                                                                           & {\semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}} \arrow[d, "{m_{\Delta'_1,\Delta}\otimes id}"]                             \\
+{\left( \semg{\Delta'_1} \otimes \LinSigTyLimit {x} {X} {\sem{A}(\gamma , x)} \right) \otimes \semg{\Delta'_2}} \arrow[d, ""]                                                                 & \left( \semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2} \arrow[l, "(id \otimes \semg{e}) \otimes id"] \\
+{\LinSigTyLimit {x} {X} {\left( \sem{\Delta'_1}(\gamma , x) \otimes
+      \sem{A}(\gamma , x) \right) \otimes \sem{\Delta'_2}(\gamma , x)}} \arrow[r, "{m^{-1}_{(\Delta'_1,a:A),\Delta'_2}}"] & {\LinSigTyLimit {x} {X} {\sem{\Delta'_1,a : A,\Delta'_2}(\gamma , x)}} \arrow[d, "{[ \sem{e'}(\gamma , x) ]_{(x : X)}}"]     \\
+\semg{C}                                                                                                                                                                                      & {\sem{C}(\gamma, x)} \arrow[l]
+\end{tikzcd}
+
+%equalizer
+Equalizer introduction,
+$\semg{\equalizerin{e}} : \semg{\Delta} \to \semg{\equalizer e f g}$ where
+$\semg{e} : \semg{\Delta} \to \semg{A}$ and
+$\semg{f} \circ \semg{e} = \semg{g} \circ \semg{e}$. By the universal property
+of the equalizer the preceding equality induces a unique morphism
+$\semg{\Delta} \to Eq(\semg{f} , \semg{g}) = \semg{\equalizer e f g}$. Define
+$\semg{\equalizerin{e}}$ to be this map.
+
+Equalizer elimination, $\semg{\equalizerpi e} : \semg{\Delta} \to \semg{A}$
+
+\[
+  \semg{\equalizerpi e} = \pi_{eq} \circ \semg{e}
+\]
 

### 2024-11-14 paper/paper.tex
appendix wip

@@
   & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ \rho^{-1} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
-  & = \semg{e'}
+  & = \semg{e'} \tag{coherence}
 \end{align*}
@@
 \end{align*}
-so without loss of generality we may take may take $e$ to be variable $a'$.
+so without loss of generality we may take may take $e$ to be variable $a'$. We
+will additionally use this style of argumentation when necessary throughout this section.
 
@@
   & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \lambda) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
-  & = \semg{e'}
+  & = \semg{e'} \tag{coherence}
 \end{align*}
@@
   \semg{e'[a'/a]}
-  &= \semg{e'} \circ \semg{a'/a} \\
-  &= \semg{e'}
+  &= \semg{e'} \circ \semg{a'/a} \tag{\cref{lem:subst}} \\
+  &= \semg{e'} \tag{coherence}
 \end{align*}
-Again by the coherence theorem, $\semg{a'a} = id$. Thus, $\eta$ law for $I$
+Again by the coherence theorem, $\semg{a'/a} = id$. Thus, $\eta$ law for $I$
 holds in the denotational semantics.
@@
 
+\begin{center}
 \begin{tikzcd}
@@
 \end{tikzcd}
+\end{center}
 
@@
 \end{tikzcd}
+\end{center}
 
 \paragraph{$\otimes\beta$}
+The desired $\beta$ equality for $\otimes$ is,
+\[
+  \semg{\letin {(a,b)} {(a',b')} {f}} = \semg{f[a'/a,b'/b]}
+\]
+
+\begin{proof}
+The left hand side reduces as follows,
+\begin{align*}
+  \semg{\letin {(a,b)} {(a',b')} {f}}
+  & = \semg{f} \circ \alpha \otimes id \circ (id \otimes \semg{(a' , b')}) \otimes id \circ m_{\Delta,\Delta'} \\
+  & = \semg{f} \circ \alpha \otimes id \circ (id \otimes \lambda \otimes \lambda \circ m_{a : A, b : B})) \otimes id \circ m_{\Delta,\Delta'} \\
+  & = \semg{f} \tag{coherence}
+\end{align*}
+
+Which is equal to the right hand side,
+\begin{align*}
+  \semg{f[a'/a,b'/b]}
+  & = \semg{f} \circ \semg{a'/a,b'/b} \\
+  & = \semg{f} \tag{coherence}
+\end{align*}
+\end{proof}
 
 \paragraph{$\otimes\eta$}
-
+The desired $\eta$ equality for $\otimes$ is,
+\[
+  \semg{\letin {(a,b)} {c'} {f[(a,b)/c]}} = \semg{f[c'/c]}
+\]
+
+\begin{proof}
+\begin{align*}
+  \semg{\letin {(a,b)} {c'} {f[(a,b)/c]}}
+  & = \semg{f[(a,b)/c]} \circ \alpha \otimes id \circ (id \otimes \semg{c'}) \otimes id \circ m_{\Delta,\Delta'} \\
+  & = \semg{f} \circ \semg{(a,b)/c} \circ \alpha \otimes id \circ (id \otimes \semg{c'}) \otimes id \circ m_{\Delta,\Delta'} \\
+  & = \semg{f} \tag{coherence}
+\end{align*}
+
+\begin{align*}
+  \semg{f[c'/c]}
+  &= \semg{f} \circ \semg{c'/c} \tag{\cref{lem:subst}} \\
+  &= \semg{f} \tag{coherence}
+\end{align*}
+\end{proof}
+
+\subsubsection{$\lto$-Functions}
 % linear function
-In the semantics $\semg{- \otimes A}$ and $\semg{A \lto -}$ are left adjoint, so
-for all $A$ and $B$ we have the natural isomorphism
-$\phi : Hom(\semg{\Delta} \otimes \semg{A}, \semg{B}) \to Hom(\semg{\Delta}, \semg{A \lto B})$.
-We use this adjunction to provide the denotation $\semg{\lamblto a e} : \semg{\Delta} \to \semg{A \lto B}$.
+\paragraph{$\lto$-Introduction}
+$\semg{\lamblto a e} : \semg{\Delta} \to \semg{A \lto B}$ is defined using the
+natural isomorphism
+$\phi : Hom(\semg{\Delta} \otimes \semg{A}, \semg{B}) \to Hom(\semg{\Delta}, \semg{A \lto B})$
+that is provided by the adjunction between $\semg{- \otimes A}$ and $\semg{A \lto -}$.
 
@@
 
-The denotation for application of a $\lto$-function,
-$\semg{\applto {e'} {e}} : \semg{\Delta,\Delta'} \to \semg{B}$ defined by the diagram
-
+\paragraph{$\lto$-Elimination}
+$\semg{\applto {e'} {e}} : \semg{\Delta,\Delta'} \to \semg{B}$ is defined by the
+diagram,
+
+\begin{center}
 \begin{tikzcd}
@@
 \end{tikzcd}
-
-$\semg{\lambtol a e} : \semg{\Delta} \to \semg{B \tol A}$ defined using the
+\end{center}
+
+\paragraph{$\lto\beta$}
+The $\beta$ rule for $\lto$ is given by,
+\[
+  \semg{\applto {(\lamblto {a} {e})} {a'}} = \semg{e[a'/a]}
+\]
+\begin{proof}
+\begin{align*}
+  \semg{\applto {(\lamblto {a} {e})} {a'}}
+  & = \phi^{-1}(\semg{\lamblto {a} {e}}) \circ id \otimes \semg{a'} \circ m_{\Delta,\Delta'} \\
+  & = \phi^{-1}(\phi(\semg{e})) \circ id \otimes \semg{a'} \circ m_{\Delta,\Delta'} \\
+  & = \semg{e} \circ id \otimes \semg{a'} \circ m_{\Delta,\Delta'} \\
+  & = \semg{e} \circ id \otimes \lambda \circ m_{\Delta,\Delta'} \\
+  & = \semg{e} \tag{coherence}
+\end{align*}
+
+\begin{align*}
+  \semg{e[a'/a]}
+  & = \semg{e} \circ \semg{a'/a} \tag{\cref{lem:subst}} \\
+  & = \semg{e} \tag{coherence}
+\end{align*}
+\end{proof}
+
+\paragraph{$\lto\eta$}
+The $\eta$ rule for $\lto$ is given by,
+\[
+  \semg{\lamblto {a} {\applto e a}} = \semg{e}
+\]
+
+\begin{proof}
+\begin{align*}
+  \semg{\lamblto {a} {\applto e a}}
+  & = \phi(\semg{\applto e a}) \\
+  & = \phi(\phi^{-1}(\semg{e}) \circ id \otimes \semg{a} \circ m_{\Delta,a:A}) \\
+  & = \phi(\phi^{-1}(\semg{e}) \circ id \otimes \lambda \circ m_{\Delta,a:A}) \\
+  & = \phi(\phi^{-1}(\semg{e})) \tag{coherence} \\
+  & = \semg{e}
+\end{align*}
+\end{proof}
+
+\subsubsection{$\tol$-Functions}
+\paragraph{$\tol$-Introduction}
+Just as with the other linear function type, we have an adjunction between
+$\semg{A \otimes -}$ and $\semg{- \tol A}$. $\semg{\lambtol a e} : \semg{\Delta} \to \semg{B \tol A}$ is defined using the
 natural isomorphism
 $\psi : Hom(\semg{A} \otimes \semg{\Delta}, \semg{B}) \to Hom(\semg{\Delta}, \semg{B \tol A})$
+induced by this adjunction. In particular, $\semg{\lambtol a e}$ is given by $\psi$
 acting on the following diagram
 
+\begin{center}
 \begin{tikzcd}
@@
 \end{tikzcd}
-
+\end{center}
+
+\paragraph{$\tol$-Elimination}
 The application of a $\tol$-function,
@@
 \end{tikzcd}
+\end{center}
+
+\paragraph{$\tol\beta$}
+
+\paragraph{$\tol\eta$}
 

### 2024-11-14 paper/paper.tex
denotations minus sum types

@@
   \semg{\Delta'} \otimes \semg{\Delta} \arrow[r , "\semg{e'} \otimes id"] &
-  \semg{A} \otimes \semg{\Delta} \arrow[r , "\phi^{-1}\left( \semg{e} \right)"] &
+  \semg{A} \otimes \semg{\Delta} \arrow[r , "\psi^{-1}\left( \semg{e} \right)"] &
   \semg{B}
@@
 \paragraph{$\tol\beta$}
+The $\beta$ rule for $\tol$ is given by,
+\[
+  \semg{\apptol {(\lambtol {a} {e})} {a'}} = \semg{e[a'/a]}
+\]
+\begin{proof}
+\begin{align*}
+  \semg{\apptol {(\lambtol {a} {e})} {a'}}
+  & = \psi^{{-1}}(\semg{\lambtol {a} {e}}) \circ \semg{a'} \otimes id \circ m_{a:A,\Delta} \\
+  & = \psi^{{-1}}(\semg{\lambtol {a} {e}}) \circ \lambda \otimes id \circ m_{a:A,\Delta} \\
+  & = \psi^{{-1}}(\psi(\semg{e} \circ m^{-1}_{a:A,\Delta} \circ \lambda^{-1} \otimes id)) \circ \lambda \otimes id \circ m_{\Delta',\Delta} \\
+  & = \semg{e} \circ m^{-1}_{a:A,\Delta} \circ \lambda^{-1} \otimes id \circ \lambda \otimes id \circ m_{a:A,\Delta} \\
+  & = \semg{e}
+\end{align*}
+
+\begin{align*}
+  \semg{e[a'/a]}
+  & = \semg{e} \circ \semg{a'/a} \tag{\cref{lem:subst}} \\
+  & = \semg{e} \tag{coherence}
+\end{align*}
+\end{proof}
 
 \paragraph{$\tol\eta$}
-
+The $\eta$ rule for $\tol$ is given by,
+\[
+  \semg{\lambtol {a} {\apptol e a}} = \semg{e}
+\]
+\begin{proof}
+\begin{align*}
+  \semg{\lambtol {a} {\apptol e a}}
+  & = \phi(\semg{\apptol e a} \circ m^{-1}_{a:A,\Delta} \circ \lambda^{-1} \otimes id) \\
+  & = \phi(\phi^{-1}(\semg{e}) \circ \semg{a} \otimes id \circ m_{a:A,\Delta} \circ m^{-1}_{a:A,\Delta} \circ \lambda^{-1} \otimes id) \\
+  & = \phi(\phi^{-1}(\semg{e}) \circ \lambda \otimes id \circ m_{a:A,\Delta} \circ m^{-1}_{a:A,\Delta} \circ \lambda^{-1} \otimes id) \\
+  & = \phi(\phi^{-1}(\semg{e})) \\
+  & = \semg{e}
+\end{align*}
+\end{proof}
+
+\subsubsection{$\bigamp$-Products}
 % Products
-Product introduction,
+\paragraph{$\bigamp$-Introduction}
 $\semg{\dlamb x e} : \semg{\Delta} \to \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)}$ is
@@
 
-Product elimination, $\semg{e.\pi~M} : \semg{\Delta} \to \sem{A}(\gamma , M)$
-
+\paragraph{$\bigamp$-Elimination}
+$\semg{e.\pi~M} : \semg{\Delta} \to \sem{A}(\gamma , M)$ is defined using the
+projection out of the product,
+
+\begin{center}
 \begin{tikzcd}
@@
 \end{tikzcd}
-
+\end{center}
+
+\paragraph{$\bigamp\beta$}
+The $\beta$ law for $\bigamp$ is given by,
+\[
+  \sem{(\dlamb x e).\pi~M}(\gamma, x) = \sem{e[M/x]}(\gamma, x)
+\]
+\begin{proof}
+\begin{align*}
+  \semg{(\dlamb x e).\pi~M}
+  & = \pi_{M} \circ \semg{\dlamb x e} \\
+  & = \pi_{M} \circ (\sem{e}(\gamma , x))_{(x : X)} \\
+  & = \sem{e}(\gamma , M)
+\end{align*}
+by the universal property of the product.
+
+\begin{align*}
+  \sem{e[M/x]}(\gamma , x)
+  & = \sem{e}(\gamma , x) \circ  \sem{M/x}(\gamma , x) \\
+  & = \sem{e}(\gamma , M)
+\end{align*}
+\end{proof}
+
+\paragraph{$\bigamp\eta$}
+The $\eta$ law for $\bigamp$ is given by,
+\[
+  \semg{(\dlamb x {e.\pi~x})} = \semg{e}
+\]
+\begin{proof}
+  \semg{(\dlamb x {e.\pi~x})}
+  & = (\semg{e.\pi~x})_{x : X} \\
+  & = (\pi_x \circ \semg{e})_{x : X} \\
+  & = \semg{e}
+\end{proof}
+by the universal property of the product.
+
+\subsubsection{$\bigoplus$-Sums}
 % Sums
-Sum introduction,
+\paragraph{$\bigoplus$-Introduction}
 $\semg{\sigma~M~e} : \semg{\Delta} \to \LinSigTyLimit {x} {X} \sem{A}(\gamma, x)$
 
+\begin{center}
 \begin{tikzcd}
@@
 \end{tikzcd}
-
-Sum elimination,
+\end{center}
+
+\paragraph{$\bigoplus$-Elimination}
 $\semg{\letin {\sigma~x~a} {e} {e'}} : \semg{\Delta'_{1},\Delta,\Delta'_{2}} \to \semg{C}$
-
+is defined in the diagram
+
+\begin{center}
 % https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAB124YBbAc2CcAIjAY46AcgD6ARlLDR4+exFjJUgEwBfEFtLpMufIRRzK1es1aIOXXgIVrpcx+K0ACThDw94nu-yCKorq2rr6IBjYeAREcjJUtIwsbJwMMABmOAAU-tyBruoyHl4+fpz5DsFqJewATlh8ABY4AJT+3li+cHn2Qari0mF6BtHGcaQaiZYpNpwAMlhgAMqNACoAnos8uO7AAB4ewAAaRxW8-SHOpHTuiO4AgsoDoVrZnHx0PDy3pO77rR0I0ihhiJhIk2myWstkWK3WWy6uwOR1OezSmRyvR4lycsjeHy+P3cfwBHTKcHYYHcNNp5xxDwJ7E+31+-3anAazTa5K65QCuMGmiZLOJpMB4VGRliKDICQs0NS7HSWVy9KqL2ctU63X8cNWfE222Rhz2aKC+WAjPezKJbIBtS5LQ57B1-MqgtekpBYxlyDkAGYoVYlZaAMIiu3i71RaXgshBhUhuYBAQR3SJGBQPjwIigDJ1CA8JBkEA4CBIORJZMgHhSYA2zWyZ4hVotvFhagMOgAI1EAAVQeMbCqcN6C0XK9Ry0gNEnZrX64VroUtKU+T0sFAQF3ewOhzKQKPx4Xi4g52WK4gAKzzmHZLe83Xq4AwLQut2b7e7vsMQe+kwj0xE9J0QANpyvAA2O82GQbFXwkSNWRJdl3Aoetsn2O53GOCUf33AC2GPYEJzPAAWCCkHA6sFx4AA9YAAFpigw5dmzoRAHjbNjOyPPc-wPQDiIiUikFvS8kAomiYR3Pjf3-OMiOAkjTyQaCJMQAB2LQKC0IA
@@
 {\semg{\Delta'_1,\Delta,\Delta'_2}} \arrow[r, "{m_{(\Delta'_1,\Delta),\Delta'_2}}"]                                                                                                           & {\semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}} \arrow[d, "{m_{\Delta'_1,\Delta}\otimes id}"]                             \\
-{\left( \semg{\Delta'_1} \otimes \LinSigTyLimit {x} {X} {\sem{A}(\gamma , x)} \right) \otimes \semg{\Delta'_2}} \arrow[d, ""]                                                                 & \left( \semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2} \arrow[l, "(id \otimes \semg{e}) \otimes id"] \\
+{\left( \semg{\Delta'_1} \otimes \LinSigTyLimit {x} {X} {\sem{A}(\gamma , x)} \right) \otimes \semg{\Delta'_2}} \arrow[d, "d"]                                                                 & \left( \semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2} \arrow[l, "(id \otimes \semg{e}) \otimes id"] \\
 {\LinSigTyLimit {x} {X} {\left( \sem{\Delta'_1}(\gamma , x) \otimes
@@
 \end{tikzcd}
-
+\end{center}
+
+where $d$ is the distributivity morphism, and the last morphism implicitly
+weakens $\sem{C}$.
+
+\paragraph{$\bigoplus\beta$}
+The $\beta$ rule for $\bigoplus$ is given by,
+\[
+  \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}} = \semg{e'[M/x,a'/a]}
+\]
+\begin{proof}
+  \begin{align*}
+  \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}}
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
+  \otimes \semg{\sigma~M~a'}) \otimes id \circ m \otimes id \circ m \\
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
+  \otimes i_M \circ \semg{a'}) \otimes id \circ m \otimes id \circ m \\
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
+  \otimes i_M \circ \lambda) \otimes id \circ m \otimes id \circ m
+  \end{align*}
+\steven{TODO need to track subscripts on $m$ in the $\bigoplus\beta$ proof, and
+  finish this part of the above part of the proof}
+
+  \begin{align*}
+    \semg{e'[M/x,a'/a]}
+    &= \semg{e'} \circ \semg{M/x,a'/a} \tag{\cref{lem:subst}} \\
+    &= \semg{e'}
+  \end{align*}
+  \steven{TODO shouldn't the semantics of $e'$ be in context $\gamma,x$ not $x$?}
+\end{proof}
+
+\paragraph{$\bigoplus\eta$}
+\steven{TODO $\eta$ for sums}
+
+\subsubsection{Equalizer}
 %equalizer
-Equalizer introduction,
+\paragraph{Equalizer Introduction}
 $\semg{\equalizerin{e}} : \semg{\Delta} \to \semg{\equalizer e f g}$ where
@@
 
-Equalizer elimination, $\semg{\equalizerpi e} : \semg{\Delta} \to \semg{A}$
+\paragraph{Equalizer Elimination}
+$\semg{\equalizerpi e} : \semg{\Delta} \to \semg{A}$ is defined using the map
+$\pi_{eq}$ from $Eq(\semg{f} , \semg{g})$ to the domain of $f$ and $g$.
 
@@
 
+\paragraph{Equalizer $\beta$}
+The $\beta$ rule for $\equalizer e f g$ is given as,
+\[
+\semg{\equalizerpi {\equalizerin e}} = \semg{e}
+\]
+where $\semg{f} \circ \semg{e} = \semg{g} \circ \semg{e}$.
+\begin{proof}
+  In $\semcat$ the universal property of $Eq(\semg{f}, \semg{g})$ implies that
+  the following diagram commutes, implying the $\beta$ rule.
+
+% https://q.uiver.app/#q=WzAsNCxbMSwwLCJcXHNlbWd7QX0iXSxbMiwwLCJcXHNlbWd7Qn0iXSxbMCwwLCJFcShcXHNlbWd7Zn0sXFxzZW1ne2d9KSJdLFswLDEsIlxcc2VtZ3tcXERlbHRhfSJdLFswLDEsIlxcc2VtZ3tmfSIsMCx7Im9mZnNldCI6LTF9XSxbMCwxLCJcXHNlbWd7Z30iLDEseyJvZmZzZXQiOjF9XSxbMiwwLCJcXHBpX3tlcX0iXSxbMywwLCJcXHNlbWd7ZX0iLDJdLFszLDIsIlxcc2VtZ3tcXGVxdWFsaXplcmluIGV9Il1d
+\begin{center}
+\begin{tikzcd}
+	{Eq(\semg{f},\semg{g})} & {\semg{A}} & {\semg{B}} \\
+	{\semg{\Delta}}
+	\arrow["{\pi_{eq}}", from=1-1, to=1-2]
+	\arrow["{\semg{f}}", shift left, from=1-2, to=1-3]
+	\arrow["{\semg{g}}", shift right, swap, from=1-2, to=1-3]
+	\arrow["{\semg{\equalizerin e}}", from=2-1, to=1-1]
+	\arrow["{\semg{e}}"', from=2-1, to=1-2]
+\end{tikzcd}
+\end{center}
+\end{proof}
+
+\paragraph{Equalizer $\eta$}
+
+The $\eta$ rule for $\equalizer e f g$ is given as,
+\[
+\semg{\equalizerin {\equalizerpi e}} = \semg{e}
+\]
+\begin{proof}
+  Likewise, the universal property of $Eq(\semg{f}, \semg{g})$ implies $\eta$
+  rule via this diagram.
+
+\begin{center}
+\begin{tikzcd}
+	{Eq(\semg{f},\semg{g})} & {\semg{A}} & {\semg{B}} \\
+	{\semg{\Delta}}
+	\arrow["{\pi_{eq}}", from=1-1, to=1-2]
+	\arrow["{\semg{f}}", shift left, from=1-2, to=1-3]
+	\arrow["{\semg{g}}", shift right, swap, from=1-2, to=1-3]
+	\arrow["{\semg{e}}", from=2-1, to=1-1]
+	\arrow["{\semg{\equalizerpi e}}"', from=2-1, to=1-2]
+\end{tikzcd}
+\begin{center}
+\end{proof}
+
 \end{document}

### 2024-11-14 paper/paper.tex
better scaffold discussion + minor edits

@@
 Prior approaches to verified parsing focus on verification of a
-particular grammar formalism such as \max{Steven write down what the
-  related work uses
-  here}\cite{lasserCoStarVerifiedALL2021,firsovCertifiedNormalizationContextFree2015,danielssonTotalParserCombinators2010}.
+particular grammar formalism such as non-left-recursive grammars or LL(1) grammars \cite{lasserCoStarVerifiedALL2021,EdelmannZippy2020,danielssonTotalParserCombinators2010}.
+\steven{These have restrictions on left recursion or LL(1)-ness. Is it the most accurate
+  to call that a difference in formalism? It would seem that these are all CFGs
+  + conditions. Maybe these aren't the best works to
+  compare to if they're verifying particular algorithms}
 Each new grammar formalism must then be extended with its own verified
@@
 
-\max{Steven, is this accurate comparison?} The design of \theoryabbv
+The design of \theoryabbv
 is an extension of Joachim Lambek's \emph{categorial grammars}
@@
 terminology would be considered a kind of \emph{non-commutative linear
-logic}, a version of linear logic where the tensor product is not
+logic} --- a version of linear logic where the tensor product is not
 commutative, reflecting the obvious property that the relative
@@
 a linear $\lambda$-term that implements the program given in the
-derivation.\max{reverse the order of these?}
+derivation.\max{reverse the order of these?} \steven{the order of what?}
 
@@
 
-\steven{TODO fig 3 sizing}
 \begin{figure}
@@
 are strongly equivalent, as the equation $\lambda b. f(g(b)) \equiv
-\lambda b. b$ follows because $\B$ is unambiguous. Therefore also
-$\String$ is strongly equivalent to $\top$. Next, since $\String$ is
+\lambda b. b$ follows because $\B$ is unambiguous. Therefore
+$\String$ is also strongly equivalent to $\top$. Next, since $\String$ is
 defined as a Kleene star, we can easily show that $\String \cong \I
@@
 with an associative operation. Here we take the alphabet to be $\{
-\stringquote{(}, \stringquote{)}, \stringquote{+} \stringquote{NUM}
+\stringquote{(}, \stringquote{)}, \stringquote{+} , \stringquote{NUM}
 \}$. In \Cref{fig:binop-inductive} we define it using two mutually
@@
 right paren, then we proceed to $C$, otherwise we proceed to $M$. Here
-$\mathsf{NotStartsWithRP}$ is defined as $\I \oplus (('(' \oplus '+'
+$\mathsf{NotStartsWithRP}$ is defined as
+$\I \oplus (({\literal{(}} \oplus {\literal {+}}
 \oplus \mathsf{NUM}) \otimes \top)$. In the ``closing'' state $C$ if
@@
   We construct a parser for $\mathsf{Exp}$ by showing it is weakly
-  equivalent to $\mathsf{O}~0~\true$ 
+  equivalent to $\mathsf{O}~0~\true$.
 \end{theorem}
@@
 \subsection{Semantics}
+\steven{Add definitions of kleene,lambek,and chomsky categories here?}
 
@@
 \label{sec:discussion}
-\paragraph{Lambek Calculus and Categorial Grammar}
+\subsection{Typed Approaches to Grammars}
+\subsubsection{Lambek Calculus and Categorial Grammar}
+
+\steven{Note that lambek calculus and its extensions don't have inductives, or
+  even kleene star as far as I can tell. }
+
+\steven{Choppy notes about lambek calc:
 In 1958 \cite{lambek1958mathematics}, Joachim Lambek introduced his syntactic
@@
 relevant presentation of Lambek's original system.
-
-\steven{Should discussion of this connection to Lambek calculus be spread
-  throughout the paper?}
 
@@
 
-Also mention here that \cite{luoLambekCalculusDependent} seems similar at first
+Also mention here that Luo's work \cite{luoLambekCalculusDependent} seems similar at first
 glance, but isn't as similar as it might look because the judgments and
@@
 understand it, it differs in x,y,z ways''
-
-\steven{TODO Edit the above}
-\steven{Even though the above sounds bad, are there Lambek/categorial things that are missed?}
+}
+
+\subsubsection{Other typed approaches}
+\steven{this subsection should be renamed}
+The usage of a simple type system to reason about regular expressions was
+introduced by Frisch and Cardelli in 2004 \cite{frischCardelli}, and later expanded
+by Hengelin and Nielsen in 2011 \cite{henglein_regular_2011} to handle proofs of regular expression
+containment.
+
+\steven{Mention Neel's typed algebraic approach paper here. Say it is similar at
+  first glance, as its parsing and it has types, but that he actually is representing
+  a system for proving the LL(1) condition }
 
@@
 
-\paragraph{Typed Approaches to Regular Expressions}
-The usage of a simple type system to reason about regular expressions was
-introduced by Frisch and Cardelli in 2004 \cite{frischCardelli}, and later expanded
-by Hengelin and Nielsen in 2011 \cite{henglein_regular_2011} to handle proofs of regular expression
-containment. In 1992, while investigating the addition of arbitrary recursion to Kleene
-algebra \cite{leiss},
-Lei{\ss} used a least fixed-point operator instead of the Kleene star. Although
-he did not explicitly make use of a type system, or have access to the view of Frisch and
-Cardelli, Lei{\ss} had added the full power of inductive types to a type system
-of regular expressions.
-
-Pratt's work on the residuation of action algebras in \cite{prattActionLogicPure1991} closely mirrors
-our observation that Brzozowski derivatives mirror linear function types.
-Moreover, he shows that his residuals form a Galois connection with sequential
-composition. This is precisely a posetal --- or \emph{thin} --- version of the
-adjunction between
-linear function types and tensor product in \theoryabbv.
-
-Our primary observation in this paper can be summarized as saying that all of
-these above works can embed naturally into a rich, unifying theory.
-
-\steven{I'd like to synthesize the discussion of these works with the categorial
-grammar works, but I am not sure how. It seems as though there are these
-parallel tracks of research that have missed each other for decades}
-
-\paragraph{Inductive Linear Types}
-
-% In \cite{firsovCertifiedNormalizationContextFree2015}, Firsov and Uustalu
-% constructed a verified normalization procedure to turn a
-% context-free grammar into an equivalent grammar in Chomsky normal form.
-% Combining their work with a CYK parser, they have built a verified context-free
-% grammar parser, \emph{up to weak equivalence}. To upgrade from a weak
-% equivalence to a strong equivalence, they suggest updating their grammar
-% formalism to treat parse trees as first-class proofs of language membership.
-% That is, without explicitly stating so, they suggest that they next natural step
-% is to reason about grammars in a system where parse trees are terms, as we have
-% given in this paper.
-
-% \steven{Make the above paragraph shorter. Also idk if its worth including, but I
-% like that they practically ask for a type theory}
-
-\steven{Mention Neel's typed algebraic approach paper here. Say it is similar at
-  first glance, as its parsing with types, but that he actually is representing
-  something quite different. Or is it so different that its not worth mentioning?}
-
-\steven{TODO add Lambek calculus/categorial grammars here}
-
-\paragraph{Kleene Algebra}
+\subsubsection{Kleene Algebra}
 Since the early works in the theory of formal languages, Kleene
@@
 
-\steven{Can we sketch here what an apt categorification of KATs might be? As a
-  poor first pass: perhaps a Kleene category with a full involutive
-  subcategory that corresponds to the Boolean algebra?.}
-\steven{What can we say about internalizing Kleene algebra proofs or extensions?}
-\steven{any KA citations needed above?}
+\subsection{Semantic Actions}
+
+Our verification has mainly focused on the verification that a parser
+outputs a correct concrete syntax tree for a grammar. However, in
+practice, parsers are combined with a \emph{semantic action} that
+emits an \emph{abstract} syntax tree that omits superflous syntactic
+details that aren't needed in later stages of the overall program. We
+can define a semantic action in \theoryabbv for a linear type $A$ with
+semantic outputs in a non-linear type $X$ to be a function $\ltonl {(A
+  \lto \bigoplus_{\_:X} \top)}$. That is, a semantic action is a
+function that produces a semantic element of $X$ from the concrete
+parses of $A$. In future work, we aim to study the question of
+verifying efficient implementations of parsers with semantic actions.
 
@@
 %%      logic?
-\paragraph{Relation to Separation Logic}
+\subsection{Relation to Separation Logic}
 
@@
 
-%% \paragraph{Linearity with Dependent Types}
-%% \steven{I don't know if it worth discussing the different influences of
-%%   categorical model here? We take Neel's model and add to it as necessary}
-
-% Vakar might be worth citing to discuss different characterizations of the
-% notion of model \cite{vakarSyntaxSemanticsLinear2015}
-
-% \cite{fu2023twolevellineardependenttype} give a compiler for a two-level
-% linear dependent type theory. This could serve as a template for a
-% compiler implementation of \theoryname that is not just inference rules
-% embedded in Agda.
-
-\paragraph{Implementations in Parsing}
-\steven{Talk about future implementation-side work}
-
-\cite{fu2023twolevellineardependenttype} give a compiler for a two-level
-linear dependent type theory. This could serve as a template for a
-compiler implementation of \theoryname that is not just inference rules
-embedded in Agda.
-
-\paragraph{Type Checking and Elaboration}
+\subsection{Implementation}
+\steven{Moving away from a prototype implementation of \theoryabbv}
+\steven{Mention performance}
+
+\subsection{Type Checking and Elaboration}
 Our focus in this work has been on the verification of parsers for
@@
 
-\subsection{Semantic Actions}
-
-Our verification has mainly focused on the verification that a parser
-outputs a correct concrete syntax tree for a grammar. However, in
-practice, parsers are combined with a \emph{semantic action} that
-emits an \emph{abstract} syntax tree that omits superflous syntactic
-details that aren't needed in later stages of the overall program. We
-can define a semantic action in \theoryabbv for a linear type $A$ with
-semantic outputs in a non-linear type $X$ to be a function $\ltonl {(A
-  \lto \bigoplus_{\_:X} \top)}$. That is, a semantic action is a
-function that produces a semantic element of $X$ from the concrete
-parses of $A$. In future work, we aim to study the question of
-verifying efficient implementations of parsers with semantic actions.
-
 \bibliographystyle{plain}
 \bibliography{refs.bib}
+
+\pagebreak
 
@@
 \end{figure}
-\steven{Do we need inverse for $\ltonl A$ in judgmental equality figure?}
-\steven{Do we need to say anything about the equalizer proofs being unique in
-  judgmental equality figure?}
+\steven{Need to include premises for the judgmental equality rules}
 
@@
 
-\begin{definition}
-  Given any $\Gamma ; \Delta \vdash e : A$ and $\delta \in
-  \Subst(\Delta',\Delta)$, we define the action of the substitution on
-  $e$ as follows, frequently using the inversion principle to split
-  the substitution into constituent components.
+\begin{figure}
   \begin{align*}
@@
   \end{align*}
+\caption{Action of substitution on linear terms}
+\label{fig:substact}
+\end{figure}
+\steven{captilizataion should be consistent over all figures. idk if all caps or
+all lowercase is beter?}
+
+\begin{definition}
+  Given any $\Gamma ; \Delta \vdash e : A$ and $\delta \in
+  \Subst(\Delta',\Delta)$, we define the action of the substitution on
+  $e$ in \cref{fig:substact}, frequently using the inversion principle to split
+  the substitution into constituent components.
   By induction on linear term and equality judgments, we establish the
@@
 \begin{proof}
+\begin{align*}
   \semg{(\dlamb x {e.\pi~x})}
@@
   & = \semg{e}
+\end{align*}
 \end{proof}
@@
 \end{tikzcd}
-\begin{center}
+\end{center}
 \end{proof}

### 2024-11-14 paper/paper.tex
addd automaton todo
two related work todos
small changes to semantics section

@@
 grammars and parse transformers, we will now define a
-\emph{denotational semantics} that makes this mathematically
-precise. To interpret our types of codes we assume a universe of
-\emph{small} sets.
+\emph{denotational semantics} that makes this mathematically precise
+by defining a notion of formal grammar and parse transformer and
+showing that our type theory can be soundly interpreted in this
+model. We then discuss how this denotational semantics provides the
+basis for our prototype implementation in Agda.
 
@@
 
-First, we need to establish a precise definition of a formal grammar
-that accounts for the possibility of ambiguity. The most common
-definition of a formal grammar is as generative grammars, defined by
-a set of non-terminals, with a specified start symbol and set of
-production rules. We instead use a more abstract formulation that is
-closer in spirit to the standard definition of a formal language:
-
+The most common definition of a formal grammar is as generative
+grammars, defined by a set of non-terminals, a specified start symbol
+and set of production rules. We instead use a more abstract
+formulation that is closer in spirit to the standard definition of a
+formal \emph{language}\cite{elliott}:
 \begin{definition}
   A \emph{formal language} $L$ is a function from strings to propositions.
-  A \emph{formal grammar} $A$ is a function from strings to small sets.
+  A (small) \emph{formal grammar} $A$ is a function from strings to (small) sets.
 \end{definition}
-
 We think of the grammar $A$ as taking a string to the set of all parse
@@
 tree, but \emph{which} parse tree it has, i.e., what the structure of
-the element of $A(w)$ is.
-
-The linear types of \theoryabbv can all be interpreted as formal
-grammars, but what about the linear terms? These can be interpreted as
-\emph{parse transformers}, which we now define.
+the element of $A(w)$ is.  To interpret our universes $U , L$ we
+assume we have a universe of \emph{small} sets. In the remainder, all
+formal grammars are assumed to be small.
+
+We then interpret linear \emph{terms} as \emph{parse transformers}:
 \begin{definition}
-  Let $L_1,L_2$ be formal languages. Then we say $L_1\subseteq L_2$ if
-  for every string $w$, $L_1(w)\Rightarrow L_2(w)$.
-
   Let $A_1,A_2$ be formal grammars. Then a parse transformer $f$ from
@@
 \end{definition}
-\pedro{Do we actually need a definition of formal language morphism?
-In case we do, it should probably be its own definition.}
-First, note that parse transformers generalize language inclusion: if
-$A_1(w), A_2(w)$ are all subsets of a one-element set, then a parse
-transformer is equivalent to showing that $A_1(w) \subseteq A_2(w)$.
-In our denotational semantics, linear terms will be interpreted as
-such parse transformers, and the notions of unambiguous grammar,
-parseability, etc, introduced in \Cref{sec:concepts} can be verified
-to correspond to their intended meanings under this interpretation.
+Just as formal grammars generalize formal languages, parse
+transformers generalize formal language inclusion: if $A_1(w), A_2(w)$
+are all subsets of a one-element set, then a parse transformer is
+equivalent to showing that $A_1(w) \subseteq A_2(w)$. In our
+denotational semantics, linear terms will be interpreted as such parse
+transformers, and the notions of unambiguous grammar, parsers,
+disjointness, etc, introduced in \Cref{sec:applications} can be
+verified to correspond to their intended meanings under this
+interpretation.
 
@@
 structures to model the linear types, terms and equalities in
-\theoryabbv.
+\theoryabbv. These categorical properties are precisely what is
+required to interpret all of the linear type and term constructors.
 %% \begin{theorem}
@@
 \subsection{Semantics}
-\steven{Add definitions of kleene,lambek,and chomsky categories here?}
-
-We can now define our denotational semantics.
+
+We now define our denotational semantics.
 \begin{definition}[Grammar Semantics]
@@
   \item If $\Gamma \vdash X \isSmall$, then $\sem X \gamma$ is a small set.
-  \item If $\Gamma \vdash X \cong X'$ then for every $\gamma$, $\sem{X}\gamma = \sem{X'}\gamma$
-  \item If $\Gamma \vdash A \cong A'$ then for every $\gamma$, $\sem{A}\gamma = \sem{A'}\gamma$
-  \item If $\Gamma \vdash M \cong M' : X$ then for every $\gamma$, $\sem{M}\gamma = \sem{M'}\gamma$
-  \item If $\Gamma;\Delta \vdash e \cong e' : A$ then for every $\gamma$, $\sem{e}\gamma = \sem{e'}\gamma$
+  \item If $\Gamma \vdash X \equiv X'$ then for every $\gamma$, $\sem{X}\gamma = \sem{X'}\gamma$
+  \item If $\Gamma \vdash A \equiv A'$ then for every $\gamma$, $\sem{A}\gamma = \sem{A'}\gamma$
+  \item If $\Gamma \vdash M \equiv M' : X$ then for every $\gamma$, $\sem{M}\gamma = \sem{M'}\gamma$
+  \item If $\Gamma;\Delta \vdash e \equiv e' : A$ then for every $\gamma$, $\sem{e}\gamma = \sem{e'}\gamma$
   \end{enumerate}
@@
 
-For reasons of brevity, we present only the cases related to the
-linear types in \Cref{fig:semantics}. The full details are in the
-appendix.\max{TODO: put the full details in the appendix}.  The
-grammar for a literal $c$ has a single parse precisely when the input
-string consists of the single character. The grammar for the unit
-similarly has a single parse for the empty string.  A parse of the
-tensor product $A \otimes B$ consists of a \emph{splitting} of the
-empty string into a prefix $w_1$ and suffix $w_2$ along with an $A$
-parse of $w_1$ and $B$ parse of $w_2$. A parse of $\bigoplus_{x:X} A$
-is a pair of an element of the set $X$ and a parse of $A(x)$, while
-dually a parse of $\bigwith_{x:X} A$ is a \emph{function} taking any
-$x:X$ to a parse of $A(x)$. A $w$-parse of $A \lto B$ is a function
-that takes an $A$ parse of some other string $w'$ to a $B$ parse of
-$ww'$, and $B \tol A$ is the same except the $B$ parse is for the
-reversed concatenation $w'w$.
+The interpretation of dependent types as sets is standard
+\cite{dependent-types-as-sets}. We present the concrete descriptions
+of the semantics of linear types, as well as our non-standard
+non-linear types in \Cref{fig:semantics}.
+The grammar for a literal $c$ has a single parse
+precisely when the input string consists of the single character. The
+grammar for the unit similarly has a single parse for the empty
+string.  A parse of the tensor product $A \otimes B$ consists of a
+\emph{splitting} of the empty string into a prefix $w_1$ and suffix
+$w_2$ along with an $A$ parse of $w_1$ and $B$ parse of $w_2$. A parse
+of $\bigoplus_{x:X} A$ is a pair of an element of the set $X$ and a
+parse of $A(x)$, while dually a parse of $\bigwith_{x:X} A$ is a
+\emph{function} taking any $x:X$ to a parse of $A(x)$. A $w$-parse of
+$A \lto B$ is a function that takes an $A$ parse of some other string
+$w'$ to a $B$ parse of $ww'$, and $B \tol A$ is the same except the
+$B$ parse is for the reversed concatenation $w'w$.
 %
@@
 $\sem{\uparrow(A \lto B)}$, i.e., parse transformers.
-
+%
 The universe $L$ of linear types is interpreted as the set of all
 (small) grammars.
-
 \begin{figure}
@@
   from $I$ to $O$ consists of a set of shapes $S$, a set of positions
-  $P$ and functions $I \leftarrow P \to S \to O$. The \emph{extension}
+  $P$ and functions $f : P \to I$, $g : P \to S$ and $h : S \to O$. The \emph{extension}
   of a polynomial is a functor $\Set/I \to \Set/O$ defined as the
@@
   are, respectively, the right and left adjoint of their pullback functors
-  $g^*$ and $h^*$.
-
-  A dependent polynomial functor from $I$ to $O$ is a functor $\Set/I
+  $g^*$ and $h^*$. A dependent polynomial functor from $I$ to $O$ is a functor $\Set/I
   \to \Set/O$ that is nautrally isomorphic to the extension of a
@@
 \end{definition}
-With this interpretation, $\el(F)$ and $\map(F)$ are just interpreted
-as the action of the functor on objects and morphisms,
-respectively. We interpret the constructors $\mathsf{K},\Var$, etc. on
-functors in the obvious way that matches the definitional behavior of
-$\el$ and $\map$. The non-trivial part of the construction is
-verifying that such functors are constructing corresponding dependent
-polynomials. The details are tedious but straightforward extension of
-prior work on dependent polynomials and indexed containers and we have
-verified the construction in Agda.
-
-The reason we use dependent polynomials functors on sets is that it is
-well-known that their initial algebras exist
-\cite{initial-algebras-of-dep-poly}. Additionally, initial algebras
-for dependent polynomials are straightforwardly constructed in Agda as
-an inductive type of ``IW'' trees, which is available as a
-construction in the cubical library of Agda \cite{iw,cubical-library}.
-\begin{theorem}
-  Every dependent polynomial functor from $I$ to $I$ has an initial
-  algebra.
-\end{theorem}
+With this interpretation of $F$ as a polynomial functor, $\el(F)$ and
+$\map(F)$ are just interpreted as the action of the functor on objects
+and morphisms, respectively. We interpret the constructors
+$\mathsf{K},\Var$, etc. on functors in the obvious way that matches
+the definitional behavior of $\el$ and $\map$. The non-trivial part of
+the construction is verifying that such such constructions are closed
+under being polynomial. The details are tedious but straightforward
+extension of prior work on dependent polynomials and indexed
+containers and we have verified the construction in Agda.
+
+We use dependent polynomials functors on sets as these are guaranteed
+to have initial algebras \cite{initial-algebras-of-dep-poly}.  Further
+these initial algebras are readily constructed in our Agda
+implementation as an inductive type of $IW$ trees which are already
+available in the cubical library of Agda \cite{iw,cubical-library}.
 Then an element $F \in \sem{X \to \SPF\,X}\gamma$ is an
@@
 used to interpret $\roll, \fold$ and the corresponding axioms.
+
+The remaining details of the interpretation of linear terms as parse
+transformers and verification of the equational axioms is a relatively
+straightforward extension of existing semantics of linear logic in
+monoidal categories, and is included in the
+appendix\cite{linearLogicMonoidalCategory}.
+
+\subsection{Agda Implementation}
+
+This denotational semantics in grammars and parse transformers serves
+as the basis for our prototype implementation of \theoryabbv in
+cubical Agda. The implementation is a shallow embedding, meaning that
+rather than formalizing a syntax of \theoryabbv types and terms, we
+work directly with Agda types and a definition of a formal grammar as
+a function $\String \to \Set$. Then we implement each of the type and
+term constructors of \theoryabbv as combinators on formal grammars or
+parse transformers, and use cubical Agda's equality type to model the
+term equalities. Cubical Agda is convenient for this purpose as it has
+built-in support for function extensionality which we use
+extensively. Axioms such as distributivity of $\oplus/\&$ and
+disjointness of constructors are then provable directly in Agda, and
+we are careful to only construct grammars, terms and proofs using
+constructs that are possible in \theoryabbv. The main difference
+between our shallow embedding and \theoryabbv is that our linear terms
+are written in a combinator-style, without being able to use named
+variables in linear terms. A benefit of this shallow embedding is that
+the parsers are immediately available to a larger Agda development, as
+they are just normal Agda code. In future work we will look to
+implement a type checker for a syntax closer to the presentation in
+this paper, but with the goal of still making it integrate with an
+existing proof assistant.
 
@@
 \subsection{Typed Approaches to Grammars}
+
+\steven{Talk about Conal Elliott's Automatic and Symbolic differentiation paper}
+\steven{Talk briefly about this recent arxiv paper? https://arxiv.org/abs/2411.06358}
+
 \subsubsection{Lambek Calculus and Categorial Grammar}

### 2024-11-14 paper/paper.tex
fix a lot of citations and references

@@
 is an extension of Joachim Lambek's \emph{categorial grammars}
-\cite{lambek}. Categorial grammars are a grammar formalism equivalent
+\cite{lambek58}. Categorial grammars are a grammar formalism equivalent
 in expressive power to context-free grammars that in modern
@@
 as a shallowly embedded domain-specific langauge in the Cubical Agda
-proof assistant \cite{agda,cubical}. We have implemented many example
+proof assistant \cite{VezzosiMortbergAbel2019}. We have implemented many example
 grammars and parsers in our system including regular expressions,
@@
 non-commutative linear logic.
-\cite{krishnaswami_integrating_2015,lambekCalculus}. As in \lnld,
+\cite{krishnaswami_integrating_2015,lambek58}. As in \lnld,
 \theoryabbv includes both non-linear dependent types, as well as
@@
 %
-The main point of departure from \lnld's design is that, as in Lambek calculus \cite{lambekCalculus}, the linear
+The main point of departure from \lnld's design is that, as in Lambek calculus \cite{lambek58}, the linear
 typing is \emph{non-commutative}, i.e., that exchange is not an
@@
 booleans, unit and empty types, and we assume function
-extensionality\cite{standard-type-theory-reference}. We present the
+extensionality\cite{Hofmann_1997}. We present the
 other non-linear type constructions in
-\Cref{fig:non-linear-stuff}. First, we include universe types $U$ of
+\Cref{fig:non-linear-types}. First, we include universe types $U$ of
 small non-linear types and $L$ of linear types. These are defined as
@@
 non-linear type is small and define the universe to internalize
-precisely this judgment \cite{universesAlaCoquand}. These universe
+precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. These universe
 types are needed so that we can define types by recursion on natural
@@
 linear logic or the persistence modality $\square$ of separation logic
-\cite{linearlogic,persistencemodalityseparationlogic}.
+\cite{girard_linear_1987,jung_higher-order_2016}.
 
@@
 \end{figure}
-\max{todo: decide if we shoudl keep all of these rules or not}
 
@@
 specified is given by a non-linear term. The rules for indexed
-$\oplus$ are analogous to a \emph{weak} $\Sigma$ type: it has an
+$\oplus$ are analogous to a ``\emph{weak}'' $\Sigma$ type: it has an
 injection introduction rule $\sigma$, but its elimination rule is
-given by \emph{pattern matching} rather than projections
-\cite{weakSigmatype}. We can define the more typical nullary and
-binary versions of these connectives by using indexing over the empty
-and boolean type respectively. We will freely use $0$ to refer to this
+given by \emph{pattern matching} rather than first and second
+projections. We can define the more typical nullary and binary
+versions of these connectives by using indexing over the empty and
+boolean type respectively. We will freely use $0$ to refer to this
 empty disjunction and $\top$ to refer to the empty conjunction, and
@@
   \caption{Linear Types (selection)}
+  \label{fig:linear-types}
 \end{figure}
@@
   \caption{Linear terms (selection)}
-  \label{fig:linsyntax}
+  \label{fig:linear-terms}
 \end{figure}
@@
 prior work on inductive types
-\cite{quantitativePolynomialFunctors,indexedContainers}. The syntax is
+\cite{nakov_quantitative_2022,altenkirch_indexed_2015}. The syntax is
 given in \Cref{fig:iilt}. First, we add a non-linear type $\SPF\,X$ of
@@
 \emph{proposition} in the terminology of homotopy type
-theory\cite{hott}. The most basic unambiguous types are $\top$ and
+theory\cite{hottbook}. The most basic unambiguous types are $\top$ and
 $0$, and in a system of classical logic all unambiguous types would
@@
 interest. Instead we might try to write a partial function as a
-$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option'' monad
-\cite{maybeMonad}. This allows for the possibility that the input
-string doesn't parse, but is far too weak as a specification: we can
-trivially implement a parser for any type by always returning
-$\inr$. The correct notion of a parser should be one that allows for
-failure, but only in the case that a parse cannot be constructed.
+$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option''
+monad. This allows for the possibility that the input string doesn't
+parse, but is far too weak as a specification: we can trivially
+implement a parser for any type by always returning $\inr$. The
+correct notion of a parser should be one that allows for failure, but
+only in the case that a parse cannot be constructed.
 \begin{definition}
@@
 star is replaced by an arbitrary fixed point
-operation\cite{muregular}.
+operation\cite{leis_towards_1992}.
 
@@
 The Dyck language is an example of an $\textrm{LL}(0)$ language, one
-that can be parsed top-down with no lookahead\cite{ll0}. This means we
-can implement it simply as an \emph{infinite} state deterministic
+that can be parsed top-down with no
+lookahead\cite{rosenkrantz_properties_1970}. This means we can
+implement it simply as an \emph{infinite} state deterministic
 automaton, in \Cref{fig:DyckAutomaton}. Here the state is a ``stack''
@@
 $parse_{M}$ and $print_{M}$ for this automaton can be defined
-analogously to \cref{fig:dfaparser,fig:printdfa}, and so $\LinSigTy
-{s} {M.states} {\LinSigTy {b}{Bool} {Trace_{M}~s~b}}$ is likewise
+analogously to the parser and printer for DFAs, and so $\LinSigTy {s}
+{M.states} {\LinSigTy {b}{Bool} {Trace_{M}~s~b}}$ is likewise
 unambiguous.
@@
 formulation that is closer in spirit to the standard definition of a
-formal \emph{language}\cite{elliott}:
+formal \emph{language}\cite{elliottSymbolicAutomaticDifferentiation2021}:
 \begin{definition}
@@
 The interpretation of dependent types as sets is standard
-\cite{dependent-types-as-sets}. We present the concrete descriptions
-of the semantics of linear types, as well as our non-standard
-non-linear types in \Cref{fig:semantics}.
-The grammar for a literal $c$ has a single parse
-precisely when the input string consists of the single character. The
-grammar for the unit similarly has a single parse for the empty
-string.  A parse of the tensor product $A \otimes B$ consists of a
-\emph{splitting} of the empty string into a prefix $w_1$ and suffix
-$w_2$ along with an $A$ parse of $w_1$ and $B$ parse of $w_2$. A parse
-of $\bigoplus_{x:X} A$ is a pair of an element of the set $X$ and a
-parse of $A(x)$, while dually a parse of $\bigwith_{x:X} A$ is a
-\emph{function} taking any $x:X$ to a parse of $A(x)$. A $w$-parse of
-$A \lto B$ is a function that takes an $A$ parse of some other string
-$w'$ to a $B$ parse of $ww'$, and $B \tol A$ is the same except the
-$B$ parse is for the reversed concatenation $w'w$.
+\cite{Hofmann_1997}. We present the concrete descriptions of the
+semantics of linear types, as well as our non-standard non-linear
+types in \Cref{fig:semantics}. The grammar for a literal $c$ has a
+single parse precisely when the input string consists of the single
+character. The grammar for the unit similarly has a single parse for
+the empty string.  A parse of the tensor product $A \otimes B$
+consists of a \emph{splitting} of the empty string into a prefix $w_1$
+and suffix $w_2$ along with an $A$ parse of $w_1$ and $B$ parse of
+$w_2$. A parse of $\bigoplus_{x:X} A$ is a pair of an element of the
+set $X$ and a parse of $A(x)$, while dually a parse of $\bigwith_{x:X}
+A$ is a \emph{function} taking any $x:X$ to a parse of $A(x)$. A
+$w$-parse of $A \lto B$ is a function that takes an $A$ parse of some
+other string $w'$ to a $B$ parse of $ww'$, and $B \tol A$ is the same
+except the $B$ parse is for the reversed concatenation $w'w$.
 %
@@
 Next, a parse in the equalizer $\equalizer{a}{f}{g}$ is defined as
-a parse in $\sem{A}$ that is mapped to the same parse by $f$ and
-$g$. Recall that $\sem{f}\gamma$ and $\sem{g}$ are elements of
-$\sem{\uparrow(A \lto B)}$, i.e., parse transformers.
+a parse in $\sem{A}$ that is mapped to the same parse by the parse transformers $\sem{f}$ and
+$\sem{g}$.
 %
 The universe $L$ of linear types is interpreted as the set of all
-(small) grammars.
+small grammars.
 \begin{figure}
@@
     \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
-    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:indexed-container}}\\
+    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma)\\
   \end{align*}
   \caption{Grammar Semantics}
+  \label{fig:semantics}
 \end{figure}
@@
 functor} on the category of sets, also sometimes called an
-\emph{indexed container} \cite{dep-poly,indexed-container}.
+\emph{indexed container} \cite{gambino_wellfounded_2004,altenkirch_indexed_2015}.
 \begin{definition}
@@
 We use dependent polynomials functors on sets as these are guaranteed
-to have initial algebras \cite{initial-algebras-of-dep-poly}.  Further
-these initial algebras are readily constructed in our Agda
-implementation as an inductive type of $IW$ trees which are already
-available in the cubical library of Agda \cite{iw,cubical-library}.
-Then an element $F \in \sem{X \to \SPF\,X}\gamma$ is an
-$\sem{X}\gamma$-indexed family of polynomial functors from
-$\sem{X}\gamma\times \Sigma^*$ to $\Sigma^*$, and taking the product
-of these constructs a polynomial functor from $\sem{X}\gamma\times
-\Sigma^*$ to itself. Then $\sem{\mu\,F}\gamma$ is defined to be the
-initial algebra of this functor, and the initial algebra structure is
-used to interpret $\roll, \fold$ and the corresponding axioms.
+to have initial algebras. Further, these initial algebras are readily
+constructed in our Agda implementation as an inductive type of $IW$
+trees which are already available in the cubical library of Agda
+\cite{The_Agda_Community_Cubical_Agda_Library_2024}.  Then an element $F \in \sem{X \to
+  \SPF\,X}\gamma$ is an $\sem{X}\gamma$-indexed family of polynomial
+functors from $\sem{X}\gamma\times \Sigma^*$ to $\Sigma^*$, and taking
+the product of these constructs a polynomial functor from
+$\sem{X}\gamma\times \Sigma^*$ to itself. Then $\sem{\mu\,F}\gamma$ is
+defined to be the initial algebra of this functor, and the initial
+algebra structure is used to interpret $\roll, \fold$ and the
+corresponding axioms.
 
@@
 straightforward extension of existing semantics of linear logic in
-monoidal categories, and is included in the
-appendix\cite{linearLogicMonoidalCategory}.
+monoidal categories, and is included in the appendix\cite{seely89}.
 
@@
 \steven{Choppy notes about lambek calc:
-In 1958 \cite{lambek1958mathematics}, Joachim Lambek introduced his syntactic
+In 1958 \cite{lambek58}, Joachim Lambek introduced his syntactic
 calculus as a logical system for linguistic derivations. In fact, one may view
@@
 theory is extremely successful, having found applications in concurrency theory
-\cite{hoare2000concurrent}, software-defined networks \cite{anderson2014netkat},
+\cite{hoare2009concurrent}, software-defined networks \cite{anderson2014netkat},
 theory of programming languages \cite{angus2001kleene}, compiler optimizations
@@
     \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
-    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:indexed-container}}\\
+    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma)
   \end{align*}

### 2024-11-14 paper/paper.tex
sum proof wip

@@
 \section{Denotational Semantics}
-\steven{Define a Kleene, Lambek, and Chomsky category}
-
-Let $\semcat$ be a Chomsky category.
-
-\steven{Finish the denotation of lin types figure for spf, el , map , mu}
-\begin{figure}
-  \begin{align*}
-    \semg{I} &= I \\
-    \semg{A \otimes B} &= \semg{A} \otimes \semg{B} \\
-    \semg{A \lto B}  &= \semg{A} \lto \semg{B} \\
-    \semg{B \tol A} &= \semg{B} \tol \semg{A} \\
-    \semg{\bigoplus_{x:X} A} &= \coprod_{x:X} \semg{A} \\
-    \semg{\bigwith_{x:X} A} &= \prod_{x : X} \semg{A} \\
-    \semg{\uparrow A} &= Hom(I,A) \\
-    \semg{\equalizer{a:A}{f}{g}} &= Eq\left( \semg{f} , \semg{g} \right) \\
-    \sem{L}\gamma &= \semcat_0\\
-    \sem{\SPF\,X}\gamma &= \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
-    \sem{\el(F)}\gamma\,G &= \sem{F}\gamma\,G\\
-    \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
-    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma)
-  \end{align*}
-  \caption{Denotation of Linear Types}
-\end{figure}
+Here we extend the denotational semantics from \cref{sec:semantics-and-metatheory} to cover all of \theoryname syntax.
 
@@
 \paragraph{$\bigamp$-Introduction}
-$\semg{\dlamb x e} : \semg{\Delta} \to \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)}$ is
+$\semg{\dlamb x e} : \semg{\Delta} \to \prod_{x : X} {\sem{A}(\gamma , x)}$ is
 defined by the universal property of the product
@@
   \semg{\Delta} \arrow[r , "\semg{e}"] &
-  \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)} \arrow[r , "\pi_M"] &
+  \prod_{x : X} {\sem{A}(\gamma , x)} \arrow[r , "\pi_M"] &
   \sem{A}(\gamma , M)
@@
 \paragraph{$\bigoplus$-Introduction}
-$\semg{\sigma~M~e} : \semg{\Delta} \to \LinSigTyLimit {x} {X} \sem{A}(\gamma, x)$
+$\semg{\sigma~M~e} : \semg{\Delta} \to \coprod_{x : X} \sem{A}(\gamma, x)$
 
@@
   \sem{A}(\gamma, M) \arrow[r , "i_M"] &
-  \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)}
+  \coprod_{x : X} {\sem{A}(\gamma , x)}
 \end{tikzcd}
@@
 {\semg{\Delta'_1,\Delta,\Delta'_2}} \arrow[r, "{m_{(\Delta'_1,\Delta),\Delta'_2}}"]                                                                                                           & {\semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}} \arrow[d, "{m_{\Delta'_1,\Delta}\otimes id}"]                             \\
-{\left( \semg{\Delta'_1} \otimes \LinSigTyLimit {x} {X} {\sem{A}(\gamma , x)} \right) \otimes \semg{\Delta'_2}} \arrow[d, "d"]                                                                 & \left( \semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2} \arrow[l, "(id \otimes \semg{e}) \otimes id"] \\
-{\LinSigTyLimit {x} {X} {\left( \sem{\Delta'_1}(\gamma , x) \otimes
-      \sem{A}(\gamma , x) \right) \otimes \sem{\Delta'_2}(\gamma , x)}} \arrow[r, "{m^{-1}_{(\Delta'_1,a:A),\Delta'_2}}"] & {\LinSigTyLimit {x} {X} {\sem{\Delta'_1,a : A,\Delta'_2}(\gamma , x)}} \arrow[d, "{[ \sem{e'}(\gamma , x) ]_{(x : X)}}"]     \\
+{\left( \semg{\Delta'_1} \otimes \coprod_{x:X} {\sem{A}(\gamma , x)} \right) \otimes \semg{\Delta'_2}} \arrow[d, "d"]                                                                 & \left( \semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2} \arrow[l, "(id \otimes \semg{e}) \otimes id"] \\
+{\coprod_{x : X} {\left( \sem{\Delta'_1}(\gamma , x) \otimes
+      \sem{A}(\gamma , x) \right) \otimes \sem{\Delta'_2}(\gamma , x)}} \arrow[r, "{m^{-1}_{(\Delta'_1,a:A),\Delta'_2}}"] & {\coprod_{x : X} {\sem{\Delta'_1,a : A,\Delta'_2}(\gamma , x)}} \arrow[d, "{[ \sem{e'}(\gamma , x) ]_{(x : X)}}"]     \\
 \semg{C}                                                                                                                                                                                      & {\sem{C}(\gamma, x)} \arrow[l]
@@
   & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
-  \otimes i_M \circ \semg{a'}) \otimes id \circ m \otimes id \circ m \\
+  \otimes (i_M \circ \semg{a'})) \otimes id \circ m \otimes id \circ m \\
   & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
-  \otimes i_M \circ \lambda) \otimes id \circ m \otimes id \circ m
+  \otimes (i_M \circ \lambda)) \otimes id \circ m \otimes id \circ m \\
+  & = ? \\
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ i_{M} \\
+  & = \sem{e'}(\gamma , x)
   \end{align*}
-\steven{TODO need to track subscripts on $m$ in the $\bigoplus\beta$ proof, and
-  finish this part of the above part of the proof}
+
+  \steven{How does the distributivity morphism handled? Seems like its a barrier to invoking coherence}
 
@@
     &= \semg{e'} \circ \semg{M/x,a'/a} \tag{\cref{lem:subst}} \\
-    &= \semg{e'}
+    &= \sem{e'}(\gamma , x)
   \end{align*}
-  \steven{TODO shouldn't the semantics of $e'$ be in context $\gamma,x$ not $x$?}
 \end{proof}

### 2024-11-14 paper/paper.tex
pass over future work

@@
 \steven{Talk about Conal Elliott's Automatic and Symbolic differentiation paper}
-\steven{Talk briefly about this recent arxiv paper? https://arxiv.org/abs/2411.06358}
 
@@
 
-\subsubsection{Kleene Algebra}
-Since the early works in the theory of formal languages, Kleene
-algebras have played an important role in its development. They
-generalize the operations known from regular languages by introducing
-operations generalizing language composition, language union and the
-Kleene star.  More generally, they are defined as inequational theory
-where the inequality is meant to capture language containment.  This
-theory is extremely successful, having found applications in concurrency theory
-\cite{hoare2009concurrent}, software-defined networks \cite{anderson2014netkat},
-theory of programming languages \cite{angus2001kleene}, compiler optimizations
-\cite{kozen2000certification} and more.
-
-A frequently fruitful research direction is exploring varying
-extensions of Kleene algebras, Kleene algebra with tests (KAT) being
-one of the most notable ones \cite{kozen1997kleene}. Our approach is radically different from
-most extensions, which usually aim at modifying or adding new
-operations to Kleene algebras, but still keeping it as an inequational
-theory. By adopting a category-theoretic treatment and allowing the
-``order structure'' to encode more information than merely
-inequalities, we were able to extend Kleene algebra to reason about
-parsing as well.
+%% \subsubsection{Kleene Algebra}
+%% Since the early works in the theory of formal languages, Kleene
+%% algebras have played an important role in its development. They
+%% generalize the operations known from regular languages by introducing
+%% operations generalizing language composition, language union and the
+%% Kleene star.  More generally, they are defined as inequational theory
+%% where the inequality is meant to capture language containment.  This
+%% theory is extremely successful, having found applications in concurrency theory
+%% \cite{hoare2009concurrent}, software-defined networks \cite{anderson2014netkat},
+%% theory of programming languages \cite{angus2001kleene}, compiler optimizations
+%% \cite{kozen2000certification} and more.
+
+%% A frequently fruitful research direction is exploring varying
+%% extensions of Kleene algebras, Kleene algebra with tests (KAT) being
+%% one of the most notable ones \cite{kozen1997kleene}. Our approach is
+%% radically different from most extensions, which usually aim at
+%% modifying or adding new operations to Kleene algebras, but still
+%% keeping it as an inequational theory. By adopting a category-theoretic
+%% treatment and allowing the ``order structure'' to encode more
+%% information than merely inequalities, we were able to extend Kleene
+%% algebra to reason about parsing as well.
+
+%%    - verified imperative implementation: Lambek logic ala separation
+%%      logic?
+\subsection{Relation to Separation Logic}
+
+\theoryabbv is similar in spirit to separation logic
+\cite{reynolds_separation_2002}. Semantically, they are closely
+related: linear types in \theoryabbv denote families of sets indexed
+by a monoid of strings, whereas separation logic formulae typically
+denote families of predicates indexed by an ordered partial
+commutative monoid of worlds \cite{jung_higher-order_2016}. The
+monoidal structure in both cases is are instances of the
+category-theoretic notion of Day convolution monoidal structure
+\cite{day1970construction}. From a separation-logic perspective, our
+notion of memory is very primitive: a memory shape is just a string of
+characters and the state of the memory is never allowed to evolve.
+
+%% Syntactically, our type theory can be viewed as a kind of ``separation
+%% type theory'' in that we care not just about provability of formulae
+%% but the constructive content of our linear terms, as these act as
+%% parsers. The design of \theoryabbv is based on the $\textrm{LNL}_D$
+%% calculus which explicitly was designed to be a dependent
+%% type-theoretic version of separation logic. Many concepts of
+%% separation logic then have direct analogues in our system. The
+%% characters $c$ are analogous to the ``points-to'' formulae $l \mapsto
+%% x$ in separation logic. Our $\otimes,\lto,\tol$ are analogous to the
+%% separating conjunction and magic-wand. The analogue of the persistence
+%% modality $\square$ of separation logic is $\bigoplus_{\_:\uparrow A} I$,
+%% which ``zeros out'' all but the parses from the empty string. If we
+%% think of the portions of the string we are parsing as our notion of
+%% resource, then these make sense as the ``resource-free'' elements of
+%% the type $A$.
+
+This semantic connection to separation logic suggests an avenue of
+future work: to develop a program logic based on non-commutative
+separation logic for verifying imperative implementations of
+parsers. This could be implemented by modifying an existing separation
+logic implementation or embedding the logic within \theoryabbv.
 
@@
 
-%%    - verified imperative implementation: Lambek logic ala separation
-%%      logic?
-\subsection{Relation to Separation Logic}
-
-\theoryabbv is similar in spirit to separation logic. Semantically,
-they are closely related: linear types in \theoryabbv denote families
-of sets indexed by a monoid of strings, whereas separation logic
-formulae typically denote families of predicates indexed by an ordered
-partial commutative monoid of worlds. The monoidal structure in both
-cases is are instances of the category-theoretic notion of Day
-convolution monoidal structure. From a separation-logic perspective,
-our notion of memory is very primitive: a memory shape is just a
-string of characters and the state of the memory is never allowed to
-evolve.
-
-Syntactically, our type theory can be viewed as a kind of ``separation
-type theory'' in that we care not just about provability of formulae
-but the constructive content of our linear terms, as these act as
-parsers. The design of \theoryabbv is based on the $\textrm{LNL}_D$
-calculus which explicitly was designed to be a dependent
-type-theoretic version of separation logic. Many concepts of
-separation logic then have direct analogues in our system. The
-characters $c$ are analogous to the ``points-to'' formulae $l \mapsto
-x$ in separation logic. Our $\otimes,\lto,\tol$ are analogous to the
-separating conjunction and magic-wand. The analogue of the persistence
-modality $\square$ of separation logic is $\bigoplus_{\_:\uparrow A} I$,
-which ``zeros out'' all but the parses from the empty string. If we
-think of the portions of the string we are parsing as our notion of
-resource, then these make sense as the ``resource-free'' elements of
-the type $A$.
-
-This semantic connection to separation logic also suggests an avenue
-of future work: rather than working with a dependent type theory, we
-could instead make a program logic based on non-commutative separation
-logic for verifying imperative implementations of parsers. This could
-be accomplished by modifying an existing separation logic
-implementation or embedding the logic within \theoryabbv.
-
 \subsection{Implementation}
-\steven{Moving away from a prototype implementation of \theoryabbv}
-\steven{Mention performance}
+
+Our Agda prototype implementation serves as a usefule proof of concept
+for showing what can be implemented in \theoryabbv, but it has
+downsides we aim to address in future work. Firstly, it would be
+preferable to work with the more intuitive type theoretic syntax we
+have used in this work, rather than the combinator-style our shallow
+embedding requires. Additionally, Agda itself does not have a
+high-performance implementation, and so the parsers we implement in
+Agda do not have competitive performance to industry parser
+generators. In future work we aim to study if we can embed a proof of
+the correctness of a parser generator that produces imperative
+programs, and to what extent the correctness of those imperative
+programs can be proven within \theoryabbv.
 
@@
 definition of arbitrarily powerful grammars, the system could also be
-used for more sophisticated semantic analysis such as scope checking
-or type checking. Alternatively, we could more directly encode type
-type systems as linear types in a modified version of \theoryabbv
-whose semantics is not in families of sets indexed by \emph{strings}
-but families of sets indexed by \emph{abstract syntax trees}.
+used in principle for more sophisticated semantic analysis such as
+scope checking or type checking. Alternatively, we could more directly
+encode type type systems as linear types in a modified version of
+\theoryabbv where linear types are not grammars over strings, but
+\emph{type systems} over trees. This could analogously serve as a
+framework for verified type checking and static analysis.
 

### 2024-11-14 paper/paper.tex
appendix stuff

@@
 
-\max{TODO: add rules for non-linear types $\Pi,\Sigma$, equality etc}
-
-\begin{figure}
-  \begin{mathpar}
-  \boxed{\Gamma \vdash M : X}
-
-  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash M : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash M : X}
-  %
-  \and
-  %
-  \inferrule{~}{\Gamma \vdash () : 1}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash M : X \\ \Gamma \vdash N : \subst Y M x}{\Gamma \vdash (M, N) : \SigTy x X Y}
-  %
-  \and
-%
-  \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_1\, M : X}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_2\, M : \subst Y {\pi_1\, M} x}
-  \and
-  \inferrule{\Gamma, x : X \vdash M : Y}{\Gamma \vdash \lamb x M : \PiTy x X Y}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash M : \PiTy x X Y \\ \Gamma \vdash N : X}{\Gamma \vdash \app M {N} : \subst Y {N} x}
-  %
-  \and
-  %
-  \inferrule{\Gamma \vdash M \equiv N : X}{\Gamma \vdash \mathsf{refl} : M =_X N}
-  \and
-  \textrm{TODO: inductive types (boolean, nat) with their eliminators allowed in }
-  \end{mathpar}
-  \caption{Intuitionistic typing}
-  \label{fig:inttyping}
-\end{figure}
+%% \begin{figure}
+%%   \begin{mathpar}
+%%   \boxed{\Gamma \vdash M : X}
+
+%%   \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{\Gamma \vdash M : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash M : X}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{~}{\Gamma \vdash () : 1}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{\Gamma \vdash M : X \\ \Gamma \vdash N : \subst Y M x}{\Gamma \vdash (M, N) : \SigTy x X Y}
+%%   %
+%%   \and
+%% %
+%%   \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_1\, M : X}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_2\, M : \subst Y {\pi_1\, M} x}
+%%   \and
+%%   \inferrule{\Gamma, x : X \vdash M : Y}{\Gamma \vdash \lamb x M : \PiTy x X Y}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{\Gamma \vdash M : \PiTy x X Y \\ \Gamma \vdash N : X}{\Gamma \vdash \app M {N} : \subst Y {N} x}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{\Gamma \vdash M \equiv N : X}{\Gamma \vdash \mathsf{refl} : M =_X N}
+%%   \and
+%%   \textrm{TODO: inductive types (boolean, nat) with their eliminators allowed in }
+%%   \end{mathpar}
+%%   \caption{Intuitionistic typing}
+%%   \label{fig:inttyping}
+%% \end{figure}
 
@@
     \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
-    %
-    \\
-
   \end{mathpar}
@@
   \begin{mathpar}
-    \boxed{\Gamma \vdash M \equiv N : X}
-
-    \inferrule{\Gamma \vdash p : M =_X N}{\Gamma \vdash M \equiv N : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \app {(\lamb x M)} {N} \equiv \subst x M {N} : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash m \equiv \lamb x {\app M x} : \PiTy x X Y}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \pi_1\, (M, N) \equiv M : X}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash \pi_2\, (M , N) \equiv N : \subst x {M} Y}
-%
-    \and
-%
-    \inferrule{~}{\Gamma \vdash M \equiv (\pi_1\, M, \pi_2\, M) : \SigTy x X Y}
-%
-    \and
-%
-    % \inferrule{~}{}
-%
-    \inferrule{~}{\Gamma \vdash M \equiv N : 1}
-%
-    % \and
-%
-    % \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
-
-    \\
-
     \boxed{\Gamma ; \Delta \vdash e \equiv e' : A}
@@
 
-  \max{do we need identity and composition of substitutions and their functoriality?}
+  %% \max{do we need identity and composition of substitutions and their functoriality?}
   %% By induction we can define an identity substitution
@@
   the type theory that are parameterized by a non-linear context
-  $\Gamma$ support an admissible action of a substitution $\gamma \in
-  \Subst(\Gamma',\Gamma)$:
+  $\Gamma$ support the admissible actions of a substitution $\gamma
+  \in \Subst(\Gamma',\Gamma)$ in \Cref{fig:non-linear-substitution}.
+  %% Further, these all satisfy identity and composition equations M[\Gamma] = M and M[\gamma][\gamma'] = M[(\gamma \circ \gamma')]
+\end{definition}
+
+\begin{figure}
   \begin{mathpar}
@@
   \end{mathpar}  
-
-  %% Further, these all satisfy identity and composition equations M[\Gamma] = M and M[\gamma][\gamma'] = M[(\gamma \circ \gamma')]
-\end{definition}
+  \caption{Non-linear substitution}
+  \label{fig:non-linear-substitution}
+\end{figure}
 
@@
 
-  \max{identity and composition?}
+  %% \max{identity and composition?}
 \end{definition}
@@
 \section{Denotational Semantics}
-Here we extend the denotational semantics from \cref{sec:semantics-and-metatheory} to cover all of \theoryname syntax.
+Here we extend the denotational semantics from
+\cref{sec:semantics-and-metatheory} to cover all of \theoryname
+syntax. Here we will freely use that the category of grammars is a
+complete, co-complete biclosed monoidal category, and use categorical
+notation for the constructions in the denotational semantics. For
+example, we will use the same notation $I,\otimes,\lto,\tol$ for the
+biclosed monoidal structure of $\Grammar$ that we do for the
+corresponding syntactic notions.
 
@@
 \subsection{Grammar Semantics for Linear Terms}
-Here we define denotations of linear terms. Note that the denotations interpret
-typing derivations, not raw terms, as the data of how contexts are
-split is needed in order to construct the correct associator
-functions. Further, we demonstrate that the denotational semantics resprects the
-equational theory of \theoryabbv.
+Here we define denotations of linear terms. Note that the denotations
+interpret typing derivations, not raw terms, as the data of how
+contexts are split is needed in order to construct the correct
+associator functions. Further, we demonstrate that the denotational
+semantics resprects the equational theory of \theoryabbv. The
+correctness of the equational theory heavily relies on the
+\emph{coherence theorem} for monoidal categories. The coherence
+theorem says that any diagram in a monoidal category constructed using
+only associators $\alpha_{A,B,C} : (A \otimes B) \otimes C \cong A
+\otimes (B \otimes C)$, unitors $\rho_A : A \otimes I \cong A$ and
+$\lambda_A : I \otimes A \cong A$ and compositions and tensor products
+of these, commutes. We call a morphism built in this way a
+\emph{generalized associator}.
 
@@
 
-\steven{Define generalized associators}
-
 Because
@@
 {\coprod_{x : X} {\left( \sem{\Delta'_1}(\gamma , x) \otimes
-      \sem{A}(\gamma , x) \right) \otimes \sem{\Delta'_2}(\gamma , x)}} \arrow[r, "{m^{-1}_{(\Delta'_1,a:A),\Delta'_2}}"] & {\coprod_{x : X} {\sem{\Delta'_1,a : A,\Delta'_2}(\gamma , x)}} \arrow[d, "{[ \sem{e'}(\gamma , x) ]_{(x : X)}}"]     \\
+      \sem{A}(\gamma , x) \right) \otimes \sem{\Delta'_2}(\gamma , x)}} \arrow[r, "\coprod_{x:X}({m^{-1}_{(\Delta'_1,a:A),\Delta'_2}})"] & {\coprod_{x : X} {\sem{\Delta'_1,a : A,\Delta'_2}(\gamma , x)}} \arrow[d, "{[ \sem{e'}(\gamma , x) ]_{(x : X)}}"]     \\
 \semg{C}                                                                                                                                                                                      & {\sem{C}(\gamma, x)} \arrow[l]
@@
   \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}}
-  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id
   \otimes \semg{\sigma~M~a'}) \otimes id \circ m \otimes id \circ m \\
-  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X}m^{-1} \circ d \circ (id
   \otimes (i_M \circ \semg{a'})) \otimes id \circ m \otimes id \circ m \\
-  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
-  \otimes (i_M \circ \lambda)) \otimes id \circ m \otimes id \circ m \\
-  & = ? \\
-  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ i_{M} \\
-  & = \sem{e'}(\gamma , x)
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X} m^{-1} \circ d \circ (id \otimes i_M) \otimes id \circ (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X} m^{-1} \circ i_M \circ (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
+  & = \sem{e'}(\gamma , M) \circ m^{-1} (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
+  & = \sem{e'}(\gamma , M) \tag{coherence}
   \end{align*}

### 2024-11-14 paper/paper.tex
remove outdated note
lambek calculus related work

@@
 The design of \theoryabbv
-is an extension of Joachim Lambek's \emph{categorial grammars}
-\cite{lambek58}. Categorial grammars are a grammar formalism equivalent
+is an extension of Joachim Lambek's \emph{syntactic calculus}
+\cite{lambek58}. Lambek calculus is a grammar formalism equivalent
 in expressive power to context-free grammars that in modern
@@
 \subsubsection{Lambek Calculus and Categorial Grammar}
-
-\steven{Note that lambek calculus and its extensions don't have inductives, or
-  even kleene star as far as I can tell. }
-
-\steven{Choppy notes about lambek calc:
-In 1958 \cite{lambek58}, Joachim Lambek introduced his syntactic
-calculus as a logical system for linguistic derivations. In fact, one may view
-the subtheory of our calculus generated over $\otimes, \lto, \tol$ to be a proof
-relevant presentation of Lambek's original system.
-
-Lambek later explicitly connects his syntactic calculus to the structure of a
-non-commutative biclosed monoidal category \cite{lambek1988categorial}. There is
-a rich line of subsequent research after these revelations by Lambek, including
-the study of pregroup grammars
-\cite{coeckeMathematicalFoundationsCompositional2010}, the presentation of
-abstract categorial grammars \cite{degrooteAbstractCategorialParsing2015}, and
-even Lambek calculus with dependent types \cite{luoLambekCalculusDependent}. As
-far as the authors can tell, these works seem to be primarily interested in natural
-language semantics in the style of Montague grammars, rather than parsing formal
-languages. That is, they seem to try to give a functorial semantic
-interpretation of natural language.
-
-There seem to be some implementations from the abstract categorial grammar
-toolkit, but again they seem natural language focused and not up to the task of
-verified formal language parsing.
-
-Also mention here that Luo's work \cite{luoLambekCalculusDependent} seems similar at first
-glance, but isn't as similar as it might look because the judgments and
-dependent types seem to
-take on a different interpretation of ours. Say something here like, ``as we
-understand it, it differs in x,y,z ways''
-}
+Lambek's original syntactic calculus \cite{lambek58} describes a logical system
+for linguistic derivations. Later, he explicitly gives semantics to his
+syntactic calculus inside of a non-commutative biclosed monoidal category
+\cite{lambek1988categorial}. Following Lambek's illuminating ideas, Buszkowski
+gives a succinct overview of the
+rich body of work studying the role of linear logic and lambda calculi in
+linguistics \cite{buszkowskiTypeLogicsGrammar2003}. \steven{Is citing a survey
+  like this okay?} Grammars based on logic in
+this manner may go by the names Lambek calculus, typelogical grammar, or
+categorial grammar. Such categorial grammars correspond to the subtheory of our
+calculus generated by $\otimes$, $\lto$, and $\tol$.
+
+While applicable to both
+formal and natural languages, the existing works on categorial grammar are
+frequently oriented toward giving functorial semantics to natural language in
+the tradition of Montague \cite{Montague1974-MONFPS}. The Abstract Categorial
+Grammar Toolkit \cite{Guillaume2024} and Grammatical Framework \cite{ranta-2011}
+each mechanize aspects of Lambek calculus ; however, to the best of our knowledge,
+\theoryabbv is the first mechanization designed for formal language parsing.
+
+\steven{Contrast their interpretation of dependent types with ours. How tf do I
+  say idk what they're doing. Perhaps just say that their dependent types take
+  on a different interpreation and no more.}
+Luo \cite{luo} describes a Lambek calculus with dependent types.
 
@@
 
-  \steven{How does the distributivity morphism handled? Seems like its a barrier to invoking coherence}
-
   \begin{align*}

### 2024-11-14 paper/paper.tex
draft of related work

@@
 
-\steven{Talk about Conal Elliott's Automatic and Symbolic differentiation paper}
-
 \subsubsection{Lambek Calculus and Categorial Grammar}
@@
 each mechanize aspects of Lambek calculus ; however, to the best of our knowledge,
-\theoryabbv is the first mechanization designed for formal language parsing.
-
-\steven{Contrast their interpretation of dependent types with ours. How tf do I
-  say idk what they're doing. Perhaps just say that their dependent types take
-  on a different interpreation and no more.}
-Luo \cite{luo} describes a Lambek calculus with dependent types.
-
-\subsubsection{Other typed approaches}
-\steven{this subsection should be renamed}
-The usage of a simple type system to reason about regular expressions was
-introduced by Frisch and Cardelli in 2004 \cite{frischCardelli}, and later expanded
-by Hengelin and Nielsen in 2011 \cite{henglein_regular_2011} to handle proofs of regular expression
-containment.
-
-\steven{Mention Neel's typed algebraic approach paper here. Say it is similar at
-  first glance, as its parsing and it has types, but that he actually is representing
-  a system for proving the LL(1) condition }
+\theoryabbv is the first mechanization designed for formal language parsing first.
+Moreover, we provide a novel use of inductive and dependent types inside of a
+Lambek calculus.
+
+Luo \cite{luo} describes a Lambek calculus with dependent types. However, they
+do not provide an analagous denotational semantics, so it is not clear what
+exactly their semantics are. Further, their contexts of non-linear variables are
+ordered, while ours are not, so they provide a distinct dependent extension
+whose types do not have a one-to-one correspondence with the dependent types in \theoryabbv.
+
+\subsubsection{Typed Grammars}
+The usage of a simple type system to reason about regular expressions up to weak
+equivalence was
+introduced by Frisch and Cardelli \cite{frischCardelli}.  Hengelin and Nielsen
+\cite{henglein_regular_2011} later adapt this approach by providing a type
+system for regular expressions whose semantics were in one-to-one correspondence
+with the set of parse trees.
+Elliott \cite{elliottSymbolicAutomaticDifferentiation2021} gives a similar
+semantic interpretation of grammars as type-level predicates on strings. The type system of \theoryabbv extends these semantic interpretations of to a broader class
+of grammars.
+
+Krishnaswami and Yallop \cite{krishnaswami_typed_2019} provide a type system for
+demonstrating that carries proves of the LL(1) condition for context-free
+grammars. Despite being a type system over formal grammars, their interpretation
+of types is very different from the semantics of \theoryabbc.
 

### 2024-11-14 paper/paper.tex
progress on sum eta
related work

@@
 Prior approaches to verified parsing focus on verification of a
-particular grammar formalism such as non-left-recursive grammars or LL(1) grammars \cite{lasserCoStarVerifiedALL2021,EdelmannZippy2020,danielssonTotalParserCombinators2010}.
-\steven{These have restrictions on left recursion or LL(1)-ness. Is it the most accurate
-  to call that a difference in formalism? It would seem that these are all CFGs
-  + conditions. Maybe these aren't the best works to
-  compare to if they're verifying particular algorithms}
-Each new grammar formalism must then be extended with its own verified
-implementation.
+particular grammar formalism such as non-left-recursive grammars or
+LL(1) grammars
+\cite{lasserCoStarVerifiedALL2021,EdelmannZippy2020,danielssonTotalParserCombinators2010}.
+Each new grammar formalism is extended with its own independent
+verified implementation.
 
@@
 syntactic calculus inside of a non-commutative biclosed monoidal category
-\cite{lambek1988categorial}. Following Lambek's illuminating ideas, Buszkowski
-gives a succinct overview of the
-rich body of work studying the role of linear logic and lambda calculi in
-linguistics \cite{buszkowskiTypeLogicsGrammar2003}. \steven{Is citing a survey
-  like this okay?} Grammars based on logic in
+\cite{lambek1988categorial}. This led to many uses of non-commutative linear logic and lambda calculi in
+linguistics \cite{buszkowskiTypeLogicsGrammar2003}. Grammars based on logic in
 this manner may go by the names Lambek calculus, typelogical grammar, or
@@
 
-Luo \cite{luo} describes a Lambek calculus with dependent types. However, they
-do not provide an analagous denotational semantics, so it is not clear what
-exactly their semantics are. Further, their contexts of non-linear variables are
-ordered, while ours are not, so they provide a distinct dependent extension
-whose types do not have a one-to-one correspondence with the dependent types in \theoryabbv.
+The most similar prior work to our own is Luo's Lambek calculus with
+dependent types \cite{luo}. Their design differs from our own in
+allowing linear types to depend on other linear types and supporting
+directed $\Pi$ and $\Sigma$ types that have no analog in our
+system. They do not provide a semantics for these connectives, and it
+is unclear how to interpret their connectives in our grammar
+semantics. Further, we provide several examples showing that
+\theoryabbv is a practical system for describing grammar formalisms
+and parsers, and it is unclear if these could be implemented in their calculus.
 
@@
 grammars. Despite being a type system over formal grammars, their interpretation
-of types is very different from the semantics of \theoryabbc.
+of types is very different from the semantics of \theoryabbv.
 
@@
 \paragraph{$\bigoplus\eta$}
-\steven{TODO $\eta$ for sums}
+It suffices to show
+\[
+  \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}} = \semg{f[c'/c]}
+\]
+\begin{proof}
+  \begin{align*}
+    \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}}
+    &= [ \sem{f[\sigma~x~a/c]} ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m
+  \end{align*}
+\end{proof}
+\max{todo: finish}
 

### 2024-11-14 paper/paper.tex
related work

@@
 Elliott \cite{elliottSymbolicAutomaticDifferentiation2021} gives a similar
-semantic interpretation of grammars as type-level predicates on strings. The type system of \theoryabbv extends these semantic interpretations of to a broader class
-of grammars.
-
-Krishnaswami and Yallop \cite{krishnaswami_typed_2019} provide a type system for
-demonstrating that carries proves of the LL(1) condition for context-free
-grammars. Despite being a type system over formal grammars, their interpretation
-of types is very different from the semantics of \theoryabbv.
+semantic interpretation of regular grammars as type-level predicates on strings.
+The type system of \theoryabbv extends these semantic interpretations of to a broader class of grammars and also gives a formal syntax and equational theory for the parse transformers that these prior works lack.
+
+%% Krishnaswami and Yallop \cite{krishnaswami_typed_2019} provide a type system for
+%% demonstrating that carries proves of the LL(1) condition for context-free
+%% grammars. Despite being a type system over formal grammars, their interpretation
+%% of types is very different from the semantics of \theoryabbv.
 

### 2024-11-14 paper/paper.tex
tweak

@@
 
-\paragraph{Weak and Strong Equivalence}
+%% \paragraph{Weak and Strong Equivalence}
 In the theory of formal grammars, there are two different notions of
@@
 these notions of equivalence in \theoryabbv.
-
 \begin{definition}
@@
 
-That is, grammars are weakly equivalent if we can construct a parse
-tree of $B$ from a parse tree of $A$ and vice-versa, whereas they are
-strongly equivalent if these processes are inverse to each other. The
-definition of retract is a useful intermediate notion. In retrospect,
-our axiom about $\String$ says that $\String$ is a retract of $\top$.
+%% That is, grammars are weakly equivalent if we can construct a parse
+%% tree of $B$ from a parse tree of $A$ and vice-versa, whereas they are
+%% strongly equivalent if these processes are inverse to each other. The
+%% definition of retract is a useful intermediate notion. In retrospect,
+%% our axiom about $\String$ says that $\String$ is a retract of $\top$.
 
@@
 
-\paragraph{Ambiguity}
+%% \paragraph{Ambiguity}
 A formal grammar $\A$ is ambiguous if there are multiple parse trees
@@
 
-\paragraph{Partial and Total Parsers}
+%% \paragraph{Partial and Total Parsers}
 

### 2024-11-14 paper/paper.tex
sum eta law

@@
 
-\subsubsection{Lambek Calculus and Categorial Grammar}
+%% \subsubsection{Lambek Calculus and Categorial Grammar}
 Lambek's original syntactic calculus \cite{lambek58} describes a logical system
@@
 
-\subsubsection{Typed Grammars}
+%% \subsubsection{Typed Grammars}
 The usage of a simple type system to reason about regular expressions up to weak
@@
 
-\subsection{Type Checking and Elaboration}
+\subsection{Type Checking and Semantic Analysis}
 Our focus in this work has been on the verification of parsers for
@@
 \[
-  \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}} = \semg{f[c'/c]}
+  \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}} = \semg{f[c'/c]} = \semg{f}
   \]
+First, expanding the left hand side, we have.
 \begin{proof}
@@
     \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}}
-    &= [ \sem{f[\sigma~x~a/c]} ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m
+    &= [ \semg{f[\sigma~x~a/c]} ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
+    &= [ \semg{f} \circ (id \otimes i_x) \otimes id ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
+    &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
+  \end{align*}
+Since the domain has the universal property of a coproduct (due to distributivity), to prove
+this is equal to $\semg{f}$, it is sufficient to prove they are equal
+when composed with the injections:
+
+\begin{align*}
+  \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id
+  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d  \circ (id \otimes i_y) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
+  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ i_y \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
+  &= \semg{f} \circ (id \otimes i_y) \otimes id \circ (m^{-1}) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \\
+  &= \semg{f} \circ (m^{-1}) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id \\
+  &= \semg{f} \circ (id \otimes i_y) \otimes id \tag{coherence}\\
 \end{align*}
 \end{proof}
-\max{todo: finish}
 

### 2024-11-14 paper/paper.tex
lambek

@@
 \subsection{Typed Approaches to Grammars}
-
-%% \subsubsection{Lambek Calculus and Categorial Grammar}
+\subsubsection{Lambek Calculus and Categorial Grammar}
 Lambek's original syntactic calculus \cite{lambek58} describes a logical system
-for linguistic derivations. Later, he explicitly gives semantics to his
-syntactic calculus inside of a non-commutative biclosed monoidal category
+for linguistic derivations, and it can be given semantics inside of a non-commutative biclosed monoidal category
 \cite{lambek1988categorial}. This led to many uses of non-commutative linear logic and lambda calculi in
-linguistics \cite{buszkowskiTypeLogicsGrammar2003}. Grammars based on logic in
-this manner may go by the names Lambek calculus, typelogical grammar, or
-categorial grammar. Such categorial grammars correspond to the subtheory of our
-calculus generated by $\otimes$, $\lto$, and $\tol$.
-
-While applicable to both
-formal and natural languages, the existing works on categorial grammar are
-frequently oriented toward giving functorial semantics to natural language in
-the tradition of Montague \cite{Montague1974-MONFPS}. The Abstract Categorial
-Grammar Toolkit \cite{Guillaume2024} and Grammatical Framework \cite{ranta-2011}
-each mechanize aspects of Lambek calculus ; however, to the best of our knowledge,
-\theoryabbv is the first mechanization designed for formal language parsing first.
-Moreover, we provide a novel use of inductive and dependent types inside of a
-Lambek calculus.
+linguistics \cite{buszkowskiTypeLogicsGrammar2003} --- including mechanized
+categorial grammar parsers \cite{Guillaume2024,ranta-2011}. This style of grammar
+formalism has gone by the names Lambek calculus or \emph{categorial grammar},
+and it is equal in expressivity to context-free grammars.
+The existing works on categorial grammar
+frequently present logical operations on linear predicates rather than the term
+language given in this paper.
+Additionally, much of the existing work is
+oriented toward computational linguistics, giving functorial semantics to natural language in
+the tradition of Montague \cite{Montague1974-MONFPS}.
 

### 2024-11-14 paper/paper.tex
shrinkage

@@
 the context $\internalize{\stringquote{ab}} = a : \literal a , b :
-\literal b$. On the left side of \cref{fig:fingram}, we give a typing
-derivation that shows that $\inl(a \otimes b)$ defines a valid parse
-tree for the finite grammar $(\literal a \otimes \literal b) \oplus
-\literal c$. Underneath the derivation in \cref{fig:fingram}, we give
-a linear $\lambda$-term that implements the program given in the
-derivation.\max{reverse the order of these?} \steven{the order of what?}
+\literal b$. In \Cref{fig:fingram}, we give a lambda term and its
+typing derivation to define a parse for a finite grammar.
 
 \begin{figure}
-\begin{mathpar}
+\begin{lstlisting}
+f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
+f (a , b) = inl (a (*@$\otimes$@*) b)
+\end{lstlisting}\begin{mathpar}
   \inferrule
@@
 \end{mathpar}
-\begin{lstlisting}
-f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
-f (a , b) = inl (a (*@$\otimes$@*) b)
-\end{lstlisting}
 \caption{\stringquote{ac} matches
@@
 string.
-
-Just as \cref{fig:kleenestarderivation} shows the \stringquote{ab} matches
-$({\literal a}^{*} \otimes \literal b) \oplus \literal c$, \cref{fig:nfaderivation} demonstrates that \stringquote{ab} is
-accepted by the NFA representation of this regular expression. In particular, we
-show that \stringquote{ab} matches the grammar $\Trace~0$ since $\0$ is the
-initial state of the NFA.\ If we construct parse transformers from
-$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ to $\Trace~\0$, and vice versa, we would demonstrate that the regular expression
-and the automaton capture the same language --- i.e.\ they recognize the same
-set of strings. As we will see in \cref{sec:applications}, we can indeed define parse
-transformers between regular expressions and their associated automaton.
-Moreover, we prove that these parse transformers define a bijective
-correspondence of parses over every input string. That is, in a strong sense the
-parse of a regular expression and the traces of their associated automata
-capture precisely the same data, i.e., that they are \emph{strongly equivalent} as grammar formalisms.
+%
+Lastly, we give a lambda term that constructs an accepting trace
+starting at the initial state for the string \stringquote{ab}.  Later
+in \Cref{sec:applications}, we will show that we can actually
+construct mutually inverse functions between the regular expression
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ and its
+corresponding NFA traces ($\Trace~\0$) demonstrating that the regular
+expression and the automaton capture the same language. Further since
+the functions are mutually inverse, this shows they are \emph{strongly
+equivalent} as grammars.
 
@@
   0to1 : (*@$\uparrow$@*)(Trace 1 (*@$\lto$@*) Trace 0)
+
+k : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*)  Trace 0)
+k (a , b) = 0to1 (1to1 a (1to2 b stop))
+
 \end{lstlisting}
@@
   \label{fig:exampleNFA}
-\end{figure}
-
-\begin{figure}
-\begin{lstlisting}
-k : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*)  Trace 0)
-k (a , b) = 0to1 (1to1 a (1to2 b stop))
-\end{lstlisting}
-\caption{\stringquote{ab} is accepted by the NFA for
-  $({\literal a}^{*} \otimes {\literal b }) \oplus \literal c$}
-\label{fig:nfaderivation}
 \end{figure}
@@
 \begin{figure}
+  \begin{footnotesize}
     \begin{mathpar}
@@
   \end{mathpar}
+  \end{footnotesize}
   \caption{Non-linear types (selection)}
@@
 
-We give an overview of the linear types in \Cref{fig:linear-types} and
-terms in \Cref{fig:linear-terms}. The equational theory for these
-types is straightforward and included in the appendix. First, the
-linear variable rule says that a linear variable can be used if it is
-the \emph{only} variable in the context.
+We give an overview of the linear types and terms in
+\Cref{fig:linear-types-and-terms}. The equational theory for these
+types is straightforward $\beta\eta$ equivalence and included in the
+appendix. First, the linear variable rule says that a linear variable
+can be used if it is the \emph{only} variable in the context.
 
@@
 \begin{figure}
+  \begin{footnotesize}
   \begin{mathpar}
@@
   \end{mathpar}
-  \caption{Linear Types (selection)}
-  \label{fig:linear-types}
-\end{figure}
-
-\begin{figure}
 \begin{mathpar}
@@
   \end{mathpar}    
-  \caption{Linear terms (selection)}
-  \label{fig:linear-terms}
+  \end{footnotesize}
+  \caption{Linear Types and Terms (selection)}
+  \label{fig:linear-types-and-terms}
 \end{figure}

### 2024-11-14 paper/paper.tex
20 pages

@@
 \label{sec:discussion}
-\subsection{Typed Approaches to Grammars}
-\subsubsection{Lambek Calculus and Categorial Grammar}
-Lambek's original syntactic calculus \cite{lambek58} describes a logical system
-for linguistic derivations, and it can be given semantics inside of a non-commutative biclosed monoidal category
-\cite{lambek1988categorial}. This led to many uses of non-commutative linear logic and lambda calculi in
-linguistics \cite{buszkowskiTypeLogicsGrammar2003} --- including mechanized
-categorial grammar parsers \cite{Guillaume2024,ranta-2011}. This style of grammar
-formalism has gone by the names Lambek calculus or \emph{categorial grammar},
-and it is equal in expressivity to context-free grammars.
-The existing works on categorial grammar
-frequently present logical operations on linear predicates rather than the term
-language given in this paper.
-Additionally, much of the existing work is
-oriented toward computational linguistics, giving functorial semantics to natural language in
-the tradition of Montague \cite{Montague1974-MONFPS}.
+\paragraph{Grammars as (Linear) Types}
+Lambek's original syntactic calculus \cite{lambek58} describes a
+logical system for linguistic derivations, and it can be given
+semantics inside of a non-commutative biclosed monoidal category
+\cite{lambek1988categorial}. This led to many uses of non-commutative
+linear logic and lambda calculi in linguistics
+\cite{buszkowskiTypeLogicsGrammar2003} --- including mechanized
+categorial grammar parsers \cite{Guillaume2024,ranta-2011}. This style
+of grammar formalism has gone by the names Lambek calculus or
+\emph{categorial grammar}, and it is equal in expressivity to
+context-free grammars.  The existing works on categorial grammar are
+different in nature to our approach: they are based on non-commutative
+linear logic, but their terms do not include elimination rules, and so
+can only express parse trees, and so cannot be used to express
+verified parsers.
 
@@
 %%      logic?
-\subsection{Relation to Separation Logic}
+\paragraph{Relation to Separation Logic}
 
@@
 
-\subsection{Semantic Actions}
+\paragraph{Semantic Actions}
 
@@
 
-\subsection{Implementation}
+\paragraph{Implementation}
 
@@
 the correctness of a parser generator that produces imperative
-programs, and to what extent the correctness of those imperative
+programs, and if the correctness of those imperative
 programs can be proven within \theoryabbv.
 
-\subsection{Type Checking and Semantic Analysis}
+\paragraph{Type Checking and Semantic Analysis}
 Our focus in this work has been on the verification of parsers for

### 2024-11-14 paper/paper.tex
small tweaks, remove comments from body of paper

@@
   automata. Using these encodings, we define parsers for regular
-  expressions using deterministic automata, as well as providing
-  examples parsers for LL(1) and LR(1) grammars.
+  expressions using deterministic automata, as well as
+  examples of verified parsers of context-free grammars.
 
@@
 
-Our contributions are then:\max{This contribution section needs to be punchier}
+Our contributions are then:% \max{This contribution section needs to be punchier}
 %
@@
 
-\max{combine these three figures into one?}
 Finite automata are precisely the class of machines that recognize regular
@@
 \begin{align*}
-  \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id
+  \semg{f} &\circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id\\
   &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d  \circ (id \otimes i_y) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\

### 2024-11-15 paper/paper.tex
jdgeq figure
jdg eq premises on all inferrules
draftfalse
nathan email
comments section 3

--- 2024-11-14 paper/paper.tex
@@
 Prior approaches to verified parsing focus on verification of a
-particular grammar formalism such as non-left-recursive grammars or
-LL(1) grammars
-\cite{lasserCoStarVerifiedALL2021,EdelmannZippy2020,danielssonTotalParserCombinators2010}.
-Each new grammar formalism is extended with its own independent
-verified implementation.
+particular grammar formalism such as non-left-recursive grammars or LL(1) grammars \cite{lasserCoStarVerifiedALL2021,EdelmannZippy2020,danielssonTotalParserCombinators2010}.
+\steven{These have restrictions on left recursion or LL(1)-ness. Is it the most accurate
+  to call that a difference in formalism? It would seem that these are all CFGs
+  + conditions. Maybe these aren't the best works to
+  compare to if they're verifying particular algorithms}
+Each new grammar formalism must then be extended with its own verified
+implementation.
 
@@
 The design of \theoryabbv
-is an extension of Joachim Lambek's \emph{syntactic calculus}
-\cite{lambek58}. Lambek calculus is a grammar formalism equivalent
+is an extension of Joachim Lambek's \emph{categorial grammars}
+\cite{lambek}. Categorial grammars are a grammar formalism equivalent
 in expressive power to context-free grammars that in modern
@@
 as a shallowly embedded domain-specific langauge in the Cubical Agda
-proof assistant \cite{VezzosiMortbergAbel2019}. We have implemented many example
+proof assistant \cite{agda,cubical}. We have implemented many example
 grammars and parsers in our system including regular expressions,
@@
 the context $\internalize{\stringquote{ab}} = a : \literal a , b :
-\literal b$. In \Cref{fig:fingram}, we give a lambda term and its
-typing derivation to define a parse for a finite grammar.
+\literal b$. On the left side of \cref{fig:fingram}, we give a typing
+derivation that shows that $\inl(a \otimes b)$ defines a valid parse
+tree for the finite grammar $(\literal a \otimes \literal b) \oplus
+\literal c$. Underneath the derivation in \cref{fig:fingram}, we give
+a linear $\lambda$-term that implements the program given in the
+derivation.\max{reverse the order of these?} \steven{the order of what?}
 
 \begin{figure}
-\begin{lstlisting}
-f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
-f (a , b) = inl (a (*@$\otimes$@*) b)
-\end{lstlisting}\begin{mathpar}
+\begin{mathpar}
   \inferrule
@@
 \end{mathpar}
+\begin{lstlisting}
+f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
+f (a , b) = inl (a (*@$\otimes$@*) b)
+\end{lstlisting}
 \caption{\stringquote{ac} matches
@@
 string.
-%
-Lastly, we give a lambda term that constructs an accepting trace
-starting at the initial state for the string \stringquote{ab}.  Later
-in \Cref{sec:applications}, we will show that we can actually
-construct mutually inverse functions between the regular expression
-$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ and its
-corresponding NFA traces ($\Trace~\0$) demonstrating that the regular
-expression and the automaton capture the same language. Further since
-the functions are mutually inverse, this shows they are \emph{strongly
-equivalent} as grammars.
+
+Just as \cref{fig:kleenestarderivation} shows the \stringquote{ab} matches
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$, \cref{fig:nfaderivation} demonstrates that \stringquote{ab} is
+accepted by the NFA representation of this regular expression. In particular, we
+show that \stringquote{ab} matches the grammar $\Trace~0$ since $\0$ is the
+initial state of the NFA.\ If we construct parse transformers from
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ to $\Trace~\0$, and vice versa, we would demonstrate that the regular expression
+and the automaton capture the same language --- i.e.\ they recognize the same
+set of strings. As we will see in \cref{sec:applications}, we can indeed define parse
+transformers between regular expressions and their associated automaton.
+Moreover, we prove that these parse transformers define a bijective
+correspondence of parses over every input string. That is, in a strong sense the
+parse of a regular expression and the traces of their associated automata
+capture precisely the same data, i.e., that they are \emph{strongly equivalent} as grammar formalisms.
 
@@
   0to1 : (*@$\uparrow$@*)(Trace 1 (*@$\lto$@*) Trace 0)
-
-k : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*)  Trace 0)
-k (a , b) = 0to1 (1to1 a (1to2 b stop))
-
 \end{lstlisting}
@@
   \label{fig:exampleNFA}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}
+k : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*)  Trace 0)
+k (a , b) = 0to1 (1to1 a (1to2 b stop))
+\end{lstlisting}
+\caption{\stringquote{ab} is accepted by the NFA for
+  $({\literal a}^{*} \otimes {\literal b }) \oplus \literal c$}
+\label{fig:nfaderivation}
 \end{figure}
@@
 non-commutative linear logic.
-\cite{krishnaswami_integrating_2015,lambek58}. As in \lnld,
+\cite{krishnaswami_integrating_2015,lambekCalculus}. As in \lnld,
 \theoryabbv includes both non-linear dependent types, as well as
@@
 %
-The main point of departure from \lnld's design is that, as in Lambek calculus \cite{lambek58}, the linear
+The main point of departure from \lnld's design is that, as in Lambek calculus \cite{lambekCalculus}, the linear
 typing is \emph{non-commutative}, i.e., that exchange is not an
@@
 booleans, unit and empty types, and we assume function
-extensionality\cite{Hofmann_1997}. We present the
+extensionality\cite{standard-type-theory-reference}. We present the
 other non-linear type constructions in
-\Cref{fig:non-linear-types}. First, we include universe types $U$ of
+\Cref{fig:non-linear-stuff}. First, we include universe types $U$ of
 small non-linear types and $L$ of linear types. These are defined as
@@
 non-linear type is small and define the universe to internalize
-precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. These universe
+precisely this judgment \cite{universesAlaCoquand}. These universe
 types are needed so that we can define types by recursion on natural
@@
 linear logic or the persistence modality $\square$ of separation logic
-\cite{girard_linear_1987,jung_higher-order_2016}.
+\cite{linearlogic,persistencemodalityseparationlogic}.
 
@@
 \begin{figure}
-  \begin{footnotesize}
   \begin{mathpar}
@@
   \end{mathpar}
-  \end{footnotesize}
   \caption{Non-linear types (selection)}
@@
 
-We give an overview of the linear types and terms in
-\Cref{fig:linear-types-and-terms}. The equational theory for these
-types is straightforward $\beta\eta$ equivalence and included in the
-appendix. First, the linear variable rule says that a linear variable
-can be used if it is the \emph{only} variable in the context.
+We give an overview of the linear types in \Cref{fig:linear-types} and
+terms in \Cref{fig:linear-terms}. The equational theory for these
+types is straightforward and included in the appendix. First, the
+linear variable rule says that a linear variable can be used if it is
+the \emph{only} variable in the context.
 
@@
 specified is given by a non-linear term. The rules for indexed
-$\oplus$ are analogous to a ``\emph{weak}'' $\Sigma$ type: it has an
+$\oplus$ are analogous to a \emph{weak} $\Sigma$ type: it has an
 injection introduction rule $\sigma$, but its elimination rule is
-given by \emph{pattern matching} rather than first and second
-projections. We can define the more typical nullary and binary
-versions of these connectives by using indexing over the empty and
-boolean type respectively. We will freely use $0$ to refer to this
+given by \emph{pattern matching} rather than projections
+\cite{weakSigmatype}. We can define the more typical nullary and
+binary versions of these connectives by using indexing over the empty
+and boolean type respectively. We will freely use $0$ to refer to this
 empty disjunction and $\top$ to refer to the empty conjunction, and
@@
 
+\pedro{imo, since we are providing an Agda implementation, we can
+  be looser with the explanation of the meaning of the disjointness axioms.
+  As it stands, it's a bit hard do follow along. Is there a simpler example
+  we could point to?}
+
 \begin{figure}
-  \begin{footnotesize}
   \begin{mathpar}
@@
   \end{mathpar}
+  \caption{Linear Types (selection)}
+\end{figure}
+
+\begin{figure}
   \begin{mathpar}
@@
   \end{mathpar}
-  \end{footnotesize}
-  \caption{Linear Types and Terms (selection)}
-  \label{fig:linear-types-and-terms}
+  \caption{Linear terms (selection)}
+  \label{fig:linsyntax}
 \end{figure}
@@
 prior work on inductive types
-\cite{nakov_quantitative_2022,altenkirch_indexed_2015}. The syntax is
+\cite{quantitativePolynomialFunctors,indexedContainers}. The syntax is
 given in \Cref{fig:iilt}. First, we add a non-linear type $\SPF\,X$ of
@@
 dependent \emph{elimination} principle, which can be used to
-implemetns functions by recursion as well as proofs by
-induction. However since linear types do not support dependency on
+implement functions by recursion as well as proofs by
+induction. Unfortunately, since linear types do not support dependency on
 linear types, we cannot directly adapt this approach. However, if we
@@
 equality by induction. That is, if our goal is to prove two functions
-$f,g : \uparrow(\mu F\,x \lto A\,x)$ are equal, it suffices to
+$f,g : \uparrow(\mu F\,x \lto A\,x)$ equal, it suffices to
 implement a function $\mathsf{ind} : \uparrow(\mu F\,x \lto \equalizer
@@
 the $\beta\eta$ principles for equalizers and inductive types, and
-this is how our most complex inductive proofs are implemented in Agda
+this is how our most complex inductive proofs are implemented in the Agda
 formalization.
+\pedro{I don't like the inductive proof explanation above, but I
+  don't know how I would phrase it}
 
@@
 and define a type $\String$ as the Kleene star of $\Char$, i.e. as an
-inductive linear types. Then we add a function $\mathsf{read} :
+inductive linear type.
+\pedro{Should we reference again figure 2 (the inductive type definition of the Kleene star)?}
+Then we add a function $\mathsf{read} :
 \ltonl{(\top\lto \String)}$ that intuitively ``reads'' the input
@@
 add an axiom that $\lambda s. \mathsf{read}(\top) \equiv \lambda
-s. s$, i.e., that if we have a string, but then throw it away and read
+s. s$, \pedro{overloaded notation: $\top$ and type and $\top$ as term}
+i.e., that if we have a string, but then throw it away and read
 it from the input, then in fact that is equivalent to the string we
@@
 
-%% \paragraph{Weak and Strong Equivalence}
+\paragraph{Weak and Strong Equivalence}
 In the theory of formal grammars, there are two different notions of
@@
 these notions of equivalence in \theoryabbv.
+
 \begin{definition}
@@
 
-%% That is, grammars are weakly equivalent if we can construct a parse
-%% tree of $B$ from a parse tree of $A$ and vice-versa, whereas they are
-%% strongly equivalent if these processes are inverse to each other. The
-%% definition of retract is a useful intermediate notion. In retrospect,
-%% our axiom about $\String$ says that $\String$ is a retract of $\top$.
+That is, grammars are weakly equivalent if we can construct a parse
+tree of $B$ from a parse tree of $A$ and vice-versa, whereas they are
+strongly equivalent if these processes are inverse to each other. The
+definition of retract is a useful intermediate notion. In retrospect,
+our axiom about $\String$ says that $\String$ is a retract of $\top$.
 
@@
 
-%% \paragraph{Ambiguity}
+\paragraph{Ambiguity}
 A formal grammar $\A$ is ambiguous if there are multiple parse trees
@@
 \emph{proposition} in the terminology of homotopy type
-theory\cite{hottbook}. The most basic unambiguous types are $\top$ and
+theory\cite{hott}. The most basic unambiguous types are $\top$ and
 $0$, and in a system of classical logic all unambiguous types would
@@
 
-%% \paragraph{Partial and Total Parsers}
+\paragraph{Partial and Total Parsers}
 
@@
 interest. Instead we might try to write a partial function as a
-$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option''
-monad. This allows for the possibility that the input string doesn't
-parse, but is far too weak as a specification: we can trivially
-implement a parser for any type by always returning $\inr$. The
-correct notion of a parser should be one that allows for failure, but
-only in the case that a parse cannot be constructed.
+$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option'' monad
+\cite{maybeMonad}. This allows for the possibility that the input
+string doesn't parse, but is far too weak as a specification: we can
+trivially implement a parser for any type by always returning
+$\inr$. The correct notion of a parser should be one that allows for
+failure, but only in the case that a parse cannot be constructed.
 \begin{definition}
@@
 $\otimes$, and Kleene star.  In \cref{sec:type-theory-examples}, we
-saw one particular NFA corresponding and its type of traces. More
+saw one particular NFA corresponding \pedro{trailing sentence?} and its type of traces. More
 generally we define the linear type of traces as in
@@
 star is replaced by an arbitrary fixed point
-operation\cite{leis_towards_1992}.
+operation\cite{muregular}.
 
@@
 The Dyck language is an example of an $\textrm{LL}(0)$ language, one
-that can be parsed top-down with no
-lookahead\cite{rosenkrantz_properties_1970}. This means we can
-implement it simply as an \emph{infinite} state deterministic
+that can be parsed top-down with no lookahead\cite{ll0}. This means we
+can implement it simply as an \emph{infinite} state deterministic
 automaton, in \Cref{fig:DyckAutomaton}. Here the state is a ``stack''
@@
 $parse_{M}$ and $print_{M}$ for this automaton can be defined
-analogously to the parser and printer for DFAs, and so $\LinSigTy {s}
-{M.states} {\LinSigTy {b}{Bool} {Trace_{M}~s~b}}$ is likewise
+analogously to \cref{fig:dfaparser,fig:printdfa}, and so $\LinSigTy
+{s} {M.states} {\LinSigTy {b}{Bool} {Trace_{M}~s~b}}$ is likewise
 unambiguous.
@@
 grammars and parse transformers, we will now define a
-\emph{denotational semantics} that makes this mathematically precise
-by defining a notion of formal grammar and parse transformer and
-showing that our type theory can be soundly interpreted in this
-model. We then discuss how this denotational semantics provides the
-basis for our prototype implementation in Agda.
+\emph{denotational semantics} that makes this mathematically
+precise. To interpret our types of codes we assume a universe of
+\emph{small} sets.
 
@@
 
-The most common definition of a formal grammar is as generative
-grammars, defined by a set of non-terminals, a specified start symbol
-and set of production rules. We instead use a more abstract
-formulation that is closer in spirit to the standard definition of a
-formal \emph{language}\cite{elliottSymbolicAutomaticDifferentiation2021}:
+First, we need to establish a precise definition of a formal grammar
+that accounts for the possibility of ambiguity. The most common
+definition of a formal grammar is as generative grammars, defined by
+a set of non-terminals, with a specified start symbol and set of
+production rules. We instead use a more abstract formulation that is
+closer in spirit to the standard definition of a formal language:
+
 \begin{definition}
   A \emph{formal language} $L$ is a function from strings to propositions.
-  A (small) \emph{formal grammar} $A$ is a function from strings to (small) sets.
+  A \emph{formal grammar} $A$ is a function from strings to small sets.
 \end{definition}
+
 We think of the grammar $A$ as taking a string to the set of all parse
@@
 tree, but \emph{which} parse tree it has, i.e., what the structure of
-the element of $A(w)$ is.  To interpret our universes $U , L$ we
-assume we have a universe of \emph{small} sets. In the remainder, all
-formal grammars are assumed to be small.
-
-We then interpret linear \emph{terms} as \emph{parse transformers}:
+the element of $A(w)$ is.
+
+The linear types of \theoryabbv can all be interpreted as formal
+grammars, but what about the linear terms? These can be interpreted as
+\emph{parse transformers}, which we now define.
 \begin{definition}
+  Let $L_1,L_2$ be formal languages. Then we say $L_1\subseteq L_2$ if
+  for every string $w$, $L_1(w)\Rightarrow L_2(w)$.
+
   Let $A_1,A_2$ be formal grammars. Then a parse transformer $f$ from
@@
 \end{definition}
-Just as formal grammars generalize formal languages, parse
-transformers generalize formal language inclusion: if $A_1(w), A_2(w)$
-are all subsets of a one-element set, then a parse transformer is
-equivalent to showing that $A_1(w) \subseteq A_2(w)$. In our
-denotational semantics, linear terms will be interpreted as such parse
-transformers, and the notions of unambiguous grammar, parsers,
-disjointness, etc, introduced in \Cref{sec:applications} can be
-verified to correspond to their intended meanings under this
-interpretation.
+\pedro{Do we actually need a definition of formal language morphism?
+In case we do, it should probably be its own definition.}
+First, note that parse transformers generalize language inclusion: if
+$A_1(w), A_2(w)$ are all subsets of a one-element set, then a parse
+transformer is equivalent to showing that $A_1(w) \subseteq A_2(w)$.
+In our denotational semantics, linear terms will be interpreted as
+such parse transformers, and the notions of unambiguous grammar,
+parseability, etc, introduced in \Cref{sec:concepts} can be verified
+to correspond to their intended meanings under this interpretation.
 
@@
 structures to model the linear types, terms and equalities in
-\theoryabbv. These categorical properties are precisely what is
-required to interpret all of the linear type and term constructors.
+\theoryabbv.
 %% \begin{theorem}
@@
 \subsection{Semantics}
-
-We now define our denotational semantics.
+\steven{Add definitions of kleene,lambek,and chomsky categories here?}
+
+We can now define our denotational semantics.
 \begin{definition}[Grammar Semantics]
@@
   \item If $\Gamma \vdash X \isSmall$, then $\sem X \gamma$ is a small set.
-  \item If $\Gamma \vdash X \equiv X'$ then for every $\gamma$, $\sem{X}\gamma = \sem{X'}\gamma$
-  \item If $\Gamma \vdash A \equiv A'$ then for every $\gamma$, $\sem{A}\gamma = \sem{A'}\gamma$
-  \item If $\Gamma \vdash M \equiv M' : X$ then for every $\gamma$, $\sem{M}\gamma = \sem{M'}\gamma$
-  \item If $\Gamma;\Delta \vdash e \equiv e' : A$ then for every $\gamma$, $\sem{e}\gamma = \sem{e'}\gamma$
+  \item If $\Gamma \vdash X \cong X'$ then for every $\gamma$, $\sem{X}\gamma = \sem{X'}\gamma$
+  \item If $\Gamma \vdash A \cong A'$ then for every $\gamma$, $\sem{A}\gamma = \sem{A'}\gamma$
+  \item If $\Gamma \vdash M \cong M' : X$ then for every $\gamma$, $\sem{M}\gamma = \sem{M'}\gamma$
+  \item If $\Gamma;\Delta \vdash e \cong e' : A$ then for every $\gamma$, $\sem{e}\gamma = \sem{e'}\gamma$
   \end{enumerate}
@@
 
-The interpretation of dependent types as sets is standard
-\cite{Hofmann_1997}. We present the concrete descriptions of the
-semantics of linear types, as well as our non-standard non-linear
-types in \Cref{fig:semantics}. The grammar for a literal $c$ has a
-single parse precisely when the input string consists of the single
-character. The grammar for the unit similarly has a single parse for
-the empty string.  A parse of the tensor product $A \otimes B$
-consists of a \emph{splitting} of the empty string into a prefix $w_1$
-and suffix $w_2$ along with an $A$ parse of $w_1$ and $B$ parse of
-$w_2$. A parse of $\bigoplus_{x:X} A$ is a pair of an element of the
-set $X$ and a parse of $A(x)$, while dually a parse of $\bigwith_{x:X}
-A$ is a \emph{function} taking any $x:X$ to a parse of $A(x)$. A
-$w$-parse of $A \lto B$ is a function that takes an $A$ parse of some
-other string $w'$ to a $B$ parse of $ww'$, and $B \tol A$ is the same
-except the $B$ parse is for the reversed concatenation $w'w$.
+For reasons of brevity, we present only the cases related to the
+linear types in \Cref{fig:semantics}. The full details are in the
+appendix.\max{TODO: put the full details in the appendix}.  The
+grammar for a literal $c$ has a single parse precisely when the input
+string consists of the single character. The grammar for the unit
+similarly has a single parse for the empty string.  A parse of the
+tensor product $A \otimes B$ consists of a \emph{splitting} of the
+empty string into a prefix $w_1$ and suffix $w_2$ along with an $A$
+parse of $w_1$ and $B$ parse of $w_2$. A parse of $\bigoplus_{x:X} A$
+is a pair of an element of the set $X$ and a parse of $A(x)$, while
+dually a parse of $\bigwith_{x:X} A$ is a \emph{function} taking any
+$x:X$ to a parse of $A(x)$. A $w$-parse of $A \lto B$ is a function
+that takes an $A$ parse of some other string $w'$ to a $B$ parse of
+$ww'$, and $B \tol A$ is the same except the $B$ parse is for the
+reversed concatenation $w'w$.
 %
@@
 Next, a parse in the equalizer $\equalizer{a}{f}{g}$ is defined as
-a parse in $\sem{A}$ that is mapped to the same parse by the parse transformers $\sem{f}$ and
-$\sem{g}$.
-%
+a parse in $\sem{A}$ that is mapped to the same parse by $f$ and
+$g$. Recall that $\sem{f}\gamma$ and $\sem{g}$ are elements of
+$\sem{\uparrow(A \lto B)}$, i.e., parse transformers.
+
 The universe $L$ of linear types is interpreted as the set of all
-small grammars.
+(small) grammars.
+
 \begin{figure}
@@
     \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
-    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma)\\
+    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:indexed-container}}\\
   \end{align*}
   \caption{Grammar Semantics}
-  \label{fig:semantics}
 \end{figure}
@@
 functor} on the category of sets, also sometimes called an
-\emph{indexed container} \cite{gambino_wellfounded_2004,altenkirch_indexed_2015}.
+\emph{indexed container} \cite{dep-poly,indexed-container}.
 \begin{definition}
@@
   from $I$ to $O$ consists of a set of shapes $S$, a set of positions
-  $P$ and functions $f : P \to I$, $g : P \to S$ and $h : S \to O$. The \emph{extension}
+  $P$ and functions $I \leftarrow P \to S \to O$. The \emph{extension}
   of a polynomial is a functor $\Set/I \to \Set/O$ defined as the
@@
   are, respectively, the right and left adjoint of their pullback functors
-  $g^*$ and $h^*$. A dependent polynomial functor from $I$ to $O$ is a functor $\Set/I
+  $g^*$ and $h^*$.
+
+  A dependent polynomial functor from $I$ to $O$ is a functor $\Set/I
   \to \Set/O$ that is nautrally isomorphic to the extension of a
@@
 \end{definition}
-With this interpretation of $F$ as a polynomial functor, $\el(F)$ and
-$\map(F)$ are just interpreted as the action of the functor on objects
-and morphisms, respectively. We interpret the constructors
-$\mathsf{K},\Var$, etc. on functors in the obvious way that matches
-the definitional behavior of $\el$ and $\map$. The non-trivial part of
-the construction is verifying that such such constructions are closed
-under being polynomial. The details are tedious but straightforward
-extension of prior work on dependent polynomials and indexed
-containers and we have verified the construction in Agda.
-
-We use dependent polynomials functors on sets as these are guaranteed
-to have initial algebras. Further, these initial algebras are readily
-constructed in our Agda implementation as an inductive type of $IW$
-trees which are already available in the cubical library of Agda
-\cite{The_Agda_Community_Cubical_Agda_Library_2024}.  Then an element $F \in \sem{X \to
-  \SPF\,X}\gamma$ is an $\sem{X}\gamma$-indexed family of polynomial
-functors from $\sem{X}\gamma\times \Sigma^*$ to $\Sigma^*$, and taking
-the product of these constructs a polynomial functor from
-$\sem{X}\gamma\times \Sigma^*$ to itself. Then $\sem{\mu\,F}\gamma$ is
-defined to be the initial algebra of this functor, and the initial
-algebra structure is used to interpret $\roll, \fold$ and the
-corresponding axioms.
-
-The remaining details of the interpretation of linear terms as parse
-transformers and verification of the equational axioms is a relatively
-straightforward extension of existing semantics of linear logic in
-monoidal categories, and is included in the appendix\cite{seely89}.
-
-\subsection{Agda Implementation}
-
-This denotational semantics in grammars and parse transformers serves
-as the basis for our prototype implementation of \theoryabbv in
-cubical Agda. The implementation is a shallow embedding, meaning that
-rather than formalizing a syntax of \theoryabbv types and terms, we
-work directly with Agda types and a definition of a formal grammar as
-a function $\String \to \Set$. Then we implement each of the type and
-term constructors of \theoryabbv as combinators on formal grammars or
-parse transformers, and use cubical Agda's equality type to model the
-term equalities. Cubical Agda is convenient for this purpose as it has
-built-in support for function extensionality which we use
-extensively. Axioms such as distributivity of $\oplus/\&$ and
-disjointness of constructors are then provable directly in Agda, and
-we are careful to only construct grammars, terms and proofs using
-constructs that are possible in \theoryabbv. The main difference
-between our shallow embedding and \theoryabbv is that our linear terms
-are written in a combinator-style, without being able to use named
-variables in linear terms. A benefit of this shallow embedding is that
-the parsers are immediately available to a larger Agda development, as
-they are just normal Agda code. In future work we will look to
-implement a type checker for a syntax closer to the presentation in
-this paper, but with the goal of still making it integrate with an
-existing proof assistant.
+With this interpretation, $\el(F)$ and $\map(F)$ are just interpreted
+as the action of the functor on objects and morphisms,
+respectively. We interpret the constructors $\mathsf{K},\Var$, etc. on
+functors in the obvious way that matches the definitional behavior of
+$\el$ and $\map$. The non-trivial part of the construction is
+verifying that such functors are constructing corresponding dependent
+polynomials. The details are tedious but straightforward extension of
+prior work on dependent polynomials and indexed containers and we have
+verified the construction in Agda.
+
+The reason we use dependent polynomials functors on sets is that it is
+well-known that their initial algebras exist
+\cite{initial-algebras-of-dep-poly}. Additionally, initial algebras
+for dependent polynomials are straightforwardly constructed in Agda as
+an inductive type of ``IW'' trees, which is available as a
+construction in the cubical library of Agda \cite{iw,cubical-library}.
+\begin{theorem}
+  Every dependent polynomial functor from $I$ to $I$ has an initial
+  algebra.
+\end{theorem}
+Then an element $F \in \sem{X \to \SPF\,X}\gamma$ is an
+$\sem{X}\gamma$-indexed family of polynomial functors from
+$\sem{X}\gamma\times \Sigma^*$ to $\Sigma^*$, and taking the product
+of these constructs a polynomial functor from $\sem{X}\gamma\times
+\Sigma^*$ to itself. Then $\sem{\mu\,F}\gamma$ is defined to be the
+initial algebra of this functor, and the initial algebra structure is
+used to interpret $\roll, \fold$ and the corresponding axioms.
 
@@
 \subsection{Typed Approaches to Grammars}
+
+\steven{Talk about Conal Elliott's Automatic and Symbolic differentiation paper}
+\steven{Talk briefly about this recent arxiv paper? https://arxiv.org/abs/2411.06358}
+
 \subsubsection{Lambek Calculus and Categorial Grammar}
-Lambek's original syntactic calculus \cite{lambek58} describes a logical system
-for linguistic derivations, and it can be given semantics inside of a non-commutative biclosed monoidal category
-\cite{lambek1988categorial}. This led to many uses of non-commutative linear logic and lambda calculi in
-linguistics \cite{buszkowskiTypeLogicsGrammar2003} --- including mechanized
-categorial grammar parsers \cite{Guillaume2024,ranta-2011}. This style of grammar
-formalism has gone by the names Lambek calculus or \emph{categorial grammar},
-and it is equal in expressivity to context-free grammars.
-The existing works on categorial grammar
-frequently present logical operations on linear predicates rather than the term
-language given in this paper.
-Additionally, much of the existing work is
-oriented toward computational linguistics, giving functorial semantics to natural language in
-the tradition of Montague \cite{Montague1974-MONFPS}.
-
-The most similar prior work to our own is Luo's Lambek calculus with
-dependent types \cite{luo}. Their design differs from our own in
-allowing linear types to depend on other linear types and supporting
-directed $\Pi$ and $\Sigma$ types that have no analog in our
-system. They do not provide a semantics for these connectives, and it
-is unclear how to interpret their connectives in our grammar
-semantics. Further, we provide several examples showing that
-\theoryabbv is a practical system for describing grammar formalisms
-and parsers, and it is unclear if these could be implemented in their calculus.
-
-%% \subsubsection{Typed Grammars}
-The usage of a simple type system to reason about regular expressions up to weak
-equivalence was
-introduced by Frisch and Cardelli \cite{frischCardelli}. Hengelin and Nielsen
-\cite{henglein_regular_2011} later adapt this approach by providing a type
-system for regular expressions whose semantics were in one-to-one correspondence
-with the set of parse trees.
-Elliott \cite{elliottSymbolicAutomaticDifferentiation2021} gives a similar
-semantic interpretation of regular grammars as type-level predicates on strings.
-The type system of \theoryabbv extends these semantic interpretations of to a broader class of grammars and also gives a formal syntax and equational theory for the parse transformers that these prior works lack.
-
-%% Krishnaswami and Yallop \cite{krishnaswami_typed_2019} provide a type system for
-%% demonstrating that carries proves of the LL(1) condition for context-free
-%% grammars. Despite being a type system over formal grammars, their interpretation
-%% of types is very different from the semantics of \theoryabbv.
+
+\steven{Note that lambek calculus and its extensions don't have inductives, or
+  even kleene star as far as I can tell. }
+
+\steven{Choppy notes about lambek calc:
+In 1958 \cite{lambek1958mathematics}, Joachim Lambek introduced his syntactic
+calculus as a logical system for linguistic derivations. In fact, one may view
+the subtheory of our calculus generated over $\otimes, \lto, \tol$ to be a proof
+relevant presentation of Lambek's original system.
+
+Lambek later explicitly connects his syntactic calculus to the structure of a
+non-commutative biclosed monoidal category \cite{lambek1988categorial}. There is
+a rich line of subsequent research after these revelations by Lambek, including
+the study of pregroup grammars
+\cite{coeckeMathematicalFoundationsCompositional2010}, the presentation of
+abstract categorial grammars \cite{degrooteAbstractCategorialParsing2015}, and
+even Lambek calculus with dependent types \cite{luoLambekCalculusDependent}. As
+far as the authors can tell, these works seem to be primarily interested in natural
+language semantics in the style of Montague grammars, rather than parsing formal
+languages. That is, they seem to try to give a functorial semantic
+interpretation of natural language.
+
+There seem to be some implementations from the abstract categorial grammar
+toolkit, but again they seem natural language focused and not up to the task of
+verified formal language parsing.
+
+Also mention here that Luo's work \cite{luoLambekCalculusDependent} seems similar at first
+glance, but isn't as similar as it might look because the judgments and
+dependent types seem to
+take on a different interpretation of ours. Say something here like, ``as we
+understand it, it differs in x,y,z ways''
+}
+
+\subsubsection{Other typed approaches}
+\steven{this subsection should be renamed}
+The usage of a simple type system to reason about regular expressions was
+introduced by Frisch and Cardelli in 2004 \cite{frischCardelli}, and later expanded
+by Hengelin and Nielsen in 2011 \cite{henglein_regular_2011} to handle proofs of regular expression
+containment.
+
+\steven{Mention Neel's typed algebraic approach paper here. Say it is similar at
+  first glance, as its parsing and it has types, but that he actually is representing
+  a system for proving the LL(1) condition }
 
@@
 
-%% \subsubsection{Kleene Algebra}
-%% Since the early works in the theory of formal languages, Kleene
-%% algebras have played an important role in its development. They
-%% generalize the operations known from regular languages by introducing
-%% operations generalizing language composition, language union and the
-%% Kleene star.  More generally, they are defined as inequational theory
-%% where the inequality is meant to capture language containment.  This
-%% theory is extremely successful, having found applications in concurrency theory
-%% \cite{hoare2009concurrent}, software-defined networks \cite{anderson2014netkat},
-%% theory of programming languages \cite{angus2001kleene}, compiler optimizations
-%% \cite{kozen2000certification} and more.
-
-%% A frequently fruitful research direction is exploring varying
-%% extensions of Kleene algebras, Kleene algebra with tests (KAT) being
-%% one of the most notable ones \cite{kozen1997kleene}. Our approach is
-%% radically different from most extensions, which usually aim at
-%% modifying or adding new operations to Kleene algebras, but still
-%% keeping it as an inequational theory. By adopting a category-theoretic
-%% treatment and allowing the ``order structure'' to encode more
-%% information than merely inequalities, we were able to extend Kleene
-%% algebra to reason about parsing as well.
-
-%%    - verified imperative implementation: Lambek logic ala separation
-%%      logic?
-\subsection{Relation to Separation Logic}
-
-\theoryabbv is similar in spirit to separation logic
-\cite{reynolds_separation_2002}. Semantically, they are closely
-related: linear types in \theoryabbv denote families of sets indexed
-by a monoid of strings, whereas separation logic formulae typically
-denote families of predicates indexed by an ordered partial
-commutative monoid of worlds \cite{jung_higher-order_2016}. The
-monoidal structure in both cases is are instances of the
-category-theoretic notion of Day convolution monoidal structure
-\cite{day1970construction}. From a separation-logic perspective, our
-notion of memory is very primitive: a memory shape is just a string of
-characters and the state of the memory is never allowed to evolve.
-
-%% Syntactically, our type theory can be viewed as a kind of ``separation
-%% type theory'' in that we care not just about provability of formulae
-%% but the constructive content of our linear terms, as these act as
-%% parsers. The design of \theoryabbv is based on the $\textrm{LNL}_D$
-%% calculus which explicitly was designed to be a dependent
-%% type-theoretic version of separation logic. Many concepts of
-%% separation logic then have direct analogues in our system. The
-%% characters $c$ are analogous to the ``points-to'' formulae $l \mapsto
-%% x$ in separation logic. Our $\otimes,\lto,\tol$ are analogous to the
-%% separating conjunction and magic-wand. The analogue of the persistence
-%% modality $\square$ of separation logic is $\bigoplus_{\_:\uparrow A} I$,
-%% which ``zeros out'' all but the parses from the empty string. If we
-%% think of the portions of the string we are parsing as our notion of
-%% resource, then these make sense as the ``resource-free'' elements of
-%% the type $A$.
-
-This semantic connection to separation logic suggests an avenue of
-future work: to develop a program logic based on non-commutative
-separation logic for verifying imperative implementations of
-parsers. This could be implemented by modifying an existing separation
-logic implementation or embedding the logic within \theoryabbv.
+\subsubsection{Kleene Algebra}
+Since the early works in the theory of formal languages, Kleene
+algebras have played an important role in its development. They
+generalize the operations known from regular languages by introducing
+operations generalizing language composition, language union and the
+Kleene star.  More generally, they are defined as inequational theory
+where the inequality is meant to capture language containment.  This
+theory is extremely successful, having found applications in concurrency theory
+\cite{hoare2000concurrent}, software-defined networks \cite{anderson2014netkat},
+theory of programming languages \cite{angus2001kleene}, compiler optimizations
+\cite{kozen2000certification} and more.
+
+A frequently fruitful research direction is exploring varying
+extensions of Kleene algebras, Kleene algebra with tests (KAT) being
+one of the most notable ones \cite{kozen1997kleene}. Our approach is radically different from
+most extensions, which usually aim at modifying or adding new
+operations to Kleene algebras, but still keeping it as an inequational
+theory. By adopting a category-theoretic treatment and allowing the
+``order structure'' to encode more information than merely
+inequalities, we were able to extend Kleene algebra to reason about
+parsing as well.
 
@@
 
+%%    - verified imperative implementation: Lambek logic ala separation
+%%      logic?
+\subsection{Relation to Separation Logic}
+
+\theoryabbv is similar in spirit to separation logic. Semantically,
+they are closely related: linear types in \theoryabbv denote families
+of sets indexed by a monoid of strings, whereas separation logic
+formulae typically denote families of predicates indexed by an ordered
+partial commutative monoid of worlds. The monoidal structure in both
+cases is are instances of the category-theoretic notion of Day
+convolution monoidal structure. From a separation-logic perspective,
+our notion of memory is very primitive: a memory shape is just a
+string of characters and the state of the memory is never allowed to
+evolve.
+
+Syntactically, our type theory can be viewed as a kind of ``separation
+type theory'' in that we care not just about provability of formulae
+but the constructive content of our linear terms, as these act as
+parsers. The design of \theoryabbv is based on the $\textrm{LNL}_D$
+calculus which explicitly was designed to be a dependent
+type-theoretic version of separation logic. Many concepts of
+separation logic then have direct analogues in our system. The
+characters $c$ are analogous to the ``points-to'' formulae $l \mapsto
+x$ in separation logic. Our $\otimes,\lto,\tol$ are analogous to the
+separating conjunction and magic-wand. The analogue of the persistence
+modality $\square$ of separation logic is $\bigoplus_{\_:\uparrow A} I$,
+which ``zeros out'' all but the parses from the empty string. If we
+think of the portions of the string we are parsing as our notion of
+resource, then these make sense as the ``resource-free'' elements of
+the type $A$.
+
+This semantic connection to separation logic also suggests an avenue
+of future work: rather than working with a dependent type theory, we
+could instead make a program logic based on non-commutative separation
+logic for verifying imperative implementations of parsers. This could
+be accomplished by modifying an existing separation logic
+implementation or embedding the logic within \theoryabbv.
+
 \subsection{Implementation}
-
-Our Agda prototype implementation serves as a usefule proof of concept
-for showing what can be implemented in \theoryabbv, but it has
-downsides we aim to address in future work. Firstly, it would be
-preferable to work with the more intuitive type theoretic syntax we
-have used in this work, rather than the combinator-style our shallow
-embedding requires. Additionally, Agda itself does not have a
-high-performance implementation, and so the parsers we implement in
-Agda do not have competitive performance to industry parser
-generators. In future work we aim to study if we can embed a proof of
-the correctness of a parser generator that produces imperative
-programs, and to what extent the correctness of those imperative
-programs can be proven within \theoryabbv.
-
-\subsection{Type Checking and Semantic Analysis}
+\steven{Moving away from a prototype implementation of \theoryabbv}
+\steven{Mention performance}
+
+\subsection{Type Checking and Elaboration}
 Our focus in this work has been on the verification of parsers for
@@
 definition of arbitrarily powerful grammars, the system could also be
-used in principle for more sophisticated semantic analysis such as
-scope checking or type checking. Alternatively, we could more directly
-encode type type systems as linear types in a modified version of
-\theoryabbv where linear types are not grammars over strings, but
-\emph{type systems} over trees. This could analogously serve as a
-framework for verified type checking and static analysis.
+used for more sophisticated semantic analysis such as scope checking
+or type checking. Alternatively, we could more directly encode type
+type systems as linear types in a modified version of \theoryabbv
+whose semantics is not in families of sets indexed by \emph{strings}
+but families of sets indexed by \emph{abstract syntax trees}.
 
@@
 
-%% \begin{figure}
-%%   \begin{mathpar}
-%%   \boxed{\Gamma \vdash M : X}
-
-%%   \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
-%%   %
-%%   \and
-%%   %
-%%   \inferrule{\Gamma \vdash M : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash M : X}
-%%   %
-%%   \and
-%%   %
-%%   \inferrule{~}{\Gamma \vdash () : 1}
-%%   %
-%%   \and
-%%   %
-%%   \inferrule{\Gamma \vdash M : X \\ \Gamma \vdash N : \subst Y M x}{\Gamma \vdash (M, N) : \SigTy x X Y}
-%%   %
-%%   \and
-%% %
-%%   \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_1\, M : X}
-%%   %
-%%   \and
-%%   %
-%%   \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_2\, M : \subst Y {\pi_1\, M} x}
-%%   \and
-%%   \inferrule{\Gamma, x : X \vdash M : Y}{\Gamma \vdash \lamb x M : \PiTy x X Y}
-%%   %
-%%   \and
-%%   %
-%%   \inferrule{\Gamma \vdash M : \PiTy x X Y \\ \Gamma \vdash N : X}{\Gamma \vdash \app M {N} : \subst Y {N} x}
-%%   %
-%%   \and
-%%   %
-%%   \inferrule{\Gamma \vdash M \equiv N : X}{\Gamma \vdash \mathsf{refl} : M =_X N}
-%%   \and
-%%   \textrm{TODO: inductive types (boolean, nat) with their eliminators allowed in }
-%%   \end{mathpar}
-%%   \caption{Intuitionistic typing}
-%%   \label{fig:inttyping}
-%% \end{figure}
+\max{TODO: add rules for non-linear types $\Pi,\Sigma$, equality etc}
+
+\begin{figure}
+  \begin{mathpar}
+  \boxed{\Gamma \vdash M : X}
+
+  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash M : X}
+  %
+  \and
+  %
+  \inferrule{~}{\Gamma \vdash () : 1}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M : X \\ \Gamma \vdash N : \subst Y M x}{\Gamma \vdash (M, N) : \SigTy x X Y}
+  %
+  \and
+%
+  \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_1\, M : X}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_2\, M : \subst Y {\pi_1\, M} x}
+  \and
+  \inferrule{\Gamma, x : X \vdash M : Y}{\Gamma \vdash \lamb x M : \PiTy x X Y}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M : \PiTy x X Y \\ \Gamma \vdash N : X}{\Gamma \vdash \app M {N} : \subst Y {N} x}
+  %
+  \and
+  %
+  \inferrule{\Gamma \vdash M \equiv N : X}{\Gamma \vdash \mathsf{refl} : M =_X N}
+  \and
+  \textrm{TODO: inductive types (boolean, nat) with their eliminators allowed in }
+  \end{mathpar}
+  \caption{Intuitionistic typing}
+  \label{fig:inttyping}
+\end{figure}
 
@@
     \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
+    %
+    \\
+
   \end{mathpar}
@@
   \begin{mathpar}
+    \boxed{\Gamma \vdash M \equiv N : X}
+
+    \inferrule{\Gamma \vdash p : M =_X N}{\Gamma \vdash M \equiv N : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \app {(\lamb x M)} {N} \equiv \subst x M {N} : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash m \equiv \lamb x {\app M x} : \PiTy x X Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_1\, (M, N) \equiv M : X}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash \pi_2\, (M , N) \equiv N : \subst x {M} Y}
+%
+    \and
+%
+    \inferrule{~}{\Gamma \vdash M \equiv (\pi_1\, M, \pi_2\, M) : \SigTy x X Y}
+%
+    \and
+%
+    % \inferrule{~}{}
+%
+    \inferrule{~}{\Gamma \vdash M \equiv N : 1}
+%
+    % \and
+%
+    % \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
+
+    \\
+
     \boxed{\Gamma ; \Delta \vdash e \equiv e' : A}
@@
 
-  %% \max{do we need identity and composition of substitutions and their functoriality?}
+  \max{do we need identity and composition of substitutions and their functoriality?}
   %% By induction we can define an identity substitution
@@
   the type theory that are parameterized by a non-linear context
-  $\Gamma$ support the admissible actions of a substitution $\gamma
-  \in \Subst(\Gamma',\Gamma)$ in \Cref{fig:non-linear-substitution}.
-  %% Further, these all satisfy identity and composition equations M[\Gamma] = M and M[\gamma][\gamma'] = M[(\gamma \circ \gamma')]
-\end{definition}
-
-\begin{figure}
+  $\Gamma$ support an admissible action of a substitution $\gamma \in
+  \Subst(\Gamma',\Gamma)$:
   \begin{mathpar}
@@
   \end{mathpar}
-  \caption{Non-linear substitution}
-  \label{fig:non-linear-substitution}
-\end{figure}
+
+  %% Further, these all satisfy identity and composition equations M[\Gamma] = M and M[\gamma][\gamma'] = M[(\gamma \circ \gamma')]
+\end{definition}
 
@@
 
-  %% \max{identity and composition?}
+  \max{identity and composition?}
 \end{definition}
@@
 \section{Denotational Semantics}
-Here we extend the denotational semantics from
-\cref{sec:semantics-and-metatheory} to cover all of \theoryname
-syntax. Here we will freely use that the category of grammars is a
-complete, co-complete biclosed monoidal category, and use categorical
-notation for the constructions in the denotational semantics. For
-example, we will use the same notation $I,\otimes,\lto,\tol$ for the
-biclosed monoidal structure of $\Grammar$ that we do for the
-corresponding syntactic notions.
+\steven{Define a Kleene, Lambek, and Chomsky category}
+
+Let $\semcat$ be a Chomsky category.
+
+\steven{Finish the denotation of lin types figure for spf, el , map , mu}
+\begin{figure}
+  \begin{align*}
+    \semg{I} &= I \\
+    \semg{A \otimes B} &= \semg{A} \otimes \semg{B} \\
+    \semg{A \lto B}  &= \semg{A} \lto \semg{B} \\
+    \semg{B \tol A} &= \semg{B} \tol \semg{A} \\
+    \semg{\bigoplus_{x:X} A} &= \coprod_{x:X} \semg{A} \\
+    \semg{\bigwith_{x:X} A} &= \prod_{x : X} \semg{A} \\
+    \semg{\uparrow A} &= Hom(I,A) \\
+    \semg{\equalizer{a:A}{f}{g}} &= Eq\left( \semg{f} , \semg{g} \right) \\
+    \sem{L}\gamma &= \semcat_0\\
+    \sem{\SPF\,X}\gamma &= \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
+    \sem{\el(F)}\gamma\,G &= \sem{F}\gamma\,G\\
+    \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
+    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:indexed-container}}\\
+  \end{align*}
+  \caption{Denotation of Linear Types}
+\end{figure}
 
@@
 \subsection{Grammar Semantics for Linear Terms}
-Here we define denotations of linear terms. Note that the denotations
-interpret typing derivations, not raw terms, as the data of how
-contexts are split is needed in order to construct the correct
-associator functions. Further, we demonstrate that the denotational
-semantics resprects the equational theory of \theoryabbv. The
-correctness of the equational theory heavily relies on the
-\emph{coherence theorem} for monoidal categories. The coherence
-theorem says that any diagram in a monoidal category constructed using
-only associators $\alpha_{A,B,C} : (A \otimes B) \otimes C \cong A
-\otimes (B \otimes C)$, unitors $\rho_A : A \otimes I \cong A$ and
-$\lambda_A : I \otimes A \cong A$ and compositions and tensor products
-of these, commutes. We call a morphism built in this way a
-\emph{generalized associator}.
+Here we define denotations of linear terms. Note that the denotations interpret
+typing derivations, not raw terms, as the data of how contexts are
+split is needed in order to construct the correct associator
+functions. Further, we demonstrate that the denotational semantics resprects the
+equational theory of \theoryabbv.
 
@@
 
+\steven{Define generalized associators}
+
 Because
@@
 \paragraph{$\bigamp$-Introduction}
-$\semg{\dlamb x e} : \semg{\Delta} \to \prod_{x : X} {\sem{A}(\gamma , x)}$ is
+$\semg{\dlamb x e} : \semg{\Delta} \to \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)}$ is
 defined by the universal property of the product
@@
   \semg{\Delta} \arrow[r , "\semg{e}"] &
-  \prod_{x : X} {\sem{A}(\gamma , x)} \arrow[r , "\pi_M"] &
+  \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)} \arrow[r , "\pi_M"] &
   \sem{A}(\gamma , M)
@@
 \paragraph{$\bigoplus$-Introduction}
-$\semg{\sigma~M~e} : \semg{\Delta} \to \coprod_{x : X} \sem{A}(\gamma, x)$
+$\semg{\sigma~M~e} : \semg{\Delta} \to \LinSigTyLimit {x} {X} \sem{A}(\gamma, x)$
 
@@
   \sem{A}(\gamma, M) \arrow[r , "i_M"] &
-  \coprod_{x : X} {\sem{A}(\gamma , x)}
+  \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)}
 \end{tikzcd}
@@
 {\semg{\Delta'_1,\Delta,\Delta'_2}} \arrow[r, "{m_{(\Delta'_1,\Delta),\Delta'_2}}"]                                                                                                           & {\semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}} \arrow[d, "{m_{\Delta'_1,\Delta}\otimes id}"]                             \\
-{\left( \semg{\Delta'_1} \otimes \coprod_{x:X} {\sem{A}(\gamma , x)} \right) \otimes \semg{\Delta'_2}} \arrow[d, "d"]                                                                 & \left( \semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2} \arrow[l, "(id \otimes \semg{e}) \otimes id"] \\
-{\coprod_{x : X} {\left( \sem{\Delta'_1}(\gamma , x) \otimes
-      \sem{A}(\gamma , x) \right) \otimes \sem{\Delta'_2}(\gamma , x)}} \arrow[r, "\coprod_{x:X}({m^{-1}_{(\Delta'_1,a:A),\Delta'_2}})"] & {\coprod_{x : X} {\sem{\Delta'_1,a : A,\Delta'_2}(\gamma , x)}} \arrow[d, "{[ \sem{e'}(\gamma , x) ]_{(x : X)}}"]     \\
+{\left( \semg{\Delta'_1} \otimes \LinSigTyLimit {x} {X} {\sem{A}(\gamma , x)} \right) \otimes \semg{\Delta'_2}} \arrow[d, "d"]                                                                 & \left( \semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2} \arrow[l, "(id \otimes \semg{e}) \otimes id"] \\
+{\LinSigTyLimit {x} {X} {\left( \sem{\Delta'_1}(\gamma , x) \otimes
+      \sem{A}(\gamma , x) \right) \otimes \sem{\Delta'_2}(\gamma , x)}} \arrow[r, "{m^{-1}_{(\Delta'_1,a:A),\Delta'_2}}"] & {\LinSigTyLimit {x} {X} {\sem{\Delta'_1,a : A,\Delta'_2}(\gamma , x)}} \arrow[d, "{[ \sem{e'}(\gamma , x) ]_{(x : X)}}"]     \\
 \semg{C}                                                                                                                                                                                      & {\sem{C}(\gamma, x)} \arrow[l]
@@
   \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}}
-  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
   \otimes \semg{\sigma~M~a'}) \otimes id \circ m \otimes id \circ m \\
-  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X}m^{-1} \circ d \circ (id
-  \otimes (i_M \circ \semg{a'})) \otimes id \circ m \otimes id \circ m \\
-  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X} m^{-1} \circ d \circ (id \otimes i_M) \otimes id \circ (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
-  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X} m^{-1} \circ i_M \circ (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
-  & = \sem{e'}(\gamma , M) \circ m^{-1} (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
-  & = \sem{e'}(\gamma , M) \tag{coherence}
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
+  \otimes i_M \circ \semg{a'}) \otimes id \circ m \otimes id \circ m \\
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
+  \otimes i_M \circ \lambda) \otimes id \circ m \otimes id \circ m
   \end{align*}
+\steven{TODO need to track subscripts on $m$ in the $\bigoplus\beta$ proof, and
+  finish this part of the above part of the proof}
 
@@
     &= \semg{e'} \circ \semg{M/x,a'/a} \tag{\cref{lem:subst}} \\
-    &= \sem{e'}(\gamma , x)
+    &= \semg{e'}
   \end{align*}
+  \steven{TODO shouldn't the semantics of $e'$ be in context $\gamma,x$ not $x$?}
 \end{proof}
@@
 \paragraph{$\bigoplus\eta$}
-It suffices to show
-\[
-  \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}} = \semg{f[c'/c]} = \semg{f}
-  \]
-First, expanding the left hand side, we have.
-\begin{proof}
-  \begin{align*}
-    \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}}
-    &= [ \semg{f[\sigma~x~a/c]} ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
-    &= [ \semg{f} \circ (id \otimes i_x) \otimes id ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
-    &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
-  \end{align*}
-Since the domain has the universal property of a coproduct (due to distributivity), to prove
-this is equal to $\semg{f}$, it is sufficient to prove they are equal
-when composed with the injections:
-
-\begin{align*}
-  \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id
-  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d  \circ (id \otimes i_y) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
-  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ i_y \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
-  &= \semg{f} \circ (id \otimes i_y) \otimes id \circ (m^{-1}) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \\
-  &= \semg{f} \circ (m^{-1}) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id \\
-  &= \semg{f} \circ (id \otimes i_y) \otimes id \tag{coherence}\\
-\end{align*}
-\end{proof}
+\steven{TODO $\eta$ for sums}
 
@@
 
+\steven{Add the arithmetic expression automaton}
+
 \end{document}

### 2024-11-15 paper/paper.tex

--- 2024-11-14 paper/paper.tex
@@
 \newif\ifdraft
-\drafttrue
+\draftfalse
 \newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven says}: #1]}\fi}
@@
 }
-\email{nvarner@umich.edu}
+\email{nmvarner@umich.edu}
 
@@
 
+\pedro{imo, since we are providing an Agda implementation, we can
+  be looser with the explanation of the meaning of the disjointness axioms.
+  As it stands, it's a bit hard do follow along. Is there a simpler example
+  we could point to?}
+
 \begin{figure}
@@
     \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \tol B \isLinTy}\and
-    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigoplus\limits_{x:X} A}\and
-    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigamp\limits_{x:X} A}\and
+    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigoplus\limits_{x:X} A \isLinTy}\and
+    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigamp\limits_{x:X} A \isLinTy}\and
     \inferrule
@@
 %
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e {e'} a : C}
+    \inferrule{\Gamma ; \Delta , a : A \vdash e : C \\ \Gamma ; \Delta' \vdash e' : A }{\Gamma; \Delta,\Delta' \vdash \app {(\lamblto a e)} {e'} \equiv \subst e {e'} a : C}
 %
@@
 %
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
+    \inferrule{\Gamma ; \Delta \vdash e : A \lto B}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
 %
@@
 %
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
+    \inferrule{\Gamma ; a : A , \Delta \vdash e : C \\ \Gamma ; \Delta' \vdash e' : A }{\Gamma; \Delta , \Delta' \vdash \app {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
 %
@@
 %
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : B \tol A}
+    \inferrule{\Gamma ; \Delta \vdash e : B \tol A}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : B \tol A}
 %
@@
 %
-    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {M} \equiv \subst {e} M x : C}
+    \inferrule{\Gamma , x : X \vdash e : A \\ \Gamma \vdash M : X}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {M} \equiv \subst {e} M x : C}
 %
@@
 %
-    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
+    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy {x} {X} {A}}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
 %
@@
 %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
+    \inferrule{\Gamma ; \Delta_1 , \Delta_2 \vdash e : C}{\Gamma; \Delta_1 , \Delta_2 \vdash \letin {()} {()} e \equiv e : C}
     %
@@
     %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} e a : C}
+    \inferrule{\Gamma ; \Delta_2 \vdash e : I \\ \Gamma ; \Delta_1 , a : A , \Delta_3 \vdash e' : C}{\Gamma; \Delta_1 , \Delta_3 \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} e a : C}
 %
@@
 %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} {e \otimes e'} e'' \equiv e'' \{ e/a , e'/b \} : C}
+    \inferrule{\Gamma ; \Delta_2 \vdash e : A \\ \Gamma ; \Delta_3 \vdash e' : B \\ \Gamma ; \Delta_1 , a : A , b : B , \Delta_4 \vdash e'' : C}{\Gamma; \Delta_1 , \Delta_2, \Delta_3 , \Delta_4 \vdash \letin {(a, b)} {(e , e')} e'' \equiv e'' \{ e/a , e'/b \} : C}
 %
@@
     %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} e c : C}
+    \inferrule{\Gamma ; \Delta_2 \vdash e : A \otimes B \\ \Gamma ; \Delta_1 , c : A \otimes B , \Delta_3 \vdash e' : C }{\Gamma; \Delta_1, \Delta_2, \Delta_3 \vdash \letin {(a , b)} e {\subst {e'} {(a, b)} c} \equiv \subst {e'} e c : C}
 %
@@
 %
-    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(M, e)} {e'} \equiv e' \{ M/x , e/a \} : C}
+    \inferrule{\Gamma \vdash M : X \\ \Gamma ; \Delta_2 \vdash e : A \\ \Gamma , x : X \vdash \Delta_1 , a : A , \Delta_3 }{\Gamma;\Delta_1 , \Delta_2 , \Delta_3 \vdash \letin {\sigma~x~a} {\sigma~M~e} {e'} \equiv e' \{ M/x , e/a \} : C}
     %
@@
     %
-    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
+    \inferrule{\Gamma ; \Delta_1, y : \LinSigTy {x} {X} {A}, \Delta_2 \vdash e' : C \\ \Gamma ; \Delta_2 \vdash e : \LinSigTy {x} {X} {A}}{\Gamma; \Delta_1 , \Delta_2 , \Delta_3 \vdash \letin {\sigma~x~a} e {\subst {e'} {\sigma~x~a} y} \equiv \subst {e'} e y : C}
     \and
@@
     \inferrule
-    {~}
+    {\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta \vdash f~e \equiv g~e}
     {\Gamma ; \Delta \vdash \equalizerpi {\equalizerin {e}} \equiv e : A}
@@
     \inferrule
-    {~}
+    {\Gamma ; \Delta \vdash e : \equalizer e f g}
     {\Gamma ; \Delta \vdash \equalizerin {\equalizerpi {e}} \equiv e :
@@
     \Subst(\Delta',\cdot) &= \{ \cdot \pipe \Delta' = \cdot \}\\
-    \Subst(\Delta',(\Delta,a:A)) &= \{ (\delta, e/a) \pipe \delta \in \Subst(\Delta_1, \Delta) \and \Delta_2 \vdash e : A \and \Delta' = \Delta_1,\Delta_2\}
+    \Subst(\Delta',(\Delta,a:A)) &= \{ (\delta, e/a) \pipe \delta \in \Subst(\Delta_1, \Delta) , \Delta_2 \vdash e : A , \Delta' = (\Delta_1,\Delta_2)\}
   \end{align*}

### 2024-11-15 paper/paper.tex
address some of pedro's comments

@@
 inductive linear type.
-\pedro{Should we reference again figure 2 (the inductive type definition of the Kleene star)?}
 Then we add a function $\mathsf{read} :
@@
 of resources, and not $\I$ which controls no resources. Further, we
-add an axiom that $\lambda s. \mathsf{read}(\top) \equiv \lambda
-s. s$, \pedro{overloaded notation: $\top$ and type and $\top$ as term}
+add an axiom that $\lambda s. \mathsf{read}(!(s)) \equiv \lambda
+s. s$ where $!$ is the unique function $\ltonl(\String\lto \top)$,
 i.e., that if we have a string, but then throw it away and read
@@
 $\otimes$, and Kleene star.  In \cref{sec:type-theory-examples}, we
-saw one particular NFA corresponding \pedro{trailing sentence?} and its type of traces. More
+saw one particular NFA and its corresponding type of traces. More
 generally we define the linear type of traces as in

### 2024-11-15 paper/paper.tex
shrink a figure to get us back under
spellcheck

@@
 To show the feasibility of our design, we have implemented \theoryabbv
-as a shallowly embedded domain-specific langauge in the Cubical Agda
+as a shallowly embedded domain-specific language in the Cubical Agda
 proof assistant \cite{VezzosiMortbergAbel2019}. We have implemented many example
@@
 $({\literal a}^{*} \otimes \literal b) \oplus \literal c$ does not give a very operational perspective of how to parse
-it. For this reason, most parsers are based on compiling a grammar to a corresponding notion of automaton, which is readily implemented. To impleemnt
+it. For this reason, most parsers are based on compiling a grammar to a corresponding notion of automaton, which is readily implemented. To implement
 these algorithms in \theoryabbv, we need a way to represent automata as types in the same way we can represent regular expressions.
@@
   $\epsilon$-closure of all the states reachable from $X$ via a transition
-  labelled with the character $\c$.
+  labeled with the character $\c$.
 
@@
 
-%% The first two equalities are by defintion. The thrid equality holds because
+%% The first two equalities are by defintion. The third equality holds because
 %% $\literal c$ is unambiguous and
@@
 their composition is defined pointwise, i.e. $(f\circ g)_w = f_w \circ g_w$.
-Furthemore, given a formal grammar $A$, its identity transformer is $id_w =
+Furthermore, given a formal grammar $A$, its identity transformer is $id_w =
 id_{A(w)}$, where $id_{A(w)}$ is the identity function on the set $A(w)$. This
@@
 \begin{figure}
+  \begin{footnotesize}
     \begin{align*}
@@
   \end{align*}
+  \end{footnotesize}
   \caption{Grammar Semantics}
@@
   \end{tikzcd}\]
-  Where $f^*$ is the pullback functor along $f$; $\Pi_g$ is the depedent
+  Where $f^*$ is the pullback functor along $f$; $\Pi_g$ is the dependent
   product operation and $\Sigma_h$ is the dependent sum operation, which
@@
   $g^*$ and $h^*$. A dependent polynomial functor from $I$ to $O$ is a functor $\Set/I
-  \to \Set/O$ that is nautrally isomorphic to the extension of a
+  \to \Set/O$ that is naturally isomorphic to the extension of a
   dependent polynomial from $I$ to $O$.
@@
 practice, parsers are combined with a \emph{semantic action} that
-emits an \emph{abstract} syntax tree that omits superflous syntactic
+emits an \emph{abstract} syntax tree that omits superfluous syntactic
 details that aren't needed in later stages of the overall program. We
@@
 
-Our Agda prototype implementation serves as a usefule proof of concept
+Our Agda prototype implementation serves as a useful proof of concept
 for showing what can be implemented in \theoryabbv, but it has
@@
 associator functions. Further, we demonstrate that the denotational
-semantics resprects the equational theory of \theoryabbv. The
+semantics respects the equational theory of \theoryabbv. The
 correctness of the equational theory heavily relies on the

### 2024-11-15 paper/paper.tex
Henglein
fixed a couple of typos

@@
 not quite appropriate
-to think of a linear term here as a parse \textit{trees}.
+to think of a linear term here as a parse \textit{tree}.
 The
@@
 Defining an indexed inductive type can be thought
-of as defining a family of mutually recursive inductive types, one for each element of the indexing type. Here $\Trace$ is 
+of as defining a family of mutually recursive inductive types, one for each element of the indexing type. Here $\Trace$ 
 uses an index $\s : \texttt{Fin 3}$ which picks out which
@@
 equivalence was
-introduced by Frisch and Cardelli \cite{frischCardelli}. Hengelin and Nielsen
+introduced by Frisch and Cardelli \cite{frischCardelli}. Henglein and Nielsen
 \cite{henglein_regular_2011} later adapt this approach by providing a type

### 2024-11-15 paper/paper.tex
add a flag for the appendix

@@
 
+\newif\ifappendix
+\appendixtrue
+
 \begin{document}
 
-\pagestyle{plain}
+%\pagestyle{plain}
 
@@
 its transition is labeled with the empty string $\epsilon$. Therefore,
-$\texttt{0to1}$ takes in a trace beginning a state $\1$ and returns a
+$\texttt{0to1}$ takes in a trace beginning at state $\1$ and returns a
 trace beginning at state $\0$ corresponding to the same underlying
@@
 corresponding NFA traces ($\Trace~\0$) demonstrating that the regular
-expression and the automaton capture the same language. Further since
+expression and the automaton capture the same language. Further, since
 the functions are mutually inverse, this shows they are \emph{strongly
@@
 
-The design of \theoryabbv is based the dependent
+The design of \theoryabbv is based on the dependent
 linear-non-linear calculus (\lnld) and Lambek calculus, also known as
@@
 The main point of departure from \lnld's design is that, as in Lambek calculus \cite{lambek58}, the linear
-typing is \emph{non-commutative}, i.e., that exchange is not an
+typing is \emph{non-commutative} --- i.e., that exchange is not an
 admissible structural rule. Furthermore, we add a general-purpose
@@
 The formation rules for the judgments of \theoryabbv are shown in
-\Cref{fig:formation}. $\Gamma$ stand for non-linear contexts, $X,Y,Z$
-stand for non-linear types, $M,N$ stand for non-linear terms, these
-act as in an ordinary dependent type theory. $\Delta$ stands for
-linear contexts, $A,B,C$ for linear types and $e,f,g$ for linear
+\Cref{fig:formation}. $\Gamma$ stands for non-linear contexts; $X,Y,Z$
+stand for non-linear types; $M,N$ stand for non-linear terms, these
+act as in an ordinary dependent type theory; $\Delta$ stands for
+linear contexts; $A,B,C$ for linear types; and, $e,f,g$ for linear
 terms. These contexts, types and terms are allowed to depend on an
@@
   \end{mathpar}
-  \caption{Formation Rules}
+  \caption{Formation rules}
   \label{fig:formation}
@@
 logic. First, we need that additive conjunction \emph{distributes}
-over additive disjunction, e.g., in the finitary case that $0 \& A
+over additive disjunction --- e.g., in the finitary case that $0 \& A
 \cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most
@@
   \end{footnotesize}
-  \caption{Linear Types and Terms (selection)}
+  \caption{Linear types and terms (selection)}
   \label{fig:linear-types-and-terms}
@@
   %% \end{footnotesize}
-  \caption{Strictly Positive Functors and Indexed Inductive Linear Types}
+  \caption{Strictly positive functors and indexed inductive linear types}
   \label{fig:iilt}
@@
 In the theory of formal grammars, there are two different notions of
-equivalence: weak generative capacity, meaning just which strings are
-accepted by the grammar, and \emph{strong} generative capacity, when
+equivalence: up to weak generative capacity, meaning just which strings are
+accepted by the grammar; and up to \emph{strong} generative capacity, when
 the parse trees of the two grammars are isomorphic
@@
   $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
-  then $\f \equiv \g$
+  then $\f \equiv \g$.
 \end{definition}
@@
   A parser for a linear type $A$ is a function $\uparrow{(\String \lto
-    A \oplus A_{\neg})}$ where $A_{\neg}$ is a linear type that is
+    A \oplus A_{\neg})}$, where $A_{\neg}$ is a linear type that is
   \emph{disjoint} from $A$ in that we can implement a function
-  $\uparrow{(A \& A_{\neg} \lto 0)}$
+  $\uparrow{(A \& A_{\neg} \lto 0)}$.
 \end{definition}
@@
 \end{lstlisting}
-\caption{Parser/Printer for DFA traces}
+\caption{Parser/printer for DFA traces}
 \label{fig:printdfa}
@@
   parse transformer.
-  Vice-versa,
 
@@
 \emph{denotational semantics} that makes this mathematically precise
-by defining a notion of formal grammar and parse transformer and
+by defining a notion of formal grammar and parse transformer then
 showing that our type theory can be soundly interpreted in this
@@
   \item If $\Gamma \vdash X \isSmall$, then $\sem X \gamma$ is a small set.
-  \item If $\Gamma \vdash X \equiv X'$ then for every $\gamma$, $\sem{X}\gamma = \sem{X'}\gamma$
-  \item If $\Gamma \vdash A \equiv A'$ then for every $\gamma$, $\sem{A}\gamma = \sem{A'}\gamma$
-  \item If $\Gamma \vdash M \equiv M' : X$ then for every $\gamma$, $\sem{M}\gamma = \sem{M'}\gamma$
-  \item If $\Gamma;\Delta \vdash e \equiv e' : A$ then for every $\gamma$, $\sem{e}\gamma = \sem{e'}\gamma$
+  \item If $\Gamma \vdash X \equiv X'$ then for every $\gamma$, $\sem{X}\gamma = \sem{X'}\gamma$.
+  \item If $\Gamma \vdash A \equiv A'$ then for every $\gamma$, $\sem{A}\gamma = \sem{A'}\gamma$.
+  \item If $\Gamma \vdash M \equiv M' : X$ then for every $\gamma$, $\sem{M}\gamma = \sem{M'}\gamma$.
+  \item If $\Gamma;\Delta \vdash e \equiv e' : A$ then for every $\gamma$, $\sem{e}\gamma = \sem{e'}\gamma$.
   \end{enumerate}
@@
 
+\ifappendix
 \appendix
@@
   \end{align*}
-  \caption{Strictly Positive Functors functorial actions}
+  \caption{Strictly positive functors functorial actions}
 \end{figure}
@@
   \end{mathpar}
-  \caption{Context Rules}
+  \caption{Context rules}
 \end{figure}
@@
   \end{mathpar}
-  \caption{Linear Terms, extended}
+  \caption{Linear terms (extended)}
 \end{figure}
@@
 \end{proof}
-
+\fi
 \end{document}

### 2024-11-15 paper/paper.tex
punctuation, captions, and typo

@@
 
-\newif\ifappendix
-\appendixtrue
-
 \begin{document}
 
-%\pagestyle{plain}
+\pagestyle{plain}
 
@@
 its transition is labeled with the empty string $\epsilon$. Therefore,
-$\texttt{0to1}$ takes in a trace beginning at state $\1$ and returns a
+$\texttt{0to1}$ takes in a trace beginning a state $\1$ and returns a
 trace beginning at state $\0$ corresponding to the same underlying
@@
 
-\ifappendix
 \appendix
@@
 \end{proof}
-\fi
+
 \end{document}

### 2025-03-17 paper/paper.tex
fixed line numbering and a typo
Artifact prep (#35)
* Add README

* Dyck example uses Automata.Deterministic

* BinOp example fixed

* Add equivalence of Term and Element

* Delete Term.Bilinear

* Dyck parser benchmarks

* Examples.Section2

* Reorganization and renamings

--- 2024-11-15 paper/paper.tex
@@
   doneBad : (*@$\uparrow$@*)(&[ n ] M (n + 1) false)
-  add : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O n b (*@$\lto$@*) M n b)
+  add : (*@$\uparrow$@*)(&[ n ] &[ b ] '+' (*@$\lto$@*) O n b (*@$\lto$@*) M n b)
   unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')'(*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) M n false)

### 2025-03-17 Everythings.hs
delete unused files + flatten directory

@@ -77,7 +77,7 @@ checkEverythings excludes dirs = do
 checkREADME :: IO ()
 checkREADME = do
   (sub_dirs', _) <- getSubDirsFiles ["."]
-  let sub_dirs = filter (\d -> d `notElem` ["Everything.hs", "_build", "README.agda", "Makefile"]) sub_dirs'
+  let sub_dirs = filter (\d -> d `notElem` ["Everything.hs", "_build", "README.agda", "Makefile", "Dockerfile"]) sub_dirs'
   imported <- getImported ["README"]
   let missing_files = fmap (\dir -> ["Everything",dir]) sub_dirs \\ imported
   if null missing_files then pure ()
@@ -115,7 +115,7 @@ helpText = unlines [
 main :: IO ()
 main = do
   all_dirs' <- filter ('.' `notElem`) <$> getDirectoryContents "."
-  let all_dirs = filter (\d -> d `notElem` ["Everything.hs", "_build", "README.agda", "Makefile"]) all_dirs'
+  let all_dirs = filter (\d -> d `notElem` ["Everything.hs", "_build", "README.agda", "Makefile", "Dockerfile"]) all_dirs'
   args <- getArgs
   case args of
     "check":dirs -> checkEverythings [] dirs

### 2025-03-19 paper/paper.tex
fix typos

@@
 not a verified parser generator that compiles regular expressions to
-deterministic finite automata, but is instead is a domain specific
+deterministic finite automata, but is instead a domain specific
 language \emph{for writing} such a verified parser generator.
@@
 a parse of $\A$ over the string $\w$ or a
-parse of $\B$ over the string $\w$. Similarly, $\w$ matches $\A \otimes \B$ if
-$\w$ can be split into two strings $\w_{\A}$ and $\w_{\B}$ that match $\A$
+parse of $\B$ over the string $\w$. Similarly, a parse of $\A \otimes \B$ for
+$\w$ is a splitting of $w$ into two strings $\w_{\A}$ and $\w_{\B}$ with
+parses for $\A$
 and $\B$, respectively.
@@
     }
-    {a : \literal a , b : \literal b \vdash a \otimes b : \literal a \otimes
+    {a : \literal a , b : \literal b \vdash (a , b) : \literal a \otimes
       \literal b}
   }
-  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a \otimes b) :
+  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a , b) :
     (\literal a \otimes \literal b) \oplus \literal c}
 \end{mathpar}
-\caption{\stringquote{ac} matches
-  $(\literal a \oplus \literal b) \otimes \literal c$}
+\caption{\stringquote{ab} is parsed by $(\literal a \oplus \literal b) \otimes \literal c$
+}
 \label{fig:fingram}
@@
 \cref{fig:kleenestarderivation} gives a derivation that shows
-\stringquote{ab} matches the regular expression $({\literal a}^{*}
+\stringquote{ab} is parsed by the regular expression $({\literal a}^{*}
 \otimes \literal b) \oplus \literal c$. The leaves of the proof tree
@@
   }
-  {a : {\literal a} , b : {\literal b} \vdash (\cons~a~\nil) \otimes b : {\literal a}^{*} \otimes {\literal b}}
+  {a : {\literal a} , b : {\literal b} \vdash (\cons~a~\nil , b) : {\literal a}^{*} \otimes {\literal b}}
 }
-{a : \literal a , b : \literal b \vdash \texttt{g} := \inl ((\cons~a~\nil)
-  \otimes b) : ({\literal a}^{*} \otimes \literal b) \oplus \literal c}
+{a : \literal a , b : \literal b \vdash \texttt{g} := \inl (\cons~a~\nil , b) : ({\literal a}^{*} \otimes \literal b) \oplus \literal c}
 \end{mathpar}
@@
 g : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*) ('a'(*@$^*$@*) (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
-g (a , b) = inl (cons a' nil (*@$\otimes$@*) b')
+g (a , b) = inl (cons a nil (*@$\otimes$@*) b)
 \end{lstlisting}
-\caption{\stringquote{ab} matches
+\caption{\stringquote{ab} is parsed by
   $({\literal a}^{*} \otimes \literal b) \oplus \literal c$}
@@
         {~}
-        {a_1 : \A, a_2 : \A , a_{*} : \A^* \vdash \cons (a_1 , \cons(a_2 , a_{*})) : \A^*}
+        {a_1 : \A, a_2 : \A , a_{*} : \A^* \vdash \cons~a_1 (\cons~a_2~a_{*}) : \A^*}
       }
-      {a' : \A \otimes \A , a_{*} : \A^* \vdash g := \letin {a_1 \otimes a_2} {a'} {\cons (a_1 , \cons(a_2 , a_{*})) : \A^*}}
+      {a' : \A \otimes \A , a_{*} : \A^* \vdash g := \letin {a_1 \otimes a_2}
+        {a'} {\cons~a_1 (\cons~a_2~a_{*}) : \A^*}}
     }
@@
 
-First, we cover the ``multiplicative'' connectives of non-commutative
+Next, we cover the ``multiplicative'' connectives of non-commutative
 linear logic. The linear unit ($\I$) and tensor product ($\otimes$)
@@
     \and
-    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta, \Delta' \vdash \applto {e'} {e} : B}
+    \inferrule{\Gamma ; \Delta \vdash e : A \lto B \\ \Gamma ; \Delta' \vdash e' : A} {\Gamma ; \Delta, \Delta' \vdash \applto {e} {e'} : B}
     \\
@@
   \label{def:weakequiv}
-  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exists
+  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exist
   parse transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B
@@
 is unambiguous if there is at most one way to transform parses of any
-other grammar $\B$ into parse of $\A$. This notion of an unambiguous
+other grammar $\B$ into parses of $\A$. This notion of an unambiguous
 type is the analog for linear types of the definition of a (homotopy)
@@
 $\uparrow{(A \& A_{\neg} \lto 0)}$ ensures that it is impossible for
-$A$ and $A_{\neg}$ to match the same input string. This means that in
+$A$ and $A_{\neg}$ to parse the same input string. This means that in
 defining a parser, we will need to define a kind of negative grammar
@@
 
-A regular expressions in \theoryabbv is a linear type constructed
+A regular expression in \theoryabbv is a linear type constructed
 using only the connectives $\literal c$, $0$, $\oplus$, $I$,
@@
 data Trace(*@$_{\N}$@*) : (s : N.states) (*@$\to$@*) L where
-  nil : (*@$\uparrow$@*)(&[ s : N.states ] N.isAcc (*@$\to$@*) Trace(*@$_{\N}$@*) s)
+  nil : (*@$\uparrow$@*)(&[ s : N.states ] N.isAcc s (*@$\to$@*) Trace(*@$_{\N}$@*) s)
   cons : (*@$\uparrow$@*)(&[ t : N.transitions ]
@@
 star is replaced by an arbitrary fixed point
-operation\cite{leis_towards_1992}.
+operation \cite{leis_towards_1992}.
 
@@
 A simple example of a CFG is the Dyck grammar of balanced
-parentheses, which we define in \Cref{fig:dyckinductive}
+parentheses, which we define in \Cref{fig:dyckinductive}.
 $\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$. The $\nil$ constructor shows that
@@
     $\gamma \in \sem\Gamma$, we define a set $\sem X \gamma$.
-  \item For each linear type $\Gamma A \isLinTy$ and element $\gamma
+  \item For each linear type $\Gamma \vdash A \isLinTy$ and element $\gamma
     \in \sem\Gamma$, we define a formal grammar $\sem{A}\gamma$. We

### 2025-03-20 paper/paper.tex
checking before meeting

@@
 axioms that do not generally hold in systems based on linear
-logic. First, we need that additive conjunction \emph{distributes}
+logic.
+
+\paragraph{Distributivity}
+First, we need that additive conjunction \emph{distributes}
 over additive disjunction --- e.g., in the finitary case that $0 \& A
@@
   \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y}$ has an
-inverse. Second, we need that the different constructors of
+inverse.
+
+\steven{Maybe this detail should go closer to the actual example
+We use distributivity to encode the lookahead of the LL(1) parser in
+\cref{thm:exp-parser}. At parse time, an LL(1) automaton peeks at the next
+character of a string and uses the peeked character to inform its next
+transition. Our automaton from \cref{fig:binop-inductive} chooses which state to
+transition to dependent on if the next character is $\stringquote{)}$.
+
+TODO
+
+the proof obligation is
+
+$\Char \otimes \LinPiTy{s}{state}{Trace s true}$ which is iso to
+$(\literal{)} \otimes \top \oplus NotStatsWithRP) \otimes
+\LinPiTy{s}{state}{Trace s true}$.
+
+However, without distributivity we get stuck and can go further to actually use
+$lookaheadRP$ or $lookaheadNot$.
+
+}
+
+\paragraph{Disjointness of Constructors}
+Second, we need that the different constructors of
 $\bigoplus$ are disjoint. We add this by adding the axiom that for any
@@
 
+\steven{To define $\DtoN$, we take the smallest trace through $\N$ subject to an
+  ordering on its states and transitions, where the ordering is supplied
+  non-linearly. In essence, the given ordering specifies a disambiguation strategy.}
+
 %% For a moment, let us turn our attention to the content behind the traces in $\D$.
@@
 \begin{theorem}[\agdalogo]
+  \label{thm:exp-parser}
   We construct a parser for $\mathsf{Exp}$ by showing it is weakly
@@
 equivalence was
-introduced by Frisch and Cardelli \cite{frischCardelli}. Henglein and Nielsen
+introduced by Frisch and Cardelli \cite{frischCardelli}. \steven{Frisch and
+  Caredilli also define a linear time greedy parser for regular expressions. We
+  have not yet tried to verify a greedy algorithm, however the approach used in
+  \cref{thm:determinization} would apply to their algorithm.
+  \cref{thm:determinization} requires a choice function for traces through an
+  NFA (i.e. a disambiguation strategy) which may be instantiated to encode greediness. } Henglein and Nielsen
 \cite{henglein_regular_2011} later adapt this approach by providing a type

### 2025-03-20 paper/paper.tex
Check in EOD

@@
 \newif\ifdraft
-\draftfalse
+\drafttrue
 \newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven says}: #1]}\fi}
@@
 \newcommand{\pipe}{\,|\,}
+
+\newtheorem{axiom}{Axiom}[section]
+\newtheorem{construction}{Construction}
 
@@
 f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
-f (a , b) = inl (a (*@$\otimes$@*) b)
+f (a , b) = inl (a , b)
 \end{lstlisting}\begin{mathpar}
@@
 g : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*) ('a'(*@$^*$@*) (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
-g (a , b) = inl (cons a nil (*@$\otimes$@*) b)
+g (a , b) = inl (cons a nil , b)
 \end{lstlisting}
@@
 h nil = nil
-h (cons (a1 (*@$\otimes$@*) a2) as) = cons a1 (cons a2 (h as))
+h (cons (a1 , a2) as) = cons a1 (cons a2 (h as))
 \end{lstlisting}
@@
 over additive disjunction --- e.g., in the finitary case that $0 \& A
-\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most
-general form, the axiom says that the definable function
-${\bigoplus\limits_{f : \prod_{x:X}
-    Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto
-  \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y}$ has an
-inverse.
-
-\steven{Maybe this detail should go closer to the actual example
-We use distributivity to encode the lookahead of the LL(1) parser in
-\cref{thm:exp-parser}. At parse time, an LL(1) automaton peeks at the next
-character of a string and uses the peeked character to inform its next
-transition. Our automaton from \cref{fig:binop-inductive} chooses which state to
-transition to dependent on if the next character is $\stringquote{)}$.
-
-TODO
-
-the proof obligation is
-
-$\Char \otimes \LinPiTy{s}{state}{Trace s true}$ which is iso to
-$(\literal{)} \otimes \top \oplus NotStatsWithRP) \otimes
-\LinPiTy{s}{state}{Trace s true}$.
-
-However, without distributivity we get stuck and can go further to actually use
-$lookaheadRP$ or $lookaheadNot$.
-
-}
+\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most general form,
+
+\begin{axiom}[Distributivity]
+\label{ax:dist}
+For any $A : (x : X) \to Y x \to L$, the function
+
+\[
+ \lamblto{e}{\letin{\sigma\,f\,e'}{e}{\withlamb{x}{e'\,.\pi\,x}}} : \ltonl {\left(  \LinSigTyLimit{f}{\prod\limits_{x : X} . Y (x)}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
+\LinPiTyLimit{x}{X}{\LinSigTyLimit{y}{Y(x)}{A\,x\,y}} \right)}
+\]
+
+has an inverse.
+\end{axiom}
+
+Our primary use of distributivity is to define an equivalence that
+expresses a linear type $A$ as a sum over which character it starts with, if any,
+\(
+A \cong \left(A \& I \right) \oplus \LinSigTyLimit{c}{\Sigma_0}{A \& \left( \literal{c} \otimes \top \right)}
+\). We use this equivalence when building a parser for the traces of the lookahead
+automaton in \cref{fig:binop-inductive}.
+
+\begin{corollary}
+  \label{cor:binary-mono}
+  Distributivity implies that the coprojections $\inl : A \to A \oplus B$,
+  $\inr : B \to A \oplus B$ of a binary sum are monomorphisms.
+\end{corollary}
 
@@
 Second, we need that the different constructors of
-$\bigoplus$ are disjoint. We add this by adding the axiom that for any
-$A : X \to L$ and $x \neq x' : X$ there is a function
-$\uparrow(\equalizer{b}{\sigma\,x \circ \pi_1}{\sigma\,x'\circ \pi_2}
-\lto 0)$ where $b: A(x) \& A(x')$, i.e., the grammar of pairs of an
+$\bigoplus$ are disjoint. That is, we want to enforce that $\sigma~x~a$ and
+$\sigma~x'~a'$ of type $\LinSigTyLimit{x}{X}{A\,x}$ are not equal whenever $x \neq x'$.
+However, because these are linear terms we cannot state their disequality
+directly. Instead, we encode the disequality via a function out of an equalizer,
+
+\begin{axiom}[$\sigma$-Disjointness]
+\label{ax:disjointness}
+For any $A : X \to L$ and $x \neq x' : X$ there is a function,
+\[
+  \uparrow(\equalizer{b}{\sigma\,x \circ \pi_1}{\sigma\,x'\circ \pi_2}
+  \lto 0)
+\]
+  where $b: A(x) \& A(x')$. That is, the grammar of pairs of an
 $a:A(x)$ and an $a': A(x')$ such that $\sigma\,x\,a = \sigma\,x'\,a'$
 is empty.
-
-\pedro{imo, since we are providing an Agda implementation, we can
-  be looser with the explanation of the meaning of the disjointness axioms.
-  As it stands, it's a bit hard do follow along. Is there a simpler example
-  we could point to?}
+\end{axiom}
+
+\steven{need to get cref to cite this as a lemma}
+We use $\sigma$-disjointness in \cref{lem:unambig-to-disjoint} to proof that
+unambiguous sums have disjoint summands.
+
+% \pedro{imo, since we are providing an Agda implementation, we can
+%   be looser with the explanation of the meaning of the disjointness axioms.
+%   As it stands, it's a bit hard do follow along. Is there a simpler example
+%   we could point to?}
 
@@
   If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
-  If a disjunction $\bigoplus\limits_{x:X} A(x)$ is unambiguous then
-  each $A(x)$ is unambiguous.
+  As a consequence of \steven{fix cref} \cref{cor:binary-mono}, if a binary disjunction $A \oplus B$ is unambiguous then
+  $A$ and $B$ are each unambiguous.
 \end{lemma}
@@
 \oplus \Char \oplus (\Char \otimes \Char \otimes \String)$. Then by
-the second principle we have that $\I$ and $\Char$ and each literal
+the second principle, we have that $\I$ and
+$\Char$ and each literal
 $\literal c$ are unambiguous as well.
@@
 only in the case that a parse cannot be constructed.
+
 \begin{definition}
-  A parser for a linear type $A$ is a function $\uparrow{(\String \lto
-    A \oplus A_{\neg})}$, where $A_{\neg}$ is a linear type that is
-  \emph{disjoint} from $A$ in that we can implement a function
-  $\uparrow{(A \& A_{\neg} \lto 0)}$.
+  Linear types $A$ and $B$ are \emph{disjoint} if there is a function
+  $\ltonl {\left( A \& B \lto 0 \right)}$.
+\end{definition}
+
+\begin{definition}
+  A parser for a linear type $A$ is the choice a type $A_{\neg}$ disjoint from
+  $A$ and function $\uparrow{(\String \lto
+    A \oplus A_{\neg})}$.
 \end{definition}
@@
 defining a parser, we will need to define a kind of negative grammar
-for strings that do not parse. Fortunately, we will see that
+for strings that do not parse.
+
+Fortunately, we will see that
 deterministic automata naturally support such a notion with no
@@
 traces that end in a rejecting state. This follows from the following
-principle, a consequence of our axiom that $\oplus$ constructors are
+principle, a consequence of \cref{ax:disjointness}.
+
+\steven{
+\begin{lemma}[\agdalogo]
+  \label{lem:unambig-to-disjoint}
+  If $\LinSigTyLimit{x}{X}{A\,x}$ is unambiguous, then for $x \neq x'$, $A\,x$ and $A,\y$ are
   disjoint.
-\begin{lemma}[\agdalogo]
-  If $A \oplus A_{\neg}$ is unambiguous, then $A$ and $A_{\neg}$ are
-  disjoint.
+
+  In particular, if the binary product $A \oplus A_{\neg}$ is unambiguous, then
+  $A$ and $A_{\neg}$ are disjoint.
 \end{lemma}
+}
 
@@
 \end{figure}
+
+\steven{Maybe this detail should go closer to the actual example
+We use distributivity to encode the of the LL(1) parser in
+\cref{thm:exp-parser}. At parse time, an LL(1) automaton peeks at the next
+character of a string and uses the peeked character to inform its next
+transition. Our automaton from \cref{fig:binop-inductive} chooses which state to
+transition to dependent on if the next character is $\stringquote{)}$.
+
+TODO
+
+the proof obligation is
+
+$\Char \otimes \LinPiTy{s}{state}{Trace s true}$ which is iso to
+$(\literal{)} \otimes \top \oplus NotStatsWithRP) \otimes
+\LinPiTy{s}{state}{Trace s true}$.
+
+However, without distributivity we get stuck and can go further to actually use
+$lookaheadRP$ or $lookaheadNot$.
+
+}
 

### 2025-03-21 paper/paper.tex
touchup axioms, fix crefs, determinization disambiguation

@@
 \usepackage{xspace}
+\usepackage{thmtools} % needed for sane cleveref naming of theorems
 
@@
 \newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
-\newcommand{\String}{\textbf{String}}
-\newcommand{\Char}{\textbf{Char}}
-\newcommand{\stringg}{\texttt{String}}
-\newcommand{\charg}{\mathtt{Char}}
+\newcommand{\StringSem}{\textbf{String}}
+\newcommand{\CharGram}{\texttt{Char}}
+\newcommand{\StringGram}{\texttt{String}}
 \newcommand{\Set}{\mathbf{Set}}
@@
 
-\newtheorem{axiom}{Axiom}[section]
-\newtheorem{construction}{Construction}
+\newcommand{\states}{\mathtt{states}}
+\newcommand{\labelt}{label}
+\newcommand{\transitions}{\texttt{transitions}}
+\newcommand{\epstransitions}{\epsilon\texttt{transitions}}
+\newcommand{\isAcc}{\mathtt{isAcc}}
+\newcommand{\init}{\mathtt{init}}
+\newcommand{\src}{\mathtt{src}}
+\newcommand{\dst}{\mathtt{dst}}
+\newcommand{\pparse}{\mathtt{parse}}
+\newcommand{\print}{\mathtt{print}}
+\newcommand{\epssrc}{\epsilon\mathtt{src}}
+\newcommand{\epsdst}{\epsilon\mathtt{dst}}
+\newcommand{\balanced}{\mathtt{balanced}}
+\newcommand{\Dyck}{\mathtt{Dyck}}
+\newcommand{\N}{\texttt{N}}
+
+% These declarations are to get cleveref to play nice with
+% different theorem-like environemnts. Otherwise \cref{lem1.1}
+% will display 'Theorem 1.1'
+
+\declaretheoremstyle[headfont=\normalfont\itshape]{defstyle}
+\declaretheoremstyle[headfont=\normalfont\scshape]{thmstyle}
+
+\declaretheorem[name=Definition, style=defstyle, numberwithin=section]{definition}
+\declaretheorem[style=thmstyle, name=Theorem, numberlike=definition]{theorem}
+\declaretheorem[style=thmstyle, name=Axiom, numberlike=definition]{axiom}
+\declaretheorem[style=thmstyle, name=Corollary, numberlike=definition]{corollary}
+\declaretheorem[style=thmstyle, name=Construction, numberlike=definition]{construction}
+
+\crefname{definition}{Definition}{Definitions}
+\crefname{theorem}{Theorem}{Theorems}
+\crefname{axiom}{Axiom}{Axioms}
+\crefname{corollary}{Corollary}{Corollarys}
+\crefname{construction}{Construction}{Constructions}
+
+% \newtheorem{axiom}{Axiom}[section]
+% \newtheorem{construction}{Construction}
 
@@
 language membership. That is, a \emph{formal grammar} $A$ is a
-function $\String \to \Set$ such that for a string $w$, $A~w$ is the
+function $\StringSem \to \Set$ such that for a string $w$, $A~w$ is the
 set of ``proofs'' showing that $w$ belongs to the language recognized
@@
 axioms that do not generally hold in systems based on linear
-logic.
-
-\paragraph{Distributivity}
-First, we need that additive conjunction \emph{distributes}
+logic. First, we need that additive conjunction \emph{distributes}
 over additive disjunction --- e.g., in the finitary case that $0 \& A
@@
 
+\begin{corollary}
+  \label{cor:binary-mono}
+  Distributivity implies that the coprojections $\inl : A \to A \oplus B$,
+  $\inr : B \to A \oplus B$ of a binary sum are monomorphisms --- i.e. if
+  $\inl\,a \equiv \inl\,a'$, then $a \equiv a'$.
+\end{corollary}
+
+\steven{cite reference for distributivity implies coprojections mono. Cockett
+  ``Introduction to Distributive Categories''? Awodey?}
+
 Our primary use of distributivity is to define an equivalence that
@@
 \(
-A \cong \left(A \& I \right) \oplus \LinSigTyLimit{c}{\Sigma_0}{A \& \left( \literal{c} \otimes \top \right)}
+A \cong \left(A \& I \right) \oplus \LinSigTyLimit{c}{\Sigma_0}{\left( A \& \left( \literal{c} \otimes \top \right) \right)}
 \). We use this equivalence when building a parser for the traces of the lookahead
@@
 
-\begin{corollary}
-  \label{cor:binary-mono}
-  Distributivity implies that the coprojections $\inl : A \to A \oplus B$,
-  $\inr : B \to A \oplus B$ of a binary sum are monomorphisms.
-\end{corollary}
-
-\paragraph{Disjointness of Constructors}
 Second, we need that the different constructors of
@@
 \[
-  \uparrow(\equalizer{b}{\sigma\,x \circ \pi_1}{\sigma\,x'\circ \pi_2}
+  \uparrow(\equalizer{b}{\left( \sigma\,x \circ \pi_1 \right)}{\left(  \sigma\,x'\circ \pi_2 \right)}
   \lto 0)
@@
 
-\steven{need to get cref to cite this as a lemma}
-We use $\sigma$-disjointness in \cref{lem:unambig-to-disjoint} to proof that
+We use $\sigma$-disjointness in \cref{lem:unambig-to-disjoint} to prove that
 unambiguous sums have disjoint summands.
@@
 alphabet, we add a corresponding linear type $c$. We can then define a
-non-linear type $\Char$ as the disjunction of all of these characters,
-and define a type $\String$ as the Kleene star of $\Char$, i.e. as an
+non-linear type $\CharGram$ as the disjunction of all of these characters,
+and define a type $\StringGram$ as the Kleene star of $\CharGram$, i.e. as an
 inductive linear type.
 Then we add a function $\mathsf{read} :
-\ltonl{(\top\lto \String)}$ that intuitively ``reads'' the input
+\ltonl{(\top\lto \StringGram)}$ that intuitively ``reads'' the input
 string from the input and makes it available. It is important that the
@@
 add an axiom that $\lambda s. \mathsf{read}(!(s)) \equiv \lambda
-s. s$ where $!$ is the unique function $\ltonl(\String\lto \top)$,
+s. s$ where $!$ is the unique function $\ltonl(\StringGram\lto \top)$,
 i.e., that if we have a string, but then throw it away and read
 it from the input, then in fact that is equivalent to the string we
-were given originally. This ensures that the elements of the $\String$
+were given originally. This ensures that the elements of the $\StringGram$
 type always stand for the actual input string in our reasoning. In the
@@
 %% definition of retract is a useful intermediate notion. In retrospect,
-%% our axiom about $\String$ says that $\String$ is a retract of $\top$.
+%% our axiom about $\StringGram$ says that $\StringGram$ is a retract of $\top$.
 
@@
 \begin{lemma}[\agdalogo]
+  \label{lem:retract}
   If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
-  As a consequence of \steven{fix cref} \cref{cor:binary-mono}, if a binary disjunction $A \oplus B$ is unambiguous then
+\end{lemma}
+\begin{lemma}[\agdalogo]
+  \label{lem:unambig-sum}
+  As a consequence of \cref{cor:binary-mono}, if a binary disjunction $A \oplus B$ is unambiguous then
   $A$ and $B$ are each unambiguous.
 \end{lemma}
-From the first principle, we can prove that $\String$ is unambiguous,
+From \cref{lem:retract}, we can prove that $\StringGram$ is unambiguous,
 since it is a retract of $\top$. In fact, observe that if $\A$ is a
@@
 \lambda b. b$ follows because $\B$ is unambiguous. Therefore
-$\String$ is also strongly equivalent to $\top$. Next, since $\String$ is
-defined as a Kleene star, we can easily show that $\String \cong \I
-\oplus \Char \oplus (\Char \otimes \Char \otimes \String)$. Then by
-the second principle, we have that $\I$ and
-$\Char$ and each literal
-$\literal c$ are unambiguous as well.
+$\StringGram$ is also strongly equivalent to $\top$. Next, since $\StringGram$ is
+defined as a Kleene star, we can easily show that $\StringGram \cong \I
+\oplus \CharGram \oplus (\CharGram \otimes \CharGram \otimes \StringGram)$. Then by
+\cref{lem:unambig-sum}, we have that $\I$ and
+$\CharGram$ --- and thus each literal
+$\literal c$ --- are unambiguous as well.
 
 % \paragraph{A Grammar of Strings}
-% Define the character grammar $\charg$ to be the disjunction over characters in the alphabet $\Sigma$,
-% $\charg = \LinSigTy {c} {\Sigma} c$. Then define a linear type of strings as a
+% Define the character grammar $\CharGram$ to be the disjunction over characters in the alphabet $\Sigma$,
+% $\CharGram = \LinSigTy {c} {\Sigma} c$. Then define a linear type of strings as a
 % linear list of characters,
-% $\stringg = \charg^{*}$.
+% $\StringGram = \CharGram^{*}$.
 
 % In the grammars model, we can build a
-% strong equivalence between $\top$ and $\stringg$. hen building parsers, we may wish to enforce this
-% syntactically by including $\top \cong \stringg$ as an axiom.
-
-% Through inclusion of this axiom, $! : \stringg \to \top$ is an isomorphism, and
-% thus also a monomorphism. That is, $\stringg$ is unambiguous.
+% strong equivalence between $\top$ and $\StringGram hen building parsers, we may wish to enforce this
+% syntactically by including $\top \cong \StringGram$ as an axiom.
+
+% Through inclusion of this axiom, $! : \StringGram \to \top$ is an isomorphism, and
+% thus also a monomorphism. That is, $\StringGram$ is unambiguous.
 % \max{we should have a section of the previous section which gives the axioms, this one and the distributivity axiom}
@@
 % We say $A$ is \emph{parseable} if there exists a grammar $B$ and a map
-% $\top \to A \oplus B$. Taking $\top \cong \stringg$, a grammar $A$ is parseable
+% $\top \to A \oplus B$. Taking $\top \cong \StringGram$, a grammar $A$ is parseable
 % exactly when we may read an arbitrary string into a parse of $A$ or a $B$-valued
@@
 $A$, a first attempt at defining a parser would be to implement a
-function $\uparrow{(\String \lto A)}$. But since our linear functions
+function $\uparrow{(\StringGram \lto A)}$. But since our linear functions
 must be total, this means that we can construct an $A$ parse for
@@
 interest. Instead we might try to write a partial function as a
-$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option''
+$\ltonl{(\StringGram \lto (A \oplus \top))}$ using the ``option''
 monad. This allows for the possibility that the input string doesn't
@@
 \begin{definition}
+  \label{def:disjoint}
   Linear types $A$ and $B$ are \emph{disjoint} if there is a function
@@
 \begin{definition}
+  \label{def:parser}
   A parser for a linear type $A$ is the choice a type $A_{\neg}$ disjoint from
-  $A$ and function $\uparrow{(\String \lto
+  $A$ and function $\uparrow{(\StringGram \lto
     A \oplus A_{\neg})}$.
@@
 
-\steven{
 \begin{lemma}[\agdalogo]
   \label{lem:unambig-to-disjoint}
-  If $\LinSigTyLimit{x}{X}{A\,x}$ is unambiguous, then for $x \neq x'$, $A\,x$ and $A,\y$ are
-  disjoint.
-
-  In particular, if the binary product $A \oplus A_{\neg}$ is unambiguous, then
+  If $\LinSigTy{x}{X}{A\,x}$ is unambiguous, then for $x \neq x'$, $A\,x$ and $A\,x'$ are
+  disjoint. In particular, if the binary product $A \oplus A_{\neg}$ is unambiguous, then
   $A$ and $A_{\neg}$ are disjoint.
 \end{lemma}
-}
-
-Writing a parser as a linear term in this way is an intrinsic
+
+\begin{proof}
+If $\LinSigTy{x}{X}{A\,x}$ is unambiguous, then all functions into it from
+$A\,x \& A\,x'$ are equal. In particular,
+$\sigma\,x\,\circ\,\pi_{1} \equiv \sigma\,x'\,\circ\,\pi_{2}$ so there is a
+function \(
+\ltonl{\left(A\,x \& A\,x' \lto
+  \equalizer{b}{\left( \sigma\,x \circ \pi_1 \right)}{\left(  \sigma\,x'\circ \pi_2 \right)}
+\right)}.
+\)
+We then compose with the function in \cref{ax:disjointness} to prove that $A\,x$ and $A\,x'$ are disjoint.
+\end{proof}
+
+Writing a parser as a linear term is an intrinsic
 verification of the \emph{soundness} of the parser completely for free
 from the typing: any $\inl$ parse that we return \emph{must}
-correspond to a parse tree of the input string. Further if we verify
-the disjointness property we then also get the \emph{completeness} of
+correspond to a parse tree of the input string. Further, if we verify
+the disjointness property of \cref{def:parser} we then also get the \emph{completeness} of
 the parser as well, that it never fails to generate an $A$ parse when
@@
   \label{lem:wk-eqv-parse}
-  If $\A$ is weakly equivalent to $\B$ then any parser for $\A$ can be
-  extended to a parser for $\B$.
+  If $\A$ is weakly equivalent to $\B$ then any parser for $\A$ extends to a
+  parser for $\B$.
 \end{lemma}
 Here we need both directions of the weak equivalence. We need $A \lto
-B$ to extend the parser from $\String \lto A \oplus A_{\neg}$ to
-$\String \lto B \oplus A_{\neg}$ but then we also need $B \lto A$ to
+B$ to extend the parser from $\StringGram \lto A \oplus A_{\neg}$ to
+$\StringGram \lto B \oplus A_{\neg}$ but then we also need $B \lto A$ to
 establish that $A_{\neg}$ is disjoint from $B$.
@@
 a verified regex parser.
-
-\newcommand{\states}{\mathtt{states}}
-\newcommand{\labelt}{label}
-\newcommand{\transitions}{\texttt{transitions}}
-\newcommand{\epstransitions}{\epsilon\texttt{transitions}}
-\newcommand{\isAcc}{\mathtt{isAcc}}
-\newcommand{\init}{\mathtt{init}}
-\newcommand{\src}{\mathtt{src}}
-\newcommand{\dst}{\mathtt{dst}}
-\newcommand{\parse}{\mathtt{parse}}
-\newcommand{\print}{\mathtt{print}}
-\newcommand{\epssrc}{\epsilon\mathtt{src}}
-\newcommand{\epsdst}{\epsilon\mathtt{dst}}
-
-\newcommand{\N}{\texttt{N}}
 
@@
 we show that for any start state $s$, ${\LinSigTy {b} {\Bool}
-  \Trace_{D}~s~b}$ is a retract of $\stringg$. That is, first we
-construct a function $\parse_{D}$ that is a parser for
+  \Trace_{D}~s~b}$ is a retract of $\StringGram$ and apply \cref{lem:retract} to
+derive unambiguity. That is, first we
+construct a function $\pparse_{D}$ that is a parser for
 $\Trace_D~{s}~{true}$, with $\Trace_D~{s}~{false}$ being the disjoint
-type used. The fact that these types are disjoint follows by showing
-that this function is part of a retraction, i.e., that there is only
-one way to trace through a deterministic automaton.
+type used. The fact that these trace types are disjoint follows from the
+unambiguity of $\LinSigTy {b} {\Bool} {\Trace_{D}~s~b}$ by \cref{lem:unambig-sum}.
 
 The parser,
-$\parse_{D}$,
-is defined by recursion on strings in \cref{fig:printdfa}. If this string is empty, then $\parse_{D}$
+$\pparse_{D}$,
+is defined by recursion on strings in \cref{fig:printdfa}. If this string is empty, then $\pparse_{D}$
 defines a linear function that terminates a trace at the input state $s$. If the
-string is nonempty, then $\parse_{D}$ walks forward in $D$ from the input state $s$ by
+string is nonempty, then $\pparse_{D}$ walks forward in $D$ from the input state $s$ by
 the character at the head of the string.
@@
 via $\nil$, then $\print_{D}$ returns the empty string. Otherwise, if the trace
-is defined by $\cons$ then $\parse_{D}$ appends the character from the most
+is defined by $\cons$ then $\pparse_{D}$ appends the character from the most
 recent transition to the output
@@
 \begin{theorem}[\agdalogo]
-  $\parse_D s$ is a parser for $\Trace_D~s~\true$.
+  \label{thm:dfa-parser}
+  $\pparse_D s$ is a parser for $\Trace_D~s~\true$.
 \end{theorem}
@@
 DFA is unambiguous even if the NFA is not.
-\begin{theorem}[Determinization, \Agda]
-  \label{thm:determinization}
-  Given an NFA $N$, there exists a DFA $D$ such that
+\begin{construction}[Determinization, \Agda]
+  \label{cons:determinization}
+  Given an NFA $N$, we construct a DFA $D$ such that
   $Parse_{N}$ is weakly equivalent to $Parse_{D}$.
-\end{theorem}
+\end{construction}
 \newcommand{\X}{\texttt{X}}
@@
 \begin{proof}
-  Define the states of $\D$ to be the $\mathbb{P}_{\varepsilon}(\N.\states)$ --- the type
+  Define the states of $\D$ to be the $\mathbb{P}_{\epsilon}(\N.\states)$ --- the type
   of $\epsilon$-closed\footnote{A subset of states $\X$ is $\epsilon$-closed if
@@
 To construct a term from DFA traces to NFA traces, we similarly strengthen our
-induction hypothesis and build a parse transformer
+inductive hypothesis and build a parse transformer
 \[
-  \DtoN : \ltonl {\left( \Trace_{\D}~\X~\true \lto \bigoplus\limits_{s : \N.\mathsf{states}}\bigoplus\limits_{\texttt{sInX} : \X \ni \s} \Trace_{\N}~s~\true \right)}
+  \DtoN : \ltonl {\left( \Trace_{\D}~\X~\true \lto \bigoplus\limits_{s : \N.\states}\bigoplus\limits_{\texttt{sInX} : \X \ni \s} \Trace_{\N}~s~\true \right)}
 \]
 
-\steven{To define $\DtoN$, we take the smallest trace through $\N$ subject to an
-  ordering on its states and transitions, where the ordering is supplied
-  non-linearly. In essence, the given ordering specifies a disambiguation strategy.}
+Because the states of $D$ are $\epsilon$-closed subsets of $N.\states$, if two
+states $s$ and $s'$ belong to the the same $\epsilon$-closed subset
+$X : \mathbb{P}_{\epsilon}(N.\states)$ then there exists a $\epsilon$-path in
+$N$ between the two, but we do not necessarily know which one. Similarly, the
+data contained in the type $\Trace_{\D}}~X~true$ is the \emph{existence} of an accepting trace in
+$\N$ beginning at some state in $\D$.
+
+To define $\DtoN$, we need a choice function that extracts out a trace in $N$
+from the mere existence of one. We achieve this by choosing the smallest trace
+through $\N$ subject to an ordering on the non-linear types $N.\states$,
+$N.\transitions$, and $N.\epstransitions$. In essence, the given orderings
+specify a global disambiguation strategy.
 
@@
 \newcommand{\R}{\texttt{R}}
-\begin{theorem}[Thompson's Construction, \Agda]
-  Given a regular expression $\R$, there exists an NFA $\N$ such that $\R$ is
+\begin{construction}[Thompson's Construction, \Agda]
+  \label{cons:thompson}
+  Given a regular expression $\R$, we build an NFA $\N$ such that $\R$ is
   strongly equivalent to $\Trace_{N}(\N.\init)$.
-\end{theorem}
+\end{construction}
 \begin{proof}
@@
 
+\begin{corollary}
+  We may build a parser for every regular expression $\R$.
+\end{corollary}
+\begin{proof}
+We combine the strong equivalence of \cref{cons:thompson} with the weak
+equivalance of \cref{cons:determinization} to show that $\R$ is weakly
+equivalent to the traces of its determinized automaton. Then, we use
+\cref{lem:wk-eqv-parse} to extend the parser from \cref{thm:dfa-parser}
+with respect to this weak equivalence.
+\end{proof}
+
 \subsection{Context-free grammars}
@@
 
-\newcommand{\balanced}{\mathtt{balanced}}
-\newcommand{\Dyck}{\mathtt{Dyck}}
 A simple example of a CFG is the Dyck grammar of balanced
@@
 
-$\Char \otimes \LinPiTy{s}{state}{Trace s true}$ which is iso to
+$\CharGram \otimes \LinPiTy{s}{state}{Trace s true}$ which is iso to
 $(\literal{)} \otimes \top \oplus NotStatsWithRP) \otimes
@@
 \subsection{Unrestricted Grammars}
-\newcommand{\stringNL}{String'}
 
@@
 demonstrate this, we show that for any \emph{non-linear} function $P :
-\stringNL \to U$, where here $\stringNL$ is the \emph{non-linear} type
+\StringSem \to U$, where here $\StringSem$ is the \emph{non-linear} type
 of strings over the alphabet, we can construct a grammar whose parses
 correspond to $P$.
-%% Define $\stringNL$ to be the non-linear list type over the alphabet $\Sigma$ and
+%% Define $\StringSem$ to be the non-linear list type over the alphabet $\Sigma$ and
 %% let  be a mapping from the set of strings to the universe of
 %% nonlinear types. The function $P$ describes a family of nonlinear types
-%% dependent on $\stringNL$.
+%% dependent on $\StringSem$.
 %% We view the image of $P$ as a set of parse trees for each input string, and thus
@@
 \[
-Reify~P = \LinSigTyLimit {w} {\stringNL} {\LinSigTyLimit {x} {P~w} {~\lceil w \rceil}}
+Reify~P = \LinSigTyLimit {w} {\StringSem} {\LinSigTyLimit {x} {P~w} {~\lceil w \rceil}}
 \]
@@
 
-This reification operation on functions $\stringNL \to U$ is incredibly expressive,
+This reification operation on functions $\StringSem \to U$ is incredibly expressive,
 as it allows to sidestep our linear typing connectives and utilize the whole of nonlinear
@@
 Turing machine $Tur$ one may
-define a non-linear predicate $accepts : \stringNL \to U$ such that $accept~w$ is equal to $\top$ if
+define a non-linear predicate $accepts : \StringSem \to U$ such that $accept~w$ is equal to $\top$ if
 $T$ accepts $w$ and equal to $0$ otherwise. Then, $Reify~accepts$ is a linear
@@
 work directly with Agda types and a definition of a formal grammar as
-a function $\String \to \Set$. Then we implement each of the type and
+a function $\StringSem \to \Set$. Then we implement each of the type and
 term constructors of \theoryabbv as combinators on formal grammars or

### 2025-03-21 paper/paper.tex
cfgs

@@
   \begin{tikzpicture}[node distance = 17mm ]
-    \node[state, initial, accepting] (0) {$0$};
+    \node[state, initial above, accepting] (0) {$0$};
     \node[state, right of=0] (1) {$1$};
@@
     \node[right of=2] (3) {$\dots$};
-    \node[state, below of=0] (fail) {$fail$};
+    \node[state, left of=0] (fail) {$fail$};
 
@@
               (3) edge[below, bend left] node{\stringquote{)}} (2)
-              (fail) edge[loop right] node{\stringquote{(}, \stringquote{)}} (fail)
-              (0) edge[left] node{\stringquote{)}} (fail);
+              (fail) edge[loop left] node{\stringquote{(}, \stringquote{)}} (fail)
+              (0) edge[above] node{\stringquote{)}} (fail);
   \end{tikzpicture}
@@
 
-\steven{Maybe this detail should go closer to the actual example
-We use distributivity to encode the of the LL(1) parser in
-\cref{thm:exp-parser}. At parse time, an LL(1) automaton peeks at the next
-character of a string and uses the peeked character to inform its next
-transition. Our automaton from \cref{fig:binop-inductive} chooses which state to
-transition to dependent on if the next character is $\stringquote{)}$.
-
-TODO
-
-the proof obligation is
-
-$\CharGram \otimes \LinPiTy{s}{state}{Trace s true}$ which is iso to
-$(\literal{)} \otimes \top \oplus NotStatsWithRP) \otimes
-\LinPiTy{s}{state}{Trace s true}$.
-
-However, without distributivity we get stuck and can go further to actually use
-$lookaheadRP$ or $lookaheadNot$.
-
-}
-
 Our final example is of a simple grammar of arithmetic expressions
@@
 \stringquote{(}, \stringquote{)}, \stringquote{+} , \stringquote{NUM}
-\}$. In \Cref{fig:binop-inductive} we define it using two mutually
+\}$. In \Cref{fig:binop-inductive} we define the expression grammar
+using two mutually
 recursive types, corresponding to the two non-terminals we would use
@@
 operation is right associative. In the same figure, we define the
-traces of an automaton with one token of lookahead. The automaton has
+traces of an automaton with one token of lookahead. The automaton\footnote{In
+  the automaton definition,
+  $\texttt{NotStartsWithLP}$ is defined as
+  \(I \oplus
+  (\literal{)} \oplus \literal{+} \oplus \mathtt{NUM}) \otimes \top
+  \). Similarly,
+  $\texttt{NotStartsWithRP}$ is defined as
+  \(I \oplus
+  (\literal{(} \oplus \literal{+} \oplus \mathtt{NUM}) \otimes \top
+  \).
+} has
 four different ``states'', each with access to a natural number
-``stack''. The ``opening'' state $O$ expects either an open paren, in
+``stack''. The ``opening'' state $O$ expects either an left paren, in
 which case it increments the stack and stays in the opening state, or
@@
 state $D$ is where lookahead is used: if the next token will be a
-right paren, then we proceed to $C$, otherwise we proceed to $M$. Here
-$\mathsf{NotStartsWithRP}$ is defined as
-$\I \oplus (({\literal{(}} \oplus {\literal {+}}
-\oplus \mathsf{NUM}) \otimes \top)$. In the ``closing'' state $C$ if
-we observe a close paren we decrement the count and continue to the
-$D$ state. In the ``multiplying'' state $M$, we succeed if the string
-ends and the count is $0$, and if we see a plus we continue to the $O$
-state. Additionally, since the automaton need also parse all of the
-incorrect strings, we add additional failing cases. It is
-straightforward to implement a parser for this lookahead automaton,
-generalizing the approach for deterministic automata.
+right paren, then we proceed to $C$; otherwise, we proceed to $A$.
+In the ``closing'' state $C$ if
+we observe a right paren, then we decrement the count and continue to the
+$D$ state. In the ``adding'' state $A$, we succeed if the string
+ends and the count is $0$; otherwise, if we see a plus we continue to the $O$
+state. Additionally, since the automaton need parse all of the
+incorrect strings, we add all of the failing cases.
 
@@
 \end{theorem}
+
+\steven{
+With \cref{ax:dist}, it is straightforward to implement a parser for this lookahead automaton,
+generalizing the approach for deterministic automata. Without access to
+distributivity, we may define a rudimentary lookahead operation via the chain of
+equivalences,
+
+\[
+  A~n~b \cong (A~n~b) \& \top \cong (A~n~b)\,\& \left(\left(\literal{)} \otimes \top \right) \oplus \texttt{NotStartsWithRP}\right)
+\]
+
+When defining a parser, if the lookahead character is a right paren we would
+like to apply $\texttt{lookAheadRP}$; otherwise, apply $\texttt{lookAheadNot}$.
+However, without distributivity we have no means of relating the bit of
+information learned by looking ahead to the input $A~n~b$ parse.
+That is, to choose which constructor to apply we would need the end of this
+chain of equivalences to be a binary sum rather than a binary product,
+necessitating the distributivity axiom.
+}
+
 \begin{figure}
@@
   left : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O (n + 1) b (*@$\lto$@*) O n b)
-  num : (*@$\uparrow$@*)(&[ n ] &[ b ] NUM (*@$\lto$@*) DO n b (*@$\lto$@*) O n b)
-  done : (*@$\uparrow$@*)(&[ n ] O n false)
-  unexpected : (*@$\uparrow$@*)(&[ n ] (')' (*@$\oplus$@*) '+') (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) O n false)
+  num : (*@$\uparrow$@*)(&[ n ] &[ b ] NUM (*@$\lto$@*) D n b (*@$\lto$@*) O n b)
+  unexpected : (*@$\uparrow$@*)(&[ n ] (NotStartsWithLP (*@$\lto$@*) O n false)
 data D : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
   lookAheadRP : (*@$\uparrow$@*)(&[ n ] &[ b ] ((')' (*@$\otimes$@*) (*@$\top$@*)) (*@$\&$@*) C n b) (*@$\lto$@*) D n b)
-  lookAheadNot  : (*@$\uparrow$@*)(&[ n ] &[ b ] (NotStartsWithRP (*@$\&$@*) M n b) (*@$\lto$@*) D n b)
+  lookAheadNot : (*@$\uparrow$@*)(&[ n ] &[ b ] (NotStartsWithRP (*@$\&$@*) A n b) (*@$\lto$@*) D n b)
 data C : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
   closeGood : (*@$\uparrow$@*)(&[ n ] &[ b ] ')' (*@$\lto$@*) D n b (*@$\lto$@*) C (n + 1) b)
-  closeBad : (*@$\uparrow$@*)(&[ n ] ')' (*@$\lto$@*) C 0 b)
-  done : (*@$\uparrow$@*)(&[ n ] C n false)
-  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) '+' (*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) C n false)
-data M : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
-  doneGood : (*@$\uparrow$@*)(M 0 true)
-  doneBad : (*@$\uparrow$@*)(&[ n ] M (n + 1) false)
-  add : (*@$\uparrow$@*)(&[ n ] &[ b ] '+' (*@$\lto$@*) O n b (*@$\lto$@*) M n b)
-  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')'(*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) M n false)
+  closeBad : (*@$\uparrow$@*)(')' (*@$\lto$@*) C 0 false)
+  unexpected : (*@$\uparrow$@*)(&[ n ] NotStartsWithRP (*@$\lto$@*) C n false)
+data A : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
+  doneGood : (*@$\uparrow$@*)(A 0 true)
+  doneBad : (*@$\uparrow$@*)(&[ n ] A (n + 1) false)
+  add : (*@$\uparrow$@*)(&[ n ] &[ b ] '+' (*@$\lto$@*) O n b (*@$\lto$@*) A n b)
+  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')'(*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) A n false)
 \end{lstlisting}

### 2025-03-21 paper/paper.tex
turing machine

@@
 $N$ between the two, but we do not necessarily know which one. Similarly, the
-data contained in the type $\Trace_{\D}}~X~true$ is the \emph{existence} of an accepting trace in
+data contained in the type $\Trace_{\D}~X~true$ is the \emph{existence} of an accepting trace in
 $\N$ beginning at some state in $\D$.
@@
 dependent type theory to define a grammar. For example, given a
-Turing machine $Tur$ one may
-define a non-linear predicate $accepts : \StringSem \to U$ such that $accept~w$ is equal to $\top$ if
-$T$ accepts $w$ and equal to $0$ otherwise. Then, $Reify~accepts$ is a linear
-type that captures precisely the strings accepted by $Tur$. That is,
+Turing machine $T$ one may
+define a non-linear predicate $accepts : \StringSem \to U$ that encodes that $T$
+halts and accepts an input string. Then, $Reify~accepts$ is a linear type that
+captures precisely the string the same language as $T$. That is,
 $Lang(Reify~accepts)$ is recursively enumerable --- the most general class of
 languages in the Chomsky hierarchy.
-\begin{theorem}[\Agda]
-  For any Turing machine, we can construct a grammar in \theoryabbv
-  that accepts the same language as the Turing machine.
-\end{theorem}
+\begin{construction}[\Agda]
+  \label{cons:turing}
+  For any Turing machine $T$, we can construct a grammar in \theoryabbv
+  that accepts the same language as $T$.
+\end{construction}
 
@@
 %% \subsubsection{Typed Grammars}
-The usage of a simple type system to reason about regular expressions up to weak
-equivalence was
-introduced by Frisch and Cardelli \cite{frischCardelli}. \steven{Frisch and
-  Caredilli also define a linear time greedy parser for regular expressions. We
-  have not yet tried to verify a greedy algorithm, however the approach used in
-  \cref{thm:determinization} would apply to their algorithm.
-  \cref{thm:determinization} requires a choice function for traces through an
-  NFA (i.e. a disambiguation strategy) which may be instantiated to encode greediness. } Henglein and Nielsen
-\cite{henglein_regular_2011} later adapt this approach by providing a type
-system for regular expressions whose semantics were in one-to-one correspondence
-with the set of parse trees.
-Elliott \cite{elliottSymbolicAutomaticDifferentiation2021} gives a similar
-semantic interpretation of regular grammars as type-level predicates on strings.
-The type system of \theoryabbv extends these semantic interpretations of to a broader class of grammars and also gives a formal syntax and equational theory for the parse transformers that these prior works lack.
-
+Frisch and Cardelli introduced the use of a simple type system to reason about regular expressions up to weak equivalence \cite{frischCardelli}.
+Henglein and Nielsen build on this type-theoretic view of grammars, pointing out that the values of the types correspond precisely to the parse trees \cite{henglein_regular_2011}.
+Our design of \theoryabbv extends this view of grammars as types to a much broader class of grammars, as well as providing a syntax and equational theory for \emph{parse transformers}, showing that the terms can be viewed not just as parse trees, but as intrinsically verified parsers.
+
+Elliott \cite{elliottSymbolicAutomaticDifferentiation2021} uses essentially the same denotational semantics as ours, interpreting regular expressions as type-valued functions on strings.
+Our denotational semantics of \theoryabbv extends this to a much broader class of grammars.
 %% Krishnaswami and Yallop \cite{krishnaswami_typed_2019} provide a type system for
@@
 %% algebra to reason about parsing as well.
+
+\paragraph{Dependent Linear Types}
+
+Our syntax for non-commutative dependent linear type theory is based on the
+dependent commutative linear type theory of
+\citet{krishnaswami_integrating_2015}, itself an extension of Benton and
+Wadler's linear-non-linear calculus \cite{bentonMixedLinearNonlinear1995}. A
+distinct feature of these systems is that linear types and non-linear types are
+distinct sorts, so the linear logic $!$ operator is not a primitive, but is
+instead definable. Vakar develops a dependent linear type theory where instead
+the non-linear types are accessed using the $!$ modality, similar to Girard's
+original approach \cite{vakar,girard_linear_1987}. Additionally, Vakar develops
+a general categorical semantics for their system, whereas we have only developed
+a single denotational model. It appears straightforward to adapt Vakar's general
+semantics approach to a non-commutative variant that would apply to
+\theoryabbv. This may have applications in finding alternative models, or
+developing logical relations proofs categorically.
 
@@
 
+\subsection{Future Work}
+
+In this work we have demonstrated feasibility of \theoryabbv for the
+verification of formal grammar theory and sound-by-construction parsers. In
+future work, we aim to extend \theoryabbv to be a practical tool for developing
+verified parsing components of larger verified software systems.
+
+\paragraph{Verified Parser Generators}
+
+We aim to extend our work on parsing to verify other algorithms. For instance,
+our regular expression parser makes somewhat arbitrary choices in what parse to
+produce when the grammar is ambiguous. In the future we aim to verify Frisch and
+Cardelli's \emph{greedy} algorithm for regular expression
+disambiguation. Defining this algorithm in \theoryabbv would provide soundness
+by construction, but it is not obvious if the greediness property could be
+verified easily as well.
+
+We also aim to adapt the approach used for our context-free grammar parserss for
+LL and LR/LALR with the aim of developing a shared library of intrinsically
+verified parsing utilities that can be usefully shared across different
+parsers. We would also like to investigate how high-performance verified parsers
+can be implemented in \theoryabbv. This might be done by developing an efficient
+compiler for \theoryabbv directly or by developing a verified compiler to an
+imperative system within \theoryabbv itself.
+
 \paragraph{Semantic Actions}
 
-Our verification has mainly focused on the verification that a parser
-outputs a correct concrete syntax tree for a grammar. However, in
-practice, parsers are combined with a \emph{semantic action} that
-emits an \emph{abstract} syntax tree that omits superfluous syntactic
-details that aren't needed in later stages of the overall program. We
-can define a semantic action in \theoryabbv for a linear type $A$ with
-semantic outputs in a non-linear type $X$ to be a function $\ltonl {(A
-  \lto \bigoplus_{\_:X} \top)}$. That is, a semantic action is a
-function that produces a semantic element of $X$ from the concrete
-parses of $A$. In future work, we aim to study the question of
-verifying efficient implementations of parsers with semantic actions.
+Our verification has mainly focused on the verification that a parser outputs a
+correct concrete syntax tree for a grammar. However, in practice, parsers are
+combined with a \emph{semantic action} that emits an \emph{abstract} syntax tree
+that omits superfluous syntactic details that are unneeded in later stages of
+the overall program. We can define a semantic action in \theoryabbv for a linear
+type $A$ with semantic outputs in a non-linear type $X$ to be a function $\ltonl
+{(A \lto \bigoplus_{\_:X} \top)}$. That is, a semantic action is a function that
+produces a semantic element of $X$ from the concrete parses of $A$. In future
+work, we aim to study the question of verifying efficient implementations of
+parsers with semantic actions, and integrating them into larger verified
+systems.
 

### 2025-03-21 paper/paper.tex
footnote clarifying non-commutativity

@@
 power to support arbitrarily powerful grammar formalisms: inductive
-linear types, as well as dependency of linear types on non-linear
+linear\footnote{Throughout, we shall use ``linear types'' to refer to the
+non-commutative linear types.} types, as well as dependency of linear types on non-linear
 data. The resulting system has two kinds of types: non-linear types

### 2025-03-21 paper/paper.tex
future work

--- 2025-03-21 paper/paper.tex
@@
 dependent type theory to define a grammar. For example, given a
-Turing machine $T$ one may
-define a non-linear predicate $accepts : \StringSem \to U$ that encodes that $T$
-halts and accepts an input string. Then, $Reify~accepts$ is a linear type that
-captures precisely the string the same language as $T$. That is,
+Turing machine $Tur$ one may
+define a non-linear predicate $accepts : \StringSem \to U$ such that $accept~w$ is equal to $\top$ if
+$T$ accepts $w$ and equal to $0$ otherwise. Then, $Reify~accepts$ is a linear
+type that captures precisely the strings accepted by $Tur$. That is,
 $Lang(Reify~accepts)$ is recursively enumerable --- the most general class of
 languages in the Chomsky hierarchy.
-\begin{construction}[\Agda]
-  \label{cons:turing}
-  For any Turing machine $T$, we can construct a grammar in \theoryabbv
-  that accepts the same language as $T$.
-\end{construction}
+\begin{theorem}[\Agda]
+  For any Turing machine, we can construct a grammar in \theoryabbv
+  that accepts the same language as the Turing machine.
+\end{theorem}
 
@@
 %% algebra to reason about parsing as well.
-
-\paragraph{Dependent Linear Types}
-
-Our syntax for non-commutative dependent linear type theory is based on the
-dependent commutative linear type theory of
-\citet{krishnaswami_integrating_2015}, itself an extension of Benton and
-Wadler's linear-non-linear calculus \cite{bentonMixedLinearNonlinear1995}. A
-distinct feature of these systems is that linear types and non-linear types are
-distinct sorts, so the linear logic $!$ operator is not a primitive, but is
-instead definable. Vakar develops a dependent linear type theory where instead
-the non-linear types are accessed using the $!$ modality, similar to Girard's
-original approach \cite{vakar,girard_linear_1987}. Additionally, Vakar develops
-a general categorical semantics for their system, whereas we have only developed
-a single denotational model. It appears straightforward to adapt Vakar's general
-semantics approach to a non-commutative variant that would apply to
-\theoryabbv. This may have applications in finding alternative models, or
-developing logical relations proofs categorically.
 

### 2025-03-21 paper/paper.tex
in progress of migrating typing figures

--- 2025-03-21 paper/paper.tex
@@
 \usepackage{xspace}
+\usepackage[nounderscore]{syntax}
 \usepackage{thmtools} % needed for sane cleveref naming of theorems
@@
 \newcommand{\SigTy}[3]{\textstyle\sum (#1 : #2). #3}
+\newcommand{\PiTyLimit}[3]{\textstyle\prod\limits_{#1 : #2} #3}
+\newcommand{\SigTyLimit}[3]{\textstyle\sum\limits_{#1 : #2} #3}
 \newcommand{\LinPiTy}[3]{\textstyle\bigamp (#1 : #2). #3}
@@
 \newcommand{\linctxwffjdg}[2]{#1 \vdash #2 \isLinTy}
+\newcommand{\nonlinterm}[3]{#1 \vdash #2 : #3}
+\newcommand{\linterm}[4]{#1 ; #2 \vdash #3 : #4}
 
@@
 
-For a type $\A$, a parse tree of a string $\w$ is represented as a
-term of type $\A$ in the context $\internalize \w$, where
-$\internalize \w$ is a context with one variable for each character of
-$\w$. For example, to define a parse tree for \stringquote{ab}, we use
+A grammar $A$ derives the word $w$ if there exists a derivation
+$\internalize{w} \vdash a : A$,
+where $\internalize \w$ is a context with one variable for each character of
+$\w$. The term $a : A$ represents a \emph{parse tree} of $w$ for the grammar $A$. For
+example, to define a parse tree for \stringquote{ab}, we use
 the context $\internalize{\stringquote{ab}} = a : \literal a , b :
@@
   \begin{mathpar}
-    \inferrule{~}{\ctxwff \Gamma}
-
-    \inferrule{\ctxwff \Gamma}{\ctxwffjdg \Gamma X}
+    \footnotesize
+    \boxed{\ctxwff \Gamma}
+
+    \inferrule{~}{\ctxwff \cdot}
+
+    \inferrule{\ctxwff \Gamma \and \ctxwffjdg \Gamma X}{\ctxwff {\Gamma , x : X}}
+
+    \\
+
+    \boxed{\linctxwff \Gamma \Delta}
+
+    \inferrule{~}{\linctxwff \Gamma \cdot}
+
+    \inferrule{\linctxwff \Gamma \Delta \and \linctxwffjdg \Gamma A}{\linctxwff
+      {\Gamma} {\Delta , a : A}}
+
+    \boxed{\Gamma \vdash X \isSmall}
 
@@
 
+    \\
+
+    \boxed{\ctxwffjdg \Gamma X}
+
+    \inferrule{~}{\ctxwffjdg \Gamma {Bool}}
+
+    \inferrule{~}{\ctxwffjdg \Gamma 1}
+
+    \inferrule{~}{\ctxwffjdg \Gamma \bot}
+
+    \inferrule{~}{\ctxwffjdg \Gamma {Nat}}
+
+    \inferrule{~}{\ctxwffjdg \Gamma {U}}
+
+    \inferrule{~}{\ctxwffjdg \Gamma {L}}
+
+    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg {\Gamma , x : X} {Y}}{\ctxwffjdg \Gamma {\PiTyLimit {x}{X}{Y}}}
+
+    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg {\Gamma , x : X} {Y}}{\ctxwffjdg \Gamma {\SigTyLimit {x}{X}{Y}}}
+
+    \inferrule{\nonlinterm \Gamma M U}{\ctxwffjdg \Gamma {\unquoteTy M}}
+
+    \\
+
+    \inferrule{\linctxwffjdg \Gamma A}{\ctxwffjdg \Gamma {\ltonl A}}
+
     \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg \Gamma Y}{\ctxwffjdg \Gamma {X \equiv Y}}
 
-    \inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash M : X}
-
-    \inferrule{\Gamma \vdash M : X \and \Gamma \vdash N : X}{\Gamma \vdash M \equiv N : X}
-
-    \inferrule{\ctxwff \Gamma}{\linctxwff \Gamma \Delta}
-
-    \inferrule{\ctxwff \Gamma}{\Gamma \vdash A \isLinTy}
-
-    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \equiv B}
-
-    \inferrule{\Gamma \vdash \Delta \isLinCtx \and \Gamma \vdash A \isLinTy}{\Gamma;\Delta\vdash e : A}
-
-    \inferrule{\Gamma; \Delta \vdash e : A \and \Gamma;\Delta \vdash f : A}{\Gamma;\Delta\vdash e \equiv f : A}
+    \inferrule{\nonlinterm \Gamma M X \and \nonlinterm \Gamma N X}{\ctxwffjdg \Gamma {M \equiv N}}
+
+    \\
+
+    \boxed{\linctxwffjdg \Gamma A}
+
+    \inferrule{~}{\linctxwffjdg \Gamma I}
+
+    \inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg
+      \Gamma {A \otimes B}}
+
+    \inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg
+      \Gamma {A \lto B}}
+
+    \inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg
+      \Gamma {A \tol B}}
+
+    \inferrule{\linctxwffjdg {\Gamma, x : X} A}{\linctxwffjdg
+      \Gamma {\LinSigTyLimit{x}{X}{A}}}
+
+    \inferrule{\linctxwffjdg {\Gamma, x : X} A}{\linctxwffjdg
+      \Gamma {\LinPiTyLimit{x}{X}{A}}}
+
+    \inferrule{\nonlinterm \Gamma f {\ltonl{(A \lto B)}} \and
+      \nonlinterm \Gamma g {\ltonl{(A \lto B)}}}{\linctxwffjdg \Gamma {\equalizer {a}{f}{g}}}
+
+    \inferrule{\nonlinterm \Gamma M L}{\linctxwffjdg \Gamma {\unquoteTy M}}
+
+    \inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg
+      \Gamma {A \equiv B}}
   \end{mathpar}
@@
 
+\begin{figure}
+  \begin{mathpar}
+  \footnotesize
+    \boxed{\nonlinterm \Gamma M X}
+
+    \inferrule{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma {\unquoteTy X} U}
+
+    \inferrule{\linctxwffjdg \Gamma A}{\nonlinterm \Gamma {\quoteTy A} L}
+
+    \inferrule{\linterm \Gamma \cdot e A}{\nonlinterm \Gamma e \ltonl{A}}
+  \end{mathpar}
+  \caption{Non-linear Typing}
+  \label{fig:non-linear-typing}
+\end{figure}
+
 \subsection{Non-linear Typing}
@@
 booleans, unit and empty types, and we assume function
-extensionality\cite{Hofmann_1997}. We present the
-other non-linear type constructions in
-\Cref{fig:non-linear-types}. First, we include universe types $U$ of
+extensionality\cite{Hofmann_1997}. We present a selection of
+the non-linear type constructions in
+\Cref{fig:non-linear-types} and provide the rest in
+\cref{fig:full-non-linear-types} \steven{Do we all need non-linear terms? That seems
+extra} of the appendix.
+First, we include universe types $U$ of
 small non-linear types and $L$ of linear types. These are defined as
@@
 \end{figure}
-
 \subsection{Linear Typing}
@@
 the introduction rule for $A \lto B$ adds a variable to the right side
-of the context, whereas the introduction rule (elided) for $B \tol A$ adds a
+of the context, whereas the introduction rule\footnote{Rules for $\tol$ elided
+  here, but found in the appendix in \cref{fig:extend-lin-terms}.} for $B \tol A$ adds a
 variable to the left side of the context. In our experience, because
@@
   \begin{mathpar}
-    \inferrule{}{\Gamma \vdash \I \isLinTy}\and
-    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \otimes B \isLinTy}\and
-    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \lto B \isLinTy}\and
-    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \tol B \isLinTy}\and
-    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigoplus\limits_{x:X} A \isLinTy}\and
-    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigamp\limits_{x:X} A \isLinTy}\and
-    \inferrule
-    {\Gamma \vdash f : \ltonl {(A \lto B)} \and \Gamma \vdash g : \ltonl { (A \lto B) }}
-    {\Gamma \vdash \equalizer {a}{f}{g} \isLinTy}
-  \end{mathpar}
-\begin{mathpar}
+    \boxed{\linterm \Gamma \Delta {e} {A}}
+
     \inferrule{~}{\Gamma ; a : A \vdash a : A}
@@
     \inferrule{\Gamma ; \Delta \vdash e : A \lto B \\ \Gamma ; \Delta' \vdash e' : A} {\Gamma ; \Delta, \Delta' \vdash \applto {e} {e'} : B}
+    \\
+    %
+    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
     \\
@@
   \end{footnotesize}
-  \caption{Linear types and terms (selection)}
+  \caption{Linear terms}
   \label{fig:linear-types-and-terms}
@@
 \section{Syntax}
+\begin{figure}
+  \begin{mathpar}
+    \inferrule{~}{\ctxwff \Gamma}
+
+    \inferrule{\ctxwff \Gamma}{\ctxwffjdg \Gamma X}
+
+    \inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash X \isSmall}
+
+    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg \Gamma Y}{\ctxwffjdg \Gamma {X \equiv Y}}
+
+    \inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash M : X}
+
+    \inferrule{\Gamma \vdash M : X \and \Gamma \vdash N : X}{\Gamma \vdash M \equiv N : X}
+
+    \inferrule{\ctxwff \Gamma}{\linctxwff \Gamma \Delta}
+
+    \inferrule{\ctxwff \Gamma}{\Gamma \vdash A \isLinTy}
+
+    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \equiv B}
+
+    \inferrule{\Gamma \vdash \Delta \isLinCtx \and \Gamma \vdash A \isLinTy}{\Gamma;\Delta\vdash e : A}
+
+    \inferrule{\Gamma; \Delta \vdash e : A \and \Gamma;\Delta \vdash f : A}{\Gamma;\Delta\vdash e \equiv f : A}
+  \end{mathpar}
+  \caption{Type formation rules}
+  \label{fig:formation}
+\end{figure}
+
+\begin{figure}
+  \begin{footnotesize}
+    \begin{mathpar}
+    \inferrule{}{\Gamma \vdash U \isTy}\and
+    \inferrule
+    {\Gamma \vdash M : U}
+    {\Gamma \vdash \unquoteTy M \isTy}\and
+    \inferrule
+    {\Gamma \vdash X \isTy}
+    {\Gamma \vdash \quoteTy X : U}\and
+    \unquoteTy{\quoteTy{X}} \equiv X \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}
+
+    \inferrule{}{\Gamma \vdash L \isTy}\and
+    \inferrule
+    {\Gamma \vdash M : L}
+    {\Gamma \vdash \unquoteTy M \isLinTy}\and
+    \inferrule
+    {\Gamma \vdash A \isLinTy}
+    {\Gamma \vdash \quoteTy A : L}\and
+    \unquoteTy{\quoteTy{A}} \equiv A \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}
+
+    \inferrule
+    {\Gamma \vdash A \isLinTy}
+    {\Gamma \vdash \ltonl { A } \isTy}
+    \and
+    \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash e : \ltonl A}\and
+    \inferrule{\Gamma \vdash M : \ltonl A}{\Gamma ; \cdot \vdash M : A} \and
+  \end{mathpar}
+  \end{footnotesize}
+  \caption{Non-linear types}
+  \label{fig:full-non-linear-types}
+\end{figure}
 
@@
     %
-    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
   \end{mathpar}
   \caption{Linear terms (extended)}
+  \label{fig:extend-lin-terms}
 \end{figure}

### 2025-03-21 paper/paper.tex
discuss Vakar's stuff (citation needed)

--- 2025-03-21 paper/paper.tex
@@
 
+\paragraph{Dependent Linear Types}
+
+Our syntax for non-commutative dependent linear type theory is based on the
+dependent commutative linear type theory of
+\citet{krishnaswami_integrating_2015}, itself an extension of Benton and
+Wadler's linear-non-linear calculus \cite{bentonMixedLinearNonlinear1995}. A
+distinct feature of these systems is that linear types and non-linear types are
+distinct sorts, so the linear logic $!$ operator is not a primitive, but is
+instead definable. Vakar develops a dependent linear type theory where instead
+the non-linear types are accessed using the $!$ modality, similar to Girard's
+original approach \cite{vakar,girard_linear_1987}. Additionally, Vakar develops
+a general categorical semantics for their system, whereas we have only developed
+a single denotational model. It appears straightforward to adapt Vakar's general
+semantics approach to a non-commutative variant that would apply to
+\theoryabbv. This may have applications in finding alternative models, or
+developing logical relations proofs categorically.
+
 %%    - verified imperative implementation: Lambek logic ala separation

### 2025-03-21 paper/paper.tex
non-linear figure work in progress

--- 2025-03-21 paper/paper.tex
@@
 
+    \\
+
+    \boxed{\ctxwffjdg \Gamma X}
+
+    \inferrule{~}{\ctxwffjdg \Gamma {Bool}}
+
+    \inferrule{~}{\ctxwffjdg \Gamma 1}
+
+    \inferrule{~}{\ctxwffjdg \Gamma \bot}
+
+    \inferrule{~}{\ctxwffjdg \Gamma {Nat}}
+
+    \inferrule{~}{\ctxwffjdg \Gamma {U}}
+
+    \inferrule{~}{\ctxwffjdg \Gamma {L}}
+
+    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg {\Gamma , x : X} {Y}}{\ctxwffjdg \Gamma {\PiTyLimit {x}{X}{Y}}}
+
+    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg {\Gamma , x : X} {Y}}{\ctxwffjdg \Gamma {\SigTyLimit {x}{X}{Y}}}
+
+    \inferrule{\nonlinterm \Gamma M U}{\ctxwffjdg \Gamma {\unquoteTy M}}
+
+    \inferrule{\linctxwffjdg \Gamma A}{\ctxwffjdg \Gamma {\ltonl A}}
+
+    \\
+
+    \boxed{\ctxwffjdg \Gamma {X \equiv Y}}
+
+    \inferrule{\nonlinterm \Gamma {X \equiv Y} {U}}{\ctxwffjdg \Gamma {X \equiv Y}}
+
     \boxed{\Gamma \vdash X \isSmall}
@@
 
-    \boxed{\ctxwffjdg \Gamma X}
-
-    \inferrule{~}{\ctxwffjdg \Gamma {Bool}}
-
-    \inferrule{~}{\ctxwffjdg \Gamma 1}
-
-    \inferrule{~}{\ctxwffjdg \Gamma \bot}
-
-    \inferrule{~}{\ctxwffjdg \Gamma {Nat}}
-
-    \inferrule{~}{\ctxwffjdg \Gamma {U}}
-
-    \inferrule{~}{\ctxwffjdg \Gamma {L}}
-
-    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg {\Gamma , x : X} {Y}}{\ctxwffjdg \Gamma {\PiTyLimit {x}{X}{Y}}}
-
-    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg {\Gamma , x : X} {Y}}{\ctxwffjdg \Gamma {\SigTyLimit {x}{X}{Y}}}
-
-    \inferrule{\nonlinterm \Gamma M U}{\ctxwffjdg \Gamma {\unquoteTy M}}
-
-    \\
-
-    \inferrule{\linctxwffjdg \Gamma A}{\ctxwffjdg \Gamma {\ltonl A}}
-
-    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg \Gamma Y}{\ctxwffjdg \Gamma {X \equiv Y}}
-
-    \inferrule{\nonlinterm \Gamma M X \and \nonlinterm \Gamma N X}{\ctxwffjdg \Gamma {M \equiv N}}
-
-    \\
-
     \boxed{\linctxwffjdg \Gamma A}
@@
     \inferrule{~}{\linctxwffjdg \Gamma I}
+
+    \inferrule{c \in \Sigma}{\linctxwffjdg \Gamma {\literal{c}}}
 
@@
 
-    \inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg
+    \\
+
+    \boxed{\linctxwffjdg \Gamma {A \equiv B}}
+
+    \inferrule{\nonlinterm \Gamma {A \equiv B} L}{\linctxwffjdg
       \Gamma {A \equiv B}}
+
+    \inferrule{\lintctxwffjdg \Gamma A}{\linctxwffjdg \Gamma {\unquoteTy
+        {\quoteTy A} \equiv A}}
   \end{mathpar}
@@
 
+\steven{need to write the rules for the equalities of codes
+  $M \equiv \quoteTy {\unquoteTy M} : U$ in an equational theory figure for
+  non-linear stuff? Same for the codes in universe $L$.
+
+  I will put the nonlinear equational rules, which are so
+  unnecessary, in the appendix}
 \begin{figure}
@@
 
-    \inferrule{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma {\unquoteTy X} U}
+    \inferrule{~}{\nonlinterm {\Gamma , x : X , \Gamma'} {x} {X}}
+
+    \inferrule{\nonlinterm \Gamma M Y \and \ctxwffjdg \Gamma {X \equiv Y}}{\nonlinterm {\Gamma} {M} {X}}
+
+    \inferrule{~}{\nonlinterm {\Gamma} {tt} {1}}
+
+    \inferrule{\nonlinterm \Gamma M X \and \nonlinterm \Gamma {N} {\subst Y {M}
+        {x}}}{\nonlinterm {\Gamma} {(M , N)} {\SigTyLimit {x} {X} {Y}}}
+
+    \inferrule{\nonlinterm {\Gamma} {M} {\SigTyLimit {x} {X} {Y}}}
+              {\nonlinterm {\Gamma} {M.fst} X}
+
+    \inferrule{\nonlinterm {\Gamma} {M} {\SigTyLimit {x} {X} {Y}}}
+              {\nonlinterm {\Gamma} {M.snd} {\subst Y {M.fst} {x}}}
+
+    \inferrule{\ctxwffjdg \Gamma {\PiTyLimit {x} {X} {Y}} \and \nonlinterm
+      {\Gamma , x : X} {M} {Y}}
+              {\nonlinterm {\Gamma} {\lamb {x} {M}} {\PiTyLimit {x} {X} {Y}}}
+
+    \inferrule{\nonlinterm {\Gamma} {M} {\PiTyLimit {x} {X} {Y}} \and
+      \nonlinterm \Gamma N X}
+              {\nonlinterm \Gamma {M~N} {\subst Y {N} {x}}}
+
+    \inferrule{\nonlinterm \Gamma {M \equiv N} X}{\nonlinterm {\Gamma} {refl} {M
+      =_{X} N}}
+
+    \inferrule{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma {\quoteTy X} U}
 
@@
 
+\steven{todo}
 We give an overview of the linear types and terms in
-\Cref{fig:linear-types-and-terms}. The equational theory for these
+\Cref{fig:linear-terms}. The equational theory for these
 types is straightforward $\beta\eta$ equivalence and included in the
@@
 \begin{figure}
-  \begin{footnotesize}
+  \footnotesize
   \begin{mathpar}
@@
     \inferrule{~}{\Gamma ; a : A \vdash a : A}
+    \and
+    \inferrule{\nonlinterm \Gamma {M} {\ltonl A}}{\linterm {\Gamma} {\cdot} {M} {A}}
     \and
@@
     \and
-    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
+    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e'
+      : B \tol A}{\Gamma ; \Delta, \Delta' \vdash \apptol {e'} {e} : B}
     \\
@@
     %
-    \and
+    \\
     %
@@
   \end{mathpar}
-  \end{footnotesize}
   \caption{Linear terms}
-  \label{fig:linear-types-and-terms}
+  \label{fig:linear-terms}
 \end{figure}

### 2025-03-22 paper/paper.tex
nonlinear figs, frisch/cardelli, axioms in appendix

@@
 \newcommand{\Subst}{\textrm{Subst}}
-\newcommand{\SPF}{\mathsf{SPF}}
+\newcommand{\SPF}{SPF}
 \newcommand{\Var}{\mathsf{Var}}
@@
 \newcommand{\linterm}[4]{#1 ; #2 \vdash #3 : #4}
+\newcommand{\nonlineq}[4]{#1 \vdash #2 \equiv #3 : #4}
+\newcommand{\lineq}[5]{#1 ; #2 \vdash #3 \equiv #4 : #5}
+
+\newcommand{\LLL}{\textrm{LL}}
+\newcommand{\LRR}{\textrm{LR}}
+\newcommand{\LALRR}{\textrm{LALR}}
+\newcommand{\LL}[1]{\LLL(#1)}
+\newcommand{\LR}[1]{\LRR(#1)}
+\newcommand{\LALR}[1]{\LALRR(#1)}
 
@@
 source program. Eventually, a correct parser was implemented using an
-automaton that is formally verified to implement an LR grammar
+automaton that is formally verified to implement an $\LR{1}$ grammar
 \cite{jourdanValidatingLRParsers2012}.
@@
 particular grammar formalism such as non-left-recursive grammars or
-LL(1) grammars
+$\LL{1}$ grammars
 \cite{lasserCoStarVerifiedALL2021,EdelmannZippy2020,danielssonTotalParserCombinators2010}.
@@
 non-deterministic and deterministic automata, as well as some example
-context-free grammars and parsers based on LL(1) and LR(1)
+context-free grammars and parsers based on $\LL{1}$ and $\LR{1}$
 automata. Throughout this paper, we will use \agdalogo~ to mark
@@
 definitional equality judgments for both kinds of type and term
-judgments as well. Additionally, we have a judgment $\Gamma\vdash X
+judgments as well. Additionally, we have a judgment $\Gamma \vdash X
 \isSmall$ which is used in the definition of universe types.
@@
 
-    \\
-
     \boxed{\linctxwff \Gamma \Delta}
@@
       {\Gamma} {\Delta , a : A}}
+  \end{mathpar}
+  \caption{Context well-formedness rules}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \footnotesize
+    \boxed{\ctxwffjdg \Gamma X}
+
+    \inferrule{~}{\ctxwffjdg \Gamma {U}}
+
+    \inferrule{~}{\ctxwffjdg \Gamma {L}}
+
+    \inferrule{~}{\ctxwffjdg \Gamma 1}
+
+    \inferrule{~}{\ctxwffjdg \Gamma \bot}
+
+    \inferrule{~}{\ctxwffjdg \Gamma {Nat}}
+
+    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg {\Gamma , x : X} {Y}}{\ctxwffjdg \Gamma {\PiTyLimit {x}{X}{Y}}}
+
+    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg {\Gamma , x : X} {Y}}{\ctxwffjdg \Gamma {\SigTyLimit {x}{X}{Y}}}
+
+    \inferrule{\nonlinterm \Gamma M U}{\ctxwffjdg \Gamma {\unquoteTy M}}
+
+    \inferrule{\linctxwffjdg \Gamma A}{\ctxwffjdg \Gamma {\ltonl A}}
+
+    \inferrule{\ctxwffjdg \Gamma X \and \nonlinterm \Gamma M X \and \nonlinterm \Gamma N X}{\ctxwffjdg \Gamma {M =_{X} N}}
 
@@
 
-    \boxed{\ctxwffjdg \Gamma X}
-
-    \inferrule{~}{\ctxwffjdg \Gamma {Bool}}
-
-    \inferrule{~}{\ctxwffjdg \Gamma 1}
-
-    \inferrule{~}{\ctxwffjdg \Gamma \bot}
-
-    \inferrule{~}{\ctxwffjdg \Gamma {Nat}}
-
-    \inferrule{~}{\ctxwffjdg \Gamma {U}}
-
-    \inferrule{~}{\ctxwffjdg \Gamma {L}}
-
-    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg {\Gamma , x : X} {Y}}{\ctxwffjdg \Gamma {\PiTyLimit {x}{X}{Y}}}
-
-    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg {\Gamma , x : X} {Y}}{\ctxwffjdg \Gamma {\SigTyLimit {x}{X}{Y}}}
-
-    \inferrule{\nonlinterm \Gamma M U}{\ctxwffjdg \Gamma {\unquoteTy M}}
-
-    \inferrule{\linctxwffjdg \Gamma A}{\ctxwffjdg \Gamma {\ltonl A}}
-
-    \\
-
     \boxed{\ctxwffjdg \Gamma {X \equiv Y}}
@@
     \inferrule{\nonlinterm \Gamma {X \equiv Y} {U}}{\ctxwffjdg \Gamma {X \equiv Y}}
+
+    \inferrule{\linctxwffjdg \Gamma X}{\linctxwffjdg \Gamma {\unquoteTy
+        {\quoteTy X} \equiv X}}
 
@@
 
-    \inferrule{\lintctxwffjdg \Gamma A}{\linctxwffjdg \Gamma {\unquoteTy
+    \inferrule{\linctxwffjdg \Gamma A}{\linctxwffjdg \Gamma {\unquoteTy
         {\quoteTy A} \equiv A}}
   \end{mathpar}
-  \caption{Formation rules}
+  \caption{Type formation rules}
   \label{fig:formation}
@@
 
-\steven{need to write the rules for the equalities of codes
-  $M \equiv \quoteTy {\unquoteTy M} : U$ in an equational theory figure for
-  non-linear stuff? Same for the codes in universe $L$.
-
-  I will put the nonlinear equational rules, which are so
-  unnecessary, in the appendix}
 \begin{figure}
@@
 
-    \inferrule{~}{\nonlinterm {\Gamma , x : X , \Gamma'} {x} {X}}
-
-    \inferrule{\nonlinterm \Gamma M Y \and \ctxwffjdg \Gamma {X \equiv Y}}{\nonlinterm {\Gamma} {M} {X}}
-
-    \inferrule{~}{\nonlinterm {\Gamma} {tt} {1}}
-
-    \inferrule{\nonlinterm \Gamma M X \and \nonlinterm \Gamma {N} {\subst Y {M}
-        {x}}}{\nonlinterm {\Gamma} {(M , N)} {\SigTyLimit {x} {X} {Y}}}
-
-    \inferrule{\nonlinterm {\Gamma} {M} {\SigTyLimit {x} {X} {Y}}}
-              {\nonlinterm {\Gamma} {M.fst} X}
-
-    \inferrule{\nonlinterm {\Gamma} {M} {\SigTyLimit {x} {X} {Y}}}
-              {\nonlinterm {\Gamma} {M.snd} {\subst Y {M.fst} {x}}}
-
-    \inferrule{\ctxwffjdg \Gamma {\PiTyLimit {x} {X} {Y}} \and \nonlinterm
-      {\Gamma , x : X} {M} {Y}}
-              {\nonlinterm {\Gamma} {\lamb {x} {M}} {\PiTyLimit {x} {X} {Y}}}
-
-    \inferrule{\nonlinterm {\Gamma} {M} {\PiTyLimit {x} {X} {Y}} \and
-      \nonlinterm \Gamma N X}
-              {\nonlinterm \Gamma {M~N} {\subst Y {N} {x}}}
-
-    \inferrule{\nonlinterm \Gamma {M \equiv N} X}{\nonlinterm {\Gamma} {refl} {M
-      =_{X} N}}
-
     \inferrule{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma {\quoteTy X} U}
@@
     \inferrule{\linterm \Gamma \cdot e A}{\nonlinterm \Gamma e \ltonl{A}}
+
+    \inferrule{\nonlinterm \Gamma {M \equiv N} {X}}
+      {\nonlinterm \Gamma {refl} {M =_{X} N}}
   \end{mathpar}
-  \caption{Non-linear Typing}
-  \label{fig:non-linear-typing}
+  \caption{Non-linear typing (selection)}
+  \label{fig:non-linear-types}
 \end{figure}
@@
 The non-linear types of \theoryabbv include the standard dependent
-types for $\Pi,\Sigma$, extensional equality, natural numbers,
-booleans, unit and empty types, and we assume function
+types for $\Pi$,$\Sigma$, extensional equality, natural numbers,
+\steven{,just take Booleans to be implemented via $1$ and $\Sigma$,} unit, and empty types, and we assume function
 extensionality\cite{Hofmann_1997}. We present a selection of
@@
 \Cref{fig:non-linear-types} and provide the rest in
-\cref{fig:full-non-linear-types} \steven{Do we all need non-linear terms? That seems
-extra} of the appendix.
+\cref{fig:full-non-linear-types, fig:jdg-eq-nonlinear} of the appendix.
 First, we include universe types $U$ of
@@
 types are needed so that we can define types by recursion on natural
-numbers. Next, we include a non-linear type $\ltonl A$ where $A$ is a
+numbers. As in \cite{krishnaswami_integrating_2015}, the extensional equality type
+$M =_{X} N$ used here has the introduction form $refl$, but no
+elimination due to the presence of equality reflection. Next, we include a non-linear type $\ltonl A$ where $A$ is a
 linear type. The intuition for this type is that its elements are the
@@
 
-%% \begin{figure}
-%%   \[
-%%   \begin{array}{rrcl}
-%%     \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
-%%     \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe
-%%                                                   \mathsf{Bool} \pipe
-%%                                                   \mathsf{Empty} \pipe
-%%                                                   \mathsf{Unit} \pipe
-%%                                                   \mathsf{SPF}\,X\pipe U \pipe L
-%%                                                   \pipe \ltonl A \pipe \sum\limits_{x:X} Y
-%%                                                   \pipe \prod\limits_{x:X} Y\pipe M =_A
-%%                                                   N
-%%   \end{array}
-%%   \]
-%%   \caption{Non-linear Types}
-%% \end{figure}
-\begin{figure}
-  \begin{footnotesize}
-    \begin{mathpar}
-    \inferrule{}{\Gamma \vdash U \isTy}\and
-    \inferrule
-    {\Gamma \vdash M : U}
-    {\Gamma \vdash \unquoteTy M \isTy}\and
-    \inferrule
-    {\Gamma \vdash X \isTy}
-    {\Gamma \vdash \quoteTy X : U}\and
-    \unquoteTy{\quoteTy{X}} \equiv X \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}
-
-    \inferrule{}{\Gamma \vdash L \isTy}\and
-    \inferrule
-    {\Gamma \vdash M : L}
-    {\Gamma \vdash \unquoteTy M \isLinTy}\and
-    \inferrule
-    {\Gamma \vdash A \isLinTy}
-    {\Gamma \vdash \quoteTy A : L}\and
-    \unquoteTy{\quoteTy{A}} \equiv A \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}
-
-    \inferrule
-    {\Gamma \vdash A \isLinTy}
-    {\Gamma \vdash \ltonl { A } \isTy}
-    \and
-    \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash e : \ltonl A}\and
-    \inferrule{\Gamma \vdash M : \ltonl A}{\Gamma ; \cdot \vdash M : A} \and
-  \end{mathpar}
-  \end{footnotesize}
-  \caption{Non-linear types (selection)}
-  \label{fig:non-linear-types}
-\end{figure}
 \subsection{Linear Typing}
-
-\steven{todo}
-We give an overview of the linear types and terms in
-\Cref{fig:linear-terms}. The equational theory for these
+We give the rules for linear type
+formation in \cref{fig:formation} and the definition of linear terms in
+\cref{fig:linear-terms}. The equational theory for these
 types is straightforward $\beta\eta$ equivalence and included in the
@@
 the introduction rule for $A \lto B$ adds a variable to the right side
-of the context, whereas the introduction rule\footnote{Rules for $\tol$ elided
-  here, but found in the appendix in \cref{fig:extend-lin-terms}.} for $B \tol A$ adds a
+of the context, whereas the introduction rule for $B \tol A$ adds a
 variable to the left side of the context. In our experience, because
@@
 
-\begin{corollary}
+\steven{I'm not sure if this is the proper way to cite the corollary, especially
+because the citation is a reference, not the originator of the idea}
+\begin{corollary}[\cite{Cockett_1993}]
   \label{cor:binary-mono}
@@
 \end{corollary}
-
-\steven{cite reference for distributivity implies coprojections mono. Cockett
-  ``Introduction to Distributive Categories''? Awodey?}
 
@@
 input type of $\mathsf{read}$ is $\top$, which can control any amount
-of resources, and not $\I$ which controls no resources. Further, we
-add an axiom that $\lambda s. \mathsf{read}(!(s)) \equiv \lambda
-s. s$ where $!$ is the unique function $\ltonl(\StringGram\lto \top)$,
-i.e., that if we have a string, but then throw it away and read
-it from the input, then in fact that is equivalent to the string we
-were given originally. This ensures that the elements of the $\StringGram$
+of resources, and not $\I$ which controls no resources.
+
+\begin{axiom}
+  \label{ax:string-top}
+   $\lambda s. \mathsf{read}(!(s)) \equiv \lambda
+s. s$ where $!$ is the unique function $\ltonl(\StringGram\lto \top)$. That is, $\StringGram$ is strongly equivalent to $\top$.
+\end{axiom}
+
+If we have a string, but then throw it away and read
+it from the input, then we, in fact, recover the original string.
+This ensures that the elements of the $\StringGram$
 type always stand for the actual input string in our reasoning. In the
@@
 
-\steven{
 With \cref{ax:dist}, it is straightforward to implement a parser for this lookahead automaton,
@@
 necessitating the distributivity axiom.
-}
 
@@
 \begin{figure}
+  \begin{minipage}[t]{.5\textwidth}
   \begin{footnotesize}
-    \begin{align*}
-    \sem{c}\gamma\,w &= \{ c | w = c\}\\
-    \sem{I}\gamma\,w &= \{ () \pipe w = \varepsilon \}\\
-    \sem{A \otimes B}\gamma\,w &= \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
-    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
-    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
-    \sem{\bigoplus_{x:X} A}\gamma\,w &= \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
-    \sem{\bigwith_{x:X} A}\gamma\,w &= \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w\\
-    \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
-    \sem{\equalizer{a}{f}{g}}\gamma\,w &= \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
-    \sem{L}\gamma &= \Grammar_0\\
-    \sem{\SPF\,X}\gamma &= \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
-    \sem{\el(F)}\gamma\,G &= \sem{F}\gamma\,G\\
-    \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
-    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma)\\
-  \end{align*}
+    \begin{flalign*}
+    &\sem{c}\gamma\,w = \{ c | w = c\}\\
+    &\sem{I}\gamma\,w = \{ () \pipe w = \varepsilon \}\\
+    &\sem{A \otimes B}\gamma\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
+    &\sem{A \lto B}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
+    &\sem{B \tol A}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
+    &\sem{\bigoplus_{x:X} A}\gamma\,w = \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
+    &\sem{\bigwith_{x:X} A}\gamma\,w = \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w
+  \end{flalign*}
   \end{footnotesize}
+  \end{minipage}%
+  \begin{minipage}[t]{.5\textwidth}
+  \begin{footnotesize}
+    \begin{flalign*}
+    &\sem{\uparrow A}\gamma = \sem{A}\gamma\,\varepsilon\\
+    &\sem{\equalizer{a}{f}{g}}\gamma\,w = \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
+    &\sem{L}\gamma = \Grammar_0\\
+    &\sem{SPF\,X}\gamma = \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
+    &\sem{\el(F)}\gamma\,G = \sem{F}\gamma\,G\\
+    &\sem{\map(F)}\gamma\,f = \sem{F}\gamma\,f\\
+    &\sem{\mu A}\gamma = \mu (\sem{A}\gamma)
+  \end{flalign*}
+  \end{footnotesize}
+  \end{minipage}
   \caption{Grammar Semantics}
@@
 The remaining details of the interpretation of linear terms as parse
-transformers and verification of the equational axioms is a relatively
+transformers and verification of the equational axioms --- as well as
+\cref{ax:dist,ax:disjointness,ax:string-top} --- is a relatively
 straightforward extension of existing semantics of linear logic in
@@
 The most similar prior work to our own is Luo's Lambek calculus with
-dependent types \cite{luo}. Their design differs from our own in
-allowing linear types to depend on other linear types and supporting
+dependent types \cite{luo}. They present two systems: one like ours where linear
+types may only depend on non-linear ones, and another that
+allows linear types to depend on other linear types and supports
 directed $\Pi$ and $\Sigma$ types that have no analog in our
@@
 %% \subsubsection{Typed Grammars}
-Frisch and Cardelli introduced the use of a simple type system to reason about regular expressions up to weak equivalence \cite{frischCardelli}.
+Frisch and Cardelli introduced the use of a simple type system to reason about
+regular expressions up to weak equivalence \cite{frischCardelli}. \steven{They
+  use their system in the design of a linear-time algorithm for greedy regular
+  expression matching. It is yet unclear if greediness can be encoded in
+  \theoryname; however, it appears that our approach in determinizing an NFA is
+  quite similar to their algorithm. The biggest difference being that our
+  approach uses a global disambiguation strategy while theirs is a local, greedy strategy.
+}
 Henglein and Nielsen build on this type-theoretic view of grammars, pointing out that the values of the types correspond precisely to the parse trees \cite{henglein_regular_2011}.
@@
 We also aim to adapt the approach used for our context-free grammar parserss for
-LL and LR/LALR with the aim of developing a shared library of intrinsically
+$\LLL$ and $\LRR$/$\LALRR$ with the aim of developing a shared library of intrinsically
 verified parsing utilities that can be usefully shared across different
@@
 \appendix
-
 \section{Syntax}
-\begin{figure}
-  \begin{mathpar}
-    \inferrule{~}{\ctxwff \Gamma}
-
-    \inferrule{\ctxwff \Gamma}{\ctxwffjdg \Gamma X}
-
-    \inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash X \isSmall}
-
-    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg \Gamma Y}{\ctxwffjdg \Gamma {X \equiv Y}}
-
-    \inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash M : X}
-
-    \inferrule{\Gamma \vdash M : X \and \Gamma \vdash N : X}{\Gamma \vdash M \equiv N : X}
-
-    \inferrule{\ctxwff \Gamma}{\linctxwff \Gamma \Delta}
-
-    \inferrule{\ctxwff \Gamma}{\Gamma \vdash A \isLinTy}
-
-    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \equiv B}
-
-    \inferrule{\Gamma \vdash \Delta \isLinCtx \and \Gamma \vdash A \isLinTy}{\Gamma;\Delta\vdash e : A}
-
-    \inferrule{\Gamma; \Delta \vdash e : A \and \Gamma;\Delta \vdash f : A}{\Gamma;\Delta\vdash e \equiv f : A}
-  \end{mathpar}
-  \caption{Type formation rules}
-  \label{fig:formation}
-\end{figure}
-
-\begin{figure}
-  \begin{footnotesize}
-    \begin{mathpar}
-    \inferrule{}{\Gamma \vdash U \isTy}\and
-    \inferrule
-    {\Gamma \vdash M : U}
-    {\Gamma \vdash \unquoteTy M \isTy}\and
-    \inferrule
-    {\Gamma \vdash X \isTy}
-    {\Gamma \vdash \quoteTy X : U}\and
-    \unquoteTy{\quoteTy{X}} \equiv X \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}
-
-    \inferrule{}{\Gamma \vdash L \isTy}\and
-    \inferrule
-    {\Gamma \vdash M : L}
-    {\Gamma \vdash \unquoteTy M \isLinTy}\and
-    \inferrule
-    {\Gamma \vdash A \isLinTy}
-    {\Gamma \vdash \quoteTy A : L}\and
-    \unquoteTy{\quoteTy{A}} \equiv A \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}
-
-    \inferrule
-    {\Gamma \vdash A \isLinTy}
-    {\Gamma \vdash \ltonl { A } \isTy}
-    \and
-    \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash e : \ltonl A}\and
-    \inferrule{\Gamma \vdash M : \ltonl A}{\Gamma ; \cdot \vdash M : A} \and
-  \end{mathpar}
-  \end{footnotesize}
-  \caption{Non-linear types}
-  \label{fig:full-non-linear-types}
-\end{figure}
-
 \begin{figure}
@@
   \begin{mathpar}
-    \inferrule{~}{\ctxwff \cdot}
-    \and
-    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}
-    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}
-
-    \boxed{\linctxwffjdg \Gamma A}
-
-    \inferrule{~}{\linctxwff \Gamma \cdot}
-    \and
-    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}
+  \footnotesize
+    \boxed{\nonlinterm \Gamma M X}
+
+    \inferrule{~}{\nonlinterm {\Gamma , x : X , \Gamma'} {x} {X}}
+
+    \inferrule{\nonlinterm \Gamma M Y \and \ctxwffjdg \Gamma {X \equiv Y}}{\nonlinterm {\Gamma} {M} {X}}
+
+    \inferrule{~}{\nonlinterm {\Gamma} {tt} {1}}
+
+    \inferrule
+    {\nonlinterm \Gamma M \bot \and \ctxwffjdg \Gamma X}
+    {\nonlinterm \Gamma {()} {\bot}}
+
+    \inferrule{~}{\nonlinterm {\Gamma} {0} {Nat}}
+
+    \inferrule{\nonlinterm {\Gamma} {n} {Nat}}{\nonlinterm {\Gamma} {suc~n} {Nat}}
+
+    \inferrule{\ctxwffjdg {\Gamma , n : Nat} {X(n)} \and
+           \nonlinterm {\Gamma} {M_0} {X(0)} \and
+           \nonlinterm {\Gamma, n : Nat, x : X(n)} {M_{suc}(n , x)} {X(suc~n)}}
+          {\nonlinterm {\Gamma} {elim_{Nat}(X,M_0,M_{suc})(n)} {X(n)}}
+
+    \inferrule{\nonlinterm \Gamma M X \and \nonlinterm \Gamma {N} {\subst Y {M}
+        {x}}}{\nonlinterm {\Gamma} {(M , N)} {\SigTyLimit {x} {X} {Y}}}
+
+    \inferrule{\nonlinterm {\Gamma} {M} {\SigTyLimit {x} {X} {Y}}}
+              {\nonlinterm {\Gamma} {M.fst} X}
+
+    \inferrule{\nonlinterm {\Gamma} {M} {\SigTyLimit {x} {X} {Y}}}
+              {\nonlinterm {\Gamma} {M.snd} {\subst Y {M.fst} {x}}}
+
+    \inferrule{\ctxwffjdg \Gamma {\PiTyLimit {x} {X} {Y}} \and \nonlinterm
+      {\Gamma , x : X} {M} {Y}}
+              {\nonlinterm {\Gamma} {\lamb {x} {M}} {\PiTyLimit {x} {X} {Y}}}
+
+    \inferrule{\nonlinterm {\Gamma} {M} {\PiTyLimit {x} {X} {Y}} \and
+      \nonlinterm \Gamma N X}
+              {\nonlinterm \Gamma {M~N} {\subst Y {N} {x}}}
+
+    \inferrule{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma {\quoteTy X} U}
+
+    \inferrule{\linctxwffjdg \Gamma A}{\nonlinterm \Gamma {\quoteTy A} L}
+
+    \inferrule{\linterm \Gamma \cdot e A}{\nonlinterm \Gamma e \ltonl{A}}
   \end{mathpar}
-  \caption{Context rules}
+  \caption{Non-linear Typing}
+  \label{fig:full-non-linear-types}
 \end{figure}
-
-%% \begin{figure}
-%%   \begin{mathpar}
-%%   \boxed{\Gamma \vdash M : X}
-
-%%   \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
-%%   %
-%%   \and
-%%   %
-%%   \inferrule{\Gamma \vdash M : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash M : X}
-%%   %
-%%   \and
-%%   %
-%%   \inferrule{~}{\Gamma \vdash () : 1}
-%%   %
-%%   \and
-%%   %
-%%   \inferrule{\Gamma \vdash M : X \\ \Gamma \vdash N : \subst Y M x}{\Gamma \vdash (M, N) : \SigTy x X Y}
-%%   %
-%%   \and
-%% %
-%%   \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_1\, M : X}
-%%   %
-%%   \and
-%%   %
-%%   \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_2\, M : \subst Y {\pi_1\, M} x}
-%%   \and
-%%   \inferrule{\Gamma, x : X \vdash M : Y}{\Gamma \vdash \lamb x M : \PiTy x X Y}
-%%   %
-%%   \and
-%%   %
-%%   \inferrule{\Gamma \vdash M : \PiTy x X Y \\ \Gamma \vdash N : X}{\Gamma \vdash \app M {N} : \subst Y {N} x}
-%%   %
-%%   \and
-%%   %
-%%   \inferrule{\Gamma \vdash M \equiv N : X}{\Gamma \vdash \mathsf{refl} : M =_X N}
-%%   \and
-%%   \textrm{TODO: inductive types (boolean, nat) with their eliminators allowed in }
-%%   \end{mathpar}
-%%   \caption{Intuitionistic typing}
-%%   \label{fig:inttyping}
-%% \end{figure}
 
@@
   \begin{mathpar}
-    %
+  \boxed{\nonlinterm \Gamma {M \equiv N} {X}}
+
+  \inferrule{\nonlinterm \Gamma M 1 \and \nonlinterm \Gamma N 1}{\nonlineq \Gamma M N 1}
+
+  \inferrule{\nonlinterm \Gamma M \bot \and \nonlinterm \Gamma N \bot}{\nonlineq \Gamma M N \bot}
+
+  \inferrule{\nonlinterm {\Gamma , x : X} {M} {Y} \and \nonlinterm {\Gamma} {N} {X}}{\nonlineq \Gamma {\left(\lamb {x} {M} \right)~N} {\subst M {N}
+      {x}} {X}}
+
+  \inferrule{\nonlinterm \Gamma M {\PiTyLimit {x}{X}{Y}}}{\nonlineq \Gamma {M} {\lamb {x} {M~x}} {\PiTyLimit{x}{X}{Y}}}
+
+  \inferrule{\nonlinterm \Gamma M {X} \and \nonlinterm {\Gamma , x : X} {N}
+    {\subst Y {M} {x}}}{\nonlineq \Gamma {(M , N).fst} {M} {X}}
+
+  \inferrule{\nonlinterm \Gamma M {X} \and \nonlinterm {\Gamma , x : X} {N}
+    {\subst Y {M} {x}}}{\nonlineq \Gamma {(M , N).snd} {N} {\subst Y {M} {x}}}
+
+  \inferrule{\nonlinterm \Gamma {M} {\SigTy {x} {X} {Y}}}{\nonlineq \Gamma {(M.fst , M.snd)} {M} {\SigTyLimit{x}{X}{Y}}}
+
+  \inferrule{\nonlinterm {\Gamma} {M_0} {X(0)} \and
+      \nonlinterm {\Gamma, n : Nat, x : X(n)} {M_{suc}(n , x)} {X(suc~n)}}
+      {\nonlineq {\Gamma} {elim_{Nat}(X,M_0,M_{suc})(0)} {M_0} {X(0)}}
+
+  \inferrule{\nonlinterm {\Gamma} {M_0} {X(0)} \and
+      \nonlinterm {\Gamma, n : Nat, x : X(n)} {M_{suc}(n , x)} {X(suc~n)} \and
+      \nonlinterm {\Gamma} {n} {Nat}}
+      {\nonlineq {\Gamma} {elim_{Nat}(X,M_0,M_{suc})(suc~n)} {M_{suc}(n ,
+          elim_{Nat}(X,M_0,M_{suc})(n))} {X(suc~n)}}
+
+  \inferrule{\nonlinterm \Gamma M U}{\nonlineq \Gamma {\quoteTy {\unquoteTy {M}}}
+  {M} {U}}
+
+  \inferrule{\nonlinterm \Gamma M L}{\nonlineq \Gamma {\quoteTy {\unquoteTy {M}}}
+  {M} {L}}
   \end{mathpar}
-  \caption{Linear terms (extended)}
-  \label{fig:extend-lin-terms}
+  \caption{Judgmental equality for non-linear terms}
+  \label{fig:jdg-eq-nonlinear}
 \end{figure}
@@
 %
-    \inferrule{\Gamma ; a : A , \Delta \vdash e : C \\ \Gamma ; \Delta' \vdash e' : A }{\Gamma; \Delta , \Delta' \vdash \app {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
+    \inferrule{\Gamma ; a : A , \Delta \vdash e : C \\ \Gamma ; \Delta' \vdash e' : A }{\Gamma; \Delta , \Delta' \vdash \apptol {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
 %
@@
 %
-    \inferrule{\Gamma ; \Delta \vdash e : B \tol A}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : B \tol A}
+    \inferrule{\Gamma ; \Delta \vdash e : B \tol A}{\Gamma; \Delta \vdash e \equiv \lambtol a {\apptol e a} : B \tol A}
 %
@@
 \end{figure}
-\steven{Need to include premises for the judgmental equality rules}
 
@@
     \inferrule
-    {\Gamma \vdash X \isTy}
-    {\Gamma' \vdash X[\gamma] \isTy}
+    {\ctxwffjdg \Gamma X}
+    {\ctxwffjdg {\Gamma'} {X[\gamma]}}
 
@@
     \inferrule
-    {\Gamma \vdash X \equiv Y}
-    {\Gamma' \vdash X[\gamma] \equiv Y[\gamma]}
+    {\ctxwffjdg \Gamma {X \equiv Y}}
+    {\ctxwffjdg {\Gamma'} {X[\gamma] \equiv Y[\gamma]}}
 
     \inferrule
-    {\Gamma \vdash M : X}
-    {\Gamma' \vdash M[\gamma] : X[\gamma]}
+    {\nonlinterm \Gamma {M} {X}}
+    {\nonlinterm {\Gamma'} {M[\gamma]} {X[\gamma]}}
 
     \inferrule
-    {\Gamma \vdash M \equiv N : X}
-    {\Gamma' \vdash M[\gamma] \equiv N[\gamma] : X[\gamma]}
+    {\nonlineq \Gamma {M} {N} {X}}
+    {\nonlineq {\Gamma'} {M[\gamma]} {N[\gamma]} {X[\gamma]}}
 
@@
     \inferrule
-    {\Gamma \vdash A \equiv B}
-    {\Gamma' \vdash A[\gamma] \equiv B[\gamma]}
+    {\linctxwffjdg \Gamma {A \equiv B}}
+    {\linctxwffjdg {\Gamma'} {A[\gamma] \equiv B[\gamma]}}
 
@@
     {\Gamma;\Delta \vdash e : A}
-    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] : A[\gamma]}j
+    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] : A[\gamma]}
 
@@
     {\Gamma;\Delta \vdash e \equiv f : A}
-    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] \equiv f[\gamma] : A[\gamma]}j
+    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] \equiv f[\gamma] : A[\gamma]}
   \end{mathpar}  
@@
 \end{figure}
-\steven{captilizataion should be consistent over all figures. idk if all caps or
-all lowercase is beter?}
 
@@
 \end{proof}
+
+\subsection{Grammar Semantics Respects Additional Axioms}
+We verify that the denotational semantics verfies each of the axioms we have
+assumed.
+
+\paragraph{\cref{ax:dist}}
+\begin{theorem}
+  In $\Grammar$, $\semg{\lamblto{e}{\letin{\sigma\,f\,e'}{e}{\withlamb{x}{e'\,.\pi\,x}}}}$ has
+  an inverse.
+\end{theorem}
+\begin{proof}
+  \steven{Double check this}.
+  The following map forms the desired inverse,
+
+  \[
+    \lamb {\gamma} {\lamb {w} {\lamb {p} {(\lamb {\lamb {x} {(p~x).fst}} , \lamb {x} {(p~x).snd})}}}
+  \]
+\end{proof}
+
+\paragraph{\cref{ax:disjointness}}
+\begin{theorem}
+  In $\Grammar$, $\semg{\sigma x}$ and $\semg{\sigma x'}$ are disjoint for
+  $x \neq x'$.
+\end{theorem}
+\begin{proof}
+  This is trivially true, as the denotation of linear sum types is as $\Sigma$
+  types in $\Grammar$. For any input, the first projections of $\semg{\sigma x}$
+  is $x$ and the first projection of
+  $\semg{\sigma x'}$ is $x'$. Because $x \neq x'$, the images of
+  $\semg{\sigma x}$ and $\semg{\sigma x'}$ cannot agree.
+\end{proof}
+
+\paragraph{\cref{ax:string-top}}
+\begin{theorem}
+  In $\Grammar$, $\semg{!}$ has an inverse.
+\end{theorem}
+\begin{proof}
+  Because $\semgg{\top}\gamma\,w$ is a singleton set, it suffices to show that
+  $\semg{\StringGram}\gamma\,w$ is likewise a singleton set for all $\gamma$ and
+  $w$.
+
+  First, we prove by induction on $w$ that $\semg{\StringGram}$ is a retract of
+  $\semg{\LinSigTyLimit{w}{\StringSem}{\internalize{w}}}$. Then, again by
+  induction on $w$, we show that
+  $\semg{\LinSigTyLimit{w}{\StringSem}{\internalize{w}}}$ is isomorphic to
+  $\semg{\top}$. Thus, $\semg{\StringGram}$ is a retract of $\semg{\top}$. In particular,
+  so $\semg{\StringGram}\gamma\,w$ is a singleton, and thus
+  $\semg{\StringGram}\gamma\,w \cong \semg{\top}\gamma\,w$ in $\Set$.
+\end{proof}
+
 \fi

### 2025-03-22 paper/paper.tex
move frisch/cardelli

@@
 Frisch and Cardelli introduced the use of a simple type system to reason about
-regular expressions up to weak equivalence \cite{frischCardelli}. \steven{They
-  use their system in the design of a linear-time algorithm for greedy regular
-  expression matching. It is yet unclear if greediness can be encoded in
-  \theoryname; however, it appears that our approach in determinizing an NFA is
-  quite similar to their algorithm. The biggest difference being that our
-  approach uses a global disambiguation strategy while theirs is a local, greedy strategy.
-}
+regular expressions up to weak equivalence \cite{frischCardelli}.
 Henglein and Nielsen build on this type-theoretic view of grammars, pointing out that the values of the types correspond precisely to the parse trees \cite{henglein_regular_2011}.
@@
 verified easily as well.
+
+\steven{
+  Is any of this detail useful here, or nah?
+
+  They
+  use their system in the design of a linear-time algorithm for greedy regular
+  expression matching. It is yet unclear if greediness can be encoded in
+  \theoryname; however, it appears that our approach in determinizing an NFA is
+  quite similar to their algorithm. The biggest difference being that our
+  approach uses a global disambiguation strategy while theirs is a local, greedy strategy.
+}
 

### 2025-03-23 paper/paper.tex
need to write up non linear inductive rules

@@
 \begin{figure}
-  \begin{tikzpicture}[node distance = 25mm ]
-    \node[state, initial] (0) {$\0$};
-    \node[state, below left of=0] (1) {$\1$};
-    \node[state, right of=1, accepting] (2) {$\2$};
-
-    \path[->] (0) edge[above] node{$\epsilon$} (1)
-              (0) edge[right] node{$\stringquote{c}$} (2)
-              (1) edge[loop left] node{$\stringquote{a}$} (1)
-              (1) edge[below] node{$\stringquote{b}$} (2);
-  \end{tikzpicture}
+  \begin{minipage}[t]{.6\textwidth}
+    \vspace{0pt}
   \begin{lstlisting}
@@
 \end{lstlisting}
+  \end{minipage}%
+  \begin{minipage}[t]{.4\textwidth}
+  \vspace{0pt}
+  \hspace{10pt}
+  \begin{tikzpicture}[node distance = 25mm ]
+    \node[state, initial] (0) {$\0$};
+    \node[state, below left of=0] (1) {$\1$};
+    \node[state, right of=1, accepting] (2) {$\2$};
+
+    \path[->] (0) edge[above] node{$\epsilon$} (1)
+              (0) edge[right] node{$\stringquote{c}$} (2)
+              (1) edge[loop left] node{$\stringquote{a}$} (1)
+              (1) edge[below] node{$\stringquote{b}$} (2);
+  \end{tikzpicture}
+  \end{minipage}%
   \caption{NFA for $(\a^{*} \otimes \b) \oplus \c$ and its corresponding type}
@@
 
+\steven{TODO define general inductives as the least fixed point of strictly
+  positive endofunctors on $U$}
+
+\steven{TODO linear $\mu$ in the type formation rules? nonlinear $\mu$ in the
+  type formation rules?}
+
 \begin{figure}
@@
 
-    \inferrule{~}{\ctxwffjdg \Gamma {Nat}}
+    % \inferrule{~}{\ctxwffjdg \Gamma {Nat}}
 
@@
 The non-linear types of \theoryabbv include the standard dependent
-types for $\Pi$,$\Sigma$, extensional equality, natural numbers,
-\steven{,just take Booleans to be implemented via $1$ and $\Sigma$,} unit, and empty types, and we assume function
-extensionality\cite{Hofmann_1997}. We present a selection of
+types for $\Pi$,$\Sigma$, extensional equality, unit and empty types, and
+inductive types that are defined strictly positively. We may also assume function
+extensionality\cite{Hofmann_1997}; Function
+extensionality is holds in Cubical Agda, and in our implementation we make heavy use of it of when
+validating that the denotational semantics respects the equational theory of
+\theoryabbv. However, it is not necessary to assume function extensionality for
+the non-linear types in \theoryabbv, as we don't utilize it in any of our constructions.
+
+We present a selection of
 the non-linear type constructions in
@@
 
-    \inferrule{~}{\nonlinterm {\Gamma} {0} {Nat}}
-
-    \inferrule{\nonlinterm {\Gamma} {n} {Nat}}{\nonlinterm {\Gamma} {suc~n} {Nat}}
-
-    \inferrule{\ctxwffjdg {\Gamma , n : Nat} {X(n)} \and
-           \nonlinterm {\Gamma} {M_0} {X(0)} \and
-           \nonlinterm {\Gamma, n : Nat, x : X(n)} {M_{suc}(n , x)} {X(suc~n)}}
-          {\nonlinterm {\Gamma} {elim_{Nat}(X,M_0,M_{suc})(n)} {X(n)}}
+    % \inferrule{~}{\nonlinterm {\Gamma} {0} {Nat}}
+
+    % \inferrule{\nonlinterm {\Gamma} {n} {Nat}}{\nonlinterm {\Gamma} {suc~n} {Nat}}
+
+    % \inferrule{\ctxwffjdg {\Gamma , n : Nat} {X(n)} \and
+    %        \nonlinterm {\Gamma} {M_0} {X(0)} \and
+    %        \nonlinterm {\Gamma, n : Nat, x : X(n)} {M_{suc}(n , x)} {X(suc~n)}}
+    %       {\nonlinterm {\Gamma} {elim_{Nat}(X,M_0,M_{suc})(n)} {X(n)}}
 
@@
 \begin{proof}
-  Because $\semgg{\top}\gamma\,w$ is a singleton set, it suffices to show that
+  Because $\semg{\top}\gamma\,w$ is a singleton set, it suffices to show that
   $\semg{\StringGram}\gamma\,w$ is likewise a singleton set for all $\gamma$ and

### 2025-03-23 paper/paper.tex
fix some type theory stuff, non-linear types intro, cite vakar, missing \otimes term syntax fixed

@@
       }
-      {a' : \A \otimes \A , a_{*} : \A^* \vdash g := \letin {a_1 \otimes a_2}
+      {a' : \A \otimes \A , a_{*} : \A^* \vdash g := \letin {(a_1 , a_2)}
         {a'} {\cons~a_1 (\cons~a_2~a_{*}) : \A^*}}
@@
 
-\steven{TODO define general inductives as the least fixed point of strictly
-  positive endofunctors on $U$}
-
-\steven{TODO linear $\mu$ in the type formation rules? nonlinear $\mu$ in the
-  type formation rules?}
+\subsection{Non-linear Typing}
 
@@
   \footnotesize
-    \boxed{\ctxwffjdg \Gamma X}
+    \boxed{\inferrule{\ctxwff\Gamma}{\ctxwffjdg \Gamma X}}
 
@@
 
-    % \inferrule{~}{\ctxwffjdg \Gamma {Nat}}
+    \inferrule{~}{\ctxwffjdg \Gamma {Nat}}
 
@@
 
-    \boxed{\ctxwffjdg \Gamma {X \equiv Y}}
+    \boxed{\inferrule{\ctxwffjdg\Gamma X \and \ctxwffjdg\Gamma Y}{\ctxwffjdg \Gamma {X \equiv Y}}}
 
@@
 
-    \boxed{\Gamma \vdash X \isSmall}
-
-    \inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash X \isSmall}
+    \boxed{\inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash X \isSmall}}
 
     \\
-
-    \boxed{\linctxwffjdg \Gamma A}
-
-    \inferrule{~}{\linctxwffjdg \Gamma I}
-
-    \inferrule{c \in \Sigma}{\linctxwffjdg \Gamma {\literal{c}}}
-
-    \inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg
-      \Gamma {A \otimes B}}
-
-    \inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg
-      \Gamma {A \lto B}}
-
-    \inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg
-      \Gamma {A \tol B}}
-
-    \inferrule{\linctxwffjdg {\Gamma, x : X} A}{\linctxwffjdg
-      \Gamma {\LinSigTyLimit{x}{X}{A}}}
-
-    \inferrule{\linctxwffjdg {\Gamma, x : X} A}{\linctxwffjdg
-      \Gamma {\LinPiTyLimit{x}{X}{A}}}
-
-    \inferrule{\nonlinterm \Gamma f {\ltonl{(A \lto B)}} \and
-      \nonlinterm \Gamma g {\ltonl{(A \lto B)}}}{\linctxwffjdg \Gamma {\equalizer {a}{f}{g}}}
-
-    \inferrule{\nonlinterm \Gamma M L}{\linctxwffjdg \Gamma {\unquoteTy M}}
-
-    \\
-
-    \boxed{\linctxwffjdg \Gamma {A \equiv B}}
-
-    \inferrule{\nonlinterm \Gamma {A \equiv B} L}{\linctxwffjdg
-      \Gamma {A \equiv B}}
-
-    \inferrule{\linctxwffjdg \Gamma A}{\linctxwffjdg \Gamma {\unquoteTy
-        {\quoteTy A} \equiv A}}
-  \end{mathpar}
-  \caption{Type formation rules}
-  \label{fig:formation}
-\end{figure}
-
-\begin{figure}
-  \begin{mathpar}
-  \footnotesize
-    \boxed{\nonlinterm \Gamma M X}
+    \boxed{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma M X}
 
@@
       {\nonlinterm \Gamma {refl} {M =_{X} N}}
+
+    \\
+    \boxed{\inferrule{\nonlinterm\Gamma M X \and \nonlinterm \Gamma N X}{\nonlinterm \Gamma {M \equiv N} X}}
+
+    \inferrule{\nonlinterm \Gamma P {M =_{X} N}}{\nonlinterm\Gamma{M \equiv N} X}
   \end{mathpar}
-  \caption{Non-linear typing (selection)}
+  \caption{Non-linear Formation and Typing Rules (selection)}
   \label{fig:non-linear-types}
@@
 
-\subsection{Non-linear Typing}
-The non-linear types of \theoryabbv include the standard dependent
-types for $\Pi$,$\Sigma$, extensional equality, unit and empty types, and
-inductive types that are defined strictly positively. We may also assume function
-extensionality\cite{Hofmann_1997}; Function
-extensionality is holds in Cubical Agda, and in our implementation we make heavy use of it of when
-validating that the denotational semantics respects the equational theory of
-\theoryabbv. However, it is not necessary to assume function extensionality for
-the non-linear types in \theoryabbv, as we don't utilize it in any of our constructions.
-
-We present a selection of
-the non-linear type constructions in
+We present a selection of the non-linear type constructions in
 \Cref{fig:non-linear-types} and provide the rest in
-\cref{fig:full-non-linear-types, fig:jdg-eq-nonlinear} of the appendix.
-First, we include universe types $U$ of
-small non-linear types and $L$ of linear types. These are defined as
+\cref{fig:full-non-linear-types, fig:jdg-eq-nonlinear} of the appendix.  First,
+we include universe types $U$ of small non-linear types and $L$ of linear
+types. These are defined as
 universes ``ala Coquand'' in that we define a judgment saying when a
@@
 types are needed so that we can define types by recursion on natural
-numbers. As in \cite{krishnaswami_integrating_2015}, the extensional equality type
-$M =_{X} N$ used here has the introduction form $refl$, but no
-elimination due to the presence of equality reflection. Next, we include a non-linear type $\ltonl A$ where $A$ is a
-linear type. The intuition for this type is that its elements are the
-linear terms that are ``resource free'': its introduction rule says we
-can construct an $\ltonl A$ when we have a linear term of type $A$
-with no free linear variables. This type is used extensively in our
-examples, playing a similar role to the $!$ modality of ordinary
-linear logic or the persistence modality $\square$ of separation logic
-\cite{girard_linear_1987,jung_higher-order_2016}.
+numbers. 
+Next, we include standard $\Sigma,\Pi$, empty, unit and natural number types.
+
+We use an \emph{extensional} equality type $M =_{X} N$ with introduction form
+$refl$, but no elimination form. Instead we have the equality reflection rule
+which allows us to conclude a definitional equality $M =_{X} N$ from an
+arbitrary typal equality proof $P$.  The usage of an extensional equality type
+matches our implementation, which interprets both judgmental and typal equality
+as Cubical Agda's \texttt{Path} type, and so naturally supports the equality
+reflection rule. Extensional equality makes type checking of our syntax as such
+undecidable because the conversion rule may require an arbitrarily complex
+equality proof with no explicit proof term, but in our Agda implementation we
+must provide all of these equalities manually, and so the extensionality does
+not raise any issues \cite{Hofmann_1997}.
+
+Lastly, we include a non-linear type $\ltonl A$ where $A$ is a linear type. The
+intuition for this type is that its elements are the linear terms that are
+``resource free'': its introduction rule says we can construct an $\ltonl A$
+when we have a linear term of type $A$ with no free linear
+variables. Semantically, this is the type of parses of the empty string. This
+type is used extensively in our examples, playing a similar role to the $!$
+modality of ordinary linear logic or the persistence modality $\square$ of
+separation logic \cite{girard_linear_1987,jung_higher-order_2016}.
 
 \subsection{Linear Typing}
+
+\begin{figure}
+  \begin{mathpar}
+    \footnotesize
+    \boxed{\inferrule{\ctxwff\Gamma}{\linctxwffjdg \Gamma A}}
+
+    \inferrule{~}{\linctxwffjdg \Gamma I}
+
+    \inferrule{c \in \Sigma}{\linctxwffjdg \Gamma {\literal{c}}}
+
+    \inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg
+      \Gamma {A \otimes B}}
+
+    \inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg
+      \Gamma {A \lto B}}
+
+    \inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg
+      \Gamma {A \tol B}}
+
+    \inferrule{\linctxwffjdg {\Gamma, x : X} A}{\linctxwffjdg
+      \Gamma {\LinSigTyLimit{x}{X}{A}}}
+
+    \inferrule{\linctxwffjdg {\Gamma, x : X} A}{\linctxwffjdg
+      \Gamma {\LinPiTyLimit{x}{X}{A}}}
+
+    \inferrule{\nonlinterm \Gamma f {\ltonl{(A \lto B)}} \and
+      \nonlinterm \Gamma g {\ltonl{(A \lto B)}}}{\linctxwffjdg \Gamma {\equalizer {a}{f}{g}}}
+
+    \inferrule{\nonlinterm \Gamma M L}{\linctxwffjdg \Gamma {\unquoteTy M}}
+
+    \\
+    \boxed{\inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg \Gamma {A \equiv B}}}
+
+    \inferrule{\nonlinterm \Gamma {A \equiv B} L}{\linctxwffjdg
+      \Gamma {A \equiv B}}
+
+    \inferrule{\linctxwffjdg \Gamma A}{\linctxwffjdg \Gamma {\unquoteTy
+        {\quoteTy A} \equiv A}}
+  \end{mathpar}
+  \caption{Linear Type Formers, Type Equivalence}
+  \label{fig:linear-formation}
+\end{figure}
+
 We give the rules for linear type
-formation in \cref{fig:formation} and the definition of linear terms in
+formation in \cref{fig:linear-formation} and the definition of linear terms in
 \cref{fig:linear-terms}. The equational theory for these
@@
 
-\steven{I'm not sure if this is the proper way to cite the corollary, especially
-because the citation is a reference, not the originator of the idea}
-\begin{corollary}[\cite{Cockett_1993}]
+The following corollary is a well known consequence of distributivity \cite{Cockett_1993}.
+\begin{corollary}
   \label{cor:binary-mono}
-  Distributivity implies that the coprojections $\inl : A \to A \oplus B$,
-  $\inr : B \to A \oplus B$ of a binary sum are monomorphisms --- i.e. if
+  Distributivity implies that the constructors $\inl : A \to A \oplus B$,
+  $\inr : B \to A \oplus B$ of a binary sum are injective --- i.e. if
   $\inl\,a \equiv \inl\,a'$, then $a \equiv a'$.
@@
 distinct sorts, so the linear logic $!$ operator is not a primitive, but is
-instead definable. Vakar develops a dependent linear type theory where instead
-the non-linear types are accessed using the $!$ modality, similar to Girard's
-original approach \cite{vakar,girard_linear_1987}. Additionally, Vakar develops
-a general categorical semantics for their system, whereas we have only developed
-a single denotational model. It appears straightforward to adapt Vakar's general
+instead definable, for instance in \theoryabbv as $!A = \bigoplus_{\_:\ltonl A}
+I$. Vakar develops a dependent linear type theory where instead the non-linear
+types are accessed using the $!$ modality, similar to Girard's original approach
+\cite{vakar,girard_linear_1987}. Additionally, Vakar develops a general
+categorical semantics for their system, whereas we have only developed a single
+denotational model. It appears straightforward to adapt Vakar's general
 semantics approach to a non-commutative variant that would apply to
@@
 
+\max{TODO: include the rest of the rules}
 \begin{figure}
@@
     \inferrule{\linterm \Gamma \cdot e A}{\nonlinterm \Gamma e \ltonl{A}}
+
   \end{mathpar}

### 2025-03-23 paper/paper.tex
fixed accents in 'V\'ak'ar'

@@
 instead definable, for instance in \theoryabbv as $!A = \bigoplus_{\_:\ltonl A}
-I$. Vakar develops a dependent linear type theory where instead the non-linear
+I$. V\'ak\'ar develops a dependent linear type theory where instead the non-linear
 types are accessed using the $!$ modality, similar to Girard's original approach
-\cite{vakar,girard_linear_1987}. Additionally, Vakar develops a general
+\cite{vakar,girard_linear_1987}. Additionally, V\'ak\'ar develops a general
 categorical semantics for their system, whereas we have only developed a single
-denotational model. It appears straightforward to adapt Vakar's general
+denotational model. It appears straightforward to adapt V\'ak\'ar's general
 semantics approach to a non-commutative variant that would apply to
@@
 }
+
+\pedr{I think that it's fine not to include these details}
 

### 2025-03-24 paper/paper.tex
fixed typo in my comment
added bool

@@
 The formation rules for the judgments of \theoryabbv are shown in
-\Cref{fig:formation}. $\Gamma$ stands for non-linear contexts; $X,Y,Z$
+\Cref{fig:contexts,fig:non-linear-types,fig:linear-formation}. $\Gamma$ stands for non-linear contexts; $X,Y,Z$
 stand for non-linear types; $M,N$ stand for non-linear terms, these
@@
   \caption{Context well-formedness rules}
+  \label{fig:contexts}
 \end{figure}
@@
 
+    \inferrule{~}{\ctxwffjdg \Gamma {Bool}}
+
     \inferrule{~}{\ctxwffjdg \Gamma {Nat}}
@@
     \\
-    \boxed{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma M X}
+    \boxed{\inferrule{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma M X}}
 
@@
 numbers. 
-Next, we include standard $\Sigma,\Pi$, empty, unit and natural number types.
+Next, we include standard $\Sigma,\Pi$, empty, unit, Boolean, and natural number types.
 
@@
 
-The following corollary is a well known consequence of distributivity \cite{Cockett_1993}.
+The following corollary is a well known consequence of distributivity \cite{Cockett_1993}, which we use in \cref{lem:unambig-to-disjoint} to prove that unambiguous binary sums have unambiguous summands.
 \begin{corollary}
@@
 
-\steven{
-  Is any of this detail useful here, or nah?
-
-  They
-  use their system in the design of a linear-time algorithm for greedy regular
-  expression matching. It is yet unclear if greediness can be encoded in
-  \theoryname; however, it appears that our approach in determinizing an NFA is
-  quite similar to their algorithm. The biggest difference being that our
-  approach uses a global disambiguation strategy while theirs is a local, greedy strategy.
-}
-
-\pedr{I think that it's fine not to include these details}
-
 We also aim to adapt the approach used for our context-free grammar parserss for
@@
     {\nonlinterm \Gamma M \bot \and \ctxwffjdg \Gamma X}
-    {\nonlinterm \Gamma {()} {\bot}}
-
-    % \inferrule{~}{\nonlinterm {\Gamma} {0} {Nat}}
-
-    % \inferrule{\nonlinterm {\Gamma} {n} {Nat}}{\nonlinterm {\Gamma} {suc~n} {Nat}}
-
-    % \inferrule{\ctxwffjdg {\Gamma , n : Nat} {X(n)} \and
-    %        \nonlinterm {\Gamma} {M_0} {X(0)} \and
-    %        \nonlinterm {\Gamma, n : Nat, x : X(n)} {M_{suc}(n , x)} {X(suc~n)}}
-    %       {\nonlinterm {\Gamma} {elim_{Nat}(X,M_0,M_{suc})(n)} {X(n)}}
+    {\nonlinterm \Gamma {elim_{\bot,X}(M)} {X}}
+
+    \inferrule{~}{\nonlinterm \Gamma {true} {Bool}}
+
+    \inferrule{~}{\nonlinterm \Gamma {false} {Bool}}
+
+    \inferrule{\ctxwffjdg {\Gamma , b : Bool} {X(b)}}
+      {\nonlinterm
+        {\Gamma, x_0 : X(false), x_1 : X(true), b : Bool}
+        {if\,b\,then\,x_0\,else\,x_1}
+        {X(b)}
+      }
+
+    \\
+
+    \inferrule{~}{\nonlinterm {\Gamma} {0} {Nat}}
+
+    \inferrule{\nonlinterm {\Gamma} {n} {Nat}}{\nonlinterm {\Gamma} {suc~n} {Nat}}
+
+    \inferrule{\ctxwffjdg {\Gamma , n : Nat} {X(n)} \and
+           \nonlinterm {\Gamma} {M_0} {X(0)} \and
+           \nonlinterm {\Gamma, n : Nat, x : X(n)} {M_{suc}(n , x)} {X(suc~n)}}
+         {\nonlinterm
+           {\Gamma, n : Nat}
+           {elim_{Nat}(X,M_0,M_{suc})(n)}
+           {X(n)}}
 
@@
 
+  \inferrule{\ctxwffjdg {\Gamma , b : Bool} {X(b)}}
+        {\nonlineq
+        {\Gamma, x_0 : X(false), x_1 : X(true)}
+        {if\,false\,then\,x_0\,else\,x_1}
+        {x_0}
+        {X(false)}
+      }
+
+  \inferrule{\ctxwffjdg {\Gamma , b : Bool} {X(b)}}
+        {\nonlineq
+        {\Gamma, x_0 : X(false), x_1 : X(true)}
+        {if\,true\,then\,x_0\,else\,x_1}
+        {x_1}
+        {X(true)}
+      }
+
   \inferrule{\nonlinterm {\Gamma} {M_0} {X(0)} \and
@@
   \inferrule{\nonlinterm {\Gamma} {M_0} {X(0)} \and
-      \nonlinterm {\Gamma, n : Nat, x : X(n)} {M_{suc}(n , x)} {X(suc~n)} \and
-      \nonlinterm {\Gamma} {n} {Nat}}
-      {\nonlineq {\Gamma} {elim_{Nat}(X,M_0,M_{suc})(suc~n)} {M_{suc}(n ,
+      \nonlinterm {\Gamma, n : Nat, x : X(n)} {M_{suc}(n , x)} {X(suc~n)}
+    }
+      {\nonlineq {\Gamma, n : Nat} {elim_{Nat}(X,M_0,M_{suc})(suc~n)} {M_{suc}(n ,
           elim_{Nat}(X,M_0,M_{suc})(n))} {X(suc~n)}}

### 2025-03-24 paper/paper.tex
lists + touch up end of appendix

@@
 \newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
-\newcommand{\StringSem}{\textbf{String}}
+\newcommand{\StringSem}{\mathbf{String}}
 \newcommand{\CharGram}{\texttt{Char}}
@@
 \newcommand{\internalize}[1]{\lceil#1\rceil}
+\newcommand{\listTy}[1]{List\left( #1 \right)}
 
@@
 
+    \inferrule{\ctxwffjdg \Gamma X}{\ctxwffjdg \Gamma {\listTy{X}}}
+
     \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg {\Gamma , x : X} {Y}}{\ctxwffjdg \Gamma {\PiTyLimit {x}{X}{Y}}}
@@
 \max{TODO: include the rest of the rules}
+\steven{I added rules for Bool, Nat, and List. What else do we need?}
+
 \begin{figure}
@@
 
+    \\
+
+    \inferrule{~}{\nonlinterm \Gamma {nil} {\listTy{X}}}
+
+    \inferrule
+    { \nonlinterm \Gamma {x} {X}
+      \and
+      \nonlinterm \Gamma {M} {\listTy{X}}}
+    {\nonlinterm \Gamma {cons(x,M)} {\listTy{X}}}
+
+    \inferrule
+    {
+      \ctxwffjdg {\Gamma, xs : \listTy{X}} {Y(xs)} \and
+      \nonlinterm {\Gamma} {M} {Y (nil)} \and
+      \nonlinterm {\Gamma} {N}
+      {\PiTyLimit {x} {X}
+        {\PiTyLimit {xs} {\listTy{X}}
+          {\PiTyLimit {ys} {Y(xs)} {Y(cons(x,xs))}}}}
+    }
+    {
+      \nonlinterm
+      {\Gamma, xs : \listTy{X}}
+      {fold(M,N)(xs)}
+      {Y(xs)}
+    }
+
+    \\
+
     \inferrule{\nonlinterm \Gamma M X \and \nonlinterm \Gamma {N} {\subst Y {M}
@@
           elim_{Nat}(X,M_0,M_{suc})(n))} {X(suc~n)}}
+
+  \inferrule
+    {
+      \nonlinterm {\Gamma} {M} {Y (nil)} \and
+      \nonlinterm {\Gamma} {N}
+      {\PiTyLimit {x} {X}
+        {\PiTyLimit {xs} {\listTy{X}}
+          {\PiTyLimit {ys} {Y(xs)} {Y(cons(x,xs))}}}}
+    }
+    {
+      \nonlineq
+      {\Gamma}
+      {fold(M,N)(nil)}
+      {y_{nil}}
+      {Y(nil)}
+    }
+
+  \inferrule
+    {
+      \nonlinterm {\Gamma} {M} {Y (nil)} \and
+      \nonlinterm {\Gamma} {N}
+      {\PiTyLimit {x} {X}
+        {\PiTyLimit {xs} {\listTy{X}}
+          {\PiTyLimit {ys} {Y(xs)} {Y(cons(x,xs))}}}}
+    }
+    {
+      \nonlineq
+      {\Gamma, x : X, xs : \listTy{X}}
+      {fold(M,N)(cons(x , xs))}
+      {N(x)(xs)(fold(M,N)(xs))}
+      {Y(cons(x , xs))}
+    }
 
@@
 
-\paragraph{\cref{ax:dist}}
-\begin{theorem}
+\paragraph{\cref{ax:dist}} Distributivity
+\begin{theorem}[\Agda]
   In $\Grammar$, $\semg{\lamblto{e}{\letin{\sigma\,f\,e'}{e}{\withlamb{x}{e'\,.\pi\,x}}}}$ has
@@
 \begin{proof}
-  \steven{Double check this}.
+  Distributivity is true in the denotational semantics, as the category $\Grammar$ is a topos, which are well-known to be distributive.
   The following map forms the desired inverse,
@@
   \[
-    \lamb {\gamma} {\lamb {w} {\lamb {p} {(\lamb {\lamb {x} {(p~x).fst}} , \lamb {x} {(p~x).snd})}}}
+    \lamb {\gamma} {\lamb {w} {\lamb {p} {(\lamb {\left( \lamb {x} {(p~x).fst} \right)} , \lamb {x} {(p~x).snd})}}}
   \]
@@
 
-\paragraph{\cref{ax:disjointness}}
-\begin{theorem}
+\paragraph{\cref{ax:disjointness}} $\sigma$-Disjointness
+\begin{theorem}[\Agda]
   In $\Grammar$, $\semg{\sigma x}$ and $\semg{\sigma x'}$ are disjoint for
@@
 
-\paragraph{\cref{ax:string-top}}
-\begin{theorem}
+\paragraph{\cref{ax:string-top}} String is strongly equivalent to $\top$.
+\begin{theorem}[\Agda]
   In $\Grammar$, $\semg{!}$ has an inverse.
@@
 \begin{proof}
-  Because $\semg{\top}\gamma\,w$ is a singleton set, it suffices to show that
-  $\semg{\StringGram}\gamma\,w$ is likewise a singleton set for all $\gamma$ and
-  $w$.
-
-  First, we prove by induction on $w$ that $\semg{\StringGram}$ is a retract of
-  $\semg{\LinSigTyLimit{w}{\StringSem}{\internalize{w}}}$. Then, again by
+  Because $\semg{\top}w$ is a singleton set for all $\gamma$ and $w$, it suffices to show that
+  $\semg{\StringGram}w$ is likewise a singleton.
+
+  First, we prove by induction on $w$ that $\semg{\StringGram}w$ is a retract of
+  $\semg{\LinSigTy{w}{\StringSem}{\internalize{w}}}w$, a sum over a nonlinear
+  type of strings. Then, again by
   induction on $w$, we show that
-  $\semg{\LinSigTyLimit{w}{\StringSem}{\internalize{w}}}$ is isomorphic to
-  $\semg{\top}$. Thus, $\semg{\StringGram}$ is a retract of $\semg{\top}$. In particular,
-  so $\semg{\StringGram}\gamma\,w$ is a singleton, and thus
-  $\semg{\StringGram}\gamma\,w \cong \semg{\top}\gamma\,w$ in $\Set$.
+  $\semg{\LinSigTy{w}{\StringSem}{\internalize{w}}}w$ is isomorphic to
+  $\semg{\top}w$.
+
+  Each $\semg{\internalize{w}}w'$ is inhabited if and only if $w$ is equal to
+  $w'$, and the parses of $w$ for $\semg{\internalize{w}}$ are unique. That is,
+  $\semg{\LinSigTy{w}{\StringSem}{\internalize{w}}}w$ is a singleton set.
+  So, $\semg{\StringGram}w$ is a retract of a singleton set, and is itself
+  singleton. Thus, $\semg{\StringGram}w \cong \semg{\top}w$ in $\Set$.
 \end{proof}

### 2025-03-24 paper/paper.tex
more on nonlinear figures

@@
 
-    \inferrule{\ctxwffjdg \Gamma X}{\ctxwffjdg \Gamma {\listTy{X}}}
-
     \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg {\Gamma , x : X} {Y}}{\ctxwffjdg \Gamma {\PiTyLimit {x}{X}{Y}}}
@@
   \begin{mathpar}
-    \boxed{\linterm \Gamma \Delta {e} {A}}
+    \boxed{
+      \inferrule
+      {\linctxwff \Gamma \Delta \and \linctxwffjdg \Gamma \A}
+      {\linterm \Gamma \Delta {e} {A}}}
 
@@
   \footnotesize
-    \boxed{\nonlinterm \Gamma M X}
+    \boxed{\inferrule{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma M X}}
 
@@
     {\nonlinterm \Gamma M \bot \and \ctxwffjdg \Gamma X}
-    {\nonlinterm \Gamma {elim_{\bot,X}(M)} {X}}
+    {\nonlinterm \Gamma {elim_{\bot}(X, M)} {X}}
 
@@
 
-    \inferrule{\ctxwffjdg {\Gamma , b : Bool} {X(b)}}
+    \inferrule
+    {
+      \ctxwffjdg {\Gamma , b : Bool} {X(b)} \and
+      \nonlinterm \Gamma {M_{0}} {X(false)} \and
+      \nonlinterm \Gamma {M_{1}} {X(true)}
+    }
     {\nonlinterm
-        {\Gamma, x_0 : X(false), x_1 : X(true), b : Bool}
-        {if\,b\,then\,x_0\,else\,x_1}
+        {\Gamma, b : Bool}
+        {elim_{Bool}(X, M_0, M_1)(b)}
         {X(b)}
@@
 
-    \inferrule{~}{\nonlinterm \Gamma {nil} {\listTy{X}}}
-
-    \inferrule
-    { \nonlinterm \Gamma {x} {X}
-      \and
-      \nonlinterm \Gamma {M} {\listTy{X}}}
-    {\nonlinterm \Gamma {cons(x,M)} {\listTy{X}}}
-
-    \inferrule
-    {
-      \ctxwffjdg {\Gamma, xs : \listTy{X}} {Y(xs)} \and
-      \nonlinterm {\Gamma} {M} {Y (nil)} \and
-      \nonlinterm {\Gamma} {N}
-      {\PiTyLimit {x} {X}
-        {\PiTyLimit {xs} {\listTy{X}}
-          {\PiTyLimit {ys} {Y(xs)} {Y(cons(x,xs))}}}}
-    }
-    {
-      \nonlinterm
-      {\Gamma, xs : \listTy{X}}
-      {fold(M,N)(xs)}
-      {Y(xs)}
-    }
-
-    \\
-
     \inferrule{\nonlinterm \Gamma M X \and \nonlinterm \Gamma {N} {\subst Y {M}
@@
   \begin{mathpar}
-  \boxed{\nonlinterm \Gamma {M \equiv N} {X}}
+    \boxed{
+      \inferrule
+      {\nonlinterm \Gamma M X \and
+       \nonlinterm \Gamma N X}
+      {\nonlinterm \Gamma {M \equiv N} {X}}}
 
@@
 
-  \inferrule{\nonlinterm \Gamma M \bot \and \nonlinterm \Gamma N \bot}{\nonlineq \Gamma M N \bot}
+  % \inferrule{\nonlinterm \Gamma M \bot \and \nonlinterm \Gamma N \bot}{\nonlineq \Gamma M N \bot}
 
@@
 
-  \inferrule{\ctxwffjdg {\Gamma , b : Bool} {X(b)}}
+  \inferrule
+  {
+    \nonlinterm \Gamma {M_{0}} {X(false)} \and
+    \nonlinterm \Gamma {M_{1}} {X(true)}
+  }
   {\nonlineq
-        {\Gamma, x_0 : X(false), x_1 : X(true)}
-        {if\,false\,then\,x_0\,else\,x_1}
-        {x_0}
-        {X(false)}
+      {\Gamma}
+      {elim_{Bool}(X, M_0,M_1)(false)}
+      {M_0}
+      {X(b)}
   }
 
-  \inferrule{\ctxwffjdg {\Gamma , b : Bool} {X(b)}}
+  \inferrule
+  {
+    \nonlinterm \Gamma {M_{0}} {X(false)} \and
+    \nonlinterm \Gamma {M_{1}} {X(true)}
+  }
   {\nonlineq
-        {\Gamma, x_0 : X(false), x_1 : X(true)}
-        {if\,true\,then\,x_0\,else\,x_1}
-        {x_1}
-        {X(true)}
+      {\Gamma}
+      {elim_{Bool}(X, M_0,M_1)(true)}
+      {M_1}
+      {X(b)}
   }
@@
           elim_{Nat}(X,M_0,M_{suc})(n))} {X(suc~n)}}
-
-  \inferrule
-    {
-      \nonlinterm {\Gamma} {M} {Y (nil)} \and
-      \nonlinterm {\Gamma} {N}
-      {\PiTyLimit {x} {X}
-        {\PiTyLimit {xs} {\listTy{X}}
-          {\PiTyLimit {ys} {Y(xs)} {Y(cons(x,xs))}}}}
-    }
-    {
-      \nonlineq
-      {\Gamma}
-      {fold(M,N)(nil)}
-      {y_{nil}}
-      {Y(nil)}
-    }
-
-  \inferrule
-    {
-      \nonlinterm {\Gamma} {M} {Y (nil)} \and
-      \nonlinterm {\Gamma} {N}
-      {\PiTyLimit {x} {X}
-        {\PiTyLimit {xs} {\listTy{X}}
-          {\PiTyLimit {ys} {Y(xs)} {Y(cons(x,xs))}}}}
-    }
-    {
-      \nonlineq
-      {\Gamma, x : X, xs : \listTy{X}}
-      {fold(M,N)(cons(x , xs))}
-      {N(x)(xs)(fold(M,N)(xs))}
-      {Y(cons(x , xs))}
-    }
 
@@
   \begin{mathpar}
-    \boxed{\Gamma ; \Delta \vdash e \equiv e' : A}
+    \boxed{
+      \inferrule
+      {\linterm \Gamma \Delta e A \and
+       \linterm \Gamma \Delta e' A}
+      {\Gamma ; \Delta \vdash e \equiv e' : A}}
 

### 2025-03-24 paper/paper.tex
agda links

@@
 \usepackage{thmtools} % needed for sane cleveref naming of theorems
+\usepackage{hyperref}
+\usepackage[capitalise]{cleveref}
 
@@
 
- %% \steven{TODO make the Agda logo clickable} %
-\newcommand{\Agda}{\agdalogo}
 \renewcommand{\Gamma}{\mathgtt{G}}
@@
 
-\usepackage{hyperref}
-\usepackage[capitalise]{cleveref}
 \usepackage{stmaryrd}
@@
 
+\newcommand{\fillerlink}{http://localhost:1313}
+\newcommand{\Agda}[1]{\href{#1}{\agdalogo}}
+
 \newif\ifdraft
@@
 context-free grammars and parsers based on $\LL{1}$ and $\LR{1}$
-automata. Throughout this paper, we will use \agdalogo~ to mark
-results that are mechanized in our Agda development.
+automata. Throughout this paper, \agdalogo~ will mark
+results that are mechanized in our Agda development and provide a link to their
+implementation.
 
@@
 \end{mathpar}
-\caption{\stringquote{ab} is parsed by $(\literal a \oplus \literal b) \otimes \literal c$
+\caption{(\Agda{fillerlink}) \stringquote{ab} is parsed by $(\literal a \oplus \literal b) \otimes \literal c$
 }
@@
 \end{lstlisting}
-\caption{\stringquote{ab} is parsed by
+\caption{(\Agda{\fillerlink}) \stringquote{ab} is parsed by
   $({\literal a}^{*} \otimes \literal b) \oplus \literal c$}
@@
 \end{lstlisting}
-\caption{A parse transformer for abstract grammars}
+\caption{(\Agda{\fillerlink}) A parse transformer for abstract grammars}
 \label{fig:kleeneabstractproof}
@@
   \end{minipage}%
-  \caption{NFA for $(\a^{*} \otimes \b) \oplus \c$ and its corresponding type}
+  \caption{(\Agda{\fillerlink}) NFA for $(\a^{*} \otimes \b) \oplus \c$ and its corresponding type}
   \label{fig:exampleNFA}
@@
 establish two useful properties of unambiguity.
-\begin{lemma}[\agdalogo]
+\begin{lemma}[\Agda{\fillerlink}]
   \label{lem:retract}
@@
 \end{lemma}
-\begin{lemma}[\agdalogo]
+\begin{lemma}[\Agda{\fillerlink}]
   \label{lem:unambig-sum}
@@
 
-\begin{lemma}[\agdalogo]
+\begin{lemma}[\Agda{\fillerlink}]
   \label{lem:unambig-to-disjoint}
@@
 following:
-\begin{lemma}[\agdalogo]
+\begin{lemma}[\Agda{\fillerlink}]
   \label{lem:wk-eqv-parse}
@@
 
-\begin{theorem}[\agdalogo]
+\begin{theorem}[\Agda{\fillerlink}]
   \label{thm:dfa-parser}
@@
 DFA is unambiguous even if the NFA is not.
-\begin{construction}[Determinization, \Agda]
+\begin{construction}[Determinization, \Agda{\fillerlink}]
   \label{cons:determinization}
@@
 \newcommand{\R}{\texttt{R}}
-\begin{construction}[Thompson's Construction, \Agda]
+\begin{construction}[Thompson's Construction, \Agda{\fillerlink}]
   \label{cons:thompson}
@@
 
-\begin{corollary}
+\begin{corollary}[\Agda{\fillerlink}]
   We may build a parser for every regular expression $\R$.
@@
 
-\begin{theorem}[\Agda]
+\begin{theorem}[\Agda{\fillerlink}]
   \label{thm:dyck}
@@
 
-\begin{theorem}[\agdalogo]
+\begin{theorem}[\Agda{\fillerlink}]
   \label{thm:exp-parser}
@@
 languages in the Chomsky hierarchy.
-\begin{construction}[\Agda]
+\begin{construction}[\Agda{\fillerlink}]
   \label{cons:turing}
@@
 term constructors of \theoryabbv as combinators on formal grammars or
-parse transformers, and use cubical Agda's equality type to model the
+parse transformers, and use Cubical Agda's equality type to model the
 term equalities. Cubical Agda is convenient for this purpose as it has
@@
 \paragraph{\cref{ax:dist}} Distributivity
-\begin{theorem}[\Agda]
+\begin{theorem}[\Agda{\fillerlink}]
   In $\Grammar$, $\semg{\lamblto{e}{\letin{\sigma\,f\,e'}{e}{\withlamb{x}{e'\,.\pi\,x}}}}$ has
@@
 \paragraph{\cref{ax:disjointness}} $\sigma$-Disjointness
-\begin{theorem}[\Agda]
+\begin{theorem}[\Agda{\fillerlink}]
   In $\Grammar$, $\semg{\sigma x}$ and $\semg{\sigma x'}$ are disjoint for
@@
 \paragraph{\cref{ax:string-top}} String is strongly equivalent to $\top$.
-\begin{theorem}[\Agda]
+\begin{theorem}[\Agda{\fillerlink}]
   In $\Grammar$, $\semg{!}$ has an inverse.

### 2025-03-24 paper/paper.tex
spelling

@@
 traces of $D(A)$ and combine this using \Cref{lem:wk-eqv-parse} to get
-a verified regex parser.
+a verified regular expression parser.
 
@@
 accepting state with the constructor $\nil$. Here we use an Agda-style
-unicode syntax for $\bigamp$, as well as using the function arrow to
+Unicode syntax for $\bigamp$, as well as using the function arrow to
 mean a non-dependent version of $\bigamp$. If we had a trace beginning
@@
 
-%% The first two equalities are by defintion. The third equality holds because
+%% The first two equalities are by definition. The third equality holds because
 %% $\literal c$ is unambiguous and
@@
 We combine the strong equivalence of \cref{cons:thompson} with the weak
-equivalance of \cref{cons:determinization} to show that $\R$ is weakly
+equivalence of \cref{cons:determinization} to show that $\R$ is weakly
 equivalent to the traces of its determinized automaton. Then, we use
@@
 four different ``states'', each with access to a natural number
-``stack''. The ``opening'' state $O$ expects either an left paren, in
+``stack''. The ``opening'' state $O$ expects either a left paren, in
 which case it increments the stack and stays in the opening state, or
@@
 
-We also aim to adapt the approach used for our context-free grammar parserss for
+We also aim to adapt the approach used for our context-free grammar parsers for
 $\LLL$ and $\LRR$/$\LALRR$ with the aim of developing a shared library of intrinsically
@@
 \subsection{Grammar Semantics Respects Additional Axioms}
-We verify that the denotational semantics verfies each of the axioms we have
+We verify that the denotational semantics validates each of the axioms we have
 assumed.

### 2025-03-24 paper/paper.tex
save spacing
cleanup and formatting

@@
 \newcommand{\inr}{\mathsf{inr}}
-\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
+\newcommand{\sem}[1]{\left\llbracket{#1}\right\rrbracket}
 \newcommand{\semg}[1]{\sem{#1}\gamma}
@@
 \newcommand{\apptol}[2]{#1 \mathop{{}^{\tol}} #2}
-\newcommand{\PiTy}[3]{\textstyle\prod (#1 : #2). #3}
-\newcommand{\SigTy}[3]{\textstyle\sum (#1 : #2). #3}
+\newcommand{\PiTy}[3]{\textstyle\prod_{#1 : #2} #3}
+\newcommand{\SigTy}[3]{\textstyle\sum_{#1 : #2} #3}
 \newcommand{\PiTyLimit}[3]{\textstyle\prod\limits_{#1 : #2} #3}
 \newcommand{\SigTyLimit}[3]{\textstyle\sum\limits_{#1 : #2} #3}
-\newcommand{\LinPiTy}[3]{\textstyle\bigamp (#1 : #2). #3}
+\newcommand{\LinPiTy}[3]{\textstyle\bigamp_{#1 : #2} #3}
 \newcommand{\LinPiTyLimit}[3]{\bigwith\limits_{#1 : #2} #3}
-\newcommand{\LinSigTy}[3]{\textstyle\bigoplus (#1 : #2). #3}
+\newcommand{\LinSigTy}[3]{\textstyle\bigoplus_{#1 : #2} #3}
 \newcommand{\LinSigTyLimit}[3]{\bigoplus\limits_{#1 : #2} #3}
@@
 Our Agda prototype is based on a \emph{denotational semantics} of
-\theoryabbv. The core idea of the denotational semantics is
-\cite{elliottSymbolicAutomaticDifferentiation2021}, Elliott describes
-a formal grammar as type-level predicates on strings that prove
-language membership. That is, a \emph{formal grammar} $A$ is a
+\theoryabbv. The core idea of the denotational semantics stems from an observation of Elliott:
+% The core idea of the denotational semantics is
+% \cite{elliottSymbolicAutomaticDifferentiation2021}, Elliott describes
+a formal grammar is a type-level predicate on strings that proves
+language membership \cite{elliottSymbolicAutomaticDifferentiation2021}. That is, a \emph{formal grammar} $A$ is a
 function $\StringSem \to \Set$ such that for a string $w$, $A~w$ is the
@@
     extends prior work on dependent linear-non-linear type theory to
-    support inductive linear types
-  \item Demonstration of how to encode many common grammar formalisms
+    support inductive linear types.
+  \item A demonstration of how to encode many common grammar formalisms
     (regular expressions, (non-)deterministic automata, context-free
@@
   \item A denotational semantics for \theoryabbv that shows that the
-    parsers are in fact verified to be correct and soundness of the
+    parsers are in fact verified to be correct, and the soundness of the
     equational theory.
@@
 example programs from \theoryabbv to build intuition.  From there, in
-\cref{sec:tt} we provide the syntax, typing and equational theory of
+\cref{sec:tt} we provide the syntax, typing, and equational theory of
 \theoryname. In \cref{sec:applications} we demonstrate the
@@
 \]
-
 %
@@
 $\B$ as output. The arrow, $\uparrow$, wrapping these
-constructors means that the constructors by themselves are not consumed
+constructors means that the constructors are not consumed
 upon usage, and so are \emph{non-linear} values themselves. That is, the
@@
 appropriate to think of it as a parse \textit{transformer} --- a function from
-parses of $(\A \otimes \A)^{*}$ to parse of $\A^{*}$.
+parses of $(\A \otimes \A)^{*}$ to parses of $\A^{*}$.
 
@@
 \begin{figure}
+\footnotesize
 \begin{mathpar}
@@
         {~}
-        {a_1 : \A, a_2 : \A , a_{*} : \A^* \vdash \cons~a_1 (\cons~a_2~a_{*}) : \A^*}
+        {a_1 : \A, a_2 : \A , as : \A^* \vdash \cons~a_1 (\cons~a_2~as) : \A^*}
       }
-      {a' : \A \otimes \A , a_{*} : \A^* \vdash g := \letin {(a_1 , a_2)}
-        {a'} {\cons~a_1 (\cons~a_2~a_{*}) : \A^*}}
+      {aa : \A \otimes \A , as : \A^* \vdash \letin {(a_1 , a_2)}
+        {aa} {\cons~a_1 (\cons~a_2~as)} : \A^*}
     }
-    {\cdot \vdash f := \lamblto a {\lamblto {a_{*}} {\app {\app g a} {a_{*}} }} : (\A \otimes \A) \lto \A^* \lto \A^*}
+    {\cdot \vdash f := \lamblto {aa} {\lamblto {as} {\letin {(a_1 , a_2)}
+        {aa} {\cons~a_1 (\cons~a_2~as)}}} : (\A \otimes \A) \lto \A^* \lto \A^*}
   }
-  {a : (\A \otimes \A)^* \vdash \texttt{h} := \fold(\nil , f)(a) : \A^* }
+  {aas : (\A \otimes \A)^* \vdash \texttt{h} := \fold(\nil , f)(aas) : \A^* }
 \end{mathpar}
@@
   \begin{minipage}[t]{.4\textwidth}
-  \vspace{0pt}
+  \vspace{8pt}
   \hspace{10pt}
@@
 linear-non-linear calculus (\lnld) and Lambek calculus, also known as
-non-commutative linear logic.
+non-commutative linear logic
 \cite{krishnaswami_integrating_2015,lambek58}. As in \lnld,
@@
 \Cref{fig:non-linear-types} and provide the rest in
-\cref{fig:full-non-linear-types, fig:jdg-eq-nonlinear} of the appendix.  First,
+\cref{fig:full-non-linear-types,fig:jdg-eq-nonlinear} of the appendix.  First,
 we include universe types $U$ of small non-linear types and $L$ of linear
@@
 For any $A : (x : X) \to Y x \to L$, the function
-
 \[
- \lamblto{e}{\letin{\sigma\,f\,e'}{e}{\withlamb{x}{e'\,.\pi\,x}}} : \ltonl {\left(  \LinSigTyLimit{f}{\prod\limits_{x : X} . Y (x)}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
+ \lamblto{e}{\letin{\sigma\,f\,e'}{e}{\withlamb{x}{e'\,.\pi\,x}}} : \ltonl {\left(  \LinSigTyLimit{f}{\PiTy{x}{X}{Y(x)}}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
 \LinPiTyLimit{x}{X}{\LinSigTyLimit{y}{Y(x)}{A\,x\,y}} \right)}
 \]
-
 has an inverse.
@@
 \(
-A \cong \left(A \& I \right) \oplus \LinSigTyLimit{c}{\Sigma_0}{\left( A \& \left( \literal{c} \otimes \top \right) \right)}
+A \cong \left(A \& I \right) \oplus \LinSigTy{c}{\Sigma_0}{\left( A \& \left( \literal{c} \otimes \top \right) \right)}
 \). We use this equivalence when building a parser for the traces of the lookahead
@@
 $\bigoplus$ are disjoint. That is, we want to enforce that $\sigma~x~a$ and
-$\sigma~x'~a'$ of type $\LinSigTyLimit{x}{X}{A\,x}$ are not equal whenever $x \neq x'$.
+$\sigma~x'~a'$ of type $\LinSigTy{x}{X}{A\,x}$ are not equal whenever $x \neq x'$.
 However, because these are linear terms we cannot state their disequality
@@
     {\Gamma \vdash f : \prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)} \and \Gamma; \Delta \vdash e : \el(F\,x)(\mu\,F)}
-    {\Gamma;\Delta \vdash \fold\,F\,f\,x\,(\roll e) \equiv f\,x\,(\map(F\,x)\,(\fold\,F\,f)) : A\,x}
+    {\Gamma;\Delta \vdash \fold\,F\,f\,x\,(\roll\,e) \equiv f\,x\,(\map(F\,x)\,(\fold\,F\,f)) : A\,x}
 
@@
      \and \Gamma \vdash e : \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}\\\\
-      \Gamma,x:X;a:\el(F\,x)(\mu F) \vdash e\,x\,(\roll a) \equiv f\,x\,(\map(F\,x)\,e) : A\,x }
+      \Gamma,x:X;a:\el(F\,x)(\mu F) \vdash e\,x\,(\roll\,a) \equiv f\,x\,(\map(F\,x)\,e) : A\,x }
     {\fold\,F\,f\equiv e' : \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}}
@@
 parse(*@$_D$@*) String.nil s = (*@$\sigma$@*) (D.isAcc s) (Trace(*@$_{\D}$@*).nil s)
-parse(*@$_D$@*) (String.cons ((*@$\sigma$@*) c a) w) s =
-   let (*@$\sigma$@*) b t = parse w (D.(*@$\delta$@*) c s) in
+parse(*@$_D$@*) (String.cons ((*@$\sigma$@*) c a) w) s = let (*@$\sigma$@*) b t = parse w (D.(*@$\delta$@*) c s) in
                                    (*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c s b a t)
@@
 \]
-
 Because the states of $D$ are $\epsilon$-closed subsets of $N.\states$, if two
@@
 equivalences,
-
 \[
@@
 \]
-
 When defining a parser, if the lookahead character is a right paren we would
@@
 formulation that is closer in spirit to the standard definition of a
-formal \emph{language}\cite{elliottSymbolicAutomaticDifferentiation2021}:
+formal \emph{language} \cite{elliottSymbolicAutomaticDifferentiation2021}:
 \begin{definition}
@@
     &\sem{B \tol A}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
-    &\sem{\bigoplus_{x:X} A}\gamma\,w = \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
-    &\sem{\bigwith_{x:X} A}\gamma\,w = \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w
+    &\sem{\bigoplus_{x:X} A}\gamma\,w = \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}
   \end{flalign*}
@@
     \begin{flalign*}
+    &\sem{\bigwith_{x:X} A}\gamma\,w = \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w \\
     &\sem{\uparrow A}\gamma = \sem{A}\gamma\,\varepsilon\\
@@
 \end{figure}
-
 The most complex part of the semantics is the interpretation of
@@
 as the basis for our prototype implementation of \theoryabbv in
-cubical Agda. The implementation is a shallow embedding, meaning that
+Cubical Agda. The implementation is a shallow embedding, meaning that
 rather than formalizing a syntax of \theoryabbv types and terms, we
@@
 %% algebra to reason about parsing as well.
-
 \paragraph{Dependent Linear Types}
-
 Our syntax for non-commutative dependent linear type theory is based on the
@@
 distinct sorts, so the linear logic $!$ operator is not a primitive, but is
-instead definable, for instance in \theoryabbv as $!A = \bigoplus_{\_:\ltonl A}
-I$. V\'ak\'ar develops a dependent linear type theory where instead the non-linear
+instead definable, for instance in \theoryabbv as $!A = \LinSigTy{\_}{\ltonl A}{I}$.
+V\'ak\'ar develops a dependent linear type theory where instead the non-linear
 types are accessed using the $!$ modality, similar to Girard's original approach
@@
 \subsection{Future Work}
-
 In this work we have demonstrated feasibility of \theoryabbv for the
@@
 \paragraph{Verified Parser Generators}
-
 We aim to extend our work on parsing to verify other algorithms. For instance,
@@
 We also aim to adapt the approach used for our context-free grammar parsers for
-$\LLL$ and $\LRR$/$\LALRR$ with the aim of developing a shared library of intrinsically
+$\LLL$ and $\LRR$/$\LALRR$ parsers, with the aim of developing a shared library of intrinsically
 verified parsing utilities that can be usefully shared across different
@@
 type $A$ with semantic outputs in a non-linear type $X$ to be a function $\ltonl
-{(A \lto \bigoplus_{\_:X} \top)}$. That is, a semantic action is a function that
+{(A \lto \LinSigTy{\_}{X}{\top})}$. That is, a semantic action is a function that
 produces a semantic element of $X$ from the concrete parses of $A$. In future
@@
 \]
-
 So for $a: A \vdash a : A$, the denotation of a single variable term
@@
 \]
-
 \begin{proof}
@@
 that is provided by the adjunction between $\semg{- \otimes A}$ and $\semg{A \lto -}$.
-
 \[
@@
 \]
-
 \paragraph{$\lto$-Elimination}
@@
 $\pi_{eq}$ from $Eq(\semg{f} , \semg{g})$ to the domain of $f$ and $g$.
-
 \[
@@
 \]
-
 \paragraph{Equalizer $\beta$}
@@
   The following map forms the desired inverse,
-
   \[

### 2025-03-25 paper/paper.tex
functorial actions side by side
some work on the bib
small tweaks

@@
 
-Writing a parser as a linear term is an intrinsic
-verification of the \emph{soundness} of the parser completely for free
-from the typing: any $\inl$ parse that we return \emph{must}
-correspond to a parse tree of the input string. Further, if we verify
-the disjointness property of \cref{def:parser} we then also get the \emph{completeness} of
-the parser as well, that it never fails to generate an $A$ parse when
-it is possible.
+Writing a parser as a linear term intrinsically verifies the \emph{soundness} of
+the parser for free from the typing: any $\inl$ parse that we return \emph{must}
+correspond to a parse tree of the input string. Further, if we verify the
+disjointness property of \cref{def:parser} we then also get the
+\emph{completeness} of the parser as well, that when the parser rejects the
+input that there are no valid parses.
 
@@
 
-\section{Discussion and Future Work}
+\section{Related and Future Work}
 \label{sec:discussion}
+
+\subsection{Related Work}
 \paragraph{Grammars as (Linear) Types}
@@
 \citet{krishnaswami_integrating_2015}, itself an extension of Benton and
-Wadler's linear-non-linear calculus \cite{bentonMixedLinearNonlinear1995}. A
+Wadler's linear-non-linear calculus \cite{benton1994}. A
 distinct feature of these systems is that linear types and non-linear types are
@@
 types are accessed using the $!$ modality, similar to Girard's original approach
-\cite{vakar,girard_linear_1987}. Additionally, V\'ak\'ar develops a general
+\cite{vakar2015,girard_linear_1987}. Additionally, V\'ak\'ar develops a general
 categorical semantics for their system, whereas we have only developed a single
@@
 \begin{figure}
+\begin{minipage}[t]{.35\textwidth}
   \begin{align*}
-    \el(\Var\,M) B &= B\,M\\
-    \el(\mathsf{K}\,A) B &= A\\
-    \el(\bigoplus A) B &= \bigoplus_{y:Y}\el(A y)B\\
-    \el(\bigamp A) B &= \bigamp_{y:Y}\el(A y)B\\
-    \el(A \otimes A') B &= \el(A)B \otimes \el(A')B\\\\
-    \map(\Var\,M)\,f &= f\,M\\
-    \map(\K\,A)\,f &= \lambda a. a\\
-    \map(\bigoplus A)\,f &= \lambda a. \letin{\oplusinj y {a_y}}{a}{\oplusinj y {\map(A\,y)\,f\,a_y}}\\
-    \map(\bigwith\,A)\,f &= \lambda a. \withlamb{y}{\map(A\,y)\,f\,(\withprj y a)}\\
-    \map(A \otimes A')\,f &= \lambda b. \letin{(a,a')}{b}{(\map(A)\,f\,a,\map(A')\,f\,a')}
+    \el\left(\Var\,M\right) B &= B\,M\\
+    \el\left(\mathsf{K}\,A\right) B &= A\\
+    \el\left(\bigoplus A\right) B &= \bigoplus_{y:Y}\el(A y)B\\
+    \el\left(\bigamp A\right) B &= \bigamp_{y:Y}\el(A y)B\\
+    \el\left(A \otimes A'\right) B &= \el(A)B \otimes \el(A')B
   \end{align*}
+\end{minipage}%
+\begin{minipage}[t]{.65\textwidth}
+\begin{align*}
+    \map\left(\Var\,M\right)\,f &= f\,M\\
+    \map\left(\K\,A\right)\,f &= \lambda a. a\\
+    \map\left(\bigoplus A\right)\,f &= \lambda a. \letin{\oplusinj y {a_y}}{a}{\oplusinj y {\map(A\,y)\,f\,a_y}}\\
+    \map\left(\bigwith\,A\right)\,f &= \lambda a. \withlamb{y}{\map(A\,y)\,f\,(\withprj y a)}\\
+    \map\left(A \otimes A'\right)\,f &= \lambda b. \letin{(a,a')}{b}{(\map(A)\,f\,a,\map(A')\,f\,a')}
+  \end{align*}
+\end{minipage}%
   \caption{Strictly positive functors functorial actions}
@@
 \begin{figure}
+  \footnotesize
   \begin{mathpar}
@@
 %
-    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy {x} {X} {A}}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
+    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTyLimit {x} {X} {A}}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTyLimit x X A}
 %
@@
     %
-    \inferrule{\Gamma ; \Delta_1, y : \LinSigTy {x} {X} {A}, \Delta_2 \vdash e' : C \\ \Gamma ; \Delta_2 \vdash e : \LinSigTy {x} {X} {A}}{\Gamma; \Delta_1 , \Delta_2 , \Delta_3 \vdash \letin {\sigma~x~a} e {\subst {e'} {\sigma~x~a} y} \equiv \subst {e'} e y : C}
+    \inferrule{\Gamma ; \Delta_1, y : \LinSigTyLimit {x} {X} {A}, \Delta_2 \vdash e' : C \\ \Gamma ; \Delta_2 \vdash e : \LinSigTyLimit {x} {X} {A}}{\Gamma; \Delta_1 , \Delta_2 , \Delta_3 \vdash \letin {\sigma~x~a} e {\subst {e'} {\sigma~x~a} y} \equiv \subst {e'} e y : C}
     \and

### 2025-03-25 paper/paper.tex
combinator stub

@@
 
-We define $\texttt{h}$ by recursion on terms of type $(\A \otimes \A)^{*}$.
+We define $h$ by recursion on terms of type $(\A \otimes \A)^{*}$.
 This recursion is expressed in the derivation tree by invoking the
 elimination principle for Kleene star, written
-as $\fold$. The parse transformer $\texttt{h}$ is more compactly presented in the pseudocode
+as $\fold$. The parse transformer $h$ is more compactly presented in the pseudocode
 of \cref{fig:kleeneabstractproof} by pattern matching on the input and making an
-explicit recursive call in the body of its definition.
+explicit recursive call in the body of its definition. At the bottom of
+\cref{fig:kleeneabstractproof}, we give a defintion of the same function as it
+appears in our current implementation. $h'$ is written via combinators,
+without any pattern matching or introduction
+of named linear variables. Further, when written in this style, we must
+explicitly invoke combinators that explicitly reassociate $\otimes$. \steven{What else to say here?}
 
@@
 \end{lstlisting}
+\begin{lstlisting}
+h' : (*@$\uparrow$@*)((A (*@$\otimes$@*) A)(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
+h' = fold*r nil (cons (*@$\circ$@*) id (*@$\otimes$@*) cons (*@$\circ$@*) assoc(*@$^{-1}$@*))
+\end{lstlisting}
 \caption{(\Agda{\fillerlink}) A parse transformer for abstract grammars}
@@
 k (a , b) = 0to1 (1to1 a (1to2 b stop))
-
 \end{lstlisting}
@@
 are written in a combinator-style, without being able to use named
-variables in linear terms. A benefit of this shallow embedding is that
+variables in linear terms. We give examples of this style in
+\cref{fig:kleeneabstractproof} and the linked Agda code \agdalogo.
+A benefit of this shallow embedding is that
 the parsers are immediately available to a larger Agda development, as

### 2025-03-25 paper/paper.tex
agda code example comment change

@@
 % -*- fill-column: 80; -*-
-\documentclass[review,anonymous,screen,acmsmall,nonacm]{acmart}
+%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
+\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+%% For final camera-ready submission, w/ required CCS and ACM Reference
+%\documentclass[acmsmall]{acmart}\settopmatter{}
+
+%% Journal information
+%% Supplied to authors by publisher for camera-ready submission;
+%% use defaults for review submission.
+\acmJournal{PACMPL}
+\acmVolume{1}
+\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
+\acmArticle{1}
+\acmYear{2018}
+\acmMonth{1}
+\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\startPage{1}
+
+%% Copyright information
+%% Supplied to authors (based on authors' rights management selection;
+%% see authors.acm.org) by publisher for camera-ready submission;
+%% use 'none' for review submission.
+\setcopyright{none}
+%\setcopyright{acmcopyright}
+%\setcopyright{acmlicensed}
+%\setcopyright{rightsretained}
+%\copyrightyear{2018}           %% If different from \acmYear
+
+%% Bibliography style
+\bibliographystyle{ACM-Reference-Format}
+%% Citation style
+%% Note: author/year citations are required for papers published as an
+%% issue of PACMPL.
+%% \citestyle{acmauthoryear}   %% For author/year citations
+\citestyle{acmnumeric}
 \usepackage{mathpartir}
@@
 of \cref{fig:kleeneabstractproof} by pattern matching on the input and making an
-explicit recursive call in the body of its definition. At the bottom of
+explicit recursive call in the body of its definition.
+
+At the bottom of
 \cref{fig:kleeneabstractproof}, we give a defintion of the same function as it
-appears in our current implementation. $h'$ is written via combinators,
-without any pattern matching or introduction
-of named linear variables. Further, when written in this style, we must
-explicitly invoke combinators that explicitly reassociate $\otimes$. \steven{What else to say here?}
+appears in our current Agda implementation \cite{?}. $h'$ is written as a
+composition of combinators,
+each without any pattern matching or introduction
+of named linear variables. More discussion of our implementation is given in \cref{subsec:impl}.
 
@@
 \subsection{Agda Implementation}
-
+\label{subsec:impl}
 This denotational semantics in grammars and parse transformers serves
@@
 
-Elliott \cite{elliottSymbolicAutomaticDifferentiation2021} uses essentially the same denotational semantics as ours, interpreting regular expressions as type-valued functions on strings.
+\citet{elliottSymbolicAutomaticDifferentiation2021} uses essentially the same denotational semantics as ours, interpreting regular expressions as type-valued functions on strings.
 Our denotational semantics of \theoryabbv extends this to a much broader class of grammars.
@@
 
-\bibliographystyle{plain}
 \bibliography{refs.bib}

### 2025-03-25 paper/paper.tex
actually follow the pacmpl template like we were supposed to this whole time whoops

@@
 of \cref{fig:kleeneabstractproof} by pattern matching on the input and making an
-explicit recursive call in the body of its definition.
-
-At the bottom of
+explicit recursive call in the body of its definition. At the bottom of
 \cref{fig:kleeneabstractproof}, we give a defintion of the same function as it
-appears in our current Agda implementation \cite{?}. $h'$ is written as a
-composition of combinators,
-each without any pattern matching or introduction
-of named linear variables. More discussion of our implementation is given in \cref{subsec:impl}.
+appears in our current implementation. $h'$ is written via combinators,
+without any pattern matching or introduction
+of named linear variables. Further, when written in this style, we must
+explicitly invoke combinators that explicitly reassociate $\otimes$. \steven{What else to say here?}
 
@@
 \subsection{Agda Implementation}
-\label{subsec:impl}
+
 This denotational semantics in grammars and parse transformers serves

### 2025-03-25 paper/paper.tex
remove unused import
save some space

--- 2025-03-25 paper/paper.tex
@@
 \usepackage{xspace}
-\usepackage[nounderscore]{syntax}
 \usepackage{thmtools} % needed for sane cleveref naming of theorems
@@
 language=Agda,
-basicstyle=\ttfamily\small,
+basicstyle=\ttfamily\footnotesize,
 keywordstyle=\color{blue},
@@
 \newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
+
+\newcommand{\String}{\textbf{String}}
+\newcommand{\Char}{\textbf{Char}}
+\newcommand{\stringg}{\texttt{String}}
+\newcommand{\charg}{\mathtt{Char}}
 \newcommand{\StringSem}{\mathbf{String}}
@@
 \newcommand{\StringGram}{\texttt{String}}
+\newcommand{\stringNL}{String'}
 \newcommand{\Set}{\mathbf{Set}}
@@
 \newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
-\newcommand{\theoryabbv}{$\textrm{Lambek}^D$~}
+\newcommand{\theoryabbv}{$\textrm{Lambek}^D$\xspace}
 \newcommand{\lnld}{$\textrm{LNL}_D$}
@@
 \end{lstlisting}\begin{mathpar}
+  \footnotesize
   \inferrule
@@
 erroneous derivations,
-\[
-  a : \literal a , b : \literal b \not \vdash a : \literal a \qquad a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a
-\]
-%
+\(
+{a : \literal a , b : \literal b \not \vdash a : \literal a}\) and
+\({a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a}
+\).
+
 Finally, the ordering of characters in a string cannot be ignored while
 parsing, so we omit the exchange rule because it would allow
-for variables in the context to be reordered,
-\[
-  a : \literal a , b : \literal b \not \vdash (b , a) : \literal b \otimes \literal a
-\]
+for variables in the context to be reordered. Likewise, we prevent the following derivation,
+\(
+  {a : \literal a , b : \literal b \not \vdash (b , a) : \literal b \otimes \literal a}
+\).
 % otherwise we would have that $\inl(x\otimes y)$ is a valid
@@
 \end{lstlisting}
-% \begin{align*}
-% \data &~A^{*} : L~\where\\
-%       & \nil : \ltonl {A^{*}} \\
-%       & \cons : \ltonl {(A \lto A^{*} \lto A^{*})}
-% \end{align*}
 \caption{Kleene Star as an inductive type}
@@
 
-At the bottom of
-\cref{fig:kleeneabstractproof}, we give a defintion of the same function as it
-appears in our current Agda implementation \cite{?}. $h'$ is written as a
-composition of combinators,
-each without any pattern matching or introduction
-of named linear variables. More discussion of our implementation is given in \cref{subsec:impl}.
-
 \begin{figure}
@@
 \end{lstlisting}
-\begin{lstlisting}
-h' : (*@$\uparrow$@*)((A (*@$\otimes$@*) A)(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
-h' = fold*r nil (cons (*@$\circ$@*) id (*@$\otimes$@*) cons (*@$\circ$@*) assoc(*@$^{-1}$@*))
-\end{lstlisting}
 \caption{(\Agda{\fillerlink}) A parse transformer for abstract grammars}
@@
     \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg {\Gamma , x : X} {Y}}{\ctxwffjdg \Gamma {\SigTyLimit {x}{X}{Y}}}
+
+    \\
 
@@
 \label{ax:dist}
-For any $A : (x : X) \to Y x \to L$, the function
-\[
- \lamblto{e}{\letin{\sigma\,f\,e'}{e}{\withlamb{x}{e'\,.\pi\,x}}} : \ltonl {\left(  \LinSigTyLimit{f}{\PiTy{x}{X}{Y(x)}}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
-\LinPiTyLimit{x}{X}{\LinSigTyLimit{y}{Y(x)}{A\,x\,y}} \right)}
-\]
+For any $A : (x : X) \to Y(x) \to L$, the definable distribuitivity function
+%%%%%%%%%% An attempt to inline
+% \(
+%  {\LinSigTy{f}{\left(\PiTy{x}{X}{Y(x)}\right)}{\LinPiTy{x}{X}{A\,x\,(f\,x)}} \lto
+% \LinPiTy{x}{X}{\LinSigTy{y}{Y(x)}{A\,x\,y}} \right)}
+% \)
+\(
+ \lamblto{e}{\letin{\sigma\,f\,e'}{e}{\withlamb{x}{e'\,.\pi\,x}}} : \LinSigTyLimit{f}{\PiTy{x}{X}{Y(x)}}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
+\LinPiTyLimit{x}{X}{\LinSigTyLimit{y}{Y(x)}{A\,x\,y}}
+\)
 has an inverse.
@@
   grammars. To build the parse transformer
-  $\ltonl{(Parse_{N} \lto Parse_{D})}$, we strengthen our inductive hypothesis
+  $\ltonl{(Parse_{N} \lto Parse_{D})}$, we strengthen our inductive hypothesis to quantify over every start state
   and build a term
@@
 \newcommand{\DtoN}{\texttt{DtoN}}
-  \[
-    \NtoD : \ltonl {\left(\Trace_{\N}~\s~\true \lto \bigwith\limits_{\X : \D.\states} \bigwith\limits_{\texttt{sInX} : \X \ni \s} \Trace_{\D}~\X~\true \right)}
-  \]
+  \(
+    \NtoD : \ltonl {\left(\Trace_{\N}~\s~\true \lto \LinPiTy{\X}{\D.\states}{\LinPiTy{\texttt{sInX}}{\X \ni \s}{ \Trace_{\D}~\X~\true}} \right) }
+  \)
   that maps a trace in $\N$ from an arbitrary state $\s$ to a trace in $\D$
@@
 To construct a term from DFA traces to NFA traces, we similarly strengthen our
-inductive hypothesis and build a parse transformer
-\[
-  \DtoN : \ltonl {\left( \Trace_{\D}~\X~\true \lto \bigoplus\limits_{s : \N.\states}\bigoplus\limits_{\texttt{sInX} : \X \ni \s} \Trace_{\N}~s~\true \right)}
-\]
+inductive hypothesis and build a parse transformer that ranges over arbitrary $\epsilon$-closed subsets $X : \mathbb{P}_{\epsilon}{\N.\states}$,
+\(
+  \DtoN : \ltonl {\left( \Trace_{\D}~\X~\true \lto \LinSigTy{s}{\N.\states}{\LinSigTy{\texttt{sInX}}{\X \ni \s}{ \Trace_{\N}~s~\true}} \right)}
+\).
 Because the states of $D$ are $\epsilon$-closed subsets of $N.\states$, if two
@@
 specify a global disambiguation strategy.
-
+\end{proof}
 %% For a moment, let us turn our attention to the content behind the traces in $\D$.
@@
 %% weakly equivalent.
-\end{proof}
 
@@
 \cite{thompsonProgrammingTechniquesRegular1968}, showing that NFAs are, up to strong equivalence, closed under each type operation for regular expressions.
-
+\end{proof}
 %% In this paper we will walk through
@@
 %% elided details may be found in the Agda formalization.
-\end{proof}
 
@@
 %% a first-class linear type in \theoryabbv. 
-\[
-Reify~P = \LinSigTyLimit {w} {\StringSem} {\LinSigTyLimit {x} {P~w} {~\lceil w \rceil}}
-\]
+\(
+Reify~P = \LinSigTy{w} {\StringSem} {\LinSigTy {x} {P~w} {~\lceil w \rceil}}
+\)
 %% \steven{Probably put this ceiling business around the discussion of the linear
@@
 equivalent to the set of parse transformers:
-\[ \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\]
+\( \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\).
+
 Next, a parse in the equalizer $\equalizer{a}{f}{g}$ is defined as
@@
 \begin{figure}
+  \footnotesize
   \begin{align*}

### 2025-03-25 paper/paper.tex
agda implementation section

@@
 \newcommand{\Set}{\mathbf{Set}}
+\newcommand{\hSet}{\mathbf{hSet}}
 \newcommand{\Syn}{\mathbf{Synx}}
@@
 \begin{figure}
+  \footnotesize
   \begin{tikzpicture}[node distance = 17mm ]
@@
 
-\subsection{Agda Implementation}
+\subsection{Agda Implementation \Agda{\fillerlink}}
 \label{subsec:impl}
@@
 Cubical Agda. The implementation is a shallow embedding, meaning that
-rather than formalizing a syntax of \theoryabbv types and terms, we
-work directly with Agda types and a definition of a formal grammar as
-a function $\StringSem \to \Set$. Then we implement each of the type and
+rather than formalizing a syntax of \theoryabbv types and terms, we interpret
+the non-linear universe $U$ as Cubical Agda's universe of (homotopy) sets (at
+some universe level $\ell$), $\hSet~\ell$, and the linear universe $L$ as the
+function type
+$\StringSem \to \hSet~\ell$. Then we implement each of the type and
 term constructors of \theoryabbv as combinators on formal grammars or
@@
 we are careful to only construct grammars, terms and proofs using
-constructs that are possible in \theoryabbv. The main difference
+constructs that are possible in \theoryabbv.
+
+The main difference
 between our shallow embedding and \theoryabbv is that our linear terms
 are written in a combinator-style, without being able to use named
-variables in linear terms. We give examples of this style in
-\cref{fig:kleeneabstractproof} and the linked Agda code \agdalogo.
+variables in linear terms. For example, the function $h$ from
+\cref{fig:kleeneabstractproof} would be written with the fold combinator applied
+to subexpressions for the $nil$ and $cons$ cases,
+\(
+h = fold~nil~(cons \circ id \otimes cons \circ assoc^{-1})
+\). Notice in the $cons$ case of the fold, we manually reassociate
+$(\literal{a} \otimes \literal{a}) \otimes \literal{a}^{*}$ to
+$\literal{a} \otimes (\literal{a} \otimes \literal{a}^{*})$, then act on the
+left by the identity and on the right by $cons$ to get a parse
+$\literal{a} \otimes \literal{a}^{*}$, and finally we act again by $cons$ to
+finally produce a parse of $\literal{a}^{*}$.
+
 A benefit of this shallow embedding is that
 the parsers are immediately available to a larger Agda development, as
-they are just normal Agda code. In future work we will look to
+they are just normal Agda code. Although, even in the small example from
+\cref{fig:kleeneabstractproof} we must manage additional complexity --- such as
+manually reassociations --- and
+this only grows more complex in larger programs. In future work we will look to
 implement a type checker for a syntax closer to the presentation in
-this paper, but with the goal of still making it integrate with an
+this paper, while maintaining the goal of integration with an
 existing proof assistant.

### 2025-03-25 paper/paper.tex
update the extensionality discussion

@@
 reflection rule. Extensional equality makes type checking of our syntax as such
-undecidable because the conversion rule may require an arbitrarily complex
+undecidable \citep{Hofmann_1997} because the conversion rule may require an arbitrarily complex
 equality proof with no explicit proof term, but in our Agda implementation we
 must provide all of these equalities manually, and so the extensionality does
-not raise any issues \cite{Hofmann_1997}.
+not raise any issues for our implementation. This makes
+\theoryabbv into an extensional theory, supporting function extensionality and
+the uniqueness of identity proofs. The development could be ported to an intensional type theory in the future, possibly requiring the use of setoids to handle function extensionality.
 
@@
 the non-linear universe $U$ as Cubical Agda's universe of (homotopy) sets (at
-some universe level $\ell$), $\hSet~\ell$, and the linear universe $L$ as the
+xsome universe level $\ell$), $\hSet~\ell$, and the linear universe $L$ as the
 function type
-$\StringSem \to \hSet~\ell$. Then we implement each of the type and
+$\StringSem \to \hSet~\ell$.
+We use $\hSet~\ell$ as it ensures that uniqueness of identity proofs holds for the interpretation of all types in \theoryabbv, as it does in any extensional type theory.
+Then we implement each of the type and
 term constructors of \theoryabbv as combinators on formal grammars or
@@
 term equalities. Cubical Agda is convenient for this purpose as it has
-built-in support for function extensionality which we use
-extensively. Axioms such as distributivity of $\oplus/\&$ and
+built-in support for function extensionality which is convenient for the verification of equality rules. Axioms such as distributivity of $\oplus/\&$ and
 disjointness of constructors are then provable directly in Agda, and

### 2025-03-25 paper/paper.tex
smaller related work discussion + remove line break in paper
tweak changed.md wording, add year to Coquand note, use authoryear cites
Benton and Wadler -> Benton
implement richer nonlinear inductive

@@
 %% issue of PACMPL.
-%% \citestyle{acmauthoryear}   %% For author/year citations
-\citestyle{acmnumeric}
+\citestyle{acmauthoryear}   %% For author/year citations
+%% \citestyle{acmnumeric}
 \usepackage{mathpartir}
@@
 Next, we include standard $\Sigma,\Pi$, empty, unit, Boolean, and natural number types.
+More complex inductive types --- such as sum types, list types, and $Fin~n$ --- can be defined in terms of these primitives, and we give their encoding in \cref{TODO} of the appendix.
 
@@
 finally produce a parse of $\literal{a}^{*}$.
-
 A benefit of this shallow embedding is that
@@
 dependent commutative linear type theory of
-\citet{krishnaswami_integrating_2015}, itself an extension of Benton and
-Wadler's linear-non-linear calculus \cite{benton1994}. A
-distinct feature of these systems is that linear types and non-linear types are
-distinct sorts, so the linear logic $!$ operator is not a primitive, but is
-instead definable, for instance in \theoryabbv as $!A = \LinSigTy{\_}{\ltonl A}{I}$.
-V\'ak\'ar develops a dependent linear type theory where instead the non-linear
-types are accessed using the $!$ modality, similar to Girard's original approach
+\citet{krishnaswami_integrating_2015}, itself an extension of Benton's
+linear-non-linear calculus \cite{benton1994}. A distinct feature of these
+systems is that linear types and non-linear types are distinct sorts, so the
+linear logic $!$ operator is not a primitive, but is instead definable, for
+instance in \theoryabbv as $!A = \LinSigTy{\_}{\ltonl A}{I}$.  V\'ak\'ar
+develops a dependent linear type theory where instead the non-linear types are
+accessed using the $!$ modality, similar to Girard's original approach
 \cite{vakar2015,girard_linear_1987}. Additionally, V\'ak\'ar develops a general
@@
 
+\subsection{Non-linear Types}
+We define sum types, list types, and $Fin~n$ from primitve non-linear types.
+
+For $X , Y : U$, define the sum type $X + Y$ as,
+
+\[
+X + Y = \SigTyLimit{b}{Bool}{elim_{Bool}(U, X, Y)(b)}
+\]
+
+For $n : Nat$, define $Fin~n$ as,
+
+\begin{align*}
+  Fin~0 &= \bot \\
+  Fin~(suc n) &= 1 + (Fin~n)
+\end{align*}
+
+For $X : U$, define $List~X$ as,
+
+\[
+  List~X = \SigTyLimit{n}{Nat}{\PiTyLimit{k}{Fin~n}{X}}
+\]
+
 \max{TODO: include the rest of the rules}
 \steven{I added rules for Bool, Nat, and List. What else do we need?}
-
 \begin{figure}
@@
 
+\subsection{Substitutions}
 \begin{definition}

### 2025-03-25 paper/paper.tex
description of figures in appendix

@@
 \newcommand{\isLinTy}{\textrm{ lin. type}}
+\newcommand{\isSmallLin}{\textrm{ small lin.}}
 \newcommand{\isLinCtx}{\textrm{ lin. ctx.}}
@@
 
-    \inferrule{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma {\quoteTy X} U}
-
-    \inferrule{\linctxwffjdg \Gamma A}{\nonlinterm \Gamma {\quoteTy A} L}
+    \inferrule{\Gamma \vdash  X \isSmall}{\nonlinterm \Gamma {\quoteTy X} U}
+
+    \inferrule{\Gamma \vdash A \isSmallLin}{\nonlinterm \Gamma {\quoteTy A} L}
 
@@
 types. These are defined as
-universes ``ala Coquand'' in that we define a judgment saying when a
-non-linear type is small and define the universe to internalize
-precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. These universe
-types are needed so that we can define types by recursion on natural
+universes ``ala Coquand'' in that we define judgments saying when
+non-linear and linear types are \emph{small} and define the universes to internalize
+precisely this judgment \cite{coquandPresheafModel,lmcs:7713}.
+The definition of smallness is simply that all types are small as long as their sub-formulae are, with the exception of the two universe types themselves.
+These universe
+types are needed so that we can define non-linear and linear types by recursion on natural
 numbers.
@@
     \inferrule{\nonlinterm \Gamma M L}{\linctxwffjdg \Gamma {\unquoteTy M}}
+
+    \boxed{\inferrule{\linctxwffjdg \Gamma A}{\Gamma \vdash A \isSmallLin}}
 
@@
 
+%% \pagebreak
+
+\begin{acks}
+  This material is based upon work supported by the
+  \grantsponsor{GS100000001}{National Science
+    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
+  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
+  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
+  conclusions or recommendations expressed in this material are those
+  of the author and do not necessarily reflect the views of the
+  National Science Foundation.
+\end{acks}
+
 \bibliography{refs.bib}
-
-\pagebreak
 
@@
   \caption{Strictly positive functors functorial actions}
+  \label{fig:spf-act}
 \end{figure}
@@
 substitution.
+
+In \cref{fig:spf-act}, we provide the functorial actions for the strictly
+positive functors defined in \cref{fig:iilt}. We also include the full set of
+rules for non-linear types in \cref{fig:full-non-linear-types}, and we provide
+their judgmental equalities in \cref{fig:jdg-eq-nonlinear}. Finally,
+\cref{fig:jdgeq} contains the judgmental equalities between linear terms in
+\theoryabbv.
 
@@
 
-\max{TODO: include the rest of the rules}
-\steven{I added rules for Bool, Nat, and List. What else do we need?}
+\begin{figure}
+  \begin{mathpar}
+    \inferrule{\Gamma \vdash M : U}{\Gamma \vdash \unquoteTy M \isSmall}
+
+    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash Y \isSmall}{\Gamma \vdash \PiTyLimit {x}{X}{Y} \isSmall}
+    \cdots
+  \end{mathpar}
+  \caption{Small Non-linear Types}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    \inferrule{\Gamma \vdash M : L}{\Gamma \vdash \unquoteTy M \isSmallLin}
+
+    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash A \isSmallLin}{\Gamma \vdash \LinPiTyLimit {x}{X}{A} \isSmallLin}
+    \cdots
+  \end{mathpar}
+  \caption{Small Linear Types}
+\end{figure}
+
 \begin{figure}
@@
   $\Gamma$ support the admissible actions of a substitution $\gamma
-  \in \Subst(\Gamma',\Gamma)$ in \Cref{fig:non-linear-substitution}.
+  \in \Subst(\Gamma',\Gamma)$ given in \cref{fig:non-linear-substitution,fig:substact}.
   %% Further, these all satisfy identity and composition equations M[\Gamma] = M and M[\gamma][\gamma'] = M[(\gamma \circ \gamma')]

### 2025-03-25 paper/paper.tex
fix typo
fix issues with smallness, start smallness fig in appendix, fix a citation

@@
   \caption{Strictly positive functors functorial actions}
-  \label{fig:spf-act}
 \end{figure}
@@
 substitution.
-
-In \cref{fig:spf-act}, we provide the functorial actions for the strictly
-positive functors defined in \cref{fig:iilt}. We also include the full set of
-rules for non-linear types in \cref{fig:full-non-linear-types}, and we provide
-their judgmental equalities in \cref{fig:jdg-eq-nonlinear}. Finally,
-\cref{fig:jdgeq} contains the judgmental equalities between linear terms in
-\theoryabbv.
 
@@
   $\Gamma$ support the admissible actions of a substitution $\gamma
-  \in \Subst(\Gamma',\Gamma)$ given in \cref{fig:non-linear-substitution,fig:substact}.
+  \in \Subst(\Gamma',\Gamma)$ in \Cref{fig:non-linear-substitution}.
   %% Further, these all satisfy identity and composition equations M[\Gamma] = M and M[\gamma][\gamma'] = M[(\gamma \circ \gamma')]

### 2025-03-25 paper/paper.tex
mention smallness in the changes
fill in smallness figures
remove todo

--- 2025-03-25 paper/paper.tex
@@
 numbers.
-Next, we include standard $\Sigma,\Pi$, empty, unit, Boolean, and natural number types.
-More complex inductive types --- such as sum types, list types, and $Fin~n$ --- can be defined in terms of these primitives, and we give their encoding in \cref{TODO} of the appendix.
+Next, we include standard $\Sigma,\Pi$, empty, unit, Boolean, and natural number
+types. More complex inductive types --- such as sum types, list types, and
+$Fin~n$ --- can be defined in terms of these primitives, and we give their
+encoding in the appendix.
 
@@
 
-%% \pagebreak
+\newpage
 
@@
 
+\newpage
 \ifappendix
@@
   Fin~0 &= \bot \\
-  Fin~(suc n) &= 1 + (Fin~n)
+  Fin~(suc~n) &= 1 + (Fin~n)
 \end{align*}
@@
 \begin{figure}
+  \footnotesize
   \begin{mathpar}
+    \inferrule{~}{\Gamma \vdash 1 \isSmall}
+
+    \inferrule{~}{\Gamma \vdash Bool \isSmall}
+
+    \inferrule{~}{\Gamma \vdash \bot \isSmall}
+
+    \inferrule{~}{\Gamma \vdash Nat \isSmall}
+
+    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash Y \isSmall}{\Gamma \vdash \PiTyLimit {x}{X}{Y} \isSmall}
+
+    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash Y \isSmall}{\Gamma \vdash \SigTyLimit {x}{X}{Y} \isSmall}
+
     \inferrule{\Gamma \vdash M : U}{\Gamma \vdash \unquoteTy M \isSmall}
 
-    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash Y \isSmall}{\Gamma \vdash \PiTyLimit {x}{X}{Y} \isSmall}
-    \cdots
+    \inferrule{\Gamma \vdash A \isSmallLin}{\Gamma \vdash {\ltonl A} \isSmall}
+
+    \inferrule{\Gamma \vdash X \isSmall \and \nonlinterm \Gamma M X \and
+      \nonlinterm \Gamma N X}{\Gamma \vdash {M =_{X} N} \isSmall}
   \end{mathpar}
@@
 \begin{figure}
+  \footnotesize
   \begin{mathpar}
+    \inferrule{~}{\Gamma \vdash I \isSmallLin}
+
+    \inferrule{c \in \Sigma}{\Gamma \vdash {\literal{c}} \isSmallLin}
+
+    \inferrule{\Gamma \vdash A \isSmallLin \and \Gamma \vdash B \isSmallLin}{
+      \Gamma \vdash {A \otimes B} \isSmallLin}
+
+    \inferrule{\Gamma \vdash A \isSmallLin \and \Gamma \vdash B \isSmallLin}{
+      \Gamma \vdash {A \lto B} \isSmallLin}
+
+    \inferrule{\Gamma \vdash A \isSmallLin \and \Gamma \vdash B \isSmallLin}{
+      \Gamma \vdash {A \tol B} \isSmallLin}
+
+    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash A \isSmallLin}{\Gamma \vdash \LinSigTyLimit {x}{X}{A} \isSmallLin}
+
+    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash A \isSmallLin}{\Gamma \vdash \LinPiTyLimit {x}{X}{A} \isSmallLin}
+
+    \inferrule{\Gamma \vdash A \isSmallLin \and
+      \Gamma \vdash B \isSmallLin \and
+      \nonlinterm \Gamma f {\ltonl{(A \lto B)}} \and
+      \nonlinterm \Gamma g {\ltonl{(A \lto B)}}}{\Gamma \vdash {\equalizer
+        {a}{f}{g}} \isSmallLin}
+
     \inferrule{\Gamma \vdash M : L}{\Gamma \vdash \unquoteTy M \isSmallLin}
-
-    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash A \isSmallLin}{\Gamma \vdash \LinPiTyLimit {x}{X}{A} \isSmallLin}
-    \cdots
   \end{mathpar}

### 2025-03-25 paper/paper.tex
fix typo

@@
 the non-linear universe $U$ as Cubical Agda's universe of (homotopy) sets (at
-xsome universe level $\ell$), $\hSet~\ell$, and the linear universe $L$ as the
+some universe level $\ell$), $\hSet~\ell$, and the linear universe $L$ as the
 function type

### 2025-03-25 paper/paper.tex
capitalize cubical

@@
 constructed in our Agda implementation as an inductive type of $IW$
-trees which are already available in the cubical library of Agda
+trees which are already available in the Cubical library of Agda
 \cite{The_Agda_Community_Cubical_Agda_Library_2024}.  Then an element $F \in \sem{X \to

### 2025-03-26 paper/paper.tex
fix phrasings from pedros comments

@@
   parse transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B
-    \lto \A)}$. $\A$ is a \emph{retract} of $\B$ if $\A$ and $\B$ are
+    \lto \A)}$. $\A$ is a \emph{retract} of $\B$ if they are
   weakly equivalent and $\lambda a. g(f(a)) \equiv \lambda a.a$. They
@@
 linear logic, but their terms do not include elimination rules, and so
-can only express parse trees, and so cannot be used to express
-verified parsers.
+can only express parse trees, not verified parsers.
 

### 2025-03-26 paper/paper.tex
comment out acks to build diff

@@
 
-\begin{acks}
-  This material is based upon work supported by the
-  \grantsponsor{GS100000001}{National Science
-    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-  conclusions or recommendations expressed in this material are those
-  of the author and do not necessarily reflect the views of the
-  National Science Foundation.
-\end{acks}
+% \begin{acks}
+%   This material is based upon work supported by the
+%   \grantsponsor{GS100000001}{National Science
+%     Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
+%   No.~\grantnum{GS100000001}{nnnnnnn} and Grant
+%   No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
+%   conclusions or recommendations expressed in this material are those
+%   of the author and do not necessarily reflect the views of the
+%   National Science Foundation.
+% \end{acks}
 

### 2025-03-27 Everythings.hs
should i flatten the directory structure?

@@ -77,7 +77,7 @@ checkEverythings excludes dirs = do
 checkREADME :: IO ()
 checkREADME = do
   (sub_dirs', _) <- getSubDirsFiles ["."]
-  let sub_dirs = filter (\d -> d `notElem` ["Everything.hs", "_build", "README.agda", "Makefile", "Dockerfile"]) sub_dirs'
+  let sub_dirs = filter (\d -> d `notElem` ["Everything.hs", "_build", "README.agda", "Makefile", "Dockerfile", "paper"]) sub_dirs'
   imported <- getImported ["README"]
   let missing_files = fmap (\dir -> ["Everything",dir]) sub_dirs \\ imported
   if null missing_files then pure ()
@@ -115,7 +115,7 @@ helpText = unlines [
 main :: IO ()
 main = do
   all_dirs' <- filter ('.' `notElem`) <$> getDirectoryContents "."
-  let all_dirs = filter (\d -> d `notElem` ["Everything.hs", "_build", "README.agda", "Makefile", "Dockerfile"]) all_dirs'
+  let all_dirs = filter (\d -> d `notElem` ["Everything.hs", "_build", "README.agda", "Makefile", "Dockerfile", "paper"]) all_dirs'
   args <- getArgs
   case args of
     "check":dirs -> checkEverythings [] dirs

### 2025-03-27 check-line-lengths.sh
make src directory

@@ -1,45 +1,0 @@
-#!/usr/bin/env bash
-# originally written by Steven Schaefer <stschaef>
-
-error=0
-
-VENV_DIR="env"
-
-# Check if the virtual environment directory exists
-if [ ! -d "$VENV_DIR" ]; then
-    echo "Virtual environment does not exist. Creating one now..."
-    # Create the virtual environment with python3
-    python3 -m venv "$VENV_DIR" || { echo "Failed to create virtual environment."; exit 1; }
-
-fi
-
-source "$VENV_DIR/bin/activate"
-
-if [ $? -ne 0 ]; then
-    echo "Error: Unable to activate the virtual environment."
-    exit 2
-fi
-
-echo "Virtual environment activated."
-
-pip3 install wcwidth || { echo "Failed to install wcwidth."; exit 3; }
-
-echo "Package 'wcwidth' has been successfully installed in the virtual environment."
-
-while IFS= read -r -d '' file; do
-  python3 -c '
-import sys
-from wcwidth import wcswidth
-filename = sys.argv[1]
-try:
-    with open(filename, "r", encoding="utf-8") as file:
-        for lineno, line in enumerate(file, 1):
-            if wcswidth(line.rstrip("\n")) > 80:
-                print(f"{filename}:{lineno}: line too long")
-                sys.exit(1)
-except Exception as e:
-    print(f"Error processing {filename}: {e}")
-    sys.exit(1)
-  ' "${file}" || error=1
-done  < <(find . -name '*.agda' -print0)
-exit ${error}

### 2025-03-27 paper/submission.tex

--- 2025-03-25 paper/paper.tex
@@
 % -*- fill-column: 80; -*-
-%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[acmsmall]{acmart}\settopmatter{}
-
-%% Journal information
-%% Supplied to authors by publisher for camera-ready submission;
-%% use defaults for review submission.
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
-\acmArticle{1}
-\acmYear{2018}
-\acmMonth{1}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission;
-%% use 'none' for review submission.
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-%% \citestyle{acmnumeric}
+\documentclass[review,anonymous,screen,acmsmall,nonacm]{acmart}
 \usepackage{mathpartir}
@@
 \usepackage{xspace}
-\usepackage{thmtools} % needed for sane cleveref naming of theorems
-\usepackage{hyperref}
-\usepackage[capitalise]{cleveref}
 
@@
 language=Agda,
-basicstyle=\ttfamily\footnotesize,
+basicstyle=\ttfamily\small,
 keywordstyle=\color{blue},
@@
 
+\usepackage{hyperref}
+\usepackage[capitalise]{cleveref}
 \usepackage{stmaryrd}
@@
 \newcommand{\Subst}{\textrm{Subst}}
-\newcommand{\SPF}{SPF}
+\newcommand{\SPF}{\mathsf{SPF}}
 \newcommand{\Var}{\mathsf{Var}}
@@
 \newcommand{\inr}{\mathsf{inr}}
-\newcommand{\sem}[1]{\left\llbracket{#1}\right\rrbracket}
+\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
 \newcommand{\semg}[1]{\sem{#1}\gamma}
@@
 \newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
-
 \newcommand{\String}{\textbf{String}}
@@
 \newcommand{\charg}{\mathtt{Char}}
-\newcommand{\StringSem}{\mathbf{String}}
-\newcommand{\CharGram}{\texttt{Char}}
-\newcommand{\StringGram}{\texttt{String}}
-\newcommand{\stringNL}{String'}
 \newcommand{\Set}{\mathbf{Set}}
-\newcommand{\hSet}{\mathbf{hSet}}
 \newcommand{\Syn}{\mathbf{Synx}}
@@
 \newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
-\newcommand{\theoryabbv}{$\textrm{Lambek}^D$\xspace}
+\newcommand{\theoryabbv}{$\textrm{Lambek}^D$~}
 \newcommand{\lnld}{$\textrm{LNL}_D$}
@@
 \newcommand{\isLinTy}{\textrm{ lin. type}}
-\newcommand{\isSmallLin}{\textrm{ small lin.}}
 \newcommand{\isLinCtx}{\textrm{ lin. ctx.}}
@@
 \newcommand{\internalize}[1]{\lceil#1\rceil}
-\newcommand{\listTy}[1]{List\left( #1 \right)}
 
@@
 \newcommand{\apptol}[2]{#1 \mathop{{}^{\tol}} #2}
-\newcommand{\PiTy}[3]{\textstyle\prod_{#1 : #2} #3}
-\newcommand{\SigTy}[3]{\textstyle\sum_{#1 : #2} #3}
-\newcommand{\PiTyLimit}[3]{\textstyle\prod\limits_{#1 : #2} #3}
-\newcommand{\SigTyLimit}[3]{\textstyle\sum\limits_{#1 : #2} #3}
-\newcommand{\LinPiTy}[3]{\textstyle\bigamp_{#1 : #2} #3}
+\newcommand{\PiTy}[3]{\textstyle\prod (#1 : #2). #3}
+\newcommand{\SigTy}[3]{\textstyle\sum (#1 : #2). #3}
+\newcommand{\LinPiTy}[3]{\textstyle\bigamp (#1 : #2). #3}
 \newcommand{\LinPiTyLimit}[3]{\bigwith\limits_{#1 : #2} #3}
-\newcommand{\LinSigTy}[3]{\textstyle\bigoplus_{#1 : #2} #3}
+\newcommand{\LinSigTy}[3]{\textstyle\bigoplus (#1 : #2). #3}
 \newcommand{\LinSigTyLimit}[3]{\bigoplus\limits_{#1 : #2} #3}
@@
 \newcommand{\linctxwffjdg}[2]{#1 \vdash #2 \isLinTy}
-\newcommand{\nonlinterm}[3]{#1 \vdash #2 : #3}
-\newcommand{\linterm}[4]{#1 ; #2 \vdash #3 : #4}
-\newcommand{\nonlineq}[4]{#1 \vdash #2 \equiv #3 : #4}
-\newcommand{\lineq}[5]{#1 ; #2 \vdash #3 \equiv #4 : #5}
-
-\newcommand{\LLL}{\textrm{LL}}
-\newcommand{\LRR}{\textrm{LR}}
-\newcommand{\LALRR}{\textrm{LALR}}
-\newcommand{\LL}[1]{\LLL(#1)}
-\newcommand{\LR}[1]{\LRR(#1)}
-\newcommand{\LALR}[1]{\LALRR(#1)}
 
@@
   \usebox{\logoagdabox}}%
-
-\newcommand{\fillerlink}{http://localhost:1313}
-\newcommand{\Agda}[1]{\href{#1}{\agdalogo}}
 
@@
 \newcommand{\pipe}{\,|\,}
-
-\newcommand{\states}{\mathtt{states}}
-\newcommand{\labelt}{label}
-\newcommand{\transitions}{\texttt{transitions}}
-\newcommand{\epstransitions}{\epsilon\texttt{transitions}}
-\newcommand{\isAcc}{\mathtt{isAcc}}
-\newcommand{\init}{\mathtt{init}}
-\newcommand{\src}{\mathtt{src}}
-\newcommand{\dst}{\mathtt{dst}}
-\newcommand{\pparse}{\mathtt{parse}}
-\newcommand{\print}{\mathtt{print}}
-\newcommand{\epssrc}{\epsilon\mathtt{src}}
-\newcommand{\epsdst}{\epsilon\mathtt{dst}}
-\newcommand{\balanced}{\mathtt{balanced}}
-\newcommand{\Dyck}{\mathtt{Dyck}}
-\newcommand{\N}{\texttt{N}}
-
-% These declarations are to get cleveref to play nice with
-% different theorem-like environemnts. Otherwise \cref{lem1.1}
-% will display 'Theorem 1.1'
-
-\declaretheoremstyle[headfont=\normalfont\itshape]{defstyle}
-\declaretheoremstyle[headfont=\normalfont\scshape]{thmstyle}
-
-\declaretheorem[name=Definition, style=defstyle, numberwithin=section]{definition}
-\declaretheorem[style=thmstyle, name=Theorem, numberlike=definition]{theorem}
-\declaretheorem[style=thmstyle, name=Axiom, numberlike=definition]{axiom}
-\declaretheorem[style=thmstyle, name=Corollary, numberlike=definition]{corollary}
-\declaretheorem[style=thmstyle, name=Construction, numberlike=definition]{construction}
-
-\crefname{definition}{Definition}{Definitions}
-\crefname{theorem}{Theorem}{Theorems}
-\crefname{axiom}{Axiom}{Axioms}
-\crefname{corollary}{Corollary}{Corollarys}
-\crefname{construction}{Construction}{Constructions}
-
-% \newtheorem{axiom}{Axiom}[section]
-% \newtheorem{construction}{Construction}
 
@@
 source program. Eventually, a correct parser was implemented using an
-automaton that is formally verified to implement an $\LR{1}$ grammar
+automaton that is formally verified to implement an LR grammar
 \cite{jourdanValidatingLRParsers2012}.
@@
 particular grammar formalism such as non-left-recursive grammars or
-$\LL{1}$ grammars
+LL(1) grammars
 \cite{lasserCoStarVerifiedALL2021,EdelmannZippy2020,danielssonTotalParserCombinators2010}.
@@
 not a verified parser generator that compiles regular expressions to
-deterministic finite automata, but is instead a domain specific
+deterministic finite automata, but is instead is a domain specific
 language \emph{for writing} such a verified parser generator.
@@
 power to support arbitrarily powerful grammar formalisms: inductive
-linear\footnote{Throughout, we shall use ``linear types'' to refer to the
-non-commutative linear types.} types, as well as dependency of linear types on non-linear
+linear types, as well as dependency of linear types on non-linear
 data. The resulting system has two kinds of types: non-linear types
@@
 non-deterministic and deterministic automata, as well as some example
-context-free grammars and parsers based on $\LL{1}$ and $\LR{1}$
-automata. Throughout this paper, \agdalogo~ will mark
-results that are mechanized in our Agda development and provide a link to their
-implementation.
+context-free grammars and parsers based on LL(1) and LR(1)
+automata. Throughout this paper, we will use \agdalogo~ to mark
+results that are mechanized in our Agda development.
 
 Our Agda prototype is based on a \emph{denotational semantics} of
-\theoryabbv. The core idea of the denotational semantics stems from an observation of Elliott:
-% The core idea of the denotational semantics is
-% \cite{elliottSymbolicAutomaticDifferentiation2021}, Elliott describes
-a formal grammar is a type-level predicate on strings that proves
-language membership \cite{elliottSymbolicAutomaticDifferentiation2021}. That is, a \emph{formal grammar} $A$ is a
-function $\StringSem \to \Set$ such that for a string $w$, $A~w$ is the
+\theoryabbv. The core idea of the denotational semantics is
+\cite{elliottSymbolicAutomaticDifferentiation2021}, Elliott describes
+a formal grammar as type-level predicates on strings that prove
+language membership. That is, a \emph{formal grammar} $A$ is a
+function $\String \to \Set$ such that for a string $w$, $A~w$ is the
 set of ``proofs'' showing that $w$ belongs to the language recognized
@@
     extends prior work on dependent linear-non-linear type theory to
-    support inductive linear types.
-  \item A demonstration of how to encode many common grammar formalisms
+    support inductive linear types
+  \item Demonstration of how to encode many common grammar formalisms
     (regular expressions, (non-)deterministic automata, context-free
@@
   \item A denotational semantics for \theoryabbv that shows that the
-    parsers are in fact verified to be correct, and the soundness of the
+    parsers are in fact verified to be correct and soundness of the
     equational theory.
@@
 example programs from \theoryabbv to build intuition.  From there, in
-\cref{sec:tt} we provide the syntax, typing, and equational theory of
+\cref{sec:tt} we provide the syntax, typing and equational theory of
 \theoryname. In \cref{sec:applications} we demonstrate the
@@
 a parse of $\A$ over the string $\w$ or a
-parse of $\B$ over the string $\w$. Similarly, a parse of $\A \otimes \B$ for
-$\w$ is a splitting of $w$ into two strings $\w_{\A}$ and $\w_{\B}$ with
-parses for $\A$
+parse of $\B$ over the string $\w$. Similarly, $\w$ matches $\A \otimes \B$ if
+$\w$ can be split into two strings $\w_{\A}$ and $\w_{\B}$ that match $\A$
 and $\B$, respectively.
 
-A grammar $A$ derives the word $w$ if there exists a derivation
-$\internalize{w} \vdash a : A$,
-where $\internalize \w$ is a context with one variable for each character of
-$\w$. The term $a : A$ represents a \emph{parse tree} of $w$ for the grammar $A$. For
-example, to define a parse tree for \stringquote{ab}, we use
+For a type $\A$, a parse tree of a string $\w$ is represented as a
+term of type $\A$ in the context $\internalize \w$, where
+$\internalize \w$ is a context with one variable for each character of
+$\w$. For example, to define a parse tree for \stringquote{ab}, we use
 the context $\internalize{\stringquote{ab}} = a : \literal a , b :
@@
 f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
-f (a , b) = inl (a , b)
+f (a , b) = inl (a (*@$\otimes$@*) b)
 \end{lstlisting}\begin{mathpar}
-  \footnotesize
   \inferrule
@@
     }
-    {a : \literal a , b : \literal b \vdash (a , b) : \literal a \otimes
+    {a : \literal a , b : \literal b \vdash a \otimes b : \literal a \otimes
       \literal b}
   }
-  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a , b) :
+  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a \otimes b) :
     (\literal a \otimes \literal b) \oplus \literal c}
 \end{mathpar}
-\caption{(\Agda{fillerlink}) \stringquote{ab} is parsed by $(\literal a \oplus \literal b) \otimes \literal c$
-}
+\caption{\stringquote{ac} matches
+  $(\literal a \oplus \literal b) \otimes \literal c$}
 \label{fig:fingram}
@@
 erroneous derivations,
-\(
-{a : \literal a , b : \literal b \not \vdash a : \literal a}\) and
-\({a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a}
-\).
-
+\[
+  a : \literal a , b : \literal b \not \vdash a : \literal a \qquad a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a
+\]
+
+%
 Finally, the ordering of characters in a string cannot be ignored while
 parsing, so we omit the exchange rule because it would allow
-for variables in the context to be reordered. Likewise, we prevent the following derivation,
-\(
-  {a : \literal a , b : \literal b \not \vdash (b , a) : \literal b \otimes \literal a}
-\).
+for variables in the context to be reordered,
+\[
+  a : \literal a , b : \literal b \not \vdash (b , a) : \literal b \otimes \literal a
+\]
 % otherwise we would have that $\inl(x\otimes y)$ is a valid
@@
 $\B$ as output. The arrow, $\uparrow$, wrapping these
-constructors means that the constructors are not consumed
+constructors means that the constructors by themselves are not consumed
 upon usage, and so are \emph{non-linear} values themselves. That is, the
@@
 \end{lstlisting}
+% \begin{align*}
+% \data &~A^{*} : L~\where\\
+%       & \nil : \ltonl {A^{*}} \\
+%       & \cons : \ltonl {(A \lto A^{*} \lto A^{*})}
+% \end{align*}
 \caption{Kleene Star as an inductive type}
@@
 \cref{fig:kleenestarderivation} gives a derivation that shows
-\stringquote{ab} is parsed by the regular expression $({\literal a}^{*}
+\stringquote{ab} matches the regular expression $({\literal a}^{*}
 \otimes \literal b) \oplus \literal c$. The leaves of the proof tree
@@
   }
-  {a : {\literal a} , b : {\literal b} \vdash (\cons~a~\nil , b) : {\literal a}^{*} \otimes {\literal b}}
+  {a : {\literal a} , b : {\literal b} \vdash (\cons~a~\nil) \otimes b : {\literal a}^{*} \otimes {\literal b}}
 }
-{a : \literal a , b : \literal b \vdash \texttt{g} := \inl (\cons~a~\nil , b) : ({\literal a}^{*} \otimes \literal b) \oplus \literal c}
+{a : \literal a , b : \literal b \vdash \texttt{g} := \inl ((\cons~a~\nil)
+  \otimes b) : ({\literal a}^{*} \otimes \literal b) \oplus \literal c}
 \end{mathpar}
@@
 g : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*) ('a'(*@$^*$@*) (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
-g (a , b) = inl (cons a nil , b)
+g (a , b) = inl (cons a' nil (*@$\otimes$@*) b')
 \end{lstlisting}
-\caption{(\Agda{\fillerlink}) \stringquote{ab} is parsed by
+\caption{\stringquote{ab} matches
   $({\literal a}^{*} \otimes \literal b) \oplus \literal c$}
@@
 appropriate to think of it as a parse \textit{transformer} --- a function from
-parses of $(\A \otimes \A)^{*}$ to parses of $\A^{*}$.
-
-We define $h$ by recursion on terms of type $(\A \otimes \A)^{*}$.
+parses of $(\A \otimes \A)^{*}$ to parse of $\A^{*}$.
+
+We define $\texttt{h}$ by recursion on terms of type $(\A \otimes \A)^{*}$.
 This recursion is expressed in the derivation tree by invoking the
 elimination principle for Kleene star, written
-as $\fold$. The parse transformer $h$ is more compactly presented in the pseudocode
+as $\fold$. The parse transformer $\texttt{h}$ is more compactly presented in the pseudocode
 of \cref{fig:kleeneabstractproof} by pattern matching on the input and making an
@@
 \begin{figure}
-\footnotesize
 \begin{mathpar}
@@
         {~}
-        {a_1 : \A, a_2 : \A , as : \A^* \vdash \cons~a_1 (\cons~a_2~as) : \A^*}
+        {a_1 : \A, a_2 : \A , a_{*} : \A^* \vdash \cons (a_1 , \cons(a_2 , a_{*})) : \A^*}
       }
-      {aa : \A \otimes \A , as : \A^* \vdash \letin {(a_1 , a_2)}
-        {aa} {\cons~a_1 (\cons~a_2~as)} : \A^*}
+      {a' : \A \otimes \A , a_{*} : \A^* \vdash g := \letin {a_1 \otimes a_2} {a'} {\cons (a_1 , \cons(a_2 , a_{*})) : \A^*}}
     }
-    {\cdot \vdash f := \lamblto {aa} {\lamblto {as} {\letin {(a_1 , a_2)}
-        {aa} {\cons~a_1 (\cons~a_2~as)}}} : (\A \otimes \A) \lto \A^* \lto \A^*}
+    {\cdot \vdash f := \lamblto a {\lamblto {a_{*}} {\app {\app g a} {a_{*}} }} : (\A \otimes \A) \lto \A^* \lto \A^*}
   }
-  {aas : (\A \otimes \A)^* \vdash \texttt{h} := \fold(\nil , f)(aas) : \A^* }
+  {a : (\A \otimes \A)^* \vdash \texttt{h} := \fold(\nil , f)(a) : \A^* }
 \end{mathpar}
@@
 h nil = nil
-h (cons (a1 , a2) as) = cons a1 (cons a2 (h as))
+h (cons (a1 (*@$\otimes$@*) a2) as) = cons a1 (cons a2 (h as))
 \end{lstlisting}
-\caption{(\Agda{\fillerlink}) A parse transformer for abstract grammars}
+\caption{A parse transformer for abstract grammars}
 \label{fig:kleeneabstractproof}
@@
 \begin{figure}
-  \begin{minipage}[t]{.6\textwidth}
-    \vspace{0pt}
+  \begin{tikzpicture}[node distance = 25mm ]
+    \node[state, initial] (0) {$\0$};
+    \node[state, below left of=0] (1) {$\1$};
+    \node[state, right of=1, accepting] (2) {$\2$};
+
+    \path[->] (0) edge[above] node{$\epsilon$} (1)
+              (0) edge[right] node{$\stringquote{c}$} (2)
+              (1) edge[loop left] node{$\stringquote{a}$} (1)
+              (1) edge[below] node{$\stringquote{b}$} (2);
+  \end{tikzpicture}
   \begin{lstlisting}
@@
 k (a , b) = 0to1 (1to1 a (1to2 b stop))
+
 \end{lstlisting}
-  \end{minipage}%
-  \begin{minipage}[t]{.4\textwidth}
-  \vspace{8pt}
-  \hspace{10pt}
-  \begin{tikzpicture}[node distance = 25mm ]
-    \node[state, initial] (0) {$\0$};
-    \node[state, below left of=0] (1) {$\1$};
-    \node[state, right of=1, accepting] (2) {$\2$};
-
-    \path[->] (0) edge[above] node{$\epsilon$} (1)
-              (0) edge[right] node{$\stringquote{c}$} (2)
-              (1) edge[loop left] node{$\stringquote{a}$} (1)
-              (1) edge[below] node{$\stringquote{b}$} (2);
-  \end{tikzpicture}
-  \end{minipage}%
-  \caption{(\Agda{\fillerlink}) NFA for $(\a^{*} \otimes \b) \oplus \c$ and its corresponding type}
+  \caption{NFA for $(\a^{*} \otimes \b) \oplus \c$ and its corresponding type}
   \label{fig:exampleNFA}
@@
 linear-non-linear calculus (\lnld) and Lambek calculus, also known as
-non-commutative linear logic
+non-commutative linear logic.
 \cite{krishnaswami_integrating_2015,lambek58}. As in \lnld,
@@
 The formation rules for the judgments of \theoryabbv are shown in
-\Cref{fig:contexts,fig:non-linear-types,fig:linear-formation}. $\Gamma$ stands for non-linear contexts; $X,Y,Z$
+\Cref{fig:formation}. $\Gamma$ stands for non-linear contexts; $X,Y,Z$
 stand for non-linear types; $M,N$ stand for non-linear terms, these
@@
 definitional equality judgments for both kinds of type and term
-judgments as well. Additionally, we have a judgment $\Gamma \vdash X
+judgments as well. Additionally, we have a judgment $\Gamma\vdash X
 \isSmall$ which is used in the definition of universe types.
@@
   \begin{mathpar}
-    \footnotesize
-    \boxed{\ctxwff \Gamma}
-
-    \inferrule{~}{\ctxwff \cdot}
-
-    \inferrule{\ctxwff \Gamma \and \ctxwffjdg \Gamma X}{\ctxwff {\Gamma , x : X}}
-
-    \boxed{\linctxwff \Gamma \Delta}
-
-    \inferrule{~}{\linctxwff \Gamma \cdot}
-
-    \inferrule{\linctxwff \Gamma \Delta \and \linctxwffjdg \Gamma A}{\linctxwff
-      {\Gamma} {\Delta , a : A}}
+    \inferrule{~}{\ctxwff \Gamma}
+
+    \inferrule{\ctxwff \Gamma}{\ctxwffjdg \Gamma X}
+
+    \inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash X \isSmall}
+
+    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg \Gamma Y}{\ctxwffjdg \Gamma {X \equiv Y}}
+
+    \inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash M : X}
+
+    \inferrule{\Gamma \vdash M : X \and \Gamma \vdash N : X}{\Gamma \vdash M \equiv N : X}
+
+    \inferrule{\ctxwff \Gamma}{\linctxwff \Gamma \Delta}
+
+    \inferrule{\ctxwff \Gamma}{\Gamma \vdash A \isLinTy}
+
+    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \equiv B}
+
+    \inferrule{\Gamma \vdash \Delta \isLinCtx \and \Gamma \vdash A \isLinTy}{\Gamma;\Delta\vdash e : A}
+
+    \inferrule{\Gamma; \Delta \vdash e : A \and \Gamma;\Delta \vdash f : A}{\Gamma;\Delta\vdash e \equiv f : A}
   \end{mathpar}
-  \caption{Context well-formedness rules}
-  \label{fig:contexts}
+  \caption{Formation rules}
+  \label{fig:formation}
 \end{figure}
@@
 \subsection{Non-linear Typing}
-
+The non-linear types of \theoryabbv include the standard dependent
+types for $\Pi,\Sigma$, extensional equality, natural numbers,
+booleans, unit and empty types, and we assume function
+extensionality\cite{Hofmann_1997}. We present the
+other non-linear type constructions in
+\Cref{fig:non-linear-types}. First, we include universe types $U$ of
+small non-linear types and $L$ of linear types. These are defined as
+universes ``ala Coquand'' in that we define a judgment saying when a
+non-linear type is small and define the universe to internalize
+precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. These universe
+types are needed so that we can define types by recursion on natural
+numbers. Next, we include a non-linear type $\ltonl A$ where $A$ is a
+linear type. The intuition for this type is that its elements are the
+linear terms that are ``resource free'': its introduction rule says we
+can construct an $\ltonl A$ when we have a linear term of type $A$
+with no free linear variables. This type is used extensively in our
+examples, playing a similar role to the $!$ modality of ordinary
+linear logic or the persistence modality $\square$ of separation logic
+\cite{girard_linear_1987,jung_higher-order_2016}.
+
+%% \begin{figure}
+%%   \[
+%%   \begin{array}{rrcl}
+%%     \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
+%%     \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe
+%%                                                   \mathsf{Bool} \pipe
+%%                                                   \mathsf{Empty} \pipe
+%%                                                   \mathsf{Unit} \pipe
+%%                                                   \mathsf{SPF}\,X\pipe U \pipe L
+%%                                                   \pipe \ltonl A \pipe \sum\limits_{x:X} Y
+%%                                                   \pipe \prod\limits_{x:X} Y\pipe M =_A
+%%                                                   N
+%%   \end{array}
+%%   \]
+%%   \caption{Non-linear Types}
+%% \end{figure}
 \begin{figure}
+  \begin{footnotesize}
     \begin{mathpar}
-  \footnotesize
-    \boxed{\inferrule{\ctxwff\Gamma}{\ctxwffjdg \Gamma X}}
-
-    \inferrule{~}{\ctxwffjdg \Gamma {U}}
-
-    \inferrule{~}{\ctxwffjdg \Gamma {L}}
-
-    \inferrule{~}{\ctxwffjdg \Gamma 1}
-
-    \inferrule{~}{\ctxwffjdg \Gamma \bot}
-
-    \inferrule{~}{\ctxwffjdg \Gamma {Bool}}
-
-    \inferrule{~}{\ctxwffjdg \Gamma {Nat}}
-
-    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg {\Gamma , x : X} {Y}}{\ctxwffjdg \Gamma {\PiTyLimit {x}{X}{Y}}}
-
-    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg {\Gamma , x : X} {Y}}{\ctxwffjdg \Gamma {\SigTyLimit {x}{X}{Y}}}
-
-    \\
-
-    \inferrule{\nonlinterm \Gamma M U}{\ctxwffjdg \Gamma {\unquoteTy M}}
-
-    \inferrule{\linctxwffjdg \Gamma A}{\ctxwffjdg \Gamma {\ltonl A}}
-
-    \inferrule{\ctxwffjdg \Gamma X \and \nonlinterm \Gamma M X \and \nonlinterm \Gamma N X}{\ctxwffjdg \Gamma {M =_{X} N}}
-
-    \\
-
-    \boxed{\inferrule{\ctxwffjdg\Gamma X \and \ctxwffjdg\Gamma Y}{\ctxwffjdg \Gamma {X \equiv Y}}}
-
-    \inferrule{\nonlinterm \Gamma {X \equiv Y} {U}}{\ctxwffjdg \Gamma {X \equiv Y}}
-
-    \inferrule{\linctxwffjdg \Gamma X}{\linctxwffjdg \Gamma {\unquoteTy
-        {\quoteTy X} \equiv X}}
-
-    \boxed{\inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash X \isSmall}}
-
-    \\
-    \boxed{\inferrule{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma M X}}
-
-    \inferrule{\Gamma \vdash  X \isSmall}{\nonlinterm \Gamma {\quoteTy X} U}
-
-    \inferrule{\Gamma \vdash A \isSmallLin}{\nonlinterm \Gamma {\quoteTy A} L}
-
-    \inferrule{\linterm \Gamma \cdot e A}{\nonlinterm \Gamma e \ltonl{A}}
-
-    \inferrule{\nonlinterm \Gamma {M \equiv N} {X}}
-      {\nonlinterm \Gamma {refl} {M =_{X} N}}
-
-    \\
-    \boxed{\inferrule{\nonlinterm\Gamma M X \and \nonlinterm \Gamma N X}{\nonlinterm \Gamma {M \equiv N} X}}
-
-    \inferrule{\nonlinterm \Gamma P {M =_{X} N}}{\nonlinterm\Gamma{M \equiv N} X}
+    \inferrule{}{\Gamma \vdash U \isTy}\and
+    \inferrule
+    {\Gamma \vdash M : U}
+    {\Gamma \vdash \unquoteTy M \isTy}\and
+    \inferrule
+    {\Gamma \vdash X \isTy}
+    {\Gamma \vdash \quoteTy X : U}\and
+    \unquoteTy{\quoteTy{X}} \equiv X \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}
+
+    \inferrule{}{\Gamma \vdash L \isTy}\and
+    \inferrule
+    {\Gamma \vdash M : L}
+    {\Gamma \vdash \unquoteTy M \isLinTy}\and
+    \inferrule
+    {\Gamma \vdash A \isLinTy}
+    {\Gamma \vdash \quoteTy A : L}\and
+    \unquoteTy{\quoteTy{A}} \equiv A \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}
+
+    \inferrule
+    {\Gamma \vdash A \isLinTy}
+    {\Gamma \vdash \ltonl { A } \isTy}
+    \and
+    \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash e : \ltonl A}\and
+    \inferrule{\Gamma \vdash M : \ltonl A}{\Gamma ; \cdot \vdash M : A} \and
   \end{mathpar}
-  \caption{Non-linear Formation and Typing Rules (selection)}
+  \end{footnotesize}
+  \caption{Non-linear types (selection)}
   \label{fig:non-linear-types}
@@
 
-We present a selection of the non-linear type constructions in
-\Cref{fig:non-linear-types} and provide the rest in
-\cref{fig:full-non-linear-types,fig:jdg-eq-nonlinear} of the appendix.  First,
-we include universe types $U$ of small non-linear types and $L$ of linear
-types. These are defined as
-universes ``ala Coquand'' in that we define judgments saying when
-non-linear and linear types are \emph{small} and define the universes to internalize
-precisely this judgment \cite{coquandPresheafModel,lmcs:7713}.
-The definition of smallness is simply that all types are small as long as their sub-formulae are, with the exception of the two universe types themselves.
-These universe
-types are needed so that we can define non-linear and linear types by recursion on natural
-numbers.
-Next, we include standard $\Sigma,\Pi$, empty, unit, Boolean, and natural number
-types. More complex inductive types --- such as sum types, list types, and
-$Fin~n$ --- can be defined in terms of these primitives, and we give their
-encoding in the appendix.
-
-We use an \emph{extensional} equality type $M =_{X} N$ with introduction form
-$refl$, but no elimination form. Instead we have the equality reflection rule
-which allows us to conclude a definitional equality $M =_{X} N$ from an
-arbitrary typal equality proof $P$.  The usage of an extensional equality type
-matches our implementation, which interprets both judgmental and typal equality
-as Cubical Agda's \texttt{Path} type, and so naturally supports the equality
-reflection rule. Extensional equality makes type checking of our syntax as such
-undecidable \citep{Hofmann_1997} because the conversion rule may require an arbitrarily complex
-equality proof with no explicit proof term, but in our Agda implementation we
-must provide all of these equalities manually, and so the extensionality does
-not raise any issues for our implementation. This makes
-\theoryabbv into an extensional theory, supporting function extensionality and
-the uniqueness of identity proofs. The development could be ported to an intensional type theory in the future, possibly requiring the use of setoids to handle function extensionality.
-
-Lastly, we include a non-linear type $\ltonl A$ where $A$ is a linear type. The
-intuition for this type is that its elements are the linear terms that are
-``resource free'': its introduction rule says we can construct an $\ltonl A$
-when we have a linear term of type $A$ with no free linear
-variables. Semantically, this is the type of parses of the empty string. This
-type is used extensively in our examples, playing a similar role to the $!$
-modality of ordinary linear logic or the persistence modality $\square$ of
-separation logic \cite{girard_linear_1987,jung_higher-order_2016}.
-
 \subsection{Linear Typing}
 
-\begin{figure}
-  \begin{mathpar}
-    \footnotesize
-    \boxed{\inferrule{\ctxwff\Gamma}{\linctxwffjdg \Gamma A}}
-
-    \inferrule{~}{\linctxwffjdg \Gamma I}
-
-    \inferrule{c \in \Sigma}{\linctxwffjdg \Gamma {\literal{c}}}
-
-    \inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg
-      \Gamma {A \otimes B}}
-
-    \inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg
-      \Gamma {A \lto B}}
-
-    \inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg
-      \Gamma {A \tol B}}
-
-    \inferrule{\linctxwffjdg {\Gamma, x : X} A}{\linctxwffjdg
-      \Gamma {\LinSigTyLimit{x}{X}{A}}}
-
-    \inferrule{\linctxwffjdg {\Gamma, x : X} A}{\linctxwffjdg
-      \Gamma {\LinPiTyLimit{x}{X}{A}}}
-
-    \inferrule{\nonlinterm \Gamma f {\ltonl{(A \lto B)}} \and
-      \nonlinterm \Gamma g {\ltonl{(A \lto B)}}}{\linctxwffjdg \Gamma {\equalizer {a}{f}{g}}}
-
-    \inferrule{\nonlinterm \Gamma M L}{\linctxwffjdg \Gamma {\unquoteTy M}}
-
-    \boxed{\inferrule{\linctxwffjdg \Gamma A}{\Gamma \vdash A \isSmallLin}}
-
-    \\
-    \boxed{\inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg \Gamma {A \equiv B}}}
-
-    \inferrule{\nonlinterm \Gamma {A \equiv B} L}{\linctxwffjdg
-      \Gamma {A \equiv B}}
-
-    \inferrule{\linctxwffjdg \Gamma A}{\linctxwffjdg \Gamma {\unquoteTy
-        {\quoteTy A} \equiv A}}
-  \end{mathpar}
-  \caption{Linear Type Formers, Type Equivalence}
-  \label{fig:linear-formation}
-\end{figure}
-
-We give the rules for linear type
-formation in \cref{fig:linear-formation} and the definition of linear terms in
-\cref{fig:linear-terms}. The equational theory for these
+We give an overview of the linear types and terms in
+\Cref{fig:linear-types-and-terms}. The equational theory for these
 types is straightforward $\beta\eta$ equivalence and included in the
@@
 
-Next, we cover the ``multiplicative'' connectives of non-commutative
+First, we cover the ``multiplicative'' connectives of non-commutative
 linear logic. The linear unit ($\I$) and tensor product ($\otimes$)
@@
 the introduction rule for $A \lto B$ adds a variable to the right side
-of the context, whereas the introduction rule for $B \tol A$ adds a
+of the context, whereas the introduction rule (elided) for $B \tol A$ adds a
 variable to the left side of the context. In our experience, because
@@
 over additive disjunction --- e.g., in the finitary case that $0 \& A
-\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most general form,
-
-\begin{axiom}[Distributivity]
-\label{ax:dist}
-For any $A : (x : X) \to Y(x) \to L$, the definable distribuitivity function
-%%%%%%%%%% An attempt to inline
-% \(
-%  {\LinSigTy{f}{\left(\PiTy{x}{X}{Y(x)}\right)}{\LinPiTy{x}{X}{A\,x\,(f\,x)}} \lto
-% \LinPiTy{x}{X}{\LinSigTy{y}{Y(x)}{A\,x\,y}} \right)}
-% \)
-\(
- \lamblto{e}{\letin{\sigma\,f\,e'}{e}{\withlamb{x}{e'\,.\pi\,x}}} : \LinSigTyLimit{f}{\PiTy{x}{X}{Y(x)}}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
-\LinPiTyLimit{x}{X}{\LinSigTyLimit{y}{Y(x)}{A\,x\,y}}
-\)
-has an inverse.
-\end{axiom}
-
-The following corollary is a well known consequence of distributivity \cite{Cockett_1993}, which we use in \cref{lem:unambig-to-disjoint} to prove that unambiguous binary sums have unambiguous summands.
-\begin{corollary}
-  \label{cor:binary-mono}
-  Distributivity implies that the constructors $\inl : A \to A \oplus B$,
-  $\inr : B \to A \oplus B$ of a binary sum are injective --- i.e. if
-  $\inl\,a \equiv \inl\,a'$, then $a \equiv a'$.
-\end{corollary}
-
-Our primary use of distributivity is to define an equivalence that
-expresses a linear type $A$ as a sum over which character it starts with, if any,
-\(
-A \cong \left(A \& I \right) \oplus \LinSigTy{c}{\Sigma_0}{\left( A \& \left( \literal{c} \otimes \top \right) \right)}
-\). We use this equivalence when building a parser for the traces of the lookahead
-automaton in \cref{fig:binop-inductive}.
-
-Second, we need that the different constructors of
-$\bigoplus$ are disjoint. That is, we want to enforce that $\sigma~x~a$ and
-$\sigma~x'~a'$ of type $\LinSigTy{x}{X}{A\,x}$ are not equal whenever $x \neq x'$.
-However, because these are linear terms we cannot state their disequality
-directly. Instead, we encode the disequality via a function out of an equalizer,
-
-\begin{axiom}[$\sigma$-Disjointness]
-\label{ax:disjointness}
-For any $A : X \to L$ and $x \neq x' : X$ there is a function,
-\[
-  \uparrow(\equalizer{b}{\left( \sigma\,x \circ \pi_1 \right)}{\left(  \sigma\,x'\circ \pi_2 \right)}
-  \lto 0)
-\]
-  where $b: A(x) \& A(x')$. That is, the grammar of pairs of an
+\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most
+general form, the axiom says that the definable function
+${\bigoplus\limits_{f : \prod_{x:X}
+    Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto
+  \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y}$ has an
+inverse. Second, we need that the different constructors of
+$\bigoplus$ are disjoint. We add this by adding the axiom that for any
+$A : X \to L$ and $x \neq x' : X$ there is a function
+$\uparrow(\equalizer{b}{\sigma\,x \circ \pi_1}{\sigma\,x'\circ \pi_2}
+\lto 0)$ where $b: A(x) \& A(x')$, i.e., the grammar of pairs of an
 $a:A(x)$ and an $a': A(x')$ such that $\sigma\,x\,a = \sigma\,x'\,a'$
 is empty.
-\end{axiom}
-
-We use $\sigma$-disjointness in \cref{lem:unambig-to-disjoint} to prove that
-unambiguous sums have disjoint summands.
-
-% \pedro{imo, since we are providing an Agda implementation, we can
-%   be looser with the explanation of the meaning of the disjointness axioms.
-%   As it stands, it's a bit hard do follow along. Is there a simpler example
-%   we could point to?}
+
+\pedro{imo, since we are providing an Agda implementation, we can
+  be looser with the explanation of the meaning of the disjointness axioms.
+  As it stands, it's a bit hard do follow along. Is there a simpler example
+  we could point to?}
 
 \begin{figure}
-  \footnotesize
+  \begin{footnotesize}
   \begin{mathpar}
-    \boxed{
+    \inferrule{}{\Gamma \vdash \I \isLinTy}\and
+    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \otimes B \isLinTy}\and
+    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \lto B \isLinTy}\and
+    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \tol B \isLinTy}\and
+    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigoplus\limits_{x:X} A \isLinTy}\and
+    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigamp\limits_{x:X} A \isLinTy}\and
     \inferrule
-      {\linctxwff \Gamma \Delta \and \linctxwffjdg \Gamma \A}
-      {\linterm \Gamma \Delta {e} {A}}}
-
+    {\Gamma \vdash f : \ltonl {(A \lto B)} \and \Gamma \vdash g : \ltonl { (A \lto B) }}
+    {\Gamma \vdash \equalizer {a}{f}{g} \isLinTy}
+  \end{mathpar}
+\begin{mathpar}
     \inferrule{~}{\Gamma ; a : A \vdash a : A}
-    \and
-    \inferrule{\nonlinterm \Gamma {M} {\ltonl A}}{\linterm {\Gamma} {\cdot} {M} {A}}
     \and
@@
     \and
-    \inferrule{\Gamma ; \Delta \vdash e : A \lto B \\ \Gamma ; \Delta' \vdash e' : A} {\Gamma ; \Delta, \Delta' \vdash \applto {e} {e'} : B}
-    \\
-    %
-    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
-    \and
-    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e'
-      : B \tol A}{\Gamma ; \Delta, \Delta' \vdash \apptol {e'} {e} : B}
+    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta, \Delta' \vdash \applto {e'} {e} : B}
     \\
@@
   \end{mathpar}
-  \caption{Linear terms}
-  \label{fig:linear-terms}
+  \end{footnotesize}
+  \caption{Linear types and terms (selection)}
+  \label{fig:linear-types-and-terms}
 \end{figure}
@@
     {\Gamma \vdash f : \prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)} \and \Gamma; \Delta \vdash e : \el(F\,x)(\mu\,F)}
-    {\Gamma;\Delta \vdash \fold\,F\,f\,x\,(\roll\,e) \equiv f\,x\,(\map(F\,x)\,(\fold\,F\,f)) : A\,x}
+    {\Gamma;\Delta \vdash \fold\,F\,f\,x\,(\roll e) \equiv f\,x\,(\map(F\,x)\,(\fold\,F\,f)) : A\,x}
 
@@
      \and \Gamma \vdash e : \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}\\\\
-      \Gamma,x:X;a:\el(F\,x)(\mu F) \vdash e\,x\,(\roll\,a) \equiv f\,x\,(\map(F\,x)\,e) : A\,x }
+      \Gamma,x:X;a:\el(F\,x)(\mu F) \vdash e\,x\,(\roll a) \equiv f\,x\,(\map(F\,x)\,e) : A\,x }
     {\fold\,F\,f\equiv e' : \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}}
@@
 alphabet, we add a corresponding linear type $c$. We can then define a
-non-linear type $\CharGram$ as the disjunction of all of these characters,
-and define a type $\StringGram$ as the Kleene star of $\CharGram$, i.e. as an
+non-linear type $\Char$ as the disjunction of all of these characters,
+and define a type $\String$ as the Kleene star of $\Char$, i.e. as an
 inductive linear type.
 Then we add a function $\mathsf{read} :
-\ltonl{(\top\lto \StringGram)}$ that intuitively ``reads'' the input
+\ltonl{(\top\lto \String)}$ that intuitively ``reads'' the input
 string from the input and makes it available. It is important that the
 input type of $\mathsf{read}$ is $\top$, which can control any amount
-of resources, and not $\I$ which controls no resources.
-
-\begin{axiom}
-  \label{ax:string-top}
-   $\lambda s. \mathsf{read}(!(s)) \equiv \lambda
-s. s$ where $!$ is the unique function $\ltonl(\StringGram\lto \top)$. That is, $\StringGram$ is strongly equivalent to $\top$.
-\end{axiom}
-
-If we have a string, but then throw it away and read
-it from the input, then we, in fact, recover the original string.
-This ensures that the elements of the $\StringGram$
+of resources, and not $\I$ which controls no resources. Further, we
+add an axiom that $\lambda s. \mathsf{read}(!(s)) \equiv \lambda
+s. s$ where $!$ is the unique function $\ltonl(\String\lto \top)$,
+i.e., that if we have a string, but then throw it away and read
+it from the input, then in fact that is equivalent to the string we
+were given originally. This ensures that the elements of the $\String$
 type always stand for the actual input string in our reasoning. In the
@@
 these notions of equivalence in \theoryabbv.
-
 \begin{definition}
   \label{def:weakequiv}
-  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exist
+  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exists
   parse transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B
@@
 %% definition of retract is a useful intermediate notion. In retrospect,
-%% our axiom about $\StringGram$ says that $\StringGram$ is a retract of $\top$.
+%% our axiom about $\String$ says that $\String$ is a retract of $\top$.
 
@@
 is unambiguous if there is at most one way to transform parses of any
-other grammar $\B$ into parses of $\A$. This notion of an unambiguous
+other grammar $\B$ into parse of $\A$. This notion of an unambiguous
 type is the analog for linear types of the definition of a (homotopy)
@@
 establish two useful properties of unambiguity.
-\begin{lemma}[\Agda{\fillerlink}]
-  \label{lem:retract}
+\begin{lemma}[\agdalogo]
   If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
+  If a disjunction $\bigoplus\limits_{x:X} A(x)$ is unambiguous then
+  each $A(x)$ is unambiguous.
 \end{lemma}
-\begin{lemma}[\Agda{\fillerlink}]
-  \label{lem:unambig-sum}
-  As a consequence of \cref{cor:binary-mono}, if a binary disjunction $A \oplus B$ is unambiguous then
-  $A$ and $B$ are each unambiguous.
-\end{lemma}
-From \cref{lem:retract}, we can prove that $\StringGram$ is unambiguous,
+From the first principle, we can prove that $\String$ is unambiguous,
 since it is a retract of $\top$. In fact, observe that if $\A$ is a
@@
 \lambda b. b$ follows because $\B$ is unambiguous. Therefore
-$\StringGram$ is also strongly equivalent to $\top$. Next, since $\StringGram$ is
-defined as a Kleene star, we can easily show that $\StringGram \cong \I
-\oplus \CharGram \oplus (\CharGram \otimes \CharGram \otimes \StringGram)$. Then by
-\cref{lem:unambig-sum}, we have that $\I$ and
-$\CharGram$ --- and thus each literal
-$\literal c$ --- are unambiguous as well.
+$\String$ is also strongly equivalent to $\top$. Next, since $\String$ is
+defined as a Kleene star, we can easily show that $\String \cong \I
+\oplus \Char \oplus (\Char \otimes \Char \otimes \String)$. Then by
+the second principle we have that $\I$ and $\Char$ and each literal
+$\literal c$ are unambiguous as well.
 
 % \paragraph{A Grammar of Strings}
-% Define the character grammar $\CharGram$ to be the disjunction over characters in the alphabet $\Sigma$,
-% $\CharGram = \LinSigTy {c} {\Sigma} c$. Then define a linear type of strings as a
+% Define the character grammar $\charg$ to be the disjunction over characters in the alphabet $\Sigma$,
+% $\charg = \LinSigTy {c} {\Sigma} c$. Then define a linear type of strings as a
 % linear list of characters,
-% $\StringGram = \CharGram^{*}$.
+% $\stringg = \charg^{*}$.
 
 % In the grammars model, we can build a
-% strong equivalence between $\top$ and $\StringGram hen building parsers, we may wish to enforce this
-% syntactically by including $\top \cong \StringGram$ as an axiom.
-
-% Through inclusion of this axiom, $! : \StringGram \to \top$ is an isomorphism, and
-% thus also a monomorphism. That is, $\StringGram$ is unambiguous.
+% strong equivalence between $\top$ and $\stringg$. hen building parsers, we may wish to enforce this
+% syntactically by including $\top \cong \stringg$ as an axiom.
+
+% Through inclusion of this axiom, $! : \stringg \to \top$ is an isomorphism, and
+% thus also a monomorphism. That is, $\stringg$ is unambiguous.
 % \max{we should have a section of the previous section which gives the axioms, this one and the distributivity axiom}
@@
 % We say $A$ is \emph{parseable} if there exists a grammar $B$ and a map
-% $\top \to A \oplus B$. Taking $\top \cong \StringGram$, a grammar $A$ is parseable
+% $\top \to A \oplus B$. Taking $\top \cong \stringg$, a grammar $A$ is parseable
 % exactly when we may read an arbitrary string into a parse of $A$ or a $B$-valued
@@
 $A$, a first attempt at defining a parser would be to implement a
-function $\uparrow{(\StringGram \lto A)}$. But since our linear functions
+function $\uparrow{(\String \lto A)}$. But since our linear functions
 must be total, this means that we can construct an $A$ parse for
@@
 interest. Instead we might try to write a partial function as a
-$\ltonl{(\StringGram \lto (A \oplus \top))}$ using the ``option''
+$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option''
 monad. This allows for the possibility that the input string doesn't
@@
 only in the case that a parse cannot be constructed.
-
 \begin{definition}
-  \label{def:disjoint}
-  Linear types $A$ and $B$ are \emph{disjoint} if there is a function
-  $\ltonl {\left( A \& B \lto 0 \right)}$.
-\end{definition}
-
-\begin{definition}
-  \label{def:parser}
-  A parser for a linear type $A$ is the choice a type $A_{\neg}$ disjoint from
-  $A$ and function $\uparrow{(\StringGram \lto
-    A \oplus A_{\neg})}$.
+  A parser for a linear type $A$ is a function $\uparrow{(\String \lto
+    A \oplus A_{\neg})}$, where $A_{\neg}$ is a linear type that is
+  \emph{disjoint} from $A$ in that we can implement a function
+  $\uparrow{(A \& A_{\neg} \lto 0)}$.
 \end{definition}
@@
 $\uparrow{(A \& A_{\neg} \lto 0)}$ ensures that it is impossible for
-$A$ and $A_{\neg}$ to parse the same input string. This means that in
+$A$ and $A_{\neg}$ to match the same input string. This means that in
 defining a parser, we will need to define a kind of negative grammar
-for strings that do not parse.
-
-Fortunately, we will see that
+for strings that do not parse. Fortunately, we will see that
 deterministic automata naturally support such a notion with no
@@
 traces that end in a rejecting state. This follows from the following
-principle, a consequence of \cref{ax:disjointness}.
-
-\begin{lemma}[\Agda{\fillerlink}]
-  \label{lem:unambig-to-disjoint}
-  If $\LinSigTy{x}{X}{A\,x}$ is unambiguous, then for $x \neq x'$, $A\,x$ and $A\,x'$ are
-  disjoint. In particular, if the binary product $A \oplus A_{\neg}$ is unambiguous, then
-  $A$ and $A_{\neg}$ are disjoint.
+principle, a consequence of our axiom that $\oplus$ constructors are
+disjoint.
+\begin{lemma}[\agdalogo]
+  If $A \oplus A_{\neg}$ is unambiguous, then $A$ and $A_{\neg}$ are
+  disjoint.
 \end{lemma}
 
-\begin{proof}
-If $\LinSigTy{x}{X}{A\,x}$ is unambiguous, then all functions into it from
-$A\,x \& A\,x'$ are equal. In particular,
-$\sigma\,x\,\circ\,\pi_{1} \equiv \sigma\,x'\,\circ\,\pi_{2}$ so there is a
-function \(
-\ltonl{\left(A\,x \& A\,x' \lto
-  \equalizer{b}{\left( \sigma\,x \circ \pi_1 \right)}{\left(  \sigma\,x'\circ \pi_2 \right)}
-\right)}.
-\)
-We then compose with the function in \cref{ax:disjointness} to prove that $A\,x$ and $A\,x'$ are disjoint.
-\end{proof}
-
-Writing a parser as a linear term intrinsically verifies the \emph{soundness} of
-the parser for free from the typing: any $\inl$ parse that we return \emph{must}
-correspond to a parse tree of the input string. Further, if we verify the
-disjointness property of \cref{def:parser} we then also get the
-\emph{completeness} of the parser as well, that when the parser rejects the
-input that there are no valid parses.
+Writing a parser as a linear term in this way is an intrinsic
+verification of the \emph{soundness} of the parser completely for free
+from the typing: any $\inl$ parse that we return \emph{must}
+correspond to a parse tree of the input string. Further if we verify
+the disjointness property we then also get the \emph{completeness} of
+the parser as well, that it never fails to generate an $A$ parse when
+it is possible.
 
@@
 following:
-\begin{lemma}[\Agda{\fillerlink}]
+\begin{lemma}[\agdalogo]
   \label{lem:wk-eqv-parse}
-  If $\A$ is weakly equivalent to $\B$ then any parser for $\A$ extends to a
-  parser for $\B$.
+  If $\A$ is weakly equivalent to $\B$ then any parser for $\A$ can be
+  extended to a parser for $\B$.
 \end{lemma}
 Here we need both directions of the weak equivalence. We need $A \lto
-B$ to extend the parser from $\StringGram \lto A \oplus A_{\neg}$ to
-$\StringGram \lto B \oplus A_{\neg}$ but then we also need $B \lto A$ to
+B$ to extend the parser from $\String \lto A \oplus A_{\neg}$ to
+$\String \lto B \oplus A_{\neg}$ but then we also need $B \lto A$ to
 establish that $A_{\neg}$ is disjoint from $B$.
@@
 traces of $D(A)$ and combine this using \Cref{lem:wk-eqv-parse} to get
-a verified regular expression parser.
-
-A regular expression in \theoryabbv is a linear type constructed
+a verified regex parser.
+
+\newcommand{\states}{\mathtt{states}}
+\newcommand{\labelt}{label}
+\newcommand{\transitions}{\texttt{transitions}}
+\newcommand{\epstransitions}{\epsilon\texttt{transitions}}
+\newcommand{\isAcc}{\mathtt{isAcc}}
+\newcommand{\init}{\mathtt{init}}
+\newcommand{\src}{\mathtt{src}}
+\newcommand{\dst}{\mathtt{dst}}
+\newcommand{\pparse}{\mathtt{parse}}
+\newcommand{\print}{\mathtt{print}}
+\newcommand{\epssrc}{\epsilon\mathtt{src}}
+\newcommand{\epsdst}{\epsilon\mathtt{dst}}
+
+\newcommand{\N}{\texttt{N}}
+
+A regular expressions in \theoryabbv is a linear type constructed
 using only the connectives $\literal c$, $0$, $\oplus$, $I$,
@@
 accepting state with the constructor $\nil$. Here we use an Agda-style
-Unicode syntax for $\bigamp$, as well as using the function arrow to
+unicode syntax for $\bigamp$, as well as using the function arrow to
 mean a non-dependent version of $\bigamp$. If we had a trace beginning
@@
 data Trace(*@$_{\N}$@*) : (s : N.states) (*@$\to$@*) L where
-  nil : (*@$\uparrow$@*)(&[ s : N.states ] N.isAcc s (*@$\to$@*) Trace(*@$_{\N}$@*) s)
+  nil : (*@$\uparrow$@*)(&[ s : N.states ] N.isAcc (*@$\to$@*) Trace(*@$_{\N}$@*) s)
   cons : (*@$\uparrow$@*)(&[ t : N.transitions ]
@@
 we show that for any start state $s$, ${\LinSigTy {b} {\Bool}
-  \Trace_{D}~s~b}$ is a retract of $\StringGram$ and apply \cref{lem:retract} to
-derive unambiguity. That is, first we
+  \Trace_{D}~s~b}$ is a retract of $\stringg$. That is, first we
 construct a function $\pparse_{D}$ that is a parser for
 $\Trace_D~{s}~{true}$, with $\Trace_D~{s}~{false}$ being the disjoint
-type used. The fact that these trace types are disjoint follows from the
-unambiguity of $\LinSigTy {b} {\Bool} {\Trace_{D}~s~b}$ by \cref{lem:unambig-sum}.
+type used. The fact that these types are disjoint follows by showing
+that this function is part of a retraction, i.e., that there is only
+one way to trace through a deterministic automaton.
 
@@
 
-\begin{theorem}[\Agda{\fillerlink}]
-  \label{thm:dfa-parser}
+\begin{theorem}[\agdalogo]
   $\pparse_D s$ is a parser for $\Trace_D~s~\true$.
@@
 parse(*@$_D$@*) String.nil s = (*@$\sigma$@*) (D.isAcc s) (Trace(*@$_{\D}$@*).nil s)
-parse(*@$_D$@*) (String.cons ((*@$\sigma$@*) c a) w) s = let (*@$\sigma$@*) b t = parse w (D.(*@$\delta$@*) c s) in
+parse(*@$_D$@*) (String.cons ((*@$\sigma$@*) c a) w) s =
+   let (*@$\sigma$@*) b t = parse w (D.(*@$\delta$@*) c s) in
    (*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c s b a t)
@@
 DFA is unambiguous even if the NFA is not.
-\begin{construction}[Determinization, \Agda{\fillerlink}]
-  \label{cons:determinization}
-  Given an NFA $N$, we construct a DFA $D$ such that
+\begin{theorem}[Determinization, \Agda]
+  \label{thm:determinization}
+  Given an NFA $N$, there exists a DFA $D$ such that
   $Parse_{N}$ is weakly equivalent to $Parse_{D}$.
-\end{construction}
+\end{theorem}
 \newcommand{\X}{\texttt{X}}
@@
 \begin{proof}
-  Define the states of $\D$ to be the $\mathbb{P}_{\epsilon}(\N.\states)$ --- the type
+  Define the states of $\D$ to be the $\mathbb{P}_{\varepsilon}(\N.\states)$ --- the type
   of $\epsilon$-closed\footnote{A subset of states $\X$ is $\epsilon$-closed if
@@
   grammars. To build the parse transformer
-  $\ltonl{(Parse_{N} \lto Parse_{D})}$, we strengthen our inductive hypothesis to quantify over every start state
+  $\ltonl{(Parse_{N} \lto Parse_{D})}$, we strengthen our inductive hypothesis
   and build a term
@@
 \newcommand{\DtoN}{\texttt{DtoN}}
-  \(
-    \NtoD : \ltonl {\left(\Trace_{\N}~\s~\true \lto \LinPiTy{\X}{\D.\states}{\LinPiTy{\texttt{sInX}}{\X \ni \s}{ \Trace_{\D}~\X~\true}} \right) }
-  \)
+  \[
+    \NtoD : \ltonl {\left(\Trace_{\N}~\s~\true \lto \bigwith\limits_{\X : \D.\states} \bigwith\limits_{\texttt{sInX} : \X \ni \s} \Trace_{\D}~\X~\true \right)}
+  \]
   that maps a trace in $\N$ from an arbitrary state $\s$ to a trace in $\D$
@@
 To construct a term from DFA traces to NFA traces, we similarly strengthen our
-inductive hypothesis and build a parse transformer that ranges over arbitrary $\epsilon$-closed subsets $X : \mathbb{P}_{\epsilon}{\N.\states}$,
-\(
-  \DtoN : \ltonl {\left( \Trace_{\D}~\X~\true \lto \LinSigTy{s}{\N.\states}{\LinSigTy{\texttt{sInX}}{\X \ni \s}{ \Trace_{\N}~s~\true}} \right)}
-\).
-Because the states of $D$ are $\epsilon$-closed subsets of $N.\states$, if two
-states $s$ and $s'$ belong to the the same $\epsilon$-closed subset
-$X : \mathbb{P}_{\epsilon}(N.\states)$ then there exists a $\epsilon$-path in
-$N$ between the two, but we do not necessarily know which one. Similarly, the
-data contained in the type $\Trace_{\D}~X~true$ is the \emph{existence} of an accepting trace in
-$\N$ beginning at some state in $\D$.
-
-To define $\DtoN$, we need a choice function that extracts out a trace in $N$
-from the mere existence of one. We achieve this by choosing the smallest trace
-through $\N$ subject to an ordering on the non-linear types $N.\states$,
-$N.\transitions$, and $N.\epstransitions$. In essence, the given orderings
-specify a global disambiguation strategy.
-\end{proof}
+induction hypothesis and build a parse transformer
+\[
+  \DtoN : \ltonl {\left( \Trace_{\D}~\X~\true \lto \bigoplus\limits_{s : \N.\mathsf{states}}\bigoplus\limits_{\texttt{sInX} : \X \ni \s} \Trace_{\N}~s~\true \right)}
+\]
+
 %% For a moment, let us turn our attention to the content behind the traces in $\D$.
@@
 %% weakly equivalent.
+\end{proof}
 
@@
 \newcommand{\R}{\texttt{R}}
-\begin{construction}[Thompson's Construction, \Agda{\fillerlink}]
-  \label{cons:thompson}
-  Given a regular expression $\R$, we build an NFA $\N$ such that $\R$ is
+\begin{theorem}[Thompson's Construction, \Agda]
+  Given a regular expression $\R$, there exists an NFA $\N$ such that $\R$ is
   strongly equivalent to $\Trace_{N}(\N.\init)$.
-\end{construction}
+\end{theorem}
 \begin{proof}
@@
 \cite{thompsonProgrammingTechniquesRegular1968}, showing that NFAs are, up to strong equivalence, closed under each type operation for regular expressions.
-\end{proof}
+
 %% In this paper we will walk through
@@
 
-%% The first two equalities are by definition. The third equality holds because
+%% The first two equalities are by defintion. The third equality holds because
 %% $\literal c$ is unambiguous and
@@
 %% elided details may be found in the Agda formalization.
-
-\begin{corollary}[\Agda{\fillerlink}]
-  We may build a parser for every regular expression $\R$.
-\end{corollary}
-\begin{proof}
-We combine the strong equivalence of \cref{cons:thompson} with the weak
-equivalence of \cref{cons:determinization} to show that $\R$ is weakly
-equivalent to the traces of its determinized automaton. Then, we use
-\cref{lem:wk-eqv-parse} to extend the parser from \cref{thm:dfa-parser}
-with respect to this weak equivalence.
 \end{proof}
@@
 star is replaced by an arbitrary fixed point
-operation \cite{leis_towards_1992}.
-
+operation\cite{leis_towards_1992}.
+
+\newcommand{\balanced}{\mathtt{balanced}}
+\newcommand{\Dyck}{\mathtt{Dyck}}
 A simple example of a CFG is the Dyck grammar of balanced
-parentheses, which we define in \Cref{fig:dyckinductive}.
+parentheses, which we define in \Cref{fig:dyckinductive}
 $\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$. The $\nil$ constructor shows that
@@
 
-\begin{theorem}[\Agda{\fillerlink}]
+\begin{theorem}[\Agda]
   \label{thm:dyck}
@@
 \begin{figure}
-  \footnotesize
   \begin{tikzpicture}[node distance = 17mm ]
-    \node[state, initial above, accepting] (0) {$0$};
+    \node[state, initial, accepting] (0) {$0$};
     \node[state, right of=0] (1) {$1$};
@@
     \node[right of=2] (3) {$\dots$};
-    \node[state, left of=0] (fail) {$fail$};
+    \node[state, below of=0] (fail) {$fail$};
 
@@
               (3) edge[below, bend left] node{\stringquote{)}} (2)
-              (fail) edge[loop left] node{\stringquote{(}, \stringquote{)}} (fail)
-              (0) edge[above] node{\stringquote{)}} (fail);
+              (fail) edge[loop right] node{\stringquote{(}, \stringquote{)}} (fail)
+              (0) edge[left] node{\stringquote{)}} (fail);
   \end{tikzpicture}
@@
 \stringquote{(}, \stringquote{)}, \stringquote{+} , \stringquote{NUM}
-\}$. In \Cref{fig:binop-inductive} we define the expression grammar
-using two mutually
+\}$. In \Cref{fig:binop-inductive} we define it using two mutually
 recursive types, corresponding to the two non-terminals we would use
@@
 operation is right associative. In the same figure, we define the
-traces of an automaton with one token of lookahead. The automaton\footnote{In
-  the automaton definition,
-  $\texttt{NotStartsWithLP}$ is defined as
-  \(I \oplus
-  (\literal{)} \oplus \literal{+} \oplus \mathtt{NUM}) \otimes \top
-  \). Similarly,
-  $\texttt{NotStartsWithRP}$ is defined as
-  \(I \oplus
-  (\literal{(} \oplus \literal{+} \oplus \mathtt{NUM}) \otimes \top
-  \).
-} has
+traces of an automaton with one token of lookahead. The automaton has
 four different ``states'', each with access to a natural number
-``stack''. The ``opening'' state $O$ expects either a left paren, in
+``stack''. The ``opening'' state $O$ expects either an open paren, in
 which case it increments the stack and stays in the opening state, or
@@
 state $D$ is where lookahead is used: if the next token will be a
-right paren, then we proceed to $C$; otherwise, we proceed to $A$.
-In the ``closing'' state $C$ if
-we observe a right paren, then we decrement the count and continue to the
-$D$ state. In the ``adding'' state $A$, we succeed if the string
-ends and the count is $0$; otherwise, if we see a plus we continue to the $O$
-state. Additionally, since the automaton need parse all of the
-incorrect strings, we add all of the failing cases.
-
-\begin{theorem}[\Agda{\fillerlink}]
-  \label{thm:exp-parser}
+right paren, then we proceed to $C$, otherwise we proceed to $M$. Here
+$\mathsf{NotStartsWithRP}$ is defined as
+$\I \oplus (({\literal{(}} \oplus {\literal {+}}
+\oplus \mathsf{NUM}) \otimes \top)$. In the ``closing'' state $C$ if
+we observe a close paren we decrement the count and continue to the
+$D$ state. In the ``multiplying'' state $M$, we succeed if the string
+ends and the count is $0$, and if we see a plus we continue to the $O$
+state. Additionally, since the automaton need also parse all of the
+incorrect strings, we add additional failing cases. It is
+straightforward to implement a parser for this lookahead automaton,
+generalizing the approach for deterministic automata.
+
+\begin{theorem}[\agdalogo]
   We construct a parser for $\mathsf{Exp}$ by showing it is weakly
@@
 \end{theorem}
-
-With \cref{ax:dist}, it is straightforward to implement a parser for this lookahead automaton,
-generalizing the approach for deterministic automata. Without access to
-distributivity, we may define a rudimentary lookahead operation via the chain of
-equivalences,
-\[
-  A~n~b \cong (A~n~b) \& \top \cong (A~n~b)\,\& \left(\left(\literal{)} \otimes \top \right) \oplus \texttt{NotStartsWithRP}\right)
-\]
-When defining a parser, if the lookahead character is a right paren we would
-like to apply $\texttt{lookAheadRP}$; otherwise, apply $\texttt{lookAheadNot}$.
-However, without distributivity we have no means of relating the bit of
-information learned by looking ahead to the input $A~n~b$ parse.
-That is, to choose which constructor to apply we would need the end of this
-chain of equivalences to be a binary sum rather than a binary product,
-necessitating the distributivity axiom.
-
 \begin{figure}
@@
   left : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O (n + 1) b (*@$\lto$@*) O n b)
-  num : (*@$\uparrow$@*)(&[ n ] &[ b ] NUM (*@$\lto$@*) D n b (*@$\lto$@*) O n b)
-  unexpected : (*@$\uparrow$@*)(&[ n ] (NotStartsWithLP (*@$\lto$@*) O n false)
+  num : (*@$\uparrow$@*)(&[ n ] &[ b ] NUM (*@$\lto$@*) DO n b (*@$\lto$@*) O n b)
+  done : (*@$\uparrow$@*)(&[ n ] O n false)
+  unexpected : (*@$\uparrow$@*)(&[ n ] (')' (*@$\oplus$@*) '+') (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) O n false)
 data D : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
   lookAheadRP  : (*@$\uparrow$@*)(&[ n ] &[ b ] ((')' (*@$\otimes$@*) (*@$\top$@*)) (*@$\&$@*) C n b) (*@$\lto$@*) D n b)
-  lookAheadNot : (*@$\uparrow$@*)(&[ n ] &[ b ] (NotStartsWithRP (*@$\&$@*) A n b) (*@$\lto$@*) D n b)
+  lookAheadNot  : (*@$\uparrow$@*)(&[ n ] &[ b ] (NotStartsWithRP (*@$\&$@*) M n b) (*@$\lto$@*) D n b)
 data C : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
   closeGood : (*@$\uparrow$@*)(&[ n ] &[ b ] ')' (*@$\lto$@*) D n b (*@$\lto$@*) C (n + 1) b)
-  closeBad : (*@$\uparrow$@*)(')' (*@$\lto$@*) C 0 false)
-  unexpected : (*@$\uparrow$@*)(&[ n ] NotStartsWithRP (*@$\lto$@*) C n false)
-data A : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
-  doneGood : (*@$\uparrow$@*)(A 0 true)
-  doneBad : (*@$\uparrow$@*)(&[ n ] A (n + 1) false)
-  add : (*@$\uparrow$@*)(&[ n ] &[ b ] '+' (*@$\lto$@*) O n b (*@$\lto$@*) A n b)
-  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')'(*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) A n false)
+  closeBad : (*@$\uparrow$@*)(&[ n ] ')' (*@$\lto$@*) C 0 b)
+  done : (*@$\uparrow$@*)(&[ n ] C n false)
+  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) '+' (*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) C n false)
+data M : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
+  doneGood : (*@$\uparrow$@*)(M 0 true)
+  doneBad : (*@$\uparrow$@*)(&[ n ] M (n + 1) false)
+  add : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O n b (*@$\lto$@*) M n b)
+  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')'(*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) M n false)
 \end{lstlisting}
@@
 \subsection{Unrestricted Grammars}
+\newcommand{\stringNL}{String'}
 
@@
 demonstrate this, we show that for any \emph{non-linear} function $P :
-\StringSem \to U$, where here $\StringSem$ is the \emph{non-linear} type
+\stringNL \to U$, where here $\stringNL$ is the \emph{non-linear} type
 of strings over the alphabet, we can construct a grammar whose parses
 correspond to $P$.
-%% Define $\StringSem$ to be the non-linear list type over the alphabet $\Sigma$ and
+%% Define $\stringNL$ to be the non-linear list type over the alphabet $\Sigma$ and
 %% let  be a mapping from the set of strings to the universe of
 %% nonlinear types. The function $P$ describes a family of nonlinear types
-%% dependent on $\StringSem$.
+%% dependent on $\stringNL$.
 %% We view the image of $P$ as a set of parse trees for each input string, and thus
@@
 %% a first-class linear type in \theoryabbv.
-\(
-Reify~P = \LinSigTy{w} {\StringSem} {\LinSigTy {x} {P~w} {~\lceil w \rceil}}
-\)
+\[
+Reify~P = \LinSigTyLimit {w} {\stringNL} {\LinSigTyLimit {x} {P~w} {~\lceil w \rceil}}
+\]
 %% \steven{Probably put this ceiling business around the discussion of the linear
@@
 
-This reification operation on functions $\StringSem \to U$ is incredibly expressive,
+This reification operation on functions $\stringNL \to U$ is incredibly expressive,
 as it allows to sidestep our linear typing connectives and utilize the whole of nonlinear
 dependent type theory to define a grammar. For example, given a
-Turing machine $T$ one may
-define a non-linear predicate $accepts : \StringSem \to U$ that encodes that $T$
-halts and accepts an input string. Then, $Reify~accepts$ is a linear type that
-captures precisely the string the same language as $T$. That is,
+Turing machine $Tur$ one may
+define a non-linear predicate $accepts : \stringNL \to U$ such that $accept~w$ is equal to $\top$ if
+$T$ accepts $w$ and equal to $0$ otherwise. Then, $Reify~accepts$ is a linear
+type that captures precisely the strings accepted by $Tur$. That is,
 $Lang(Reify~accepts)$ is recursively enumerable --- the most general class of
 languages in the Chomsky hierarchy.
-\begin{construction}[\Agda{\fillerlink}]
-  \label{cons:turing}
-  For any Turing machine $T$, we can construct a grammar in \theoryabbv
-  that accepts the same language as $T$.
-\end{construction}
+\begin{theorem}[\Agda]
+  For any Turing machine, we can construct a grammar in \theoryabbv
+  that accepts the same language as the Turing machine.
+\end{theorem}
 
@@
 formulation that is closer in spirit to the standard definition of a
-formal \emph{language} \cite{elliottSymbolicAutomaticDifferentiation2021}:
+formal \emph{language}\cite{elliottSymbolicAutomaticDifferentiation2021}:
 \begin{definition}
@@
     $\gamma \in \sem\Gamma$, we define a set $\sem X \gamma$.
-  \item For each linear type $\Gamma \vdash A \isLinTy$ and element $\gamma
+  \item For each linear type $\Gamma A \isLinTy$ and element $\gamma
     \in \sem\Gamma$, we define a formal grammar $\sem{A}\gamma$. We
@@
 equivalent to the set of parse transformers:
-\( \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\).
-
+\[ \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\]
 Next, a parse in the equalizer $\equalizer{a}{f}{g}$ is defined as
@@
 \begin{figure}
-  \begin{minipage}[t]{.5\textwidth}
   \begin{footnotesize}
-    \begin{flalign*}
-    &\sem{c}\gamma\,w = \{ c | w = c\}\\
-    &\sem{I}\gamma\,w = \{ () \pipe w = \varepsilon \}\\
-    &\sem{A \otimes B}\gamma\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
-    &\sem{A \lto B}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
-    &\sem{B \tol A}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
-    &\sem{\bigoplus_{x:X} A}\gamma\,w = \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}
-  \end{flalign*}
+    \begin{align*}
+    \sem{c}\gamma\,w &= \{ c | w = c\}\\
+    \sem{I}\gamma\,w &= \{ () \pipe w = \varepsilon \}\\
+    \sem{A \otimes B}\gamma\,w &= \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
+    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
+    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
+    \sem{\bigoplus_{x:X} A}\gamma\,w &= \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
+    \sem{\bigwith_{x:X} A}\gamma\,w &= \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w\\
+    \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
+    \sem{\equalizer{a}{f}{g}}\gamma\,w &= \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
+    \sem{L}\gamma &= \Grammar_0\\
+    \sem{\SPF\,X}\gamma &= \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
+    \sem{\el(F)}\gamma\,G &= \sem{F}\gamma\,G\\
+    \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
+    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma)\\
+  \end{align*}
   \end{footnotesize}
-  \end{minipage}%
-  \begin{minipage}[t]{.5\textwidth}
-  \begin{footnotesize}
-    \begin{flalign*}
-    &\sem{\bigwith_{x:X} A}\gamma\,w = \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w \\
-    &\sem{\uparrow A}\gamma = \sem{A}\gamma\,\varepsilon\\
-    &\sem{\equalizer{a}{f}{g}}\gamma\,w = \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
-    &\sem{L}\gamma = \Grammar_0\\
-    &\sem{SPF\,X}\gamma = \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
-    &\sem{\el(F)}\gamma\,G = \sem{F}\gamma\,G\\
-    &\sem{\map(F)}\gamma\,f = \sem{F}\gamma\,f\\
-    &\sem{\mu A}\gamma = \mu (\sem{A}\gamma)
-  \end{flalign*}
-  \end{footnotesize}
-  \end{minipage}
   \caption{Grammar Semantics}
@@
 \end{figure}
+
 The most complex part of the semantics is the interpretation of
@@
 constructed in our Agda implementation as an inductive type of $IW$
-trees which are already available in the Cubical library of Agda
+trees which are already available in the cubical library of Agda
 \cite{The_Agda_Community_Cubical_Agda_Library_2024}.  Then an element $F \in \sem{X \to
@@
 The remaining details of the interpretation of linear terms as parse
-transformers and verification of the equational axioms --- as well as
-\cref{ax:dist,ax:disjointness,ax:string-top} --- is a relatively
+transformers and verification of the equational axioms is a relatively
 straightforward extension of existing semantics of linear logic in
@@
 
-\subsection{Agda Implementation \Agda{\fillerlink}}
-\label{subsec:impl}
+\subsection{Agda Implementation}
+
 This denotational semantics in grammars and parse transformers serves
 as the basis for our prototype implementation of \theoryabbv in
-Cubical Agda. The implementation is a shallow embedding, meaning that
-rather than formalizing a syntax of \theoryabbv types and terms, we interpret
-the non-linear universe $U$ as Cubical Agda's universe of (homotopy) sets (at
-some universe level $\ell$), $\hSet~\ell$, and the linear universe $L$ as the
-function type
-$\StringSem \to \hSet~\ell$.
-We use $\hSet~\ell$ as it ensures that uniqueness of identity proofs holds for the interpretation of all types in \theoryabbv, as it does in any extensional type theory.
-Then we implement each of the type and
+cubical Agda. The implementation is a shallow embedding, meaning that
+rather than formalizing a syntax of \theoryabbv types and terms, we
+work directly with Agda types and a definition of a formal grammar as
+a function $\String \to \Set$. Then we implement each of the type and
 term constructors of \theoryabbv as combinators on formal grammars or
-parse transformers, and use Cubical Agda's equality type to model the
+parse transformers, and use cubical Agda's equality type to model the
 term equalities. Cubical Agda is convenient for this purpose as it has
-built-in support for function extensionality which is convenient for the verification of equality rules. Axioms such as distributivity of $\oplus/\&$ and
+built-in support for function extensionality which we use
+extensively. Axioms such as distributivity of $\oplus/\&$ and
 disjointness of constructors are then provable directly in Agda, and
 we are careful to only construct grammars, terms and proofs using
-constructs that are possible in \theoryabbv.
-
-The main difference
+constructs that are possible in \theoryabbv. The main difference
 between our shallow embedding and \theoryabbv is that our linear terms
 are written in a combinator-style, without being able to use named
-variables in linear terms. For example, the function $h$ from
-\cref{fig:kleeneabstractproof} would be written with the fold combinator applied
-to subexpressions for the $nil$ and $cons$ cases,
-\(
-h = fold~nil~(cons \circ id \otimes cons \circ assoc^{-1})
-\). Notice in the $cons$ case of the fold, we manually reassociate
-$(\literal{a} \otimes \literal{a}) \otimes \literal{a}^{*}$ to
-$\literal{a} \otimes (\literal{a} \otimes \literal{a}^{*})$, then act on the
-left by the identity and on the right by $cons$ to get a parse
-$\literal{a} \otimes \literal{a}^{*}$, and finally we act again by $cons$ to
-finally produce a parse of $\literal{a}^{*}$.
-A benefit of this shallow embedding is that
+variables in linear terms. A benefit of this shallow embedding is that
 the parsers are immediately available to a larger Agda development, as
-they are just normal Agda code. Although, even in the small example from
-\cref{fig:kleeneabstractproof} we must manage additional complexity --- such as
-manually reassociations --- and
-this only grows more complex in larger programs. In future work we will look to
+they are just normal Agda code. In future work we will look to
 implement a type checker for a syntax closer to the presentation in
-this paper, while maintaining the goal of integration with an
+this paper, but with the goal of still making it integrate with an
 existing proof assistant.
@@
 
-\section{Related and Future Work}
+\section{Discussion and Future Work}
 \label{sec:discussion}
-
-\subsection{Related Work}
 \paragraph{Grammars as (Linear) Types}
@@
 The most similar prior work to our own is Luo's Lambek calculus with
-dependent types \cite{luo}. They present two systems: one like ours where linear
-types may only depend on non-linear ones, and another that
-allows linear types to depend on other linear types and supports
+dependent types \cite{luo}. Their design differs from our own in
+allowing linear types to depend on other linear types and supporting
 directed $\Pi$ and $\Sigma$ types that have no analog in our
@@
 %% \subsubsection{Typed Grammars}
-Frisch and Cardelli introduced the use of a simple type system to reason about
-regular expressions up to weak equivalence \cite{frischCardelli}.
-Henglein and Nielsen build on this type-theoretic view of grammars, pointing out that the values of the types correspond precisely to the parse trees \cite{henglein_regular_2011}.
-Our design of \theoryabbv extends this view of grammars as types to a much broader class of grammars, as well as providing a syntax and equational theory for \emph{parse transformers}, showing that the terms can be viewed not just as parse trees, but as intrinsically verified parsers.
-
-\citet{elliottSymbolicAutomaticDifferentiation2021} uses essentially the same denotational semantics as ours, interpreting regular expressions as type-valued functions on strings.
-Our denotational semantics of \theoryabbv extends this to a much broader class of grammars.
+The usage of a simple type system to reason about regular expressions up to weak
+equivalence was
+introduced by Frisch and Cardelli \cite{frischCardelli}. Henglein and Nielsen
+\cite{henglein_regular_2011} later adapt this approach by providing a type
+system for regular expressions whose semantics were in one-to-one correspondence
+with the set of parse trees.
+Elliott \cite{elliottSymbolicAutomaticDifferentiation2021} gives a similar
+semantic interpretation of regular grammars as type-level predicates on strings.
+The type system of \theoryabbv extends these semantic interpretations of to a broader class of grammars and also gives a formal syntax and equational theory for the parse transformers that these prior works lack.
+
 %% Krishnaswami and Yallop \cite{krishnaswami_typed_2019} provide a type system for
@@
 %% algebra to reason about parsing as well.
-\paragraph{Dependent Linear Types}
-Our syntax for non-commutative dependent linear type theory is based on the
-dependent commutative linear type theory of
-\citet{krishnaswami_integrating_2015}, itself an extension of Benton's
-linear-non-linear calculus \cite{benton1994}. A distinct feature of these
-systems is that linear types and non-linear types are distinct sorts, so the
-linear logic $!$ operator is not a primitive, but is instead definable, for
-instance in \theoryabbv as $!A = \LinSigTy{\_}{\ltonl A}{I}$.  V\'ak\'ar
-develops a dependent linear type theory where instead the non-linear types are
-accessed using the $!$ modality, similar to Girard's original approach
-\cite{vakar2015,girard_linear_1987}. Additionally, V\'ak\'ar develops a general
-categorical semantics for their system, whereas we have only developed a single
-denotational model. It appears straightforward to adapt V\'ak\'ar's general
-semantics approach to a non-commutative variant that would apply to
-\theoryabbv. This may have applications in finding alternative models, or
-developing logical relations proofs categorically.
 
@@
 
-\subsection{Future Work}
-In this work we have demonstrated feasibility of \theoryabbv for the
-verification of formal grammar theory and sound-by-construction parsers. In
-future work, we aim to extend \theoryabbv to be a practical tool for developing
-verified parsing components of larger verified software systems.
-
-\paragraph{Verified Parser Generators}
-We aim to extend our work on parsing to verify other algorithms. For instance,
-our regular expression parser makes somewhat arbitrary choices in what parse to
-produce when the grammar is ambiguous. In the future we aim to verify Frisch and
-Cardelli's \emph{greedy} algorithm for regular expression
-disambiguation. Defining this algorithm in \theoryabbv would provide soundness
-by construction, but it is not obvious if the greediness property could be
-verified easily as well.
-
-We also aim to adapt the approach used for our context-free grammar parsers for
-$\LLL$ and $\LRR$/$\LALRR$ parsers, with the aim of developing a shared library of intrinsically
-verified parsing utilities that can be usefully shared across different
-parsers. We would also like to investigate how high-performance verified parsers
-can be implemented in \theoryabbv. This might be done by developing an efficient
-compiler for \theoryabbv directly or by developing a verified compiler to an
-imperative system within \theoryabbv itself.
-
 \paragraph{Semantic Actions}
 
-Our verification has mainly focused on the verification that a parser outputs a
-correct concrete syntax tree for a grammar. However, in practice, parsers are
-combined with a \emph{semantic action} that emits an \emph{abstract} syntax tree
-that omits superfluous syntactic details that are unneeded in later stages of
-the overall program. We can define a semantic action in \theoryabbv for a linear
-type $A$ with semantic outputs in a non-linear type $X$ to be a function $\ltonl
-{(A \lto \LinSigTy{\_}{X}{\top})}$. That is, a semantic action is a function that
-produces a semantic element of $X$ from the concrete parses of $A$. In future
-work, we aim to study the question of verifying efficient implementations of
-parsers with semantic actions, and integrating them into larger verified
-systems.
+Our verification has mainly focused on the verification that a parser
+outputs a correct concrete syntax tree for a grammar. However, in
+practice, parsers are combined with a \emph{semantic action} that
+emits an \emph{abstract} syntax tree that omits superfluous syntactic
+details that aren't needed in later stages of the overall program. We
+can define a semantic action in \theoryabbv for a linear type $A$ with
+semantic outputs in a non-linear type $X$ to be a function $\ltonl {(A
+  \lto \bigoplus_{\_:X} \top)}$. That is, a semantic action is a
+function that produces a semantic element of $X$ from the concrete
+parses of $A$. In future work, we aim to study the question of
+verifying efficient implementations of parsers with semantic actions.
 
@@
 
-\newpage
-
-\begin{acks}
-  This material is based upon work supported by the
-  \grantsponsor{GS100000001}{National Science
-    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-  conclusions or recommendations expressed in this material are those
-  of the author and do not necessarily reflect the views of the
-  National Science Foundation.
-\end{acks}
-
+\bibliographystyle{plain}
 \bibliography{refs.bib}
 
-\newpage
+\pagebreak
+
 \ifappendix
 \appendix
+
 \section{Syntax}
+
 \begin{figure}
-\begin{minipage}[t]{.35\textwidth}
   \begin{align*}
-    \el\left(\Var\,M\right) B &= B\,M\\
-    \el\left(\mathsf{K}\,A\right) B &= A\\
-    \el\left(\bigoplus A\right) B &= \bigoplus_{y:Y}\el(A y)B\\
-    \el\left(\bigamp A\right) B &= \bigamp_{y:Y}\el(A y)B\\
-    \el\left(A \otimes A'\right) B &= \el(A)B \otimes \el(A')B
+    \el(\Var\,M) B &= B\,M\\
+    \el(\mathsf{K}\,A) B &= A\\
+    \el(\bigoplus A) B &= \bigoplus_{y:Y}\el(A y)B\\
+    \el(\bigamp A) B &= \bigamp_{y:Y}\el(A y)B\\
+    \el(A \otimes A') B &= \el(A)B \otimes \el(A')B\\\\
+    \map(\Var\,M)\,f &= f\,M\\
+    \map(\K\,A)\,f &= \lambda a. a\\
+    \map(\bigoplus A)\,f &= \lambda a. \letin{\oplusinj y {a_y}}{a}{\oplusinj y {\map(A\,y)\,f\,a_y}}\\
+    \map(\bigwith\,A)\,f &= \lambda a. \withlamb{y}{\map(A\,y)\,f\,(\withprj y a)}\\
+    \map(A \otimes A')\,f &= \lambda b. \letin{(a,a')}{b}{(\map(A)\,f\,a,\map(A')\,f\,a')}
   \end{align*}
-\end{minipage}%
-\begin{minipage}[t]{.65\textwidth}
-\begin{align*}
-    \map\left(\Var\,M\right)\,f &= f\,M\\
-    \map\left(\K\,A\right)\,f &= \lambda a. a\\
-    \map\left(\bigoplus A\right)\,f &= \lambda a. \letin{\oplusinj y {a_y}}{a}{\oplusinj y {\map(A\,y)\,f\,a_y}}\\
-    \map\left(\bigwith\,A\right)\,f &= \lambda a. \withlamb{y}{\map(A\,y)\,f\,(\withprj y a)}\\
-    \map\left(A \otimes A'\right)\,f &= \lambda b. \letin{(a,a')}{b}{(\map(A)\,f\,a,\map(A')\,f\,a')}
-  \end{align*}
-\end{minipage}%
   \caption{Strictly positive functors functorial actions}
-  \label{fig:spf-act}
 \end{figure}
@@
 
-In \cref{fig:spf-act}, we provide the functorial actions for the strictly
-positive functors defined in \cref{fig:iilt}. We also include the full set of
-rules for non-linear types in \cref{fig:full-non-linear-types}, and we provide
-their judgmental equalities in \cref{fig:jdg-eq-nonlinear}. Finally,
-\cref{fig:jdgeq} contains the judgmental equalities between linear terms in
-\theoryabbv.
-
-\subsection{Non-linear Types}
-We define sum types, list types, and $Fin~n$ from primitve non-linear types.
-
-For $X , Y : U$, define the sum type $X + Y$ as,
-
-\[
-X + Y = \SigTyLimit{b}{Bool}{elim_{Bool}(U, X, Y)(b)}
-\]
-
-For $n : Nat$, define $Fin~n$ as,
-
-\begin{align*}
-  Fin~0 &= \bot \\
-  Fin~(suc~n) &= 1 + (Fin~n)
-\end{align*}
-
-For $X : U$, define $List~X$ as,
-
-\[
-  List~X = \SigTyLimit{n}{Nat}{\PiTyLimit{k}{Fin~n}{X}}
-  \]
-
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    \inferrule{~}{\Gamma \vdash 1 \isSmall}
-
-    \inferrule{~}{\Gamma \vdash Bool \isSmall}
-
-    \inferrule{~}{\Gamma \vdash \bot \isSmall}
-
-    \inferrule{~}{\Gamma \vdash Nat \isSmall}
-
-    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash Y \isSmall}{\Gamma \vdash \PiTyLimit {x}{X}{Y} \isSmall}
-
-    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash Y \isSmall}{\Gamma \vdash \SigTyLimit {x}{X}{Y} \isSmall}
-
-    \inferrule{\Gamma \vdash M : U}{\Gamma \vdash \unquoteTy M \isSmall}
-
-    \inferrule{\Gamma \vdash A \isSmallLin}{\Gamma \vdash {\ltonl A} \isSmall}
-
-    \inferrule{\Gamma \vdash X \isSmall \and \nonlinterm \Gamma M X \and
-      \nonlinterm \Gamma N X}{\Gamma \vdash {M =_{X} N} \isSmall}
-  \end{mathpar}
-  \caption{Small Non-linear Types}
-\end{figure}
-
-\begin{figure}
-  \footnotesize
-  \begin{mathpar}
-    \inferrule{~}{\Gamma \vdash I \isSmallLin}
-
-    \inferrule{c \in \Sigma}{\Gamma \vdash {\literal{c}} \isSmallLin}
-
-    \inferrule{\Gamma \vdash A \isSmallLin \and \Gamma \vdash B \isSmallLin}{
-      \Gamma \vdash {A \otimes B} \isSmallLin}
-
-    \inferrule{\Gamma \vdash A \isSmallLin \and \Gamma \vdash B \isSmallLin}{
-      \Gamma \vdash {A \lto B} \isSmallLin}
-
-    \inferrule{\Gamma \vdash A \isSmallLin \and \Gamma \vdash B \isSmallLin}{
-      \Gamma \vdash {A \tol B} \isSmallLin}
-
-    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash A \isSmallLin}{\Gamma \vdash \LinSigTyLimit {x}{X}{A} \isSmallLin}
-
-    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash A \isSmallLin}{\Gamma \vdash \LinPiTyLimit {x}{X}{A} \isSmallLin}
-
-    \inferrule{\Gamma \vdash A \isSmallLin \and
-      \Gamma \vdash B \isSmallLin \and
-      \nonlinterm \Gamma f {\ltonl{(A \lto B)}} \and
-      \nonlinterm \Gamma g {\ltonl{(A \lto B)}}}{\Gamma \vdash {\equalizer
-        {a}{f}{g}} \isSmallLin}
-
-    \inferrule{\Gamma \vdash M : L}{\Gamma \vdash \unquoteTy M \isSmallLin}
-  \end{mathpar}
-  \caption{Small Linear Types}
-\end{figure}
-
 \begin{figure}
   \begin{mathpar}
-  \footnotesize
-    \boxed{\inferrule{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma M X}}
-
-    \inferrule{~}{\nonlinterm {\Gamma , x : X , \Gamma'} {x} {X}}
-
-    \inferrule{\nonlinterm \Gamma M Y \and \ctxwffjdg \Gamma {X \equiv Y}}{\nonlinterm {\Gamma} {M} {X}}
-
-    \inferrule{~}{\nonlinterm {\Gamma} {tt} {1}}
-
-    \inferrule
-    {\nonlinterm \Gamma M \bot \and \ctxwffjdg \Gamma X}
-    {\nonlinterm \Gamma {elim_{\bot}(X, M)} {X}}
-
-    \inferrule{~}{\nonlinterm \Gamma {true} {Bool}}
-
-    \inferrule{~}{\nonlinterm \Gamma {false} {Bool}}
-
-    \inferrule
-    {
-      \ctxwffjdg {\Gamma , b : Bool} {X(b)} \and
-      \nonlinterm \Gamma {M_{0}} {X(false)} \and
-      \nonlinterm \Gamma {M_{1}} {X(true)}
-    }
-    {\nonlinterm
-        {\Gamma, b : Bool}
-        {elim_{Bool}(X, M_0, M_1)(b)}
-        {X(b)}
-    }
-
-    \\
-
-    \inferrule{~}{\nonlinterm {\Gamma} {0} {Nat}}
-
-    \inferrule{\nonlinterm {\Gamma} {n} {Nat}}{\nonlinterm {\Gamma} {suc~n} {Nat}}
-
-    \inferrule{\ctxwffjdg {\Gamma , n : Nat} {X(n)} \and
-           \nonlinterm {\Gamma} {M_0} {X(0)} \and
-           \nonlinterm {\Gamma, n : Nat, x : X(n)} {M_{suc}(n , x)} {X(suc~n)}}
-         {\nonlinterm
-           {\Gamma, n : Nat}
-           {elim_{Nat}(X,M_0,M_{suc})(n)}
-           {X(n)}}
-
-    \\
-
-    \inferrule{\nonlinterm \Gamma M X \and \nonlinterm \Gamma {N} {\subst Y {M}
-        {x}}}{\nonlinterm {\Gamma} {(M , N)} {\SigTyLimit {x} {X} {Y}}}
-
-    \inferrule{\nonlinterm {\Gamma} {M} {\SigTyLimit {x} {X} {Y}}}
-              {\nonlinterm {\Gamma} {M.fst} X}
-
-    \inferrule{\nonlinterm {\Gamma} {M} {\SigTyLimit {x} {X} {Y}}}
-              {\nonlinterm {\Gamma} {M.snd} {\subst Y {M.fst} {x}}}
-
-    \inferrule{\ctxwffjdg \Gamma {\PiTyLimit {x} {X} {Y}} \and \nonlinterm
-      {\Gamma , x : X} {M} {Y}}
-              {\nonlinterm {\Gamma} {\lamb {x} {M}} {\PiTyLimit {x} {X} {Y}}}
-
-    \inferrule{\nonlinterm {\Gamma} {M} {\PiTyLimit {x} {X} {Y}} \and
-      \nonlinterm \Gamma N X}
-              {\nonlinterm \Gamma {M~N} {\subst Y {N} {x}}}
-
-    \inferrule{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma {\quoteTy X} U}
-
-    \inferrule{\linctxwffjdg \Gamma A}{\nonlinterm \Gamma {\quoteTy A} L}
-
-    \inferrule{\linterm \Gamma \cdot e A}{\nonlinterm \Gamma e \ltonl{A}}
-
+    \inferrule{~}{\ctxwff \cdot}
+    \and
+    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}
+    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}
+
+    \boxed{\linctxwffjdg \Gamma A}
+
+    \inferrule{~}{\linctxwff \Gamma \cdot}
+    \and
+    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}
   \end{mathpar}
-  \caption{Non-linear Typing}
-  \label{fig:full-non-linear-types}
+  \caption{Context rules}
 \end{figure}
 
+%% \begin{figure}
+%%   \begin{mathpar}
+%%   \boxed{\Gamma \vdash M : X}
+
+%%   \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{\Gamma \vdash M : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash M : X}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{~}{\Gamma \vdash () : 1}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{\Gamma \vdash M : X \\ \Gamma \vdash N : \subst Y M x}{\Gamma \vdash (M, N) : \SigTy x X Y}
+%%   %
+%%   \and
+%% %
+%%   \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_1\, M : X}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{\Gamma \vdash M : \SigTy x X Y}{\Gamma \vdash \pi_2\, M : \subst Y {\pi_1\, M} x}
+%%   \and
+%%   \inferrule{\Gamma, x : X \vdash M : Y}{\Gamma \vdash \lamb x M : \PiTy x X Y}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{\Gamma \vdash M : \PiTy x X Y \\ \Gamma \vdash N : X}{\Gamma \vdash \app M {N} : \subst Y {N} x}
+%%   %
+%%   \and
+%%   %
+%%   \inferrule{\Gamma \vdash M \equiv N : X}{\Gamma \vdash \mathsf{refl} : M =_X N}
+%%   \and
+%%   \textrm{TODO: inductive types (boolean, nat) with their eliminators allowed in }
+%%   \end{mathpar}
+%%   \caption{Intuitionistic typing}
+%%   \label{fig:inttyping}
+%% \end{figure}
+
 \begin{figure}
-  \footnotesize
   \begin{mathpar}
-    \boxed{
-      \inferrule
-      {\nonlinterm \Gamma M X \and
-       \nonlinterm \Gamma N X}
-      {\nonlinterm \Gamma {M \equiv N} {X}}}
-
-  \inferrule{\nonlinterm \Gamma M 1 \and \nonlinterm \Gamma N 1}{\nonlineq \Gamma M N 1}
-
-  % \inferrule{\nonlinterm \Gamma M \bot \and \nonlinterm \Gamma N \bot}{\nonlineq \Gamma M N \bot}
-
-  \inferrule{\nonlinterm {\Gamma , x : X} {M} {Y} \and \nonlinterm {\Gamma} {N} {X}}{\nonlineq \Gamma {\left(\lamb {x} {M} \right)~N} {\subst M {N}
-      {x}} {X}}
-
-  \inferrule{\nonlinterm \Gamma M {\PiTyLimit {x}{X}{Y}}}{\nonlineq \Gamma {M} {\lamb {x} {M~x}} {\PiTyLimit{x}{X}{Y}}}
-
-  \inferrule{\nonlinterm \Gamma M {X} \and \nonlinterm {\Gamma , x : X} {N}
-    {\subst Y {M} {x}}}{\nonlineq \Gamma {(M , N).fst} {M} {X}}
-
-  \inferrule{\nonlinterm \Gamma M {X} \and \nonlinterm {\Gamma , x : X} {N}
-    {\subst Y {M} {x}}}{\nonlineq \Gamma {(M , N).snd} {N} {\subst Y {M} {x}}}
-
-  \inferrule{\nonlinterm \Gamma {M} {\SigTy {x} {X} {Y}}}{\nonlineq \Gamma {(M.fst , M.snd)} {M} {\SigTyLimit{x}{X}{Y}}}
-
-  \inferrule
-  {
-    \nonlinterm \Gamma {M_{0}} {X(false)} \and
-    \nonlinterm \Gamma {M_{1}} {X(true)}
-  }
-  {\nonlineq
-      {\Gamma}
-      {elim_{Bool}(X, M_0,M_1)(false)}
-      {M_0}
-      {X(b)}
-  }
-
-  \inferrule
-  {
-    \nonlinterm \Gamma {M_{0}} {X(false)} \and
-    \nonlinterm \Gamma {M_{1}} {X(true)}
-  }
-  {\nonlineq
-      {\Gamma}
-      {elim_{Bool}(X, M_0,M_1)(true)}
-      {M_1}
-      {X(b)}
-  }
-
-  \inferrule{\nonlinterm {\Gamma} {M_0} {X(0)} \and
-      \nonlinterm {\Gamma, n : Nat, x : X(n)} {M_{suc}(n , x)} {X(suc~n)}}
-      {\nonlineq {\Gamma} {elim_{Nat}(X,M_0,M_{suc})(0)} {M_0} {X(0)}}
-
-  \inferrule{\nonlinterm {\Gamma} {M_0} {X(0)} \and
-      \nonlinterm {\Gamma, n : Nat, x : X(n)} {M_{suc}(n , x)} {X(suc~n)}
-    }
-      {\nonlineq {\Gamma, n : Nat} {elim_{Nat}(X,M_0,M_{suc})(suc~n)} {M_{suc}(n ,
-          elim_{Nat}(X,M_0,M_{suc})(n))} {X(suc~n)}}
-
-  \inferrule{\nonlinterm \Gamma M U}{\nonlineq \Gamma {\quoteTy {\unquoteTy {M}}}
-  {M} {U}}
-
-  \inferrule{\nonlinterm \Gamma M L}{\nonlineq \Gamma {\quoteTy {\unquoteTy {M}}}
-  {M} {L}}
+    %
+    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
+    \and
+    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
   \end{mathpar}
-  \caption{Judgmental equality for non-linear terms}
-  \label{fig:jdg-eq-nonlinear}
+  \caption{Linear terms (extended)}
 \end{figure}
@@
 \begin{figure}
-  \footnotesize
   \begin{mathpar}
-    \boxed{
-      \inferrule
-      {\linterm \Gamma \Delta e A \and
-       \linterm \Gamma \Delta e' A}
-      {\Gamma ; \Delta \vdash e \equiv e' : A}}
+    \boxed{\Gamma ; \Delta \vdash e \equiv e' : A}
 
@@
 %
-    \inferrule{\Gamma ; a : A , \Delta \vdash e : C \\ \Gamma ; \Delta' \vdash e' : A }{\Gamma; \Delta , \Delta' \vdash \apptol {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
+    \inferrule{\Gamma ; a : A , \Delta \vdash e : C \\ \Gamma ; \Delta' \vdash e' : A }{\Gamma; \Delta , \Delta' \vdash \app {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
 %
@@
 %
-    \inferrule{\Gamma ; \Delta \vdash e : B \tol A}{\Gamma; \Delta \vdash e \equiv \lambtol a {\apptol e a} : B \tol A}
+    \inferrule{\Gamma ; \Delta \vdash e : B \tol A}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : B \tol A}
 %
@@
 %
-    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTyLimit {x} {X} {A}}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTyLimit x X A}
+    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy {x} {X} {A}}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
 %
@@
     %
-    \inferrule{\Gamma ; \Delta_1, y : \LinSigTyLimit {x} {X} {A}, \Delta_2 \vdash e' : C \\ \Gamma ; \Delta_2 \vdash e : \LinSigTyLimit {x} {X} {A}}{\Gamma; \Delta_1 , \Delta_2 , \Delta_3 \vdash \letin {\sigma~x~a} e {\subst {e'} {\sigma~x~a} y} \equiv \subst {e'} e y : C}
+    \inferrule{\Gamma ; \Delta_1, y : \LinSigTy {x} {X} {A}, \Delta_2 \vdash e' : C \\ \Gamma ; \Delta_2 \vdash e : \LinSigTy {x} {X} {A}}{\Gamma; \Delta_1 , \Delta_2 , \Delta_3 \vdash \letin {\sigma~x~a} e {\subst {e'} {\sigma~x~a} y} \equiv \subst {e'} e y : C}
     \and
@@
 \end{figure}
-
-\subsection{Substitutions}
+\steven{Need to include premises for the judgmental equality rules}
+
 \begin{definition}
@@
 \begin{figure}
-  \footnotesize
   \begin{mathpar}
     \inferrule
-    {\ctxwffjdg \Gamma X}
-    {\ctxwffjdg {\Gamma'} {X[\gamma]}}
+    {\Gamma \vdash X \isTy}
+    {\Gamma' \vdash X[\gamma] \isTy}
 
@@
     \inferrule
-    {\ctxwffjdg \Gamma {X \equiv Y}}
-    {\ctxwffjdg {\Gamma'} {X[\gamma] \equiv Y[\gamma]}}
+    {\Gamma \vdash X \equiv Y}
+    {\Gamma' \vdash X[\gamma] \equiv Y[\gamma]}
 
     \inferrule
-    {\nonlinterm \Gamma {M} {X}}
-    {\nonlinterm {\Gamma'} {M[\gamma]} {X[\gamma]}}
+    {\Gamma \vdash M : X}
+    {\Gamma' \vdash M[\gamma] : X[\gamma]}
 
     \inferrule
-    {\nonlineq \Gamma {M} {N} {X}}
-    {\nonlineq {\Gamma'} {M[\gamma]} {N[\gamma]} {X[\gamma]}}
+    {\Gamma \vdash M \equiv N : X}
+    {\Gamma' \vdash M[\gamma] \equiv N[\gamma] : X[\gamma]}
 
@@
     \inferrule
-    {\linctxwffjdg \Gamma {A \equiv B}}
-    {\linctxwffjdg {\Gamma'} {A[\gamma] \equiv B[\gamma]}}
+    {\Gamma \vdash A \equiv B}
+    {\Gamma' \vdash A[\gamma] \equiv B[\gamma]}
 
@@
     {\Gamma;\Delta \vdash e : A}
-    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] : A[\gamma]}
+    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] : A[\gamma]}j
 
@@
     {\Gamma;\Delta \vdash e \equiv f : A}
-    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] \equiv f[\gamma] : A[\gamma]}
+    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] \equiv f[\gamma] : A[\gamma]}j
   \end{mathpar}
@@
 \begin{figure}
-  \footnotesize
   \begin{align*}
@@
 \end{figure}
+\steven{captilizataion should be consistent over all figures. idk if all caps or
+all lowercase is beter?}
 
@@
 \]
+
 So for $a: A \vdash a : A$, the denotation of a single variable term
@@
 \]
+
 \begin{proof}
@@
 that is provided by the adjunction between $\semg{- \otimes A}$ and $\semg{A \lto -}$.
+
 \[
@@
 \]
+
 \paragraph{$\lto$-Elimination}
@@
 $\pi_{eq}$ from $Eq(\semg{f} , \semg{g})$ to the domain of $f$ and $g$.
+
 \[
@@
 \]
+
 \paragraph{Equalizer $\beta$}
@@
 \end{proof}
-
-\subsection{Grammar Semantics Respects Additional Axioms}
-We verify that the denotational semantics validates each of the axioms we have
-assumed.
-
-\paragraph{\cref{ax:dist}} Distributivity
-\begin{theorem}[\Agda{\fillerlink}]
-  In $\Grammar$, $\semg{\lamblto{e}{\letin{\sigma\,f\,e'}{e}{\withlamb{x}{e'\,.\pi\,x}}}}$ has
-  an inverse.
-\end{theorem}
-\begin{proof}
-  Distributivity is true in the denotational semantics, as the category $\Grammar$ is a topos, which are well-known to be distributive.
-  The following map forms the desired inverse,
-  \[
-    \lamb {\gamma} {\lamb {w} {\lamb {p} {(\lamb {\left( \lamb {x} {(p~x).fst} \right)} , \lamb {x} {(p~x).snd})}}}
-  \]
-\end{proof}
-
-\paragraph{\cref{ax:disjointness}} $\sigma$-Disjointness
-\begin{theorem}[\Agda{\fillerlink}]
-  In $\Grammar$, $\semg{\sigma x}$ and $\semg{\sigma x'}$ are disjoint for
-  $x \neq x'$.
-\end{theorem}
-\begin{proof}
-  This is trivially true, as the denotation of linear sum types is as $\Sigma$
-  types in $\Grammar$. For any input, the first projections of $\semg{\sigma x}$
-  is $x$ and the first projection of
-  $\semg{\sigma x'}$ is $x'$. Because $x \neq x'$, the images of
-  $\semg{\sigma x}$ and $\semg{\sigma x'}$ cannot agree.
-\end{proof}
-
-\paragraph{\cref{ax:string-top}} String is strongly equivalent to $\top$.
-\begin{theorem}[\Agda{\fillerlink}]
-  In $\Grammar$, $\semg{!}$ has an inverse.
-\end{theorem}
-\begin{proof}
-  Because $\semg{\top}w$ is a singleton set for all $\gamma$ and $w$, it suffices to show that
-  $\semg{\StringGram}w$ is likewise a singleton.
-
-  First, we prove by induction on $w$ that $\semg{\StringGram}w$ is a retract of
-  $\semg{\LinSigTy{w}{\StringSem}{\internalize{w}}}w$, a sum over a nonlinear
-  type of strings. Then, again by
-  induction on $w$, we show that
-  $\semg{\LinSigTy{w}{\StringSem}{\internalize{w}}}w$ is isomorphic to
-  $\semg{\top}w$.
-
-  Each $\semg{\internalize{w}}w'$ is inhabited if and only if $w$ is equal to
-  $w'$, and the parses of $w$ for $\semg{\internalize{w}}$ are unique. That is,
-  $\semg{\LinSigTy{w}{\StringSem}{\internalize{w}}}w$ is a singleton set.
-  So, $\semg{\StringGram}w$ is a retract of a singleton set, and is itself
-  singleton. Thus, $\semg{\StringGram}w \cong \semg{\top}w$ in $\Set$.
-\end{proof}
-
 \fi

### 2025-03-28 src/Everythings.hs
Merge artifact + paper revisions (#37)
* Remove files that are not needed for the claims in the paper

* Spelling in README

* delete unused files + flatten directory

* dockerfile path

* remove workflows and Helper.agda

* readme spelling

* readme docker instructions

* *.tar added to gitignore

* regex parser + parser levels

* spelling, wording tweak

* rewording

* need to prove that axioms imply unamibuity of eps and literal then im done

* checkin

* final checkin

* update names in Dockerfile

* final code clean up

* docker image is wayyy smaller now

* Delete LICS draft

* fix typos

@@ -1,0 +1,133 @@
+import System.Environment ( getArgs )
+import System.Directory   ( getDirectoryContents )
+import System.Exit        ( exitFailure )
+import Control.Monad      ( forM, forM_ )
+import Data.Maybe         ( listToMaybe, maybeToList, mapMaybe )
+import Data.List          ( (\\), delete, find, intercalate, sort, stripPrefix, isSuffixOf )
+
+stripSuffix :: (Eq a) => [a] -> [a] -> Maybe [a]
+stripSuffix sfx = fmap reverse . stripPrefix (reverse sfx) . reverse
+
+splitOn :: Char -> String -> [String]
+splitOn d = foldr go []
+  where go :: Char -> [String] -> [String]
+        go x acc | d == x = []:acc
+        go x (a:acc) = (x:a):acc
+        go x [] = [[x]]
+
+
+type SplitFilePath = [String]
+
+showFP :: Char -> SplitFilePath -> String
+showFP c fp = intercalate [c] (reverse fp)
+
+addToFP :: SplitFilePath -> String -> SplitFilePath
+addToFP fp dir = dir : fp
+
+-- Given a path to a directory, returns a pair containing the list of all its
+--  subdirectories and the list of all agda files it contains
+getSubDirsFiles :: SplitFilePath -> IO ([String], [String])
+getSubDirsFiles fp = do
+  ls <- getDirectoryContents ("./" ++ showFP '/' fp)
+  let sub_dirs = filter ('.' `notElem`) ls
+      files    = mapMaybe (stripSuffix ".agda") ls
+  pure (sub_dirs, files)
+
+-- Given a path to an agda file, returns the list of all files it imports
+getImported :: SplitFilePath -> IO [SplitFilePath]
+getImported fp = do
+  ls <- fmap words . lines <$> readFile ("./" ++ showFP '/' fp ++ ".agda")
+  pure $ fmap (reverse . splitOn '.') (mapMaybe f ls)
+  where f :: [String] -> Maybe String
+        f ("open":"import":x:_) = Just x
+        f ("import":x:_) = Just x
+        f _ = Nothing
+
+-- Given a path to a directory $fp and a path to an agda file $fileToCheck.agda,
+--  returns the list of all files in* $fp not imported in $fileToCheck.agda
+-- * recursively
+getMissingFiles :: SplitFilePath -> Maybe SplitFilePath -> IO [SplitFilePath]
[…]

### 2025-04-03 paper/paper.tex
use flag for references to the appendix

@@
   %
-  \raisebox{-2pt}{\includegraphics[height=1em]{logo-agda.pdf}}%
+  \raisebox{-2pt}{\includegraphics[height=1em,natheight=500,natwidth=500]{logo-agda.pdf}}%
 }
@@
 \newcommand{\pipe}{\,|\,}
+
+\newif\ifappendix
+% \appendixfalse
+\appendixtrue
 
@@
 
-% \newtheorem{axiom}{Axiom}[section]
-% \newtheorem{construction}{Construction}
-
-\newif\ifappendix
-\appendixtrue
-
 \begin{document}
-
-%\pagestyle{plain}
 
@@
 \Cref{fig:non-linear-types} and provide the rest in
-\cref{fig:full-non-linear-types,fig:jdg-eq-nonlinear} of the appendix.  First,
+\ifappendix{\cref{fig:full-non-linear-types,fig:jdg-eq-nonlinear} of
+    \cref{sec:syntax}}\else{the appendix}\fi.  First,
 we include universe types $U$ of small non-linear types and $L$ of linear
@@
 $Fin~n$ --- can be defined in terms of these primitives, and we give their
-encoding in the appendix.
+encoding in \ifappendix{\cref{sec:syntax}}\else{the appendix}\fi.
 
@@
 \cref{fig:linear-terms}. The equational theory for these
-types is straightforward $\beta\eta$ equivalence and included in the
-appendix. First, the linear variable rule says that a linear variable
+types is straightforward $\beta\eta$ equivalence and is included in
+\ifappendix{\cref{fig:jdgeq} of \cref{sec:denotational}}\else{the appendix}\fi. First, the
+linear variable rule says that a linear variable
 can be used if it is the \emph{only} variable in the context.
@@
 products and additive conjunction and disjunction of type expressions.
-We additionally add equations in the appendix that say that the
+Further, we add equations in \ifappendix{\cref{fig:spf-act} of \cref{sec:syntax}
+  }\else{the appendix }\fi that say that the
 $\el$/$\map$ operations correspond to these descriptions of the
@@
 The remaining details of the interpretation of linear terms as parse
-transformers and verification of the equational axioms --- as well as
-\cref{ax:dist,ax:disjointness,ax:string-top} --- is a relatively
-straightforward extension of existing semantics of linear logic in
-monoidal categories, and is included in the appendix\cite{seely89}.
+transformers is a relatively straightforward extension of existing semantics for
+linear logic in monoidal categories \cite{seely89}. In
+\ifappendix{\cref{sec:denotational}}\else{the appendix}\fi, we include the denotations
+of linear terms and prove that our semantics respects the equational theory of
+\theoryabbv --- as well as \cref{ax:dist,ax:disjointness,ax:string-top}.
 
@@
 \section{Syntax}
+\label{sec:syntax}
+
 \begin{figure}
@@
 \subsection{Non-linear Types}
+\label{subsec:nonlinty}
 We define sum types, list types, and $Fin~n$ from primitve non-linear types.
@@
 \section{Denotational Semantics}
+\label{sec:denotational}
 Here we extend the denotational semantics from

### 2025-04-03 paper/paper.tex
more appendix references

@@
 
-We present a selection of the non-linear type constructions in
+We present a selection of the non-linear type constructors in
 \Cref{fig:non-linear-types} and provide the rest in
@@
 non-linear and linear types are \emph{small} and define the universes to internalize
-precisely this judgment \cite{coquandPresheafModel,lmcs:7713}.
-The definition of smallness is simply that all types are small as long as their sub-formulae are, with the exception of the two universe types themselves.
-These universe
+precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. The definition of
+smallness is simply that all types are small as long as their sub-formulae are,
+with the exception of the two universe types themselves. A formal
+description of smallness is given in
+\ifappendix{\cref{fig:small-nonlin-ty,fig:small-lin-ty} of
+    \cref{sec:syntax}}\else{the appendix}\fi. These universe
 types are needed so that we can define non-linear and linear types by recursion on natural
@@
 
-In \cref{fig:spf-act}, we provide the functorial actions for the strictly
-positive functors defined in \cref{fig:iilt}. We also include the full set of
-rules for non-linear types in \cref{fig:full-non-linear-types}, and we provide
-their judgmental equalities in \cref{fig:jdg-eq-nonlinear}. Finally,
-\cref{fig:jdgeq} contains the judgmental equalities between linear terms in
-\theoryabbv.
+\begin{itemize}
+\item In \cref{fig:spf-act}, we provide the functorial actions of the $\el$
+  and $\map$ operations used to define the indexed inductive types in
+  \cref{fig:iilt}.
+\item In \cref{fig:small-nonlin-ty,fig:small-lin-ty} we give the inference rules
+  for the smallness judgements on nonlinear types and linear types, respectively.
+\item In \cref{fig:full-non-linear-types}, we include the full set of rules for
+  non-linear types, and in \cref{fig:jdg-eq-nonlinear} we provide their
+  judgmental equalities.
+\item In \cref{fig:jdgeq}, we give the judgmental equalities between
+  linear terms in \theoryabbv.
+\end{itemize}
 
@@
   \caption{Small Non-linear Types}
+  \label{fig:small-nonlin-ty}
 \end{figure}
@@
   \caption{Small Linear Types}
+  \label{fig:small-lin-ty}
 \end{figure}

### 2025-04-03 paper/paper.tex
zenodo link + shepherd comment + camera-ready documentclass

@@
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[acmsmall]{acmart}\settopmatter{}
+% \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+% For final camera-ready submission, w/ required CCS and ACM Reference
+\documentclass[acmsmall]{acmart}\settopmatter{}
 
@@
   %
-  \raisebox{-2pt}{\includegraphics[height=1em,natheight=500,natwidth=500]{logo-agda.pdf}}%
+  \raisebox{-2pt}{\includegraphics[height=1em]{logo-agda.pdf}}%
 }
@@
 
-\newcommand{\fillerlink}{http://localhost:1313}
-\newcommand{\Agda}[1]{\href{#1}{\agdalogo}}
+\newcommand{\zenodolink}{https://zenodo.org/records/15049780}
+\newcommand{\Agda}{\href{\zenodolink}{\agdalogo}}
 
@@
 \end{mathpar}
-\caption{(\Agda{fillerlink}) \stringquote{ab} is parsed by $(\literal a \oplus \literal b) \otimes \literal c$
+\caption{(\Agda) \stringquote{ab} is parsed by $(\literal a \oplus \literal b) \otimes \literal c$
 }
@@
 \end{lstlisting}
-\caption{(\Agda{\fillerlink}) \stringquote{ab} is parsed by
+\caption{(\Agda) \stringquote{ab} is parsed by
   $({\literal a}^{*} \otimes \literal b) \oplus \literal c$}
@@
 \end{lstlisting}
-\caption{(\Agda{\fillerlink}) A parse transformer for abstract grammars}
+\caption{(\Agda) A parse transformer for abstract grammars}
 \label{fig:kleeneabstractproof}
@@
   \end{minipage}%
-  \caption{(\Agda{\fillerlink}) NFA for $(\a^{*} \otimes \b) \oplus \c$ and its corresponding type}
+  \caption{(\Agda) NFA for $(\a^{*} \otimes \b) \oplus \c$ and its corresponding type}
   \label{fig:exampleNFA}
@@
 over additive disjunction --- e.g., in the finitary case that $0 \& A
-\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most general form,
-
+\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. More generally, we assume
+\cref{ax:dist}.
+% Comment from shepherd
+% - The pagebreaks and figures make the explanation of axiom 3.1 a
+% little confusing to read (with the text "In its most general form,"
+% separated from the axiom it is intended to introduce by a pagebreak
+% and a figure). You could fiddle with the layout a bit, or just revise
+% that sentence to "More generally, we assume Axiom 3.1.", or similar.
+%
+% - If I understand it correctly, axiom 3.1 itself is using "e" as a variable name, which is confusing since it's also used as a metavariable for terms - I suggest picking a different variable name.
 \begin{axiom}[Distributivity]
@@
 \(
- \lamblto{e}{\letin{\sigma\,f\,e'}{e}{\withlamb{x}{e'\,.\pi\,x}}} : \LinSigTyLimit{f}{\PiTy{x}{X}{Y(x)}}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
+ \lamblto{b}{\letin{\sigma\,f\,b'}{b}{\withlamb{x}{b'\,.\pi\,x}}} : \LinSigTyLimit{f}{\PiTy{x}{X}{Y(x)}}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
 \LinPiTyLimit{x}{X}{\LinSigTyLimit{y}{Y(x)}{A\,x\,y}}
@@
 establish two useful properties of unambiguity.
-\begin{lemma}[\Agda{\fillerlink}]
+\begin{lemma}[\Agda]
   \label{lem:retract}
@@
 \end{lemma}
-\begin{lemma}[\Agda{\fillerlink}]
+\begin{lemma}[\Agda]
   \label{lem:unambig-sum}
@@
 
-\begin{lemma}[\Agda{\fillerlink}]
+\begin{lemma}[\Agda]
   \label{lem:unambig-to-disjoint}
@@
 following:
-\begin{lemma}[\Agda{\fillerlink}]
+\begin{lemma}[\Agda]
   \label{lem:wk-eqv-parse}
@@
 
-\begin{theorem}[\Agda{\fillerlink}]
+\begin{theorem}[\Agda]
   \label{thm:dfa-parser}
@@
 DFA is unambiguous even if the NFA is not.
-\begin{construction}[Determinization, \Agda{\fillerlink}]
+\begin{construction}[Determinization, \Agda]
   \label{cons:determinization}
@@
 \newcommand{\R}{\texttt{R}}
-\begin{construction}[Thompson's Construction, \Agda{\fillerlink}]
+\begin{construction}[Thompson's Construction, \Agda]
   \label{cons:thompson}
@@
 
-\begin{corollary}[\Agda{\fillerlink}]
+\begin{corollary}[\Agda]
   We may build a parser for every regular expression $\R$.
@@
 
-\begin{theorem}[\Agda{\fillerlink}]
+\begin{theorem}[\Agda]
   \label{thm:exp-parser}
@@
 languages in the Chomsky hierarchy.
-\begin{construction}[\Agda{\fillerlink}]
+\begin{construction}[\Agda]
   \label{cons:turing}
@@
 
-\subsection{Agda Implementation \Agda{\fillerlink}}
+\subsection{Agda Implementation \Agda}
 \label{subsec:impl}
@@
 \paragraph{\cref{ax:dist}} Distributivity
-\begin{theorem}[\Agda{\fillerlink}]
+\begin{theorem}[\Agda]
   In $\Grammar$, $\semg{\lamblto{e}{\letin{\sigma\,f\,e'}{e}{\withlamb{x}{e'\,.\pi\,x}}}}$ has
@@
 \paragraph{\cref{ax:disjointness}} $\sigma$-Disjointness
-\begin{theorem}[\Agda{\fillerlink}]
+\begin{theorem}[\Agda]
   In $\Grammar$, $\semg{\sigma x}$ and $\semg{\sigma x'}$ are disjoint for
@@
 \paragraph{\cref{ax:string-top}} String is strongly equivalent to $\top$.
-\begin{theorem}[\Agda{\fillerlink}]
+\begin{theorem}[\Agda]
   In $\Grammar$, $\semg{!}$ has an inverse.

### 2025-04-03 paper/paper.tex
orcids + acks

@@
 \newif\ifappendix
-% \appendixfalse
 \appendixtrue
@@
 \email{stschaef@umich.edu}
+\orcid{0009-0007-1258-9501}
 
@@
 \email{nmvarner@umich.edu}
+\orcid{0009-0000-3031-4930}
 
@@
 \email{pedro.azevedo.de.amorim@cs.ox.ac.uk}
+\orcid{0000-0002-8338-8973}
 
@@
 \email{maxsnew@umich.edu}
+\orcid{0000-0001-8141-195X}
 
@@
 context-free grammars and parsers based on $\LL{1}$ and $\LR{1}$
-automata. Throughout this paper, \agdalogo~ will mark
+automata. Throughout this paper, \Agda~ will mark
 results that are mechanized in our Agda development and provide a link to their
@@
 
-% \begin{acks}
-%   This material is based upon work supported by the
-%   \grantsponsor{GS100000001}{National Science
-%     Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-%   No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-%   No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-%   conclusions or recommendations expressed in this material are those
-%   of the author and do not necessarily reflect the views of the
-%   National Science Foundation.
-% \end{acks}
+\begin{acks}
+  This material is based upon work supported by the
+  \grantsponsor{GS100000001}{National Science
+    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
+  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
+  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
+  conclusions or recommendations expressed in this material are those
+  of the author and do not necessarily reflect the views of the
+  National Science Foundation.
+\end{acks}
 

### 2025-04-03 paper/paper.tex
grammar + spacing

@@
 \newif\ifappendix
-\appendixtrue
+\appendixfalse
 
@@
 an extensive experiment finding bugs in C compilers
-\cite{yangFindingUnderstandingBugs}, an early version of the formally
-verified CompCert C compiler only contained bugs in the then
+\cite{yangFindingUnderstandingBugs}, an early version of CompCert, the formally
+verified C compiler, only contained bugs in the then
 unverified parsing component \cite{leroy_formal_2009}. Bugs in parsers
 undermine the overall correctness theorem for a verified system: an
-incorrectly parsed C program will be compiled correctly but this is
+incorrectly parsed C program will be compiled correctly, but this is
 not very useful if it did not correctly correspond to the actual
@@
 The substructural nature of \theoryabbv is well-aligned with the
-requirements intrinsic to parsing and to the theory of formal
-languages, where strings constitute a very clear notion of resource
+requirements intrinsic to parsing and the theory of formal
+languages, where strings constitute a clear notion of resource
 that cannot be duplicated, reordered, or dropped. Moreover, the constructive
 aspect of \theoryabbv ensures that verification of parsers written in the
-calculus are \emph{correct-by-construction}. The type system is powerful enough
-that derivations of a term type checking carry intrinsic proofs of correctness.
+calculus are \emph{orrect-by-construction}. That is, our type system is rich
+enough that typing derivations carry intrinsic proofs of correctness.
 Parsers written in \theoryabbv take on a linear functional style, which makes
@@
 context-free grammars and parsers based on $\LL{1}$ and $\LR{1}$
-automata. Throughout this paper, \Agda~ will mark
+automata. Throughout this paper \Agda~ will mark
 results that are mechanized in our Agda development and provide a link to their
@@
     support inductive linear types.
-  \item A demonstration of how to encode many common grammar formalisms
+  \item A demonstration of how to encode many common grammar and parser formalisms
     (regular expressions, (non-)deterministic automata, context-free
-    grammars) and parser formalisms within our type theory.
+    grammars) within our type theory.
   \item A prototype implementation of \theoryabbv in Agda with all
@@
 but an expression such as
-$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ does not give a very operational perspective of how to parse
-it. For this reason, most parsers are based on compiling a grammar to a corresponding notion of automaton, which is readily implemented. To implement
-these algorithms in \theoryabbv, we need a way to represent automata as types in the same way we can represent regular expressions.
+$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ does not give an
+operationalized method for parsing. For this reason, most parsers are
+implemented by
+compiling a grammar to a corresponding automaton, which is readily
+implemented. To implement
+these algorithms in \theoryabbv, we represent automata as types in the same way we can represent regular expressions.
 
@@
 undecidable \citep{Hofmann_1997} because the conversion rule may require an arbitrarily complex
-equality proof with no explicit proof term, but in our Agda implementation we
-must provide all of these equalities manually, and so the extensionality does
-not raise any issues for our implementation. This makes
+equality proof with no explicit proof term. However, in our Agda implementation we
+must provide all of these equalities manually, so the extensionality does
+not raise any issues. This makes
 \theoryabbv into an extensional theory, supporting function extensionality and
@@
 types is straightforward $\beta\eta$ equivalence and is included in
-\ifappendix{\cref{fig:jdgeq} of \cref{sec:denotational}}\else{the appendix}\fi. First, the
+\ifappendix{\cref{fig:jdgeq} of \cref{sec:denotational}}\else{the appendix}\fi.
+
+First, the
 linear variable rule says that a linear variable
-can be used if it is the \emph{only} variable in the context.
-
-Next, we cover the ``multiplicative'' connectives of non-commutative
+can be used if it is the \emph{only} variable in the context. Next, we cover the
+``multiplicative'' connectives of non-commutative
 linear logic. The linear unit ($\I$) and tensor product ($\otimes$)
@@
   \Trace_{D}~s~b}$ is a retract of $\StringGram$ and apply \cref{lem:retract} to
-derive unambiguity. That is, first we
+derive unambiguity. That is, we first
 construct a function $\pparse_{D}$ that is a parser for
 $\Trace_D~{s}~{true}$, with $\Trace_D~{s}~{false}$ being the disjoint
-type used. The fact that these trace types are disjoint follows from the
+type used, with disjointness following from the
 unambiguity of $\LinSigTy {b} {\Bool} {\Trace_{D}~s~b}$ by \cref{lem:unambig-sum}.
@@
 \begin{proof}
-  Define the states of $\D$ to be the $\mathbb{P}_{\epsilon}(\N.\states)$ --- the type
+  Define the states of $\D$ to be $\mathbb{P}_{\epsilon}(\N.\states)$ --- the type
   of $\epsilon$-closed\footnote{A subset of states $\X$ is $\epsilon$-closed if
@@
   from $\N$. Construct the initial state of $\D$ as the
-  $\epsilon$-closure of $\N.\init$. Lastly define the transition function
+  $\epsilon$-closure of $\N.\init$. Lastly, define the transition function
   of $\D$ to send the subset $X$ under the character $\c$ to the
@@
     similarly define a formal grammar $\sem\Delta\gamma$ for each
-    linear context $\Gamma \Delta\isLinCtx$.
+    linear context $\linctxwff \Gamma \Delta$.
   \item For each non-linear term $\Gamma \vdash M : X$ and $\gamma \in \sem{\Gamma}$, we define an element $\sem{M}\gamma \in \sem{X}\gamma$.
-  \item For each linear term $\Gamma; \Delta \vdash e : A$ and $\gamma \in \sem{\Gamma}$ we define a parse transformer from $\sem{\Delta}\gamma$ to $\sem{A}\gamma$.
+  \item For each linear term $\Gamma; \Delta \vdash e : A$ and $\gamma \in \sem{\Gamma}$, we define a parse transformer from $\sem{\Delta}\gamma$ to $\sem{A}\gamma$.
   \end{enumerate}
@@
 \begin{figure}
-  \begin{minipage}[t]{.5\textwidth}
+  \begin{minipage}[t]{.3\textwidth}
   \begin{footnotesize}
@@
     &\sem{I}\gamma\,w = \{ () \pipe w = \varepsilon \}\\
-    &\sem{A \otimes B}\gamma\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
+    &\sem{\uparrow A}\gamma = \sem{A}\gamma\,\varepsilon\\
     &\sem{A \lto B}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
     &\sem{B \tol A}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
-    &\sem{\bigoplus_{x:X} A}\gamma\,w = \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}
+    &\sem{\el(F)}\gamma\,G = \sem{F}\gamma\,G\\
+    &\sem{\map(F)}\gamma\,f = \sem{F}\gamma\,f\\
+    &\sem{\mu A}\gamma = \mu (\sem{A}\gamma)
   \end{flalign*}
@@
   \end{minipage}%
-  \begin{minipage}[t]{.5\textwidth}
+  \begin{minipage}[t]{.7\textwidth}
   \begin{footnotesize}
     \begin{flalign*}
+    &\sem{A \otimes B}\gamma\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
+    &\sem{\bigoplus_{x:X} A}\gamma\,w = \{ (x, a) \pipe x \in \sem{X}\gamma
+      \wedge a \in \sem{A}(\gamma,x)w \} \\
     &\sem{\bigwith_{x:X} A}\gamma\,w = \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w \\
-    &\sem{\uparrow A}\gamma = \sem{A}\gamma\,\varepsilon\\
     &\sem{\equalizer{a}{f}{g}}\gamma\,w = \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
     &\sem{L}\gamma = \Grammar_0\\
-    &\sem{SPF\,X}\gamma = \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
-    &\sem{\el(F)}\gamma\,G = \sem{F}\gamma\,G\\
-    &\sem{\map(F)}\gamma\,f = \sem{F}\gamma\,f\\
-    &\sem{\mu A}\gamma = \mu (\sem{A}\gamma)
+    &\sem{SPF\,X}\gamma = \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)
   \end{flalign*}
@@
 we are careful to only construct grammars, terms and proofs using
-constructs that are possible in \theoryabbv.
+constructs from \theoryabbv.
 

### 2025-04-04 paper/paper.tex
Added Pedro's funding information

@@
   No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-  conclusions or recommendations expressed in this material are those
-  of the author and do not necessarily reflect the views of the
-  National Science Foundation.
+  No.~\grantnum{GS100000001}{mmmmmmm}. This work is also supported by
+  the ERC Consolidator Grant BLAST and the ARIA programme on Safeguarded AI.
+  Any opinions, findings, and conclusions or recommendations expressed in 
+  this material are those of the author and do not necessarily reflect the 
+  views of the National Science Foundation, the European Research Council or
+  the Advanced Research and Invention Agency.
 \end{acks}

### 2025-04-04 paper/paper.tex
author info, AFOSR grant acknowledgment

@@
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-% \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-% For final camera-ready submission, w/ required CCS and ACM Reference
+%% \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+%% For final camera-ready submission, w/ required CCS and ACM Reference
 \documentclass[acmsmall]{acmart}\settopmatter{}
@@
 \author{Max S. New}
+%% \authornote{with author1 note}          %% \authornote is optional;
+                                        %% can be repeated if necessary
+\orcid{0000-0001-8141-195X}
 \affiliation{
-  \department{Electrical Engineering and Computer Science}
-  \institution{University of Michigan}
-  \country{USA}
+  \position{Assistant Professor}
+  \department{Computer Science and Engineering}              %% \department is recommended
+  \institution{University of Michigan}            %% \institution is required
+  % \streetaddress{2260 Hayward St.}
+  \city{Ann Arbor}
+  \state{MI}
+  % \postcode{48109-2121}
+  \country{USA}                    %% \country is recommended
 }
-\email{maxsnew@umich.edu}
-\orcid{0000-0001-8141-195X}
+\email{maxsnew@umich.edu} %% \email is recommended
 
@@
 \end{abstract}
+
+%% 2012 ACM Computing Classification System (CSS) concepts
+%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
+
+%% Keywords
+%% comma separated list
+\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
 
@@
 \begin{acks}
-  This material is based upon work supported by the
-  \grantsponsor{GS100000001}{National Science
-    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}. This work is also supported by
-  the ERC Consolidator Grant BLAST and the ARIA programme on Safeguarded AI.
-  Any opinions, findings, and conclusions or recommendations expressed in 
-  this material are those of the author and do not necessarily reflect the 
-  views of the National Science Foundation, the European Research Council or
-  the Advanced Research and Invention Agency.
+  This material is based upon work supported by the Air Force Office of
+  Scientific Research under Grant No.~\grantnum{}{FA9550-23-1-0760}.  Any
+  opinions, findings and conclusions or recommendations expressed in this
+  material are those of the authors and do not necessarily reflect the views of
+  the U.S. Department of Defense.
+  %
+  This work is also supported by the ERC Consolidator Grant BLAST and the ARIA
+  programme on Safeguarded AI.  Any opinions, findings, and conclusions or
+  recommendations expressed in this material are those of the author and do not
+  necessarily reflect the views of the National Science Foundation, the European
+  Research Council or the Advanced Research and Invention Agency.
 \end{acks}

### 2025-04-04 paper/paper.tex
more grammar + spacing

@@
 \newcommand{\cons}{\texttt{cons}}
-\newcommand{\epscons}{\varepsilon\texttt{cons}}
+\newcommand{\epscons}{\epsilon\texttt{cons}}
 \newcommand{\data}{\mathsf{data}}
@@
 \newif\ifappendix
-\appendixfalse
+\appendixtrue
 
@@
 not very useful if it did not correctly correspond to the actual
-source program. Eventually, a correct parser was implemented using an
-automaton that is formally verified to implement an $\LR{1}$ grammar
+source program. Eventually, CompCert
+was updated to use a verified parser that implements
+an $\LR{1}$ grammar
 \cite{jourdanValidatingLRParsers2012}.
@@
 \emph{why} verified parsing was not part of the initial releases of
-CompCert: parsing algorithms and formal grammars are a complex area,
+CompCert: parsing algorithms are a complex area,
 featuring a variety of domain-specific formalisms such as context-free
@@
 Prior approaches to verified parsing focus on verification of a
-particular grammar formalism such as non-left-recursive grammars or
+particular grammar formalism, such as non-left-recursive grammars or
 $\LL{1}$ grammars
 \cite{lasserCoStarVerifiedALL2021,EdelmannZippy2020,danielssonTotalParserCombinators2010}.
-Each new grammar formalism is extended with its own independent
+This has led to a series of isolated solutions, as each new grammar formalism is extended with its own independent
 verified implementation.
@@
 aspect of \theoryabbv ensures that verification of parsers written in the
-calculus are \emph{orrect-by-construction}. That is, our type system is rich
-enough that typing derivations carry intrinsic proofs of correctness.
+calculus are \emph{correct-by-construction}. That is, our type system is rich
+enough that typing derivations carry intrinsic proofs of parser correctness.
 Parsers written in \theoryabbv take on a linear functional style, which makes
@@
 definitional equality judgments for both kinds of type and term
-judgments as well. Additionally, we have a judgment $\Gamma \vdash X
-\isSmall$ which is used in the definition of universe types.
+judgments as well. Additionally, we have judgments $\Gamma \vdash X
+\isSmall$ and $\Gamma \vdash A \isSmallLin$ which are used in the definition of
+universe types.
 
@@
 grammar theory and define parsers, we need only add a few
-grammar-specific constructions. First for each character in our
+grammar-specific constructions. First, for each character in our
 alphabet, we add a corresponding linear type $c$. We can then define a
@@
 \emph{proposition} in the terminology of homotopy type
-theory\cite{hottbook}. The most basic unambiguous types are $\top$ and
+theory \cite{hottbook}. The most basic unambiguous types are $\top$ and
 $0$, and in a system of classical logic all unambiguous types would
@@
 derive unambiguity. That is, we first
-construct a function $\pparse_{D}$ that is a parser for
-$\Trace_D~{s}~{true}$, with $\Trace_D~{s}~{false}$ being the disjoint
-type used, with disjointness following from the
+construct a function $\pparse_{D}$ which is a parser for
+$\Trace_D~{s}~{true}$ with $\Trace_D~{s}~{false}$ being the disjoint
+type used, and disjointness follows from the
 unambiguity of $\LinSigTy {b} {\Bool} {\Trace_{D}~s~b}$ by \cref{lem:unambig-sum}.
@@
 equivalent to the set of parse transformers:
-\( \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\).
+\( \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\epsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\).
 
@@
     &\sem{c}\gamma\,w = \{ c | w = c\}\\
-    &\sem{I}\gamma\,w = \{ () \pipe w = \varepsilon \}\\
-    &\sem{\uparrow A}\gamma = \sem{A}\gamma\,\varepsilon\\
+    &\sem{I}\gamma\,w = \{ () \pipe w = \epsilon \}\\
+    &\sem{\uparrow A}\gamma = \sem{A}\gamma\,\epsilon\\
     &\sem{A \lto B}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
@@
 Henglein and Nielsen build on this type-theoretic view of grammars, pointing out that the values of the types correspond precisely to the parse trees \cite{henglein_regular_2011}.
-Our design of \theoryabbv extends this view of grammars as types to a much broader class of grammars, as well as providing a syntax and equational theory for \emph{parse transformers}, showing that the terms can be viewed not just as parse trees, but as intrinsically verified parsers.
+\theoryabbv extends this view of grammars as types to a much broader class of grammars, as well as providing a syntax and equational theory for \emph{parse transformers}, showing that the terms can be viewed not just as parse trees, but as intrinsically verified parsers.
 
@@
 We aim to extend our work on parsing to verify other algorithms. For instance,
-our regular expression parser makes somewhat arbitrary choices in what parse to
-produce when the grammar is ambiguous. In the future we aim to verify Frisch and
+our regular expression parser makes somewhat arbitrary choices
+when the grammar is ambiguous. In the future we aim to verify Frisch and
 Cardelli's \emph{greedy} algorithm for regular expression
@@
 
-We also aim to adapt the approach used for our context-free grammar parsers for
-$\LLL$ and $\LRR$/$\LALRR$ parsers, with the aim of developing a shared library of intrinsically
-verified parsing utilities that can be usefully shared across different
-parsers. We would also like to investigate how high-performance verified parsers
+We also aim to adapt the approach used for our context-free grammar parsers to
+$\LLL$ and $\LRR$/$\LALRR$ parsers, with the aim of developing a shared library
+of intrinsically verified parsing utilities.
+We would also like to investigate how high-performance verified parsers
 can be implemented in \theoryabbv. This might be done by developing an efficient

### 2025-04-04 paper/paper.tex
finite alphabet
fix appendix figures from clipping into margins

@@
 grammar theory and define parsers, we need only add a few
-grammar-specific constructions. First, for each character in our
-alphabet, we add a corresponding linear type $c$. We can then define a
+grammar-specific constructions.
+
+\theoryabbv is parameterized by a fixed, finite alphabet $\Sigma$ from which we
+build our strings.  For each character $c \in \Sigma$,
+we add a corresponding linear type $\literal{c}$. We can then define a
 non-linear type $\CharGram$ as the disjunction of all of these characters,
@@
 \begin{figure}
-\begin{minipage}[t]{.35\textwidth}
   \begin{align*}
@@
     \el\left(\bigamp A\right) B &= \bigamp_{y:Y}\el(A y)B\\
-    \el\left(A \otimes A'\right) B &= \el(A)B \otimes \el(A')B
-  \end{align*}
-\end{minipage}%
-\begin{minipage}[t]{.65\textwidth}
-\begin{align*}
+    \el\left(A \otimes A'\right) B &= \el(A)B \otimes \el(A')B \\
     \map\left(\Var\,M\right)\,f &= f\,M\\
@@
   \end{align*}
-\end{minipage}%
   \caption{Strictly positive functors functorial actions}
@@
   \begin{align*}
-  \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}}
+  \llbracket \mathsf{let}\, {\sigma~x~a} = & {\sigma~M~a'} \, \mathsf{in}\, {e'} \rrbracket \gamma \\
+  % \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}}
   & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id
@@
 \[
-  \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}} = \semg{f[c'/c]} = \semg{f}
+  \semg{\letin {\sigma~x~a} {c'} {f[(\sigma~x~a)/c]}} = \semg{f[c'/c]} = \semg{f}
   \]
 First, expanding the left hand side, we have.
+% \newcommand{\sem}[1]{\left\llbracket{#1}\right\rrbracket}
+% \newcommand{\semg}[1]{\sem{#1}\gamma}
+% \newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
 \begin{proof}
   \begin{align*}
-    \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}}
-    &= [ \semg{f[\sigma~x~a/c]} ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
+    \llbracket \mathsf{let}\, {\sigma~x~a} = {c'} & \, \mathsf{in}\, {f[(\sigma~x~a)/c]} \rrbracket \gamma \\
+    &= [ \semg{f[(\sigma~x~a)/c]} ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
     &= [ \semg{f} \circ (id \otimes i_x) \otimes id ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
@@
 \end{document}
+
+%%% Local Variables:
+%%% mode: LaTeX
+%%% TeX-master: t
+%%% End:

### 2025-04-04 paper/paper.tex
fix clipping in appendix

@@
     &\sem{A \otimes B}\gamma\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
-    &\sem{\bigoplus_{x:X} A}\gamma\,w = \{ (x, a) \pipe x \in \sem{X}\gamma
-      \wedge a \in \sem{A}(\gamma,x)w \} \\
+    % &\sem{\bigoplus_{x:X} A}\gamma\,w = \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \} \\
+    &\sem{\bigoplus_{x:X} A}\gamma\,w = \coprod_{x \in \sem{X}\gamma} \sem{A}(\gamma, x)\,w \\
     &\sem{\bigwith_{x:X} A}\gamma\,w = \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w \\
@@
 \begin{center}
-\begin{tikzcd}
+\begin{tikzcd}[column sep = 1in]
   \semg{\Delta,\Delta'} \arrow[r, "m_{\Delta,\Delta'}"] &
@@
 \begin{center}
-\begin{tikzcd}
+\begin{tikzcd}[column sep=.75in]
   \semg{\Delta,\Delta'} \arrow[r , "m_{\Delta,\Delta'}"] &
@@
 \begin{center}
-\begin{tikzcd}
+\begin{tikzcd}[column sep=.75in]
   \semg{A} \otimes \semg{\Delta} \arrow[r , "\lambda^{-1} \otimes id"] &
@@
 \begin{center}
-\begin{tikzcd}
+\begin{tikzcd}[column sep=.75in]
   \semg{\Delta',\Delta} \arrow[r , "m_{\Delta',\Delta}"] &
@@
 \paragraph{$\bigamp$-Introduction}
-$\semg{\dlamb x e} : \semg{\Delta} \to \prod_{x : X} {\sem{A}(\gamma , x)}$ is
+$\semg{\dlamb x e} : \semg{\Delta} \to \prod_{x : \semg{X}} {\sem{A}(\gamma , x)}$ is
 defined by the universal property of the product
 \[
-\semg{\dlamb x e} = \left( \sem{e}(\gamma , x) \right)_{(x : X)}
+\semg{\dlamb x e} = \left( \sem{e}(\gamma , x) \right)_{(x : \semg{X})}
 \]
@@
   \semg{\Delta} \arrow[r , "\semg{e}"] &
-  \prod_{x : X} {\sem{A}(\gamma , x)} \arrow[r , "\pi_M"] &
+  \prod_{x : \semg{X}} {\sem{A}(\gamma , x)} \arrow[r , "\pi_M"] &
   \sem{A}(\gamma , M)
@@
   & = \pi_{M} \circ \semg{\dlamb x e} \\
-  & = \pi_{M} \circ (\sem{e}(\gamma , x))_{(x : X)} \\
+  & = \pi_{M} \circ (\sem{e}(\gamma , x))_{(x : \semg{X})} \\
   & = \sem{e}(\gamma , M)
@@
   \semg{(\dlamb x {e.\pi~x})}
-  & = (\semg{e.\pi~x})_{x : X} \\
-  & = (\pi_x \circ \semg{e})_{x : X} \\
+  & = (\semg{e.\pi~x})_{x : \semg{X}} \\
+  & = (\pi_x \circ \semg{e})_{x : \semg{X}} \\
   & = \semg{e}
@@
 \paragraph{$\bigoplus$-Introduction}
-$\semg{\sigma~M~e} : \semg{\Delta} \to \coprod_{x : X} \sem{A}(\gamma, x)$
+$\semg{\sigma~M~e} : \semg{\Delta} \to \coprod_{x : \semg{X}} \sem{A}(\gamma, x)$
 
@@
   \sem{A}(\gamma, M) \arrow[r , "i_M"] &
-  \coprod_{x : X} {\sem{A}(\gamma , x)}
+  \coprod_{x : \semg{X}} {\sem{A}(\gamma , x)}
 \end{tikzcd}
@@
 {\semg{\Delta'_1,\Delta,\Delta'_2}} \arrow[r, "{m_{(\Delta'_1,\Delta),\Delta'_2}}"]                                                                                                           & {\semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}} \arrow[d, "{m_{\Delta'_1,\Delta}\otimes id}"]                             \\
-{\left( \semg{\Delta'_1} \otimes \coprod_{x:X} {\sem{A}(\gamma , x)} \right) \otimes \semg{\Delta'_2}} \arrow[d, "d"]                                                                 & \left( \semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2} \arrow[l, "(id \otimes \semg{e}) \otimes id"] \\
-{\coprod_{x : X} {\left( \sem{\Delta'_1}(\gamma , x) \otimes
-      \sem{A}(\gamma , x) \right) \otimes \sem{\Delta'_2}(\gamma , x)}} \arrow[r, "\coprod_{x:X}({m^{-1}_{(\Delta'_1,a:A),\Delta'_2}})"] & {\coprod_{x : X} {\sem{\Delta'_1,a : A,\Delta'_2}(\gamma , x)}} \arrow[d, "{[ \sem{e'}(\gamma , x) ]_{(x : X)}}"]     \\
+{\left( \semg{\Delta'_1} \otimes \coprod_{x:\semg{X}} {\sem{A}(\gamma , x)} \right) \otimes \semg{\Delta'_2}} \arrow[d, "d"]                                                                 & \left( \semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2} \arrow[l, "(id \otimes \semg{e}) \otimes id"] \\
+{\coprod_{x : \semg{X}} {\left( \sem{\Delta'_1}(\gamma , x) \otimes
+      \sem{A}(\gamma , x) \right) \otimes \sem{\Delta'_2}(\gamma , x)}} \arrow[r, outer sep=.15cm, "\coprod_{x:\semg{X}}({m^{-1}_{(\Delta'_1,a:A),\Delta'_2}})"] & {\coprod_{x : \semg{X}} {\sem{\Delta'_1,a : A,\Delta'_2}(\gamma , x)}} \arrow[d, "{[ \sem{e'}(\gamma , x) ]_{(x : \semg{X})}}"]     \\
 \semg{C}                                                                                                                                                                                      & {\sem{C}(\gamma, x)} \arrow[l]
@@
   % \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}}
-  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : \semg{X})}} \circ \coprod_{x:\semg{X}}(m^{-1}) \circ d \circ (id
   \otimes \semg{\sigma~M~a'}) \otimes id \circ m \otimes id \circ m \\
-  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X}m^{-1} \circ d \circ (id
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : \semg{X})}} \circ \coprod_{x:\semg{X}}m^{-1} \circ d \circ (id
   \otimes (i_M \circ \semg{a'})) \otimes id \circ m \otimes id \circ m \\
-  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X} m^{-1} \circ d \circ (id \otimes i_M) \otimes id \circ (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
-  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X} m^{-1} \circ i_M \circ (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : \semg{X})}} \circ \coprod_{x:\semg{X}} m^{-1} \circ d \circ (id \otimes i_M) \otimes id \circ (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
+  & = {[ \sem{e'}(\gamma , x) ]_{(x : \semg{X})}} \circ \coprod_{x:\semg{X}} m^{-1} \circ i_M \circ (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
   & = \sem{e'}(\gamma , M) \circ m^{-1} (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
@@
     \llbracket \mathsf{let}\, {\sigma~x~a} = {c'} & \, \mathsf{in}\, {f[(\sigma~x~a)/c]} \rrbracket \gamma \\
-    &= [ \semg{f[(\sigma~x~a)/c]} ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
-    &= [ \semg{f} \circ (id \otimes i_x) \otimes id ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
-    &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
+    &= [ \semg{f[(\sigma~x~a)/c]} ]_{(x : \semg{X})} \circ \coprod_{x:\semg{X}}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
+    &= [ \semg{f} \circ (id \otimes i_x) \otimes id ]_{(x : \semg{X})} \circ \coprod_{x:\semg{X}}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
+    &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : \semg{X})} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
   \end{align*}
@@
 \begin{align*}
-  \semg{f} &\circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id\\
-  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d  \circ (id \otimes i_y) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
-  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ i_y \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
+  \semg{f} &\circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : \semg{X})} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id\\
+  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : \semg{X})} \circ d  \circ (id \otimes i_y) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
+  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : \semg{X})} \circ i_y \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
   &= \semg{f} \circ (id \otimes i_y) \otimes id \circ (m^{-1}) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \\

### 2025-04-04 paper/paper.tex
arxiv flags

@@
 % -*- fill-column: 80; -*-
-%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-%% \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-\documentclass[acmsmall]{acmart}\settopmatter{}
+\newif\ifarxiv
+\arxivtrue
+
+\ifarxiv
+\documentclass[acmsmall,nonacm]{acmart}
+\else
+\documentclass[acmsmall]{acmart}
+\fi
+
+\settopmatter{}
 
@@
 
-\newif\ifappendix
-\appendixtrue
-
 \newcommand{\states}{\mathtt{states}}
@@
 %% comma separated list
+\ifarxiv
+\else
 \keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
+\fi
 
@@
 that makes precise the connection between \theoryabbv syntax and
-formal grammars. Finally in \Cref{sec:discussion} we discuss related
+formal grammars. Finally, in \Cref{sec:discussion} we discuss related
 and future work.
+
+\ifarxiv\else {\textbf{Extended version:} An extended version of this paper is available in \citet{extendedLambekD}.}\fi
 
@@
 \Cref{fig:non-linear-types} and provide the rest in
-\ifappendix{\cref{fig:full-non-linear-types,fig:jdg-eq-nonlinear} of
+\ifarxiv{\cref{fig:full-non-linear-types,fig:jdg-eq-nonlinear} of
     \cref{sec:syntax}}\else{the appendix}\fi.  First,
@@
 description of smallness is given in
-\ifappendix{\cref{fig:small-nonlin-ty,fig:small-lin-ty} of
+\ifarxiv{\cref{fig:small-nonlin-ty,fig:small-lin-ty} of
     \cref{sec:syntax}}\else{the appendix}\fi. These universe
@@
 $Fin~n$ --- can be defined in terms of these primitives, and we give their
-encoding in \ifappendix{\cref{sec:syntax}}\else{the appendix}\fi.
+encoding in \ifarxiv{\cref{sec:syntax}}\else{the appendix}\fi.
 
@@
 types is straightforward $\beta\eta$ equivalence and is included in
-\ifappendix{\cref{fig:jdgeq} of \cref{sec:denotational}}\else{the appendix}\fi.
+\ifarxiv{\cref{fig:jdgeq} of \cref{sec:denotational}}\else{the appendix}\fi.
 
@@
 products and additive conjunction and disjunction of type expressions.
-Further, we add equations in \ifappendix{\cref{fig:spf-act} of \cref{sec:syntax}
+Further, we add equations in \ifarxiv{\cref{fig:spf-act} of \cref{sec:syntax}
   }\else{the appendix }\fi that say that the
@@
 data Atom : L where
-  num : (*@$\uparrow$@*)(NUM (*@$\lto$@*) Atom)
+  num : (*@$\uparrow$@*)('NUM' (*@$\lto$@*) Atom)
   parens : (*@$\uparrow$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ')' (*@$\lto$@*) Atom)
@@
   left : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O (n + 1) b (*@$\lto$@*) O n b)
-  num : (*@$\uparrow$@*)(&[ n ] &[ b ] NUM (*@$\lto$@*) D n b (*@$\lto$@*) O n b)
+  num : (*@$\uparrow$@*)(&[ n ] &[ b ] 'NUM' (*@$\lto$@*) D n b (*@$\lto$@*) O n b)
   unexpected : (*@$\uparrow$@*)(&[ n ] (NotStartsWithLP (*@$\lto$@*) O n false)
@@
   add : (*@$\uparrow$@*)(&[ n ] &[ b ] '+' (*@$\lto$@*) O n b (*@$\lto$@*) A n b)
-  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')'(*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) A n false)
+  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')'(*@$\oplus$@*) 'NUM') (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) A n false)
 \end{lstlisting}
@@
 linear logic in monoidal categories \cite{seely89}. In
-\ifappendix{\cref{sec:denotational}}\else{the appendix}\fi, we include the denotations
+\ifarxiv{\cref{sec:denotational}}\else{the appendix}\fi, we include the denotations
 of linear terms and prove that our semantics respects the equational theory of
@@
 \newpage
-\ifappendix
+\ifarxiv
 \appendix

### 2025-04-04 paper/paper.tex
syntax highlighting

@@
 \usepackage{xspace}
+\usepackage[dvipsnames]{xcolor}
 \usepackage{thmtools} % needed for sane cleveref naming of theorems
@@
 
-\definecolor{codegreen}{rgb}{0,0.6,0}
-\definecolor{codegray}{rgb}{0.5,0.5,0.5}
-\definecolor{codepurple}{rgb}{0.58,0,0.82}
+\colorlet{keywordcolor}{blue}
+\colorlet{nonlintycolor}{RubineRed}
+\colorlet{lintycolor}{Green}
+\colorlet{nonlinconstructorcolor}{nonlintycolor}
+\colorlet{linconstructorcolor}{lintycolor}
+\colorlet{alphabetcolor}{black}
+\colorlet{funarrowcolor}{black}
 \definecolor{backcolour}{rgb}{0.95,0.95,0.92}
@@
 \lstdefinelanguage{Agda}{
+alsoletter={012=\&},
 keywords={data, where, module, import, open, public,
@@
 Renamer},
-morekeywords=[2]{Set, Type, Prop, moduleFuncs, instance, Named},
+morekeywords=[2]{L, Bool, Nat, Fin},
+morekeywords=[3]{String, Char, Dyck, Trace, Exp, Atom, O, D, C, A, NotStartsWithLP, NotStartsWithRP},
+morekeywords=[4]{bal, .nil, .cons, cons, nil, inl, inr, stop, 1to1, 1to2, 0to2, 0to1},
+morekeywords=[5]{.states, states, .isAcc, isAcc, transitions, src, dst, .transitions, .dst,
+  .src, label},
+morekeywords=[6]{NUM},
+morekeywords=[7]{= , \&},
 sensitive=true,
@@
 mathescape=true,
-escapeinside={(*@}{@*)}
+escapeinside={(*@}{@*)},
+escapebegin=\color{funarrowcolor},
+escapeend={}
 }
@@
 basicstyle=\ttfamily\footnotesize,
-keywordstyle=\color{blue},
-keywordstyle=[2]\color{teal},
+keywordstyle=\color{keywordcolor},
+keywordstyle=[2]\color{nonlintycolor},
+keywordstyle=[3]\color{lintycolor},
+keywordstyle=[4]\color{linconstructorcolor},
+keywordstyle=[5]\color{nonlinconstructorcolor},
+keywordstyle=[6]\color{alphabetcolor},
+keywordstyle=[7]\color{funarrowcolor},
 identifierstyle=\color{black},
@@
 \begin{lstlisting}
-f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
+f : (*@$\uparrow$@*)('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)' (*@$\lto$@*) ('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\oplus$@*) '(*@\color{alphabetcolor}c@*)')
 f (a , b) = inl (a , b)
@@
 \begin{lstlisting}
-g : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*) ('a'(*@$^*$@*) (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
+g : (*@$\uparrow$@*)(('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\lto$@*) ('(*@\color{alphabetcolor}a@*)'(*@$^*$@*) (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\oplus$@*) '(*@\color{alphabetcolor}c@*)')
 g (a , b) = inl (cons a nil , b)
@@
   stop : (*@$\uparrow$@*)(Trace 2)
-  1to1 : (*@$\uparrow$@*)('a' (*@$\lto$@*) Trace 1 (*@$\lto$@*) Trace 1)
-  1to2 : (*@$\uparrow$@*)('b' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 1)
-  0to2 : (*@$\uparrow$@*)('c' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 0)
+  1to1 : (*@$\uparrow$@*)('(*@\color{alphabetcolor}a@*)' (*@$\lto$@*) Trace 1 (*@$\lto$@*) Trace 1)
+  1to2 : (*@$\uparrow$@*)('(*@\color{alphabetcolor}b@*)' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 1)
+  0to2 : (*@$\uparrow$@*)('(*@\color{alphabetcolor}c@*)' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 0)
   0to1 : (*@$\uparrow$@*)(Trace 1 (*@$\lto$@*) Trace 0)
 
-k : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*)  Trace 0)
+k : (*@$\uparrow$@*)(('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\lto$@*)  Trace 0)
 k (a , b) = 0to1 (1to1 a (1to2 b stop))
@@
 \begin{lstlisting}
-data Trace(*@$_{\N}$@*) : (s : N.states) (*@$\to$@*) L where
-  nil : (*@$\uparrow$@*)(&[ s : N.states ] N.isAcc s (*@$\to$@*) Trace(*@$_{\N}$@*) s)
-  cons : (*@$\uparrow$@*)(&[ t : N.transitions ]
-    ('N.label t' (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.dst t) (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.src t)))
-  (*@$\epsilon$@*)cons : (*@$\uparrow$@*)(&[ t : N.(*@$\epsilon$@*)transitions ]
-    (Trace(*@$_{\N}$@*) (N.(*@$\epsilon$@*)dst t) (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)src t)))
-
-data Trace(*@$_{\D}$@*) : (s : D.states) (b : Bool) (*@$\to$@*) L where
-  nil : (*@$\uparrow$@*)(&[ s : D.states ] Trace(*@$_{\D}$@*) s D.isAcc s)
-  cons : (*@$\uparrow$@*)(&[ c : Char ] &[ s : D.states ] &[ b : Bool ]
-  ('c' (*@$\lto$@*) Trace(*@$_{\D}$@*) (D.(*@$\delta$@*) c s) b (*@$\lto$@*) Trace(*@$_{\D}$@*) s b))
+data Trace(*@$_{\color{black}\N}$@*) : (s : N .states) (*@$\to$@*) L where
+  nil : (*@$\uparrow$@*)(&[ s : N .states ] N .isAcc s (*@$\to$@*) Trace(*@$_{\color{black}\N}$@*) s)
+  cons : (*@$\uparrow$@*)(&[ t : N .transitions ]
+    ('N .label t' (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .dst t) (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .src t)))
+  (*@$\color{linconstructorcolor}\epsilon$@*)cons : (*@$\uparrow$@*)(&[ t : N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)transitions ]
+    (Trace(*@$_{\color{black}\N}$@*) (N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)dst t) (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)src t)))
+
+data Trace(*@$_{\color{black}\D}$@*) : (s : (*@\color{black}D@*) .states) (b : Bool) (*@$\to$@*) L where
+  nil : (*@$\uparrow$@*)(&[ s : (*@\color{black}D@*) .states ] Trace(*@$_{\color{black}\D}$@*) s (*@\color{black}D@*) .isAcc s)
+  cons : (*@$\uparrow$@*)(&[ c : Char ] &[ s : (*@\color{black}D@*) .states ] &[ b : Bool ]
+  ('(*@\color{alphabetcolor}c@*)' (*@$\lto$@*) Trace(*@$_{\color{black}\D}$@*) ((*@\color{black}D@*) .(*@$\color{nonlinconstructorcolor}\delta$@*) c s) b (*@$\lto$@*) Trace(*@$_{\color{black}\D}$@*) s b))
 \end{lstlisting}
@@
 \begin{lstlisting}
-parse(*@$_D$@*) : (*@$\uparrow$@*)(String (*@$\lto$@*) &[ s : D.states ] (*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\D}$@*) s b)
-parse(*@$_D$@*) String.nil s = (*@$\sigma$@*) (D.isAcc s) (Trace(*@$_{\D}$@*).nil s)
-parse(*@$_D$@*) (String.cons ((*@$\sigma$@*) c a) w) s = let (*@$\sigma$@*) b t = parse w (D.(*@$\delta$@*) c s) in
-                                   (*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c s b a t)
-
-print(*@$_D$@*) : (s : D.states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\D}$@*) s b) (*@$\lto$@*) String)
-print(*@$_D$@*) s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).nil .s)) = String.nil
-print(*@$_D$@*) s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c (D.(*@$\delta$@*) c .s) b a trace)) =
-  String.cons ((*@$\sigma$@*) c a) (print(*@$_D$@*) (D.(*@$\delta$@*) c s) ((*@$\sigma$@*) b trace))
+parse(*@$\color{black}_D$@*) : (*@$\uparrow$@*)(String (*@$\lto$@*) &[ s : (*@\color{black}D@*) .states ] (*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\color{black}\D}$@*) s b)
+parse(*@$\color{black}_D$@*) String .nil s = (*@$\color{linconstructorcolor}\sigma$@*) ((*@\color{black}D@*) .isAcc s) (Trace(*@$_{\color{black}\D}$@*) .nil s)
+parse(*@$\color{black}_D$@*) (String .cons ((*@$\color{linconstructorcolor}\sigma$@*) c a) w) s = let (*@$\color{linconstructorcolor}\sigma$@*) b t = parse w ((*@\color{black}D@*) .(*@$\color{nonlinconstructorcolor}\delta$@*) c s) in
+                                    (*@$\color{linconstructorcolor}\sigma$@*) b (Trace(*@$_{\color{black}\D}$@*) .cons c s b a t)
+
+print(*@$\color{black}_D$@*) : (s : (*@\color{black}D@*) .states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\color{black}\D}$@*) s b) (*@$\lto$@*) String)
+print(*@$\color{black}_D$@*) s ((*@$\color{linconstructorcolor}\sigma$@*) b (Trace(*@$_{\color{black}\D}$@*) .nil .s)) = String .nil
+print(*@$\color{black}_D$@*) s ((*@$\color{linconstructorcolor}\sigma$@*) b (Trace(*@$_{\color{black}\D}$@*) .cons c ((*@\color{black}D@*) .(*@$\color{nonlinconstructorcolor}\delta$@*) c .s) b a trace)) =
+  String .cons ((*@$\color{linconstructorcolor}\sigma$@*) c a) (print(*@$\color{black}_D$@*) ((*@\color{black}D@*) .(*@$\color{nonlinconstructorcolor}\delta$@*) c s) ((*@$\color{linconstructorcolor}\sigma$@*) b trace))
 \end{lstlisting}
@@
 analogously to the parser and printer for DFAs, and so $\LinSigTy {s}
-{M.states} {\LinSigTy {b}{Bool} {Trace_{M}~s~b}}$ is likewise
+{M .states} {\LinSigTy {b}{Bool} {Trace_{M}~s~b}}$ is likewise
 unambiguous.
@@
   \(I \oplus
-  (\literal{)} \oplus \literal{+} \oplus \mathtt{NUM}) \otimes \top
+  (\literal{)} \oplus \literal{+} \oplus \literal{NUM}) \otimes \top
   \). Similarly,
@@
   \(I \oplus
-  (\literal{(} \oplus \literal{+} \oplus \mathtt{NUM}) \otimes \top
+  (\literal{(} \oplus \literal{+} \oplus \literal{NUM}) \otimes \top
   \).
-} has
-four different ``states'', each with access to a natural number
+}
+has four different ``states'', each with access to a natural number
 ``stack''. The ``opening'' state $O$ expects either a left paren, in
@@
 data D : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
-  lookAheadRP : (*@$\uparrow$@*)(&[ n ] &[ b ] ((')' (*@$\otimes$@*) (*@$\top$@*)) (*@$\&$@*) C n b) (*@$\lto$@*) D n b)
+  lookAheadRP : (*@$\uparrow$@*)(&[ n ] &[ b ] ((')' (*@$\otimes$@*) (*@$\color{lintycolor}\top$@*)) (*@$\&$@*) C n b) (*@$\lto$@*) D n b)
   lookAheadNot : (*@$\uparrow$@*)(&[ n ] &[ b ] (NotStartsWithRP (*@$\&$@*) A n b) (*@$\lto$@*) D n b)
@@
   add : (*@$\uparrow$@*)(&[ n ] &[ b ] '+' (*@$\lto$@*) O n b (*@$\lto$@*) A n b)
-  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')'(*@$\oplus$@*) 'NUM') (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) A n false)
+  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')' (*@$\oplus$@*) 'NUM') (*@$\lto$@*) (*@$\color{lintycolor} \top$@*) (*@$\lto$@*) A n false)
 \end{lstlisting}

### 2025-04-08 paper/paper.tex
arxiv xcolor weirdness
spelling + arxiv bibtex

@@
 \newif\ifarxiv
-\arxivtrue
+\arxivfalse
+
+\PassOptionsToPackage{prologue,dvipsnames}{xcolor}
 
@@
 \usepackage{xspace}
-\usepackage[dvipsnames]{xcolor}
 \usepackage{thmtools} % needed for sane cleveref naming of theorems
@@
 
+\ifarxiv
+\title{Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent
+  Lambek Calculus (Extended Version)}
+\else
 \title{Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus}
+\fi
 
@@
 
-\ifarxiv\else {\textbf{Extended version:} An extended version of this paper is available in \citet{extendedLambekD}.}\fi
+\ifarxiv\else {Proofs and syntactic forms elided from this paper are presented
+    in full in the extended version of this paper \cite{schaefer2025intrinsicverificationparsersformal}.}\fi
 
@@
 \label{ax:dist}
-For any $A : (x : X) \to Y(x) \to L$, the definable distribuitivity function
+For any $A : (x : X) \to Y(x) \to L$, the definable distributivity function
 %%%%%%%%%% An attempt to inline
@@
 \item In \cref{fig:small-nonlin-ty,fig:small-lin-ty} we give the inference rules
-  for the smallness judgements on nonlinear types and linear types, respectively.
+  for the smallness judgments on nonlinear types and linear types, respectively.
 \item In \cref{fig:full-non-linear-types}, we include the full set of rules for
@@
 \label{subsec:nonlinty}
-We define sum types, list types, and $Fin~n$ from primitve non-linear types.
+We define sum types, list types, and $Fin~n$ from primitive non-linear types.
 

### 2025-04-08 paper/paper.tex
arbitrary linear term is an e not an a

@@
 A grammar $A$ derives the word $w$ if there exists a derivation
-$\internalize{w} \vdash a : A$,
+$\internalize{w} \vdash e : A$,
 where $\internalize \w$ is a context with one variable for each character of
-$\w$. The term $a : A$ represents a \emph{parse tree} of $w$ for the grammar $A$. For
+$\w$. The term $e : A$ represents a \emph{parse tree} of $w$ for the grammar $A$. For
 example, to define a parse tree for \stringquote{ab}, we use

### 2025-04-08 paper/paper.tex
fix error in distributivity and make it fit the width

@@
 %
-Lastly, we give a lambda term that constructs an accepting trace
+Lastly, we give a $\lambda$ term that constructs an accepting trace
 starting at the initial state for the string \stringquote{ab}.  Later
@@
 \label{ax:dist}
-For any $A : (x : X) \to Y(x) \to L$, the definable distributivity function
-%%%%%%%%%% An attempt to inline
-% \(
-%  {\LinSigTy{f}{\left(\PiTy{x}{X}{Y(x)}\right)}{\LinPiTy{x}{X}{A\,x\,(f\,x)}} \lto
-% \LinPiTy{x}{X}{\LinSigTy{y}{Y(x)}{A\,x\,y}} \right)}
-% \)
-\(
- \lamblto{b}{\letin{\sigma\,f\,b'}{b}{\withlamb{x}{b'\,.\pi\,x}}} : \LinSigTyLimit{f}{\PiTy{x}{X}{Y(x)}}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
+For any $A : (x : X) \to Y(x) \to L$, the definable function distributing
+conjunction over disjunction
+% Just type
+\(\LinSigTyLimit{f}{\PiTy{x}{X}{Y(x)}}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
 \LinPiTyLimit{x}{X}{\LinSigTyLimit{y}{Y(x)}{A\,x\,y}}
 \)
+% Term and Type
+% \(
+%  \lamblto{b}{\letin{\sigma\,f\,b'}{b}{\withlamb{x}{\sigma\,(f\,x)\,(b'\,.\pi\,x)}}} : \LinSigTyLimit{f}{\PiTy{x}{X}{Y(x)}}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
+% \LinPiTyLimit{x}{X}{\LinSigTyLimit{y}{Y(x)}{A\,x\,y}}
+% \)
 has an inverse.

### 2025-04-08 paper/paper.tex
punctuation + note on finite alphabet and unambiguity

@@
 that specify that this makes the family $\mu F$ into an \emph{initial
-algebra} for the functor $\el(F)$. That is the $\beta$ rule says that
+algebra} for the functor $\el(F)$. That is, the $\beta$ rule says that
 a $\mathsf{fold}$ applied to a $\mathsf{roll}$ is equivalent to
@@
 \cref{lem:unambig-sum}, we have that $\I$ and
-$\CharGram$ --- and thus each literal
-$\literal c$ --- are unambiguous as well.
+$\CharGram$ are unambiguous as well. By the finiteness of the alphabet $\Sigma$,
+we additionally have that each literal $\literal c$ is likewise unambiguous.
 
@@
 generally we define the linear type of traces as in
-\Cref{fig:nfatrace}
+\Cref{fig:nfatrace}.
 

### 2025-04-08 paper/paper.tex
finiteness comment + save a line of space

@@
 \cref{lem:unambig-sum}, we have that $\I$ and
-$\CharGram$ are unambiguous as well. By the finiteness of the alphabet $\Sigma$,
-we additionally have that each literal $\literal c$ is likewise unambiguous.
+$\CharGram$ are unambiguous as well. Using the finiteness of the alphabet
+$\Sigma$ and the unambiguity of $\CharGram$,
+we have that each literal $\literal c$ is likewise unambiguous.
 
@@
 that can be parsed top-down with no
-lookahead\cite{rosenkrantz_properties_1970}. This means we can
+lookahead \cite{rosenkrantz_properties_1970}. This means we can
 implement it simply as an \emph{infinite} state deterministic
@@
   \label{thm:dyck}
-  $Dyck$ and $Parse_{M}$ are strongly equivalent. And therefore we can construct a parser for $\Dyck$.
+  $Dyck$ and $Parse_{M}$ are strongly equivalent, so we may build a parser for $\Dyck$.
 \end{theorem}

### 2025-04-09 paper/paper.tex
phrasing + spacing

@@
 B$ to extend the parser from $\StringGram \lto A \oplus A_{\neg}$ to
-$\StringGram \lto B \oplus A_{\neg}$ but then we also need $B \lto A$ to
+$\StringGram \lto B \oplus A_{\neg}$, but then we also need $B \lto A$ to
 establish that $A_{\neg}$ is disjoint from $B$.
@@
 saw one particular NFA and its corresponding type of traces. More
-generally we define the linear type of traces as in
-\Cref{fig:nfatrace}.
-
-%% The definition of the linear type of traces
-%% through this
-%% NFA, as
-%% presented in \cref{fig:nfainductive}, inspects the concrete structure of the
-%% NFA.\ When generalizing from this example to an arbitrary NFA, we do not have the
-%% luxury of inspecting a concrete definition. Therefore, we abstract over the
-%% specification of an arbitrary NFA and define traces with respect to that
-%% specification.
-
-%% Let the data of an NFA $\N$ be given by the following nonlinear data,
-%% \begin{itemize}
-%% \item $\N.\states$ --- a finite set of states
-%% \item $\N.\init : \N.\states$ --- the initial state
-%% \item $\N.\isAcc : \N.\states \to \Bool$ --- a mapping marking which states are
-%%         accepting
-%% \item $\N.\transitions$ --- a finite set of labeled transitions
-%% \item $\N.\labelt : \N.\transitions \to \Sigma$ --- a mapping from $N.\transitions$ to the character labeling the transition
-%% \item $\N.\src , \N.\dst : \N.\transitions \to \N.\states$ --- mappings from
-%%         $\N.\transitions$ to the source and target states, respectively, of a labeled transition
-%% \item $\N.\epstransitions$ --- a finite set of $\varepsilon$-transitions
-%% \item $\N.\epssrc, \N.\epsdst : \N.\epstransitions \to \N.\states$ --- mappings
-%%         from $\N.\epstransitions$ to the source and target states of an $\epsilon$-transition
-%% \end{itemize}
-
-In \cref{fig:nfatrace} we define a linear type of traces through an
-NFA $\N$. $\Trace_{\N}$ is an inductive type indexed by the starting
-state of the trace $\s : \N.\states$. Traces in $\N$ may be built
+generally, in \cref{fig:nfatrace} we define a linear type of traces through an
+arbitrary NFA $\N$.
+
+$\Trace_{\N}$ is an inductive type indexed by the starting
+state of the trace $\s : \N.\states$, and it may be built
 through one of three constructors. We may terminate a trace at an
@@
 at the destination state of a transition, then we may use the $\cons$
-constructor to linearly combine that trace with a parse of the label
+constructor to combine that trace with a parse of the label
 of the transition to build a trace beginning at the source of the
@@
 along the $\epsilon$-transition and construct a trace beginning at the
-source of the $\epsilon$-transition. As a shorthand, write $Parse_{N}$
+source of the $\epsilon$-transition. As a shorthand, we write $Parse_{N}$
 for the accepting traces out of $N.init$.
@@
 is defined by recursion on strings in \cref{fig:printdfa}. If this string is empty, then $\pparse_{D}$
-defines a linear function that terminates a trace at the input state $s$. If the
+terminates a trace at the input state $s$. If the
 string is nonempty, then $\pparse_{D}$ walks forward in $D$ from the input state $s$ by
@@
 via $\nil$, then $\print_{D}$ returns the empty string. Otherwise, if the trace
-is defined by $\cons$ then $\pparse_{D}$ appends the character from the most
+is defined by $\cons$ then $\print_{D}$ appends the character from the most
 recent transition to the output
@@
 Next, we give two examples for parsing context-free
-grammars. Context-free grammars (CFG) can be encoded in our type
-theory in a similar way to regular expressions, as CFGs are equivalent
+grammars (CFGs). CFGs can be encoded in our type
+theory similarly to regular expressions, as CFGs are equivalent
 to the formalism of \emph{$\mu$}-regular expressions, where the Kleene
@@
 $parse_{M}$ and $print_{M}$ for this automaton can be defined
-analogously to the parser and printer for DFAs, and so $\LinSigTy {s}
+analogously to those for DFAs, and so $\LinSigTy {s}
 {M .states} {\LinSigTy {b}{Bool} {Trace_{M}~s~b}}$ is likewise

### 2025-04-09 paper/paper.tex
no widows/orphans

@@
 \settopmatter{}
+
+\clubpenalty = 10000
+\widowpenalty = 10000
+\displaywidowpenalty = 10000
 
@@
 accepting traces of $D(A)$. Then we can easily construct a parser for
-traces of $D(A)$ and combine this using \Cref{lem:wk-eqv-parse} to get
+traces of $D(A)$ and apply \cref{lem:wk-eqv-parse} to get
 a verified regular expression parser.
@@
 This reification operation on functions $\StringSem \to U$ is incredibly expressive,
-as it allows to sidestep our linear typing connectives and utilize the whole of nonlinear
+as it sidesteps our linear typing connectives and utilizes the whole of nonlinear
 dependent type theory to define a grammar. For example, given a

### 2025-04-10 paper/paper.tex
copyright info filler
include publisher copyright info, journal info, and authors
ccs + keywords

@@
 
-%% Journal information
-%% Supplied to authors by publisher for camera-ready submission;
-%% use defaults for review submission.
+%%% The following is specific to PLDI '25 and the paper
+%%% 'Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus'
+%%% by Steven Schaefer, Nathan Varner, Pedro Henrique Azevedo de Amorim, and Max S. New.
+%%%
+\setcopyright{cc}
+\setcctype{by}
+\acmDOI{10.1145/3729281}
+\acmYear{2025}
 \acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
-\acmArticle{1}
-\acmYear{2018}
-\acmMonth{1}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission;
-%% use 'none' for review submission.
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
+\acmVolume{9}
+\acmNumber{PLDI}
+\acmArticle{178}
+\acmMonth{6}
+\received{2024-11-14}
+\received[accepted]{2025-03-06}
 
@@
 \newcommand{\LinSigTyLimit}[3]{\bigoplus\limits_{#1 : #2} #3}
-%% \newcommand{\DepWith}[2]{{\textstyle\bigamp}\limits_{#1}{#2}
-%% \newcommand{\DepPlus}[2]{{\textstyle\bigoplus}\limits_{#1}{#2}
 \newcommand{\GrTy}{\mathsf{Gr}}
@@
 \author{Steven Schaefer}
-\affiliation{\department{Electrical Engineering and Computer Science}
+\orcid{0009-0007-1258-9501}
+\affiliation{%
   \institution{University of Michigan}
+  \city{Ann Arbor}
   \country{USA}
@@
 \email{stschaef@umich.edu}
-\orcid{0009-0007-1258-9501}
 
 \author{Nathan Varner}
-\affiliation{\department{Electrical Engineering and Computer Science}
+\orcid{0009-0000-3031-4930}
+\affiliation{%
   \institution{University of Michigan}
+  \city{Ann Arbor}
   \country{USA}
@@
 \email{nmvarner@umich.edu}
-\orcid{0009-0000-3031-4930}
-
-\author{Pedro H. Azevedo de Amorim}
-\affiliation{
-  \department{Department of Computer Science}
+
+\author{Pedro Henrique Azevedo de Amorim}
+\orcid{0000-0002-8338-8973}
+\affiliation{%
   \institution{University of Oxford}
-  \country{UK}
+  \city{Oxford}
+  \country{United Kingdom}
 }
-\email{pedro.azevedo.de.amorim@cs.ox.ac.uk}
-\orcid{0000-0002-8338-8973}
+\email{pedro.azevedodeamorim@cs.ox.ac.uk}
 
 \author{Max S. New}
-%% \authornote{with author1 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
 \orcid{0000-0001-8141-195X}
-\affiliation{
-  \position{Assistant Professor}
-  \department{Computer Science and Engineering}              %% \department is recommended
-  \institution{University of Michigan}            %% \institution is required
-  % \streetaddress{2260 Hayward St.}
+\affiliation{%
+  \institution{University of Michigan}
   \city{Ann Arbor}
-  \state{MI}
-  % \postcode{48109-2121}
-  \country{USA}                    %% \country is recommended
+  \country{USA}
 }
-\email{maxsnew@umich.edu} %% \email is recommended
+\email{maxsnew@umich.edu}
 
@@
 %% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10003752.10010124.10010131.10010133</concept_id>
+       <concept_desc>Theory of computation~Denotational semantics</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011050.10011017</concept_id>
+       <concept_desc>Software and its engineering~Domain specific languages</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10003752.10003766.10003771</concept_id>
+       <concept_desc>Theory of computation~Grammars and context-free languages</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011041.10011688</concept_id>
+       <concept_desc>Software and its engineering~Parsers</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011074.10011099.10011692</concept_id>
+       <concept_desc>Software and its engineering~Formal software verification</concept_desc>
+       <concept_significance>100</concept_significance>
+       </concept>
+ </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[300]{Theory of computation~Denotational semantics}
+\ccsdesc[300]{Software and its engineering~Domain specific languages}
+\ccsdesc[500]{Theory of computation~Grammars and context-free languages}
+\ccsdesc[500]{Software and its engineering~Parsers}
+\ccsdesc[100]{Software and its engineering~Formal software verification}
 
@@
 \else
-\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
+\keywords{Linear types, Lambek calculus, Intrinsic verification}  %% \keywords are mandatory in final camera-ready submission
 \fi

### 2025-04-10 paper/paper.tex
citation to code

@@
 as the basis for our prototype implementation of \theoryabbv in
-Cubical Agda. The implementation is a shallow embedding, meaning that
+Cubical Agda \cite{schaefer_2025_15049780}. The implementation is a shallow embedding, meaning that
 rather than formalizing a syntax of \theoryabbv types and terms, we interpret

### 2025-04-10 paper/paper.tex
data availability

@@
 
+\section*{Data-Availability Statement}
+For reproduction, the software supporting the claims in this paper is archived on Zenodo \cite{schaefer_2025_15049780}.
+For reuse, the Agda implementation of \theoryabbv is available \href{https://github.com/maxsnew/grammars-and-semantic-actions/tree/PLDI25Artifact}{on GitHub}.
+
 \bibliography{refs.bib}

### 2025-04-10 paper/paper.tex
a more human data availability statement
clean up the acks

--- 2025-04-10 paper/paper.tex
@@
   This material is based upon work supported by the Air Force Office of
-  Scientific Research under Grant No.~\grantnum{}{FA9550-23-1-0760}.  Any
-  opinions, findings and conclusions or recommendations expressed in this
+  Scientific Research under Grant No.~\grantnum{}{FA9550-23-1-0760}, the ERC
+  Consolidator Grant BLAST and the ARIA programme on Safeguarded AI. Any
+  opinions, findings, and conclusions or recommendations expressed in this
   material are those of the authors and do not necessarily reflect the views of
-  the U.S. Department of Defense.
-  %
-  This work is also supported by the ERC Consolidator Grant BLAST and the ARIA
-  programme on Safeguarded AI.  Any opinions, findings, and conclusions or
-  recommendations expressed in this material are those of the author and do not
-  necessarily reflect the views of the National Science Foundation, the European
-  Research Council or the Advanced Research and Invention Agency.
+  the U.S. Department of Defense, the European Research Council or the Advanced
+  Research and Invention Agency.
 \end{acks}
+
+\section*{Data-Availability Statement}
+The Agda implementation of \theoryabbv and the formalization of our proofs are archived on Zenodo \cite{schaefer_2025_15049780} and the source code is available \href{https://github.com/maxsnew/grammars-and-semantic-actions/tree/PLDI25Artifact}{on GitHub}.
 

### 2025-04-10 paper/paper.tex
Remove the non-proof of thompsons construction for space

@@
 \label{sec:applications}
-This section explores the applications of \theoryabbv to conducting formal
+This section explores the applications of \theoryabbv to formal
 grammar theory. We demonstrate that several classical notions and constructions
@@
   $\N.\states$. A subset is accepting in $\D$ if it contains an accepting state
-  from $\N$. Construct the initial state of $\D$ as the
-  $\epsilon$-closure of $\N.\init$. Lastly, define the transition function
-  of $\D$ to send the subset $X$ under the character $\c$ to the
+  from $\N$. The initial state of $\D$ is the
+  $\epsilon$-closure of $\N.\init$. Lastly, the transition function
+  of $\D$ sends the subset $X$ under the character $\c$ to the
   $\epsilon$-closure of all the states reachable from $X$ via a transition
@@
 \newcommand{\R}{\texttt{R}}
-\begin{construction}[Thompson's Construction, \Agda]
+\begin{construction}[Thompson's Construction \cite{thompsonProgrammingTechniquesRegular1968}, \Agda]
   \label{cons:thompson}
@@
 \end{construction}
-\begin{proof}
-We use a variant of Thompson's construction
-\cite{thompsonProgrammingTechniquesRegular1968}, showing that NFAs are, up to strong equivalence, closed under each type operation for regular expressions.
-\end{proof}
+% \begin{proof}
+% We use a variant of Thompson's construction
+% \cite{thompsonProgrammingTechniquesRegular1968}, showing that NFAs are, up to strong equivalence, closed under each type operation for regular expressions.
+% \end{proof}
 %% In this paper we will walk through

### 2025-04-18 paper/paper.tex
Dependent type keyword
code formatting

@@
 \colorlet{nonlintycolor}{RubineRed}
-\colorlet{lintycolor}{Green}
+\colorlet{lintycolor}{PineGreen}
 \colorlet{nonlinconstructorcolor}{nonlintycolor}
@@
 \colorlet{alphabetcolor}{black}
-\colorlet{funarrowcolor}{black}
+\colorlet{funarrowcolor}{lintycolor}
 \definecolor{backcolour}{rgb}{0.95,0.95,0.92}
+
+\newcommand{\uparrowcode}{\color{nonlinconstructorcolor} \uparrow}
+\newcommand{\tocode}{\color{nonlinconstructorcolor} \to}
+\newcommand{\tocodelin}{\color{linconstructorcolor} \to}
 
@@
 \lstdefinelanguage{Agda}{
-alsoletter={012=\&},
+alsoletter={0123\&'},
 keywords={data, where, module, import, open, public,
@@
 morekeywords=[2]{L, Bool, Nat, Fin},
-morekeywords=[3]{String, Char, Dyck, Trace, Exp, Atom, O, D, C, A, NotStartsWithLP, NotStartsWithRP},
-morekeywords=[4]{bal, .nil, .cons, cons, nil, inl, inr, stop, 1to1, 1to2, 0to2, 0to1},
+morekeywords=[3]{String, Dyck, Trace, Exp, Atom, O, D, C, A, NotStartsWithLP, NotStartsWithRP},
+morekeywords=[4]{bal, .nil, .cons, cons, nil, inl, inr, stop, 1to1, 1to2, 0to2, 0to1,
+  done, add, num, parens, left, unexpected, lookAheadRP, lookAheadNot, closeGood, closeBad,
+  doneGood, doneBad},
 morekeywords=[5]{.states, states, .isAcc, isAcc, transitions, src, dst, .transitions, .dst,
-  .src, label},
+  .src, .label, label, 0 , 1 , 2, 3, true, false},
 morekeywords=[6]{NUM},
-morekeywords=[7]{= , \&},
+morekeywords=[7]{\&, '},
 sensitive=true,
@@
 mathescape=true,
+moredelim={[is][]{\#}{\#}},
 escapeinside={(*@}{@*)},
@@
 \else
-\keywords{Linear types, Lambek calculus, Intrinsic verification}  %% \keywords are mandatory in final camera-ready submission
+\keywords{Linear types, Dependent Types, Lambek calculus, Intrinsic verification}  %% \keywords are mandatory in final camera-ready submission
 \fi
@@
 \begin{lstlisting}
-f : (*@$\uparrow$@*)('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)' (*@$\lto$@*) ('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\oplus$@*) '(*@\color{alphabetcolor}c@*)')
+f : (*@$\uparrowcode$@*)('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)' (*@$\lto$@*) ('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\oplus$@*) '(*@\color{alphabetcolor}c@*)')
 f (a , b) = inl (a , b)
@@
 \end{mathpar}
-\caption{(\Agda) \stringquote{ab} is parsed by $(\literal a \oplus \literal b) \otimes \literal c$
+\caption{(\Agda) \stringquote{ab} is parsed by $(\literal a \otimes \literal b) \oplus \literal c$
 }
@@
 data A(*@$^*$@*) : L where
-  nil : (*@$\uparrow$@*)(A(*@$^*$@*))
-  cons : (*@$\uparrow$@*)(A (*@$\lto$@*) A(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
+  nil : (*@$\uparrowcode$@*)(A(*@$^*$@*))
+  cons : (*@$\uparrowcode$@*)(A (*@$\lto$@*) A(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
 \end{lstlisting}
@@
 \begin{lstlisting}
-g : (*@$\uparrow$@*)(('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\lto$@*) ('(*@\color{alphabetcolor}a@*)'(*@$^*$@*) (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\oplus$@*) '(*@\color{alphabetcolor}c@*)')
+g : (*@$\uparrowcode$@*)(('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\lto$@*) ('(*@\color{alphabetcolor}a@*)'(*@$^*$@*) (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\oplus$@*) '(*@\color{alphabetcolor}c@*)')
 g (a , b) = inl (cons a nil , b)
@@
 \begin{lstlisting}
-h : (*@$\uparrow$@*)((A (*@$\otimes$@*) A)(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
+h : (*@$\uparrowcode$@*)((A (*@$\otimes$@*) A)(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
 h nil = nil
@@
   \begin{lstlisting}
-data Trace : (s : Fin 3) (*@$\to$@*) L where
-  stop : (*@$\uparrow$@*)(Trace 2)
-  1to1 : (*@$\uparrow$@*)('(*@\color{alphabetcolor}a@*)' (*@$\lto$@*) Trace 1 (*@$\lto$@*) Trace 1)
-  1to2 : (*@$\uparrow$@*)('(*@\color{alphabetcolor}b@*)' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 1)
-  0to2 : (*@$\uparrow$@*)('(*@\color{alphabetcolor}c@*)' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 0)
-  0to1 : (*@$\uparrow$@*)(Trace 1 (*@$\lto$@*) Trace 0)
-
-k : (*@$\uparrow$@*)(('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\lto$@*)  Trace 0)
+data Trace : (s : Fin 3) (*@$\tocode$@*) L where
+  stop : (*@$\uparrowcode$@*)(Trace 2)
+  1to1 : (*@$\uparrowcode$@*)('(*@\color{alphabetcolor}a@*)' (*@$\lto$@*) Trace 1 (*@$\lto$@*) Trace 1)
+  1to2 : (*@$\uparrowcode$@*)('(*@\color{alphabetcolor}b@*)' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 1)
+  0to2 : (*@$\uparrowcode$@*)('(*@\color{alphabetcolor}c@*)' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 0)
+  0to1 : (*@$\uparrowcode$@*)(Trace 1 (*@$\lto$@*) Trace 0)
+
+k : (*@$\uparrowcode$@*)(('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\lto$@*)  Trace 0)
 k (a , b) = 0to1 (1to1 a (1to2 b stop))
@@
 \begin{lstlisting}
-data Trace(*@$_{\color{black}\N}$@*) : (s : N .states) (*@$\to$@*) L where
-  nil : (*@$\uparrow$@*)(&[ s : N .states ] N .isAcc s (*@$\to$@*) Trace(*@$_{\color{black}\N}$@*) s)
-  cons : (*@$\uparrow$@*)(&[ t : N .transitions ]
-    ('N .label t' (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .dst t) (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .src t)))
-  (*@$\color{linconstructorcolor}\epsilon$@*)cons : (*@$\uparrow$@*)(&[ t : N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)transitions ]
-    (Trace(*@$_{\color{black}\N}$@*) (N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)dst t) (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)src t)))
-
-data Trace(*@$_{\color{black}\D}$@*) : (s : (*@\color{black}D@*) .states) (b : Bool) (*@$\to$@*) L where
-  nil : (*@$\uparrow$@*)(&[ s : (*@\color{black}D@*) .states ] Trace(*@$_{\color{black}\D}$@*) s (*@\color{black}D@*) .isAcc s)
-  cons : (*@$\uparrow$@*)(&[ c : Char ] &[ s : (*@\color{black}D@*) .states ] &[ b : Bool ]
-  ('(*@\color{alphabetcolor}c@*)' (*@$\lto$@*) Trace(*@$_{\color{black}\D}$@*) ((*@\color{black}D@*) .(*@$\color{nonlinconstructorcolor}\delta$@*) c s) b (*@$\lto$@*) Trace(*@$_{\color{black}\D}$@*) s b))
+data Trace(*@$_{\color{black}\N}$@*) : (s : N .states) (*@$\tocode$@*) L where
+  nil : (*@$\uparrowcode$@*)(&[ s : N .states ] N .isAcc s (*@$\tocodelin$@*) Trace(*@$_{\color{black}\N}$@*) s)
+  cons : (*@$\uparrowcode$@*)(&[ t : N .transitions ] ('#N .label t#' (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .dst t)
+                                                 (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .src t)))
+  (*@$\color{linconstructorcolor}\epsilon$@*)cons : (*@$\uparrowcode$@*)(&[ t : N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)transitions ] (Trace(*@$_{\color{black}\N}$@*) (N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)dst t) (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)src t)))
+
+data Trace(*@$_{\color{black}\D}$@*) : (s : (*@\color{black}D@*) .states) (b : Bool) (*@$\tocode$@*) L where
+  nil : (*@$\uparrowcode$@*)(&[ s : (*@\color{black}D@*) .states ] Trace(*@$_{\color{black}\D}$@*) s (*@\color{black}D@*) .isAcc s)
+  cons : (*@$\uparrowcode$@*)(&[ c : (*@$\color{nonlintycolor}\Sigma$@*) ] &[ s : (*@\color{black}D@*) .states ]
+           &[ b : Bool ] ('(*@\color{alphabetcolor}c@*)' (*@$\lto$@*) Trace(*@$_{\color{black}\D}$@*) ((*@\color{black}D@*) .(*@$\color{nonlinconstructorcolor}\delta$@*) c s) b (*@$\lto$@*) Trace(*@$_{\color{black}\D}$@*) s b))
 \end{lstlisting}
@@
 \begin{lstlisting}
-parse(*@$\color{black}_D$@*) : (*@$\uparrow$@*)(String (*@$\lto$@*) &[ s : (*@\color{black}D@*) .states ] (*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\color{black}\D}$@*) s b)
+parse(*@$\color{black}_D$@*) : (*@$\uparrowcode$@*)(String (*@$\lto$@*) &[ s : (*@\color{black}D@*) .states ] (*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\color{black}\D}$@*) s b)
 parse(*@$\color{black}_D$@*) String .nil s = (*@$\color{linconstructorcolor}\sigma$@*) ((*@\color{black}D@*) .isAcc s) (Trace(*@$_{\color{black}\D}$@*) .nil s)
@@
 
-print(*@$\color{black}_D$@*) : (s : (*@\color{black}D@*) .states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\color{black}\D}$@*) s b) (*@$\lto$@*) String)
+print(*@$\color{black}_D$@*) : (s : (*@\color{black}D@*) .states) (*@$\tocode$@*) (*@$\uparrowcode$@*)(((*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\color{black}\D}$@*) s b) (*@$\lto$@*) String)
 print(*@$\color{black}_D$@*) s ((*@$\color{linconstructorcolor}\sigma$@*) b (Trace(*@$_{\color{black}\D}$@*) .nil .s)) = String .nil
@@
 data Dyck : L where
-  nil : (*@$\uparrow$@*) Dyck
-  bal : (*@$\uparrow$@*)('(' (*@$\lto$@*) Dyck (*@$\lto$@*) ')' (*@$\lto$@*) Dyck (*@$\lto$@*) Dyck)
+  nil : (*@$\uparrowcode$@*) Dyck
+  bal : (*@$\uparrowcode$@*)('(' (*@$\lto$@*) Dyck (*@$\lto$@*) ')' (*@$\lto$@*) Dyck (*@$\lto$@*) Dyck)
 \end{lstlisting}
@@
 data Exp : L where
-  done : (*@$\uparrow$@*)(Atom (*@$\lto$@*) Exp)
-  add : (*@$\uparrow$@*)(Atom (*@$\lto$@*) '+' (*@$\lto$@*) Exp (*@$\lto$@*) Exp)
+  done : (*@$\uparrowcode$@*)(Atom (*@$\lto$@*) Exp)
+  add : (*@$\uparrowcode$@*)(Atom (*@$\lto$@*) '+' (*@$\lto$@*) Exp (*@$\lto$@*) Exp)
 data Atom : L where
-  num : (*@$\uparrow$@*)('NUM' (*@$\lto$@*) Atom)
-  parens : (*@$\uparrow$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ')' (*@$\lto$@*) Atom)
-
-data O : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
-  left : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O (n + 1) b (*@$\lto$@*) O n b)
-  num : (*@$\uparrow$@*)(&[ n ] &[ b ] 'NUM' (*@$\lto$@*) D n b (*@$\lto$@*) O n b)
-  unexpected : (*@$\uparrow$@*)(&[ n ] (NotStartsWithLP (*@$\lto$@*) O n false)
-data D : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
-  lookAheadRP : (*@$\uparrow$@*)(&[ n ] &[ b ] ((')' (*@$\otimes$@*) (*@$\color{lintycolor}\top$@*)) (*@$\&$@*) C n b) (*@$\lto$@*) D n b)
-  lookAheadNot : (*@$\uparrow$@*)(&[ n ] &[ b ] (NotStartsWithRP (*@$\&$@*) A n b) (*@$\lto$@*) D n b)
-data C : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
-  closeGood : (*@$\uparrow$@*)(&[ n ] &[ b ] ')' (*@$\lto$@*) D n b (*@$\lto$@*) C (n + 1) b)
-  closeBad : (*@$\uparrow$@*)(')' (*@$\lto$@*) C 0 false)
-  unexpected : (*@$\uparrow$@*)(&[ n ] NotStartsWithRP (*@$\lto$@*) C n false)
-data A : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
-  doneGood : (*@$\uparrow$@*)(A 0 true)
-  doneBad : (*@$\uparrow$@*)(&[ n ] A (n + 1) false)
-  add : (*@$\uparrow$@*)(&[ n ] &[ b ] '+' (*@$\lto$@*) O n b (*@$\lto$@*) A n b)
-  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')' (*@$\oplus$@*) 'NUM') (*@$\lto$@*) (*@$\color{lintycolor} \top$@*) (*@$\lto$@*) A n false)
+  num : (*@$\uparrowcode$@*)('#NUM#' (*@$\lto$@*) Atom)
+  parens : (*@$\uparrowcode$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ')' (*@$\lto$@*) Atom)
+
+data O : Nat (*@$\tocode$@*) Bool (*@$\tocode$@*) L where
+  left : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] '(' (*@$\lto$@*) O (n + 1) b (*@$\lto$@*) O n b)
+  num : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] '#NUM#' (*@$\lto$@*) D n b (*@$\lto$@*) O n b)
+  unexpected : (*@$\uparrowcode$@*)(&[ n : Nat ] (NotStartsWithLP (*@$\lto$@*) O n false)
+data D : Nat (*@$\tocode$@*) Bool (*@$\tocode$@*) L where
+  lookAheadRP : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] ((')' (*@$\otimes$@*) (*@$\color{lintycolor}\top$@*)) (*@$\&$@*) C n b) (*@$\lto$@*) D n b)
+  lookAheadNot : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] (NotStartsWithRP (*@$\&$@*) A n b) (*@$\lto$@*) D n b)
+data C : Nat (*@$\tocode$@*) Bool (*@$\tocode$@*) L where
+  closeGood : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] ')' (*@$\lto$@*) D n b (*@$\lto$@*) C (n + 1) b)
+  closeBad : (*@$\uparrowcode$@*)(')' (*@$\lto$@*) C 0 false)
+  unexpected : (*@$\uparrowcode$@*)(&[ n : Nat ] NotStartsWithRP (*@$\lto$@*) C n false)
+data A : Nat (*@$\tocode$@*) Bool (*@$\tocode$@*) L where
+  doneGood : (*@$\uparrowcode$@*)(A 0 true)
+  doneBad : (*@$\uparrowcode$@*)(&[ n : Nat ] A (n + 1) false)
+  add : (*@$\uparrowcode$@*)(&[ n : Nat ] &[ b : Bool ] '+' (*@$\lto$@*) O n b (*@$\lto$@*) A n b)
+  unexpected : (*@$\uparrowcode$@*)(&[ n : Nat ] ('(' (*@$\oplus$@*) ')' (*@$\oplus$@*) '#NUM#') (*@$\lto$@*) (*@$\color{lintycolor} \top$@*) (*@$\lto$@*) A n false)
 \end{lstlisting}

### 2025-04-18 paper/paper.tex
Keyword case
Remove comments of drafts that did not make it into final version of paper

@@
 \else
-\keywords{Linear types, Dependent Types, Lambek calculus, Intrinsic verification}  %% \keywords are mandatory in final camera-ready submission
+\keywords{Linear types, Dependent types, Lambek calculus, Intrinsic verification}  %% \keywords are mandatory in final camera-ready submission
 \fi
@@
 \theoryabbv. The core idea of the denotational semantics stems from an observation of Elliott:
-% The core idea of the denotational semantics is
-% \cite{elliottSymbolicAutomaticDifferentiation2021}, Elliott describes
 a formal grammar is a type-level predicate on strings that proves
@@
 
-%% Formal verification encompasses a broad suite of techniques one may use to
-%% mathematically guarantee the correctness of software systems. We can only
-%% reap the benefits of
-%% any proof effort if its underlying assumptions hold. For instance, verifying the
-%% functional correctness of an algorithm is no use if it is run on buggy hardware.
-%% Likewise, the promised guarantees of a program may be jeopardized if it is
-%% parsed incorrectly.
-
-Our contributions are then:% \max{This contribution section needs to be punchier}
+Our contributions are then:
 %
@@
 \).
-% otherwise we would have that $\inl(x\otimes y)$ is a valid
-% term in context $y:b,x:a$, but there should be no parses of $ba$ in
-% this grammar.
 
@@
 
-%% 3. Third example: NFA traces to show the indexed inductive types
-
 \section{Syntax and Typing for \theoryname}
 \label{sec:tt}
-
-%% OUTLINE
-%%
-%% 1. Overview of syntax: what are the judgments
-%% 2. Cover the linear types except for inductives
-%% 3. Cover non-linear types, mainly focus on the universes
-%% 4. Cover indexed inductive linear types
-%% 5. Cover the "axioms" we add
 
@@
 \cref{ax:dist}.
-% Comment from shepherd
-% - The pagebreaks and figures make the explanation of axiom 3.1 a
-% little confusing to read (with the text "In its most general form,"
-% separated from the axiom it is intended to introduce by a pagebreak
-% and a figure). You could fiddle with the layout a bit, or just revise
-% that sentence to "More generally, we assume Axiom 3.1.", or similar.
-%
-% - If I understand it correctly, axiom 3.1 itself is using "e" as a variable name, which is confusing since it's also used as a metavariable for terms - I suggest picking a different variable name.
+
 \begin{axiom}[Distributivity]
@@
 unambiguous sums have disjoint summands.
-
-% \pedro{imo, since we are providing an Agda implementation, we can
-%   be looser with the explanation of the meaning of the disjointness axioms.
-%   As it stands, it's a bit hard do follow along. Is there a simpler example
-%   we could point to?}
 
@@
   \end{footnotesize}
-  %% \begin{footnotesize}
-  %%   \begin{align*}
-  %%   \mathsf{Var} &: \prod_{X:U} X \to \SPF\,X\\
-  %%   \mathsf{K} &: \prod_{X:U} L \to \SPF\,X\\
-  %%   \mathsf{\bigoplus} &: \prod_{X:U}\prod_{Y:U}(Y \to \SPF\,X) \to \SPF\,X\\
-  %%   \mathsf{\bigamp} &: \prod_{X:U}\prod_{Y:U}(Y \to \SPF\,X) \to \SPF\,X\\
-  %%   \mathsf{\otimes} &: \prod_{X:U}\SPF\,X \to \SPF\,X \to \SPF\,X\\
-  %%   \el &: \prod_{X:U}\SPF\,X \to (X \to A) \to L\\
-  %%   \map &: \prod_{X:U}\prod_{F : \SPF\,X}\prod_{A,B:X\to L}{\left(\prod_{x:X}\uparrow(\unquoteTy{A\,x}\lto \unquoteTy{B\,x})\right)} \to {\uparrow(\unquoteTy{\el(F)(A)} \lto \unquoteTy{\el(F)(B)})}\\\\
-  %%   \roll &: \prod_{X:U}\prod_{F:X \to \SPF\,X}\prod_{x:X}\ltonl{(\el(F\,x)(\mu\,F))}\\
-  %%   \fold &: \prod_{X:U}\prod_{F:X\to\SPF\,X}\prod_{A:X \to L}
-  %%   \left(\prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)}\right)
-  %%   \to \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}
-  %% \end{align*}
-  %% \end{footnotesize}
   \caption{Strictly positive functors and indexed inductive linear types}
@@
 
-% \steven{TODO the intro to this section is to be condensed based on what is
-%   included in the examples section as motivation and intuition building}
-
-% Omission of the structural rules of a deductive system, such as in
-% linear logic \cite{GIRARD19871}, offers precise control over how a value is used
-% in a derivation. Linear logic omits the weakening and contraction rules
-% to ensure that every value in context is used exactly once. This control enables
-% \emph{resource-sensitive} reasoning, where we may treat a resource as
-% \emph{consumed} after usage. This viewpoint is amenable to parsing applications,
-% as we may treat characters of a string as finite resources that are consumed at
-% parse-time.
-That is, when reading a string, the occurrence of any character is read once.
-Freely duplicating or dropping characters from a string changes the meaning of that string.
-\max{what about a backtracking parser?}.
-One may then envision a linear type system where the types comprise
-% formal grammars generated over some alphabet $\Sigma$, and the type constructors
-% correspond precisely to inductive constructions on grammars --- such as
-% conjunction, disjunction, concatenation, etc.
-
-% Programming in a purely linear term language is limiting due to the variable
-% usage restrictions. Code in such a language can
-% become unnecessarily verbose when ensuring the linearity invariant.
-% To alleviate this
-% concern, in 1995 Benton et al.\ proposed an alternative categorical
-% semantics of linear logic based on adjoint interactions between linear
-% and non-linear logics \cite{bentonMixedLinearNonlinear1995} ---
-% appropriately referred to as a \emph{linear-non-linear} (LNL)
-% system. This work is simply typed, so the boundary between linear and
-% non-linear subtheories is entirely characterized via a monoidal
-% adjunction between linear terms and non-linear terms.
-
-% \steven{Describe this monoidal adjunction with $\ltonl A$ and the derivable
-%   version of $F$}
-
-% Inside of an LNL system, linearity may be thought of as an option that users can
-% choose to deliberately invoke at deliberate points in their developments in an
-% otherwise intuitionistic system. However, if we are wishing to treat parsers as
-% linear terms over input strings, the non-linear fragment of an LNL theory does
-% not really assist in the development of parsers. It is instead the case that
-% parsers may benefit from a \emph{dependence} on non-linear terms.
-% Through the approach described by Krishnaswami et al.\ in
-% \cite{krishnaswami_integrating_2015},
-% we may define a restricted form of dependent types.
-
-% In particular, dependence
-% on linear terms
-% is disallowed; however, through dependence of a linear term on a non-linear
-% index, we can recover the definition of Aho's indexed grammars \cite{AhoIndexed}
-% internal to \theoryabbv. Although, we can define strictly more things, as Aho's
-% indexed grammars provide restricted access to the index whereas we can do
-% whatever with it.
-
-% \steven{elaborate on indexed grammars in a more articulate way in the last sentence}
-% \steven{Put the aho stuff in the subtheory section}
-%
-
-% \paragraph{Regular Expressions}
-% We may realize regular expression as a subtheory
-% of our language by restricting the type constructors to the
-% linear unit, characters, $\otimes$, $\oplus$, and Kleene star. Classically, the
-% languages recognized by these are exactly the regular languages --- the lowest
-% on the Chomsky hierarchy.
-
-% \paragraph{$\mu$-Regular Expressions} Similarly, we may restrict the connectives
-% to be linear unit, characters, $\otimes$, $\oplus$, and arbitrary recursion via
-% left-fixed point rather than only Kleene star.
-% Instead of regular expressions these correspond to Lei{\ss}'s $\mu$-regular
-% expressions, which are known to be equivalent to context-free grammars
-% \cite{leiss,krishnaswami_typed_2019}.
-
-% \paragraph{Beyond Context-Free Grammars} The previous two subtheories induce as semantics
-% regular grammars and context-free grammars, respectively; however, by including
-% the LNL dependent types we may actually express the entirety of the Chomsky
-% hierarchy. Through use of the LNL dependent types, we may encode indexed
-% grammars, which are known to be properly between context-free and
-% context-sensitive grammars within the Chomsky hierarchy \cite{AhoIndexed}. We
-% will further see in \cref{subsubsec:tm} how to use dependence to encode Turing machines internal to our
-% calculus, which of course induces a semantics of unrestricted grammars.
-
-% \paragraph{Indexed Grammars}
-% The LNL dependent types can be used to encode Aho's indexed grammars
-% \cite{AhoIndexed} internal to \theoryabbv. Although, we can define strictly more
-% things with this non-linear index, as we have unrestricted access to the index
-% whereas Aho's grammars can only manipulate it in a particular way.
-
-% \steven{Elaborate on these, and note that they are bw context free and context
-%   sensitive on the Chomsky hierarchy}
-
 \section{Formal Grammar Theory in \theoryname}
@@
 
-%% That is, grammars are weakly equivalent if we can construct a parse
-%% tree of $B$ from a parse tree of $A$ and vice-versa, whereas they are
-%% strongly equivalent if these processes are inverse to each other. The
-%% definition of retract is a useful intermediate notion. In retrospect,
-%% our axiom about $\StringGram$ says that $\StringGram$ is a retract of $\top$.
-
-%% \newcommand{\lang}{\texttt{Lang}}
-%% A parse transformer $\f : \ltonl{(\A \lto \B)}$ takes as input parses of $\A$ for string
-%% $\w$ and returns a parse of $\B$ for $\w$. Therefore, $\f$ demonstrates an
-%% inclusion of languages $\lang(\A) \subset \lang(\B)$. Similarly, the existence of
-%% $\g$ shows that $\lang(\B) \subset \lang(\A)$, and thus $\lang(\A) = \lang(\B)$.
-
-%% \steven{Need better notation for language that is cohesive throughout the paper}
-
-%% \begin{definition}
-%%   \label{def:strongequiv}
-%%   Grammars $\A$ and $\B$ are \emph{strongly equivalent} if there exists parse
-%%   transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B \lto \A)}$ such
-%%   that $\f \circ \g \equiv \ident$ and $\g \circ \f \equiv \ident$.
-%% \end{definition}
-
-%% For a string $\w$, \cref{def:strongequiv} induces a bijection between the
-%% $\A$-parses of $\w$ and the $\B$-parses of $\w$, thus demonstrating that $\A$
-%% and $\B$ agree in their strong generative capacity.
-
-%% \paragraph{Ambiguity}
 A formal grammar $\A$ is ambiguous if there are multiple parse trees
@@
 
-% \paragraph{A Grammar of Strings}
-% Define the character grammar $\CharGram$ to be the disjunction over characters in the alphabet $\Sigma$,
-% $\CharGram = \LinSigTy {c} {\Sigma} c$. Then define a linear type of strings as a
-% linear list of characters,
-% $\StringGram = \CharGram^{*}$.
-
-% In the grammars model, we can build a
-% strong equivalence between $\top$ and $\StringGram hen building parsers, we may wish to enforce this
-% syntactically by including $\top \cong \StringGram$ as an axiom.
-
-% Through inclusion of this axiom, $! : \StringGram \to \top$ is an isomorphism, and
-% thus also a monomorphism. That is, $\StringGram$ is unambiguous.
-% \max{we should have a section of the previous section which gives the axioms, this one and the distributivity axiom}
-
-% \paragraph{Parseability}
-% We say $A$ is \emph{parseable} if there exists a grammar $B$ and a map
-% $\top \to A \oplus B$. Taking $\top \cong \StringGram$, a grammar $A$ is parseable
-% exactly when we may read an arbitrary string into a parse of $A$ or a $B$-valued
-% refutation.
-
-% \paragraph{Decidability}
-% In the case that $A$ is parseable with respect to $\neg A$ --- i.e.\
-% there's a map $\top \to A \oplus \neg A$ --- then call $A$ \emph{decidable}.
-
-%% \paragraph{Partial and Total Parsers}
-
 We now turn to our main task, which is using our linear type system to
@@
 \end{figure}
-
-%% \paragraph{Deterministic Finite Automata}
-%% Just as the above defines an NFA, we provide a specification for an arbitrary
-%% deterministic finite automaton.\ A DFA $D$ is
-%% given by the following data,
-
-%% \begin{itemize}
-%% \item $\D.\states$ --- a finite set of states
-%% \item $\D.\init : \D.\states$ --- the initial state
-%%   \item $\D.\isAcc : \D.\states \to \Bool$ --- a mapping marking which states
-%%         are accepting
-%% \item $\D.\delta : \Sigma \to \D.\states \to \D.\states$ --- a deterministic transition function. Given a character $\c$ and a state
-%%         $\s$, $\D.\delta~\c~\s$ is the state for which there is a transition $\s \overset{\stringquote{c}}{\to} (\D.\delta~\c~\s)$
-%% \end{itemize}
-
-%% \begin{figure}
-
-%% \caption{Traces through the DFA $D$ as an indexed inductive type}
-%% \label{fig:dfatrace}
-%% \end{figure}
 
@@
 
-%% \begin{figure}
-%% \begin{lstlisting}
-%% NtoD : (s : N.states) (*@$\to$@*)
-%%        (*@$\uparrow$@*)(Trace(*@$_{\N}$@*) s true (*@$\lto$@*)
-%%            &[ X : D.states ] &[ sInX : (X (*@$\ni$@*) s) ] Trace(*@$_{\D}$@*) X true)
-%% NtoD .s (nil s) X sInX = nil s
-%% NtoD .(N.src t) (N.cons trans .true a dstTrace) X sInX =
-%%   D.cons (N.label t) X true a
-%%     (((NtoD s dstTrace).(*@$\pi$@*) (D.(*@$\delta$@*) (N.label t) X)).(*@$\pi$@*) (N.dst t))
-%% NtoD .(N.(*@$\epsilon$@*)src t) ((*@$\epsilon$@*)cons (*@$\epsilon$@*)trans .true (*@$\epsilon$@*)DstTrace) X sInX =
-%%   ((NtoD s (*@$\epsilon$@*)DstTrace).(*@$\pi$@*) X).(*@$\pi$@*) (N.(*@$\epsilon$@*)dst t)
-%% \end{lstlisting}
-%% \caption{Parse transformer from NFA traces to Powerset DFA traces}
-%% \label{fig:NtoD}
-%% \end{figure}
-
-%% Because the state $\s$ is now arbitrary, we can define $\NtoD$ via
-%% recursion on the trace through the NFA. Code for this is given in \cref{fig:NtoD}.
-%% If the NFA trace is terminating at the accepting state $\s$ and $\s \in \X$,
-%% then we return a trace in $\D$ that terminates at
-%% $\X$, which is accepting because it contains the accepting state $\s$.
-
-%% If the NFA trace is defined via $\cons$ for a labeled transition
-%% $t : N.transitions$ such that $N.src~t \in \X$, then
-%% we first decompose the trace into a parse of $N.label~t$ and a
-%% trace starting from $N.dst~t$. We then recursively call $\NtoD$
-%% on this trace which returns a family of traces parameterized by $\epsilon$-closed
-%% subsets containing $N.\dst~t$. We instantiate this family at the
-%% set of states
-%% $\D.\delta~(N.label~t)~X$. Because $\N.\dst~t$ is
-%% reachable from $\N.\src~\texttt{trans}$ by a single transition labeled by
-%% $N.label~t$ and $N.src~t \in X$ we have that
-%% $N.dst~t \in \D.\delta~(N.label~t)~X$. Therefore
-%% we can linearly combine the parse of the label and the trace built from the
-%% recursive call to get a trace in $\D$ rooted at $X$.
-
-%% Lastly, if the NFA trace is defined via $\epsilon\cons$ for an $\epsilon$-transition
-%% $t : \N.\epstransitions$ such that
-%% $\N.\epssrc~t \in \X$, then the construction is quite
-%% similar. In this case we again recursively call $\NtoD$, except afterwards we
-%% instantiate the family of grammars at the set $X$ itself. We may do this because
-%% all states in $\D$ are taken to be $\epsilon$-closed, therefore
-%% $\N.\epsdst~t \in \X$ because
-%% $\N.\epssrc~t$ is in $X$. This concludes the construction of $\NtoD$.
-
 To construct a term from DFA traces to NFA traces, we similarly strengthen our
@@
 \end{proof}
-%% For a moment, let us turn our attention to the content behind the traces in $\D$.
-%% Traces from subset $\X$ to subset $\Y$ describes the \emph{existence} of some trace
-%% in the $\N$ from some $\x \in \X$ to some $\y \in \Y$. In general it is not clear how
-%% to extract out a choice of trace in $\N$, which is what $\DtoN$ is asking of us.
-%% However, we may further take the states and transitions of $\N$ to be \emph{ordered}.
-%% This allows us to
-%% induce an order on traces in $\D$ from $\X$ to $\Y$. From these orderings we can then
-%% define the necessary choice functions on each type by taking the smallest with
-%% respect to the ordering.
-
-%% We define $\DtoN$ by recursion on traces through $\D$. First if the trace stops at
-%% some accepting $\epsilon$-closed subset $\X$, then we may choose some state
-%% $\s \in X$ such that $\s$ is accepting and terminate the trace in $\N$ at $\s$.
-
-%% Next if the trace in $\D$ is defined via transition
-%% $\X \overset{\stringquote{c}}{\to} (\D.\delta~\c~\X)$ then we may choose some state in
-%% $\s \in \D.\delta~\c~\X$. Recursively, $\DtoN$ sends the trace rooted at
-%% $\D.\delta~\c~\X$ to one rooted at $\s$. By definition of $\D.\delta~\c~\X$,
-%% there exists some $\s' \in \X$ such that $\s'$ transitions to $\s$ by some number of
-%% $\epsilon$-transitions followed by a transition labeled by $\c$.
-%% We order all
-%% such traces of this form and choose the smallest one, which is then linearly
-%% combined with the parse of $\c$ to build an $\N$ trace. This concludes the
-%% definition of $\DtoN$, thus showing that $Parse_{N}$ and $Parse_{D}$ are
-%% weakly equivalent.
 
@@
 \end{construction}
-% \begin{proof}
-% We use a variant of Thompson's construction
-% \cite{thompsonProgrammingTechniquesRegular1968}, showing that NFAs are, up to strong equivalence, closed under each type operation for regular expressions.
-% \end{proof}
-%% In this paper we will walk through
-%% some of the cases of this proof. The remaining cases
-%% follow
-%% similar reasoning and the details may be found in the formalized artifact.
-
-%% \begin{figure}
-%%   \begin{tikzpicture}[node distance = 17mm ]
-%%     \node[state, initial] (0) {$\0$};
-%%     \node[state, right of=0, accepting] (1) {$\1$};
-
-%%     \path[->] (0) edge[above] node{$\stringquote{c}$} (1);
-%%   \end{tikzpicture}
-%%   \caption{NFA $\N_{\c}$ for a single character $\c$}
-%%   \label{fig:litnfa}
-%% \end{figure}
-
-%% First handle the case where $R$ is a literal character $\literal c$. The NFA $N_{c}$ pictured in
-%% \cref{fig:litnfa} is strongly equivalent to the grammar $c$. To demonstrate this
-%% strong equivalence, we build functions between $\literal c$ and $Parse_{N_{c}}$
-%% and show they are mutually inverse.
-
-%% \begin{figure}
-%% \begin{lstlisting}
-%% A : (s : Fin 2) (*@$\to$@*) L
-%% A 0 = 'c'
-%% A 1 = I
-
-%% toN(*@$_c$@*) : (s : Fin 2) (*@$\to$@*) (*@$\uparrow$@*)(A s (*@$\lto$@*) Trace(*@$_{N_c}$@*) s true)
-%% toN(*@$_c$@*) 0 c = cons t true c (nil 1)
-%% toN(*@$_c$@*) 1 () = nil 1
-
-%% fromN(*@$_c$@*) : (s : Fin 2) (*@$\to$@*) (*@$\uparrow$@*)(Trace(*@$_{N_c}$@*) s true (*@$\lto$@*) A s)
-%% fromN(*@$_c$@*) 0 (cons t .true c tr) = let a (*@$\otimes$@*) () = c (*@$\otimes$@*) fromN(*@$_c$@*) 1 tr in a
-%% fromN(*@$_c$@*) 1 (nil .1) = ()
-%% \end{lstlisting}
-%% \caption{Parse transformer out of $Trace_{N_{c}}~s~true$}
-%% \label{fig:litNFAterms}
-%% \end{figure}
-
-%% In \cref{fig:litNFAterms} we define a family of grammars $A$ that serve as
-%% interpretations for each state in $N_{c}$. We then prove that each $A~s$ is
-%% equivalent to the accepting traces through $N_{c}$ out of state $s$.
-
-%% When
-%% defining $fromN_{c}$ we pattern match on the accepting traces beginning in each
-%% state. Out of state $0$ the only accepting traces are built via $cons$ over the
-%% single transition $t$, and out
-%% of state $1$ the only accepting trace is built via $nil$.
-
-%% Each $A~s$ is unambiguous, so $(fromN_{c}~s) \circ (toN_{c}~s)$ is equal to the
-%% identity on each $A~s$. To complete the proof of strong equivalence it remains
-%% to show that $(toN_{c}~s) \circ (fromN_{c}~s)$ is equal to the identity on
-%% $Trace~s~true$, which we show by using the equalizer type.
-
-%% \begin{figure}
-%% \begin{lstlisting}
-%% eqHom(*@$_{N_c}$@*) : (s : N(*@$_c$@*).state) (*@$\to$@*)
-%%       (*@$\uparrow$@*)(Trace(*@$_{N_c}$@*) s true (*@$\lto$@*) {tr | (toN(*@$_c$@*) s (*@$\circ$@*) fromN(*@$_{c}$@*) s) tr = id tr})
-%% \end{lstlisting}
-%% \caption{Signature for the equalizer term}
-%% \label{fig:litequalizer}
-%% \end{figure}
-
-%% The type signature for the family of functions, $eqHom_{N_{c}}$, into the equalizer type is given in
-%% \cref{fig:litequalizer}. By defining an $Trace_{N_{c}}$-algebra homomorphism
-%% into the equalizer, we enable a
-%% proof strategy for equality by induction on traces through $N_{c}$. First consider the traces
-%% out of state $1$. The only case to handle here is $nil 1$, and by definition
-%% $toN_{c}~1$ and $fromN_{c}~1$ invert each other.
-%% When defining $eqHom_{N_{c}}$ for traces out of state $0$, we must leverage an
-%% inductive hypothesis,
-
-%% \begin{align*}
-%%   (toN_{c}~0 \circ fromN_{c}~0) (cons~t~true~c~tr)
-%%   & \equiv (toN_{c}~0) (let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a) \\
-%%   & \equiv cons~t~true~(let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a)~(nil~1) \\
-%%   & \equiv cons~t~true~c~(nil~1) \\
-%%   & \equiv cons~t~true~c~tr \\
-%% \end{align*}
-
-%% The first two equalities are by definition. The third equality holds because
-%% $\literal c$ is unambiguous and
-%% $let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a$ and $c$ have type
-%% $\literal c$. The final equality holds because our inductive hypothesis states
-%% that $nil~1 \equiv tr$. Thus, each $A s$ and $Trace_{N_{c}}~s~true$ are strongly
-%% equivalent --- in particular, $\literal c$ and $Parse_{N_{c}}$.
-
-%% \steven{Rewrite the disjunction case with equalizers}
-%% Now consider the case where $R$ is the disjunction of two regular expressions
-%% $R_{1}$ and $R_{2}$. Inductively, we may build NFAs $N_{1}$ and $N_{2}$ that are
-%% strongly equivalent to $R_{1}$ and $R_{2}$, respectively. We define $N_{\oplus}$
-%% as a new NFA built as the disjunction of $N_{1}$ and $N_{2}$.
-%% $N_{\oplus}.\states$ comprises the states from $N_{1}$, the states from
-%% $N_{2}$, and a new marked initial state. The accepting states are precisely
-%% those that were accepting in the subautomata. The initial state has an
-%% $\varepsilon$-transition to each of the initial states of the subautomata. Once in
-%% $N_{1}.\init$, the downstream transitions are a copy of those in $N_{1}$,
-%% likewise for $N_{2}.\init$.  That is, $N_{\oplus}$ contains a copy of each
-%% $N_{1}$ and $N_{2}$, and we choose which copy to inhabit by deciding which
-%% $\varepsilon$-transition out of the initial state to take. A schematic
-%% representation of this automaton is given in \cref{fig:disjunctionnfa}.
-
-%% \begin{figure}
-%%   \begin{tikzpicture}[node distance = 17mm ]
-%%     \node[state, initial] (init) {$\init$};
-%%     \node[state, above right of=init] (init1) {$N_{1}.\init$};
-%%     \node[state, below right of=init] (init2) {$N_{2}.\init$};
-%%     \node[right of=init1] (dots1) {$\dots$};
-%%     \node[right of=init2] (dots2) {$\dots$};
-%%     \node[state, accepting, right of=dots1] (acc1) {$s_{1}$};
-%%     \node[state, accepting, right of=dots2] (acc2) {$s_{2}$};
-
-%%     \path[->] (init) edge[below] node{$\varepsilon$} (init1)
-%%               (init) edge[above] node{$\varepsilon$} (init2)
-%%               (init1) edge[above] node{} (dots1)
-%%               (init2) edge[above] node{} (dots2)
-%%               (dots1) edge[above] node{} (acc1)
-%%               (dots2) edge[above] node{} (acc2);
-%%   \end{tikzpicture}
-%%   \caption{NFA $N_{\oplus}$ as a disjunction of $N_{1}$ and $N_{2}$}
-%%   \label{fig:disjunctionnfa}
-%% \end{figure}
-
-%% Because $N_{1}$ and $N_{2}$ are each strongly equivalent to $R_{1}$ and $R_{2}$,
-%% it suffices to show that
-%% $Parse_{N_{1}} \oplus Parse_{N_{2}}$ is strongly
-%% equivalent to $Parse_{N_{\oplus}}$.
-
-%% Just like the case of a single character, when building the maps in either
-%% direction, we strengthen the induction hypothesis by mapping out of traces
-%% beginning with an arbitrary state then proceed by induction on traces.
-
-%% First, define the
-%% mapping out of $\Trace_{N_{\oplus}}(s)$. If $s$ is a state from $N_{1}$ then
-%% send the trace to a trace in $N_{1}$ beginning at $s$. Likewise, if $s$ is from
-%% $N_{2}$, then we map to a trace in $N_{2}$ beginning at $s$. If $s$ is
-%% $N_{\oplus}.\init$, then we map the trace to a disjunction of traces
-%% $\Trace_{N_{1}}(N_{1}.\init) \oplus \Trace_{N_{2}}(N_{2}.\init)$.
-
-%% In the other direction, we need to map a disjunction of traces through $N_{1}$
-%% or $N_{2}$ to a trace in $N_{\oplus}$. Without loss of generality, assume that
-%% the trace came from $N_{1}$. That is, assume we currently have a term of type
-%% $\Trace_{{N_{1}}}(N_{1}.\init)$ and we aim to construct a term of type
-%% $\Trace_{N_{\oplus}}(N_{\oplus}.\init)$. We induct on the trace $N$ trace an map
-%% it over to a trace beginning at the copy of $N_{1}.\init$ in $N_{\oplus}$, and
-%% then we conclude by using the $\epscons$ constructor to turn this trace into one
-%% beginning at $N_{\oplus}.\init$.
-
-%% We show that these maps mutually invert each other by the same sort of argument
-%% we used with the single character NFA.\ That is, we prove that the each map is a
-%% homomorphism of algebras and then appeal to the initiality of each of
-%% $\Trace_{N_{1}}$, $\Trace_{N_{2}}$, and $\Trace_{N_{\oplus}}$ to prove that the
-%% relevant maps compose to the identity.
-
-%% In this sketch, we elide the proof that these maps do indeed form homomorphisms.
-%% Further, we are eliding the needed NFAs for the cases where the regular
-%% expression is $I$, $0$, a Kleene star, or a concatenation. All of these
-%% elided details may be found in the Agda formalization.
 
@@
 \end{theorem}
-
-%% We can build a $Dyck$ parser by utilizing
-%% this artifact, and the
-%% runtime of the $Dyck$ parser is about 30 seconds for input that is 25,000
-%% characters long.
-%% \pedro{It might be a good idea to mention explicitly in the introduction
-%%   that we are not claiming that our verified parsers are competitive with
-%%   the state of the art.}
 
@@
 \end{figure}
-% \subsection{LL(1)}
-% %% 4. LL(1) parser
-
-% \steven{TODO impute LL(1) example, mention that its weak equivalence is mechanized}
 
@@
 correspond to $P$.
-%% Define $\StringSem$ to be the non-linear list type over the alphabet $\Sigma$ and
-%% let  be a mapping from the set of strings to the universe of
-%% nonlinear types. The function $P$ describes a family of nonlinear types
-%% dependent on $\StringSem$.
-%% We view the image of $P$ as a set of parse trees for each input string, and thus
-%% $P$ carries the structure of a formal grammar. However, even though $P$ describes this grammatical data, it is not
-%% a first-class linear type in \theoryabbv.
 \(
@@
 \)
-%% \steven{Probably put this ceiling business around the discussion of the linear
-%%   type of strings. Is this in section 3?}
 where $\lceil \stringquote{} \rceil = I$ and
@@
 required to interpret all of the linear type and term constructors.
-%% \begin{theorem}
-%%   The category $\Grammar$ is complete, co-complete and cartesian
-%%   closed. Furthermore, it is monoidal biclosed with the monoidal
-%%   product given by
-%%   \[ (A \otimes B) w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in A w_1 \wedge b \in B w_2\} \]
-%%   and unit by
-%%   \[ I w = \{ () \pipe w = \varepsilon \} \]
-%% \end{theorem}
-
-%% \max{todo: need to decide if we need this section here or if it's in the intro already}.
-% The linear typing judgment in our syntax takes on the following schematic form
-% $\Gamma ; \Delta \vdash a : A$. First, $A$ represents a \emph{linear type} in
-% our syntax. The intended semantics of these linear types are formal grammars.
-% That is, the linear typing system is designed to syntactically reflect the
-% behavior of formal grammars. For this reason, we may often interchangeably use
-% the terms ``linear type'' and ``grammar''.
-
-% The term $a$ is an inhabitant of type $A$, which is thought of as a parse tree of
-% the grammar $A$. The core idea of this entire paper follows precisely from this
-% single correspondence: grammars are types, and the inhabitants of these types
-% are parse trees for the grammars.
-
-% $\Gamma$ represents a non-linear context, while $\Delta$ represents a
-% \emph{linear context} dependent on $\Gamma$. These linear contexts behave
-% substructurally. As stated earlier, they are linear --- they do not obey
-% weakening or contraction --- because a character is exhausted once read by a
-% parsing procedure. Moreover, the characters in strings appear in the order in
-% which we read them. We do not have the freedom to freely permute characters,
-% therefore any type theory that is used to reason about formal grammars ought to
-% omit the structural rule of exchange as well. This means that every variable
-% within $\Delta$ must be used \emph{exactly once} and \emph{in order of
-%   occurrence}. Thus, we can think of the linear contexts as an ordered list of
-% limited resources. Once a resource is consumed, we cannot make reference to it
-% again. Variables in a linear context then act like building
-% blocks for constructing patterns over strings.
-
-% We give the base types and type constructors for linear terms. As the
-% interpretation of types as grammars in $\Grammar$ serves as our intended
-% semantics, we simultaneously give the interpretation $\sem \cdot$ of the
-% semantics as grammars.
-
-% \pedro{The paragraphs above are a bit circular, what about the following alternative?}
-% \steven{I am in favor of cutting what's above, but we haven't introduced the
-%   syntax in the main body of the paper yet, even if the figures appeared
-%   earlier. So what you write needs to take that into account}
-% \pedro{In which case we should introduce it in the previous section. I suggest
-%   adding this right after ``and the intuitionistic typing rules in fig 3''}
+
 \subsection{Semantics}
@@
 
-%% That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
-%% complete to parses of $B$. In this manner, the linear function types generalize
-%% Brzozowksi's notion of derivative
-%% \cite{brzozowskiDerivativesRegularExpressions1964}.
-%% Brzozowski initially only gave an accounting of this operation for
-%% generalized regular expressions, but later work from Might et al.\ demonstrates that the same
-%% construction can be generalized to context free grammars
-%% \cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
-%% types, the same notion of derivative extends to the grammars of \theoryabbv.
-
-%% Note, to ensure that the linear function types do indeed generalize Brzozowski
-%% derivatives, we must include the equalities in \cref{fig:brzozowskideriv} as axioms\max{we must? why must we}.
-%% \pedro{We should further explain why these equations are required, or at the very
-%% least give some intuition}
-
-%% Of course, all the above statements for the left function type also have
-%% corresponding analogues for the
-%% right-handed counterpart.
-
-%% \begin{figure}
-%% \begin{align*}
-%%   c\lto c &\equiv I\\
-%%   c\lto d &\equiv 0\\
-%%   c\lto I &\equiv 0\\
-%%   c\lto 0 &\equiv 0\\
-%%   c\lto (A \otimes B) & \equiv (c\lto A) \otimes B + (A \amp I) \otimes (c\lto B)\\
-%%   c\lto A^* &\equiv (A \amp I)^{*} \otimes (c \lto A) \otimes A^* \\
-%%   c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
-%% \end{align*}
-%% \caption{Equality for Brzozowski Derivatives}
-%% \label{fig:brzozowskideriv}
-%% \end{figure}
-
-%% \paragraph{LNL Dependent Types}
-%% \steven{talk about how binary sum and binary with are definable over $\Bool$}
-%% Given a non-linear type $X$, we may form both the dependent product type
-%% $\LinPiTy x X A$ and the dependent pair type $\LinSigTy x X A$ as linear types
-%% where $x$ is free in $A$.
-
-%% $\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-
-%% $\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
-
-%% The grammar semantics of the linear product type is
-%% indeed a dependent function out of the
-%% semantics of $X$. Likewise, the grammar semantics of the linear dependent pair type is a dependent
-%% pair in $\Set$.\pedro{The connection to grammar semantics needs to be better explained. For instance,
-%% we should explain that disjunction corresponds to ``or'' of grammars}
-
-%% Note that even though we do not take the binary additive conjunction and
-%% disjunction as primitive, we may define them via these LNL dependent types. In
-%% particular, via a dependence on $\Bool$.
-%% %
-%% \[
-%%   A \amp B := \LinPiTy b \Bool {C(b)}
-%% \]
-%% \[
-%%   A \oplus B := \LinSigTy b \Bool {C(b)},
-%% \]
-%% where $C(\true) = A, C(\false) = B$.
-
-%% \paragraph{Universal Type}\max{this is absolutely not the right terminology. Universal type means multiple other completely different things}
-%% \steven{Definable as nullary product}
-%% The universal type $\top$ may be formed in any context.
-%% %
-%% \[ \sem{\top} \gamma w = \{ \ast \}\]
-%% %
-%% Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.
-
-%% \paragraph{Empty Type}
-%% \steven{Definable as nullary sum}
-%% The empty type $0$ has no inhabitants. The elimination for the empty type
-%% witnesses the principle of explosion --- i.e.\ from a term of type $0$ we may
-%% introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.
-%% %
-%% \[ \sem {0} \gamma w = \emptyset \]
-%% %
-%% \paragraph{The $G$ Modality}
-%% \steven{Rename this}
-%% \steven{Compare to persistence modalities from separation logic}
-%% The $G$ modality provides the embedding of linear types in the empty context
-%% into non-linear types. The introduction and elimination forms for $G A$ obey the
-%% same rules as given in \cite{bentonMixedLinearNonlinear1995} and \cite{krishnaswami_integrating_2015}.
-
-%% The left adjoint $F$ from non-linear types back to linear types may be defined
-%% as,
-%% %
-%% \[
-%%   F X := \LinSigTy a A I
-%% \]
-%% %
-%% \steven{Elaborate on $G$ more and rename $G A$to $\ltonl A$ in this section}
-
-%% \pedro{Agreed :) maybe mention connections to persistent propositions from the
-%% separation logic literature}
-
-%% The semantics of $G A$ are exactly the semantics of $A$ at the empty word $\varepsilon$.
-%% %
-%% \[ \sem{G A} \gamma = \sem{A} \gamma \varepsilon \]
-%% %
-%% %
-%% \paragraph{Recursive Types}
-%% \steven{reflavor this section as ``indexed inductive'' instead of ``recursive'',
-%% and entirely rework with the new inductive definition}
-%% \steven{The non-indexed inductive are definable as being trivially indexed over unit}
-%% Recursive linear types may be defined via the least-fixed point operator $\mu$.
-% The grammar semantics of which are the fixed-point of sets induced by the
-% grammar semantics of $A$.
-% %
-% \[ \sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x) \]
-% %
-% \pedro{We should justify, even if briefly, the existence of this
-%   fixed-point}
-
-% Observe that we need not take the Kleene star as a primitive
-% grammar constructor, as it is definable as a fixed point.
-% The Kleene star of a grammar $g$ is given as,
-
-% \[
-%   g^{*} := \mu X . I \oplus (g \otimes X)
-% \]
-
-% \begin{figure}[h!]
-% \begin{mathpar}
-%   \inferrule
-%   {\Gamma ; \Delta \vdash p : I}
-%   {\Gamma ; \Delta \vdash \mathsf{nil} : g^{*}}
-
-%   \and
-
-%   \inferrule
-%   {\Gamma ; \Delta \vdash p : g \\ \Gamma ; \Delta' \vdash q
-%   : g^{*}}
-%   {\Gamma ; \Delta \vdash \mathsf{cons}(p , q) : g^{*}}
-
-%   \\
-
-%   \inferrule
-%   {
-%     \Gamma ; \Delta \vdash p : g^{*} \\
-%     \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
-%     \Gamma ; x : g , y : h \vdash p_{\ast} : h
-%   }
-%   {\Gamma ; \Delta \vdash \mathsf{foldr}(p_{\varepsilon} , p_{\ast}) : g^{*}}
-% \end{mathpar}
-% \caption{Kleene Star Rules}
-% \label{fig:star}
-% \end{figure}
-
-% Likewise, $g^{*}$ has admissible introduction and
-% elimination rules, shown in \cref{fig:star}. Note that this
-% definition of $g^{*}$ and these
-% rules arbitrarily assigns a handedness to the Kleene star.
-% We could have just as well took it to be a fixed point of
-% $I \oplus (X \otimes g)$. In fact, the definitions are
-% equivalent, as the existence of the $\mathsf{foldl}$ term below
-% shows that $g^{*}$ is also a fixed point of
-% $I \oplus (X \otimes g)$.
-% This is a marked point of difference from Kleene
-% algebra with recursion, where the fixed points for the left and right variants
-% of Kleene star need not agree \cite{leiss}.
-
-% \begin{equation}
-%   \label{eq:foldl}
-%   \inferrule
-%   {
-%     \Gamma ; \Delta \vdash p : g^{*} \\
-%     \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
-%% %     \Gamma ; y : h , x : h \vdash p_{\ast} : h
-%% %   }
-%% %   {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
-%% % \end{equation}
-
-%% % In fact, the $\mathsf{foldl}$ term is defined using
-%% % $\mathsf{foldr}$ --- much in the same way one
-%% % may define a left fold over lists in terms of a right fold
-%% % in a functional programming language.
-%% % The underlying trick is to fold over a list of linear functions
-%% % instead of the original string. We curry each character $c$
-%% % of the string into a function that concatenates $c$, and
-%% % right fold over this list of linear functions. Because function
-%% % application is left-associative, this results in a left
-%% % fold over the original string.
-
-%% % We only take fixed points of a single variable as a
-%% % primitive operation in the type theory, but we may apply
-%% % Beki\`c's theorem \cite{Bekić1984} to define an admissible
-%% % notion of multivariate fixed point. This is particularly
-%% % useful for defining grammars that encode the states of an
-%% % automaton, as we will see in \cref{sec:automata}. In \cref{fig:multifix} we provide the
-%% % introduction and elimination principles for such a fixed
-%% % point, where $\sigma$ is the substitution that unrolls the
-%% % mutually recursive definitions one level. That is,
-
-%% % \begin{align*}
-%% %   \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
-%% % \end{align*}
-
-%% % \begin{figure}
-%% % \begin{mathpar}
-%% %   \inferrule
-%% %   {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
-%% %   {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
-
-%% %   \and
-
-%% %   \inferrule
-%% %   {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
-%% %              \Gamma ; x_{j} : \simulsubst {A_{j}}{\sigma} \vdash e_{j} : B_{j}\quad \forall j
-%% %   }
-%% %   {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
-%% % \end{mathpar}
-%% % \caption{Multi-fixed Points}
-%% % \label{fig:multifix}
-%% % \end{figure}
-
-%% %% \section{Categorical Semantics of \theoryabbv}
-%% %% \label{sec:categorify}
-
-%% As described in previous sections, \theoryabbv has a ``standard''
-%% interpretation where non-linear types denote sets, linear types denote
-%% formal grammars, non-linear terms denote functions and linear terms
-%% denote parse transformers.
-%% %
-%% Further, as is typical for type theory, every type constructor in the
-%% calculus is interpreted by using a \emph{universal construction} in
-%% either the category of sets or the category of grammars\footnote{the
-%% only exception is that the axioms for Brzozowski derivatives we
-%% considered do not follow solely from universal properties. In this
-%% section we focus on models that do not necessarily satisfy these
-%% axioms.}.
-%% %
-%% This leads to an immediate opportunity for generalization: the type
-%% theory has in addition to the standard grammar-theoretic
-%% interpretation, an interpretation in any categorical structure that
-%% exhibits the same universal constructions.
-%% %
-%% On the one hand this gives us a structured way to formalize the
-%% standard semantics precisely, but additionally it enables us to
-%% consider \emph{non-standard} models in the next section that point to
-%% further applications as well as meta-theoretic results.
-
-%% In developing this categorical semantics, we will start with the
-%% closest analogue from formal language theory: Kleene algebra.  Kleene
-%% algebras are an important tool in the theory of regular languages
-%% serving as a bridge between algebraic reasoning and equivalence of
-%% regular expressions. More broadly, through various extensions, they
-%% serve as a theoretical substrate to studying different kinds of formal
-%% languages.
-%% %
-%% We can then see that the ``regular fragment'' of our type theory
-%% (i.e., just characters, $\otimes$, $\oplus$, $0,1$ and a Kleene star)
-%% has models in what we call a \emph{Kleene category} a categorification
-%% of Kleene algebra from posets to categories.
-%% %
-%% We then further develop this into our final notion of model, which we
-%% call a \emph{Chomsky category} as it can model not just regular
-%% languages, but the full Chomsky hierarchy.
-
-%% \subsection{Comparison to Kleene Algebra}
-%% \subsection{Kleene Algebra and Kleene Categories}
-%% A Kleene algebra is a tuple $(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is
-%% a set, $+$ and $\cdot$ are binary operations over $A$, $(-)^*$ is a
-%% function over $A$, and $1$ and $0$ are constants. These structures
-%% satisfy the axioms depicted in Figure~\ref{fig:axioms}.
-
-%% \begin{figure}
-%%   \begin{align*}
-%%     x + (y + z) &= (x + y) + z & x + y &= y + x\\
-%%     x + 0 &= x & x + x &= x\\
-%%     x(yz) &= (xy)z & x1 &= 1x = x\\
-%%     x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
-%%     x0 &= 0x = x & & \\
-%%     1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
-%%      b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
-%%   \end{align*}
-%%   \caption{Kleene algebra axioms}
-%%   \label{fig:axioms}
-%% \end{figure}
-
-%% The addition operation can be used to define the partial order
-%% structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
-%% formal languages, this order structure can be used to model language
-%% containment. In this section, we categorify the concept of Kleene
-%% algebra and build on top of it in order to define an abstract theory
-%% of parsing. We start by defining \emph{Kleene categories}.
-
-%% \begin{definition}
-%%   A Kleene category is a distributive monoidal category $\cat{K}$
-%%   such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
-%%   = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
-%%   algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
-%%   X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogous isomorphism
-%%   for $G_{A,B}$ also holds.
-%% \end{definition}
-
-%% As a sanity check, note that Kleene algebras are indeed examples of
-%% Kleene categories.
-
-%% \begin{example}
-%%   Every Kleene algebra, seen a posetal category, is a Kleene category.
-%%   The product $\cdot$ is a monoidal product and the addition is a
-%%   least-upper bound, which corresponds to a coproduct. Lastly, the
-%%   axioms of the Kleene star have a direct correspondence to the
-%%   coherence conditions postulated by the initial algebras of Kleene
-%%   categories.
-%% \end{example}
-
-%% This example provides a neat categorical justification to how
-%% restrictive Kleene algebras are in terms of reasoning about
-%% languages. By only having at most one morphism between objects, there
-%% is not a lot of information they can convey. In this case, the only
-%% information you get is language containment. As demonstrated by
-%% $\mathbf{Gr}$, the extra degrees of freedom granted by having more
-%% morphisms gives you more algebraic structure for reasoning about
-%% languages.
-
-%% For the next example, we see an unexpected connection with the theory
-%% of substructural logics.
-
-%% \begin{example}
-%%   The opposite category of every Kleene category is a model of a variant of
-%%   conjunctive ordered logic, where the Kleene star plays the role of the ``of
-%%   course'' modality from substructural logics which allows hypotheses to
-%%   be discarded or duplicated.
-%% \end{example}
-
-%% As we have seen, the proposed axioms are a direct translation of the
-%% Kleene algebra axioms to a categorical setting. Its most unusual aspect is the
-%% axiomatization of the Kleene star as a family of initial algebras
-%% satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
-%% more structure, then these isomorphisms hold ``for free''.
-
-%% \begin{theorem}
-%%   \label{th:kleeneclosed}
-%%   Let $\cat{K}$ be a Kleene category such that it is also monoidal
-%%   closed.  Then, the initial algebras isomorphisms hold automatically.
-%% \end{theorem}
-%% \begin{proof}
-%%   We prove this by the unicity (up-to isomorphism) of initial
-%%   algebras. Let $[hd, tl]: I + (\mu X.\, F_{A, I}(X)) \otimes A \to
-%%   (\mu X.\, F_{A, I}(X))$ be the initial algebra structure of $(\mu
-%%   X.\, F_{A, I}(X))$ and consider the map $[hd, tl] : B + B \otimes
-%%   (\mu X.\, F_{A, I}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
-%%     I}(X))$.
-
-%%   Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
-%%   we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,I} \to B \lto Y$. We can show existence and
-%%   uniqueness by showing that the diagram on top commutes if, and
-%%   only if, the diagram on the bottom commutes:
-
-%% % https://q.uiver.app/#q=WzAsOCxbMCwwLCJCICsgQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFsyLDAsIkIgKyBZIFxcb3RpbWVzIEEiXSxbMiwyLCJZIl0sWzAsMiwiQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDMsIjEgKyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDUsIlxcbXUgWC5cXCwgMSArIFggXFxvdGltZXMgQSJdLFsyLDMsIjEgKyAoIEIgXFxsdG8gWSkgXFxvdGltZXMgQSJdLFsyLDUsIkIgXFxsdG8gWSJdLFswLDEsImlkICsgKGlkIFxcb3RpbWVzIGg7IGV2KSBcXG90aW1lcyBpZF9BIl0sWzEsMiwiW2YsZ10iXSxbMywyLCJpZCBcXG90aW1lcyBoOyBldiIsMl0sWzAsMywiW2lkIFxcb3RpbWVzIGgsIGlkIFxcb3RpbWVzIHRsXSIsMl0sWzQsNiwiaWQgKyAoaCBcXG90aW1lcyBYKVxcb3RpbWVzIGlkIl0sWzUsNywiaCIsMl0sWzYsNywiW2YnLCBnJ10iXSxbNCw1LCJbaGQsIHRsXSIsMl1d
-%% \[\begin{tikzcd}
-%% 	{B + B \otimes (\mu X.\, I + X \otimes A)} && {B + Y \otimes A} \\
-%% 	\\
-%% 	{B \otimes (\mu X.\, I + X \otimes A)} && Y \\
-%% 	{I + (\mu X.\, I + X \otimes A)} && {I + ( B \lto Y) \otimes A} \\
-%% 	\\
-%% 	{\mu X.\, I + X \otimes A} && {B \lto Y}
-%% 	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
-%% 	\arrow["{[f,g]}", from=1-3, to=3-3]
-%% 	\arrow["{id \otimes h; ev}"', from=3-1, to=3-3]
-%% 	\arrow["{[id \otimes h, id \otimes tl]}"', from=1-1, to=3-1]
-%% 	\arrow["{id + (h \otimes X)\otimes id}", from=4-1, to=4-3]
-%% 	\arrow["h"', from=6-1, to=6-3]
-%% 	\arrow["{[f', g']}", from=4-3, to=6-3]
-%% 	\arrow["{[hd, tl]}"', from=4-1, to=6-1]
-%% \end{tikzcd}\]
-%%   This equivalence follows by using the adjunction structure given
-%%   by the monoidal closed structure of $\cat{K}$. A completely analogous
-%%   argument for $G_{A,B}$ also holds. Furthermore, by generalizing the
-%%   construction of \Cref{sec:formaltype}, we can also show that from
-%%   the monoidal closed assumption it follows that $\mu X.\, F_{A, I}(X) \cong \mu X.\, G_{A, I}(X)$
-%% \end{proof}
-
-%% Something surprising about this lemma is that it provides an alternative
-%% perspective on the observation that if a Kleene algebra has an
-%% residuation operation, also called action algebra \cite{kozen1994action},
-%% then the Kleene star admits a simpler axiomatization.
-
-%% Since we want Kleene categories to generalize our notion of formal
-%% grammars as presheaves $\String \to \Set$, we prove that they do
-%% indeed form a Kleene category. We start by presenting a well-known
-%% construction from presheaf categories.
-
-%% \begin{definition}
-%%   Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
-%%   two functors $\cat{C} \to \Set$. Their Day convolution tensor
-%%   product is defined as the following coend formula:
-%%   \[
-%%   (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z)
-%%   \]
-%%   Dually, its internal hom is given by the following end formula:
-%%   \[
-%%   (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
-%%   \]
-%% \end{definition}
-
-%% \begin{lemma}[Day \cite{day1970construction}]
-%%   Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
-%%   monoidal closed.
-%% \end{lemma}
-
-%% %% \begin{theorem}
-%% %%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
-%% %%   The functor category $[A, \cat{K}]$.
-%% %%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
-%% %%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
-%% %% \end{theorem}
-%% \begin{theorem}
-%%   If $\cat{C}$ is a locally small monoidal category, then
-%%   $\Set^{\cat{C}}$ is a Kleene category.
-%% \end{theorem}
-%% \begin{proof}
-
-%%   By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
-%%   is a presheaf category, it has coproducts. Furthermore, the tensor
-%%   is a left adjoint, i.e. it preserves colimits and, therefore, it is
-%%   a distributive category.
-
-%%   As for the Kleene star, since presheaf categories admit small colimits,
-%%   the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
-%%   defined as the filtered colimit of the diagrams:
-
-%%   From Theorem~\ref{th:kleeneclosed} it follows that these initial
-%%   algebras satisfy the required isomorphisms and this concludes the
-%%   proof.
-%% \end{proof}
-
-%% \begin{corollary}
-%%   For every alphabet $\Sigma$, the presheaf category $\Set^{\cat{\Sigma^*}}$
-%%   is a Kleene category.
-%% \end{corollary}
-%% \begin{proof}
-%%   Note that string concatenation and the empty string make the
-%%   discrete category $\Sigma^*$ a strict monoidal category.
-%% \end{proof}
-
-%% Much like in the posetal case, the abstract structure of a Kleene
-%% category is expressive enough to synthetically reason about regular
-%% languages. A significant difference between them is that while Kleene
-%% algebras can reason about language containment, Kleene categories can
-%% reason about \emph{ambiguity}, \emph{strong equivalence} of grammars.
-
-%% \subsection{Lambek Hyperdoctrines and Chomsky Hyperdoctrines}
-
-%% Though Kleene categories are expressive enough to reason about
-%% concepts that are outside of reach of Kleene algebras, their
-%% simply-typed nature makes them not so expressive from a type theoretic
-%% point of view. This is limiting because type theories are successful
-%% syntactic frameworks for manipulating complicated categorical
-%% structures while avoiding some issues common in category theory, such
-%% as coherence issues.
-
-%% With this in mind, we want to design a categorical semantics that
-%% builds on top of Kleene categories with the goal of extending them
-%% with dependent types and making them capable reasoning about languages
-%% and their parsers. This leads us to the abstract notion of model we
-%% are interested in capturing with \theoryabbv: a \emph{Chomsky
-%% category}.
-
-%% We do this in two stages: first we define a \emph{Lambek hyperdoctrine} to be a
-%% notion of model for the judgmental structure of \theoryabbv: that is we have
-%% notions of linear and non-linear type, contexts, substitution and terms, but do
-%% not assume any particular type constructions exist. Then we will define when a
-%% Lambek hyperdoctrine is a Chomsky hyperdoctrine, meaning it can interpret all
-%% the type formers of \theoryabbv and therefore arbitrary strength grammars.
-
-%% \begin{definition}[Lambek Hyperdoctrine]
-%%   A \emph{Lambek hyperdoctrine} consists of
-%%   \begin{enumerate}
-%%   \item A category $\mathcal C$ with a terminal object.
-%%   \item A category with families structure over $\mathcal C$.
-%%   \item A contravariant functor $L : \mathcal C^{o} \to \textrm{MultiCat}$ from
-%%     $\mathcal C$ to the category of Multicategories.
-%%   \end{enumerate}
-%% \end{definition}
-%% Here the objects of $\mathcal C$ model the non-linear contexts, and morphisms
-%% model non-linear substitutions. The category with families structure over
-%% $\mathcal C$ models the dependent non-linear types. Finally the
-%% ``hyperdoctrine'' of multicategories $L$ models the linear types and terms. The
-%% fact that this is functorial in $\mathcal C$ corresponds to the fact that linear
-%% types are all relative to a non-linear context and that linear variables and
-%% composition commute with non-linear substitution.
-
-%% We will have three main Lambek hyperdoctrines of interest in this paper: the
-%% ``standard model'' of sets and grammars, the ``syntactic model'' given by our
-%% type theory itself and lastly a gluing model introduced in
-%% Section~\ref{sec:canonicity} to prove the canonicity theorem.
-%% \begin{example}
-%%   \begin{enumerate}
-%%   \item Let $\mathcal C$ be the category of sets, equipped with its usual
-%%     category with families structures of families. Define $L : \mathcal C^{o}
-%%     \to \textrm{MultiCat}$ to map $L(X)$ to the representable multicategory
-%%     $(\Set^{\Sigma^*})^X$ of $X$-indexed families of grammars where the monoidal
-%%     structure is given pointwise by the Day convolution monoidal structure.
-%%   \item We can build a ``syntactic'' Lambek hyperdoctrine $\Syn(\Sigma)$ from the syntax
-%%     itself: the category $\mathcal C$ is given by non-linear contexts and
-%%     substitutions, the category with families by the non-linear types and terms
-%%     and the hyperdoctrine of multicategories by the linear types and terms.
-%%   \end{enumerate}
-%% \end{example}
-
-%% Next each type linear and non-linear type constructor corresponds to extra data
-%% on a Lambek hyperdoctrine.
-%% \begin{definition}
-%%   \label{def:chomsky-data}
-%%   \begin{enumerate}
-%%   \item Dependent type structure: standard (say extensional type theory, give a reference)
-%%   \item Universes
-%%   \item Inductive grammars?
-%%   \item Non-inductive Grammar constructors: standard monoidal category stuff
-%%   \end{enumerate}
-%% \end{definition}
-%% \begin{definition}[Chomsky Hyperdoctrine]
-%%   A Chomsky Hyperdoctrine is a Lambek hyperdoctrine equipped with all of the data in Definition~\ref{def:chomsky-data}.
-%% \end{definition}
-%% \pedro{We should probably define some of the words in this definition}
-%% \begin{definition}
-%%   A Chomsky category is a locally Cartesian category with two hierarchies of
-%%   universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
-%%   every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
-%%   $U_i$ to be closed under dependent products and sum,
-%%   $L_i$ to be closed under the Kleene category connectives,
-%%   dependent products, left and right closed structures, with
-%%   a type constructor $G : L_i \to U_i$ and a linear dependent sum
-%%   going the other direction.
-%% \end{definition}
-
-%% \steven{Max suggests augmenting the definition of a Chomsky category to
-%%   something like two categories $L$ and $U$, and $U$ is Cartesian closed I don't
-%%   fully recall the rest.
-
-%%   My best guess is that you take $L$ a Kleene category with a hierarchy of universes
-%%   two, and you further require that $L$ is $Psh(U)$-enriched, except he
-%%   suggested a further adjective on these presheaves. Perhaps representability?
-%% }
-
-%% \begin{theorem}
-%%   The presheaf category $\Grammar = \Set^{\cat{\Sigma^*}}$ is a Chomsky category.
-%% \end{theorem}
-
-%% Further, the syntactic category of \theoryname is manifestly a Chomsky category.
-
-%% \pedro{This is likely true, but if we explicitly say so, this warrants a proof. I think that
-%% if we don't say anything about the syntactic category, reviewers won't mind.}
-
-%% \steven{I agree that we don't want to say anything that opens unnecessary
-%%   questions for proofs we haven't written. However, it seems hard to make the
-%%   case that we have the right categorical model of the syntax if this statement
-%%   isn't true. By restating our definition of Chomsky category, this should be
-%%   obvious or a quick proof}
-
-%% \pedro{I agree, then we should add the quick proof :) }
-
-%% \subsection{Concrete Models of \theoryabbv}
-%% \label{sec:othermodels}
-
-%% \steven{Because we can define a version of semantic actions internally, we
-%%   shouldn't put it as a separate model}
-
-%% One of the powers of type theories is that it can be profitable to interpret them
-%% in various models. In this section, by using our just-defined Chomsky categories,
-%% we show how other useful concepts from formal language theory can also be organized
-%% as models of \theoryabbv. We illustrate this point by providing two examples that
-%% are closely related to the theory of formal languages: language equivalence and
-%% semantic actions. Furthermore, in order to justify how $\mathbf{Gr}$ relates to
-%% more traditional notions of parsing, we define a glued model that proves a
-%% canonicity property of grammar terms.
-
-%% \subsection{Language Semantics}
-%% Every grammar induces a language semantics. Also languages can be taken as a
-%% propositionally truncated view of the syntax. Logical equivalence should induce
-%% weak equivalence, and thus even give a syntactic way to reason about language equivalence.
-%% \steven{TODO language semantics}
-
-%% % \subsection{Semantic Actions}
-%% % \steven{Tentatively planning to cut this subsection for an internal representation
-%% %   of semantic actions}
-%% % Returning to the problem of parsing, the output of a parse usually is not the
-%% % literal parse tree. Rather, the output is the result of some \emph{semantic
-%% %   actions} ran on the parse tree, which usually serve to remove some syntactic
-%% % details that are unnecessary for later processing.
-
-%% % Given a grammar $G : \String \to \Set$, we define a semantic action to be a set
-%% % $X$ with a function $f$ that produces a semantic element from any parse of $G$.
-
-%% % \[
-%% %   f : \PiTy w \String {G w \to X}
-%% % \]
-
-%% % Further, semantic actions can be arranged into a structured category.
-%% % Define $\SemAct$ as the comma
-%% % category $\Grammar / \Delta$, where $\Delta : \Set \to \Grammar$ defines a
-%% % discrete presheaf. That is, for a set $X$, $\Delta (X)(w) = X$ for all
-%% % $w \in \String$. As $\SemAct$ is defined as a comma category, it has a forgetful
-%% % functor into $\Grammar$. That is, $\SemAct$ serves as a notion of formal
-%% % grammar. Moreover, $\SemAct$ is a model of \theoryabbv.
-
-%% % \steven{It being a notion of formal grammar is distinct from being a model. This
-%% % probably warrants a proof}
-
-%% % \steven{TODO semantic actions}
-
-%% % \pedro{This is a very nice opportunity of showing off the supremacy of denotational
-%% %   reasoning ;) We should probably prove the gluing lemma in the previous section
-%% %   and apply it here and in the canonicity section. The actual proof might have to be
-%% %   moved to the appendix, though}
-
-%% \subsection{Parse Canonicity}
-%% Canonicity is an important metatheoretic theorem in the type theory
-%% literature.  It provides insight on the normal forms of terms and,
-%% therefore, on its computational aspects. Frequently, proving
-%% canonicity for boolean types, i.e. every closed term of type bool
-%% reduces to either true or false, is enough to justify that the type
-%% theory being studied is well-behaved. In our case, however, since we
-%% want to connect \theoryabbv to parsers, we must provide a more
-%% detailed account of canonicity. In particular, we give a nonstandard semantics
-%% of \theoryabbv that carries a proof of canonicity along with it.
-
-%% If $\cdot \vdash A$ is a closed linear type then there are
-%% two obvious notions of what constitutes a ``parse'' of a string w
-%% according to the grammar $A$:
-%% \begin{enumerate}
-%% \item On the one hand we have the set-theoretic semantics just
-%%   defined, $\llbracket A \rrbracket \cdot w$
-%% \item On the other hand, we can view the string $w = c_1c_2\cdots$ as
-%%   a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
-%%   define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
-%%   \lceil w \rceil \vdash e : G$.
-%% \end{enumerate}
-%% It is not difficult to see that at least for the ``purely positive''
-%% formulae (those featuring only the positive connectives
-%% $0,+,I,\otimes,\mu, \overline\Sigma,c$) that
-%% every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
-%% that the nodes of the tree correspond precisely to the introduction
-%% forms of the type. However it is far less obvious that \emph{every}
-%% linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
-%% sequence of introduction forms since proofs can include elimination
-%% forms as well. To show that this is indeed the case we give a
-%% \emph{canonicity} result for the calculus: that the parses for .
-
-%% \begin{definition}
-%%   A non-linear type $X$ is purely positive if it is built up using
-%%   only finite sums, finite products and least fixed points.
-
-%%   A linear type is purely positive if it is built up using only finite
-%%   sums, tensor products, generators $c$, least fixed points and linear
-%%   sigma types over purely positive non-linear types.
-%% \end{definition}
-
-%% \begin{definition}
-%%   %% Let $X$ be a closed non-linear type. The closed elements $\textrm{Cl}(X)$ of $X$ are the definitional equivalence classes of terms $\cdot \vdash e : X$.
-
-%%   Let $A$ be a closed linear type. The nerve $N(A)$ is a presheaf on
-%%   strings that takes a string $w$ to the definitional equivalence
-%%   classes of terms $\cdot; \lceil w\rceil \vdash e: N(A)$.
-%% \end{definition}
-
-%% \begin{theorem}[Canonicity]
-%%   Let $A$ be a closed, purely positive linear type. Then there is an
-%%   isomorphism between $\llbracket A\rrbracket$ and $N(A)$.
-%% \end{theorem}
-%% \begin{proof}
-%%   We outline the proof here, more details are in the appendix. The
-%%   proof proceeds first by a standard logical families construction
-%%   that combines canonicity arguments for dependent type theory
-%%   TODO cite coquand
-%%   % \cite{coquand,etc}
-%%   with logical relations constructions for linear
-%%   types
-%%   TODO cite hylandschalk
-%%   % \cite{hylandschalk}
-%%   . It is easy to see by induction that the
-%%   logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
-%%   \rrbracket$ and the fundamental lemma proves that the projection
-%%   morphism $p : \hat A \to N(A)$ has a section, the canonicalization
-%%   procedure. Then we establish again by induction that
-%%   canonicalization is also a retraction by observing that introduction
-%%   forms are taken to constructors.
-%% \end{proof}
-
-%% \begin{enumerate}
-%% \item Every term $\lceil w \rceil \vdash p : G + H$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
-%% \item There are no terms $\lceil w \rceil \vdash p : 0$
-%% \item If there is a term $\lceil w \rceil \vdash p : c$ then $w = c$ and $p = x$.
-%% \item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
-%% \item Every term $\lceil w \rceil \vdash p : \varepsilon$ is equal to $()$
-%% \item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
-%% \item Every term $\lceil w \rceil \vdash p : \mu X. G$ is equal to $\textrm{roll}(q)$ where $q : G(\mu X.G/X)$
-%% \item Every term $\lceil w \rceil \vdash p : (x:A) \times G$ is equal
-%%   to $(M,q)$ where $\cdot \vdash M : A$
-%% \end{enumerate}
-
-%% To prove this result we will use a logical families model. We give a
-%% brief overview of this model concretely:
-%% \begin{enumerate}
-%% \item A context $\Gamma$ denotes a family of sets indexed by closing substitutions $\hat\Gamma : (\cdot \vdash \Gamma) \Rightarrow \Set_i$
-%% \item A type $\Gamma \vdash X : U_i$ denotes a family of sets $\hat X : \Pi(\gamma:\cdot \vdash \Gamma) \hat\Gamma \Rightarrow (\cdot \vdash \simulsubst X \gamma) \Rightarrow \Set_i$
-%% \item A term $\Gamma \vdash e : X$ denotes a section $\hat e : \Pi(\gamma)\Pi(\hat\gamma)\hat X \gamma \hat\gamma (\simulsubst e \gamma)$
-%% \item A linear type $\Gamma \vdash A : L_i$ denotes a family of grammars $\hat A : \Pi(\gamma:\cdot\vdash\Gamma)\,\hat\Gamma \Rightarrow \Pi(w:\Sigma^*) (\cdot;\lceil w\rceil \vdash A[\gamma])\Rightarrow \Set_i$, and the denotation of a linear context $\Delta$ is similar.
-%% \item A linear term $\Gamma;\Delta \vdash e : A$ denotes a function \[\hat e : \Pi(\gamma)\Pi(\hat\gamma)\Pi(w)\Pi(\delta : \lceil w \rceil \vdash \simulsubst \Delta \gamma) \hat\Delta \gamma \hat\gamma \delta \Rightarrow \hat A \gamma \hat\gamma w {(\simulsubst {\simulsubst e \gamma} \delta)}\]
-%% \end{enumerate}
-%% And some of the constructions:
-%% \begin{enumerate}
-%% \item $\widehat {(G A)} \gamma \hat\gamma e = \hat A \gamma \hat\gamma \varepsilon (G^{-1}e)$
-%% \item $\widehat {(A \otimes B)} \gamma \hat\gamma w e = \Sigma(w_Aw_B = w)\Sigma(e_A)\Sigma(e_B) (e_A,e_B) = e \wedge \hat A \gamma \hat \gamma w_A e_A \times \hat B \gamma \hat \gamma w_B e_B$
-%% \item $\widehat {(A \lto B)} \gamma \hat\gamma w e = \Pi(w_A)\Pi(e_A) \hat A \gamma\hat\gamma w_A e_A \Rightarrow \hat B \gamma\hat\gamma (ww_A) (\applto {e_A} e)$
-%% \end{enumerate}
-
-%% First, the category with families will be
-%% the category of logical families over set contexts/types
-%% $\Delta$/$A$. Then the propositional portion will be defined by
-%% mapping a logical family $\hat \Gamma \to \Gamma$
-
-%% First, let $L$ be the category of BI formulae and proofs (quotiented
-%% by $\beta\eta$ equality). Define a functor $N : L \to \Set^{\Sigma^*}$ by
-%% \[ N(\phi)(w) = L(w,\phi) \]
-
-%% Then define the gluing category $\mathcal G$ as the comma category
-%% $\Set^{\Sigma^*}/N$. That is, an object of this category is a pair of
-%% a formula $\phi \in L$ and an object $S \in \mathcal
-%% P(\Sigma^*)/N(\phi)$. We can then use the equivalence $\mathcal
-%% P(\Sigma^*)/N(\phi) \cong \mathcal P(\int N(\phi))$ to get a simple
-%% description of such an $S$: it is simply a family of sets indexed by
-%% proofs $L(w,\phi)$:
-%% \[ \prod_{w\in\Sigma^*} L(w,\phi) \to \Set \]
-%% This category clearly comes with a projection functor $\pi : \mathcal
-%% G \to \mathcal L$ and then our goal is to define a section by using
-%% the universal property of $\mathcal L$.
-
-%% To this end we define
-%% \begin{enumerate}
-%% \item $(\phi, S) \otimes (\psi, T) = (\phi \otimes \psi, S\otimes T)$ where
-%%   \[ (S \otimes T)(w, p) = (w_1w_2 = w) \times (q_1,q_2 = p) \times S\,w_1\,q_1 \times T\,w_2\,q_2\]
-%% \item $(\phi, S) \multimap (\psi, T) = (\phi \multimap \psi, S \multimap T)$ where
-%%   \[ (S \multimap T)(w,p) = w' \to q \to S\,w'\,q \to T (ww') (p\,q) \]
-%% \item $\mu X. ??$ ??
-%% \end{enumerate}
-
-%% \pedro{We should conclude this section by explaining the relevance of the canonicity
-%% theorem. Could also be done before stating the theorem.}
-
 \section{Related and Future Work}
@@
 Our denotational semantics of \theoryabbv extends this to a much broader class of grammars.
-%% Krishnaswami and Yallop \cite{krishnaswami_typed_2019} provide a type system for
-%% demonstrating that carries proves of the LL(1) condition for context-free
-%% grammars. Despite being a type system over formal grammars, their interpretation
-%% of types is very different from the semantics of \theoryabbv.
-
-%% 1. More practical: integrating it into a larger verified development
-%%    - verified imperative implementation: Lambek logic ala separation
-%%      logic?
-%% 2. Semantic actions
-%% 3. Type systems as tree grammars
-
-%% \subsubsection{Kleene Algebra}
-%% Since the early works in the theory of formal languages, Kleene
-%% algebras have played an important role in its development. They
-%% generalize the operations known from regular languages by introducing
-%% operations generalizing language composition, language union and the
-%% Kleene star.  More generally, they are defined as inequational theory
-%% where the inequality is meant to capture language containment.  This
-%% theory is extremely successful, having found applications in concurrency theory
-%% \cite{hoare2009concurrent}, software-defined networks \cite{anderson2014netkat},
-%% theory of programming languages \cite{angus2001kleene}, compiler optimizations
-%% \cite{kozen2000certification} and more.
-
-%% A frequently fruitful research direction is exploring varying
-%% extensions of Kleene algebras, Kleene algebra with tests (KAT) being
-%% one of the most notable ones \cite{kozen1997kleene}. Our approach is
-%% radically different from most extensions, which usually aim at
-%% modifying or adding new operations to Kleene algebras, but still
-%% keeping it as an inequational theory. By adopting a category-theoretic
-%% treatment and allowing the ``order structure'' to encode more
-%% information than merely inequalities, we were able to extend Kleene
-%% algebra to reason about parsing as well.
+
 \paragraph{Dependent Linear Types}
@@
 
-%%    - verified imperative implementation: Lambek logic ala separation
-%%      logic?
 \paragraph{Relation to Separation Logic}
@@
 characters and the state of the memory is never allowed to evolve.
-
-%% Syntactically, our type theory can be viewed as a kind of ``separation
-%% type theory'' in that we care not just about provability of formulae
-%% but the constructive content of our linear terms, as these act as
-%% parsers. The design of \theoryabbv is based on the $\textrm{LNL}_D$
-%% calculus which explicitly was designed to be a dependent
-%% type-theoretic version of separation logic. Many concepts of
-%% separation logic then have direct analogues in our system. The
-%% characters $c$ are analogous to the ``points-to'' formulae $l \mapsto
-%% x$ in separation logic. Our $\otimes,\lto,\tol$ are analogous to the
-%% separating conjunction and magic-wand. The analogue of the persistence
-%% modality $\square$ of separation logic is $\bigoplus_{\_:\uparrow A} I$,
-%% which ``zeros out'' all but the parses from the empty string. If we
-%% think of the portions of the string we are parsing as our notion of
-%% resource, then these make sense as the ``resource-free'' elements of
-%% the type $A$.
 
@@
 framework for verified type checking and static analysis.
-
-%% While parsing typically refers to the generation of semantic
-%% objects from string input, many tasks in programming can be
-%% viewed as parsing of objects with more structure, such as
-%% trees with binding structure or graphs. Fundamental to the
-%% frontend of many
-%% programming language implementations are type systems. In
-%% particular, \emph{type checking}
-%% --- analogous to language recognition --- and \emph{typed
-%%   elaboration} --- analogous to parsing --- arise when
-%% producing a semantic object subject to some analysis. Just
-%% as our string grammars were given as functors from $\String$
-%% to $\Set$, we envision adapting the same philosophy
-%% to functors from a \emph{category of trees} to $\Set$ to craft a syntax
-%% that natively captures typed elaboration. This suggests an
-%% unusual sort of bunched type theory, where context extension
-%% no longer resembles concatenation of strings but instead
-%% takes on the form of tree constructors.
-
-% type system              ~~ Formal Grammar
-% typing derivation        ~~ parse tree
-% algorithmic type system  ~~ LL(1), LR(1) grammar
-% Uniqueness of derivations ~~ unambiguous grammar
-% type elaborator           ~~ semantic actions
 
@@
 
-  %% \max{do we need identity and composition of substitutions and their functoriality?}
-  %% By induction we can define an identity substitution
-  %% $\textrm{Subst}(\Gamma,\Gamma)$ which maps each variable to
-  %% itself. We write this substitution simply as $\Gamma$.
-  %% \begin{align*}
-  %%   \cdot \circ \gamma' &= \cdot\\
-  %%   (\gamma,M/x) \circ \gamma' &= (\gamma\circ \gamma') , M[\gamma']/x
-  %% \end{align*}
-
   It is straightforward, but laborious to establish that all forms in
@@
   \in \Subst(\Gamma',\Gamma)$ given in \cref{fig:non-linear-substitution,fig:substact}.
-  %% Further, these all satisfy identity and composition equations M[\Gamma] = M and M[\gamma][\gamma'] = M[(\gamma \circ \gamma')]
 \end{definition}
@@
   \Delta_2)$.
-  %% Again for any $\Gamma \vdash \Delta$, we can define an identity
-  %% substitution mapping each variable to itself. Additionally we can define composition of
 \end{definition}
@@
   \end{mathpar}
-
-  %% \max{identity and composition?}
 \end{definition}
@@
 First, expanding the left hand side, we have.
-% \newcommand{\sem}[1]{\left\llbracket{#1}\right\rrbracket}
-% \newcommand{\semg}[1]{\sem{#1}\gamma}
-% \newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
 \begin{proof}

### 2025-04-18 paper/paper.tex
artifact ref version

@@
 as the basis for our prototype implementation of \theoryabbv in
-Cubical Agda \cite{schaefer_2025_15049780}. The implementation is a shallow embedding, meaning that
+Cubical Agda \cite{schaefer_2025_15243560}. The implementation is a shallow embedding, meaning that
 rather than formalizing a syntax of \theoryabbv types and terms, we interpret
@@
 \section*{Data-Availability Statement}
-The Agda implementation of \theoryabbv and the formalization of our proofs are archived on Zenodo \cite{schaefer_2025_15049780} and the source code is available \href{https://github.com/maxsnew/grammars-and-semantic-actions/tree/PLDI25Artifact}{on GitHub}.
+The Agda implementation of \theoryabbv and the formalization of our proofs are archived on Zenodo \cite{schaefer_2025_15243560} and the source code is available \href{https://github.com/maxsnew/grammars-and-semantic-actions/tree/PLDI25Artifact}{on GitHub}.
 

### 2025-04-19 paper/paper.tex
Keep listings within margins

@@
 \usepackage[capitalise]{cleveref}
+
+% \usepackage{showframe}
 
@@
 backgroundcolor=\color{backcolour},
-frame=single
+frame=single,
+xleftmargin = 1.5mm ,
+xrightmargin = 1.5mm
 }
+
+\lstnewenvironment{floatlisting}{
+  \FloatBarrier
+  \lstset{
+    aboveskip=2mm
+  }
+}{}
 
@@
 \begin{figure}
-\begin{lstlisting}
+\begin{floatlisting}
 f : (*@$\uparrowcode$@*)('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)' (*@$\lto$@*) ('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\oplus$@*) '(*@\color{alphabetcolor}c@*)')
 f (a , b) = inl (a , b)
-\end{lstlisting}\begin{mathpar}
+\end{floatlisting}\begin{mathpar}
   \footnotesize
@@
     \vspace{0pt}
-  \begin{lstlisting}
+  \begin{floatlisting}
 data Trace : (s : Fin 3) (*@$\tocode$@*) L where
@@
 k (a , b) = 0to1 (1to1 a (1to2 b stop))
-\end{lstlisting}
+\end{floatlisting}
   \end{minipage}%
@@
 \begin{figure}
-\begin{lstlisting}
+\begin{floatlisting}
 data Trace(*@$_{\color{black}\N}$@*) : (s : N .states) (*@$\tocode$@*) L where
@@
   (*@$\color{linconstructorcolor}\epsilon$@*)cons : (*@$\uparrowcode$@*)(&[ t : N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)transitions ] (Trace(*@$_{\color{black}\N}$@*) (N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)dst t) (*@$\lto$@*) Trace(*@$_{\color{black}\N}$@*) (N .(*@$\color{nonlinconstructorcolor}\epsilon$@*)src t)))
-
 data Trace(*@$_{\color{black}\D}$@*) : (s : (*@\color{black}D@*) .states) (b : Bool) (*@$\tocode$@*) L where
@@
            &[ b : Bool ] ('(*@\color{alphabetcolor}c@*)' (*@$\lto$@*) Trace(*@$_{\color{black}\D}$@*) ((*@\color{black}D@*) .(*@$\color{nonlinconstructorcolor}\delta$@*) c s) b (*@$\lto$@*) Trace(*@$_{\color{black}\D}$@*) s b))
-\end{lstlisting}
+\end{floatlisting}
 \caption{Traces of an NFA $N$ and a DFA $D$}
@@
                                     (*@$\color{linconstructorcolor}\sigma$@*) b (Trace(*@$_{\color{black}\D}$@*) .cons c s b a t)
-
 print(*@$\color{black}_D$@*) : (s : (*@\color{black}D@*) .states) (*@$\tocode$@*) (*@$\uparrowcode$@*)(((*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\color{black}\D}$@*) s b) (*@$\lto$@*) String)
@@
 \begin{figure}
-\begin{lstlisting}
+\begin{floatlisting}
 data Dyck : L where
@@
   bal : (*@$\uparrowcode$@*)('(' (*@$\lto$@*) Dyck (*@$\lto$@*) ')' (*@$\lto$@*) Dyck (*@$\lto$@*) Dyck)
-\end{lstlisting}
+\end{floatlisting}
 \caption{The Dyck grammar as an inductive linear type}
@@
 \begin{figure}
-\begin{lstlisting}
+\begin{floatlisting}
 data Exp : L where
@@
   unexpected : (*@$\uparrowcode$@*)(&[ n : Nat ] ('(' (*@$\oplus$@*) ')' (*@$\oplus$@*) '#NUM#') (*@$\lto$@*) (*@$\color{lintycolor} \top$@*) (*@$\lto$@*) A n false)
-\end{lstlisting}
+\end{floatlisting}
 \caption{Associative arithmetic expressions and a corresponding lookahead automaton}

### 2025-04-21 paper/paper.tex
add missing dependence for margin fixes
final pass on intro and abstract

@@
 
+\usepackage{placeins} % for \FloatBarrier
+
 \lstnewenvironment{floatlisting}{
@@
 
-  We present a denotational semantics of our type theory that
-  interprets the types as a mathematical notion of formal
-  grammars. Based on this denotational semantics, we have made a
-  prototype implementation of \theoryabbv using a shallow embedding in
-  the Agda proof assistant. All of our examples parsers have been
-  implemented in this prototype implementation.
+  We present a denotational semantics of our type theory that interprets the
+  linear types as functions from strings to sets of abstract parse trees and
+  terms as parse transformers. Based on this denotational semantics, we have
+  made a prototype implementation of \theoryabbv using a shallow embedding in
+  the Agda proof assistant. All of our examples parsers have been implemented in
+  this prototype implementation.
 \end{abstract}
@@
 Parsing structured data from untrusted input is a ubiquitous task in
-computing. Any formally verified software system that interacts with
-the outside world must contain some parsing component. For example, in
-an extensive experiment finding bugs in C compilers
+computing. Any formally verified software system that interacts with the outside
+world must contain some parsing component, and unverified parsers can undermine the overall correctnes theorem for the system.
+For example, in an extensive experiment finding bugs in C compilers
 \cite{yangFindingUnderstandingBugs}, an early version of CompCert, the formally
-verified C compiler, only contained bugs in the then
-unverified parsing component \cite{leroy_formal_2009}. Bugs in parsers
-undermine the overall correctness theorem for a verified system: an
-incorrectly parsed C program will be compiled correctly, but this is
+verified C compiler, was found to have bugs only in the then
+unverified parsing component \cite{leroy_formal_2009}.
+The correctness of the compiler only showed that the abstract syntax tree would be compiled correctly, but this is
 not very useful if it did not correctly correspond to the actual
@@
 is a domain-specific language in which many grammar formalisms and
-their verified parsers can be implemented. For example, \theoryabbv is
-not a verified parser generator that compiles regular expressions to
+their verified parsers can be implemented. For example, \theoryabbv itself is
+not a verified parser generator compiling regular expressions to
 deterministic finite automata, but is instead a domain specific
-language \emph{for writing} such a verified parser generator.
+language in which we can write such a verified parser generator.
 
@@
 languages, where strings constitute a clear notion of resource
-that cannot be duplicated, reordered, or dropped. Moreover, the constructive
-aspect of \theoryabbv ensures that verification of parsers written in the
-calculus are \emph{correct-by-construction}. That is, our type system is rich
+that cannot be duplicated, reordered, or dropped. Moreover, \theoryabbv ensures that parsers written in the calculus are \emph{correct-by-construction}. That is, our type system is rich
 enough that typing derivations carry intrinsic proofs of parser correctness.
@@
 \theoryabbv. The core idea of the denotational semantics stems from an observation of Elliott:
-a formal grammar is a type-level predicate on strings that proves
-language membership \cite{elliottSymbolicAutomaticDifferentiation2021}. That is, a \emph{formal grammar} $A$ is a
+an abstract notion of formal grammar can be given by a ``proof-relevant'' predicate on strings \cite{elliottSymbolicAutomaticDifferentiation2021}. That is, a \emph{formal grammar} $A$ is a
 function $\StringSem \to \Set$ such that for a string $w$, $A~w$ is the
@@
 interpreted as an abstract formal grammar in this sense, and that
-linear terms are a kind of \emph{parse transformer}, a function that
+linear terms can be interpreted as a kind of \emph{parse transformer}, a function that
 takes a parse tree from one grammar to a parse tree in a different
@@
   \item A denotational semantics for \theoryabbv that shows that the
-    parsers are in fact verified to be correct, and the soundness of the
-    equational theory.
+    parsers are in fact verified to be correct, and the equational theory is sound.
 \end{itemize}

### 2025-04-21 paper/paper.tex
more tweaks, one in intro and one in section 2

@@
 \ifarxiv\else {Proofs and syntactic forms elided from this paper are presented
-    in full in the extended version of this paper \cite{schaefer2025intrinsicverificationparsersformal}.}\fi
+    in full in the extended version \cite{schaefer2025intrinsicverificationparsersformal}.}\fi
 
@@
 a parse of $\A$ over the string $\w$ or a
-parse of $\B$ over the string $\w$. Similarly, a parse of $\A \otimes \B$ for
+parse of $\B$ over the string $\w$, along with a tag \texttt{inl} or \texttt{inr} indicating which case was taken. A parse of $\A \otimes \B$ for
 $\w$ is a splitting of $w$ into two strings $\w_{\A}$ and $\w_{\B}$ with
@@
 
-A grammar $A$ derives the word $w$ if there exists a derivation
+A derivation of a word $w$ in a grammar $A$ is given by a term in our calculus that satisfies the typing
 $\internalize{w} \vdash e : A$,
@@
 type theory: weakening, contraction and exchange. Weakening allows
-for variables to go unused, while contraction allows for the same
-variable to be used twice, but in a parse tree, every character must
-be accounted for exactly once. That is, we want to prevent the following
-erroneous derivations,
-\(
-{a : \literal a , b : \literal b \not \vdash a : \literal a}\) and
+for variables to go unused, would allow for a character in an input string to be ignored, yielding the erroneous parse tree \( {a : \literal a , b : \literal b \not \vdash a : \literal a} \). 
+Contraction allows for the same variable to be used twice, yielding the erroneous parse
 \({a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a}
 \).
-
 Finally, the ordering of characters in a string cannot be ignored while
 parsing, so we omit the exchange rule because it would allow
-for variables in the context to be reordered. Likewise, we prevent the following derivation,
+for variables in the context to be reordered. This prevents the erroneous derivation,
 \(
@@
 \emph{inductive linear type} of linear lists, as shown in
-\cref{fig:kleenestarinductive}. Here $\A^{*} : \L$ means we are defining
-a \emph{linear} type. $\A^{*}$ has two constructors: $\nil$, which
+\cref{fig:kleenestarinductive}. Here $\A^{*} : \L$ means we are defining $\A^*$
+to be a \emph{linear} type. $\A^{*}$ has two constructors: $\nil$, which
 builds a parse of type $\A^{*}$ from nothing; and $\cons$, which
@@
 elimination principle for Kleene star, written
-as $\fold$. The parse transformer $h$ is more compactly presented in the pseudocode
+as $\fold$. The parse transformer $h$ is more intuitively presented in the pseudocode
 of \cref{fig:kleeneabstractproof} by pattern matching on the input and making an
@@
 implemented. To implement
-these algorithms in \theoryabbv, we represent automata as types in the same way we can represent regular expressions.
+these algorithms in \theoryabbv, we represent automata as types, just as we did with regular expressions.
 

### 2025-04-21 paper/paper.tex
appendix -> extended version

@@
 \ifarxiv{\cref{fig:full-non-linear-types,fig:jdg-eq-nonlinear} of
-    \cref{sec:syntax}}\else{the appendix}\fi.  First,
+    \cref{sec:syntax}}\else{the extended version of the paper}\fi.  First,
 we include universe types $U$ of small non-linear types and $L$ of linear
@@
 \ifarxiv{\cref{fig:small-nonlin-ty,fig:small-lin-ty} of
-    \cref{sec:syntax}}\else{the appendix}\fi. These universe
+    \cref{sec:syntax}}\else{the extended version of the paper}\fi. These universe
 types are needed so that we can define non-linear and linear types by recursion on natural
@@
 $Fin~n$ --- can be defined in terms of these primitives, and we give their
-encoding in \ifarxiv{\cref{sec:syntax}}\else{the appendix}\fi.
+encoding in \ifarxiv{\cref{sec:syntax}}\else{the extended version of the paper}\fi.
 
@@
 types is straightforward $\beta\eta$ equivalence and is included in
-\ifarxiv{\cref{fig:jdgeq} of \cref{sec:denotational}}\else{the appendix}\fi.
+\ifarxiv{\cref{fig:jdgeq} of \cref{sec:denotational}}\else{the extended version of the paper}\fi.
 
@@
 Further, we add equations in \ifarxiv{\cref{fig:spf-act} of \cref{sec:syntax}
-  }\else{the appendix }\fi that say that the
+  }\else{the extended version of the paper}\fi that say that the
 $\el$/$\map$ operations correspond to these descriptions of the
@@
 linear logic in monoidal categories \cite{seely89}. In
-\ifarxiv{\cref{sec:denotational}}\else{the appendix}\fi, we include the denotations
+\ifarxiv{\cref{sec:denotational}}\else{the extended version of the paper}\fi, we include the denotations
 of linear terms and prove that our semantics respects the equational theory of

### 2025-04-21 paper/paper.tex
small tweaks to section 3, remove contractions

@@
 constructors means that the constructors are not consumed
-upon usage, and so are \emph{non-linear} values themselves. That is, the
+upon usage, and so are \emph{non-linear} values themselves, which are not small. That is, the
 names $\nil$ and $\cons$ are function symbols that may be reused as many times
@@
 since they can be constructed using the equality type as $\sum_{x:X}
-f\,x=_Y g\,x$, but this construction can't be used for linear types
+f\,x=_Y g\,x$, but this construction cannot be used for linear types
 because it uses a \emph{dependent} version of the equality type, which
@@
 $\Var~x$ to one of the linear type variables, a constant expression
-that doesn't mention any type variables $K$, as well as tensor
+that does not mention any type variables $K$, as well as tensor
 products and additive conjunction and disjunction of type expressions.
 Further, we add equations in \ifarxiv{\cref{fig:spf-act} of \cref{sec:syntax}
-  }\else{the extended version of the paper}\fi that say that the
+  }\else{the extended version\fi that say that the
 $\el$/$\map$ operations correspond to these descriptions of the
@@
 $\ltonl{(\StringGram \lto (A \oplus \top))}$ using the ``option''
-monad. This allows for the possibility that the input string doesn't
+monad. This allows for the possibility that the input string does not
 parse, but is far too weak as a specification: we can trivially

### 2025-04-21 paper/paper.tex
"correctness"

@@
 computing. Any formally verified software system that interacts with the outside
-world must contain some parsing component, and unverified parsers can undermine the overall correctnes theorem for the system.
+world must contain some parsing component, and unverified parsers can undermine the overall correctness theorem for the system.
 For example, in an extensive experiment finding bugs in C compilers

### 2025-04-21 paper/paper.tex
more tweaks. One more section left

@@
 non-deterministic and deterministic automata, as well as some example
-context-free grammars and parsers based on $\LL{1}$ and $\LR{1}$
+$\LL{1}$ context-free grammars and parsers using stack-based
 automata. Throughout this paper \Agda~ will mark
@@
    $\lambda s. \mathsf{read}(!(s)) \equiv \lambda
-s. s$ where $!$ is the unique function $\ltonl(\StringGram\lto \top)$. That is, $\StringGram$ is strongly equivalent to $\top$.
+s. s$ where $!$ is the unique function $\ltonl(\StringGram\lto \top)$.
 \end{axiom}
@@
 integral to the theory of formal languages are faithfully represented
-in \theoryname. By encoding well-established formal grammar concepts, we ensure
-that our framework remains grounded in the
-foundational principles of formal language theory while opening the door to
-compositional formal verification of parsers.
+in \theoryname.
 
@@
 grammar is unambiguous when there is at most one parse tree for any
-input string. We can capture this notion as a type in \theoryabbv in a
-clever way:
+input string. We can capture this notion as a type in \theoryabbv as follows:
 \begin{definition}
@@
 
-In this section, we describe how to construct an intrinsically
+Next, we describe how to construct an intrinsically
 verified parser for regular expressions by compiling it to an NFA and
@@
 
-Because DFAs are deterministic, we are able to prove that their type
-of traces are unambiguous and define a parser for them. In particular
+Because DFAs are deterministic, we are able to prove that their types
+of traces are unambiguous and define a parser for them directly. In particular
 we show that for any start state $s$, ${\LinSigTy {b} {\Bool}
@@
 
-The Dyck language is an example of an $\textrm{LL}(0)$ language, one
-that can be parsed top-down with no
-lookahead \cite{rosenkrantz_properties_1970}. This means we can
-implement it simply as an \emph{infinite} state deterministic
+We implement the parser for the Dyck language using an \emph{infinite} state deterministic
 automaton, in \Cref{fig:DyckAutomaton}. Here the state is a ``stack''
@@
 
-This reification operation on functions $\StringSem \to U$ is incredibly expressive,
+This reification operation on functions $\StringSem \to U$ is very expressive,
 as it sidesteps our linear typing connectives and utilizes the whole of nonlinear
@@
 halts and accepts an input string. Then, $Reify~accepts$ is a linear type that
-captures precisely the string the same language as $T$. That is,
+captures precisely the string the same language as $T$. In general,
 $Lang(Reify~accepts)$ is recursively enumerable --- the most general class of
@@
 This category is equivalent to the slice category $\Set/\Sigma^*$ and
-as such is very well-behaved. It is complete, co-complete, Cartesian
+is very well-behaved. It is complete, co-complete, Cartesian
 closed and carries a monoidal biclosed structure. We will use these
 structures to model the linear types, terms and equalities in
-\theoryabbv. These categorical properties are precisely what is
-required to interpret all of the linear type and term constructors.
+\theoryabbv.
 
@@
 finally produce a parse of $\literal{a}^{*}$.
-A benefit of this shallow embedding is that
-the parsers are immediately available to a larger Agda development, as
-they are just normal Agda code. Although, even in the small example from
+Even in the small example from
 \cref{fig:kleeneabstractproof} we must manage additional complexity --- such as
 manually reassociations --- and
-this only grows more complex in larger programs. In future work we will look to
+this only grows more complex in larger programs. On the other hand, a benefit of this shallow embedding is that
+the parsers are immediately available to a larger Agda development, as
+they are just normal Agda code. In future work we will look to
 implement a type checker for a syntax closer to the presentation in
-this paper, while maintaining the goal of integration with an
+this paper, while maintaining the easy integration with an
 existing proof assistant.

### 2025-04-21 paper/paper.tex
one last tweak
remove comma for dangling clause

@@
 not a verified parser generator compiling regular expressions to
-deterministic finite automata, but is instead a domain specific
+deterministic finite automata but is instead a domain specific
 language in which we can write such a verified parser generator.
@@
 categorical semantics for their system, whereas we have only developed a single
-denotational model. It appears straightforward to adapt V\'ak\'ar's general
+denotational model. It should be straightforward to adapt V\'ak\'ar's general
 semantics approach to a non-commutative variant that would apply to

### 2025-04-21 paper/paper.tex
add which

@@
 type theory: weakening, contraction and exchange. Weakening allows
-for variables to go unused, would allow for a character in an input string to be ignored, yielding the erroneous parse tree \( {a : \literal a , b : \literal b \not \vdash a : \literal a} \).
+for variables to go unused, which would allow for a character in an input string to be ignored, yielding the erroneous parse tree \( {a : \literal a , b : \literal b \not \vdash a : \literal a} \).
 Contraction allows for the same variable to be used twice, yielding the erroneous parse

### 2025-05-13 src/runDyck.sh
fix nat error and add bash scripts

@@ -1,0 +1,3 @@
+agda --compile Examples/Dyck.agda
+ghc -iMAlonzo MAlonzo/Code/Examples/Dyck.hs
+./Dyck

### 2025-05-13 src/runNat.sh

@@ -1,0 +1,3 @@
+agda --compile Examples/Unicode/Nat.agda
+ghc -iMAlonzo MAlonzo/Code/Examples/Unicode/Nat.hs
+./Nat

### 2025-07-05 kleenecat/kleenecat.tex
Kleene categories

\documentclass[acmsmall,anonymous,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{quiver}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{amsmath}

\DeclareFontFamily{U}{dmjhira}{}
\DeclareFontShape{U}{dmjhira}{m}{n}{ <-> dmjhira }{}

\DeclareRobustCommand{\yo}{\text{\usefont{U}{dmjhira}{m}{n}\symbol{"48}}}

\usepackage{xstring} % Need it for strlen
\newcommand{\cat}[1]{
  \relax
  \StrLen{#1}[\arglength]
  \ifnum\arglength=1
  \mathcal{#1}
  \else
  \mathbf{#1}
  \fi
}

\newcommand{\Set}{\cat{Set}}
\newcommand{\lto}{\multimap}
\newcommand{\grammar}{\Set^{\cat{\Sigma^*}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

\begin{document}
\title{Kleene Categories}
\author{Steven Schaefer}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{stschaef@umich.edu}

\author{Max S. New}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{maxsnew@umich.edu}

\author{Pedro H. Azevedo de Amorim}
\affiliation{
  \department{Department of Computer Science}
  \institution{University of Oxford}
  \country{UK}
}
\email{pedro.azevedo.de.amorim@cs.ox.ac.uk}

\begin{abstract}
  We propose a categorification of Kleene algebras and study some of its basic properties.
\end{abstract}

\maketitle

\section{Kleene Category}

Kleene algebras are an important tool in the theory of regular
languages. More broadly, they serve as a theoretical substrate to
studying various kinds of formal languages. Formally, they are a tuple
$(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is a set, $+$ and $\cdot$
are binary operations over $A$, $(-)^*$ is a function over $A$, and
$1$ and $0$ are constants. These structures satisfy the axoims depicted
in Figure~\ref{fig:axioms}.

\begin{figure}
  \begin{align*}
    x + (y + z) &= (x + y) + z & x + y &= y + x\\
    x + 0 &= x & x + x &= x\\
    x(yz) &= (xy)z & x1 &= 1x = x\\
    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
    x0 &= 0x = x & & \\
    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
  \end{align*}
  \label{fig:axioms}
  \caption{Kleene algebra axioms}
\end{figure}

The addition operation can be used to define the partial order
structure $a \leq b$ if $a + b = b$. In the theory of formal languages
this order structure can be used to model language containment. In this
section, we want to categorify the concept of Kleene algebra and
build on top of it in order to define an abstract theory of parsing.
We start by defining \emph{Kleene categories}.

\begin{definition}
  A Kleene category is a distributive monoidal category $\cat{K}$
  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that the canonical
  morphisms $\mu X.\, F_{A, B}(X) \to B \otimes (\mu   X.\, F_{A, 1})$
  $\mu X.\, G_{A, B}(X) \to (\mu   X.\, G_{A, 1})\otimes B $ are iso.
\end{definition}

As a sanity check, note that Kleene algebras are indeed examples of
Kleene categories.

\begin{example}
  Every Kleene algebra, seen a posetal category, is a Kleene category.
\end{example}

An unexpected example comes from the theory of substructural logics.

\begin{example}
  The opposite category of every Kleene category is a model of a variant of
  conjunctive ordered logic, where the Kleene star plays the role of the ``of
  course'' modality from substructural logics.
\end{example}

This proposed definition is sensible. After all, the proposed axioms
are a direct translation of the Kleene algebra axioms to a categorical
setting. Its most convoluted aspect is the axiomatization of the Kleene
star as a family os initial algebras satisfying certain
isomorphisms. If the Kleene category $\cat{K}$ has more structure,
then these isomorphisms hold ``for free''.

\begin{theorem}
  \label{th:kleeneclosed}
  Let $\cat{K}$ be a Kleene category such that it is also monoidal
  closed.  Then, the initial algebras isomorphisms hold automatically.
\end{theorem}
\begin{proof}
  We prove this by the unicity (up-to isomorphism) of initial
  algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
  (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
  X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
  (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
    1}(X))$.

  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
  we want to show that there is a unique algebra morphism $h : B
  \otimes \mu X.\, F_{A,1} \to Y$. We can show existence and uniqueness
  by showing that the diagram on the left commute if, and only if,
  the diagram on the right commutes:

  This equivalence follows by using the adjunction structure given
  by the monoidal closed structure of $\cat{K}$. A completely analogous
  argument for $G_{A,B}$ also holds.
\end{proof}

This result feels similar in spirit to the definition of action
algebras, which are algebras where the product also has adjoint
operations which results in the Kleene star being more easily
axiomatized \cite{kozen1994}. We are now ready to prove that our
concept of formal grammars fit nicely within our categorical
framework. We start by presenting a well-known construction
from presheaf categories.

\begin{definition}
  Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
  two functors $\cat{C} \to \Set$. Their Day convolution tensor
  product is defined as the following coend formula:
  \[
  (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z) 
  \]
  Dually, its internal hom is given by the following end formula:
  \[
  (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
  \]
\end{definition}

\begin{lemma}[\cite{day1970}]
  Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
  monoidal closed.
\end{lemma}

%% \begin{theorem}
%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
%%   The functor category $[A, \cat{K}]$.
%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
%% \end{theorem}
\begin{theorem}
  If $\cat{C}$ is a locally small monoidal category, then
  $\Set^{\cat{C}}$ is a Kleene category.
\end{theorem}
\begin{proof}

  By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
  is a presheaf category, it has coproducts, and since the tensor
  is a left adjoint, it preserves colimits and, therefore, it is
  a distributive category.

  As for the Kleene star, since presheaf categories admit small colimits,
  the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
  defined as the filtered colimit of the diagrams:

  From Theorem~\ref{th:kleeneclosed} it follows that these initial
  algebras satisfy the required isomorphisms and this concludes the
  proof.
\end{proof}

\begin{corollary}
  For every alphabet $\Sigma$, the presheaf category $\grammar$
  is a Kleene category.
\end{corollary}
\begin{proof}
  Note that string concatenation and the empty string make the
  discrete category $\Sigma^*$ a strict monoidal category.
\end{proof}

\begin{definition}
  Let $K$ be a Kleene category over alphabet $\Sigma$. This
  induces a strong monoidal functor $\Sigma^* \to K$. The
  nerve functor $w \mapsto K(w, -)$ is the grammar interpretation
  functor of $K$.
\end{definition}

\section{Proof-Relevant Deterministic Automata}

Now, we will consider a categorification of deterministic automata.

\begin{definition}
  A proof-relevant DFA is an $F$-coalgebra, for the functor
  $F(\cat{C}) = \Set \times (\Sigma \Rightarrow \cat{C})$.
\end{definition}

Assuming that the underlying category $\cat{C}$ is not discrete,
arrows $f : X \to Y$ in $\cat{C}$ should be thought of as a proof that
the state $X$ is related to the $Y$. This can be seen as a
generalization of simulation relations. After all, functoriality of $F
: \cat{C} \times \Sigma \to \cat{C}$ implies that $F(f, id_c): F(X, c)
\to F(Y, c)$, i.e. after one transition, the end states will also be
related.

\begin{definition}
  We can equip $\grammar$ with the $F$-coalgebra structure $\langle \alpha_1, \alpha_2\rangle$,
  where $\alpha_1(F) = F(\varepsilon)$ and $\alpha_2(F, c) = \yo(c) \lto F$.
\end{definition}

\begin{theorem}
  The morphism $\alpha : \grammar \to F(\grammar)$ is the final
  $F$-coalgebra.
\end{theorem}
\begin{proof}
  Given an $F$-coalgebra $\langle \alpha, \delta\rangle : \cat{C} \to \Set \times \cat{C}^\Sigma$, we can define the functor $\cat{C} \times \Sigma^* \to \Set$ much like in the DFA case:
  by induction on $\Sigma^*$.

\end{proof}

\begin{comment}
\subsection{DFAs with simulation relation}

Let $\langle \tau, \delta\rangle : X \to 2 \times X^\Sigma$ be a DFA
over an alphabet $\Sigma$.

\begin{definition}
  A binary relation $\mathcal{R} \subseteq X \times X$ is a simulation if
  \begin{itemize}
  \item It is a preorder.
  \item If $x_1 \mathcal{R} x_2$ and $x_1$ is an accept state, so is $x_2$.
  \item If $x_1 \mathcal{R} x_2$ then for every $a : \Sigma$,
    $\delta(x_1, a) \mathcal{R} \delta(x_2, a)$.
  \end{itemize}
\end{definition}

We refine the notion of DFA morphism so that it also preserves the
simulation relation.

\begin{definition}
A morphism between DFAs with simulation $f : (X, \tau, \delta,
\mathcal{R}) \to (Y, \tau', \delta', \mathcal{S})$ is a function $f :
X \to Y$ such that it is a DFA morphism and if $x_1 \mathcal{R} x_2$,
then $f(x_1) \mathcal{S} f(x_2)$.
\end{definition}

These can be naturally organized as a category, which we denote
$\cat{SimDFA}$.

\begin{theorem}
  The forgetful functor $U : \cat{SimDFA} \to \cat{DFA}$ is a fibration.
\end{theorem}
\begin{proof}
  Since this functor is faithful, proving the universal property of
  fibrations is simpler. Consider the following lifting problem:

  We define the DFA with relation $(X, \mathcal{S}^*)$, where
  $\mathcal{S} = \{ (x_1, x_2) \mid f(x_1) \mathcal{S} f(x_2)\}$.
  We now prove that $\mathcal{S}^*$ is a simulation relation.
  \begin{description}
  \item[Preorder:] It follows directly from the fact that $\mathcal{S}$
    is a preorder.
  \item[Preserves acceptance:] Assume that $f(x_1) \mathcal{S} f(x_2)$
    and that $x_1$ is an accepting state. Since $f$ is a DFA morphism,
    it follows that $f(x)$ is an accepting state if, and only if, $x$
    is accepting. Since by assumption $\mathcal{S}$ is a simulation
    relation, we can conclude that $f(x_2)$ is an accepting state and
    so is $x_2$.
  \item[Stable under transition:] Assume that $f(x_1) \mathcal{S} f(x_2)$,
    $\delta_a(x_1) = x'_1$ and $\delta_a(x_2) = x'_2$. We want to
    show that $f(x'_1) \mathcal{S} f(x'_2)$. By assumption that $f$ is
    a DFA morphism, we have $f \circ \delta_a = \delta'_a \circ f$,
    for every character $a : \Sigma$. Therefore, we have to prove that
    $\delta'_a(f(x_1)) \mathcal{S} \delta'_a(f(x_2))$. This follows
    by the assumption that $\mathcal{S}$ is a simulation relation.
  \end{description}

  Therefore $(X, \mathcal{S}^*)$ is indeed a lifting of $X$.
  Showing that it is the Cartesian lifting follows by unfolding
  the definitions and using the fact that $U$ is faithful.
\end{proof}
\end{comment}

\section{Proposed construction for free Kleene categories}

For now, we ignore the subtleties behind pseudomonads, 2-monads,
lax algebras, etc. We begin by defining pre Kleene categories.

\begin{definition}
  A pre Kleene category is a distributive monoidal category such that
  the functors $F_{A, B}$ and $G_{A, B}$ have initial algebras, but
  they do not necessarily satisfy the isomorphism condition.
\end{definition}

These objects can pretty naturally be organized as a 2-category
$\cat{PKC}$: 1-cells are functors that strictly preserves all of the
structure while 2-cells are just natural transformations.

Let $\cat{C}$ be a category. It is more or less straightforward to
define free pre Kleene category $PKC(\cat{C})$. We can do that by
using the internal language of pre Kleene categories. Concretely, it
will be a type theory with (non-commutative) tensors, sum types and
inductive types corresponding to the initial algebras for $F_{A, B}$
and $G_{A, B}$, for every $A$, $B$. By type theoretic reasons, this
structure forces the coproduct to distribute over the tensor, where we
can write the morphism $A \otimes (B + C) \to (A \otimes B) + (A
\otimes C) $ as the term
\begin{align*}
  &z \vdash \mathsf{let}\, x \otimes y = z\, \mathsf{in}\\
  &\mathsf{case}\, y \, \mathsf{with}\\
  &| \mathsf{in}_1\, y_1 \Rightarrow \mathsf{in}_1\, (x \otimes y_1)\\
  &| \mathsf{in}_2\,y_2\Rightarrow \mathsf{in}_2 \,(x \otimes y_2)\\
  &\mathsf{end}
\end{align*}
%
We can prove directly by the equational theory of this type theory
that this is an inverse to the canonical map going in the other
direction. The inductive nature of type theories should make the
following theorem pretty straightforward to prove:
%
\begin{theorem}
  The forgetful functor $\cat{PKC} \to \cat{Cat}$ has a left adjoint
  in $\cat{2Cat}$.
\end{theorem}

Unfortunately, we can see that such a free construction will not be a
Kleene category, since there is no way of constructing the Kleene star
isomorphisms. That being said, there is a comparison map
$\mu F_{A,B} \to A \mu F_{1, B}$ for $F$ and $G$. This leads me to conjecture
the following:
%
\begin{conjecture}
  The forgetful functor $\cat{KC} \to \cat{PKC}$ is a reflective
  subcategory
\end{conjecture}

If the theorem above is true, we get a Kleene category
monad on $\cat{Cat}$ by composing the adjunctions above.
While I'm not sure how to define the reflection, Theorem~\ref{th:kleeneclosed}
might be important in our construction. I remember asking Daniel Gratzer
about this and he told me something about mapping a pre Kleene
category $\cat{C}$ to $\widehat{\cat{C}}$ and ``carving it out'' in order
to obtain an appropriate subcategory. Note that $\widehat{C}$ is monoidal
closed, so it will be Kleene as well.

\section{Revisiting Brzozowski's Distributive Law}

It has been observed that Brzozowski's derivative can be more conceptually
explained as a distributive law between the Kleene algebra monad and the
DFA functor $2 \times (\Sigma \Rightarrow -)$.

\begin{definition}
  We can equip the two element set $2 = \{0, 1 \}$ with the following KA structure
  \begin{itemize}
    \item $b_1 + b_2 = b_1 \lor b_2$
    \item $0 = 0$
    \item $b_1 b_2 = b_1 \land b_2$
    \item $1 = 1$
    \item $b^* = 1$
  \end{itemize}
\end{definition}

Since the axiomatization of Kleene algebras over an alphabet $\Sigma$
is a Horn theory, the category of Kleene algebras and Kleene algebra
morphisms is monadic over $\Set$. Let $E_\Sigma : \Set \to \Set$. be the 
Kleene algebra monad. It is possible to show the following theorem.

\begin{theorem}
  There is a distributive law $E_\Sigma(2 \times (\Sigma \Rightarrow X)) \to 2 \times (\Sigma \Rightarrow E_\Sigma(X))$.
\end{theorem}
\begin{proof}
  The component $E_\Sigma(2 \times (\Sigma \Rightarrow X)) \to 2$ is
  defined using the universal property of free algebraic structures,
  i.e. it suffices to define a morphism $2 \times (\Sigma \Rightarrow
  X)) \to 2$, which we choose it to be the first projection
  $\pi_1$. The component $E_\Sigma(2 \times (\Sigma \Rightarrow X))
  \to (\Sigma \Rightarrow E_\Sigma(X))$ requires us to be more
  clever. There is a ``non-standard'' Kleene algebra structure on
  $(\Sigma \Rightarrow E_\Sigma(X)) \times E_\Sigma(X)$ that makes
  this possible.
\end{proof}

\section{Generalization}

We now want to prove that the distributive law above generalizes to
the Kleene category case. We begin by showing that the ``non-standard''
structure defined in the previous section can be extended to the
proof-relevant setting.

\begin{lemma}
  Let $\cat{K}$ and $\cat{L}$ be two Kleene categories, $F : \cat{L}
  \to \cat{K}$ and $G : \cat{L} \to \cat{K}$ two Kleene category
  morphisms. We can equip $\cat{K} \times \cat{L}$ with the following
  Kleene category structure:
  \begin{itemize}
  \item $(A, X) + (B, Y) = (A + B, X + Y)$
  \item $0 = (0, 0)$
  \item $(A, X) \otimes (B, Y) = (A \otimes F(Y) + G(X) \otimes B, X \otimes Y)$
  \item $(A, X)^* = (A \otimes F(X^*), X^*)$
  \end{itemize}
\end{lemma}
%% \begin{proof}
%%   Since the morphism structure is defined componentwise, it is
%%   immediate to show that the proposed cocartesian structure is
%%   valid. For the monoidal structure, the bifunctoriality of $\otimes$
%%   follows from the bifunctoriality of the monoidal structures
%%   $\otimes$ and $+$ in $\cat{K}$. Next, we define the coherence
%%   isomorphisms. As suggested by the action on the $\cat{L}$ component,
%%   such isomorphisms are the same as in $\cat{L}$. The $\cat{K}$
%%   components of these morphisms are more interesting:
%% \end{proof}

With this lemma and the 2-monadicity of Kleene categories over $\cat{Cat}$,
we can define the distributive law between the proof-relevant
DFA functor and the Kleene category 2-monad.

\paragraph{Possible motivations}

People have considered derivative-based parsing before, this would provide
a semantic counterpart to those implementations.

\section{Gluing Kleene categories}

The following theorems should be true
\begin{theorem}
  Given Kleene categories $\cat{L}$, $\cat{K}$, $\cat{K}'$
  and functors $F : \cat{L} \to \cat{K}$, $p : \cat{K}' \to \cat{K}$
  such that $F$ is lax monoidal such that it laxly preserves the initial
  algebras and $p$ is a monoidal fibration strictly preserving the
  Kleene category structure, then in the pullback diagram
  % https://q.uiver.app/#q=WzAsNCxbMCwxLCJcXGNhdHtMfSJdLFsxLDEsIlxcY2F0e0t9Il0sWzEsMCwiXFxjYXR7S30nIl0sWzAsMCwiXFxjYXR7TH0nIl0sWzAsMSwiRiIsMl0sWzIsMSwicCJdLFszLDJdLFszLDBdLFszLDEsIiIsMSx7InN0eWxlIjp7Im5hbWUiOiJjb3JuZXIifX1dXQ==
\[\begin{tikzcd}
	{\cat{L}'} & {\cat{K}'} \\
	{\cat{L}} & {\cat{K}}
	\arrow[from=1-1, to=1-2]
	\arrow[from=1-1, to=2-1]
	\arrow["\lrcorner"{anchor=center, pos=0.125}, draw=none, from=1-1, to=2-2]
	\arrow["p", from=1-2, to=2-2]
	\arrow["F"', from=2-1, to=2-2]
\end{tikzcd}\]
$\cat{L}'$ is a Kleene category and the functor $\cat{L}'\to \cat{L}$ is a
monoidal fibration strictly preserving the Kleene category structure.
\end{theorem}
\begin{proof}
  It is known from linear logic reasons that the pullback is distributive monoidal.
  The Kleene star should be defined much like the exponential in glued models of
  linear logic. And the projection functor is a monoidal fibration because those
  are stable under pullback.
\end{proof}

Recently I have been enjoying comma categories a bit more, which leads me to the follwing
conjecture:

\begin{conjecture}
  The 2-category of Kleene categories has comma objects.
\end{conjecture}

\end{document}

### 2025-07-05 kleenecat/kleenecat.tex
Different axiomatization of Kleene categories using parameterized list objects

@@
 \newcommand{\grammar}{\Set^{\cat{\Sigma^*}}}
+\newtheorem*{construction}{Construction}
 
@@
 
-\section{Kleene Category}
+\section{Kleene Algebra and Kleene Categories}
 
@@
 languages. More broadly, they serve as a theoretical substrate to
-studying various kinds of formal languages. Formally, they are a tuple
+studying various kinds of formal languages. Formally, a Kleene algebra is a tuple
 $(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is a set, $+$ and $\cdot$
@@
 
-The addition operation can be used to define the partial order
-structure $a \leq b$ if $a + b = b$. In the theory of formal languages
-this order structure can be used to model language containment. In this
-section, we want to categorify the concept of Kleene algebra and
-build on top of it in order to define an abstract theory of parsing.
-We start by defining \emph{Kleene categories}.
+The addition operation induces a partial order structure where $a \leq
+b$ if $a + b = b$. In the theory of formal languages this order
+structure models language containment.
+
+\begin{example}
+  The free Kleene algebra over a set $\Sigma$ is given by regular
+  expressions with alphabet $\Sigma$, quotiented by the Kleene algebra
+  axioms.
+\end{example}
+
+Kleene algebra provides an algebraic foundation for the implementation
+of langauge \emph{recognizers}, i.e., programs that determine whether
+or not a string is a member of a formal language. On the other hand,
+the fact that $+$ is idempotent means that a Kleene algebra cannot
+distinguish between multiple parses of the same string, and so Kleene
+algebra is insufficient for the discussion of \emph{parsers} which may
+select one of several different possible parses in the case of
+ambiguous grammars.
+
+Our goal is to provide a generalization of Kleene algebra that is
+``proof-relevant'', and so can provide a foundation for parsing
+regular expressions in the same way that Kleene algebra does for
+recognition. To do this we need to move beyond the view of a Kleene
+algebra as algebraic structure over sets, and instead as a kind of
+structured \emph{poset}. We then generalize this to a Kleene
+\emph{category}, where the morphisms of the category correspond to not
+just language inclusion but \emph{formal reductions} of grammars, and
+isomorphism is not language equivalence but instead \emph{strong
+equivalence} of grammars.
+
+The most complex part of this generalization is the appropriate
+generalization of the Kleene star. For that we make the following
+definitions
+\begin{definition}
+  Let $\cat C$ be a monoidal category. Given an object $A \in \cat C$,
+  a \textbf{parameterized cons-list object} consists of
+  \begin{enumerate}
+  \item an object $L \in \cat C$
+  \item with morphisms $\textrm{nil} : I \to L$ and $\textrm{cons} : A \otimes L \to L$
+  \item such that for every $P,B \in \cat C$, and morphisms $n : P \to
+    B$ and $c : A \otimes B \to B$ there is a unique morphism $\textrm{foldr}(n,c) : L
+    \otimes P \to B$ satisfying
+
+    % https://q.uiver.app/#q=WzAsNixbMSwwLCJMIFxcb3RpbWVzIFAiXSxbMCwwLCJJIFxcb3RpbWVzIFAiXSxbMCwxLCJQIl0sWzIsMCwiQSBcXG90aW1lcyBMIFxcb3RpbWVzIFAiXSxbMiwxLCJBIFxcb3RpbWVzIEIiXSxbMSwxLCJCIl0sWzEsMiwiIiwwLHsibGV2ZWwiOjIsInN0eWxlIjp7ImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMiw1LCJuIl0sWzQsNSwiYyIsMl0sWzMsNCwiQSBcXG90aW1lcyBcXHRleHRybXtmb2xkcn0obixjKSJdLFszLDAsIlxcdGV4dHJte2NvbnN9IFxcb3RpbWVzIFAiLDJdLFsxLDAsIlxcdGV4dHJte25pbH0gXFxvdGltZXMgUCJdLFswLDUsIlxcdGV4dHJte2ZvbGRyfShuLGMpIl1d
+\[\begin{tikzcd}[ampersand replacement=\&]
+	{I \otimes P} \& {L \otimes P} \& {A \otimes L \otimes P} \\
+	P \& B \& {A \otimes B}
+	\arrow["{\textrm{nil} \otimes P}", from=1-1, to=1-2]
+	\arrow[equals, from=1-1, to=2-1]
+	\arrow["{\textrm{foldr}(n,c)}", from=1-2, to=2-2]
+	\arrow["{\textrm{cons} \otimes P}"', from=1-3, to=1-2]
+	\arrow["{A \otimes \textrm{foldr}(n,c)}", from=1-3, to=2-3]
+	\arrow["n", from=2-1, to=2-2]
+	\arrow["c"', from=2-3, to=2-2]
+\end{tikzcd}\]
+  \end{enumerate}
+
+  A \textbf{cons-list object} is like a parameterized cons-list object but where $P$ can only be $I$.
+
+  A \textbf{parameterized snoc-list object} is a cons-list object in
+  the dual monoidal structure $\cat C^{co}$. In this case we call the
+  $\textrm{cons}$ instead $\textrm{snoc}$ and the $\textrm{foldr}$
+  morphism $\textrm{foldl}$. A non-parameterized snoc-list object is
+  defined similarly.
+\end{definition}
+
+\begin{construction}
+  Given a cons-list object for $A$, we can define a morphism $\textrm{snoc} : L \otimes A \to L$ as $\textrm{foldr}(\eta, \textrm{cons})$
+  where $\eta : A \to L$ is the composition
+  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBIl0sWzEsMCwiQSBcXG90aW1lcyBJIl0sWzIsMCwiQSBcXG90aW1lcyBMIl0sWzMsMCwiTCJdLFswLDEsIiIsMCx7ImxldmVsIjoyLCJzdHlsZSI6eyJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMiwiQSBcXG90aW1lcyBcXHRleHRybXtuaWx9Il0sWzIsMywiXFx0ZXh0cm17Y29uc30iXV0=
+\[\begin{tikzcd}[ampersand replacement=\&]
+	A \& {A \otimes I} \& {A \otimes L} \& L
+	\arrow[equals, from=1-1, to=1-2]
+	\arrow["{A \otimes \textrm{nil}}", from=1-2, to=1-3]
+	\arrow["{\textrm{cons}}", from=1-3, to=1-4]
+\end{tikzcd}\]
+\end{construction}
 
 \begin{definition}
-  A Kleene category is a distributive monoidal category $\cat{K}$
-  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
-  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
-  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that the canonical
-  morphisms $\mu X.\, F_{A, B}(X) \to B \otimes (\mu   X.\, F_{A, 1})$
-  $\mu X.\, G_{A, B}(X) \to (\mu   X.\, G_{A, 1})\otimes B $ are iso.
+  A Kleene category is a distributive monoidal category $\cat{K}$ such
+  that for every object $A$, there is a cons-list object
+  $(A^*,\textrm{nil},\textrm{cons})$ such that $(A^*, \textrm{nil},
+  \textrm{snoc})$ is a snoc-list object.
 \end{definition}
@@
 
-\begin{example}
-  Every Kleene algebra, seen a posetal category, is a Kleene category.
-\end{example}
+\begin{lemma}
+  A Kleene algebra is equivalent to a posetal Kleene category.
+\end{lemma}
 
+% This isn't an example of a Kleene category, it's an example of a model of conjunctive ordered logic
 An unexpected example comes from the theory of substructural logics.
@@
 are a direct translation of the Kleene algebra axioms to a categorical
-setting. Its most convoluted aspect is the axiomatization of the Kleene
-star as a family os initial algebras satisfying certain
-isomorphisms. If the Kleene category $\cat{K}$ has more structure,
-then these isomorphisms hold ``for free''.
+setting. Its most convoluted aspect is the axiomatization of the
+Kleene star as a parameterized list and snoc object. If the Kleene
+category $\cat{K}$ has more structure, then these isomorphisms hold
+``for free''.
+
+\begin{definition}
+  A pre-Kleene category is a distributive monoidal category with a (non-parameterized) cons-list object.
+\end{definition}
 
 \begin{theorem}
-  \label{th:kleeneclosed}
-  Let $\cat{K}$ be a Kleene category such that it is also monoidal
-  closed.  Then, the initial algebras isomorphisms hold automatically.
+  If $\cat K$ is a pre-Kleene category that is also biclosed, then it is a Kleene category.
 \end{theorem}
 \begin{proof}
-  We prove this by the unicity (up-to isomorphism) of initial
-  algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
-  (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
-  X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
-  (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
-    1}(X))$.
+  TODO: closed implies parameterized for free and that cons-list and snoc-list objects are equivalent.
+\end{proof}
 
-  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
-  we want to show that there is a unique algebra morphism $h : B
-  \otimes \mu X.\, F_{A,1} \to Y$. We can show existence and uniqueness
-  by showing that the diagram on the left commute if, and only if,
-  the diagram on the right commutes:
+%% \begin{theorem}
+%%   \label{th:kleeneclosed}
+%%   Let $\cat{K}$ be a Kleene category such that it is also monoidal
+%%   closed.  Then, the initial algebras isomorphisms hold automatically.
+%% \end{theorem}
+%% \begin{proof}
+%%   We prove this by the unicity (up-to isomorphism) of initial
+%%   algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
+%%   (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
+%%   X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
+%%   (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
+%%     1}(X))$.
 
-  This equivalence follows by using the adjunction structure given
-  by the monoidal closed structure of $\cat{K}$. A completely analogous
-  argument for $G_{A,B}$ also holds.
-\end{proof}
+%%   Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
+%%   we want to show that there is a unique algebra morphism $h : B
+%%   \otimes \mu X.\, F_{A,1} \to Y$. We can show existence and uniqueness
+%%   by showing that the diagram on the left commute if, and only if,
+%%   the diagram on the right commutes:
+
+%%   This equivalence follows by using the adjunction structure given
+%%   by the monoidal closed structure of $\cat{K}$. A completely analogous
+%%   argument for $G_{A,B}$ also holds.
+%% \end{proof}
 
@@
 \end{definition}
+Note that in the case the category $\cat C$ is the discrete category
+of a monoid, the Day convolution admits a simpler description where
+the coend degenerates to a coproduct and the end ot a product.
 
@@
 \begin{proof}
-
   By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
@@
   by induction on $\Sigma^*$.
-
 \end{proof}
@@
 For now, we ignore the subtleties behind pseudomonads, 2-monads,
-lax algebras, etc. We begin by defining pre Kleene categories.
-
-\begin{definition}
-  A pre Kleene category is a distributive monoidal category such that
-  the functors $F_{A, B}$ and $G_{A, B}$ have initial algebras, but
-  they do not necessarily satisfy the isomorphism condition.
-\end{definition}
-
-These objects can pretty naturally be organized as a 2-category
-$\cat{PKC}$: 1-cells are functors that strictly preserves all of the
-structure while 2-cells are just natural transformations.
+lax algebras, etc.
 
 Let $\cat{C}$ be a category. It is more or less straightforward to
-define free pre Kleene category $PKC(\cat{C})$. We can do that by
-using the internal language of pre Kleene categories. Concretely, it
+define the free Kleene category $PKC(\cat{C})$. We can do that by
+using the internal language of Kleene categories. Concretely, it
 will be a type theory with (non-commutative) tensors, sum types and
-inductive types corresponding to the initial algebras for $F_{A, B}$
-and $G_{A, B}$, for every $A$, $B$. By type theoretic reasons, this
+a primitive type of lists $L(A)$ for every $A$. By type theoretic reasons, this
 structure forces the coproduct to distribute over the tensor, where we
@@
 %
+The list type includes primitive terms for $\textrm{cons}$,
+$\textrm{nil}$, $\textrm{foldr}$ and $\textrm{foldl}$ where
+$\textrm{foldr}$ and $\textrm{foldl}$ allow for a context to be used
+in the construction of the nil case, but no free variables can be used
+in the cons case:
+\begin{mathpar}
+  \inferrule
+  {\Gamma \vdash L : A^* \and \Delta \vdash N : B \and a:A,b:B \vdash M : B
+  }
+  {\Gamma,\Delta \vdash \textrm{foldr}(L,N,a.b.M) : B}
+\end{mathpar}
+
 We can prove directly by the equational theory of this type theory
@@
 \begin{theorem}
-  The forgetful functor $\cat{PKC} \to \cat{Cat}$ has a left adjoint
+  The forgetful functor $\cat{KC} \to \cat{Cat}$ has a left adjoint
   in $\cat{2Cat}$.
 \end{theorem}
-
-Unfortunately, we can see that such a free construction will not be a
-Kleene category, since there is no way of constructing the Kleene star
-isomorphisms. That being said, there is a comparison map
-$\mu F_{A,B} \to A \mu F_{1, B}$ for $F$ and $G$. This leads me to conjecture
-the following:
-%
-\begin{conjecture}
-  The forgetful functor $\cat{KC} \to \cat{PKC}$ is a reflective
-  subcategory
-\end{conjecture}
-
-If the theorem above is true, we get a Kleene category
-monad on $\cat{Cat}$ by composing the adjunctions above.
-While I'm not sure how to define the reflection, Theorem~\ref{th:kleeneclosed}
-might be important in our construction. I remember asking Daniel Gratzer
-about this and he told me something about mapping a pre Kleene
-category $\cat{C}$ to $\widehat{\cat{C}}$ and ``carving it out'' in order
-to obtain an appropriate subcategory. Note that $\widehat{C}$ is monoidal
-closed, so it will be Kleene as well.
 

### 2025-07-10 kleenecat/kleenecat.tex

@@
 \usepackage{amsmath}
+\usepackage{stmaryrd}
 
@@
 
+\newcommand{\FKC}{\textrm{F}_{\textrm{KC}}}
+\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
 \newcommand{\Set}{\cat{Set}}
@@
 \newcommand{\grammar}{\Set^{\cat{\Sigma^*}}}
+\newcommand{\id}{\textrm{id}}
 \newtheorem*{construction}{Construction}
@@
 Let $\cat{C}$ be a category. It is more or less straightforward to
-define the free Kleene category $PKC(\cat{C})$. We can do that by
+define the free Kleene category $\FKC(\cat{C})$. We can do that by
 using the internal language of Kleene categories. Concretely, it
@@
 
-\section{Gluing Kleene categories}
+This should imply that the algebraic and coalgebraic semantics of regular grammars coincides?
 
-The following theorems should be true
-\begin{theorem}
-  Given Kleene categories $\cat{L}$, $\cat{K}$, $\cat{K}'$
-  and functors $F : \cat{L} \to \cat{K}$, $p : \cat{K}' \to \cat{K}$
-  such that $F$ is lax monoidal such that it laxly preserves the initial
-  algebras and $p$ is a monoidal fibration strictly preserving the
-  Kleene category structure, then in the pullback diagram
-  % https://q.uiver.app/#q=WzAsNCxbMCwxLCJcXGNhdHtMfSJdLFsxLDEsIlxcY2F0e0t9Il0sWzEsMCwiXFxjYXR7S30nIl0sWzAsMCwiXFxjYXR7TH0nIl0sWzAsMSwiRiIsMl0sWzIsMSwicCJdLFszLDJdLFszLDBdLFszLDEsIiIsMSx7InN0eWxlIjp7Im5hbWUiOiJjb3JuZXIifX1dXQ==
-\[\begin{tikzcd}
-	{\cat{L}'} & {\cat{K}'} \\
-	{\cat{L}} & {\cat{K}}
-	\arrow[from=1-1, to=1-2]
-	\arrow[from=1-1, to=2-1]
-	\arrow["\lrcorner"{anchor=center, pos=0.125}, draw=none, from=1-1, to=2-2]
-	\arrow["p", from=1-2, to=2-2]
-	\arrow["F"', from=2-1, to=2-2]
-\end{tikzcd}\]
-$\cat{L}'$ is a Kleene category and the functor $\cat{L}'\to \cat{L}$ is a
-monoidal fibration strictly preserving the Kleene category structure.
+\section{Canonicity for Free Kleene Categories}
+
+Let $\Sigma$ be a set. Then the ``standard semantics'' of the free
+Kleene category $\FKC(\Sigma)$ is the unique functor $\sem{\cdot} :
+\FKC(\Sigma) \to \Set^{\Sigma^*}$ preserving the (Day convolution)
+Kleene category structure.
+
+We first address whether we can generalize Kozen's completeness
+theorems for Kleene algebras. Firstly, the fact that the Free Kleene
+category is complete for \emph{language} inclusion and \emph{weak}
+equivalence follows immediately from the fact that the free Kleene
+algebra is equivalent to the posetal reflection of the free Kleene
+category and the completeness theorem for Kleene algebras. That is,
+for any objects $A, B \in \FKC(\Sigma)$, if $|\sem{A}| \subseteq
+|\sem{B}|$ then there is a morphism $A \to B$ in $\FKC(\Sigma)$.
+
+The immediate generalization to strong equivalence is false. For
+example, we might ask that $\sem{\cdot}$ is a \emph{full} functor,
+that any semantic reduction of the formal grammars is the denotation
+of a morphism in the free Kleene category. But consider that the
+regular grammar $1^*$ denotes a discrete grammar of natural numbers
+and so morphisms $1^* \otimes \cdots \to 1^*$ are mapped into a set of
+multi-argument functions on natural numbers. So completeness would
+imply that all functions on natural numbers can be defined in the
+syntax of the free Kleene category, which fails as all such functions
+should be primitive recursive\footnote{flesh out the
+details?}.
+
+Failing completeness for definability of reductions, we could ask that
+the \emph{equational theory} of free Kleene categories is
+complete. This means that $\sem{\cdot}$ is a \emph{faithful} functor,
+i.e., if two syntactic morphisms denote the same reduction $\sem{M} =
+\sem{N}$, then they are provably equal $M = N$. I don't know if this
+is true yet, but it seems unlikely to me.
+
+However there is a (very) restricted form of completeness that is
+true. The denotation $\sem{\cdot}$ is fully faithful when restricting
+to hom sets of the form $\FKC(\Sigma)(\lceil w \rceil, A)$. The
+significance of this is that the denotation of such a morphism is
+equivalent to $\sem{A}\,w$, the set of ``parse trees'' for the grammar
+$A$. Then fullness says that the term syntax is at least complete
+enough to describe all parse trees, and faithfulness says that the
+equational theory is at least complete enough to prove that terms that
+represent the same parse tree are provably equal. In other words, for
+every parse tree, there is a unique equivalence class of terms that
+map to it. This theorem is called canonicity because it says that all
+terms of a certain type are provably equal to a term in a
+``canonical'' form.
+
+\begin{theorem}{Canonicity}
+  The denotation $\sem{\cdot}$ is bijective when restricting to hom
+  sets of the form $\FKC(\Sigma)(\lceil w \rceil, A)$.
+\end{theorem}
+
+To prove this theorem, we first change perspective. The function
+$\lceil \cdot \rceil : \Sigma^* \to \FKC(\Sigma)$ induces a restricted
+Yoneda embedding called the ``nerve'' functor $N : \FKC(\Sigma) \to
+\Set^{\Sigma^*}$ defined by
+\[ N(A)w = \FKC(\Sigma)(\lceil w \rceil, A) \]
+The action of $\sem{\cdot}$ on the restricted hom sets then induces a
+natural transformation
+\[ u : N \Rightarrow \sem{\cdot} \]
+defined as
+\[ u_w(f : \FKC(\Sigma)(\lceil w \rceil , A)) = \sem{f}(\iota_w) : A w \]
+where $\iota_w : \sem{\lceil w \rceil} w$ is the unique parse tree
+witnessing that the grammar $Y c_1 \otimes \cdots \otimes \varepsilon$
+matches the string $c_1\cdots$. This $\iota_w$ is a universal element
+of $\sem{\lceil w \rceil}$, meaning that evaluating at $\iota_w$ is a
+bijection between $\Set^{\Sigma^*}(\sem{\lceil w \rceil}, A)$ and $A w$. So we
+have that $\sem{\cdot}$ is fully faithful on these hom sets if and
+only if $u$ is a natural isomorphism.
+
+To do this we will explicitly construct an inverse natural
+transformation $q : \sem{\cdot} \to N$ which we call ``quote'' because it quotes
+a semantic parse tree back into the syntax. $q$ can be defined
+explicitly by recursion on syntactic types and naturality proven by
+induction over syntactic terms but we will develop a more structured
+proof using the universal property of the free Kleene category.  To do
+this we need to develop a Kleene category for which constructing a
+functor from $\FKC(\Sigma)$ into it is equivalent to defining such a
+natural transformation. We can do this by using the following
+properties of \emph{comma categories}.
+
+\begin{lemma}
+  Let $F : \cat C \to \cat E$ and $G : \cat D \to \cat E$ and let $F
+  \downarrow G$ be the comma category (i.e., whose objects are
+  morphisms $F c \to G d$) with projections $p_C,p_D$.
+  \begin{enumerate}
+  \item If $\cat C, \cat D, \cat E$ have a monoidal structure for
+    which $F$ is oplax and $G$ is lax monoidal, then $F \downarrow G$
+    can be equipped with a monoidal structure such that $p_C,p_D$ are
+    strict monoidal.
+  \item If $\cat C, \cat D$ have (specified) colimits of shape
+    $\mathcal I$ and $F$ preserves colimits of shape $\mathcal I$ then
+    $F \downarrow G$ has colimits of shape $\mathcal I$ and the
+    projections (strictly) preserve them.
+  \end{enumerate}
+\end{lemma}
+
+Then we can prove the following
+\begin{theorem}{Gluing for Kleene categories}
+  Let $\cat K, \cat L, \cat M$ be Kleene categories and $\Phi : \cat K
+  \to \cat M$ be a Kleene functor and $G : \cat L \to \cat M$ be a lax
+  monoidal functor. Then $\Phi \downarrow G$ is a Kleene category, and
+  the projections are (strict) Kleene functors.
 \end{theorem}
 \begin{proof}
-  It is known from linear logic reasons that the pullback is distributive monoidal.
-  The Kleene star should be defined much like the exponential in glued models of
-  linear logic. And the projection functor is a monoidal fibration because those
-  are stable under pullback.
+  The monoidal and coproduct aspects of Kleene categories are handled
+  by the previous lemma. We need to show that the monoidal structure
+  is distributive and construct the Kleene star.
+
+  For the Kleene star object, given $f :
+  \Phi(k) \Rightarrow G l$ we define the carrier as
+  \[ \Phi(k^*) \cong \Phi(k)^* \to (G l)^* \to G l^* \]
+  where $f^* : \Phi(k)^* \to (G l)^*$ and $g : (G l)^* \to G l^*$ is
+  defined by the induced cons-structure $G l \otimes G l^* \to G (l
+  \otimes l^*) \to G l^*$ and $I \to G I \to G l^*$. Then it is
+  routine to check that the nil, cons and fold all lift to commutative
+  squares.
+
+  TODO: address the snoc-part as well, and distributivity
 \end{proof}
 
-Recently I have been enjoying comma categories a bit more, which leads me to the follwing
-conjecture:
+Then $\sem {\cdot} : \FKC(\Sigma) \to \Set^{\Sigma^*}$ is a strict
+Kleene functor and $N : \FKC(\Sigma) \to \Set^{\Sigma^*}$ is a lax
+monoidal functor, and so $\sem{\cdot}\downarrow N$ is a Kleene
+category and the projection to $\FKC(\Sigma)^2$ is a strict Kleene
+functor. Then by the universal property of $\FKC(\Sigma)$, there is a
+unique strict Kleene functor $q$ from $\FKC(\Sigma)$ to
+$\sem{\cdot}\downarrow N$ and further the composition with the
+projection to $\FKC(\Sigma)^2$ is strict Kleene and so $q$ is
+equivalent to a natural transformation $\sem{\cdot} \Rightarrow N$.
 
-\begin{conjecture}
-  The 2-category of Kleene categories has comma objects.
-\end{conjecture}
+\begin{lemma}
+  $q \circ u$ is the identity
+\end{lemma}
+\begin{proof}
+  Let $f : \FKC(\Sigma)(\lceil w \rceil, A)$, then by naturality,
+  $q(\sem{f}(\iota)) = f \circ q(\sem{\id}(\iota))$ so it suffices to
+  show that $q(\iota) = \id$. This can be confirmed by calculation by
+  induction on the length of $w$.
+\end{proof}
+
+\begin{lemma}
+  $u \circ q$ is the identity
+\end{lemma}
+\begin{proof}
+  By the universal property of $\FKC(\Sigma)$, it suffices to show
+  that $u \circ q$ and $\id$ correspond to (strict?) Kleene functors
+  from $\FKC(\Sigma)$ to $\sem{\cdot} \downarrow \sem{\cdot}$. The
+  identity part should be a general theorem, and verifying the Kleene
+  category properties is done by cases.
+\end{proof}
 

### 2025-07-15 kleenecat/kleenecat.tex
progress on distributive law

@@
 }
+
+\usepackage{mdframed}
+\usepackage{xcolor}
+\usepackage{tcolorbox}
+\newcommand\makenote[4]{\par\noindent\begin{tcolorbox}
+[size=small,colback=#2]\color{#3}\textbf{#4: }#1\end{tcolorbox}\noindent}
+\definecolor{cornellred}{RGB}{196,18,48}
+\definecolor{umichblue}{HTML}{00274C}
+
+\newcommand\pa[1]{\makenote{#1}{cornellred}{white}{PA}}
+\newcommand\msn[1]{\makenote{#1}{umichblue}{white}{MSN}}
 
@@
     x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
-    x0 &= 0x = x & & \\
+    x0 &= 0x = 0 & & \\
     1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
@@
 
+\begin{definition}
+  A distributive law between a monad $T$ and an endofunctor $T$ is a
+  natural transformation $\lambda : TF \to F T$ making the following
+  diagrams commute:
+% https://q.uiver.app/#q=WzAsOCxbMSwwLCJGIl0sWzAsMSwiVEYiXSxbMiwxLCJGVCJdLFswLDIsIlRURiJdLFsyLDIsIkZUVCJdLFsxLDIsIlRGVCJdLFswLDMsIlRGIl0sWzIsMywiRlQiXSxbMSwyLCJcXGxhbWJkYSIsMl0sWzAsMSwiXFxldGFfRiIsMl0sWzAsMiwiRlxcZXRhIl0sWzMsNiwiXFxtdV9GIiwyXSxbNiw3LCJcXGxhbWJkYSJdLFszLDUsIlRcXGxhbWJkYSJdLFs1LDQsIlxcbGFtYmRhIFQiXSxbNCw3LCJGIFxcbXUiXV0=
+\[\begin{tikzcd}
+	& F \\
+	TF && FT \\
+	TTF & TFT & FTT \\
+	TF && FT
+	\arrow["{\eta_F}"', from=1-2, to=2-1]
+	\arrow["{F\eta}", from=1-2, to=2-3]
+	\arrow["\lambda"', from=2-1, to=2-3]
+	\arrow["{T\lambda}", from=3-1, to=3-2]
+	\arrow["{\mu_F}"', from=3-1, to=4-1]
+	\arrow["{\lambda T}", from=3-2, to=3-3]
+	\arrow["{F \mu}", from=3-3, to=4-3]
+	\arrow["\lambda", from=4-1, to=4-3]
+\end{tikzcd}\]
+\end{definition}
+
 It has been observed that Brzozowski's derivative can be more conceptually
 explained as a distributive law between the Kleene algebra monad and the
-DFA functor $2 \times (\Sigma \Rightarrow -)$.
+DFA functor $2 \times (-)^\Sigma$.
 
@@
 
-Since the axiomatization of Kleene algebras over an alphabet $\Sigma$
+\pa{I can't help but think that there's more to the construction
+below... It feels like $K^\Sigma \times K$ is ``fibered'' over $K$}
+
+\begin{definition}
+  \label{def:nonstandard}
+  If $K$ is a Kleene algebra over $\Sigma$, we can equip
+  $K^\Sigma \times K$ with the following Kleene algebra
+  structure.
+    \begin{align*}
+    (f_1, e_1) + (f_2, e_2) &= (f_1 + f_2, e_1 + e_2)\\
+    0 &= (\lambda c.\, 0, 0)\\
+    (f_1, e_1) \cdot (f_2, e_2) &= (\lambda c.\, f_1(c)\cdot e_2 + \alpha(e_1)\cdot f_2(c), e_1\cdot e_2)\\
+    1 &= (\lambda c.\, 0, 1)\\
+    (f, e)^* &= (\lambda c.\, f(c)\cdot e^*, e^*)
+    \end{align*}
+\end{definition}
+
+The proof that this is a Kleene algebra is boring, but unsurprising.
+Because the axiomatization of Kleene algebras over an alphabet $\Sigma$
 is a Horn theory, the category of Kleene algebras and Kleene algebra
-morphisms is monadic over $\Set$. Let $E_\Sigma : \Set \to \Set$. be the 
-Kleene algebra monad. It is possible to show the following theorem.
+morphisms is monadic over $\Set$. Let $E_\Sigma : \Set \to \Set$ be the 
+Kleene algebra monad.
+
+It is possible to show that there is a distributive law
+$E_\Sigma(2 \times X^\Sigma) \to 2 \times (E_\Sigma(X)^\Sigma)$.
+Its recursive definition is depicted in Figure~\ref{fig:distlaw}. 
+%
+\begin{figure}
+  \begin{align*}
+    \alpha : E_\Sigma(2 \times X^\Sigma) \to 2 \quad\quad & \delta : E_\Sigma(2 \times X^\Sigma) \to E_\Sigma(X)^\Sigma\\
+  \alpha(e_1 + e_2) = \alpha(e_1) + \alpha(e_2) \quad \quad & \delta_c(e_1 + e_2) = \delta_c(e_1) + \delta_c(e_2)\\ 
+  \alpha(e_1\cdot e_2) = \alpha(e_1) \land \alpha(e_2) \quad \quad& \delta_c(e_1\cdot e_2) = \delta_c(e_1)\cdot e_2 + \alpha(e_1)\cdot \delta_c(e_2)\\ 
+  \alpha(e^*) = 1 \quad \quad& \delta_c(e^*) = \delta_c(e)e^*\\ 
+  \alpha(1) = 1 \quad \quad& \delta_c(1) = 0\\ 
+  \alpha(0) = 0 \quad \quad& \delta_c(0) = 0\\ 
+  \alpha(d : \Sigma) = 0 \quad \quad& \delta_c(d : \Sigma) = [c = d]\\ 
+  \alpha(b, f) = b \quad \quad& \delta_c(b, f) = f(c)  
+\end{align*}
+\caption{Distributive law for Kleene algebras}
+\label{fig:distlaw}
+\end{figure}
 
 \begin{theorem}
-  There is a distributive law $E_\Sigma(2 \times (\Sigma \Rightarrow X)) \to 2 \times (\Sigma \Rightarrow E_\Sigma(X))$.
+  There is a distributive law $E_\Sigma(2 \times (X^\Sigma)) \to 2 \times (E_\Sigma(X)^\Sigma)$.
 \end{theorem}
 \begin{proof}
-  The component $E_\Sigma(2 \times (\Sigma \Rightarrow X)) \to 2$ is
+  The component $E_\Sigma(2 \times (X^\Sigma)) \to 2$ is
   defined using the universal property of free algebraic structures,
-  i.e. it suffices to define a morphism $2 \times (\Sigma \Rightarrow
-  X)) \to 2$, which we choose it to be the first projection
-  $\pi_1$. The component $E_\Sigma(2 \times (\Sigma \Rightarrow X))
-  \to (\Sigma \Rightarrow E_\Sigma(X))$ requires us to be more
-  clever. There is a ``non-standard'' Kleene algebra structure on
-  $(\Sigma \Rightarrow E_\Sigma(X)) \times E_\Sigma(X)$ that makes
-  this possible.
+  i.e. it suffices to define a morphism $2 \times X^\Sigma \to 2$,
+  which we choose it to be the first projection
+  $\pi_1$. The component $E_\Sigma(2 \times (X^\Sigma)) \to (E_\Sigma(X)^\Sigma)$
+  requires us to be more
+  clever. Instantiating the ``non-standard'' Kleene algebra structure from
+  Definition~\ref{def:nonstandard} to $E_\Sigma(X)$
+  gives us a morphism $\psi : E_\Sigma(2 \times X^\Sigma) \to E_\Sigma(X)^\Sigma \times E_\Sigma(X)$
+  by using the universal property of free Kleene
+  algebras and lifting the function
+  \[\lambda (b, f).\, (\lambda c.\, \eta(f(c)), 1) : 2 \times X^\Sigma \to E_\Sigma(X)^\Sigma \times E_\Sigma(X)\]
+  The derivative $\delta$ is the postcomposition of $\psi$ with the
+  projection function.
 \end{proof}
@@
 We now want to prove that the distributive law above generalizes to
-the Kleene category case. We begin by showing that the ``non-standard''
+the Kleene category case. First, we have to define a natural
+transformation $K(\Set \times (-)^\Sigma) \to \Set \times (K -
+)^\Sigma$ making the appropriate diagrams commute. We decompose the
+natural transformation into two components.  The component $K(\Set
+\times (-)^\Sigma) \to \Set$ is by equipping $\Set$ with a Kleene
+category structure and then using the universal property of the free
+Kleene category. The Kleene category structure on $\Set$ is the usual
+(co)cartesian structure and $X^*$ are the lists on $X$. Therefore, it
+suffices to define a natural transformation $\Set \times (-)^\Sigma \to \Set$,
+which we define as the first projection.
+
+For the component $K(\Set \times (-)^\Sigma) \to (K - )^\Sigma$,
+we will once again use the universal property of free Kleene
+categories. We begin by showing that the ``non-standard''
 structure defined in the previous section can be extended to the
@@
 
+\pa{This is where a better understanding of what this ``non-standard''
+  construction is actually doing could come in handy. Showing that the
+  monoidal product satisfies all of the coherence diagrams is really gnarly.}
 \begin{lemma}
-  Let $\cat{K}$ and $\cat{L}$ be two Kleene categories, $F : \cat{L}
-  \to \cat{K}$ and $G : \cat{L} \to \cat{K}$ two Kleene category
-  morphisms. We can equip $\cat{K} \times \cat{L}$ with the following
-  Kleene category structure:
+  Let $\cat{K}$ and $\cat{L}$ be two Kleene categories, $F : \cat{L} \to \cat{K}$
+  and $G : \cat{L} \to \cat{K}$ two Kleene functors.
+  We can equip $\cat{K} \times \cat{L}$ with the following Kleene
+  category structure:
   \begin{itemize}
@@
   \item $(A, X) \otimes (B, Y) = (A \otimes F(Y) + G(X) \otimes B, X \otimes Y)$
+  \item $I = (0, I)$
   \item $(A, X)^* = (A \otimes F(X^*), X^*)$
@@
 \end{lemma}
+
+Therefore we equip $K(\cat{C})^\Sigma \times K(\cat{C})$ with the
+Kleene category structure given by the lemma above.
 %% \begin{proof}
@@
 %% \end{proof}
-
-With this lemma and the 2-monadicity of Kleene categories over $\cat{Cat}$,
-we can define the distributive law between the proof-relevant
-DFA functor and the Kleene category 2-monad.
 
@@
 Then we can prove the following
-\begin{theorem}{Gluing for Kleene categories}
+\begin{theorem}[Gluing for Kleene categories]
   Let $\cat K, \cat L, \cat M$ be Kleene categories and $\Phi : \cat K

### 2025-11-11 kleenecat/kleenecat.tex
edits on distributive law section

@@
-\documentclass[acmsmall,anonymous,review,screen]{acmart}
+\documentclass[sigplan,anonymous,review,screen]{acmart}
 \usepackage{mathpartir}
@@
 \begin{document}
-\title{Kleene Categories}
-\author{Steven Schaefer}
-\affiliation{
-  \department{Electrical Engineering and Computer Science}
-  \institution{University of Michigan}
-  \country{USA}
-}
-\email{stschaef@umich.edu}
-
-\author{Max S. New}
-\affiliation{
-  \department{Electrical Engineering and Computer Science}
-  \institution{University of Michigan}
-  \country{USA}
-}
-\email{maxsnew@umich.edu}
-
-\author{Pedro H. Azevedo de Amorim}
-\affiliation{
-  \department{Department of Computer Science}
-  \institution{University of Oxford}
-  \country{UK}
-}
-\email{pedro.azevedo.de.amorim@cs.ox.ac.uk}
+\title{Kozen Categories}
 
@@
 
-\section{Kleene Algebra and Kleene Categories}
+\section{Kleene Algebra and Kozen Categories}
 
@@
 algebra as algebraic structure over sets, and instead as a kind of
-structured \emph{poset}. We then generalize this to a Kleene
+structured \emph{poset}. We then generalize this to a Kozen
 \emph{category}, where the morphisms of the category correspond to not
@@
 \begin{definition}
-  A Kleene category is a distributive monoidal category $\cat{K}$ such
+  A Kozen category is a distributive monoidal category $\cat{K}$ such
   that for every object $A$, there is a cons-list object
@@
 As a sanity check, note that Kleene algebras are indeed examples of
-Kleene categories.
+Kozen categories.
 
 \begin{lemma}
-  A Kleene algebra is equivalent to a posetal Kleene category.
+  A Kleene algebra is equivalent to a posetal Kozen category.
 \end{lemma}
 
-% This isn't an example of a Kleene category, it's an example of a model of conjunctive ordered logic
+% This isn't an example of a Kozen category, it's an example of a model of conjunctive ordered logic
 An unexpected example comes from the theory of substructural logics.
@@
 \begin{example}
-  The opposite category of every Kleene category is a model of a variant of
+  The opposite category of every Kozen category is a model of a variant of
   conjunctive ordered logic, where the Kleene star plays the role of the ``of
@@
 setting. Its most convoluted aspect is the axiomatization of the
-Kleene star as a parameterized list and snoc object. If the Kleene
+Kleene star as a parameterized list and snoc object. If the Kozen
 category $\cat{K}$ has more structure, then these isomorphisms hold
@@
 \begin{definition}
-  A pre-Kleene category is a distributive monoidal category with a (non-parameterized) cons-list object.
+  A pre-Kozen category is a distributive monoidal category with a (non-parameterized) cons-list object.
 \end{definition}
@@
 \begin{theorem}
-  If $\cat K$ is a pre-Kleene category that is also biclosed, then it is a Kleene category.
+  If $\cat K$ is a pre-Kozen category that is also biclosed, then it is a Kozen category.
 \end{theorem}
@@
 %%   \label{th:kleeneclosed}
-%%   Let $\cat{K}$ be a Kleene category such that it is also monoidal
+%%   Let $\cat{K}$ be a Kozen category such that it is also monoidal
 %%   closed.  Then, the initial algebras isomorphisms hold automatically.
@@
   If $\cat{C}$ is a locally small monoidal category, then
-  $\Set^{\cat{C}}$ is a Kleene category.
+  $\Set^{\cat{C}}$ is a Kozen category.
 \end{theorem}
@@
   For every alphabet $\Sigma$, the presheaf category $\grammar$
-  is a Kleene category.
+  is a Kozen category.
 \end{corollary}
@@
 \begin{definition}
-  Let $K$ be a Kleene category over alphabet $\Sigma$. This
+  Let $K$ be a Kozen category over alphabet $\Sigma$. This
   induces a strong monoidal functor $\Sigma^* \to K$. The
@@
 arrows $f : X \to Y$ in $\cat{C}$ should be thought of as a proof that
-the state $X$ is related to the $Y$. This can be seen as a
+the state $X$ is related to $Y$. This can be seen as a
 generalization of simulation relations. After all, functoriality of $F
@@
 \to F(Y, c)$, i.e. after one transition, the end states will also be
-related.
+related. We now formalize this intuition
+
+\begin{example}[Automata with simulation relation]
+  Various notions of simulation for automata can be formulated using
+  the language of fibered category theory and liftings of functors.
+  Assume that we have a determinisitc automaton over a state space $S$
+  with a simulation relation $\mathcal{R}$. We can see $S$ and $\mathcal{R}$
+  as a thin category and equip it with a canonical $F$-coalgebra structure.
+  \begin{itemize}
+  \item The component $(S, \mathcal{R}) \to \Set$ maps an element of $S$ to the
+    singleton set $\{*\}$ if it is an accepting state and to the empty set
+    otherwise. The action on morphisms is defined using the simulation
+    acceptance condition, i.e. if $(x, y) \in \mathcal{R}$, then
+    $\alpha(x) \leq \alpha(y)$. Functoriality follows from initiality/terminality
+    of the empty and singleton sets, respectively.
+  \item The component $(S, \mathcal{R}) \times \Sigma \to (S, \mathcal{R})$
+    is, as expected, the transition function $\delta : S \times \Sigma \to S$.
+    The action on morphisms is given by the transition condition of simulation
+    relations, i.e. if $(x, y) \in \mathcal{R})$ then for every $c \in \Sigma$
+    $(\delta(x, c), \delta(y, c)) \in \mathcal{R}$.
+  \end{itemize}
+\end{example}
 
@@
 
-\section{Proposed construction for free Kleene categories}
+\section{Proposed construction for free Kozen categories}
 
@@
 Let $\cat{C}$ be a category. It is more or less straightforward to
-define the free Kleene category $\FKC(\cat{C})$. We can do that by
-using the internal language of Kleene categories. Concretely, it
+define the free Kozen category $\FKC(\cat{C})$. We can do that by
+using the internal language of Kozen categories. Concretely, it
 will be a type theory with (non-commutative) tensors, sum types and
@@
 
-The proof that this is a Kleene algebra is boring, but unsurprising.
+The proof that this is a Kleene algebra is boring and unsurprising.
 Because the axiomatization of Kleene algebras over an alphabet $\Sigma$
@@
 We now want to prove that the distributive law above generalizes to
-the Kleene category case. First, we have to define a natural
+the Kozen category case. First, we have to define a natural
 transformation $K(\Set \times (-)^\Sigma) \to \Set \times (K -
@@
 natural transformation into two components.  The component $K(\Set
-\times (-)^\Sigma) \to \Set$ is by equipping $\Set$ with a Kleene
+\times (-)^\Sigma) \to \Set$ is by equipping $\Set$ with a Kozen
 category structure and then using the universal property of the free
-Kleene category. The Kleene category structure on $\Set$ is the usual
+Kozen category. The Kozen category structure on $\Set$ is the usual
 (co)cartesian structure and $X^*$ are the lists on $X$. Therefore, it
@@
 For the component $K(\Set \times (-)^\Sigma) \to (K - )^\Sigma$,
-we will once again use the universal property of free Kleene
+we will once again use the universal property of free Kozen
 categories. We begin by showing that the ``non-standard''
@@
 \begin{lemma}
-  Let $\cat{K}$ and $\cat{L}$ be two Kleene categories, $F : \cat{L} \to \cat{K}$
-  and $G : \cat{L} \to \cat{K}$ two Kleene functors.
-  We can equip $\cat{K} \times \cat{L}$ with the following Kleene
+  Let $\cat{K}$ and $\cat{L}$ be two Kozen categories, $F : \cat{L} \to \cat{K}$
+  and $G : \cat{L} \to \cat{K}$ two Kozen functors.
+  We can equip $\cat{K} \times \cat{L}$ with the following Kozen
   category structure:
@@
 Therefore we equip $K(\cat{C})^\Sigma \times K(\cat{C})$ with the
-Kleene category structure given by the lemma above.
+Kozen category structure given by the lemma above.
+
+  In our case, we want to instantiate $F$ with the unit for the reader
+  monad $(-)^\Sigma$. The generalization of $G$ is a bit subtle, as it
+  is not enough to simply output weather a state accepts the empty
+  string or not, we have to carry around the the witnesses of such
+  parse trees. Usin the fact that $2$ is the free Kleene algebra with
+  zero generators on the empty alphabet, we instantiate $G$ to be a
+  ``prunning'' operation that factors through the free Kozen category
+  on the empty category. More concretely, we define a functor
+  $K(\cat{C}) \to K(\cat{C})^\Sigma$ by first defining a functor
+  $K(\cat{C}) \to K(\emptyset)$ by using the universal property of free
+  Kozen categories and mapping $\cat{C}$ to the constant $0$ object of
+  $K(\emptyset)$. The functor $K(\emptyset) \to K(\cat{C})^\Sigma$ is defined
+  using the fact that left adjoints preserve colimits, so $K(\emptyset)$
+  is initial in the category of Kozen categories.
+
 %% \begin{proof}
@@
 
-\section{Canonicity for Free Kleene Categories}
+\section{Canonicity for Free Kozen Categories}
 
 Let $\Sigma$ be a set. Then the ``standard semantics'' of the free
-Kleene category $\FKC(\Sigma)$ is the unique functor $\sem{\cdot} :
+Kozen category $\FKC(\Sigma)$ is the unique functor $\sem{\cdot} :
 \FKC(\Sigma) \to \Set^{\Sigma^*}$ preserving the (Day convolution)
-Kleene category structure.
+Kozen category structure.
 
 We first address whether we can generalize Kozen's completeness
-theorems for Kleene algebras. Firstly, the fact that the Free Kleene
+theorems for Kleene algebras. Firstly, the fact that the Free Kozen
 category is complete for \emph{language} inclusion and \emph{weak}
 equivalence follows immediately from the fact that the free Kleene
-algebra is equivalent to the posetal reflection of the free Kleene
+algebra is equivalent to the posetal reflection of the free Kozen
 category and the completeness theorem for Kleene algebras. That is,
@@
 that any semantic reduction of the formal grammars is the denotation
-of a morphism in the free Kleene category. But consider that the
+of a morphism in the free Kozen category. But consider that the
 regular grammar $1^*$ denotes a discrete grammar of natural numbers
@@
 imply that all functions on natural numbers can be defined in the
-syntax of the free Kleene category, which fails as all such functions
+syntax of the free Kozen category, which fails as all such functions
 should be primitive recursive\footnote{flesh out the
@@
 Failing completeness for definability of reductions, we could ask that
-the \emph{equational theory} of free Kleene categories is
+the \emph{equational theory} of free Kozen categories is
 complete. This means that $\sem{\cdot}$ is a \emph{faithful} functor,
@@
 induction over syntactic terms but we will develop a more structured
-proof using the universal property of the free Kleene category.  To do
-this we need to develop a Kleene category for which constructing a
+proof using the universal property of the free Kozen category.  To do
+this we need to develop a Kozen category for which constructing a
 functor from $\FKC(\Sigma)$ into it is equivalent to defining such a
@@
 Then we can prove the following
-\begin{theorem}[Gluing for Kleene categories]
-  Let $\cat K, \cat L, \cat M$ be Kleene categories and $\Phi : \cat K
-  \to \cat M$ be a Kleene functor and $G : \cat L \to \cat M$ be a lax
-  monoidal functor. Then $\Phi \downarrow G$ is a Kleene category, and
-  the projections are (strict) Kleene functors.
+\begin{theorem}[Gluing for Kozen categories]
+  Let $\cat K, \cat L, \cat M$ be Kozen categories and $\Phi : \cat K
+  \to \cat M$ be a Kozen functor and $G : \cat L \to \cat M$ be a lax
+  monoidal functor. Then $\Phi \downarrow G$ is a Kozen category, and
+  the projections are (strict) Kozen functors.
 \end{theorem}
 \begin{proof}
-  The monoidal and coproduct aspects of Kleene categories are handled
+  The monoidal and coproduct aspects of Kozen categories are handled
   by the previous lemma. We need to show that the monoidal structure
@@
 Then $\sem {\cdot} : \FKC(\Sigma) \to \Set^{\Sigma^*}$ is a strict
-Kleene functor and $N : \FKC(\Sigma) \to \Set^{\Sigma^*}$ is a lax
-monoidal functor, and so $\sem{\cdot}\downarrow N$ is a Kleene
-category and the projection to $\FKC(\Sigma)^2$ is a strict Kleene
+Kozen functor and $N : \FKC(\Sigma) \to \Set^{\Sigma^*}$ is a lax
+monoidal functor, and so $\sem{\cdot}\downarrow N$ is a Kozen
+category and the projection to $\FKC(\Sigma)^2$ is a strict Kozen
 functor. Then by the universal property of $\FKC(\Sigma)$, there is a
-unique strict Kleene functor $q$ from $\FKC(\Sigma)$ to
+unique strict Kozen functor $q$ from $\FKC(\Sigma)$ to
 $\sem{\cdot}\downarrow N$ and further the composition with the
-projection to $\FKC(\Sigma)^2$ is strict Kleene and so $q$ is
+projection to $\FKC(\Sigma)^2$ is strict Kozen and so $q$ is
 equivalent to a natural transformation $\sem{\cdot} \Rightarrow N$.
@@
   By the universal property of $\FKC(\Sigma)$, it suffices to show
-  that $u \circ q$ and $\id$ correspond to (strict?) Kleene functors
+  that $u \circ q$ and $\id$ correspond to (strict?) Kozen functors
   from $\FKC(\Sigma)$ to $\sem{\cdot} \downarrow \sem{\cdot}$. The
-  identity part should be a general theorem, and verifying the Kleene
+  identity part should be a general theorem, and verifying the Kozen
   category properties is done by cases.
